### 2016-02-16 frac.tex
fractional story

\documentclass{article}
\usepackage{fullpage}
\usepackage{bbold}
\usepackage{url}

\newcommand{\inl}[1]{\textsf{inl}(#1)}
\newcommand{\inr}[1]{\textsf{inr}(#1)}
\newcommand{\zt}{\mathbb{0}}
\newcommand{\ot}{\mathbb{1}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\fv}[2]{\langle #1 ~|~ #2 \rangle}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Zn}{\mathbb{Z}_n}
\newcommand{\Zvn}{\mathbb{Z}^v_n}
\newcommand{\pt}[2]{#1_{\bullet}^{#2}}
\newcommand{\cycle}{\textsf{cycle}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Action Groupoids and Fractional Types}
\author{Everyone}
\begin{document}
\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}

Our starting point is $\Pi$:
\begin{itemize}
\item We have finite types $\zt$, $\ot$, $\tau_1\oplus\tau_2$, and $\tau_1\otimes\tau_2$
\item The types have points $()$, $\inl{v}$, $\inr{v}$, and $(v_1,v_2)$
\item A type $\tau$ has $|\tau|$ points
\item We have combinators $c : \tau_1\leftrightarrow\tau_2$ between the types which witness type isomorphisms and which correspond to the axioms of commutative rigs
\item If we have combinators $c_1, c_2 : \tau_1\leftrightarrow\tau_2$, we have level-2 combinators
$\alpha : c_1 \Leftrightarrow c_2$ which look quite messy but they are essentially witnesses of
the coherence conditions for rig groupoids.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Generalization}

Now we do the following:
\begin{itemize}
\item We generalize the syntax of types to include fractional types $\tau_1/\tau_2$ 
\item The elements of $\tau_1/\tau_2$ will be denoted $\fv{v}{\G}$ where $v : \tau_1$ and $\G$ is essentially the cyclic group $\Z_n$ of order $n=|\tau_2|$. More precisely, we will think of $\G$ as a particular enumeration of the elements of $\tau_2$ in some canonical order allowing us to cycle through the elements
\item \textbf{Note:} The types $\zt/\tau$ (including $\zt/\zt$) are all empty
\item Each type $\ot/\tau$ has one value $\fv{()}{\G_\tau}$. This group allows us to cycle through the elements of $\tau$.
\item \textbf{Note:} If the group happens to be isomorphic to $\Z_1$ it has no effect and we recover the plain types from before; the types $\tau/\ot$ are essentially identical to $\tau$
\item \textbf{Note:} According to our convention, the type $\ot/\zt$ would have one value $\fv{()}{\G_\zt}$ where $\G_\zt$ is isomorphic to $\Z_0$; the latter is, by convention, the infinite cyclic group $\Z$. There is probably no harm in this.
\item The semantic justification for using division is the cardinality of $\tau_1/\tau_2$ is $|\tau_1|/|\tau_2|$. The reason is that if the elements of $\tau_1$ are $\{v_1,\ldots,v_{|\tau_1|}\}$, the elements of $\tau_1/\tau_2$ are $\{ \fv{v_1}{\G_{\tau_2}}, \ldots, \fv{v_{|\tau_1|}}{\G_{\tau_2}} \}$. This type isomorphic to $\bigoplus_{|\tau_1|} 1/\tau_2$
\item We can combine types using $\oplus$ and $\otimes$ in ways that satisfy the familiar algebraic identities
for the rational numbers
\item We now introduce the idea of a \emph{pointed type} $\pt{\tau}{v}$ which is a non-empty type $\tau$ with one value $v : \tau$ in focus
\item A pointed type $\pt{\tau}{v}$ can be used anywhere $\tau$ can be used but we must keep track of what happens to $v$; a transformation $\tau \rightarrow \tau'$ when acting on the pointed type $\pt{\tau}{v}$ will map $v$ to some element $v' : \tau'$ and we must keep track of that in the type.
\item Semantically when we have a type $1/\pt{\tau}{v}$, we have the group $\G_\tau$ which cycles through the elements of $\tau$ with one particular value $v$ in focus. We will denote this as $\G_\tau^v$
\item We can ``create'' and ``cancel'' fractional pointed types using $\eta_{\pt{\tau}{v}}$ and $\epsilon_{\pt{\tau}{v}}$ as follows: 
\[\begin{array}{rcl}
\eta_{\pt{\tau}{v}} &:& \ot \rightarrow \pt{\tau}{v} \otimes 1/\pt{\tau}{v} \\
\eta_{\pt{\tau}{v}}~() &=& (v , \fv{()}{\G^v_{|\tau|}}) \\
\\
\epsilon_{\pt{\tau}{v}} &:& \pt{\tau}{v} \otimes 1/\pt{\tau}{v} \rightarrow \ot \\
\epsilon_{\pt{\tau}{v}}~(v , \fv{()}{\G^v_{|\tau|}}) &=& () 
\end{array}\]
\item Another crucial operation we can do is to use the group to cycle through the values:
\[\begin{array}{rcl}
\cycle &:& \pt{\tau}{v} \otimes 1/\pt{\tau}{v} \rightarrow \pt{\tau}{v'} \otimes 1/\pt{\tau}{v'} \\
\cycle~(v, \fv{()}{\G^v_{|\tau|}}) &=& (v', \fv{()}{\G^{v'}_{|\tau|}})
  \quad \mbox{if~$v'$~is~the~next~value~after~$v$~in~the~cycle~order~of~the~group}
\end{array}\]
\item Let's consider the following algebraic identity and how it would execute in our setting. For $a \neq 0$:
\[\begin{array}{rcl}
a &=& a * 1 \\
&=& a * (1/a * a) \\
&=& (a * 1/a) * a \\
&=& 1 * a \\
&=& a
\end{array}\]
We want to correspond to some transformation $a \rightarrow a$. If $a$ is the empty type, we can't apply this transformation to anything. Otherwise, we start with a value $v : a$, convert it to the pair $(v, ())$, then use $\eta$ to produce $(v , (v' , \fv{()}{\G_a^{v'}}))$ for some value $v' : a$. We know nothing about $v'$; it may be the identical to $v$ or not. Then we reassociate to get $((v , \fv{()}{\G_a^{v'}}), v')$. If $v$ is identical to $v'$ we can use $\epsilon$ to cancel the first pair; if not, we have to re-reassociate, cycle to choose another value and until $v'$ becomes identical to $v$ and then cancel. To summarize:
\[\begin{array}{rcl}
v &\mapsto& (v , ()) \\
&\mapsto& (v , (v' , \fv{()}{\G_a^{v'}})) \\
&\mapsto& ((v , \fv{()}{\G_a^{v'}}), v')  \quad \mbox{stuck~because~} v \neq v' \\
&\mapsto& (v , (v' , \fv{()}{\G_a^{v'}})) \\
&\mapsto& (v , (v , \fv{()}{\G_a^{v}})) \quad \mbox{using~cycle} \\
&\mapsto& ((v , \fv{()}{\G_a^{v}}), v)  \\
&\mapsto& ((), v) \\
&\mapsto& v
\end{array}\]
To make sense of this story, consider that there are two sites; one site has a value $v$ that it wants to communicate to another site. In a conventional situation, the two sites must synchronize but here we have an alternative idea. The second site can speculatively proceed with a guess $v'$ and produce some constraint that can propagate independently that recalls the guess. The second site can in principle proceed further with its guessed value. Meanwhile the constraint reaches the first site and we discover that there is a mismatch. The only
course of action is for the constraint to travel back to the second site, adjust the guess, and continue after the guessed value matches the original value. This idea is reminiscent of our ``reversible concurrency'' paper which discusses much related work. 
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Further Thoughts}

\begin{itemize}
\item If one thinks of 2D types, i.e., types with equivalence
  relations on them, there are many possible equivalence relations one
  can think of. Cyclic groups are a special case: a nice one.

\item This special case leads to action groupoids which we interpret
  as fractional types.

\item An important question arises: what other kinds of finite
  groupoids are out there; this almost calls for a classification of
  finite groupoids which is probably just as bad as classifying finite
  groups. However there are some interesting perspectives from the
  point of complexity theory
  \url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.68.7980&rep=rep1&type=pdf}

\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2016-02-16 frac.tex
organizing the flow in the paper better and starting to formalize things a bit

@@
 \usepackage{url}
+\usepackage{xcolor}
+\usepackage{amssymb}
 
+\newcommand{\alt}{~|~}
 \newcommand{\inl}[1]{\textsf{inl}(#1)}
@@
 \newcommand{\cycle}{\textsf{cycle}}
+\newcommand{\twod}{\mathbb{T}}
+\newcommand{\fract}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}
 
@@
 \item The types have points $()$, $\inl{v}$, $\inr{v}$, and $(v_1,v_2)$
-\item A type $\tau$ has $|\tau|$ points
-\item We have combinators $c : \tau_1\leftrightarrow\tau_2$ between the types which witness type isomorphisms and which correspond to the axioms of commutative rigs
-\item If we have combinators $c_1, c_2 : \tau_1\leftrightarrow\tau_2$, we have level-2 combinators
-$\alpha : c_1 \Leftrightarrow c_2$ which look quite messy but they are essentially witnesses of
-the coherence conditions for rig groupoids.
+\item A type $\tau$ has $|\tau|$ points:
+\[\begin{array}{rcl}
+|\zt| &=& 0 \\
+|\ot| &=& 1 \\
+|\tau_1 \oplus \tau_2| &=& |\tau_1| + |\tau_2| \\
+|\tau_1 \otimes \tau_2| &=& |\tau_1| * |\tau_2| 
+\end{array}\]
+\item We have combinators $c : \tau_1\leftrightarrow\tau_2$ between
+  the types which witness type isomorphisms and which correspond to
+  the axioms of commutative rigs
+\item If we have combinators $c_1, c_2 : \tau_1\leftrightarrow\tau_2$,
+  we have level-2 combinators $\alpha : c_1 \Leftrightarrow c_2$ which
+  are (quite messy) equivalences of isomorphisms, and which happen to
+  correspond to the coherence conditions for rig groupoids.
 \end{itemize}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Generalization}
+\section{2D-types: Intuition}
 
-Now we do the following:
+In HoTT, types are groupoids. One can visualize a type as a collection
+of points connected by paths. The paths can be arbitrary subject to
+three conditions: there is an associative operation which composes
+paths; there is an identity path; and every path has an inverse.  This
+most general form is difficult to capture structurally and
+computationally but there are some interesting special cases, one of
+which we explore in detail in this paper. The special case we study in
+detail is that of an \emph{action groupoid} $S \rtimes \G$. This is a
+groupoid in which the paths are generated by group.
+
+Give lots of examples of action groupoids. Explain cardinality.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{2D-types: Syntax}
+
 \begin{itemize}
-\item We generalize the syntax of types to include fractional types $\tau_1/\tau_2$ 
+\item We generalize the syntax of types to include fractional types
+  $\tau_1/\tau_2$ which we will write as $\fract{\tau_1}{\tau_2}$ with
+  the numerator in white and the denominator in grey. The idea will be
+  that $\tau_1$ denotes a set $S$ and $\tau_2$ denotes a group $\G$
+  and that $\fract{\tau_1}{\tau_2}$ denotes the action groupoid
+  $S \rtimes \G$.
+
+\item We actually have two levels of types:
+\[\begin{array}{rcl} 
+\tau &::=& \zt \alt \ot \alt \tau_1\oplus\tau_2 \alt \tau_1\otimes\tau_2 \\
+\twod &::=& \fract{\tau_1}{\tau_2} \alt \twod_1 \boxplus \twod_2
+            \alt \twod_1 \boxtimes \twod_2  
+\end{array}\]
+The $\tau$ level describes plain sets. The $\twod$ level describes
+``two-dimensional types'' which will be explained in more detail
+later. 
+\item Note that 2D types are closed under sums and products but
+  \emph{not} under ``division.'' In other words, we cannot form types
+  $\fract{\fract{\tau_1}{\tau_2}}{\fract{\tau_3}{\tau_4}}$. This is
+  why we call our types 2D as we are restricted to two levels.
+
+\item Pointed types\ldots
+
 \item The elements of $\tau_1/\tau_2$ will be denoted $\fv{v}{\G}$ where $v : \tau_1$ and $\G$ is essentially the cyclic group $\Z_n$ of order $n=|\tau_2|$.

### 2016-06-09 pibackground.tex
ok now all changes in sec 3 should remain local to sec 3

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}

The main syntactic vehicle for the technical developments in this paper is a
simple language called $\Pi$ whose only computations are isomorphisms between
finite types~\cite{James:2012:IE:2103656.2103667}. After reviewing the
motivation for this language and its relevance to HoTT, we present its syntax
and semantics.

\begin{figure*}[ht]
\[\begin{array}{cc}
\begin{array}{rrcll}
\unitepl :&  \zt \oplus \tau & \iso & \tau &: \unitipl \\
\unitepr :&  \tau \oplus \zt & \iso & \tau &: \unitipr \\
\swapp :&  \tau_1 \oplus \tau_2 & \iso & \tau_2 \oplus \tau_1 &: \swapp \\
\assoclp :&  \tau_1 \oplus (\tau_2 \oplus \tau_3) & \iso & (\tau_1 \oplus \tau_2) \oplus \tau_3 
  &: \assocrp \\
\\
\unitetl :&  \ot \otimes \tau & \iso & \tau &: \unititl \\
\unitetr :&  \tau \otimes \ot & \iso & \tau &: \unititr \\
\swapt :&  \tau_1 \otimes \tau_2 & \iso & \tau_2 \otimes \tau_1 &: \swapt \\
\assoclt :&  \tau_1 \otimes (\tau_2 \otimes \tau_3) & \iso & (\tau_1 \otimes \tau_2) \otimes \tau_3 
  &: \assocrt \\
\\
\absorbr :&~ \zt \otimes \tau & \iso & \zt &: \factorzl \\
\absorbl :&~ \tau \otimes \zt & \iso & \zt &: \factorzr \\
\dist :&~ (\tau_1 \oplus \tau_2) \otimes \tau_3 & 
  \iso & (\tau_1 \otimes \tau_3) \oplus (\tau_2 \otimes \tau_3)~ &: \factor \\
\distl :&~ \tau_1 \otimes (\tau_2 \oplus \tau_3) & 
  \iso & (\tau_1 \otimes \tau_2) \oplus (\tau_1 \otimes \tau_3)~ &: \factorl 
\end{array}
& 
\begin{minipage}{0.4\textwidth}
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso : \tau \iso \tau}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 \oplus \tau_3 \iso \tau_2 \oplus \tau_4}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 \otimes \tau_3 \iso \tau_2 \otimes \tau_4}}
{}
\end{array}\]
\end{minipage}
\end{array}\]
Each 1-combinator $c$ has an inverse $!~c$, e.g, $!~\unitepl=\unitipl$,
$!(c_1 \odot c_2) = ~!c_2 \odot~ !c_1$, etc.
\caption{$\Pi$-combinators~\cite{James:2012:IE:2103656.2103667}
\label{pi-combinators}}
\end{figure*}

\begin{figure*}[ht]
\[\begin{array}{c}
\Rule{}
{c_1 : \tau_1 \iso \tau_2 \quad c_2 : \tau_2 \iso \tau_3 \quad c_3 : \tau_3 \iso \tau_4}
{\jdg{}{}{\assocdl : c_1 \odot (c_2 \odot c_3) \isotwo (c_1 \odot c_2) \odot c_3 : \assocdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idldl : \idiso \odot c \isotwo c : \idldr}}
{}
\quad
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idrdl : c \odot \idiso \isotwo c : \idrdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\rinvdl : ~! c \odot c \isotwo \idiso : \rinvdr}}
{}
\quad
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\linvdl : c ~\odot ~! c \isotwo \idiso : \linvdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idisotwo : c \isotwo c}}
{}
\quad
\Rule{}
{c_1,c_2,c_3 : \tau_1 \iso \tau_2 \quad \alpha_1 : c_1 \isotwo c_2 \quad \alpha_2 : c_2 \isotwo c_3}
{\jdg{}{}{\transtwo ~\alpha_1 ~\alpha_2 : c_1 \isotwo c_3}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\alpha_1 \respstwo \alpha_2 : c_1 \odot c_2 \isotwo c_3 \odot c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respptwo ~\alpha_1 ~\alpha_2 : c_1 \oplus c_2 \isotwo c_3 \oplus c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respttwo ~\alpha_1 ~\alpha_2 : c_1 \otimes c_2 \isotwo c_3 \otimes c_4}}
{}
\end{array}\]

Each 2-combinator $\alpha$ has an inverse $2!~\alpha$, e.g, $2!~\assocdl=\assocdr$,
$2!(\transtwo~\alpha_1~\alpha_2) = \transtwo~(2!~\alpha_2)~(2!~\alpha_1)$, etc.

\caption{$\Pi$-combinators~\cite{James:2012:IE:2103656.2103667}
\label{pi-combinators2}}
\end{figure*}

\begin{figure*}[ht]
\begin{multicols}{2}
\[\begin{array}{r@{\!}lcl}
%% \evalone{\unitepl}{&(\inl{()})} \\
\evalone{\unitepl}{&(\inr{v})} &=& v \\
\evalone{\unitipl}{&v} &=& \inr{v} \\
\evalone{\unitepr}{&(\inl{v})} &=& v \\
%% \evalone{\unitepr}{&(\inr{()})} \\
\evalone{\unitipr}{&v} &=& \inl{v} \\
\evalone{\swapp}{&(\inl{v})} &=& \inr{v} \\
\evalone{\swapp}{&(\inr{v})} &=& \inl{v} \\
\evalone{\assoclp}{&(\inl{v})} &=& \inl{(\inl{v})} \\
\evalone{\assoclp}{&(\inr{(\inl{v})})} &=& \inl{(\inr{v})} \\
\evalone{\assoclp}{&(\inr{(\inr{v})})} &=& \inr{v} \\
\evalone{\assocrp}{&(\inl{(\inl{v})})} &=& \inl{v} \\
\evalone{\assocrp}{&(\inl{(\inr{v})})} &=& \inr{(\inl{v})} \\
\evalone{\assocrp}{&(\inr{v})} &=& \inr{(\inr{v})} \\
\\
\evalone{\unitetl}{&(\unitv,v)} &=& v \\
\evalone{\unititl}{&v} &=& (\unitv,v) \\
\evalone{\unitetr}{&(v,\unitv)} &=& v \\
\evalone{\unititr}{&v} &=& (v,\unitv) \\
\evalone{\swapt}{&(v_1,v_2)} &=& (v_2,v_1) \\
\evalone{\assoclt}{&(v_1,(v_2,v_3))} &=& ((v_1,v_2),v_3) \\
\evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3)) \\
\end{array}\]
\[\begin{array}{r@{\!}lcl}
\evalone{\absorbr}{&(v,\_)} &=& v \\
\evalone{\absorbl}{&(\_,v)} &=& v \\
%% \evalone{\factorzl}{&()} \\
%% \evalone{\factorzr}{&()} \\
\evalone{\dist}{&(\inl{v_1},v_3)} &=& \inl{(v_1,v_3)} \\
\evalone{\dist}{&(\inr{v_2},v_3)} &=& \inr{(v_2,v_3)} \\
\evalone{\factor}{&\inl{(v_1,v_3)}} &=& (\inl{v_1},v_3) \\
\evalone{\factor}{&\inr{(v_2,v_3)}} &=& (\inr{v_2},v_3) \\
\evalone{\distl}{&(v_1,\inl{v_3})} &=& \inl{(v_1,v_3)} \\
\evalone{\distl}{&(v_2,\inr{v_3})} &=& \inr{(v_2,v_3)} \\
\evalone{\factorl}{&\inl{(v_1,v_3)}} &=& (v_1,\inl{v_3}) \\
\evalone{\factorl}{&\inr{(v_2,v_3)}} &=& (v_2,\inr{v_3}) \\
\\
\evalone{\idiso}{&v} &=& v \\
\evalone{(c_1 \odot c_2)}{&v} &=& 
  \evalone{c_2}{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inl{v})} &=& 
  \inl{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inr{v})} &=& 
  \inr{(\evalone{c_2}{v})} \\
\evalone{(c_1 \otimes c_2)}{&(v_1,v_2)} &=& 
  (\evalone{c_1}v_1, \evalone{c_2}v_2) 
\end{array}\]
\end{multicols}
\caption{\label{opsem}Operational Semantics}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%
\subsection{Reversibility}

The relevance of reversibility to HoTT is based on the following
analysis. The conventional HoTT approach starts with two, a priori, different
notions: functions and paths, and then postulates an equivalence between a
particular class of functions and paths. As illustrated above, some functions
like \AgdaFunction{not} correspond to paths. Most functions, however, are
evidently unrelated to paths. In particular, any function of type
\AgdaBound{A}~\AgdaSymbol{→}~\AgdaBound{B} that does not have an inverse of
type \AgdaBound{B}~\AgdaSymbol{→}~\AgdaBound{A} cannot have any direct
correspondence to paths as all paths have inverses. An interesting question
then poses itself: since reversible computational models --- in which all
functions have inverses --- are known to be universal computational models,
what would happen if we considered a variant of HoTT based exclusively on
reversible functions?  Presumably in such a variant, all functions --- being
reversible --- would potentially correspond to paths and the distinction
between the two notions would vanish making the univalence postulate
unnecessary. This is the precise technical idea we investigate in detail in
the remainder of the paper.

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Syntax of $\Pi$}
\label{opsempi}

The $\Pi$ family of languages is based on type isomorphisms. In the variant
we consider, the set of types $\tau$ includes the empty type $\zt$, the unit type
$\ot$, and conventional sum and product types. The values classified by these
types are the conventional ones: $\unitv$ of type $\ot$, $\inl{v}$ and $\inr{v}$ for
injections into sum types, and $(v_1,v_2)$ for product types:
\[\begin{array}{lrcl}
(\textrm{Types}) & 
  \tau &::=& \zt \alt \ot \alt \tau_1 \oplus \tau_2 \alt \tau_1 \otimes \tau_2 \\
(\textrm{Values}) & 
  v &::=& \unitv \alt \inl{v} \alt \inr{v} \alt (v_1,v_2) \\
(\textrm{1-combinators}) & 
  c &:& \tau_1 \iso \tau_2 ~ [\textit{see Fig.~\ref{pi-combinators}}] \\
(\textrm{2-combinators}) &
  \alpha &:& c_1 \isotwo c_2 \mbox{~where~} c_1, c_2 : \tau_1 \iso \tau_2 \\
  & && [\textit{see Fig.~\ref{pi-combinators2}}]
\end{array}\]
The interesting syntactic category of $\Pi$ is that of \emph{combinators}
which are witnesses for type isomorphisms $\tau_1 \iso \tau_2$. They consist
of base combinators (on the left side of Fig.~\ref{pi-combinators}) and
compositions (on the right side of the same figure). Each line of the figure
on the left introduces a pair of dual constants\footnote{where $\swapp$ and
$\swapt$ are self-dual.} that witness the type isomorphism in the
middle. This set of isomorphisms is known to be
complete~\cite{Fiore:2004,fiore-remarks} and the language is universal for
hardware combinational
circuits~\cite{James:2012:IE:2103656.2103667}.\footnote{If recursive types
and a trace operator are added, the language becomes Turing
complete~\cite{James:2012:IE:2103656.2103667,rc2011}. We will not be
concerned with this extension in the main body of this paper but it will be
briefly discussed in the conclusion.}

For $c_1, c_2 : \tau_1\iso\tau_2$, we have level-2 combinators $\alpha
: c_1 \isotwo c_2$ which are equivalences of isomorphisms, and which
happen to correspond to the coherence conditions for rig
groupoids. Many of the level 2 combinators not used. We only present
the ones we use in this paper.

\begin{proposition}
For any $c : \tau_1 \iso \tau_2$, we have $c \isotwo ~!~(!~c)$.
\end{proposition}

\begin{proposition}
For any $c_1,c_2 : \tau_1 \iso \tau_2$, we have $c_1 \isotwo c_2$ implies
$!c_1 \isotwo ~!c_2$.
\end{proposition}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Operational Semantics}

We give an operational semantics for $\Pi$.  Operationally, the
semantics consists of a pair of mutually recursive evaluators that
take a combinator and a value and propagate the value in the
``forward'' $\triangleright$ direction or in the
``backwards''~$\triangleleft$ direction. We show the complete forward
evaluator in Fig.~\ref{opsem}; the backwards evaluator differs in
trivial ways.

Every 1-combinator has an \emph{order} $n$: applying this combinator $n$ times is equivalent to the identity

A couple of examples: compute order

The type $(\ot \oplus \ot) \oplus \ot$ has cardinality 3 and hence has
6 automorphisms on it:

\[\begin{array}{rcl}
a_1 &=& \idiso \\
a_2 &=& \swapp \oplus \idiso \\
a_3 &=& \assocrp \odot (\idiso \oplus \swapp) \odot \assoclp \\
a_4 &=& a_2 \odot a_3 \\
a_5 &=& a_3 \odot a_2 \\
a_6 &=& a_4 \odot a_2
\end{array}\]

The combinators realize all the 3! permutations on the set of
cardinality 3. The permutation $a_1$ is the identity permutation; the
permutations $a_2$, $a_3$, and $a_6$ swap two of the three elements
leaving the third intact; the permutations $a_4$ and $a_5$ rotate the
three elements. A crucial property of combinators is their
\emph{order} which classifies them according to the number of
iterations needed for them to compose to the identity. In our case,
the reader is invited to check the following: $\mathit{order}(a_1)=1$,
$\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
$\mathit{order}(a_4)=\mathit{order}(a_5)=3$. Generally speaking to
calculate the order of a combinator $c$, we apply $c$ to every value
and observe the length of the orbit. The order is the lcm of all the
orbits.

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Groupoid Model}

From the perspective of category theory, the language $\Pi$ models what is
called a \emph{symmetric bimonoidal category} or a \emph{commutative rig
category}. These are categories with two binary operations and satisfying the
axioms of a commutative rig (i.e., a commutative ring without negative
elements also known as a commutative semiring) up to coherent
isomorphisms. And indeed the types of the $\Pi$-combinators are precisely the
commutative semiring axioms. A formal way of saying this is that $\Pi$ is the
\emph{categorification}~\cite{math/9802029} of the natural numbers. A simple
(slightly degenerate) example of such categories is the category of finite
sets and permutations in which we interpret every $\Pi$-type as a finite set,
interpret the values as elements in these finite sets, and interpret the
combinators as permutations.

Explain that we only reach a trivial class of groupoids

introduce discreteC and discreteG

discrete category: give a set $S$ objects are $S$, morphisms are only
identity morphisms; each object has exactly one morphism id. That is a
groupoid, trivially, because id is symmetric

Instead of modeling the types of $\Pi$ using sets and the combinators using
permutations we use a semantics that identifies $\Pi$-combinators with
\emph{paths}. More precisely, we model the universe of $\Pi$-types as a space
\AgdaFunction{U} whose points are the individual $\Pi$-types (which are
themselves spaces \AgdaBound{t} containing points). We then postulate that
there is path between the spaces \AgdaBound{t₁} and \AgdaBound{t₂} if there
is a $\Pi$-combinator $c : t_1 \iso t_2$. Our postulate is similar in spirit
to the univalence axiom but, unlike the latter, it has a simple computational
interpretation. A path directly corresponds to a type isomorphism with a
clear operational semantics as presented in the previous section. As we will
explain in more detail below, this approach replaces the datatype
\AgdaSymbol{≡} modeling propositional equality with the datatype
\AgdaSymbol{⟷} modeling type isomorphisms. With this switch, the
$\Pi$-combinators of Fig.~\ref{pi-combinators} become \emph{syntax} for the
paths in the space $U$. Put differently, instead of having exactly one
constructor \AgdaInductiveConstructor{refl} for paths with all other paths
discovered by proofs (see Secs. 2.5--2.12 of the HoTT
book~\citeyearpar{hottbook}) or postulated by the univalence axiom, we have
an \emph{inductive definition} that completely specifies all the paths in the
space $U$.

We begin with the datatype definition of the universe \AgdaFunction{U} of finite
types which are constructed using \AgdaInductiveConstructor{ZERO},
\AgdaInductiveConstructor{ONE}, \AgdaInductiveConstructor{PLUS}, and
\AgdaInductiveConstructor{TIMES}. Each of these finite types will correspond
to a set of points with paths connecting some of the points. The underlying
set of points is computed by \AgdaSymbol{⟦}\_\AgdaSymbol{⟧} as follows:
\AgdaInductiveConstructor{ZERO} maps to the empty set~\AgdaSymbol{⊥},
\AgdaInductiveConstructor{ONE} maps to the singleton set \AgdaSymbol{⊤},
\AgdaInductiveConstructor{PLUS} maps to the disjoint union \AgdaSymbol{⊎},
and \AgdaInductiveConstructor{TIMES} maps to the cartesian product
\AgdaSymbol{×}.

We note that the refinement of the $\Pi$-combinators to combinators on
pointed spaces is given by an inductive family for
\emph{heterogeneous} equality that generalizes the usual inductive
family for propositional equality. Put differently, what used to be
the only constructor for paths \AgdaInductiveConstructor{refl} is now
just one of the many constructors (named
\AgdaInductiveConstructor{id⟷} in the figure). Among the new
constructors we have \AgdaInductiveConstructor{◎} that constructs path
compositions. By construction, every combinator has an inverse
calculated as shown in Fig.~\ref{sym}. These constructions are
sufficient to guarantee that the universe~\AgdaFunction{U} is a
groupoid. Additionally, we have paths that connect values in different
but isomorphic spaces like \ldots

The example \AgdaFunction{notpath} which earlier required the use of the
univalence axiom can now be directly defined using
\AgdaInductiveConstructor{swap1₊} and \AgdaInductiveConstructor{swap2₊}. To
see this, note that \AgdaPrimitiveType{Bool} can be viewed as a shorthand for
\AgdaInductiveConstructor{PLUS} \AgdaInductiveConstructor{ONE}
\AgdaInductiveConstructor{ONE} with \AgdaInductiveConstructor{true} and
\AgdaInductiveConstructor{false} as shorthands for
\AgdaInductiveConstructor{inj₁} \AgdaInductiveConstructor{tt} and
\AgdaInductiveConstructor{inj₂} \AgdaInductiveConstructor{tt}. With this in
mind, the path corresponding to boolean negation consists of two ``fibers'',
one for each boolean value as shown below:

\noindent In other words, a path between spaces is really a collection of
paths connecting the various points. Note however that we never need to
``collect'' these paths using a universal quantification.

Shouldn't we also show that \AgdaPrimitiveType{Bool} contains exactly
$2$ things, and that TRUE and FALSE are ``different''? The other thing
is, whereas not used to be a path between Bool and Bool, we no longer
have that.  Shouldn't we show that, somehow, BOOL and BOOL.F 'union'
BOOL.T are somehow ``equivalent''?  And there there is a coherent
notpath built the same way?  Especially since I am sure it is quite
easy to build incoherent sets of paths!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-06-09 pibackground.tex
making progress on sec 3

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Background}
+\section{Programming with Equivalences}
 
-The main syntactic vehicle for the technical developments in this paper is a
-simple language called $\Pi$ whose only computations are isomorphisms between
-finite types~\cite{James:2012:IE:2103656.2103667}. After reviewing the
-motivation for this language and its relevance to HoTT, we present its syntax
-and semantics.
+The main syntactic vehicle for the technical developments in this
+paper is the language $\Pi$ whose only computations are isomorphisms
+between finite types and equivalences between these
+isomorphisms~\cite{James:2012:IE:2103656.2103667,Carette2016}. We
+present the syntax and operational and denotational semantics of
+the parts of the language relevant to our development.
 
@@
 $!(c_1 \odot c_2) = ~!c_2 \odot~ !c_1$, etc.
-\caption{$\Pi$-combinators~\cite{James:2012:IE:2103656.2103667}
+\caption{$\Pi$ 1-combinators~\cite{James:2012:IE:2103656.2103667}
 \label{pi-combinators}}
@@
 \[\begin{array}{c}
+\Rule{}
+{c : \tau_1 \iso \tau_2}
+{\jdg{}{}{\idisotwo : c \isotwo c}}
+{}
+~
+\Rule{}
+{c_1,c_2,c_3 : \tau_1 \iso \tau_2 \quad \alpha_1 : c_1 \isotwo c_2 \quad \alpha_2 : c_2 \isotwo c_3}
+{\jdg{}{}{\transtwo ~\alpha_1 ~\alpha_2 : c_1 \isotwo c_3}}
+{}
+~
 \Rule{}
@@
 {}
-\quad
+~
 \Rule{}
@@
 {}
-\\
-\\
+~
 \Rule{}
@@
 \Rule{}
-{c : \tau_1 \iso \tau_2}
-{\jdg{}{}{\idisotwo : c \isotwo c}}
-{}
-\quad
-\Rule{}
-{c_1,c_2,c_3 : \tau_1 \iso \tau_2 \quad \alpha_1 : c_1 \isotwo c_2 \quad \alpha_2 : c_2 \isotwo c_3}
-{\jdg{}{}{\transtwo ~\alpha_1 ~\alpha_2 : c_1 \isotwo c_3}}
-{}
-\\
-\\
-\Rule{}
 {c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
   \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
-{\jdg{}{}{\alpha_1 \respstwo \alpha_2 : c_1 \odot c_2 \isotwo c_3 \odot c_4}}
+{\jdg{}{}{\alpha_1 ~\respstwo~ \alpha_2 : c_1 \odot c_2 \isotwo c_3 \odot c_4}}
 {}
@@
 
-\caption{$\Pi$-combinators~\cite{James:2012:IE:2103656.2103667}
+\caption{$\Pi$ 2-combinators (excerpt)~\cite{Carette2016}
 \label{pi-combinators2}}
@@
 \end{multicols}
-\caption{\label{opsem}Operational Semantics}
+\caption{\label{opsem}$\Pi$ operational semantics}
 \end{figure*}
-
-%%%%%%%%%%%%%%%%%%%%%
-\subsection{Reversibility}
-
-The relevance of reversibility to HoTT is based on the following
-analysis. The conventional HoTT approach starts with two, a priori, different
-notions: functions and paths, and then postulates an equivalence between a
-particular class of functions and paths. As illustrated above, some functions
-like \AgdaFunction{not} correspond to paths. Most functions, however, are
-evidently unrelated to paths. In particular, any function of type
-\AgdaBound{A}~\AgdaSymbol{→}~\AgdaBound{B} that does not have an inverse of
-type \AgdaBound{B}~\AgdaSymbol{→}~\AgdaBound{A} cannot have any direct
-correspondence to paths as all paths have inverses. An interesting question
-then poses itself: since reversible computational models --- in which all
-functions have inverses --- are known to be universal computational models,
-what would happen if we considered a variant of HoTT based exclusively on
-reversible functions?  Presumably in such a variant, all functions --- being
-reversible --- would potentially correspond to paths and the distinction
-between the two notions would vanish making the univalence postulate
-unnecessary. This is the precise technical idea we investigate in detail in
-the remainder of the paper.
 
@@
 
-The $\Pi$ family of languages is based on type isomorphisms. In the variant
-we consider, the set of types $\tau$ includes the empty type $\zt$, the unit type
-$\ot$, and conventional sum and product types. The values classified by these
-types are the conventional ones: $\unitv$ of type $\ot$, $\inl{v}$ and $\inr{v}$ for
-injections into sum types, and $(v_1,v_2)$ for product types:
+The $\Pi$ family of languages is based on type isomorphisms. In the
+variant we consider, the set of types $\tau$ includes the empty
+type~$\zt$, the unit type $\ot$, and sum $\oplus$ and product
+$\otimes$ types. The values classified by these types are the
+conventional ones: $\unitv$ of type~$\ot$, $\inl{v}$ and $\inr{v}$ for
+injections into sum types, and $(v_1,v_2)$ for product types. The
+language has two other syntactic categories of programs to be
+described in detail:
 \[\begin{array}{lrcl}
@@
 \end{array}\]
-The interesting syntactic category of $\Pi$ is that of \emph{combinators}
-which are witnesses for type isomorphisms $\tau_1 \iso \tau_2$. They consist
-of base combinators (on the left side of Fig.~\ref{pi-combinators}) and
-compositions (on the right side of the same figure). Each line of the figure
-on the left introduces a pair of dual constants\footnote{where $\swapp$ and
-$\swapt$ are self-dual.} that witness the type isomorphism in the
-middle. This set of isomorphisms is known to be
-complete~\cite{Fiore:2004,fiore-remarks} and the language is universal for
-hardware combinational
-circuits~\cite{James:2012:IE:2103656.2103667}.\footnote{If recursive types
-and a trace operator are added, the language becomes Turing
-complete~\cite{James:2012:IE:2103656.2103667,rc2011}. We will not be
-concerned with this extension in the main body of this paper but it will be
-briefly discussed in the conclusion.}
 
-For $c_1, c_2 : \tau_1\iso\tau_2$, we have level-2 combinators $\alpha
-: c_1 \isotwo c_2$ which are equivalences of isomorphisms, and which
-happen to correspond to the coherence conditions for rig
-groupoids. Many of the level 2 combinators not used. We only present
-the ones we use in this paper.
+Both classes of programs, 1-combinators $c$ and
+2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT
+sense. The elements $c$ of 1-combinators denote type isomorphisms. The
+elements $\alpha$ of 2-combinators denote equivalences between these
+type isomorphisms. Using the 1-combinators, it is possible to write
+any reversible boolean function and hence encode arbitrary boolean
+functions by a technique that goes back to \citet{Toffoli:1980}. The
+2-combinators provide a layer of programs that computes semantic
+preserving transformations of 1-combinators. As a small example, let
+us abbreviate $\ot \oplus \ot$ as $\mathsf{Bool}$ and examine two
+possible implementations of boolean negation. The first directly uses
+the primitive combinator
+$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to swap the
+two values of type $\mathsf{Bool}$; the second use three consecutive
+copies of $\swapp$ to achieve the same effect:
+\[\begin{array}{rcl}
+\mathsf{not_1} &=& \swapp \\
+\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp 
+\end{array}\]
+We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
+\isotwo \mathsf{not_1}$:
+\[
+\transtwo~(\idldl ~\respstwo~ \idisotwo)~\idldl
+\]
+which not only shows the equivalence of the two implementations of
+negation but also shows \emph{how} to transform one to the other.
+
+Fig.~\ref{pi-combinators} lists all the 1-combinators which consist of
+base combinators (on the left) and compositions (on the right). Each
+line of the base combinators introduces a pair of dual
+constants\footnote{where $\swapp$ and $\swapt$ are self-dual.} that
+witness the type isomorphism in the middle. This set of isomorphisms
+is known to be sound and
+complete~\cite{Fiore:2004,fiore-remarks}. Fig.~\ref{pi-combinators2}
+lists a few of the 2-combinators that we use in this paper. Each
+2-combinator relates two 1-combinators of the same type and witnesses
+their equivalence. The 2-equivalences behave as expected with respect
+to inverses of 1-combinators as shown below.
 
@@
 semantics consists of a pair of mutually recursive evaluators that
-take a combinator and a value and propagate the value in the
-``forward'' $\triangleright$ direction or in the
-``backwards''~$\triangleleft$ direction. We show the complete forward
-evaluator in Fig.~\ref{opsem}; the backwards evaluator differs in
-trivial ways.
+take a combinator and a value and propagate the value in the forward
+direction~$\triangleright$ or in the backward
+direction~$\triangleleft$. We show the complete forward evaluator in
+Fig.~\ref{opsem}; the backward evaluator is easy to infer.
 
-Every 1-combinator has an \emph{order} $n$: applying this combinator $n$ times is equivalent to the identity
-
-A couple of examples: compute order
-
-The type $(\ot \oplus \ot) \oplus \ot$ has cardinality 3 and hence has
-6 automorphisms on it:
+As an example, let $\mathbb{3}$ abbreviate the type
+$(\ot \oplus \ot) \oplus \ot$ and consider the following six
+combinators of type $\mathbb{3} \iso \mathbb{3}$:
 
@@
 \end{array}\]
+
+There are three values of type $\mathbb{3}$ which are
+$\inl{\inl{\unitv}}$, $\inl{\inr{\unitv}}$, and
+$\inr{\unitv}$. Evaluting $a_2$ on each of these values, we get:
+\[\begin{array}{rcl}
+\evalone{(\swapp\oplus\idiso)}{\inl{\inl{\unitv}}} &=& \inl{\evalone{\swapp}{\inl{\unitv}}} \\
+&=& \inl{\inr{\unitv}} \\
+\\
+\evalone{(\swapp\oplus\idiso)}{\inl{\inr{\unitv}}} &=& \inl{\evalone{\swapp}{\inr{\unitv}}} \\
+&=& \inl{\inl{\unitv}} \\
+\\
+\evalone{(\swapp\oplus\idiso)}{\inr{\unitv}} &=& \inr{\evalone{\idiso}{\unitv}} \\
+&=& \inr{\unitv}
+\end{array}\]
+
+Every 1-combinator has an \emph{order} $n$: applying this combinator
+$n$ times is equivalent to the identityA couple of examples: compute order
+The type $(\ot \oplus \ot) \oplus \ot$ has cardinality 3 and hence has
+6 automorphisms on it:
 
@@
 \subsection{Groupoid Model}
+
+For $c_1, c_2 : \tau_1\iso\tau_2$, we have level-2 combinators $\alpha
+: c_1 \isotwo c_2$ which are equivalences of isomorphisms, and which
+happen to correspond to the coherence conditions for rig
+groupoids. Many of the level 2 combinators not used. We only present
+the ones we use in this paper.
 

### 2016-06-10 pibackground.tex
done with sec 3 for now; will start sec 4

@@
 isomorphisms~\cite{James:2012:IE:2103656.2103667,Carette2016}. We
-present the syntax and operational and denotational semantics of
-the parts of the language relevant to our development.
+present the syntax and operational semantics of the parts of the
+language relevant to our development.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Operational Semantics}
+\subsection{Operational Semantics and the Order of 1-Combinators}
 
-We give an operational semantics for $\Pi$.  Operationally, the
-semantics consists of a pair of mutually recursive evaluators that
-take a combinator and a value and propagate the value in the forward
+We give an operational semantics for the 1-combinators of $\Pi$ which
+represent ``conventional'' programs.  Operationally, the semantics
+consists of a pair of mutually recursive evaluators that take a
+combinator and a value and propagate the value in the forward
 direction~$\triangleright$ or in the backward
@@
 As an example, let $\mathbb{3}$ abbreviate the type
-$(\ot \oplus \ot) \oplus \ot$ and consider the following six
-combinators of type $\mathbb{3} \iso \mathbb{3}$:
-
+$(\ot \oplus \ot) \oplus \ot$. There are three values of type
+$\mathbb{3}$ which are $\inl{\inl{\unitv}}$, $\inl{\inr{\unitv}}$, and
+$\inr{\unitv}$. There are six combinators of type
+$\mathbb{3} \iso \mathbb{3}$, each representing a different
+permutation of three elements, which could be written as follows:
 \[\begin{array}{rcl}
@@
 \end{array}\]
-
-There are three values of type $\mathbb{3}$ which are
-$\inl{\inl{\unitv}}$, $\inl{\inr{\unitv}}$, and
-$\inr{\unitv}$. Evaluting $a_2$ on each of these values, we get:
+Tracing the evaluation of $a_2$ on each of the possible inputs yields:
 \[\begin{array}{rcl}
@@
 \end{array}\]
+Thus the effect of combinator $a_2$ is to swap the values
+$\inl{\inl{\unitv}}$ and $\inl{\inr{\unitv}}$ leaving the value
+$\inr{\unitv}$ intact. Iterating $a_2$ again is therefore equivalent
+to the identity permutation, which can be verified using the full set of
+2-combinators~\cite{Carette2016}:
+\[\begin{array}{rcl}
+a_2 \odot a_2 &=& (\swapp \oplus \idiso) \odot (\swapp \oplus \idiso) \\
+&\isotwo& (\swapp \odot \swapp) \oplus (\idiso \odot \idiso) \\
+&\isotwo& \idiso \oplus \idiso \\
+&=& \idiso
+\end{array}\]
 
-Every 1-combinator has an \emph{order} $n$: applying this combinator
-$n$ times is equivalent to the identityA couple of examples: compute order
-The type $(\ot \oplus \ot) \oplus \ot$ has cardinality 3 and hence has
-6 automorphisms on it:
-
-The combinators realize all the 3! permutations on the set of
-cardinality 3. The permutation $a_1$ is the identity permutation; the
-permutations $a_2$, $a_3$, and $a_6$ swap two of the three elements
-leaving the third intact; the permutations $a_4$ and $a_5$ rotate the
-three elements. A crucial property of combinators is their
-\emph{order} which classifies them according to the number of
-iterations needed for them to compose to the identity. In our case,
-the reader is invited to check the following: $\mathit{order}(a_1)=1$,
+A similar calculation shows that the combinator $a_1$ is the identity
+permutation; the combinators $a_3$ and $a_6$ also swap two of the
+three elements leaving the third intact; and the combinators $a_4$ and
+$a_5$ rotate the three elements. We therefore have
+$\mathit{order}(a_1)=1$,
 $\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
-$\mathit{order}(a_4)=\mathit{order}(a_5)=3$. Generally speaking to
-calculate the order of a combinator $c$, we apply $c$ to every value
-and observe the length of the orbit. The order is the lcm of all the
-orbits.
-
-%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Groupoid Model}
-
-For $c_1, c_2 : \tau_1\iso\tau_2$, we have level-2 combinators $\alpha
-: c_1 \isotwo c_2$ which are equivalences of isomorphisms, and which
-happen to correspond to the coherence conditions for rig
-groupoids. Many of the level 2 combinators not used. We only present
-the ones we use in this paper.
-
- From the perspective of category theory, the language $\Pi$ models what is
-called a \emph{symmetric bimonoidal category} or a \emph{commutative rig
-category}. These are categories with two binary operations and satisfying the
-axioms of a commutative rig (i.e., a commutative ring without negative
-elements also known as a commutative semiring) up to coherent
-isomorphisms. And indeed the types of the $\Pi$-combinators are precisely the
-commutative semiring axioms. A formal way of saying this is that $\Pi$ is the
-\emph{categorification}~\cite{math/9802029} of the natural numbers. A simple
-(slightly degenerate) example of such categories is the category of finite
-sets and permutations in which we interpret every $\Pi$-type as a finite set,
-interpret the values as elements in these finite sets, and interpret the
-combinators as permutations.
-
-Explain that we only reach a trivial class of groupoids
-
-introduce discreteC and discreteG
-
-discrete category: give a set $S$ objects are $S$, morphisms are only
-identity morphisms; each object has exactly one morphism id. That is a
-groupoid, trivially, because id is symmetric
-
-Instead of modeling the types of $\Pi$ using sets and the combinators using
-permutations we use a semantics that identifies $\Pi$-combinators with
-\emph{paths}. More precisely, we model the universe of $\Pi$-types as a space
-\AgdaFunction{U} whose points are the individual $\Pi$-types (which are
-themselves spaces \AgdaBound{t} containing points). We then postulate that
-there is path between the spaces \AgdaBound{t₁} and \AgdaBound{t₂} if there
-is a $\Pi$-combinator $c : t_1 \iso t_2$. Our postulate is similar in spirit
-to the univalence axiom but, unlike the latter, it has a simple computational
-interpretation. A path directly corresponds to a type isomorphism with a
-clear operational semantics as presented in the previous section. As we will
-explain in more detail below, this approach replaces the datatype
-\AgdaSymbol{≡} modeling propositional equality with the datatype
-\AgdaSymbol{⟷} modeling type isomorphisms. With this switch, the
-$\Pi$-combinators of Fig.~\ref{pi-combinators} become \emph{syntax} for the
-paths in the space $U$. Put differently, instead of having exactly one
-constructor \AgdaInductiveConstructor{refl} for paths with all other paths
-discovered by proofs (see Secs. 2.5--2.12 of the HoTT
-book~\citeyearpar{hottbook}) or postulated by the univalence axiom, we have
-an \emph{inductive definition} that completely specifies all the paths in the
-space $U$.
-
-We begin with the datatype definition of the universe \AgdaFunction{U} of finite
-types which are constructed using \AgdaInductiveConstructor{ZERO},
-\AgdaInductiveConstructor{ONE}, \AgdaInductiveConstructor{PLUS}, and
-\AgdaInductiveConstructor{TIMES}. Each of these finite types will correspond
-to a set of points with paths connecting some of the points. The underlying
-set of points is computed by \AgdaSymbol{⟦}\_\AgdaSymbol{⟧} as follows:
-\AgdaInductiveConstructor{ZERO} maps to the empty set~\AgdaSymbol{⊥},
-\AgdaInductiveConstructor{ONE} maps to the singleton set \AgdaSymbol{⊤},
-\AgdaInductiveConstructor{PLUS} maps to the disjoint union \AgdaSymbol{⊎},
-and \AgdaInductiveConstructor{TIMES} maps to the cartesian product
-\AgdaSymbol{×}.
-
-We note that the refinement of the $\Pi$-combinators to combinators on
-pointed spaces is given by an inductive family for
-\emph{heterogeneous} equality that generalizes the usual inductive
-family for propositional equality. Put differently, what used to be
-the only constructor for paths \AgdaInductiveConstructor{refl} is now
-just one of the many constructors (named
-\AgdaInductiveConstructor{id⟷} in the figure). Among the new
-constructors we have \AgdaInductiveConstructor{◎} that constructs path
-compositions. By construction, every combinator has an inverse
-calculated as shown in Fig.~\ref{sym}. These constructions are
-sufficient to guarantee that the universe~\AgdaFunction{U} is a
-groupoid. Additionally, we have paths that connect values in different
-but isomorphic spaces like \ldots
-
-The example \AgdaFunction{notpath} which earlier required the use of the
-univalence axiom can now be directly defined using
-\AgdaInductiveConstructor{swap1₊} and \AgdaInductiveConstructor{swap2₊}. To
-see this, note that \AgdaPrimitiveType{Bool} can be viewed as a shorthand for
-\AgdaInductiveConstructor{PLUS} \AgdaInductiveConstructor{ONE}
-\AgdaInductiveConstructor{ONE} with \AgdaInductiveConstructor{true} and
-\AgdaInductiveConstructor{false} as shorthands for
-\AgdaInductiveConstructor{inj₁} \AgdaInductiveConstructor{tt} and
-\AgdaInductiveConstructor{inj₂} \AgdaInductiveConstructor{tt}. With this in
-mind, the path corresponding to boolean negation consists of two ``fibers'',
-one for each boolean value as shown below:
-
-\noindent In other words, a path between spaces is really a collection of
-paths connecting the various points. Note however that we never need to
-``collect'' these paths using a universal quantification.
-
-Shouldn't we also show that \AgdaPrimitiveType{Bool} contains exactly
-$2$ things, and that TRUE and FALSE are ``different''? The other thing
-is, whereas not used to be a path between Bool and Bool, we no longer
-have that.  Shouldn't we show that, somehow, BOOL and BOOL.F 'union'
-BOOL.T are somehow ``equivalent''?  And there there is a coherent
-notpath built the same way?  Especially since I am sure it is quite
-easy to build incoherent sets of paths!
+$\mathit{order}(a_4)=\mathit{order}(a_5)=3$.
 

### 2016-06-19 pibackground.tex
sec 5.1 help

@@
 \subsection{Operational Semantics and the Order of 1-Combinators}
+\label{sec:pisem}
 
@@
 
+\begin{definition}[Power of a permutation]
+  The $k^{th}$ power of a 1-combinator $p : \tau \iso \tau$, for
+  $k \in \Z$ is
+  \[
+    p^k =
+  \begin{cases}
+    \idiso & k = 0 \\
+    p \odot p^{k - 1} & k > 0 \\
+    (!~p) \odot p^{k + 1} & k < 0 \\
+  \end{cases}
+  \]
+\end{definition}
+
+\begin{definition}[Order of a permutation]
+  The order of a 1-combinator $p : \tau \iso \tau$, $\ord{p}$, is the
+  least postitive natural number $k \in \N^+$ such that
+  $p^k \isotwo \idiso$.
+\end{definition}
+
+% \begin{lemma}
+%   Every $p : \tau \iso \tau$ has an order.
+% \end{lemma}
+
+% \begin{proof}
+%   By cases.
+%   \begin{enumerate}
+%   \item $\ord{\idiso} = 1$
+%   \item $\ord{\swapp} = \ord{\swapt} = 2$
+%   \item $\ord{p_1 \odot p_2} = ???$
+%   \item $\ord{p_1 \oplus p_2} = \ord{p_1 \otimes p_2} = \mathsf{lcm}(\ord{p_1}, \ord{p_2})$
+%   \end{enumerate}
+% \end{proof}
+
+\begin{lemma}
+  For $p : \tau \iso \tau$, $n \in \Z$, $p^{k + n} \isotwo p^n$ where
+  $k = \ord{p}$.
+\end{lemma}
+
+\begin{proof}
+  Trivial.
+\end{proof}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-06-24 pibackground.tex

@@
 $\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
-$\mathit{order}(a_4)=\mathit{order}(a_5)=3$.
+$\mathit{order}(a_4)=\mathit{order}(a_5)=3$. More generally, we define
+the power and order of 1-combinators as follows.
 
-\begin{definition}[Power of a permutation]
+\begin{definition}[Power of a 1-combinator]
   The $k^{th}$ power of a 1-combinator $p : \tau \iso \tau$, for
@@
 
-\begin{definition}[Order of a permutation]
+\begin{definition}[Order of a 1-combinator]
   The order of a 1-combinator $p : \tau \iso \tau$, $\ord{p}$, is the
@@
 
-\begin{proof}
-  Trivial.
-\end{proof}
+% \begin{proof}
+%   Trivial.
+% \end{proof}
 

### 2016-06-26 pibackground.tex
4.5
done with first pass over sec 4

@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Operational Semantics and the Order of 1-Combinators}
+\subsection{Semantics}
 \label{sec:pisem}
@@
 
+\begin{lemma}
+\label{lem:distiterplus}
+  For $p : \tau\iso\tau$, $m,n\in\Z$, we have a 2-combinator
+  $\distiterplus{p}{m}{n} : (p^m \odot p^n) \isotwo p ^{m + n}$.
+\end{lemma}
+
+For an information-theoretic perspective on the language $\Pi$, we
+think of a type containing $N$ values as an abstract system that
+has~$N$ distinguishable states. According to the conventional theory
+of information~\cite{Shannon1948}, the amount of information contained
+in each state of a system with $N$ distinguishable states is $\log N$
+bits of information. For example, the type $\mathsf{Bool}$ can be
+thought of as an abstract system with two distinguishable states
+labeled $\mathsf{true}$ and $\mathsf{false}$ each containing
+$\log 2 = 1$ bit. Similarly, the type $\mathbb{3}$ can be thought of
+as an abstract system with three distinguishable states each
+containing $\log 3$ bits. The logarithmic map implies that information
+contained in a composite state is the sum of the information contained
+in its constituents. For example, the type
+$\mathsf{Bool} \otimes \mathbb{3}$ can be thought of a composite
+system consisting of two independent unrelated subsystems. Each state
+of the composite system therefore contains
+$\log (2 * 3) = \log 2 + \log 3$ bits which is the sum of the
+information contained in each subsystem. Since all the combinators
+preserve cardinality, they are also information-preserving.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-07-01 pibackground.tex
messing with agda style etc
updated sec 3 a bit

@@
 \section{Programming with Equivalences} 
+\label{sec:pi}
 
@@
 
-Both classes of programs, 1-combinators $c$ and
+\noindent Both classes of programs, 1-combinators $c$ and
 2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT
@@
 functions by a technique that goes back to \citet{Toffoli:1980}. The
-2-combinators provide a layer of programs that computes semantic
-preserving transformations of 1-combinators. As a small example, let
-us abbreviate $\ot \oplus \ot$ as $\mathsf{Bool}$ and examine two
-possible implementations of boolean negation. The first directly uses
-the primitive combinator
+2-combinators provide a layer of programs that computes
+semantic-preserving transformations of 1-combinators. As a small
+example, let us abbreviate $\ot \oplus \ot$ as $\mathsf{Bool}$ and
+examine two possible implementations of boolean negation. The first
+directly uses the primitive combinator
 $\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to swap the
@@
 
-A similar calculation shows that the combinator $a_1$ is the identity
-permutation; the combinators $a_3$ and $a_6$ also swap two of the
-three elements leaving the third intact; and the combinators $a_4$ and
-$a_5$ rotate the three elements. We therefore have
-$\mathit{order}(a_1)=1$,
-$\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
-$\mathit{order}(a_4)=\mathit{order}(a_5)=3$. More generally, we define
-the power and order of 1-combinators as follows.
+More generally we can iterate 1-combinators to produce different
+reversible functions between finite sets, eventually wrapping around
+at some number which represents the \emph{order} of the underlying
+permutation.
 
@@
 
+For our example combinators on the type $\mathbb{3}$, simple traces
+using the operational semantics show the combinator $a_1$ is the
+identity permutation; the combinators $a_3$ and $a_6$ swap two of the
+three elements leaving the third intact; and the combinators $a_4$ and
+$a_5$ rotate the three elements. We therefore have
+$\mathit{order}(a_1)=1$,
+$\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
+$\mathit{order}(a_4)=\mathit{order}(a_5)=3$.
+
 % \begin{lemma}
@@
 
-\begin{lemma}
-  For $p : \tau \iso \tau$, $n \in \Z$, $p^{k + n} \isotwo p^n$ where
-  $k = \ord{p}$.
-\end{lemma}
+The 2-combinators, being complete equivalences between
+1-combinators~\cite{Carette2016}, also capture equivalences regarding
+power of combinators and their order.
 
@@
 
-For an information-theoretic perspective on the language $\Pi$, we
-think of a type containing $N$ values as an abstract system that
-has~$N$ distinguishable states. According to the conventional theory
-of information~\cite{Shannon1948}, the amount of information contained
-in each state of a system with $N$ distinguishable states is $\log N$
-bits of information. For example, the type $\mathsf{Bool}$ can be
-thought of as an abstract system with two distinguishable states
-labeled $\mathsf{true}$ and $\mathsf{false}$ each containing
-$\log 2 = 1$ bit. Similarly, the type $\mathbb{3}$ can be thought of
-as an abstract system with three distinguishable states each
-containing $\log 3$ bits. The logarithmic map implies that information
-contained in a composite state is the sum of the information contained
-in its constituents. For example, the type
-$\mathsf{Bool} \otimes \mathbb{3}$ can be thought of a composite
+\begin{lemma}
+  For $p : \tau \iso \tau$, $n \in \Z$, $p^{k + n} \isotwo p^n$ where
+  $k = \ord{p}$.
+\end{lemma}
+
+\paragraph{Intermezzo.} For an information-theoretic perspective on
+the language $\Pi$, we think of a type containing $N$ values as an
+abstract system that has~$N$ distinguishable states. According to the
+conventional theory of information~\cite{Shannon1948}, the amount of
+information contained in each state of a system with $N$
+distinguishable states is $\log N$ bits of information. For example,
+the type $\mathsf{Bool}$ can be thought of as an abstract system with
+two distinguishable states labeled $\mathsf{true}$ and
+$\mathsf{false}$ each containing $\log 2 = 1$ bit. Similarly, the type
+$\mathbb{3}$ can be thought of as an abstract system with three
+distinguishable states each containing $\log 3$ bits. The logarithmic
+map implies that information contained in a composite state is the sum
+of the information contained in its constituents. For example, the
+type $\mathsf{Bool} \otimes \mathbb{3}$ can be thought of a composite
 system consisting of two independent unrelated subsystems. Each state

### 2016-07-01 pibackground.tex
more on sec 3

@@
 {c_1,c_2,c_3 : \tau_1 \iso \tau_2 \quad \alpha_1 : c_1 \isotwo c_2 \quad \alpha_2 : c_2 \isotwo c_3}
-{\jdg{}{}{\transtwo ~\alpha_1 ~\alpha_2 : c_1 \isotwo c_3}}
+{\jdg{}{}{\alpha_1~\transtwo~\alpha_2 : c_1 \isotwo c_3}}
 {}
@@
 \Rule{}
+{}
+{\jdg{}{}{\sumid : \idiso\oplus\idiso \isotwo \idiso : \splitid}}
+{}
+~
+\Rule{}
+{c_1 : \tau_5\iso\tau_1 \quad c_2 : \tau_6\iso\tau_2 \quad c_3 :
+    \tau_1\iso\tau_3 \quad c_4 : \tau_2\iso\tau_4}
+{\jdg{}{}{\homps : (c_1\odot c_3)\oplus(c_2\odot c_4) \isotwo
+    (c_1\oplus c_2) \odot (c_3\oplus c_4) : \homsp }}
+{}
+\\
+\\
+\Rule{}
 {c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
@@
 \end{array}\]
-
 Each 2-combinator $\alpha$ has an inverse $2!~\alpha$, e.g, $2!~\assocdl=\assocdr$,
-$2!(\transtwo~\alpha_1~\alpha_2) = \transtwo~(2!~\alpha_2)~(2!~\alpha_1)$, etc.
-
+$2!(\alpha_1~\transtwo~\alpha_2) = (2!~\alpha_2)~\transtwo~(2!~\alpha_1)$, etc. 
 \caption{$\Pi$ 2-combinators (excerpt)~\cite{Carette2016}
@@
 \[
-\transtwo~(\idldl ~\respstwo~ \idisotwo)~\idldl
+(\idldl ~\respstwo~ \idisotwo)~\transtwo~\idldl
 \]
@@
 $\inr{\unitv}$ intact. Iterating $a_2$ again is therefore equivalent
-to the identity permutation, which can be verified using the full set of
-2-combinators~\cite{Carette2016}:
+to the identity permutation, which can be verified using the 
+2-combinators:
 \[\begin{array}{rcl}

### 2016-07-02 pibackground.tex

@@
 
-\noindent Both classes of programs, 1-combinators $c$ and
+\noindent Both classes of programs, 1-combinators $c$, and
 2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT
@@
 elements $\alpha$ of 2-combinators denote equivalences between these
-type isomorphisms. Using the 1-combinators, it is possible to write
-any reversible boolean function and hence encode arbitrary boolean
+type isomorphisms which are sound and complete 1-combinator program
+transformations. Using the 1-combinators, it is possible to write any
+reversible boolean function and hence encode arbitrary boolean
 functions by a technique that goes back to \citet{Toffoli:1980}. The
@@
 semantic-preserving transformations of 1-combinators. As a small
-example, let us abbreviate $\ot \oplus \ot$ as $\mathsf{Bool}$ and
-examine two possible implementations of boolean negation. The first
-directly uses the primitive combinator
+example, let us abbreviate $\ot \oplus \ot$ as the type $\mathbb{2}$
+of booleans and examine two possible implementations of boolean
+negation. The first directly uses the primitive combinator
 $\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to swap the
-two values of type $\mathsf{Bool}$; the second use three consecutive
+two values of type $\mathbb{2}$; the second use three consecutive
 copies of $\swapp$ to achieve the same effect:
@@
 2-combinator relates two 1-combinators of the same type and witnesses
-their equivalence. The 2-equivalences behave as expected with respect
-to inverses of 1-combinators as shown below.
+their equivalence. Both 1-combinators and 2-combinators are invertible
+and the 2-combinators behave as expected with respect to inverses of
+1-combinators as shown below.
 
@@
 We give an operational semantics for the 1-combinators of $\Pi$ which
-represent ``conventional'' programs.  Operationally, the semantics
-consists of a pair of mutually recursive evaluators that take a
-combinator and a value and propagate the value in the forward
+represent the conventional layer of programs.  Operationally, the
+semantics consists of a pair of mutually recursive evaluators that
+take a combinator and a value and propagate the value in the forward
 direction~$\triangleright$ or in the backward
@@
 distinguishable states is $\log N$ bits of information. For example,
-the type $\mathsf{Bool}$ can be thought of as an abstract system with
+the type $\mathbb{2}$ can be thought of as an abstract system with
 two distinguishable states labeled $\mathsf{true}$ and
@@
 of the information contained in its constituents. For example, the
-type $\mathsf{Bool} \otimes \mathbb{3}$ can be thought of a composite
+type $\mathbb{2} \otimes \mathbb{3}$ can be thought of a composite
 system consisting of two independent unrelated subsystems. Each state

### 2016-07-05 pibackground.tex
secs 1 2 3 4

@@
 $\mathbb{3}$ which are $\inl{\inl{\unitv}}$, $\inl{\inr{\unitv}}$, and
-$\inr{\unitv}$. There are six combinators of type
+$\inr{\unitv}$. There are, up to equivalence, six combinators of type
 $\mathbb{3} \iso \mathbb{3}$, each representing a different

### 2016-07-06 pibackground.tex
small changes in first sections

@@
 information contained in each state of a system with $N$
-distinguishable states is $\log N$ bits of information. For example,
+distinguishable states is $(\log N)$ bits of information. For example,
 the type $\mathbb{2}$ can be thought of as an abstract system with

### 2016-10-12 pibackground.tex
move forward with LIPIcs format

@@
 reversible boolean function and hence encode arbitrary boolean
-functions by a technique that goes back to \citet{Toffoli:1980}. The
+functions by a technique that goes back to \cite{Toffoli:1980}. The
 2-combinators provide a layer of programs that computes
-semantic-preserving transformations of 1-combinators. As a small
+semantics-preserving transformations of 1-combinators. As a small
 example, let us abbreviate $\ot \oplus \ot$ as the type $\mathbb{2}$

### 2016-11-28 A.hs
Add a dir-locals and enforce formatting everywhere

@@ -1,4 +1,3 @@
-
 module A where
 
 data Digit =
@@ -15,16 +14,16 @@ data Digit =
   deriving (Show,Eq)
 
 sucD :: Digit -> (Digit,Digit)
-sucD Zero   = (Zero,One)
-sucD One    = (Zero,Two)
-sucD Two    = (Zero,Three)
-sucD Three  = (Zero,Four)
-sucD Four   = (Zero,Five)
-sucD Five   = (Zero,Six)
-sucD Six    = (Zero,Seven)
-sucD Seven  = (Zero,Eight)
-sucD Eight  = (Zero,Nine)
-sucD Nine   = (One,Zero)
+sucD Zero  = (Zero,One)
+sucD One   = (Zero,Two)
+sucD Two   = (Zero,Three)
+sucD Three = (Zero,Four)
+sucD Four  = (Zero,Five)
+sucD Five  = (Zero,Six)
+sucD Six   = (Zero,Seven)
+sucD Seven = (Zero,Eight)
+sucD Eight = (Zero,Nine)
+sucD Nine  = (One,Zero)
 
 suc :: (Digit,Digit) -> (Digit,Digit)
 suc (d,u) = let (carry,u') = sucD u
@@ -34,4 +33,3 @@ suc (d,u) = let (carry,u') = sucD u
                     in if carry' == Zero
                        then (d',u')
                        else (Zero,Zero)
-

### 2016-11-28 AM.hs

@@ -5,45 +5,45 @@ import Debug.Trace
 
 -- Abstract Machine for PI/
 
-data Combinator a = 
+data Combinator a =
                   Zeroe | Zeroi
-                | SwapPlus 
+                | SwapPlus
                 | AssoclPlus | AssocrPlus
                 | Unite a | Uniti a
-                | SwapTimes 
+                | SwapTimes
                 | AssoclTimes | AssocrTimes
                 | DistribZero | FactorZero
                 | Distrib | Factor
-                | Id | Sym (Combinator a) 
+                | Id | Sym (Combinator a)
                 | Seq (Combinator a) (Combinator a)
-                | Plus (Combinator a) (Combinator a) 
+                | Plus (Combinator a) (Combinator a)
                 | Times (Combinator a) (Combinator a)
                 | Eta (Combinator a) | Epsilon a
   deriving (Eq,Show)
 
 -- a Pale imitation of <=>.  But enough for the tests.
 ceq :: Eq a => Combinator a -> Combinator a -> Bool
-ceq (Seq Id Id) Id = True
-ceq Id (Seq Id Id) = True
+ceq (Seq Id Id) Id               = True
+ceq Id (Seq Id Id)               = True
 ceq (Seq SwapTimes SwapTimes) Id = True
 ceq Id (Seq SwapTimes SwapTimes) = True
-ceq c1 c2 = c1 == c2
+ceq c1 c2                        = c1 == c2
 
-data Value a = Unit a | Inj1 (Value a) | Inj2 (Value a) 
+data Value a = Unit a | Inj1 (Value a) | Inj2 (Value a)
              | Pair (Value a) (Value a)
              | Comb (Combinator a) | Recip (Combinator a)
   deriving (Eq,Show)
 
-data Context a = 
+data Context a =
     Empty
   | Fst (Context a) (Combinator a) | Snd (Combinator a) (Context a)
-  | LeftTimes (Context a) (Combinator a) (Value a) 
+  | LeftTimes (Context a) (Combinator a) (Value a)
   | RightTimes (Combinator a) (Value a) (Context a)
   | LeftPlus (Context a) (Combinator a) | RightPlus (Combinator a) (Context a)
   deriving (Eq,Show)
 
-data State a = 
[…]

### 2016-11-28 AM2.hs

@@ -5,18 +5,18 @@ import Debug.Trace
 
 -- Abstract Machine for PI/
 
-data Combinator a = 
+data Combinator a =
                   Zeroe | Zeroi
-                | SwapPlus 
+                | SwapPlus
                 | AssoclPlus | AssocrPlus
                 | Unite a | Uniti a
-                | SwapTimes 
+                | SwapTimes
                 | AssoclTimes | AssocrTimes
                 | DistribZero | FactorZero
                 | Distrib | Factor
-                | Id | Sym (Combinator a) 
+                | Id | Sym (Combinator a)
                 | Seq (Combinator a) (Combinator a)
-                | Plus (Combinator a) (Combinator a) 
+                | Plus (Combinator a) (Combinator a)
                 | Times (Combinator a) (Combinator a)
                 | Eta (Combinator a) | Epsilon a CEq
   deriving (Eq,Show)
@@ -24,21 +24,21 @@ data Combinator a =
 data CEq = IdC | SwapSwapTimesL | SwapSwapTimesR
   deriving (Eq, Show)
 
-data Value a = Unit a | Inj1 (Value a) | Inj2 (Value a) 
+data Value a = Unit a | Inj1 (Value a) | Inj2 (Value a)
              | Pair (Value a) (Value a)
              | Comb (Combinator a) | Recip (Combinator a)
   deriving (Eq,Show)
 
-data Context a = 
+data Context a =
     Empty
   | Fst (Context a) (Combinator a) | Snd (Combinator a) (Context a)
-  | LeftTimes (Context a) (Combinator a) (Value a) 
+  | LeftTimes (Context a) (Combinator a) (Value a)
   | RightTimes (Combinator a) (Value a) (Context a)
   | LeftPlus (Context a) (Combinator a) | RightPlus (Combinator a) (Context a)
   deriving (Eq,Show)
 
-data State a = 
-    Enter (Combinator a) (Value a) (Context a) 
+data State a =
+    Enter (Combinator a) (Value a) (Context a)
   | Exit (Combinator a) (Value a) (Context a)
   deriving (Eq,Show)
 
@@ -46,16 +46,16 @@ data Dir = Forward | Back
   deriving (Eq,Show)
[…]

### 2016-12-01 pibackground.tex
remove math from section titles, for now.
working on sec 3 background
Add a dir-locals and enforce formatting everywhere
sec 2

@@
 
-\begin{figure*}[ht]
-\[\begin{array}{cc}
-\begin{array}{rrcll}
+\begin{figure}[ht]
+\[\begin{array}{rrcll}
 \unitepl :&  \zt \oplus \tau & \iso & \tau &: \unitipl \\
@@
   \iso & (\tau_1 \otimes \tau_2) \oplus (\tau_1 \otimes \tau_3)~ &: \factorl
-\end{array}
-&
-\begin{minipage}{0.4\textwidth}
+\end{array}\]
+
 \[\begin{array}{c}
@@
 {}
-\\
-\\
+\qquad\qquad
 \Rule{}
@@
 \end{array}\]
-\end{minipage}
-\end{array}\]
+
 Each 1-combinator $c$ has an inverse $!~c$, e.g, $!~\unitepl=\unitipl$,
@@
 \label{pi-combinators}}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
 \[\begin{array}{c}
@@
 {}
-~
+\\
+\\
 \Rule{}
@@
 \label{pi-combinators2}}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[ht]
-\begin{multicols}{2}
-\[\begin{array}{r@{\!}lcl}
+\begin{figure}[ht]
+\[\begin{array}{rlcl}
 %% \evalone{\unitepl}{&(\inl{()})} \\
@@
 \evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3)) \\
-\end{array}\]
-\[\begin{array}{r@{\!}lcl}
+\\
 \evalone{\absorbr}{&(v,\_)} &=& v \\
@@
 \end{array}\]
-\end{multicols}
 \caption{\label{opsem}$\Pi$ operational semantics}
-\end{figure*}
+\end{figure}
 
@@
 (\textrm{2-combinators}) &
-  \alpha &:& c_1 \isotwo c_2 \mbox{~where~} c_1, c_2 : \tau_1 \iso \tau_2 \\
-  & && [\textit{see Fig.~\ref{pi-combinators2}}]
+  \alpha &:& c_1 \isotwo c_2 \mbox{~where~} c_1, c_2 : \tau_1 \iso \tau_2
+  ~[\textit{see Fig.~\ref{pi-combinators2}}]
 \end{array}\]
@@
 \noindent Both classes of programs, 1-combinators $c$, and
-2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT
-sense. The elements $c$ of 1-combinators denote type isomorphisms. The
-elements $\alpha$ of 2-combinators denote equivalences between these
-type isomorphisms which are sound and complete 1-combinator program
-transformations. Using the 1-combinators, it is possible to write any
-reversible boolean function and hence encode arbitrary boolean
-functions by a technique that goes back to \cite{Toffoli:1980}. The
-2-combinators provide a layer of programs that computes
-semantics-preserving transformations of 1-combinators. As a small
-example, let us abbreviate $\ot \oplus \ot$ as the type $\mathbb{2}$
-of booleans and examine two possible implementations of boolean
-negation. The first directly uses the primitive combinator
-$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to swap the
-two values of type $\mathbb{2}$; the second use three consecutive
-copies of $\swapp$ to achieve the same effect:
+2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT sense. The
+elements $c$ of 1-combinators denote type isomorphisms. The elements $\alpha$ of
+2-combinators denote the set of sound and complete equivalences between these
+type isomorphisms. Using the 1-combinators, it is possible to write any
+reversible boolean function and hence encode arbitrary boolean functions by a
+technique that goes back to Toffoli~\cite{Toffoli:1980}. The 2-combinators
+provide a layer of programs that computes semantics-preserving transformations
+of 1-combinators. As a small example, let us abbreviate $\ot \oplus \ot$ as the
+type $\mathbb{2}$ of booleans and examine two possible implementations of
+boolean negation. The first directly uses the primitive combinator
+$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to exchange the two
+values of type $\mathbb{2}$; the second use three consecutive copies of $\swapp$
+to achieve the same effect:
 \[\begin{array}{rcl}
@@
 \[
-(\idldl ~\respstwo~ \idisotwo)~\transtwo~\idldl
+(\linvdl ~\respstwo~ \idisotwo)~\transtwo~\idldl
 \]
-which not only shows the equivalence of the two implementations of
-negation but also shows \emph{how} to transform one to the other.
+which not only shows the equivalence of the two implementations of negation but
+also shows \emph{how} to transform one to the other. In this example, we focus
+on the first two occurrences of $\swapp$ and use $\linvdl$ to reduce them to
+$\idiso$ since they are inverses. We then use $\idldl$ to simplify the
+composition of $\idiso$ with $\swapp$ to just $\swapp$.
 
@@
 
-\paragraph{Intermezzo.} For an information-theoretic perspective on
+\paragraph*{Intermezzo.} For an information-theoretic perspective on
 the language $\Pi$, we think of a type containing $N$ values as an

### 2016-12-07 pibackground.tex
figures in sec 2; will start sec 3 next
working on sec 3
sec 3 working on it
renamed a2

@@
 \begin{figure}[ht]
-\[\begin{array}{rlcl}
+{\footnotesize
+\[\begin{array}{cc}
+\begin{array}[t]{rlcl}
 %% \evalone{\unitepl}{&(\inl{()})} \\
@@
 \evalone{\assocrp}{&(\inl{(\inr{v})})} &=& \inr{(\inl{v})} \\
-\evalone{\assocrp}{&(\inr{v})} &=& \inr{(\inr{v})} \\
-\\
+\evalone{\assocrp}{&(\inr{v})} &=& \inr{(\inr{v})}
+\end{array} &
+\begin{array}[t]{rlcl}
 \evalone{\unitetl}{&(\unitv,v)} &=& v \\
@@
 \evalone{\assoclt}{&(v_1,(v_2,v_3))} &=& ((v_1,v_2),v_3) \\
-\evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3)) \\
+\evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3))
+\end{array} \\
 \\
+\begin{array}[t]{rlcl}
 \evalone{\absorbr}{&(v,\_)} &=& v \\
@@
 \evalone{\factorl}{&\inl{(v_1,v_3)}} &=& (v_1,\inl{v_3}) \\
-\evalone{\factorl}{&\inr{(v_2,v_3)}} &=& (v_2,\inr{v_3}) \\
-\\
+\evalone{\factorl}{&\inr{(v_2,v_3)}} &=& (v_2,\inr{v_3})
+\end{array} &
+\begin{array}[t]{rlcl}
 \evalone{\idiso}{&v} &=& v \\
@@
   (\evalone{c_1}v_1, \evalone{c_2}v_2)
+\end{array}
 \end{array}\]
 \caption{\label{opsem}$\Pi$ operational semantics}
+}
 \end{figure}
@@
 As an example, let $\mathbb{3}$ abbreviate the type
-$(\ot \oplus \ot) \oplus \ot$. There are three values of type
-$\mathbb{3}$ which are $\inl{\inl{\unitv}}$, $\inl{\inr{\unitv}}$, and
-$\inr{\unitv}$. There are, up to equivalence, six combinators of type
-$\mathbb{3} \iso \mathbb{3}$, each representing a different
-permutation of three elements, which could be written as follows:
+$(\ot \oplus \ot) \oplus \ot$. There are three values of type $\mathbb{3}$ which
+are $ll=\inl{\inl{\unitv}}$, $lr=\inl{\inr{\unitv}}$, and
+$r=\inr{\unitv}$. There are, up to equivalence, six combinators of type
+$\mathbb{3} \iso \mathbb{3}$, each representing a different permutation of three
+elements, which could be written as follows:
 \[\begin{array}{rcl}
-a_1 &=& \idiso \\
-a_2 &=& \swapp \oplus \idiso \\
-a_3 &=& \assocrp \odot (\idiso \oplus \swapp) \odot \assoclp \\
-a_4 &=& a_2 \odot a_3 \\
-a_5 &=& a_3 \odot a_2 \\
-a_6 &=& a_4 \odot a_2
+\permone &=& \idiso \\
+\permtwo &=& \swapp \oplus \idiso \\
+\permthree &=& \assocrp \odot (\idiso \oplus \swapp) \odot \assoclp \\
+\permfour &=& \permtwo \odot \permthree \\
+\permfive &=& \permthree \odot \permtwo \\
+\permsix &=& \permfour \odot \permtwo
 \end{array}\]
-Tracing the evaluation of $a_2$ on each of the possible inputs yields:
+Tracing the evaluation of $\permtwo$ on each of the possible inputs yields:
 \[\begin{array}{rcl}
@@
 \end{array}\]
-Thus the effect of combinator $a_2$ is to swap the values
+Thus the effect of combinator $\permtwo$ is to swap the values
 $\inl{\inl{\unitv}}$ and $\inl{\inr{\unitv}}$ leaving the value
-$\inr{\unitv}$ intact. Iterating $a_2$ again is therefore equivalent
+$\inr{\unitv}$ intact. Iterating $\permtwo$ again is therefore equivalent
 to the identity permutation, which can be verified using the
@@
 \[\begin{array}{rcl}
-a_2 \odot a_2 &=& (\swapp \oplus \idiso) \odot (\swapp \oplus \idiso) \\
+\permtwo \odot \permtwo &=& (\swapp \oplus \idiso) \odot (\swapp \oplus \idiso) \\
 &\isotwo& (\swapp \odot \swapp) \oplus (\idiso \odot \idiso) \\
@@
 
-For our example combinators on the type $\mathbb{3}$, simple traces
-using the operational semantics show the combinator $a_1$ is the
-identity permutation; the combinators $a_3$ and $a_6$ swap two of the
-three elements leaving the third intact; and the combinators $a_4$ and
-$a_5$ rotate the three elements. We therefore have
-$\mathit{order}(a_1)=1$,
-$\mathit{order}(a_2)=\mathit{order}(a_3)=\mathit{order}(a_6)=2$, and
-$\mathit{order}(a_4)=\mathit{order}(a_5)=3$.
+For our example combinators on the type $\mathbb{3}$, simple traces using the
+operational semantics show the combinator $\permone$ is the identity
+permutation; the combinators $\permthree$ and $\permsix$ swap two of the three
+elements leaving the third intact; and the combinators $\permfour$ and
+$\permfive$ rotate the three elements. We therefore have
+$\mathit{order}(\permone)=1$,
+$\mathit{order}(\permtwo)=\mathit{order}(\permthree)=\mathit{order}(\permsix)=2$,
+and $\mathit{order}(\permfour)=\mathit{order}(\permfive)=3$.
 
@@
 \begin{lemma}
+\label{lem:ordertwo}
   For $p : \tau \iso \tau$, $n \in \Z$, $p^{k + n} \isotwo p^n$ where

### 2016-12-08 pibackground.tex
restructure sec 3

@@
 \begin{definition}[Order of a 1-combinator]
+\label{def:order}
   The order of a 1-combinator $p : \tau \iso \tau$, $\ord{p}$, is the

### 2016-12-15 typespp.tex
new clean file with no agda

\documentclass[a4paper,USenglish]{lipics-v2016-utf8x}

\usepackage{amsthm,amstext,amssymb,amsmath}
\usepackage{datetime}
\usepackage{bbold}
\usepackage{url}
\usepackage{xcolor}
\usepackage{multicol}
\usepackage{proof}
\usepackage{stmaryrd}
\usepackage{bussproofs}
\usepackage{tikz}
\usepackage{tikz-cd}
\usetikzlibrary{quotes}
\usepackage[section]{placeins}
\usepackage{fdsymbol}

\newcommand{\amr}[1]{\fbox{\begin{minipage}{0.4\textwidth}\color{purple}{Amr says: #1}\end{minipage}}}
\newcommand{\vic}[1]{\fbox{\begin{minipage}{0.4\textwidth}\color{purple}{Vikraman says: #1}\end{minipage}}}
\newcommand{\jc}[1]{\fbox{\begin{minipage}{\textwidth}\color{purple}{Jacques says: #1}\end{minipage}}}

\newcommand{\hide}[1]{}

\newcommand{\textpi}{$\pi$}
\newcommand{\ft}{\mathbb{T}}
\newcommand{\hash}{\#}
\newcommand{\dg}{conjugation groupoid}
\newcommand{\DG}{Conjugation Groupoid}
\newcommand{\pifrac}{\ensuremath{\Pi^/}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\isotwo}{\Leftrightarrow}
\newcommand{\alt}{~|~}
\newcommand{\ag}[2]{#1 \sslash #2}
\newcommand{\agp}[3]{#1 \sslash_{#3} #2}
\newcommand{\order}[1]{\hash #1}
\newcommand{\iorder}[1]{1/\hash #1}
\newcommand{\oneg}[1]{\mathbb{1}_{#1}}
\newcommand{\divgl}[2]{\hash{#1} \div_{l} \hash{#2}}
\newcommand{\divgr}[2]{\hash{#1} \div_{r} \hash{#2}}
\newcommand{\ord}[1]{\ensuremath{\mathsf{order}(#1)}}
\newcommand{\inl}[1]{\textsf{inl}(#1)}
\newcommand{\inr}[1]{\textsf{inr}(#1)}
\newcommand{\zt}{\mathbb{0}}
\newcommand{\ot}{\mathbb{1}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Zn}{\mathbb{Z}_n}
\newcommand{\Zvn}{\mathbb{Z}^v_n}
\newcommand{\N}{\mathbb{N}}
\newcommand{\cycle}{\textsf{cycle}}
\newcommand{\twod}{\mathbb{T}}
\newcommand{\fract}[2]{#1/#2}
\newcommand{\fv}[2]{\fcolorbox{black}{white}{\strut $#1$}\fcolorbox{black}{gray!20}{$\strut #2$}}
\newcommand{\pt}[2]{\bullet[#1,#2]}
\newcommand{\sing}[1]{\textsc{Sing}(#1)}
\newcommand{\singi}[2]{\textsc{SingI}_{{#1}}(#2)}
\newcommand{\iter}[1]{\textsc{Iter}(#1)}
\newcommand{\pair}[2]{\langle #1,#2 \rangle}
\newcommand{\triple}[3]{\langle #1,#2,#3 \rangle}
\newcommand{\distiterplus}[3]{\mathsf{dist}~#1~#2~#3}

\newcommand{\permone}{\mathit{perm}_{...}}
\newcommand{\permtwo}{\mathit{perm}_{\times\!\times.}}
\newcommand{\permthree}{\mathit{perm}_{.\times\!\times}}
\newcommand{\permfour}{\mathit{perm}_{\rightarrow}}
\newcommand{\permfive}{\mathit{perm}_{\leftarrow}}
\newcommand{\permsix}{\mathit{perm}_{\times.\times}}

\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\proves}{\vdash}
\newcommand{\jdg}[3]{{#1} #3}
\newcommand{\evalone}[2]{#1~\triangleright~#2}
\newcommand{\evaloneb}[2]{#1~\triangleleft~#2}
\newcommand{\unitv}{()}
\newcommand{\unitepl}{\mathsf{{unite_+l}}}
\newcommand{\unitipl}{\mathsf{{uniti_+l}}}
\newcommand{\unitepr}{\mathsf{{unite_+r}}}
\newcommand{\unitipr}{\mathsf{{uniti_+r}}}
\newcommand{\swapp}{\mathsf{{swap_+}}}
\newcommand{\assoclp}{\mathsf{{assocl_+}}}
\newcommand{\assocrp}{\mathsf{{assocr_+}}}
\newcommand{\unitetl}{\mathsf{{unite_{\star}l}}}
\newcommand{\unititl}{\mathsf{{uniti_{\star}l}}}
\newcommand{\unitetr}{\mathsf{{unite_{\star}r}}}
\newcommand{\unititr}{\mathsf{{uniti_{\star}r}}}
\newcommand{\swapt}{\mathsf{{swap_{\star}}}}
\newcommand{\assoclt}{\mathsf{{assocl_{\star}}}}
\newcommand{\assocrt}{\mathsf{{assocr_{\star}}}}
\newcommand{\absorbr}{\mathsf{{absorbr}}}
\newcommand{\absorbl}{\mathsf{{absorbl}}}
\newcommand{\factorzr}{\mathsf{{factorzr}}}
\newcommand{\factorzl}{\mathsf{{factorzl}}}
\newcommand{\dist}{\mathsf{{dist}}}
\newcommand{\factor}{\mathsf{{factor}}}
\newcommand{\distl}{\mathsf{{distl}}}
\newcommand{\factorl}{\mathsf{{factorl}}}
\newcommand{\idiso}{\mathsf{{id}}}

\newcommand{\assocdl}{\mathsf{{assoc_{\odot}l}}}
\newcommand{\assocdr}{\mathsf{{assoc_{\odot}r}}}
\newcommand{\idldl}{\mathsf{{idl_{\odot}l}}}
\newcommand{\idldr}{\mathsf{{idl_{\odot}r}}}
\newcommand{\idrdl}{\mathsf{{idr_{\odot}l}}}
\newcommand{\idrdr}{\mathsf{{idr_{\odot}r}}}
\newcommand{\rinvdl}{\mathsf{{rinv_{\odot}l}}}
\newcommand{\rinvdr}{\mathsf{{rinv_{\odot}r}}}
\newcommand{\linvdl}{\mathsf{{linv_{\odot}l}}}
\newcommand{\linvdr}{\mathsf{{linv_{\odot}r}}}
\newcommand{\idisotwo}{\mathsf{{id}}}
\newcommand{\transtwo}{\bullet}
\newcommand{\respstwo}{\mathsf{{\boxdot}}}
\newcommand{\respptwo}{\mathsf{{resp_{\oplus\Leftrightarrow}}}}
\newcommand{\respttwo}{\mathsf{{resp_{\otimes\Leftrightarrow}}}}
\newcommand{\sumid}{\mathsf{sumid}}
\newcommand{\splitid}{\mathsf{splitid}}
\newcommand{\homps}{\mathsf{hom_{\oplus\odot}}}
\newcommand{\homsp}{\mathsf{hom_{\odot\oplus}}}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}

\newtheorem{proposition}{Proposition}

\makeatletter
\tikzset{my loop/.style =  {to path={
  \pgfextra{\let\tikztotarget=\tikztostart}
  [looseness=8,min distance=5mm, in=240,out=300]
  \tikz@to@curve@path}
  }}
\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\title{Fractional Types}
\titlerunning{Fractional Types}
\author[1]{Jacques Carette}
\author[2]{Chao-Hong Chen}
\author[3]{Vikraman Choudhury}
\author[4]{Amr Sabry}
\affil[1]{Computing and Software Department, McMaster University,
  Hamilton, Ontario, Canada \\ \texttt{carette@mcmaster.ca}}
\affil[2]{Computer Science Department, Indiana University,
  Bloomington, Indiana, USA \\ \texttt{chen464@indiana.edu}}
\affil[3]{Computer Science Department, Indiana University,
  Bloomington, Indiana, USA \\ \texttt{vikraman@indiana.edu}}
\affil[4]{Computer Science Department, Indiana University,
  Bloomington, Indiana, USA \\ \texttt{sabry@indiana.edu}}

\authorrunning{J. Carette, C.-H. Chen, V. Choudhury and A. Sabry}

\Copyright{Jacques Carette, Chao-Hong Chen, Vikraman Choudhury and Amr Sabry}

\keywords{dummy}
\subjclass{dummy}

\maketitle

\begin{abstract}

We exhibit types whose natural cardinality is fractional.  More
precisely, we show that the groupoid cardinality (as defined by
Baez-Dolan) of the denotation of the type of a singleton reversible
program $p$ with exactly $k$ distinct proofs of reversibility has
cardinality $1/k$.  We further show that this type is naturally a
multiplicative inverse to the type of all iterates $p ^ i$ of that
reversible program.

We situate this work as an extension of a larger reversible
programming language ($\Pi$), and show that this extension is also
reversible.  Interestingly, this extension supports first-class
functions as well as some natural analogues of operations coming from
compact closed categories.  We emphasize that the key ingredients are
reversibility and proof relevance: cardinality $1/k$ arises from
having exactly $k$ proofs of reversibility.

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

In Homotopy Type Theory (HoTT)~\cite{hottbook}, types have the
structure of \emph{weak $\omega$-groupoids}. As a first approximation,
we can think of such structures as sets with points (objects) and
paths (equivalences) between the points and higher paths between these
paths and so on. Here are two simple but non-trivial examples:

\begin{center}
\begin{tabular}{c@{\kern -4pt}c}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]
  \draw[dashed] (0,0) ellipse (3cm and 1.2cm);
  \node[below] (A) at (-1.7,0.3) {\texttt{a}};
  \node[below] (B) at (-1.1,0.3) {\texttt{b}};
  \node[below] (C) at (-0.5,0.3) {\texttt{c}};
  \node[below] (D) at (0.5,0.3) {\texttt{d}};
  \node[below] (E) at (1.1,0.3) {\texttt{e}};
  \node[below] (F) at (1.7,0.3) {\texttt{f}};
  \draw[fill] (-1.7,0.3) circle [radius=0.05];
  \draw[fill] (-1.1,0.3) circle [radius=0.05];
  \draw[fill] (-0.5,0.3) circle [radius=0.05];
  \draw[fill] (0.5,0.3) circle [radius=0.05];
  \draw[fill] (1.1,0.3) circle [radius=0.05];
  \draw[fill] (1.7,0.3) circle [radius=0.05];
  \draw (-1.7,0.3) -- (-1.1,0.3);
  \draw (-1.1,0.3) -- (-0.5,0.3);
  \draw (-1.7,0.3) to[bend left] (-0.5,0.3) ;
  \draw (0.5,0.3) -- (1.1,0.3);
  \draw (1.1,0.3) -- (1.7,0.3);
  \draw (0.5,0.3) to[bend left] (1.7,0.3) ;
  \path (A) edge [loop below] node[below] {\texttt{id}} (B);
  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
  \path (C) edge [loop below] node[below] {\texttt{id}} (B);
  \path (D) edge [loop below] node[below] {\texttt{id}} (B);
  \path (E) edge [loop below] node[below] {\texttt{id}} (B);
  \path (F) edge [loop below] node[below] {\texttt{id}} (B);
\end{tikzpicture}
&
\begin{tikzpicture}[scale=0.5,every node/.style={scale=0.5}]
  \draw[dashed] (0,0) ellipse (2cm and 2.4cm);
  \node[below] (B) at (0,-1.1) {\texttt{*}};
  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
  \path (B) edge [loop above, looseness=15, in=48, out=132] node[above] {$p$} (B);
  \path (B) edge [loop above, looseness=25, in=40, out=140] node[above] {$q$} (B);
  \path (B) edge [loop above, looseness=40, in=32, out=148] node[above] {$q'$} (B);
  \path (-1,0.4) edge[double, bend left] node[left] {$\alpha$} (-1,1.0);
\end{tikzpicture}
\end{tabular}
\end{center}

\medskip\noindent Baez and Dolan~\cite{groupoidcard} assign to each groupoid a
\emph{cardinality} that counts the objects up to equivalences. The
groupoid on the left has six points \texttt{a}, \texttt{b}, \texttt{c},
\texttt{d}, \texttt{e}, and \texttt{f} with two groups of three
points clustered in an equivalence class and hence the groupoid has
cardinality~2. The (2-)groupoid on the right has one point $\ast$ with
four equivalences \texttt{id}, $p$, $q$, and~$q'$ on it. The
equivalences $q$ and $q'$ are however identified by~$\alpha$ leaving
only three distinct equivalence classes and hence making the
cardinality $\frac{1}{3}$.

Both groupoids involve some notion of ``division'' that can be captured at the
level of types using a syntactic notion of \emph{fractional types}. Our aim is
to explore this notion of types and its associated operational semantics. The
remainder of the paper is organized as follows. We start by reviewing the
necessary background consisting of the language $\Pi$ for programming with
isomorphisms or equivalences in a reversible information-preserving way. Sec.~4
explains the main novel semantic ideas of using $\Pi$ programs to generate
non-trivial groupoids with fractional cardinality. Sec.~5 translates the
semantic ideas into an extension of $\Pi$ with new type constructors denoting
non-trivial groupoids and new programs that manipulate such types. Sec.~6
presents the operational semantics of the extended language. The last two
sections put our work in perspective and conclude.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Programming with Equivalences}
\label{sec:pi}

The main syntactic vehicle for the technical developments
is the language $\Pi$ whose only computations are isomorphisms
between finite types and equivalences between these
isomorphisms~\cite{Carette2016,James:2012:IE:2103656.2103667}. We
present the syntax and operational semantics of the parts of the
language relevant to our work.

\begin{figure}[ht]
\[\begin{array}{rrcll}
\unitepl :&  \zt \oplus \tau & \iso & \tau &: \unitipl \\
\unitepr :&  \tau \oplus \zt & \iso & \tau &: \unitipr \\
\swapp :&  \tau_1 \oplus \tau_2 & \iso & \tau_2 \oplus \tau_1 &: \swapp \\
\assoclp :&  \tau_1 \oplus (\tau_2 \oplus \tau_3) & \iso & (\tau_1 \oplus \tau_2) \oplus \tau_3
  &: \assocrp \\
\\
\unitetl :&  \ot \otimes \tau & \iso & \tau &: \unititl \\
\unitetr :&  \tau \otimes \ot & \iso & \tau &: \unititr \\
\swapt :&  \tau_1 \otimes \tau_2 & \iso & \tau_2 \otimes \tau_1 &: \swapt \\
\assoclt :&  \tau_1 \otimes (\tau_2 \otimes \tau_3) & \iso & (\tau_1 \otimes \tau_2) \otimes \tau_3
  &: \assocrt \\
\\
\absorbr :&~ \zt \otimes \tau & \iso & \zt &: \factorzl \\
\absorbl :&~ \tau \otimes \zt & \iso & \zt &: \factorzr \\
\dist :&~ (\tau_1 \oplus \tau_2) \otimes \tau_3 &
  \iso & (\tau_1 \otimes \tau_3) \oplus (\tau_2 \otimes \tau_3)~ &: \factor \\
\distl :&~ \tau_1 \otimes (\tau_2 \oplus \tau_3) &
  \iso & (\tau_1 \otimes \tau_2) \oplus (\tau_1 \otimes \tau_3)~ &: \factorl
\end{array}\]

\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso : \tau \iso \tau}}
{}
\qquad\qquad
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
{\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
{}
\\
\\
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \oplus c_2 : \tau_1 \oplus \tau_3 \iso \tau_2 \oplus \tau_4}}
{}
\qquad\qquad
\Rule{}
{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
{\jdg{}{}{c_1 \otimes c_2 : \tau_1 \otimes \tau_3 \iso \tau_2 \otimes \tau_4}}
{}
\end{array}\]

Each 1-combinator $c$ has an inverse $!~c$, e.g, $!~\unitepl=\unitipl$,
$!(c_1 \odot c_2) = ~!c_2 \odot~ !c_1$, etc.
\caption{$\Pi$ 1-combinators~\cite{James:2012:IE:2103656.2103667}
\label{pi-combinators}}
\end{figure}

\begin{figure}[ht]
\[\begin{array}{c}
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idisotwo : c \isotwo c}}
{}
~
\Rule{}
{c_1,c_2,c_3 : \tau_1 \iso \tau_2 \quad \alpha_1 : c_1 \isotwo c_2 \quad \alpha_2 : c_2 \isotwo c_3}
{\jdg{}{}{\alpha_1~\transtwo~\alpha_2 : c_1 \isotwo c_3}}
{}
\\
\\
\Rule{}
{c_1 : \tau_1 \iso \tau_2 \quad c_2 : \tau_2 \iso \tau_3 \quad c_3 : \tau_3 \iso \tau_4}
{\jdg{}{}{\assocdl : c_1 \odot (c_2 \odot c_3) \isotwo (c_1 \odot c_2) \odot c_3 : \assocdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idldl : \idiso \odot c \isotwo c : \idldr}}
{}
~
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\idrdl : c \odot \idiso \isotwo c : \idrdr}}
{}
\\
\\
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\rinvdl : ~! c \odot c \isotwo \idiso : \rinvdr}}
{}
~
\Rule{}
{c : \tau_1 \iso \tau_2}
{\jdg{}{}{\linvdl : c ~\odot ~! c \isotwo \idiso : \linvdr}}
{}
\\
\\
\Rule{}
{}
{\jdg{}{}{\sumid : \idiso\oplus\idiso \isotwo \idiso : \splitid}}
{}
\\
\\
\Rule{}
{c_1 : \tau_5\iso\tau_1 \quad c_2 : \tau_6\iso\tau_2 \quad c_3 :
    \tau_1\iso\tau_3 \quad c_4 : \tau_2\iso\tau_4}
{\jdg{}{}{\homps : (c_1\odot c_3)\oplus(c_2\odot c_4) \isotwo
    (c_1\oplus c_2) \odot (c_3\oplus c_4) : \homsp }}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\alpha_1 ~\respstwo~ \alpha_2 : c_1 \odot c_2 \isotwo c_3 \odot c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respptwo ~\alpha_1 ~\alpha_2 : c_1 \oplus c_2 \isotwo c_3 \oplus c_4}}
{}
\\
\\
\Rule{}
{c_1,c_3 : \tau_1 \iso \tau_2 \quad c_2,c_4 : \tau_2 \iso \tau_3 \quad
  \alpha_1 : c_1 \isotwo c_3 \quad \alpha_2 : c_2 \isotwo c_4}
{\jdg{}{}{\respttwo ~\alpha_1 ~\alpha_2 : c_1 \otimes c_2 \isotwo c_3 \otimes c_4}}
{}
\end{array}\]
Each 2-combinator $\alpha$ has an inverse $2!~\alpha$, e.g, $2!~\assocdl=\assocdr$,
$2!(\alpha_1~\transtwo~\alpha_2) = (2!~\alpha_2)~\transtwo~(2!~\alpha_1)$, etc.
\caption{$\Pi$ 2-combinators (excerpt)~\cite{Carette2016}
\label{pi-combinators2}}
\end{figure}

\begin{figure}[ht]
{\footnotesize
\[\begin{array}{cc}
\begin{array}[t]{rlcl}
\evalone{\unitepl}{&(\inr{v})} &=& v \\
\evalone{\unitipl}{&v} &=& \inr{v} \\
\evalone{\unitepr}{&(\inl{v})} &=& v \\
\evalone{\unitipr}{&v} &=& \inl{v} \\
\evalone{\swapp}{&(\inl{v})} &=& \inr{v} \\
\evalone{\swapp}{&(\inr{v})} &=& \inl{v} \\
\evalone{\assoclp}{&(\inl{v})} &=& \inl{(\inl{v})} \\
\evalone{\assoclp}{&(\inr{(\inl{v})})} &=& \inl{(\inr{v})} \\
\evalone{\assoclp}{&(\inr{(\inr{v})})} &=& \inr{v} \\
\evalone{\assocrp}{&(\inl{(\inl{v})})} &=& \inl{v} \\
\evalone{\assocrp}{&(\inl{(\inr{v})})} &=& \inr{(\inl{v})} \\
\evalone{\assocrp}{&(\inr{v})} &=& \inr{(\inr{v})}
\end{array} &
\begin{array}[t]{rlcl}
\evalone{\unitetl}{&(\unitv,v)} &=& v \\
\evalone{\unititl}{&v} &=& (\unitv,v) \\
\evalone{\unitetr}{&(v,\unitv)} &=& v \\
\evalone{\unititr}{&v} &=& (v,\unitv) \\
\evalone{\swapt}{&(v_1,v_2)} &=& (v_2,v_1) \\
\evalone{\assoclt}{&(v_1,(v_2,v_3))} &=& ((v_1,v_2),v_3) \\
\evalone{\assocrt}{&((v_1,v_2),v_3)} &=& (v_1,(v_2,v_3))
\end{array} \\
\\
\begin{array}[t]{rlcl}
\evalone{\absorbr}{&(v,\_)} &=& v \\
\evalone{\absorbl}{&(\_,v)} &=& v \\
\evalone{\dist}{&(\inl{v_1},v_3)} &=& \inl{(v_1,v_3)} \\
\evalone{\dist}{&(\inr{v_2},v_3)} &=& \inr{(v_2,v_3)} \\
\evalone{\factor}{&\inl{(v_1,v_3)}} &=& (\inl{v_1},v_3) \\
\evalone{\factor}{&\inr{(v_2,v_3)}} &=& (\inr{v_2},v_3) \\
\evalone{\distl}{&(v_1,\inl{v_3})} &=& \inl{(v_1,v_3)} \\
\evalone{\distl}{&(v_2,\inr{v_3})} &=& \inr{(v_2,v_3)} \\
\evalone{\factorl}{&\inl{(v_1,v_3)}} &=& (v_1,\inl{v_3}) \\
\evalone{\factorl}{&\inr{(v_2,v_3)}} &=& (v_2,\inr{v_3})
\end{array} &
\begin{array}[t]{rlcl}
\evalone{\idiso}{&v} &=& v \\
\evalone{(c_1 \odot c_2)}{&v} &=&
  \evalone{c_2}{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inl{v})} &=&
  \inl{(\evalone{c_1}{v})} \\
\evalone{(c_1 \oplus c_2)}{&(\inr{v})} &=&
  \inr{(\evalone{c_2}{v})} \\
\evalone{(c_1 \otimes c_2)}{&(v_1,v_2)} &=&
  (\evalone{c_1}v_1, \evalone{c_2}v_2)
\end{array}
\end{array}\]
\caption{\label{opsem}$\Pi$ operational semantics}
}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Syntax of $\Pi$}
\label{opsempi}

The $\Pi$ family of languages is based on type isomorphisms. In the
variant we consider, the set of types $\tau$ includes the empty
type~$\zt$, the unit type $\ot$, and sum $\oplus$ and product
$\otimes$ types. The values classified by these types are the
conventional ones: $\unitv$ of type~$\ot$, $\inl{v}$ and $\inr{v}$ for
injections into sum types, and $(v_1,v_2)$ for product types. The
language has two other syntactic categories of programs to be
described in detail:
\[\begin{array}{lrcl}
(\textrm{Types}) &
  \tau &::=& \zt \alt \ot \alt \tau_1 \oplus \tau_2 \alt \tau_1 \otimes \tau_2 \\
(\textrm{Values}) &
  v &::=& \unitv \alt \inl{v} \alt \inr{v} \alt (v_1,v_2) \\
(\textrm{1-combinators}) &
  c &:& \tau_1 \iso \tau_2 ~ [\textit{see Fig.~\ref{pi-combinators}}] \\
(\textrm{2-combinators}) &
  \alpha &:& c_1 \isotwo c_2 \mbox{~where~} c_1, c_2 : \tau_1 \iso \tau_2
  ~[\textit{see Fig.~\ref{pi-combinators2}}]
\end{array}\]

\noindent Both classes of programs, 1-combinators $c$, and
2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT sense. The
elements $c$ of 1-combinators denote type isomorphisms. The elements $\alpha$ of
2-combinators denote the set of sound and complete equivalences between these
type isomorphisms. Using the 1-combinators, it is possible to write any
reversible boolean function and hence encode arbitrary boolean functions by a
technique that goes back to Toffoli~\cite{Toffoli:1980}. The 2-combinators
provide a layer of programs that computes semantics-preserving transformations
of 1-combinators. As a small example, let us abbreviate $\ot \oplus \ot$ as the
type $\mathbb{2}$ of booleans and examine two possible implementations of
boolean negation. The first directly uses the primitive combinator
$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to exchange the two
values of type~$\mathbb{2}$; the second use three consecutive uses of $\swapp$
to achieve the same effect:
\[\begin{array}{rcl}
\mathsf{not_1} &=& \swapp \\
\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp
\end{array}\]
We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
\isotwo \mathsf{not_1}$:
\[
(\linvdl ~\respstwo~ \idisotwo)~\transtwo~\idldl
\]
which not only shows the equivalence of the two implementations of negation but
also shows \emph{how} to transform one to the other. This rewriting focuses
on the first two occurrences of $\swapp$ and uses $\linvdl$ to reduce them to
$\idiso$ since they are inverses. It then uses $\idldl$ to simplify the
composition of $\idiso$ with $\swapp$ to just $\swapp$.

Fig.~\ref{pi-combinators} lists all the 1-combinators which consist of base
combinators (top) and compositions (bottom). Each line of the base combinators
introduces a pair of dual constants\footnote{where $\swapp$ and $\swapt$ are
self-dual.} that witness the type isomorphism in the middle. This set of
isomorphisms is known to be sound and
complete~\cite{Fiore:2004,fiore-remarks}. As the full set of 2-combinators has
$113$ entries, Fig.~\ref{pi-combinators2} lists a few of the 2-combinators that
we use in this paper. Each 2-combinator relates two 1-combinators of the same
type and witnesses their equivalence. Both 1-combinators and 2-combinators are
invertible and the 2-combinators behave as expected with respect to inverses of
1-combinators.

\begin{proposition}
For any $c : \tau_1 \iso \tau_2$, we have $c \isotwo ~!~(!~c)$.
\end{proposition}

\begin{proposition}
For any $c_1,c_2 : \tau_1 \iso \tau_2$, we have $c_1 \isotwo c_2$ implies
$!~c_1 \isotwo ~!~c_2$.
\end{proposition}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Semantics}
\label{sec:pisem}

We give an operational semantics for the 1-combinators of $\Pi$ which
represent the conventional layer of programs.  Operationally, the
semantics consists of a pair of evaluators that
take a combinator and a value and propagate the value in the forward
direction~$\triangleright$ or in the backward
direction~$\triangleleft$. We show the complete forward evaluator in
Fig.~\ref{opsem}; the backward evaluator is easy to infer.

As an example, let $\mathbb{3}$ abbreviate the type
$(\ot \oplus \ot) \oplus \ot$. There are three values of
type~$\mathbb{3}$ which are $ll=\inl{\inl{\unitv}}$,
$lr=\inl{\inr{\unitv}}$, and $r=\inr{\unitv}$. Pictorially, the type
$\mathbb{3}$ with its three inhabitants can be represented as the
left-leaning tree:
\begin{center}
\begin{tikzpicture}[level distance=0.5cm]
  \node {$\cdot$}
    child {node {$\cdot$}
      child {node {$0$}}
      child {node {$1$}}
    }
    child {node {$2$}} ;
\end{tikzpicture}
\end{center}
Note that the values of type $\mathbb{3}$ are the names of
the paths from the root to each of the leaves.  We use
$0$, $1$ and $2$ as ordinals, to give an order to each
of the values.

There are, up to equivalence, six combinators of type
$\mathbb{3} \iso \mathbb{3}$, each representing a different
permutation of three elements that leave the \emph{shape} of the three
unchanged. The six permutations on $\mathbb{3}$ can be written as
$\Pi$-terms:
\[\begin{array}{rcl}
\permone &=& \idiso \\
\permtwo &=& \swapp \oplus \idiso \\
\permthree &=& \assocrp \odot (\idiso \oplus \swapp) \odot \assoclp \\
\permfour &=& \permtwo \odot \permthree \\
\permfive &=& \permthree \odot \permtwo \\
\permsix &=& \permfour \odot \permtwo
\end{array}\]
Tracing the evaluation of $\permtwo$ on each of the possible inputs yields:
\[\begin{array}{rcl}
\evalone{(\swapp\oplus\idiso)}{\inl{\inl{\unitv}}} &=& \inl{\evalone{\swapp}{\inl{\unitv}}} \\
&=& \inl{\inr{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inl{\inr{\unitv}}} &=& \inl{\evalone{\swapp}{\inr{\unitv}}} \\
&=& \inl{\inl{\unitv}} \\
\\
\evalone{(\swapp\oplus\idiso)}{\inr{\unitv}} &=& \inr{\evalone{\idiso}{\unitv}} \\
&=& \inr{\unitv}
\end{array}\]
Thus the effect of combinator $\permtwo$ is to swap the values
$\inl{\inl{\unitv}}$ and $\inl{\inr{\unitv}}$ leaving the value
$\inr{\unitv}$ intact. In other words, the effect of $\permtwo$
can be visualized as giving the tree:
\begin{center}
\begin{tikzpicture}[level distance=0.5cm]
  \node {$\cdot$}
    child {node {$\cdot$}
      child {node {$1$}}
      child {node {$0$}}
    }
    child {node {$2$}} ;
\end{tikzpicture}
\end{center}

These trees should also make it clear why mathematicians shorten
their notation to
$\begin{pmatrix}
0 & 1 & 2 \\
1 & 0 & 2 \\
\end{pmatrix}$
for the same permutation.  We will not do so, as this
notation is \emph{untyped}, as it does not enforce that
the shape of the tree is preserved.

Iterating $\permtwo$ again is equivalent to the identity permutation, which can
be verified using 2-combinators: \[\begin{array}{rcl} \permtwo \odot \permtwo
&=& (\swapp \oplus \idiso) \odot (\swapp \oplus \idiso) \\ &\isotwo& (\swapp
\odot \swapp) \oplus (\idiso \odot \idiso) \\ &\isotwo& \idiso \oplus \idiso \\
&=& \idiso \end{array}\]

More generally we can iterate 1-combinators to produce different
reversible functions between finite sets, eventually wrapping around
at some number which represents the \emph{order} of the underlying
permutation.

\begin{definition}[Power of a 1-combinator]
  The $k^{th}$ power of a 1-combinator $p : \tau \iso \tau$, for
  $k \in \Z$ is
  \[
    p^k =
  \begin{cases}
    \idiso & k = 0 \\
    p \odot p^{k - 1} & k > 0 \\
    (!~p) \odot p^{k + 1} & k < 0 \\
  \end{cases}
  \]
\end{definition}

\begin{definition}[Order of a 1-combinator]
\label{def:order}
  The order of a 1-combinator $p : \tau \iso \tau$, $\ord{p}$, is the
  least postitive natural number $k \in \N^+$ such that
  $p^k \isotwo \idiso$.
\end{definition}

For our example combinators on the type $\mathbb{3}$, simple traces using the
operational semantics show the combinator $\permone$ is the identity
permutation; the combinators $\permthree$ and $\permsix$ swap two of the three
elements leaving the third intact; and the combinators $\permfour$ and
$\permfive$ rotate the three elements. We therefore have:
\[\begin{array}{rcl}
\mathit{order}(\permone) &=& 1 \\
\mathit{order}(\permtwo) = \mathit{order}(\permthree) = \mathit{order}(\permsix) &=& 2 \\
\mathit{order}(\permfour) = \mathit{order}(\permfive) &=& 3
\end{array}\]
We should note that the above definition is the only one in this
paper which is not \emph{effective};  in other words, we do not have an
algorithm for computing it.  While there is an obvious method to
compute it using the action of a 1-combinator on the elements of the
type it acts on, this is extremely inefficient.  What we mean is that
we know of no effective algorithm which works on the syntax of
combinators.  The (only) difficulty is $\odot$, which can have an
arbitrary effect on the order.

The 2-combinators, being complete equivalences between
1-combinators~\cite{Carette2016}, also capture equivalences regarding
power of combinators and their order.

\begin{lemma}
\label{lem:distiterplus}
  For $p : \tau\iso\tau$, $m,n\in\Z$, we have a 2-combinator
  $\distiterplus{p}{m}{n} : (p^m \odot p^n) \isotwo p ^{m + n}$.
\end{lemma}

\begin{lemma}
\label{lem:ordercancel}
  For $p : \tau \iso \tau$, $n \in \Z$, $p^{k + n} \isotwo p^n$ where
  $k = \ord{p}$.
\end{lemma}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{From Sets to Groupoids}
\label{sec:groupoids}

From a denotational perspective, a $\Pi$ type $\tau$ denotes a finite set, a
$\Pi$ 1-combinator denotes a permutation on finite sets, and the 2-combinators
denote coherence conditions on these permutations~\cite{Carette2016}. Formally,
the language $\Pi$ is a \emph{categorification}~\cite{math/9802029} of the
natural numbers as a \emph{symmetric rig groupoid}~\cite{nlabrig}. This
structure is a \emph{symmetric bimonoidal category} or a \emph{commutative rig
category} in which every morphism is invertible. The underlying category
consists of two symmetric monoidal structures~\cite{nla.cat-vn1051288}
separately induced by the properties of addition and multiplication of the
natural numbers. The monoidal structures are then augmented with distributivity
and absorption natural isomorphisms~\cite{laplaza} to model the full commutative
semiring (aka, commutative rig) of the natural numbers. Despite this rich
structure, the individual objects in the category for $\Pi$ are just plain sets
with no interesting structure. In this section we introduce, in the denotation
of $\Pi$, some non-trivial groupoids which we call \emph{iteration groupoids}
and \emph{symmetry groupoids}. Under certain conditions, sums and products of
these groupoids behave as expected which ensures that a sensible compositional
programming language can be designed around them.

%%%%%
\subsection{$\Pi$ Types as Sets (Discrete Groupoids)}

Each $\Pi$ type $\tau$ denotes a (structured) finite set $\sem{\tau}$ as follows:

\[\begin{array}{rcl}
\sem{\zt} &=& \bot \\
\sem{\ot} &=& \top \\
\sem{\tau_1 \oplus \tau_2} &=& \sem{\tau_1} \uplus \sem{\tau_2} \\
\sem{\tau_1 \otimes \tau_2} &=& \sem{\tau_1} \times \sem{\tau_2}
\end{array}\]

\noindent where we use $\bot$ to denote the empty set, $\top$ to denote a set
with one element, and $\uplus$ and~$\times$ to denote the disjoint union of sets
and the cartesian product of sets respectively. Each set can be viewed as a
groupoid whose objects are the set elements and with only identity morphisms on
each object. Nevertheless, the denotations of $\ot \oplus (\ot \oplus \ot)$ of
$(\ot \oplus \ot) \oplus \ot$ are not in fact equal, although they are
trivially isomorphic.

By only being able to express types whose denotations are trivial
groupoids, $\Pi$ leaves untapped an enormous amount of combinatorial structure
that is expressible in type theory. We show that with a small but deep technical
insight, it is possible to extend~$\Pi$ with types whose denotations are
not discrete.

%%%%%
\subsection{Groupoids and Groupoid Cardinality}

There are many definitions of groupoids that provide complementary perspectives
and insights. Perhaps the simplest definition to state, and the one which is
most immediately useful for our work, is that a groupoid is a category
in which every morphism has an inverse. Intuitively, such a category consists of
clusters of connected objects where each cluster is equivalent (in the
category-theoretic sense) to a group, viewed as a $1$-object category. Thus an
alternative definition of a groupoid is as a generalization of a group that
allows for individual elements
to have ``internal symmetries''~\cite{groupoidintro}. Baez et
al.~\cite{2009arXiv0908.4305B} associate with each groupoid a cardinality that
counts the elements up to these ``internal symmetries''.

\begin{definition}[Groupoid Cardinality~\cite{2009arXiv0908.4305B}]
  The cardinality of a groupoid $G$ is the (positive) real number:
  \[
    |G| = \sum_{[x]} \frac{1}{|\textsf{Aut}(x)|}
  \]
  provided the sum converges. The summation is over \emph{isomorphism classes}
  $[x]$ of objects $x$ and $|\textsf{Aut}(x)|$ is the number of \emph{distinct}
  automorphisms of $x$.
\end{definition}

\begin{figure}[t]
\begin{center}
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.8}]
  \draw[dashed] (0,-0.3) ellipse (1.5cm and 2.1cm);
  \node[below] (A) at (-0.5,0) {$a$};
  \node[below] (C) at (0.5,0) {$c$};
  \path (A) edge [loop below] node[below] {\texttt{id}} (A);
  \path (C) edge [loop below] node[below] {\texttt{id}} (C);
  \path (C) edge [out=140, in=40, looseness=4] (C);
\end{tikzpicture}
\qquad \qquad \qquad
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.8}]
  \draw[dashed] (0,-0.3) ellipse (3cm and 2.5cm);
  \node[below] (A) at (-2,0) {$a$};
  \node[below] (B) at (0,0) {$b$};
  \node[below] (C) at (1.5,0) {$c$};
  \path (A) edge [bend left=50] (B);
  \path (C) edge [out=140, in=40, looseness=4] (C);
  \path (A) edge [loop below] node[below] {\texttt{id}} (A);
  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
  \path (C) edge [loop below] node[below] {\texttt{id}} (C);
\end{tikzpicture}
\qquad \qquad \qquad
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.8}]
  \draw[dashed] (0,-0.3) ellipse (2.8cm and 2.5cm);
  \node[below] (A) at (-1.6,0) {$a$};
  \node[below] (B) at (0,0) {$b$};
  \node[below] (C) at (1.6,0) {$c$};
  \path (A) edge [loop below] node[below] {\texttt{id}} (A);
  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
  \path (C) edge [loop below] node[below] {\texttt{id}} (C);
  \path (A) edge [loop above, looseness=3, in=48, out=132] (A);
  \path (B) edge [loop above, looseness=3, in=48, out=132] (B);
  \path (C) edge [loop above, looseness=3, in=48, out=132] (C);
\end{tikzpicture}
\end{center}
\caption{\label{fig:groupoids2}Example groupoids $G_1$, $G_2$, and $G_3$.}
\end{figure}

For plain sets, the definition just counts the elements as each element is its
own equivalence class and has exactly one automorphism (the identity). Without
quite formalizing them and relying on the informal diagrams until the next
section, we argue that each of the groupoids $G_1$, $G_2$, and $G_3$ in
Fig.~\ref{fig:groupoids2} has cardinality $\frac{3}{2}$. Groupoid~$G_1$ consists
of two isomorphism classes: class~$a$ has one object with one automorphism (the
identity) and class~$c$ has one object with two distinct automorphisms; the
cardinality is $\frac{1}{1} + \frac{1}{2} = \frac{3}{2}$. For groupoid~$G_2$, we
also have two isomorphism classes with representatives $a$ and $c$; the class
containing $a$ has two automorphisms starting from $a$: the identity and the
loop going from $a$ to~$b$ and back. By the groupoid axioms, this loop is
equivalent to the identity which means that the class containing $a$ has just
one automorphism. The isomorphism class of $c$ has two non-equivalent
automorphisms on it and hence the cardinality of $G_2$ is also
$\frac{1}{1} + \frac{1}{2} = \frac{3}{2}$. For~$G_3$, we have three isomorphism
classes, each with two non-equivalent automorphisms, and hence the cardinality
of $G_3$ is $\frac{1}{2} + \frac{1}{2} + \frac{1}{2} = \frac{3}{2}$.
It is important to note that $G_1$ and $G_2$ are (categorically) equivalent
groupoids, but that $G_3$ is not equivalent to either $G_1$ or $G_2$.
Roughly speaking this is because the number of connected components is also
an invariant of a groupoid, and here $G_1$ and $G_2$ have $2$ whilst $G_3$ has
$3$.

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{$\Pi$-Combinators as Automorphism Classes}

To formalize the counting above, we need, in the context of $\Pi$, a precise
definition of what it means for automorphisms to be ``distinct''. We start with
an example. Recall the type $\mathbb{3}$ with its three elements
$ll=\inl{\inl{\unitv}}$, $lr=\inl{\inr{\unitv}}$, and $r=\inr{\unitv}$. One of
the combinators of type $\mathbb{3} \iso \mathbb{3}$ is $\permtwo$. Observing
the results of applying the iterates $(\permtwo)^k$ for $k \in \Z$ on the three
elements we find:
\[\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\evalone{(\permtwo)^{2k}}{ll} &=& ll \\
\evalone{(\permtwo)^{2k}}{lr} &=& lr \\
\evalone{(\permtwo)^{2k}}{r} &=& r
\end{array} &
\begin{array}{rcl}
\evalone{(\permtwo)^{2k+1}}{ll} &=& lr \\
\evalone{(\permtwo)^{2k+1}}{lr} &=& ll \\
\evalone{(\permtwo)^{2k+1}}{r} &=& r
\end{array}
\end{array}\]
Furthermore, Lem.~\ref{lem:ordercancel} gives us the following families of
2-combinators $\alpha_{2k} : \idiso \isotwo (\permtwo)^{2k}$ and
$\alpha_{2k+1} : \permtwo \isotwo (\permtwo)^{2k+1}$. We can put these facts together to
construct a groupoid whose objects are the elements of $\mathbb{3}$, whose
1-morphisms relate $v_i$ and~$v_j$ if $\evalone{(\permtwo)^k}{v_i} = v_j$ for some
$k \in \Z$, and whose 2-morphisms are the families $\alpha_{2k}$ and
$\alpha_{2k+1}$ above. Such a construction produces the following groupoid where
each family of 1-morphisms that are identified by a family of 2-morphisms is
drawn using a thick line:

\begin{center}
\begin{tikzpicture}[scale=0.8,every node/.style={scale=0.8}]
  \draw[dashed] (0,-0.3) ellipse (3cm and 2.5cm);
  \node[below] (A) at (-2,0) {$ll$};
  \node[below] (B) at (0,0) {$lr$};
  \node[below] (C) at (1.5,0) {$r$};
  \path[ultra  thick] (A) edge [bend left=50] node[above] {$\permtwo$} (B);
  \path[ultra  thick] (C) edge [out=140, in=40, looseness=4] node[above] {$\permtwo$}  (C);
  \path[ultra  thick] (A) edge [my loop] node[below] {\texttt{id}} (A);
  \path[ultra  thick] (B) edge [my loop] node[below] {\texttt{id}} (B);
  \path[ultra  thick] (C) edge [my loop] node[below] {\texttt{id}} (C);
\end{tikzpicture}
\end{center}

Clearly, the resulting groupoid is a reconstruction of $G_2$ in
Fig.~\ref{fig:groupoids2} using $\Pi$ types and combinators. As analyzed
earlier, this groupoid has cardinality $\frac{3}{2}$. From the perspective
of~$\Pi$, this cardinality corresponds to the number of elements in the
underlying set which is $3$ divided by the order of the combinator $\permtwo$
which is 2. It is important to note that, as Def.~\ref{def:order} states, the
calculation of the order of a 1-combinator is defined up to the equivalence
induced by 2-combinators.

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Iteration Groupoids $\order{p}$}

The previous construction is quite useful: it allows us to take a set of
cardinality $N$ and a permutation on that set of order $P$ to construct a
groupoid of cardinality $\frac{N}{P}$. Although this idea allows us to construct
a groupoid of cardinality $\frac{3}{2}$ as shown above, it is not expressive
enough to construct a groupoid of cardinality $\frac{1}{3}$. Indeed if the
underlying set has only one element, the only permutation is the identity and
$P$ must be 1. The construction, however, already contains the main ingredient
needed for the construction of more general groupoids with fractional
cardinality. This key piece is the set of iterates of a combinator which we will
use in two different ways: (i) a perspective of combinators-as-data in which the
iterates of a combinator $p$ become the objects of groupoid $\order{p}$ of
cardinality $\ord{p}$, and (ii) a perspective of combinators-as-symmetries in
which the iterates of a combinator $p$ become the internal symmetries of a
one-element groupoid $\iorder{p}$ of cardinality $\frac{1}{\ord{p}}$. We explain
these two perspectives in this section and the next.

We define the $k^{\text{th}}$ iterate of $p$ as:

and then collect all such iterates into a type

\noindent For example, the zeroth and first iterate of a combinator are represented as

These ingredients are sufficient to construct a groupoid $\order{p}$ from the
iterates of a 1-combinator $p : \tau\iso\tau$. The objects will be
the triples $\triple{k}{q}{\alpha}$ indexed by integers $k$, 1-combinators $q :
\tau\iso\tau$, and 2-combinators $\alpha : q \isotwo p^k$. This triple encodes
our knowledge that we have some (arbitrary) iterate $q$ of $p$; we do not have
any a priori knowledge of the actual syntactic structure of $q$, but we do
know that it is equivalent to $p ^ k$.

We then add (reversible!) morphisms between any iterates related by
2-combinators; categorically, this will make any such objects equivalent.
If $p$ has order $o$, Lem.~\ref{lem:ordercancel} gives us a
2-combinator $\alpha$ which witnesses that $p^i \isotwo p^{i+o}$.
Thus given two iterates $\triple{i}{q_i}{\alpha_i}$ and
$\triple{i+o}{q_j}{\alpha_j}$, they must be equivalent since
$\alpha_i~\bullet~\alpha~\bullet~!\,\alpha_j$ shows that $q_i \isotwo q_j$.
In other words $p^j$ will be equivalent to $p^k$ exactly when $j$ and $k$
differ by $o$.  This informal description formalizes straightforwardly:

Despite its involved internal structure, the groupoid $\order{p}$ is essentially
a set of cardinality $\ord{p}$.

\begin{lemma}
  $|\order{p}| = \ord{p}$
\end{lemma}
\begin{proof}
  Let $o = \ord{p}$. There are $o$ isomorphism classes of
  objects. Consider an object $x = \triple{k}{q}{\alpha}$, its
  isomorphism class $[x] = \triple{k+io}{q_i}{\alpha_i}$ where
  $i \in \Z$. The group $\textsf{Aut}(x)$ is the group generated by
  $\idisotwo$ and has order 1. Hence
  $|\order{p}| = \sum\limits_{1}^{o}\frac{1}{1} = o$.
\end{proof}

As an example, the groupoid $\order{(\permtwo)}$ can be represented as follows. Up to equivalence, this groupoid is indeed equivalent to a set with two elements $\idiso$ and $\permtwo$.

\begin{center}
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.6}]
  \draw[dashed] (0,-0.3) ellipse (9cm and 2.5cm);

  \node[below] at (-8,0) {$\ldots$};
  \node[below] (A) at (-6,0) {$< -2 , \idiso , \ldots >$};
  \node[below] (B) at (-3,0) {$< -1 , \permtwo , \ldots >$};
  \node[below] (C) at (0,0) {$< 0, \idiso, \idisotwo >$};
  \node[below] (D) at (3,0) {$< 1 , \permtwo , \ldots >$};
  \node[below] (E) at (6,0) {$< 2 , \idiso , \ldots >$};
  \node[below] at (8,0) {$\ldots$};

  \path[ultra  thick] (A) edge [my loop] node[below] {\texttt{id}} (A);
  \path[ultra  thick] (B) edge [my loop] node[above] {\texttt{id}} (B);
  \path[ultra  thick] (C) edge [my loop] node[below] {\texttt{id}} (C);
  \path[ultra  thick] (D) edge [my loop] node[above] {\texttt{id}} (D);
  \path[ultra  thick] (E) edge [my loop] node[below] {\texttt{id}} (E);

  \path[ultra  thick] (A) edge [bend left=50] node[above] {$\alpha_{-2,0}$} (C);
  \path[ultra  thick] (C) edge [bend left=50] node[above] {$\alpha_{0,2}$} (E);
  \path[ultra  thick] (B) edge [bend left=-50] node[below] {$\alpha_{-1,1}$} (D);

\end{tikzpicture}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Symmetry Groupoids $\iorder{p}$}

The elements of $\iter{p}$ form a group under the following operation:
\[
\triple{k_1}{p_1}{\alpha_1} ~\circ~ \triple{k_2}{p_2}{\alpha_2} =
 \triple{k_1+k_2}{p_1 \odot p_2}{(\alpha_1 ~\respstwo~
    \alpha_2)~\transtwo~(\distiterplus{p}{k_1}{k_2})}
\]
where $\distiterplus{p}{k_1}{k_2}$ is defined in
Lem.~\ref{lem:distiterplus}. The common categorical representation of
a group is a category with one trivial object and the group elements
as morphisms on that trivial object. Our construction is essentially
the same except that the trivial object is represented as the iterates
of the identity over a singleton type.

Formally, given a 1-combinator $p : \tau\iso\tau$, the groupoid
$\iorder{p}$ is the groupoid (2-groupoid technically) whose objects
are the iterates in $\iter{\idiso_{\top}}$. Every pair of objects is
connected by the morphisms in $\iter{p}$. To capture the relationship
between these morphisms, we also have 2-morphisms relating the
distinct iterates as in the previous section.

\begin{lemma}
  $|\iorder{p}| = \frac{1}{\ord{p}}$
\end{lemma}
\begin{proof}
  Let $o = \ord{p}$. The objects form one isomorphism class
  $[p]$. There are, up to equivalence, exactly $\ord{p}$ distinct
  morphisms on this equivalence class. Hence, the group
  $\textsf{Aut}([p])$ is the group generated by
  $p^0, p^1 \dots p^{o-1}$, and the cardinality $|\iorder{p}|$ is
  $\frac{1}{o}$.
\end{proof}

Note that for each power $p ^ i$ of $p$, there is a morphism
$\triple{k}{q}{\alpha}$ in $\iter{p}$ such that $q$ annihilates $p^i$
to the identity. Note also that everything is well-defined even if we
choose $p : \zt\iso\zt$. In that case, the cardinality is 1.

\begin{center}
\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.6}]
  \draw[dashed] (0,-0.3) ellipse (9cm and 2.5cm);

  \node[below] at (-8,0) {$\ldots$};
  \node[below] (A) at (-6,0) {$< -2 , \idiso , \ldots >$};
  \node[below] (B) at (-3,0) {$< -1 , \idiso , \ldots >$};
  \node[below] (C) at (0,0) {$< 0, \idiso, \idisotwo >$};
  \node[below] (D) at (3,0) {$< 1 , \idiso , \ldots >$};
  \node[below] (E) at (6,0) {$< 2 , \idiso , \ldots >$};
  \node[below] at (8,0) {$\ldots$};

  \path[ultra  thick] (A) edge [my loop] node[below] {$\idiso$} (A);
  \path[ultra  thick] (A) edge [out=-140, in=-40, looseness=20] node[below] {$\permtwo$} (A);

  \path[ultra  thick] (A) edge [bend left=50] node[above] {$\alpha_{-2,0}$} (B);
  \path[ultra  thick] (B) edge [bend left=50] node[above] {$\alpha_{0,2}$} (C);
  \path[ultra  thick] (C) edge [bend left=50] node[above] {$\alpha_{0,2}$} (D);
  \path[ultra  thick] (D) edge [bend left=50] node[above] {$\alpha_{0,2}$} (E);
\end{tikzpicture}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sums and Products of Iteration and Symmetry Groupoids}

If we are to build a compositional programming language around iteration and
symmetry groupoids, we need to ensure that they compose sensibly with the
existing type formers. Groupoids, viewed as categories, come associated with
natural notions of sums and products, and one might expect or hope that
identities which hold for rational numbers lift to identities in our
situation. The most prominent and important identity we seek is:
such as:
\[
\order{p} \otimes \iorder{p} \simeq \mathbb{1}
\]
The groupoids on either side are not categorically equivalent but they
do have the same cardinality. Thus if we focus on
cardinality-preserving morphisms, it is sensible to consider them
equivalent. The situation is subtle, however, and studied in detail in
the next section.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^/$}

We are now ready to turn the groupoid constructions from the previous section
into an actual programming language. The language $\Pi^/$ will be an extension
of $\Pi$ with new type constructors and combinators for creating and
manipulating syntactic counterparts to $\order{p}$ and $\iorder{p}$.

We begin by adding four new syntactic categories to the definition of $\Pi$ in
Sec.~\ref{opsempi}:

\[\begin{array}{lrcl}
(\textrm{Types}) & \tau &::=& \ldots \\
(\textrm{Values}) & v &::=& \ldots \\
(\textrm{1-combinators}) & c &:& \tau_1 \iso \tau_2 ~::= \ldots \\
(\textrm{2-combinators}) & \alpha &:& c_1 \isotwo c_2 \mbox{~where~}
  c_1, c_2 : \tau_1 \iso \tau_2 ~::= \ldots \\
\\
(\textrm{Fractional Types}) & \mathbb{T} &::=&
  \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
  \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
 (\textrm{Fractional Values}) & \mathbb{V} &::=&
  c^k \alt \alpha^k \alt \langle \mathbb{V},\mathbb{V} \rangle \\
 (\textrm{/-combinators}) & \rho &:& \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::= \\
& &&     \idiso/ \alt \rho \circledcirc \rho \alt \rho \circledast \rho \\
& &\alt& \unitetl/ \alt \unititl/ \alt \unitetr/ \alt \unititr/ \\
& &\alt& \swapt/ \alt \assoclt/ \alt \assocrt/ \\
& &\alt& \order{\alpha} \alt \eta- \alt \eta+ \alt \epsilon+ \alt \epsilon- \\
(\textrm{//-combinators}) & \varrho &:& \rho_1 \leftrightblackspoon \rho_2 \mbox{~where~}
  \rho_1, \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::= \ldots
\end{array}\]

The syntactic category $\ft$ of fractional types introduces type expressions for
iteration groupoids, symmetry groupoids, and their products. The combinators
that relate these fractional types fall into several groups. The first line
ensures that the combinators include the identity and sequential composition
$\circledcirc$ and can be applied anywhere inside a product $\circledast$. The
next two lines ensure that $\circledast$ carries a symmetric monoidal structure
with combinators witnessing unit, commutativity, and associativity. The last
line relates two types $\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$
whenever $\alpha : c_1 \isotwo c_2$. Indeed if $c_1$ and $c_2$ are considered
equivalent via a 2-combinator then we also consider their iteration groupoids to
be equivalent. The last two combinators $\eta_{\ft}$ and $\epsilon_{\ft}$ are
inspired from the definition of compact closed categories~\cite{ccc} which are
symmetric monoidal categories in which every object has a dual. In that setting,
the dual $A^*$ of an object $A$ comes equipped with two morphisms:
\begin{itemize}
\item a \emph{unit} $\eta_A : 1 \rightarrow A^* \otimes A$, and
\item a \emph{counit} $\epsilon_A : A \otimes A^* \rightarrow 1$
\end{itemize}
that satisfy appropriate coherence conditions. These conherence conditions are
added as terms in the $\varrho$ syntactic category.

\paragraph*{Values.} When the types denote sets, it is evident what it means to have a
value of a given type: it is just an element of the set. When types
denote groupoids, it is less clear what it means to have a value,
especially when the total number of values, as reported by the
groupoid cardinality, is a proper fraction. We obviously cannot list
``half a value'' but what we \emph{can} do is to list an integral
number of values and provide an equivalence relation that specifies
which values are distinguishable such that the ultimate counting of
distinguishable values is a fractional amount. The idea is not
uncommon: in the conventional $\lambda$-calculus, we list $\lambda
x.x$ and $\lambda y.y$ as separate values of type $\tau \rightarrow
\tau$ and then provide a separate equivalence relation
($\alpha$-equivalence) to express the fact that these two values are
indistinguishable. The treatment in our setting is similar but richer
as in some cases the equivalence relation is not external but is
itself part of the value and the resulting count may be fractional.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{$\Pi^/$: Operational Semantics}

The operational semantics for all the primitive combinators is a
simple transliteration of Fig.~\ref{opsem}. We omit the
straightforward implementation of prim and its inverse
below:

The interesting combinators operationally are the unit/counit combinators $\eta$
and $\epsilon$, and the synchronization combinators. Their semantics is defined
by the following pair of mutually recursive interpreters:

The unit combinators $\eta$ simply generate unit tangles, and since unit tangles
are equivalent to the \emph{identity}, they can be freely eliminated by the
counit combinators $\epsilon$. The two synchronization operations exchange the
tangled product with an iterate. The two interpreters satisfy several properties
described below: First they are congruences for the equivalence $≈$ on values.
We can further prove that both are \emph{reversible}.  Furthermore, we can show
two coherence conditions: first, that the reverse interpreter gives equivalent
results to the forward interpreter applied to a reverse combinator, and second,
that two equivalent combinators (as given by a 2-combinator) will evaluate to
equivalent values.

The operational definition of synchr⋆ might seem at odds with our claim that it is related to associativity while it looks like commutativity.  The way to understand it is as follows: if indeed all values were indepedendent, it would be associativity.  However, in this case, $p // q$ is really two \emph{tangled} values, which we do not have access to.  To keep these synchronized and yet to achieve the given type, the only choice (operationally) is to swap.  This is where the ``action at a distance'' occurs.

We implement an example zig-zag, which use synchr⋆
to synchronize the \emph{tangled} values created by $\eta$ and $\epsilon$.
The circuit of zig-zag is shown in figure \ref{fig:zigzag} which
shows how the input of type $\#c$ is swapped by synchr⋆.

\begin{figure}[bht]
\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.9},highlight/.style={line width=4pt,orange!70}]
  \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
  \node at (0.5,1) {$\textsf{unit}_\times$};
  \path (-1.1,1) edge node[above] {${\textsf{\#c}}$} (0,1);
  \path (1,1.75) edge node[above] {$\mathbb{1}$} (1.6,1.75);
  \path [highlight] (1,0.25) edge node[above] {} (4,0.25);
  \path (1,0.25) edge node[above] {${\textsf{\#c}}$} (4,0.25);
  \draw (1.6,1) -- (2.6,1) -- (2.6,2.8) -- (1.6,2.8) -- cycle;
  \node at (2.1,1.9) {$\eta_{\textsf{c}}$};
  \draw[dashed] (2.8,2.6) -- (2.8,1.25);
  \path [highlight] (2.6,2.6) edge node[above] {} (4,2.6);
  \path (2.6,2.6) edge node[above] {${\textsf{\#c}}$} (4,2.6);
  \path [highlight] (2.6,1.25) edge node[above] {} (4,1.25);
  \path (2.6,1.25) edge node[above] {$1/\textsf{\#c}$} (4,1.25);
  \draw (4,0) -- (5.5,0) -- (5.5,3) -- (4,3) -- cycle;
  \node at (4.75,1.5) {$synch_{\textsf{c}}$};
  \path [highlight] (5.5,1.25) edge node[above] {} (7,1.25);
  \path (5.5,1.25) edge node[above] {$1/\#c$} (7,1.25);
  \path [highlight] (5.5,2.6) edge node[above] {} (9,2.6);
  \path (5.5,2.6) edge node[above] {$\#c$} (9,2.6);
  \draw (7,0) -- (8,0) -- (8,1.6) -- (7,1.6) -- cycle;
  \node at (7.5,0.8) {$\epsilon_c$};
  \draw[dashed] (6.8,0.2) -- (6.8,1.25);
  \path (8,1) edge node[above] {$\mathbb{1}$} (9,1);
  \path [highlight] (5.5,0.2) edge node[above] {} (7,0.2);
  \path (5.5,0.2) edge node[above] {$\#c$} (7,0.2);
  \draw (9,0.8) -- (10,0.8) -- (10,2.8) -- (9,2.8) -- cycle;
  \node at (9.5,1.8) {$\textsf{unit}_\times$};
  \path (10,1.8) edge node [above] {$\#c$} (11,1.8);
\end{tikzpicture}
\caption{\label{fig:zigzag} Zig-zag}
\end{figure}

The following is an example usage of zig-zag, which uses NOT combinator.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Limitations}

Given any non-negative rational number, we can form a type whose cardinality is
that number. And yet, our types do not capture the full structure of the
non-negative rational numbers, as these form a commutative semifield. There are
a number of operations and properties which need to be added to complete this.
Most seem quite straightforward.  We can however single one out which may not
be: what we are missing is a multiplicative inverse for every type, and not just
$\order{p}$. In particular, we would like to form the type
$\iorder{(\iorder{p})}$ and have it be equivalent to $\order{p}$.  The
relationship between $\order{p}$ and $\iorder{p}$ in $\Pi^/$ reveals a pattern
that can be exploited to achieve this generalization. The multiplicative inverse
of $\order{p}$ is obtained by promoting the objects in $\order{p}$ to morphisms
in $1/\hash p$, which is a general process called \emph{delooping}. It is clear
we can go in the inverse direction by demoting morphisms to objects which is a
process called \emph{looping}. We leave a possible extension with a general
looping/delooping process to future work.

A second more immediate concern is to discover the abstract monadic and/or
comonadic structure underlying the operational semantics. It is clear that the
operational interpretation of fractional types involves a computational effect
that requires synchronization with the ``future''. Implementing this effect
using backtracking, reference cells, or dataflow constraints is possible but
each implementation choice is ad hoc. We conjecture that with the right monadic
and/or comonadic abstraction, the operational semantics will be simpler and
expose still richer properties.  We chose the dependently-typed dataflow
semantics because we are able to prove that it is reversible and coherent,
unlike our previous attempts with explicit effects, none of which were provably
reversible and/or coherent.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}

We have presented a natural notion of \emph{fractional types} that
enriches a class of reversible programming languages in several
dimensions.  Further research might show how to use fractional
types in a conventional (i.e., irreversible) programming language, their full
potential is only achieved when the ambient language guarantees that
no information is created or erased.

The key semantic insight is that iterating a reversible program $p$ on
a finite type must eventually reach the identity in $\ord{p}$
steps. By being careful not to collapse proofs, each such reversible
program has $\ord{p}$ distinct proofs of reversibility and hence gives
rise to a groupoid with cardinality $\frac{1}{\ord{p}}$. Going from
this observation to a full programming language required several
difficult and subtle design choices which we have explored to produce
$\Pi^/$. The latter language has a natural denotational semantics
where types denote groupoids. Its operational semantics requires a
mechanism to express a computational effect which enables spatially
separated parts of the program to communicate in a way that is
reminiscent of entanglement in quantum mechanics. It is possible to
realize such an operational semantics using global reference cells,
backtracking, or other conventional technique. A more enlightening and
less ad hoc implementation encodes the required dependency in dataflow
constraints encoded in dependent types. The key idea is to generalize
the usual cartesian product to a \emph{tangled product} that allows the
components to interact at synchronization points.

Our fractional types extend the natural denotation of types from sets
to non-trivial groupoids but they only scratch the surface of the
tower of weak $\omega$-groupoids that is expressible in HoTT. A long
term goal of our research is to find natural type constructors
inspired by the rich combinatorial structure of weak
$\omega$-groupoids and that provide novel programming abstractions.

For an information-theoretic perspective on the language $\Pi$, we think of a
type containing $N$ values as an abstract system that has~$N$ distinguishable
states. According to the conventional theory of information~\cite{Shannon1948},
the amount of information contained in each state of a system with $N$
distinguishable states is $(\log N)$ bits of information. For example, the type
$\mathbb{2}$ can be thought of as an abstract system with two distinguishable
states labeled $\mathsf{true}$ and $\mathsf{false}$ each containing $\log 2 = 1$
bit. Similarly, the type $\mathbb{3}$ can be thought of as an abstract system
with three distinguishable states each containing $\log 3$ bits. The logarithmic
map implies that information contained in a composite state is the sum of the
information contained in its constituents. For example, the type $\mathbb{2}
\otimes \mathbb{3}$ can be thought of a composite system consisting of two
independent unrelated subsystems. Each state of the composite system therefore
contains $\log (2 * 3) = \log 2 + \log 3$ bits which is the sum of the
information contained in each subsystem. Since all the combinators preserve
cardinality, they are also information-preserving.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plainurl}
\bibliography{cites}

\end{document}

### 2016-12-15 typespp.tex
clean up more and up to sec 2 small revisions

@@
 \usepackage{datetime}
+\usepackage{fdsymbol} %% must be loaded before bbold because it interferes with numbers
 \usepackage{bbold}
@@
 \usepackage[section]{placeins}
-\usepackage{fdsymbol}
-
-\newcommand{\amr}[1]{\fbox{\begin{minipage}{0.4\textwidth}\color{purple}{Amr says: #1}\end{minipage}}}
-\newcommand{\vic}[1]{\fbox{\begin{minipage}{0.4\textwidth}\color{purple}{Vikraman says: #1}\end{minipage}}}
-\newcommand{\jc}[1]{\fbox{\begin{minipage}{\textwidth}\color{purple}{Jacques says: #1}\end{minipage}}}
-
-\newcommand{\hide}[1]{}
 
@@
 \newcommand{\hash}{\#}
-\newcommand{\dg}{conjugation groupoid}
-\newcommand{\DG}{Conjugation Groupoid}
 \newcommand{\pifrac}{\ensuremath{\Pi^/}}
@@
 
-\keywords{dummy}
-\subjclass{dummy}
-
 \maketitle
@@
 \begin{abstract}
-
 We exhibit types whose natural cardinality is fractional.  More
@@
 having exactly $k$ proofs of reversibility.
-
 \end{abstract}
@@
 \begin{center}
-\begin{tabular}{c@{\kern -4pt}c}
+\begin{tabular}{c@{\qquad\qquad}c}
 \begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]
-  \draw[dashed] (0,0) ellipse (3cm and 1.2cm);
+  \draw[dashed] (0,0) ellipse (3cm and 1.5cm);
   \node[below] (A) at (-1.7,0.3) {\texttt{a}};
@@
   \draw (0.5,0.3) to[bend left] (1.7,0.3) ;
-  \path (A) edge [loop below] node[below] {\texttt{id}} (B);
-  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
-  \path (C) edge [loop below] node[below] {\texttt{id}} (B);
-  \path (D) edge [loop below] node[below] {\texttt{id}} (B);
-  \path (E) edge [loop below] node[below] {\texttt{id}} (B);
-  \path (F) edge [loop below] node[below] {\texttt{id}} (B);
+  \path (A) edge [my loop] node[below] {\texttt{id}} (B);
+  \path (B) edge [my loop] node[below] {\texttt{id}} (B);
+  \path (C) edge [my loop] node[below] {\texttt{id}} (B);
+  \path (D) edge [my loop] node[below] {\texttt{id}} (B);
+  \path (E) edge [my loop] node[below] {\texttt{id}} (B);
+  \path (F) edge [my loop] node[below] {\texttt{id}} (B);
 \end{tikzpicture}
 &
-\begin{tikzpicture}[scale=0.5,every node/.style={scale=0.5}]
-  \draw[dashed] (0,0) ellipse (2cm and 2.4cm);
+\begin{tikzpicture}[scale=0.6,every node/.style={scale=0.6}]
+  \draw[dashed] (0,0) ellipse (2cm and 2.6cm);
   \node[below] (B) at (0,-1.1) {\texttt{*}};
-  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
+  \path (B) edge [my loop] node[below] {\texttt{id}} (B);
   \path (B) edge [loop above, looseness=15, in=48, out=132] node[above] {$p$} (B);
@@
 \texttt{d}, \texttt{e}, and \texttt{f} with two groups of three
-points clustered in an equivalence class and hence the groupoid has
+points each clustered in an equivalence class and hence the groupoid has
 cardinality~2. The (2-)groupoid on the right has one point $\ast$ with
@@
 \end{array}\]
-We should note that the above definition is the only one in this
-paper which is not \emph{effective};  in other words, we do not have an
-algorithm for computing it.  While there is an obvious method to
-compute it using the action of a 1-combinator on the elements of the
-type it acts on, this is extremely inefficient.  What we mean is that
-we know of no effective algorithm which works on the syntax of
-combinators.  The (only) difficulty is $\odot$, which can have an
-arbitrary effect on the order.
+We should note that the above definition is the only one in this paper which is not \emph{effective}.
+While there is an obvious method to compute it using the action of a 1-combinator on the elements of the type it acts on, this is extremely inefficient.
+We do not have an effective algorithm for computing it that works on the syntax of combinators.
+The (only) difficulty is $\odot$, which can have an arbitrary effect on the order.
 
@@
   \node[below] (C) at (0.5,0) {$c$};
-  \path (A) edge [loop below] node[below] {\texttt{id}} (A);
-  \path (C) edge [loop below] node[below] {\texttt{id}} (C);
+  \path (A) edge [my loop] node[below] {\texttt{id}} (A);
+  \path (C) edge [my loop] node[below] {\texttt{id}} (C);
   \path (C) edge [out=140, in=40, looseness=4] (C);
@@
   \path (C) edge [out=140, in=40, looseness=4] (C);
-  \path (A) edge [loop below] node[below] {\texttt{id}} (A);
-  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
-  \path (C) edge [loop below] node[below] {\texttt{id}} (C);
+  \path (A) edge [my loop] node[below] {\texttt{id}} (A);
+  \path (B) edge [my loop] node[below] {\texttt{id}} (B);
+  \path (C) edge [my loop] node[below] {\texttt{id}} (C);
 \end{tikzpicture}
@@
   \node[below] (C) at (1.6,0) {$c$};
-  \path (A) edge [loop below] node[below] {\texttt{id}} (A);
-  \path (B) edge [loop below] node[below] {\texttt{id}} (B);
-  \path (C) edge [loop below] node[below] {\texttt{id}} (C);
+  \path (A) edge [my loop] node[below] {\texttt{id}} (A);
+  \path (B) edge [my loop] node[below] {\texttt{id}} (B);
+  \path (C) edge [my loop] node[below] {\texttt{id}} (C);
   \path (A) edge [loop above, looseness=3, in=48, out=132] (A);

### 2016-12-15 typespp.tex
up to end sec 3

@@
 
-\begin{definition}[Power of a 1-combinator]
-  The $k^{th}$ power of a 1-combinator $p : \tau \iso \tau$, for
+\begin{definition}[Iterated Power of a 1-combinator]
+  The $k^{th}$ iterated power of a 1-combinator $p : \tau \iso \tau$, for
   $k \in \Z$ is
@@
 
-From a denotational perspective, a $\Pi$ type $\tau$ denotes a finite set, a
-$\Pi$ 1-combinator denotes a permutation on finite sets, and the 2-combinators
-denote coherence conditions on these permutations~\cite{Carette2016}. Formally,
-the language $\Pi$ is a \emph{categorification}~\cite{math/9802029} of the
-natural numbers as a \emph{symmetric rig groupoid}~\cite{nlabrig}. This
-structure is a \emph{symmetric bimonoidal category} or a \emph{commutative rig
-category} in which every morphism is invertible. The underlying category
-consists of two symmetric monoidal structures~\cite{nla.cat-vn1051288}
-separately induced by the properties of addition and multiplication of the
-natural numbers. The monoidal structures are then augmented with distributivity
-and absorption natural isomorphisms~\cite{laplaza} to model the full commutative
-semiring (aka, commutative rig) of the natural numbers. Despite this rich
-structure, the individual objects in the category for $\Pi$ are just plain sets
-with no interesting structure. In this section we introduce, in the denotation
-of $\Pi$, some non-trivial groupoids which we call \emph{iteration groupoids}
-and \emph{symmetry groupoids}. Under certain conditions, sums and products of
-these groupoids behave as expected which ensures that a sensible compositional
-programming language can be designed around them.
+From a denotational perspective, a $\Pi$ type $\tau$ denotes a finite set, a $\Pi$ 1-combinator denotes a permutation on finite sets, and the 2-combinators denote coherence conditions on these permutations~\cite{Carette2016}.
+Formally, the language $\Pi$ is a \emph{categorification}~\cite{math/9802029} of the natural numbers as a \emph{symmetric rig groupoid}~\cite{nlabrig}.
+This structure is a \emph{symmetric bimonoidal category} or a \emph{commutative rig category} in which every morphism is invertible.
+The underlying category consists of two symmetric monoidal structures~\cite{nla.cat-vn1051288} separately induced by the properties of addition and multiplication of the natural numbers.
+The monoidal structures are then augmented with distributivity and absorption natural isomorphisms~\cite{laplaza} to model the full commutative semiring (aka, commutative rig) of the natural numbers.
+Despite this rich structure, the individual objects in the category for $\Pi$ are just plain sets with no interesting structure.
+In this section we introduce, in the denotation of $\Pi$, some non-trivial groupoids which we call \emph{iteration groupoids} and \emph{symmetry groupoids}.
+Products of these groupoids behave as expected which ensures that a sensible compositional programming language can be designed around them.
 
@@
 a groupoid of cardinality $\frac{3}{2}$ as shown above, it is not expressive
-enough to construct a groupoid of cardinality $\frac{1}{3}$. Indeed if the
+enough to construct a groupoid of cardinality, say, $\frac{1}{3}$. Indeed if the
 underlying set has only one element, the only permutation is the identity and
-$P$ must be 1. The construction, however, already contains the main ingredient
-needed for the construction of more general groupoids with fractional
-cardinality. This key piece is the set of iterates of a combinator which we will
-use in two different ways: (i) a perspective of combinators-as-data in which the
-iterates of a combinator $p$ become the objects of groupoid $\order{p}$ of
-cardinality $\ord{p}$, and (ii) a perspective of combinators-as-symmetries in
-which the iterates of a combinator $p$ become the internal symmetries of a
-one-element groupoid $\iorder{p}$ of cardinality $\frac{1}{\ord{p}}$. We explain
-these two perspectives in this section and the next.
-
-We define the $k^{\text{th}}$ iterate of $p$ as:
-
-and then collect all such iterates into a type
-
-\noindent For example, the zeroth and first iterate of a combinator are represented as
-
-These ingredients are sufficient to construct a groupoid $\order{p}$ from the
-iterates of a 1-combinator $p : \tau\iso\tau$. The objects will be
-the triples $\triple{k}{q}{\alpha}$ indexed by integers $k$, 1-combinators $q :
-\tau\iso\tau$, and 2-combinators $\alpha : q \isotwo p^k$. This triple encodes
-our knowledge that we have some (arbitrary) iterate $q$ of $p$; we do not have
-any a priori knowledge of the actual syntactic structure of $q$, but we do
-know that it is equivalent to $p ^ k$.
-
-We then add (reversible!) morphisms between any iterates related by
-2-combinators; categorically, this will make any such objects equivalent.
-If $p$ has order $o$, Lem.~\ref{lem:ordercancel} gives us a
-2-combinator $\alpha$ which witnesses that $p^i \isotwo p^{i+o}$.
-Thus given two iterates $\triple{i}{q_i}{\alpha_i}$ and
-$\triple{i+o}{q_j}{\alpha_j}$, they must be equivalent since
-$\alpha_i~\bullet~\alpha~\bullet~!\,\alpha_j$ shows that $q_i \isotwo q_j$.
-In other words $p^j$ will be equivalent to $p^k$ exactly when $j$ and $k$
-differ by $o$.  This informal description formalizes straightforwardly:
+$P$ must be 1.
+
+The construction, however, already contains the main ingredient needed for the construction of more general groupoids with fractional cardinality.
+This key piece is the set of iterates of a combinator which we formally define as follows.
+
+\begin{definition}[$\iter{p}$]
+For each 1-combinator $p : \tau\iso\tau$, we form the set $\iter{p}$ whose elements are triples consisting of an integer $k$, a 1-combinator $q : \tau\iso\tau$ and a 2-combinator $\alpha : q \isotwo p^k$.
+\end{definition}
+Each triple encodes our knowledge that we have some (arbitrary) iterate $q$ of $p$; we do not have any a priori knowledge of the actual syntactic structure of $q$, but we do know that it is equivalent to $p^k$.
+For example, we have:
+\[\begin{array}{rcl}
+ \iter{\permtwo} &=& \{ \triple{0}{\idiso}{\idisotwo},\triple{1}{\permtwo}{\idrdr}, \triple{-1}{\permtwo}{\idisotwo}, \ldots \}
+\end{array}\]
+The idea is that $\iter{\permtwo}$ is, up to equivalence, the set of all distinct iterates $(\permtwo)^k$ of $\permtwo$.
+Because of the underlying group structure of automorphisms, there are, up to equivalence, only $\ord{\permtwo}$ distinct iterates in $\iter{\permtwo}$.
+In a proof-irrelevant setting, $\iter{\permtwo}$ is simply $\{ \idiso, \permtwo \}$.
+In this section and the next, we will use the elements of $\iter{p}$ in two groupoid constructions as either objects (emphasizing their ``data'' aspect) or morphisms (emphasizing their ``symmetry'' aspect).
+
+Given a 1-combinator $p : \tau\iso\tau$, we define the groupoid $\order{p}$ as follows.
+The objects are the elements of $\iter{p}$, i.e., the triples $\triple{k}{q}{\alpha}$ indexed by integers $k$, 1-combinators $q : \tau\iso\tau$, and 2-combinators $\alpha : q \isotwo p^k$.
+We then add (reversible) morphisms between any iterates related by 2-combinators; categorically, this will make any such objects equivalent.
+If $p$ has order $o$, Lem.~\ref{lem:ordercancel} gives us a 2-combinator $\alpha$ which witnesses that $p^i \isotwo p^{i+o}$.
+Thus given two iterates $\triple{i}{q_i}{\alpha_i}$ and $\triple{i+o}{q_j}{\alpha_j}$, they must be equivalent since $\alpha_i~\bullet~\alpha~\bullet~!\,\alpha_j$ shows that $q_i \isotwo q_j$.
+In other words $p^j$ will be equivalent to $p^k$ exactly when $j$ and $k$ differ by $o$.
+This informal description formalizes straightforwardly.
+
+\begin{definition}[$\order{p}$] For each 1-combinator $p : \tau\iso\tau$, we form the groupoid $\order{p}$ as follows:
+\begin{itemize}
+\item objects are the elements $\triple{k}{q}{\alpha}$ of $\iter{p}$;
+\item there is a morphism between $\triple{k_1}{q_1}{\alpha_1}$ and $\triple{k_2}{q_2}{\alpha_2}$ for each $\alpha : q_1 \isotwo q_2$
+\end{itemize}
+\end{definition}
 
@@
   \path[ultra  thick] (A) edge [my loop] node[below] {\texttt{id}} (A);
-  \path[ultra  thick] (B) edge [my loop] node[above] {\texttt{id}} (B);
+  \path[ultra  thick] (B) edge [my loop] node[below] {\texttt{id}} (B);
   \path[ultra  thick] (C) edge [my loop] node[below] {\texttt{id}} (C);
-  \path[ultra  thick] (D) edge [my loop] node[above] {\texttt{id}} (D);
+  \path[ultra  thick] (D) edge [my loop] node[below] {\texttt{id}} (D);
   \path[ultra  thick] (E) edge [my loop] node[below] {\texttt{id}} (E);
@@
 \]
-where $\distiterplus{p}{k_1}{k_2}$ is defined in
-Lem.~\ref{lem:distiterplus}. The common categorical representation of
-a group is a category with one trivial object and the group elements
-as morphisms on that trivial object. Our construction is essentially
-the same except that the trivial object is represented as the iterates
-of the identity over a singleton type.
-
-Formally, given a 1-combinator $p : \tau\iso\tau$, the groupoid
-$\iorder{p}$ is the groupoid (2-groupoid technically) whose objects
-are the iterates in $\iter{\idiso_{\top}}$. Every pair of objects is
-connected by the morphisms in $\iter{p}$. To capture the relationship
-between these morphisms, we also have 2-morphisms relating the
-distinct iterates as in the previous section.
+where $\distiterplus{p}{k_1}{k_2}$ is defined in Lem.~\ref{lem:distiterplus}.
+The common categorical representation of a group is a category with one trivial object and the group elements as morphisms on that trivial object.
+Our construction of the groupoid $\iorder{p}$ is essentially the same.
+
+\begin{definition}[$\iorder{p}$] For each 1-combinator $p : \tau\iso\tau$, we form the 2-groupoid $\iorder{p}$ as follows:
+\begin{itemize}
+\item the objects are the iterates of the identity combinator on $\tau$;
+\item the morphisms between every pair of objects are the elements of $\iter{p}$;
+\item there is a 2-morphisms between between 1-morphisms $\triple{k_1}{q_1}{\alpha_1}$ and $\triple{k_2}{q_2}{\alpha_2}$ for each $\alpha : q_1 \isotwo q_2$ for every $\alpha : q_1 \isotwo q_2$
+\end{itemize}
+\end{definition}
+
+Note that for each power $p ^ i$ of $p$, there is a morphism $\triple{k}{q}{\alpha}$ in $\iter{p}$ such that $q$ annihilates $p^i$ to the identity.
+Note also that everything is well-defined even if we choose $p : \zt\iso\zt$.
+In that case, the cardinality is 1.
 
@@
 
-Note that for each power $p ^ i$ of $p$, there is a morphism
-$\triple{k}{q}{\alpha}$ in $\iter{p}$ such that $q$ annihilates $p^i$
-to the identity. Note also that everything is well-defined even if we
-choose $p : \zt\iso\zt$. In that case, the cardinality is 1.
+As an example, the groupoid $\iorder{(\permtwo)}$ can be represented as follows.
 
@@
 \begin{tikzpicture}[scale=0.6,every node/.style={scale=0.6}]
-  \draw[dashed] (0,-0.3) ellipse (9cm and 2.5cm);
+  \draw[dashed] (0,-0.3) ellipse (9cm and 3cm);
 
@@
   \path[ultra  thick] (A) edge [my loop] node[below] {$\idiso$} (A);
-  \path[ultra  thick] (A) edge [out=-140, in=-40, looseness=20] node[below] {$\permtwo$} (A);
+  \path[ultra  thick] (B) edge [my loop] node[below] {$\idiso$} (B);
+  \path[ultra  thick] (C) edge [my loop] node[below] {$\idiso$} (C);
+  \path[ultra  thick] (D) edge [my loop] node[below] {$\idiso$} (D);
+  \path[ultra  thick] (E) edge [my loop] node[below] {$\idiso$} (E);
 
@@
   \path[ultra  thick] (D) edge [bend left=50] node[above] {$\alpha_{0,2}$} (E);
+
+  \path[ultra  thick] (A) edge [out=-140, in=-40, looseness=10] node[below] {$\permtwo$} (A);
+  \path[ultra  thick] (B) edge [out=-140, in=-40, looseness=10] node[below] {$\permtwo$} (B);
+  \path[ultra  thick] (C) edge [out=-140, in=-40, looseness=10] node[below] {$\permtwo$} (C);
+  \path[ultra  thick] (D) edge [out=-140, in=-40, looseness=10] node[below] {$\permtwo$} (D);
+  \path[ultra  thick] (E) edge [out=-140, in=-40, looseness=10] node[below] {$\permtwo$} (E);
 \end{tikzpicture}
@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sums and Products of Iteration and Symmetry Groupoids}
+\subsection{Products of Iteration and Symmetry Groupoids}
 
@@
 \]
-The groupoids on either side are not categorically equivalent but they
-do have the same cardinality. Thus if we focus on
-cardinality-preserving morphisms, it is sensible to consider them
-equivalent. The situation is subtle, however, and studied in detail in
-the next section.
+The groupoids on either side are not categorically equivalent but they do have the same cardinality.
+Thus if we focus on some notion of cardinality-preserving morphisms, it is sensible to consider them equivalent.
+The situation is however subtle and does not directly support a notion of sums.
+It is studied in detail in the next section.
 
@@
 \section{$\Pi^/$}
+
+***WAVEFRONT***
 

### 2016-12-15 typespp.tex
working on sec 4

@@
 \Rule{}
-{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_2 \iso \tau_3}
+{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad c_2 : \tau_2 \iso \tau_3}
 {\jdg{}{}{c_1 \odot c_2 : \tau_1 \iso \tau_3}}
@@
 \Rule{}
-{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
+{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad c_2 : \tau_3 \iso \tau_4}
 {\jdg{}{}{c_1 \oplus c_2 : \tau_1 \oplus \tau_3 \iso \tau_2 \oplus \tau_4}}
@@
 \Rule{}
-{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad \vdash c_2 : \tau_3 \iso \tau_4}
+{\jdg{}{}{c_1 : \tau_1 \iso \tau_2} \quad c_2 : \tau_3 \iso \tau_4}
 {\jdg{}{}{c_1 \otimes c_2 : \tau_1 \otimes \tau_3 \iso \tau_2 \otimes \tau_4}}
@@
 language has two other syntactic categories of programs to be
-described in detail:
+described in detail.
+
+\begin{definition}[$\Pi$]
+\label{def:pi}
+The syntax of $\Pi$ is given by the following categories:
 \[\begin{array}{lrcl}
@@
 \end{array}\]
+\end{definition}
 
@@
 
-***WAVEFRONT***
-
-We are now ready to turn the groupoid constructions from the previous section
-into an actual programming language. The language $\Pi^/$ will be an extension
-of $\Pi$ with new type constructors and combinators for creating and
-manipulating syntactic counterparts to $\order{p}$ and $\iorder{p}$.
-
-We begin by adding four new syntactic categories to the definition of $\Pi$ in
-Sec.~\ref{opsempi}:
-
+We are now ready to turn the groupoid constructions from the previous section into an extension of $\Pi$ with new type constructors and combinators for creating and manipulating syntactic counterparts to $\order{p}$ and $\iorder{p}$.
+
+\begin{figure}[ht]
+\[\begin{array}{rrcll}
+\unitetl/ :&  \order{\idiso} \circledast \mathbb{T} & \leftrightspoon & \mathbb{T} &: \unititl/ \\
+\unitetr/ :&  \mathbb{T} \circledast \order{\idiso} & \leftrightspoon & \mathbb{T} &: \unititr/ \\
+\swapt/ :&  \mathbb{T}_1 \circledast \mathbb{T}_2 & \leftrightspoon & \mathbb{T}_2 \circledast \mathbb{T}_1 &: \swapt/ \\
+\assoclt/ :&  \mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3) & \leftrightspoon & (\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3
+  &: \assocrt/ \\
+\\
+\eta- :& \order{\idiso} & \leftrightspoon & \iorder{c} \circledast \order{c} &: \epsilon- \\
+\eta+ :& \order{\idiso} & \leftrightspoon & \order{c} \circledast \iorder{c} &: \epsilon+ \\
+\end{array}\]
+
+\[\begin{array}{c}
+\Rule{}
+{}
+{\jdg{}{}{\idiso/ : \mathbb{T} \leftrightspoon \mathbb{T}}}
+{}
+\qquad
+\Rule{}
+{\jdg{}{}{\rho_1 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2} \quad \rho_2 : \mathbb{T}_2 \leftrightspoon \mathbb{T}_3}
+{\jdg{}{}{\rho_1 \circledcirc \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_3}}
+{}
+\\
+\\
+\Rule{}
+{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
+{\jdg{}{}{\order{\alpha} : \order{c_1}  \leftrightspoon \order{c_2}}}
+{}
+\qquad
+\Rule{}
+{\jdg{}{}{\rho_1 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2} \quad \rho_2 : \mathbb{T}_3 \leftrightspoon \mathbb{T}_4}
+{\jdg{}{}{\rho_1 \circledast \rho_2 : \mathbb{T}_1 \circledast \mathbb{T}_3 \leftrightspoon \mathbb{T}_2 \circledast \mathbb{T}_4}}
+{}
+\end{array}\]
+
+Each combinator $\rho$ has an inverse.
+\caption{$\Pi^/$ fractional combinators}
+\label{pifrac:comb}
+\end{figure}
+
+We call the new extension $\Pi^/$. We begin by adding four new syntactic categories to the definition of $\Pi$ in Sec.~\ref{opsempi}:
+
+\begin{definition}[$\Pi^/$]
+\label{def:pifrac}
+The syntax of $\Pi^/$ is given by the following categories:
 \[\begin{array}{lrcl}
-(\textrm{Types}) & \tau &::=& \ldots \\
-(\textrm{Values}) & v &::=& \ldots \\
-(\textrm{1-combinators}) & c &:& \tau_1 \iso \tau_2 ~::= \ldots \\
+(\textrm{Types}) & \tau &::=& [\textit{see Def.~\ref{def:pi}}] \\
+(\textrm{Values}) & v &::=& [\textit{see Def.~\ref{def:pi}}] \\
+(\textrm{1-combinators}) & c &:& \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
 (\textrm{2-combinators}) & \alpha &:& c_1 \isotwo c_2 \mbox{~where~}
-  c_1, c_2 : \tau_1 \iso \tau_2 ~::= \ldots \\
+  c_1, c_2 : \tau_1 \iso \tau_2 ~::= [\textit{see Def.~\ref{def:pi}}] \\
 \\
@@
   \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
-  \order{c} \alt \iorder{c} \alt \mathbb{T}_1 \circledast \mathbb{T}_2 \\
  (\textrm{Fractional Values}) & \mathbb{V} &::=&
-  c^k \alt \alpha^k \alt \langle \mathbb{V},\mathbb{V} \rangle \\
- (\textrm{/-combinators}) & \rho &:& \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::= \\
-& &&     \idiso/ \alt \rho \circledcirc \rho \alt \rho \circledast \rho \\
-& &\alt& \unitetl/ \alt \unititl/ \alt \unitetr/ \alt \unititr/ \\
-& &\alt& \swapt/ \alt \assoclt/ \alt \assocrt/ \\
-& &\alt& \order{\alpha} \alt \eta- \alt \eta+ \alt \epsilon+ \alt \epsilon- \\
+  c^k \alt 1/c^k \alt \langle \mathbb{V},\mathbb{V} \rangle \\
+ (\textrm{/-combinators}) & \rho &:& \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
+  [\textit{see Fig.~\ref{pifrac:comb}}] \\
 (\textrm{//-combinators}) & \varrho &:& \rho_1 \leftrightblackspoon \rho_2 \mbox{~where~}
-  \rho_1, \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::= \ldots
-\end{array}\]
+  \rho_1, \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
+  [\textit{see Fig.~\ref{pifrac:twocomb}}]
+\end{array}\]
+\end{definition}
 

### 2016-12-16 typespp.tex
typos
sec 4

@@
 
-We call the new extension $\Pi^/$. We begin by adding four new syntactic categories to the definition of $\Pi$ in Sec.~\ref{opsempi}:
+We call the new extension $\Pi^/$. We begin by adding three new syntactic categories to the definition of $\Pi$ in Sec.~\ref{opsempi}:
 
@@
  (\textrm{/-combinators}) & \rho &:& \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
-  [\textit{see Fig.~\ref{pifrac:comb}}] \\
-(\textrm{//-combinators}) & \varrho &:& \rho_1 \leftrightblackspoon \rho_2 \mbox{~where~}
-  \rho_1, \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2 ~::=
-  [\textit{see Fig.~\ref{pifrac:twocomb}}]
+  [\textit{see Fig.~\ref{pifrac:comb}}]
 \end{array}\]
@@
 
-The syntactic category $\ft$ of fractional types introduces type expressions for
-iteration groupoids, symmetry groupoids, and their products. The combinators
-that relate these fractional types fall into several groups. The first line
-ensures that the combinators include the identity and sequential composition
-$\circledcirc$ and can be applied anywhere inside a product $\circledast$. The
-next two lines ensure that $\circledast$ carries a symmetric monoidal structure
-with combinators witnessing unit, commutativity, and associativity. The last
-line relates two types $\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$
-whenever $\alpha : c_1 \isotwo c_2$. Indeed if $c_1$ and $c_2$ are considered
-equivalent via a 2-combinator then we also consider their iteration groupoids to
-be equivalent. The last two combinators $\eta_{\ft}$ and $\epsilon_{\ft}$ are
-inspired from the definition of compact closed categories~\cite{ccc} which are
-symmetric monoidal categories in which every object has a dual. In that setting,
-the dual $A^*$ of an object $A$ comes equipped with two morphisms:
-\begin{itemize}
-\item a \emph{unit} $\eta_A : 1 \rightarrow A^* \otimes A$, and
-\item a \emph{counit} $\epsilon_A : A \otimes A^* \rightarrow 1$
-\end{itemize}
-that satisfy appropriate coherence conditions. These conherence conditions are
-added as terms in the $\varrho$ syntactic category.
-
-\paragraph*{Values.} When the types denote sets, it is evident what it means to have a
-value of a given type: it is just an element of the set. When types
-denote groupoids, it is less clear what it means to have a value,
-especially when the total number of values, as reported by the
-groupoid cardinality, is a proper fraction. We obviously cannot list
-``half a value'' but what we \emph{can} do is to list an integral
-number of values and provide an equivalence relation that specifies
-which values are distinguishable such that the ultimate counting of
-distinguishable values is a fractional amount. The idea is not
-uncommon: in the conventional $\lambda$-calculus, we list $\lambda
-x.x$ and $\lambda y.y$ as separate values of type $\tau \rightarrow
-\tau$ and then provide a separate equivalence relation
-($\alpha$-equivalence) to express the fact that these two values are
-indistinguishable. The treatment in our setting is similar but richer
-as in some cases the equivalence relation is not external but is
-itself part of the value and the resulting count may be fractional.
+The syntactic category $\ft$ of fractional types introduces type expressions for iteration groupoids, symmetry groupoids, and their products.
+The combinators that relate these fractional types are in Fig.~\ref{pifrac:comb}.
+The bottom group includes identity and sequential composition $\circledcirc$ combinators and ensures that the combinators can be applied anywhere inside a product $\circledast$.
+There is also a combinator that relates two types $\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo c_2$.
+Indeed if $c_1$ and $c_2$ are considered equivalent via a 2-combinator then we also consider their iteration groupoids to be equivalent.
+The top group ensures that $\circledast$ carries a symmetric monoidal structure with combinators witnessing the unit, commutativity, and associativity properties.
+The last two pairs combinators $\eta-/\epsilon-$ and $\eta+/\epsilon+$ are inspired from the definition of compact closed categories~\cite{ccc} which are
+symmetric monoidal categories in which every object has a dual.
+They are the ones that witness the ``fractional'' nature of symmetry groupoids: they are indeed the multiplicative inverses of iteration groupoids.
+
+The last new syntactic category in $\Pi^/$ is that of values which deserves some background discussion.
+When the types denote sets, it is evident what it means to have a value of a given type: it is just an element of the set.
+When types denote groupoids, it is less clear what it means to have a value, especially when the total number of values, as reported by the groupoid cardinality, is a proper fraction.
+In that case we the values up to an equivalence relation.
+The idea is not uncommon: in the conventional $\lambda$-calculus, we list $\lambda x.x$ and $\lambda y.y$ as separate values of type $\tau \rightarrow \tau$ and then provide a separate equivalence relation ($\alpha$-equivalence) to express the fact that these two values are indistinguishable.
+The treatment in our setting is similar: for iteration groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate value with the understanding that some iterates are $\idisotwo$-equivalent.
+Similarly for symmetry groupoids $\iorder{c}$, every morphism is listed as a value $1/c^k$ with the same understanding that that some of these morphisms are $\idisotwo$-equivalent.
 

### 2016-12-16 typespp.tex
sec 4

@@
 The idea is not uncommon: in the conventional $\lambda$-calculus, we list $\lambda x.x$ and $\lambda y.y$ as separate values of type $\tau \rightarrow \tau$ and then provide a separate equivalence relation ($\alpha$-equivalence) to express the fact that these two values are indistinguishable.
-The treatment in our setting is similar: for iteration groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate value with the understanding that some iterates are $\idisotwo$-equivalent.
-Similarly for symmetry groupoids $\iorder{c}$, every morphism is listed as a value $1/c^k$ with the same understanding that that some of these morphisms are $\idisotwo$-equivalent.
+The treatment in our setting is similar: for iteration groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate value with the understanding that some iterates are $\isotwo$-equivalent.
+Similarly for symmetry groupoids $\iorder{c}$, every morphism is listed as a value $1/c^k$ with the same understanding that that some of these morphisms are $\isotwo$-equivalent.
 

### 2016-12-16 typespp.tex
structure of paper

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Looping and Delooping}
+
+add here
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{$\Pi^/$: Operational Semantics}
+
+Only non-trivial pairs are created by eta; they create ``synchronized'' uses combinators $c^k$ and $1/c^k$.
+The rest of the combinators can only replace these combinators by $\isotwo$-equivalent ones.
+Assoc is subtle: what if we have $\langle \order{c_1} , \langle \iorder{c_2} , \order{c_2} \rangle \rangle$; naive assoc would allow us to create a pair with unrelated combinators which cannot be eliminated using epsilon.
+But if assoc is constrained to maintain the invariant that only ``synchronized'' uses of combinators are ever paired together, everything works great.
 
@@
 
+ccc have a notion of higher-order functions, trace, etc. Talk about these examples.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-12-16 typespp.tex
looping and delooping subsection

@@
 
-add here
+Our constructions of $\order{p}$ and $\iorder{p}$ are instances of a general pattern that we have instantiated to the setting of $\Pi$ and adapted to explicitly refer to weak (i.e., up to $\isotwo$ in our case) equivalences.
+In the general construction~\cite{looping}, a group $G$ is known to induce two groupoids, a looping groupoid $LG$ and a delooping groupoid $BG$.
+These are defined as follows.
+
+The looping groupoid $LG$ has the group elements as objects and, for every group elements~$g$ and $h$, there is a morphism between $g$ and $h^-1 \cdot g \cdot h$.
+If we fix a type $\tau$ and a particular 1-combinator $p : \tau \iso \tau$, we get a group whose elements are the iterates $p^k$.
+The corresponding looping groupoid would have all the iterates as objects and morphisms between $p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that $p^{k_2}$ is equivalent to $p^{-j} \odot p^{k_1} \odot p^{j}$.
+By Lem.~\ref{lem:distiterplus}, the latter term is $\isotwo$-equivalent to $p^{-j+k_1+j}$ which is $\isotwo$-equivalent to $p{k_1}$.
+In other words, we have a morphism between $p^{k_1}$ and $p^{k_2}$ whenever they are $\isotwo$-equivalent which is consistent with our definition of iteration groupoids.
+The delooping groupoid $BG$ has one trivial element and for each group element $g$ a morphism from that element to itself.
+Modulo $\isotwo$-equivalences, this is consistent with our definition of symmetry groupoids.
 

### 2016-12-16 typespp.tex
small changes

@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Products of Iteration and Symmetry Groupoids}
-
-If we are to build a compositional programming language around iteration and
-symmetry groupoids, we need to ensure that they compose sensibly with the
-existing type formers. Groupoids, viewed as categories, come associated with
-natural notions of sums and products, and one might expect or hope that
-identities which hold for rational numbers lift to identities in our
-situation. The most prominent and important identity we seek is:
-such as:
-\[
-\order{p} \otimes \iorder{p} \simeq \mathbb{1}
-\]
-The groupoids on either side are not categorically equivalent but they do have the same cardinality.
-Thus if we focus on some notion of cardinality-preserving morphisms, it is sensible to consider them equivalent.
-The situation is however subtle and does not directly support a notion of sums.
-It is studied in detail in the next section.
-
-%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Looping and Delooping}
@@
 Modulo $\isotwo$-equivalences, this is consistent with our definition of symmetry groupoids.
+
+%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Products of Iteration and Symmetry Groupoids}
+
+If we are to build a compositional programming language around iteration and
+symmetry groupoids, we need to ensure that they compose sensibly with the
+existing type formers. Groupoids, viewed as categories, come associated with
+natural notions of sums and products, and one might expect or hope that
+identities which hold for rational numbers lift to identities in our
+situation. The most prominent and important identity we seek is:
+\[
+\order{p} \otimes \iorder{p} \simeq \mathbb{1}
+\]
+for some appropriate unit object $\mathbb{1}$.
+The groupoids on either side are not categorically equivalent but they do have the same cardinality.
+Furthermore the iteration groupoid and the symmetry groupoid naturally ``cancel out'' in the sense that the iteration groupoid introduces the iterates of a combinator $c$ as separate objects while the symmetry groupoid uses these same iterates to express internal symmetries.
+Thus if we focus on the right notion of morphisms, it is sensible to consider them equivalent.
+The situation is however subtle and does not directly support a notion of sums.
+It is studied in detail in the next section.
 

### 2016-12-16 typespp.tex
small things

@@
 The combinators that relate these fractional types are in Fig.~\ref{pifrac:comb}.
-The bottom group includes identity and sequential composition $\circledcirc$ combinators and ensures that the combinators can be applied anywhere inside a product $\circledast$.
+The bottom group includes identity and sequential composition~$\circledcirc$ combinators and ensures that the combinators can be applied anywhere inside a product $\circledast$.
 There is also a combinator that relates two types $\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo c_2$.

### 2016-12-17 typespp.tex
make fractional types pointed and use point to keep track of dependency
plan for end of sec 5 and sec 6

@@
 
+at end of previous section; add a new subsection explaining functor showing
+equivalence of 1 and #p x 1/#p. Morita functor. Pullback condition. there is a
+choice for eta but there is an extra condition that the choices are all
+consistent. The big question is how to encode this global condition; side
+effect; synchronization; etc. so naturally lead to pointed types and revise types
+
 Only non-trivial pairs are created by eta; they create ``synchronized'' uses combinators $c^k$ and $1/c^k$.
@@
 But if assoc is constrained to maintain the invariant that only ``synchronized'' uses of combinators are ever paired together, everything works great.
+
+Idea is to change the types to be of the form $\Sigma_{\overline{k}}\mathbb{T}$
+where we keep track of the values (pointed types). When we take a product we
+concatenate the two sequences of integers.
+\[
+(\textrm{Fractional Types})  \mathbb{T} ::=
+  \Sigma_k\order{c} \alt \Sigma_k\iorder{c} \alt
+  \Sigma_{\overline{j},\overline{k}}\mathbb{T}_1 \circledast \mathbb{T}_2
+\]
 

### 2016-12-17 typespp.tex
fix compile error
a little on sec 3.7 but not done

@@
 classes, each with two non-equivalent automorphisms, and hence the cardinality
-of $G_3$ is $\frac{1}{2} + \frac{1}{2} + \frac{1}{2} = \frac{3}{2}$.
-It is important to note that $G_1$ and $G_2$ are (categorically) equivalent
-groupoids, but that $G_3$ is not equivalent to either $G_1$ or $G_2$.
-Roughly speaking this is because the number of connected components is also
-an invariant of a groupoid, and here $G_1$ and $G_2$ have $2$ whilst $G_3$ has
-$3$.
+of $G_3$ is $\frac{1}{2} + \frac{1}{2} + \frac{1}{2} = \frac{3}{2}$.  It is
+important to note that $G_1$ and $G_2$ are (categorically) equivalent groupoids,
+but that $G_3$ is not equivalent to either $G_1$ or $G_2$.  Roughly speaking
+this is because the number of connected components is also a categorical
+invariant of a groupoid, and here $G_1$ and $G_2$ have $2$ whilst $G_3$ has $3$.
 
@@
 
-Given a 1-combinator $p : \tau\iso\tau$, we define the groupoid $\order{p}$ as follows.
-The objects are the elements of $\iter{p}$, i.e., the triples $\triple{k}{q}{\alpha}$ indexed by integers $k$, 1-combinators $q : \tau\iso\tau$, and 2-combinators $\alpha : q \isotwo p^k$.
-We then add (reversible) morphisms between any iterates related by 2-combinators; categorically, this will make any such objects equivalent.
-If $p$ has order $o$, Lem.~\ref{lem:ordercancel} gives us a 2-combinator $\alpha$ which witnesses that $p^i \isotwo p^{i+o}$.
-Thus given two iterates $\triple{i}{q_i}{\alpha_i}$ and $\triple{i+o}{q_j}{\alpha_j}$, they must be equivalent since $\alpha_i~\bullet~\alpha~\bullet~!\,\alpha_j$ shows that $q_i \isotwo q_j$.
-In other words $p^j$ will be equivalent to $p^k$ exactly when $j$ and $k$ differ by $o$.
-This informal description formalizes straightforwardly.
+Given a 1-combinator $p : \tau\iso\tau$, we define the groupoid $\order{p}$ as
+follows. The objects are the elements of $\iter{p}$, i.e., the triples
+$\triple{k}{q}{\alpha}$ indexed by integers $k$, 1-combinators
+$q : \tau\iso\tau$, and 2-combinators $\alpha : q \isotwo p^k$. We then add
+(reversible) morphisms between any iterates related by 2-combinators;
+categorically, this will make any such objects equivalent.  If $p$ has order
+$o$, Lem.~\ref{lem:ordercancel} gives us a 2-combinator $\alpha$ which witnesses
+that $p^i \isotwo p^{i+o}$.  Thus given two iterates $\triple{i}{q_i}{\alpha_i}$
+and $\triple{i+o}{q_j}{\alpha_j}$, they must be equivalent since
+$\alpha_i~\bullet~\alpha~\bullet~!\,\alpha_j$ shows that $q_i \isotwo q_j$.  In
+other words $p^j$ will be equivalent to $p^k$ exactly when $j$ and $k$ differ by
+$o$.  This informal description formalizes straightforwardly.
 
@@
 \]
-where $\distiterplus{p}{k_1}{k_2}$ is defined in Lem.~\ref{lem:distiterplus}.
-The common categorical representation of a group is a category with one trivial object and the group elements as morphisms on that trivial object.
-Our construction of the groupoid $\iorder{p}$ is essentially the same.
+where $\distiterplus{p}{k_1}{k_2}$ is defined in
+Lem.~\ref{lem:distiterplus}. The common categorical representation of a group is
+a category with one trivial object and the group elements as morphisms on that
+trivial object. Our construction of the groupoid $\iorder{p}$ is essentially the
+same.
 
@@
 \[
-\order{p} \otimes \iorder{p} \simeq \mathbb{1}
+\order{p} \otimes \iorder{p} \simeq \order{\idiso}
 \]
-for some appropriate unit object $\mathbb{1}$.
-The groupoids on either side are not categorically equivalent but they do have the same cardinality.
-Furthermore the iteration groupoid and the symmetry groupoid naturally ``cancel out'' in the sense that the iteration groupoid introduces the iterates of a combinator $c$ as separate objects while the symmetry groupoid uses these same iterates to express internal symmetries.
-Thus if we focus on the right notion of morphisms, it is sensible to consider them equivalent.
-The situation is however subtle and does not directly support a notion of sums.
-It is studied in detail in the next section.
+The left hand side is the categorical product of the iteration groupoid and
+symmetry groupoid for some arbitrary 1-combinator $p$. The cardinality of this
+groupoid is 1. The right hand side is the iteration groupoid for the identity
+1-combinator which also has cardinality 1. The groupoids on either side are not
+categorically equivalent and hence $\simeq$ cannot be categorical equivalence.
+There are however other notions of groupoid equivalence such as Morita
+equivalence~\cite{looping} that identifies the two groupoids above.  Although
+the notion of equivalence we use in the next section appears novel, it is
+related ``in spirit'' to Morita equivalence and we find it useful to give a
+small motivating example instantitating $p$ above to $\permfive$ (which has
+order 3). Essentially we are verifying that the product of the groupoid
+$\order{\permfive}$ of cardinality 3 and the groupoid $\iorder{\permfive}$ of
+cardinality $\frac{1}{3}$ is the trivial groupoid with only one distinct object
+with one trivial automorphism.
 
@@
 at end of previous section; add a new subsection explaining functor showing
-equivalence of 1 and #p x 1/#p. Morita functor. Pullback condition. there is a
+equivalence of 1 and $\order{p} \times \iorder{p}$. Morita functor. Pullback condition. there is a
 choice for eta but there is an extra condition that the choices are all

### 2016-12-17 typespp.tex
3.7

@@
 \medskip\noindent Baez and Dolan~\cite{groupoidcard} assign to each groupoid a
-\emph{cardinality} that counts the objects up to equivalences. The
-groupoid on the left has six points \texttt{a}, \texttt{b}, \texttt{c},
-\texttt{d}, \texttt{e}, and \texttt{f} with two groups of three
-points each clustered in an equivalence class and hence the groupoid has
-cardinality~2. The (2-)groupoid on the right has one point $\ast$ with
-four equivalences \texttt{id}, $p$, $q$, and~$q'$ on it. The
-equivalences $q$ and $q'$ are however identified by~$\alpha$ leaving
-only three distinct equivalence classes and hence making the
-cardinality $\frac{1}{3}$.
+\emph{cardinality} that counts the objects up to equivalences. The groupoid on
+the left has six points \texttt{a}, \texttt{b}, \texttt{c}, \texttt{d},
+\texttt{e}, and \texttt{f} with two groups of three points each clustered in an
+equivalence class and hence the groupoid has cardinality~2. The (2-)groupoid on
+the right has one point $\ast$ with four equivalences \texttt{id}, $p$, $q$,
+and~$q'$ on it. The equivalences $q$ and $q'$ are however identified by~$\alpha$
+leaving only three distinct isomorphism classes and hence making the cardinality
+$\frac{1}{3}$.
 
@@
 \[
-\order{p} \otimes \iorder{p} \simeq \order{\idiso}
+\order{p} \circledast \iorder{p} \simeq \order{\idiso}
 \]
@@
 groupoid is 1. The right hand side is the iteration groupoid for the identity
-1-combinator which also has cardinality 1. The groupoids on either side are not
-categorically equivalent and hence $\simeq$ cannot be categorical equivalence.
-There are however other notions of groupoid equivalence such as Morita
-equivalence~\cite{looping} that identifies the two groupoids above.  Although
-the notion of equivalence we use in the next section appears novel, it is
-related ``in spirit'' to Morita equivalence and we find it useful to give a
-small motivating example instantitating $p$ above to $\permfive$ (which has
-order 3). Essentially we are verifying that the product of the groupoid
+1-combinator which also has cardinality 1. The groupoids on either side are
+however not categorically equivalent and hence $\simeq$ cannot be categorical
+equivalence.  There are however other notions of groupoid equivalence, such as
+Morita equivalence~\cite{looping}, that identify the two groupoids above.
+Although the notion of equivalence we use in the next section appears novel, it
+is related ``in spirit'' to Morita equivalence and we find it useful to give a
+small example. We will verify that the product of the groupoid
 $\order{\permfive}$ of cardinality 3 and the groupoid $\iorder{\permfive}$ of
 cardinality $\frac{1}{3}$ is the trivial groupoid with only one distinct object
-with one trivial automorphism.
+and one trivial automorphism. To avoid clutter we assume the groupoids are
+strict, e.g., that the groupoid $\order{\permfive}$ has three elements (instead
+of 3 isomorphism classes of elements). The definition of Morita equivalence in
+this case reduces to to finding a Morita morphism from $\order{\idiso}$ to
+$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor from the
+first groupoid to the next with an additional condition that it is a
+\emph{pullback}. The intuition is that the codomain consists of three objects
+with associated morphisms, each of them counting as a ``third.'' The functor
+from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
+include an arbitrary choice of ``which third'' to target and the pullback
+condition ensures that the choice is ultimately insignificant as it leads to the
+same result as every other possible functor. To summarize the equivalence
+between the two groupoids requires a \emph{local} map that chooses a target and
+a \emph{global} condition ensuring that the choice ultimately annhilates. The
+challenge we are facing is to turn this idea into an operational semantics that
+somehow combines the local and global aspects of the equivalence.
 

### 2016-12-17 typespp.tex
question in email

@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
 \[\begin{array}{rrcll}
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
 \[\begin{array}{c}
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
 {\footnotesize
@@
 \subsection{Products of Iteration and Symmetry Groupoids}
+\label{sub:products}
 
@@
 Each combinator $\rho$ has an inverse.
-\caption{$\Pi^/$ fractional combinators}
+\caption{$\Pi^/$ fractional combinators (non-dependent version)}
 \label{pifrac:comb}
@@
 
-We call the new extension $\Pi^/$. We begin by adding three new syntactic categories to the definition of $\Pi$ in Sec.~\ref{opsempi}:
-
-\begin{definition}[$\Pi^/$]
+%%%%%
+\subsection{Non-Dependent Version}
+
+As a first approximation to the language we seek, we consider adding three new
+syntactic categories to the definition of $\Pi$ in Sec.~\ref{opsempi}:
+
+\begin{definition}[Non-dependent $\Pi^/$]
 \label{def:pifrac}
-The syntax of $\Pi^/$ is given by the following categories:
+The syntax of (non-dependent) $\Pi^/$ is given by the following categories:
 \[\begin{array}{lrcl}
@@
 
-The syntactic category $\ft$ of fractional types introduces type expressions for iteration groupoids, symmetry groupoids, and their products.
-The combinators that relate these fractional types are in Fig.~\ref{pifrac:comb}.
-The bottom group includes identity and sequential composition~$\circledcirc$ combinators and ensures that the combinators can be applied anywhere inside a product $\circledast$.
-There is also a combinator that relates two types $\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo c_2$.
-Indeed if $c_1$ and $c_2$ are considered equivalent via a 2-combinator then we also consider their iteration groupoids to be equivalent.
-The top group ensures that $\circledast$ carries a symmetric monoidal structure with combinators witnessing the unit, commutativity, and associativity properties.
-The last two pairs combinators $\eta-/\epsilon-$ and $\eta+/\epsilon+$ are inspired from the definition of compact closed categories~\cite{ccc} which are
-symmetric monoidal categories in which every object has a dual.
-They are the ones that witness the ``fractional'' nature of symmetry groupoids: they are indeed the multiplicative inverses of iteration groupoids.
-
-The last new syntactic category in $\Pi^/$ is that of values which deserves some background discussion.
-When the types denote sets, it is evident what it means to have a value of a given type: it is just an element of the set.
-When types denote groupoids, it is less clear what it means to have a value, especially when the total number of values, as reported by the groupoid cardinality, is a proper fraction.
-In that case we the values up to an equivalence relation.
-The idea is not uncommon: in the conventional $\lambda$-calculus, we list $\lambda x.x$ and $\lambda y.y$ as separate values of type $\tau \rightarrow \tau$ and then provide a separate equivalence relation ($\alpha$-equivalence) to express the fact that these two values are indistinguishable.
-The treatment in our setting is similar: for iteration groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate value with the understanding that some iterates are $\isotwo$-equivalent.
-Similarly for symmetry groupoids $\iorder{c}$, every morphism is listed as a value $1/c^k$ with the same understanding that that some of these morphisms are $\isotwo$-equivalent.
+The syntactic category $\ft$ of fractional types introduces type expressions for
+iteration groupoids, symmetry groupoids, and their products. The combinators
+that relate these fractional types are in Fig.~\ref{pifrac:comb}. The bottom
+group includes identity and sequential composition~$\circledcirc$ combinators
+and ensures that the combinators can be applied anywhere inside a product
+$\circledast$. There is also a combinator that relates two types $\order{c_1}$
+and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo
+c_2$.
+Indeed if $c_1$ and $c_2$ are considered equivalent via a 2-combinator then we
+also consider their iteration groupoids to be equivalent. The top group ensures
+that $\circledast$ carries a symmetric monoidal structure with combinators
+witnessing the unit, commutativity, and associativity properties. The last two
+pairs combinators $\eta-/\epsilon-$ and $\eta+/\epsilon+$ are inspired from the
+definition of compact closed categories~\cite{ccc} which are symmetric monoidal
+categories in which every object has a dual. They are the ones that witness the
+``fractional'' nature of symmetry groupoids as motivated in Sec.~\ref{sub:products}.
+
+The last new syntactic category in $\Pi^/$ is that of values which deserves some
+background discussion. When the types denote sets, it is evident what it means
+to have a value of a given type: it is just an element of the set. When types
+denote groupoids, it is less clear what it means to have a value, especially
+when the total number of values, as reported by the groupoid cardinality, is a
+proper fraction. In that case we the values up to an equivalence relation. The
+idea is not uncommon: in the conventional $\lambda$-calculus, we list
+$\lambda x.x$ and $\lambda y.y$ as separate values of type
+$\tau \rightarrow \tau$ and then provide a separate equivalence relation
+($\alpha$-equivalence) to express the fact that these two values are
+indistinguishable. The treatment in our setting is similar: for iteration
+groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate
+value with the understanding that some iterates are
+$\isotwo$-equivalent. Similarly for symmetry groupoids $\iorder{c}$, every
+morphism is listed as a value $1/c^k$ with the same understanding that that some
+of these morphisms are $\isotwo$-equivalent.
+
+%%%%%
+\subsection{Pointed Types}
+
+To understand the challenge in designing an operational semantics for $\eta$ and
+$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
+Fig.~\ref{fig:zigzag}. Let the combinator $c$ to be of order greater than or
+equal to 2, i.e., the type $\order{c}$ has at least two distinct values $c^i$
+and $c^j$ and the type $\iorder{c}$ has at least two distinct values $1/c^i$ and
+$1/c^j$. The use of $\eta+$ has then at least four choices; as discussed at the
+end of Sec.~\ref{sub:products}, $\eta+$ will internalize some arbitrary but
+fixed choice and might for example generate the pair
+$\langle c^i , 1/c^i \rangle$. No matter what choice is internalized in $\eta+$,
+there is at least one different incoming value $c^j$ of type $\order{c}$. In
+that case, the combinator $\epsilon-$ encounters $\langle 1/c^i , c^j
+\rangle$. If our operational semantics allows $\epsilon-$ to annihilate
+arbitrary pairs, the net effect of the circuit would be to map $c^j$ to $c^i$
+for arbitrary $i$ and $j$.
+
+\begin{figure}[bht]
+\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+  \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+  \node at (0.5,1) {\footnotesize{$\unitetl/$}};
+  \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (0,1);
+  \path (1,1.75) edge node[above] {\footnotesize{$\order{\idiso}$}} (1.6,1.75);
+  \path [highlight] (1,0.25) edge node[above] {} (4,0.25);
+  \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,0.25);
+  \draw (1.6,1) -- (2.6,1) -- (2.6,2.8) -- (1.6,2.8) -- cycle;
+  \node at (2.1,1.9) {\footnotesize{$\eta+$}};
+  \draw[dashed] (2.8,2.6) -- (2.8,1.25);
+  \path [highlight] (2.6,2.6) edge node[above] {} (4,2.6);
+  \path (2.6,2.6) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (4,2.6);
+  \path [highlight] (2.6,1.25) edge node[above] {} (4,1.25);
+  \path (2.6,1.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.25);
+  \draw (4,0) -- (5.5,0) -- (5.5,3) -- (4,3) -- cycle;
+  \node at (4.75,1.5) {$\footnotesize{\assocrt/}$};
+  \path [highlight] (5.5,1.25) edge node[above] {} (7,1.25);
+  \path (5.5,1.25) edge node[above] {$\footnotesize{1/\#c}$} (7,1.25);
+  \path [highlight] (5.5,2.6) edge node[above] {} (9,2.6);
+  \path (5.5,2.6) edge node[above] {\footnotesize{$\#c$}} (9,2.6);
+  \draw (7,0) -- (8,0) -- (8,1.6) -- (7,1.6) -- cycle;
+  \node at (7.5,0.8) {\footnotesize{$\epsilon-$}};
+  \draw[dashed] (6.8,0.2) -- (6.8,1.25);
+  \path (8,1) edge node[above] {\footnotesize{$\order{\idiso}$}} (9,1);
+  \path [highlight] (5.5,0.2) edge node[above] {} (7,0.2);
+  \path (5.5,0.2) edge node[above] {\footnotesize{$\#c$}} (7,0.2);
+  \draw (9,0.8) -- (10,0.8) -- (10,2.8) -- (9,2.8) -- cycle;
+  \node at (9.5,1.8) {\footnotesize{$\unitetr/$}};
+  \path (10,1.8) edge node [above] {\footnotesize{$\#c$}} (11,1.8);
+\end{tikzpicture}
+\caption{\label{fig:zigzag} Zig-zag}
+\end{figure}
 
@@
 This is where the ``action at a distance'' occurs.
-
-We implement an example zig-zag, which use synchr⋆
-to synchronize the \emph{tangled} values created by $\eta$ and $\epsilon$.
-The circuit of zig-zag is shown in figure \ref{fig:zigzag} which
-shows how the input of type $\#c$ is swapped by synchr⋆.
-
-\begin{figure}[bht]
-\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.9},highlight/.style={line width=4pt,orange!70}]
-  \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
-  \node at (0.5,1) {$\textsf{unit}_\times$};
-  \path (-1.1,1) edge node[above] {${\textsf{\#c}}$} (0,1);
-  \path (1,1.75) edge node[above] {$\mathbb{1}$} (1.6,1.75);
-  \path [highlight] (1,0.25) edge node[above] {} (4,0.25);
-  \path (1,0.25) edge node[above] {${\textsf{\#c}}$} (4,0.25);
-  \draw (1.6,1) -- (2.6,1) -- (2.6,2.8) -- (1.6,2.8) -- cycle;
-  \node at (2.1,1.9) {$\eta_{\textsf{c}}$};
-  \draw[dashed] (2.8,2.6) -- (2.8,1.25);
-  \path [highlight] (2.6,2.6) edge node[above] {} (4,2.6);
-  \path (2.6,2.6) edge node[above] {${\textsf{\#c}}$} (4,2.6);
-  \path [highlight] (2.6,1.25) edge node[above] {} (4,1.25);
-  \path (2.6,1.25) edge node[above] {$1/\textsf{\#c}$} (4,1.25);
-  \draw (4,0) -- (5.5,0) -- (5.5,3) -- (4,3) -- cycle;
-  \node at (4.75,1.5) {$synch_{\textsf{c}}$};
-  \path [highlight] (5.5,1.25) edge node[above] {} (7,1.25);
-  \path (5.5,1.25) edge node[above] {$1/\#c$} (7,1.25);
-  \path [highlight] (5.5,2.6) edge node[above] {} (9,2.6);
-  \path (5.5,2.6) edge node[above] {$\#c$} (9,2.6);
-  \draw (7,0) -- (8,0) -- (8,1.6) -- (7,1.6) -- cycle;
-  \node at (7.5,0.8) {$\epsilon_c$};
-  \draw[dashed] (6.8,0.2) -- (6.8,1.25);
-  \path (8,1) edge node[above] {$\mathbb{1}$} (9,1);
-  \path [highlight] (5.5,0.2) edge node[above] {} (7,0.2);
-  \path (5.5,0.2) edge node[above] {$\#c$} (7,0.2);
-  \draw (9,0.8) -- (10,0.8) -- (10,2.8) -- (9,2.8) -- cycle;
-  \node at (9.5,1.8) {$\textsf{unit}_\times$};
-  \path (10,1.8) edge node [above] {$\#c$} (11,1.8);
-\end{tikzpicture}
-\caption{\label{fig:zigzag} Zig-zag}
-\end{figure}
 

### 2016-12-17 typespp.tex
Add a diagram and few lines about morita equivalence

@@
 \usepackage[section]{placeins}
+\usepackage{braket}
 
@@
 strict, e.g., that the groupoid $\order{\permfive}$ has three elements (instead
-of 3 isomorphism classes of elements). The definition of Morita equivalence in
-this case reduces to to finding a Morita morphism from $\order{\idiso}$ to
-$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor from the
-first groupoid to the next with an additional condition that it is a
-\emph{pullback}. The intuition is that the codomain consists of three objects
-with associated morphisms, each of them counting as a ``third.'' The functor
-from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
-include an arbitrary choice of ``which third'' to target and the pullback
-condition ensures that the choice is ultimately insignificant as it leads to the
+of 3 isomorphism classes of elements). The definition of \emph{Morita
+  equivalence} in this case reduces to to finding a \emph{Morita morphism} from
+$\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$. This is
+a functor $\psi$ from the unit groupoid to the product groupoid, say $G$, with
+an additional condition that the commutative diagram in figure~\ref{fig:morita}
+is a \emph{pullback}. The intuition is that the codomain consists of three
+objects with associated morphisms, each of them counting as a ``third.'' The
+functor from $\order{\idiso}$ to
+$\order{\permfive} \circledast \iorder{\permfive}$ must include an arbitrary
+choice of ``which third'' to target and the universality condition of the
+pullback ensures that the choice is ultimately insignificant as it leads to the
 same result as every other possible functor. To summarize the equivalence
@@
 somehow combines the local and global aspects of the equivalence.
+
+\begin{figure}[h]
+  \centering
+  \begin{tikzcd}
+    Q
+    \arrow[drr, bend left, "{\xi_1}"]
+    \arrow[ddr, bend right, "{(s,t)}"]
+    \arrow[dr, dotted] & & \\
+    & \set{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
+    & G_1 \arrow[d, "{(s,t)}"] \\
+    & \set{\ast} \times \set{\ast} \arrow[r, "{\psi_0 \times \psi_0}"]
+    & G_0 \times G_0
+  \end{tikzcd}
+  \caption{
+    \label{fig:morita}
+    Pullback diagram for a Morita morphism,
+    $\order{\idiso} \rightarrow \order{\permfive} \circledast
+    \iorder{\permfive}$
+  }
+\end{figure}
 

### 2016-12-17 typespp.tex
Add citation for morita equivalence

@@
 equivalence.  There are however other notions of groupoid equivalence, such as
-Morita equivalence~\cite{looping}, that identify the two groupoids above.
-Although the notion of equivalence we use in the next section appears novel, it
-is related ``in spirit'' to Morita equivalence and we find it useful to give a
-small example. We will verify that the product of the groupoid
-$\order{\permfive}$ of cardinality 3 and the groupoid $\iorder{\permfive}$ of
-cardinality $\frac{1}{3}$ is the trivial groupoid with only one distinct object
-and one trivial automorphism. To avoid clutter we assume the groupoids are
-strict, e.g., that the groupoid $\order{\permfive}$ has three elements (instead
-of 3 isomorphism classes of elements). The definition of \emph{Morita
-  equivalence} in this case reduces to to finding a \emph{Morita morphism} from
-$\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$. This is
-a functor $\psi$ from the unit groupoid to the product groupoid, say $G$, with
-an additional condition that the commutative diagram in figure~\ref{fig:morita}
-is a \emph{pullback}. The intuition is that the codomain consists of three
-objects with associated morphisms, each of them counting as a ``third.'' The
-functor from $\order{\idiso}$ to
-$\order{\permfive} \circledast \iorder{\permfive}$ must include an arbitrary
-choice of ``which third'' to target and the universality condition of the
-pullback ensures that the choice is ultimately insignificant as it leads to the
-same result as every other possible functor. To summarize the equivalence
-between the two groupoids requires a \emph{local} map that chooses a target and
-a \emph{global} condition ensuring that the choice ultimately annhilates. The
-challenge we are facing is to turn this idea into an operational semantics that
-somehow combines the local and global aspects of the equivalence.
+Morita equivalence\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that
+identify the two groupoids above.  Although the notion of equivalence we use in
+the next section appears novel, it is related ``in spirit'' to Morita
+equivalence and we find it useful to give a small example. We will verify that
+the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
+groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is the trivial
+groupoid with only one distinct object and one trivial automorphism. To avoid
+clutter we assume the groupoids are strict, e.g., that the groupoid
+$\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
+elements). The definition of \emph{Morita equivalence} in this case reduces to
+to finding a \emph{Morita morphism} from $\order{\idiso}$ to
+$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
+from the unit groupoid to the product groupoid, say $G$, with an additional
+condition that the commutative diagram in figure~\ref{fig:morita} is a
+\emph{pullback}. The intuition is that the codomain consists of three objects
+with associated morphisms, each of them counting as a ``third.'' The functor
+from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
+include an arbitrary choice of ``which third'' to target and the universality
+condition of the pullback ensures that the choice is ultimately insignificant as
+it leads to the same result as every other possible functor. To summarize the
+equivalence between the two groupoids requires a \emph{local} map that chooses a
+target and a \emph{global} condition ensuring that the choice ultimately
+annhilates. The challenge we are facing is to turn this idea into an operational
+semantics that somehow combines the local and global aspects of the equivalence.
 

### 2016-12-18 typespp.tex
op sem

--- 2016-12-17 typespp.tex
@@
 
-\begin{definition}[Groupoid Cardinality~\cite{2009arXiv0908.4305B}]
+\begin{definition}[Groupoid cardinality~\cite{2009arXiv0908.4305B}]
   The cardinality of a groupoid $G$ is the (positive) real number:
@@
 a functor $\psi$ from the unit groupoid to the product groupoid, say $G$, with
-an additional condition that the commutative diagram in figure~\ref{fig:morita}
-is a \emph{pullback}. The intuition is that the codomain consists of three
-objects with associated morphisms, each of them counting as a ``third.'' The
-functor from $\order{\idiso}$ to
-$\order{\permfive} \circledast \iorder{\permfive}$ must include an arbitrary
-choice of ``which third'' to target and the universality condition of the
-pullback ensures that the choice is ultimately insignificant as it leads to the
-same result as every other possible functor. To summarize the equivalence
-between the two groupoids requires a \emph{local} map that chooses a target and
-a \emph{global} condition ensuring that the choice ultimately annhilates. The
-challenge we are facing is to turn this idea into an operational semantics that
-somehow combines the local and global aspects of the equivalence.
-
-\begin{figure}[h]
+an additional condition that the commutative diagram in Fig.~\ref{fig:morita} is
+a \emph{pullback}. The intuition is that the codomain consists of three objects
+with associated morphisms, each of them counting as a ``third.'' The functor
+from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
+include an arbitrary choice of ``which third'' to target and the universality
+condition of the pullback ensures that the choice is ultimately insignificant as
+it leads to the same result as every other possible functor. To summarize the
+equivalence between the two groupoids requires a \emph{local} map that chooses a
+target and a \emph{global} condition ensuring that the choice ultimately
+annihilates. The challenge we are facing is to turn this idea into an operational
+semantics that somehow combines the local and global aspects of the equivalence.
+
+\begin{figure}[t]
   \centering
@@
 
-%%%%%
-\subsection{Pointed Types}
-
-To understand the challenge in designing an operational semantics for $\eta$ and
-$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
-Fig.~\ref{fig:zigzag}. Let the combinator $c$ to be of order greater than or
-equal to 2, i.e., the type $\order{c}$ has at least two distinct values $c^i$
-and $c^j$ and the type $\iorder{c}$ has at least two distinct values $1/c^i$ and
-$1/c^j$. The use of $\eta+$ has then at least four choices; as discussed at the
-end of Sec.~\ref{sub:products}, $\eta+$ will internalize some arbitrary but
-fixed choice and might for example generate the pair
-$\langle c^i , 1/c^i \rangle$. No matter what choice is internalized in $\eta+$,
-there is at least one different incoming value $c^j$ of type $\order{c}$. In
-that case, the combinator $\epsilon-$ encounters $\langle 1/c^i , c^j
-\rangle$. If our operational semantics allows $\epsilon-$ to annihilate
-arbitrary pairs, the net effect of the circuit would be to map $c^j$ to $c^i$
-for arbitrary $i$ and $j$.
-
-\begin{figure}[bht]
+\begin{figure}[t]
 \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
@@
 
+\begin{figure}[t]
+\[\begin{array}{rrcll}
+\unitetl/ :&  \order{\idiso} \circledast \mathbb{T} & \leftrightspoon & \mathbb{T} &: \unititl/ \\
+\unitetr/ :&  \mathbb{T} \circledast \order{\idiso} & \leftrightspoon & \mathbb{T} &: \unititr/ \\
+\swapt/ :&  \mathbb{T}_1 \circledast \mathbb{T}_2 & \leftrightspoon & \mathbb{T}_2 \circledast \mathbb{T}_1 &: \swapt/ \\
+\assoclt/ :&  \mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3) & \leftrightspoon & (\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3
+  &: \assocrt/ \\
+\\
+\eta- :& \order{\idiso} & \leftrightspoon & \iorder{c} \circledast \order{c} &: \epsilon- \\
+\eta+ :& \order{\idiso} & \leftrightspoon & \order{c} \circledast \iorder{c} &: \epsilon+ \\
+\end{array}\]
+
+\[\begin{array}{c}
+\Rule{}
+{}
+{\jdg{}{}{\idiso/ : \mathbb{T} \leftrightspoon \mathbb{T}}}
+{}
+\qquad
+\Rule{}
+{\jdg{}{}{\rho_1 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2} \quad \rho_2 : \mathbb{T}_2 \leftrightspoon \mathbb{T}_3}
+{\jdg{}{}{\rho_1 \circledcirc \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_3}}
+{}
+\\
+\\
+\Rule{}
+{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
+{\jdg{}{}{\order{\alpha} : \order{c_1}  \leftrightspoon \order{c_2}}}
+{}
+\qquad
+\Rule{}
+{\jdg{}{}{\rho_1 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2} \quad \rho_2 : \mathbb{T}_3 \leftrightspoon \mathbb{T}_4}
+{\jdg{}{}{\rho_1 \circledast \rho_2 : \mathbb{T}_1 \circledast \mathbb{T}_3 \leftrightspoon \mathbb{T}_2 \circledast \mathbb{T}_4}}
+{}
+\end{array}\]
+
+Each combinator $\rho$ has an inverse.
+\caption{$\Pi^/$ fractional combinators with pointed types (dependent version)}
+\label{pifracd:comb}
+\end{figure}
+
+%%%%%
+\subsection{Pointed Types}
+
+To understand the challenge in designing an operational semantics for $\eta$ and
+$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
+Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
+manipulation on types:
+\[
+x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
+\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
+\]
+and by the coherence conditions of compact closed categories, we expect this
+circuit, when run in either direction, to be equivalent to the
+identity. Consider now the case of a combinator $c$ whose order is greater than
+or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
+$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
+$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
+$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
+either end. The problem becomes apparent when one observes that the uses of
+$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
+right-to-left execution must behave differently depending on the incoming
+value. But the ``local'' information available to either of these combinators
+does not include the ``global'' information about the incoming value and there
+is no way for them to make consistent guesses locally. It is possible to modify
+the operational semantics to use a mechanism to express a computational effect
+which enables spatially separated parts of the program to communicate in a way
+that is reminiscent of entanglement in quantum mechanics. Possibilities to
+realize this communication are global reference cells, backtracking, logical
+variables with unification. We choose a more abstract and enlightening approach
+that encodes the required dependency in dataflow constraints using dependent
+types.
+
+\begin{definition}[Dependent $\Pi^/$ with pointed types]
+\label{def:pifracd}
+The syntax of dependent $\Pi^/$ with pointed types differs from
+Def.~\ref{def:pifrac} as follows:
+\[\begin{array}{lrcl}
+(\textrm{Pointed Fractional Types}) & \mathbb{T_\bullet} &::=& [\mathbb{T},v] \\
+ (\textrm{Pointed /-combinators}) & \rho_\bullet &:&
+  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2] ~::=
+  [\textit{see Fig.~\ref{pifracd:comb}}]
+\end{array}\]
+\end{definition}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The operational definition of synchr⋆ might seem at odds with our claim that it is related to associativity while it looks like commutativity.
-The way to understand it is as follows: if indeed all values were indepedendent, it would be associativity.
+The way to understand it is as follows: if indeed all values were independent, it would be associativity.
 However, in this case, $p // q$ is really two \emph{tangled} values, which we do not have access to.

### 2016-12-18 typespp.tex

@@
 equivalence.  There are however other notions of groupoid equivalence, such as
-Morita equivalence~\cite{looping}, that identify the two groupoids above.
-Although the notion of equivalence we use in the next section appears novel, it
-is related ``in spirit'' to Morita equivalence and we find it useful to give a
-small example. We will verify that the product of the groupoid
-$\order{\permfive}$ of cardinality 3 and the groupoid $\iorder{\permfive}$ of
-cardinality $\frac{1}{3}$ is the trivial groupoid with only one distinct object
-and one trivial automorphism. To avoid clutter we assume the groupoids are
-strict, e.g., that the groupoid $\order{\permfive}$ has three elements (instead
-of 3 isomorphism classes of elements). The definition of \emph{Morita
-  equivalence} in this case reduces to to finding a \emph{Morita morphism} from
-$\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$. This is
-a functor $\psi$ from the unit groupoid to the product groupoid, say $G$, with
-an additional condition that the commutative diagram in Fig.~\ref{fig:morita} is
-a \emph{pullback}. The intuition is that the codomain consists of three objects
+Morita equivalence\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that
+identify the two groupoids above.  Although the notion of equivalence we use in
+the next section appears novel, it is related ``in spirit'' to Morita
+equivalence and we find it useful to give a small example. We will verify that
+the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
+groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is the trivial
+groupoid with only one distinct object and one trivial automorphism. To avoid
+clutter we assume the groupoids are strict, e.g., that the groupoid
+$\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
+elements). The definition of \emph{Morita equivalence} in this case reduces to
+to finding a \emph{Morita morphism} from $\order{\idiso}$ to
+$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
+from the unit groupoid to the product groupoid, say $G$, with an additional
+condition that the commutative diagram in figure~\ref{fig:morita} is a
+\emph{pullback}. The intuition is that the codomain consists of three objects
 with associated morphisms, each of them counting as a ``third.'' The functor
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
   \centering

### 2016-12-18 typespp.tex
synchronized pointed types

@@
 \\
-\eta- :& \order{\idiso} & \leftrightspoon & \iorder{c} \circledast \order{c} &: \epsilon- \\
+\eta-_k :& [\order{\idiso},\idiso] & \leftrightblackspoon &
+  [\iorder{c},1/c^k] \circledast [\order{c},c^k] &: \epsilon-_k \\
 \eta+ :& \order{\idiso} & \leftrightspoon & \order{c} \circledast \iorder{c} &: \epsilon+ \\
@@
 Each combinator $\rho$ has an inverse.
-\caption{$\Pi^/$ fractional combinators with pointed types (dependent version)}
+\caption{$\Pi^/$ fractional combinators (dependent version)}
 \label{pifracd:comb}
@@
 %%%%%
-\subsection{Pointed Types}
+\subsection{Dependent Pointed Types}
 
@@
 realize this communication are global reference cells, backtracking, logical
-variables with unification. We choose a more abstract and enlightening approach
-that encodes the required dependency in dataflow constraints using dependent
-types.
-
-\begin{definition}[Dependent $\Pi^/$ with pointed types]
+variables with unification. We choose a more abstract approach that encodes the
+required dependency in dataflow constraints using dependent types.
+
+\begin{definition}[Dependent $\Pi^/$]
 \label{def:pifracd}
-The syntax of dependent $\Pi^/$ with pointed types differs from
-Def.~\ref{def:pifrac} as follows:
+The syntax of dependent $\Pi^/$ differs from Def.~\ref{def:pifrac} as follows:
 \[\begin{array}{lrcl}
-(\textrm{Pointed Fractional Types}) & \mathbb{T_\bullet} &::=& [\mathbb{T},v] \\
- (\textrm{Pointed /-combinators}) & \rho_\bullet &:&
-  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2] ~::=
+(\textrm{Fractional Types}) & \mathbb{T_\bullet} &::=&
+  [\order{c},c^k] \alt
+  [\iorder{c},1/c^k] \alt
+  [\mathbb{T}_1 \circledast \mathbb{T}_2,\langle v_1,v_2 \rangle] \alt
+  \exists k.\mathbb{T_\bullet} \\
+ (\textrm{/-combinators}) & \rho_\bullet &:&
+  \mathbb{T}_{\bullet 1} \leftrightblackspoon \mathbb{T}_{\bullet 2} ~::=
   [\textit{see Fig.~\ref{pifracd:comb}}]
@@
 \end{definition}
+
+The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
+where each type $\mathbb{T}$ refers to a particular value $v$ in
+$\mathbb{T}$. We then allow indices of values $k$ to be range over a collection
+of types. In particular, the type
+$\exists k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
+imposes a constraint on the two values associated with the types $\order{c}$ and
+$\iorder{c}$; they must be ``synchronized'' sharing the same $k$.
 

### 2016-12-18 typespp.tex
fix typos in 3.6

@@
 
-The looping groupoid $LG$ has the group elements as objects and, for every group elements~$g$ and $h$, there is a morphism between $g$ and $h^-1 \cdot g \cdot h$.
-If we fix a type $\tau$ and a particular 1-combinator $p : \tau \iso \tau$, we get a group whose elements are the iterates $p^k$.
-The corresponding looping groupoid would have all the iterates as objects and morphisms between $p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that $p^{k_2}$ is equivalent to $p^{-j} \odot p^{k_1} \odot p^{j}$.
-By Lem.~\ref{lem:distiterplus}, the latter term is $\isotwo$-equivalent to $p^{-j+k_1+j}$ which is $\isotwo$-equivalent to $p{k_1}$.
-In other words, we have a morphism between $p^{k_1}$ and $p^{k_2}$ whenever they are $\isotwo$-equivalent which is consistent with our definition of iteration groupoids.
-The delooping groupoid $BG$ has one trivial element and for each group element $g$ a morphism from that element to itself.
-Modulo $\isotwo$-equivalences, this is consistent with our definition of symmetry groupoids.
+The looping groupoid $LG$ has the group elements as objects and, for every
+group elements~$g$ and $h$, there is a morphism between $g$ and $h^{-1} \cdot g
+\cdot h$. If we fix a type $\tau$ and a particular 1-combinator $p : \tau \iso
+\tau$, we get a group whose elements are the iterates $p^k$. The corresponding
+looping groupoid would have all the iterates as objects and morphisms between
+$p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that
+$p^{k_2}$ is equivalent to $p^{-j} \odot p^{k_1} \odot p^{j}$. By
+Lem.~\ref{lem:distiterplus}, the latter term is $\isotwo$-equivalent to
+$p^{-j+k_1+j}$ which is $\isotwo$-equivalent to $p^{k_1}$. In other words, we
+have a morphism between $p^{k_1}$ and $p^{k_2}$ whenever they are
+$\isotwo$-equivalent which is consistent with our definition of iteration
+groupoids. The delooping groupoid $BG$ has one trivial element and for each
+group element $g$ a morphism from that element to itself. Modulo
+$\isotwo$-equivalences, this is consistent with our definition of symmetry
+groupoids.
 

### 2016-12-18 typespp.tex
Some more on looping and delooping

@@
 \newcommand{\distiterplus}[3]{\mathsf{dist}~#1~#2~#3}
+\newcommand{\looping}[1]{\mathcal{L}{#1}}
+\newcommand{\delooping}[1]{\mathcal{B}{#1}}
 
@@
 
-Our constructions of $\order{p}$ and $\iorder{p}$ are instances of a general pattern that we have instantiated to the setting of $\Pi$ and adapted to explicitly refer to weak (i.e., up to $\isotwo$ in our case) equivalences.
-In the general construction~\cite{looping}, a group $G$ is known to induce two groupoids, a looping groupoid $LG$ and a delooping groupoid $BG$.
-These are defined as follows.
-
-The looping groupoid $LG$ has the group elements as objects and, for every
-group elements~$g$ and $h$, there is a morphism between $g$ and $h^{-1} \cdot g
-\cdot h$. If we fix a type $\tau$ and a particular 1-combinator $p : \tau \iso
-\tau$, we get a group whose elements are the iterates $p^k$. The corresponding
-looping groupoid would have all the iterates as objects and morphisms between
-$p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that
+Our constructions of $\order{p}$ and $\iorder{p}$ are instances of a general
+pattern that we have instantiated to the setting of $\Pi$ and adapted to
+explicitly refer to weak (i.e., up to $\isotwo$ in our case) equivalences. In
+the general construction~\cite{looping}, a group $G$ is known to induce two
+groupoids, a looping groupoid $\looping{G}$ and a delooping groupoid
+$\delooping{G}$. These are defined as follows.
+
+The looping groupoid $\looping{G}$ has the group elements as objects and
+morphisms between elements in the same conjugacy class, that is, there is a
+morphism between elements $g$ and $h$ iff there exists a $x \in G$ such that
+$g = x^{-1} \cdot g \cdot x$. Fixing a type $\tau$ and a particular 1-combinator
+$p : \tau \iso \tau$, we get a group whose elements are the iterates $p^k$. The
+corresponding looping groupoid has all the iterates as objects and morphisms
+between $p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that
 $p^{k_2}$ is equivalent to $p^{-j} \odot p^{k_1} \odot p^{j}$. By
@@
 $\isotwo$-equivalent which is consistent with our definition of iteration
-groupoids. The delooping groupoid $BG$ has one trivial element and for each
-group element $g$ a morphism from that element to itself. Modulo
+groupoids.
+
+The delooping groupoid $\delooping{G}$ has one trivial element $\ast$, and for
+each group element $g$ a morphism (loop) from $\ast$ to $\ast$. Modulo
 $\isotwo$-equivalences, this is consistent with our definition of symmetry

### 2016-12-18 typespp.tex
spacing

--- 2016-12-18 typespp.tex
@@
 equivalence.  There are however other notions of groupoid equivalence, such as
-Morita equivalence\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that
+Morita equivalence~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that
 identify the two groupoids above.  Although the notion of equivalence we use in

### 2016-12-18 typespp.tex

@@
 
-Our constructions of $\order{p}$ and $\iorder{p}$ are instances of a general pattern that we have instantiated to the setting of $\Pi$ and adapted to explicitly refer to weak (i.e., up to $\isotwo$ in our case) equivalences.
-In the general construction~\cite{looping}, a group $G$ is known to induce two groupoids, a looping groupoid $LG$ and a delooping groupoid $BG$.
-These are defined as follows.
+Our constructions of $\order{p}$ and $\iorder{p}$ are instances of a general
+pattern that we have instantiated to the setting of $\Pi$ and adapted to
+explicitly refer to weak (i.e., up to $\isotwo$ in our case) equivalences. In
+the general construction~\cite{looping}, a group $G$ is known to induce two
+groupoids, a looping groupoid $\looping{G}$ and a delooping groupoid
+$\delooping{G}$. These are defined as follows.
 
@@
 $\isotwo$-equivalent which is consistent with our definition of iteration
-groupoids. The delooping groupoid $BG$ has one trivial element and for each
-group element $g$ a morphism from that element to itself. Modulo
-$\isotwo$-equivalences, this is consistent with our definition of symmetry
 groupoids.

### 2016-12-18 typespp.tex
opsem
Fix references to opsem

@@
 \end{array}\]
-\caption{\label{opsem}$\Pi$ operational semantics}
+\caption{\label{fig:opsem}$\Pi$ operational semantics}
 }
@@
 direction~$\triangleleft$. We show the complete forward evaluator in
-Fig.~\ref{opsem}; the backward evaluator is easy to infer.
+Fig.~\ref{fig:opsem}; the backward evaluator is easy to infer.
 
@@
 The operational semantics for all the primitive combinators is a
-simple transliteration of Fig.~\ref{opsem}. We omit the
+simple transliteration of Fig.~\ref{fig:opsem}. We omit the
 straightforward implementation of prim and its inverse
@@
 ccc have a notion of higher-order functions, trace, etc. Talk about these examples.
+
+\begin{figure}[t]
+{\footnotesize
+\[\begin{array}{cc}
+\begin{array}[t]{rlcl}
+  \evalone{\unitetl/}{&\langle id^{k},\mathbb{V} \rangle} &=& \mathbb{V} \\
+  \evalone{\unititl/}{& \mathbb{V}} &=& \langle id^{k_1},\mathbb{V} \rangle \\
+  \evalone{\unitetr/}{&\langle \mathbb{V},id^{k} \rangle} &=& \mathbb{V} \\
+  \evalone{\unititr/}{& \mathbb{V}} &=& \langle \mathbb{V},id^{k} \rangle\\
+  \evalone{\swapt}{&\langle \mathbb{V}_1, \mathbb{V}_2 \rangle} &=& \langle \mathbb{V}_2, \mathbb{V}_1 \rangle \\
+  \evalone{\assoclt/}{&\langle c^k , \langle 1/c^{k}, c^{k} \rangle\rangle} &=& \langle \langle c^k , 1/c^{k} \rangle, c^{k} \rangle \\
+  \evalone{\assocrt/}{&\langle \langle c^k , 1/c^{k} \rangle, c^{k} \rangle} &=& \langle c^k , \langle 1/c^{k}, c^{k} \rangle\rangle \\
+\end{array}
+&
+\begin{array}[t]{rlcl}
+  \evalone{\eta-}{& id^{k_1}} &=& \langle 1/c^{k_2},c^{k_2} \rangle\\
+  \evalone{\eta+}{& id^{k_1}} &=& \langle c^{k_2},1/c^{k_2} \rangle\\
+  \evalone{\epsilon-}{& \langle 1/c^{k_1},c^{k_1} \rangle} &=& id^{k_2} \\
+  \evalone{\epsilon+}{& \langle c^{k_1},1/c^{k_1} \rangle} &=& id^{k_2} \\
+\end{array}
+
+\end{array}\]
+\caption{\label{fig:opsem-frac}$\Pi^/$ operational semantics}
+}
+\end{figure}
 

### 2016-12-18 typespp.tex
pointed combinators

@@
 \newcommand{\unitv}{()}
+\newcommand{\synch}{\mathsf{{synch}}}
 \newcommand{\unitepl}{\mathsf{{unite_+l}}}
@@
 \begin{figure}[t]
+{\footnotesize
 \[\begin{array}{rrcll}
-\unitetl/ :&  \order{\idiso} \circledast \mathbb{T} & \leftrightspoon & \mathbb{T} &: \unititl/ \\
-\unitetr/ :&  \mathbb{T} \circledast \order{\idiso} & \leftrightspoon & \mathbb{T} &: \unititr/ \\
-\swapt/ :&  \mathbb{T}_1 \circledast \mathbb{T}_2 & \leftrightspoon & \mathbb{T}_2 \circledast \mathbb{T}_1 &: \swapt/ \\
-\assoclt/ :&  \mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3) & \leftrightspoon & (\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3
+\unitetl/ :&  \exists\overline{k}.[\order{\idiso} \circledast \mathbb{T},
+                                                   \langle \idiso,v \rangle]
+  & \leftrightblackspoon &
+                    \exists\overline{k}.[\mathbb{T},v]
+  &: \unititl/ \\
+\unitetr/ :&  \exists\overline{k}.[\mathbb{T} \circledast \order{\idiso},
+                                                    \langle v,\idiso]
+  & \leftrightblackspoon &
+                    \exists\overline{k}.[\mathbb{T},v]
+  &: \unititr/ \\
+\swapt/ :&  \exists\overline{k}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
+                                                   \langle v_1,v_2 \rangle]
+ & \leftrightblackspoon &
+                   \exists\overline{k}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
+                                                  \langle v_2,v_1 \rangle]
+  &: \swapt/ \\
+  \assoclt/ :&  \exists\overline{k}.[\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
+                                        \langle v_1,\langle v_2,v_3 \rangle]
+  & \leftrightblackspoon &
+                       \exists\overline{k}.[(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
+                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
   &: \assocrt/ \\
 \\
-\eta-_k :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  [\iorder{c},1/c^k] \circledast [\order{c},c^k] &: \epsilon-_k \\
-\eta+ :& \order{\idiso} & \leftrightspoon & \order{c} \circledast \iorder{c} &: \epsilon+ \\
-\end{array}\]
-
+\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
+  \exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] &: \epsilon- \\
+\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
+  \exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] &: \epsilon+ \\
+\end{array}\]
 \[\begin{array}{c}
@@
 {}
-{\jdg{}{}{\idiso/ : \mathbb{T} \leftrightspoon \mathbb{T}}}
-{}
-\qquad
-\Rule{}
-{\jdg{}{}{\rho_1 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2} \quad \rho_2 : \mathbb{T}_2 \leftrightspoon \mathbb{T}_3}
-{\jdg{}{}{\rho_1 \circledcirc \rho_2 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_3}}
-{}
-\\
-\\
+{\jdg{}{}{\idiso/ : \exists\overline{k}.[\mathbb{T},v] \leftrightblackspoon
+  \exists\overline{k}.[\mathbb{T},v]}}
+{}
+~~
 \Rule{}
 {\jdg{}{}{\alpha : c_1 \isotwo c_2}}
-{\jdg{}{}{\order{\alpha} : \order{c_1}  \leftrightspoon \order{c_2}}}
-{}
-\qquad
-\Rule{}
-{\jdg{}{}{\rho_1 : \mathbb{T}_1 \leftrightspoon \mathbb{T}_2} \quad \rho_2 : \mathbb{T}_3 \leftrightspoon \mathbb{T}_4}
-{\jdg{}{}{\rho_1 \circledast \rho_2 : \mathbb{T}_1 \circledast \mathbb{T}_3 \leftrightspoon \mathbb{T}_2 \circledast \mathbb{T}_4}}
-{}
-\end{array}\]
-
-Each combinator $\rho$ has an inverse.
+{\jdg{}{}{(\order{\alpha})_k :
+ [\order{c_1},c_1^k]  \leftrightblackspoon
+ [\order{c_2}.c_2^k]}}
+{}
+\\
+\\
+\Rule{}
+{\jdg{}{}{\rho_1 : \exists{\overline{k_1}}.
+  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
+  \rho_2 : \exists{\overline{k_2}}.
+  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
+{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
+ \exists{\overline{k_1}\overline{k_2}}.
+ [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
+{}
+\\
+\\
+\Rule{}
+{\jdg{}{}{\rho_1 :
+  \exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
+  \exists{\overline{k_2}}.[\mathbb{T}_2,v_2]}
+ \qquad \rho_2 :
+  \exists{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
+  \exists{\overline{k_4}}.[\mathbb{T}_4,v_4]}
+{\jdg{}{}{\rho_1 \circledast \rho_2 :
+  \exists{\overline{k_1k_2k_3k_4}}.
+  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
+  \leftrightblackspoon
+  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
+{}
+\\
+\\
+\Rule{}
+{\jdg{}{}{}}
+{\jdg{}{}{\synch : \exists{\overline{k}k'}.[\mathbb{T},v]
+  \leftrightblackspoon
+   [\mathbb{T},v[k'\mapsto ??]]}}
+{}
+\end{array}\]
+Each combinator $\rho_\bullet$ has an inverse.
+}
 \caption{$\Pi^/$ fractional combinators (dependent version)}
@@
 imposes a constraint on the two values associated with the types $\order{c}$ and
-$\iorder{c}$; they must be ``synchronized'' sharing the same $k$.
+$\iorder{c}$; they must be ``synchronized'' sharing the same $k$. In general, a
+pointed type will be of the form $\exists\overline{k}.[\mathbb{T},v]$ for some
+sequence of integers $\overline{k}$ and some value $v$ in $\mathbb{T}$ which
+might of course refer to the integers $\overline{k}$ to enforce some
+synchronization among its sub-values.
+
+Here is how the circuit in Fig.~\ref{fig:zigzag} evaluates in the forward and
+backwards directions:
+\[\begin{array}{rcl}
+[\order{c},c^i]
+&\stackrel{\unitetl/}{\longmapsto}&
+  [\order{id}\circledast\order{c}, \langle \idiso,c^i \rangle] \\
+&\stackrel{\eta+\circledast\idiso/}{\longmapsto}&
+  \exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
+  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
+&\stackrel{\assocrt/}{\longmapsto}&
+  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
+  \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
+&\stackrel{\synch}{\longmapsto}&
+  [\order{c}\circledast(\iorder{c}\circledast\order{c}),
+  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
+&\stackrel{\idiso/\circledast\epsilon-}{\longmapsto}&
+  [\order{c}\circledast\order{\idiso},
+  \langle c^i, \idiso \rangle] \\
+&\stackrel{\unitetr/}{\longmapsto}&
+  [\order{c}, c^i]
+\end{array}\]
+
+\[\begin{array}{rcl}
+  [\order{c}, c^i]
+&\stackrel{\unitetr/}{\longmapsfrom}&
+  [\order{c}\circledast\order{\idiso},
+  \langle c^i, \idiso \rangle] \\
+&\stackrel{\idiso/\circledast\epsilon-}{\longmapsfrom}&
+  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
+  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
+&\stackrel{\synch}{\longmapsfrom}&
+ [\order{c}\circledast(\iorder{c}\circledast\order{c}),
+  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
+&\stackrel{\assocrt/}{\longmapsfrom}&
+  [(\order{c}\circledast\iorder{c})\circledast\order{c},
+  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
+&\stackrel{\eta+\circledast\idiso/}{\longmapsfrom}&
+  [\order{c}\circledast\order{\idiso}, \langle c^i, \idiso] \\
+&\stackrel{\unitetl/}{\longmapsfrom}&
+  [\order{c},c^i]
+\end{array}\]
 

### 2016-12-18 typespp.tex
subst is pullback; exists should be forall

@@
 annihilates. The challenge we are facing is to turn this idea into an operational
-semantics that somehow combines the local and global aspects of the equivalence.
+semantics that somehow combines the local and global aspects of the
+equivalence. We will be inspired by the folklore result that ``substitution into a
+dependent type is a pullback''~\cite{subpullback} as explained in the next section.
 
@@
 \[\begin{array}{rrcll}
-\unitetl/ :&  \exists\overline{k}.[\order{\idiso} \circledast \mathbb{T},
+\unitetl/ :&  \forall\overline{k}.[\order{\idiso} \circledast \mathbb{T},
                                                    \langle \idiso,v \rangle]
   & \leftrightblackspoon &
-                    \exists\overline{k}.[\mathbb{T},v]
+                    \forall\overline{k}.[\mathbb{T},v]
   &: \unititl/ \\
-\unitetr/ :&  \exists\overline{k}.[\mathbb{T} \circledast \order{\idiso},
+\unitetr/ :&  \forall\overline{k}.[\mathbb{T} \circledast \order{\idiso},
                                                     \langle v,\idiso]
   & \leftrightblackspoon &
-                    \exists\overline{k}.[\mathbb{T},v]
+                    \forall\overline{k}.[\mathbb{T},v]
   &: \unititr/ \\
-\swapt/ :&  \exists\overline{k}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
+\swapt/ :&  \forall\overline{k}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                    \langle v_1,v_2 \rangle]
  & \leftrightblackspoon &
-                   \exists\overline{k}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
+                   \forall\overline{k}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                   \langle v_2,v_1 \rangle]
   &: \swapt/ \\
-  \assoclt/ :&  \exists\overline{k}.[\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
+  \assoclt/ :&  \forall\overline{k}.[\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
                                         \langle v_1,\langle v_2,v_3 \rangle]
   & \leftrightblackspoon &
-                       \exists\overline{k}.[(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
+                       \forall\overline{k}.[(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
                                        \langle \langle v_1,v_2 \rangle , v_3 \rangle]
@@
 \eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  \exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] &: \epsilon- \\
+  \forall{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] &: \epsilon- \\
 \eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  \exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] &: \epsilon+ \\
+  \forall{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] &: \epsilon+ \\
 \end{array}\]
@@
 {}
-{\jdg{}{}{\idiso/ : \exists\overline{k}.[\mathbb{T},v] \leftrightblackspoon
-  \exists\overline{k}.[\mathbb{T},v]}}
+{\jdg{}{}{\idiso/ : \forall\overline{k}.[\mathbb{T},v] \leftrightblackspoon
+  \forall\overline{k}.[\mathbb{T},v]}}
 {}
@@
 \Rule{}
-{\jdg{}{}{\rho_1 : \exists{\overline{k_1}}.
+{\jdg{}{}{\rho_1 : \forall{\overline{k_1}}.
   [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
-  \rho_2 : \exists{\overline{k_2}}.
+  \rho_2 : \forall{\overline{k_2}}.
   [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
 {\jdg{}{}{\rho_1 \circledcirc \rho_2 :
- \exists{\overline{k_1}\overline{k_2}}.
+ \forall{\overline{k_1}\overline{k_2}}.
  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
@@
 {\jdg{}{}{\rho_1 :
-  \exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
-  \exists{\overline{k_2}}.[\mathbb{T}_2,v_2]}
+  \forall{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
+  \forall{\overline{k_2}}.[\mathbb{T}_2,v_2]}
  \qquad \rho_2 :
-  \exists{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
-  \exists{\overline{k_4}}.[\mathbb{T}_4,v_4]}
+  \forall{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
+  \forall{\overline{k_4}}.[\mathbb{T}_4,v_4]}
 {\jdg{}{}{\rho_1 \circledast \rho_2 :
-  \exists{\overline{k_1k_2k_3k_4}}.
+  \forall{\overline{k_1k_2k_3k_4}}.
   [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
@@
 {\jdg{}{}{}}
-{\jdg{}{}{\synch : \exists{\overline{k}k'}.[\mathbb{T},v]
+{\jdg{}{}{\synch : \forall{\overline{k}k'}.[\mathbb{T},v]
   \leftrightblackspoon
@@
   [\mathbb{T}_1 \circledast \mathbb{T}_2,\langle v_1,v_2 \rangle] \alt
-  \exists k.\mathbb{T_\bullet} \\
+  \forall k.\mathbb{T_\bullet} \\
  (\textrm{/-combinators}) & \rho_\bullet &:&
@@
 of types. In particular, the type
-$\exists k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
+$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
 imposes a constraint on the two values associated with the types $\order{c}$ and
 $\iorder{c}$; they must be ``synchronized'' sharing the same $k$. In general, a
-pointed type will be of the form $\exists\overline{k}.[\mathbb{T},v]$ for some
+pointed type will be of the form $\forall\overline{k}.[\mathbb{T},v]$ for some
 sequence of integers $\overline{k}$ and some value $v$ in $\mathbb{T}$ which
@@
 &\stackrel{\eta+\circledast\idiso/}{\longmapsto}&
-  \exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
+  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
   \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
 &\stackrel{\assocrt/}{\longmapsto}&
-  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
+  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
   \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
@@
 &\stackrel{\idiso/\circledast\epsilon-}{\longmapsfrom}&
-  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
+  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
   \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\

### 2016-12-18 typespp.tex
finished example
tweak 3.6, and expand explanation in 3.7

@@
 
-The looping groupoid $\looping{G}$ has the group elements as objects and
-morphisms between elements in the same conjugacy class, that is, there is a
+The looping groupoid $\looping{G}$ has the elements $g$ of the group $G$ as
+objects and morphisms between elements in the same conjugacy class; that is,
+there is a
 morphism between elements $g$ and $h$ iff there exists a $x \in G$ such that
@@
 $p : \tau \iso \tau$, we get a group whose elements are the iterates $p^k$. The
-corresponding looping groupoid has all the iterates as objects and morphisms
+corresponding looping groupoid has all the iterates as objects, and morphisms
 between $p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that
@@
 the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
-groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is the trivial
+groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is equivalent
+to the trivial
 groupoid with only one distinct object and one trivial automorphism. To avoid
@@
 condition that the commutative diagram in figure~\ref{fig:morita} is a
-\emph{pullback}. The intuition is that the codomain consists of three objects
+\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps,
+and $\mathbb{1}$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids
+defined previously.
+The intuition is that the codomain consists of three objects
 with associated morphisms, each of them counting as a ``third.'' The functor
@@
 equivalence. We will be inspired by the folklore result that ``substitution into a
-dependent type is a pullback''~\cite{subpullback} as explained in the next section.
+dependent type is a pullback''~\cite{subpullback} (but which turns out to have
+been fully explained in~\cite{practicalfoundations}) as explained in the next
+section.
 
@@
     \arrow[dr, dotted] & & \\
-    & \set{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
-    & G_1 \arrow[d, "{(s,t)}"] \\
-    & \set{\ast} \times \set{\ast} \arrow[r, "{\psi_0 \times \psi_0}"]
-    & G_0 \times G_0
+    & \order{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
+    & G \arrow[d, "{(s,t)}"] \\
+    & \mathbb{1} \times \mathbb{1} \arrow[r, "{\psi_0 \times \psi_0}"]
+    & \mathbb{3} \times \mathbb{3}
   \end{tikzcd}
@@
 {\jdg{}{}{}}
-{\jdg{}{}{\synch : \forall{\overline{k}k'}.[\mathbb{T},v]
+{\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
   \leftrightblackspoon
-   [\mathbb{T},v[k'\mapsto ??]]}}
+   \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
 {}
@@
   \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
-&\stackrel{\synch}{\longmapsto}&
+&\stackrel{\synch_i}{\longmapsto}&
   [\order{c}\circledast(\iorder{c}\circledast\order{c}),
@@
   \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
-&\stackrel{\synch}{\longmapsfrom}&
+&\stackrel{\synch_i}{\longmapsfrom}&
  [\order{c}\circledast(\iorder{c}\circledast\order{c}),

### 2016-12-18 typespp.tex
edit 4.1 and move Figure 8

@@
 
-We are now ready to turn the groupoid constructions from the previous section into an extension of $\Pi$ with new type constructors and combinators for creating and manipulating syntactic counterparts to $\order{p}$ and $\iorder{p}$.
+We are now ready to turn the groupoid constructions from the previous section
+into an extension of $\Pi$ with new type constructors and combinators for
+creating and manipulating syntactic counterparts to $\order{p}$ and
+$\iorder{p}$.  Just adding the ``obvious'' types and combinators does
+not work; however, as the failures are quite informative, we nevertheless go
+through this first attempt.  This will help illustrate why we need extra
+machinery in Sec.~\ref{sec:deppointedtypes}.
 
@@
 %%%%%
-\subsection{Non-Dependent Version}
+\subsection{First attempt}
 
@@
 
-The syntactic category $\ft$ of fractional types introduces type expressions for
-iteration groupoids, symmetry groupoids, and their products. The combinators
-that relate these fractional types are in Fig.~\ref{pifrac:comb}. The bottom
-group includes identity and sequential composition~$\circledcirc$ combinators
-and ensures that the combinators can be applied anywhere inside a product
-$\circledast$. There is also a combinator that relates two types $\order{c_1}$
-and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo
-c_2$.
-Indeed if $c_1$ and $c_2$ are considered equivalent via a 2-combinator then we
-also consider their iteration groupoids to be equivalent. The top group ensures
-that $\circledast$ carries a symmetric monoidal structure with combinators
-witnessing the unit, commutativity, and associativity properties. The last two
-pairs combinators $\eta-/\epsilon-$ and $\eta+/\epsilon+$ are inspired from the
-definition of compact closed categories~\cite{ccc} which are symmetric monoidal
-categories in which every object has a dual. They are the ones that witness the
-``fractional'' nature of symmetry groupoids as motivated in Sec.~\ref{sub:products}.
+The syntactic category $\ft$ of fractional types introduces type expressions
+for iteration groupoids, symmetry groupoids, and their products. The
+combinators that relate these fractional types are in Fig.~\ref{pifrac:comb}.
+The bottom group includes identity and sequential composition~$\circledcirc$
+combinators and ensures that the combinators can be applied anywhere inside a
+product $\circledast$. There is also a combinator that relates two types
+$\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1
+\isotwo c_2$.  Indeed if $c_1$ and $c_2$ are considered equivalent via a
+$2$-combinator then we also consider their iteration groupoids to be equivalent.
+The top group ensures that $\circledast$ carries a symmetric monoidal structure
+with combinators witnessing the unit, commutativity, and associativity
+properties. The last two pairs of combinators $\eta\!- / \epsilon-$ and
+$\eta\!+ / \epsilon+$ are inspired from the definition of compact closed
+categories~\cite{ccc} which are symmetric monoidal categories in which every
+object has a dual. They are the ones that witness the ``fractional'' nature of
+symmetry groupoids as motivated in Sec.~\ref{sub:products}.
 
 The last new syntactic category in $\Pi^/$ is that of values which deserves some
-background discussion. When the types denote sets, it is evident what it means
-to have a value of a given type: it is just an element of the set. When types
-denote groupoids, it is less clear what it means to have a value, especially
-when the total number of values, as reported by the groupoid cardinality, is a
-proper fraction. In that case we the values up to an equivalence relation. The
-idea is not uncommon: in the conventional $\lambda$-calculus, we list
+additional discussion. When types denote sets, values of a type are clear:
+they are just the elements of the set.  In the case when types are groupoids,
+this is less clear, especially when the groupoid cardinality is a proper
+fraction.  What we get instead are equivalence classes of values.
+The idea is not uncommon: in the conventional $\lambda$-calculus, we list
 $\lambda x.x$ and $\lambda y.y$ as separate values of type
@@
 
-\begin{figure}[t]
-{\footnotesize
-\[\begin{array}{rrcll}
-\unitetl/ :&  \forall\overline{k}.[\order{\idiso} \circledast \mathbb{T},
-                                                   \langle \idiso,v \rangle]
-  & \leftrightblackspoon &
-                    \forall\overline{k}.[\mathbb{T},v]
-  &: \unititl/ \\
-\unitetr/ :&  \forall\overline{k}.[\mathbb{T} \circledast \order{\idiso},
-                                                    \langle v,\idiso]
-  & \leftrightblackspoon &
-                    \forall\overline{k}.[\mathbb{T},v]
-  &: \unititr/ \\
-\swapt/ :&  \forall\overline{k}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
-                                                   \langle v_1,v_2 \rangle]
- & \leftrightblackspoon &
-                   \forall\overline{k}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
-                                                  \langle v_2,v_1 \rangle]
-  &: \swapt/ \\
-  \assoclt/ :&  \forall\overline{k}.[\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
-                                        \langle v_1,\langle v_2,v_3 \rangle]
-  & \leftrightblackspoon &
-                       \forall\overline{k}.[(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
-                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
-  &: \assocrt/ \\
-\\
-\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  \forall{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] &: \epsilon- \\
-\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  \forall{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] &: \epsilon+ \\
-\end{array}\]
-\[\begin{array}{c}
-\Rule{}
-{}
-{\jdg{}{}{\idiso/ : \forall\overline{k}.[\mathbb{T},v] \leftrightblackspoon
-  \forall\overline{k}.[\mathbb{T},v]}}
-{}
-~~
-\Rule{}
-{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
-{\jdg{}{}{(\order{\alpha})_k :
- [\order{c_1},c_1^k]  \leftrightblackspoon
- [\order{c_2}.c_2^k]}}
-{}
-\\
-\\
-\Rule{}
-{\jdg{}{}{\rho_1 : \forall{\overline{k_1}}.
-  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
-  \rho_2 : \forall{\overline{k_2}}.
-  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
-{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
- \forall{\overline{k_1}\overline{k_2}}.
- [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
-{}
-\\
-\\
-\Rule{}
-{\jdg{}{}{\rho_1 :
-  \forall{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
-  \forall{\overline{k_2}}.[\mathbb{T}_2,v_2]}
- \qquad \rho_2 :
-  \forall{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
-  \forall{\overline{k_4}}.[\mathbb{T}_4,v_4]}
-{\jdg{}{}{\rho_1 \circledast \rho_2 :
-  \forall{\overline{k_1k_2k_3k_4}}.
-  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
-  \leftrightblackspoon
-  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
-{}
-\\
-\\
-\Rule{}
-{\jdg{}{}{}}
-{\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
-  \leftrightblackspoon
-   \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
-{}
-\end{array}\]
-Each combinator $\rho_\bullet$ has an inverse.
-}
-\caption{$\Pi^/$ fractional combinators (dependent version)}
-\label{pifracd:comb}
-\end{figure}
-
 %%%%%
-\subsection{Dependent Pointed Types}
+\subsection{Dependent Pointed Types}\label{sec:deppointedtypes}
 
@@
 
+\begin{figure}[t]
+{\footnotesize
+\[\begin{array}{rrcll}
+\unitetl/ :&  \forall\overline{k}.[\order{\idiso} \circledast \mathbb{T},
+                                                   \langle \idiso,v \rangle]
+  & \leftrightblackspoon &
+                    \forall\overline{k}.[\mathbb{T},v]
+  &: \unititl/ \\
+\unitetr/ :&  \forall\overline{k}.[\mathbb{T} \circledast \order{\idiso},
+                                                    \langle v,\idiso]
+  & \leftrightblackspoon &
+                    \forall\overline{k}.[\mathbb{T},v]
+  &: \unititr/ \\
+\swapt/ :&  \forall\overline{k}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
+                                                   \langle v_1,v_2 \rangle]
+ & \leftrightblackspoon &
+                   \forall\overline{k}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
+                                                  \langle v_2,v_1 \rangle]
+  &: \swapt/ \\
+  \assoclt/ :&  \forall\overline{k}.[\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
+                                        \langle v_1,\langle v_2,v_3 \rangle]
+  & \leftrightblackspoon &
+                       \forall\overline{k}.[(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
+                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
+  &: \assocrt/ \\
+\\
+\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
+  \forall{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] &: \epsilon- \\
+\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
+  \forall{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] &: \epsilon+ \\
+\end{array}\]
+\[\begin{array}{c}
+\Rule{}
+{}
+{\jdg{}{}{\idiso/ : \forall\overline{k}.[\mathbb{T},v] \leftrightblackspoon
+  \forall\overline{k}.[\mathbb{T},v]}}
+{}
+~~
+\Rule{}
+{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
+{\jdg{}{}{(\order{\alpha})_k :
+ [\order{c_1},c_1^k]  \leftrightblackspoon
+ [\order{c_2}.c_2^k]}}
+{}
+\\
+\\
+\Rule{}
+{\jdg{}{}{\rho_1 : \forall{\overline{k_1}}.
+  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
+  \rho_2 : \forall{\overline{k_2}}.
+  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
+{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
+ \forall{\overline{k_1}\overline{k_2}}.
+ [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
+{}
+\\
+\\
+\Rule{}
+{\jdg{}{}{\rho_1 :
+  \forall{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
+  \forall{\overline{k_2}}.[\mathbb{T}_2,v_2]}
+ \qquad \rho_2 :
+  \forall{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
+  \forall{\overline{k_4}}.[\mathbb{T}_4,v_4]}
+{\jdg{}{}{\rho_1 \circledast \rho_2 :
+  \forall{\overline{k_1k_2k_3k_4}}.
+  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
+  \leftrightblackspoon
+  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
+{}
+\\
+\\
+\Rule{}
+{\jdg{}{}{}}
+{\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
+  \leftrightblackspoon
+   \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
+{}
+\end{array}\]
+Each combinator $\rho_\bullet$ has an inverse.
+}
+\caption{$\Pi^/$ fractional combinators (dependent version)}
+\label{pifracd:comb}
+\end{figure}
+
 The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$

### 2016-12-18 typespp.tex
edit parts of 4.2, and use the right latex syntax in 3.7

@@
 %%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Symmetry Groupoids $\iorder{p}$}
+\subsection{Symmetry Groupoids $\iorder{p}$}\label{sec:symmetryg}
 
@@
 
-Note that for each power $p ^ i$ of $p$, there is a morphism $\triple{k}{q}{\alpha}$ in $\iter{p}$ such that $q$ annihilates $p^i$ to the identity.
-Note also that everything is well-defined even if we choose $p : \zt\iso\zt$.
-In that case, the cardinality is 1.
+Note that for each power $p ^ i$ of $p$, there is a morphism
+$\triple{k}{q}{\alpha}$ in $\iter{p}$ such that $q$ annihilates $p^i$ to the
+identity. Note also that everything is well-defined even if we choose $p :
+\zt\iso\zt$. In that case, the cardinality is 1.
 
@@
 \emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps,
-and $\mathbb{1}$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids
+and $\ot$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids
 defined previously.
@@
     & G \arrow[d, "{(s,t)}"] \\
-    & \mathbb{1} \times \mathbb{1} \arrow[r, "{\psi_0 \times \psi_0}"]
+    & \ot \times \ot \arrow[r, "{\psi_0 \times \psi_0}"]
     & \mathbb{3} \times \mathbb{3}
@@
 that is reminiscent of entanglement in quantum mechanics. Possibilities to
-realize this communication are global reference cells, backtracking, logical
-variables with unification. We choose a more abstract approach that encodes the
-required dependency in dataflow constraints using dependent types.
+realize this communication are global reference cells, backtracking, or logical
+variables with unification. Instead, we choose a more abstract approach that
+encodes the required dependency in dataflow constraints using dependent types.
 
@@
 where each type $\mathbb{T}$ refers to a particular value $v$ in
-$\mathbb{T}$. We then allow indices of values $k$ to be range over a collection
-of types. In particular, the type
+$\mathbb{T}$. We then allow universal quantification over indices $k$
+associated to values, over a collection of types.
+In particular, the type
 $\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
 imposes a constraint on the two values associated with the types $\order{c}$ and
-$\iorder{c}$; they must be ``synchronized'' sharing the same $k$. In general, a
-pointed type will be of the form $\forall\overline{k}.[\mathbb{T},v]$ for some
-sequence of integers $\overline{k}$ and some value $v$ in $\mathbb{T}$ which
-might of course refer to the integers $\overline{k}$ to enforce some
-synchronization among its sub-values.
+$\iorder{c}$; they must be ``synchronized'' by sharing the same $k$. In
+general, a pointed type will be of the form
+$\forall\overline{k}.[\mathbb{T},v]$ for some sequence of integers
+$\overline{k}$ and some value $v$ in $\mathbb{T}$, which might of course refer
+to the integers of $\overline{k}$ to enforce some synchronization among its
+sub-values.  It is good to remember that there is no loss in generality
+in going to pointed types, as all fractional types are inhabited (see
+remark in Sec.~\cite{sec:symmetryg}.
 

### 2016-12-18 typespp.tex
ref, not cite.

@@
 in going to pointed types, as all fractional types are inhabited (see
-remark in Sec.~\cite{sec:symmetryg}.
+remark in Sec.~\ref{sec:symmetryg}.
 

### 2016-12-18 typespp.tex
forall is dual to exists

--- 2016-12-18 typespp.tex
@@
 
-The looping groupoid $\looping{G}$ has the elements $g$ of the group $G$ as
-objects and morphisms between elements in the same conjugacy class; that is,
-there is a
+The looping groupoid $\looping{G}$ has the group elements as objects and
+morphisms between elements in the same conjugacy class, that is, there is a
 morphism between elements $g$ and $h$ iff there exists a $x \in G$ such that
@@
 $p : \tau \iso \tau$, we get a group whose elements are the iterates $p^k$. The
-corresponding looping groupoid has all the iterates as objects, and morphisms
+corresponding looping groupoid has all the iterates as objects and morphisms
 between $p^{k_1}$ and $p^{k_2}$ whenever there exists an integer $j$ such that
@@
 the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
-groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is equivalent
-to the trivial
+groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is the trivial
 groupoid with only one distinct object and one trivial automorphism. To avoid
@@
 condition that the commutative diagram in figure~\ref{fig:morita} is a
-\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps,
-and $\mathbb{1}$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids
-defined previously.
-The intuition is that the codomain consists of three objects
+\emph{pullback}. The intuition is that the codomain consists of three objects
 with associated morphisms, each of them counting as a ``third.'' The functor
@@
 equivalence. We will be inspired by the folklore result that ``substitution into a
-dependent type is a pullback''~\cite{subpullback} (but which turns out to have
-been fully explained in~\cite{practicalfoundations}) as explained in the next
-section.
+dependent type is a pullback''~\cite{subpullback} as explained in the next section.
 
@@
     \arrow[dr, dotted] & & \\
-    & \order{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
-    & G \arrow[d, "{(s,t)}"] \\
-    & \mathbb{1} \times \mathbb{1} \arrow[r, "{\psi_0 \times \psi_0}"]
-    & \mathbb{3} \times \mathbb{3}
+    & \set{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
+    & G_1 \arrow[d, "{(s,t)}"] \\
+    & \set{\ast} \times \set{\ast} \arrow[r, "{\psi_0 \times \psi_0}"]
+    & G_0 \times G_0
   \end{tikzcd}
@@
 
-We are now ready to turn the groupoid constructions from the previous section
-into an extension of $\Pi$ with new type constructors and combinators for
-creating and manipulating syntactic counterparts to $\order{p}$ and
-$\iorder{p}$.  Just adding the ``obvious'' types and combinators does
-not work; however, as the failures are quite informative, we nevertheless go
-through this first attempt.  This will help illustrate why we need extra
-machinery in Sec.~\ref{sec:deppointedtypes}.
+We are now ready to turn the groupoid constructions from the previous section into an extension of $\Pi$ with new type constructors and combinators for creating and manipulating syntactic counterparts to $\order{p}$ and $\iorder{p}$.
 
@@
 %%%%%
-\subsection{First attempt}
+\subsection{Non-Dependent Version}
 
@@
 
-The syntactic category $\ft$ of fractional types introduces type expressions
-for iteration groupoids, symmetry groupoids, and their products. The
-combinators that relate these fractional types are in Fig.~\ref{pifrac:comb}.
-The bottom group includes identity and sequential composition~$\circledcirc$
-combinators and ensures that the combinators can be applied anywhere inside a
-product $\circledast$. There is also a combinator that relates two types
-$\order{c_1}$ and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1
-\isotwo c_2$.  Indeed if $c_1$ and $c_2$ are considered equivalent via a
-$2$-combinator then we also consider their iteration groupoids to be equivalent.
-The top group ensures that $\circledast$ carries a symmetric monoidal structure
-with combinators witnessing the unit, commutativity, and associativity
-properties. The last two pairs of combinators $\eta\!- / \epsilon-$ and
-$\eta\!+ / \epsilon+$ are inspired from the definition of compact closed
-categories~\cite{ccc} which are symmetric monoidal categories in which every
-object has a dual. They are the ones that witness the ``fractional'' nature of
-symmetry groupoids as motivated in Sec.~\ref{sub:products}.
+The syntactic category $\ft$ of fractional types introduces type expressions for
+iteration groupoids, symmetry groupoids, and their products. The combinators
+that relate these fractional types are in Fig.~\ref{pifrac:comb}. The bottom
+group includes identity and sequential composition~$\circledcirc$ combinators
+and ensures that the combinators can be applied anywhere inside a product
+$\circledast$. There is also a combinator that relates two types $\order{c_1}$
+and $\order{c_2}$ via $\order{\alpha}$ whenever $\alpha : c_1 \isotwo
+c_2$.
+Indeed if $c_1$ and $c_2$ are considered equivalent via a 2-combinator then we
+also consider their iteration groupoids to be equivalent. The top group ensures
+that $\circledast$ carries a symmetric monoidal structure with combinators
+witnessing the unit, commutativity, and associativity properties. The last two
+pairs combinators $\eta-/\epsilon-$ and $\eta+/\epsilon+$ are inspired from the
+definition of compact closed categories~\cite{ccc} which are symmetric monoidal
+categories in which every object has a dual. They are the ones that witness the
+``fractional'' nature of symmetry groupoids as motivated in Sec.~\ref{sub:products}.
 
 The last new syntactic category in $\Pi^/$ is that of values which deserves some
-additional discussion. When types denote sets, values of a type are clear:
-they are just the elements of the set.  In the case when types are groupoids,
-this is less clear, especially when the groupoid cardinality is a proper
-fraction.  What we get instead are equivalence classes of values.
-The idea is not uncommon: in the conventional $\lambda$-calculus, we list
+background discussion. When the types denote sets, it is evident what it means
+to have a value of a given type: it is just an element of the set. When types
+denote groupoids, it is less clear what it means to have a value, especially
+when the total number of values, as reported by the groupoid cardinality, is a
+proper fraction. In that case we the values up to an equivalence relation. The
+idea is not uncommon: in the conventional $\lambda$-calculus, we list
 $\lambda x.x$ and $\lambda y.y$ as separate values of type
@@
 
+\begin{figure}[t]
+{\footnotesize
+\[\begin{array}{rrcll}
+\unitetl/ :& [\order{\idiso} \circledast \mathbb{T},
+                                                   \langle \idiso,v \rangle]
+  & \leftrightblackspoon &
+                   [\mathbb{T},v]
+  &: \unititl/ \\
+\unitetr/ :& [\mathbb{T} \circledast \order{\idiso},
+                                                    \langle v,\idiso]
+  & \leftrightblackspoon &
+                   [\mathbb{T},v]
+  &: \unititr/ \\
+\swapt/ :& [\mathbb{T}_1 \circledast \mathbb{T}_2,
+                                                   \langle v_1,v_2 \rangle]
+ & \leftrightblackspoon &
+                  [\mathbb{T}_2 \circledast \mathbb{T}_1,
+                                                  \langle v_2,v_1 \rangle]
+  &: \swapt/ \\
+  \assoclt/ :& [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
+                                        \langle v_1,\langle v_2,v_3 \rangle]
+  & \leftrightblackspoon &
+                      [(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
+                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
+  &: \assocrt/ \\
+\\
+\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
+  \forall{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] \\
+\epsilon-: & \exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle]
+  & \leftrightblackspoon & [\order{\idiso},\idiso] \\
+\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
+  \forall{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] \\
+\epsilon+ : & \exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
+    & \leftrightblackspoon & [\order{\idiso},\idiso]
+\end{array}\]
+\[\begin{array}{c}
+\Rule{}
+{}
+{\jdg{}{}{\idiso/ :[\mathbb{T},v] \leftrightblackspoon
+ [\mathbb{T},v]}}
+{}
+~~
+\Rule{}
+{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
+{\jdg{}{}{(\order{\alpha})_k :
+ [\order{c_1},c_1^k]  \leftrightblackspoon
+ [\order{c_2}.c_2^k]}}
+{}
+\\
+\\
+\Rule{}
+{\jdg{}{}{\rho_1 :
+  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
+  \rho_2 :
+  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
+{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
+ [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
+{}
+\\
+\\
+\Rule{}
+{\jdg{}{}{\rho_1 :
+ [\mathbb{T}_1,v_1] \leftrightblackspoon
+ [\mathbb{T}_2,v_2]}
+ \qquad \rho_2 :
+ [\mathbb{T}_3,v_3] \leftrightblackspoon
+ [\mathbb{T}_4,v_4]}
+{\jdg{}{}{\rho_1 \circledast \rho_2 :
+  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
+  \leftrightblackspoon
+  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
+{}
+% \\
+% \\
+% \Rule{}
+% {\jdg{}{}{}}
+% {\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
+%   \leftrightblackspoon
+%    \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
+% {}
+\end{array}\]
+Each combinator $\rho_\bullet$ has an inverse.
+}
+\caption{$\Pi^/$ fractional combinators (dependent version)}
+\label{pifracd:comb}
+\end{figure}
+
 %%%%%
-\subsection{Dependent Pointed Types}\label{sec:deppointedtypes}
+\subsection{Dependent Pointed Types}
 
@@
   [\mathbb{T}_1 \circledast \mathbb{T}_2,\langle v_1,v_2 \rangle] \alt
-  \forall k.\mathbb{T_\bullet} \\
+  \forall k.\mathbb{T_\bullet} \alt \exists k.\mathbb{T_\bullet} \\
  (\textrm{/-combinators}) & \rho_\bullet &:&
@@
 \end{definition}
-
-\begin{figure}[t]
-{\footnotesize
-\[\begin{array}{rrcll}
-\unitetl/ :&  \forall\overline{k}.[\order{\idiso} \circledast \mathbb{T},
-                                                   \langle \idiso,v \rangle]
-  & \leftrightblackspoon &
-                    \forall\overline{k}.[\mathbb{T},v]
-  &: \unititl/ \\
-\unitetr/ :&  \forall\overline{k}.[\mathbb{T} \circledast \order{\idiso},
-                                                    \langle v,\idiso]
-  & \leftrightblackspoon &
-                    \forall\overline{k}.[\mathbb{T},v]
-  &: \unititr/ \\
-\swapt/ :&  \forall\overline{k}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
-                                                   \langle v_1,v_2 \rangle]
- & \leftrightblackspoon &
-                   \forall\overline{k}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
-                                                  \langle v_2,v_1 \rangle]
-  &: \swapt/ \\
-  \assoclt/ :&  \forall\overline{k}.[\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
-                                        \langle v_1,\langle v_2,v_3 \rangle]
-  & \leftrightblackspoon &
-                       \forall\overline{k}.[(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
-                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
-  &: \assocrt/ \\
-\\
-\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  \forall{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] &: \epsilon- \\
-\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  \forall{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] &: \epsilon+ \\
-\end{array}\]
-\[\begin{array}{c}
-\Rule{}
-{}
-{\jdg{}{}{\idiso/ : \forall\overline{k}.[\mathbb{T},v] \leftrightblackspoon
-  \forall\overline{k}.[\mathbb{T},v]}}
-{}
-~~
-\Rule{}
-{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
-{\jdg{}{}{(\order{\alpha})_k :
- [\order{c_1},c_1^k]  \leftrightblackspoon
- [\order{c_2}.c_2^k]}}
-{}
-\\
-\\
-\Rule{}
-{\jdg{}{}{\rho_1 : \forall{\overline{k_1}}.
-  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
-  \rho_2 : \forall{\overline{k_2}}.
-  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
-{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
- \forall{\overline{k_1}\overline{k_2}}.
- [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
-{}
-\\
-\\
-\Rule{}
-{\jdg{}{}{\rho_1 :
-  \forall{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
-  \forall{\overline{k_2}}.[\mathbb{T}_2,v_2]}
- \qquad \rho_2 :
-  \forall{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
-  \forall{\overline{k_4}}.[\mathbb{T}_4,v_4]}
-{\jdg{}{}{\rho_1 \circledast \rho_2 :
-  \forall{\overline{k_1k_2k_3k_4}}.
-  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
-  \leftrightblackspoon
-  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
-{}
-\\
-\\
-\Rule{}
-{\jdg{}{}{}}
-{\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
-  \leftrightblackspoon
-   \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
-{}
-\end{array}\]
-Each combinator $\rho_\bullet$ has an inverse.
-}
-\caption{$\Pi^/$ fractional combinators (dependent version)}
-\label{pifracd:comb}
-\end{figure}
 
@@
 imposes a constraint on the two values associated with the types $\order{c}$ and
-$\iorder{c}$; they must be ``synchronized'' sharing the same $k$. In general, a
-pointed type will be of the form $\forall\overline{k}.[\mathbb{T},v]$ for some
-sequence of integers $\overline{k}$ and some value $v$ in $\mathbb{T}$ which
-might of course refer to the integers $\overline{k}$ to enforce some
-synchronization among its sub-values.
+$\iorder{c}$; they must be ``synchronized'' sharing the same $k$.
 
@@
   \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
-&\stackrel{\synch_i}{\longmapsto}&
+%% &\stackrel{\synch_i}{\longmapsto}&
+&k\mapsto i&
   [\order{c}\circledast(\iorder{c}\circledast\order{c}),
   \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
+&\exists&
+  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
+  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
 &\stackrel{\idiso/\circledast\epsilon-}{\longmapsto}&
@@
   \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
-&\stackrel{\synch_i}{\longmapsfrom}&
- [\order{c}\circledast(\iorder{c}\circledast\order{c}),
-  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
 &\stackrel{\assocrt/}{\longmapsfrom}&
+  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
+  \langle \langle c^i,1/c^k \rangle,c^k \rangle] \\
+&k\mapsto i&
   [(\order{c}\circledast\iorder{c})\circledast\order{c},
   \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
+&\exists&
+  \exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
+  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
 &\stackrel{\eta+\circledast\idiso/}{\longmapsfrom}&
-  [\order{c}\circledast\order{\idiso}, \langle c^i, \idiso] \\
+  [\order{\idiso}\circledast\order{c}, \langle \idiso,c^i\rangle] \\
 &\stackrel{\unitetl/}{\longmapsfrom}&

### 2016-12-18 typespp.tex

@@
 condition that the commutative diagram in figure~\ref{fig:morita} is a
-\emph{pullback}. The intuition is that the codomain consists of three objects
+\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps,
+and $\ot$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids
+defined previously.
+The intuition is that the codomain consists of three objects
 with associated morphisms, each of them counting as a ``third.'' The functor
@@
     \arrow[dr, dotted] & & \\
-    & \set{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
-    & G_1 \arrow[d, "{(s,t)}"] \\
-    & \set{\ast} \times \set{\ast} \arrow[r, "{\psi_0 \times \psi_0}"]
-    & G_0 \times G_0
+    & \order{\idiso} \arrow[r, "{\psi_1}"] \arrow[d, "{(s,t)}"]
+    & G \arrow[d, "{(s,t)}"] \\
+    & \ot \times \ot \arrow[r, "{\psi_0 \times \psi_0}"]
+    & \mathbb{3} \times \mathbb{3}
   \end{tikzcd}
@@
 
+%%%%%
+\subsection{Dependent Pointed Types}\label{sec:deppointedtypes}
+
+To understand the challenge in designing an operational semantics for $\eta$ and
+$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
+Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
+manipulation on types:
+\[
+x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
+\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
+\]
+and by the coherence conditions of compact closed categories, we expect this
+circuit, when run in either direction, to be equivalent to the
+identity. Consider now the case of a combinator $c$ whose order is greater than
+or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
+$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
+$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
+$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
+either end. The problem becomes apparent when one observes that the uses of
+$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
+right-to-left execution must behave differently depending on the incoming
+value. But the ``local'' information available to either of these combinators
+does not include the ``global'' information about the incoming value and there
+is no way for them to make consistent guesses locally. It is possible to modify
+the operational semantics to use a mechanism to express a computational effect
+which enables spatially separated parts of the program to communicate in a way
+that is reminiscent of entanglement in quantum mechanics. Possibilities to
+realize this communication are global reference cells, backtracking, or logical
+variables with unification. Instead, we choose a more abstract approach that
+encodes the required dependency in dataflow constraints using dependent types.
+
+\begin{definition}[Dependent $\Pi^/$]
+\label{def:pifracd}
+The syntax of dependent $\Pi^/$ differs from Def.~\ref{def:pifrac} as follows:
+\[\begin{array}{lrcl}
+(\textrm{Fractional Types}) & \mathbb{T_\bullet} &::=&
+  [\order{c},c^k] \alt
+  [\iorder{c},1/c^k] \alt
+  [\mathbb{T}_1 \circledast \mathbb{T}_2,\langle v_1,v_2 \rangle] \alt
+  \forall k.\mathbb{T_\bullet} \\
+ (\textrm{/-combinators}) & \rho_\bullet &:&
+  \mathbb{T}_{\bullet 1} \leftrightblackspoon \mathbb{T}_{\bullet 2} ~::=
+  [\textit{see Fig.~\ref{pifracd:comb}}]
+\end{array}\]
+\end{definition}
+
 \begin{figure}[t]
@@
 where each type $\mathbb{T}$ refers to a particular value $v$ in
-$\mathbb{T}$. We then allow indices of values $k$ to be range over a collection
-of types. In particular, the type
+$\mathbb{T}$. We then allow universal quantification over indices $k$
+associated to values, over a collection of types.
+In particular, the type
 $\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
@@
 $\iorder{c}$; they must be ``synchronized'' sharing the same $k$.
+It is important to remember that there is no loss in generality
+in going to pointed types, as all fractional types are inhabited (see
+remark in Sec.~\ref{sec:symmetryg}.
 

### 2016-12-18 typespp.tex
more comments in 4.2

--- 2016-12-18 typespp.tex
@@
 to the integers of $\overline{k}$ to enforce some synchronization among its
-sub-values.  It is good to remember that there is no loss in generality
-in going to pointed types, as all fractional types are inhabited (see
-remark in Sec.~\ref{sec:symmetryg}.
+sub-values.  The resulting combinators are shown in Fig.~\ref{pifrac:comb}.
+We omit the quantifier from the notation when it is over the empty sequence
+integers.  We recall that, as all fractional types are inhabited (see remark in
+Sec.~\ref{sec:symmetryg}), and thus there is no loss in generality in going to
+pointed types.
 

### 2016-12-18 typespp.tex

--- 2016-12-18 typespp.tex
@@
 imposes a constraint on the two values associated with the types $\order{c}$ and
-$\iorder{c}$; they must be ``synchronized'' sharing the same $k$.
-It is important to remember that there is no loss in generality
-in going to pointed types, as all fractional types are inhabited (see
-remark in Sec.~\ref{sec:symmetryg}.
+$\iorder{c}$; they must be ``synchronized'' by sharing the same $k$. In
+general, a pointed type will be of the form
+$\forall\overline{k}.[\mathbb{T},v]$ for some sequence of integers
+$\overline{k}$ and some value $v$ in $\mathbb{T}$, which might of course refer
+to the integers of $\overline{k}$ to enforce some synchronization among its
+sub-values.  The resulting combinators are shown in Fig.~\ref{pifrac:comb}.
+We omit the quantifier from the notation when it is over the empty sequence
+integers.  We recall that, as all fractional types are inhabited (see remark in
+Sec.~\ref{sec:symmetryg}), and thus there is no loss in generality in going to
+pointed types.
 

### 2016-12-18 typespp.tex
removed redundant def
actually there was a redundant section for some reason; fixed
a few thoughts

@@
 
-%%%%%
-\subsection{Dependent Pointed Types}\label{sec:deppointedtypes}
-
-To understand the challenge in designing an operational semantics for $\eta$ and
-$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
-Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
-manipulation on types:
-\[
-x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
-\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
-\]
-and by the coherence conditions of compact closed categories, we expect this
-circuit, when run in either direction, to be equivalent to the
-identity. Consider now the case of a combinator $c$ whose order is greater than
-or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
-$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
-$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
-$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
-either end. The problem becomes apparent when one observes that the uses of
-$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
-right-to-left execution must behave differently depending on the incoming
-value. But the ``local'' information available to either of these combinators
-does not include the ``global'' information about the incoming value and there
-is no way for them to make consistent guesses locally. It is possible to modify
-the operational semantics to use a mechanism to express a computational effect
-which enables spatially separated parts of the program to communicate in a way
-that is reminiscent of entanglement in quantum mechanics. Possibilities to
-realize this communication are global reference cells, backtracking, or logical
-variables with unification. Instead, we choose a more abstract approach that
-encodes the required dependency in dataflow constraints using dependent types.
-
-\begin{definition}[Dependent $\Pi^/$]
-\label{def:pifracd}
-The syntax of dependent $\Pi^/$ differs from Def.~\ref{def:pifrac} as follows:
-\[\begin{array}{lrcl}
-(\textrm{Fractional Types}) & \mathbb{T_\bullet} &::=&
-  [\order{c},c^k] \alt
-  [\iorder{c},1/c^k] \alt
-  [\mathbb{T}_1 \circledast \mathbb{T}_2,\langle v_1,v_2 \rangle] \alt
-  \forall k.\mathbb{T_\bullet} \\
- (\textrm{/-combinators}) & \rho_\bullet &:&
-  \mathbb{T}_{\bullet 1} \leftrightblackspoon \mathbb{T}_{\bullet 2} ~::=
-  [\textit{see Fig.~\ref{pifracd:comb}}]
-\end{array}\]
-\end{definition}
-
 \begin{figure}[t]
@@
 %%%%%
-\subsection{Dependent Pointed Types}
+\subsection{Dependent Pointed Types}\label{sec:deppointedtypes}
 
@@
 
+Need to explain that a $\forall$ in the forward direction is a choice and
+becomes $\exists$ in the backwards direction.
+
+Rules for moving quantifiers?
+
+Combinators for packing (creating an existential) and synchronization
+(instantiating a universal)??
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-12-18 typespp.tex
Fix a typo

@@
 the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
-groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is equivalent
-to the trivial
-groupoid with only one distinct object and one trivial automorphism. To avoid
-clutter we assume the groupoids are strict, e.g., that the groupoid
+groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is equivalent to the
+trivial groupoid with only one distinct object and one trivial automorphism. To
+avoid clutter we assume the groupoids are strict, e.g., that the groupoid
 $\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
@@
 condition that the commutative diagram in figure~\ref{fig:morita} is a
-\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps,
-and $\ot$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids
-defined previously.
-The intuition is that the codomain consists of three objects
-with associated morphisms, each of them counting as a ``third.'' The functor
-from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
+\emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps, and
+$\ot$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids defined
+previously.  The intuition is that the codomain consists of three objects with
+associated morphisms, each of them counting as a ``third.'' The functor from
+$\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
 include an arbitrary choice of ``which third'' to target and the universality
@@
 target and a \emph{global} condition ensuring that the choice ultimately
-annihilates. The challenge we are facing is to turn this idea into an operational
-semantics that somehow combines the local and global aspects of the
-equivalence. We will be inspired by the folklore result that ``substitution into a
-dependent type is a pullback''~\cite{subpullback} (but which turns out to have
+annihilates. The challenge we are facing is to turn this idea into an
+operational semantics that somehow combines the local and global aspects of the
+equivalence. We will be inspired by the folklore result that ``substitution into
+a dependent type is a pullback''~\cite{subpullback} (but which turns out to have
 been fully explained in~\cite{practicalfoundations}) as explained in the next
@@
 The last new syntactic category in $\Pi^/$ is that of values which deserves some
-additional discussion. When types denote sets, values of a type are clear:
-they are just the elements of the set.  In the case when types are groupoids,
-this is less clear, especially when the groupoid cardinality is a proper
-fraction.  What we get instead are equivalence classes of values.
-The idea is not uncommon: in the conventional $\lambda$-calculus, we list
-$\lambda x.x$ and $\lambda y.y$ as separate values of type
-$\tau \rightarrow \tau$ and then provide a separate equivalence relation
-($\alpha$-equivalence) to express the fact that these two values are
-indistinguishable. The treatment in our setting is similar: for iteration
-groupoids $\order{c}$, every distinct iterate $c^k$ is a listed as a separate
-value with the understanding that some iterates are
+additional discussion. When types denote sets, values of a type are clear: they
+are just the elements of the set.  In the case when types are groupoids, this is
+less clear, especially when the groupoid cardinality is a proper fraction.  What
+we get instead are equivalence classes of values.  The idea is not uncommon: in
+the conventional $\lambda$-calculus, we list $\lambda x.x$ and $\lambda y.y$ as
+separate values of type $\tau \rightarrow \tau$ and then provide a separate
+equivalence relation ($\alpha$-equivalence) to express the fact that these two
+values are indistinguishable. The treatment in our setting is similar: for
+iteration groupoids $\order{c}$, every distinct iterate $c^k$ is listed as a
+separate value with the understanding that some iterates are
 $\isotwo$-equivalent. Similarly for symmetry groupoids $\iorder{c}$, every
-morphism is listed as a value $1/c^k$ with the same understanding that that some
-of these morphisms are $\isotwo$-equivalent.
+morphism is listed as a value $1/c^k$ with the same understanding that some of
+these morphisms are $\isotwo$-equivalent.
 

### 2016-12-18 typespp.tex
move things around in sec 4
forall exists combined rule

@@
 
+To understand the challenge in designing an operational semantics for $\eta$ and
+$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
+Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
+manipulation on types:
+\[
+x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
+\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
+\]
+and by the coherence conditions of compact closed categories, we expect this
+circuit, when run in either direction, to be equivalent to the
+identity. Consider now the case of a combinator $c$ whose order is greater than
+or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
+$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
+$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
+$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
+either end. The problem becomes apparent when one observes that the uses of
+$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
+right-to-left execution must behave differently depending on the incoming
+value. But the ``local'' information available to either of these combinators
+does not include the ``global'' information about the incoming value and there
+is no way for them to make consistent guesses locally. It is possible to modify
+the operational semantics to use a mechanism to express a computational effect
+which enables spatially separated parts of the program to communicate in a way
+that is reminiscent of entanglement in quantum mechanics. Possibilities to
+realize this communication are global reference cells, backtracking, logical
+variables with unification. We present in the next section a more abstract
+approach that encodes the required dependency in dataflow constraints using
+dependent types.
+
 \begin{figure}[t]
@@
 \eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  \forall{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle] \\
-\epsilon-: & \exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k \rangle]
-  & \leftrightblackspoon & [\order{\idiso},\idiso] \\
+  \forall\exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k  \rangle]
+  & : \epsilon- \\
 \eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  \forall{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle] \\
-\epsilon+ : & \exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
-    & \leftrightblackspoon & [\order{\idiso},\idiso]
+  \forall\exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
+  & : \epsilon+
 \end{array}\]
@@
 \end{array}\]
-Each combinator $\rho_\bullet$ has an inverse.
+\\
+Each combinator $\rho_\bullet$ has an inverse. A rule $\mathbb{T_1}_\bullet \leftrightblackspoon
+\forall\exists{k}.\mathbb{T_2}_\bullet$ uses the $\forall$ in the left-to-right
+direction and uses the $\exists$ in the right-to-left direction.
 }
@@
 
-To understand the challenge in designing an operational semantics for $\eta$ and
-$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
-Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
-manipulation on types:
-\[
-x \mapsto (1 \times x) \mapsto ((x \times 1/x) \times x)
-\mapsto (x \times (1/x \times x)) \mapsto (x \times 1) \mapsto x
-\]
-and by the coherence conditions of compact closed categories, we expect this
-circuit, when run in either direction, to be equivalent to the
-identity. Consider now the case of a combinator $c$ whose order is greater than
-or equal to 2, i.e., the type $\order{c}$ has at least two distinct values
-$\idiso$ and $c$ and the type $\iorder{c}$ has at least two distinct values
-$1/\idiso$ and $1/c$. We require this circuit to produce $\idiso$ when given
-$\idiso$ from either end, and similarly to produce $c$ when given $c$ from
-either end. The problem becomes apparent when one observes that the uses of
-$\eta+$ in the left-to-right execution and the use of $\epsilon-$ in the
-right-to-left execution must behave differently depending on the incoming
-value. But the ``local'' information available to either of these combinators
-does not include the ``global'' information about the incoming value and there
-is no way for them to make consistent guesses locally. It is possible to modify
-the operational semantics to use a mechanism to express a computational effect
-which enables spatially separated parts of the program to communicate in a way
-that is reminiscent of entanglement in quantum mechanics. Possibilities to
-realize this communication are global reference cells, backtracking, logical
-variables with unification. We choose a more abstract approach that encodes the
-required dependency in dataflow constraints using dependent types.
+The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
+where each type $\mathbb{T}$ refers to a particular value $v$ in
+$\mathbb{T}$. We recall that, as all fractional types are inhabited (see remark
+in Sec.~\ref{sec:symmetryg}), and thus there is no loss in generality in going
+to pointed types. We then allow universal quantification over indices $k$
+associated to values, over a collection of types.  In particular, the type
+$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
+imposes a constraint on the two values associated with the types $\order{c}$ and
+$\iorder{c}$; they must be ``synchronized'' by sharing the same $k$ but are
+otherwise arbitrary. Once introduce a universal quantifier in one direction of
+execution, we are led to also include an existential quantifier for the reverse
+execution. Intuitively, a forward evaluation step that is at liberty to choose a
+$k$ will, when reversed encounter a particular, i.e., existentially quantified,
+$k$. The syntax is therefore as follows.
 
@@
 \end{definition}
-
-The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
-where each type $\mathbb{T}$ refers to a particular value $v$ in
-$\mathbb{T}$. We then allow universal quantification over indices $k$
-associated to values, over a collection of types.
-In particular, the type
-$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
-imposes a constraint on the two values associated with the types $\order{c}$ and
-$\iorder{c}$; they must be ``synchronized'' by sharing the same $k$. In
-general, a pointed type will be of the form
-$\forall\overline{k}.[\mathbb{T},v]$ for some sequence of integers
-$\overline{k}$ and some value $v$ in $\mathbb{T}$, which might of course refer
-to the integers of $\overline{k}$ to enforce some synchronization among its
-sub-values.  The resulting combinators are shown in Fig.~\ref{pifrac:comb}.
-We omit the quantifier from the notation when it is over the empty sequence
-integers.  We recall that, as all fractional types are inhabited (see remark in
-Sec.~\ref{sec:symmetryg}), and thus there is no loss in generality in going to
-pointed types.
 

### 2016-12-18 typespp.tex
clarified example

@@
 imposes a constraint on the two values associated with the types $\order{c}$ and
-$\iorder{c}$; they must be ``synchronized'' by sharing the same $k$ but are
-otherwise arbitrary. Once introduce a universal quantifier in one direction of
+$\iorder{c}$; they must be ``synchronized'' by sharing the same~$k$ but are
+otherwise arbitrary. Once we introduce a universal quantifier in one direction of
 execution, we are led to also include an existential quantifier for the reverse
@@
  (\textrm{/-combinators}) & \rho_\bullet &:&
-  \mathbb{T}_{\bullet 1} \leftrightblackspoon \mathbb{T}_{\bullet 2} ~::=
+  \mathbb{T_1}_{\bullet} \leftrightblackspoon \mathbb{T_2}_{\bullet} ~::=
   [\textit{see Fig.~\ref{pifracd:comb}}]
@@
 Here is how the circuit in Fig.~\ref{fig:zigzag} evaluates in the forward and
-backwards directions:
+backwards directions where we diambiguate the uses of $\forall$ and $\exists$
+for clarity.
+
 \[\begin{array}{rcl}
@@
 
-Need to explain that a $\forall$ in the forward direction is a choice and
-becomes $\exists$ in the backwards direction.
-
 Rules for moving quantifiers?

### 2016-12-18 typespp.tex
explicit quantifiers everywhere

@@
 $\triple{k}{q}{\alpha}$ in $\iter{p}$ such that $q$ annihilates $p^i$ to the
-identity. Note also that everything is well-defined even if we choose $p :
-\zt\iso\zt$. In that case, the cardinality is 1.
+identity.
+
+\paragraph*{Remark.} Note also that everything is well-defined even if we choose
+$p : \zt\iso\zt$. In that case, the cardinality is 1.
 
@@
 \[\begin{array}{rrcll}
-\unitetl/ :& [\order{\idiso} \circledast \mathbb{T},
+\unitetl/ :& \forall\exists{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
                                                    \langle \idiso,v \rangle]
   & \leftrightblackspoon &
-                   [\mathbb{T},v]
+                   \forall\exists{\overline{k}}.[\mathbb{T},v]
   &: \unititl/ \\
-\unitetr/ :& [\mathbb{T} \circledast \order{\idiso},
+\unitetr/ :& \forall\exists{\overline{k}}.[\mathbb{T} \circledast \order{\idiso},
                                                     \langle v,\idiso]
   & \leftrightblackspoon &
-                   [\mathbb{T},v]
+                   \forall\exists{\overline{k}}.[\mathbb{T},v]
   &: \unititr/ \\
-\swapt/ :& [\mathbb{T}_1 \circledast \mathbb{T}_2,
+\swapt/ :& \forall\exists{\overline{k}}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                    \langle v_1,v_2 \rangle]
  & \leftrightblackspoon &
-                  [\mathbb{T}_2 \circledast \mathbb{T}_1,
+                  \forall\exists{\overline{k}}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                   \langle v_2,v_1 \rangle]
   &: \swapt/ \\
-  \assoclt/ :& [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
+  \assoclt/ :& \forall\exists{\overline{k}}.
+               [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
                                         \langle v_1,\langle v_2,v_3 \rangle]
   & \leftrightblackspoon &
+                      \forall\exists{\overline{k}}.
                [(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
@@
 {}
-{\jdg{}{}{\idiso/ :[\mathbb{T},v] \leftrightblackspoon
- [\mathbb{T},v]}}
+{\jdg{}{}{\idiso/ : \forall\exists{\overline{k}}.[\mathbb{T},v] \leftrightblackspoon
+   \forall\exists{\overline{k}}.[\mathbb{T},v]}}
 {}
@@
 {\jdg{}{}{\rho_1 :
-  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
+  \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
+  \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2]} \quad
   \rho_2 :
-  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
+  \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2] \leftrightblackspoon
+  \forall\exists{\overline{k_3}}.[\mathbb{T}_3,v_3]}
 {\jdg{}{}{\rho_1 \circledcirc \rho_2 :
- [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
+ \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
+ \forall\exists{\overline{k_3}}.[\mathbb{T}_3},v_3]}
 {}
@@
 {\jdg{}{}{\rho_1 :
- [\mathbb{T}_1,v_1] \leftrightblackspoon
- [\mathbb{T}_2,v_2]}
+ \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
+ \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2]}
  \qquad \rho_2 :
- [\mathbb{T}_3,v_3] \leftrightblackspoon
- [\mathbb{T}_4,v_4]}
+ \forall\exists{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
+ \forall\exists{\overline{k_4}}.[\mathbb{T}_4,v_4]}
 {\jdg{}{}{\rho_1 \circledast \rho_2 :
-  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
+  \forall\exists{\overline{k_1k_3}}.[\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
   \leftrightblackspoon
-  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
+  \forall\exists{\overline{k_2k_4}}.[\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
 {}
@@
 where each type $\mathbb{T}$ refers to a particular value $v$ in
-$\mathbb{T}$. We recall that, as all fractional types are inhabited (see remark
+$\mathbb{T}$. We recall that, as all fractional types are inhabited (see Remark
 in Sec.~\ref{sec:symmetryg}), and thus there is no loss in generality in going
@@
 $\iorder{c}$; they must be ``synchronized'' by sharing the same~$k$ but are
-otherwise arbitrary. Once we introduce a universal quantifier in one direction of
-execution, we are led to also include an existential quantifier for the reverse
-execution. Intuitively, a forward evaluation step that is at liberty to choose a
-$k$ will, when reversed encounter a particular, i.e., existentially quantified,
-$k$. The syntax is therefore as follows.
+otherwise arbitrary. Once we introduce a universal quantifier in one direction
+of execution, we are led to also include an existential quantifier for the
+reverse execution. Intuitively, a forward evaluation step that is at liberty to
+choose a $k$ will, when reversed encounter a particular, i.e., existentially
+quantified, $k$. The syntax is therefore as follows.
 
@@
 
+A type $\mathbb{T}_\bullet$ is generally of the form
+$\forall\exists{\overline{k}}.[\mathbb{T},v]$ where all the quantifiers are
+lifted to the top leaving a non-dependent type and a value in that type. The
+combinators in Fig.~\ref{pifracd:comb} also serve as the operational semantics
+for the language as they explain how the value currently in focus changes during
+evaluation.
+
 Here is how the circuit in Fig.~\ref{fig:zigzag} evaluates in the forward and
@@
 [\order{c},c^i]
-&\stackrel{\unitetl/}{\longmapsto}&
+&\stackrel{\unitetl/}{\leftrightblackspoon}&
   [\order{id}\circledast\order{c}, \langle \idiso,c^i \rangle] \\
-&\stackrel{\eta+\circledast\idiso/}{\longmapsto}&
+&\stackrel{\eta+\circledast\idiso/}{\leftrightblackspoon}&
   \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
   \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
-&\stackrel{\assocrt/}{\longmapsto}&
+&\stackrel{\assocrt/}{\leftrightblackspoon}&
   \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
   \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
-%% &\stackrel{\synch_i}{\longmapsto}&
+%% &\stackrel{\synch_i}{\leftrightblackspoon}&
 &k\mapsto i&
@@
   \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
-&\stackrel{\idiso/\circledast\epsilon-}{\longmapsto}&
+&\stackrel{\idiso/\circledast\epsilon-}{\leftrightblackspoon}&
   [\order{c}\circledast\order{\idiso},
   \langle c^i, \idiso \rangle] \\
-&\stackrel{\unitetr/}{\longmapsto}&
+&\stackrel{\unitetr/}{\leftrightblackspoon}&
   [\order{c}, c^i]
@@
   [\order{c}, c^i]
-&\stackrel{\unitetr/}{\longmapsfrom}&
+&\stackrel{\unitetr/}{\leftrightblackspoon}&
   [\order{c}\circledast\order{\idiso},
   \langle c^i, \idiso \rangle] \\
-&\stackrel{\idiso/\circledast\epsilon-}{\longmapsfrom}&
+&\stackrel{\idiso/\circledast\epsilon-}{\leftrightblackspoon}&
   \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
   \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
-&\stackrel{\assocrt/}{\longmapsfrom}&
+&\stackrel{\assocrt/}{\leftrightblackspoon}&
   \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
@@
   \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
-&\stackrel{\eta+\circledast\idiso/}{\longmapsfrom}&
+&\stackrel{\eta+\circledast\idiso/}{\leftrightblackspoon}&
   [\order{\idiso}\circledast\order{c}, \langle \idiso,c^i\rangle] \\
-&\stackrel{\unitetl/}{\longmapsfrom}&
+&\stackrel{\unitetl/}{\leftrightblackspoon}&
   [\order{c},c^i]

### 2016-12-18 typespp.tex
did all updates to sec 4

@@
 \newcommand{\unitv}{()}
-\newcommand{\synch}{\mathsf{{synch}}}
+\newcommand{\synchl}{\mathsf{{synchl}}}
+\newcommand{\synchr}{\mathsf{{synchr}}}
+\newcommand{\packl}{\mathsf{{packl}}}
+\newcommand{\packr}{\mathsf{{packr}}}
 \newcommand{\unitepl}{\mathsf{{unite_+l}}}
@@
   \forall\exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
-  & : \epsilon+
+  & : \epsilon+ \\
+\synchl :& \forall\exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
+  \langle \langle c^k,1/c^k \rangle, c^i]
+  & \leftrightblackspoon &
+  [(\order{c}\circledast\iorder{c})\circledast\order{c},
+    \langle \langle c^i,1/c^i \rangle, c^i]
+  & : \packl \\
+\synchr :& \forall\exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
+  \langle c^i, \langle 1/c^k, c^k \rangle \rangle]
+  & \leftrightblackspoon &
+   [\order{c}\circledast(\iorder{c}\circledast\order{c}),
+   \langle c^i, \langle 1/c^i, c^i \rangle \rangle]
+  & : \packr
 \end{array}\]
@@
 {}
-% \\
-% \\
-% \Rule{}
-% {\jdg{}{}{}}
-% {\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
-%   \leftrightblackspoon
-%    \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
-% {}
 \end{array}\]
@@
 Each combinator $\rho_\bullet$ has an inverse. A rule $\mathbb{T_1}_\bullet \leftrightblackspoon
-\forall\exists{k}.\mathbb{T_2}_\bullet$ uses the $\forall$ in the left-to-right
-direction and uses the $\exists$ in the right-to-left direction.
+\forall\exists{k}.\mathbb{T_2}_\bullet$ introduces a $\forall$ in the left-to-right
+direction and eliminates an $\exists$ in the right-to-left direction. A rule
+$\forall\exists{k}.\mathbb{T_1}_\bullet \leftrightblackspoon
+\mathbb{T_2}_\bullet$ eliminates a $\forall$ in the left-to-right direction and
+introduces an $\exists$ in the right-to-left direction.
 }
@@
 for the language as they explain how the value currently in focus changes during
-evaluation.
-
-Here is how the circuit in Fig.~\ref{fig:zigzag} evaluates in the forward and
-backwards directions where we diambiguate the uses of $\forall$ and $\exists$
-for clarity.
-
+evaluation. Given the new combinators, we can modify the the circuit in
+Fig.~\ref{fig:zigzag} by adding explicit synchronization primitives. The
+evaluation proceeds as follows where we diambiguate the uses of $\forall$ and
+$\exists$ for clarity.
+
+Forward execution:
 \[\begin{array}{rcl}
@@
   \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
+&\stackrel{\synchl}{\leftrightblackspoon}&
+  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
 &\stackrel{\assocrt/}{\leftrightblackspoon}&
-  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
-  \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
-%% &\stackrel{\synch_i}{\leftrightblackspoon}&
-&k\mapsto i&
   [\order{c}\circledast(\iorder{c}\circledast\order{c}),
   \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
-&\exists&
+&\stackrel{\packr}{\leftrightblackspoon}&
   \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
@@
 
+Reverse execution:
 \[\begin{array}{rcl}
@@
   \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
-&\stackrel{\assocrt/}{\leftrightblackspoon}&
-  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
-  \langle \langle c^i,1/c^k \rangle,c^k \rangle] \\
-&k\mapsto i&
+&\stackrel{\synchr}{\leftrightblackspoon}&
  [(\order{c}\circledast\iorder{c})\circledast\order{c},
   \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
-&\exists&
+&\stackrel{\assocrt/}{\leftrightblackspoon}&
+  [(\order{c}\circledast\iorder{c})\circledast\order{c},
+  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
+&\stackrel{\packl}{\leftrightblackspoon}&
   \exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
@@
 \end{array}\]
-
-Rules for moving quantifiers?
-
-Combinators for packing (creating an existential) and synchronization
-(instantiating a universal)??
 

### 2016-12-18 typespp.tex
major revisions; sec 5 completely blank

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{$\Pi^/$}
+\section{$\Pi^/$: Syntax and Semantics}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{$\Pi^/$: Operational Semantics}
-
-at end of previous section; add a new subsection explaining functor showing
-equivalence of 1 and $\order{p} \times \iorder{p}$. Morita functor. Pullback condition. there is a
-choice for eta but there is an extra condition that the choices are all
-consistent. The big question is how to encode this global condition; side
-effect; synchronization; etc. so naturally lead to pointed types and revise types
-
-Only non-trivial pairs are created by eta; they create ``synchronized'' uses combinators $c^k$ and $1/c^k$.
-The rest of the combinators can only replace these combinators by $\isotwo$-equivalent ones.
-Assoc is subtle: what if we have $\langle \order{c_1} , \langle \iorder{c_2} , \order{c_2} \rangle \rangle$; naive assoc would allow us to create a pair with unrelated combinators which cannot be eliminated using epsilon.
-But if assoc is constrained to maintain the invariant that only ``synchronized'' uses of combinators are ever paired together, everything works great.
-
-Idea is to change the types to be of the form $\Sigma_{\overline{k}}\mathbb{T}$
-where we keep track of the values (pointed types). When we take a product we
-concatenate the two sequences of integers.
-\[
-(\textrm{Fractional Types})  \mathbb{T} ::=
-  \Sigma_k\order{c} \alt \Sigma_k\iorder{c} \alt
-  \Sigma_{\overline{j},\overline{k}}\mathbb{T}_1 \circledast \mathbb{T}_2
-\]
-
-The operational semantics for all the primitive combinators is a
-simple transliteration of Fig.~\ref{fig:opsem}. We omit the
-straightforward implementation of prim and its inverse
-below:
-
-The interesting combinators operationally are the unit/counit combinators $\eta$
-and $\epsilon$, and the synchronization combinators. Their semantics is defined
-by the following pair of mutually recursive interpreters:
-
-The unit combinators $\eta$ simply generate unit tangles, and since unit tangles
-are equivalent to the \emph{identity}, they can be freely eliminated by the
-counit combinators $\epsilon$. The two synchronization operations exchange the
-tangled product with an iterate. The two interpreters satisfy several properties
-described below: First they are congruences for the equivalence $≈$ on values.
-We can further prove that both are \emph{reversible}.  Furthermore, we can show
-two coherence conditions: first, that the reverse interpreter gives equivalent
-results to the forward interpreter applied to a reverse combinator, and second,
-that two equivalent combinators (as given by a 2-combinator) will evaluate to
-equivalent values.
-
-The operational definition of synchr⋆ might seem at odds with our claim that it is related to associativity while it looks like commutativity.
-The way to understand it is as follows: if indeed all values were independent, it would be associativity.
-However, in this case, $p // q$ is really two \emph{tangled} values, which we do not have access to.
-To keep these synchronized and yet to achieve the given type, the only choice (operationally) is to swap.
-This is where the ``action at a distance'' occurs.
-
-The following is an example usage of zig-zag, which uses NOT combinator.
+\section{$\Pi^/$: Additional Examples}
 
 ccc have a notion of higher-order functions, trace, etc. Talk about these examples.
-
-\begin{figure}[t]
-{\footnotesize
-\[\begin{array}{cc}
-\begin{array}[t]{rlcl}
-  \evalone{\unitetl/}{&\langle id^{k},\mathbb{V} \rangle} &=& \mathbb{V} \\
-  \evalone{\unititl/}{& \mathbb{V}} &=& \langle id^{k_1},\mathbb{V} \rangle \\
-  \evalone{\unitetr/}{&\langle \mathbb{V},id^{k} \rangle} &=& \mathbb{V} \\
-  \evalone{\unititr/}{& \mathbb{V}} &=& \langle \mathbb{V},id^{k} \rangle\\
-  \evalone{\swapt}{&\langle \mathbb{V}_1, \mathbb{V}_2 \rangle} &=& \langle \mathbb{V}_2, \mathbb{V}_1 \rangle \\
-  \evalone{\assoclt/}{&\langle c^k , \langle 1/c^{k}, c^{k} \rangle\rangle} &=& \langle \langle c^k , 1/c^{k} \rangle, c^{k} \rangle \\
-  \evalone{\assocrt/}{&\langle \langle c^k , 1/c^{k} \rangle, c^{k} \rangle} &=& \langle c^k , \langle 1/c^{k}, c^{k} \rangle\rangle \\
-\end{array}
-&
-\begin{array}[t]{rlcl}
-  \evalone{\eta-}{& id^{k_1}} &=& \langle 1/c^{k_2},c^{k_2} \rangle\\
-  \evalone{\eta+}{& id^{k_1}} &=& \langle c^{k_2},1/c^{k_2} \rangle\\
-  \evalone{\epsilon-}{& \langle 1/c^{k_1},c^{k_1} \rangle} &=& id^{k_2} \\
-  \evalone{\epsilon+}{& \langle c^{k_1},1/c^{k_1} \rangle} &=& id^{k_2} \\
-\end{array}
-
-\end{array}\]
-\caption{\label{fig:opsem-frac}$\Pi^/$ operational semantics}
-}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Limitations}
-
-Given any non-negative rational number, we can form a type whose cardinality is
-that number. And yet, our types do not capture the full structure of the
-non-negative rational numbers, as these form a commutative semifield. There are
-a number of operations and properties which need to be added to complete this.
-Most seem quite straightforward.  We can however single one out which may not
-be: what we are missing is a multiplicative inverse for every type, and not just
-$\order{p}$. In particular, we would like to form the type
-$\iorder{(\iorder{p})}$ and have it be equivalent to $\order{p}$.  The
-relationship between $\order{p}$ and $\iorder{p}$ in $\Pi^/$ reveals a pattern
-that can be exploited to achieve this generalization. The multiplicative inverse
-of $\order{p}$ is obtained by promoting the objects in $\order{p}$ to morphisms
-in $1/\hash p$, which is a general process called \emph{delooping}. It is clear
-we can go in the inverse direction by demoting morphisms to objects which is a
-process called \emph{looping}. We leave a possible extension with a general
-looping/delooping process to future work.
-
-A second more immediate concern is to discover the abstract monadic and/or
-comonadic structure underlying the operational semantics. It is clear that the
-operational interpretation of fractional types involves a computational effect
-that requires synchronization with the ``future''. Implementing this effect
-using backtracking, reference cells, or dataflow constraints is possible but
-each implementation choice is ad hoc. We conjecture that with the right monadic
-and/or comonadic abstraction, the operational semantics will be simpler and
-expose still richer properties.  We chose the dependently-typed dataflow
-semantics because we are able to prove that it is reversible and coherent,
-unlike our previous attempts with explicit effects, none of which were provably
-reversible and/or coherent.
 
@@
 
-We have presented a natural notion of \emph{fractional types} that
-enriches a class of reversible programming languages in several
-dimensions.  Further research might show how to use fractional
-types in a conventional (i.e., irreversible) programming language, their full
-potential is only achieved when the ambient language guarantees that
-no information is created or erased.
-
-The key semantic insight is that iterating a reversible program $p$ on
-a finite type must eventually reach the identity in $\ord{p}$
-steps. By being careful not to collapse proofs, each such reversible
-program has $\ord{p}$ distinct proofs of reversibility and hence gives
-rise to a groupoid with cardinality $\frac{1}{\ord{p}}$. Going from
-this observation to a full programming language required several
-difficult and subtle design choices which we have explored to produce
-$\Pi^/$. The latter language has a natural denotational semantics
-where types denote groupoids. Its operational semantics requires a
-mechanism to express a computational effect which enables spatially
-separated parts of the program to communicate in a way that is
-reminiscent of entanglement in quantum mechanics. It is possible to
-realize such an operational semantics using global reference cells,
-backtracking, or other conventional technique. A more enlightening and
-less ad hoc implementation encodes the required dependency in dataflow
-constraints encoded in dependent types. The key idea is to generalize
-the usual cartesian product to a \emph{tangled product} that allows the
-components to interact at synchronization points.
-
-Our fractional types extend the natural denotation of types from sets
-to non-trivial groupoids but they only scratch the surface of the
-tower of weak $\omega$-groupoids that is expressible in HoTT. A long
-term goal of our research is to find natural type constructors
-inspired by the rich combinatorial structure of weak
-$\omega$-groupoids and that provide novel programming abstractions.
-
-For an information-theoretic perspective on the language $\Pi$, we think of a
-type containing $N$ values as an abstract system that has~$N$ distinguishable
-states. According to the conventional theory of information~\cite{Shannon1948},
-the amount of information contained in each state of a system with $N$
-distinguishable states is $(\log N)$ bits of information. For example, the type
-$\mathbb{2}$ can be thought of as an abstract system with two distinguishable
-states labeled $\mathsf{true}$ and $\mathsf{false}$ each containing $\log 2 = 1$
-bit. Similarly, the type $\mathbb{3}$ can be thought of as an abstract system
-with three distinguishable states each containing $\log 3$ bits. The logarithmic
-map implies that information contained in a composite state is the sum of the
-information contained in its constituents. For example, the type $\mathbb{2}
-\otimes \mathbb{3}$ can be thought of a composite system consisting of two
-independent unrelated subsystems. Each state of the composite system therefore
-contains $\log (2 * 3) = \log 2 + \log 3$ bits which is the sum of the
-information contained in each subsystem. Since all the combinators preserve
-cardinality, they are also information-preserving.
+We have presented a natural notion of \emph{fractional types} that enriches a
+class of reversible programming languages in several dimensions.  Further
+research might show how to use fractional types in a conventional (i.e.,
+irreversible) programming language, their full potential is only achieved when
+the ambient language guarantees that no information is created or erased. The
+key semantic insight is that iterating a reversible program $p$ on a finite type
+must eventually reach the identity in $\ord{p}$ steps. By being careful not to
+collapse proofs, each such reversible program has $\ord{p}$ distinct proofs of
+reversibility and hence gives rise to a groupoid with cardinality
+$\frac{1}{\ord{p}}$. Going from this observation to a full programming language
+required several difficult and subtle design choices which we have explored to
+produce $\Pi^/$. The latter language has a natural denotational semantics where
+types denote groupoids. Its operational semantics requires a mechanism to
+express a computational effect which enables spatially separated parts of the
+program to communicate in a way that is reminiscent of entanglement in quantum
+mechanics. It appears possible to realize such an operational semantics using
+global reference cells, backtracking, or other conventional technique.
+
+Our fractional types extend the natural denotation of types from sets to
+non-trivial groupoids but they only scratch the surface of the tower of weak
+$\omega$-groupoids that is expressible in HoTT. A long term goal of our research
+is to find natural type constructors inspired by the rich combinatorial
+structure of weak $\omega$-groupoids and that provide novel programming
+abstractions. We conclude by outlining several interesting connections and
+potential avenues for future work.
+
+\paragraph*{Quotient Types.} Certain groupoids naturally correspond to
+conventional \emph{quotient types}. Traditionally~\cite{quotient}, a quotient
+type $\ag{T}{E}$ combines a type $T$ with an equivalence relation $E$ that
+serves as the equality relation on the elements of $T$. Our notion of fractional
+types in $\pifrac$ appears to subsume conventional quotient types and their
+applications~\cite{Cohen2013} (e.g., defining fractions, multivariate
+polynomials, field extensions, algebraic numbers, etc.) and it would interesting
+to explicitly formalize this connection.
+
+\paragraph*{Conservation of Information and Negative Entropy.} According to the
+conventional theory of communication~\cite{Shannon1948}, a type with~$N$ values
+is viewed as an abstract system that has~$N$ distinguishable states where the
+amount of information contained in each state is $(\log{N})$. This entropy is a
+measure of information which materializes itself in memory or bandwidth
+requirements when storing or transmitting elements of this type. Thus a type
+with 8 elements needs 3 bits of memory for storage or 3 bits of bandwidth for
+communication. The logarithmic map implies that information contained in a
+composite state is the sum of the information contained in its constituents. For
+example, the type $\mathbb{2} \otimes \mathbb{3}$ can be thought of a composite
+system consisting of two independent unrelated subsystems. Each state of the
+composite system therefore contains $\log (2 * 3) = \log 2 + \log 3$ bits which
+is the sum of the information contained in each subsystem. If quantum field
+theory is correct (as it so far seems to be) then information, during any
+physical process, is neither created nor
+destroyed. Landauer~\cite{Landauer:1961,bennett1985fundamental,Landauer},
+Bennet~\cite{Bennett:1973:LRC,bennett2003notes,bennett2010notes},
+Fredkin~\cite{fredkin1982conservative} and others made compelling arguments that
+this physical principle induces a corresponding computational principle of
+``conservation of information.'' In the context of finite types, generated from
+the empty type $\zt$, the unit type $\ot$, and sums and products $\oplus$ and
+$\otimes$, this principle states that the foundational (i.e., physical) notion
+of computation is computation via type
+isomorphisms~\cite{James:2012:IE:2103656.2103667} or type
+equivalences~\cite{Carette2016}, which are both sound and complete with respect
+to cardinality-preserving maps. The introduction, in $\pifrac$, of types
+(groupoids) with fractional cardinalities introduces types with \emph{negative
+  entropy}. For example, a type with cardinality $\frac{1}{8}$ has entropy
+$\log{\frac{1}{8}} = -3$. It is natural to interpret this negative entropy just
+like we interpret ``negative money,'' as a debt to be repaid by some other part
+of the system. In fact, the zig-zag example can be viewed as modeling a credit
+card transaction where the money is produced at the output site by generating a
+corresponding debt that is reconciled at the input site.
+
+\paragraph*{Correspondence with Commutative Semifields.} Computations over
+finite types naturally emerge from viewing types as syntax for semiring
+elements, semiring identities as type isomorphisms, and justifications for
+semiring identities as program transformations and
+optimizations~\cite{Carette2016}. This correspondence provides a rich
+proof-relevant version of the Curry-Howard correspondence between algebra and
+reversible programming languages. The addition of full fractional types to the
+mix would enrich the correspondence to commutative semifields, providing a
+categorification~\cite{math/9802029} of the non-negative rational numbers in a
+computational setting. This correspondence in $\Pi^/$ is still lacking,
+however. Given any non-negative rational number, we can form a type whose
+cardinality is that number. And yet, our types do not capture the full structure
+of the non-negative rational numbers, as these form a commutative
+semifield. There are a number of operations and properties which need to be
+added to complete this.  Most seem quite straightforward.  We can however single
+one out which may not be: what we are missing is a multiplicative inverse for
+every type, and not just $\order{p}$. In particular, we would like to form the
+type $\iorder{(\iorder{p})}$ and have it be equivalent to $\order{p}$.  As
+explained in Sec.~\ref{sec:groupoids}, this would require a futher
+generalization of looping and delooping that can be iterated. We leave a
+possible extension with a general looping/delooping process to future work.
 

### 2016-12-19 typespp.tex
bibtex
Add dual of (#α)ₖ rule
tweak one sentence

@@
 non-trivial groupoids and new programs that manipulate such types. Sec.~6
-presents the operational semantics of the extended language. The last two
-sections put our work in perspective and conclude.
+presents the operational semantics of the extended language. The last section
+puts our work in perspective and concludes.
 
@@
 where each type $\mathbb{T}$ refers to a particular value $v$ in
-$\mathbb{T}$. We recall that, as all fractional types are inhabited (see Remark
-in Sec.~\ref{sec:symmetryg}), and thus there is no loss in generality in going
+$\mathbb{T}$. As all fractional types are inhabited (see Remark
+in Sec.~\ref{sec:symmetryg}), there is no loss in generality in going
 to pointed types. We then allow universal quantification over indices $k$

### 2016-12-19 typespp.tex
adjust introduction to fit the paper.

@@
 
-In Homotopy Type Theory (HoTT)~\cite{hottbook}, types have the
+In modern treatments of type theory, types have the
 structure of \emph{weak $\omega$-groupoids}. As a first approximation,
@@
 
-Both groupoids involve some notion of ``division'' that can be captured at the
-level of types using a syntactic notion of \emph{fractional types}. Our aim is
-to explore this notion of types and its associated operational semantics. The
-remainder of the paper is organized as follows. We start by reviewing the
-necessary background consisting of the language $\Pi$ for programming with
-isomorphisms or equivalences in a reversible information-preserving way. Sec.~4
-explains the main novel semantic ideas of using $\Pi$ programs to generate
-non-trivial groupoids with fractional cardinality. Sec.~5 translates the
-semantic ideas into an extension of $\Pi$ with new type constructors denoting
-non-trivial groupoids and new programs that manipulate such types. Sec.~6
-presents the operational semantics of the extended language. The last section
-puts our work in perspective and concludes.
+Both groupoids involve some notion of (semantic) ``equivalence'', which we
+would like to capture as first-class entities in a programming language.
+In other words, we would like to have some syntactic notion of a type, whose
+denotation would be a groupoid with fractional cardinality.  In other words,
+a syntactic notion of \emph{fractional types}. Our aim then is to create a
+language which possesses ``fractional types'', explore this notion of types,
+and its associated operational semantics.
+
+The remainder of the paper is organized as follows. We start by reviewing
+necessary background material, consisting of the language $\Pi$ for programming
+with isomorphisms or equivalences in a reversible information-preserving way.
+Sec.~4 explains the main novel semantic ideas of using $\Pi$ programs to
+generate non-trivial groupoids with fractional cardinality, as well as
+translating the semantic ideas into an extension of $\Pi$ with new type
+constructors denoting non-trivial groupoids and new programs that manipulate
+such types.  Sec.~5 presents some example programs.  The last section puts our
+work in perspective and concludes.
 
@@
 {}
-~~
+\\
+\\
 \Rule{}
@@
  [\order{c_1},c_1^k]  \leftrightblackspoon
- [\order{c_2}.c_2^k]}}
+ [\order{c_2},c_2^k]}}
+{}
+~~
+\Rule{}
+{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
+{\jdg{}{}{(\iorder{\alpha})_k :
+ [\iorder{c_1},1/c_1^k]  \leftrightblackspoon
+ [\iorder{c_2},1/c_2^k]}}
 {}

### 2016-12-19 typespp.tex
edit conclusion (and insert HoTT reference earlier)

@@
 \noindent Both classes of programs, 1-combinators $c$, and
-2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT sense. The
+2-combinators~$\alpha$, denote \emph{equivalences} in the Homotopy Type
+Theory (HoTT) sense~\cite{hottbook}. The
 elements $c$ of 1-combinators denote type isomorphisms. The elements $\alpha$ of
@@
 We have presented a natural notion of \emph{fractional types} that enriches a
-class of reversible programming languages in several dimensions.  Further
+class of reversible programming languages in several dimensions.  While further
 research might show how to use fractional types in a conventional (i.e.,
@@
 required several difficult and subtle design choices which we have explored to
-produce $\Pi^/$. The latter language has a natural denotational semantics where
+produce $\Pi^/$. As with most functional languages, it has 
+a natural denotational semantics where
 types denote groupoids. Its operational semantics requires a mechanism to
@@
 mechanics. It appears possible to realize such an operational semantics using
-global reference cells, backtracking, or other conventional technique.
-
-Our fractional types extend the natural denotation of types from sets to
-non-trivial groupoids but they only scratch the surface of the tower of weak
-$\omega$-groupoids that is expressible in HoTT. A long term goal of our research
+global reference cells, backtracking, or other conventional technique.  Rather
+than going via effects, we chose dependent types to track the required
+constraints via dataflow.  Of our various attempts at an operational semantics,
+this one turned out simplest.
+
+Our fractional types have natural denotations which are non-trivial groupoids,
+but they only scratch the surface of the tower of weak $\omega$-groupoids that
+is expressible in HoTT~\cite{hottbook}. A long term goal of our research
 is to find natural type constructors inspired by the rich combinatorial
-structure of weak $\omega$-groupoids and that provide novel programming
+structure of weak $\omega$-groupoids that provide novel programming
 abstractions. We conclude by outlining several interesting connections and
@@
 requirements when storing or transmitting elements of this type. Thus a type
-with 8 elements needs 3 bits of memory for storage or 3 bits of bandwidth for
-communication. The logarithmic map implies that information contained in a
+with $8$ elements needs $3$ bits of memory for storage or $3$ bits of bandwidth
+for communication. The logarithmic map implies that information contained in a
 composite state is the sum of the information contained in its constituents. For
@@
 isomorphisms~\cite{James:2012:IE:2103656.2103667} or type
-equivalences~\cite{Carette2016}, which are both sound and complete with respect
+equivalences~\cite{Carette2016}; these are both sound and complete with respect
 to cardinality-preserving maps. The introduction, in $\pifrac$, of types

### 2016-12-19 typespp.tex
fig:cong
Circuits for name and coname
Some more on name and coname

@@
 
-ccc have a notion of higher-order functions, trace, etc. Talk about these examples.
+ccc have a notion of higher-order functions, trace, etc. Talk about these
+examples.
+
+\subsection{\texttt{name} and \texttt{coname}}
+
+Given a \textrm{/-combinator} $p : \#c \leftrightblackspoon \#d$ which is a
+morphism in the compact closed category, we can build circuits that construct
+the \texttt{name} $\ulcorner{p}\urcorner$ and \texttt{coname}
+$\llcorner{p}\lrcorner$ of $p$. These are shown in figure~\ref{fig:name-coname}.
+
+\begin{figure}[ht]
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+      \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{[\#id,id]}}$}} (0,1);
+
+      \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+      \node at (0.5,1) {\footnotesize{$\eta -$}};
+
+      \path [highlight] (1,1.75) edge node[above] {} (2,1.75);
+      \path (1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (2,1.75);
+      \path [highlight] (1,0.25) edge node[above] {} (2,0.25);
+      \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (2,0.25);
+      \draw[dashed] (1.25,1.75) -- (1.25,0.25);
+
+      \draw (2,0) -- (3,0) -- (3,2) -- (2,2) -- cycle;
+      \node at (2.5,1) {\footnotesize{$\texttt{cong}$}};
+
+      \path (3,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.75);
+      \path (3,0.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (4,0.25);
+
+      \path (2.5,0) edge node[left] {\footnotesize{$p : \#c \leftrightblackspoon \#d$}} (2.5,-0.5);
+    \end{tikzpicture}
+  \end{subfigure}
+  \begin{subfigure}{.5\textwidth}
+    \centering
+    \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+
+      \path (-1.1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (0,1.75);
+      \path (-1.1,0.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (0,0.25);
+
+      \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+      \node at (0.5,1) {\footnotesize{\texttt{cong}}};
+
+      \path (0.5,0) edge node[right] {\footnotesize{$p : \#c \leftrightblackspoon \#d$}} (0.5,-0.5);
+
+      \path [highlight] (1,1.75) edge node[above] {} (2,1.75);
+      \path (1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (2,1.75);
+      \path [highlight] (1,0.25) edge node[above] {} (2,0.25);
+      \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (2,0.25);
+      \draw[dashed] (1.75,1.75) -- (1.75,0.25);
+
+      \draw (2,0) -- (3,0) -- (3,2) -- (2,2) -- cycle;
+      \node at (2.5,1) {\footnotesize{$\epsilon +$}};
+
+      \path (3,1) edge node[above] {\footnotesize{${\textsf{[\#id,id]}}$}} (4,1);
+    \end{tikzpicture}
+  \end{subfigure}
+  \caption{\label{fig:name-coname} \texttt{name} and \texttt{coname}}
+\end{figure}
 

### 2016-12-19 typespp.tex
(1/#α)ₖ rule is derivable from dual and inverse
fig:dual
About cong

@@
 {}
-\\
-\\
+~~
 \Rule{}
@@
  [\order{c_2},c_2^k]}}
-{}
-~~
-\Rule{}
-{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
-{\jdg{}{}{(\iorder{\alpha})_k :
- [\iorder{c_1},1/c_1^k]  \leftrightblackspoon
- [\iorder{c_2},1/c_2^k]}}
 {}
@@
 
+\begin{lemma}
+  \label{lem:cong}
+  For $\mathbb{v} : \mathbb{T}_1 \circledast \mathbb{T}_2$,
+  $p : \mathbb{T}_1 \leftrightblackspoon \mathbb{T}$, we have
+  $\texttt{cong}^{l}_{p}(\mathbb{v}) : \mathbb{T} \circledast \mathbb{T}_2$, and
+  for $q : \mathbb{T}_2 \leftrightblackspoon \mathbb{T}$, we have
+  $\texttt{cong}^{r}_{q}(\mathbb{v}) : \mathbb{T}_1 \circledast \mathbb{T}$.
+  \begin{proof}
+    It follows from the fact that $\circledast$ is an endo-bi-functor.
+  \end{proof}
+\end{lemma}
+
 \subsection{\texttt{name} and \texttt{coname}}
@@
       \draw (2,0) -- (3,0) -- (3,2) -- (2,2) -- cycle;
-      \node at (2.5,1) {\footnotesize{$\texttt{cong}$}};
+      \node at (2.5,1) {\footnotesize{$\texttt{cong}^{r}$}};
 
@@
       \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
-      \node at (0.5,1) {\footnotesize{\texttt{cong}}};
+      \node at (0.5,1) {\footnotesize{$\texttt{cong}^{r}$}};
 
@@
 
+\begin{figure}
+    \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+      \path (-1,1) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (0,1);
+
+      \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+      \node at (0.5,1) {\footnotesize{$\unitetl/$}};
+
+      \path (1,1.75) edge node[above] {\footnotesize{$\#id$}} (2,1.75);
+      \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#c}}$}} (4,0.25);
+
+      \draw (2,0.75) -- (3,0.75) -- (3,2.75) -- (2,2.75) -- cycle;
+      \node at (2.5,1.75) {\footnotesize{$\eta +$}};
+
+      \path (3,2.25) edge node[above] {\footnotesize{$1/\textsf{\#d}$}} (6,2.25);
+      \path (3,1.25) edge node[above] {\footnotesize{$\textsf{\#d}$}} (4,1.25);
+
+      \draw (4,-0.25) -- (5,-0.25) -- (5,1.75) -- (4,1.75) -- cycle;
+      \node at (4.5,0.75) {\footnotesize{\texttt{cong}}};
+      \path (4.5,-0.25) edge node[right] {\footnotesize{$p : 1/\#c \leftrightblackspoon 1/\#d$}} (4.5,-0.75);
+
+      \path (5,0.25) edge node[above] {\footnotesize{$1/\textsf{\#d}$}} (6,0.25);
+      \path (5,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,1.25);
+
+      \draw (6,0) -- (7,0) -- (7,2.75) -- (6,2.75) -- cycle;
+      \node at (6.5,1.375) {\footnotesize{$\synchr$}};
+
+      \path (7,2.25) edge node[above] {\footnotesize{$1/{\textsf{\#d}}$}} (8,2.25);
+
+      \path (7,0.25) edge node[above] {\footnotesize{$1/{\textsf{\#d}}$}} (10,0.25);
+      \path (7,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (8,1.25);
+
+      \draw (8,0.75) -- (9,0.75) -- (9,2.75) -- (8,2.75) -- cycle;
+      \node at (8.5,1.75) {\footnotesize{$\epsilon +$}};
+
+      \path (9,1.75) edge node[above] {\footnotesize{$\#id$}} (10,1.75);
+
+      \draw (10,0) -- (11,0) -- (11,2) -- (10,2) -- cycle;
+      \node at (10.5,1) {\footnotesize{$\unititl/$}};
+
+      \path (11,1) edge node[above] {\footnotesize{$1/{\textsf{\#d}}$}} (12,1);
+    \end{tikzpicture}
+    \caption{\label{fig:dual} Dual}
+\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-12-19 typespp.tex
fig:dual
Make figure consistent
Make dual circuit consistent

--- 2016-12-19 typespp.tex
@@
     \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
-      \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{[\#id,id]}}$}} (0,1);
+      \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (0,1);
 
@@
       \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
-      \node at (0.5,1) {\footnotesize{\texttt{cong}}};
-
-      \path (0.5,0) edge node[right] {\footnotesize{$p : \#c \leftrightblackspoon \#d$}} (0.5,-0.5);
+      \node at (0.5,1) {\footnotesize{$\texttt{cong}^{r}$}};
+
+      \path (0.5,0) edge node[right] {\footnotesize{$p^{-1} : \#d \leftrightblackspoon \#c$}} (0.5,-0.5);
 
@@
 
-      \path (3,1) edge node[above] {\footnotesize{${\textsf{[\#id,id]}}$}} (4,1);
+      \path (3,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (4,1);
     \end{tikzpicture}
@@
 
+\begin{figure}
+  \centering
+  \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+    \path (-1,1) edge node[above] {\footnotesize{$1/{\textsf{\#d}}$}} (0,1);
+
+    \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+    \node at (0.5,1) {\footnotesize{$\unitetl/$}};
+
+    \path (1,1.75) edge node[above] {\footnotesize{${\#id}$}} (2,1.75);
+    \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (4,0.25);
+
+    \draw (2,0.75) -- (3,0.75) -- (3,2.75) -- (2,2.75) -- cycle;
+    \node at (2.5,1.75) {\footnotesize{$\eta +$}};
+
+    \path (3,2.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (8,2.25);
+    \path (3,1.25) edge node[above] {\footnotesize{$\textsf{\#c}$}} (4,1.25);
+
+    \draw (4,-0.25) -- (5,-0.25) -- (5,1.75) -- (4,1.75) -- cycle;
+    \node at (4.5,0.75) {\footnotesize{$\texttt{cong}^{l}$}};
+    \path (4.5,-0.25) edge node[right] {\footnotesize{$p : \#c \leftrightblackspoon \#d$}} (4.5,-0.75);
+
+    \path (5,0.25) edge node[above] {\footnotesize{$1/\textsf{\#d}$}} (6,0.25);
+    \path (5,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,1.25);
+
+    \draw (6,-0.25) -- (7,-0.25) -- (7,1.75) -- (6,1.75) -- cycle;
+    \node at (6.5,0.75) {\footnotesize{$\epsilon +$}};
+
+    \path (7,0.75) edge node[above] {\footnotesize{${\#id}$}} (8,0.75);
+
+    \draw (8,0.5) -- (9,0.5) -- (9,2.5) -- (8,2.5) -- cycle;
+    \node at (8.5,1.5) {\footnotesize{$\unititl/$}};
+
+    \path (9,1.5) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (10,1.5);
+  \end{tikzpicture}
+  \caption{\label{fig:dual} Dual}
+\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-12-19 typespp.tex
Add some text for dual

@@
 
+\subsection{\texttt{dual}}
+
+Given a \textrm{/-combinator} $p : \#c \leftrightblackspoon \#d$ which is a
+morphism in the compact closed category, we have a dual morphism
+$p^{\ast} : 1/\#d \leftrightblackspoon 1/\#c$. This is shown by building a
+reversible circuit in figure~\ref{fig:dual}.
+
+\begin{figure}[htb]
+  \centering
+  \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+    \path (-1,1) edge node[above] {\footnotesize{$1/{\textsf{\#d}}$}} (0,1);
+
+    \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+    \node at (0.5,1) {\footnotesize{$\unitetl/$}};
+
+    \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
+    \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (4,0.25);
+
+    \draw (2,0.75) -- (3,0.75) -- (3,2.75) -- (2,2.75) -- cycle;
+    \node at (2.5,1.75) {\footnotesize{$\eta +$}};
+
+    \draw[dashed] (3.2,1.25) -- (3.2,2.25);
+    \path [highlight] (3,2.25) edge node[above] {} (8,2.25);
+    \path (3,2.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (8,2.25);
+    \path [highlight] (3,1.25) edge node[above] {} (4,1.25);
+    \path (3,1.25) edge node[above] {\footnotesize{$\textsf{\#c}$}} (4,1.25);
+
+    \draw (4,-0.25) -- (5,-0.25) -- (5,1.75) -- (4,1.75) -- cycle;
+    \node at (4.5,0.75) {\footnotesize{$\texttt{cong}^{l}$}};
+    \path (4.5,-0.25) edge node[left] {\footnotesize{$p : \#c \leftrightblackspoon \#d$}} (4.5,-0.75);
+
+    \draw[dashed] (5.8,0.25) -- (5.8,1.25);
+    \path [highlight] (5,0.25) edge node[above] {} (6,0.25);
+    \path (5,0.25) edge node[above] {\footnotesize{$1/\textsf{\#d}$}} (6,0.25);
+    \path [highlight] (5,1.25) edge node[above] {} (6,1.25);
+    \path (5,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,1.25);
+
+    \draw (6,-0.25) -- (7,-0.25) -- (7,1.75) -- (6,1.75) -- cycle;
+    \node at (6.5,0.75) {\footnotesize{$\epsilon +$}};
+
+    \path (7,0.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (8,0.75);
+
+    \draw (8,0.5) -- (9,0.5) -- (9,2.5) -- (8,2.5) -- cycle;
+    \node at (8.5,1.5) {\footnotesize{$\unititl/$}};
+
+    \path (9,1.5) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (10,1.5);
+  \end{tikzpicture}
+  \caption{\label{fig:dual} Dual}
+\end{figure}
+
 \subsection{\texttt{name} and \texttt{coname}}
 
-Given a \textrm{/-combinator} $p : \#c \leftrightblackspoon \#d$ which is a
-morphism in the compact closed category, we can build circuits that construct
-the \texttt{name} $\ulcorner{p}\urcorner$ and \texttt{coname}
-$\llcorner{p}\lrcorner$ of $p$. These are shown in figure~\ref{fig:name-coname}.
-
-\begin{figure}[ht]
+We also show circuits that construct the \texttt{name} $\ulcorner{p}\urcorner$
+and \texttt{coname} $\llcorner{p}\lrcorner$ of $p$, in
+figure~\ref{fig:name-coname}.
+
+\begin{figure}[htb]
   \begin{subfigure}{.5\textwidth}
@@
 
-\begin{figure}
-  \centering
-  \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
-    \path (-1,1) edge node[above] {\footnotesize{$1/{\textsf{\#d}}$}} (0,1);
-
-    \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
-    \node at (0.5,1) {\footnotesize{$\unitetl/$}};
-
-    \path (1,1.75) edge node[above] {\footnotesize{${\#id}$}} (2,1.75);
-    \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (4,0.25);
-
-    \draw (2,0.75) -- (3,0.75) -- (3,2.75) -- (2,2.75) -- cycle;
-    \node at (2.5,1.75) {\footnotesize{$\eta +$}};
-
-    \path (3,2.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (8,2.25);
-    \path (3,1.25) edge node[above] {\footnotesize{$\textsf{\#c}$}} (4,1.25);
-
-    \draw (4,-0.25) -- (5,-0.25) -- (5,1.75) -- (4,1.75) -- cycle;
-    \node at (4.5,0.75) {\footnotesize{$\texttt{cong}^{l}$}};
-    \path (4.5,-0.25) edge node[right] {\footnotesize{$p : \#c \leftrightblackspoon \#d$}} (4.5,-0.75);
-
-    \path (5,0.25) edge node[above] {\footnotesize{$1/\textsf{\#d}$}} (6,0.25);
-    \path (5,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,1.25);
-
-    \draw (6,-0.25) -- (7,-0.25) -- (7,1.75) -- (6,1.75) -- cycle;
-    \node at (6.5,0.75) {\footnotesize{$\epsilon +$}};
-
-    \path (7,0.75) edge node[above] {\footnotesize{${\#id}$}} (8,0.75);
-
-    \draw (8,0.5) -- (9,0.5) -- (9,2.5) -- (8,2.5) -- cycle;
-    \node at (8.5,1.5) {\footnotesize{$\unititl/$}};
-
-    \path (9,1.5) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (10,1.5);
-  \end{tikzpicture}
-  \caption{\label{fig:dual} Dual}
-\end{figure}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2016-12-19 typespp.tex
Minor copy-editing

@@
 1-combinator which also has cardinality 1. The groupoids on either side are
-however not categorically equivalent and hence $\simeq$ cannot be categorical
-equivalence.  There are however other notions of groupoid equivalence, such as
-Morita equivalence~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that
-identify the two groupoids above.  Although the notion of equivalence we use in
-the next section appears novel, it is related ``in spirit'' to Morita
-equivalence and we find it useful to give a small example. We will verify that
-the product of the groupoid $\order{\permfive}$ of cardinality 3 and the
-groupoid $\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is equivalent to the
+however not isomorphic, there do not exist full and faithful functors between
+them, and hence $\simeq$ cannot be categorical equivalence.
+
+There are however weaker notions of groupoid equivalence, such as \emph{Morita
+  equivalence}~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that identify
+the two groupoids above.  Although the notion of equivalence we use in the next
+section appears novel, it is related ``in spirit'' to Morita equivalence and we
+find it useful to give a small example. We will verify that the product of the
+groupoid $\order{\permfive}$ of cardinality 3 and the groupoid
+$\iorder{\permfive}$ of cardinality $\frac{1}{3}$ is Morita equivalent to the
 trivial groupoid with only one distinct object and one trivial automorphism. To
-avoid clutter we assume the groupoids are strict, e.g., that the groupoid
+avoid clutter, we assume the groupoids are strict, e.g., that the groupoid
 $\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
-elements). The definition of \emph{Morita equivalence} in this case reduces to
-to finding a \emph{Morita morphism} from $\order{\idiso}$ to
+elements). Getting an equivalence in this case reduces to finding a \emph{Morita
+  morphism} from $\order{\idiso}$ to
 $\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
@@
 condition of the pullback ensures that the choice is ultimately insignificant as
-it leads to the same result as every other possible functor. To summarize the
-equivalence between the two groupoids requires a \emph{local} map that chooses a
-target and a \emph{global} condition ensuring that the choice ultimately
-annihilates. The challenge we are facing is to turn this idea into an
+it leads to the same result as every other possible functor.
+
+To summarize, the equivalence between the two groupoids requires a \emph{local}
+map that chooses a target and a \emph{global} condition ensuring that the choice
+ultimately annihilates. The challenge we are facing is to turn this idea into an
 operational semantics that somehow combines the local and global aspects of the
-equivalence. We will be inspired by the folklore result that ``substitution into
-a dependent type is a pullback''~\cite{subpullback} (but which turns out to have
-been fully explained in~\cite{practicalfoundations}) as explained in the next
-section.
+equivalence. We will be inspired by the folklore result in categorical semantics
+of dependent type theory, that ``substitution into a dependent type is a
+pullback''~\cite{subpullback}, (but which turns out to have been fully explained
+in~\cite{practicalfoundations}) as explained in the next section.
 

### 2016-12-19 typespp.tex
spell check and clarification about quantifiers

@@
 \begin{definition}[Iterated Power of a 1-combinator]
-  The $k^{th}$ iterated power of a 1-combinator $p : \tau \iso \tau$, for
+  The $k^{\mbox{th}}$ iterated power of a 1-combinator $p : \tau \iso \tau$, for
   $k \in \Z$ is
@@
 \mathbb{T_2}_\bullet$ eliminates a $\forall$ in the left-to-right direction and
-introduces an $\exists$ in the right-to-left direction.
+introduces an $\exists$ in the right-to-left direction. A rule
+$\forall\exists{k}.\mathbb{T_1}_\bullet \leftrightblackspoon
+\forall\exists{k}.\mathbb{T_2}_\bullet$ maintains (think of elimination then
+introduction) the universal quantifier in the left-to-right direction and
+the existential quantifier in the right-to-left direction. We omit the
+quantifiers when the sequence of bound variables is empty.
 }
@@
 Fig.~\ref{fig:zigzag} by adding explicit synchronization primitives. The
-evaluation proceeds as follows where we diambiguate the uses of $\forall$ and
+evaluation proceeds as follows where we disambiguate the uses of $\forall$ and
 $\exists$ for clarity.
@@
 type $\iorder{(\iorder{p})}$ and have it be equivalent to $\order{p}$.  As
-explained in Sec.~\ref{sec:groupoids}, this would require a futher
+explained in Sec.~\ref{sec:groupoids}, this would require a further
 generalization of looping and delooping that can be iterated. We leave a

### 2016-12-19 typespp.tex
tweaks to intro

@@
 
-Both groupoids involve some notion of (semantic) ``equivalence'', which we
-would like to capture as first-class entities in a programming language.
-In other words, we would like to have some syntactic notion of a type, whose
-denotation would be a groupoid with fractional cardinality.  In other words,
-a syntactic notion of \emph{fractional types}. Our aim then is to create a
-language which possesses ``fractional types'', explore this notion of types,
-and its associated operational semantics.
+Both groupoids involve some notion of (semantic) ``equivalence'', which we would
+like to capture as first-class entities in a programming language.  In other
+words, we would like to have some syntactic notion of a type, a \emph{fractional
+  type}, whose denotation would be a groupoid with fractional cardinality.  Our
+aim then is to create a language which possesses these fractional types, explore
+this notion of types, their equivalence, and their associated operational semantics.
 
@@
 constructors denoting non-trivial groupoids and new programs that manipulate
-such types.  Sec.~5 presents some example programs.  The last section puts our
-work in perspective and concludes.
+such types.  Sec.~5 presents some example programs to illustrate the
+expressiveness of the language.  The last section puts our work in perspective
+and concludes.
 

### 2016-12-19 typespp.tex
simplify diagrams (by omitting, like in the rest of the circuits, applications of id combinators).  Add more words to their description.  These are now quite simple.

@@
 
-ccc have a notion of higher-order functions, trace, etc. Talk about these
-examples.
-
-\begin{lemma}
-  \label{lem:cong}
-  For $\mathbb{v} : \mathbb{T}_1 \circledast \mathbb{T}_2$,
-  $p : \mathbb{T}_1 \leftrightblackspoon \mathbb{T}$, we have
-  $\texttt{cong}^{l}_{p}(\mathbb{v}) : \mathbb{T} \circledast \mathbb{T}_2$, and
-  for $q : \mathbb{T}_2 \leftrightblackspoon \mathbb{T}$, we have
-  $\texttt{cong}^{r}_{q}(\mathbb{v}) : \mathbb{T}_1 \circledast \mathbb{T}$.
-  \begin{proof}
-    It follows from the fact that $\circledast$ is an endo-bi-functor.
-  \end{proof}
-\end{lemma}
+Compact closed categories have various interesting notions, such as
+duals, name and coname, which eventually lead to a certain notion of 
+higher order functions.  Here we show that the first three (duals,
+name and coname) are definable circuits in $\Pi^/$.
 
@@
 
-Given a \textrm{/-combinator} $p : \#c \leftrightblackspoon \#d$ which is a
-morphism in the compact closed category, we have a dual morphism
+Given a \textrm{/-combinator} $p : \#c \leftrightblackspoon \#d$,
+we can define a \emph{dual morphism}
 $p^{\ast} : 1/\#d \leftrightblackspoon 1/\#c$. This is shown by building a
-reversible circuit in figure~\ref{fig:dual}.
+reversible circuit in Figure~\ref{fig:dual}.  The highlights that seem to
+``escape'' the circuit indicate that \emph{dual} can only be used in a
+context with at least one quantifier being present.
 
@@
   \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+    \path [highlight] (-1,1) edge node[above] {} (0,1);
     \path (-1,1) edge node[above] {\footnotesize{$1/{\textsf{\#d}}$}} (0,1);
@@
     \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
-    \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (4,0.25);
 
@@
 
-    \draw (4,-0.25) -- (5,-0.25) -- (5,1.75) -- (4,1.75) -- cycle;
-    \node at (4.5,0.75) {\footnotesize{$\texttt{cong}^{l}$}};
-    \path (4.5,-0.25) edge node[left] {\footnotesize{$p : \#c \leftrightblackspoon \#d$}} (4.5,-0.75);
+    \draw (4,0.75) -- (5,0.75) -- (5,1.75) -- (4,1.75) -- cycle;
+    \node at (4.5,1.25) {\footnotesize{$p$}};
 
     \draw[dashed] (5.8,0.25) -- (5.8,1.25);
-    \path [highlight] (5,0.25) edge node[above] {} (6,0.25);
-    \path (5,0.25) edge node[above] {\footnotesize{$1/\textsf{\#d}$}} (6,0.25);
+    \path [highlight] (1,0.25) edge node[above] {} (6,0.25);
+    \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (6,0.25);
     \path [highlight] (5,1.25) edge node[above] {} (6,1.25);
@@
 
+    \path [highlight] (9,1.5) edge node[above] {} (10,1.5);
     \path (9,1.5) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (10,1.5);
@@
 and \texttt{coname} $\llcorner{p}\lrcorner$ of $p$, in
-figure~\ref{fig:name-coname}.
+Figure~\ref{fig:name-coname}.  Like with dual, we need to insert a combinator
+$p$ \emph{under} a quantifier.
 
@@
 
-      \path [highlight] (1,1.75) edge node[above] {} (2,1.75);
-      \path (1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (2,1.75);
+      \path [highlight] (1,1.75) edge node[above] {} (4,1.75);
+      \path (1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.75);
+
       \path [highlight] (1,0.25) edge node[above] {} (2,0.25);
@@
 
-      \draw (2,0) -- (3,0) -- (3,2) -- (2,2) -- cycle;
-      \node at (2.5,1) {\footnotesize{$\texttt{cong}^{r}$}};
-
-      \path (3,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.75);
+      \draw (2,0) -- (3,0) -- (3,0.5) -- (2,0.5) -- cycle;
+      \node at (2.5,0.25) {\footnotesize{$p$}};
+
+      \path [highlight] (3,0.25) edge node[above] {} (4,0.25);
       \path (3,0.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (4,0.25);
 
-      \path (2.5,0) edge node[left] {\footnotesize{$p : \#c \leftrightblackspoon \#d$}} (2.5,-0.5);
     \end{tikzpicture}
@@
 
-      \path (-1.1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (0,1.75);
+      \path [highlight] (-1.1,1.75) edge node[above] {} (2,1.75);
+      \path (-1.1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (2,1.75);
+
+      \path [highlight] (-1.1,0.25) edge node[above] {} (0,0.25);
       \path (-1.1,0.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (0,0.25);
 
-      \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
-      \node at (0.5,1) {\footnotesize{$\texttt{cong}^{r}$}};
-
-      \path (0.5,0) edge node[right] {\footnotesize{$p^{-1} : \#d \leftrightblackspoon \#c$}} (0.5,-0.5);
-
-      \path [highlight] (1,1.75) edge node[above] {} (2,1.75);
-      \path (1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (2,1.75);
+      \draw (0,0) -- (1,0) -- (1,0.5) -- (0,0.5) -- cycle;
+      \node at (0.50,0.25) {\footnotesize{$p^{-1}$}};
+
       \path [highlight] (1,0.25) edge node[above] {} (2,0.25);
@@
   \end{subfigure}
-  \caption{\label{fig:name-coname} \texttt{name} and \texttt{coname}}
+  \caption{\label{fig:name-coname} \texttt{name} and \texttt{coname},
+  where $p : \#c \leftrightblackspoon \#d$}
 \end{figure}

### 2016-12-19 typespp.tex
global revisions; sec 5 is the only remaining thing to do

--- 2016-12-19 typespp.tex
@@
 (\textrm{1-combinators}) &
-  c &:& \tau_1 \iso \tau_2 ~ [\textit{see Fig.~\ref{pi-combinators}}] \\
+  c,p &:& \tau_1 \iso \tau_2 ~ [\textit{see Fig.~\ref{pi-combinators}}] \\
 (\textrm{2-combinators}) &
@@
 \noindent Both classes of programs, 1-combinators $c$, and
-2-combinators~$\alpha$, denote \emph{equivalences} in the Homotopy Type
-Theory (HoTT) sense~\cite{hottbook}. The
-elements $c$ of 1-combinators denote type isomorphisms. The elements $\alpha$ of
-2-combinators denote the set of sound and complete equivalences between these
-type isomorphisms. Using the 1-combinators, it is possible to write any
-reversible boolean function and hence encode arbitrary boolean functions by a
-technique that goes back to Toffoli~\cite{Toffoli:1980}. The 2-combinators
-provide a layer of programs that computes semantics-preserving transformations
-of 1-combinators. As a small example, let us abbreviate $\ot \oplus \ot$ as the
-type $\mathbb{2}$ of booleans and examine two possible implementations of
-boolean negation. The first directly uses the primitive combinator
+2-combinators~$\alpha$, denote \emph{equivalences} in the Homotopy Type Theory
+(HoTT) sense~\cite{hottbook}. The elements $c$ or $p$ of 1-combinators denote
+type isomorphisms. The elements $\alpha$ of 2-combinators denote the set of
+sound and complete equivalences between these type isomorphisms. Using the
+1-combinators, it is possible to write any reversible boolean function and hence
+encode arbitrary boolean functions by a technique that goes back to
+Toffoli~\cite{Toffoli:1980}. The 2-combinators provide a layer of programs that
+computes semantics-preserving transformations of 1-combinators. As a small
+example, let us abbreviate $\ot \oplus \ot$ as the type $\mathbb{2}$ of booleans
+and examine two possible implementations of boolean negation. The first directly
+uses the primitive combinator
 $\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to exchange the two
-values of type~$\mathbb{2}$; the second use three consecutive uses of $\swapp$
-to achieve the same effect:
+values of type~$\mathbb{2}$; the second uses three consecutive $\swapp$s to
+achieve the same effect:
 \[\begin{array}{rcl}
@@
 of $G_3$ is $\frac{1}{2} + \frac{1}{2} + \frac{1}{2} = \frac{3}{2}$.  It is
-important to note that $G_1$ and $G_2$ are (categorically) equivalent groupoids,
-but that $G_3$ is not equivalent to either $G_1$ or $G_2$.  Roughly speaking
+important to note that $G_1$ and $G_2$ are categorically equivalent groupoids,
+but that $G_3$ is not categorically equivalent to either $G_1$ or $G_2$.  Roughly speaking
 this is because the number of connected components is also a categorical
@@
 
-The previous construction is quite useful: it allows us to take a set of
+The previous construction known in the literature as an \emph{action
+  groupoid}~\cite{groupoidintro} is quite useful: it allows us to take a set of
 cardinality $N$ and a permutation on that set of order $P$ to construct a
@@
 a groupoid of cardinality $\frac{3}{2}$ as shown above, it is not expressive
-enough to construct a groupoid of cardinality, say, $\frac{1}{3}$. Indeed if the
-underlying set has only one element, the only permutation is the identity and
-$P$ must be 1.
+enough to construct a groupoid of cardinality, say, $\frac{1}{3}$. Indeed, if the
+underlying set has only one element ($N=1$) the only permutation is the identity
+and $P$ must be 1.
 
@@
 \item the morphisms between every pair of objects are the elements of $\iter{p}$;
-\item there is a 2-morphisms between between 1-morphisms $\triple{k_1}{q_1}{\alpha_1}$ and $\triple{k_2}{q_2}{\alpha_2}$ for each $\alpha : q_1 \isotwo q_2$ for every $\alpha : q_1 \isotwo q_2$
+\item there is a 2-morphism between 1-morphisms $\triple{k_1}{q_1}{\alpha_1}$ and $\triple{k_2}{q_2}{\alpha_2}$ for every $\alpha : q_1 \isotwo q_2$.
 \end{itemize}
@@
 there is a
-morphism between elements $g$ and $h$ iff there exists a $x \in G$ such that
-$g = x^{-1} \cdot g \cdot x$. Fixing a type $\tau$ and a particular 1-combinator
+morphism between elements~$g$ and $h$ iff there exists an $x \in G$ such that
+$h = x^{-1} \cdot g \cdot x$. Fixing a type $\tau$ and a particular 1-combinator
 $p : \tau \iso \tau$, we get a group whose elements are the iterates $p^k$. The
@@
 identities which hold for rational numbers lift to identities in our
-situation. The most prominent and important identity we seek is:
+situation. The integration of these groupoids with the categorical sum is
+complicated and left for future work. For products, the most prominent identity,
+from which many other identities follow, is:
 \[
@@
 There are however weaker notions of groupoid equivalence, such as \emph{Morita
-  equivalence}~\cite[C5.3]{johnstone2002sketches}~\cite{looping}, that identify
+  equivalence}~\cite{looping}\cite[C5.3]{johnstone2002sketches}, that identify
 the two groupoids above.  Although the notion of equivalence we use in the next
@@
 $\order{\permfive}$ has three elements (instead of 3 isomorphism classes of
-elements). Getting an equivalence in this case reduces to finding a \emph{Morita
-  morphism} from $\order{\idiso}$ to
-$\order{\permfive} \circledast \iorder{\permfive}$. This is a functor $\psi$
-from the unit groupoid to the product groupoid, say $G$, with an additional
-condition that the commutative diagram in figure~\ref{fig:morita} is a
+elements).
+
+Getting an equivalence in this case reduces to finding a \emph{Morita morphism}
+from $\order{\idiso}$ to $\order{\permfive} \circledast
+\iorder{\permfive}$.
+This is a functor $\psi$ from the unit groupoid to the product groupoid, say
+$G$, with an additional condition that the commutative diagram below is a
 \emph{pullback}, where $s,t$ are the \emph{source} and \emph{target} maps, and
 $\ot$, $\mathbb{3}$ are the $1$- and $3$-point discrete groupoids defined
-previously.  The intuition is that the codomain consists of three objects with
-associated morphisms, each of them counting as a ``third.'' The functor from
-$\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
-include an arbitrary choice of ``which third'' to target and the universality
-condition of the pullback ensures that the choice is ultimately insignificant as
-it leads to the same result as every other possible functor.
-
-To summarize, the equivalence between the two groupoids requires a \emph{local}
-map that chooses a target and a \emph{global} condition ensuring that the choice
-ultimately annihilates. The challenge we are facing is to turn this idea into an
-operational semantics that somehow combines the local and global aspects of the
-equivalence. We will be inspired by the folklore result in categorical semantics
-of dependent type theory, that ``substitution into a dependent type is a
-pullback''~\cite{subpullback}, (but which turns out to have been fully explained
-in~\cite{practicalfoundations}) as explained in the next section.
-
-\begin{figure}[h]
-  \centering
+previously.
+
+\begin{center}
 \begin{tikzcd}
@@
 \end{tikzcd}
-  \caption{
-    \label{fig:morita}
-    Pullback diagram for a Morita morphism,
-    $\order{\idiso} \rightarrow \order{\permfive} \circledast
-    \iorder{\permfive}$
-  }
-\end{figure}
+\end{center}
+
+\medskip\noindent The intuition is that the codomain consists of three objects
+with associated morphisms, each of them counting as a ``third.'' The functor
+from $\order{\idiso}$ to $\order{\permfive} \circledast \iorder{\permfive}$ must
+include an arbitrary choice of ``which third'' to target and the universality
+condition of the pullback ensures that the choice is ultimately insignificant as
+it leads to the same result as every other possible functor.
+
+To summarize, the equivalence between the two groupoids requires a \emph{local}
+map that chooses a target and a \emph{global} condition ensuring that the choice
+ultimately annihilates. The challenge we aim to solve in the next section is to
+turn this idea into an operational semantics that somehow combines the local and
+global aspects of the equivalence. We will be inspired by the folklore result in
+the categorical semantics of dependent type theory~\cite{subpullback}, first
+explained by Paul Taylor~\cite{practicalfoundations}, that ``substitution into a
+dependent type is a pullback''.
 
@@
 %%%%%
-\subsection{First attempt}
+\subsection{First attempt: non-dependent version}
 
@@
 To understand the challenge in designing an operational semantics for $\eta$ and
-$\epsilon$ in the non-dependent setting, we consider the ``zig-zag'' circuit in
+$\epsilon$ in the non-dependent setting, we consider the ``zigzag'' circuit in
 Fig.~\ref{fig:zigzag}. Algebraically this circuit corresponds to the following
@@
 does not include the ``global'' information about the incoming value and there
-is no way for them to make consistent guesses locally. It is possible to modify
-the operational semantics to use a mechanism to express a computational effect
-which enables spatially separated parts of the program to communicate in a way
-that is reminiscent of entanglement in quantum mechanics. Possibilities to
-realize this communication are global reference cells, backtracking, logical
-variables with unification. We present in the next section a more abstract
+is no way for them to make consistent guesses locally. It is possible to design
+an operational semantics that uses computational effects to enable spatially
+separated parts of the program to communicate in a way that is reminiscent of
+entanglement in quantum mechanics. Possibilities to realize this communication
+are global reference cells, backtracking, and logical variables with
+unification. We choose however to present in the next section a more abstract
 approach that encodes the required dependency in dataflow constraints using
@@
 \end{tikzpicture}
-\caption{\label{fig:zigzag} Zig-zag}
+\caption{\label{fig:zigzag} Zigzag}
 \end{figure}
@@
 
-The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
-where each type $\mathbb{T}$ refers to a particular value $v$ in
-$\mathbb{T}$. As all fractional types are inhabited (see Remark
-in Sec.~\ref{sec:symmetryg}), there is no loss in generality in going
-to pointed types. We then allow universal quantification over indices $k$
-associated to values, over a collection of types.  In particular, the type
-$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
-imposes a constraint on the two values associated with the types $\order{c}$ and
-$\iorder{c}$; they must be ``synchronized'' by sharing the same~$k$ but are
-otherwise arbitrary. Once we introduce a universal quantifier in one direction
-of execution, we are led to also include an existential quantifier for the
-reverse execution. Intuitively, a forward evaluation step that is at liberty to
-choose a $k$ will, when reversed encounter a particular, i.e., existentially
-quantified, $k$. The syntax is therefore as follows.
+We modify the syntax of $\Pi^/$ as follows.
 
@@
 
-A type $\mathbb{T}_\bullet$ is generally of the form
+In the new definition, we enrich the types to pointed types $\mathbb{T_\bullet}$
+where each type $\mathbb{T}$ refers to a particular value $v$ in
+$\mathbb{T}$. As all fractional types are inhabited (see Remark in
+Sec.~\ref{sec:symmetryg}), there is no loss in generality in going to pointed
+types. We then allow universal quantification over indices $k$ associated to
+values, over a collection of types.  In particular, the type
+$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
+imposes a constraint on the two values associated with the types $\order{c}$ and
+$\iorder{c}$; they must be ``synchronized'' by sharing the same~$k$ but are
+otherwise arbitrary. Once we introduce a universal quantifier in one direction
+of execution, we are led to also include an existential quantifier for the
+reverse execution. Intuitively, a forward evaluation step that is at liberty to
+choose a $k$ will, when reversed encounter a particular, i.e., existentially
+quantified, $k$. A type $\mathbb{T}_\bullet$ is therefore generally of the form
 $\forall\exists{\overline{k}}.[\mathbb{T},v]$ where all the quantifiers are
-lifted to the top leaving a non-dependent type and a value in that type. The
-combinators in Fig.~\ref{pifracd:comb} also serve as the operational semantics
-for the language as they explain how the value currently in focus changes during
-evaluation. Given the new combinators, we can modify the the circuit in
-Fig.~\ref{fig:zigzag} by adding explicit synchronization primitives. The
-evaluation proceeds as follows where we disambiguate the uses of $\forall$ and
-$\exists$ for clarity.
-
-Forward execution:
+lifted to the top followed by a non-dependent type and a value in that type.
+
+The adaptation of the combinators to the pointed dependent setting in Fig .~\ref
+{pifracd:comb} is relatively straightforward. First, the types of the
+combinators must keep track of how the value currently in focus changes during
+evaluation, and hence directly encode the operational semantics of the
+language. As suggested above, the quantifiers are interpreted differently
+depending on which direction the rule is used. We use a $\forall\exists$
+notation to express the presence of \emph{one} quantifier that is interpreted
+ddifferently in each direction. The notation has the advantage of being compact,
+avoiding duplication, and making the symmetry of the rules explicit. We also
+include two new pairs of combinators $\synchl/packl$ and $\synchr/packr$ that
+serve as explicit quantifier introduction and elimination rules: they serve as
+explicit ``synchronization'' points as illustrated in the example below.
+
+The example revisits the zigzag circuit in Fig.~\ref{fig:zigzag} using the
+dependent types and the new explicit synchronization primitives. The evaluation
+proceeds as follows where we disambiguate the uses of $\forall$ and $\exists$
+in each direction for clarity:
+
+\paragraph*{Forward execution of zigzag circuit with synchronization.}
 \[\begin{array}{rcl}
@@
 
-Reverse execution:
+\paragraph*{Reverse execution of zigzag circuit with synchronization.}
 \[\begin{array}{rcl}
@@
 $p^{\ast} : 1/\#d \leftrightblackspoon 1/\#c$. This is shown by building a
-reversible circuit in figure~\ref{fig:dual}.
+reversible circuit in Fig.~\ref{fig:dual}.
 
@@
 and \texttt{coname} $\llcorner{p}\lrcorner$ of $p$, in
-figure~\ref{fig:name-coname}.
+Fig.~\ref{fig:name-coname}.
 
@@
 Our fractional types have natural denotations which are non-trivial groupoids,
-but they only scratch the surface of the tower of weak $\omega$-groupoids that
-is expressible in HoTT~\cite{hottbook}. A long term goal of our research
-is to find natural type constructors inspired by the rich combinatorial
-structure of weak $\omega$-groupoids that provide novel programming
-abstractions. We conclude by outlining several interesting connections and
-potential avenues for future work.
+but they cannot be composed with sums and they only scratch the surface of the
+tower of weak $\omega$-groupoids that is expressible in HoTT~\cite{hottbook}. A
+long term goal of our research is to find natural type constructors inspired by
+the rich combinatorial structure of weak $\omega$-groupoids that provide novel
+programming abstractions.
+
+We conclude by outlining several interesting connections and potential avenues
+for future work.
 
@@
 like we interpret ``negative money,'' as a debt to be repaid by some other part
-of the system. In fact, the zig-zag example can be viewed as modeling a credit
+of the system. In fact, the zigzag example can be viewed as modeling a credit
 card transaction where the money is produced at the output site by generating a

### 2016-12-19 typespp.tex

@@
 
-ccc have a notion of higher-order functions, trace, etc. Talk about these
-examples.
-
-\begin{lemma}
-  \label{lem:cong}
-  For $\mathbb{v} : \mathbb{T}_1 \circledast \mathbb{T}_2$,
-  $p : \mathbb{T}_1 \leftrightblackspoon \mathbb{T}$, we have
-  $\texttt{cong}^{l}_{p}(\mathbb{v}) : \mathbb{T} \circledast \mathbb{T}_2$, and
-  for $q : \mathbb{T}_2 \leftrightblackspoon \mathbb{T}$, we have
-  $\texttt{cong}^{r}_{q}(\mathbb{v}) : \mathbb{T}_1 \circledast \mathbb{T}$.
-  \begin{proof}
-    It follows from the fact that $\circledast$ is an endo-bi-functor.
-  \end{proof}
-\end{lemma}
+Compact closed categories have various interesting notions, such as
+duals, name and coname, which eventually lead to a certain notion of
+higher order functions.  Here we show that the first three (duals,
+name and coname) are definable circuits in $\Pi^/$.
 
@@
 
-Given a \textrm{/-combinator} $p : \#c \leftrightblackspoon \#d$ which is a
-morphism in the compact closed category, we have a dual morphism
+Given a \textrm{/-combinator} $p : \#c \leftrightblackspoon \#d$,
+we can define a \emph{dual morphism}
 $p^{\ast} : 1/\#d \leftrightblackspoon 1/\#c$. This is shown by building a
-reversible circuit in Fig.~\ref{fig:dual}.
+reversible circuit in Fig.~\ref{fig:dual}.  The highlights that seem to
+``escape'' the circuit indicate that \emph{dual} can only be used in a
+context with at least one quantifier being present.
 
@@
 and \texttt{coname} $\llcorner{p}\lrcorner$ of $p$, in
-Fig.~\ref{fig:name-coname}.
+Fig.~\ref{fig:name-coname}.  Like with dual, we need to insert a combinator $p$
+\emph{under} a quantifier.
 

### 2016-12-19 typespp.tex
typo
missing '['
sec 5 changes

@@
 \\
-Each combinator $\rho_\bullet$ has an inverse. A rule $\mathbb{T_1}_\bullet \leftrightblackspoon
+Each combinator $\rho_\bullet$ has an inverse $\rho_\bullet^{-1}$.
+A rule $\mathbb{T_1}_\bullet \leftrightblackspoon
 \forall\exists{k}.\mathbb{T_2}_\bullet$ introduces a $\forall$ in the left-to-right
@@
 
-Compact closed categories have various interesting notions, such as
-duals, name and coname, which eventually lead to a certain notion of
-higher order functions.  Here we show that the first three (duals,
-name and coname) are definable circuits in $\Pi^/$.
-
+Compact closed categories support various interesting constructions, such as
+duals, name and coname, and trace, which eventually lead to a certain notion of
+recursive higher-order functions~\cite{abramsky-2008}.  We illustrate the
+expressiveness of $\Pi^/$ by defining these constructions.
+
+%%%%%%%%%%
 \subsection{\texttt{dual}}
@@
 
+%%%
 \subsection{\texttt{name} and \texttt{coname}}
 
-We also show circuits that construct the \texttt{name} $\ulcorner{p}\urcorner$
-and \texttt{coname} $\llcorner{p}\lrcorner$ of $p$, in
-Fig.~\ref{fig:name-coname}.  Like with dual, we need to insert a combinator $p$
-\emph{under} a quantifier.
+Each morphism $\rho_\bullet : \order{c} \leftrightblackspoon \order{d}$ can be
+given a \emph{name} and a \emph{coname}. As shown in Fig.~\ref{fig:name-coname}.
+the name of $\rho_\bullet$ denoted $\ulcorner{\rho_\bullet}\urcorner$ has type
+$\order{\idiso} \leftrightblackspoon \iorder{c} \circledast \order{d}$ and the
+coname of $\rho_\bullet$ denoted $\llcorner{\rho_\bullet}\lrcorner$ has type
+$\iorder{c} \circledast \order{d} \leftrightblackspoon \order{\idiso}$.
+Intuitively, this means that for each morphism of type
+$\order{c} \leftrightblackspoon \order{d}$, it is possible to construct from
+``nothing'' an object $\iorder{c} \circledast \order{d}$ in the category that
+denotes this morphism and dually to eliminate this object. Like with the case
+for the dual construction, we need to insert a combinator \emph{under} a
+quantifier.
 
@@
       \draw (2,0) -- (3,0) -- (3,0.5) -- (2,0.5) -- cycle;
-      \node at (2.5,0.25) {\footnotesize{$p$}};
+      \node at (2.5,0.25) {\footnotesize{$\rho_\bullet$}};
 
@@
       \draw (0,0) -- (1,0) -- (1,0.5) -- (0,0.5) -- cycle;
-      \node at (0.50,0.25) {\footnotesize{$p^{-1}$}};
+      \node at (0.50,0.25) {\footnotesize{$\rho_\bullet^{-1}$}};
 
@@
 \end{figure}
+
+%%%
+\subsection{Higher-Order Functions}
+
+As shown above, names and conames encode a morphism of type
+$\order{c} \leftrightblackspoon \order{d}$ using morphisms between $\order{id}$
+and $\iorder{c} \circledast \order{d}$. These names and conames give rise to a
+notion of higher-order function that can be composed and applied. Given two
+morphisms of types $\order{c} \leftrightblackspoon \order{d}$ and
+$\order{d} \leftrightblackspoon \order{d}$, we can compose their names to get
+the name of their sequential composition. Given the name of a morphism
+$\order{c} \leftrightblackspoon \order{d}$ and given a value of type
+$\order{c}$, we can produce a value of type $\order{d}$ effectively ``applying''
+the morphism.
 

### 2016-12-19 typespp.tex
fig:apply
fig:compose
trace

@@
 \[\begin{array}{rrcll}
-\unitetl/ :& \forall\exists{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
+\unitetl/ :& \mbox{\AE}{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
                                                    \langle \idiso,v \rangle]
@@
 notation to express the presence of \emph{one} quantifier that is interpreted
-ddifferently in each direction. The notation has the advantage of being compact,
+differently in each direction. The notation has the advantage of being compact,
 avoiding duplication, and making the symmetry of the rules explicit. We also
@@
 &\stackrel{\synchl}{\leftrightblackspoon}&
-  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
+  [\langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
 &\stackrel{\assocrt/}{\leftrightblackspoon}&
@@
 the morphism.
+
+%%%
+\subsection{Trace}
+
+Traced categories~\cite{joyal1996traced} have proved useful for modeling
+recursion~\cite{Hasegawa:1997:RCS:645893.671607}. It is interesting that every
+compact closed category admits a trace which we can implement as shown below. In
+the figure, we show that given a combinator
+$\rho_\bullet : \order{c} \circledast \order{d} \leftrightblackspoon \order{c}
+\circledast \order{e}$
+we can construct a circuit of type $\order{d} \leftrightblackspoon \order{e}$
+that typechecks iff we have a fixed point for $f$ for the argument of type
+$\order{c}$.
+
+FIX TRACE
+
+\begin{figure}[htb]
+  \centering
+  \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+    \path [highlight] (-1,1) edge node[above] {} (0,1);
+    \path (-1,1) edge node[above] {\footnotesize{$\order{d}$}} (0,1);
+
+    \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+    \node at (0.5,1) {\footnotesize{$\unitetl/$}};
+
+    \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
+
+    \draw (2,0.75) -- (3,0.75) -- (3,2.75) -- (2,2.75) -- cycle;
+    \node at (2.5,1.75) {\footnotesize{$\eta +$}};
+
+    \draw[dashed] (3.2,1.25) -- (3.2,2.25);
+    \path [highlight] (3,2.25) edge node[above] {} (8,2.25);
+    \path (3,2.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (8,2.25);
+    \path [highlight] (3,1.25) edge node[above] {} (4,1.25);
+    \path (3,1.25) edge node[above] {\footnotesize{$\textsf{\#c}$}} (4,1.25);
+
+    \draw (4,0.75) -- (5,0.75) -- (5,1.75) -- (4,1.75) -- cycle;
+    \node at (4.5,1.25) {\footnotesize{$p$}};
+
+    \draw[dashed] (5.8,0.25) -- (5.8,1.25);
+    \path [highlight] (1,0.25) edge node[above] {} (6,0.25);
+    \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (6,0.25);
+    \path [highlight] (5,1.25) edge node[above] {} (6,1.25);
+    \path (5,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,1.25);
+
+    \draw (6,-0.25) -- (7,-0.25) -- (7,1.75) -- (6,1.75) -- cycle;
+    \node at (6.5,0.75) {\footnotesize{$\epsilon +$}};
+
+    \path (7,0.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (8,0.75);
+
+    \draw (8,0.5) -- (9,0.5) -- (9,2.5) -- (8,2.5) -- cycle;
+    \node at (8.5,1.5) {\footnotesize{$\unititl/$}};
+
+    \path [highlight] (9,1.5) edge node[above] {} (10,1.5);
+    \path (9,1.5) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (10,1.5);
+  \end{tikzpicture}
+  \caption{\label{fig:trace} Trace}
+\end{figure}
 

### 2016-12-19 typespp.tex
Fix packl/packr typesetting

@@
 avoiding duplication, and making the symmetry of the rules explicit. We also
-include two new pairs of combinators $\synchl/packl$ and $\synchr/packr$ that
+include two new pairs of combinators $\synchl/\packl$ and $\synchr/\packr$ that
 serve as explicit quantifier introduction and elimination rules: they serve as
@@
   \caption{\label{fig:trace} Trace}
+\end{figure}
+
+\begin{figure}[htb]
+  \centering
+  \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+    \path (-1,2) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (2,2);
+    \path (-1,0.5) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (0,0.5);
+
+    \draw (0,-0.25) -- (1,-0.25) -- (1,1.5) -- (0,1.5) -- cycle;
+    \node at (0.5,0.6) {\footnotesize{$\textsf{name}_p$}};
+
+    \path (1,1.25) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (2,1.25);
+    \path (1,0) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (4,0);
+
+    \draw (2,1) -- (3,1) -- (3,2.25) -- (2,2.25) -- cycle;
+    \node at (2.5,1.625) {\footnotesize{$\epsilon +$}};
+
+    \path (3,1.625) edge node[above] {\footnotesize{$\textsf{\#id}$}} (4,1.625);
+
+    \draw (4,-0.25) -- (5,-0.25) -- (5,2) -- (4,2) -- cycle;
+    \node at (4.5,0.85) {\footnotesize{$\unititl/$}};
+
+    \path (5,0.85) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,0.85);
+  \end{tikzpicture}
+  \caption{\label{fig:apply} apply}
+\end{figure}
+
+\begin{figure}[htb]
+  \centering
+  \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+    \path (-1,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (0,1);
+
+    \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+    \node at (0.5,1) {\footnotesize{$\unitetl/$}};
+
+    \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
+    \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,0.25);
+
+    \draw (2,1.25) -- (3,1.25) -- (3,2.25) -- (2,2.25) -- cycle;
+    \node at (2.5,1.75) {\footnotesize{$\textsf{name}_{p_1}$}};
+
+    \path (3,2.125) edge node[above] {\footnotesize{${1/\textsf{\#c}}$}} (8,2.125);
+    \path (3,1.375) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (4,1.375);
+
+    \draw (2,0.75) -- (3,0.75) -- (3,-0.25) -- (2,-0.25) -- cycle;
+    \node at (2.5,0.25) {\footnotesize{$\textsf{name}_{p_2}$}};
+
+    \path (3,0.625) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (4,0.625);
+    \path (3,-0.125) edge node[above] {\footnotesize{${\textsf{\#e}}$}} (6,-0.125);
+
+    \draw (4,1.5) -- (5,1.5) -- (5,0.5) -- (4,0.5) -- cycle;
+    \node at (4.5,1) {\footnotesize{$\epsilon +$}};
+
+    \path (5,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (6,1);
+
+    \draw (6,1.125) -- (7,1.125) -- (7,-0.25) -- (6,-0.25) -- cycle;
+    \node at (6.5,0.5) {\footnotesize{$\unititl/$}};
+
+    \path (7,0.5) edge node[above] {\footnotesize{${\textsf{\#e}}$}} (8,0.5);
+  \end{tikzpicture}
+  \caption{\label{fig:compose} compose}
 \end{figure}

### 2016-12-19 typespp.tex
before merge

@@
 \[\begin{array}{rrcll}
-\unitetl/ :& \mbox{\AE}{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
+\unitetl/ :& \forall\exists{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
                                                    \langle \idiso,v \rangle]
@@
 avoiding duplication, and making the symmetry of the rules explicit. We also
-include two new pairs of combinators $\synchl/\packl$ and $\synchr/\packr$ that
+include two new pairs of combinators $\synchl/packl$ and $\synchr/packr$ that
 serve as explicit quantifier introduction and elimination rules: they serve as
@@
 recursive higher-order functions~\cite{abramsky-2008}.  We illustrate the
-expressiveness of $\Pi^/$ by defining these constructions.
+expressiveness of $\Pi^/$ by defining duals, name and coname, and a higher-order
+\emph{apply} combinator.
 

### 2016-12-19 typespp.tex
Introduce notation for ∀∃

--- 2016-12-19 typespp.tex
@@
 \usepackage{braket}
+\usepackage{mathtools}
 
@@
 \newcommand{\permsix}{\mathit{perm}_{\times.\times}}
+
+\newcommand{\quant}[0]{\mbox{\ensuremath{\mathrlap{\diamond}\square}}}
 
@@
 \[\begin{array}{rrcll}
-\unitetl/ :& \mbox{\AE}{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
+\unitetl/ :& \quant{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
                                                    \langle \idiso,v \rangle]
   & \leftrightblackspoon &
-                   \forall\exists{\overline{k}}.[\mathbb{T},v]
+                   {\overline{k}}.[\mathbb{T},v]
   &: \unititl/ \\
-\unitetr/ :& \forall\exists{\overline{k}}.[\mathbb{T} \circledast \order{\idiso},
+\unitetr/ :& \quant{\overline{k}}.[\mathbb{T} \circledast \order{\idiso},
                                                     \langle v,\idiso]
   & \leftrightblackspoon &
-                   \forall\exists{\overline{k}}.[\mathbb{T},v]
+                   \quant{\overline{k}}.[\mathbb{T},v]
   &: \unititr/ \\
-\swapt/ :& \forall\exists{\overline{k}}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
+\swapt/ :& \quant{\overline{k}}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                    \langle v_1,v_2 \rangle]
  & \leftrightblackspoon &
-                  \forall\exists{\overline{k}}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
+                  \quant{\overline{k}}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                   \langle v_2,v_1 \rangle]
   &: \swapt/ \\
-  \assoclt/ :& \forall\exists{\overline{k}}.
+  \assoclt/ :& \quant{\overline{k}}.
                [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
@@
   & \leftrightblackspoon &
-                      \forall\exists{\overline{k}}.
+                      \quant{\overline{k}}.
                [(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
@@
 \eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  \forall\exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k  \rangle]
+  \quant{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k  \rangle]
   & : \epsilon- \\
 \eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
-  \forall\exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
+  \quant{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
   & : \epsilon+ \\
-\synchl :& \forall\exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
+\synchl :& \quant{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
   \langle \langle c^k,1/c^k \rangle, c^i]
@@
   & : \packl \\
-\synchr :& \forall\exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
+\synchr :& \quant{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
   \langle c^i, \langle 1/c^k, c^k \rangle \rangle]
@@
 {}
-{\jdg{}{}{\idiso/ : \forall\exists{\overline{k}}.[\mathbb{T},v] \leftrightblackspoon
-   \forall\exists{\overline{k}}.[\mathbb{T},v]}}
+{\jdg{}{}{\idiso/ : \quant{\overline{k}}.[\mathbb{T},v] \leftrightblackspoon
+   \quant{\overline{k}}.[\mathbb{T},v]}}
 {}
@@
 {\jdg{}{}{\rho_1 :
-  \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
-  \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2]} \quad
+  \quant{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
+  \quant{\overline{k_2}}.[\mathbb{T}_2,v_2]} \quad
   \rho_2 :
-  \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2] \leftrightblackspoon
-  \forall\exists{\overline{k_3}}.[\mathbb{T}_3,v_3]}
+  \quant{\overline{k_2}}.[\mathbb{T}_2,v_2] \leftrightblackspoon
+  \quant{\overline{k_3}}.[\mathbb{T}_3,v_3]}
 {\jdg{}{}{\rho_1 \circledcirc \rho_2 :
- \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
- \forall\exists{\overline{k_3}}.[\mathbb{T}_3},v_3]}
+ \quant{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
+ \quant{\overline{k_3}}.[\mathbb{T}_3},v_3]}
 {}
@@
 {\jdg{}{}{\rho_1 :
- \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
- \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2]}
+ \quant{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
+ \quant{\overline{k_2}}.[\mathbb{T}_2,v_2]}
  \qquad \rho_2 :
- \forall\exists{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
- \forall\exists{\overline{k_4}}.[\mathbb{T}_4,v_4]}
+ \quant{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
+ \quant{\overline{k_4}}.[\mathbb{T}_4,v_4]}
 {\jdg{}{}{\rho_1 \circledast \rho_2 :
-  \forall\exists{\overline{k_1k_3}}.[\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
+  \quant{\overline{k_1k_3}}.[\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
   \leftrightblackspoon
-  \forall\exists{\overline{k_2k_4}}.[\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
+  \quant{\overline{k_2k_4}}.[\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
 {}
@@
 A rule $\mathbb{T_1}_\bullet \leftrightblackspoon
-\forall\exists{k}.\mathbb{T_2}_\bullet$ introduces a $\forall$ in the left-to-right
+\quant{k}.\mathbb{T_2}_\bullet$ introduces a $\forall$ in the left-to-right
 direction and eliminates an $\exists$ in the right-to-left direction. A rule
-$\forall\exists{k}.\mathbb{T_1}_\bullet \leftrightblackspoon
+$\quant{k}.\mathbb{T_1}_\bullet \leftrightblackspoon
 \mathbb{T_2}_\bullet$ eliminates a $\forall$ in the left-to-right direction and
 introduces an $\exists$ in the right-to-left direction. A rule
-$\forall\exists{k}.\mathbb{T_1}_\bullet \leftrightblackspoon
-\forall\exists{k}.\mathbb{T_2}_\bullet$ maintains (think of elimination then
+$\quant{k}.\mathbb{T_1}_\bullet \leftrightblackspoon
+\quant{k}.\mathbb{T_2}_\bullet$ maintains (think of elimination then
 introduction) the universal quantifier in the left-to-right direction and
@@
 quantified, $k$. A type $\mathbb{T}_\bullet$ is therefore generally of the form
-$\forall\exists{\overline{k}}.[\mathbb{T},v]$ where all the quantifiers are
+$\quant{\overline{k}}.[\mathbb{T},v]$ where all the quantifiers are
 lifted to the top followed by a non-dependent type and a value in that type.
@@
 language. As suggested above, the quantifiers are interpreted differently
-depending on which direction the rule is used. We use a $\forall\exists$
+depending on which direction the rule is used. We use a $\quant$
 notation to express the presence of \emph{one} quantifier that is interpreted

### 2016-12-19 typespp.tex
removed sec 5 because too complex to deal with synch in the time we have

--- 2016-12-19 typespp.tex
@@
 constructors denoting non-trivial groupoids and new programs that manipulate
-such types.  Sec.~5 presents some example programs to illustrate the
-expressiveness of the language.  The last section puts our work in perspective
-and concludes.
+such types. The last section puts our work in perspective and concludes.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{$\Pi^/$: Additional Examples}
-
-Compact closed categories support various interesting constructions, such as
-duals, name and coname, and trace, which eventually lead to a certain notion of
-recursive higher-order functions~\cite{abramsky-2008}.  We illustrate the
-expressiveness of $\Pi^/$ by defining these constructions.
-
-%%%%%%%%%%
-\subsection{\texttt{dual}}
-
-Given a \textrm{/-combinator} $p : \#c \leftrightblackspoon \#d$,
-we can define a \emph{dual morphism}
-$p^{\ast} : 1/\#d \leftrightblackspoon 1/\#c$. This is shown by building a
-reversible circuit in Fig.~\ref{fig:dual}.  The highlights that seem to
-``escape'' the circuit indicate that \emph{dual} can only be used in a
-context with at least one quantifier being present.
-
-\begin{figure}[htb]
-  \centering
-  \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
-    \path [highlight] (-1,1) edge node[above] {} (0,1);
-    \path (-1,1) edge node[above] {\footnotesize{$1/{\textsf{\#d}}$}} (0,1);
-
-    \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
-    \node at (0.5,1) {\footnotesize{$\unitetl/$}};
-
-    \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
-
-    \draw (2,0.75) -- (3,0.75) -- (3,2.75) -- (2,2.75) -- cycle;
-    \node at (2.5,1.75) {\footnotesize{$\eta +$}};
-
-    \draw[dashed] (3.2,1.25) -- (3.2,2.25);
-    \path [highlight] (3,2.25) edge node[above] {} (8,2.25);
-    \path (3,2.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (8,2.25);
-    \path [highlight] (3,1.25) edge node[above] {} (4,1.25);
-    \path (3,1.25) edge node[above] {\footnotesize{$\textsf{\#c}$}} (4,1.25);
-
-    \draw (4,0.75) -- (5,0.75) -- (5,1.75) -- (4,1.75) -- cycle;
-    \node at (4.5,1.25) {\footnotesize{$p$}};
-
-    \draw[dashed] (5.8,0.25) -- (5.8,1.25);
-    \path [highlight] (1,0.25) edge node[above] {} (6,0.25);
-    \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (6,0.25);
-    \path [highlight] (5,1.25) edge node[above] {} (6,1.25);
-    \path (5,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,1.25);
-
-    \draw (6,-0.25) -- (7,-0.25) -- (7,1.75) -- (6,1.75) -- cycle;
-    \node at (6.5,0.75) {\footnotesize{$\epsilon +$}};
-
-    \path (7,0.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (8,0.75);
-
-    \draw (8,0.5) -- (9,0.5) -- (9,2.5) -- (8,2.5) -- cycle;
-    \node at (8.5,1.5) {\footnotesize{$\unititl/$}};
-
-    \path [highlight] (9,1.5) edge node[above] {} (10,1.5);
-    \path (9,1.5) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (10,1.5);
-  \end{tikzpicture}
-  \caption{\label{fig:dual} Dual}
-\end{figure}
-
-%%%
-\subsection{\texttt{name} and \texttt{coname}}
-
-Each morphism $\rho_\bullet : \order{c} \leftrightblackspoon \order{d}$ can be
-given a \emph{name} and a \emph{coname}. As shown in Fig.~\ref{fig:name-coname}.
-the name of $\rho_\bullet$ denoted $\ulcorner{\rho_\bullet}\urcorner$ has type
-$\order{\idiso} \leftrightblackspoon \iorder{c} \circledast \order{d}$ and the
-coname of $\rho_\bullet$ denoted $\llcorner{\rho_\bullet}\lrcorner$ has type
-$\iorder{c} \circledast \order{d} \leftrightblackspoon \order{\idiso}$.
-Intuitively, this means that for each morphism of type
-$\order{c} \leftrightblackspoon \order{d}$, it is possible to construct from
-``nothing'' an object $\iorder{c} \circledast \order{d}$ in the category that
-denotes this morphism and dually to eliminate this object. Like with the case
-for the dual construction, we need to insert a combinator \emph{under} a
-quantifier.
-
-\begin{figure}[htb]
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
-      \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (0,1);
-
-      \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
-      \node at (0.5,1) {\footnotesize{$\eta -$}};
-
-      \path [highlight] (1,1.75) edge node[above] {} (4,1.75);
-      \path (1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.75);
-
-      \path [highlight] (1,0.25) edge node[above] {} (2,0.25);
-      \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (2,0.25);
-      \draw[dashed] (1.25,1.75) -- (1.25,0.25);
-
-      \draw (2,0) -- (3,0) -- (3,0.5) -- (2,0.5) -- cycle;
-      \node at (2.5,0.25) {\footnotesize{$\rho_\bullet$}};
-
-      \path [highlight] (3,0.25) edge node[above] {} (4,0.25);
-      \path (3,0.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (4,0.25);
-
-    \end{tikzpicture}
-  \end{subfigure}
-  \begin{subfigure}{.5\textwidth}
-    \centering
-    \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
-
-      \path [highlight] (-1.1,1.75) edge node[above] {} (2,1.75);
-      \path (-1.1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (2,1.75);
-
-      \path [highlight] (-1.1,0.25) edge node[above] {} (0,0.25);
-      \path (-1.1,0.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (0,0.25);
-
-      \draw (0,0) -- (1,0) -- (1,0.5) -- (0,0.5) -- cycle;
-      \node at (0.50,0.25) {\footnotesize{$\rho_\bullet^{-1}$}};
-
-      \path [highlight] (1,0.25) edge node[above] {} (2,0.25);
-      \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (2,0.25);
-      \draw[dashed] (1.75,1.75) -- (1.75,0.25);
-
-      \draw (2,0) -- (3,0) -- (3,2) -- (2,2) -- cycle;
-      \node at (2.5,1) {\footnotesize{$\epsilon +$}};
-
-      \path (3,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (4,1);
-    \end{tikzpicture}
-  \end{subfigure}
-  \caption{\label{fig:name-coname} \texttt{name} and \texttt{coname},
-  where $p : \#c \leftrightblackspoon \#d$}
-\end{figure}
-
-%%%
-\subsection{Higher-Order Functions}
-
-As shown above, names and conames encode a morphism of type
-$\order{c} \leftrightblackspoon \order{d}$ using morphisms between $\order{id}$
-and $\iorder{c} \circledast \order{d}$. These names and conames give rise to a
-notion of higher-order function that can be composed and applied. Given two
-morphisms of types $\order{c} \leftrightblackspoon \order{d}$ and
-$\order{d} \leftrightblackspoon \order{d}$, we can compose their names to get
-the name of their sequential composition. Given the name of a morphism
-$\order{c} \leftrightblackspoon \order{d}$ and given a value of type
-$\order{c}$, we can produce a value of type $\order{d}$ effectively ``applying''
-the morphism.
-
-%%%
-\subsection{Trace}
-
-Traced categories~\cite{joyal1996traced} have proved useful for modeling
-recursion~\cite{Hasegawa:1997:RCS:645893.671607}. It is interesting that every
-compact closed category admits a trace which we can implement as shown below. In
-the figure, we show that given a combinator
-$\rho_\bullet : \order{c} \circledast \order{d} \leftrightblackspoon \order{c}
-\circledast \order{e}$
-we can construct a circuit of type $\order{d} \leftrightblackspoon \order{e}$
-that typechecks iff we have a fixed point for $f$ for the argument of type
-$\order{c}$.
-
-FIX TRACE
-
-\begin{figure}[htb]
-  \centering
-  \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
-    \path [highlight] (-1,1) edge node[above] {} (0,1);
-    \path (-1,1) edge node[above] {\footnotesize{$\order{d}$}} (0,1);
-
-    \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
-    \node at (0.5,1) {\footnotesize{$\unitetl/$}};
-
-    \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
-
-    \draw (2,0.75) -- (3,0.75) -- (3,2.75) -- (2,2.75) -- cycle;
-    \node at (2.5,1.75) {\footnotesize{$\eta +$}};
-
-    \draw[dashed] (3.2,1.25) -- (3.2,2.25);
-    \path [highlight] (3,2.25) edge node[above] {} (8,2.25);
-    \path (3,2.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (8,2.25);
-    \path [highlight] (3,1.25) edge node[above] {} (4,1.25);
-    \path (3,1.25) edge node[above] {\footnotesize{$\textsf{\#c}$}} (4,1.25);
-
-    \draw (4,0.75) -- (5,0.75) -- (5,1.75) -- (4,1.75) -- cycle;
-    \node at (4.5,1.25) {\footnotesize{$p$}};
-
-    \draw[dashed] (5.8,0.25) -- (5.8,1.25);
-    \path [highlight] (1,0.25) edge node[above] {} (6,0.25);
-    \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (6,0.25);
-    \path [highlight] (5,1.25) edge node[above] {} (6,1.25);
-    \path (5,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,1.25);
-
-    \draw (6,-0.25) -- (7,-0.25) -- (7,1.75) -- (6,1.75) -- cycle;
-    \node at (6.5,0.75) {\footnotesize{$\epsilon +$}};
-
-    \path (7,0.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (8,0.75);
-
-    \draw (8,0.5) -- (9,0.5) -- (9,2.5) -- (8,2.5) -- cycle;
-    \node at (8.5,1.5) {\footnotesize{$\unititl/$}};
-
-    \path [highlight] (9,1.5) edge node[above] {} (10,1.5);
-    \path (9,1.5) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (10,1.5);
-  \end{tikzpicture}
-  \caption{\label{fig:trace} Trace}
-\end{figure}
-
-\begin{figure}[htb]
-  \centering
-  \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
-    \path (-1,2) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (2,2);
-    \path (-1,0.5) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (0,0.5);
-
-    \draw (0,-0.25) -- (1,-0.25) -- (1,1.5) -- (0,1.5) -- cycle;
-    \node at (0.5,0.6) {\footnotesize{$\textsf{name}_p$}};
-
-    \path (1,1.25) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (2,1.25);
-    \path (1,0) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (4,0);
-
-    \draw (2,1) -- (3,1) -- (3,2.25) -- (2,2.25) -- cycle;
-    \node at (2.5,1.625) {\footnotesize{$\epsilon +$}};
-
-    \path (3,1.625) edge node[above] {\footnotesize{$\textsf{\#id}$}} (4,1.625);
-
-    \draw (4,-0.25) -- (5,-0.25) -- (5,2) -- (4,2) -- cycle;
-    \node at (4.5,0.85) {\footnotesize{$\unititl/$}};
-
-    \path (5,0.85) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,0.85);
-  \end{tikzpicture}
-  \caption{\label{fig:apply} apply}
-\end{figure}
-
-\begin{figure}[htb]
-  \centering
-  \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
-    \path (-1,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (0,1);
-
-    \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
-    \node at (0.5,1) {\footnotesize{$\unitetl/$}};
-
-    \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
-    \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,0.25);
-
-    \draw (2,1.25) -- (3,1.25) -- (3,2.25) -- (2,2.25) -- cycle;
-    \node at (2.5,1.75) {\footnotesize{$\textsf{name}_{p_1}$}};
-
-    \path (3,2.125) edge node[above] {\footnotesize{${1/\textsf{\#c}}$}} (8,2.125);
-    \path (3,1.375) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (4,1.375);
-
-    \draw (2,0.75) -- (3,0.75) -- (3,-0.25) -- (2,-0.25) -- cycle;
-    \node at (2.5,0.25) {\footnotesize{$\textsf{name}_{p_2}$}};
-
-    \path (3,0.625) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (4,0.625);
-    \path (3,-0.125) edge node[above] {\footnotesize{${\textsf{\#e}}$}} (6,-0.125);
-
-    \draw (4,1.5) -- (5,1.5) -- (5,0.5) -- (4,0.5) -- cycle;
-    \node at (4.5,1) {\footnotesize{$\epsilon +$}};
-
-    \path (5,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (6,1);
-
-    \draw (6,1.125) -- (7,1.125) -- (7,-0.25) -- (6,-0.25) -- cycle;
-    \node at (6.5,0.5) {\footnotesize{$\unititl/$}};
-
-    \path (7,0.5) edge node[above] {\footnotesize{${\textsf{\#e}}$}} (8,0.5);
-  \end{tikzpicture}
-  \caption{\label{fig:compose} compose}
-\end{figure}
+%%%%
+\subsection{Additional Examples}
+
+Compact closed categories support various interesting
+constructions~\cite{abramsky-2008}, such as duals, name and coname, and
+trace~\cite{joyal1996traced,Hasegawa:1997:RCS:645893.671607}, which eventually
+lead to a certain notion of recursive higher-order functions. These
+constructions do not immediately transfer to the context of $\Pi^/$ as we have
+explicit quantifiers and synchronization primitives that are not present in the
+conventional setting. Resolving these subtleties will provide the first
+operational interpretation known to us of multiplicative higher-order recursive
+functions.
+
+% \begin{figure}[htb]
+%   \centering
+%   \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+%     \path [highlight] (-1,1) edge node[above] {} (0,1);
+%     \path (-1,1) edge node[above] {\footnotesize{$1/{\textsf{\#d}}$}} (0,1);
+
+%     \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+%     \node at (0.5,1) {\footnotesize{$\unitetl/$}};
+
+%     \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
+
+%     \draw (2,0.75) -- (3,0.75) -- (3,2.75) -- (2,2.75) -- cycle;
+%     \node at (2.5,1.75) {\footnotesize{$\eta +$}};
+
+%     \draw[dashed] (3.2,1.25) -- (3.2,2.25);
+%     \path [highlight] (3,2.25) edge node[above] {} (8,2.25);
+%     \path (3,2.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (8,2.25);
+%     \path [highlight] (3,1.25) edge node[above] {} (4,1.25);
+%     \path (3,1.25) edge node[above] {\footnotesize{$\textsf{\#c}$}} (4,1.25);
+
+%     \draw (4,0.75) -- (5,0.75) -- (5,1.75) -- (4,1.75) -- cycle;
+%     \node at (4.5,1.25) {\footnotesize{$p$}};
+
+%     \draw[dashed] (5.8,0.25) -- (5.8,1.25);
+%     \path [highlight] (1,0.25) edge node[above] {} (6,0.25);
+%     \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (6,0.25);
+%     \path [highlight] (5,1.25) edge node[above] {} (6,1.25);
+%     \path (5,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,1.25);
+
+%     \draw (6,-0.25) -- (7,-0.25) -- (7,1.75) -- (6,1.75) -- cycle;
+%     \node at (6.5,0.75) {\footnotesize{$\epsilon +$}};
+
+%     \path (7,0.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (8,0.75);
+
+%     \draw (8,0.5) -- (9,0.5) -- (9,2.5) -- (8,2.5) -- cycle;
+%     \node at (8.5,1.5) {\footnotesize{$\unititl/$}};
+
+%     \path [highlight] (9,1.5) edge node[above] {} (10,1.5);
+%     \path (9,1.5) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (10,1.5);
+%   \end{tikzpicture}
+%   \caption{\label{fig:dual} Dual}
+% \end{figure}
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \section{$\Pi^/$: Additional Examples}
+
+% Compact closed categories support various interesting constructions, such as
+% duals, name and coname, and trace, which eventually lead to a certain notion of
+% recursive higher-order functions~\cite{abramsky-2008}.  We illustrate the
+% expressiveness of $\Pi^/$ by defining duals, name and coname, and a higher-order
+% \emph{apply} combinator.
+
+% %%%%%%%%%%
+% \subsection{\texttt{dual}}
+
+% Given a \textrm{/-combinator} $p : \#c \leftrightblackspoon \#d$,
+% we can define a \emph{dual morphism}
+% $p^{\ast} : 1/\#d \leftrightblackspoon 1/\#c$. This is shown by building a
+% reversible circuit in Fig.~\ref{fig:dual}.  The highlights that seem to
+% ``escape'' the circuit indicate that \emph{dual} can only be used in a
+% context with at least one quantifier being present.
+
+% \begin{figure}[htb]
+%   \centering
+%   \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+%     \path [highlight] (-1,1) edge node[above] {} (0,1);
+%     \path (-1,1) edge node[above] {\footnotesize{$1/{\textsf{\#d}}$}} (0,1);
+
+%     \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+%     \node at (0.5,1) {\footnotesize{$\unitetl/$}};
+
+%     \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
+
+%     \draw (2,0.75) -- (3,0.75) -- (3,2.75) -- (2,2.75) -- cycle;
+%     \node at (2.5,1.75) {\footnotesize{$\eta +$}};
+
+%     \draw[dashed] (3.2,1.25) -- (3.2,2.25);
+%     \path [highlight] (3,2.25) edge node[above] {} (8,2.25);
+%     \path (3,2.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (8,2.25);
+%     \path [highlight] (3,1.25) edge node[above] {} (4,1.25);
+%     \path (3,1.25) edge node[above] {\footnotesize{$\textsf{\#c}$}} (4,1.25);
+
+%     \draw (4,0.75) -- (5,0.75) -- (5,1.75) -- (4,1.75) -- cycle;
+%     \node at (4.5,1.25) {\footnotesize{$p$}};
+
+%     \draw[dashed] (5.8,0.25) -- (5.8,1.25);
+%     \path [highlight] (1,0.25) edge node[above] {} (6,0.25);
+%     \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (6,0.25);
+%     \path [highlight] (5,1.25) edge node[above] {} (6,1.25);
+%     \path (5,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,1.25);
+
+%     \draw (6,-0.25) -- (7,-0.25) -- (7,1.75) -- (6,1.75) -- cycle;
+%     \node at (6.5,0.75) {\footnotesize{$\epsilon +$}};
+
+%     \path (7,0.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (8,0.75);
+
+%     \draw (8,0.5) -- (9,0.5) -- (9,2.5) -- (8,2.5) -- cycle;
+%     \node at (8.5,1.5) {\footnotesize{$\unititl/$}};
+
+%     \path [highlight] (9,1.5) edge node[above] {} (10,1.5);
+%     \path (9,1.5) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (10,1.5);
+%   \end{tikzpicture}
+%   \caption{\label{fig:dual} Dual}
+% \end{figure}
+
+% %%%
+% \subsection{\texttt{name} and \texttt{coname}}
+
+% Each morphism $\rho_\bullet : \order{c} \leftrightblackspoon \order{d}$ can be
+% given a \emph{name} and a \emph{coname}. As shown in Fig.~\ref{fig:name-coname}.
+% the name of $\rho_\bullet$ denoted $\ulcorner{\rho_\bullet}\urcorner$ has type
+% $\order{\idiso} \leftrightblackspoon \iorder{c} \circledast \order{d}$ and the
+% coname of $\rho_\bullet$ denoted $\llcorner{\rho_\bullet}\lrcorner$ has type
+% $\iorder{c} \circledast \order{d} \leftrightblackspoon \order{\idiso}$.
+% Intuitively, this means that for each morphism of type
+% $\order{c} \leftrightblackspoon \order{d}$, it is possible to construct from
+% ``nothing'' an object $\iorder{c} \circledast \order{d}$ in the category that
+% denotes this morphism and dually to eliminate this object. Like with the case
+% for the dual construction, we need to insert a combinator \emph{under} a
+% quantifier.
+
+% \begin{figure}[htb]
+%   \begin{subfigure}{.5\textwidth}
+%     \centering
+%     \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+%       \path (-1.1,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (0,1);
+
+%       \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+%       \node at (0.5,1) {\footnotesize{$\eta -$}};
+
+%       \path [highlight] (1,1.75) edge node[above] {} (4,1.75);
+%       \path (1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (4,1.75);
+
+%       \path [highlight] (1,0.25) edge node[above] {} (2,0.25);
+%       \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (2,0.25);
+%       \draw[dashed] (1.25,1.75) -- (1.25,0.25);
+
+%       \draw (2,0) -- (3,0) -- (3,0.5) -- (2,0.5) -- cycle;
+%       \node at (2.5,0.25) {\footnotesize{$\rho_\bullet$}};
+
+%       \path [highlight] (3,0.25) edge node[above] {} (4,0.25);
+%       \path (3,0.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (4,0.25);
+
+%     \end{tikzpicture}
+%   \end{subfigure}
+%   \begin{subfigure}{.5\textwidth}
+%     \centering
+%     \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+
+%       \path [highlight] (-1.1,1.75) edge node[above] {} (2,1.75);
+%       \path (-1.1,1.75) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (2,1.75);
+
+%       \path [highlight] (-1.1,0.25) edge node[above] {} (0,0.25);
+%       \path (-1.1,0.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (0,0.25);
+
+%       \draw (0,0) -- (1,0) -- (1,0.5) -- (0,0.5) -- cycle;
+%       \node at (0.50,0.25) {\footnotesize{$\rho_\bullet^{-1}$}};
+
+%       \path [highlight] (1,0.25) edge node[above] {} (2,0.25);
+%       \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (2,0.25);
+%       \draw[dashed] (1.75,1.75) -- (1.75,0.25);
+
+%       \draw (2,0) -- (3,0) -- (3,2) -- (2,2) -- cycle;
+%       \node at (2.5,1) {\footnotesize{$\epsilon +$}};
+
+%       \path (3,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (4,1);
+%     \end{tikzpicture}
+%   \end{subfigure}
+%   \caption{\label{fig:name-coname} \texttt{name} and \texttt{coname},
+%   where $p : \#c \leftrightblackspoon \#d$}
+% \end{figure}
+
+% %%%
+% \subsection{Higher-Order Functions}
+
+% As shown above, names and conames encode a morphism of type
+% $\order{c} \leftrightblackspoon \order{d}$ using morphisms between $\order{id}$
+% and $\iorder{c} \circledast \order{d}$. These names and conames give rise to a
+% notion of higher-order function that can be composed and applied. Given two
+% morphisms of types $\order{c} \leftrightblackspoon \order{d}$ and
+% $\order{d} \leftrightblackspoon \order{d}$, we can compose their names to get
+% the name of their sequential composition. Given the name of a morphism
+% $\order{c} \leftrightblackspoon \order{d}$ and given a value of type
+% $\order{c}$, we can produce a value of type $\order{d}$ effectively ``applying''
+% the morphism.
+
+% %%%
+% \subsection{Trace}
+
+% Traced categories~\cite{joyal1996traced} have proved useful for modeling
+% recursion~\cite{Hasegawa:1997:RCS:645893.671607}. It is interesting that every
+% compact closed category admits a trace which we can implement as shown below. In
+% the figure, we show that given a combinator
+% $\rho_\bullet : \order{c} \circledast \order{d} \leftrightblackspoon \order{c}
+% \circledast \order{e}$
+% we can construct a circuit of type $\order{d} \leftrightblackspoon \order{e}$
+% that typechecks iff we have a fixed point for $f$ for the argument of type
+% $\order{c}$.
+
+% FIX TRACE
+
+% \begin{figure}[htb]
+%   \centering
+%   \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+%     \path [highlight] (-1,1) edge node[above] {} (0,1);
+%     \path (-1,1) edge node[above] {\footnotesize{$\order{d}$}} (0,1);
+
+%     \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+%     \node at (0.5,1) {\footnotesize{$\unitetl/$}};
+
+%     \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
+
+%     \draw (2,0.75) -- (3,0.75) -- (3,2.75) -- (2,2.75) -- cycle;
+%     \node at (2.5,1.75) {\footnotesize{$\eta +$}};
+
+%     \draw[dashed] (3.2,1.25) -- (3.2,2.25);
+%     \path [highlight] (3,2.25) edge node[above] {} (8,2.25);
+%     \path (3,2.25) edge node[above] {\footnotesize{$1/\textsf{\#c}$}} (8,2.25);
+%     \path [highlight] (3,1.25) edge node[above] {} (4,1.25);
+%     \path (3,1.25) edge node[above] {\footnotesize{$\textsf{\#c}$}} (4,1.25);
+
+%     \draw (4,0.75) -- (5,0.75) -- (5,1.75) -- (4,1.75) -- cycle;
+%     \node at (4.5,1.25) {\footnotesize{$p$}};
+
+%     \draw[dashed] (5.8,0.25) -- (5.8,1.25);
+%     \path [highlight] (1,0.25) edge node[above] {} (6,0.25);
+%     \path (1,0.25) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (6,0.25);
+%     \path [highlight] (5,1.25) edge node[above] {} (6,1.25);
+%     \path (5,1.25) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,1.25);
+
+%     \draw (6,-0.25) -- (7,-0.25) -- (7,1.75) -- (6,1.75) -- cycle;
+%     \node at (6.5,0.75) {\footnotesize{$\epsilon +$}};
+
+%     \path (7,0.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (8,0.75);
+
+%     \draw (8,0.5) -- (9,0.5) -- (9,2.5) -- (8,2.5) -- cycle;
+%     \node at (8.5,1.5) {\footnotesize{$\unititl/$}};
+
+%     \path [highlight] (9,1.5) edge node[above] {} (10,1.5);
+%     \path (9,1.5) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (10,1.5);
+%   \end{tikzpicture}
+%   \caption{\label{fig:trace} Trace}
+% \end{figure}
+
+% \begin{figure}[htb]
+%   \centering
+%   \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+%     \path (-1,2) edge node[above] {\footnotesize{${\textsf{\#c}}$}} (2,2);
+%     \path (-1,0.5) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (0,0.5);
+
+%     \draw (0,-0.25) -- (1,-0.25) -- (1,1.5) -- (0,1.5) -- cycle;
+%     \node at (0.5,0.6) {\footnotesize{$\textsf{name}_p$}};
+
+%     \path (1,1.25) edge node[above] {\footnotesize{$1/{\textsf{\#c}}$}} (2,1.25);
+%     \path (1,0) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (4,0);
+
+%     \draw (2,1) -- (3,1) -- (3,2.25) -- (2,2.25) -- cycle;
+%     \node at (2.5,1.625) {\footnotesize{$\epsilon +$}};
+
+%     \path (3,1.625) edge node[above] {\footnotesize{$\textsf{\#id}$}} (4,1.625);
+
+%     \draw (4,-0.25) -- (5,-0.25) -- (5,2) -- (4,2) -- cycle;
+%     \node at (4.5,0.85) {\footnotesize{$\unititl/$}};
+
+%     \path (5,0.85) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (6,0.85);
+%   \end{tikzpicture}
+%   \caption{\label{fig:apply} apply}
+% \end{figure}
+
+% \begin{figure}[htb]
+%   \centering
+%   \begin{tikzpicture}[scale=0.9,every node/.style={scale=0.7},highlight/.style={line width=4pt,orange!70}]
+%     \path (-1,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (0,1);
+
+%     \draw (0,0) -- (1,0) -- (1,2) -- (0,2) -- cycle;
+%     \node at (0.5,1) {\footnotesize{$\unitetl/$}};
+
+%     \path (1,1.75) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,1.75);
+%     \path (1,0.25) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (2,0.25);
+
+%     \draw (2,1.25) -- (3,1.25) -- (3,2.25) -- (2,2.25) -- cycle;
+%     \node at (2.5,1.75) {\footnotesize{$\textsf{name}_{p_1}$}};
+
+%     \path (3,2.125) edge node[above] {\footnotesize{${1/\textsf{\#c}}$}} (8,2.125);
+%     \path (3,1.375) edge node[above] {\footnotesize{${\textsf{\#d}}$}} (4,1.375);
+
+%     \draw (2,0.75) -- (3,0.75) -- (3,-0.25) -- (2,-0.25) -- cycle;
+%     \node at (2.5,0.25) {\footnotesize{$\textsf{name}_{p_2}$}};
+
+%     \path (3,0.625) edge node[above] {\footnotesize{${1/\textsf{\#d}}$}} (4,0.625);
+%     \path (3,-0.125) edge node[above] {\footnotesize{${\textsf{\#e}}$}} (6,-0.125);
+
+%     \draw (4,1.5) -- (5,1.5) -- (5,0.5) -- (4,0.5) -- cycle;
+%     \node at (4.5,1) {\footnotesize{$\epsilon +$}};
+
+%     \path (5,1) edge node[above] {\footnotesize{${\textsf{\#id}}$}} (6,1);
+
+%     \draw (6,1.125) -- (7,1.125) -- (7,-0.25) -- (6,-0.25) -- cycle;
+%     \node at (6.5,0.5) {\footnotesize{$\unititl/$}};
+
+%     \path (7,0.5) edge node[above] {\footnotesize{${\textsf{\#e}}$}} (8,0.5);
+%   \end{tikzpicture}
+%   \caption{\label{fig:compose} compose}
+% \end{figure}
 

### 2016-12-19 typespp.tex
Enable microtype

@@
 \usepackage{braket}
+\usepackage{mathtools}
+\usepackage{microtype}
 
@@
 \[\begin{array}{rrcll}
-\unitetl/ :& \forall\exists{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
+\unitetl/ :& \quant{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
                                                    \langle \idiso,v \rangle]
   & \leftrightblackspoon &
-                   \forall\exists{\overline{k}}.[\mathbb{T},v]
+                   {\overline{k}}.[\mathbb{T},v]
   &: \unititl/ \\
-\unitetr/ :& \forall\exists{\overline{k}}.[\mathbb{T} \circledast \order{\idiso},
+\unitetr/ :& \quant{\overline{k}}.[\mathbb{T} \circledast \order{\idiso},
                                                     \langle v,\idiso]
   & \leftrightblackspoon &
-                   \forall\exists{\overline{k}}.[\mathbb{T},v]
+                   \quant{\overline{k}}.[\mathbb{T},v]
   &: \unititr/ \\
-\swapt/ :& \forall\exists{\overline{k}}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
+\swapt/ :& \quant{\overline{k}}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                    \langle v_1,v_2 \rangle]
  & \leftrightblackspoon &
-                  \forall\exists{\overline{k}}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
+                  \quant{\overline{k}}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                   \langle v_2,v_1 \rangle]
   &: \swapt/ \\
-  \assoclt/ :& \forall\exists{\overline{k}}.
+  \assoclt/ :& \quant{\overline{k}}.
                [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),

### 2016-12-19 typespp.tex
small tweak

@@
 constructions do not immediately transfer to the context of $\Pi^/$ as we have
-explicit quantifiers and synchronization primitives that are not present in the
-conventional setting. Resolving these subtleties will provide the first
-operational interpretation known to us of multiplicative higher-order recursive
-functions.
+pointed types with explicit quantifiers and synchronization primitives that are
+not present in the conventional setting. Adapting the conventional constructions
+requires resolving some subtleties and will provide the first operational
+interpretation known to us of multiplicative higher-order recursive functions.
 

### 2017-02-09 upi.tex
correspondence between pi and univalent universes

\documentclass[format=acmlarge,review,natbib]{acmart}

% \usepackage{fdsymbol}
\usepackage{bbold}
\usepackage{bussproofs}
\usepackage{keystroke}
\usepackage{comment}

%% \usepackage{hyphenat}
%% \usepackage{color}
%% \usepackage{url}
%% \usepackage{amssymb}
%% \usepackage{amsmath}
%% \usepackage{mdwtab}
%% \usepackage{mdwlist}
%% \usepackage{listings}
%% \lstMakeShortInline[columns=fullflexible]|
%% \lstnewenvironment{code}{\lstset{basicstyle={\sffamily\footnotesize}}}{}

\newcommand{\unitepl}{\texttt{unitepl}}
\newcommand{\unitipl}{\texttt{unitipl}}
\newcommand{\unitepr}{\texttt{unitepr}}
\newcommand{\unitipr}{\texttt{unitipr}}
\newcommand{\swap}{\texttt{swap}}
\newcommand{\swapp}{\texttt{swapp}}
\newcommand{\assoclp}{\texttt{assoclp}}
\newcommand{\assocrp}{\texttt{assocrp}}
\newcommand{\unitetl}{\texttt{unitetl}}
\newcommand{\unititl}{\texttt{unititl}}
\newcommand{\unitetr}{\texttt{unitetr}}
\newcommand{\unititr}{\texttt{unititr}}
\newcommand{\swapt}{\texttt{swapt}}
\newcommand{\assoclt}{\texttt{assoclt}}
\newcommand{\assocrt}{\texttt{assocrt}}
\newcommand{\absorbr}{\texttt{absorbr}}
\newcommand{\absorbl}{\texttt{absorbl}}
\newcommand{\factorzr}{\texttt{factorzr}}
\newcommand{\factorzl}{\texttt{factorzl}}
\newcommand{\factor}{\texttt{factor}}
\newcommand{\distl}{\texttt{distl}}
\newcommand{\dist}{\texttt{dist}}
\newcommand{\factorl}{\texttt{factorl}}
\newcommand{\id}{\texttt{id}}
\newcommand{\compc}[2]{#1 \circ #2}
\newcommand{\compcc}[2]{#1 \bullet #2}
\newcommand{\respcomp}[2]{#1 \odot #2}

\newcommand{\alt}{~\mid~}
\newcommand{\patht}[1]{\textsc{PATHS}(#1,#1)}
\newcommand{\fpatht}[1]{\textsc{FREEPATHS}(#1,\Box)}
\newcommand{\fpathp}[2]{\textsc{freepath}~#1~#2}
\newcommand{\pathind}[2]{\textsc{pathind}~#1~#2}
\newcommand{\invc}[1]{!\;#1}
\newcommand{\evalone}[2]{eval(#1,#2)}
\newcommand{\evalbone}[2]{evalB(#1,#2)}
\newcommand{\reflp}{\textsc{refl}}
\newcommand{\reflh}{\mathit{refl}_{\sim}}
\newcommand{\symh}[1]{\mathit{sym}_{\sim}~#1}
\newcommand{\transh}[2]{\mathit{trans}_{\sim}~#1~#2}
\newcommand{\reflq}{\mathit{refl}_{\simeq}}
\newcommand{\symq}[1]{\mathit{sym}_{\simeq}~#1}
\newcommand{\transq}[2]{\mathit{trans}_{\simeq}~#1~#2}
\newcommand{\isequiv}[1]{\mathit{isequiv}(#1)}
\newcommand{\idc}{\mathit{id}_{\boolt}}
\newcommand{\swapc}{\mathit{swap}_{\boolt}}
\newcommand{\assocc}{\mathit{assoc}}
\newcommand{\invl}{\mathit{invl}}
\newcommand{\invr}{\mathit{invr}}
\newcommand{\invinv}{\mathit{inv}^2}
\newcommand{\idlc}{\mathit{idl}}
\newcommand{\idrc}{\mathit{idr}}
\newcommand{\swapswap}{\swapc^2}
\newcommand{\compsim}{\compc_{\isotwo}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\isotwo}{\Leftrightarrow}
\newcommand{\piso}{\multimapdotbothB~~}
\newcommand{\zt}{\mathbb{0}}
\newcommand{\ot}{\mathbb{1}}
\newcommand{\fc}{\mathit{false}}
\newcommand{\tc}{\mathit{true}}
\newcommand{\boolt}{\mathbb{B}}
\newcommand{\univ}{\mathcal{U}}
\newcommand{\uzero}{\mathcal{U}_0}
\newcommand{\uone}{\mathcal{U}_1}
\newcommand{\Rule}[2]{
\makebox{
$\displaystyle
\frac{\begin{array}{l}#1\\\end{array}}
{\begin{array}{l}#2\\\end{array}}$}}
\newcommand{\proves}{\vdash}
\newcommand{\jdgg}[3]{#1 \proves #2 : #3}
\newcommand{\jdg}[2]{\proves #1 : #2}
\newcommand{\jdge}[3]{\proves #1 = #2 : #3}
%% codes
%% denotations

\newcommand{\amr}[1]{\fbox{\begin{minipage}{0.8\textwidth}\color{red}{Amr says: {#1}}\end{minipage}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

\title{Univalent Universes and Completness of Reversible Programs}

\author{X}
\affiliation{
  \institution{Y}
  \country{Z}}
\email{A@B.C}

\begin{abstract}
\end{abstract}

\maketitle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Story and Conjecture}

There is a cottage industry of reversible programming languages, reversible logic, programming applications of type isomorphisms, etc. This work seems it should be connected to HoTT and univalence but there aren't any precise connections or theorems.

Our conjecture:

\begin{itemize}
\item Let $\mathcal{U}$ be the univalent subuniverse generated by $0$, $1$, $+$, and $*$ suitably 1-truncated.
\item Let $\Pi$ be the previously  described reversible programming with 1-paths and 2-paths
\item We conjecture that $\Pi$ includes codes for \emph{all} paths in $\mathcal{U}$
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Bool Case}

To get started let's look at the case of $\mathbb{2}$ instead of the entire set of finite types defined using $0$, $1$, $+$, and $*$.

%%%%%
\subsection{$\Pi$}

The restriction of $\Pi$ is the following:

\[\begin{array}{rcl}
\tau &::=& \mathbb{2} \\
\\
v &::=& \begin{array}[t]{lrcl}
                    & \fc &:& \mathbb{2} \\
              \alt & \tc &:& \mathbb{2}
               \end{array} \\
\\
c &::=& \begin{array}[t]{lrcl}
              & \id &:& \tau \iso \tau \\
               \alt & \swap &:& \mathbb{2} \iso \mathbb{2} \\
               \alt & \circ &:& (\tau_1 \iso \tau_2) \rightarrow (\tau_2 \iso \tau_3)
                              \rightarrow (\tau_1 \iso \tau_3)
               \end{array} \\
\\
! &:& (\tau_1 \iso \tau_2) \rightarrow (\tau_2 \iso \tau_1) \\
\invc{\id} &=& \id \\
\invc{\swap} &=& \swap \\
\invc{(\compc{c_1}{c_2})} &=& \compc{\invc{c_2}\;}{\;\invc{c_1}} \\
\\
\alpha &::=& \begin{array}[t]{lrcl}
               & \id &:& c \isotwo c \\
               \alt & \assocc &:& \compc{c_1}{(\compc{c_2}{c_3})} \isotwo
                                              \compc{(\compc{c_1}{c_2})}{c_3} \\
               \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
               \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
               \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
               \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
               \alt & \bullet &:& (c_1 \isotwo c_2) \rightarrow (c_2 \isotwo c_3)
                                            \rightarrow (c_1 \isotwo c_3) \\
               \alt & \odot &:& (c_1 \isotwo c_1') \rightarrow (c_2 \isotwo c_2')
                                            \rightarrow (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
             \end{array}
\end{array}\]
The 2-combinators $\alpha$ also have inverses \ldots

%%%%%
\subsection{$\mathcal{U}$}

The univalent universe is $\Sigma_{X:\mathbf{Type}} \| X \equiv \mathbb{2} \|_1$ where \ldots

%%%%%
\subsection{Equivalence}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%
\section{Generalizations}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{acm}
{\footnotesize
\bibliography{cites}
}
\end{document}

### 2017-02-09 upi.tex
more on equivalence statement

@@
 
-The univalent universe is $\Sigma_{X:\mathbf{Type}} \| X \equiv \mathbb{2} \|_1$ where \ldots
+We assume an ambient univalent universe defined like in the HoTT book and including dependent pairs, a unit type, coproducts, identity types, and constants $\textsc{funext}$ and $\textsc{univalence}$.
+Our universe $\mathcal{U}$ is the subuniverse of this ambient universe defined as $\Sigma_{X:\mathbf{Type}} \| X \equiv \mathbb{2} \|_1$ where:
+
+\begin{itemize}
+
+\item $\Sigma$ is the type former for dependent pairs: it comes equipped with an introduction rule, an induction principle, and a computational rule.
+
+\item $\mathbb{2}$ is the type of booleans which comes with the two points, an induction principle, and computational rules.
+
+\item $\equiv$ is the identity type former which comes with $\reflp$, an induction principle $J$, and a computation rule.
+
+\item $\| A \|_1$ is the 1-truncation of $A$ defined as follows: for all $x,y:A$ and $p,q : x \equiv y$ and $\alpha,\beta : p \equiv q$, we have $\alpha=\beta$.
+
+\end{itemize}
 
@@
 \subsection{Equivalence}
+
+Now we claim that $\mathcal{U}_\Pi$ is equivalent to $\mathcal{U}$ where $\mathcal{U}_\Pi$ is a universe which includes types $\tau$ as a points, combinators $c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators $\alpha : c_1 \isotwo c_2$ as 2-paths.
 

### 2017-02-10 upi.tex
fixed truncation

@@
 \begin{itemize}
-\item Let $\mathcal{U}$ be the univalent subuniverse generated by $0$, $1$, $+$, and $*$ suitably 1-truncated.
+\item Let $\mathcal{U}$ be the univalent subuniverse generated by $0$, $1$, $+$,
+  and $*$ suitably truncated.
 \item Let $\Pi$ be the previously  described reversible programming with 1-paths and 2-paths
@@
 We assume an ambient univalent universe defined like in the HoTT book and including dependent pairs, a unit type, coproducts, identity types, and constants $\textsc{funext}$ and $\textsc{univalence}$.
-Our universe $\mathcal{U}$ is the subuniverse of this ambient universe defined as $\Sigma_{X:\mathbf{Type}} \| X \equiv \mathbb{2} \|_1$ where:
+Our universe $\mathcal{U}$ is the subuniverse of this ambient universe defined as $\Sigma_{X:\mathbf{Type}} \| X \equiv \mathbb{2} \|$ where:
 
@@
 
-\item $\| A \|_1$ is the 1-truncation of $A$ defined as follows: for all $x,y:A$ and $p,q : x \equiv y$ and $\alpha,\beta : p \equiv q$, we have $\alpha=\beta$.
+\item $\| A \|$ is the propositional truncation of $A$ defined as follows: for
+  any $x:A$ we have $|x| : \|A\|$, and for any $x,y : \|A\|$, we have $x \equiv y$.
 

### 2017-02-11 upi.tex
more details on upi and accompanying code

@@
 \documentclass[format=acmlarge,review,natbib]{acmart}
+
+%% Amr
+%% words to remember :-)
+%% sublime unfathomable
+%% ---
 
@@
 
+\newcommand{\Typ}{\mathbf{Type}}
 \newcommand{\alt}{~\mid~}
@@
 
-\title{Univalent Universes and Completness of Reversible Programs}
+\title{Univalent Universes and Completness of Reversible Programs: \\
+  Featherweight HoTT}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Univalence}
+
+Let $X : \Typ$. We have:
+\[
+\textit{transport-equiv} : \Pi_{P : X \to \Typ}~ \Pi_{x,y:X}~
+    x \equiv y  \to P x \simeq P y
+\]
+For a fibration with base $X$, define:
+\[\begin{array}{l}
+\textit{is-univalent-fibration} : (X \to \Typ) \to \Typ \\
+\textit{is-univalent-fibration}~P =
+    \Pi_{x,y:X}~\textit{is-equiv} (\textit{transport-equiv}~P~x~y)
+\end{array}\]
+
+\noindent\paragraph*{Examples}
+\begin{itemize}
+\item Take $X = \Typ$ and $P = \mathit{Id}_{\Typ}$. The statement
+  $\textit{is-univalent-fibration}~P$ reduces to $\Pi_{x,y:\Typ}~x \equiv y \to
+  x \simeq y$ which is the usual definition of univalence.
+\item Take $X = \mathbb{1}$ with element $\star$ and
+  $P = \lambda \_. \mathbb{1}$. The statement
+  $\textit{is-univalent-fibration}~P$ reduces to
+  $(\star\equiv\star)\simeq(\mathbb{1}\simeq\mathbb{1})$ which is true.
+\item Take $X = \mathbb{1}$ and $P = \lambda \_. \mathbb{0}$. The statement
+  $\textit{is-univalent-fibration}~P$ reduces to
+  $(\star\equiv\star)\simeq(\mathbb{0}\simeq\mathbb{0})$ which is also true.
+\item Take $X = \mathbb{1}$. For no $P$ other than the above two instances will
+  we have a univalent fibration.
+\item Take $X = \mathbb{2}$ with elements $\fc$ and $\tc$ and
+  $P = \lambda \_. \mathbb{1}$. The statement
+  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
+  (\mathbb{1}\simeq\mathbb{1})$ which is false.
+\item Take $X = \mathbb{2}$ with elements $\fc$ and $\tc$ and
+  $P = \lambda \{ \fc \to \mathbb{0}; \tc \to \mathbb{1} \}$ The statement
+  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
+  (\mathbb{0}\simeq\mathbb{1})$ which is true. That is the only choice of $P$
+  that works (modulo symmetry).
+\item For any $n > 2$, taking $X$ as sum of $n$ copies of $\mathbb{1}$ can never
+  give a univalent fibration.
+\end{itemize}
+
+\begin{theorem}
+  Let $X$ be path-connected. Then $X$ is the base of a univalent fibration only
+  if there exists an $F$ such that $X \simeq \Sigma_{Y : \Typ} \| Y \equiv F \|$
+\end{theorem}
+
+\noindent In particular $\Sigma_{Y : \Typ} \| Y \equiv \mathbb{2} \|$ is the base of a
+univalent fibration. In some sense this is the simplest univalent universe, our
+featherweight HoTT that we aim to characterize as a reversible programming
+language based on type isomorphisms.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{The Bool Case}
@@
                \alt & \swap &:& \mathbb{2} \iso \mathbb{2} \\
-               \alt & \circ &:& (\tau_1 \iso \tau_2) \rightarrow (\tau_2 \iso \tau_3)
-                              \rightarrow (\tau_1 \iso \tau_3)
+               \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3)
+                              \to (\tau_1 \iso \tau_3)
                \end{array} \\
 \\
-! &:& (\tau_1 \iso \tau_2) \rightarrow (\tau_2 \iso \tau_1) \\
+! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
 \invc{\id} &=& \id \\
@@
                \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
-               \alt & \bullet &:& (c_1 \isotwo c_2) \rightarrow (c_2 \isotwo c_3)
-                                            \rightarrow (c_1 \isotwo c_3) \\
-               \alt & \odot &:& (c_1 \isotwo c_1') \rightarrow (c_2 \isotwo c_2')
-                                            \rightarrow (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
+               \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3)
+                                            \to (c_1 \isotwo c_3) \\
+               \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2')
+                                            \to (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
              \end{array}
 \end{array}\]
-The 2-combinators $\alpha$ also have inverses \ldots
+
+The 2-combinators $\alpha$ also have inverses. The above forms a groupoid (see
+accompanying code).
 
@@
 
-Now we claim that $\mathcal{U}_\Pi$ is equivalent to $\mathcal{U}$ where $\mathcal{U}_\Pi$ is a universe which includes types $\tau$ as a points, combinators $c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators $\alpha : c_1 \isotwo c_2$ as 2-paths.
+Now we claim that $\mathcal{U}_\Pi$ is equivalent to $\mathcal{U}$ where
+$\mathcal{U}_\Pi$ is a universe which includes types $\tau$ as a points,
+combinators $c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators $\alpha : c_1
+\isotwo c_2$ as 2-paths. The proof is in a meta-logic which is itself univalent.
 
@@
 %%%%%%%%%
-\section{Generalizations}
+\section{Generalization I}
+
+Use all finite types instead of just Bool
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%
+\section{Generalization II}
+
+Add a HIT to the univalent universe, perhaps something to do with fractionals.
 

### 2017-02-21 upi.tex
unpacking {2} a little

@@
 \newcommand{\reflp}{\textsc{refl}}
+\newcommand{\notp}{\textsc{not}}
+\newcommand{\gluep}{\textsc{glue}}
 \newcommand{\reflh}{\mathit{refl}_{\sim}}
@@
 \newcommand{\ot}{\mathbb{1}}
+\newcommand{\bt}{\mathbb{2}}
 \newcommand{\fc}{\mathit{false}}
@@
   x \simeq y$ which is the usual definition of univalence.
-\item Take $X = \mathbb{1}$ with element $\star$ and
-  $P = \lambda \_. \mathbb{1}$. The statement
+\item Take $X = \ot$ with element $\star$ and $P = \lambda \_. \ot$. The
+  statement $\textit{is-univalent-fibration}~P$ reduces to
+  $(\star\equiv\star)\simeq(\ot\simeq\ot)$ which is true.
+\item Take $X = \ot$ and $P = \lambda \_. \zt$. The statement
   $\textit{is-univalent-fibration}~P$ reduces to
-  $(\star\equiv\star)\simeq(\mathbb{1}\simeq\mathbb{1})$ which is true.
-\item Take $X = \mathbb{1}$ and $P = \lambda \_. \mathbb{0}$. The statement
-  $\textit{is-univalent-fibration}~P$ reduces to
-  $(\star\equiv\star)\simeq(\mathbb{0}\simeq\mathbb{0})$ which is also true.
-\item Take $X = \mathbb{1}$. For no $P$ other than the above two instances will
+  $(\star\equiv\star)\simeq(\zt\simeq\zt)$ which is also true.
+\item Take $X = \ot$. For no $P$ other than the above two instances will
   we have a univalent fibration.
-\item Take $X = \mathbb{2}$ with elements $\fc$ and $\tc$ and
-  $P = \lambda \_. \mathbb{1}$. The statement
+\item Take $X = \bt$ with elements $\fc$ and $\tc$ and
+  $P = \lambda \_. \ot$. The statement
   $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
-  (\mathbb{1}\simeq\mathbb{1})$ which is false.
-\item Take $X = \mathbb{2}$ with elements $\fc$ and $\tc$ and
-  $P = \lambda \{ \fc \to \mathbb{0}; \tc \to \mathbb{1} \}$ The statement
+  (\ot\simeq\ot)$ which is false.
+\item Take $X = \bt$ with elements $\fc$ and $\tc$ and
+  $P = \lambda \{ \fc \to \zt; \tc \to \ot \}$ The statement
   $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
-  (\mathbb{0}\simeq\mathbb{1})$ which is true. That is the only choice of $P$
+  (\zt\simeq\ot)$ which is true. That is the only choice of $P$
   that works (modulo symmetry).
-\item For any $n > 2$, taking $X$ as sum of $n$ copies of $\mathbb{1}$ can never
+\item For any $n > 2$, taking $X$ as sum of $n$ copies of $\ot$ can never
   give a univalent fibration.
@@
   Let $X$ be path-connected. Then $X$ is the base of a univalent fibration only
-  if there exists an $F$ such that $X \simeq \Sigma_{Y : \Typ} \| Y \equiv F \|$
+  if there exists an $F$ such that $X \simeq \Sigma_{Y : \Typ} \| Y \equiv F
+  \|$. We will refer to this type as $\{F\}$.
 \end{theorem}
 
-\noindent In particular $\Sigma_{Y : \Typ} \| Y \equiv \mathbb{2} \|$ is the base of a
-univalent fibration. In some sense this is the simplest univalent universe, our
+\noindent In particular $\{\bt\}$, i.e.,
+$\Sigma_{Y : \Typ} \| Y \equiv \bt \|$ is the base of a univalent
+fibration. In some sense this is the simplest univalent universe, our
 featherweight HoTT that we aim to characterize as a reversible programming
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Some Simple Cases}
+
+Pick $F = \ot$. We have $\{\ot\} \simeq \ot$. Another
+simple but less trivial case is $F = S^1$. In that case we also have
+$\{S^1\} \simeq S^1$.p
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{The Bool Case}
 
-To get started let's look at the case of $\mathbb{2}$ instead of the entire set of finite types defined using $0$, $1$, $+$, and $*$.
+To get started let's look at the case of $\bt$ instead of the entire set of finite types defined using $0$, $1$, $+$, and $*$.
 
@@
 \[\begin{array}{rcl}
-\tau &::=& \mathbb{2} \\
+\tau &::=& \bt \\
 \\
 v &::=& \begin{array}[t]{lrcl}
-                    & \fc &:& \mathbb{2} \\
-              \alt & \tc &:& \mathbb{2}
+                    & \fc &:& \bt \\
+              \alt & \tc &:& \bt
                \end{array} \\
@@
               & \id &:& \tau \iso \tau \\
-               \alt & \swap &:& \mathbb{2} \iso \mathbb{2} \\
+               \alt & \swap &:& \bt \iso \bt \\
                \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3)
@@
 
-The 2-combinators $\alpha$ also have inverses. The above forms a groupoid (see
-accompanying code).
+The 2-combinators $\alpha$ also have inverses. The above forms a 2-groupoid (see
+accompanying code once it is fixed).
 
@@
 We assume an ambient univalent universe defined like in the HoTT book and including dependent pairs, a unit type, coproducts, identity types, and constants $\textsc{funext}$ and $\textsc{univalence}$.
-Our universe $\mathcal{U}$ is the subuniverse of this ambient universe defined as $\Sigma_{X:\mathbf{Type}} \| X \equiv \mathbb{2} \|$ where:
+Our universe $\mathcal{U}$ is the subuniverse of this ambient universe defined as $\Sigma_{X:\mathbf{Type}} \| X \equiv \bt \|$ where:
 
@@
 
-\item $\mathbb{2}$ is the type of booleans which comes with the two points, an induction principle, and computational rules.
+\item $\bt$ is the type of booleans which comes with the two points, an induction principle, and computational rules.
 
@@
 \item $\| A \|$ is the propositional truncation of $A$ defined as follows: for
-  any $x:A$ we have $|x| : \|A\|$, and for any $x,y : \|A\|$, we have $x \equiv y$.
+  any $x:A$ we have $|x| : \|A\|$, and for any $x,y : \|A\|$, we have
+  $\gluep : x \equiv y$.
 
@@
 $\mathcal{U}_\Pi$ is a universe which includes types $\tau$ as a points,
-combinators $c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators $\alpha : c_1
-\isotwo c_2$ as 2-paths. The proof is in a meta-logic which is itself univalent.
+combinators $c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators
+$\alpha : c_1 \isotwo c_2$ as 2-paths. The proof is in a meta-logic which is
+itself univalent.
+
+We begin by unpacking the objects, 1-paths, and 2-paths of $\{\bt\}$.
+
+\paragraph*{Objects in $\{\bt\}$.} The only term in $\{\bt\}$ is
+$(\bt,\|\reflp_{\bt}\|)$ which we call $`\bt$. Formally we can prove
+$\|(X,p) \equiv `\bt\|$ is inhabited for all $X$ and $p$.
+
+\paragraph*{1-Paths in $\{\bt\}$.} These are paths between $`\bt$ and
+$`\bt$. Generally speaking, paths between elements of a $\Sigma$-type are pairs
+of paths with a transport in the second component. If we have a type $A$ with
+points $a$ and $b$ and a path $p : a \equiv b$. Fix some $c$ and consider the
+dependent function $(x:A) \rightarrow (x \equiv c)$. This forms the types
+$a\equiv c$ and $b\equiv c$ with elements $ua$ and $ub$. Now a path between
+$(a,ua)$ and $(b,ub)$ is a pair $(p,\alpha)$ where $\alpha : p_* ua \equiv
+ub$. Hence paths between $`\bt$ and $`\bt$ are going to be for the form
+$(p,\alpha)$ where $p : \bt \equiv \bt$ and $\alpha$ is essentially $\gluep$
+from the definition of truncation. So we have two paths:
+\[\begin{array}{rcl}
+`\mathbf{id} &=& (\reflp,\gluep) \\
+`\mathbf{not} &=& (\notp,\gluep)
+\end{array}\]
+
+\paragraph*{2-Paths in $\{\bt\}$.} Now are considering paths between
+$`\mathbf{id}$ and $`\mathbf{id}$, and between $`\mathbf{not}$ and
+$`\mathbf{not}$. We have at least one path at this level which is
+$(\reflp,\gluep)$ but to show that that this is the only path we will have to
+first prove that $`\mathbf{not} \circ `\mathbf{not} \equiv `\mathbf{id}$.
 

### 2017-03-02 upi.tex
starting to explain things carefully

@@
 \usepackage{comment}
+\usepackage{tikz}
 
@@
 \end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Preliminaries}
+
+[Following the HoTT book and the slides on ``A Characterization of Univalent
+Fibrations'' by Dan Christensen.]
+
+We think of a type family $P : A \to \mathcal{U}$ as a \emph{fibration} with
+base space $A$, with $P(x)$ being the fiber over $x$, and with
+$\Sigma_{(x:A)} P(x)$ being the \emph{total space} of the fibration, and with
+first projection $\Sigma_{(x:A)} P(x) \to A$. In the figure below, the pairs
+$(x,u)$ and $(y,p_*(y))$ are in the total space. A critical property of
+fibrations is that a path $p : x \equiv y$ in the base space $A$ and a point $u$
+in the fiber over $x$ induce a function that \emph{lifts} the path to the total
+space: the lifted path starts at $(x,u)$ and ends at $(y,p_*(u))$ for some point
+$p_*(u)$ in the fiber over $y$ that depends on $p$ and $u$.
+
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
+  \draw (-3,0) ellipse (1.5cm and 3cm);
+  \draw (3,2) ellipse (0.5cm and 1cm);
+  \draw (3,-2) ellipse (0.5cm and 1cm);
+  \node[blue,ultra thick,above] at (-3,3) {$A$};
+  \node[blue,ultra thick,above] at (3,3) {$P(x)$ (fiber over $x$)};
+  \node[blue,ultra thick,below] at (3,-3) {$P(y)$ (fiber over $y$)};
+  \draw[fill] (-3,1.5) circle [radius=0.025];
+  \draw[fill] (-3,-1.5) circle [radius=0.025];
+  \draw[left,cyan,thick] (-3,1.5) -- (-3,-1.5);
+  \node[above] at (-3,1.5) {$x$};
+  \node[below] at (-3,-1.5) {$y$};
+  \draw[fill] (3,1.8) circle [radius=0.025];
+  \draw[fill] (3,-1.8) circle [radius=0.025];
+  \node[above] at (3,1.8) {$u$};
+  \node[below] at (3,-1.8) {$p_*(u)$};
+  \node[left,cyan] at (-3,0) {$p$};
+  \draw[->,red,dashed,ultra thick] (-3,1.5) to [out=45, in=135] (2.4,2.5);
+  \draw[->,red,dashed,ultra thick] (-3,-1.5) to [out=-45, in=-135] (2.4,-2.5);
+\end{tikzpicture}
+\end{center}
 

### 2017-03-02 upi.tex
a few more definitions

@@
 
-We think of a type family $P : A \to \mathcal{U}$ as a \emph{fibration} with
-base space $A$, with $P(x)$ being the fiber over $x$, and with
-$\Sigma_{(x:A)} P(x)$ being the \emph{total space} of the fibration, and with
-first projection $\Sigma_{(x:A)} P(x) \to A$. In the figure below, the pairs
-$(x,u)$ and $(y,p_*(y))$ are in the total space. A critical property of
-fibrations is that a path $p : x \equiv y$ in the base space $A$ and a point $u$
-in the fiber over $x$ induce a function that \emph{lifts} the path to the total
-space: the lifted path starts at $(x,u)$ and ends at $(y,p_*(u))$ for some point
-$p_*(u)$ in the fiber over $y$ that depends on $p$ and $u$.
+\paragraph*{Type Family.} A dependent function $P : A \to \mathcal{U}$ defines a
+\emph{type family}, $P(x)$ indexed by $x:A$.
+
+\paragraph*{Transport.} Given a type family $P : A \to \mathcal{U}$, a path
+$p : x \equiv y$ in $A$ induces a function $p_* : P(x) \to P(y)$ which
+transports points in $P(x)$ to points in $P(y)$.
+
+\paragraph*{Fibration.} We think of a type family $P : A \to \mathcal{U}$ as a
+\emph{fibration} with base space $A$, with $P(x)$ being the fiber over $x$, and
+with $\Sigma_{(x:A)} P(x)$ being the \emph{total space} of the fibration, and
+with first projection $\Sigma_{(x:A)} P(x) \to A$. In the figure below, $p_*(u)$
+is the transport of $u$ along $p$ and both pairs $(x,u)$ and $(y,p_*(y))$ are in
+the total space. A critical property of fibrations is that a path
+$p : x \equiv y$ in the base space $A$ and a point $u$ in the fiber over $x$
+induce a function that \emph{lifts} the path to the total space: the lifted path
+starts at $(x,u)$ and ends at $(y,p_*(u))$. A function $f : \Pi_{(x:A)} P(x)$ is
+sometimes called a \emph{section} of the fibration $P$.
 
@@
 \end{center}
+
+\paragraph*{Univalent Type Family.} A type family $P : A \to \mathcal{U}$ is
+\emph{univalent} if the following function is an equivalence:
+\[\begin{array}{rcl}
+f & : & (x \equiv y) \to (P(x) \simeq P(y)) \\
+\end{array}\]
+where $x,y:A$.
 

### 2017-03-02 upi.tex
a few more definitions

@@
 \[\begin{array}{rcl}
-f & : & (x \equiv y) \to (P(x) \simeq P(y)) \\
+V_P & : & (x \equiv y) \to (P(x) \simeq P(y)) \\
+V_P &=& \lambda p. \mathit{idtoequiv}(\mathit{ap}_{P}(p))
 \end{array}\]
-where $x,y:A$.
+where $x,y:A$ and the standard constructions from the book
+$\mathit{ap} : (x \equiv y) \to (P(x) \equiv P(y)$ and
+$\mathit{idtoequiv} : (A \equiv B) \to (A \simeq B)$.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Univalence}
+\paragraph*{Example I.} Consider the type family
+$\mathit{Id} : \mathcal{U} \to \mathcal{U}$. In this case, the function
+$V_{\mathit{Id}}$ is $\mathit{idtoequiv}$ and the definition of univalent type
+family reduces to the conventional univalence axiom.
+
+\newpage
 

### 2017-03-06 upi.tex
cleaning up definitions

--- 2017-03-02 upi.tex
@@
   \node[below] at (-3,-1.5) {$y$};
-  \draw[fill] (3,1.8) circle [radius=0.025];
+  \draw[fill] (3,1.5) circle [radius=0.025];
+  \draw[fill] (3,2.1) circle [radius=0.025];
   \draw[fill] (3,-1.8) circle [radius=0.025];
-  \node[above] at (3,1.8) {$u$};
+  \node[above] at (3,1.5) {$u$};
+  \node[above] at (3,2.1) {$v$};
   \node[below] at (3,-1.8) {$p_*(u)$};
@@
 
-\paragraph*{Univalent Type Family.} A type family $P : A \to \mathcal{U}$ is
-\emph{univalent} if the following function is an equivalence:
+It is worth repeating Remark 2.7.1 from the book. In the figure above, we could
+have $(x,u) \equiv (x,v)$ without having $u \equiv v$. The only thing we can
+conclude is that there is a (possibly non-trivial) path path $q : x \equiv x$
+such that $q_*(u) \equiv v$.
+
+\paragraph*{Univalent Type Family.}
+Given standard constructions in the book, we have a function:
 \[\begin{array}{rcl}
-f & : & (x \equiv y) \to (P(x) \simeq P(y)) \\
+\textit {transport-equiv} &:& \Pi_{P : A \to \mathcal{U}}~ \Pi_{x,y:A}~
+    x \equiv y  \to P x \simeq P y \\
+\textit{transport-equiv}~P~x~y &=& \lambda p. \mathit{idtoequiv}(\mathit{ap}_{P}(p))
 \end{array}\]
-where $x,y:A$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Univalence}
-
-Let $X : \Typ$. We have:
-\[
-\textit{transport-equiv} : \Pi_{P : X \to \Typ}~ \Pi_{x,y:X}~
-    x \equiv y  \to P x \simeq P y
-\]
-For a fibration with base $X$, define:
+with the standard definitions for
+$\mathit{ap} : (x \equiv y) \to (P(x) \equiv P(y))$ and
+$\mathit{idtoequiv} : (A \equiv B) \to (A \simeq B)$. We say a type family
+$P : A \to \mathcal{U}$ is \emph{univalent} if for all $x,y:A$
+the function $\textit{transport-equiv}~P~x~y$ is an equivalence. Formally:
 \[\begin{array}{l}
-\textit{is-univalent-fibration} : (X \to \Typ) \to \Typ \\
-\textit{is-univalent-fibration}~P =
-    \Pi_{x,y:X}~\textit{is-equiv} (\textit{transport-equiv}~P~x~y)
+\textit{is-univalent-fibration} : (A \to \mathcal{U}) \to \mathcal{U} \\
+\textit{is-univalent-fibration}~P = \Pi_{(x,y:A)} \textit{is-equiv}~(\textit{transport-equiv}~P~x~y)
 \end{array}\]
@@
 \begin{itemize}
-\item Take $X = \Typ$ and $P = \mathit{Id}_{\Typ}$. The statement
-  $\textit{is-univalent-fibration}~P$ reduces to $\Pi_{x,y:\Typ}~x \equiv y \to
+\item Take $A = \mathcal{U}$ and $P = \mathit{Id}_{\mathcal{U}}$. The statement
+  $\textit{is-univalent-fibration}~P$ reduces to $\Pi_{x,y:\mathcal{U}}~x \equiv y \to
   x \simeq y$ which is the usual definition of univalence.
+
+\newpage
+
 \item Take $X = \ot$ with element $\star$ and $P = \lambda \_. \ot$. The
@@
   Let $X$ be path-connected. Then $X$ is the base of a univalent fibration only
-  if there exists an $F$ such that $X \simeq \Sigma_{Y : \Typ} \| Y \equiv F
+  if there exists an $F$ such that $X \simeq \Sigma_{Y : \mathcal{U}} \| Y \equiv F
   \|$. We will refer to this type as $\{F\}$.
@@
 \noindent In particular $\{\bt\}$, i.e.,
-$\Sigma_{Y : \Typ} \| Y \equiv \bt \|$ is the base of a univalent
+$\Sigma_{Y : \mathcal{U}} \| Y \equiv \bt \|$ is the base of a univalent
 fibration. In some sense this is the simplest univalent universe, our

### 2017-03-06 upi.tex
better figure

@@
 \usepackage{tikz}
+
+\makeatletter
+\tikzset{my loop/.style =  {to path={
+  \pgfextra{\let\tikztotarget=\tikztostart}
+%%  [looseness=100, in=60,out=120]
+%%  [looseness=8,min distance=5mm, in=60,out=120]
+  \tikz@to@curve@path}
+  }}
+\makeatletter
 
@@
   \draw (-3,0) ellipse (1.5cm and 3cm);
-  \draw (3,2) ellipse (0.5cm and 1cm);
+  \draw (3,2) ellipse (0.8cm and 2cm);
   \draw (3,-2) ellipse (0.5cm and 1cm);
   \node[blue,ultra thick,above] at (-3,3) {$A$};
-  \node[blue,ultra thick,above] at (3,3) {$P(x)$ (fiber over $x$)};
+  \node[blue,ultra thick,above] at (3,3.9) {$P(x)$ (fiber over $x$)};
   \node[blue,ultra thick,below] at (3,-3) {$P(y)$ (fiber over $y$)};
@@
   \draw[left,cyan,thick] (-3,1.5) -- (-3,-1.5);
-  \node[above] at (-3,1.5) {$x$};
-  \node[below] at (-3,-1.5) {$y$};
-  \draw[fill] (3,1.5) circle [radius=0.025];
-  \draw[fill] (3,2.1) circle [radius=0.025];
+  \node[left] (X) at (-3,1.5) {$x$};
+  \path[cyan,thick] (X) edge [loop above, looseness=8, in=40, out=140] node[above] {$q$} (X);
+  \node[left] at (-3,-1.5) {$y$};
   \draw[fill] (3,-1.8) circle [radius=0.025];
-  \node[above] at (3,1.5) {$u$};
-  \node[above] at (3,2.1) {$v$};
+  \draw[fill] (3,3) circle [radius=0.025];
+  \node[above] at (3,3) {$u$};
+  \draw[fill] (3,2.2) circle [radius=0.025];
+  \node[below] at (3,2.2) {$q_*(u)$};
+  \draw[fill] (3,1) circle [radius=0.025];
+  \node[below] at (3,1) {$v$};
   \node[below] at (3,-1.8) {$p_*(u)$};
   \node[left,cyan] at (-3,0) {$p$};
-  \draw[->,red,dashed,ultra thick] (-3,1.5) to [out=45, in=135] (2.4,2.5);
-  \draw[->,red,dashed,ultra thick] (-3,-1.5) to [out=-45, in=-135] (2.4,-2.5);
+  \draw[->,red,dashed,ultra thick] (-3,1.5) to [out=45, in=135] (2.2,2.5);
+  \draw[->,red,dashed,ultra thick] (-3,-1.5) to [out=-45, in=-135] (2.5,-2.5);
 \end{tikzpicture}
@@
 
-\item Take $X = \ot$ with element $\star$ and $P = \lambda \_. \ot$. The
+\item Take $A = \ot$ with element $\star$ and $P = \lambda \_. \ot$. The
   statement $\textit{is-univalent-fibration}~P$ reduces to
   $(\star\equiv\star)\simeq(\ot\simeq\ot)$ which is true.
-\item Take $X = \ot$ and $P = \lambda \_. \zt$. The statement
+\item Take $A = \ot$ and $P = \lambda \_. \zt$. The statement
   $\textit{is-univalent-fibration}~P$ reduces to
   $(\star\equiv\star)\simeq(\zt\simeq\zt)$ which is also true.
-\item Take $X = \ot$. For no $P$ other than the above two instances will
+\item Take $A = \ot$. For no $P$ other than the above two instances will
   we have a univalent fibration.
-\item Take $X = \bt$ with elements $\fc$ and $\tc$ and
+\item Take $A = \bt$ with elements $\fc$ and $\tc$ and
   $P = \lambda \_. \ot$. The statement
@@
   (\ot\simeq\ot)$ which is false.
-\item Take $X = \bt$ with elements $\fc$ and $\tc$ and
+\item Take $A = \bt$ with elements $\fc$ and $\tc$ and
   $P = \lambda \{ \fc \to \zt; \tc \to \ot \}$ The statement
@@
   that works (modulo symmetry).
-\item For any $n > 2$, taking $X$ as sum of $n$ copies of $\ot$ can never
+\item For any $n > 2$, taking $A$ as sum of $n$ copies of $\ot$ can never
   give a univalent fibration.

### 2017-03-06 upi.tex
correcting notation

@@
 \usepackage{tikz}
-
-\makeatletter
-\tikzset{my loop/.style =  {to path={
-  \pgfextra{\let\tikztotarget=\tikztostart}
-%%  [looseness=100, in=60,out=120]
-%%  [looseness=8,min distance=5mm, in=60,out=120]
-  \tikz@to@curve@path}
-  }}
-\makeatletter
 
@@
 \[\begin{array}{rcl}
-\textit {transport-equiv} &:& \Pi_{P : A \to \mathcal{U}}~ \Pi_{x,y:A}~
+\textit {transport-equiv} &:& \Pi_{(P : A \to \mathcal{U})}~ \Pi_{(x,y:A)}~
     x \equiv y  \to P x \simeq P y \\
-\textit{transport-equiv}~P~x~y &=& \lambda p. \mathit{idtoequiv}(\mathit{ap}_{P}(p))
+\textit{transport-equiv}~P~x~y &=& \lambda p. \mathit{idtoeqv}(\mathit{ap}_{P}(p))
 \end{array}\]
@@
 $\mathit{ap} : (x \equiv y) \to (P(x) \equiv P(y))$ and
-$\mathit{idtoequiv} : (A \equiv B) \to (A \simeq B)$. We say a type family
+$\mathit{idtoeqv} : (A \equiv B) \to (A \simeq B)$. We say a type family
 $P : A \to \mathcal{U}$ is \emph{univalent} if for all $x,y:A$
@@
 \item Take $A = \mathcal{U}$ and $P = \mathit{Id}_{\mathcal{U}}$. The statement
-  $\textit{is-univalent-fibration}~P$ reduces to $\Pi_{x,y:\mathcal{U}}~x \equiv y \to
-  x \simeq y$ which is the usual definition of univalence.
+  $\textit{is-univalent-fibration}~P$ reduces to
+  $\Pi_{(x,y:\mathcal{U})}~\textit{is-equiv}~\mathit {idtoeqv}$ which is the
+  usual definition of univalence.
 

### 2017-03-07 upi.tex
cleaned up definitions

@@
 
-\title{Univalent Universes and Completness of Reversible Programs: \\
+\title[Featherweight HoTT]{Univalent Universes and Completness of Reversible Programs: \\
   Featherweight HoTT}
@@
 
-[Following the HoTT book and the slides on ``A Characterization of Univalent
-Fibrations'' by Dan Christensen.]
+Following the HoTT book and the slides on ``A Characterization of Univalent
+Fibrations'' by Dan Christensen.
+
+%%%%%
+\subsection{Definitions}
 
@@
 
-\paragraph*{Transport.} Given a type family $P : A \to \mathcal{U}$, a path
-$p : x \equiv y$ in $A$ induces a function $p_* : P(x) \to P(y)$ which
-transports points in $P(x)$ to points in $P(y)$.
+\paragraph*{Functors (Lemma 2.2.1 in the book)} Suppose $f : A \to B$ is a
+function. Then for any $x,y:A$ there is an operation
+$\mathit{ap}_f : (x \equiv y) \to (f(x) \equiv f(y))$.
 
-\paragraph*{Fibration.} We think of a type family $P : A \to \mathcal{U}$ as a
-\emph{fibration} with base space $A$, with $P(x)$ being the fiber over $x$, and
-with $\Sigma_{(x:A)} P(x)$ being the \emph{total space} of the fibration, and
-with first projection $\Sigma_{(x:A)} P(x) \to A$. In the figure below, $p_*(u)$
-is the transport of $u$ along $p$ and both pairs $(x,u)$ and $(y,p_*(y))$ are in
-the total space. A critical property of fibrations is that a path
-$p : x \equiv y$ in the base space $A$ and a point $u$ in the fiber over $x$
-induce a function that \emph{lifts} the path to the total space: the lifted path
-starts at $(x,u)$ and ends at $(y,p_*(u))$. A function $f : \Pi_{(x:A)} P(x)$ is
-sometimes called a \emph{section} of the fibration $P$.
+\paragraph*{Equivalences (Chapter 4 in the book)} For any two types $A$ and $B$
+we can form the type $A \simeq B = \Sigma_{(f : A \to B)} \mathit{isequiv}~f$ of
+equivalences. Each equivalence consists of a function $f : A \to B$ together
+with evidence that this function is an equivalence.
+
+\paragraph*{Paths to Equivalences (Lemma 2.10.1 in the book)} Given two types
+$A$ and $B$, there is a function
+$\mathit{idtoeqv} : (A \equiv B) \to (A \simeq B)$ mapping paths to
+equivalences.
+
+\paragraph*{Transport (Lemma 2.3.1 in the book)} Given a type family
+$P : A \to \mathcal{U}$, a path $p : x \equiv y$ in $A$ induces a function
+$p_* : P(x) \to P(y)$ which transports points in $P(x)$ to points in $P(y)$.
+
+\paragraph*{Fibration (Sections 2.3 and 2.7 in the book)} We think of a type
+family $P : A \to \mathcal{U}$ as a \emph{fibration} with base space $A$, with
+$P(x)$ being the fiber over $x$, and with $\Sigma_{(x:A)} P(x)$ being the
+\emph{total space} of the fibration, and with first projection
+$\Sigma_{(x:A)} P(x) \to A$. In the figure below, $p_*(u)$ is the transport of
+$u$ along $p$ and both pairs $(x,u)$ and $(y,p_*(y))$ are in the total space. A
+critical property of fibrations is that a path $p : x \equiv y$ in the base
+space $A$ and a point $u$ in the fiber over $x$ induce a function that
+\emph{lifts} the path to the total space: the lifted path starts at $(x,u)$ and
+ends at $(y,p_*(u))$. A function $f : \Pi_{(x:A)} P(x)$ is sometimes called a
+\emph{section} of the fibration $P$.
 
@@
 
-It is worth repeating Remark 2.7.1 from the book. In the figure above, we could
-have $(x,u) \equiv (x,v)$ without having $u \equiv v$. The only thing we can
-conclude is that there is a (possibly non-trivial) path path $q : x \equiv x$
-such that $q_*(u) \equiv v$.
+\noindent It is worth repeating Remark 2.7.1 from the book. In the figure above,
+we could have $(x,u) \equiv (x,v)$ without necessarily having $u \equiv v$. The
+only thing we can conclude regarding $u$ and $v$ is that there is a (possibly
+non-trivial) path $q : x \equiv x$ such that $q_*(u) \equiv v$.
+
+\paragraph*{Propositional Truncation.} Given a type $A : \mathcal{U}$, the type
+$\|A\|$ is the propositional truncation of $A$ defined as follows: for
+  any $x:A$ we have $|x| : \|A\|$, and for any $x,y : \|A\|$, we have
+  $\gluep : x \equiv y$.
 
 \paragraph*{Univalent Type Family.}
-Given standard constructions in the book, we have a function:
+For each fibration, we have a function:
 \[\begin{array}{rcl}
 \textit {transport-equiv} &:& \Pi_{(P : A \to \mathcal{U})}~ \Pi_{(x,y:A)}~
-    x \equiv y  \to P x \simeq P y \\
+    x \equiv y  \to P(x) \simeq P(y) \\
 \textit{transport-equiv}~P~x~y &=& \lambda p. \mathit{idtoeqv}(\mathit{ap}_{P}(p))
 \end{array}\]
-with the standard definitions for
-$\mathit{ap} : (x \equiv y) \to (P(x) \equiv P(y))$ and
-$\mathit{idtoeqv} : (A \equiv B) \to (A \simeq B)$. We say a type family
-$P : A \to \mathcal{U}$ is \emph{univalent} if for all $x,y:A$
-the function $\textit{transport-equiv}~P~x~y$ is an equivalence. Formally:
+We say a type family $P : A \to \mathcal{U}$ is \emph{univalent} if for all
+$x,y:A$ the function $\textit{transport-equiv}~P~x~y$ is an
+equivalence. Formally:
 \[\begin{array}{l}
 \textit{is-univalent-fibration} : (A \to \mathcal{U}) \to \mathcal{U} \\
-\textit{is-univalent-fibration}~P = \Pi_{(x,y:A)} \textit{is-equiv}~(\textit{transport-equiv}~P~x~y)
+\textit{is-univalent-fibration}~P = \Pi_{(x,y:A)} \textit{isequiv}~(\textit{transport-equiv}~P~x~y)
 \end{array}\]
 
-\noindent\paragraph*{Examples}
+%%%%%
+\subsection{Examples}
+
+We will choose particular base spaces $A$ and fibrations $P : A \to \mathcal{U}$
+below and reason about whether they are univalent or not. For each fibration,
+the question of whether it is univalent reduces to whether, for all $x,y:A$, the
+function $\textit{transport-equiv}~P~x~y$ is an equivalence. In other words, the
+question we are asking is whether we have an equivalence
+$(x \equiv y) \simeq (P(x) \simeq P(y))$ witnessed by the function
+$\textit{transport-equiv}~P~x~y$.
+
 \begin{itemize}
-\item Take $A = \mathcal{U}$ and $P = \mathit{Id}_{\mathcal{U}}$. The statement
-  $\textit{is-univalent-fibration}~P$ reduces to
-  $\Pi_{(x,y:\mathcal{U})}~\textit{is-equiv}~\mathit {idtoeqv}$ which is the
-  usual definition of univalence.
 
-\newpage
+\item Take $A = \mathcal{U}$ and $P = \mathit{Id}_{\mathcal{U}}$. By definition,
+  this fibration is univalent if
+  $\textit{transport-equiv}~\mathit{Id}_{\mathcal{U}}~x~y = \mathit{idtoeqv}$ of
+    type $x \equiv y \to x \simeq y$ is an equivalence. As Axiom 2.10.3 in the
+    book states this is exactly the statement of the conventional univalence
+    axiom for universe $\mathcal{U}$.
 
-\item Take $A = \ot$ with element $\star$ and $P = \lambda \_. \ot$. The
-  statement $\textit{is-univalent-fibration}~P$ reduces to
-  $(\star\equiv\star)\simeq(\ot\simeq\ot)$ which is true.
-\item Take $A = \ot$ and $P = \lambda \_. \zt$. The statement
-  $\textit{is-univalent-fibration}~P$ reduces to
-  $(\star\equiv\star)\simeq(\zt\simeq\zt)$ which is also true.
-\item Take $A = \ot$. For no $P$ other than the above two instances will
-  we have a univalent fibration.
+\item Take $A = \ot$ with element $\star$ and $P = \lambda \_. \ot$. This
+    fibration $P$ is univalent if we have an equivalence
+    $(\star\equiv\star) \simeq (\ot\simeq\ot)$ witnessed by
+    $\textit{transport-equiv}~P~x~y$. The two types are clearly equivalent in
+    this case.
+
+\item Take $A = \ot$ and $P = \lambda \_. \zt$. This fibration is univalent if
+    we have an equivalence $(\star\equiv\star)\simeq(\zt\simeq\zt)$ witnessed by
+    $\textit{transport-equiv}~P~x~y$. The two types are also clearly equivalent
+    in this case.
+
+\item Take $A = \ot$. For no $P$ other than the above two instances will we
+    have a univalent fibration.
+
 \item Take $A = \bt$ with elements $\fc$ and $\tc$ and
@@
   (\ot\simeq\ot)$ which is false.
+
 \item Take $A = \bt$ with elements $\fc$ and $\tc$ and
-  $P = \lambda \{ \fc \to \zt; \tc \to \ot \}$ The statement
-  $\textit{is-univalent-fibration}~P$ reduces to $(\fc \equiv \tc) \simeq
-  (\zt\simeq\ot)$ which is true. That is the only choice of $P$
-  that works (modulo symmetry).
+  $P = \lambda \{ \fc \to \zt; \tc \to \ot \}$. This fibration is univalent if
+  we have an equivalence $(\fc \equiv \tc) \simeq (\zt\simeq\ot)$ which is
+  true. That is the only choice of $P$ that works (modulo symmetry).
+
 \item For any $n > 2$, taking $A$ as sum of $n$ copies of $\ot$ can never
   give a univalent fibration.
+
+\item For any type $F$ we define
+  $\{F\} = \Sigma_{A : \mathcal{U}} \| F \simeq A \|$. By a proposition of
+  Christensen, any fibration with base space $\{F\}$, i.e., any fibration
+  $\{F\} \to \mathcal{U}$, is univalent. In particular, we can take the base
+  space to be:
+  \begin{itemize}
+  \item $\{\ot\}$. The type $\{\ot\}$ is actually the same as $\ot$ and this was
+    covered by previous examples.
+  \item $\{S^1\}$. The type $\{S^1\}$ is the same as $S^1$. This is perhaps the
+    simplest base space to choose but it does not naturally correspond to a
+    conventional programming language.
+  \item $\{\bt\}$. In some sense this is the simplest
+    programming-language--relevant universe --- our featherweight HoTT. We aim
+    to characterize this base space as a reversible programming language based
+    on type isomorphisms.
   \end{itemize}
 
-\begin{theorem}
-  Let $X$ be path-connected. Then $X$ is the base of a univalent fibration only
-  if there exists an $F$ such that $X \simeq \Sigma_{Y : \mathcal{U}} \| Y \equiv F
-  \|$. We will refer to this type as $\{F\}$.
-\end{theorem}
-
-\noindent In particular $\{\bt\}$, i.e.,
-$\Sigma_{Y : \mathcal{U}} \| Y \equiv \bt \|$ is the base of a univalent
-fibration. In some sense this is the simplest univalent universe, our
-featherweight HoTT that we aim to characterize as a reversible programming
-language based on type isomorphisms.
+\end{itemize}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Some Simple Cases}
+\section{Analyzing $\{\bt\}$ and its Connection to $\Pi$}
 
-Pick $F = \ot$. We have $\{\ot\} \simeq \ot$. Another
-simple but less trivial case is $F = S^1$. In that case we also have
-$\{S^1\} \simeq S^1$.p
+%%%%%
+\subsection{$\{\bt\}$}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{The Bool Case}
-
-To get started let's look at the case of $\bt$ instead of the entire set of finite types defined using $0$, $1$, $+$, and $*$.
+Recall that $\{\bt\} = \Sigma_{(A:\mathcal{U})} \|A \simeq \bt\|$.
 
@@
 
-The restriction of $\Pi$ is the following:
+The restriction of $\Pi$ to the case of just booleans is the following:
 
@@
 %%%%%
-\subsection{$\mathcal{U}$}
-
-We assume an ambient univalent universe defined like in the HoTT book and including dependent pairs, a unit type, coproducts, identity types, and constants $\textsc{funext}$ and $\textsc{univalence}$.
-Our universe $\mathcal{U}$ is the subuniverse of this ambient universe defined as $\Sigma_{X:\mathbf{Type}} \| X \equiv \bt \|$ where:
-
-\begin{itemize}
-
-\item $\Sigma$ is the type former for dependent pairs: it comes equipped with an introduction rule, an induction principle, and a computational rule.
-
-\item $\bt$ is the type of booleans which comes with the two points, an induction principle, and computational rules.
-
-\item $\equiv$ is the identity type former which comes with $\reflp$, an induction principle $J$, and a computation rule.
-
-\item $\| A \|$ is the propositional truncation of $A$ defined as follows: for
-  any $x:A$ we have $|x| : \|A\|$, and for any $x,y : \|A\|$, we have
-  $\gluep : x \equiv y$.
-
-\end{itemize}
-
-%%%%%
 \subsection{Equivalence}
 
-Now we claim that $\mathcal{U}_\Pi$ is equivalent to $\mathcal{U}$ where
-$\mathcal{U}_\Pi$ is a universe which includes types $\tau$ as a points,
-combinators $c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators
+Now we claim that $\{\bt\}$ is equivalent to $\mathcal{U}_\Pi$ which is the
+universe which includes types $\tau$ as a points, combinators
+$c : \tau_1 \iso \tau_2$ as 1-paths, and 2-combinators
 $\alpha : c_1 \isotwo c_2$ as 2-paths. The proof is in a meta-logic which is

### 2017-05-02 upi.tex
removed obsolete fwhott
added many 2 combinators

@@
 %% sublime unfathomable
+%% path categorical semantics
 %% ---
 
-% \usepackage{fdsymbol}
 \usepackage{bbold}
@@
 \usepackage{tikz}
-
-%% \usepackage{hyphenat}
-%% \usepackage{color}
-%% \usepackage{url}
-%% \usepackage{amssymb}
-%% \usepackage{amsmath}
-%% \usepackage{mdwtab}
-%% \usepackage{mdwlist}
-%% \usepackage{listings}
-%% \lstMakeShortInline[columns=fullflexible]|
-%% \lstnewenvironment{code}{\lstset{basicstyle={\sffamily\footnotesize}}}{}
 
@@
 
-\title[Featherweight HoTT]{Univalent Universes and Completness of Reversible Programs: \\
-  Featherweight HoTT}
-
-\author{X}
+\title{From Reversible Programs to Univalent Universes and Back}
+\author{Jacques Carette}
 \affiliation{
-  \institution{Y}
-  \country{Z}}
-\email{A@B.C}
+  \institution{McMaster University}
+  \country{Canada}}
+  \email{carette@mcmaster.ca}
+\author{Chao-Hong Chen}
+\author{Vikraman Choudhury}
+\author{Robert Rose}
+\author{Amr Sabry}
+\affiliation{
+  \institution{Indiana University}
+  \country{United States}}
+\email{\{chen464,vikraman,rrose1,sabry\}@indiana.edu}
 
 \begin{abstract}
+  We establish a close connection between a reversible programming language
+  based on type isomorphisms and a univalent (in the sense of Homotopy Type
+  Theory) universe. The correspondence relates combinators witnessing type
+  isomorphisms in the programming language to paths in the univalent universe,
+  and combinator optimizations in the programming language to 2-paths in the
+  univalent universe. The result provides a simple computational interpretation
+  of paths, 2-paths, and univalence in terms of simple programming
+  constructs.
 \end{abstract}
@@
 
-There is a cottage industry of reversible programming languages, reversible logic, programming applications of type isomorphisms, etc.
-This work seems it should be connected to HoTT and univalence but there aren't any precise connections or theorems.
+There is a cottage industry of reversible programming languages, reversible
+logic, programming applications of type isomorphisms, etc. This work seems it
+should be connected to HoTT and univalence but there aren't any precise
+connections or theorems.
 
-Our conjecture:
+We propose a featherweight very specialized version of HoTT: a smaller
+collection of types and only reversible functions. No dependent types!  More
+precisely, our conjecture:
 
@@
 Following the HoTT book and the slides on ``A Characterization of Univalent
-Fibrations'' by Dan Christensen.
+Fibrations'' by Dan Christensen. All the definitions and most of the examples in
+this section are formalized in the accompanying Agda code.
 
@@
 equivalences. Each equivalence consists of a function $f : A \to B$ together
-with evidence that this function is an equivalence.
+with evidence $\mathit{equiv}~f$ that this function is an equivalence.
 
@@
 \item For any type $F$ we define
-  $\{F\} = \Sigma_{A : \mathcal{U}} \| F \simeq A \|$. By a proposition of
+  $\{F\} = \Sigma_{(A : \mathcal{U})} \| F \simeq A \|$. By a proposition of
   Christensen, any fibration with base space $\{F\}$, i.e., any fibration
@@
 %%%%%
-\subsection{$\{\bt\}$}
+\subsection{The Subuniverse $\{\bt\}$}
 
-Recall that $\{\bt\} = \Sigma_{(A:\mathcal{U})} \|A \simeq \bt\|$.
+Recall that $\{\bt\} = \Sigma_{(A:\mathcal{U})} \|A \simeq \bt\|$. We begin by
+unpacking the objects, 1-paths, and 2-paths of $\{\bt\}$.
+
+\paragraph*{Objects in $\{\bt\}$.} The only term in $\{\bt\}$ is
+$(\bt,\|\reflp_{\bt}\|)$ which we call $`\bt$. Formally we can prove
+$\|(X,p) \equiv `\bt\|$ is inhabited for all $X$ and $p$.
+
+\paragraph*{1-Paths in $\{\bt\}$.} These are paths between $`\bt$ and
+$`\bt$. Generally speaking, paths between elements of a $\Sigma$-type are pairs
+of paths with a transport in the second component. If we have a type $A$ with
+points $a$ and $b$ and a path $p : a \equiv b$. Fix some $c$ and consider the
+dependent function $(x:A) \rightarrow (x \equiv c)$. This forms the types
+$a\equiv c$ and $b\equiv c$ with elements $ua$ and $ub$. Now a path between
+$(a,ua)$ and $(b,ub)$ is a pair $(p,\alpha)$ where $\alpha : p_* ua \equiv
+ub$. Hence paths between $`\bt$ and $`\bt$ are going to be for the form
+$(p,\alpha)$ where $p : \bt \equiv \bt$ and $\alpha$ is essentially $\gluep$
+from the definition of truncation. So we have two paths:
+\[\begin{array}{rcl}
+`\mathbf{id} &=& (\reflp,\gluep) \\
+`\mathbf{not} &=& (\notp,\gluep)
+\end{array}\]
+
+\paragraph*{2-Paths in $\{\bt\}$.} Now are considering paths between
+$`\mathbf{id}$ and $`\mathbf{id}$, and between $`\mathbf{not}$ and
+$`\mathbf{not}$. We have at least one path at this level which is
+$(\reflp,\gluep)$ but to show that that this is the only path we will have to
+first prove that $`\mathbf{not} \circ `\mathbf{not} \equiv `\mathbf{id}$.
 
 %%%%%
-\subsection{$\Pi$}
+\subsection{The Language $\Pi$ of Type Isomorphisms}
 
@@
 
-We begin by unpacking the objects, 1-paths, and 2-paths of $\{\bt\}$.
-
-\paragraph*{Objects in $\{\bt\}$.} The only term in $\{\bt\}$ is
-$(\bt,\|\reflp_{\bt}\|)$ which we call $`\bt$. Formally we can prove
-$\|(X,p) \equiv `\bt\|$ is inhabited for all $X$ and $p$.
-
-\paragraph*{1-Paths in $\{\bt\}$.} These are paths between $`\bt$ and
-$`\bt$. Generally speaking, paths between elements of a $\Sigma$-type are pairs
-of paths with a transport in the second component. If we have a type $A$ with
-points $a$ and $b$ and a path $p : a \equiv b$. Fix some $c$ and consider the
-dependent function $(x:A) \rightarrow (x \equiv c)$. This forms the types
-$a\equiv c$ and $b\equiv c$ with elements $ua$ and $ub$. Now a path between
-$(a,ua)$ and $(b,ub)$ is a pair $(p,\alpha)$ where $\alpha : p_* ua \equiv
-ub$. Hence paths between $`\bt$ and $`\bt$ are going to be for the form
-$(p,\alpha)$ where $p : \bt \equiv \bt$ and $\alpha$ is essentially $\gluep$
-from the definition of truncation. So we have two paths:
-\[\begin{array}{rcl}
-`\mathbf{id} &=& (\reflp,\gluep) \\
-`\mathbf{not} &=& (\notp,\gluep)
-\end{array}\]
-
-\paragraph*{2-Paths in $\{\bt\}$.} Now are considering paths between
-$`\mathbf{id}$ and $`\mathbf{id}$, and between $`\mathbf{not}$ and
-$`\mathbf{not}$. We have at least one path at this level which is
-$(\reflp,\gluep)$ but to show that that this is the only path we will have to
-first prove that $`\mathbf{not} \circ `\mathbf{not} \equiv `\mathbf{id}$.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-05-03 upi.tex
started writing

@@
   univalent universe. The result provides a simple computational interpretation
-  of paths, 2-paths, and univalence in terms of simple programming
-  constructs.
+  of paths, 2-paths, and univalence in terms of familiar programming constructs.
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Story and Conjecture}
+\section{Introduction}
 
@@
 \end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{A Simple Reversible Programming Language}
+
+The practice of programming languages is replete with \emph{ad hoc} instances of
+reversible computations such as database transactions, mechanisms for data
+provenance, checkpoints, stack and exception traces, logs, backups, rollback
+recoveries, version control systems, reverse engineering, software transactional
+memories, continuations, backtracking search, and multiple-level ``undo''
+features in commercial applications. In the early nineties,
+Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
+treatment of reversibility but intensive research in full-fledged reversible
+models of computations and reversible programming languages was only sparked by
+the discovery of deep connections between physics and
+computation~\cite{Landauer:1961, PhysRevA.32.3266,Toffoli:1980,
+  bennett1985fundamental,Frank:1999:REC:930275} and by the potential for
+efficient quantum computation~\cite{springerlink:10.1007/BF02650179}.
+
+The early developments of reversible programming languages started with a
+conventional programming language, e.g., an extended λ-calculus, and either
+extended the language with a history
+mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
+imposed constraints on the conventional control flow constructs to make them
+reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
+recognize that reversible programming requires a fresh approach and should be
+designed from first principles without the detour via conventional irreversible
+languages~\cite{Yokoyama:2008:PRP, Mu:2004:ILRC, abramsky2005structural,
+  DiPierro:2006:RCL:1166042.1166047}.
+
+\begin{verbatim}
+Pi... isos, recursion, finite types, Turing complete... nice blah popl 12
+
+ESOP shows a natural 2-layer structure
+
+Examples
+
+conjecture it is related to HoTT
+
+study that language and formalize the connections between its combinators and
+between HoTT
+
+stud the smallest core that can explain the idea: restrict to just B; essence
+not losing anything really
+\end{verbatim}
 

### 2017-05-03 upi.tex

@@
 The early developments of reversible programming languages started with a
-conventional programming language, e.g., an extended λ-calculus, and either
-extended the language with a history
+conventional programming language, e.g., extended $\lambda$-calculi, and either
+(i) extended the language with a history
 mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
-imposed constraints on the conventional control flow constructs to make them
+(ii) imposed constraints on the control flow constructs to make them
 reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
-recognize that reversible programming requires a fresh approach and should be
-designed from first principles without the detour via conventional irreversible
-languages~\cite{Yokoyama:2008:PRP, Mu:2004:ILRC, abramsky2005structural,
-  DiPierro:2006:RCL:1166042.1166047}.
+recognize that reversible programming languages require a fresh approach and
+should be designed from first principles without the detour via conventional
+irreversible languages~\cite{Yokoyama:2008:PRP, Mu:2004:ILRC,
+  abramsky2005structural, DiPierro:2006:RCL:1166042.1166047}.
+
+%%%%%
+\subsection{The $\Pi$ Family of Languages}
+
+In previous work, Carette, Bowman, James, and
+Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
+the~$\Pi$ family of reversible languages based on type isomorphisms and
+commutative semiring identities. The fragment without recursive types is
+universal for reversible boolean
+circuits~\citep{James:2012:IE:2103656.2103667}. With the addition of recursive
+types and trace operators~\citep{Hasegawa:1997:RCS:645893.671607}, the variant
+of $\Pi$ becomes a Turing-complete reversible
+language~\citep{James:2012:IE:2103656.2103667,rc2011}. Before identifying a
+``featherweight'' representative to use for the formal development in this
+paper, we illustrate the main underlying ideas for the entire family of
+languages using some examples.
+
+\amr{EXAMPLES OF CIRCUITS}
+
+Although writing circuits using the raw syntax for combinators is tedious, the
+examples illustrate the programming language nature of this family of $\Pi$
+languages. In other work, one can find a compiler from a conventional functional
+language to circuits~\citep{James:2012:IE:2103656.2103667}, a systematic
+technique to translate abstract machines to $\Pi$~\citep{rc2012}, and a
+Haskell-like surface language~\citep{theseus} which can ease writing
+circuits. All that reinforces the claim that this is a practical approach to
+reversible programming.
+
+\amr{EXAMPLES OF OPTIMIZATIONS}
 
@@
 Examples
+
+\end{verbatim}
+
+%%%%%
+\subsection{$\Pi_{\mathbb{2}}$}
+
+\begin{verbatim}
 

### 2017-05-03 upi.tex
entcs style...

@@
-\documentclass[format=acmlarge,review,natbib]{acmart}
+\documentclass{entcs}
+\usepackage{prentcsmacro}
+
+\def\lastname{Carette, Chen, Choudhury, Rose, and Sabry}
 
@@
 
-\title{From Reversible Programs to Univalent Universes and Back}
+\begin{frontmatter}
+\title{From Reversible Programs to \\ Univalent Universes and Back}
 \author{Jacques Carette}
-\affiliation{
-  \institution{McMaster University}
-  \country{Canada}}
-  \email{carette@mcmaster.ca}
+\address{McMaster University}
 \author{Chao-Hong Chen}
+\address{Indiana University}
 \author{Vikraman Choudhury}
@@
 \author{Amr Sabry}
-\affiliation{
-  \institution{Indiana University}
-  \country{United States}}
-\email{\{chen464,vikraman,rrose1,sabry\}@indiana.edu}
+\address{Indiana University}
 
@@
 
-\maketitle
+\end{frontmatter}
 
@@
 the discovery of deep connections between physics and
-computation~\cite{Landauer:1961, PhysRevA.32.3266,Toffoli:1980,
-  bennett1985fundamental,Frank:1999:REC:930275} and by the potential for
-efficient quantum computation~\cite{springerlink:10.1007/BF02650179}.
+computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275}
+and by the potential for efficient quantum
+computation~\cite{springerlink:10.1007/BF02650179}.
 
@@
 should be designed from first principles without the detour via conventional
-irreversible languages~\cite{Yokoyama:2008:PRP, Mu:2004:ILRC,
-  abramsky2005structural, DiPierro:2006:RCL:1166042.1166047}.
+irreversible languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.
 
@@
 universal for reversible boolean
-circuits~\citep{James:2012:IE:2103656.2103667}. With the addition of recursive
-types and trace operators~\citep{Hasegawa:1997:RCS:645893.671607}, the variant
+circuits~\cite{James:2012:IE:2103656.2103667}. With the addition of recursive
+types and trace operators~\cite{Hasegawa:1997:RCS:645893.671607}, the variant
 of $\Pi$ becomes a Turing-complete reversible
-language~\citep{James:2012:IE:2103656.2103667,rc2011}. Before identifying a
+language~\cite{James:2012:IE:2103656.2103667,rc2011}. Before identifying a
 ``featherweight'' representative to use for the formal development in this
@@
 languages. In other work, one can find a compiler from a conventional functional
-language to circuits~\citep{James:2012:IE:2103656.2103667}, a systematic
-technique to translate abstract machines to $\Pi$~\citep{rc2012}, and a
-Haskell-like surface language~\citep{theseus} which can ease writing
+language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
+technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
+Haskell-like surface language~\cite{theseus} which can ease writing
 circuits. All that reinforces the claim that this is a practical approach to

### 2017-05-04 upi.tex
examples; notation needs to be unified one way or the ohter; either all math; all haskell; all agda

@@
 The early developments of reversible programming languages started with a
-conventional programming language, e.g., extended $\lambda$-calculi, and either
+conventional programming language, e.g., an extended $\lambda$-calculus, and either
 (i) extended the language with a history
@@
 commutative semiring identities. The fragment without recursive types is
-universal for reversible boolean
-circuits~\cite{James:2012:IE:2103656.2103667}. With the addition of recursive
-types and trace operators~\cite{Hasegawa:1997:RCS:645893.671607}, the variant
-of $\Pi$ becomes a Turing-complete reversible
-language~\cite{James:2012:IE:2103656.2103667,rc2011}. Before identifying a
-``featherweight'' representative to use for the formal development in this
-paper, we illustrate the main underlying ideas for the entire family of
-languages using some examples.
+universal for reversible boolean circuits~\cite{James:2012:IE:2103656.2103667}
+and the extension with recursive types and trace
+operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
+language~\cite{James:2012:IE:2103656.2103667,rc2011}. In this paper we will
+identify a ``featherweight'' representative to use for the formal
+development. But first, we illustrate the flavor of the family of languages with
+some examples written in a Haskell syntax.
 
-\amr{EXAMPLES OF CIRCUITS}
+First consider the following program:
+\begin{verbatim}
+addSub1 :: (Int, Int) :<=> (Int, Int)
+addSub1 = CommuteTimes
+      :.: (UnfoldN :*: Id)
+      :.: Distribute
+      :.: (Id :+: CommuteTimes)
+      :.: Factor
+      :.: (FoldN :*: Id)
+\end{verbatim}
+which takes a pair of natural numbers and processes them as follows depending on
+whether the second number is 0 or not. If the second number is 0, the evaluation
+proceeds as follows:
+$(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
+\mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
+second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
+$(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
+\mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
+(\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
+applications of this program would increment the first number and decrement the
+second number maintaining the sum at all times. When the second number reaches
+0, the numbers are swapped.
 
-Although writing circuits using the raw syntax for combinators is tedious, the
-examples illustrate the programming language nature of this family of $\Pi$
+As another example, consider the following expressions culminating with the
+implementation of the Toffoli gate, whichi is universal for combinational
+circuits.
+
+\begin{verbatim}
+inot :: Bool :<=> Bool
+inot = UnfoldB :.: CommutePlus :.: FoldB
+
+cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
+cond f g = (UnfoldB :*: Id)
+       :.: Distribute
+       :.: ((Id :*: f) :+: (Id :*: g))
+       :.: Factor
+       :.: (FoldB :*: Id)
+
+controlled :: (a :<=> a) -> ((Bool, a) :<=> (Bool, a))
+controlled f = cond f Id
+
+cnot :: (Bool, Bool) :<=> (Bool, Bool)
+cnot = controlled inot
+
+toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
+toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
+\end{verbatim}
+
+Although writing circuits using the raw syntax for combinators above is tedious,
+the examples illustrate the programming language nature of this family of $\Pi$
 languages. In other work, one can find a compiler from a conventional functional
@@
 Haskell-like surface language~\cite{theseus} which can ease writing
-circuits. All that reinforces the claim that this is a practical approach to
+programs. All that reinforces the claim that this is a practical approach to
 reversible programming.
 
-\amr{EXAMPLES OF OPTIMIZATIONS}
+\begin{verbatim}
+neg₁ neg₂ neg₃ neg₄ neg₅ : BOOL ⟷ BOOL
+neg₁ = swap₊
+neg₂ = id⟷ ◎ swap₊
+neg₃ = swap₊ ◎ swap₊ ◎ swap₊
+neg₄ = swap₊ ◎ id⟷
+neg₅ = uniti⋆l ◎ swap⋆ ◎ (swap₊ ⊗ id⟷) ◎ swap⋆ ◎ unite⋆l
+neg₆ = uniti⋆r ◎ (swap₊ {ONE} {ONE} ⊗ id⟷) ◎ unite⋆r
+
+negEx : neg₅ ⇔ neg₁
+negEx = uniti⋆l ◎ (swap⋆ ◎ ((swap₊ ⊗ id⟷) ◎ (swap⋆ ◎ unite⋆l)))
+          ⇔⟨ id⇔ ⊡ assoc◎l ⟩
+        uniti⋆l ◎ ((swap⋆ ◎ (swap₊ ⊗ id⟷)) ◎ (swap⋆ ◎ unite⋆l))
+          ⇔⟨ id⇔ ⊡ (swapl⋆⇔ ⊡ id⇔) ⟩
+        uniti⋆l ◎ (((id⟷ ⊗ swap₊) ◎ swap⋆) ◎ (swap⋆ ◎ unite⋆l))
+          ⇔⟨ id⇔ ⊡ assoc◎r ⟩
+        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ (swap⋆ ◎ (swap⋆ ◎ unite⋆l)))
+          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ assoc◎l) ⟩
+        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ ((swap⋆ ◎ swap⋆) ◎ unite⋆l))
+          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ (linv◎l ⊡ id⇔)) ⟩
+        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ (id⟷ ◎ unite⋆l))
+          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ idl◎l) ⟩
+        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ unite⋆l)
+          ⇔⟨ assoc◎l ⟩
+        (uniti⋆l ◎ (id⟷ ⊗ swap₊)) ◎ unite⋆l
+          ⇔⟨ unitil⋆⇔l ⊡ id⇔ ⟩
+        (swap₊ ◎ uniti⋆l) ◎ unite⋆l
+          ⇔⟨ assoc◎r ⟩
+        swap₊ ◎ (uniti⋆l ◎ unite⋆l)
+          ⇔⟨ id⇔ ⊡ linv◎l ⟩
+        swap₊ ◎ id⟷
+          ⇔⟨ idr◎l ⟩
+        swap₊ ▤
+\end{verbatim}
 

### 2017-05-04 upi.tex
code in haskell syntax for now

@@
 First consider the following program:
+
+{\footnotesize
 \begin{verbatim}
@@
       :.: (FoldN :*: Id)
-\end{verbatim}
+\end{verbatim}}
+%
 which takes a pair of natural numbers and processes them as follows depending on
@@
 implementation of the Toffoli gate, whichi is universal for combinational
-circuits.
+circuits. We have two implementations of simple boolean negation: a direct one
+and a more convoluted one. These two implementations are ``equivalent'' in the
+sense to be explained below. The expression \verb|cond| is the main reversible
+control flow mechanism: it allows for the conditional execution of one of two
+branches but retains the choice of which branch. This expression can be used
+twice to implement the Toffoli gate, which performs a conditional negation on
+one bit if two other bits are both true:
 
+{\footnotesize
 \begin{verbatim}
 inot :: Bool :<=> Bool
-inot = UnfoldB :.: CommutePlus :.: FoldB
+inot = CommutePlus
+
+inot2 :: Bool :<=> Bool
+inot2 = UnitTimesI
+    :.: CommuteTimes
+    :.: (CommutePlus :*: Id)
+    :.: CommuteTimes
+    :.: UnitTimesE
 
 cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
-cond f g = (UnfoldB :*: Id)
-       :.: Distribute
+cond f g = Distribute
        :.: ((Id :*: f) :+: (Id :*: g))
        :.: Factor
-       :.: (FoldB :*: Id)
 
@@
 toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
-\end{verbatim}
+\end{verbatim}}
 
@@
 
+The raw syntax for combinators has, however, a significant benefit: it allows
+algebraic manipulation of the programs using a second level of combinators. As
+an example it is possible to show the the two implementations of boolean
+negation are equivalent as follows:
+
+{\footnotesize
 \begin{verbatim}
-neg₁ neg₂ neg₃ neg₄ neg₅ : BOOL ⟷ BOOL
-neg₁ = swap₊
-neg₂ = id⟷ ◎ swap₊
-neg₃ = swap₊ ◎ swap₊ ◎ swap₊
-neg₄ = swap₊ ◎ id⟷
-neg₅ = uniti⋆l ◎ swap⋆ ◎ (swap₊ ⊗ id⟷) ◎ swap⋆ ◎ unite⋆l
-neg₆ = uniti⋆r ◎ (swap₊ {ONE} {ONE} ⊗ id⟷) ◎ unite⋆r
+negOpt :: inot2 :<===> inot
+negOpt = (Id :>: AssocLeftEquiv)
+    :..: (Id :>: (SlideCommuteTimes :>: Id))
+    :..: (Id :>: AssocRightEquiv)
+    :..: (Id :>: (Id :>: AssocLeftEquiv))
+    :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
+    :..: (Id :>: (Id :>: IdElimL))
+    :..: AssocLeftEquiv
+    :..: (SlideUnitI :>: Id)
+    :..: AssocRightEquiv
+    :..: (Id :>: InvInvEquivId)
+    :..: IdElimR
+\end{verbatim}}
 
-negEx : neg₅ ⇔ neg₁
-negEx = uniti⋆l ◎ (swap⋆ ◎ ((swap₊ ⊗ id⟷) ◎ (swap⋆ ◎ unite⋆l)))
-          ⇔⟨ id⇔ ⊡ assoc◎l ⟩
-        uniti⋆l ◎ ((swap⋆ ◎ (swap₊ ⊗ id⟷)) ◎ (swap⋆ ◎ unite⋆l))
-          ⇔⟨ id⇔ ⊡ (swapl⋆⇔ ⊡ id⇔) ⟩
-        uniti⋆l ◎ (((id⟷ ⊗ swap₊) ◎ swap⋆) ◎ (swap⋆ ◎ unite⋆l))
-          ⇔⟨ id⇔ ⊡ assoc◎r ⟩
-        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ (swap⋆ ◎ (swap⋆ ◎ unite⋆l)))
-          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ assoc◎l) ⟩
-        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ ((swap⋆ ◎ swap⋆) ◎ unite⋆l))
-          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ (linv◎l ⊡ id⇔)) ⟩
-        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ (id⟷ ◎ unite⋆l))
-          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ idl◎l) ⟩
-        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ unite⋆l)
-          ⇔⟨ assoc◎l ⟩
-        (uniti⋆l ◎ (id⟷ ⊗ swap₊)) ◎ unite⋆l
-          ⇔⟨ unitil⋆⇔l ⊡ id⇔ ⟩
-        (swap₊ ◎ uniti⋆l) ◎ unite⋆l
-          ⇔⟨ assoc◎r ⟩
-        swap₊ ◎ (uniti⋆l ◎ unite⋆l)
-          ⇔⟨ id⇔ ⊡ linv◎l ⟩
-        swap₊ ◎ id⟷
-          ⇔⟨ idr◎l ⟩
-        swap₊ ▤
-\end{verbatim}
+This program, which can be thought of as an optimization or a proof, relates the
+two implementations of boolean negation via a series of elementary
+optimizations of which the following are the most interesting:
 
+{\footnotesize
 \begin{verbatim}
-Pi... isos, recursion, finite types, Turing complete... nice blah popl 12
-
-ESOP shows a natural 2-layer structure
-
-Examples
-
-\end{verbatim}
+AssocLeftEquiv    ::           (c1 :.: (c2 :.: c3)) :<===> ((c1 :.: c2) :.: c3)
+SlideCommuteTimes :: (CommuteTimes :.: (c1 :*: c2)) :<===> ((c2 :*: c1) :.: CommuteTimes)
+IdElimL           ::                     (Id :.: c) :<===> c
+SlideUnitI        ::   (UnitTimesI :.: (c1 :*: c2)) :<===> (c2 :.: UnitTimesI)
+\end{verbatim}}
 

### 2017-05-04 upi.tex

@@
 development. But first, we illustrate the flavor of the family of languages with
-some examples written in a Haskell syntax.
-
-First consider the following program:
+some examples written in a Haskell syntax. First consider the following program:
 
@@
 %
-which takes a pair of natural numbers and processes them as follows depending on
+\noindent which takes a pair of natural numbers and processes them as follows depending on
 whether the second number is 0 or not. If the second number is 0, the evaluation
@@
 As another example, consider the following expressions culminating with the
-implementation of the Toffoli gate, whichi is universal for combinational
+implementation of the Toffoli gate, which is universal for combinational
 circuits. We have two implementations of simple boolean negation: a direct one
@@
 control flow mechanism: it allows for the conditional execution of one of two
-branches but retains the choice of which branch. This expression can be used
-twice to implement the Toffoli gate, which performs a conditional negation on
-one bit if two other bits are both true:
+branches but retains the choice of which branch was executed. This mechanism can
+be used twice to implement the Toffoli gate, which performs a conditional
+negation on one bit if two other bits are both true:
 
@@
 \end{verbatim}}
-
-This program, which can be thought of as an optimization or a proof, relates the
+%
+\noindent This program, which can be thought of as an optimization or a proof, relates the
 two implementations of boolean negation via a series of elementary
@@
 
-\begin{verbatim}
+In previous work, on $\Pi$ we noted a possible connection with HoTT:
+\begin{quote}
+  It is, therefore, at least plausible that a variant of HoTT based exclusively
+  on reversible functions that directly correspond to equivalences would have
+  better computational properties. Our current result is a step, albeit
+  preliminary, in that direction as it only applies to finite
+  types~\cite{Carette2016}.
+\end{quote}
+Formalizing, in a precise sense, the connection between reversible programs
+based on combinators and paths in HoTT, as intuitive as it may seem, is however
+difficult. Paths in HoTT come equipped with principles like the
+``contractibility of singletons'', ``transport'', and ``path induction.'' None
+of these principles seems to have any direct counterpart in the world of
+reversible programming.
 
-conjecture it is related to HoTT
+We will however show, in a very clear sense, how a $\Pi$-based language is
+``equivalent'' to a univalent HoTT universe that possesses all the required
+properties of paths. The essence of the construction can be demonstrated with a
+tiny ``featherweight'' variant of $\Pi$ that includes just one type: booleans,
+and a tiny ``featherweight'' univalent universe to be formalized in the next
+section. Here we give a formalization of the small $\Pi$-based language which we
+will use in the formalization in the rest of the paper.
 
-study that language and formalize the connections between its combinators and
-between HoTT
+The restriction of $\Pi$ to the case of just booleans is the following:
 
-stud the smallest core that can explain the idea: restrict to just B; essence
-not losing anything really
-\end{verbatim}
+\[\begin{array}{rcl}
+\tau &::=& \bt \\
+\\
+v &::=& \begin{array}[t]{lrcl}
+                    & \fc &:& \bt \\
+              \alt & \tc &:& \bt
+               \end{array} \\
+\\
+c &::=& \begin{array}[t]{lrcl}
+              & \id &:& \tau \iso \tau \\
+               \alt & \swap &:& \bt \iso \bt \\
+               \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3)
+                              \to (\tau_1 \iso \tau_3)
+               \end{array} \\
+\\
+! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
+\invc{\id} &=& \id \\
+\invc{\swap} &=& \swap \\
+\invc{(\compc{c_1}{c_2})} &=& \compc{\invc{c_2}\;}{\;\invc{c_1}} \\
+\\
+\alpha &::=& \begin{array}[t]{lrcl}
+               & \id &:& c \isotwo c \\
+               \alt & \assocc &:& \compc{c_1}{(\compc{c_2}{c_3})} \isotwo
+                                              \compc{(\compc{c_1}{c_2})}{c_3} \\
+               \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
+               \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
+               \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
+               \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
+               \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3)
+                                            \to (c_1 \isotwo c_3) \\
+               \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2')
+                                            \to (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
+             \end{array}
+\end{array}\]
+
+The 2-combinators $\alpha$ also have inverses. The above forms a 2-groupoid.
 
@@
 %%%%%
-\subsection{The Language $\Pi$ of Type Isomorphisms}
-
-The restriction of $\Pi$ to the case of just booleans is the following:
-
-\[\begin{array}{rcl}
-\tau &::=& \bt \\
-\\
-v &::=& \begin{array}[t]{lrcl}
-                    & \fc &:& \bt \\
-              \alt & \tc &:& \bt
-               \end{array} \\
-\\
-c &::=& \begin{array}[t]{lrcl}
-              & \id &:& \tau \iso \tau \\
-               \alt & \swap &:& \bt \iso \bt \\
-               \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3)
-                              \to (\tau_1 \iso \tau_3)
-               \end{array} \\
-\\
-! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
-\invc{\id} &=& \id \\
-\invc{\swap} &=& \swap \\
-\invc{(\compc{c_1}{c_2})} &=& \compc{\invc{c_2}\;}{\;\invc{c_1}} \\
-\\
-\alpha &::=& \begin{array}[t]{lrcl}
-               & \id &:& c \isotwo c \\
-               \alt & \assocc &:& \compc{c_1}{(\compc{c_2}{c_3})} \isotwo
-                                              \compc{(\compc{c_1}{c_2})}{c_3} \\
-               \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
-               \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
-               \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
-               \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
-               \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3)
-                                            \to (c_1 \isotwo c_3) \\
-               \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2')
-                                            \to (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
-             \end{array}
-\end{array}\]
-
-The 2-combinators $\alpha$ also have inverses. The above forms a 2-groupoid (see
-accompanying code once it is fixed).
-
-%%%%%
 \subsection{Equivalence}

### 2017-05-11 upi.tex
abstract

@@
   We establish a close connection between a reversible programming language
-  based on type isomorphisms and a univalent (in the sense of Homotopy Type
-  Theory) universe. The correspondence relates combinators witnessing type
-  isomorphisms in the programming language to paths in the univalent universe,
-  and combinator optimizations in the programming language to 2-paths in the
-  univalent universe. The result provides a simple computational interpretation
-  of paths, 2-paths, and univalence in terms of familiar programming constructs.
+  based on type isomorphisms and a formally presented univalent universe. The
+  correspondence relates combinators witnessing type isomorphisms in the
+  programming language to paths in the univalent universe, and combinator
+  optimizations in the programming language to 2-paths in the univalent
+  universe. The result suggests a simple computational interpretation of paths
+  and of univalence in terms of familiar programming constructs whenever the
+  universe in question is computable.
 \end{abstract}

### 2017-05-12 upi.tex
fixed pi2
figure fibration
better fib figure and some direction for sec 3

@@
 \\
-c &::=& \begin{array}[t]{lrcl}
+c &::=& \begin{array}[t]{llcl}
               & \id &:& \tau \iso \tau \\
                \alt & \swap &:& \bt \iso \bt \\
-               \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3)
-                              \to (\tau_1 \iso \tau_3)
+               \alt & ! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
+               \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)
                \end{array} \\
-\\
-! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
-\invc{\id} &=& \id \\
-\invc{\swap} &=& \swap \\
-\invc{(\compc{c_1}{c_2})} &=& \compc{\invc{c_2}\;}{\;\invc{c_1}} \\
 \\
@@
                & \id &:& c \isotwo c \\
-               \alt & \assocc &:& \compc{c_1}{(\compc{c_2}{c_3})} \isotwo
-                                              \compc{(\compc{c_1}{c_2})}{c_3} \\
             \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
@@
             \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
-               \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3)
-                                            \to (c_1 \isotwo c_3) \\
-               \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2')
-                                            \to (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
+            \alt & \rho &:& \swap \circ \swap \isotwo \id \\
+            \alt & \assocc &:& \compc{(\compc{c_1}{c_2})}{c_3} \isotwo \compc{c_1}{(\compc{c_2}{c_3})} \\
+            \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
+            \alt & !! &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_1) \\
+            \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3) \to (c_1 \isotwo c_3)
              \end{array}
@@
 
-The 2-combinators $\alpha$ also have inverses. The above forms a 2-groupoid.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Univalent Type Families}
+
+We will consider a general statement of \emph{univalence} for type families. We
+start with a some preliminaries. Fig.~\ref{fig:fib} depicts a general type
+family $P : A \rightarrow \mathcal{U}$ mapping each point $x : A$ to a space
+$P(x)$. As everything is continuous, we expect this type family to respect
+identities. In other words if $x:A$ and $y:A$ are identified, it must be the
+case that the spaces $P(x)$ and $P(y)$ are ``equivalent.''
+
+\begin{figure}
+\begin{tabular}{c@{\qquad\qquad}c}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
+  \draw (0,-5) ellipse (2cm and 0.8cm);
+  \node[below] at (0,-6) {Base Space $A$};
+  \draw[fill] (-1,-5) circle [radius=0.025];
+  \node[below] at (-1,-5) {$x$};
+  \draw[fill] (1,-5) circle [radius=0.025];
+  \node[below] at (1,-5) {$y$};
+  \draw (-1,-2) ellipse (0.5cm and 2cm);
+  \node[left] at (-1.5,-2) {Fiber $P(x)$};
+  \draw (1,-2) ellipse (0.5cm and 2cm);
+  \node[right] at (1.5,-2) {Fiber $P(y)$};
+\end{tikzpicture}
+&
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
+  \draw (0,-5) ellipse (2cm and 0.8cm);
+  \node[below] at (0,-6) {Base Space $A$};
+  \draw[fill] (-1,-5) circle [radius=0.025];
+  \node[below] at (-1,-5) {$x$};
+  \draw[fill] (1,-5) circle [radius=0.025];
+  \node[below] at (1,-5) {$y$};
+  \draw (-1.3,-2) ellipse (0.5cm and 2cm);
+  \node[left] at (-1.8,-2) {Fiber $P(x)$};
+  \draw (1.3,-2) ellipse (0.5cm and 2cm);
+  \node[right] at (1.8,-2) {Fiber $P(y)$};
+  \draw[below,cyan,thick] (-1,-5) -- (1,-5);
+  \node[below,cyan,thick] at (0,-5) {$p$};
+  \draw[->,red,ultra thick] (-0.8,-1.7) to [out=45, in=135] (0.8,-1.7);
+  \draw[->,red,ultra thick] (0.8,-2.3) to [out=-135, in=-45] (-0.8,-2.3);
+  \node[red,ultra thick] at (0,-2) {$\simeq$};
+\end{tikzpicture}
+\end{tabular}
+\caption{\label{fig:fib}(left)Type family $P : A \rightarrow \mathcal{U}$ as a
+  fibration with total space $\Sigma_{(x:A)} P(x)$; (right) a path $x \equiv y$
+  induces an equivalence between the spaces $P(x)$ and $P(y)$}
+\end{figure}
+
+The converse statement is whether an equivalence between $P(x)$ and $P(y)$
+necessarily induces an identification $x \equiv y$. It is easy to construct
+several examples in which this converse statement fails and it instructive to
+examine such examples.
 

### 2017-05-12 upi.tex
univalent fibration non-example

@@
 We will consider a general statement of \emph{univalence} for type families. We
-start with a some preliminaries. Fig.~\ref{fig:fib} depicts a general type
-family $P : A \rightarrow \mathcal{U}$ mapping each point $x : A$ to a space
-$P(x)$. As everything is continuous, we expect this type family to respect
-identities. In other words if $x:A$ and $y:A$ are identified, it must be the
-case that the spaces $P(x)$ and $P(y)$ are ``equivalent.''
+start with a some preliminaries. Fig.~\ref{fig:fib} depicts a type family
+$P : A \rightarrow \mathcal{U}$ mapping each point $x : A$ to a space $P(x)$. As
+everything is continuous, we expect this type family to respect identities. In
+other words if $x:A$ and $y:A$ are identified using a path $x \equiv y$, it must
+be the case that the spaces $P(x)$ and $P(y)$ become equivalent
+$P(x) \simeq P(y)$ as a consequence. Generally speaking, there is however no
+reason to expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ to be
+themselves equivalent, and indeed we will provide a few examples in which they
+are not. When they are indeed equivalent, the fibration is a \emph{univalent
+  fibration}.
 
@@
   \draw[below,cyan,thick] (-1,-5) -- (1,-5);
-  \node[below,cyan,thick] at (0,-5) {$p$};
+  \node[below,cyan,thick] at (0,-5) {$\equiv$};
   \draw[->,red,ultra thick] (-0.8,-1.7) to [out=45, in=135] (0.8,-1.7);
@@
 
-The converse statement is whether an equivalence between $P(x)$ and $P(y)$
-necessarily induces an identification $x \equiv y$. It is easy to construct
-several examples in which this converse statement fails and it instructive to
-examine such examples.
+As a simple example of a non-univalent fibration consider $P : \mathbb{2}
+\rightarrow \mathbb{U}$ defined as $P = \lambda \_. \mathbb{1}$ illustrated below:
+
+\bigskip
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
+  \draw (0,-5) ellipse (2cm and 0.8cm);
+  \node[below] at (0,-6) {Base Space $\mathbb{2}$};
+  \draw[fill] (-1,-5) circle [radius=0.025];
+  \node[below] at (-1,-5) {$\fc$};
+  \draw[fill] (1,-5) circle [radius=0.025];
+  \node[below] at (1,-5) {$\tc$};
+  \draw (0,-2.7) ellipse (0.5cm and 1cm);
+  \node[right] at (0.5,-2.7) {Fiber $P(\fc)$ = Fiber $P(\tc)$ = $\mathbb{1}$};
+  \draw[fill] (0,-2.7) circle [radius=0.025];
+  \node[below] at (0,-2.7) {$\star$};
+\end{tikzpicture}
+
+\medskip\noindent There is one equivalence
+$P(\fc) = \mathbb{1} \simeq \mathbb{1} = P(\tc)$ but this certainly does not
+induce a path $\fc \equiv \tc$.
 

### 2017-05-12 upi.tex
more on univalent fibrations

@@
 
-As a simple example of a non-univalent fibration consider $P : \mathbb{2}
-\rightarrow \mathbb{U}$ defined as $P = \lambda \_. \mathbb{1}$ illustrated below:
+As two simple examples of non-univalent fibrations consider
+$P : \mathbb{1} \rightarrow \mathbb{U}$ defined as $P = \lambda\_.\mathbb{2}$
+and $Q : \mathbb{2} \rightarrow \mathbb{U}$ defined as
+$Q = \lambda \_. \mathbb{1}$ illustrated below:
+
+\begin{tabular}{c@{\qquad\qquad}c}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
+  \draw (0,-5) ellipse (2cm and 0.8cm);
+  \node[below] at (0,-6) {Base Space $\mathbb{1}$};
+  \draw[fill] (0,-5) circle [radius=0.025];
+  \node[below] at (0,-5) {$\star$};
+  \draw (0,-2.7) ellipse (0.5cm and 1cm);
+  \node[right] at (0.5,-2.7) {Fiber $P(\star) = \mathbb{2}$};
+  \draw[fill] (-0.2,-2.7) circle [radius=0.025];
+  \draw[fill] (0.2,-2.7) circle [radius=0.025];
+\end{tikzpicture}
+&
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
+  \draw (0,-5) ellipse (2cm and 0.8cm);
+  \node[below] at (0,-6) {Base Space $\mathbb{2}$};
+  \draw[fill] (-1,-5) circle [radius=0.025];
+  \node[below] at (-1,-5) {$\fc$};
+  \draw[fill] (1,-5) circle [radius=0.025];
+  \node[below] at (1,-5) {$\tc$};
+  \draw (0,-2.7) ellipse (0.5cm and 1cm);
+  \node[right] at (0.5,-2.7) {Fiber $Q(\fc)$ = Fiber $Q(\tc)$ = $\mathbb{1}$};
+  \draw[fill] (0,-2.7) circle [radius=0.025];
+  \node[below] at (0,-2.7) {$\star$};
+\end{tikzpicture}
+\end{tabular}
+
+\medskip\noindent The identity function and boolean negation induce two distinct
+equivalences in the space $P(\star) = \mathbb{2} \simeq \mathbb{2} =
+P(\star)$. As there is only one path $\star \equiv star$ in the base space, the
+fibration $P$ is not univalent. Similarly, there is one equivalence
+$Q(\fc) = \mathbb{1} \simeq \mathbb{1} = Q(\tc)$ but this certainly does not
+induce a path $\fc \equiv \tc$. As the examples illustrate there is a delicate
+balance required for the fibration to be univalent; otherwise some distinct
+equivalences would be identified (like for $P$) or distinct points in the base
+space would be identified (like for $Q$).
+
+An example of a univalent fibration is $P : \mathbb{2} \rightarrow \mathbb{U}$
+defined as $P(\fc) = \mathbb{0}$ and $P(\tc) = \mathbb{1}$ illustrated below:
 
@@
   \node[below] at (1,-5) {$\tc$};
-  \draw (0,-2.7) ellipse (0.5cm and 1cm);
-  \node[right] at (0.5,-2.7) {Fiber $P(\fc)$ = Fiber $P(\tc)$ = $\mathbb{1}$};
-  \draw[fill] (0,-2.7) circle [radius=0.025];
-  \node[below] at (0,-2.7) {$\star$};
+  \draw (-1.3,-2) ellipse (0.5cm and 2cm);
+  \node[left] at (-1.8,-2) {Fiber $P(\fc) = \mathbb{0}$};
+  \draw (1.3,-2) ellipse (0.5cm and 2cm);
+  \node[right] at (1.8,-2) {Fiber $P(\tc) = \mathbb{1}$};
+  \draw[fill] (1.8,-2) circle [radius=0.025];
+  \draw[below,cyan,dashed,thick] (-1,-5) -- (1,-5);
+  \node[below,cyan,dashed,thick] at (0,-5) {$\equiv$};
+  \draw[->,red,dashed,ultra thick] (-0.8,-1.7) to [out=45, in=135] (0.8,-1.7);
+  \draw[->,red,dashed,ultra thick] (0.8,-2.3) to [out=-135, in=-45] (-0.8,-2.3);
+  \node[red,ultra thick] at (0,-2) {$\simeq$};
 \end{tikzpicture}
 
-\medskip\noindent There is one equivalence
-$P(\fc) = \mathbb{1} \simeq \mathbb{1} = P(\tc)$ but this certainly does not
-induce a path $\fc \equiv \tc$.
+\medskip\noindent In this case, the space of equivalences
+$P(\fc) = \mathbb{0} \simeq \mathbb{1} = P(\tc)$ is empty and so is the space of
+paths $\fc \equiv \tc$ and hence these two spaces are vacuously equivalent.
 

### 2017-05-13 upi.tex
question
sec 3 shaping up in content

@@
   \draw[fill] (0,-2.7) circle [radius=0.025];
-  \node[below] at (0,-2.7) {$\star$};
 \end{tikzpicture}
@@
   \node[below] at (1,-5) {$\tc$};
-  \draw (-1.3,-2) ellipse (0.5cm and 2cm);
+  \draw (-1.3,-2) ellipse (0.5cm and 1cm);
   \node[left] at (-1.8,-2) {Fiber $P(\fc) = \mathbb{0}$};
-  \draw (1.3,-2) ellipse (0.5cm and 2cm);
+  \draw (1.3,-2) ellipse (0.5cm and 1cm);
+  \draw[fill] (1.3,-2) circle [radius=0.025];
   \node[right] at (1.8,-2) {Fiber $P(\tc) = \mathbb{1}$};
-  \draw[fill] (1.8,-2) circle [radius=0.025];
   \draw[below,cyan,dashed,thick] (-1,-5) -- (1,-5);
@@
 paths $\fc \equiv \tc$ and hence these two spaces are vacuously equivalent.
+
+%%%
+\subsection{A Special Class}
+
+Instead of guessing, there is a special class of univalent fibrations:
+\[\begin{array}{rcl}
+\{F\} &:& \Sigma_{A:\mathcal{U}} \| F \equiv A \| \rightarrow \mathcal{U} \\
+\{F\}(A,|p|) &=& A
+\end{array}\]
+
+This construction is always a univalent fibration~\cite{XXX}. We can understand
+the basic idea as follows. Take two points $(A,|p|)$ and $(B,|q|)$ in the base
+space. We want to compare the space $A \simeq B$ with the space
+$(A,|p|) \equiv (B,|q|)$.  In one direction, we have $p : F \equiv A$ and
+$q : F \equiv B$ and hence $A \equiv B$ and by univalence $A \simeq B$. In the
+other direction \ldots
+
+Also explain base space has higher paths \ldots
 

### 2017-05-25 upi.tex
more on slides
minor tweaks

@@
 \newcommand{\amr}[1]{\fbox{\begin{minipage}{0.8\textwidth}\color{red}{Amr says: {#1}}\end{minipage}}}
+\newcommand{\jacques}[1]{\fbox{\begin{minipage}{0.8\textwidth}\color{red}{Jacques says: {#1}}\end{minipage}}}
 
@@
   correspondence relates combinators witnessing type isomorphisms in the
-  programming language to paths in the univalent universe, and combinator
+  programming language to paths in the univalent universe; combinator
   optimizations in the programming language to 2-paths in the univalent
@@
 There is a cottage industry of reversible programming languages, reversible
-logic, programming applications of type isomorphisms, etc. This work seems it
-should be connected to HoTT and univalence but there aren't any precise
+logic, programming applications of type isomorphisms, etc. It seems that this
+work should be connected to HoTT and univalence, but there aren't any precise
 connections or theorems.
 
-We propose a featherweight very specialized version of HoTT: a smaller
+We propose a very specialized version of HoTT, which we dub
+``featherweight HoTT''; it sports a smaller
 collection of types and only reversible functions. No dependent types!  More
@@
 \end{itemize}
+
+\jacques{Just to make sure: we're aiming to prove a much much smaller version of
+this here, right?  Where in fact the only type is Bool?}
 
@@
 
-Recall that $\{\bt\} = \Sigma_{(A:\mathcal{U})} \|A \simeq \bt\|$. We begin by
-unpacking the objects, 1-paths, and 2-paths of $\{\bt\}$.
+\textbf{Theorem.} For any type $F$, the type $\Sigma_{A:\mathcal{U}} \| F \equiv A \|$ is the base
+of a univalent fibration.
+
+Recall that
+$\{\bt\} = \Sigma_{(A:\mathcal{U})} \|A \simeq \bt\| = \Sigma_{(A:\mathcal{U})}
+\| \mathit{Bool} \equiv A\|$. We begin by unpacking the objects, 1-paths, and 2-paths of
+$\{\bt\}$.
 

### 2017-05-26 upi.tex
expand intro to Pi subsection.  Will next work on not introducing a new syntax...

@@
 \usepackage{tikz}
+\usepackage[inline]{enumitem}
 
@@
 The practice of programming languages is replete with \emph{ad hoc} instances of
-reversible computations such as database transactions, mechanisms for data
+reversible computations: database transactions, mechanisms for data
 provenance, checkpoints, stack and exception traces, logs, backups, rollback
@@
 Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
-treatment of reversibility but intensive research in full-fledged reversible
+treatment of reversibility. But intensive research in full-fledged reversible
 models of computations and reversible programming languages was only sparked by
 the discovery of deep connections between physics and
-computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275}
+computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
 and by the potential for efficient quantum
@@
 conventional programming language, e.g., an extended $\lambda$-calculus, and either
-(i) extended the language with a history
-mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
-(ii) imposed constraints on the control flow constructs to make them
-reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
-recognize that reversible programming languages require a fresh approach and
-should be designed from first principles without the detour via conventional
-irreversible languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.
+\begin{enumerate*}[(i)]
+\item extended the language with a history
+mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
+\item imposed constraints on the control flow constructs to make them
+reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. 
+\end{enumerate*}
+More modern approaches recognize that reversible programming languages require
+a fresh approach and should be designed from first principles without the
+detour via conventional irreversible
+languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.
 
@@
 operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
-language~\cite{James:2012:IE:2103656.2103667,rc2011}. In this paper we will
+language~\cite{James:2012:IE:2103656.2103667,rc2011}. While at first sight,
+$\Pi$ might appear \emph{ad hoc},~\cite{Carette2016} shows that it arises naturally
+from an ``extended'' view of the Curry-Howard Correspondance: rather than looking
+at mere \emph{inhabitation} as the main source of analogy between logic
+and computation, \emph{type equivalences} becomes the source of analogy.  This
+allows one to see an analogy between algebra and reversible computation.
+Furthermore, this works at multiple levels: that of $1$-algebra (types form a
+semiring under isomorphism), but also $2$-algebra (types and equivalences form a 
+weak Rig Groupoid).  In other words, by taking ``weak Rig Groupoid'' as the 
+starting semantics, one naturally gets $\Pi$ as the syntax for the language of
+proofs of isomorphisms -- in the same way that many terms of the
+$\lambda$-calculus arise from Cartesian Closed Categories.
+
+By restricting to finite types, two operational semantics emerge:
+\begin{enumerate*}{(i)}
+\item a (reversible) operational semantics of programs over finite types, and
+\item a (reversible) operational semantics of program transformers of such programs.
+\end{enumerate*}
+Such programs can be interpreted as generalized permutations, as they work on
+structure as well as the underlying skeleton of finite types, aka the natural numbers.
+Many of the transformers can be \emph{oriented}, and thus be interpreted as
+\emph{optimizations} in some sense.
+
+In this paper we will
 identify a ``featherweight'' representative to use for the formal

### 2017-06-09 upi.tex
comments

@@
 
+\begin{verbatim}
+Thm : Let P : Type -> Type such that for all X : Type, P(X) is a proposition. Then Σ(X,P) is the base of a univalent fibration.
+
+Furthermore, we have that *all* univalent fibrations over Type arise this way.
+We don't need this fact directly, but it may help in the explanation.
+(It's certainly cleaner than making the connection to classifying bundles in geometry).
+
+Here are some more examples.
+
+is-prop : Type -> Type has the property that is-prop(X) is a proposition for all types X.
+So Σ(X , is-prop) is the base of a univalent fibration.
+In other words, this is the subuniverse consisting of all and only propositions.
+
+is-set : Type -> Type has the property that is-set(X) is a proposition for all types X.
+So Σ(X , is-set) is the base of a univalent fibration.
+In other words, this is the subuniverse consisting of all and only sets.
+
+is-1-type : Type -> Type has the property that is-1-type(X) is a proposition for
+all types X. So Σ(X , is-1-type) is the base of a univalent fibration. In other
+words, this is the subuniverse consisting of all and only 1-types. (Externally,
+these are the 1-groupoids embedded in inf-groupoids.)
+
+But here's a summary of what I wrote last week, which I think should suffice for this project.
+Designating a collection of types as a univalent universe boils down to defining a mere predicate on the universe.
+One approach to defining such a predicate is to generate a set of names and an interpretation El (a type family over these names), and ask that the disjunction "is equal to the interpretation of a name" be exclusive (i.e., a proposition).
+This last step requires that the set of names be representatives of distinct normal forms.
+Repackage these data --- the collection of types, their names, and the proof that the predicate on the universe is a proposition -- as a semantic object admitting various presentations, which when computable, can be called a "reversible programming language".
+
+Because "the proof that the predicate on the universe is a proposition" (and univalence), we naturally get a family of equivalences arising.
+At various levels.
+When reflected syntactically, we get multiple presentations of the same semantic object (where 'same' is really up to things 1 level up).
+Because all we can really talk about are equivalences, when this syntax is interpreted as a programming language, it is unavoidably reversible.
+
+Right, the predicate being a proposition suffices (and is necessary) to define an embedding into a univalent universe.
+
+(Univalence is used axiomatically, as is the rest of the type theory, to define the universe.
+But you're welcome to interpret ITT + univalence in simplicial sets and forget about univalence altogether: the univalent universe is a particular simplicial set defined in classical set theory, as are the rest of the subuniverses serving as denotational models.)
+
+By "admitting various presentations", I mean presentation in the sense of group presentation.
+For example, the real functions on R^2 which are symmetries of the square centered at the origin can be viewed as the dihedral group on 8 elements: but we're free to present this group in an infinite number of ways.
+These presentations can be taken to be syntax for a language.
+If a presentation has a decidable word problem, and you have the algorithm, then you could be said to have an operational semantics, and so a programming language of a sort.
+This is what, I take it, you and Amr did with Laplaza's work on presenting free distributive categories in the first place.
+There your semantic object is the syntactic category of his theory of distributive categories.
+
+What using HoTT as a metatheory facilitates here is the construction of models which admit presentations of infinity-groupoids.
+(You could do it without HoTT.
+But the results would likely be less general, say if the work was done directly in simplicial sets or cubical sets.) Using this approach, you have a lot more options now for cooking up reversible PLs now; just as many as you have decidable presentations of inf-groupoids.
+
+The 'exclusive disjunction' aspect of being a proposition seems to have more content: it's a sort of decidability / normal form property all rolled into one.
+
+Yes, I was struck by this too.
+At least insofar as we continue with this style of defining predicates on the universe to obtain the intended interpretation of the types in a rev PL, it seems that a decidable enumeration of normal forms of the types will already be needed.
+But this doesn't seem like a huge constraint to me; most of the action is at the level of the paths....
+
+I guess that you assumed in "This last step requires that the set of names be representatives of distinct normal forms" that these were also complete?
+That we have a decidable enumeration of all normal forms?
+
+I haven't said anything about an actual presentation of the inf-groupoid which is the univalent universe of finite types.
+You want Pi to do this work.
+So to get started, we should define a function from the BNF grammar corresponding to a syntax for a semiring to the terms of that universe.
+As it stands, we already know that those terms are reflected in Names, so it would be sufficient to define that function into Names.
+Ah, is this what you meant by "complete"?
+Yes, we have a name for everything in the extension of that predicate (by construction).
+
+\end{verbatim}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-06-10 upi.tex
slides

@@
 conventional programming language, e.g., an extended $\lambda$-calculus, and either
-\begin{enumerate*}[(i)]
+\begin{enumerate}
 \item extended the language with a history
@@
 reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.
-\end{enumerate*}
+\end{enumerate}
 More modern approaches recognize that reversible programming languages require
@@
 By restricting to finite types, two operational semantics emerge:
-\begin{enumerate*}{(i)}
+\begin{enumerate}
 \item a (reversible) operational semantics of programs over finite types, and
 \item a (reversible) operational semantics of program transformers of such programs.
-\end{enumerate*}
+\end{enumerate}
 Such programs can be interpreted as generalized permutations, as they work on
@@
 be the case that the spaces $P(x)$ and $P(y)$ become equivalent
-$P(x) \simeq P(y)$ as a consequence. Generally speaking, there is however no
-reason to expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ to be
-themselves equivalent, and indeed we will provide a few examples in which they
-are not. When they are indeed equivalent, the fibration is a \emph{univalent
-  fibration}.
+$P(x) \simeq P(y)$ as a consequence. Generally speaking, there is no reason to
+expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ to be themselves
+equivalent for all $x$ and $y$. And indeed they are most frequently not
+equivalent. In the rare cases in which they are equivalent, the fibration is a
+\emph{univalent fibration}.
 
@@
 \end{tabular}
-\caption{\label{fig:fib}(left)Type family $P : A \rightarrow \mathcal{U}$ as a
+\caption{\label{fig:fib}(left) Type family $P : A \rightarrow \mathcal{U}$ as a
   fibration with total space $\Sigma_{(x:A)} P(x)$; (right) a path $x \equiv y$
-  induces an equivalence between the spaces $P(x)$ and $P(y)$}
+  in the base space induces an equivalence between the spaces $P(x)$ and $P(y)$}
 \end{figure}
@@
 equivalences in the space $P(\star) = \mathbb{2} \simeq \mathbb{2} =
-P(\star)$. As there is only one path $\star \equiv star$ in the base space, the
+P(\star)$. As there is only one path $\star \equiv \star$ in the base space, the
 fibration $P$ is not univalent. Similarly, there is one equivalence

### 2017-06-23 upi.tex
outline for paper
my intent for section 2

@@
 
+POPL 2012: Information effects; Canonicity for 2-dimensional type theory; look
+different but we show are the ``same''.
+
 There is a cottage industry of reversible programming languages, reversible
@@
 \end{enumerate}
-Such programs can be interpreted as generalized permutations, as they work on
+$\Pi$ programs can be interpreted as generalized permutations, as they work on
 structure as well as the underlying skeleton of finite types, aka the natural numbers.
 Many of the transformers can be \emph{oriented}, and thus be interpreted as
-\emph{optimizations} in some sense.
+\emph{optimizations}.  As with traditional languages, there are different aspects
+we can optimize for (program size, number of steps to completion, etc); we will
+not touch on this further in this paper.
 
+We will later identify a ``featherweight'' version of $\Pi$ to use in our formal
+development.  But first we illustrate the flavor of the family of languages with
+some examples.
+
+\jacques{Example 1: using the combinators from pi-dual, some an equational
+derivation in type theory (which?), in equational-derivation style.  Point out
+that reading ``down'' the justifications is a $\Pi$ program which is also a
+constructive proof of the equivalence of the types.  As it is reversible, we
+can also read the reverse program by reading ``up''.}
+
+\jacques{Example 2: like example 1, but this time at the next level (i.e. a
+program equivalence).  Same remark old.  Refer to Shulmann notes, and point out
+that such equational proofs at this level can actually be interpreted in appropriate
+categories as commutative diagrams.}
+
+\jacques{Refer to some Sabry et al. papers regarding Toffoli gates.}
+
+\begin{comment}
 In this paper we will
@@
 %
-\noindent This program, which can be thought of as an optimization or a proof, relates the
-two implementations of boolean negation via a series of elementary
+\noindent This program, which can be thought of as an optimization or a proof,
+relates the two implementations of boolean negation via a series of elementary
 optimizations of which the following are the most interesting:
@@
 
+\end{comment}
 %%%%%
@@
 \end{array}\]
+
+Canonical forms: given a level 1 program p, the level 2 combinators are
+complete; they can transform p to either id or not:
+\begin{verbatim}
+cmpl2-lem : (p : B <->1 B) -> (p <->2 id) + (p <->2 not)
+\end{verbatim}
 
@@
 
+That space {2} is a higher groupoid. Let's look at its objects; its 1-paths; its
+2-paths etc.
+
+Show Agda code
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Preliminaries}
+\section{Correspondence}
+
+Soundness; completeness; etc.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Discussion}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
+~
+\newpage
+\section{Old}
 

### 2017-06-26 upi.tex
working on intro etc

@@
   correspondence relates combinators witnessing type isomorphisms in the
-  programming language to paths in the univalent universe; combinator
+  programming language to paths in the univalent universe and combinator
   optimizations in the programming language to 2-paths in the univalent
@@
 
-POPL 2012: Information effects; Canonicity for 2-dimensional type theory; look
-different but we show are the ``same''.
-
-There is a cottage industry of reversible programming languages, reversible
-logic, programming applications of type isomorphisms, etc. It seems that this
-work should be connected to HoTT and univalence, but there aren't any precise
-connections or theorems.
-
-We propose a very specialized version of HoTT, which we dub
-``featherweight HoTT''; it sports a smaller
-collection of types and only reversible functions. No dependent types!  More
-precisely, our conjecture:
-
-\begin{itemize}
-\item Let $\mathcal{U}$ be the univalent subuniverse generated by $0$, $1$, $+$,
-  and $*$ suitably truncated.
-\item Let $\Pi$ be the previously  described reversible programming with 1-paths and 2-paths
-\item We conjecture that $\Pi$ includes codes for \emph{all} paths in $\mathcal{U}$
-\end{itemize}
-
-\jacques{Just to make sure: we're aiming to prove a much much smaller version of
-this here, right?  Where in fact the only type is Bool?}
+The proceedings of the 2012 Symposium on Principles of Programming
+Languages~\cite{Field:2012:2103656} included two apparently unrelated papers:
+\emph{Information Effects} by James and Sabry and \emph{Canonicity for
+  2-dimensional type theory} by Licata and Harper. The first paper, motivated by
+the physical nature of
+computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
+proposed a reversible language $\Pi$ in which every program is a type
+isomorphism. The second paper, motivated by the topological nature of
+computation~\cite{vv06,hottbook}, proposed a judgmental formulation of
+intensional dependent type theory with a twice-iterated identity type. During
+the conference, it became apparent, at an intuitive and informal level, that the
+two papers had strong similarities. Formalizing the precise connection was far
+from obvious, however. In this paper we report on a formal connection between
+appropriately formulated reversible languages on one hand and univalent
+universes on the other.
 
@@
 
+The first step was a refinement and extension of the
+language $\Pi$ with a layer of program transformations between the programs
+witnessing type isomorphisms~\cite{Carette2016}. The second crucial result was
+the formal presention
+
+Second,
+
+POPL 2012: Information effects; Canonicity for 2-dimensional type theory; look
+different but we show are the ``same''.
+
+There is a cottage industry of reversible programming languages, reversible
+logic, programming applications of type isomorphisms, etc. It seems that this
+work should be connected to HoTT and univalence, but there aren't any precise
+connections or theorems.
+
+We propose a very specialized version of HoTT, which we dub
+``featherweight HoTT''; it sports a smaller
+collection of types and only reversible functions. No dependent types!  More
+precisely, our conjecture:
+
+\begin{itemize}
+\item Let $\mathcal{U}$ be the univalent subuniverse generated by $0$, $1$, $+$,
+  and $*$ suitably truncated.
+\item Let $\Pi$ be the previously  described reversible programming with 1-paths and 2-paths
+\item We conjecture that $\Pi$ includes codes for \emph{all} paths in $\mathcal{U}$
+\end{itemize}
+
+\jacques{Just to make sure: we're aiming to prove a much much smaller version of
+this here, right?  Where in fact the only type is Bool?}
+
  Following the HoTT book and the slides on ``A Characterization of Univalent

### 2017-06-26 upi.tex
reorganizing

@@
 computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
-proposed a reversible language $\Pi$ in which every program is a type
-isomorphism. The second paper, motivated by the topological nature of
-computation~\cite{vv06,hottbook}, proposed a judgmental formulation of
-intensional dependent type theory with a twice-iterated identity type. During
-the conference, it became apparent, at an intuitive and informal level, that the
-two papers had strong similarities. Formalizing the precise connection was far
-from obvious, however. In this paper we report on a formal connection between
-appropriately formulated reversible languages on one hand and univalent
-universes on the other.
+proposed, among other results, a reversible language $\Pi$ in which every
+program is a type isomorphism. The second paper, motivated by the connections
+between homotopy theory and type theory~\cite{vv06,hottbook}, proposed a
+judgmental formulation of intensional dependent type theory with a
+twice-iterated identity type. During the presentations and ensuing discussions
+at the conference, it became apparent, at an intuitive and informal level, that
+the two papers had strong similarities. Formalizing the precise connection was
+far from obvious, however.
+
+In this paper we report on a formal connection between appropriately formulated
+reversible languages on one hand and univalent universes on the other. In the
+next section, we give a rational reconstruction of $\Pi$ focusing on a small
+``featherweight'' fragment. In Sec.~\ref{sec:univalent}, we review
+\emph{univalent fibrations} which allow us to give formal presentations of
+``small'' univalent universes. In Sec.~\ref{sec:connection} we state and prove
+the formal connection between the systems presented in the preceding two
+sections. Sec.~\ref{sec:conclusion} puts our work in a larger context, discusses
+related and future work, and concludes.
 
@@
 
-The practice of programming languages is replete with \emph{ad hoc} instances of
-reversible computations: database transactions, mechanisms for data
-provenance, checkpoints, stack and exception traces, logs, backups, rollback
-recoveries, version control systems, reverse engineering, software transactional
-memories, continuations, backtracking search, and multiple-level ``undo''
-features in commercial applications. In the early nineties,
-Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
-treatment of reversibility. But intensive research in full-fledged reversible
-models of computations and reversible programming languages was only sparked by
-the discovery of deep connections between physics and
-computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
-and by the potential for efficient quantum
-computation~\cite{springerlink:10.1007/BF02650179}.
+Starting from the physical principle of ``conservation of
+information''~\cite{Hey:1999:FCE:304763,fredkin1982conservative}, James and
+Sabry~\cite{James:2012:IE:2103656.2103667} propose a family of programming
+languages $\Pi$ in which computation preserve information. Technically,
+computations are \emph{type isomorphisms} which, at least in the case of finite
+types, clearly preserve entropy in the information-theoretic sense.
 
-The early developments of reversible programming languages started with a
-conventional programming language, e.g., an extended $\lambda$-calculus, and either
-\begin{enumerate}
-\item extended the language with a history
-mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
-\item imposed constraints on the control flow constructs to make them
-reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.
-\end{enumerate}
-More modern approaches recognize that reversible programming languages require
-a fresh approach and should be designed from first principles without the
-detour via conventional irreversible
-languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.
+We will identify a ``featherweight'' version of $\Pi$ to use in our formal
+development.  But first we illustrate the flavor of the family of languages with
+some examples.
 
 %%%%%
-\subsection{The $\Pi$ Family of Languages}
-
-In previous work, Carette, Bowman, James, and
-Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
-the~$\Pi$ family of reversible languages based on type isomorphisms and
-commutative semiring identities. The fragment without recursive types is
-universal for reversible boolean circuits~\cite{James:2012:IE:2103656.2103667}
-and the extension with recursive types and trace
-operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
-language~\cite{James:2012:IE:2103656.2103667,rc2011}. While at first sight,
-$\Pi$ might appear \emph{ad hoc},~\cite{Carette2016} shows that it arises naturally
-from an ``extended'' view of the Curry-Howard Correspondance: rather than looking
-at mere \emph{inhabitation} as the main source of analogy between logic
-and computation, \emph{type equivalences} becomes the source of analogy.  This
-allows one to see an analogy between algebra and reversible computation.
-Furthermore, this works at multiple levels: that of $1$-algebra (types form a
-semiring under isomorphism), but also $2$-algebra (types and equivalences form a
-weak Rig Groupoid).  In other words, by taking ``weak Rig Groupoid'' as the
-starting semantics, one naturally gets $\Pi$ as the syntax for the language of
-proofs of isomorphisms -- in the same way that many terms of the
-$\lambda$-calculus arise from Cartesian Closed Categories.
-
-By restricting to finite types, two operational semantics emerge:
-\begin{enumerate}
-\item a (reversible) operational semantics of programs over finite types, and
-\item a (reversible) operational semantics of program transformers of such programs.
-\end{enumerate}
-$\Pi$ programs can be interpreted as generalized permutations, as they work on
-structure as well as the underlying skeleton of finite types, aka the natural numbers.
-Many of the transformers can be \emph{oriented}, and thus be interpreted as
-\emph{optimizations}.  As with traditional languages, there are different aspects
-we can optimize for (program size, number of steps to completion, etc); we will
-not touch on this further in this paper.
-
-We will later identify a ``featherweight'' version of $\Pi$ to use in our formal
-development.  But first we illustrate the flavor of the family of languages with
-some examples.
+\subsection{Examples}
 
@@
 
-\begin{comment}
-In this paper we will
-identify a ``featherweight'' representative to use for the formal
-development. But first, we illustrate the flavor of the family of languages with
-some examples written in a Haskell syntax. First consider the following program:
+% \begin{comment}
+% In this paper we will
+% identify a ``featherweight'' representative to use for the formal
+% development. But first, we illustrate the flavor of the family of languages with
+% some examples written in a Haskell syntax. First consider the following program:
 
-{\footnotesize
-\begin{verbatim}
-addSub1 :: (Int, Int) :<=> (Int, Int)
-addSub1 = CommuteTimes
-      :.: (UnfoldN :*: Id)
-      :.: Distribute
-      :.: (Id :+: CommuteTimes)
-      :.: Factor
-      :.: (FoldN :*: Id)
-\end{verbatim}}
-%
-\noindent which takes a pair of natural numbers and processes them as follows depending on
-whether the second number is 0 or not. If the second number is 0, the evaluation
-proceeds as follows:
-$(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
-\mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
-second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
-$(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
-\mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
-(\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
-applications of this program would increment the first number and decrement the
-second number maintaining the sum at all times. When the second number reaches
-0, the numbers are swapped.
+% {\footnotesize
+% \begin{verbatim}
+% addSub1 :: (Int, Int) :<=> (Int, Int)
+% addSub1 = CommuteTimes
+%       :.: (UnfoldN :*: Id)
+%       :.: Distribute
+%       :.: (Id :+: CommuteTimes)
+%       :.: Factor
+%       :.: (FoldN :*: Id)
+% \end{verbatim}}
+% %
+% \noindent which takes a pair of natural numbers and processes them as follows depending on
+% whether the second number is 0 or not. If the second number is 0, the evaluation
+% proceeds as follows:
+% $(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
+% \mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
+% second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
+% $(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
+% \mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
+% (\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
+% applications of this program would increment the first number and decrement the
+% second number maintaining the sum at all times. When the second number reaches
+% 0, the numbers are swapped.
 
-As another example, consider the following expressions culminating with the
-implementation of the Toffoli gate, which is universal for combinational
-circuits. We have two implementations of simple boolean negation: a direct one
-and a more convoluted one. These two implementations are ``equivalent'' in the
-sense to be explained below. The expression \verb|cond| is the main reversible
-control flow mechanism: it allows for the conditional execution of one of two
-branches but retains the choice of which branch was executed. This mechanism can
-be used twice to implement the Toffoli gate, which performs a conditional
-negation on one bit if two other bits are both true:
+% As another example, consider the following expressions culminating with the
+% implementation of the Toffoli gate, which is universal for combinational
+% circuits. We have two implementations of simple boolean negation: a direct one
+% and a more convoluted one. These two implementations are ``equivalent'' in the
+% sense to be explained below. The expression \verb|cond| is the main reversible
+% control flow mechanism: it allows for the conditional execution of one of two
+% branches but retains the choice of which branch was executed. This mechanism can
+% be used twice to implement the Toffoli gate, which performs a conditional
+% negation on one bit if two other bits are both true:
 
-{\footnotesize
-\begin{verbatim}
-inot :: Bool :<=> Bool
-inot = CommutePlus
+% {\footnotesize
+% \begin{verbatim}
+% inot :: Bool :<=> Bool
+% inot = CommutePlus
 
-inot2 :: Bool :<=> Bool
-inot2 = UnitTimesI
-    :.: CommuteTimes
-    :.: (CommutePlus :*: Id)
-    :.: CommuteTimes
-    :.: UnitTimesE
+% inot2 :: Bool :<=> Bool
+% inot2 = UnitTimesI
+%     :.: CommuteTimes
+%     :.: (CommutePlus :*: Id)
+%     :.: CommuteTimes
+%     :.: UnitTimesE
 
-cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
-cond f g = Distribute
-       :.: ((Id :*: f) :+: (Id :*: g))
-       :.: Factor
+% cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
+% cond f g = Distribute
+%        :.: ((Id :*: f) :+: (Id :*: g))
+%        :.: Factor
 
-controlled :: (a :<=> a) -> ((Bool, a) :<=> (Bool, a))
-controlled f = cond f Id
+% controlled :: (a :<=> a) -> ((Bool, a) :<=> (Bool, a))
+% controlled f = cond f Id
 
-cnot :: (Bool, Bool) :<=> (Bool, Bool)
-cnot = controlled inot
+% cnot :: (Bool, Bool) :<=> (Bool, Bool)
+% cnot = controlled inot
 
-toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
-toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
-\end{verbatim}}
+% toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
+% toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
+% \end{verbatim}}
 
-Although writing circuits using the raw syntax for combinators above is tedious,
-the examples illustrate the programming language nature of this family of $\Pi$
-languages. In other work, one can find a compiler from a conventional functional
-language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
-technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
-Haskell-like surface language~\cite{theseus} which can ease writing
-programs. All that reinforces the claim that this is a practical approach to
-reversible programming.
+% Although writing circuits using the raw syntax for combinators above is tedious,
+% the examples illustrate the programming language nature of this family of $\Pi$
+% languages. In other work, one can find a compiler from a conventional functional
+% language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
+% technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
+% Haskell-like surface language~\cite{theseus} which can ease writing
+% programs. All that reinforces the claim that this is a practical approach to
+% reversible programming.
 
-The raw syntax for combinators has, however, a significant benefit: it allows
-algebraic manipulation of the programs using a second level of combinators. As
-an example it is possible to show the the two implementations of boolean
-negation are equivalent as follows:
+% The raw syntax for combinators has, however, a significant benefit: it allows
+% algebraic manipulation of the programs using a second level of combinators. As
+% an example it is possible to show the the two implementations of boolean
+% negation are equivalent as follows:
 
-{\footnotesize
-\begin{verbatim}
-negOpt :: inot2 :<===> inot
-negOpt = (Id :>: AssocLeftEquiv)
-    :..: (Id :>: (SlideCommuteTimes :>: Id))
-    :..: (Id :>: AssocRightEquiv)
-    :..: (Id :>: (Id :>: AssocLeftEquiv))
-    :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
-    :..: (Id :>: (Id :>: IdElimL))
-    :..: AssocLeftEquiv
-    :..: (SlideUnitI :>: Id)
-    :..: AssocRightEquiv
-    :..: (Id :>: InvInvEquivId)
-    :..: IdElimR
-\end{verbatim}}
-%
-\noindent This program, which can be thought of as an optimization or a proof,
-relates the two implementations of boolean negation via a series of elementary
-optimizations of which the following are the most interesting:
+% {\footnotesize
+% \begin{verbatim}
+% negOpt :: inot2 :<===> inot
+% negOpt = (Id :>: AssocLeftEquiv)
+%     :..: (Id :>: (SlideCommuteTimes :>: Id))
+%     :..: (Id :>: AssocRightEquiv)
+%     :..: (Id :>: (Id :>: AssocLeftEquiv))
+%     :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
+%     :..: (Id :>: (Id :>: IdElimL))
+%     :..: AssocLeftEquiv
+%     :..: (SlideUnitI :>: Id)
+%     :..: AssocRightEquiv
+%     :..: (Id :>: InvInvEquivId)
+%     :..: IdElimR
+% \end{verbatim}}
+% %
+% \noindent This program, which can be thought of as an optimization or a proof,
+% relates the two implementations of boolean negation via a series of elementary
+% optimizations of which the following are the most interesting:
 
-{\footnotesize
-\begin{verbatim}
-AssocLeftEquiv    ::           (c1 :.: (c2 :.: c3)) :<===> ((c1 :.: c2) :.: c3)
-SlideCommuteTimes :: (CommuteTimes :.: (c1 :*: c2)) :<===> ((c2 :*: c1) :.: CommuteTimes)
-IdElimL           ::                     (Id :.: c) :<===> c
-SlideUnitI        ::   (UnitTimesI :.: (c1 :*: c2)) :<===> (c2 :.: UnitTimesI)
-\end{verbatim}}
+% {\footnotesize
+% \begin{verbatim}
+% AssocLeftEquiv    ::           (c1 :.: (c2 :.: c3)) :<===> ((c1 :.: c2) :.: c3)
+% SlideCommuteTimes :: (CommuteTimes :.: (c1 :*: c2)) :<===> ((c2 :*: c1) :.: CommuteTimes)
+% IdElimL           ::                     (Id :.: c) :<===> c
+% SlideUnitI        ::   (UnitTimesI :.: (c1 :*: c2)) :<===> (c2 :.: UnitTimesI)
+% \end{verbatim}}
 
-\end{comment}
+% \end{comment}
+
 %%%%%
@@
 
-In previous work, on $\Pi$ we noted a possible connection with HoTT:
-\begin{quote}
-  It is, therefore, at least plausible that a variant of HoTT based exclusively
-  on reversible functions that directly correspond to equivalences would have
-  better computational properties. Our current result is a step, albeit
-  preliminary, in that direction as it only applies to finite
-  types~\cite{Carette2016}.
-\end{quote}
-Formalizing, in a precise sense, the connection between reversible programs
-based on combinators and paths in HoTT, as intuitive as it may seem, is however
-difficult. Paths in HoTT come equipped with principles like the
-``contractibility of singletons'', ``transport'', and ``path induction.'' None
-of these principles seems to have any direct counterpart in the world of
-reversible programming.
-
-We will however show, in a very clear sense, how a $\Pi$-based language is
-``equivalent'' to a univalent HoTT universe that possesses all the required
-properties of paths. The essence of the construction can be demonstrated with a
-tiny ``featherweight'' variant of $\Pi$ that includes just one type: booleans,
-and a tiny ``featherweight'' univalent universe to be formalized in the next
-section. Here we give a formalization of the small $\Pi$-based language which we
-will use in the formalization in the rest of the paper.
-
-The restriction of $\Pi$ to the case of just booleans is the following:
+We present a small $\Pi$-based language which we will use in the formalization
+in the rest of the paper. The restriction of $\Pi$ to the case of just booleans
+is the following:
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Univalent Type Families}
+\section{Univalent Fibrations}
+\label{sec:univalent}
 
@@
 \begin{verbatim}
-Thm : Let P : Type -> Type such that for all X : Type, P(X) is a proposition. Then Σ(X,P) is the base of a univalent fibration.
+Thm: Let P : Type -> Type such that for all X : Type, P(X)
+is a proposition. Then Σ(X,P) is the base of a univalent
+fibration.
 
-Furthermore, we have that *all* univalent fibrations over Type arise this way.
-We don't need this fact directly, but it may help in the explanation.
-(It's certainly cleaner than making the connection to classifying bundles in geometry).
+Furthermore, we have that *all* univalent fibrations
+over Type arise this way. We don't need this fact directly,
+but it may help in the explanation. (It's certainly cleaner
+than making the connection to classifying bundles in
+geometry).
 
@@
 
-is-prop : Type -> Type has the property that is-prop(X) is a proposition for all types X.
-So Σ(X , is-prop) is the base of a univalent fibration.
-In other words, this is the subuniverse consisting of all and only propositions.
+is-prop : Type -> Type has the property that is-prop(X)
+is a proposition for all types X. So Σ(X , is-prop) is
+the base of a univalent fibration. In other words, this
+is the subuniverse consisting of all and only propositions.
 
-is-set : Type -> Type has the property that is-set(X) is a proposition for all types X.
-So Σ(X , is-set) is the base of a univalent fibration.
-In other words, this is the subuniverse consisting of all and only sets.
+is-set : Type -> Type has the property that is-set(X)
+is a proposition for all types X. So Σ(X , is-set) is
+the base of a univalent fibration. In other words, this
+is the subuniverse consisting of all and only sets.
 
-is-1-type : Type -> Type has the property that is-1-type(X) is a proposition for
-all types X. So Σ(X , is-1-type) is the base of a univalent fibration. In other
-words, this is the subuniverse consisting of all and only 1-types. (Externally,
-these are the 1-groupoids embedded in inf-groupoids.)
+is-1-type : Type -> Type has the property that
+is-1-type(X) is a proposition for all types X. So
+Σ(X , is-1-type) is the base of a univalent fibration.
+In other words, this is the subuniverse consisting of all
+and only 1-types. (Externally, these are the 1-groupoids
+embedded in inf-groupoids.)
 
-But here's a summary of what I wrote last week, which I think should suffice for this project.
-Designating a collection of types as a univalent universe boils down to defining a mere predicate on the universe.
-One approach to defining such a predicate is to generate a set of names and an interpretation El (a type family over these names), and ask that the disjunction "is equal to the interpretation of a name" be exclusive (i.e., a proposition).
-This last step requires that the set of names be representatives of distinct normal forms.
-Repackage these data --- the collection of types, their names, and the proof that the predicate on the universe is a proposition -- as a semantic object admitting various presentations, which when computable, can be called a "reversible programming language".
+But here's a summary of what I wrote last week, which
+I think should suffice for this project. Designating a
+collection of types as a univalent universe boils down
+to defining a mere predicate on the universe. One approach
+to defining such a predicate is to generate a set of
+names and an interpretation El (a type family over
+these names), and ask that the disjunction "is equal
+to the interpretation of a name" be exclusive (i.e.,
+a proposition). This last step requires that the set
+of names be representatives of distinct normal forms.
+Repackage these data --- the collection of types, their
+names, and the proof that the predicate on the universe
+is a proposition -- as a semantic object admitting
+various presentations, which when computable, can be
+called a "reversible programming language".
 
-Because "the proof that the predicate on the universe is a proposition" (and univalence), we naturally get a family of equivalences arising.
-At various levels.
-When reflected syntactically, we get multiple presentations of the same semantic object (where 'same' is really up to things 1 level up).
-Because all we can really talk about are equivalences, when this syntax is interpreted as a programming language, it is unavoidably reversible.
+Because "the proof that the predicate on the universe
+is a proposition" (and univalence), we naturally get a
+family of equivalences arising.  At various levels.
+When reflected syntactically, we get multiple
+presentations of the same semantic object (where
+'same' is really up to things 1 level up).  Because
+all we can really talk about are equivalences, when
+this syntax is interpreted as a programming language,
+it is unavoidably reversible.
 
-Right, the predicate being a proposition suffices (and is necessary) to define an embedding into a univalent universe.
+Right, the predicate being a proposition suffices
+(and is necessary) to define an embedding into a
+univalent universe.
 
-(Univalence is used axiomatically, as is the rest of the type theory, to define the universe.
-But you're welcome to interpret ITT + univalence in simplicial sets and forget about univalence altogether: the univalent universe is a particular simplicial set defined in classical set theory, as are the rest of the subuniverses serving as denotational models.)
+(Univalence is used axiomatically, as is the rest of
+the type theory, to define the universe. But you're
+welcome to interpret ITT + univalence in simplicial
+sets and forget about univalence altogether: the
+univalent universe is a particular simplicial set
+defined in classical set theory, as are the rest
+of the subuniverses serving as denotational models.)
 
-By "admitting various presentations", I mean presentation in the sense of group presentation.
-For example, the real functions on R^2 which are symmetries of the square centered at the origin can be viewed as the dihedral group on 8 elements: but we're free to present this group in an infinite number of ways.
-These presentations can be taken to be syntax for a language.
-If a presentation has a decidable word problem, and you have the algorithm, then you could be said to have an operational semantics, and so a programming language of a sort.
-This is what, I take it, you and Amr did with Laplaza's work on presenting free distributive categories in the first place.
-There your semantic object is the syntactic category of his theory of distributive categories.
+By "admitting various presentations", I mean
+presentation in the sense of group presentation.
+For example, the real functions on R^2 which are
+symmetries of the square centered at the origin
+can be viewed as the dihedral group on 8 elements:
+but we're free to present this group in an infinite
+number of ways. These presentations can be taken to
+be syntax for a language. If a presentation has a
+decidable word problem, and you have the algorithm,
+then you could be said to have an operational
+semantics, and so a programming language of a sort.
+This is what, I take it, you and Amr did with Laplaza's
+work on presenting free distributive categories
+in the first place. There your semantic object is
+the syntactic category of his theory of distributive
+categories.
 
-What using HoTT as a metatheory facilitates here is the construction of models which admit presentations of infinity-groupoids.
-(You could do it without HoTT.
-But the results would likely be less general, say if the work was done directly in simplicial sets or cubical sets.) Using this approach, you have a lot more options now for cooking up reversible PLs now; just as many as you have decidable presentations of inf-groupoids.
+What using HoTT as a metatheory facilitates here is
+the construction of models which admit presentations
+of infinity-groupoids. (You could do it without HoTT.
+But the results would likely be less general, say if
+the work was done directly in simplicial sets or
+cubical sets.) Using this approach, you have a lot
+more options now for cooking up reversible PLs now;
+just as many as you have decidable presentations of
+inf-groupoids.
 
-The 'exclusive disjunction' aspect of being a proposition seems to have more content: it's a sort of decidability / normal form property all rolled into one.
+The 'exclusive disjunction' aspect of being a
+proposition seems to have more content: it's a
+sort of decidability / normal form property all
+rolled into one.
 
-Yes, I was struck by this too.
-At least insofar as we continue with this style of defining predicates on the universe to obtain the intended interpretation of the types in a rev PL, it seems that a decidable enumeration of normal forms of the types will already be needed.
-But this doesn't seem like a huge constraint to me; most of the action is at the level of the paths....
+Yes, I was struck by this too. At least insofar
+as we continue with this style of defining predicates
+on the universe to obtain the intended interpretation
+of the types in a rev PL, it seems that a decidable
+enumeration of normal forms of the types will already
+be needed. But this doesn't seem like a huge constraint
+to me; most of the action is at the level of the paths....
 
-I guess that you assumed in "This last step requires that the set of names be representatives of distinct normal forms" that these were also complete?
-That we have a decidable enumeration of all normal forms?
+I guess that you assumed in "This last step requires
+that the set of names be representatives of distinct
+normal forms" that these were also complete?  That we
+have a decidable enumeration of all normal forms?
 
-I haven't said anything about an actual presentation of the inf-groupoid which is the univalent universe of finite types.
-You want Pi to do this work.
-So to get started, we should define a function from the BNF grammar corresponding to a syntax for a semiring to the terms of that universe.
-As it stands, we already know that those terms are reflected in Names, so it would be sufficient to define that function into Names.
-Ah, is this what you meant by "complete"?
-Yes, we have a name for everything in the extension of that predicate (by construction).
+I haven't said anything about an actual presentation
+of the inf-groupoid which is the univalent universe
+of finite types. You want Pi to do this work. So to
+get started, we should define a function from the
+BNF grammar corresponding to a syntax for a semiring
+to the terms of that universe. As it stands, we
+already know that those terms are reflected in Names,
+so it would be sufficient to define that function
+into Names. Ah, is this what you meant by "complete"?
+Yes, we have a name for everything in the extension of
+that predicate (by construction).
 
@@
 \section{Correspondence}
+\label{sec:connection}
+
+In previous work, on $\Pi$ we noted a possible connection with HoTT:
+\begin{quote}
+  It is, therefore, at least plausible that a variant of HoTT based exclusively
+  on reversible functions that directly correspond to equivalences would have
+  better computational properties. Our current result is a step, albeit
+  preliminary, in that direction as it only applies to finite
+  types~\cite{Carette2016}.
+\end{quote}
+Formalizing, in a precise sense, the connection between reversible programs
+based on combinators and paths in HoTT, as intuitive as it may seem, is however
+difficult. Paths in HoTT come equipped with principles like the
+``contractibility of singletons'', ``transport'', and ``path induction.'' None
+of these principles seems to have any direct counterpart in the world of
+reversible programming.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
+\section{Discussion, Related Work, and Conclusion}
+\label{sec:conclusion}
+
+\paragraph*{Reversible Languages.}
+\noindent The practice of programming languages is replete with \emph{ad hoc}
+instances of reversible computations: database transactions, mechanisms for data
+provenance, checkpoints, stack and exception traces, logs, backups, rollback
+recoveries, version control systems, reverse engineering, software transactional
+memories, continuations, backtracking search, and multiple-level ``undo''
+features in commercial applications. In the early nineties,
+Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
+treatment of reversibility. But intensive research in full-fledged reversible
+models of computations and reversible programming languages was only sparked by
+the discovery of deep connections between physics and
+computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
+and by the potential for efficient quantum
+computation~\cite{springerlink:10.1007/BF02650179}.
+
+The early developments of reversible programming languages started with a
+conventional programming language, e.g., an extended $\lambda$-calculus, and either
+\begin{enumerate}
+\item extended the language with a history
+mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
+\item imposed constraints on the control flow constructs to make them
+reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.
+\end{enumerate}
+More modern approaches recognize that reversible programming languages require
+a fresh approach and should be designed from first principles without the
+detour via conventional irreversible
+languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.
+
+\paragraph*{The $\Pi$ Family of Languages}
+\noindent In previous work, Carette, Bowman, James, and
+Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
+the~$\Pi$ family of reversible languages based on type isomorphisms and
+commutative semiring identities. The fragment without recursive types is
+universal for reversible boolean circuits~\cite{James:2012:IE:2103656.2103667}
+and the extension with recursive types and trace
+operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
+language~\cite{James:2012:IE:2103656.2103667,rc2011}. While at first sight,
+$\Pi$ might appear \emph{ad hoc},~\cite{Carette2016} shows that it arises
+naturally from an ``extended'' view of the Curry-Howard correspondance: rather
+than looking at mere \emph{inhabitation} as the main source of analogy between
+logic and computation, \emph{type equivalences} becomes the source of analogy.
+This allows one to see an analogy between algebra and reversible computation.
+Furthermore, this works at multiple levels: that of $1$-algebra (types form a
+semiring under isomorphism), but also $2$-algebra (types and equivalences form a
+weak Rig Groupoid).  In other words, by taking ``weak Rig Groupoid'' as the
+starting semantics, one naturally gets $\Pi$ as the syntax for the language of
+proofs of isomorphisms -- in the same way that many terms of the
+$\lambda$-calculus arise from Cartesian Closed Categories.
+
+By restricting to finite types, two operational semantics emerge:
+\begin{enumerate}
+\item a (reversible) operational semantics of programs over finite types, and
+\item a (reversible) operational semantics of program transformers of such programs.
+\end{enumerate}
+$\Pi$ programs can be interpreted as generalized permutations, as they work on
+structure as well as the underlying skeleton of finite types, aka the natural numbers.
+Many of the transformers can be \emph{oriented}, and thus be interpreted as
+\emph{optimizations}.  As with traditional languages, there are different aspects
+we can optimize for (program size, number of steps to completion, etc); we will
+not touch on this further in this paper.
 

### 2017-06-26 upi.tex
toffoli

@@
 \usepackage[inline]{enumitem}
+\usepackage{agda}
 
@@
 computations are \emph{type isomorphisms} which, at least in the case of finite
-types, clearly preserve entropy in the information-theoretic sense.
-
-We will identify a ``featherweight'' version of $\Pi$ to use in our formal
-development.  But first we illustrate the flavor of the family of languages with
-some examples.
+types, clearly preserve entropy in the information-theoretic
+sense~\cite{James:2012:IE:2103656.2103667}. We will identify a ``featherweight''
+version of $\Pi$ to use in our formal development.  But first we illustrate the
+general flavor of the family of languages with some examples.
 
@@
 can also read the reverse program by reading ``up''.}
+
+\newcommand{\byiso}[1]{{\leftrightarrow}{\langle} ~#1~ \rangle}
+
+{\small
+\[\def\arraystretch{1.2}\begin{array}{rcl}
+\AgdaFunction{toffoli} &:& \bt \otimes (\bt \otimes \bt)
+                           \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
+\AgdaFunction{toffoli} &=&
+      \bt \otimes (\bt \otimes \bt) \\
+&& \qquad\byiso{\AgdaFunction{unfoldBool} \otimes \AgdaFunction{id}} \\
+&& (\ot \oplus \ot) \otimes (\bt \otimes \bt) \\
+&& \qquad\byiso{\AgdaFunction{dist}} \\
+&& (\ot \otimes (\bt \otimes \bt)) \oplus (\ot \otimes (\bt \otimes \bt)) \\
+&& \qquad\byiso{\AgdaFunction{id} \oplus (\AgdaFunction{id} \otimes \AgdaFunction{cnot})} \\
+&& (\ot \otimes (\bt \otimes \bt)) \oplus (\ot \otimes (\bt \otimes \bt)) \\
+&& \qquad\byiso{\AgdaFunction{factor}} \\
+&& (\ot \oplus \ot) \otimes (\bt \otimes \bt) \\
+&& \qquad\byiso{\AgdaFunction{foldBool} \otimes \AgdaFunction{id}} \\
+&& \bt \otimes (\bt \otimes \bt)
+\end{array}\]}
 

### 2017-06-26 upi.tex
code

@@
 \[\def\arraystretch{1.2}\begin{array}{rcl}
+\AgdaFunction{controlled} &:& (a \leftrightarrow a) \rightarrow 
+                              \bt \otimes a \leftrightarrow \bt \otimes a \\
+\AgdaFunction{controlled}~\AgdaFunction{f} &=& 
+      \bt \otimes a \\
+&& \qquad\byiso{\AgdaFunction{unfoldBool} \otimes \AgdaFunction{id}} \\
+&& (\ot \oplus \ot) \otimes a \\
+&& \qquad\byiso{\AgdaFunction{dist}} \\
+&& (\ot \otimes a) \oplus (\ot \otimes a) \\
+&& \qquad\byiso{\AgdaFunction{id} \oplus (\AgdaFunction{id} \otimes \AgdaFunction{f})} \\
+&& (\ot \otimes a) \oplus (\ot \otimes a) \\
+&& \qquad\byiso{\AgdaFunction{factor}} \\
+&& (\ot \oplus \ot) \otimes a \\
+&& \qquad\byiso{\AgdaFunction{foldBool} \otimes \AgdaFunction{id}} \\
+&& \bt \otimes a \\
+\\ 
+\AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
+\AgdaFunction{not} &=& 
+  \AgdaFunction{unfoldBool} \odot \AgdaFunction{swap₊} \odot \AgdaFunction{foldBool} \\
+\\ 
+\AgdaFunction{cnot} &:& \bt \otimes \bt \leftrightarrow \bt \otimes \bt \\
+\AgdaFunction{cnot} &=& \AgdaFunction{controlled}~\AgdaFunction{not} \\
+\\ 
 \AgdaFunction{toffoli} &:& \bt \otimes (\bt \otimes \bt)
                            \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
-\AgdaFunction{toffoli} &=&
-      \bt \otimes (\bt \otimes \bt) \\
-&& \qquad\byiso{\AgdaFunction{unfoldBool} \otimes \AgdaFunction{id}} \\
-&& (\ot \oplus \ot) \otimes (\bt \otimes \bt) \\
-&& \qquad\byiso{\AgdaFunction{dist}} \\
-&& (\ot \otimes (\bt \otimes \bt)) \oplus (\ot \otimes (\bt \otimes \bt)) \\
-&& \qquad\byiso{\AgdaFunction{id} \oplus (\AgdaFunction{id} \otimes \AgdaFunction{cnot})} \\
-&& (\ot \otimes (\bt \otimes \bt)) \oplus (\ot \otimes (\bt \otimes \bt)) \\
-&& \qquad\byiso{\AgdaFunction{factor}} \\
-&& (\ot \oplus \ot) \otimes (\bt \otimes \bt) \\
-&& \qquad\byiso{\AgdaFunction{foldBool} \otimes \AgdaFunction{id}} \\
-&& \bt \otimes (\bt \otimes \bt)
+\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot}
 \end{array}\]}

### 2017-06-27 upi.tex
not versions
not opt

@@
 
+\newcommand{\byiso}[1]{{\leftrightarrow}{\langle} ~#1~ \rangle}
+\newcommand{\byisotwo}[1]{{\Leftrightarrow}{\langle} ~#1~ \rangle}
 \newcommand{\unitepl}{\texttt{unitepl}}
@@
 
-\jacques{Example 1: using the combinators from pi-dual, some an equational
-derivation in type theory (which?), in equational-derivation style.  Point out
-that reading ``down'' the justifications is a $\Pi$ program which is also a
-constructive proof of the equivalence of the types.  As it is reversible, we
-can also read the reverse program by reading ``up''.}
-
-\newcommand{\byiso}[1]{{\leftrightarrow}{\langle} ~#1~ \rangle}
-
 {\small
 \[\def\arraystretch{1.2}\begin{array}{rcl}
-\AgdaFunction{controlled} &:& (a \leftrightarrow a) \rightarrow 
-                              \bt \otimes a \leftrightarrow \bt \otimes a \\
+\AgdaFunction{controlled} &:& \forall a.~ (a \leftrightarrow a) \rightarrow 
+                              (\bt \otimes a \leftrightarrow \bt \otimes a) \\
 \AgdaFunction{controlled}~\AgdaFunction{f} &=& 
@@
 \\ 
+\AgdaFunction{not}~\AgdaFunction{not₁}~\AgdaFunction{not₂}~\AgdaFunction{not₃} &:& 
+     \bt \leftrightarrow \bt \\
+\AgdaFunction{not} &=& 
+  \AgdaFunction{unfoldBool} \odot \AgdaFunction{swap₊} \odot \AgdaFunction{foldBool} \\
+\AgdaFunction{not₁} &=& 
+  \AgdaFunction{id} \odot \AgdaFunction{not} \\
+\AgdaFunction{not₂} &=& 
+  \AgdaFunction{not} \odot \AgdaFunction{not} \odot \AgdaFunction{not} \\
+\AgdaFunction{not₃} &=& 
+  \AgdaFunction{uniti⋆} \odot \AgdaFunction{swap⋆} \odot 
+                        (\AgdaFunction{not} \otimes \AgdaFunction{id}) \odot 
+                        \AgdaFunction{swap⋆} \odot 
+                        \AgdaFunction{unite⋆} \\
+\\ 
 \AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
@@
                            \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
-\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot}
+\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot} \\
+\\
+\AgdaFunction{notOpt} &:& \AgdaFunction{not₃} \Leftrightarrow \AgdaFunction{not} \\
+\AgdaFunction{notOpt} &=& 
+  \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
+                        ((\AgdaFunction{not} \otimes \AgdaFunction{id}) \odot 
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}))) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocLeft}} \\
+&& \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
+                        (\AgdaFunction{not} \otimes \AgdaFunction{id})) \odot 
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{swapLeft} 
+                                  \boxdot \AgdaFunction{id})} \\
+&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
+                      \odot \AgdaFunction{swap⋆}) \odot 
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocRight}} \\
+&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
+                      \odot (\AgdaFunction{swap⋆} \odot 
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}))) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
+                                  \boxdot \AgdaFunction{assocLeft})} \\
+&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
+                      \odot ((\AgdaFunction{swap⋆} \odot 
+                      \AgdaFunction{swap⋆}) \odot \AgdaFunction{unite⋆})) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
+                                  \boxdot (\AgdaFunction{leftInv} \boxdot \AgdaFunction{id}))} \\
+&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
+                      \odot (\AgdaFunction{id} \odot \AgdaFunction{unite⋆})) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
+                                  \boxdot \AgdaFunction{idLeft})} \\
+&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
+                      \odot \AgdaFunction{unite⋆}) \\
+&& \qquad\byisotwo{\AgdaFunction{assocLeft}} \\
+&& (\AgdaFunction{uniti⋆} \odot (\AgdaFunction{id} \otimes \AgdaFunction{not})) 
+                      \odot \AgdaFunction{unite⋆} \\
+&& \qquad\byisotwo{\AgdaFunction{unitiLeft} \boxdot \AgdaFunction{id}} \\
+&& (\AgdaFunction{not} \otimes \AgdaFunction{uniti⋆}) \odot \AgdaFunction{unite⋆} \\    
+&& \qquad\byisotwo{\AgdaFunction{assocRight}} \\
+&& \AgdaFunction{not} \otimes (\AgdaFunction{uniti⋆} \odot \AgdaFunction{unite⋆}) \\    
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{leftInv}} \\
+&& \AgdaFunction{not} \otimes \AgdaFunction{id} \\
+&& \qquad\byisotwo{\AgdaFunction{idRight}} \\
+&& \AgdaFunction{not} 
 \end{array}\]}
+
+\jacques{Example 1: using the combinators from pi-dual, some an equational
+derivation in type theory (which?), in equational-derivation style.  Point out
+that reading ``down'' the justifications is a $\Pi$ program which is also a
+constructive proof of the equivalence of the types.  As it is reversible, we
+can also read the reverse program by reading ``up''.}
 

### 2017-06-27 upi.tex
progress on sec 2

--- 2017-06-26 upi.tex
@@
 
-\jacques{Example 1: using the combinators from pi-dual, some an equational
-derivation in type theory (which?), in equational-derivation style.  Point out
-that reading ``down'' the justifications is a $\Pi$ program which is also a
-constructive proof of the equivalence of the types.  As it is reversible, we
-can also read the reverse program by reading ``up''.}
-
-\newcommand{\byiso}[1]{{\leftrightarrow}{\langle} ~#1~ \rangle}
+The examples below assume a representation of the type of booleans
+$\bt$ as the disjoint union $\ot \oplus \ot$ with the left injection
+representing $\mathsf{false}$ and the right injection representing
+$\mathsf{true}$. Given an arbitrary reversible function
+$\AgdaFunction{f}$ of type $a \leftrightarrow a$, we can build the
+reversible function $\AgdaFunction{controlled}~\AgdaFunction{f}$ that
+takes a pair of type $\bt \otimes a$ and checks the incoming boolean;
+if it is false (i.e., we are in the left injection), the function
+behaves like the identity; otherwise the function applies
+$\AgdaFunction{f}$ to the second argument. In either case, the
+incoming boolean is reconstituted to maintain reversibility. Using the
+\AgdaFunction{controlled} building block, we can build a
+controlled-not $\AgdaFunction{cnot}$ gate and a
+controlled-controlled-not gate, also known as the
+\AgdaFunction{toffoli} gate. The latter gate is a universal function
+for combinational boolean circuits thus showing the expressiveness of
+the language:
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.9}]
+  \draw (1,2) ellipse (0.5cm and 0.5cm);
+  \draw[fill] (1,2) circle [radius=0.025];
+  \node[below] at (1,2) {*};
+
+  \draw (0,0) ellipse (0.5cm and 1cm);
+  \draw[fill] (0,0.5) circle [radius=0.025];
+  \node[below] at (0,0.5) {F};
+  \draw[fill] (0,-0.5) circle [radius=0.025];
+  \node[below] at (0,-0.5) {T};
+
+  \draw     (1,2)    -- (2,2)      ; %% ()
+  \draw     (0,0.5)  -- (2,0.5)    ; %% F
+  \draw     (0,-0.5) -- (2,-0.5)   ; %% T
+
+  \draw     (2,2)    -- (3,-0.5)   ;
+  \draw     (2,0.5)  -- (3,2)      ;
+  \draw     (2,-0.5) -- (3,1)      ;
+
+  \draw     (3,2)    -- (3.5,2)    ;
+  \draw     (3,1)    -- (3.5,1)    ;
+  \draw     (3,-0.5) -- (3.5,-0.5) ;
+
+  \draw     (3.5,2)    -- (4.5,1)    ;
+  \draw     (3.5,1)    -- (4.5,2)    ;
+  \draw     (3.5,-0.5) -- (4.5,-0.5) ;
+
+  \draw     (4.5,2)    -- (5,2)    ;
+  \draw     (4.5,1)    -- (5,1)    ;
+  \draw     (4.5,-0.5) -- (5,-0.5) ;
+
+  \draw     (5,2)    -- (6,0.5)  ;
+  \draw     (5,1)    -- (6,-0.5) ;
+  \draw     (5,-0.5) -- (6,2)    ;
+
+  \draw     (6,2)    -- (7,2)    ;
+  \draw     (6,0.5)  -- (8,0.5)  ;
+  \draw     (6,-0.5) -- (8,-0.5) ;
+
+  \draw (7,2) ellipse (0.5cm and 0.5cm);
+  \draw[fill] (7,2) circle [radius=0.025];
+  \node[below] at (7,2) {*};
+
+  \draw (8,0) ellipse (0.5cm and 1cm);
+  \draw[fill] (8,0.5) circle [radius=0.025];
+  \node[below] at (8,0.5) {F};
+  \draw[fill] (8,-0.5) circle [radius=0.025];
+  \node[below] at (8,-0.5) {T};
+
+\end{tikzpicture}
+\end{center}
+\caption{\label{fig:not}Graphical representation of \AgdaFunction{not₃}}
+\end{figure}
 
@@
 \[\def\arraystretch{1.2}\begin{array}{rcl}
-\AgdaFunction{controlled} &:& (a \leftrightarrow a) \rightarrow 
-                              \bt \otimes a \leftrightarrow \bt \otimes a \\
+\AgdaFunction{controlled} &:& \forall a.~ (a \leftrightarrow a) \rightarrow 
+                              (\bt \otimes a \leftrightarrow \bt \otimes a) \\
 \AgdaFunction{controlled}~\AgdaFunction{f} &=& 
@@
 && (\ot \oplus \ot) \otimes a \\
-&& \qquad\byiso{\AgdaFunction{dist}} \\
+&& \qquad\byiso{\AgdaFunction{distribute}} \\
 && (\ot \otimes a) \oplus (\ot \otimes a) \\
@@
                            \leftrightarrow  \bt \otimes (\bt \otimes \bt) \\
-\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot}
+\AgdaFunction{toffoli} &=& \AgdaFunction{controlled}~\AgdaFunction{cnot} \\
 \end{array}\]}
+%%%
+
+The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
+constructive evidence (i.e., a program, a logic gate, or a hardware
+circuit) for an automorphism on $\bt \otimes a$: it can be read
+top-down or bottom-up to go back and forth. This perspective naturally
+raises the questions of when two programs are ``equivalent.'' Consider
+the six programs of type $\bt \leftrightarrow \bt$ below:
+
+{\small 
+\[\def\arraystretch{1.2}\begin{array}{rcl}
+\AgdaFunction{id₁}~\AgdaFunction{id₂}~\AgdaFunction{id₃}~ 
+  \AgdaFunction{not₁}~\AgdaFunction{not₂}~\AgdaFunction{not₃} &:& \bt \leftrightarrow \bt \\
+\AgdaFunction{id₁} &=& 
+  \AgdaFunction{id} \odot \AgdaFunction{id} \\
+\AgdaFunction{id₂} &=& 
+  \AgdaFunction{not} \odot \AgdaFunction{id} \odot \AgdaFunction{not} \\
+\AgdaFunction{id₃} &=& 
+  \AgdaFunction{uniti⋆} \odot \AgdaFunction{swap⋆} \odot 
+                        (\AgdaFunction{id} \otimes \AgdaFunction{id}) \odot 
+                        \AgdaFunction{swap⋆} \odot 
+                        \AgdaFunction{unite⋆} \\
+\AgdaFunction{not₁} &=& 
+  \AgdaFunction{id} \odot \AgdaFunction{not} \\
+\AgdaFunction{not₂} &=& 
+  \AgdaFunction{not} \odot \AgdaFunction{not} \odot \AgdaFunction{not} \\
+\AgdaFunction{not₃} &=& 
+  \AgdaFunction{uniti⋆} \odot \AgdaFunction{swap⋆} \odot 
+                        (\AgdaFunction{not} \otimes \AgdaFunction{id}) \odot 
+                        \AgdaFunction{swap⋆} \odot 
+                        \AgdaFunction{unite⋆} 
+\end{array}\]}
+
+\noindent We neither expect, nor desire, that all these six programs
+be equivalent. However each group of three programs arguably
+represents ``the same'' program. Thinking extensionally, it is indeed
+easy to verify this fact but we can do better. We can provide
+\emph{evidence} (i.e., a program that manipulates programs) that can
+constructively transform one program to an equivalent one. We show
+such a level-2 program proving that $\AgdaFunction{not₃}$ is
+equivalent to $\AgdaFunction{not}$:
+
+{\small 
+\[\def\arraystretch{1.2}\begin{array}{rcl}
+\AgdaFunction{notOpt} &:& \AgdaFunction{not₃} \Leftrightarrow \AgdaFunction{not} \\
+\AgdaFunction{notOpt} &=& 
+  \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
+                        ((\AgdaFunction{not} \otimes \AgdaFunction{id}) \odot 
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}))) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocLeft}} \\
+&& \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
+                        (\AgdaFunction{not} \otimes \AgdaFunction{id})) \odot 
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{swapLeft} 
+                                  \boxdot \AgdaFunction{id})} \\
+&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
+                      \odot \AgdaFunction{swap⋆}) \odot 
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocRight}} \\
+&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
+                      \odot (\AgdaFunction{swap⋆} \odot 
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}))) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
+                                  \boxdot \AgdaFunction{assocLeft})} \\
+&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
+                      \odot ((\AgdaFunction{swap⋆} \odot 
+                      \AgdaFunction{swap⋆}) \odot \AgdaFunction{unite⋆})) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
+                                  \boxdot (\AgdaFunction{leftInv} \boxdot \AgdaFunction{id}))} \\
+&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
+                      \odot (\AgdaFunction{id} \odot \AgdaFunction{unite⋆})) \\
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
+                                  \boxdot \AgdaFunction{idLeft})} \\
+&& \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
+                      \odot \AgdaFunction{unite⋆}) \\
+&& \qquad\byisotwo{\AgdaFunction{assocLeft}} \\
+&& (\AgdaFunction{uniti⋆} \odot (\AgdaFunction{id} \otimes \AgdaFunction{not})) 
+                      \odot \AgdaFunction{unite⋆} \\
+&& \qquad\byisotwo{\AgdaFunction{unitiLeft} \boxdot \AgdaFunction{id}} \\
+&& (\AgdaFunction{not} \otimes \AgdaFunction{uniti⋆}) \odot \AgdaFunction{unite⋆} \\    
+&& \qquad\byisotwo{\AgdaFunction{assocRight}} \\
+&& \AgdaFunction{not} \otimes (\AgdaFunction{uniti⋆} \odot \AgdaFunction{unite⋆}) \\    
+&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{leftInv}} \\
+&& \AgdaFunction{not} \otimes \AgdaFunction{id} \\
+&& \qquad\byisotwo{\AgdaFunction{idRight}} \\
+&& \AgdaFunction{not} 
+\end{array}\]}
+
+%% \jacques{Example 1: using the combinators from pi-dual, some an equational
+%% derivation in type theory (which?), in equational-derivation style.  Point out
+%% that reading ``down'' the justifications is a $\Pi$ program which is also a
+%% constructive proof of the equivalence of the types.  As it is reversible, we
+%% can also read the reverse program by reading ``up''.}
 
@@
 
-\jacques{Refer to some Sabry et al. papers regarding Toffoli gates.}
+%% \jacques{Refer to some Sabry et al. papers regarding Toffoli gates.}
 

### 2017-06-27 upi.tex
pi2

@@
 top-down or bottom-up to go back and forth. This perspective naturally
-raises the questions of when two programs are ``equivalent.'' Consider
+raises the question of when two programs are ``equivalent.'' Consider
 the six programs of type $\bt \leftrightarrow \bt$ below:
@@
 
-\noindent We neither expect, nor desire, that all these six programs
-be equivalent. However each group of three programs arguably
-represents ``the same'' program. Thinking extensionally, it is indeed
-easy to verify this fact but we can do better. We can provide
-\emph{evidence} (i.e., a program that manipulates programs) that can
-constructively transform one program to an equivalent one. We show
-such a level-2 program proving that $\AgdaFunction{not₃}$ is
-equivalent to $\AgdaFunction{not}$:
+\noindent We neither expect, nor desire, that all these programs of
+the same type to be be equivalent. In the case above, each group of
+three programs does represent ``the same'' program but the two groups
+represent distinct programs. Thinking extensionally, i.e., by looking
+at all possible input-output pairs, it is indeed easy to verify this
+fact but we can do better. We can provide \emph{evidence} (i.e., a
+program that manipulates programs) that can constructively transform
+every program to an equivalent one. We show such a level-2 program
+proving that $\AgdaFunction{not₃}$ is equivalent to
+$\AgdaFunction{not}$. For illustration, the program for
+$\AgdaFunction{not₃}$ is depicted in Fig.~\ref{fig:not}. We encourage
+the reader to map the steps below to manipulations on the diagram that
+would incrementally simplify it:
 
@@
 
-%% \jacques{Example 1: using the combinators from pi-dual, some an equational
-%% derivation in type theory (which?), in equational-derivation style.  Point out
-%% that reading ``down'' the justifications is a $\Pi$ program which is also a
-%% constructive proof of the equivalence of the types.  As it is reversible, we
-%% can also read the reverse program by reading ``up''.}
-
-\jacques{Example 2: like example 1, but this time at the next level (i.e. a
-program equivalence).  Same remark old.  Refer to Shulmann notes, and point out
-that such equational proofs at this level can actually be interpreted in appropriate
-categories as commutative diagrams.}
-
-%% \jacques{Refer to some Sabry et al. papers regarding Toffoli gates.}
-
-% \begin{comment}
-% In this paper we will
-% identify a ``featherweight'' representative to use for the formal
-% development. But first, we illustrate the flavor of the family of languages with
-% some examples written in a Haskell syntax. First consider the following program:
-
-% {\footnotesize
-% \begin{verbatim}
-% addSub1 :: (Int, Int) :<=> (Int, Int)
-% addSub1 = CommuteTimes
-%       :.: (UnfoldN :*: Id)
-%       :.: Distribute
-%       :.: (Id :+: CommuteTimes)
-%       :.: Factor
-%       :.: (FoldN :*: Id)
-% \end{verbatim}}
-% %
-% \noindent which takes a pair of natural numbers and processes them as follows depending on
-% whether the second number is 0 or not. If the second number is 0, the evaluation
-% proceeds as follows:
-% $(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
-% \mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
-% second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
-% $(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
-% \mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
-% (\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
-% applications of this program would increment the first number and decrement the
-% second number maintaining the sum at all times. When the second number reaches
-% 0, the numbers are swapped.
-
-% As another example, consider the following expressions culminating with the
-% implementation of the Toffoli gate, which is universal for combinational
-% circuits. We have two implementations of simple boolean negation: a direct one
-% and a more convoluted one. These two implementations are ``equivalent'' in the
-% sense to be explained below. The expression \verb|cond| is the main reversible
-% control flow mechanism: it allows for the conditional execution of one of two
-% branches but retains the choice of which branch was executed. This mechanism can
-% be used twice to implement the Toffoli gate, which performs a conditional
-% negation on one bit if two other bits are both true:
-
-% {\footnotesize
-% \begin{verbatim}
-% inot :: Bool :<=> Bool
-% inot = CommutePlus
-
-% inot2 :: Bool :<=> Bool
-% inot2 = UnitTimesI
-%     :.: CommuteTimes
-%     :.: (CommutePlus :*: Id)
-%     :.: CommuteTimes
-%     :.: UnitTimesE
-
-% cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
-% cond f g = Distribute
-%        :.: ((Id :*: f) :+: (Id :*: g))
-%        :.: Factor
-
-% controlled :: (a :<=> a) -> ((Bool, a) :<=> (Bool, a))
-% controlled f = cond f Id
-
-% cnot :: (Bool, Bool) :<=> (Bool, Bool)
-% cnot = controlled inot
-
-% toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
-% toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
-% \end{verbatim}}
-
-% Although writing circuits using the raw syntax for combinators above is tedious,
-% the examples illustrate the programming language nature of this family of $\Pi$
-% languages. In other work, one can find a compiler from a conventional functional
-% language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
-% technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
-% Haskell-like surface language~\cite{theseus} which can ease writing
-% programs. All that reinforces the claim that this is a practical approach to
-% reversible programming.
-
-% The raw syntax for combinators has, however, a significant benefit: it allows
-% algebraic manipulation of the programs using a second level of combinators. As
-% an example it is possible to show the the two implementations of boolean
-% negation are equivalent as follows:
-
-% {\footnotesize
-% \begin{verbatim}
-% negOpt :: inot2 :<===> inot
-% negOpt = (Id :>: AssocLeftEquiv)
-%     :..: (Id :>: (SlideCommuteTimes :>: Id))
-%     :..: (Id :>: AssocRightEquiv)
-%     :..: (Id :>: (Id :>: AssocLeftEquiv))
-%     :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
-%     :..: (Id :>: (Id :>: IdElimL))
-%     :..: AssocLeftEquiv
-%     :..: (SlideUnitI :>: Id)
-%     :..: AssocRightEquiv
-%     :..: (Id :>: InvInvEquivId)
-%     :..: IdElimR
-% \end{verbatim}}
-% %
-% \noindent This program, which can be thought of as an optimization or a proof,
-% relates the two implementations of boolean negation via a series of elementary
-% optimizations of which the following are the most interesting:
-
-% {\footnotesize
-% \begin{verbatim}
-% AssocLeftEquiv    ::           (c1 :.: (c2 :.: c3)) :<===> ((c1 :.: c2) :.: c3)
-% SlideCommuteTimes :: (CommuteTimes :.: (c1 :*: c2)) :<===> ((c2 :*: c1) :.: CommuteTimes)
-% IdElimL           ::                     (Id :.: c) :<===> c
-% SlideUnitI        ::   (UnitTimesI :.: (c1 :*: c2)) :<===> (c2 :.: UnitTimesI)
-% \end{verbatim}}
-
-% \end{comment}
-
 %%%%%
@@
 
-We present a small $\Pi$-based language which we will use in the formalization
-in the rest of the paper. The restriction of $\Pi$ to the case of just booleans
-is the following:
-
-\[\begin{array}{rcl}
+Having illustrated the general flavor of the $\Pi$ family of
+languages, we present in full detail a small $\Pi$-based language
+which we will use in the formalization in the rest of the paper. The
+language below is the restriction of $\Pi$ to the case of just one
+type, the type of booleans.
+
+\[\def\arraystretch{0.8}\begin{array}{rcl}
 \tau &::=& \bt \\
@@
 
+%% \jacques{Example 1: using the combinators from pi-dual, some an equational
+%% derivation in type theory (which?), in equational-derivation style.  Point out
+%% that reading ``down'' the justifications is a $\Pi$ program which is also a
+%% constructive proof of the equivalence of the types.  As it is reversible, we
+%% can also read the reverse program by reading ``up''.}
+
+%% \jacques{Refer to some Sabry et al. papers regarding Toffoli gates.}
+
+% \begin{comment}
+% In this paper we will
+% identify a ``featherweight'' representative to use for the formal
+% development. But first, we illustrate the flavor of the family of languages with
+% some examples written in a Haskell syntax. First consider the following program:
+
+% {\footnotesize
+% \begin{verbatim}
+% addSub1 :: (Int, Int) :<=> (Int, Int)
+% addSub1 = CommuteTimes
+%       :.: (UnfoldN :*: Id)
+%       :.: Distribute
+%       :.: (Id :+: CommuteTimes)
+%       :.: Factor
+%       :.: (FoldN :*: Id)
+% \end{verbatim}}
+% %
+% \noindent which takes a pair of natural numbers and processes them as follows depending on
+% whether the second number is 0 or not. If the second number is 0, the evaluation
+% proceeds as follows:
+% $(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
+% \mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
+% second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
+% $(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
+% \mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
+% (\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
+% applications of this program would increment the first number and decrement the
+% second number maintaining the sum at all times. When the second number reaches
+% 0, the numbers are swapped.
+
+% As another example, consider the following expressions culminating with the
+% implementation of the Toffoli gate, which is universal for combinational
+% circuits. We have two implementations of simple boolean negation: a direct one
+% and a more convoluted one. These two implementations are ``equivalent'' in the
+% sense to be explained below. The expression \verb|cond| is the main reversible
+% control flow mechanism: it allows for the conditional execution of one of two
+% branches but retains the choice of which branch was executed. This mechanism can
+% be used twice to implement the Toffoli gate, which performs a conditional
+% negation on one bit if two other bits are both true:
+
+% {\footnotesize
+% \begin{verbatim}
+% inot :: Bool :<=> Bool
+% inot = CommutePlus
+
+% inot2 :: Bool :<=> Bool
+% inot2 = UnitTimesI
+%     :.: CommuteTimes
+%     :.: (CommutePlus :*: Id)
+%     :.: CommuteTimes
+%     :.: UnitTimesE
+
+% cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
+% cond f g = Distribute
+%        :.: ((Id :*: f) :+: (Id :*: g))
+%        :.: Factor
+
+% controlled :: (a :<=> a) -> ((Bool, a) :<=> (Bool, a))
+% controlled f = cond f Id
+
+% cnot :: (Bool, Bool) :<=> (Bool, Bool)
+% cnot = controlled inot
+
+% toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
+% toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
+% \end{verbatim}}
+
+% Although writing circuits using the raw syntax for combinators above is tedious,
+% the examples illustrate the programming language nature of this family of $\Pi$
+% languages. In other work, one can find a compiler from a conventional functional
+% language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
+% technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
+% Haskell-like surface language~\cite{theseus} which can ease writing
+% programs. All that reinforces the claim that this is a practical approach to
+% reversible programming.
+
+% The raw syntax for combinators has, however, a significant benefit: it allows
+% algebraic manipulation of the programs using a second level of combinators. As
+% an example it is possible to show the the two implementations of boolean
+% negation are equivalent as follows:
+
+% {\footnotesize
+% \begin{verbatim}
+% negOpt :: inot2 :<===> inot
+% negOpt = (Id :>: AssocLeftEquiv)
+%     :..: (Id :>: (SlideCommuteTimes :>: Id))
+%     :..: (Id :>: AssocRightEquiv)
+%     :..: (Id :>: (Id :>: AssocLeftEquiv))
+%     :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
+%     :..: (Id :>: (Id :>: IdElimL))
+%     :..: AssocLeftEquiv
+%     :..: (SlideUnitI :>: Id)
+%     :..: AssocRightEquiv
+%     :..: (Id :>: InvInvEquivId)
+%     :..: IdElimR
+% \end{verbatim}}
+% %
+% \noindent This program, which can be thought of as an optimization or a proof,
+% relates the two implementations of boolean negation via a series of elementary
+% optimizations of which the following are the most interesting:
+
+% {\footnotesize
+% \begin{verbatim}
+% AssocLeftEquiv    ::           (c1 :.: (c2 :.: c3)) :<===> ((c1 :.: c2) :.: c3)
+% SlideCommuteTimes :: (CommuteTimes :.: (c1 :*: c2)) :<===> ((c2 :*: c1) :.: CommuteTimes)
+% IdElimL           ::                     (Id :.: c) :<===> c
+% SlideUnitI        ::   (UnitTimesI :.: (c1 :*: c2)) :<===> (c2 :.: UnitTimesI)
+% \end{verbatim}}
+
+% \end{comment}
+
+\jacques{Example 2: like example 1, but this time at the next level (i.e. a
+program equivalence).  Same remark old.  Refer to Shulmann notes, and point out
+that such equational proofs at this level can actually be interpreted in appropriate
+categories as commutative diagrams.}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-06-27 upi.tex
done with sec 2 for now

@@
 \newcommand{\unitipr}{\texttt{unitipr}}
-\newcommand{\swap}{\texttt{swap}}
+\newcommand{\swap}{\textit{swap}}
 \newcommand{\swapp}{\texttt{swapp}}
@@
 \newcommand{\factorl}{\texttt{factorl}}
-\newcommand{\id}{\texttt{id}}
+\newcommand{\id}{\textit{id}}
 \newcommand{\compc}[2]{#1 \circ #2}
@@
 language below is the restriction of $\Pi$ to the case of just one
-type, the type of booleans.
-
-\[\def\arraystretch{0.8}\begin{array}{rcl}
-\tau &::=& \bt \\
+type, the type of booleans:
+
+\[\def\arraystretch{0.8}\begin{array}{l@{\quad}rclrl}
+(\textit{Types}) & \tau &::=& \bt \\
 \\
-v &::=& \begin{array}[t]{lrcl}
-                    & \fc &:& \bt \\
-              \alt & \tc &:& \bt
-               \end{array} \\
+(\textit{Terms}) &  v &::=& \fc &:& \bt \\
+              && \alt & \tc &:& \bt \\
 \\
-c &::=& \begin{array}[t]{llcl}
-              & \id &:& \tau \iso \tau \\
-               \alt & \swap &:& \bt \iso \bt \\
-               \alt & ! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
-               \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)
-               \end{array} \\
+ (\textit{1-combinators}) &  c &::=& \id &:& \tau \iso \tau \\
+               && \alt & \swap &:& \bt \iso \bt \\
+               && \alt & ! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
+               && \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)  \\
 \\
-\alpha &::=& \begin{array}[t]{lrcl}
-               & \id &:& c \isotwo c \\
-            \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
-            \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
-            \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
-            \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
-            \alt & \rho &:& \swap \circ \swap \isotwo \id \\
-            \alt & \assocc &:& \compc{(\compc{c_1}{c_2})}{c_3} \isotwo \compc{c_1}{(\compc{c_2}{c_3})} \\
-            \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
-            \alt & !! &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_1) \\
-            \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3) \to (c_1 \isotwo c_3)
-             \end{array}
+(\textit{2-combinators}) & \alpha &::=& \id &:& c \isotwo c \\
+            && \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
+            && \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
+            && \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
+            && \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
+            && \alt & \rho &:& \swap \circ \swap \isotwo \id \\
+            && \alt & \assocc &:& 
+                                  \compc{(\compc{c_1}{c_2})}{c_3} \isotwo \compc{c_1}{(\compc{c_2}{c_3})} \\
+            && \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to 
+                             (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
+            && \alt & !! &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_1) \\
+            && \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3) \to (c_1 \isotwo c_3)
 \end{array}\]
 
-Canonical forms: given a level 1 program p, the level 2 combinators are
-complete; they can transform p to either id or not:
-\begin{verbatim}
-cmpl2-lem : (p : B <->1 B) -> (p <->2 id) + (p <->2 not)
-\end{verbatim}
+The syntactic category $c$ is that of 1-combinators denoting
+reversible programs, type isomorphisms, permutations, or equivalences
+depending on one's favorite interpretation. There are two primitive
+combinators $\id$ and $\swap$ which are closed under inverses $!$ and
+sequential composition $\circ$. The syntactic category $\alpha$ is
+that of 2-combinators denoting reversible program transfomations,
+coherence conditions on type isomorphisms, equivalences between
+permutations, or program optimizations depending on one's favorite
+interpretation.
+
+It is relatively easy to think of a few sound program transformations
+to include in the category $\alpha$ of 2-combinators. The set above
+however is \emph{complete}.
+
+\begin{lemma}[Canonical Forms]
+  Given a 1-combinator $c : \tau \leftrightarrow \tau$, we either have
+  a normalization $c \Leftrightarrow \id$ or a normalization
+  $c \Leftrightarrow \swap$.
+\end{lemma}
 

### 2017-06-27 upi.tex
sec 3 in progress

@@
 sequential composition $\circ$. The syntactic category $\alpha$ is
-that of 2-combinators denoting reversible program transfomations,
+that of 2-combinators denoting reversible program transformations,
 coherence conditions on type isomorphisms, equivalences between
@@
 \label{sec:univalent}
+
+We now switch gear and focus on constructions of \emph{univalent
+  universes}. In order to state the definition of such universes we
+need to introduce two notions: propositional equality
+$\AgdaDatatype{≡}$ and equivalence $\AgdaDatatype{≃}$.
+
+Propositional equality, or identity types in intensional dependent
+type theory, or path types under the HoTT-interpretation come equipped
+with a single constructor \AgdaInductiveConstructor{refl} and an
+induction principle $\AgdaFunction{J}$. A peculiar aspect of
+$\AgdaFunction{J}$ is that it does \emph{not} enforce that every
+element of an identity type
+$\AgdaFunction{x}~\AgdaDatatype{≡}~\AgdaFunction{y}$ is
+\AgdaInductiveConstructor{refl}. This ``weakness'' has a beautiful
+consequence: instead of identifying all proofs and collapsing the
+structure of proofs in the process, it instead permits a rich
+combinatorial structure of proofs to emerge. The second notion we need
+is that of equivalence. We say that two types or spaces $A$ and $B$
+are equivalent if there exist functions $f : A \rightarrow B$ and
+$g : B \rightarrow A$ that compose to the identity function. A notable
+point is that equivalences do not come with an induction principle. We
+say a universe $\mathcal{U}$ is univalent if its space of identities
+$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of
+equivalences $\AgdaFunction{A} ≃ \AgdaFunction{B}$.
+
+Instead of talking about univalent universes, we generalize the
+discussion to univalent type families which will allow the
+constructions of custom universes.
+
+\newpage
 
@@
 Thm: Let P : Type -> Type such that for all X : Type, P(X)
-is a proposition. Then Σ(X,P) is the base of a univalent
+is a proposition. Then Sigma(X,P) is the base of a univalent
 fibration.
@@
 is-prop : Type -> Type has the property that is-prop(X)
-is a proposition for all types X. So Σ(X , is-prop) is
+is a proposition for all types X. So Sigma(X , is-prop) is
 the base of a univalent fibration. In other words, this
@@
 is-set : Type -> Type has the property that is-set(X)
-is a proposition for all types X. So Σ(X , is-set) is
+is a proposition for all types X. So Sigma(X , is-set) is
 the base of a univalent fibration. In other words, this
@@
 is-1-type(X) is a proposition for all types X. So
-Σ(X , is-1-type) is the base of a univalent fibration.
+Sigma(X , is-1-type) is the base of a univalent fibration.
 In other words, this is the subuniverse consisting of all

### 2017-06-30 upi.tex
expand on first part of section 2.1

@@
 $\AgdaFunction{f}$ to the second argument. In either case, the
-incoming boolean is reconstituted to maintain reversibility. Using the
+incoming boolean is reconstituted to maintain reversibility.
+
+{\small
+\[\def\arraystretch{1.2}\begin{array}{rcll}
+\AgdaFunction{controlled}  &:& \forall a.~ (a \leftrightarrow a) \rightarrow 
+                            & (\bt \otimes a \leftrightarrow \bt \otimes a) \\
+\AgdaFunction{controlled}~\AgdaFunction{f} &=&
+
+  \bt \otimes a 
+    & \byiso{\AgdaFunction{unfoldBool} \otimes \AgdaFunction{id}} \\
+&& (\ot \oplus \ot) \otimes a 
+    & \byiso{\AgdaFunction{distribute}} \\
+&& (\ot \otimes a) \oplus (\ot \otimes a) 
+    & \byiso{\AgdaFunction{id} \oplus (\AgdaFunction{id} \otimes \AgdaFunction{f})} \\
+&& (\ot \otimes a) \oplus (\ot \otimes a)
+    & \byiso{\AgdaFunction{factor}} \\
+&& (\ot \oplus \ot) \otimes a
+    & \byiso{\AgdaFunction{foldBool} \otimes \AgdaFunction{id}} \\
+&& \bt \otimes a & ~ \\
+\end{array}
+\]}
+
+The $\AgdaFunction{not}$ function is a simple lifting of
+\AgdaFunction{swap₊}. Using the
 \AgdaFunction{controlled} building block, we can build a
@@
 for combinational boolean circuits thus showing the expressiveness of
-the language:
+the language.
 
@@
 {\small
-\[\def\arraystretch{1.2}\begin{array}{rcl}
-\AgdaFunction{controlled} &:& \forall a.~ (a \leftrightarrow a) \rightarrow 
-                              (\bt \otimes a \leftrightarrow \bt \otimes a) \\
-\AgdaFunction{controlled}~\AgdaFunction{f} &=& 
-      \bt \otimes a \\
-&& \qquad\byiso{\AgdaFunction{unfoldBool} \otimes \AgdaFunction{id}} \\
-&& (\ot \oplus \ot) \otimes a \\
-&& \qquad\byiso{\AgdaFunction{distribute}} \\
-&& (\ot \otimes a) \oplus (\ot \otimes a) \\
-&& \qquad\byiso{\AgdaFunction{id} \oplus (\AgdaFunction{id} \otimes \AgdaFunction{f})} \\
-&& (\ot \otimes a) \oplus (\ot \otimes a) \\
-&& \qquad\byiso{\AgdaFunction{factor}} \\
-&& (\ot \oplus \ot) \otimes a \\
-&& \qquad\byiso{\AgdaFunction{foldBool} \otimes \AgdaFunction{id}} \\
-&& \bt \otimes a \\
-\\ 
+\[\begin{array}{rcl}
 \AgdaFunction{not} &:& \bt \leftrightarrow \bt \\
@@
 %%%
+
+While we wrote \AgdaFunction{controlled} in equation-reasoning style,
+\AgdaFunction{not} is written in combinator style.  These are
+equivalent as $\byiso{-}$ is defined in terms of $\odot$.
 

### 2017-06-30 upi.tex
second derivation in 2.1

@@
 the language.
-
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.9}]
-  \draw (1,2) ellipse (0.5cm and 0.5cm);
-  \draw[fill] (1,2) circle [radius=0.025];
-  \node[below] at (1,2) {*};
-
-  \draw (0,0) ellipse (0.5cm and 1cm);
-  \draw[fill] (0,0.5) circle [radius=0.025];
-  \node[below] at (0,0.5) {F};
-  \draw[fill] (0,-0.5) circle [radius=0.025];
-  \node[below] at (0,-0.5) {T};
-
-  \draw     (1,2)    -- (2,2)      ; %% ()
-  \draw     (0,0.5)  -- (2,0.5)    ; %% F
-  \draw     (0,-0.5) -- (2,-0.5)   ; %% T
-
-  \draw     (2,2)    -- (3,-0.5)   ;
-  \draw     (2,0.5)  -- (3,2)      ;
-  \draw     (2,-0.5) -- (3,1)      ;
-
-  \draw     (3,2)    -- (3.5,2)    ;
-  \draw     (3,1)    -- (3.5,1)    ;
-  \draw     (3,-0.5) -- (3.5,-0.5) ;
-
-  \draw     (3.5,2)    -- (4.5,1)    ;
-  \draw     (3.5,1)    -- (4.5,2)    ;
-  \draw     (3.5,-0.5) -- (4.5,-0.5) ;
-
-  \draw     (4.5,2)    -- (5,2)    ;
-  \draw     (4.5,1)    -- (5,1)    ;
-  \draw     (4.5,-0.5) -- (5,-0.5) ;
-
-  \draw     (5,2)    -- (6,0.5)  ;
-  \draw     (5,1)    -- (6,-0.5) ;
-  \draw     (5,-0.5) -- (6,2)    ;
-
-  \draw     (6,2)    -- (7,2)    ;
-  \draw     (6,0.5)  -- (8,0.5)  ;
-  \draw     (6,-0.5) -- (8,-0.5) ;
-
-  \draw (7,2) ellipse (0.5cm and 0.5cm);
-  \draw[fill] (7,2) circle [radius=0.025];
-  \node[below] at (7,2) {*};
-
-  \draw (8,0) ellipse (0.5cm and 1cm);
-  \draw[fill] (8,0.5) circle [radius=0.025];
-  \node[below] at (8,0.5) {F};
-  \draw[fill] (8,-0.5) circle [radius=0.025];
-  \node[below] at (8,-0.5) {T};
-
-\end{tikzpicture}
-\end{center}
-\caption{\label{fig:not}Graphical representation of \AgdaFunction{not₃}}
-\end{figure}
-
 {\small
@@
 raises the question of when two programs are ``equivalent.'' Consider
-the six programs of type $\bt \leftrightarrow \bt$ below:
+the follwing six programs of type~$\bt \leftrightarrow \bt$
 
@@
 
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[scale=0.9,every node/.style={scale=0.9}]
+  \draw (1,2) ellipse (0.5cm and 0.5cm);
+  \draw[fill] (1,2) circle [radius=0.025];
+  \node[below] at (1,2) {*};
+
+  \draw (0,0) ellipse (0.5cm and 1cm);
+  \draw[fill] (0,0.5) circle [radius=0.025];
+  \node[below] at (0,0.5) {F};
+  \draw[fill] (0,-0.5) circle [radius=0.025];
+  \node[below] at (0,-0.5) {T};
+
+  \draw     (1,2)    -- (2,2)      ; %% ()
+  \draw     (0,0.5)  -- (2,0.5)    ; %% F
+  \draw     (0,-0.5) -- (2,-0.5)   ; %% T
+
+  \draw     (2,2)    -- (3,-0.5)   ;
+  \draw     (2,0.5)  -- (3,2)      ;
+  \draw     (2,-0.5) -- (3,1)      ;
+
+  \draw     (3,2)    -- (3.5,2)    ;
+  \draw     (3,1)    -- (3.5,1)    ;
+  \draw     (3,-0.5) -- (3.5,-0.5) ;
+
+  \draw     (3.5,2)    -- (4.5,1)    ;
+  \draw     (3.5,1)    -- (4.5,2)    ;
+  \draw     (3.5,-0.5) -- (4.5,-0.5) ;
+
+  \draw     (4.5,2)    -- (5,2)    ;
+  \draw     (4.5,1)    -- (5,1)    ;
+  \draw     (4.5,-0.5) -- (5,-0.5) ;
+
+  \draw     (5,2)    -- (6,0.5)  ;
+  \draw     (5,1)    -- (6,-0.5) ;
+  \draw     (5,-0.5) -- (6,2)    ;
+
+  \draw     (6,2)    -- (7,2)    ;
+  \draw     (6,0.5)  -- (8,0.5)  ;
+  \draw     (6,-0.5) -- (8,-0.5) ;
+
+  \draw (7,2) ellipse (0.5cm and 0.5cm);
+  \draw[fill] (7,2) circle [radius=0.025];
+  \node[below] at (7,2) {*};
+
+  \draw (8,0) ellipse (0.5cm and 1cm);
+  \draw[fill] (8,0.5) circle [radius=0.025];
+  \node[below] at (8,0.5) {F};
+  \draw[fill] (8,-0.5) circle [radius=0.025];
+  \node[below] at (8,-0.5) {T};
+
+\end{tikzpicture}
+\end{center}
+\caption{\label{fig:not}Graphical representation of \AgdaFunction{not₃}}
+\end{figure}
+
 \noindent We neither expect, nor desire, that all these programs of
-the same type to be be equivalent. In the case above, each group of
-three programs does represent ``the same'' program but the two groups
+the same type to be be equivalent. In the above, each group of
+three does represent ``the same'' program but the two groups
 represent distinct programs. Thinking extensionally, i.e., by looking
 at all possible input-output pairs, it is indeed easy to verify this
-fact but we can do better. We can provide \emph{evidence} (i.e., a
+fact, but we can do better. We can provide \emph{evidence} (i.e., a
 program that manipulates programs) that can constructively transform
@@
 {\small 
-\[\def\arraystretch{1.2}\begin{array}{rcl}
+\[\def\arraystretch{1.2}\begin{array}{rcll}
 \AgdaFunction{notOpt} &:& \AgdaFunction{not₃} \Leftrightarrow \AgdaFunction{not} \\
@@
                         ((\AgdaFunction{not} \otimes \AgdaFunction{id}) \odot 
-                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}))) \\
-&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocLeft}} \\
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆})))
+ & \quad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocLeft}} \\
 && \AgdaFunction{uniti⋆} \odot (\AgdaFunction{swap⋆} \odot 
                         (\AgdaFunction{not} \otimes \AgdaFunction{id})) \odot 
-                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}) \\
-&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{swapLeft} 
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆})
+ & \quad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{swapLeft} 
                                   \boxdot \AgdaFunction{id})} \\
@@
                       \odot \AgdaFunction{swap⋆}) \odot 
-                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}) \\
-&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocRight}} \\
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆})
+ & \quad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{assocRight}} \\
 && \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
                       \odot (\AgdaFunction{swap⋆} \odot 
-                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆}))) \\
-&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
+                        (\AgdaFunction{swap⋆} \odot \AgdaFunction{unite⋆})))
+ & \quad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                   \boxdot \AgdaFunction{assocLeft})} \\
@@
                       \odot ((\AgdaFunction{swap⋆} \odot 
-                      \AgdaFunction{swap⋆}) \odot \AgdaFunction{unite⋆})) \\
-&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
+                      \AgdaFunction{swap⋆}) \odot \AgdaFunction{unite⋆}))
+ & \quad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                   \boxdot (\AgdaFunction{leftInv} \boxdot \AgdaFunction{id}))} \\
 && \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
-                      \odot (\AgdaFunction{id} \odot \AgdaFunction{unite⋆})) \\
-&& \qquad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
+                      \odot (\AgdaFunction{id} \odot \AgdaFunction{unite⋆}))
+ & \quad\byisotwo{\AgdaFunction{id} \boxdot (\AgdaFunction{id}
                                   \boxdot \AgdaFunction{idLeft})} \\
 && \AgdaFunction{uniti⋆} \odot ((\AgdaFunction{id} \otimes \AgdaFunction{not})
-                      \odot \AgdaFunction{unite⋆}) \\
-&& \qquad\byisotwo{\AgdaFunction{assocLeft}} \\
+                      \odot \AgdaFunction{unite⋆})
+ & \quad\byisotwo{\AgdaFunction{assocLeft}} \\
 && (\AgdaFunction{uniti⋆} \odot (\AgdaFunction{id} \otimes \AgdaFunction{not})) 
-                      \odot \AgdaFunction{unite⋆} \\
-&& \qquad\byisotwo{\AgdaFunction{unitiLeft} \boxdot \AgdaFunction{id}} \\
-&& (\AgdaFunction{not} \otimes \AgdaFunction{uniti⋆}) \odot \AgdaFunction{unite⋆} \\    
-&& \qquad\byisotwo{\AgdaFunction{assocRight}} \\
-&& \AgdaFunction{not} \otimes (\AgdaFunction{uniti⋆} \odot \AgdaFunction{unite⋆}) \\    
-&& \qquad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{leftInv}} \\
-&& \AgdaFunction{not} \otimes \AgdaFunction{id} \\
-&& \qquad\byisotwo{\AgdaFunction{idRight}} \\
+                      \odot \AgdaFunction{unite⋆}
+ & \quad\byisotwo{\AgdaFunction{unitiLeft} \boxdot \AgdaFunction{id}} \\
+&& (\AgdaFunction{not} \otimes \AgdaFunction{uniti⋆}) \odot \AgdaFunction{unite⋆}
+ & \quad\byisotwo{\AgdaFunction{assocRight}} \\
+&& \AgdaFunction{not} \otimes (\AgdaFunction{uniti⋆} \odot \AgdaFunction{unite⋆})
+ & \quad\byisotwo{\AgdaFunction{id} \boxdot \AgdaFunction{leftInv}} \\
+&& \AgdaFunction{not} \otimes \AgdaFunction{id}
+ & \quad\byisotwo{\AgdaFunction{idRight}} \\
 && \AgdaFunction{not} 

### 2017-06-30 upi.tex
add a note regarding the categorical link.
minor tweaks to abstract and introduction.

@@
   correspondence relates combinators witnessing type isomorphisms in the
-  programming language to paths in the univalent universe and combinator
+  programming language to paths in the univalent universe; and combinator
   optimizations in the programming language to 2-paths in the univalent
@@
 
-The proceedings of the 2012 Symposium on Principles of Programming
+The proceedings of the $2012$ Symposium on Principles of Programming
 Languages~\cite{Field:2012:2103656} included two apparently unrelated papers:
@@
 
+It is worthwhile mentioning that the above derivation could also be drawn as 
+one (large!) commutative diagram in an appropriate category, with each
+$\byisotwo{-}$ as a $2$-arrow (and representing a natural isomorphism).
+See~\cite{Shulmann} for that interpretation.
+
 %%%%%

### 2017-06-30 upi.tex
add a sentence about the canonical form being constructive.  Add an edcomm about the inconsistent notation.

@@
 which we will use in the formalization in the rest of the paper. The
-language below is the restriction of $\Pi$ to the case of just one
+language is the restriction of $\Pi$ to the case of just one
 type, the type of booleans:
 
+\jacques{the code above uses $\odot$ for 1-composition, 
+$\boxdot$ for parallel 2-composition of $\odot$, while the
+code below uses $\circ$ and $\odot$ respectively, which is 
+quite confusing.  We should pick one notation.}
 \[\def\arraystretch{0.8}\begin{array}{l@{\quad}rclrl}
@@
 to include in the category $\alpha$ of 2-combinators. The set above
-however is \emph{complete}.
+however is \emph{complete}. We also have canonical forms (constructively).
 

### 2017-06-30 upi.tex
work on intro to section 3

@@
 
-We now switch gear and focus on constructions of \emph{univalent
+We now switch gear and focus on the construction of \emph{univalent
   universes}. In order to state the definition of such universes we
@@
 Propositional equality, or identity types in intensional dependent
-type theory, or path types under the HoTT-interpretation come equipped
+type theory, or path types under the HoTT-interpretation, come equipped
 with a single constructor \AgdaInductiveConstructor{refl} and an
@@
 structure of proofs in the process, it instead permits a rich
-combinatorial structure of proofs to emerge. The second notion we need
+combinatorial structure to emerge.
+
+The second notion we need
 is that of equivalence. We say that two types or spaces $A$ and $B$
 are equivalent if there exist functions $f : A \rightarrow B$ and
-$g : B \rightarrow A$ that compose to the identity function. A notable
-point is that equivalences do not come with an induction principle. We
-say a universe $\mathcal{U}$ is univalent if its space of identities
+$g : B \rightarrow A$ that compose to the identity function%
+\footnote{Strictly speaking, this is a \emph{quasi-equivalence}, an
+equivalence also requires a certain coherence between the compositions.}.
+A notable point is that equivalences do not come with an induction principle.
+We say a universe $\mathcal{U}$ is univalent if its space of identities
 $\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of
@@
 
-Instead of talking about univalent universes, we generalize the
-discussion to univalent type families which will allow the
-constructions of custom universes.
+\jacques{We should probably be precise with our definition of 
+equivalence.}
+
+While univalent universes came first, it was later realised that
+the more general univalent type families are more convenient to
+work with, and allow the constructions of custom universes. 
+We thus need to build up some definitions for what this means.
+Note that below we will use ``point'' and ``member of a type'',
+``space'' and ``type'', ``path'' and ``equivalence'' as
+synonyms.  Furthermore our setting is an intuinistic type
+theory (with no classical axioms such a choice or excluded
+middle added); this implies that all mappings are
+\emph{continuous}.
+
+It is very important to remember that while $A$ and $B$ are spaces,
+so are $\AgdaFunction{A} ≡ \AgdaFunction{B}$ and
+$\AgdaFunction{A} ≃ \AgdaFunction{B}$.  Thus whatever meaningful
+questions we can ask of $A$, we can also ask about
+$\AgdaFunction{A} ≡ \AgdaFunction{B}$ and
+$\AgdaFunction{A} ≃ \AgdaFunction{B}$.
 
@@
 
-We will consider a general statement of \emph{univalence} for type families. We
-start with a some preliminaries. Fig.~\ref{fig:fib} depicts a type family
-$P : A \rightarrow \mathcal{U}$ mapping each point $x : A$ to a space $P(x)$. As
-everything is continuous, we expect this type family to respect identities. In
-other words if $x:A$ and $y:A$ are identified using a path $x \equiv y$, it must
-be the case that the spaces $P(x)$ and $P(y)$ become equivalent
-$P(x) \simeq P(y)$ as a consequence. Generally speaking, there is no reason to
+Fig.~\ref{fig:fib} depicts a type family $P : A \rightarrow \mathcal{U}$
+mapping each member $x : A$ to a type $P(x)$. As everything is continuous,
+we expect this type family to
+respect identities. In other words if $x:A$ and $y:A$ are identified using a
+path $x \equiv y$, it must
+be the case that the spaces $P(x)$ and $P(y)$ are equivalent
+$P(x) \simeq P(y)$. Generally speaking, there is no reason to
 expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ to be themselves

### 2017-06-30 upi.tex
move the wavefront in section 3

@@
 
-\newpage
-
-Fig.~\ref{fig:fib} depicts a type family $P : A \rightarrow \mathcal{U}$
+Fig.~\ref{fig:fib} on the left depicts a type family $P : A \rightarrow \mathcal{U}$
 mapping each member $x : A$ to a type $P(x)$. As everything is continuous,
-we expect this type family to
-respect identities. In other words if $x:A$ and $y:A$ are identified using a
-path $x \equiv y$, it must
+we expect this type family to respect identities. In other words if $x:A$ and
+$y:A$ are identified using a path $x \equiv y$ (depicted on the right), it must
 be the case that the spaces $P(x)$ and $P(y)$ are equivalent
-$P(x) \simeq P(y)$. Generally speaking, there is no reason to
-expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ to be themselves
-equivalent for all $x$ and $y$. And indeed they are most frequently not
-equivalent. In the rare cases in which they are equivalent, the fibration is a
-\emph{univalent fibration}.
+$P(x) \simeq P(y)$. Although we depict the path ``in'' the Base Space $A$
+(and that is indeed the case in classical homotopy theory), it is best to
+think of the evidence given by $\equiv$ as being outside $A$ (but in $\mathcal{U}$).
+
+Generally speaking, there is no reason to
+expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ of
+``evidence'' that identify $x$ and $y$, and $P(x)$ and $P(y)$ respectively,
+are themselves related, never mind equivalent, for all or even any $x$ and $y$.
+In the rare cases in which these ``spaces of evidence'' are equivalent, such a
+fibration is a \emph{univalent fibration}.
 
@@
 \caption{\label{fig:fib}(left) Type family $P : A \rightarrow \mathcal{U}$ as a
-  fibration with total space $\Sigma_{(x:A)} P(x)$; (right) a path $x \equiv y$
+  fibration with total space $\Sigma_{(x:A)} P(x)$;\newline
+ (right) a path $x \equiv y$
   in the base space induces an equivalence between the spaces $P(x)$ and $P(y)$}
 \end{figure}
+
+\newpage
 

### 2017-06-30 upi.tex
move wavefront deeper in section 3.  Split the two non-examples (easier to explain).  Add more words.

@@
 
-\newpage
-
-As two simple examples of non-univalent fibrations consider
-$P : \mathbb{1} \rightarrow \mathbb{U}$ defined as $P = \lambda\_.\mathbb{2}$
-and $Q : \mathbb{2} \rightarrow \mathbb{U}$ defined as
-$Q = \lambda \_. \mathbb{1}$ illustrated below:
-
-\begin{tabular}{c@{\qquad\qquad}c}
+As is often the case when confronted with new definitions, it is
+instructive to disect \emph{non-examples}, as a means to understand
+what ``goes wrong''.  As a first example of a non-univalent fibration,
+consider $P : \mathbb{1} \rightarrow \mathbb{U}$ defined as $P =
+\lambda\_.\mathbb{2}$.
+
 \begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
@@
   \node[below] at (0,-5) {$\star$};
-  \draw (0,-2.7) ellipse (0.5cm and 1cm);
-  \node[right] at (0.5,-2.7) {Fiber $P(\star) = \mathbb{2}$};
-  \draw[fill] (-0.2,-2.7) circle [radius=0.025];
-  \draw[fill] (0.2,-2.7) circle [radius=0.025];
+  \draw (-1.3,-2.7) ellipse (0.5cm and 1cm);
+  \draw[fill] (-1.5,-2.7) circle [radius=0.025];
+  \draw[fill] (-1.1,-2.7) circle [radius=0.025];
+  \node[left] at (-1.8,-2.7) {Fiber $P(\star) = \mathbb{2}$};
+  \draw (1.3,-2.7) ellipse (0.5cm and 1cm);
+  \draw[fill] (1.1,-2.7) circle [radius=0.025];
+  \draw[fill] (1.5,-2.7) circle [radius=0.025];
+  \node[right] at (1.8,-2.7) {Fiber $P(\star) = \mathbb{2}$};
+  \node[below,cyan,dashed,thick] at (0,-5) {$\equiv$};
+  \draw[->,red,dashed,ultra thick] (-0.8,-2.3) to [out=45, in=135] (0.8,-2.3);
+  \draw[->,red,dashed,ultra thick] (0.8,-3.1) to [out=-135, in=-45] (-0.8,-3.1);
+  \node[red,ultra thick] at (0,-2.7) {$\simeq$};
 \end{tikzpicture}
-&
+
+\medskip\noindent There is actually \emph{a single fiber}, but
+the identity function and boolean negation induce \emph{two distinct
+equivalences} of the space $P(\star) = \mathbb{2}$ with itself.
+In other words, $\mathbb{2} \simeq \mathbb{2}$ has two distinct inhabitants.
+As there is only one path $\star \equiv \star$ in the base space, the
+fibration $P$ is not univalent.
+
+Another example is given by $Q : \mathbb{2} \rightarrow \mathbb{U}$ defined as
+$Q = \lambda \_. \mathbb{1}$ illustrated below
+
 \begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
@@
 \end{tikzpicture}
-\end{tabular}
-
-\medskip\noindent The identity function and boolean negation induce two distinct
-equivalences in the space $P(\star) = \mathbb{2} \simeq \mathbb{2} =
-P(\star)$. As there is only one path $\star \equiv \star$ in the base space, the
-fibration $P$ is not univalent. Similarly, there is one equivalence
-$Q(\fc) = \mathbb{1} \simeq \mathbb{1} = Q(\tc)$ but this certainly does not
-induce a path $\fc \equiv \tc$. As the examples illustrate there is a delicate
+
+The situation here is different.  There is one equivalence
+$Q(\fc) = \mathbb{1} \simeq \mathbb{1} = Q(\tc)$ but it does not come
+from a path (such as $\fc \equiv \tc$, which does not exist).  And that
+equivalence does not ``induce'' such a path either.
+
+As the examples illustrate there is a delicate
 balance required for the fibration to be univalent; otherwise some distinct
@@
 space would be identified (like for $Q$).
+
+\newpage
 

### 2017-06-30 upi.tex
move wavefront to end of section 3 introction by fleshing out the vacuous example, and add the example with only a single inhabitant.

@@
 
-\newpage
-
-An example of a univalent fibration is $P : \mathbb{2} \rightarrow \mathbb{U}$
-defined as $P(\fc) = \mathbb{0}$ and $P(\tc) = \mathbb{1}$ illustrated below:
+On the other hand, for $P : \mathbb{2} \rightarrow \mathbb{U}$
+defined as $P(\fc) = \mathbb{0}$ and $P(\tc) = \mathbb{1}$, we do
+obtain a univalent fibration.
 
@@
 \medskip\noindent In this case, the space of equivalences
-$P(\fc) = \mathbb{0} \simeq \mathbb{1} = P(\tc)$ is empty and so is the space of
-paths $\fc \equiv \tc$ and hence these two spaces are vacuously equivalent.
+$\mathbb{0} \simeq \mathbb{1}$ is empty, as is the space of
+paths $\fc \equiv \tc$. These two are thus vacuously equivalent.
+
+Of course, we can obtain another (trivial) example from $\mathbb{1}$:
+$P : \mathbb{1} \rightarrow \mathbb{U}$, defined as
+$P(\star) = \mathbb{1}$. $P$ is univalent as both $\star \equiv \star$ and
+$\mathbb{1} \simeq \mathbb{1}$ each have a single inhabitant.
+
+\newpage
 

### 2017-06-30 upi.tex
remove some old material which was entirely subsumed and no longer served any purpose.

@@
 
-%% \jacques{Example 1: using the combinators from pi-dual, some an equational
-%% derivation in type theory (which?), in equational-derivation style.  Point out
-%% that reading ``down'' the justifications is a $\Pi$ program which is also a
-%% constructive proof of the equivalence of the types.  As it is reversible, we
-%% can also read the reverse program by reading ``up''.}
-
-%% \jacques{Refer to some Sabry et al. papers regarding Toffoli gates.}
-
-% \begin{comment}
-% In this paper we will
-% identify a ``featherweight'' representative to use for the formal
-% development. But first, we illustrate the flavor of the family of languages with
-% some examples written in a Haskell syntax. First consider the following program:
-
-% {\footnotesize
-% \begin{verbatim}
-% addSub1 :: (Int, Int) :<=> (Int, Int)
-% addSub1 = CommuteTimes
-%       :.: (UnfoldN :*: Id)
-%       :.: Distribute
-%       :.: (Id :+: CommuteTimes)
-%       :.: Factor
-%       :.: (FoldN :*: Id)
-% \end{verbatim}}
-% %
-% \noindent which takes a pair of natural numbers and processes them as follows depending on
-% whether the second number is 0 or not. If the second number is 0, the evaluation
-% proceeds as follows:
-% $(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
-% \mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
-% second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
-% $(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
-% \mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
-% (\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
-% applications of this program would increment the first number and decrement the
-% second number maintaining the sum at all times. When the second number reaches
-% 0, the numbers are swapped.
-
-% As another example, consider the following expressions culminating with the
-% implementation of the Toffoli gate, which is universal for combinational
-% circuits. We have two implementations of simple boolean negation: a direct one
-% and a more convoluted one. These two implementations are ``equivalent'' in the
-% sense to be explained below. The expression \verb|cond| is the main reversible
-% control flow mechanism: it allows for the conditional execution of one of two
-% branches but retains the choice of which branch was executed. This mechanism can
-% be used twice to implement the Toffoli gate, which performs a conditional
-% negation on one bit if two other bits are both true:
-
-% {\footnotesize
-% \begin{verbatim}
-% inot :: Bool :<=> Bool
-% inot = CommutePlus
-
-% inot2 :: Bool :<=> Bool
-% inot2 = UnitTimesI
-%     :.: CommuteTimes
-%     :.: (CommutePlus :*: Id)
-%     :.: CommuteTimes
-%     :.: UnitTimesE
-
-% cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
-% cond f g = Distribute
-%        :.: ((Id :*: f) :+: (Id :*: g))
-%        :.: Factor
-
-% controlled :: (a :<=> a) -> ((Bool, a) :<=> (Bool, a))
-% controlled f = cond f Id
-
-% cnot :: (Bool, Bool) :<=> (Bool, Bool)
-% cnot = controlled inot
-
-% toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
-% toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
-% \end{verbatim}}
-
-% Although writing circuits using the raw syntax for combinators above is tedious,
-% the examples illustrate the programming language nature of this family of $\Pi$
-% languages. In other work, one can find a compiler from a conventional functional
-% language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
-% technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
-% Haskell-like surface language~\cite{theseus} which can ease writing
-% programs. All that reinforces the claim that this is a practical approach to
-% reversible programming.
-
-% The raw syntax for combinators has, however, a significant benefit: it allows
-% algebraic manipulation of the programs using a second level of combinators. As
-% an example it is possible to show the the two implementations of boolean
-% negation are equivalent as follows:
-
-% {\footnotesize
-% \begin{verbatim}
-% negOpt :: inot2 :<===> inot
-% negOpt = (Id :>: AssocLeftEquiv)
-%     :..: (Id :>: (SlideCommuteTimes :>: Id))
-%     :..: (Id :>: AssocRightEquiv)
-%     :..: (Id :>: (Id :>: AssocLeftEquiv))
-%     :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
-%     :..: (Id :>: (Id :>: IdElimL))
-%     :..: AssocLeftEquiv
-%     :..: (SlideUnitI :>: Id)
-%     :..: AssocRightEquiv
-%     :..: (Id :>: InvInvEquivId)
-%     :..: IdElimR
-% \end{verbatim}}
-% %
-% \noindent This program, which can be thought of as an optimization or a proof,
-% relates the two implementations of boolean negation via a series of elementary
-% optimizations of which the following are the most interesting:
-
-% {\footnotesize
-% \begin{verbatim}
-% AssocLeftEquiv    ::           (c1 :.: (c2 :.: c3)) :<===> ((c1 :.: c2) :.: c3)
-% SlideCommuteTimes :: (CommuteTimes :.: (c1 :*: c2)) :<===> ((c2 :*: c1) :.: CommuteTimes)
-% IdElimL           ::                     (Id :.: c) :<===> c
-% SlideUnitI        ::   (UnitTimesI :.: (c1 :*: c2)) :<===> (c2 :.: UnitTimesI)
-% \end{verbatim}}
-
-% \end{comment}
-
-\jacques{Example 2: like example 1, but this time at the next level (i.e. a
-program equivalence).  Same remark old.  Refer to Shulmann notes, and point out
-that such equational proofs at this level can actually be interpreted in appropriate
-categories as commutative diagrams.}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newpage
-~
-\newpage
 \section{Old}
-
-The first step was a refinement and extension of the
-language $\Pi$ with a layer of program transformations between the programs
-witnessing type isomorphisms~\cite{Carette2016}. The second crucial result was
-the formal presention
-
-Second,
-
-POPL 2012: Information effects; Canonicity for 2-dimensional type theory; look
-different but we show are the ``same''.
-
-There is a cottage industry of reversible programming languages, reversible
-logic, programming applications of type isomorphisms, etc. It seems that this
-work should be connected to HoTT and univalence, but there aren't any precise
-connections or theorems.
-
-We propose a very specialized version of HoTT, which we dub
-``featherweight HoTT''; it sports a smaller
-collection of types and only reversible functions. No dependent types!  More
-precisely, our conjecture:
-
-\begin{itemize}
-\item Let $\mathcal{U}$ be the univalent subuniverse generated by $0$, $1$, $+$,
-  and $*$ suitably truncated.
-\item Let $\Pi$ be the previously  described reversible programming with 1-paths and 2-paths
-\item We conjecture that $\Pi$ includes codes for \emph{all} paths in $\mathcal{U}$
-\end{itemize}
-
-\jacques{Just to make sure: we're aiming to prove a much much smaller version of
-this here, right?  Where in fact the only type is Bool?}
-
- Following the HoTT book and the slides on ``A Characterization of Univalent
-Fibrations'' by Dan Christensen. All the definitions and most of the examples in
-this section are formalized in the accompanying Agda code.
 

### 2017-06-30 upi.tex
start working on section 3.  Introduce some macros so that we are not so display driven.  Edit well beyond the wavefront, but 3.1 is mysterious to me, so work on 3.2.

@@
 
+% TODO: give this a better name!
+\newcommand{\bracket}[1]{\ensuremath{\{#1\}}}
+\newcommand{\ptrunc}[1]{\ensuremath{\| #1 \|}}
+
+\newcommand{\PiTwo}{\ensuremath{\Pi_{\mathbb{2}}}}
+
 \newcommand{\amr}[1]{\fbox{\begin{minipage}{0.8\textwidth}\color{red}{Amr says: {#1}}\end{minipage}}}
@@
 %%%%%
-\subsection{$\Pi_{\mathbb{2}}$}
+\subsection{\PiTwo}{\label{sec:pi2}}
 
@@
 
-\newpage
-
 %%%
@@
 
-Instead of guessing, there is a special class of univalent fibrations:
-\[\begin{array}{rcl}
-\{F\} &:& \Sigma_{A:\mathcal{U}} \| F \equiv A \| \rightarrow \mathcal{U} \\
-\{F\}(A,|p|) &=& A
+It turns out that there is a special class of fibrations.  For its
+definition, we need to first introduce the idea of
+\emph{propositional truncation}, and a special map from
+paths to equivalences.
+
+\begin{defn}
+Given a type $A : \mathcal{U}$, the type
+$\ptrunc{A}$ is the propositional truncation of $A$ defined as: 
+\begin{itemize}
+\item for $x : A$, we have $|x| : \ptrunc{A}$,
+\item for any $x,y : \ptrunc{A}$, we have $\gluep : x \equiv y$.
+\end{itemize}
+\end{defn}
+\noindent We can understand $\ptrunc{A}$ as a space which has
+$|a|$ for all $a : A$ as its points, but where all such points are
+(propositionally) identified via $\gluep$.  In other others,
+by continuity, $\ptrunc{A}$ behaves the same as $\mathbb{1}$,
+in that all inhabitants of $\ptrunc{A}$ give rise to the same
+behaviour in \emph{all} situations.
+
+\begin{defn}[Lemma 2.10.1 of~\cite{HoTT}]
+Given two types $A$ and $B$, there is a function
+$\mathit{idtoeqv} : (A \equiv B) \to (A \simeq B)$.
+\end{defn}
+
+For any type $F$, let us define $\bracket{F}$ via
+\[\begin{array}{rcllcl}
+\bracket{F} &:& \Sigma_{A:\mathcal{U}} & \ptrunc{F \equiv A} & \rightarrow & \mathcal{U} \\
+\bracket{F} & & (A, & |p|) &=& A
 \end{array}\]
 
-This construction is always a univalent fibration~\cite{XXX}. We can understand
+\newpage
+
+\jacques{The following is starting to be quite fuzzy}
+This construction always gives a univalent fibration~\cite{Christiansen},
+whenever the universe $\mathcal{U}$ is univalent. We can understand
 the basic idea as follows. Take two points $(A,|p|)$ and $(B,|q|)$ in the base
 space. We want to compare the space $A \simeq B$ with the space
-$(A,|p|) \equiv (B,|q|)$.  In one direction, we have $p : F \equiv A$ and
-$q : F \equiv B$ and hence $A \equiv B$ and by univalence $A \simeq B$. In the
-other direction \ldots
+$(A,|p|) \equiv (B,|q|)$.  
+\begin{itemize}
+\item Given $(A,|p|)$ and $(B,|q|)$, we have $p : F \equiv A$ and $q : F \equiv B$.
+By symmetry and transitivity of $\equiv$, we have $A \equiv B$; by
+$\mathit{ittoequiv}$, $A \simeq B$.
+\item Given $A \simeq B$, by univalence of $\mathcal{U}$, we get that
+$A \equiv B$.  
+\end{itemize}
 
@@
 \end{verbatim}
+
+\subsection{Looking at $\bracket{\bt}$}
+
+We claim that there is a link between $\PiTwo$ (defined in Sec.~\ref{sec:pi2})
+and $\bracket{\bt}$.  To see this, we begin by unpacking the points, 1-paths,
+and 2-paths of $\bracket{\bt}$.
+
+\paragraph*{Points in $\bracket{\bt}$.} The only term of type $\bracket{\bt}$ is
+$(\bt,\|\reflp_{\bt}\|)$ which we call $`\bt$. Formally we can prove that
+$\ptrunc{(X,p) \equiv `\bt}$ is inhabited for all $X$ and $p$.
+
+\paragraph*{1-Paths in $\{\bt\}$.} These are paths between $`\bt$ and
+$`\bt$. Generally speaking, paths between elements of a $\Sigma$-type are pairs
+of paths with a transport in the second component. If we have a type $A$ with
+points $a$ and $b$ and a path $p : a \equiv b$. Fix some $c$ and consider the
+dependent function $(x:A) \rightarrow (x \equiv c)$. This forms the types
+$a\equiv c$ and $b\equiv c$ with elements $ua$ and $ub$. Now a path between
+$(a,ua)$ and $(b,ub)$ is a pair $(p,\alpha)$ where $\alpha : p_* ua \equiv
+ub$. Hence paths between $`\bt$ and $`\bt$ are going to be for the form
+$(p,\alpha)$ where $p : \bt \equiv \bt$ and $\alpha$ is essentially $\gluep$
+from the definition of truncation. So we have two paths:
+\[\begin{array}{rcl}
+`\mathbf{id} &=& (\reflp,\gluep) \\
+`\mathbf{not} &=& (\notp,\gluep)
+\end{array}\]
+
+\paragraph*{2-Paths in $\{\bt\}$.} Now are considering paths between
+$`\mathbf{id}$ and $`\mathbf{id}$, and between $`\mathbf{not}$ and
+$`\mathbf{not}$. We have at least one path at this level which is
+$(\reflp,\gluep)$ but to show that that this is the only path we will have to
+first prove that $`\mathbf{not} \circ `\mathbf{not} \equiv `\mathbf{id}$.
 

### 2017-07-02 upi.tex
fixed a few typos etc

@@
 raises the question of when two programs are ``equivalent.'' Consider
-the follwing six programs of type~$\bt \leftrightarrow \bt$
+the following six programs of type~$\bt \leftrightarrow \bt$
 
@@
 \noindent We neither expect, nor desire, that all these programs of
-the same type to be be equivalent. In the above, each group of
+the same type to be equivalent. In the above, each group of
 three does represent ``the same'' program but the two groups
@@
 
-The second notion we need
-is that of equivalence. We say that two types or spaces $A$ and $B$
-are equivalent if there exist functions $f : A \rightarrow B$ and
-$g : B \rightarrow A$ that compose to the identity function%
-\footnote{Strictly speaking, this is a \emph{quasi-equivalence}, an
-equivalence also requires a certain coherence between the compositions.}.
-A notable point is that equivalences do not come with an induction principle.
-We say a universe $\mathcal{U}$ is univalent if its space of identities
-$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of
-equivalences $\AgdaFunction{A} ≃ \AgdaFunction{B}$.
+The second notion we need is that of equivalence. We say that two types or
+spaces $A$ and $B$ are equivalent if there exist functions $f : A \rightarrow B$
+and $g : B \rightarrow A$ that compose to the identity function.
+\footnote{Strictly speaking, this is a \emph{quasi-equivalence}, an equivalence
+  also requires a certain coherence between the compositions.}  A notable point
+is that equivalences do not come with an induction principle.  We say a universe
+$\mathcal{U}$ is univalent if its space of identities
+$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of equivalences
+$\AgdaFunction{A} ≃ \AgdaFunction{B}$.
 
@@
 
-While univalent universes came first, it was later realised that
+While univalent universes came first, it was later realized that
 the more general univalent type families are more convenient to
@@
 ``space'' and ``type'', ``path'' and ``equivalence'' as
-synonyms.  Furthermore our setting is an intuinistic type
+synonyms.  Furthermore our setting is an Intuitionistic type
 theory (with no classical axioms such a choice or excluded
@@
 As is often the case when confronted with new definitions, it is
-instructive to disect \emph{non-examples}, as a means to understand
+instructive to dissect \emph{non-examples}, as a means to understand
 what ``goes wrong''.  As a first example of a non-univalent fibration,

### 2017-07-03 upi.tex
state theorems

@@
 \usepackage{agda}
+\usepackage{stmaryrd}
 
@@
 \newcommand{\respcomp}[2]{#1 \odot #2}
+\newcommand{\trunc}{\textit{trunc}}
 
@@
 \newcommand{\isotwo}{\Leftrightarrow}
+\newcommand{\isothree}{\Lleftarrow \! \! \! \! \Rrightarrow}
 \newcommand{\piso}{\multimapdotbothB~~}
@@
 \newcommand{\ptrunc}[1]{\ensuremath{\| #1 \|}}
-
+\newcommand{\dbracket}[1]{\ensuremath{\llbracket \; #1 \; \rrbracket}}
 \newcommand{\PiTwo}{\ensuremath{\Pi_{\mathbb{2}}}}
@@
 
+We add a new level to $\{\PiTwo\}$ to prove the full correspondence of the first 2-levels of $\{\PiTwo\}$ and $\{\bt\}$:
+\[\def\arraystretch{0.8}\begin{array}{l@{\quad}rclrl}
+(\textit{3-combinators}) & \xi &::=& \trunc &:& \alpha \isothree \alpha
+\end{array}\]
+
+\begin{theorem}[Soundness and completeness for $\{\PiTwo\}$]
+  There exist functions $\dbracket{ \_ }_n$ which map $n-$combinator in $\PiTwo$ to $n-$path in $\{\bt\}$,
+  such that for all $n-$combinators $p, q$, $p \leftrightarrow_n q$ iff $\dbracket{ p }_n = \dbracket{ q }_n$.
+\end{theorem}
+
+\begin{theorem}[Soundness and completeness for $\{\bt\}$]
+  There exist functions $\dbracket{ \_ }_n^{-1}$ which map $n-$path in $\{\bt\}$ to $n-$combinator in $\PiTwo$,
+  such that for all $n-$path $p, q$, $p \equiv q$ iff $\dbracket{ p }_n^{-1} \isotwo \dbracket{ q }_n^{-1}$.
+\end{theorem}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-07-03 upi.tex
fix typo

@@
 
-We add a new level to $\{\PiTwo\}$ to prove the full correspondence of the first 2-levels of $\{\PiTwo\}$ and $\{\bt\}$:
+We add a new level to $\PiTwo$ to prove the full correspondence of the first 2-levels of $\PiTwo$ and $\{\bt\}$:
 \[\def\arraystretch{0.8}\begin{array}{l@{\quad}rclrl}
@@
 
-\begin{theorem}[Soundness and completeness for $\{\PiTwo\}$]
+\begin{theorem}[Soundness and completeness for $\PiTwo$]
   There exist functions $\dbracket{ \_ }_n$ which map $n-$combinator in $\PiTwo$ to $n-$path in $\{\bt\}$,

### 2017-07-04 upi.tex
lemma

--- 2017-07-02 upi.tex
@@
 
+We add a new level to $\PiTwo$ to prove the full correspondence of the first 2-levels of $\PiTwo$ and $\{\bt\}$:
+\[\def\arraystretch{0.8}\begin{array}{l@{\quad}rclrl}
+(\textit{3-combinators}) & \xi &::=& \trunc &:& (\alpha, \beta : c_1 \isotwo c_2) \to \alpha \isothree \beta
+\end{array}\]
+
+\begin{theorem}[Soundness and completeness for $\PiTwo$]
+  There exist functions $\dbracket{ \_ }_n$ which map n-combinator in $\PiTwo$ to n-path in $\{\bt\}$,
+  such that
+  \begin{enumerate}
+  \item For all 1-combinators $p, q$, $p \iso q$ iff $\dbracket{ p }_1 = \dbracket{ q }_1$.
+  \item For all 2-combinators $\alpha, \beta$, $\alpha \isotwo \beta$ iff $\dbracket{ \alpha }_2 = \dbracket{ \beta }_2$.
+  \end{enumerate}
+\end{theorem}
+
+\begin{theorem}[Soundness and completeness for $\{\bt\}$]
+  There exist functions $\dbracket{ \_ }_n^{-1}$ which map n-path in $\{\bt\}$ to n-combinator in $\PiTwo$,
+  such that
+  \begin{enumerate}
+  \item For all 1-paths $p, q$, $p \equiv q$ iff $\dbracket{ p }_1^{-1} \isotwo \dbracket{ q }_1^{-1}$.
+  \item For all 2-paths $\alpha, \beta$, $\alpha \equiv \beta$ iff $\dbracket{ \alpha }_2^{-1} \isothree \dbracket{ \beta }_2^{-1}$.
+  \end{enumerate}
+\end{theorem}
+
+\begin{lemma}
+  For all 1-combinators $p$, $p \isotwo \dbracket{ \dbracket{ p }_1 }_1^{-1}$.
+\end{lemma}
+
+\begin{lemma}
+  For all 2-combinators $\alpha : p \isotwo q$, $\alpha \isothree \dbracket{ \dbracket{ p }_1 }_1^{-1} \bullet \dbracket{ \dbracket{ \alpha }_2 }_2^{-1} \bullet (!!
+\; \dbracket{ \dbracket{ q }_1 }_1^{-1})$.
+\end{lemma}
+
+\begin{lemma}
+  For all 1-path $p$, $p \equiv \dbracket{ \dbracket{ p }_1^{-1} }_1$.
+\end{lemma}
+
+\begin{lemma}
+  For all 2-path $\alpha : p \equiv q$, $\alpha \equiv \dbracket{ \dbracket{ p }_1^{-1} }_1 \circ \dbracket{ \dbracket{ \alpha }_2^{-1} }_2 \circ (\dbracket{ \dbracket{ q }_1^{-1} }_1)^{-1}$.
+\end{lemma}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-07-04 upi.tex
tweak syntax of Π₂
sec 1 and 2

@@
 types, clearly preserve entropy in the information-theoretic
-sense~\cite{James:2012:IE:2103656.2103667}. We will identify a ``featherweight''
-version of $\Pi$ to use in our formal development.  But first we illustrate the
-general flavor of the family of languages with some examples.
+sense~\cite{James:2012:IE:2103656.2103667}. We illustrate the general flavor of
+the family of languages with some examples and then identify a ``featherweight''
+version of $\Pi$ to use in our formal development.
 
@@
 
-The examples below assume a representation of the type of booleans
-$\bt$ as the disjoint union $\ot \oplus \ot$ with the left injection
-representing $\mathsf{false}$ and the right injection representing
-$\mathsf{true}$. Given an arbitrary reversible function
-$\AgdaFunction{f}$ of type $a \leftrightarrow a$, we can build the
-reversible function $\AgdaFunction{controlled}~\AgdaFunction{f}$ that
-takes a pair of type $\bt \otimes a$ and checks the incoming boolean;
-if it is false (i.e., we are in the left injection), the function
-behaves like the identity; otherwise the function applies
-$\AgdaFunction{f}$ to the second argument. In either case, the
-incoming boolean is reconstituted to maintain reversibility.
+The examples below assume a representation of the type of booleans $\bt$ as the
+disjoint union $\ot \oplus \ot$ with the left injection representing
+$\mathsf{false}$ and the right injection representing $\mathsf{true}$. Given an
+arbitrary reversible function $\AgdaFunction{f}$ of type $a \leftrightarrow a$,
+we can build the reversible function
+$\AgdaFunction{controlled}~\AgdaFunction{f}$ that takes a pair of type
+$\bt \otimes a$ and checks the incoming boolean; if it is false (i.e., we are in
+the left injection), the function behaves like the identity; otherwise the
+function applies $\AgdaFunction{f}$ to the second argument. The incoming boolean
+is then reconstituted to maintain reversibility:
 
@@
 \[\def\arraystretch{1.2}\begin{array}{rcll}
-\AgdaFunction{controlled}  &:& \forall a.~ (a \leftrightarrow a) \rightarrow
-                            & (\bt \otimes a \leftrightarrow \bt \otimes a) \\
+\AgdaFunction{controlled}  &:& \forall a.~ (a \leftrightarrow a) \quad\rightarrow
+                            & ~(\bt \otimes a \leftrightarrow \bt \otimes a) \\
 \AgdaFunction{controlled}~\AgdaFunction{f} &=&
@@
 
-The $\AgdaFunction{not}$ function is a simple lifting of
-\AgdaFunction{swap₊}. Using the
-\AgdaFunction{controlled} building block, we can build a
-controlled-not $\AgdaFunction{cnot}$ gate and a
-controlled-controlled-not gate, also known as the
-\AgdaFunction{toffoli} gate. The latter gate is a universal function
-for combinational boolean circuits thus showing the expressiveness of
-the language.
+\noindent The left column shows the sequence of types that are visited during
+the computation; the right column shows the names of the combinators\footnote{We
+  use names that are hopefully quite mnemonic; for the precise definitions of
+  the combinators see the $\Pi$-papers.} that witness the corresponding type
+isomorphism. The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
+constructive evidence (i.e., a program, a logic gate, or a hardware circuit) for
+an automorphism on $\bt \otimes a$: it can be read top-down or bottom-up to go
+back and forth.
+
+The $\AgdaFunction{not}$ function below is a simple lifting of
+\AgdaFunction{swap₊} which swaps the left and right injections in a sum
+type. Using the \AgdaFunction{controlled} building block, we can build a
+controlled-not ($\AgdaFunction{cnot}$) gate and a controlled-controlled-not
+gate, also known as the \AgdaFunction{toffoli} gate. The latter gate is a
+universal function for combinational boolean circuits thus showing the
+expressiveness of the language:
+
 {\small
@@
 
-While we wrote \AgdaFunction{controlled} in equation-reasoning style,
-\AgdaFunction{not} is written in combinator style.  These are
-equivalent as $\byiso{-}$ is defined in terms of $\odot$.
-
-The code for $\AgdaFunction{controlled}~\AgdaFunction{f}$ provides
-constructive evidence (i.e., a program, a logic gate, or a hardware
-circuit) for an automorphism on $\bt \otimes a$: it can be read
-top-down or bottom-up to go back and forth. This perspective naturally
-raises the question of when two programs are ``equivalent.'' Consider
-the following six programs of type~$\bt \leftrightarrow \bt$
+\noindent While we wrote \AgdaFunction{controlled} in equational-reasoning
+style, \AgdaFunction{not} is written in the point-free combinator style.  These
+are equivalent as $\byiso{-}$ is defined in terms of the sequential composition
+combinator $\odot$.
+
+As is customary in any semantic perspective on programming languages, we are
+interested in the question of when two programs are ``equivalent.'' Consider the
+following six programs of type~$\bt \leftrightarrow \bt$:
 
@@
 
-\noindent We neither expect, nor desire, that all these programs of
-the same type to be equivalent. In the above, each group of
-three does represent ``the same'' program but the two groups
-represent distinct programs. Thinking extensionally, i.e., by looking
-at all possible input-output pairs, it is indeed easy to verify this
-fact, but we can do better. We can provide \emph{evidence} (i.e., a
-program that manipulates programs) that can constructively transform
-every program to an equivalent one. We show such a level-2 program
-proving that $\AgdaFunction{not₃}$ is equivalent to
-$\AgdaFunction{not}$. For illustration, the program for
-$\AgdaFunction{not₃}$ is depicted in Fig.~\ref{fig:not}. We encourage
-the reader to map the steps below to manipulations on the diagram that
-would incrementally simplify it:
+\noindent The programs are all of the same type but this is clearly not a
+sufficient condition for ``equivalence.'' Thinking extensionally, i.e., by
+looking at all possible input-output pairs, it is easy to verify that the six
+programs split into two classes: one equivalent to the identity function and one
+equivalent to the boolean negation. In the context of $\Pi$, we can better: we
+can provide \emph{evidence} (i.e., a program that manipulates programs) that can
+constructively transform every program to an equivalent one. We show such a
+level-2 program proving that $\AgdaFunction{not₃}$ is equivalent to
+$\AgdaFunction{not}$. For illustration, the program for $\AgdaFunction{not₃}$ is
+depicted in Fig.~\ref{fig:not}. We encourage the reader to map the steps below
+to manipulations on the diagram that would incrementally simplify it:
 
@@
 
-It is worthwhile mentioning that the above derivation could also be drawn as
-one (large!) commutative diagram in an appropriate category, with each
+\noindent It is worthwhile mentioning that the above derivation could also be
+drawn as one (large!) commutative diagram in an appropriate category, with each
 $\byisotwo{-}$ as a $2$-arrow (and representing a natural isomorphism).
@@
 
-\jacques{the code above uses $\odot$ for 1-composition,
-$\boxdot$ for parallel 2-composition of $\odot$, while the
-code below uses $\circ$ and $\odot$ respectively, which is
-quite confusing.  We should pick one notation.}
+% \jacques{the code above uses $\odot$ for 1-composition,
+% $\boxdot$ for parallel 2-composition of $\odot$, while the
+% code below uses $\circ$ and $\odot$ respectively, which is
+% quite confusing.  We should pick one notation.}
+
 \[\def\arraystretch{0.8}\begin{array}{l@{\quad}rclrl}
@@
                && \alt & ! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
-               && \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)  \\
+               && \alt & \odot &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)  \\
 \\
@@
                                   \compc{(\compc{c_1}{c_2})}{c_3} \isotwo \compc{c_1}{(\compc{c_2}{c_3})} \\
-            && \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to
+            && \alt & \boxdot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to
                              (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
@@
 combinators $\id$ and $\swap$ which are closed under inverses $!$ and
-sequential composition $\circ$. The syntactic category $\alpha$ is
+sequential composition $\odot$. The syntactic category $\alpha$ is
 that of 2-combinators denoting reversible program transformations,
@@
 
-It is relatively easy to think of a few sound program transformations
-to include in the category $\alpha$ of 2-combinators. The set above
-however is \emph{complete}. We also have canonical forms (constructively).
+It is relatively easy to think of a few sound program transformations to include
+in the category $\alpha$ of 2-combinators. But, as the following lemma
+establishes, the above set is \emph{complete}:
 
 \begin{lemma}[Canonical Forms]
-  Given a 1-combinator $c : \tau \leftrightarrow \tau$, we either have
-  a normalization $c \Leftrightarrow \id$ or a normalization
-  $c \Leftrightarrow \swap$.
+  Given a 1-combinator $c : \tau \leftrightarrow \tau$, we either have a
+  2-combinator of type $c \Leftrightarrow \id$ or a 2-combinator of type
+  $c \Leftrightarrow \swap$. In other words, every 1-combinator has a canonical
+  representation as either $\id$ or $\swap$ and the set of 2-combinators is rich
+  enough to normalize $c$ to its canonical representation.
 \end{lemma}
@@
 
+{\small
+\[\def\arraystretch{1.2}\begin{array}{rcll}
+\AgdaFunction{$\dbracket{ \_ }_1$}  &:& (\bt \leftrightarrow \bt) \rightarrow `\bt \equiv `\bt \\
+\AgdaFunction{$\dbracket{ \id }_1$} &=& `\mathbf{id}\\
+\AgdaFunction{$\dbracket{ \swap }_1$} &=& `\mathbf{not}\\
+\AgdaFunction{$\dbracket{ ! p }_1$} &=& (\dbracket{ p }_1)^{-1} \\
+\AgdaFunction{$\dbracket{ p \odot q }_1$} &=& \dbracket{ p }_1 \circ \dbracket{ p }_1\\
+\end{array}
+\]}
+
 \begin{lemma}

### 2017-07-05 upi.tex
citation
minor tweaks

@@
 drawn as one (large!) commutative diagram in an appropriate category, with each
-$\byisotwo{-}$ as a $2$-arrow (and representing a natural isomorphism).
-See~\cite{Shulmann} for that interpretation.
+$\byisotwo{-}$ as a $2$-arrow (and representing a natural isomorphism).  See
+Shulman's draft book~\cite{shulman} for that interpretation.
 
@@
 
-Of course, we can obtain another (trivial) example from $\mathbb{1}$:
+Of course, we can obtain another (trivial) example from the unit type:
 $P : \mathbb{1} \rightarrow \mathbb{U}$, defined as
@@
 
-{\small
 \[\def\arraystretch{1.2}\begin{array}{rcll}
@@
 \end{array}
-\]}
+\]
 

### 2017-07-05 upi.tex
add some discussion, remove some material that seemed to veer too far off.  Split discussion and conclusion.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion, Related Work, and Conclusion}
-\label{sec:conclusion}
+\section{Discussion and Related Work}
+\label{sec:discussion}
 
@@
 
-By restricting to finite types, two operational semantics emerge:
-\begin{enumerate}
-\item a (reversible) operational semantics of programs over finite types, and
-\item a (reversible) operational semantics of program transformers of such programs.
-\end{enumerate}
-$\Pi$ programs can be interpreted as generalized permutations, as they work on
-structure as well as the underlying skeleton of finite types, aka the natural numbers.
-Many of the transformers can be \emph{oriented}, and thus be interpreted as
-\emph{optimizations}.  As with traditional languages, there are different aspects
-we can optimize for (program size, number of steps to completion, etc); we will
-not touch on this further in this paper.
+On can also flip this around, and use the $\lambda$-calculus as the
+internal language for Cartesian Closed Categories.  However, as Shulman
+explains well in his draft book~\cite{Shullman} on approaching Categorical
+Logic via Type Theory, this works for many other kinds of categories.  As
+we are interested in \emph{reversibility}, it is most natural to look at
+Groupoids.  Thus $\PiTwo$ represents the simplest non-trivial case of
+a (reversible) programming language distilled from such ideas.
+
+What is more surprising is how this also turns out to be a sound
+and complete language for describing the univalent universe $\bracket{\bt}$.
+
+\section{Conclusion}
+\label{sec:conclusion}
+
+What is $\bracket{S^1}$?  Is it useful for programming?  What about $\bracket{\mathbb{N}}$?
+
+What is the ``right'' generalization of $\bracket{-}$ so that we may have all
+the usual finite types (such as the ones available in $\Pi$) properly
+represented?
+
+\jacques{It is not clear to me that just taking a disjoint union over all the
+  types gives the correct generalization.}
+
+Looking at this from the other end: given some ``exotic'' (but finitely
+presented) Groupoid $\mathfrak{G}$, is there always a programming language
+which is sound and complete for $\mathfrak{G}$ ?
 

### 2017-07-09 upi.tex
remove lemmas
Some comments in sec 3

@@
 
+\newcommand{\newnote}[2]{{\sf {#1} $\clubsuit$ {#2} $\clubsuit$}}
+\newcommand{\VC}[1]{{\color{orange}{\newnote{VC}{#1}}}}
+
+\newcommand{\newtext}[1]{{\color{purple}{#1}}}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Propositional equality, or identity types in intensional dependent
-type theory, or path types under the HoTT-interpretation, come equipped
-with a single constructor \AgdaInductiveConstructor{refl} and an
-induction principle $\AgdaFunction{J}$. A peculiar aspect of
-$\AgdaFunction{J}$ is that it does \emph{not} enforce that every
-element of an identity type
+Propositional equality \newtext{is embodied by} identity types in
+intensional dependent type theory, or path types under the
+HoTT-interpretation. The identity type comes equipped with a single
+constructor \AgdaInductiveConstructor{refl} and an induction principle
+$\AgdaFunction{J}$. A peculiar aspect of $\AgdaFunction{J}$ is that it
+does \emph{not} enforce that every element of an identity type
 $\AgdaFunction{x}~\AgdaDatatype{≡}~\AgdaFunction{y}$ is
-\AgdaInductiveConstructor{refl}. This ``weakness'' has a beautiful
-consequence: instead of identifying all proofs and collapsing the
-structure of proofs in the process, it instead permits a rich
-combinatorial structure to emerge.
-
-The second notion we need is that of equivalence. We say that two types or
-spaces $A$ and $B$ are equivalent if there exist functions $f : A \rightarrow B$
-and $g : B \rightarrow A$ that compose to the identity function.
-\footnote{Strictly speaking, this is a \emph{quasi-equivalence}, an equivalence
-  also requires a certain coherence between the compositions.}  A notable point
-is that equivalences do not come with an induction principle.  We say a universe
-$\mathcal{U}$ is univalent if its space of identities
-$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of equivalences
-$\AgdaFunction{A} ≃ \AgdaFunction{B}$.
+\AgdaInductiveConstructor{refl}. \VC{why is that peculiar?}
+\newtext{However, unlike other types, it is not necessary to have a
+  uniqueness principle ($K$) for it, which facilitates the homotopy
+  interpretation of intensional type theory.} This ``weakness'' has a
+beautiful consequence: instead of identifying all proofs and collapsing
+the structure of proofs in the process, it instead permits a rich
+combinatorial structure to emerge. \VC{cite Homotopy Theoretic Aspects
+  of Constructive Type Theory by Michael Warren}
+
+The second notion we need is that of equivalence. We say that two types
+or spaces $A$ and $B$ are \newtext{(homotopy)} equivalent if there exist
+functions $f : A \rightarrow B$ and $g : B \rightarrow A$ that compose
+to the identity function. \footnote{Strictly speaking, this is a
+  \emph{quasi-equivalence}, an equivalence also requires a certain
+  coherence between the compositions.}  A notable point is that
+equivalences do not come with an induction principle. \VC{What does this
+  mean exactly? Maybe you're trying to say that it does not satisfy a
+  principle like path induction, but in a univalent universe, it will!}
+We say a universe $\mathcal{U}$ is univalent if its space of identities
+$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of
+equivalences $\AgdaFunction{A} ≃ \AgdaFunction{B}$.
 
@@
   equivalence.}
+
+\VC{Explain universal fibrations, object classifier?}
 
@@
 
-\[\def\arraystretch{1.2}\begin{array}{rcll}
-\AgdaFunction{$\dbracket{ \_ }_1$}  &:& (\bt \leftrightarrow \bt) \rightarrow `\bt \equiv `\bt \\
-\AgdaFunction{$\dbracket{ \id }_1$} &=& `\mathbf{id}\\
-\AgdaFunction{$\dbracket{ \swap }_1$} &=& `\mathbf{not}\\
-\AgdaFunction{$\dbracket{ ! p }_1$} &=& (\dbracket{ p }_1)^{-1} \\
-\AgdaFunction{$\dbracket{ p \odot q }_1$} &=& \dbracket{ p }_1 \circ \dbracket{ p }_1\\
-\end{array}
-\]
-
-\begin{lemma}
-  For all 1-combinators $p$, $p \isotwo \dbracket{ \dbracket{ p }_1 }_1^{-1}$.
-\end{lemma}
-
-\begin{lemma}
-  For all 2-combinators $\alpha : p \isotwo q$, $\alpha \isothree \dbracket{ \dbracket{ p }_1 }_1^{-1} \bullet \dbracket{ \dbracket{ \alpha }_2 }_2^{-1} \bullet (!!
-\; \dbracket{ \dbracket{ q }_1 }_1^{-1})$.
-\end{lemma}
-
-\begin{lemma}
-  For all 1-path $p$, $p \equiv \dbracket{ \dbracket{ p }_1^{-1} }_1$.
-\end{lemma}
-
-\begin{lemma}
-  For all 2-path $\alpha : p \equiv q$, $\alpha \equiv \dbracket{ \dbracket{ p }_1^{-1} }_1 \circ \dbracket{ \dbracket{ \alpha }_2^{-1} }_2 \circ (\dbracket{ \dbracket{ q }_1^{-1} }_1)^{-1}$.
-\end{lemma}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-10-13 PiFin/paper/main.tex
Start a document to outline what we have

\documentclass{article}

\usepackage{parskip}

\usepackage{macros}

\title{Reversible Programming with Finite Types}
\author{}

\begin{document}

\maketitle

We have the following type families:

\begin{enumerate}
  \item $Fin : \Nat \to \U$
  \item $El : \Nat \to \U$
  \item $CPerm : \Nat \to \Nat \to \U$
\end{enumerate}

We have the following equivalences:

\begin{gather*}
  \equiv{Fin(n)}{El(n)} \\
  \equiv{(Fin(n) \to A)}{List(A, n)} \\
  \equiv{CPerm(n,m)}({\equiv{El(n)}{El(m)}}) \text{ (unproven)} \\
\end{gather*}

We can characterize the path spaces of:

\begin{gather*}
  \Id{\U}{Fin(n)}{Fin(m)} \to \Id{\Nat}{m}{n}
\end{gather*}

We need to show that the subset of Tarski's high school algebra identities
(without exponentials), that is, in the signature $(\Nat, +, \times)$ is
complete.

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2021-01-27 Pi+/paper/univalent.tex
Setup latex template

\section{Univalent Subuniverses}~\label{sec:univalent}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2021-01-30 Pi+/paper/finite/main.tex
Starting

\documentclass[a4paper]{article}

\usepackage{amssymb,mathrsfs,amsmath,amscd,amsthm}
\usepackage[mathcal]{euscript}
\usepackage{stmaryrd}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{graphics}
\usepackage{amsmath}

\newcommand{\N}{\mathbb{N}}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}

%%%%%%% makra do notacji

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{example}
\newtheorem{example}{Example}

\begin{document}
\section{Permutations}
\subsection{Lehmer codes}

\begin{definition}
Let $n \in \N$ and a permutation $p$ be on a set $X = \{1, 2, ... n\}$. A Lehmer code $l(p)$ is an $n$-tuple $(r_0, r_1, ... r_{n-1})\in \N^n$ where the $r_k$-th number satisfies $r_k \leq k$. 
\end{definition}

Thus, for any given $n$, there are $n!$ possible Lehmer codes of length $n$. A the same time, there are also $n!$ permutations on a set of $n$ elements. Itmeans there is a bijection $f$ between those two sets, allowing for an encoding a permutation by a corresponting Lehmer code.

We are interested in a particular choice for $f$. To define it, we need to first introduce a notion of an inversion. 
...

The interpretation we have arrived is this: the number $r_k$ in a Lehmer code $c$ denotes the number of inversions of the element $k$ (right inversion count).

\subsection{Symmetric group}

\begin{definition}
A presentation of a group, denoted by $<S | R>$ consists of a set of generators $S$ and a set of equivalence relations $R$.
Relations are defined on the set of words $(S \cup S^{-1})^*$, where $S^{-1}$ denotes a set of formal inverses of elements in $S$.
\end{definition}

\begin{definition}
For a group $G = (G, \dot)$, we say that $\la S | R \ra$ is a presentation of $G$, if:
\begin{itemize}
	\item There is a function $f$ mapping elements of $S$ to elements of $G$.
	\item There is a surjective extension $f^*$ of $f$ to $(S \cup S^{-1})^*$ satisfying
		\begin{equation}
		\begin{cases}
		f^*(s) = f(h) & \text{ for } s \in S \\
		f^*(s^{-1}) = f(s)^{-1} & \text{ for } s \in S^{-1} \\
		f^*(w w') = f^*(w) \dot f^*(w') & \text{ for } w, w' \in (S \cup S^{-1})^*
		\end{cases}
		\end{equation}
	\item Function $f^*$ preserves the congruence closure of the set of relations $R \cup \{s = s^{-1}\}$.
\end{itemize}

For a number $n \in \N$, $n>0$, by $S_n$ we denote the symmetric group on the set $\{1, 2, ... n\}$ - elements of the group are permutations on this set. We are interested in a particular presentation of $S_n$, called Coxeter presentation.
\end{definition}

\begin{definition}
Coxeter presentation of $S_n$ is defined as $\la S|R \ra$, where $S = \{\tau_i : i \in \{1, 2, ... n - 1\}\}$ and
	\begin{equation*}
		R =
		    \begin{cases}
		      \tau_i\tau_j = \tau_j\tau_i & \text{for } i < j + 1\\
		      (\tau_i)^2 = 1 & \text{for } i \in \{1, 2, ... n - 1\}\\
		      (\tau_i\tau_{i+1})^3 = 1 & \text{for} i \in \{1, 2, ... n - 2\}\\
		    \end{cases}
	\end{equation*}
The third equation can be also written in a form $\tau_i\tau_{i+1}\tau_i = \tau_{i+1}\tau_i\tau_{i+1}$.
\end{definition}
\end{document}

### 2021-01-31 Pi+/paper/finite.tex
Work

--- 2021-01-27 Pi+/paper/univalent.tex
@@
-\section{Univalent Subuniverses}~\label{sec:univalent}
+\section{The groupoid of finite types}~\label{sec:finite}
+
+In this section, we describe the algebraic structure of the groupoid of finite types by characterising its paths.
+
+\subsection{Permutations}
+
+In the previous~\cref{sec:univalent}, we established that paths in $\UFin$ are equivalent to families of automorphisms of $\Fin{n}$ for every $n:\Nat$.
+These are permutations on finite sets of size $n$. In the following sections, we will characterise these permutations, going through a number of intermediate step. \jk{I don't like the wording}
+
+\subsubsection{Lehmer codes}
+
+There are $\fac{n}$ permutations on a set of $n$ elements.
+Thus, if we construct a type with $\fac{n}$ inhabitants, there should be a bijection of that type with the type of permutations $\Aut[\Fin[n]]$.
+To construct this bijection, we will borrow the idea of Lehmer codes~\cite{lehmerTeachingCombinatorialTricks1960a} from Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960}.
+The type family $\Lehmer$ over $\Nat$ is defined as follows.
+
+\todo{Understand the proof first and then revisit this paragraph}
+
+\begin{definition}
+    \begin{align*}
+         & \Lehmer : \Nat \to \UU                             \\
+         & \Lehmer[0] = \unit                                 \\
+         & \Lehmer[\suc[n]] = \Fin[\suc[n]] \times \Lehmer[n]
+    \end{align*}
+\end{definition}
+
+Now we construct the equivalence.
+
+\begin{proposition}
+    For all $n:\Nat$,
+    \[
+        \Lehmer[n] \eqv \Aut[\Fin[n]]
+    \]
+\end{proposition}
+
+\subsubsection{Symmetric groups}
+
+There is an obvious group structure on $\Aut[\Fin[n]]$ given by identity, composition, and inverse.
+This is the symmetric group $S_n$ on $n$ symbols.
+In the rest of the section we will construct a convenient presentation of this group.
+
+\todo{Find the reference for group presentations as coequalisers (possibly Mac Lane)}
+
+First, we formally define a presentation of a group.
+
+\begin{definition}
+    A presentation of a group $G$ is a type $FR$ where
+    $FR === Free n \to \Aut[\Fin[n]]$
+    is a coequaliser.
+\end{definition}
+
+\begin{definition}
+    Here we'll define a type family $FT n$ to be a free group generated by a type $T n$ with the following constructors
+    \begin{align*}
+        & cancel : i : Fin (S n) -> T \\ 
+        & swap : (i : Fin (S n)) -> (j : Fin (S n)) -> (i + 1 < j) -> T \\
+        & braid : (i : Fin n) -> T \\
+    \end{align*}
+\end{definition}
+
+\begin{definition}[Adjacent transposition]
+    \begin{align*}
+        transpose : (n : \Nat) \to (k : \Fin[S n]) \to \Aut[\Fin[S n]]
+    \end{align*}
+    by double induction on $n$ and $k$, where
+    \begin{align*}
+        transpose n 0 = \lambda 
+            0 -> 1
+            1 -> 0
+            m -> m
+        transpose (S n) (S k) = \lambda
+            0 -> 0
+            S m -> S ((transpose n k) m)
+    \end{align*}
+\end{definition}
+
+\begin{proposition}
+    $FR$ is a presentation of the symmetrc group, with $f$ and $g$ as follows:
+    \begin{align*}
+        f inr (cancel i) = inr (i :: i :: [])
+        f inl (cancel i) = inl (i :: i :: [])
+        f inr (swap i j) = i :: j :: i :: j :: []
+        f braid i = i :: S i :: i :: S i :: i :: S i :: []
+    \end{align*}
+    and
+    \begin{align*}
+        g inl _ = inl [] 
+        g inr _ = inr [] 
+    \end{align*}
+\end{proposition}
+
+We have : Aut(Fin n) ~ Sn
+It is required to prove that 
+    FR is presentation of Aut(Fin n)
+
+% \begin{definition}
+% A presentation of a group, denoted by $<S | R>$ consists of a set of generators $S$ and a set of equivalence relations $R$.
+% Relations are defined on the set of words $(S \cup S^{-1})^*$, where $S^{-1}$ denotes a set of formal inverses of elements in $S$.
+% \end{definition}
+
+% \begin{definition}
+% For a group $G = (G, \dot)$, we say that $\langle S | R \rangle$ is a presentation of $G$, if:
+% \begin{itemize}
+% 	\item There is a function $f$ mapping elements of $S$ to elements of $G$.
+% 	\item There is a surjective extension $f^*$ of $f$ to $(S \cup S^{-1})^*$ satisfying
+% 		\begin{equation}
+% 		\begin{cases}
+% 		f^*(s) = f(h) & \text{ for } s \in S \\
+% 		f^*(s^{-1}) = f(s)^{-1} & \text{ for } s \in S^{-1} \\
+% 		f^*(w w') = f^*(w) \dot f^*(w') & \text{ for } w, w' \in (S \cup S^{-1})^*
+% 		\end{cases}
+% 		\end{equation}
+% 	\item Function $f^*$ preserves the congruence closure of the set of relations $R \cup \{s = s^{-1}\}$.
+% \end{itemize}
+
+% For a number $n \in \Nat$, $n>0$, by $S_n$ we denote the symmetric group on the set $\{1, 2, ...
+n\}$ - elements of the group are permutations on this set.
+We are interested in a particular presentation of $S_n$, called Coxeter presentation.
+% \end{definition}
+
+% \begin{definition}
+% Coxeter presentation of $S_n$ is defined as $\langle S|R \rangle$, where $S = \{\tau_i : i \in \{1, 2, ... n - 1\}\}$ and
+% 	\begin{equation*}
+% 		R =
+% 		    \begin{cases}
+% 		      \tau_i\tau_j = \tau_j\tau_i & \text{for } i < j + 1\\
+% 		      (\tau_i)^2 = 1 & \text{for } i \in \{1, 2, ... n - 1\}\\
+% 		      (\tau_i\tau_{i+1})^3 = 1 & \text{for} i \in \{1, 2, ... n - 2\}\\
+% 		    \end{cases}
+% 	\end{equation*}
+% The third equation can be also written in a form $\tau_i\tau_{i+1}\tau_i = \tau_{i+1}\tau_i\tau_{i+1}$.
+% \end{definition}
+
+We introduce a presentation of the symmetric group
 

### 2021-02-20 Pi+/paper/finite.tex
Add text in section 2
Fixes #72

--- 2021-01-27 Pi+/paper/univalent.tex
@@
-\section{Univalent Subuniverses}~\label{sec:univalent}
+\section{The groupoid of finite types}~\label{sec:finite}
+
+In this section, we describe the algebraic structure of the groupoid of finite types by characterising its paths.
+
+\subsection{Permutations}
+
+In the previous~\cref{sec:univalent}, we established that paths in $\UFin$ are equivalent to families of automorphisms
+of $\Fin{n}$ for every $n:\Nat$. These are permutations on finite sets of size $n$. In the following sections, we will
+characterise these permutations, going through a number of intermediate step.
+
+\jk{I don't like the wording}
+
+\subsubsection{Lehmer codes}
+
+There are $\fac{n}$ permutations on a set of $n$ elements. Thus, if we construct a type with $\fac{n}$ inhabitants,
+there should be a bijection of that type with the type of permutations $\Aut[\Fin[n]]$. To construct this bijection, we
+will borrow the idea of Lehmer codes~\cite{lehmerTeachingCombinatorialTricks1960a} from Combinatorial
+Analysis~\cite{bellmanCombinatorialAnalysis1960}. The type family $\Lehmer$ over $\Nat$ is defined as follows.
+
+\todo{Understand the proof first and then revisit this paragraph}
+
+\begin{definition}
+    \begin{align*}
+         & \Lehmer : \Nat \to \UU                             \\
+         & \Lehmer[0] = \unit                                 \\
+         & \Lehmer[\suc[n]] = \Fin[\suc[n]] \times \Lehmer[n]
+    \end{align*}
+\end{definition}
+
+Now we construct the equivalence.
+
+\begin{proposition}
+    For all $n:\Nat$,
+    \[
+        \Lehmer[n] \eqv \Aut[\Fin[n]]
+    \]
+\end{proposition}
+
+\subsubsection{Symmetric groups}
+
+There is an obvious group structure on $\Aut[\Fin[n]]$ given by identity, composition, and inverse. This is the
+symmetric group $S_n$ on $n$ symbols. In the rest of the section we will construct a convenient presentation of this
+group.
+
+\todo{Reference T-algebra presentations as coequalisers (Mac Lane 6.7)}
+
+First, we formally define a presentation of a group.
+
+\begin{definition}
+  A presentation of a group $G$ is a type $FR$ such that \ldots is a coequaliser.
+\end{definition}
+
+\vc{this is just a rough draft for now}
+
+\begin{definition}
+  Here we'll define a type family $FT n$ to be a free group generated by a type $T n$ with the following constructors
+  \begin{align*}
+    & cancel : i : \Fin[\suc[n]] \to T \\
+    & swap : (i : \Fin[\suc[n]]) \to (j : \Fin[\suc[n]]) \to (i + 1 < j) \to T \\
+    & braid : (i : \Fin[n]) \to T \\
+  \end{align*}
+\end{definition}
+
+\begin{definition}[Adjacent transposition]
+  \begin{align*}
+    transpose : (n : \Nat) \to (k : \Fin[\suc[n]]) \to \Aut[\Fin[\suc[n]]]
+  \end{align*}
+  by double induction on $n$ and $k$, where
+  \begin{align*}
+    transpose (n) (0) & = \lambda
+    0 \to 1;
+    1 \to 0;
+    m \to m
+    \\
+    transpose (\suc[n]) (\suc[k]) & = \lambda
+    0 \to 0;
+    S m \to S ((transpose (n)(k)) m)
+  \end{align*}
+\end{definition}
+
+\begin{proposition}
+  $FR$ is a presentation of the symmetrc group, with $f$ and $g$ as follows:
+  \begin{align*}
+    f (inr (cancel (i))) & = inr (i :: i :: []) \\
+    f (inl (cancel (i))) & = inl (i :: i :: []) \\
+    f (inr (swap (i,j))) & = i :: j :: i :: j :: [] \\
+    f (braid (i)) & = i :: S i :: i :: S i :: i :: S i :: []
+  \end{align*}
+  and
+  \begin{align*}
+    g (inl (\_)) & = inl ([]) \\
+    g (inr (\_)) & = inr ([])
+  \end{align*}
+\end{proposition}
+
+We now introduce a presentation of the symmetric group.
 

### 2021-02-21 Pi+/paper/univalent.tex

@@
 \section{Univalent Subuniverses}~\label{sec:univalent}
+
+In this section, we introduce some basic concepts and notation that we use in
+Homotopy Type Theory. Then, we define univalent subuniverses and discuss some
+specific examples.
+
+\subsection{The Type Theory}~\label{subsec:type-theory}
+
+We work in Homotopy Type Theory, in particular, intensional Martin-L\"{o}f Type
+Theory, with a univalent universe, and Higher Inductive Types (HITs) for
+propositional truncations and set quotients. We recall a few basic facts.
+
+\subsubsection{Identity Types}
+
+\subsubsection{Univalence}
+
+\subsubsection{Higher Inductive Types}
+
+\subsection{Univalent Subuniverses}
 

### 2021-04-12 Pi+/paper/univalent.tex
Outline some text in section 3

@@
 
-In this section, we introduce some basic concepts and notation that we use in
-Homotopy Type Theory. Then, we define univalent subuniverses and discuss some
-specific examples.
+In this section, we introduce some basic concepts and notation that we use in Homotopy Type Theory. Then, we define
+univalent subuniverses and discuss some specific examples.
 
@@
 
-We work in Homotopy Type Theory, in particular, intensional Martin-L\"{o}f Type
-Theory, with a univalent universe, and Higher Inductive Types (HITs) for
-propositional truncations and set quotients. We recall a few basic facts.
+We work in Homotopy Type Theory, in particular, intensional Martin-L\"{o}f Type Theory, with a univalent universe, and
+Higher Inductive Types (HITs) for propositional truncations and set quotients. We recall a few basic facts.
 
@@
 
+\begin{definition}[Type families are fibrations]
+  A type family $P : A \to \UU$ gives a fibration $\fst : \dsum{x:A}{P(x)} \to A$ with total space $\dsum{x:A}{P(x)}$
+  and base space $A$ with a lifting property.
+\end{definition}
+
+\begin{definition}[$\term{transport-equiv}$]
+  Given a type family $P : A \to \UU$, we define
+  \[
+    \term{transport-equiv} : x \id_{A} y \to P(x) \eqv P(y)
+  \]
+\end{definition}
+
+\begin{definition}[Univalent Fibration]
+  $P$ is a univalent fibration if $\term{transport-equiv}$ is an equivalence.
+\end{definition}
+
+\begin{definition}[Universe]
+  A universe \`{a} la Tarski is given by a code $U : \UU$ and a decoding function $\El : U \to \UU$. If $\El$ is a
+  univalent fibration, $U$ is a univalent universe.
+\end{definition}
+
+\begin{definition}[Subuniverse]
+  A subtype is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $\forall x, \isProp{P(x)}$. A
+  subuniverse generated by a subtype has $U \defeq \dsum{X:\UU}{P(X)}$ and $\El \defeq \fst$.
+\end{definition}
+
+\begin{proposition}[Univalent Subuniverse]
+  Subuniverses generated by subtypes are univalent.
+\end{proposition}
+
+\begin{example}[$\Aut$]
+  \[
+    \Aut[T] \defeq T \eqv T
+  \]
+\end{example}
+
+\begin{example}[$\BAut$]
+  \[
+    \BAut[T] \defeq \Sub{T}
+  \]
+\end{example}
+
+\begin{example}
+  \[
+    P(X) \defeq \SubP{X}{\Bool}
+  \]
+\end{example}
+
+\begin{example}
+  For any $n : \Nat$, we have
+  \[
+    P(X) \defeq \SubP{X}{\Fin[n]}
+  \]
+  \[
+    \UFin[n] \defeq \Sub{\Fin[n]}
+  \]
+  is the univalent subuniverse of types with cardinality $n$.
+\end{example}
+
+\begin{definition}[$\isFin$]
+  \[
+    \isFin[X] \defeq \dsum{n:\Nat}{\SubP{X}{\Fin[n]}}
+  \]
+\end{definition}
+
+\begin{proposition}
+  $\isFin$ is a proposition.
+\end{proposition}
+
+\begin{example}
+  \[
+    \UFin \defeq \dsum{X:\UU}{\isFin[X]}
+  \]
+  is the univalent subuniverse of all finite types.
+\end{example}
+
+We characterise the path space of subunivalent universes.
+
+\begin{proposition}
+  If $T$ is an $n$-type, $\BAut[T]$ is an $(n+1)$-type.
+\end{proposition}
+
+\begin{proposition}
+  $\UFin$ is a 1-type, that is, a 1-groupoid, or has h-level 3.
+\end{proposition}
+
+\begin{proposition}
+  \[
+    \Omega\BAut[T] \eqv \Aut[T]
+  \]
+\end{proposition}
+
+\begin{proposition}
+  For every $n:\Nat$,
+  \[
+    \Omega\UFin[n] \eqv \Aut[\Fin[n]]
+  \]
+\end{proposition}
+
 %%% Local Variables:

### 2021-04-19 Pi+/paper/discussion.tex
Some more outline

--- 2021-01-27 Pi+/paper/univalent.tex
@@
-\section{Univalent Subuniverses}~\label{sec:univalent}
+\section{Discussion \& Related Work}~\label{sec:discussion}
+
+Algebraic theories, 2-algebraic theories, Syntax of Coherence (Yanofsky)
+
+Higher order rewriting theory in constructive type theory.
+
+Free monoids, free commutative monoids, normalisation of monoids.
+
+Coherence problems in type theory, coherence via Well-Foundedness.
+
+Computational group theory (in HoTT?)
+
+Presentation of the free symmetric monoidal groupoid on one generator, proof of Mac Lane's coherence theorem in HoTT.
+
+Alternatively, we can describe them using HITs, prove they have the universal property and hence equivalent.
+The full Pi is FSMG 1, and the normalised fragment of Pi is M1.
+
+Applications of FSMG and history of the coherence theorems.
+
+Other proofs of coherence theorems, Joyal-Street.
+
+Pi has other extensions (fractional/negative/recursive types).
+What are the free X monoidal structures they're describing?
+
+Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
+should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.
+
+Other applications of symmetric groups.
+
+Actions of symmetric groups, permutation groupoids.
 

### 2021-04-19 Pi+/paper/equivalence.tex
Work
Fix build
Some more outline

--- 2021-01-27 Pi+/paper/univalent.tex
@@
-\section{Univalent Subuniverses}~\label{sec:univalent}
+\section{Equivalence between \texorpdfstring{$\PiPlusLang$}{Pi} and \texorpdfstring{$\UFin$}{UFin}}~\label{sec:equivalence}
+
+First, we normalise $\PiPlusLang$ to $\PiHatLang$, and show that they are equivalent.
+
+Then we establish the completeness of $\PiHatLang$ with respect to $\UFin$.
+
+We use this to establish completeness for $\PiPlusLang$.
+
+Finally, we show how to translate $\PiLang$ programs to $\PiPlusLang$ programs.
 

### 2021-04-19 Pi+/paper/finite.tex
Some more outline

@@
 
-In this section, we describe the algebraic structure of the groupoid of finite types by characterising its paths.
+In this section, we describe the algebraic structure of the groupoid of finite types, and give a computable presentation
+for it.
 
@@
 
+\subsection{Symmetric Monoidal structure}
+
+$\UFin$ is a symmetric monoidal groupoid with unit and tensor given by
+
+\begin{definition}
+  \begin{align*}
+    I & \defeq \Fin[0] \\
+    X \otimes Y & \defeq X \sqcup Y
+  \end{align*}
+\end{definition}
+
+with unitors, associator, and braiding satisfying Mac Lane's triangle, pentagon, and hexagon coherence laws, and
+symmetry. It is the free symmetric monoidal groupoid on one generator.
+
 %%% Local Variables:

### 2021-05-02 Pi+/paper/univalent.tex
More text in section 3
Fixes #73

@@
 We work in Homotopy Type Theory, in particular, intensional Martin-L\"{o}f Type Theory, with a univalent universe, and
-Higher Inductive Types (HITs) for propositional truncations and set quotients. We recall a few basic facts.
+Higher Inductive Types (HITs) for propositional truncations and set quotients. We recall a few basic facts to
+familiarise the reader with the notation we use. For more details about basics of HoTT, we refer the reader to the HoTT
+book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.
 
@@
 
+In HoTT, the intensional identity type is the type of paths between two terms of the same type. Given two terms $x:A$
+and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the equality type between them. In book HoTT, the identity
+type is generated by reflexivity $\refl_{x}$, and the eliminator for the identity type is given by path induction or the
+$J$-rule. The identity type equips each type with the structure of an infinity groupoid, or a homotopy type.
+
+Functions between types are functors between groupoids. Given a function $f : A \to B$, the functorial action is given
+by
+
+\[
+  \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
+\]
+
+Type families are functions from a type to the universe, that is, an indexed family of groupoids. The $\term{transport}$
+operation gives the functorial action of paths in the indexing type, which is defined by path induction. If
+$P : A \to \UU$ is a type family, then for a path $x \id_{A} y$, we have
+
+\[
+  \transport{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
+\]
+
+From the topological viewpoint, a type family can also be seen as a fibration. For a type family $P : A \to \UU$ and a
+point $x : A$, $P(x)$ gives the fiber over $a$. For a path $p : x \id_{A} y$, $\transport{P}{p}$ gives the path lifting
+operation. The total space is given by $\dsum{x:A}{P(x)}$ and the first projection $\pi_1 : \dsum{x:A}{P(x)} \to A$ to
+the base space is the fibration. The lifting operation lifts paths in the base space to paths in the total space. If
+$p : x \id_{A} y$ is a path in the base space, and $u : P(x)$, we have
+
+\[
+  \term{lift}(u,p) : (x , u) \id_{\dsum{x:A}{P(x)}} (y , \tr{p}{u})
+\]
+
+where $\tr{p}{u}$ is shorthand for $\transport{P,p}(u)$.
+
+Further, using the groupoid structure of $A$, we can show that transport lifts paths to equivalences, we define
+$\term{transport-equiv}$.
+
+\[
+  \term{transport-equiv}(P) : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
+\]
+
+\begin{definition}[Univalent Fibration]
+  $P$ is a univalent type family (or simply a univalent fibration) if
+  $\term{transport-equiv}$ is an equivalence.
+\end{definition}
+
 \subsubsection{Univalence}
+
+The univalence principle characterises paths in the universe. It says that equivalent types are equal, or the following
+function is an equivalence.
+
+\[
+  \ua : A \id_{\UU} B \to A \eqv B
+\]
+
+Alternatively, one can say that the identity type family $\term{id} : \UU \to \UU$ is univalent.
 
@@
 \subsection{Univalent Subuniverses}
-
-\begin{definition}[Type families are fibrations]
-  A type family $P : A \to \UU$ gives a fibration $\fst : \dsum{x:A}{P(x)} \to A$ with total space $\dsum{x:A}{P(x)}$
-  and base space $A$ with a lifting property.
-\end{definition}
-
-\begin{definition}[$\term{transport-equiv}$]
-  Given a type family $P : A \to \UU$, we define
-  \[
-    \term{transport-equiv} : x \id_{A} y \to P(x) \eqv P(y)
-  \]
-\end{definition}
-
-\begin{definition}[Univalent Fibration]
-  $P$ is a univalent fibration if $\term{transport-equiv}$ is an equivalence.
-\end{definition}
 

### 2021-05-03 Pi+/paper/univalent.tex
Outline text in section 5
Add text in section 3.1.3

@@
 
+\begin{definition}[Propositional Truncation]
+  Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, or simply $\Trunc{A}$, is a higher inductive type
+  generated by the following constructors,
+  \begin{itemize}
+    \item an inclusion function $\trunc{\blank} : A \to \Trunc{A}$,
+    \item for each $x, y : \Trunc{A}$, a path $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$,
+  \end{itemize}
+  such that, given any type $B$ with
+  \begin{itemize}
+    \item a function $g : A \to B$,
+    \item for each $x, y : B$, a path $\term{trunc*}(x,y) : x \id_{B} y$,
+  \end{itemize}
+  there is a unique function $f : \Trunc{A} \to B$ such that,
+  \begin{itemize}
+    \item $f(\trunc{a}) \equiv g(a)$
+    \item for each $x, y : \Trunc{A}$, $\ap{f}{\term{trunc}(x,y)} \id_{B} \term{trunc*}(f(x),f(y))$.
+  \end{itemize}
+\end{definition}
+
+\begin{definition}[Set Quotient]
+  Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
+  higher inductive type generated by
+  \begin{itemize}
+    \item an inclusion function $q : A \to \quot{A}{R}$,
+    \item for each $x, y : A$ such that $R(x,y)$, a path $q(x) \id_{\quot{A}{R}} q(y)$,
+    \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
+  \end{itemize}
+  with an appropriate induction principle.
+\end{definition}
+
+We recall that quotients in HoTT are \emph{effective}, that is, if $R$ is an equivalence relation, we have
+$R(x,y) \eqv (q(x) \id_{\quot{A}{R}} q(y))$.
+
 \subsection{Univalent Subuniverses}
@@
 \begin{definition}[Universe]
-  A universe \`{a} la Tarski is given by a code $U : \UU$ and a decoding function $\El : U \to \UU$. If $\El$ is a
-  univalent fibration, $U$ is a univalent universe.
+  A universe \`{a} la Tarski is given by the following pieces of data,
+  \begin{itemize}
+    \item a code $U : \UU$,
+    \item a decoding function $\El : U \to \UU$.
+  \end{itemize}
+  If $\El$ is a univalent fibration, $U$ is a univalent universe.
 \end{definition}
@@
 \end{proposition}
+
+\begin{proof}
+
+\end{proof}
 

### 2021-05-03 Pi+/paper/univalent.tex
Add more in sec 3

@@
 Further, using the groupoid structure of $A$, we can show that transport lifts paths to equivalences, we define
-$\term{transport-equiv}$.
 
 \[
-  \term{transport-equiv}(P) : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
+  \tptEqv{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
 \]
@@
 \begin{definition}[Univalent Fibration]
-  $P$ is a univalent type family (or simply a univalent fibration) if
-  $\term{transport-equiv}$ is an equivalence.
+  $P$ is a univalent type family (or simply a univalent fibration) if $\tptEqv{P}$ is an equivalence.
 \end{definition}
@@
 
-The univalence principle characterises paths in the universe. It says that equivalent types are equal, or the following
-function is an equivalence.
+Voevodsky's \emph{univalence} principle characterises paths in the universe. It says that equivalent types are equal, or
+the following function is an equivalence.
 
@@
 \begin{proof}
-
+  Suppose $(U, \El) \defeq (\dsum{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
+  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
+  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
+  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
+  inverse follows by calculation using the appropriate computation rules.
 \end{proof}
 
-\begin{example}[$\Aut$]
-  \[
-    \Aut[T] \defeq T \eqv T
-  \]
-\end{example}
-
 \begin{example}[$\BAut$]
-  \[
-    \BAut[T] \defeq \Sub{T}
-  \]
+  The type of self-equivalences on a type $T$ is defined as $\Aut[T] \defeq T \eqv T$. The subuniverse of types that are
+  merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the delooping of $\Aut[T]$, as
+  justified by~\cref{lem:loop-deloop}.
 \end{example}
@@
 \begin{example}
-  \[
-    P(X) \defeq \SubP{X}{\Bool}
-  \]
+  The univalent subuniverse of types merely equal to $\Bool$ is given by the subtype $P(X) \defeq \SubP{X}{\Bool}$. This
+  can be used to give the denotational semantics of a 2-bit reversible
+  language~\cite{caretteReversibleProgramsUnivalent2018}.
 \end{example}
 
+\begin{definition}[$\Fin$]
+  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by cardinality. It is defined equivalently in
+  two different ways,
+  \begin{gather*}
+    \begin{aligned}
+      \Fin[n] & \defeq \dsum{k:\Nat}{k \leq n}
+    \end{aligned}
+    \begin{aligned}
+      & \Fin[0] \defeq \bot \\
+      & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
+    \end{aligned}
+  \end{gather*}
+  Note that $\Fin[n]$ is a $\hSet$, and we use the definitions interchangeably.
+\end{definition}
+
 \begin{example}
-  For any $n : \Nat$, we have
-  \[
-    P(X) \defeq \SubP{X}{\Fin[n]}
-  \]
-  \[
-    \UFin[n] \defeq \Sub{\Fin[n]}
-  \]
-  is the univalent subuniverse of types with cardinality $n$.
+  For any $n : \Nat$, we define the subtype of types merely equal to $\Fin[n]$ as $P(X) \defeq \SubP{X}{\Fin[n]}$. This
+  gives the univalent subuniverse $\UFin[n] \defeq \Sub{\Fin[n]}$ of finite types with cardinality equal to $n$.
 \end{example}
@@
 \begin{definition}[$\isFin$]
+  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n$.
   \[
@@
   \]
+  Note that the natural number $n$ need not be truncated, as justified below.
 \end{definition}
@@
 
+\begin{proof}
+  Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
+  enough to show that $n \id m$. Since $\Nat$ is a set, this is a proposition, so we can use the induction principle of
+  propositional truncation to eliminate to $n \id m$, applying it on $\phi$ and $\psi$ respectively. This gives us the
+  equalities $X \id \Fin[n]$ and $X \id \Fin[m]$, which gives us $\Fin[n] \id \Fin[m]$, from which $n \id m$ follows by
+  taking the first projection.
+\end{proof}
+
 \begin{example}
+  The univalent subuniverse of \emph{all finite types} is given by
   \[
-    \UFin \defeq \dsum{X:\UU}{\isFin[X]}
+    \UFin \defeq \dsum{X:\UU}{\isFin[X]}.
   \]
-  is the univalent subuniverse of all finite types.
+  We use this as the denotational semantics to interpret our reversible programming language $\PiPlusLang$.
 \end{example}
@@
 
-\begin{proposition}
+\begin{lemma}~\label{lem:loop-deloop}
   \[
@@
   \]
-\end{proposition}
+\end{lemma}
 

### 2021-05-06 Pi+/paper/univalent.tex
Add all proofs in section 3.2
Fixes #104

@@
 \begin{example}[$\BAut$]
-  The type of self-equivalences on a type $T$ is defined as $\Aut[T] \defeq T \eqv T$. The subuniverse of types that are
-  merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the delooping of $\Aut[T]$, as
-  justified by~\cref{lem:loop-deloop}.
+  The type of self-equivalences on a type $T$ is defined as $\Aut[T] \defeq T \eqv T$, which is an $\infty$-group. The
+  subuniverse of types that are merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the
+  classifying space of $\Aut[T]$, as justified by~\cref{lem:loop-deloop}. We write $T_{o} \defeq (T, \trunc{\refl_{T}})$
+  for the image of the inclusion of $T$.
 \end{example}
@@
 \begin{example}
-  The univalent subuniverse of types merely equal to $\Bool$ is given by the subtype $P(X) \defeq \SubP{X}{\Bool}$. This
-  can be used to give the denotational semantics of a 2-bit reversible
-  language~\cite{caretteReversibleProgramsUnivalent2018}.
+  The univalent subuniverse of types merely equal to $\Bool$ is given by $\BAut[\Bool]$. This can be used to give the
+  denotational semantics of a 2-bit reversible language~\cite{caretteReversibleProgramsUnivalent2018}.
 \end{example}
@@
 \begin{example}
-  For any $n : \Nat$, we define the subtype of types merely equal to $\Fin[n]$ as $P(X) \defeq \SubP{X}{\Fin[n]}$. This
-  gives the univalent subuniverse $\UFin[n] \defeq \Sub{\Fin[n]}$ of finite types with cardinality equal to $n$.
+  For any $n : \Nat$, we define the univalent subuniverse of types merely equal to $\Fin[n]$ as $\BAut[\Fin[n]]$. This
+  is the universe of finite sets with cardinality equal to $n$. We write $F_{n} \defeq (\Fin[n], \trunc{\refl})$ for the
+  image of the inclusion of $\Fin[n]$.
 \end{example}
@@
   equalities $X \id \Fin[n]$ and $X \id \Fin[m]$, which gives us $\Fin[n] \id \Fin[m]$, from which $n \id m$ follows by
-  taking the first projection.
+  applying the first projection.
 \end{proof}
@@
   \]
-  We use this as the denotational semantics to interpret our reversible programming language $\PiPlusLang$.
+  We write $F_{n} \defeq (n, \Fin[n], \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$. We use this as the
+  denotational semantics to interpret our reversible programming language $\PiPlusLang$.
 \end{example}
 
-We characterise the path space of subunivalent universes.
+We characterise the path space of univalent subuniverses.
 
@@
 
-\begin{proposition}
+\begin{proof}
+  We need to show that the equality type of $\BAut[T]$ is an $n$-type. Assume $X, Y : \BAut[T]$. Since $\BAut[T]$ is a
+  univalent subuniverse, we have $(X \id Y) \eqv (\fst(X) \eqv \fst(Y))$. Note that being an $n$-type is a proposition.
+  Since $T$ is an $n$-type, and $\fst(X)$ and $\fst(Y)$ are merely equal to $T$, they're also $n$-types. It follows that
+  $\fst(X) \eqv \fst(Y)$ is an $n$-type, and hence $X \id Y$ is an $n$-type.
+\end{proof}
+
+\begin{corollary}
   $\UFin$ is a 1-type, that is, a 1-groupoid, or has h-level 3.
-\end{proposition}
+\end{corollary}
 
@@
   \[
-    \Omega\BAut[T] \eqv \Aut[T]
+    \loopspace[\BAut[T],T_{0}] \eqv \Aut[T]
   \]
@@
 
-\begin{proposition}
+\begin{proof}
+  Since $\BAut[T]$ is a univalent universe, it follows that
+  \[
+    (T_{0} \id_{\BAut[T]} T_{0}) \eqv (\fst(T_{0}) \eqv \fst(T_{0})) \equiv (T \eqv T) \equiv \Aut[T].
+  \]
+\end{proof}
+
+\begin{corollary}
   For every $n:\Nat$,
   \[
-    \Omega\UFin[n] \eqv \Aut[\Fin[n]]
+    \loopspace[\UFin[n],\Fin[n]] \eqv \Aut[\Fin[n]]
   \]
-\end{proposition}
+\end{corollary}
 

### 2021-05-10 Pi+/paper/equivalence.tex
Outline text in section 5
Add more text in sec 5
Fixes #105

@@
 
-First, we normalise $\PiPlusLang$ to $\PiHatLang$, and show that they are equivalent.
+\subsection{$\PiLang$ to $\PiPlusLang$}
+
+We show how to translate $\PiLang$ programs to $\PiPlusLang$ programs.
+
+\subsection{$\PiPlusLang$ to $\PiHatLang$}
+
+We normalise $\PiPlusLang$ to $\PiHatLang$, and show that they are equivalent.
+
+\begin{definition}
+  \begin{align*}
+    \evalh_{0} & : \UPlus n \to \UHat n \\
+    \evalh_{1} & : (c : t_{1} \iso t_{2}) \to \evalh_{0}(t_{1}) \isoh \evalh_{0}(t_{2}) \\
+    \evalh_{2} & : (\alpha : c_{1} \Iso c_{2}) \to \evalh_{1}(c_{1}) \Isoh \evalh_{1}(c_{2}) \\
+  \end{align*}
+\end{definition}
+
+\begin{definition}
+  \begin{align*}
+    \quoteh_{0} & : \UHat n \to \UPlus n \\
+    \quoteh_{1} & : (p : X_{1} \isoh X_{2}) \to \quoteh_{0}(X_{1}) \iso \quoteh_{0}(X_{2}) \\
+    \quoteh_{2} & : (\alpha : p_{1} \Isoh p_{2}) \to \quoteh_{1}(p_{1}) \Iso \quoteh_{1}(p_{2}) \\
+  \end{align*}
+\end{definition}
+
+\subsection{$\PiHatLang$ to $\UFin$}
 
@@
 
+\begin{definition}
+  \begin{align*}
+    \evalt_{0} & : \UHat n \to \UFin[n] \\
+    \evalt_{1} & : (c : t_{1} \isoh t_{2}) \to \evalh_{0}(t_{1}) \id \evalh_{0}(t_{2}) \\
+    \evalt_{2} & : (\alpha : c_{1} \Isoh c_{2}) \to \evalt_{1}(c_{1}) \id \evalt_{1}(c_{2}) \\
+  \end{align*}
+\end{definition}
+
+\begin{definition}
+  \begin{align*}
+    \quotet_{0} & : \UFin[n] \to \UHat n \\
+    \quotet_{1} & : (p : X_{1} \id X_{2}) \to \quoteh_{0}(X_{1}) \isoh \quoteh_{0}(X_{2}) \\
+    \quotet_{2} & : (\alpha : p_{1} \id p_{2}) \to \quoteh_{1}(p_{1}) \Isoh \quotet_{1}(p_{2}) \\
+  \end{align*}
+\end{definition}
+
 We use this to establish completeness for $\PiPlusLang$.
 
-Finally, we show how to translate $\PiLang$ programs to $\PiPlusLang$ programs.
+\begin{definition}
+  \begin{align*}
+    \evalt_{0} & : \UPlus n \to \UFin[n] \\
+    \evalt_{1} & : (c : t_{1} \iso t_{2}) \to \evalt_{0}(t_{1}) \id \evalt_{0}(t_{2}) \\
+    \evalt_{2} & : (\alpha : c_{1} \Iso c_{2}) \to \evalt_{1}(c_{1}) \id \evalt_{1}(c_{2}) \\
+  \end{align*}
+\end{definition}
+
+\begin{definition}
+  \begin{align*}
+    \quotet_{0} & : \UFin[n] \to U n \\
+    \quotet_{1} & : (p : X_{1} \id X_{2}) \to \quotet_{0}(X_{1}) \iso \quotet_{0}(X_{2}) \\
+    \quotet_{2} & : (\alpha : p_{1} \id p_{2}) \to \quotet_{1}(p_{1}) \Iso \quotet_{1}(p_{2}) \\
+  \end{align*}
+\end{definition}
+
+\begin{theorem}
+  ${\quotet}/{\evalt}$ gives a symmetric monoidal equivalence between $\UPlus$
+  and $\UFin$.
+\end{theorem}
 

### 2021-05-19 Pi+/paper/finite.tex
Add some explanation about Lehmer codes
Progress on #108 and #109

@@
 
-In this section, we describe the algebraic structure of the groupoid of finite types, and give a computable presentation
-for it.
+In this section, we describe the algebraic structure of the groupoid of finite
+types, and give a computable presentation for it. To do so, we will characterise
+the automorphisms on finite sets of cardinality $n$, and show them to be
+equivalent to the symmetric group $\Sn$, via the Coxeter presentation. We will
+do that in two steps,
+in~\cref*{subsec:permutations,subsec:lehmer,subsec:symmetric}.
 
-\subsection{Permutations}
+%% Coxeter presentation of $\Sn$ $\eqv$ $\Lehmer[n]$ $\eqv$ $\Aut[\Fin[n]]$
 
-In the previous~\cref{sec:univalent}, we established that paths in $\UFin$ are equivalent to families of automorphisms
-of $\Fin{n}$ for every $n:\Nat$. These are permutations on finite sets of size $n$. In the following sections, we will
-characterise these permutations, going through a number of intermediate step.
+\subsection{Permutations}~\label{subsec:permutations}
 
-\jk{I don't like the wording}
+In the previous~\cref{sec:univalent}, we established that paths in $\UFin$ are
+equivalent to families of automorphisms of $\Fin{n}$ for every $n:\Nat$, that
+is, bijections on finite sets of size $n$. This is the extensional view of
+permutations. In the following sections, we will characterise these
+permutations, going through two intermediate steps.
 
-\subsubsection{Lehmer codes}
+\vc{This is obvious, maybe add something more here.}
 
-There are $\fac{n}$ permutations on a set of $n$ elements. Thus, if we construct a type with $\fac{n}$ inhabitants,
-there should be a bijection of that type with the type of permutations $\Aut[\Fin[n]]$. To construct this bijection, we
-will borrow the idea of Lehmer codes~\cite{lehmerTeachingCombinatorialTricks1960a} from Combinatorial
-Analysis~\cite{bellmanCombinatorialAnalysis1960}. The type family $\Lehmer$ over $\Nat$ is defined as follows.
+\subsection{Lehmer codes}~\label{subsec:lehmer}
 
-\todo{Understand the proof first and then revisit this paragraph}
+From grade school combinatorics, we know that there are $\fac{n}$ permutations
+on a finite set with $n$ elements. The factorial function is defined by
+recursion on natural numbers. However, now, for every $n$, we want to produce a
+type, which is a finite set, with cardinality $\fac{n}$. And, to characterise
+$\Aut[\Fin[n]]$, we further need to construct a bijection between this type and
+$\Aut[\Fin[n]]$.
+
+First, let's define this type with $\fac{n}$ elements, we name this type family
+$\Lehmer : \Nat \to \UU$, which is defined by recursion on $\Nat$ as follows.
+This is the obvious definition of factorials by recursion, but categorified from
+natural numbers to sets.
 
@@
   \begin{align*}
-         & \Lehmer : \Nat \to \UU                             \\
-         & \Lehmer[0] = \unit                                 \\
-         & \Lehmer[\suc[n]] = \Fin[\suc[n]] \times \Lehmer[n]
+    \Lehmer[0]       & \defeq \unit                           \\
+    \Lehmer[\suc[n]] & \defeq \Fin[\suc[n]] \times \Lehmer[n]
   \end{align*}
@@
 
-Now we construct the equivalence.
+\todo{Subexcedant sequences and factorial definitions are equivalent, explain
+  this!}
+
+The name Lehmer comes from Lehmer
+codes~\cite{lehmerTeachingCombinatorialTricks1960a} which are known in
+Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960}. There are many
+ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer
+codes are a particularly convenient way to represent permutations on a
+computer,~\review{they are compact and have exactly the right cardinality.
+  $\Lehmer[n]$ is a $n+1$-element tuple, where the position $k \leq n$ has an
+  element of $\Fin[k]$. The 0-th position is trivial, so we ignore it, and in
+  both the example below and the Agda proof, consider only the remaining
+  $n$-element tuple.}
+
+\vc{This is just the classical algorithm to explain the example, not the actual
+  type-theoretic proof.}
+
+Suppose we have a permutation $p$ on an $n$-element set
+$\{\el{0}, \el{1}, \el{2}, \el{3}, \el{4}\}$, we encode it as follows.
+$\Lehmer[n]$ is a $n$-element tuple. At position $k$, we put the number of
+inversions of the element $\el{k}$ in $p$, i.e. the number of elements smaller
+than $\el{k}$ occurring after $\el{k}$.
+
+As an example, consider the following tabulated presentation of the permutation:
+
+\todo{fix this figure}
+
+\[
+  p =
+  \begin{array}{ccccccccccccccc}
+    | & 0      & | & 1      & | & 2      & | & 3      & | & 4      & | \\
+    \hline                                                             \\
+    | & \el{2} & | & \el{0} & | & \el{1} & | & \el{4} & | & \el{3} & | \\
+    \hline                                                             \\
+  \end{array}
+\]
+
+%  0 1 2 3 4
+% -----------
+% |2|0|1|4|3|
+% -----------
+
+The element $\el{0}$ has 0 inversions, because there are no elements smaller
+than $\el{0}$ occurring after it. In fact, there can be no elements smaller than
+$\el{0}$ at all, so the type at the first position of the Lehmer code tuple is
+$\unit$.
+
+The element $\el{1}$ has 0 inversions as well, since elements occurring after it
+in the permutation are $\el{4}$ and $\el{2}$. There is only one different case,
+if $\el{1}$ appeared before $\el{0}$, it would have 1 inversion. This is why the
+type of the second component of the Lehmer code is $\Fin[2]$.
+
+The element $\el{2}$ has 2 inversions, because both $\el{0}$ and $\el{1}$ occur
+after it in the permutation. The element $\el{3}$ occurs as the last one, so it
+has 0 inversions. The element $\el{4}$ has 1 inversion, with the element
+$\el{3}$.
+
+Thus, the Lehmer code for the permutation $p$ is the 5-tuple
+$l = (0, 0, 2, 0, 1)$.
+
+To reconstruct the tabulated presentation of the permutation from the Lehmer
+code, we perform an algorithm similar to \emph{insertion sort}. Starting from
+the left-most position of the tuple $l$, we'll read the value $v$, insert the
+new element at the end of the newly created list, and shift it backward $v$
+places.
+
+\begin{center}
+  \begin{tabular}{c|p{0.75\linewidth}}
+    (0, 0, 2, 0, 1)               & We start from an empty list $[]$                                                                 \\
+    (\highlight{{0}}, 0, 2, 0, 1) & We read 0 as the left-most value from $l$. Thus, we append the element $\el{0}$ to our
+                                    list, getting $[\el{0}]$. The element is shifted $0$ places, so it remains in the
+                                    same place.                                                                                                                      \\
+    (0, \highlight{{0}}, 2, 0, 1) & Then, similarly, we read another 0 for the element $\el{1}$, append it to the
+                                    list getting $[\el{0}, \el{1}]$, and don't shift it either.                                                                      \\
+    (0, 0, \highlight{{2}}, 0, 1) & We read 2 for the next the element $\el{2}$ - we append $\el{2}$ to our list, getting
+                                    $[\el{0}, \el{1}, \el{2}]$, and shift it 2 places right, which results in a list $[\el{0}, \el{2}, \el{1}]$
+                                    \todo{Typeset it nicely, with arrows showing the shifting}.                                                                      \\
+    (0, 0, 2, \highlight{{0}}, 1) & Then we read 0 - appending $\el{3}$ and not shifting, getting $[\el{0}, \el{2}, \el{1}, \el{3}]$ \\
+    (0, 0, 2, 0, \highlight{{1}}) & Finally, reading 1 for element $\el{4}$ - appending $\el{4}$ to the list and shifting it
+                                    one place right results in the final list $[\el{0}, \el{2}, \el{1}, \el{4}, \el{3}]$                                             \\
+  \end{tabular}
+\end{center}
+\todo{figure}
+
+Using this Lehmer encoding algorithm, we can now construct the equivalence
+between these types.
 
@@
 
-\subsubsection{Symmetric groups}
+\begin{proof}
+  We have to turn this algorithm into a constructive proof in type theory where
+  we only use functions and recursion. \todo{Describe the agda code.}
+\end{proof}
 
-There is an obvious group structure on $\Aut[\Fin[n]]$ given by identity, composition, and inverse. This is the
-symmetric group $S_n$ on $n$ symbols. In the rest of the section we will construct a convenient presentation of this
+\subsection{Symmetric groups}~\label{subsec:symmetric}
+
+There is an obvious group structure on $\Aut[\Fin[n]]$ given by identity,
+composition, and inverse. This is the symmetric group $S_n$ on $n$ symbols. In
+the rest of the section we will construct a convenient presentation of this
 group.
@@
 
+\begin{theorem}
+  For all $n : \Nat$,
+  \[
+    Sn(n) \eqv \Lehmer[n]
+  \]
+\end{theorem}
+
 \subsection{Symmetric Monoidal structure}

### 2021-05-24 Pi+/paper/finite.tex
Add notes and comments

@@
 
-In this section, we describe the algebraic structure of the groupoid of finite
-types, and give a computable presentation for it. To do so, we will characterise
-the automorphisms on finite sets of cardinality $n$, and show them to be
-equivalent to the symmetric group $\Sn$, via the Coxeter presentation. We will
-do that in two steps,
+In this section, we describe \review{the algebraic structure} of the groupoid of
+finite types, and give \review{a computable presentation} for it.
+
+\vc{The groupoid of finite types is the free symmetric monoidal groupoid on one
+  generator. This can be presented as an algebraic 2-theory, which is our syntax
+  for $\PiHatLang$. Vertical categorification of natural numbers as a free
+  commutative monoid. See groupoidification.}
+
+\todo{Check Brent Yorgey's thesis?}
+
+To do so, we will characterise the automorphisms on finite sets of cardinality
+$n$, and show them to be equivalent to the symmetric group $\Sn$, via the
+Coxeter presentation. We will do that in two steps,
 in~\cref*{subsec:permutations,subsec:lehmer,subsec:symmetric}.
@@
   \begin{align*}
-    I           & \defeq \Fin[0]    \\
+    I           & \defeq F_{0}    \\
     X \otimes Y & \defeq X \sqcup Y
@@
 %%% TeX-master: "main"
+%%% fill-column: 120
 %%% End:

### 2021-05-26 Pi+/paper/reversible.tex

--- 2021-01-27 Pi+/paper/univalent.tex
@@
-\section{Univalent Subuniverses}~\label{sec:univalent}
+\section{Reversible Programming Languages}~\label{sec:reversible}
+
+\todo{Not the right title.}
+
+\note{This section should explain the main technical parts of the paper
+  informally, without using any technology. Use an example, such as, a
+  reversible language with $\leq 5$ bits, and examples of permutations and
+  transpositions, and when they're equal.}
+
+\note{This might be followed by a section which explains the syntax of Pi.}
 

### 2021-05-26 Pi+/paper/univalent.tex
Add some explanation about Lehmer codes
Progress on #108 and #109

@@
 
+\note{It is important that we work in HoTT using it as a metatheory, we care about proof relevance, constructivity, and
+  working with higher-dimensional algebraic structures, such as groupoids. Our results are also formalised using general
+  tools and principles from HoTT, and computer checked using a proof assistant, using Agda and the HoTT-Agda library.}
+
 \subsubsection{Identity Types}
+
+\vc{I'm explaining everything starting from the categorical language. We do care more about the groupoid structure of
+  types, and how to encode groupoids using types, since that's what we use in~\cref{sec:finite}.}
 
@@
 $J$-rule. The identity type equips each type with the structure of an infinity groupoid, or a homotopy type.
+
+\todo{Say why it is a groupoid by listing some groupoid laws.}
 
@@
 
+\todo{Why is the topological viewpoint important, or saying fibrations? Explain the intuition using classifying spaces,
+  maybe draw some diagrams.}
+
 From the topological viewpoint, a type family can also be seen as a fibration. For a type family $P : A \to \UU$ and a
 point $x : A$, $P(x)$ gives the fiber over $a$. For a path $p : x \id_{A} y$, $\transport{P}{p}$ gives the path lifting
-operation. The total space is given by $\dsum{x:A}{P(x)}$ and the first projection $\pi_1 : \dsum{x:A}{P(x)} \to A$ to
+operation. The total space is given by $\dsum{x:A}{P(x)}$ and the first projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ to
 the base space is the fibration. The lifting operation lifts paths in the base space to paths in the total space. If
@@
 
+\todo{Explain motivation.}
+
 \begin{definition}[Univalent Fibration]
@@
 
+\todo{Revise.}
+
 Alternatively, one can say that the identity type family $\term{id} : \UU \to \UU$ is univalent.
@@
 \subsubsection{Higher Inductive Types}
+
+\todo{Explain homotopy types, $\hProp$, $\hSet$, etc.}
+
+\vc{These are placeholder definitions, need informal explanations and references to the book.}
 
@@
   \end{itemize}
-  with an appropriate induction principle.
+  with \review{an appropriate induction principle.}
 \end{definition}
@@
 \subsection{Univalent Subuniverses}
+
+\todo{Explain motivation. Starting from a univalent universe which classifies all types, we want to define a subuniverse
+  which classifies only certain types, for example, types that satisfy some property that we want.}
 
@@
     \begin{aligned}
-      \Fin[n] & \defeq \dsum{k:\Nat}{k \leq n}
+      \Fin[n] & \defeq \dsum{k:\Nat}{k < n}
     \end{aligned}
@@
 
-We characterise the path space of univalent subuniverses.
+\review{We characterise the path space of univalent subuniverses.}
 
@@
   \[
-    \loopspace[\UFin[n],\Fin[n]] \eqv \Aut[\Fin[n]]
+    \loopspace[\UFin[n],F_{n}] \eqv \Aut[\Fin[n]]
   \]

### 2021-06-04 Pi+/paper/reversible.tex
Add some figures for section 2
Notes on section 2

@@
 
+\note{Motivation: There are two reversible circuits which describe the following permutation. They can be shown to be
+  equal using the 2-combinators.}
+
+\[
+  \begin{tikzpicture}
+    \begin{knot}[clip width=5]
+      \filldraw (0,5) circle (2pt) node[above] {0};
+      \filldraw (1,5) circle (2pt) node[above] {1};
+      \filldraw (2,5) circle (2pt) node[above] {2};
+      \filldraw (3,5) circle (2pt) node[above] {3};
+      \filldraw (4,5) circle (2pt) node[above] {4};
+      \filldraw (0,0) circle (2pt) node[below] {1};
+      \filldraw (1,0) circle (2pt) node[below] {4};
+      \filldraw (2,0) circle (2pt) node[below] {0};
+      \filldraw (3,0) circle (2pt) node[below] {3};
+      \filldraw (4,0) circle (2pt) node[below] {2};
+      \strand (0,5) .. controls (0.5,0.5) and (1.5,3.5) .. (2,0);
+      \strand (1,5) .. controls (0.75,0.5) and (0.25,3.5) .. (0,0);
+      \strand (2,5) .. controls (2.5,2.5) and (3.5,1.5) .. (4,0);
+      \strand (3,5) .. controls (4.5,2.5) and (4,1.5) .. (3,0);
+      \strand (4,5) .. controls (3.5,2.5) and (1.5,2.5) .. (1,0);
+      \flipcrossings{4,5};
+    \end{knot}
+  \end{tikzpicture}
+\]
+
+\note{Example: We reduce $\mathsf{swap} : 2 + 2 \leftrightarrow 2 + 2$ to a sequence of adjacent swaps. This is an
+  example of a translation from $\PiPlusLang$ to $\PiHatLang$.}
+
+\[
+  \begin{tikzpicture}
+    \begin{knot}[clip width=4]
+      \filldraw (0,4) circle (2pt) node[above] {0};
+      \filldraw (1,4) circle (2pt) node[above] {1};
+      \filldraw (2,4) circle (2pt) node[above] {2};
+      \filldraw (3,4) circle (2pt) node[above] {3};
+      \filldraw (0,0) circle (2pt) node[below] {2};
+      \filldraw (1,0) circle (2pt) node[below] {3};
+      \filldraw (2,0) circle (2pt) node[below] {0};
+      \filldraw (3,0) circle (2pt) node[below] {1};
+      \strand (0,4) .. controls (0.5,1.5) and (1.5,2.5) .. (2,0);
+      \strand (1,4) .. controls (1.5,1.5) and (2.5,2.5) .. (3,0);
+      \strand (2,4) .. controls (1.5,1.5) and (1.5,2.5) .. (0,0);
+      \strand (3,4) .. controls (2.5,1.5) and (2.5,2.5) .. (1,0);
+    \end{knot}
+  \end{tikzpicture}
+\]
+
+\begin{align*}
+  \begin{tikzpicture}
+    \begin{knot}[clip width=4]
+      \filldraw (0,4) circle (2pt) node[above] {0};
+      \filldraw (1,4) circle (2pt) node[above] {1};
+      \filldraw (2,4) circle (2pt) node[above] {2};
+      \filldraw (3,4) circle (2pt) node[above] {3};
+      \filldraw (0,0) circle (2pt) node[below] {0};
+      \filldraw (1,0) circle (2pt) node[below] {2};
+      \filldraw (2,0) circle (2pt) node[below] {1};
+      \filldraw (3,0) circle (2pt) node[below] {3};
+      \strand (0,4) to (0,0);
+      \strand (1,4) .. controls (0.5,2) and (2.5,2) .. (2,0);
+      \strand (2,4) .. controls (2.5,2) and (0.5,2) .. (1,0);
+      \strand (3,4) to (3,0);
+    \end{knot}
+  \end{tikzpicture}
+  &&
+    \begin{tikzpicture}
+      \begin{knot}[clip width=4]
+        \filldraw (0,4) circle (2pt) node[above] {0};
+        \filldraw (1,4) circle (2pt) node[above] {2};
+        \filldraw (2,4) circle (2pt) node[above] {1};
+        \filldraw (3,4) circle (2pt) node[above] {3};
+        \filldraw (0,0) circle (2pt) node[below] {2};
+        \filldraw (1,0) circle (2pt) node[below] {0};
+        \filldraw (2,0) circle (2pt) node[below] {1};
+        \filldraw (3,0) circle (2pt) node[below] {3};
+        \strand (0,4) .. controls (-0.5,2) and (1.5,2) .. (1,0);
+        \strand (1,4) .. controls (1.5,2) and (-0.5,2) .. (0,0);
+        \strand (2,4) to (2,0);
+        \strand (3,4) to (3,0);
+      \end{knot}
+    \end{tikzpicture}
+  \\
+  \begin{tikzpicture}
+    \begin{knot}[clip width=4]
+      \filldraw (0,4) circle (2pt) node[above] {2};
+      \filldraw (1,4) circle (2pt) node[above] {0};
+      \filldraw (2,4) circle (2pt) node[above] {1};
+      \filldraw (3,4) circle (2pt) node[above] {3};
+      \filldraw (0,0) circle (2pt) node[below] {2};
+      \filldraw (1,0) circle (2pt) node[below] {0};
+      \filldraw (2,0) circle (2pt) node[below] {3};
+      \filldraw (3,0) circle (2pt) node[below] {1};
+      \strand (0,4) to (0,0);
+      \strand (1,4) to (1,0);
+      \strand (2,4) .. controls (1.5,2) and (3.5,2) .. (3,0);
+      \strand (3,4) .. controls (3.5,2) and (1.5,2) .. (2,0);
+    \end{knot}
+  \end{tikzpicture}
+  &&
+    \begin{tikzpicture}
+      \begin{knot}[clip width=4]
+        \filldraw (0,4) circle (2pt) node[above] {2};
+        \filldraw (1,4) circle (2pt) node[above] {0};
+        \filldraw (2,4) circle (2pt) node[above] {3};
+        \filldraw (3,4) circle (2pt) node[above] {1};
+        \filldraw (0,0) circle (2pt) node[below] {2};
+        \filldraw (1,0) circle (2pt) node[below] {3};
+        \filldraw (2,0) circle (2pt) node[below] {0};
+        \filldraw (3,0) circle (2pt) node[below] {1};
+        \strand (0,4) to (0,0);
+        \strand (1,4) .. controls (0.5,2) and (2.5,2) .. (2,0);
+        \strand (2,4) .. controls (2.5,2) and (0.5,2) .. (1,0);
+        \strand (3,4) to (3,0);
+      \end{knot}
+    \end{tikzpicture}
+\end{align*}
+
 \note{This might be followed by a section which explains the syntax of Pi.}

### 2021-06-05 Pi+/paper/finite.tex
Add some text in section 4.3

@@
 
+\vc{this is just a rough draft for now}
+
 \todo{Reference T-algebra presentations as coequalisers (Mac Lane 6.7)}
@@
 \begin{definition}
-  A presentation of a group $G$ is a type $FR$ such that \ldots is a coequaliser.
+  Let $A$ be a type, and $\List(A)$ the free monoid on $A$. The free group $F(A)$ on $A$ is the set-quotient of
+  $\List(A + A)$ by the congruence closure of the relation $(\inl(a) \cons \inr(a) \cons \nil) \sim \nil$.
 \end{definition}
 
-\vc{this is just a rough draft for now}
+\begin{proposition}
+  Let $\eta : A \to F(A)$ be given by $a \mapsto \inl(a) \cons \nil$. Given any group $G$ with a map $f : A \to G$,
+  there is a canonical lifting of $f$ to a group homomorphism $\extend{f} : F(A) \to G$, such that
+  $\extend{f} \comp \eta_{A} \htpy f$, that is, the type of group homomorphisms $h : \Grp(F(A),G)$ satisfying
+  $h \comp \eta_{A} \htpy f$ is contractible. Equivalently, composition by $\eta_{A}$ is an equivalence
+  $\Grp(F(A),G) \eqv (A \to G)$.
+\end{proposition}
+
+\begin{definition}
+  Let $A$ be a type and $R : A \to A \to \UU$ a binary relation on $A$. The group $G$ presented by
+  $\langle A ; R \rangle$ is given by the set-quotient of the free group $F(A)$ by the normal closure $N_{R}$, or
+  equivalently, as the coequaliser
+  \[\begin{tikzcd}
+      FR && FA && G
+      \arrow[shift right=2, from=1-1, to=1-3]
+      \arrow[shift left=2, from=1-1, to=1-3]
+      \arrow[two heads, from=1-3, to=1-5]
+    \end{tikzcd}\]
+\end{definition}
+
+\todo{Universal property}
 
@@
 
-We now introduce a presentation of the symmetric group.
+We now introduce a presentation of the symmetric group $\Sn$.
 
@@
   \[
-    Sn(n) \eqv \Lehmer[n]
+    \Sn \eqv \Lehmer[n]
   \]

### 2021-06-09 Pi+/paper/finite.tex
Add text in section 4.3
Fixes #114

@@
 
-\begin{definition}
-  Here we'll define a type family $FT n$ to be a free group generated by a type $T n$ with the following constructors
+\subsubsection{Coxeter Relations}
+
+We define a binary relation on $\List(\Fin[n])$.
+
+\begin{definition}[Coxeter relation]
   \begin{align*}
-     & cancel : i : \Fin[\suc[n]] \to T                                         \\
-     & swap : (i : \Fin[\suc[n]]) \to (j : \Fin[\suc[n]]) \to (i + 1 < j) \to T \\
-     & braid : (i : \Fin[n]) \to T                                              \\
+    & \mathsf{cancel} : (n \cons n \cons \nil) \cx \nil \\
+    & \mathsf{swap} : (\suc[k] < n) \to (n \cons k \cons \nil) \cx (k \cons n \cons \nil) \\
+    & \mathsf{braid} : (\suc[n] \cons k \cons \suc[n] \cons \nil) \cx (n \cons \suc[k] \cons n \cons \nil)
   \end{align*}
@@
 
-\begin{definition}[Adjacent transposition]
-  \begin{align*}
-    transpose : (n : \Nat) \to (k : \Fin[\suc[n]]) \to \Aut[\Fin[\suc[n]]]
-  \end{align*}
-  by double induction on $n$ and $k$, where
-  \begin{align*}
-    transpose (n) (0)             & = \lambda
-    0 \to 1;
-    1 \to 0;
-    m \to m
-    \\
-    transpose (\suc[n]) (\suc[k]) & = \lambda
-    0 \to 0;
-    S m \to S ((transpose (n)(k)) m)
-  \end{align*}
+The type $\Sn$ is defined as the set-quotient of $\List(\Fin[n])$ by the congruence closure of $\cx$.
+
+\begin{definition}[$\Sn$]
+  \(\Sn \defeq \quot{\List(\Fin[n])}{\cx^{\ast}}\)
 \end{definition}
@@
 \begin{proposition}
-  $FR$ is a presentation of the symmetrc group, with $f$ and $g$ as follows:
-  \begin{align*}
-    f (inr (cancel (i))) & = inr (i :: i :: [])                     \\
-    f (inl (cancel (i))) & = inl (i :: i :: [])                     \\
-    f (inr (swap (i,j))) & = i :: j :: i :: j :: []                 \\
-    f (braid (i))        & = i :: S i :: i :: S i :: i :: S i :: []
-  \end{align*}
-  and
-  \begin{align*}
-    g (inl (\_)) & = inl ([]) \\
-    g (inr (\_)) & = inr ([])
-  \end{align*}
+  There is a group structure on $\Sn$, where the identity element is $\nil$, multiplication is given by list append, and
+  inverse is given by list reversal.
 \end{proposition}
 
-We now introduce a presentation of the symmetric group $\Sn$.
+\begin{proposition}
+  $\Sn$ is equivalent to the generated group given by the normal closure of $\cx$ extended to $\List(\Fin[n] + \Fin[n])$
+  along the codiagonal map $[\term{id},\term{id}] : A + A \to A$.
+\end{proposition}
+
+We will now show how to generate a Lehmer code from a word in $\Sn$ and back.
+
+\todo{immersion}
 

### 2021-06-09 Pi+/paper/finite.tex
Add some text and figures in section 4.4

@@
 
-$\UFin$ is a symmetric monoidal groupoid with unit and tensor given by
+We describe the symmetric monoidal structure of the groupoid $\UFin$.
+
+First, we observe a few equivalences.
+
+\begin{proposition}
+  For any $n, m : \Nat$,
+  \begin{align*}
+    \Fin[0]                & \eqv \bot \\
+    \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m] \\
+  \end{align*}
+  and for any types $X, Y, Z$,
+  \begin{align*}
+    \bot \sqcup X          & \eqv X \\
+    X \sqcup \bot          & \eqv X \\
+    (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
+    X \sqcup Y             & \eqv Y \sqcup Y \\
+  \end{align*}
+\end{proposition}
+
+We lift these equivalences to $\UFin$ giving it the (additive) symmetric monoidal structure $(I, \oplus)$, with natural
+isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the symmetry isomorphism $\mathcal{B}_{X,Y}$. Note that
+types in $\UFin$ are $\hSet$s since they're merely equivalent to $\Fin[n]$ for some $n : \Nat$.
 
@@
     I           & \defeq F_{0}    \\
-    X \otimes Y & \defeq X \sqcup Y
+    X \oplus Y & \defeq X \sqcup Y \\
+    \lambda_{X} & : I \oplus X \eqv X \\
+    \rho_{X} & : X \oplus I \eqv X \\
+    \alpha_{X,Y,Z} & : (X \oplus Y) \oplus Z \eqv X \oplus (Y \oplus Z) \\
+    \mathcal{B}_{X,Y} & : X \oplus Y \eqv Y \oplus X
   \end{align*}
@@
 
-with unitors, associator, and braiding satisfying Mac Lane's triangle, pentagon, and hexagon coherence laws, and
-symmetry. It is the free symmetric monoidal groupoid on one generator.
+These isomorphisms satisfy the Mac Lane coherence conditions for symmetric monoidal categories, that is, the triangle,
+pentagon, and hexagon identities, and the syllepsis of the braiding, upto 2-paths in $\UFin$.
+
+\begin{proposition}
+  % https://q.uiver.app/?q=WzAsNCxbMCwwLCIoWCBcXG9wbHVzIEkpIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyAoSSBcXG9wbHVzIFkpIl0sWzEsMSwiWCBcXG9wbHVzIFkiXSxbMCwxXSxbMCwxLCJcXGFscGhhX3tYLEksWX0iXSxbMCwyLCJcXHJob197WH0gXFxvcGx1cyAxX3tZfSIsMl0sWzEsMiwiMV97WH0gXFxvcGx1cyBcXGxhbWJkYV97WX0iXSxbNSw2LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+  \[\begin{tikzcd}
+      {(X \oplus I) \oplus Y} && {X \oplus (I \oplus Y)} \\
+      {} & {X \oplus Y}
+      \arrow["{\alpha_{X,I,Y}}", from=1-1, to=1-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\rho_{X} \oplus 1_{Y}}"', from=1-1, to=2-2]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{1_{X} \oplus \lambda_{Y}}", from=1-3, to=2-2]
+      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
+    \end{tikzcd}\]
+  % https://q.uiver.app/?q=WzAsNSxbMCwxLCIoKFcgXFxvcGx1cyBYKSBcXG9wbHVzIFkpIFxcb3BsdXMgWiJdLFsxLDAsIihXIFxcb3BsdXMgWCkgXFxvcGx1cyAoWSBcXG9wbHVzIFopIl0sWzIsMSwiVyBcXG9wbHVzIChYIFxcb3BsdXMgKFkgXFxvcGx1cyBaKSkiXSxbMiwzLCJXIFxcb3BsdXMgKChYIFxcb3BsdXMgWSkgXFxvcGx1cyBaKSJdLFswLDMsIihXIFxcb3BsdXMgKFggXFxvcGx1cyBZKSkgXFxvcGx1cyBaIl0sWzAsMSwiXFxhbHBoYV97VyBcXG9wbHVzIFgsIFksIFp9Il0sWzEsMiwiXFxhbHBoYV97VyxYLFkgXFxvcGx1cyBafSJdLFszLDIsIjFfe1d9IFxcb3BsdXMgXFxhbHBoYV97WCxZLFp9IiwyXSxbMCw0LCJcXGFscGhhX3tXLFgsWX0gXFxvcGx1cyAxX3tafSIsMl0sWzQsMywiXFxhbHBoYV97VyxYIFxcb3BsdXMgWSxafSIsMl0sWzAsMiwiXFxpZCIsMSx7Im9mZnNldCI6NSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+  \[\begin{tikzcd}
+      & {(W \oplus X) \oplus (Y \oplus Z)} \\
+      {((W \oplus X) \oplus Y) \oplus Z} && {W \oplus (X \oplus (Y \oplus Z))} \\
+      \\
+      {(W \oplus (X \oplus Y)) \oplus Z} && {W \oplus ((X \oplus Y) \oplus Z)}
+      \arrow["{\alpha_{W \oplus X, Y, Z}}", from=2-1, to=1-2]
+      \arrow["{\alpha_{W,X,Y \oplus Z}}", from=1-2, to=2-3]
+      \arrow["{1_{W} \oplus \alpha_{X,Y,Z}}"', from=4-3, to=2-3]
+      \arrow["{\alpha_{W,X,Y} \oplus 1_{Z}}"', from=2-1, to=4-1]
+      \arrow["{\alpha_{W,X \oplus Y,Z}}"', from=4-1, to=4-3]
+      \arrow["\id"{description}, shift right=5, draw=none, from=2-1, to=2-3]
+    \end{tikzcd}\]
+  % https://q.uiver.app/?q=WzAsNixbMSwwLCJYIFxcb3BsdXMgKFkgXFxvcGx1cyBaKSJdLFswLDEsIihYIFxcb3BsdXMgWSkgXFxvcGx1cyBaIl0sWzAsMiwiKFkgXFxvcGx1cyBYKSBcXG9wbHVzIFoiXSxbMSwzLCJZIFxcb3BsdXMgKFggXFxvcGx1cyBaKSJdLFsyLDIsIlkgXFxvcGx1cyAoWiBcXG9wbHVzIFgpIl0sWzIsMSwiKFkgXFxvcGx1cyBaKSBcXG9wbHVzIFgiXSxbMSwwLCJcXGFscGhhX3tYLFksWn0iXSxbMSwyLCJcXG1hdGhjYWx7Qn1fe1gsWX0gXFxvcGx1cyAxX3tafSIsMl0sWzIsMywiXFxhbHBoYV97WSxYLFp9IiwyXSxbMyw0LCIxX3tZfSBcXG9wbHVzIFxcbWF0aGNhbHtCfV97WCxafSIsMl0sWzUsNCwiXFxhbHBoYV97WSxaLFh9Il0sWzAsNSwiXFxtYXRoY2Fse0J9X3tYLFkgXFxvcGx1cyBafSJdLFs3LDEwLCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+  \[\begin{tikzcd}
+      & {X \oplus (Y \oplus Z)} \\
+      {(X \oplus Y) \oplus Z} && {(Y \oplus Z) \oplus X} \\
+      {(Y \oplus X) \oplus Z} && {Y \oplus (Z \oplus X)} \\
+      & {Y \oplus (X \oplus Z)}
+      \arrow["{\alpha_{X,Y,Z}}", from=2-1, to=1-2]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y} \oplus 1_{Z}}"', from=2-1, to=3-1]
+      \arrow["{\alpha_{Y,X,Z}}"', from=3-1, to=4-2]
+      \arrow["{1_{Y} \oplus \mathcal{B}_{X,Z}}"', from=4-2, to=3-3]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{\alpha_{Y,Z,X}}", from=2-3, to=3-3]
+      \arrow["{\mathcal{B}_{X,Y \oplus Z}}", from=1-2, to=2-3]
+      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
+    \end{tikzcd}\]
+  % https://q.uiver.app/?q=WzAsMyxbMCwwLCJYIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyBZIl0sWzEsMSwiWSBcXG9wbHVzIFgiXSxbMCwxLCIxX3tYIFxcb3BsdXMgWX0iLDAseyJsZXZlbCI6Miwic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsIlxcbWF0aGNhbHtCfV97WCxZfSIsMl0sWzIsMSwiXFxtYXRoY2Fse0J9X3tZLFh9IiwyXSxbNCw1LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+  \[\begin{tikzcd}
+      {X \oplus Y} && {X \oplus Y} \\
+      & {Y \oplus X}
+      \arrow["{1_{X \oplus Y}}", Rightarrow, no head, from=1-1, to=1-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y}}"', from=1-1, to=2-2]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{\mathcal{B}_{Y,X}}"', from=2-2, to=1-3]
+      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
+    \end{tikzcd}\]
+\end{proposition}
 

### 2021-06-10 Pi+/paper/finite.tex
Add sketch of Fin-Lehmer proof

@@
 
-Using this Lehmer encoding algorithm, we can now construct the equivalence
-between these types.
+Using this Lehmer encoding algorithm, we can now construct the equivalence between these types.
+
+We define a type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except the one
+provided. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence an $\hSet$.
+
+\begin{definition}
+  \( \FinExcept{n}[i] \defeq \dsum{j : \Fin[n]}{i \neq j} \).
+\end{definition}
 
 \begin{proposition}
-  For all $n:\Nat$,
-  \[
-    \Lehmer[n] \eqv \Aut[\Fin[n]]
-  \]
+  For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$.
+\end{proposition}
+
+\begin{proposition}
+  For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$.
+\end{proposition}
+
+\begin{proposition}
+  For any $n : \Nat$,
+  \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}{k}} \).
+\end{proposition}
+
+\begin{proposition}
+  For all $n:\Nat$, \( \Lehmer[n] \eqv \Aut[\Fin[n]] \).
 \end{proposition}
@@
 \begin{proof}
-  We have to turn this algorithm into a constructive proof in type theory where
-  we only use functions and recursion. \todo{Describe the agda code.}
+  For $n = 0$, note that $\Lehmer[0]$ is contractible, and so is $\Aut[\Fin[0]]$. For $n = \suc[m]$, we have the
+  following chain of equivalences.
+  \[\arraycolsep=0.5em\def\arraystretch{1.5}
+    \begin{array}{rl}
+      & \Aut[\Fin[\suc[m]]] \\
+      \eqv & \dsum{k : \Fin[\suc[m]]}{\FinExcept{\suc[m]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} \\
+      \eqv & \dsum{k : \Fin[\suc[m]]}{\FinExcept{\suc[m]}[\fzero] \eqv \Fin[m]} \\
+      \eqv & \dsum{k : \Fin[\suc[m]]}{\Fin[m] \eqv \Fin[m]} \\
+      \eqv & \Fin[\suc[m]] \times \Aut[\Fin[m]] \\
+      \eqv & \Fin[\suc[m]] \times \Lehmer[m] \\
+    \end{array}
+  \]
 \end{proof}

### 2021-06-12 Pi+/paper/finite.tex
Update references from zotero

@@
 The name Lehmer comes from Lehmer
-codes~\cite{lehmerTeachingCombinatorialTricks1960a} which are known in
+codes~\cite{lehmerTeachingCombinatorialTricks1960} which are known in
 Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960}. There are many

### 2021-06-13 Pi+/paper/finite.tex
Define the coxeter relations and closure

@@
   Let $A$ be a type and $R : A \to A \to \UU$ a binary relation on $A$. The group $G$ presented by
-  $\langle A ; R \rangle$ is given by the set-quotient of the free group $F(A)$ by the normal closure $N_{R}$, or
+  $\langle A ; R \rangle$ is given by the set-quotient of the free group $F(A)$ by the normal closure $N_{R}$ of $R$, or
   equivalently, as the coequaliser
@@
 
-\begin{definition}[Coxeter relation]
+\begin{definition}[$\cox$]
   \begin{align*}
-    & \mathsf{cancel} : (n \cons n \cons \nil) \cx \nil \\
-    & \mathsf{swap} : (\suc[k] < n) \to (n \cons k \cons \nil) \cx (k \cons n \cons \nil) \\
-    & \mathsf{braid} : (\suc[n] \cons k \cons \suc[n] \cons \nil) \cx (n \cons \suc[k] \cons n \cons \nil)
+    \cancel
+    & : \forall n \to (n \cons n \cons \nil) \cox \nil \\
+    \swap
+    & : \forall k, n \to (\suc[k] < n) \to (n \cons k \cons \nil) \cox (k \cons n \cons \nil) \\
+    \braid
+    & : \forall k, n \to (\suc[n] \cons k \cons \suc[n] \cons \nil) \cox (n \cons \suc[k] \cons n \cons \nil) \\
   \end{align*}
@@
 
-The type $\Sn$ is defined as the set-quotient of $\List(\Fin[n])$ by the congruence closure of $\cx$.
+We define $\cox*$ as the congruence closure of $\cox$.
 
+\vc{This could be simplified a bit depending on our formalisation.}
+
+\begin{definition}[$\cox*$]
+  \begin{align*}
+    \reflr{\cox}
+    & : \forall w \to w \cox* w \\
+    \symr{\cox}
+    & : \forall w_{1}, w_{2} \to w_{1} \cox* w_{2} \to w_{2} \cox* w_{1} \\
+    \transr{\cox}
+    & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \cox* w_{2} \to w_{2} \cox* w_{3} \to w_{1} \cox* w_{3} \\
+    \congrf{\cox}{\append}
+    & : \forall w_{1}, w_{2}, w_{3}, w_{4} \to  w_{1} \cox* w_{2} \to w_{3} \cox* w_{4} \to w_{1} \append w_{3} \cox* w_{2} \append w_{4} \\
+    \relr{\cox}
+    & : \forall w_{1}, w_{2} \to w_{1} \cox w_{2} \to w_{1} \cox* w_{2} \\
+  \end{align*}
+\end{definition}
+
+The type $\Sn$ is defined as the set-quotient of $\List(\Fin[n])$ by $\cox*$.
+q
 \begin{definition}[$\Sn$]
-  \(\Sn \defeq \quot{\List(\Fin[n])}{\cx^{\ast}}\)
+  \(\Sn \defeq \quot{\List(\Fin[n])}{\cox*}\)
 \end{definition}

### 2021-06-13 Pi+/paper/finite.tex
More on the coxeter relation

@@
 \begin{definition}
-  Let $A$ be a type, and $\List(A)$ the free monoid on $A$. The free group $F(A)$ on $A$ is the set-quotient of
-  $\List(A + A)$ by the congruence closure of the relation $(\inl(a) \cons \inr(a) \cons \nil) \sim \nil$.
+  Let $A$ be a type, and $\List[A]$ the free monoid on $A$. The free group $F(A)$ on $A$ is the set-quotient of
+  $\List[A + A]$ by the congruence closure of the relation $(\inl(a) \cons \inr(a) \cons \nil) \sim \nil$.
 \end{definition}
@@
 
-We define a binary relation on $\List(\Fin[n])$.
+We define a binary relation on $\List[\Fin[n]]$.
 
@@
 We define $\cox*$ as the congruence closure of $\cox$.
-
-\vc{This could be simplified a bit depending on our formalisation.}
+\vc{This could be simplified a bit depending on our formalisation. $\cox$ is directed, use a different symbol?}
 
@@
 
-The type $\Sn$ is defined as the set-quotient of $\List(\Fin[n])$ by $\cox*$.
-q
+We state and prove two desirable properties of this relation on $\List[\Fin[n]]$ from the point of view of an abstract
+rewriting system. We follow the terminology of~\cite{krausCoherenceWellFoundednessTaming2020}.
+\vc{Not completely formalised.}
+
+\begin{proposition}
+  \leavevmode
+  \begin{enumerate}
+    \item $\cox$ is (locally) confluent. For every span $w_{1} \cox w \cox w_{2}$, there is a matching extended cospan
+          $w_{1} \cox* v \cox* w_{2}$.
+    \item $\cox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \cox* v$. Equivalently,
+          the type $\dsum{v:\List[\Fin[n]]}{w \cox* v}$ is contractible, and $w \cox* v$ is a proposition for every $w$
+          and $v$.
+  \end{enumerate}
+\end{proposition}
+
+The type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.
+
 \begin{definition}[$\Sn$]
@@
 \begin{proposition}
-  $\Sn$ is equivalent to the generated group given by the normal closure of $\cx$ extended to $\List(\Fin[n] + \Fin[n])$
+  $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to $\List(\Fin[n] + \Fin[n])$
   along the codiagonal map $[\term{id},\term{id}] : A + A \to A$.

### 2021-06-14 Pi+/paper/finite.tex

@@
 
+\todo{Long form of Coxeter relations and their equivalence.}
+
 We state and prove two desirable properties of this relation on $\List[\Fin[n]]$ from the point of view of an abstract
-rewriting system. We follow the terminology of~\cite{krausCoherenceWellFoundednessTaming2020}.
-\vc{Not completely formalised.}
+rewriting system. We follow the terminology of~\cite{krausCoherenceWellFoundednessTaming2020}. \vc{Not completely
+  formalised. Is this with Coxeter or LongCoxeter?}
 
@@
           $w_{1} \cox* v \cox* w_{2}$.
-    \item $\cox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \cox* v$. Equivalently,
-          the type $\dsum{v:\List[\Fin[n]]}{w \cox* v}$ is contractible, and $w \cox* v$ is a proposition for every $w$
-          and $v$.
+    \item $\cox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \cox* v$.
+          \review{Equivalently, the type $\dsum{v:\List[\Fin[n]]}{w \cox* v}$ is contractible, and $w \cox* v$ is a
+          proposition for every $w$ and $v$.}
   \end{enumerate}
@@
 
+We will now prove the group structure of $\Sn$.
+
 \begin{proposition}
@@
 \begin{proposition}
-  $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to $\List(\Fin[n] + \Fin[n])$
-  along the codiagonal map $[\term{id},\term{id}] : A + A \to A$.
+  $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to
+  $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $[\term{id},\term{id}] : A + A \to A$.
 \end{proposition}

### 2021-06-15 Pi+/paper/extra.tex
structure

\section{Extras}

Shor's quantum factoring algorithm requires, at its core, a classical reversible function for modular exponentiation. As
a small example, consider the function $f(r) = 11^{r} \mod 15$. Using the Toffoli construction~\cite{Toffoli:1980}, we
can embed this function $f$ into a reversible function $g$, shown below, such that $g(r,0) = (r,f(r))$:
\[\begin{array}{rcll}
g(r,h) &=& (r,h+1) & \mbox{when~$r$~even~and~$h$~even} \\
g(r,h) &=& (r,h-1) & \mbox{when~$r$~even~and~$h$~odd} \\
g(r,h) &=& (r,11-h) & \mbox{when~$r$~odd~and~$4 > h \geq 0$~or~$12 > h \geq 8$} \\
g(r,h) &=& (r,19-h) & \mbox{when~$r$~odd~and~$8 > h \geq 4$~or~$16 > h \geq 12$}
  \end{array}\]
In Shor's algorithm, $r$ is of the form $2^j$ and to satisfy the complexity requirements of the algorithm, it is
necessary to implement the function $g$ in a way that grows \emph{polynomially} with $j$. As explained in standard
accounts of the algorithm (e.g., the Qiskit implementation), such efficient modular exponentiation circuits are not
straightforward and are the bottleneck in Shor’s algorithm.

P~\cite{shorefficient}

something that uses level 2 is better

Tabulating this function, we observe that it is periodic with
$f(0) = 1, f(1) = 11, f(2) = 0, f(3) = 11, f(4) = 0, \cdots$ or more succinctly:
\[\begin{array}{rcll}
f(r) &=& 1 & \mbox{if~$r$~even} \\
f(r) &=& 11 & \mbox{if~$r$~odd}
\end{array}\]

If $h=0$, we get:

\begin{verbatim}
g(r,0) = (r, 1) when r even
g(r,0) = (r, 11) when r odd
\end{verbatim}

Homotopy type theory (HoTT) introduced the concept of \emph{univalent universes
  of types.} We argue that this concept should be considered a benchmark
property of programming languages generalizing the full abstraction property. To
properly explain the idea and introduce our main results in context, we begin by
recalling the general definition of full abstraction and then specialize it to a small
language of type isomorphisms.

\paragraph*{Full Abstraction.} Consider a programming language equipped with an
operational semantics $\mathcal{O}$ and a denotational semantics
$\mathcal{M}$. As is well-established, each of these semantics induces a notion
of equivalence which we call $\simeq_{\mathcal{O}}$ and $\simeq_{\mathcal{M}}$
respectively.

\begin{definition}[Full Abstraction] A denotational semantics $\mathcal{M}$ is
  \emph{fully abstract} with respect to an operational semantics $\mathcal{O}$
  if the induced equivalences $\simeq_{\mathcal{M}}$ and $\simeq_{\mathcal{O}}$
  coincide.
\end{definition}

In many languages, the correspondence between operational and denotational
equivalences does not hold and when it does it is quite complicated to
establish. However, if we restrict our attention to the language of type
isomorphisms between finite types, the situation is much simpler.

\paragraph*{Type Isomorphisms.} Consider a small universe of types built from
the empty type ($\emptyt$) and the unit type ($\onet$), and closed under sums
$(\sumt)$ and products $(\prodt)$. The elements of these types are the expected
values $v$: the unit value, the left and right injections of values, and the
pairing of values. It is well-known that, under the conventional interpretation
of types, this universe forms a commutative semiring (see for example, the Agda
standard library). It is straightforward (see Sec.~\ref{sec:pi} for details) to
provide a syntax these type isomorphisms as a inductive collection of
combinators $c$ and an operational semantics $\Downarrow$ that reduces an
application $c(v)$ to an answer $v'$. This semantics induces the following
operational equivalence relation.

\begin{definition}[Operational Equivalence]
  Two combinators $c_1$ and $c_2$ are operationally equivalent
  $c_1 \simeq_{\mathcal{O}} c_2$, if for all values $v$, we have
  $c_1(v) \Downarrow v'$ iff $c_2(v) \Downarrow v'$.
\end{definition}

Independently of this operational interpretation of type isomorphisms, a natural
denotational model is to interpret each type $A$ as a natural number $|A|$ and
interpret each combinator $c$ as a permutation $\llbracket c \rrbracket$. This
model induces the following denotational equivalence.

\begin{definition}[Denotational Equivalence]
  Two combinators $c_1$ and $c_2$ are denotationally equivalent
  $c_1 \simeq_{\mathcal{M}} c_2$, if the denoted permutations are equal:
  $\llbracket c_1 \rrbracket = \llbracket c_2 \rrbracket$.
\end{definition}

The full abstraction property for this language follows from a well-established
result by Fiore, Di Cosmo, and Balat~\cite{fioreRemarksIsomorphismsTyped2002}.

\paragraph*{Function Extensionality.} The previous (standard) definitions of
operational and denotational equivalences rely on function extensionality as
they express the equivalence of two combinators $c_1$ and $c_2$ by reasoning
about their behavior at every argument. Generally speaking, extensional equality
can be seen as a logical principle that is invoked when we lack a way to prove
equality between two indistinguishable objects.

add variables to stand for the ``all v'' and try to prove equivalences: ex
morphims from (x+y) to (y+x). Id not there

Denotational: build meaning of a phrase by composing meanings of subphrases (set theory)

Operational build meaning of a phrase by reasoning about interactions with
environment (category theory)

categorial semantics can give you both

quantum: observational equiv says |s1> is the same as |s2>
do we have a denotational way (inside Hilbert space) to show that these two
states are the same?

\[\begin{array}{rcl}
    A \sumt B &\isot& B \sumt A \\
    A \prodt B &\isot& B \prodt A \\
    A \sumt \zerot &\isot& A \\
    A \prodt \zerot &\isot& \zerot \\
    A \prodt \onet &\isot& A \\
    A \prodt (B \prodt C) &\isot& (A \prodt B) \prodt C \\
    A \prodt (B \sumt C) &\isot& (A \prodt B) \sumt (A \prodt C)
  \end{array}\]

something about level 2 term rewriting

No variables in pi so statement much simpler

Fiore

some existing work on a language with just bool; this extends it to a more
realistic language

An operational semantics typically formalizes
program execution as a partial function $e \Downarrow v$ that maps programs $e$
to observable values $v$. This partial function induces a notion of operational
equivalence defined as follows.

An operational semantics typically formalizes
program execution as a partial function $e \Downarrow v$ that maps programs $e$
to observable values $v$. This partial function induces a notion of operational
equivalence defined as follows.

Programs of type $A \rightarrow B$ where $A \isot B$ can be given both an
operational semantics and a denotational semantics.

%% show three programs of type 1 + 1
%% id, id . id, and swap

We may choose to define this notion of ``same'' operationally or
denotationally.

\begin{definition}[Denotational Equivalence]
  Two expressions $e_1$ and $e_2$ are denotationally equivalent
  $e_1 \simeq_{\mathcal{M}} e_2$, if they denote the same element in the
  mathematical domain, i.e., if $\mathcal{M}(e_1) = \mathcal{M}(e_2)$.
\end{definition}

For

give meaning to programs, based on their execution.

The standard operational semantics of In addition to an operational understanding of type isomorphisms

If we interpret a type $A$ as denoting the natural number $\| A \|$, then a
program $p : A \rightarrow B$ where $A \isot B$ denotes a permutation on
$\| A \| (= \| B \|)$ elements. Full abstraction states that the operational
interpretation of type isomorphisms precisely matches the denotational one,
i.e.,
\[
XX
\]

As a tool for investigating program properties, full abstraction can be seen as
a completeness property of denotational semantics: every equivalence of programs
that can be proved operationally, can also be proved by denotational
means. Equivalently, a denotational proof that two terms are not equivalent will
be enough to show that they are not interchangeable in every program context.

There are at least two distinct ways to reason about such type isomorphisms: (i)
operationally in a programming language by writing programs whose types match
the isomorphisms, and denotationally by interpreting the types as mathematical
objects (e.g. sets).

In a programming language, there will be infinite ways of
expressing the isomorphism as a program. Denotationally, for the special case
under consideration, each type could be interpreted as a natural number and each
isomorphism as a permutation.

As an illustration, here is the Agda proof of the first isomorphism above:

%% ⊎-comm : ∀ {a b} (A : Set a) (B : Set b) → (A ⊎ B) ↔ (B ⊎ A)
%% ⊎-comm _ _ = inverse swap swap swap-involutive swap-involutive

% \begin{figure}[t]
% \[
% \begin{array}{rrcll}
% \idc :& t & \iso & t &: \idc \\
% \\
% \identlp :&  0 \sumtype t & \iso & t &: \identrp \\
% \swapp :&  t_1 \sumtype t_2 & \iso & t_2 \sumtype t_1 &: \swapp \\
% \assoclp :&  t_1 \sumtype (t_2 \sumtype t_3) & \iso & (t_1 \sumtype t_2) \sumtype t_3 &: \assocrp \\
% \\
% \identlt :&  1 {\prodtype} t & \iso & t &: \identrt \\
% \swapt :&  t_1 {\prodtype} t_2 & \iso & t_2 {\prodtype} t_1 &: \swapt \\
% \assoclt :&  t_1 {\prodtype} (t_2 {\prodtype} t_3) & \iso & (t_1 {\prodtype} t_2) {\prodtype} t_3 &: \assocrt \\
% \\
% \absorbr :&~ 0 {\prodtype} t & \iso & 0 ~ &: \factorzl \\
% \dist :&~ (t_1 \sumtype t_2) {\prodtype} t_3 & \iso & (t_1 {\prodtype} t_3) \sumtype (t_2 {\prodtype} t_3)~ &: \factor
% \end{array}
% \]
% \caption{$\Pi$-terms.}
% \label{pi-terms}
% \end{figure}

% We now have in our hands our desired denotational semantics for types.
% We want to create a programming language, which we call $\Pi$, such
% that the types and type combinators map to $\bot, \top, \presumtype,
% \preprodtype$, and such that we have ground terms whose denotation are
% all $16$ type isomorphisms of Fig.~\ref{type-isos}. This is rather
% straightforward, as we can simply do this literally. To make the
% analogy with commutative semirings stand out even more, we will use
% $0, 1, \sumtype$, and ${\prodtype}$ at the type level, and will denote
% ``equivalence'' by $\iso$.  Thus Fig.~\ref{pi-terms} shows the
% ``constants'' of the language.  As these all come in symmetric pairs
% (some of which are self-symmetric), we give names for both directions.
% Note how we have continued with the spirit of Curry-Howard: the terms
% of $\Pi$ are \emph{proof terms}, but rather than being witnesses of
% inhabitation, they are witnesses of equivalences. Thus we get an
% unexpected programming language design:

% \begin{center}
% \fbox{ The proof terms denoting commutative semiring equivalences
%   induce the terms of $\Pi$.}
% \end{center}
% \vspace*{3mm}

% \begin{figure}[t]
% \[
% \Rule{}
% {\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_2 \iso t_3}
% {\jdg{}{}{c_1 \odot c_2 : t_1 \iso t_3}}
% {}
% \qquad
% \Rule{}
% {\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_3 \iso t_4}
% {\jdg{}{}{c_1 \oplus c_2 : t_1 \sumtype t_3 \iso t_2 \sumtype t_4}}
% {}
% \]
% \[
% \Rule{}
% {\jdg{}{}{c_1 : t_1 \iso t_2} \quad \vdash c_2 : t_3 \iso t_4}
% {\jdg{}{}{c_1 \otimes c_2 : t_1 {\prodtype} t_3 \iso t_2 {\prodtype} t_4}}
% {}
% \]
% \caption{$\Pi$-combinators.}
% \label{pi-combinators}
% \end{figure}

% \noindent
% Of course, one does not get a programming language with just typed
% constants! There is a need to perform multiple equivalences. There are
% in fact three ways to do this: sequential composition $\odot$, choice
% composition $\oplus$ (sometimes called juxtaposition), and parallel
% composition $\otimes$. See Fig.~\ref{pi-combinators} for the
% signatures. The construction $c_1 \odot c_2$ corresponds to performing
% $c_1$ first, then $c_2$, and is the usual notion of composition -- and
% corresponds to $\fatsemi$ of the language of permutations of
% Sec.~\ref{sec:dataone}. The construction $c_1 \oplus c_2$ chooses to
% perform $c_1$ or $c_2$ depending on whether the input is labelled
% $\textsf{left}$ or $\textsf{right}$ respectively. Finally the
% construction $c_1 \otimes c_2$ operates on a product structure, and
% applies $c_1$ to the first component and $c_2$ to the second. The
% language of permutations lacked the ability to combine permutations by
% taking sums and products, which led to the awkward non-compositional
% programming style illustrated in the full adder
% example~(Eq.~\ref{eq:adder}).

% Thus the denotation of the $\Pi$ terms \emph{should} be
% permutations. But given types $A$ and $B$ denoting $\fin{m}$
% and~$\fin{n}$ respectively, what are $A \presumtype B$ and $A \preprodtype B$ ?
% They correspond exactly to $\fin{m+n}$ and $\fin{m*n}$!
% Geometrically, this corresponds to concatenation for $A + B$,
% i.e. lining up the elements of $A$ first, and then those of~$B$. For
% $A * B$, one can picture this as lining up the elements of $A$
% horizontally, those of $B$ vertically and perpendicular to those of
% $A$, and filling in the square with pairs of elements from $A$ and
% $B$; if one re-numbers these sequentially, reading row-wise, this
% gives an enumeration of $\fin{m*n}$.

% From here, it is easy to see what, for example, $c_1 \oplus c_2$ must be,
% operationally: from a permutation on $\fin{m}$ and another on $\fin{n}$,
% create a permutation on $\fin{m+n}$ by having $c_1$ operate on the first
% $m$ elements of $A+B$, and $c_2$ operate on the last $n$ elements.
% Similarly, $\swapp$ switches the roles of $A$ and $B$, and thus corresponds
% to $\fin{n+m}$. Note how we ``recover'' the commutativity of
% natural number addition from this type isomorphism. Geometrically, $\swapt$
% is also rather interesting: it corresponds to matrix transpose!
% Furthermore, in this representations, some combinators like
% $\identlp$ and $\assoclp$ are identity operations: the underlying representations
% are not merely isomorphic, they are definitionally equal.
% In other words, the passage to $\Nat$ erases some structural information.

% \begin{figure}[t]
% \[
% \Rule{}
% {\jdg{}{}{c_1 : t_1 \iso t_2}}
% {\jdg{}{}{\ !\ c_1 : t_2 \iso t_1}}
% {}
% \]
% \caption{Derived $\Pi$-combinator.}
% \label{derived-pi-combinator}
% \end{figure}

% Embedded in our definition of $\Pi$ is a conscious design decision: to make the
% terms of $\Pi$ \emph{syntactically} reversible. In other words, to
% every $\Pi$ constant, there is another $\Pi$ constant which is its
% inverse. As this is used frequently, we give it the short name $!$,
% and its type is given in Fig.~\ref{derived-pi-combinator}. This
% combinator is \emph{defined}, by pattern matching on the syntax of
% its argument and structural recursion.

% This is not the only choice.  Another would be to add a
% $\mathit{flip}$ combinator to the language; we could then remove
% quite a few combinators as redundant. The drawback is that many
% programs in $\Pi$ become longer. Furthermore, some of the symmetry
% at ``higher levels'' (see next section) is also lost. Since the
% extra burden of language definition and of proofs is quite low, we
% prefer the structural symmetry over a minimalistic language definition.

% \begin{figure}[t]
% \[
% \begin{array}{rrcll}
% \identlsp :&  t \sumtype 0 & \iso & t &: \identrsp \\
% \identlst :&  t {\prodtype} 1 & \iso & t &: \identrst \\
% \\
% \absorbl :&~ t {\prodtype} 0 & \iso & 0 ~ &: \factorzr \\
% \distl :&~ t_1 {\prodtype} (t_2 \sumtype t_3) & \iso & (t_1 {\prodtype} t_2) \sumtype (t_1 {\prodtype} t_3)~ &: \factorl
% \end{array}
% \]
% \caption{Additional $\Pi$-terms.}
% \label{more-pi}
% \end{figure}

% We also make a second design decision, which is to make the $\Pi$
% language itself symmetric in another sense: we want both left
% and right introduction/elimination rules for units, $0$ absorption
% and distributivity. Specifically, we add the $\Pi$-terms of
% Fig.~\ref{more-pi} to our language. These are redundant because
% of $\swapp$ and $\swapt$, but will later enable shorter programs
% and more elegant presentation of program transformations.

This set of isomorphisms is known to be sound and
complete~\cite*{fioreIsomorphismsGenericRecursive2004,fioreRemarksIsomorphismsTyped2002} for isomorphisms of finite
types. Furthermore, it is also universal for hardware combinational circuits~\cite{jamesInformationEffects2012}.

\paragraph*{Type Isomorphisms.} For concreteness consider a small universe of
types consisting of the empty type ($\emptyt$) and the unit type ($\onet$), and
closed under sums $(\sumt)$ and products $(\prodt)$. Under the conventional
interpretation of types, this universe would include familiar type isomorphisms
such as:
\[\begin{array}{rcl}
    A \sumt \zerot &\isot& A \\
    A \prodt \zerot &\isot& \zerot \\
    A \prodt \onet &\isot& A \\
    A \prodt B &\isot& B \prodt A \\
    A \prodt (B \prodt C) &\isot& (A \prodt B) \prodt C \\
    A \prodt (B \sumt C) &\isot& (A \prodt B) \sumt (A \prodt C)
\end{array}\]

There are at least two distinct ways to reason about such type isomorphisms: (i)
operationally in a programming language by writing programs whose types match
the isomorphisms, and denotationally by interpreting the types as mathematical
objects (e.g. sets). In a programming language, there will be infinite ways of
expressing the isomorphism as a program. Denotationally, for the special case
under consideration, each type could be interpreted as a natural number and each
isomorphism as a permutation.

%% code from T.lagda

Types can be isomorphic for example (A * B) and (B * A) but
there are many ways of writing a program witnessing that isomorphism:

f1 (a,b) = (b,a)

f2 (a,b) = (a, (b,())) = ((b,()),a) = (b,((),a)) = (b,a)

AxB => Ax(Bx1) => ((Bx1)xA) => (Bx(1xA)) => (BxA)

etc.

or (1+(1+1)) iso to or (1+(1+1)) in 6 different ways semantically and infinitely
many ways as programs

Even more extreme: 0 is isomorphic to 0; there is only one semantic map id but
many many programs

0 => 0

0 => 0xA => 0x(A+0) => (0xA) + 0x0 => 0x0 + 0xA => 0 + 0xA => 0 + 0 => 0

important to have programs because they are intentional; they display different
tradeoffs: space-time; etc

--

For simple types we would also have a simple set-theoretic way of checking
whether two types are isomorphic: check that their sizes are equal! Then any
program that preserves the sizes is a semantic permutation: for types of size n,
we have n! permutations but many many programs. But is it the case that for
every two types of the same size, we have a program that identifies the two
types.  Full abstraction would give us that for all programs P : A iso B we have
that A is equivalent to B. i.e. any program between types of the same size
defines a permutation and for every permutation there is a program that
implements it. For finite types, that's not too hard.

%%

However that story relies on extensional equality. Consider P1 and P2 that
define the same permutation; their equivalence is via the model. Can we have a
set of rewrite rules that we can use to rewrite P1 to P2... get into the
importance of level 2... can we also have different tradeoffs: how do we prove
P1 is the same as P2: we may have a direct rewrite or a very complicated proof

%%

univalent universe gives you all that in one go: summarize result
level 0
level 1
level 2

establish the result for small language (finite types) and focus on reversible
core because that's the relevant part of the result.

how result connects with other things in category theory, etc.

foundation that can be used for other programming languages with different
types, constructs, etc.

%%

Semantically, obvious that only permutation on empty set is id.

Syntactically, we have 0, 1, +, *  and an infinite number of programs of type 0
<-> 0. How do we prove that they are all equivalent to id?

Obvious induction does not work because after a sequential composition step, the
intermediate type is no longer syntactically 0.

So we add size indices to types. Separate U and UHat. Now the proof is obvious.

%%

Add full pi to pi+ translation; show injective?

\begin{verbatim}
Combinators in the (List 3) are acting on << 4 >>
so

(1 :: 2 :: 0 :: 1 :: nil) of type List 3

corresponds to the sequence:

swap 1 2 ;
swap 2 3 ;
swap 0 1;
swap 1 2

and swap 1 2 is then
   A + (B + (C + (D + 0)) <—> A + (C + (B + (D + 0))

also, swap 2 3 is
   A + (B + (C + (D + 0))) <—> A + (B + (D + (C + 0)))
\end{verbatim}

To cite:

Earliest reference we know about the free symmetric strict monoidal category (as an example of 2-monads):
Two-dimensional monad theory~\cite{blackwellTwodimensionalMonadTheory1989}.

%%

\paragraph*{Categorical Connections.}

Natural numbers under addition form the free commutative monoid on one generator. Categorifying this, we get the free
symmetric monoidal groupoid on generators. This is the groupoid of finite sets and bijections.

More generally, we can construct the free symmetric monoidal groupoid on a groupoid, by taking the action of the
groupoid of finite sets and bijections. We can construct this in HoTT by using the classifying space of the symmetric
group.

The free symmetric monoidal category on a category is a 2-monad on
Cat~\cite{blackwellTwodimensionalMonadTheory1989,abramskyAbstractScalarsLoops2005,leinsterHigherOperadsHigher2004}.
Other applications are Fock Spaces, Generalised Species, Abstract Syntax.

We establish a Curry-Howard-Lambek correspondence for Pi, with 0, 1, +. Categorically, the syntactic groupoid of Pi is
the free symmetric monoidal groupoid on one generator. The groupoid of finite sets and bijections, with coproducts, is
equivalent to the free symmetric monoidal groupoid on one generator, making Pi fully adequate with respect to this
semantics.

The logic part of this is in superstructural reversible logic, which is the free commutative semirig, since there are no
equations.

\note{This is just a collection of notes for now, and most of it might move into discussion.}

### 2021-06-16 Pi+/paper/reversible.tex
structure
pi section: rough draft

--- 2021-01-27 Pi+/paper/univalent.tex
@@
-\section{Univalent Subuniverses}~\label{sec:univalent}
+\section{Reversible Programming Languages}~\label{sec:reversible}
+
+The practice of programming languages is replete with \emph{ad hoc} instances of reversible computations: database
+transactions, mechanisms for data provenance, checkpoints, stack and exception traces, logs, backups, rollback
+recoveries, version control systems, reverse engineering, software transactional memories, continuations, backtracking
+search, and multiple-level undo features in commercial applications. In the early nineties,
+\citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class, treatment of reversibility. But intensive
+research in full-fledged reversible models of computations and reversible programming languages was only sparked by the
+discovery of deep connections between physics and
+computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275, Hey:1999:FCE:304763,fredkin1982conservative}, and by the
+potential for efficient quantum computation~\cite{springerlink:10.1007/BF02650179}.
+
+The early developments of reversible programming languages started
+with a conventional programming language, e.g., an extended
+$\lambda$-calculus, and either:
+\begin{enumerate}
+\item extended the language with a history
+mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
+\item imposed constraints on the control flow constructs to make them
+reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.
+\end{enumerate}
+More modern approaches recognize that reversible programming languages require a fresh approach and should be designed
+from first principles without the detour via conventional irreversible
+languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047,
+  rc2011,James:2012:IE:2103656.2103667,Carette2016}.
+
+%%%%%%%%%%%%%%%%%
+\subsection{The $\Pi$ Family of Languages}
+
+A natural candidate for a semantic foundation for reversible programming languages is the notion of type isomorphism
+(equivalence). Indeed, the type isomorphisms among finite types represent a universal language for reversible boolean
+circuits~\cite{James:2012:IE:2103656.2103667} and the extension with recursive types and trace
+operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
+language~\cite{James:2012:IE:2103656.2103667,rc2011}.
+
+Focusing on finite types, the building blocks of type theory are: the
+empty type ($\mathbb{0}$), the unit type ($\mathbb{1}$) containing
+just one value $\Acon{tt}$, the sum type ($+$) containing values of
+the form $\inlv{v}$ and $\inrv{v}$, and the product ($\times$) type
+containing pairs of values $(v_1,v_2)$. One may view each type $A$
+as a collection of physical wires that can transmit $|A|$ distinct
+values where $|A|$ is a natural number that indicates the size of a
+type, computed as: $| \mathbb{0} | = 0$; $| \mathbb{1} | = 1$;
+$| A + B | = | A | + |B |$; and
+$| A \times B | = | A | * |B |$.  Thus the type
+$\mathbb{B} = \mathbb{1} + \mathbb{1}$ corresponds to a wire that can
+transmit one of two values, i.e., bits, with the convention that
+$\inlv{\Acon{tt}}$ represents \AgdaFunction{𝔽} and
+$\inrv{\Acon{tt}}$ represents \AgdaFunction{𝕋}.  The type
+$\mathbb{B} \times \mathbb{B} \times \mathbb{B}$ corresponds to a
+collection of wires that can transmit three bits. From that
+perspective, a type isomorphism between types $A$ and $B$
+(such that $|A|=|B|=n$) models a \emph{reversible}
+combinational circuit that \emph{permutes} the $n$ different
+values. These type isomorphisms are collected in Fig.~\ref{pi-terms}.
+\begin{figure}[t]
+{\scalebox{\scalef}{$%
+%%\noindent\begin{minipage}{.7\linewidth}
+\begin{array}{rrcll}
+\idc :& A & \iso & A &: \idc \\
+\\
+\identlp :&  \mathbb{0} + A & \iso & A &: \identrp \\
+\swapp :&  A + B & \iso & B + A &: \swapp \\
+\assoclp :&  A + (B + C) & \iso & (A + B) + C &: \assocrp \\ [1.5ex]
+\identlt :&  \mathbb{1} \times A & \iso & A &: \identrt \\
+\swapt :&  A \times B & \iso & B \times A &: \swapt \\
+\assoclt :&  A \times (B \times C) & \iso & (A \times B) \times C &: \assocrt \\ [1.5ex]
+\absorbr :&~ \mathbb{0} \times A & \iso & \mathbb{0} ~ &: \factorzl \\
+\dist :&~ (A + B) \times C & \iso & (A \times C) + (B \times C)~ &: \factor
+  \end{array}$}}
+
+{\scalebox{\scalef}{%
+\Rule{}
+{\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : B \iso C}
+{\jdg{}{}{c_1 \fatsemi c_2 : A \iso C}}
+{}
+
+\Rule{}
+{\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : C \iso D}
+{\jdg{}{}{c_1 \oplus c_2 : A + C \iso B + D}}
+{}
+
+\Rule{}
+{\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : C \iso D}
+{\jdg{}{}{c_1 \otimes c_2 : A \times C \iso B \times D}}
+{}
+}}
+\caption{$\Pi$-terms, combinators, and their types.}
+\label{pi-terms}
+\end{figure}
+
+Each line in the top part of the figure introduces a pair of dual constants that witness the type isomorphism in the
+middle.  These are the \emph{base} (non-reducible) terms of $\Pi$. The isomorphisms are extended to form a congruence
+relation by adding the three constructors at the bottom of the figure that witness equivalence and compatible
+closure. Although austere, this combinator-based language has the advantage of being amenable to formal analysis for at
+least two reasons: (i) it is conceptually simple and small, and (ii) it has direct and evident connections to type
+theory and category theory and hence possesses a rich algebraic structure which we exploit in the remainder of the
+paper.
+
+% Specifically, the type isomorphisms of~$\Pi$ are sound and complete for all permutations on finite
+% types~\cite{Fiore:2004,fiore-remarks} and hence they are \emph{complete} for expressing reversible combinational
+% circuits~\cite{fredkin1982conservative, James:2012:IE:2103656.2103667,Toffoli:1980}. Algebraically, these types and
+% combinators form a \emph{commutative semiring} (up to type isomorphism). Logically, they form a superstructural logic
+% capturing space-time trade-offs~\cite{superstructural}. Categorically, they form a \emph{distributive bimonoidal
+%   category}~\cite{laplaza72}.
+
+% For the fragment over finite types, the syntax of the language $\Pi$ consists of several sorts:
+
+%  {\scalebox{\scalef}{$%
+% \begin{array}{lrcl}
+% \textit{Value types} & A,B,C,D &::=& \mathbb{0} \alt \mathbb{1} \alt A+B \alt A\times B \\
+% \textit{Values}      & v,w,x,y &::=& \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
+% \textit{Program types} &&& A \leftrightarrow B \\
+% \textit{Programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})
+% \end{array}$}}
+
+\begin{figure}[t]
+{\scalebox{\scalef}{%
+$\begin{array}{rclcrcl}
+\delta(\identlp,~\inrv{v}) & = & v & & \delta(\identrp,~v)        & = & \inrv{v}\\
+\delta(\swapp,~\inlv{v})     & = & \inrv{v}\\
+\delta(\swapp,~\inrv{v})     & = & \inlv{v}\\
+\delta(\assoclp,~\inlv{v})          & = & \inlv{(\inlv{v})} & & \delta(\assocrp,~\inlv{(\inlv{v})}) & = & \inlv{v}         \\
+\delta(\assoclp,~\inrv{(\inlv{v})}) & = & \inlv{(\inrv{v})} & & \delta(\assocrp,~\inlv{(\inrv{v})}) & = & \inrv{(\inlv{v})}\\
+\delta(\assoclp,~\inrv{(\inrv{v})}) & = & \inrv{v}          & & \delta(\assocrp,~\inrv{v})          & = & \inrv{(\inrv{v})}\\
+\delta(\identlt,~(\Acon{tt},v)) & = & v & & \delta(\identrt,~v)             & = & (\Acon{tt},v)\\
+\delta(\swapt,~(x,y))         & = & (y,x)    \\
+\delta(\assoclt,~(x,(y,z))) & = & ((x,y),z) & & \delta(\assocrt,~((x,y),z)) & = & (x,(y,z))\\
+\delta(\dist,~(\inlv{x},z)) & = & \inlv{(x,z)} & & \delta(\factor,~\inlv{(x,z)}) & = & (\inlv{x},z)\\
+\delta(\dist,~(\inrv{y},z)) & = & \inrv{(y,z)} & & \delta(\factor,~\inrv{(y,z)}) & = & (\inrv{y},z)
+\end{array}$}}
+\caption{Semantics of base combinators}\label{fig:delta}
+\end{figure}
+
+\begin{figure}[t]
+  {\scalebox{\scalef}{%
+ $\begin{array}{c}
+ \infer{\interp{c}{v} \Downarrow \delta(c,v)}{c ~\mbox{base combinator}}
+\qquad
+\infer{\interp{c_1\oplus c_2}{\inlv{v}} \Downarrow \inlv{w}}
+    {\interp{c_1}{v} \Downarrow w}
+\qquad
+\infer{\interp{c_1\oplus c_2}{\inrv{v}} \Downarrow \inrv{w}}
+    {\interp{c_2}{v} \Downarrow w} \\
+\\
+ \infer{\interp{\idc}{v} \Downarrow v}{}
+\qquad
+\infer{\interp{c_1\fatsemi c_2}{v} \Downarrow v_2}
+    {\interp{c_1}{v} \Downarrow v_1 \quad \interp{c_2}{v_1} \Downarrow v_2}
+\qquad
+\infer{\interp{c_1\otimes c_2}{(v_1 , v_2)} \Downarrow (w_1 , w_2)}
+    {\interp{c_1}{v_1} \Downarrow w_1 \quad \interp{c_2}{v_2} \Downarrow w_2}
+\end{array}$}}
+\caption{$\Pi$-interpreter}\label{fig:pi:interp}
+\end{figure}
+
+\noindent To formalize the semantics of $\Pi$, we first specify the
+semantics of the base combinators in Fig.~\ref{fig:delta} using a
+function $\delta$. It is clear that this function has an inverse
+$\delta^\dagger$ such that $\delta^\dagger(c,\delta(c,v))=v$. The semantics of full programs is given by the interpreter
+in Fig.~\ref{fig:pi:interp}.
+
+%%%%%%%%%
+\subsection{Examples}
+\label{sec:langRev-examples}
+\label{examples}
+
+\note{Revisit to pick examples relevant to the rest of the paper; also unify notation}
+
+\paragraph*{Booleans}
+Let us start with encoding booleans. We use the type \ensuremath{1{\sumtype}1} to
+represent booleans with \ensuremath{\mathit{left} ~()} representing \ensuremath{\mathit{true}} and
+\ensuremath{\mathit{right}~()} representing \ensuremath{\mathit{false}}.
+Boolean negation is straightforward to define:
+
+\ensuremath{\mathit{not} : \mathit{bool} \leftrightarrow \mathit{bool}}
+
+\ensuremath{\mathit{not} = \swapp}
+
+\noindent
+It is easy to verify that \ensuremath{\mathit{not}} changes \ensuremath{\mathit{true}} to \ensuremath{\mathit{false}} and
+vice versa.
+
+\paragraph*{Bit Vectors.}
+We can represent $n$-bit words using an n-ary product of
+\ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
+using the type \ensuremath{\mathit{bool} {\prodtype} (\mathit{bool} {\prodtype}  \mathit{bool})}.  We can perform various
+operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
+instance the bitwise \ensuremath{\mathit{not}} operation is the parallel composition of
+three \ensuremath{\mathit{not}} operations:
+
+\ensuremath{\mathit{not}_{\mathit{word}_3} :: \mathit{word}_3 \leftrightarrow \mathit{word}_3}
+
+\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not}  {\prodtype}  (\mathit{not}  {\prodtype}  \mathit{not})}
+
+\noindent We can express a 3-bit word reversal operation as follows:
+
+\ensuremath{\mathit{reverse} : \mathit{word}_3 \leftrightarrow \mathit{word}_3}
+
+\ensuremath{\mathit{reverse} = \swapt \odot (\swapt  \otimes  \idc)~ \odot \assocrt}
+
+\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
+applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
+derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
+many others we will see in this paper, is formed by sequentially
+composing several simpler combinators. Instead of presenting the
+operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
+for presentation reasons) to flatten the tree into a sequence of
+reductions as caused by each component. Such a sequence of reductions
+is given below:
+\[\begin{array}{rlr}
+ & (v_1, (v_2, v_3)) \\
+ \swapt & ((v_2, v_3), v_1) \\
+ \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
+ \assocrt & (v_3, (v_2, v_1)) \\
+ \end{array}\]
+%subcode source isomorphisms.tex:979
+
+\noindent On the first line is the initial value. On each subsequent
+line is a fragment of the \ensuremath{\mathit{reverse}} combinator and the value that
+results from applying this combinator to the value on the previous
+line. For example, \ensuremath{\swapt} transforms \ensuremath{(v_1, (v_2, v_3))} to
+\ensuremath{((v_2,v_3),v_1)}.  On the last line we see the expected result with
+the bits in reverse order.
+
+We can also draw out the graphical representation of the 3-bit reverse
+combinator. In the graphical representation, it is clear that the
+combinator achieves the required shuffling.
+
+\inkscape{reverse-3-bit.pdf}
+
+\paragraph*{Conditionals.}
+Even though \ensuremath{\Pi } lacks conditional expressions, they are
+expressible using the distributivity and factoring laws. The
+diagrammatic representation of \ensuremath{\dist} shows that it redirects the flow
+of a value \ensuremath{v:b} based on the value of another one of type
+\ensuremath{b_1{\sumtype}b_2}. If we choose \ensuremath{1{\sumtype}1} to be
+\ensuremath{\mathit{bool}} and apply either \ensuremath{c_1:b_1\leftrightarrow
+b_2} or \ensuremath{c_2:b_1\leftrightarrow b_2} to the value \ensuremath{v},
+then we essentially have an `if' expression.
+
+\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool}  {\prodtype}  b_1 \leftrightarrow \mathit{bool}  {\prodtype}  b_2}
+
+\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  \otimes\  c_1) {\sumtype} (\idc \otimes\  c_2)) \odot \factor}
+
+\inkscape{if-c1-c2.pdf}
+
+The diagram above shows the input value of type \ensuremath{(1{\sumtype}1) {\prodtype}  b_1}
+processed by the distribute operator \ensuremath{\dist}, which converts it into
+a value of type \ensuremath{(1 {\prodtype}  b_1){\sumtype}(1 {\prodtype}  b_1)}. In the
+\ensuremath{\mathit{left}} branch, which corresponds to the
+case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was
+\ensuremath{\mathit{left} ~()}), the combinator~\ensuremath{c_1} is applied to
+the value of type~\ensuremath{b_1}. The right
+branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
+the value of type \ensuremath{b_1} through the combinator \ensuremath{c_2}.
+The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied
+to get the final result of type \ensuremath{(1{\sumtype}1) {\prodtype} b_2}.
+
+\paragraph*{Logic Gates}
+There are several universal primitives for conventional (irreversible)
+hardware circuits, such as \ensuremath{\mathit{nand}} and \ensuremath{\mathit{fanout}}. In the case
+of reversible hardware circuits, the canonical universal primitive is
+the Toffoli gate~\cite{Toffoli:1980}. The Toffoli gate takes three
+boolean inputs: if the first two inputs are \ensuremath{\mathit{true}} then the third
+bit is negated. In a traditional language, the Toffoli gate would be
+most conveniently expressed as a conditional expression like:
+
+\noindent
+\ensuremath{ \mathit{toffoli}(v_1,v_2,v_3) = \mathit{if} ~(v_1 ~\mathit{and} ~v_2) ~\mathit{then} ~(v_1, v_2, \mathit{not}(v_3)) ~\mathit{else} ~(v_1, v_2, v_3)}
+
+We will derive Toffoli gate in \ensuremath{\Pi } by first deriving a simpler
+logic gate called \ensuremath{\mathit{cnot}}.  Consider a one-armed version, \ensuremath{\mathit{if}_c},
+of the conditional derived above. If the \ensuremath{\mathit{bool}} is
+\ensuremath{\mathit{true}}, the value of type \ensuremath{b} is modified by the operator \ensuremath{c}.
+
+\begin{center}
+\scalebox{1.5}{
+%%subcode-line{pdfimage}[diagrams/if_c.pdf]
+\includegraphics{inkscape/cnot.pdf}
+}
+\end{center}
+
+By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
+combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool} {\prodtype}  \mathit{bool}\leftrightarrow \mathit{bool} {\prodtype}  \mathit{bool}} which negates its
+second argument if the first argument is \ensuremath{\mathit{true}}. This gate
+\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.
+
+If we iterate this construction once more, the resulting combinator
+\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})\leftrightarrow \mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})}.
+The
+resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
+proceeds to check the second argument. If that is also \ensuremath{\mathit{true}} then
+it will negate the third argument. Thus \ensuremath{\mathit{if}_{\mathit{cnot}}} is the required
+Toffoli gate.
+
+\begin{center}
+\scalebox{1.6}{
+\includegraphics{inkscape/toffoli.pdf}
+}
+\end{center}
+
+%%%%%%%%%
+\subsection{A Language of Equivalences between Type Equivalences}
+\label{langeqeq}
+
+In the previous section, we examined equivalences between conventional data structures, i.e., structured trees of
+values. We now consider a richer but foundational notion of data: programs themselves. Indeed, universal computation
+models crucially rely on the fact that \emph{programs are (or can be encoded as) data}, e.g., a Turing machine can be
+encoded as a string that another Turing machine (or even the same machine) can manipulate. Similarly, first-class
+functions are the \emph{only} values in the $\lambda$-calculus.  In our setting, we ask whether the programs developed
+in the previous section can themselves be subject to (higher-level) equivalences?
+
+We will explain the ideas using two small exaamples. Consider the following two programs mapping between the types
+$A + B$ and $C+D$:
+
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
+  \draw[fill] (-2,-1.5) circle [radius=0.025];
+  \node[below] at (-2.1,-1.5) {$A$};
+  \node[below] at (-2.1,-1.9) {$+$};
+  \draw[fill] (-2,-2.5) circle [radius=0.025];
+  \node[below] at (-2.1,-2.5) {$B$};
+
+  \draw[fill] (6.5,-1.5) circle [radius=0.025];
+  \node[below] at (6.7,-1.5) {$C$};
+  \node[below] at (6.7,-1.9) {$+$};
+  \draw[fill] (6.5,-2.5) circle [radius=0.025];
+  \node[below] at (6.7,-2.5) {$D$};
+
+  \draw[<-] (-2,-1.5) to[bend left] (1,0.5) ;
+  \draw[<-] (-2,-2.5) to[bend left] (1,-0.5) ;
+  \draw[->] (3.5,0.5) to[bend left] (6.5,-1.45) ;
+  \draw[->] (3.5,-0.5) to[bend left] (6.5,-2.45) ;
+
+  \draw[<-] (-2,-1.5) to[bend right] (1,-3.5) ;
+  \draw[<-] (-2,-2.5) to[bend right] (1,-4.5) ;
+  \draw[->] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
+  \draw[->] (3.5,-4.5) to[bend right] (6.5,-2.55) ;
+
+  \draw     (2,0.5)  -- (2.5,0.5)  ;
+  \draw     (2,-0.5) -- (2.5,-0.5) ;
+
+  \draw     (2.5,0.5)  -- (3.5,-0.5)  ;
+  \draw     (2.5,-0.5) -- (3.5,0.5) ;
+
+  \draw     (1,-3.5)  -- (2,-4.5)    ;
+  \draw     (1,-4.5) -- (2,-3.5)   ;
+
+  \draw     (2,-3.5)  -- (2.5,-3.5)    ;
+  \draw     (2,-4.5) -- (2.5,-4.5)   ;
+
+  \path (1.5,0.5) node (tc1) [func] {$c_1$};
+  \path (1.5,-0.5) node (tc2) [func] {$c_2$};
+  \path (3,-4.5) node (bc1) [func] {$c_1$};
+  \path (3,-3.5) node (bc2) [func] {$c_2$};
+\end{tikzpicture}
+\end{center}
+The top path is the $\Pi$ program $(c_1~\oplus~c_2)~\odot~\swapp$ which acts on the type $A$ by $c_1$, acts on the type
+$B$ by $c_2$, and acts on the resulting value by a twist that exchanges the two injections into the sum type. The bottom
+path performs the twist first and then acts on the type $A$ by $c_1$ and on the type $B$ by $c_2$ as before. One could
+imagine the paths are physical \emph{elastic} wires in $3$ space, where the programs $c_1$ and $c_2$ as arbitrary
+deformations on these wires, and the twists do not touch but are in fact well-separated. Then, holding the points $A$,
+$B$, $C$, and $D$ fixed, it is possible to imagine sliding $c_1$ and $c_2$ from the top wire rightward past the twist,
+and then using the elasticity of the wires, pull the twist back to line up with that of the bottom --- thus making both
+parts of the diagram identical.  Each of these moves can be undone (reversed), and doing so would take the bottom part
+of the diagram into the top part.  In other words, there exists an equivalence of the program
+$(c_1~\oplus~c_2)~\odot~\swapp$ to the program $\swapp \odot (c_2~\oplus~c_1)$. We can also show that this means that,
+as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and $\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not all
+programs between the same types can be deformed into one another. The simplest example of inequivalent deformations are
+the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.
+
+As another example, consider consider a circuit that takes an input type consisting of three values \Tree [ {\small a} [
+{\small b} {\small c} ] ]~ and swaps the leftmost value with the rightmost value to produce \Tree [ {\small c} [ {\small
+  b} {\small a} ] ]~.  We can implement two such circuits using our Agda library for $\Pi$:
+
+\begin{code}%
+\>[0]\AgdaFunction{swap{-}fl1}\AgdaSpace{}%
+\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{a}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaDatatype{U}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{a}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaDatatype{⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{a}\AgdaSymbol{)}\<%
+\\
+\>[0]\AgdaFunction{swap{-}fl1}\AgdaSpace{}%
+\AgdaSymbol{=}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\<%
+\\
+%
+\\[\AgdaEmptyExtraSkip]%
+\>[0]\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
+\AgdaSymbol{=}%
+\>[52I]\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+\>[.]\<[52I]%
+\>[11]\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\<%
+\end{code}
+
+\noindent The first implementation rewrites the incoming values as follows:
+\[
+\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+\Tree [ [ {\small a} {\small b} ] {\small c} ] ~\to~
+\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+\]
+\noindent
+The second implementation rewrites the incoming values as follows:
+\[
+\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+\Tree [ {\small a} [ {\small c} {\small b} ] ] ~\to~
+\Tree [ [ {\small a} {\small c} ] {\small b} ] ~\to~
+\Tree [ [ {\small c} {\small a} ] {\small b} ] ~\to~
+\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+\]
+\noindent The two circuits are extensionally equal. Using the level-2
+isomorphisms we can \emph{explicitly} construct a sequence of
+rewriting steps that transforms the second circuit to the first.
+
+Recalling that the $\lambda$-calculus arises as the internal language of Cartesian Closed Categories
+(Elliott~\cite{Elliott-2017} gives a particularly readable account of this), we can think of $\Pi$ in similar terms, but
+for symmetric Rig Groupoids instead. For example, we can ask what does the equivalence above represent? It is actually a
+``linear'' representation of a 2-categorial commutative diagram! In fact, it is a painfully verbose version thereof, as
+it includes many \emph{refocusing} steps because our language does not build associativity into its syntax. Categorical
+diagrams usually do.  Thus if we rewrite the example in diagrammatic form, eliding all uses of associativity, but
+keeping explicit uses of identity transformations, we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents
+
+\vspace*{3mm}
+\begin{tikzcd}[column sep=normal, row sep=normal]
+ && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
+ & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
+a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
+  \arrow [dr, "\assoclp"]
+  \arrow [ddr, swap, "\assoclp"]
+    & (a+b)+c \arrow [r, "\swapp"] &
+    c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}]
+    & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
+ & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
+ & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
+             \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] &&
+    c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
+ && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
+ \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
+ \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
+\end{tikzcd}
 

### 2021-06-16 Pi+/paper/finite.tex
Remarks in sec 4.3
Some work on explaining immersion
Explain groups and deloopings

@@
 
+\todo{Big example: Start from a listed permutation, show Lehmer code, then adjacent swaps.}
+\todo{Justification for why we need group theory.}
+
 %% Coxeter presentation of $\Sn$ $\eqv$ $\Lehmer[n]$ $\eqv$ $\Aut[\Fin[n]]$
 
+\subsection{Groups}
+
+From universal algebra, a group is simply a set with a 0-ary constant, the neutral element, a binary operation for group
+multiplication, and a unary inverse operation. A simple example is $\mathbb{Z}/n\mathbb{Z}$, where the neutral element
+is 0, the inverse of $k$ is $-k$, and the group multiplication is given by addition modulo $n$. The neutral element has
+to satisfy unit and inverse laws, and the multiplication has to be associative.
+
+In type theory, a group $G$ can be defined as an $\hSet$ $S$ with the following pieces of data:
+
+\begin{enumerate}
+  \item a unit or neutral element $e : S$
+  \item a multiplication function $m : S \times S \to S$ written as $(g_{1}, g_{2}) \mapsto g_{1} \mult g_{2}$, that satisfies
+  \begin{enumerate}
+    \item the unit laws, for all $g : S$, that \( g \mult e \id g \) and \( e \mult g \id g \)
+    \item the associativity law, for all $g_{1}, g_{2}, g_{3} : S$, that \( g_{1} \mult (g_{2} \mult g_{3}) \id (g_{1} \mult g_{2}) \mult g_{3} \)
+  \end{enumerate}
+  \item an inversion function $i : S \to S$ written as $g \mapsto \inv{g}$, that satisfies
+  \begin{enumerate}
+    \item the inverse laws, for all $g : S$, that \( g \mult \inv{g} \id e \) and \( \inv{g} \mult g \id e \)
+  \end{enumerate}
+\end{enumerate}
+
+However, more conveniently, in HoTT, we can instead use groupoids to talk about groups. A group can be identified with a
+1-object groupoid, using a technique called delooping. The delooping of a group $G$ is a groupoid $\B{G}$ given by a
+unique object $\pt$ with self-loops that are 1-paths $\pt \id_{\B{G}} \pt$ corresponding to the elements of $G$. Note
+that the group operations are given by operations on the identity type automatically, with $\refl_{\pt}$ for the neutral
+element, path composition for the group multiplication, and path inverse for the group inverse. These satisfy the group
+laws as well, up to the identity type, using the groupoid coherence laws. Moreover, for 1-groups which are supposed to
+be sets, these 2-paths should be propositions, so we have to restrict $\B{G}$ to be a 1-groupoid. Hence, a group is
+simply given by a pointed, connected 1-type~\cite*{buchholtzHigherGroupsHomotopy2018,symmetryBook2021}.
+
+For example, given a pointed type $(A:\UU, a:A)$, the automorphism group structure at $a$ is given by $a \id_{A} a$. Of
+course, for 1-groups we will require that $a \id_{A} a$ is an $\hSet$, which is enforced by having $A$ be a groupoid. In
+our running example for the permutation group on finite sets, we have that
+$\Aut[\Fin[n]] \defeq (\Fin[n] \eqv \Fin[n]) \eqv (\Fin[n] \id_{\UU} \Fin[n])$, and the corresponding pointed, connected
+groupoid is given by the classifying type $\UFin[n] \defeq \BAut[\Fin[n]]$, with
+$\loopspace[\BAut[\Fin[n]],F_{n}] \eqv \Aut[\Fin[n]]$ as the corresponding automorphism group. \vc{revisit}
+
+Alternatively, groups can also be presented using generators and relations, or a quotient of the free group \ldots
 \subsection{Permutations}~\label{subsec:permutations}
@@
 
+\todo{Free 1-group, needs to be truncated}
+
 \begin{definition}
@@
 \begin{proposition}
-  Let $\eta : A \to F(A)$ be given by $a \mapsto \inl(a) \cons \nil$. Given any group $G$ with a map $f : A \to G$,
+  Let $\eta_{A} : A \to F(A)$ be given by $a \mapsto \inl(a) \cons \nil$. Given any group $G$ with a map $f : A \to G$,
   there is a canonical lifting of $f$ to a group homomorphism $\extend{f} : F(A) \to G$, such that
@@
 \end{proposition}
+
+\todo{diagram}
+\todo{Plus/Minus}
 
@@
 \todo{Universal property}
+\todo{Examples: empty relation, van Kampen of $\pi_{1}$}
 
@@
     \braid
-    & : \forall k, n \to (\suc[n] \cons k \cons \suc[n] \cons \nil) \cox (n \cons \suc[k] \cons n \cons \nil) \\
+    & : \forall n \to (\suc[n] \cons n \cons \suc[n] \cons \nil) \cox (n \cons \suc[n] \cons n \cons \nil) \\
   \end{align*}
@@
 
-\todo{immersion}
+\begin{definition}
+  \[
+    \encode{\List} : \List[\Fin[n]] \to \Lehmer[n]
+  \]
+\end{definition}
 
-\begin{theorem}
+\begin{proposition}
+  \leavevmode
+  \begin{enumerate}
+    \item \( l_{1} \cox* l_{2} \to \encode{\List}(l_{1}) \id \encode{\List}(l_{1}) \)
+  \end{enumerate}
+\end{proposition}
+
+\begin{definition}
+  \begin{align*}
+    \encodeSn & : \Sn[n] \to \Lehmer[n] \\
+    \decodeSn & : \Lehmer[n] \to \Sn[n]
+  \end{align*}
+\end{definition}
+
+\begin{proposition}
+  For all $n : \Nat$, $(\encodeSn, \decodeSn)$ is an equivalence.
+\end{proposition}
+
+\begin{corollary}
   For all $n : \Nat$,
-  \[
-    \Sn \eqv \Lehmer[n]
-  \]
-\end{theorem}
+  \(
+    \Sn \eqv \Lehmer[n] \eqv \Aut[\Fin[n]]
+  \).
+\end{corollary}
 

### 2021-06-17 Pi+/paper/finite.tex

@@
 unique object $\pt$ with self-loops that are 1-paths $\pt \id_{\B{G}} \pt$ corresponding to the elements of $G$. Note
-that the group operations are given by operations on the identity type automatically, with $\refl_{\pt}$ for the neutral
+that the group operations are automatically given by operations on the identity type, with $\refl_{\pt}$ for the neutral
 element, path composition for the group multiplication, and path inverse for the group inverse. These satisfy the group
@@
 course, for 1-groups we will require that $a \id_{A} a$ is an $\hSet$, which is enforced by having $A$ be a groupoid. In
-our running example for the permutation group on finite sets, we have that
-$\Aut[\Fin[n]] \defeq (\Fin[n] \eqv \Fin[n]) \eqv (\Fin[n] \id_{\UU} \Fin[n])$, and the corresponding pointed, connected
-groupoid is given by the classifying type $\UFin[n] \defeq \BAut[\Fin[n]]$, with
-$\loopspace[\BAut[\Fin[n]],F_{n}] \eqv \Aut[\Fin[n]]$ as the corresponding automorphism group. \vc{revisit}
+our running example for the permutation group on finite sets, we have that $\Fin[n]$ is an $\hSet$, and hence,
+$\UFin[n] \defeq \BAut[\Fin[n]]$ is a pointed, connected 1-type, whose loopspace $\loopspace[\BAut[\Fin[n]],F_{n}]$ is
+equivalent to $\Aut[\Fin[n]] \defeq (\Fin[n] \eqv \Fin[n]) \eqv (\Fin[n] \id_{\UU} \Fin[n])$, which has the
+corresponding automorphism group structure.
 

### 2021-06-17 Pi+/paper/finite.tex
Outline from Jacek's text

@@
 Alternatively, groups can also be presented using generators and relations, or a quotient of the free group \ldots
+
+2. In the usual setting, there are are many relations that are true for the elements of the group. For example, in the
+(Z mod 3), we have a relation 1 + 1 + 1 = 0.
+
+3. A group that has no other relations than the ones directly implied by
+the definition of a group is called free.
+
+4. A group is said to be generated by a set of its elemenst (e1, ..., en)
+if every element in the group can be constructed from these elements
+composing operations of + and inverse. For example, Z is a free group,
+generated by one-element set (1).
+4a. Thus, a free group g can be thought as sequences of letters from the
+generating set, divided by the congruence closure of the group relation.
+Definition: Free group.
+
+5. One useful way of defining groups is by their presentations.
+Presentation of a group builds a group by specifying what is the set of
+generators - thus making a free group - a collection of equations that have
+to be satisfied in the resulting group. For example, we could take the
+generators to be the set (1) (giving out Z), and then add an equation (1 +
+1 + 1 = 0) - thus, getting (Z mod 3).
+
+5a. The generatos can be thought of as primitive operations in a language,
+and the group structure gives us the way of composing these operations and
+inverting them.
+
+6. Building a group like that is possible because of the universal property
+of the free group.
+Theorem: Universal property of a free group.
+Definition: formal definition of a free group.
+
+7. Why do it like that? Because it is a very syntactic approach to doing
+the theory. The way described previously operates directly in the semantics
+of the group operation - equality of elements (such as the result of adding
+two elements) has to be computed externally.
+Here, the equality can be computed using a rewriting system specified by
+the group axioms and the additional equations provided. However, because
+equations are not directed, it is not always possible to construct a
+well-behaving rewriting system. In general, the word problem (i.e. deciding
+if two words are equal) is proven to be undecidable.
+
+\subsection{Permutation Groups}
+
+8. The group that we are interested in is the group of permutations on a
+fininte set. We can define it formally in two different ways:
+8a. Taking the type Aut (Fin n) and showing that it satisfies the group
+axiom.
+8b. Writing up a presentation - choosing generating set and equations that
+have to be satisfied.
+
+9. By proving these two to be equivalent, we are bridging the gap between
+syntactic and semantic side in the whole proof - by assinging the meaning
+(a bijection) to a word (a program).
+9a. However, proving them equivalent will require, in particular, solving
+the word problem (in this specific instance).
+
+10. However, in the point 8b, there is an element of choice. One group can
+be presented in many different ways. For example, we could generate the
+permutations by using generators that:
+ - swap i-th element with j-th element
+ - swap i-th element with the first element
+ - shift elements in a cyclic way: operation with label "i" leaves elements
+with numbers > i in place, and element j <= i goes to (j + 1) mod i.
+
+11. One way of thinking about these presentations is to compare them to
+sorting algorithms with different primitive operations. If using a set of
+operations, one can sort any list, then these operations generate the
+permutations group. One weird example: pancake sorting.
+
+\subsection{Coxeter Presentation}
+
+12. Here, we are choosing a presentation based on the adjacent
+transpositions generators - i.e. primitive operations of the form "swap
+i-th and i+1 elements". In the language of sorting, this corresponds to
+what the insertion sort or bubble sort are doing.
+In the explanations below, we will use diagrams to draw the permutations.
+
+12a. There two very obvious laws that such operation has to preserve:
+ - Swapping the same two elements two times in a row does not change
+anything (cancel)
+TODO diagram.
+ - When swapping two pairs of elements on a positions far away, it does not
+matter in which order one does it (swap). (We can slide the wires).
+TODO diagram.
+But there is also one more law that these operations have to satisfy, in
+the situation when we are swapping adjacent pairs of elements. It is best
+seen on a diagram (braid)
+TODO diagram.
+
+This construction is called a Coxeter presentation. Writing it formally, we
+get the following definition
+Definition: Coxeter presentation.
+
+13. There are two main considerations to this choice:
+13a. How well the presentation matches the syntactic side of Pi? This is,
+how closely does the operational semantics built from these particular
+primitive operations match the natural operational semantics of Pi?
+13b. How convenient is it to turn the presentation it into rewriting system
+and solve the word problem for it?
+
+Thus, the argument:
+14. TODO I don't know what to write here, we have to think.
+
+15a. The equations that define Coxeter presentation have a nice property:
+as it is written, the LHS is smaller (lexicographically) than the RHS.
+Thus, if we could direct them from right to left, it would have a
+termination property out of the box.
+15b. However, it is not at all obvious that the system would be confluent.
+There are many possible critical pairs, such as
+TODO diagram 1212 (braid, braid)
+or
+TODO diagram 3233 (braid, cancel)
+As it turns out, there are instances of critical pairs that do not resolve.
+A simplest example is the following:
+TODO diagram 3213
+
+16. The equations have to be changed. We thus propose a new presentation.
+We keep the (cancel) and (swap) equations, but replace (braid) with a more
+general:
+TODO diagram:
+
+Definition: Long Coxeter presentation.
+
+17. In return for increased complexity of the equations, it satisfies the
+diamond property (if we direct the equation). Thus, together with the
+(preserved) termination, it creates a confluent rewriting system.
+Proof: All the critical pairs resolve, in the appendix.
+Definition: A normal form is a list that cannot be reduced any further.
+
+18. We define the equivalence classes on List (Fin n) to be precisely these
+elements that have the same normal form.
+
+19. However, the Long Coxeter relations are unwieldly and have no
+connectino to the broader scope of computational group theory - while the
+usual Coxeter presentation is to-go choice in this area (cite all these
+papers in Coq, this guy from Sweden(?) etc.).
+Thus, we prove:
+Proposition: Long Coxeter is equivalent to normal Coxeter.
+
+19a. (Note: here we try to convince the reader about reusability badge) So,
+the Long Coxeter can be regarded as a trick to get the proof to pass - but
+the external user-facing interface is still that of the usual Coxeter
+presentation.
+
+20. Now, to prove the equivalence of 9, on one side we have Aut(Fin n), and
+on the other, equivalence classes of List (Fin n) wrt to the usual Coxeter
+group, we would define two functions from and back. The easiest way to
+define a function out of the group presentation is to define them on the
+equivalence classes' representatives. But to do that, we have to know how
+do these representatives look like exactly.
+
+\subsection{Lehmer Codes}
+
+21. Enter: Lehmer codes.
+TODO diagram with List (Fin n) being divided into equivalence classes, and
+Lehmer codes being an image of immersion, being the normal form
+(representative) in each class.
+
 \subsection{Permutations}~\label{subsec:permutations}

### 2021-06-17 Pi+/paper/ufin.tex
beginning of new section on ufin

\section{The Universe of Finite Types}
\label{sec:ufin}

In order to defiine the semantics, we seek a mathematical structure satisfying the following properties: (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust enough to ensure that equivalent encodings of finite types are identified. \note{Is this the correct informal explanation of Ufin?}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% fill-column: 120
%%% End:

### 2021-06-18 Pi+/paper/reversible.tex
Comment out the missing figures
denot semantics: where is the def of eval2 ?

@@
 %%%%%%%%%
+\subsection{Denotational Semantics}
+
+We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
+compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
+$\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:
+
+\[\begin{array}{rcll}
+\sem{A} &=& (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
+\end{array}\]
+
+\noindent The denotation $\sem{c}$ of a combinator $c : A \isot B$ is a path between $\sem{A}$ and $\sem{B}$. If the
+size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
+$(\mathsf{Fin}~n, [ n , \mathsf{refl} ])$. This path is directly constructed using $\mathit{ap}$ and the fact that $m=n$
+since combinators are always between types of the same size.
+
+\noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \iso_2 q$, the denotation
+$\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.
+
+%%%%%%%%%
 \subsection{Examples}
@@
 
-\inkscape{reverse-3-bit.pdf}
+% \inkscape{reverse-3-bit.pdf}
 
@@
 
-\inkscape{if-c1-c2.pdf}
+% \inkscape{if-c1-c2.pdf}
 
@@
 %%subcode-line{pdfimage}[diagrams/if_c.pdf]
-\includegraphics{inkscape/cnot.pdf}
+% \includegraphics{inkscape/cnot.pdf}
 }
@@
 \scalebox{1.6}{
-\includegraphics{inkscape/toffoli.pdf}
+% \includegraphics{inkscape/toffoli.pdf}
 }

### 2021-06-18 Pi+/paper/reversible1.tex
restructure and work on denot sem

--- 2021-06-16 Pi+/paper/reversible.tex
@@
-\section{Reversible Programming Languages}~\label{sec:reversible}
+\section{A Reversible Programming Language: Set-Theoretic Semantics}~\label{sec:reversibleone}
 
@@
 \end{enumerate}
-More modern approaches recognize that reversible programming languages require a fresh approach and should be designed
-from first principles without the detour via conventional irreversible
+More foundational approaches recognize that reversible programming languages require a fresh approach and should be
+designed from first principles without the detour via conventional irreversible
 languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047,
@@
 
-A natural candidate for a semantic foundation for reversible programming languages is the notion of type isomorphism
-(equivalence). Indeed, the type isomorphisms among finite types represent a universal language for reversible boolean
+A natural candidate for a semantic foundation for reversible programming languages is the notion of type
+isomorphism. Indeed, the type isomorphisms among finite types represent a universal language for reversible boolean
 circuits~\cite{James:2012:IE:2103656.2103667} and the extension with recursive types and trace
@@
 
-\begin{figure}[t]
-  {\scalebox{\scalef}{%
- $\begin{array}{c}
- \infer{\interp{c}{v} \Downarrow \delta(c,v)}{c ~\mbox{base combinator}}
-\qquad
-\infer{\interp{c_1\oplus c_2}{\inlv{v}} \Downarrow \inlv{w}}
-    {\interp{c_1}{v} \Downarrow w}
-\qquad
-\infer{\interp{c_1\oplus c_2}{\inrv{v}} \Downarrow \inrv{w}}
-    {\interp{c_2}{v} \Downarrow w} \\
+%%%%%%%%%
+\subsection{Denotational Semantics}
+
+\noindent A simple denotational semantics of $\Pi$ can be specified by mapping each type to a finite set and each
+combinator to a permutation between finite sets.
+
+Fig.~\ref{fig:delta} specifies the semantics of the base combinators
+using a function $\delta$. It is clear that this function has an inverse $\delta^\dagger$ such that
+$\delta^\dagger(c,\delta(c,v))=v$. The semantics of full programs is given by the interpreter in the second figure.
+
+We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
+compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
+$\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:
+
+\[\begin{array}{rcll}
+    \sem{A} &=& \mathsf{Fin}~\sizet{A}
 \\
- \infer{\interp{\idc}{v} \Downarrow v}{}
-\qquad
-\infer{\interp{c_1\fatsemi c_2}{v} \Downarrow v_2}
-    {\interp{c_1}{v} \Downarrow v_1 \quad \interp{c_2}{v_1} \Downarrow v_2}
-\qquad
-\infer{\interp{c_1\otimes c_2}{(v_1 , v_2)} \Downarrow (w_1 , w_2)}
-    {\interp{c_1}{v_1} \Downarrow w_1 \quad \interp{c_2}{v_2} \Downarrow w_2}
-\end{array}$}}
-\caption{$\Pi$-interpreter}\label{fig:pi:interp}
-\end{figure}
-
-\noindent To formalize the semantics of $\Pi$, we first specify the
-semantics of the base combinators in Fig.~\ref{fig:delta} using a
-function $\delta$. It is clear that this function has an inverse
-$\delta^\dagger$ such that $\delta^\dagger(c,\delta(c,v))=v$. The semantics of full programs is given by the interpreter
-in Fig.~\ref{fig:pi:interp}.
+    \sem{c : A \isot B} &:& \mathsf{Fin}~\sizet{A} \rightarrow \mathsf{Fin}~\sizet{B} \\
+    %% base combinators
+    \sem{c_1\oplus c_2}_{A+B \isot C+D}~\inlv{v} &=& \sem{c_1}_{A \isot C}~v \\
+    \sem{c_1\oplus c_2}_{A+B \isot C+D}~\inrv{v} &=& \sizet{A} + \sem{c_2}_{B \isot D}~v \\
+    \sem{\idc}~v &=& v \\
+    \sem{c_1\fatsemi c_2}~v &=& \sem{c_2} (\sem{c_1}~v) \\
+    \sem{c_1\otimes c_2}~(v_1,v_2) &=& (\sem{c_1}~v_1 , \sem{c_2}~v_2)
+\end{array}\]
 
@@
 \scalebox{1.6}{
-\includegraphics{inkscape/toffoli.pdf}
+% \includegraphics{inkscape/toffoli.pdf}
 }
 \end{center}
-
-%%%%%%%%%
-\subsection{A Language of Equivalences between Type Equivalences}
-\label{langeqeq}
-
-In the previous section, we examined equivalences between conventional data structures, i.e., structured trees of
-values. We now consider a richer but foundational notion of data: programs themselves. Indeed, universal computation
-models crucially rely on the fact that \emph{programs are (or can be encoded as) data}, e.g., a Turing machine can be
-encoded as a string that another Turing machine (or even the same machine) can manipulate. Similarly, first-class
-functions are the \emph{only} values in the $\lambda$-calculus.  In our setting, we ask whether the programs developed
-in the previous section can themselves be subject to (higher-level) equivalences?
-
-We will explain the ideas using two small exaamples. Consider the following two programs mapping between the types
-$A + B$ and $C+D$:
-
-\begin{center}
-\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
-  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
-  \draw[fill] (-2,-1.5) circle [radius=0.025];
-  \node[below] at (-2.1,-1.5) {$A$};
-  \node[below] at (-2.1,-1.9) {$+$};
-  \draw[fill] (-2,-2.5) circle [radius=0.025];
-  \node[below] at (-2.1,-2.5) {$B$};
-
-  \draw[fill] (6.5,-1.5) circle [radius=0.025];
-  \node[below] at (6.7,-1.5) {$C$};
-  \node[below] at (6.7,-1.9) {$+$};
-  \draw[fill] (6.5,-2.5) circle [radius=0.025];
-  \node[below] at (6.7,-2.5) {$D$};
-
-  \draw[<-] (-2,-1.5) to[bend left] (1,0.5) ;
-  \draw[<-] (-2,-2.5) to[bend left] (1,-0.5) ;
-  \draw[->] (3.5,0.5) to[bend left] (6.5,-1.45) ;
-  \draw[->] (3.5,-0.5) to[bend left] (6.5,-2.45) ;
-
-  \draw[<-] (-2,-1.5) to[bend right] (1,-3.5) ;
-  \draw[<-] (-2,-2.5) to[bend right] (1,-4.5) ;
-  \draw[->] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
-  \draw[->] (3.5,-4.5) to[bend right] (6.5,-2.55) ;
-
-  \draw     (2,0.5)  -- (2.5,0.5)  ;
-  \draw     (2,-0.5) -- (2.5,-0.5) ;
-
-  \draw     (2.5,0.5)  -- (3.5,-0.5)  ;
-  \draw     (2.5,-0.5) -- (3.5,0.5) ;
-
-  \draw     (1,-3.5)  -- (2,-4.5)    ;
-  \draw     (1,-4.5) -- (2,-3.5)   ;
-
-  \draw     (2,-3.5)  -- (2.5,-3.5)    ;
-  \draw     (2,-4.5) -- (2.5,-4.5)   ;
-
-  \path (1.5,0.5) node (tc1) [func] {$c_1$};
-  \path (1.5,-0.5) node (tc2) [func] {$c_2$};
-  \path (3,-4.5) node (bc1) [func] {$c_1$};
-  \path (3,-3.5) node (bc2) [func] {$c_2$};
-\end{tikzpicture}
-\end{center}
-The top path is the $\Pi$ program $(c_1~\oplus~c_2)~\odot~\swapp$ which acts on the type $A$ by $c_1$, acts on the type
-$B$ by $c_2$, and acts on the resulting value by a twist that exchanges the two injections into the sum type. The bottom
-path performs the twist first and then acts on the type $A$ by $c_1$ and on the type $B$ by $c_2$ as before. One could
-imagine the paths are physical \emph{elastic} wires in $3$ space, where the programs $c_1$ and $c_2$ as arbitrary
-deformations on these wires, and the twists do not touch but are in fact well-separated. Then, holding the points $A$,
-$B$, $C$, and $D$ fixed, it is possible to imagine sliding $c_1$ and $c_2$ from the top wire rightward past the twist,
-and then using the elasticity of the wires, pull the twist back to line up with that of the bottom --- thus making both
-parts of the diagram identical.  Each of these moves can be undone (reversed), and doing so would take the bottom part
-of the diagram into the top part.  In other words, there exists an equivalence of the program
-$(c_1~\oplus~c_2)~\odot~\swapp$ to the program $\swapp \odot (c_2~\oplus~c_1)$. We can also show that this means that,
-as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and $\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not all
-programs between the same types can be deformed into one another. The simplest example of inequivalent deformations are
-the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.
-
-As another example, consider consider a circuit that takes an input type consisting of three values \Tree [ {\small a} [
-{\small b} {\small c} ] ]~ and swaps the leftmost value with the rightmost value to produce \Tree [ {\small c} [ {\small
-  b} {\small a} ] ]~.  We can implement two such circuits using our Agda library for $\Pi$:
-
-\begin{code}%
-\>[0]\AgdaFunction{swap{-}fl1}\AgdaSpace{}%
-\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
-\AgdaSymbol{:}\AgdaSpace{}%
-\AgdaSymbol{\{}\AgdaBound{a}\AgdaSpace{}%
-\AgdaBound{b}\AgdaSpace{}%
-\AgdaBound{c}\AgdaSpace{}%
-\AgdaSymbol{:}\AgdaSpace{}%
-\AgdaDatatype{U}\AgdaSymbol{\}}\AgdaSpace{}%
-\AgdaSymbol{→}\AgdaSpace{}%
-\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
-\AgdaBound{a}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
-\AgdaBound{b}\AgdaSpace{}%
-\AgdaBound{c}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaDatatype{⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
-\AgdaBound{c}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
-\AgdaBound{b}\AgdaSpace{}%
-\AgdaBound{a}\AgdaSymbol{)}\<%
-\\
-\>[0]\AgdaFunction{swap{-}fl1}\AgdaSpace{}%
-\AgdaSymbol{=}\AgdaSpace{}%
-\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\AgdaSpace{}%
-\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\<%
-\\
-%
-\\[\AgdaEmptyExtraSkip]%
-\>[0]\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
-\AgdaSymbol{=}%
-\>[52I]\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\<%
-\\
-\>[.]\<[52I]%
-\>[11]\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\<%
-\\
-%
-\>[11]\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{id⟷}\AgdaSymbol{)}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\<%
-\\
-%
-\>[11]\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
-\AgdaInductiveConstructor{◎}\<%
-\\
-%
-\>[11]\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
-\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
-\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\<%
-\end{code}
-
-\noindent The first implementation rewrites the incoming values as follows:
-\[
-\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
-\Tree [ [ {\small a} {\small b} ] {\small c} ] ~\to~
-\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
-\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
-\]
-\noindent
-The second implementation rewrites the incoming values as follows:
-\[
-\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
-\Tree [ {\small a} [ {\small c} {\small b} ] ] ~\to~
-\Tree [ [ {\small a} {\small c} ] {\small b} ] ~\to~
-\Tree [ [ {\small c} {\small a} ] {\small b} ] ~\to~
-\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
-\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
-\]
-\noindent The two circuits are extensionally equal. Using the level-2
-isomorphisms we can \emph{explicitly} construct a sequence of
-rewriting steps that transforms the second circuit to the first.
-
-Recalling that the $\lambda$-calculus arises as the internal language of Cartesian Closed Categories
-(Elliott~\cite{Elliott-2017} gives a particularly readable account of this), we can think of $\Pi$ in similar terms, but
-for symmetric Rig Groupoids instead. For example, we can ask what does the equivalence above represent? It is actually a
-``linear'' representation of a 2-categorial commutative diagram! In fact, it is a painfully verbose version thereof, as
-it includes many \emph{refocusing} steps because our language does not build associativity into its syntax. Categorical
-diagrams usually do.  Thus if we rewrite the example in diagrammatic form, eliding all uses of associativity, but
-keeping explicit uses of identity transformations, we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents
-
-\vspace*{3mm}
-\begin{tikzcd}[column sep=normal, row sep=normal]
- && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
- & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
-a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
-  \arrow [dr, "\assoclp"]
-  \arrow [ddr, swap, "\assoclp"]
-    & (a+b)+c \arrow [r, "\swapp"] &
-    c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}]
-    & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
- & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
- & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
-             \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] &&
-    c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
- && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
- \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
- \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
-\end{tikzcd}
 

### 2021-06-18 Pi+/paper/reversible2.tex
restructure and work on denot sem

--- 2021-06-04 Pi+/paper/reversible.tex
@@
-\section{Reversible Programming Languages}~\label{sec:reversible}
+\section{A Reversible Programming Language: Groupoid Semantics}
+\label{sec:reversibletwo}
+\label{langeqeq}
 
-\todo{Not the right title.}
+In the previous section, we examined equivalences between conventional data structures, i.e., structured trees of
+values. We now consider a richer but foundational notion of data: programs themselves. Indeed, universal computation
+models crucially rely on the fact that \emph{programs are (or can be encoded as) data}, e.g., a Turing machine can be
+encoded as a string that another Turing machine (or even the same machine) can manipulate. Similarly, first-class
+functions are the \emph{only} values in the $\lambda$-calculus.  In our setting, we ask whether the programs developed
+in the previous section can themselves be subject to (higher-level) equivalences?
 
-\note{This section should explain the main technical parts of the paper
-  informally, without using any technology. Use an example, such as, a
-  reversible language with $\leq 5$ bits, and examples of permutations and
-  transpositions, and when they're equal.}
+We will explain the ideas using two small exaamples. Consider the following two programs mapping between the types
+$A + B$ and $C+D$:
 
-\note{Motivation: There are two reversible circuits which describe the following permutation. They can be shown to be
-  equal using the 2-combinators.}
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
+  \draw[fill] (-2,-1.5) circle [radius=0.025];
+  \node[below] at (-2.1,-1.5) {$A$};
+  \node[below] at (-2.1,-1.9) {$+$};
+  \draw[fill] (-2,-2.5) circle [radius=0.025];
+  \node[below] at (-2.1,-2.5) {$B$};
 
+  \draw[fill] (6.5,-1.5) circle [radius=0.025];
+  \node[below] at (6.7,-1.5) {$C$};
+  \node[below] at (6.7,-1.9) {$+$};
+  \draw[fill] (6.5,-2.5) circle [radius=0.025];
+  \node[below] at (6.7,-2.5) {$D$};
+
+  \draw[<-] (-2,-1.5) to[bend left] (1,0.5) ;
+  \draw[<-] (-2,-2.5) to[bend left] (1,-0.5) ;
+  \draw[->] (3.5,0.5) to[bend left] (6.5,-1.45) ;
+  \draw[->] (3.5,-0.5) to[bend left] (6.5,-2.45) ;
+
+  \draw[<-] (-2,-1.5) to[bend right] (1,-3.5) ;
+  \draw[<-] (-2,-2.5) to[bend right] (1,-4.5) ;
+  \draw[->] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
+  \draw[->] (3.5,-4.5) to[bend right] (6.5,-2.55) ;
+
+  \draw     (2,0.5)  -- (2.5,0.5)  ;
+  \draw     (2,-0.5) -- (2.5,-0.5) ;
+
+  \draw     (2.5,0.5)  -- (3.5,-0.5)  ;
+  \draw     (2.5,-0.5) -- (3.5,0.5) ;
+
+  \draw     (1,-3.5)  -- (2,-4.5)    ;
+  \draw     (1,-4.5) -- (2,-3.5)   ;
+
+  \draw     (2,-3.5)  -- (2.5,-3.5)    ;
+  \draw     (2,-4.5) -- (2.5,-4.5)   ;
+
+  \path (1.5,0.5) node (tc1) [func] {$c_1$};
+  \path (1.5,-0.5) node (tc2) [func] {$c_2$};
+  \path (3,-4.5) node (bc1) [func] {$c_1$};
+  \path (3,-3.5) node (bc2) [func] {$c_2$};
+\end{tikzpicture}
+\end{center}
+The top path is the $\Pi$ program $(c_1~\oplus~c_2)~\odot~\swapp$ which acts on the type $A$ by $c_1$, acts on the type
+$B$ by $c_2$, and acts on the resulting value by a twist that exchanges the two injections into the sum type. The bottom
+path performs the twist first and then acts on the type $A$ by $c_1$ and on the type $B$ by $c_2$ as before. One could
+imagine the paths are physical \emph{elastic} wires in $3$ space, where the programs $c_1$ and $c_2$ as arbitrary
+deformations on these wires, and the twists do not touch but are in fact well-separated. Then, holding the points $A$,
+$B$, $C$, and $D$ fixed, it is possible to imagine sliding $c_1$ and $c_2$ from the top wire rightward past the twist,
+and then using the elasticity of the wires, pull the twist back to line up with that of the bottom --- thus making both
+parts of the diagram identical.  Each of these moves can be undone (reversed), and doing so would take the bottom part
+of the diagram into the top part.  In other words, there exists an equivalence of the program
+$(c_1~\oplus~c_2)~\odot~\swapp$ to the program $\swapp \odot (c_2~\oplus~c_1)$. We can also show that this means that,
+as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and $\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not all
+programs between the same types can be deformed into one another. The simplest example of inequivalent deformations are
+the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.
+
+As another example, consider consider a circuit that takes an input type consisting of three values \Tree [ {\small a} [
+{\small b} {\small c} ] ]~ and swaps the leftmost value with the rightmost value to produce \Tree [ {\small c} [ {\small
+  b} {\small a} ] ]~.  We can implement two such circuits using our Agda library for $\Pi$:
+
+\begin{code}%
+\>[0]\AgdaFunction{swap{-}fl1}\AgdaSpace{}%
+\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{a}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaDatatype{U}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{a}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaDatatype{⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{a}\AgdaSymbol{)}\<%
+\\
+\>[0]\AgdaFunction{swap{-}fl1}\AgdaSpace{}%
+\AgdaSymbol{=}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\<%
+\\
+%
+\\[\AgdaEmptyExtraSkip]%
+\>[0]\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
+\AgdaSymbol{=}%
+\>[52I]\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+\>[.]\<[52I]%
+\>[11]\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\<%
+\end{code}
+
+\noindent The first implementation rewrites the incoming values as follows:
 \[
-  \begin{tikzpicture}
-    \begin{knot}[clip width=5]
-      \filldraw (0,5) circle (2pt) node[above] {0};
-      \filldraw (1,5) circle (2pt) node[above] {1};
-      \filldraw (2,5) circle (2pt) node[above] {2};
-      \filldraw (3,5) circle (2pt) node[above] {3};
-      \filldraw (4,5) circle (2pt) node[above] {4};
-      \filldraw (0,0) circle (2pt) node[below] {1};
-      \filldraw (1,0) circle (2pt) node[below] {4};
-      \filldraw (2,0) circle (2pt) node[below] {0};
-      \filldraw (3,0) circle (2pt) node[below] {3};
-      \filldraw (4,0) circle (2pt) node[below] {2};
-      \strand (0,5) .. controls (0.5,0.5) and (1.5,3.5) .. (2,0);
-      \strand (1,5) .. controls (0.75,0.5) and (0.25,3.5) .. (0,0);
-      \strand (2,5) .. controls (2.5,2.5) and (3.5,1.5) .. (4,0);
-      \strand (3,5) .. controls (4.5,2.5) and (4,1.5) .. (3,0);
-      \strand (4,5) .. controls (3.5,2.5) and (1.5,2.5) .. (1,0);
-      \flipcrossings{4,5};
-    \end{knot}
-  \end{tikzpicture}
+\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+\Tree [ [ {\small a} {\small b} ] {\small c} ] ~\to~
+\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
 \]
+\noindent
+The second implementation rewrites the incoming values as follows:
+\[
+\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+\Tree [ {\small a} [ {\small c} {\small b} ] ] ~\to~
+\Tree [ [ {\small a} {\small c} ] {\small b} ] ~\to~
+\Tree [ [ {\small c} {\small a} ] {\small b} ] ~\to~
+\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+\]
+\noindent The two circuits are extensionally equal. Using the level-2
+isomorphisms we can \emph{explicitly} construct a sequence of
+rewriting steps that transforms the second circuit to the first.
 
-\note{Example: We reduce $\mathsf{swap} : 2 + 2 \leftrightarrow 2 + 2$ to a sequence of adjacent swaps. This is an
-  example of a translation from $\PiPlusLang$ to $\PiHatLang$.}
+Recalling that the $\lambda$-calculus arises as the internal language of Cartesian Closed Categories
+(Elliott~\cite{Elliott-2017} gives a particularly readable account of this), we can think of $\Pi$ in similar terms, but
+for symmetric Rig Groupoids instead. For example, we can ask what does the equivalence above represent? It is actually a
+``linear'' representation of a 2-categorial commutative diagram! In fact, it is a painfully verbose version thereof, as
+it includes many \emph{refocusing} steps because our language does not build associativity into its syntax. Categorical
+diagrams usually do.  Thus if we rewrite the example in diagrammatic form, eliding all uses of associativity, but
+keeping explicit uses of identity transformations, we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents
 
-\[
-  \begin{tikzpicture}
-    \begin{knot}[clip width=4]
-      \filldraw (0,4) circle (2pt) node[above] {0};
-      \filldraw (1,4) circle (2pt) node[above] {1};
-      \filldraw (2,4) circle (2pt) node[above] {2};
-      \filldraw (3,4) circle (2pt) node[above] {3};
-      \filldraw (0,0) circle (2pt) node[below] {2};
-      \filldraw (1,0) circle (2pt) node[below] {3};
-      \filldraw (2,0) circle (2pt) node[below] {0};
-      \filldraw (3,0) circle (2pt) node[below] {1};
-      \strand (0,4) .. controls (0.5,1.5) and (1.5,2.5) .. (2,0);
-      \strand (1,4) .. controls (1.5,1.5) and (2.5,2.5) .. (3,0);
-      \strand (2,4) .. controls (1.5,1.5) and (1.5,2.5) .. (0,0);
-      \strand (3,4) .. controls (2.5,1.5) and (2.5,2.5) .. (1,0);
-    \end{knot}
-  \end{tikzpicture}
-\]
+\vspace*{3mm}
+\begin{tikzcd}[column sep=normal, row sep=normal]
+ && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
+ & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
+a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
+  \arrow [dr, "\assoclp"]
+  \arrow [ddr, swap, "\assoclp"]
+    & (a+b)+c \arrow [r, "\swapp"] &
+    c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}]
+    & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
+ & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
+ & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
+             \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] &&
+    c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
+ && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
+ \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
+ \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
+\end{tikzcd}
 
-\begin{align*}
-  \begin{tikzpicture}
-    \begin{knot}[clip width=4]
-      \filldraw (0,4) circle (2pt) node[above] {0};
-      \filldraw (1,4) circle (2pt) node[above] {1};
-      \filldraw (2,4) circle (2pt) node[above] {2};
-      \filldraw (3,4) circle (2pt) node[above] {3};
-      \filldraw (0,0) circle (2pt) node[below] {0};
-      \filldraw (1,0) circle (2pt) node[below] {2};
-      \filldraw (2,0) circle (2pt) node[below] {1};
-      \filldraw (3,0) circle (2pt) node[below] {3};
-      \strand (0,4) to (0,0);
-      \strand (1,4) .. controls (0.5,2) and (2.5,2) .. (2,0);
-      \strand (2,4) .. controls (2.5,2) and (0.5,2) .. (1,0);
-      \strand (3,4) to (3,0);
-    \end{knot}
-  \end{tikzpicture}
-  &&
-    \begin{tikzpicture}
-      \begin{knot}[clip width=4]
-        \filldraw (0,4) circle (2pt) node[above] {0};
-        \filldraw (1,4) circle (2pt) node[above] {2};
-        \filldraw (2,4) circle (2pt) node[above] {1};
-        \filldraw (3,4) circle (2pt) node[above] {3};
-        \filldraw (0,0) circle (2pt) node[below] {2};
-        \filldraw (1,0) circle (2pt) node[below] {0};
-        \filldraw (2,0) circle (2pt) node[below] {1};
-        \filldraw (3,0) circle (2pt) node[below] {3};
-        \strand (0,4) .. controls (-0.5,2) and (1.5,2) .. (1,0);
-        \strand (1,4) .. controls (1.5,2) and (-0.5,2) .. (0,0);
-        \strand (2,4) to (2,0);
-        \strand (3,4) to (3,0);
-      \end{knot}
-    \end{tikzpicture}
-  \\
-  \begin{tikzpicture}
-    \begin{knot}[clip width=4]
-      \filldraw (0,4) circle (2pt) node[above] {2};
-      \filldraw (1,4) circle (2pt) node[above] {0};
-      \filldraw (2,4) circle (2pt) node[above] {1};
-      \filldraw (3,4) circle (2pt) node[above] {3};
-      \filldraw (0,0) circle (2pt) node[below] {2};
-      \filldraw (1,0) circle (2pt) node[below] {0};
-      \filldraw (2,0) circle (2pt) node[below] {3};
-      \filldraw (3,0) circle (2pt) node[below] {1};
-      \strand (0,4) to (0,0);
-      \strand (1,4) to (1,0);
-      \strand (2,4) .. controls (1.5,2) and (3.5,2) .. (3,0);
-      \strand (3,4) .. controls (3.5,2) and (1.5,2) .. (2,0);
-    \end{knot}
-  \end{tikzpicture}
-  &&
-    \begin{tikzpicture}
-      \begin{knot}[clip width=4]
-        \filldraw (0,4) circle (2pt) node[above] {2};
-        \filldraw (1,4) circle (2pt) node[above] {0};
-        \filldraw (2,4) circle (2pt) node[above] {3};
-        \filldraw (3,4) circle (2pt) node[above] {1};
-        \filldraw (0,0) circle (2pt) node[below] {2};
-        \filldraw (1,0) circle (2pt) node[below] {3};
-        \filldraw (2,0) circle (2pt) node[below] {0};
-        \filldraw (3,0) circle (2pt) node[below] {1};
-        \strand (0,4) to (0,0);
-        \strand (1,4) .. controls (0.5,2) and (2.5,2) .. (2,0);
-        \strand (2,4) .. controls (2.5,2) and (0.5,2) .. (1,0);
-        \strand (3,4) to (3,0);
-      \end{knot}
-    \end{tikzpicture}
-\end{align*}
+%%%%%%%%%
+\subsection{Denotational Semantics}
 
-\note{This might be followed by a section which explains the syntax of Pi.}
+We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
+compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
+$\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:
+
+\[\begin{array}{rcll}
+\sem{A} &=& (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
+\end{array}\]
+
+\noindent The denotation $\sem{c}$ of a combinator $c : A \isot B$ is a path between $\sem{A}$ and $\sem{B}$. If the
+size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
+$(\mathsf{Fin}~n, [ n , \mathsf{refl} ])$. This path is directly constructed using $\mathit{ap}$ and the fact that $m=n$
+since combinators are always between types of the same size.
+
+\noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \iso_2 q$, the denotation
+$\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.
 

### 2021-06-18 Pi+/paper/ufin.tex
restructure and work on denot sem

@@
 
-In order to defiine the semantics, we seek a mathematical structure satisfying the following properties: (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust enough to ensure that equivalent encodings of finite types are identified.
+Previous semantics adequate for reasoning about evaluation and for checking the equivalence of two circuits.
+
+Want NBE, want rewriting system of program equivalence, what else? So we need to move to proof-relevant higher-category
+groupoid semantics instead of set-theoretic.
+
+The lambda calculus is classically presented with types, terms, and
+$\beta\eta$-equations. This can be turned into a 1-category, with 0-cells for
+types, and 1-cells for programs quotiented by congruence closure of the
+$\beta\eta$-equations.
+
+Then there is the free CCC, which is a 1-category constructed using
+0-cells and 1-cells with universal properties that hold upto the notion
+of equality that from the ambient metatheory.
+
+Lambek/Scott show that these are equivalent. In modern language, this is
+a 1-equivalence of (1,1)-categories.
+
+Now, lets work without quotients in a proof-relevant metatheory.
+
+The lambda calculus can now be presented with 0-cells for types, 1-cells
+for terms, and 2-cells for equations/rewrites. This is a
+(2,1)-category. Then, you can give a description of the free 2-CCC,
+where universal properties hold upto canonical isomorphism.
+
+Then we can show that there is a 2-equivalence of these
+(2,1)-categories. (Actually, it can be weaker than (2,1)-categories, in
+fact it holds for bicategories.)
+
+Now, lets go to our result.
+
+For Pi/Pi+, we present it as a (2,0)-category (or 1-groupoid à la HoTT),
+with 0-cells for types, 1-cells for reversible programs, 2-cells for
+rewrites/equations. And we present different descriptions (Pi+, Ufin)
+of the free (2,0)-SMC (or 1-SMG) and show they're 2-equivalent.
+
+So this is just a difference of presentation, in going from the (1,r) to
+the (2,r) category. We have simply made the equations/rewrites
+first-class in the presentation.
+
+Now the question is, what does it mean for a compiler?
+
+My opinion is that it is still the same, what rewrites to apply depends
+on the choices made by the compiler writer, but we can think about that.
+
+%%%
+\subsection{Univalent Subuniverses}
+\label{sec:univalent}
+
+In order to define the semantics, we seek a mathematical structure satisfying the following properties: (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust enough to ensure that equivalent encodings of finite types are identified.
 \note{Is this the correct informal explanation of Ufin?}
+
+ In this section, we introduce some basic concepts and notation that we use in Homotopy Type Theory. Then, we define
+univalent subuniverses and discuss some specific examples.
+
+\subsection{The Type Theory}~\label{subsec:type-theory}
+
+We work in Homotopy Type Theory, in particular, intensional Martin-L\"{o}f Type Theory, with a univalent universe, and
+Higher Inductive Types (HITs) for propositional truncations and set quotients. We recall a few basic facts to
+familiarise the reader with the notation we use. For more details about basics of HoTT, we refer the reader to the HoTT
+book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.
+
+\note{It is important that we work in HoTT using it as a metatheory, we care about proof relevance, constructivity, and
+  working with higher-dimensional algebraic structures, such as groupoids. Our results are also formalised using general
+  tools and principles from HoTT, and computer checked using a proof assistant, using Agda and the HoTT-Agda library.}
+
+\subsubsection{Identity Types}
+
+\vc{I'm explaining everything starting from the categorical language. We do care more about the groupoid structure of
+  types, and how to encode groupoids using types, since that's what we use in~\cref{sec:finite}.}
+
+In HoTT, the intensional identity type is the type of paths between two terms of the same type. Given two terms $x:A$
+and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the equality type between them. In book HoTT, the identity
+type is generated by reflexivity $\refl_{x}$, and the eliminator for the identity type is given by path induction or the
+$J$-rule. The identity type equips each type with the structure of an infinity groupoid, or a homotopy type.
+
+\todo{Say why it is a groupoid by listing some groupoid laws.}
+
+Functions between types are functors between groupoids. Given a function $f : A \to B$, the functorial action is given
+by
+
+\[
+  \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
+\]
+
+Type families are functions from a type to the universe, that is, an indexed family of groupoids. The $\term{transport}$
+operation gives the functorial action of paths in the indexing type, which is defined by path induction. If
+$P : A \to \UU$ is a type family, then for a path $x \id_{A} y$, we have
+
+\[
+  \transport{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
+\]
+
+\todo{Why is the topological viewpoint important, or saying fibrations? Explain the intuition using classifying spaces,
+  maybe draw some diagrams.}
+
+From the topological viewpoint, a type family can also be seen as a fibration. For a type family $P : A \to \UU$ and a
+point $x : A$, $P(x)$ gives the fiber over $a$. For a path $p : x \id_{A} y$, $\transport{P}{p}$ gives the path lifting
+operation. The total space is given by $\dsum{x:A}{P(x)}$ and the first projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ to
+the base space is the fibration. The lifting operation lifts paths in the base space to paths in the total space. If
+$p : x \id_{A} y$ is a path in the base space, and $u : P(x)$, we have
+
+\[
+  \term{lift}(u,p) : (x , u) \id_{\dsum{x:A}{P(x)}} (y , \tr{p}{u})
+\]
+
+where $\tr{p}{u}$ is shorthand for $\transport{P,p}(u)$.
+
+Further, using the groupoid structure of $A$, we can show that transport lifts paths to equivalences, we define
+
+\[
+  \tptEqv{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
+\]
+
+\todo{Explain motivation.}
+
+\begin{definition}[Univalent Fibration]
+  $P$ is a univalent type family (or simply a univalent fibration) if $\tptEqv{P}$ is an equivalence.
+\end{definition}
+
+\subsubsection{Univalence}
+
+Voevodsky's \emph{univalence} principle characterises paths in the universe. It says that equivalent types are equal, or
+the following function is an equivalence.
+
+\[
+  \ua : A \id_{\UU} B \to A \eqv B
+\]
+
+\todo{Revise.}
+
+Alternatively, one can say that the identity type family $\term{id} : \UU \to \UU$ is univalent.
+
+\subsubsection{Higher Inductive Types}
+
+\todo{Explain homotopy types, $\hProp$, $\hSet$, etc.}
+
+\vc{These are placeholder definitions, need informal explanations and references to the book.}
+
+\begin{definition}[Propositional Truncation]
+  Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, or simply $\Trunc{A}$, is a higher inductive type
+  generated by the following constructors,
+  \begin{itemize}
+    \item an inclusion function $\trunc{\blank} : A \to \Trunc{A}$,
+    \item for each $x, y : \Trunc{A}$, a path $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$,
+  \end{itemize}
+  such that, given any type $B$ with
+  \begin{itemize}
+    \item a function $g : A \to B$,
+    \item for each $x, y : B$, a path $\term{trunc*}(x,y) : x \id_{B} y$,
+  \end{itemize}
+  there is a unique function $f : \Trunc{A} \to B$ such that,
+  \begin{itemize}
+    \item $f(\trunc{a}) \equiv g(a)$
+    \item for each $x, y : \Trunc{A}$, $\ap{f}{\term{trunc}(x,y)} \id_{B} \term{trunc*}(f(x),f(y))$.
+  \end{itemize}
+\end{definition}
+
+\begin{definition}[Set Quotient]
+  Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
+  higher inductive type generated by
+  \begin{itemize}
+    \item an inclusion function $q : A \to \quot{A}{R}$,
+    \item for each $x, y : A$ such that $R(x,y)$, a path $q(x) \id_{\quot{A}{R}} q(y)$,
+    \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,
+  \end{itemize}
+  with \review{an appropriate induction principle.}
+\end{definition}
+
+We recall that quotients in HoTT are \emph{effective}, that is, if $R$ is an equivalence relation, we have
+$R(x,y) \eqv (q(x) \id_{\quot{A}{R}} q(y))$.
+
+\subsection{Univalent Subuniverses}
+
+\todo{Explain motivation. Starting from a univalent universe which classifies all types, we want to define a subuniverse
+  which classifies only certain types, for example, types that satisfy some property that we want.}
+
+\begin{definition}[Universe]
+  A universe \`{a} la Tarski is given by the following pieces of data,
+  \begin{itemize}
+    \item a code $U : \UU$,
+    \item a decoding function $\El : U \to \UU$.
+  \end{itemize}
+  If $\El$ is a univalent fibration, $U$ is a univalent universe.
+\end{definition}
+
+\begin{definition}[Subuniverse]
+  A subtype is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $\forall x, \isProp{P(x)}$. A
+  subuniverse generated by a subtype has $U \defeq \dsum{X:\UU}{P(X)}$ and $\El \defeq \fst$.
+\end{definition}
+
+\begin{proposition}[Univalent Subuniverse]
+  Subuniverses generated by subtypes are univalent.
+\end{proposition}
+
+\begin{proof}
+  Suppose $(U, \El) \defeq (\dsum{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
+  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
+  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
+  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
+  inverse follows by calculation using the appropriate computation rules.
+\end{proof}
+
+\begin{example}[$\BAut$]
+  The type of self-equivalences on a type $T$ is defined as $\Aut[T] \defeq T \eqv T$, which is an $\infty$-group. The
+  subuniverse of types that are merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the
+  classifying space of $\Aut[T]$, as justified by~\cref{lem:loop-deloop}. We write $T_{o} \defeq (T, \trunc{\refl_{T}})$
+  for the image of the inclusion of $T$.
+\end{example}
+
+\begin{example}
+  The univalent subuniverse of types merely equal to $\Bool$ is given by $\BAut[\Bool]$. This can be used to give the
+  denotational semantics of a 2-bit reversible language~\cite{caretteReversibleProgramsUnivalent2018}.
+\end{example}
+
+\begin{definition}[$\Fin$]
+  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by cardinality. It is defined equivalently in
+  two different ways,
+  \begin{gather*}
+    \begin{aligned}
+      \Fin[n] & \defeq \dsum{k:\Nat}{k < n}
+    \end{aligned}
+    \begin{aligned}
+      & \Fin[0] \defeq \bot \\
+      & \Fin[\suc[n]] \defeq \top \sqcup \Fin[n]
+    \end{aligned}
+  \end{gather*}
+  Note that $\Fin[n]$ is a $\hSet$, and we use the definitions interchangeably.
+\end{definition}
+
+\begin{example}
+  For any $n : \Nat$, we define the univalent subuniverse of types merely equal to $\Fin[n]$ as $\BAut[\Fin[n]]$. This
+  is the universe of finite sets with cardinality equal to $n$. We write $F_{n} \defeq (\Fin[n], \trunc{\refl})$ for the
+  image of the inclusion of $\Fin[n]$.
+\end{example}
+
+\begin{definition}[$\isFin$]
+  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n$.
+  \[
+    \isFin[X] \defeq \dsum{n:\Nat}{\SubP{X}{\Fin[n]}}
+  \]
+  Note that the natural number $n$ need not be truncated, as justified below.
+\end{definition}
+
+\begin{proposition}
+  For any type $X$, $\isFin[X]$ is a proposition.
+\end{proposition}
+
+\begin{proof}
+  Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
+  enough to show that $n \id m$. Since $\Nat$ is a set, this is a proposition, so we can use the induction principle of
+  propositional truncation to eliminate to $n \id m$, applying it on $\phi$ and $\psi$ respectively. This gives us the
+  equalities $X \id \Fin[n]$ and $X \id \Fin[m]$, which gives us $\Fin[n] \id \Fin[m]$, from which $n \id m$ follows by
+  applying the first projection.
+\end{proof}
+
+\begin{example}
+  The univalent subuniverse of \emph{all finite types} is given by
+  \[
+    \UFin \defeq \dsum{X:\UU}{\isFin[X]}.
+  \]
+  We write $F_{n} \defeq (\Fin[n], n, \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$. We use this as the
+  denotational semantics to interpret our reversible programming language $\PiPlusLang$.
+\end{example}
+
+\review{We characterise the path space of univalent subuniverses.}
+
+\begin{proposition}
+  If $T$ is an $n$-type, $\BAut[T]$ is an $(n+1)$-type.
+\end{proposition}
+
+\begin{proof}
+  We need to show that the equality type of $\BAut[T]$ is an $n$-type. Assume $X, Y : \BAut[T]$. Since $\BAut[T]$ is a
+  univalent subuniverse, we have $(X \id Y) \eqv (\fst(X) \eqv \fst(Y))$. Note that being an $n$-type is a proposition.
+  Since $T$ is an $n$-type, and $\fst(X)$ and $\fst(Y)$ are merely equal to $T$, they're also $n$-types. It follows that
+  $\fst(X) \eqv \fst(Y)$ is an $n$-type, and hence $X \id Y$ is an $n$-type.
+\end{proof}
+
+\begin{proposition}
+  For any $T : \UU$, $\BAut[T]$ is 0-connected.
+\end{proposition}
+
+\begin{corollary}
+  $\UFin$ is a pointed, connected, 1-type, that is, a 1-groupoid, or has h-level 3.
+\end{corollary}
+
+\begin{lemma}~\label{lem:loop-deloop}
+  \[
+    \loopspace[\BAut[T],T_{0}] \eqv \Aut[T]
+  \]
+\end{lemma}
+
+\begin{proof}
+  Since $\BAut[T]$ is a univalent universe, it follows that
+  \[
+    (T_{0} \id_{\BAut[T]} T_{0}) \eqv (\fst(T_{0}) \eqv \fst(T_{0})) \equiv (T \eqv T) \equiv \Aut[T].
+  \]
+\end{proof}
+
+\begin{corollary}
+  For every $n:\Nat$,
+  \[
+    \loopspace[\UFin[n],F_{n}] \eqv \Aut[\Fin[n]]
+  \]
+\end{corollary}
 

### 2021-06-19 Pi+/paper/reversible1.tex
done with set based semantics

@@
 
-Fig.~\ref{fig:delta} specifies the semantics of the base combinators
-using a function $\delta$. It is clear that this function has an inverse $\delta^\dagger$ such that
-$\delta^\dagger(c,\delta(c,v))=v$. The semantics of full programs is given by the interpreter in the second figure.
+\note{Explain code below as best as you can}
 
-We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
-compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
-$\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:
+% semT : (t : U) → Set
+% semT t = Fin ∣ t ∣
 
-\[\begin{array}{rcll}
-    \sem{A} &=& \mathsf{Fin}~\sizet{A}
-\\
-    \sem{c : A \isot B} &:& \mathsf{Fin}~\sizet{A} \rightarrow \mathsf{Fin}~\sizet{B} \\
-    %% base combinators
-    \sem{c_1\oplus c_2}_{A+B \isot C+D}~\inlv{v} &=& \sem{c_1}_{A \isot C}~v \\
-    \sem{c_1\oplus c_2}_{A+B \isot C+D}~\inrv{v} &=& \sizet{A} + \sem{c_2}_{B \isot D}~v \\
-    \sem{\idc}~v &=& v \\
-    \sem{c_1\fatsemi c_2}~v &=& \sem{c_2} (\sem{c_1}~v) \\
-    \sem{c_1\otimes c_2}~(v_1,v_2) &=& (\sem{c_1}~v_1 , \sem{c_2}~v_2)
-\end{array}\]
+% semC : {t₁ t₂ : U} → (c : t₁ ⟷₁ t₂) → Fin ∣ t₁ ∣ → Fin ∣ t₂ ∣
+% semC unite₊l = idf _
+% semC uniti₊l = idf _
+% semC {I × t₂} {t₂} unite⋆l = transport Fin (N.+-unit-r ∣ t₂ ∣)
+% semC {t₁} {I × t₁} uniti⋆l = transport Fin (! (N.+-unit-r ∣ t₁ ∣))
+% semC {t₁ + t₂} {t₂ + t₁} swap₊ = (comp+ t₂ t₁) ∘ (swap+Fin {t₁} {t₂}) ∘ (decomp+ t₁ t₂)
+% semC {t₁ × t₂} {t₂ × t₁} swap⋆ = (comp* t₂ t₁) ∘ (swap*Fin {t₁} {t₂}) ∘ (decomp* t₁ t₂)
+% semC {t₁ + (t₂ + t₃)} {(t₁ + t₂) + t₃} assocl₊ = transport Fin (! (N.+-assoc (∣ t₁ ∣) (∣ t₂ ∣) (∣ t₃ ∣)))
+% semC {(t₁ + t₂) + t₃} {t₁ + (t₂ + t₃)} assocr₊ = transport Fin (N.+-assoc (∣ t₁ ∣) (∣ t₂ ∣) (∣ t₃ ∣))
+% semC {t₁ × (t₂ × t₃)} {(t₁ × t₂) × t₃} assocl⋆ = transport Fin (! (*-assoc (∣ t₁ ∣) (∣ t₂ ∣) (∣ t₃ ∣)))
+% semC {(t₁ × t₂) × t₃} {t₁ × (t₂ × t₃)} assocr⋆ = transport Fin (*-assoc (∣ t₁ ∣) (∣ t₂ ∣) (∣ t₃ ∣))
+% semC absorbr ()
+% semC {t × O} {O} absorbl = transport Fin (*-unit-r ∣ t ∣)
+% semC factorzr ()
+% semC factorzl ()
+% semC {(t₁ + t₂) × t₃} {(t₁ × t₃) + (t₂ × t₃)} dist =
+%   comp+ (t₁ × t₃) (t₂ × t₃) ∘
+%   map+ (comp* t₁ t₃) (comp* t₂ t₃) ∘
+%   distNative ∘
+%   map* (decomp+ t₁ t₂) (idf _) ∘
+%   decomp* (t₁ + t₂) t₃
+% semC {(t₁ × t₃) + (t₂ × t₃)} {(t₁ + t₂) × t₃} factor =
+%   comp* (t₁ + t₂) t₃ ∘
+%   map* (comp+ t₁ t₂) (idf _) ∘
+%   factorNative ∘
+%   map+ (decomp* t₁ t₃) (decomp* t₂ t₃) ∘
+%   decomp+ (t₁ × t₃) (t₂ × t₃)
+% semC id⟷₁ = idf _
+% semC (c₁ ◎ c₂) = semC c₂ ∘ semC c₁
+% semC {t₁ + t₂} {t₃ + t₄} (c₁ ⊕ c₂) = (comp+ t₃ t₄) ∘ map+ (semC c₁) (semC c₂) ∘ (decomp+ t₁ t₂)
+% semC {t₁ × t₂} {t₃ × t₄} (c₁ ⊗ c₂) = (comp* t₃ t₄) ∘ map* (semC c₁) (semC c₂) ∘ (decomp* t₁ t₂)
 

### 2021-06-19 Pi+/paper/reversible1.tex
Text about denotational semantics

@@
 \noindent A simple denotational semantics of $\Pi$ can be specified by mapping each type to a finite set and each
-combinator to a permutation between finite sets.
+combinator to a bijection between finite sets, that is, a permutation.
+
+Functions with finite sets as domains are finitely supported, that is, their outputs can be tabulated using the
+canonical ordering on $\Fin[n]$. For bijections, this gives a listed permutation. By observing the action of the
+combinators on the values of the finite set, we can define a denotational semantics which constructs the bijection.
+
+\note{\(\Fin[n] \to A \eqv \type{Vec_{n}}(A)\)}
+
+An alternative point of view is to work in the category of finite sets and functions, $\SetFin$, which is the category
+freely generated by finite coproduct completion of the terminal category. Objects of $\SetFin$ can be identified with
+sets of fixed cardinality, that is, $\Fin[n] \defeq \Set{0,1,\ldots,n-1}$. $\SetFin$ has finite coproducts and products,
+which lets us interpret the types of $\PiLang$. Combinators are intepreted as morphisms in $\SetFin$, but we have to
+restrict to invertible morphisms, that is, isomorphisms. This gives the \emph{groupoid} of finite sets and bijections,
+$\BFin \defeq \mathsf{core}(\SetFin)$. The isomorphisms satisfied by coproducts and products in $\SetFin$ lift to
+$\BFin$, but they're no longer categorical coproducts and products. They give two symmetric monoidal tensor products on
+$\BFin$, the additive and multiplicative ones, with the multiplicative tensor distributing over the additive tensor.
 

### 2021-06-19 Pi+/paper/finite.tex

@@
 
-Alternatively, groups can also be presented using generators and relations, or a quotient of the free group \ldots
+\subsection{Free groups}
 
-2. In the usual setting, there are are many relations that are true for the elements of the group. For example, in the
-(Z mod 3), we have a relation 1 + 1 + 1 = 0.
+Given a generating set $A$, we can naively construct a \emph{free group on $A$},
+whose elements are drawn from the alphabet $A$, and closed under the group
+operations of multiplication and inverse, identified by the group axioms.
+For example, the singleton set generates the additive group of integers,
+$\mathbb{Z}$.
 
-3. A group that has no other relations than the ones directly implied by
-the definition of a group is called free.
+Usually, there are are many equations, besides the group axioms, that hold for
+the elements of the group. For example, in the group $\mathbb{Z}/3\mathbb{Z}$,
+we have an equation $1 + 1 + 1 = 0$. A free group has the property that has no
+other relations than the ones directly implied by the group axioms.
 
-4. A group is said to be generated by a set of its elemenst (e1, ..., en)
-if every element in the group can be constructed from these elements
-composing operations of + and inverse. For example, Z is a free group,
-generated by one-element set (1).
-4a. Thus, a free group g can be thought as sequences of letters from the
-generating set, divided by the congruence closure of the group relation.
-Definition: Free group.
+\todo{What's the best notation for HITs?}
+\begin{definition}
+  Given an $\hSet$ $A$, the free group $F(A)$ on it is given by a higher
+  inductive type with the following point and path constructors. Notice the
+  similarity with the definition of a group structure in~\ref{subsec:groups},
+  but note that each operation here is a generator for the type $F(A)$\dots.
+  \begin{itemize}
+    \item An inclusion function $\eta_{A} : A \to F(A)$
+    \item A multiplication function $m : F(A) \times F(A) \to F(A)$
+    \item An element $e : F(A)$
+    \item An inverse function $i : F(A) \to F(A)$
+  \end{itemize}
+  \smallskip
+  \begin{itemize}
+    \item For every $x, y, z : F(A)$, a path $\term{assoc} : m(x, m(y, z)) \id m(m(x, y), z)$
+    \item For every $x : F(A)$, paths $\term{unitr} : m(x, e) \id x$ and $\term{unitl} : m(e, x) \id x$
+    \item For every $x : F(A)$, paths $\term{invr} : m(x, i(x)) \id e$ and $\term{invl} : m(i(x), x) \id e$
+    \item A 0-truncation, for every $x, y : F(A)$ and $p, q : x \id y$, a 2-path $\term{trunc} : p \id q$
+  \end{itemize}
+\end{definition}
 
-5. One useful way of defining groups is by their presentations.
-Presentation of a group builds a group by specifying what is the set of
-generators - thus making a free group - a collection of equations that have
-to be satisfied in the resulting group. For example, we could take the
-generators to be the set (1) (giving out Z), and then add an equation (1 +
-1 + 1 = 0) - thus, getting (Z mod 3).
+A group homomorphism between groups is a function between the underlying sets
+that preserves the group structure, which we write as $G_1 \to^G G_2$. With
+this, we can state the universal property of free groups.
 
-5a. The generatos can be thought of as primitive operations in a language,
-and the group structure gives us the way of composing these operations and
-inverting them.
+\begin{proposition}[Universal Property of $F(A)$]~\label{prop:free-groups}
+  Given a group $G$ and a map $f : A \to G$, there is a unique group
+  homomorphism $\extend{f} : F(A) \to^G G$ such that $\extend{f} \comp \eta_A
+  \htpy f$. Equivalently, composition with $\eta_A$ gives an equivalence $F(A)
+  \to^G G \eqv A \to G$. Alternatively, the type of group homomorphisms $h :
+  F(A) \to^G G$ satisfying $h \comp \eta_A \htpy f$ is contractible.
 
-6. Building a group like that is possible because of the universal property
-of the free group.
-Theorem: Universal property of a free group.
-Definition: formal definition of a free group.
+  % https://q.uiver.app/?q=WzAsMyxbMCwyLCJBIl0sWzAsMCwiRihBKSJdLFsyLDAsIkciXSxbMCwxLCJcXGV0YV9BIl0sWzAsMiwiZiIsMl0sWzEsMiwiXFxleHRlbmR7Zn0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkYXNoZWQifX19XSxbMyw0LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+\[\begin{tikzcd}
+	{F(A)} && G \\
+	\\
+	A
+	\arrow[""{name=0, anchor=center, inner sep=0}, "{\eta_A}", from=3-1, to=1-1]
+	\arrow[""{name=1, anchor=center, inner sep=0}, "f"', from=3-1, to=1-3]
+	\arrow["{\extend{f}}", dashed, from=1-1, to=1-3]
+	\arrow["\id", Rightarrow, draw=none, from=0, to=1]
+\end{tikzcd}\]
+\end{proposition}
 
-7. Why do it like that? Because it is a very syntactic approach to doing
-the theory. The way described previously operates directly in the semantics
-of the group operation - equality of elements (such as the result of adding
-two elements) has to be computed externally.
-Here, the equality can be computed using a rewriting system specified by
-the group axioms and the additional equations provided. However, because
-equations are not directed, it is not always possible to construct a
-well-behaving rewriting system. In general, the word problem (i.e. deciding
-if two words are equal) is proven to be undecidable.
+This definition of $F(A)$ however has lots of path constructors corresponding to
+each group axiom. Alternatively, the free group construction can be thought of
+as drawing letters from the generating set, while adding formal inverses, and
+constructing words which are sequences of these letters.
 
-\subsection{Permutation Groups}
+Each element in the sequence can either be an element drawn from the generating
+set $a$, or an inverse of such an element $\inv{a}$. If we take the disjoint
+union of $A$ with itself, that is, $A + A$ as the underlying set, we can use
+$\inl/\inr$ to mark the elements. Hence, we can encode the free group using the
+free monoid, that is, lists.
 
-8. The group that we are interested in is the group of permutations on a
+\begin{definition}
+  Let $A$ be an $\hSet$, and $\List[\blank]$ the free monoid. The free group
+  $F(A)$ on $A$ is the set-quotient of $\List[A + A]$ by the congruence closure
+  of the relation $a \cons \inv{a} \cons \nil \sim \nil$.
+\end{definition}
+
+\begin{proposition}
+  $F(A) \defeq \quot{\List[A + A]}{\sim^{\ast}}$ has a group structure, with the
+  empty list $\nil$ for the neutral element, multiplication given by list append
+  $\append$, and inverse given by list reversal. Further, $F(A)$ with $\eta_A :
+  A \to F(A) \defeq \inl(a) \cons \nil$ satisfies the universal property of free
+  groups.~\cref{prop:free-groups}
+\end{proposition}
+
+\subsubsection{Group presentations}
+
+One useful way of defining groups is by their presentations. A presentation of a
+group builds it by starting from the free group $F(A)$ and introducing a
+collection of equations that have to be satisfied in the resulting group. For
+example, if we take $F(\unit) \defeq \mathbb{Z}$ and add an equation $1 + 1 + 1
+= 0$, the resulting group would be $\mathbb{Z}/3\mathbb{Z}$.
+
+The generators can be thought of as primitive operations in a (reversible)
+programming language, group structure gives the way of composing these
+operations and inverting them, and relations describe how these primitive
+operations interact with each other.
+
+Instead of opereating directly in the semantics of the group operation, here the
+focus is on the syntax. While before, the equality of elements (such as the
+result of multiplication of two elements) had to be computed externally, now it is
+reduced to a \emph{word problem}, i.e., deciding one word can be reduced to another
+using group's equations.
+
+However, because equations are not directed, is not always possible to construct a
+well-behaving rewriting system. In general, the word problem is proven to be undecidable.
+
+\jk{Example?}
+
+\begin{definition}
+  Let $A$ be an $\hSet$ and $R : (A + A) \to (A + A) \to \UU$ a binary relation
+  on $A + A$. The group $G$ presented by $\langle A ; R \rangle$ is given by the
+  set-quotient of the free group $F(A)$ by the closure of $R$, or equivalently,
+  as the coequaliser
+  \[\begin{tikzcd}
+      FR && FA && G
+      \arrow[shift right=2, from=1-1, to=1-3]
+      \arrow[shift left=2, from=1-1, to=1-3]
+      \arrow[two heads, from=1-3, to=1-5]
+    \end{tikzcd}\]
+\end{definition}
+
+\vc{do we need this level of detail?}
+\todo{Universal property}
+\todo{Examples: empty relation, van Kampen of $\pi_{1}$}
+
+The above definition is correct because the universal property of the free group allows 
+for properly extending the relation on the generating set to the whole group.
+\jk{Turn that into a propostion?}
+
+\subsection{Permutation groups}
+
+Because $\Pi$ The group that we are interested in is the group of permutations on a
 fininte set. We can define it formally in two different ways:
@@
 (a bijection) to a word (a program).
-9a. However, proving them equivalent will require, in particular, solving
-the word problem (in this specific instance).
 
@@
 
-\begin{definition}
-  Let $A$ be a type, and $\List[A]$ the free monoid on $A$. The free group $F(A)$ on $A$ is the set-quotient of
-  $\List[A + A]$ by the congruence closure of the relation $(\inl(a) \cons \inr(a) \cons \nil) \sim \nil$.
-\end{definition}
-
-\begin{proposition}
-  Let $\eta_{A} : A \to F(A)$ be given by $a \mapsto \inl(a) \cons \nil$. Given any group $G$ with a map $f : A \to G$,
-  there is a canonical lifting of $f$ to a group homomorphism $\extend{f} : F(A) \to G$, such that
-  $\extend{f} \comp \eta_{A} \htpy f$, that is, the type of group homomorphisms $h : \Grp(F(A),G)$ satisfying
-  $h \comp \eta_{A} \htpy f$ is contractible. Equivalently, composition by $\eta_{A}$ is an equivalence
-  $\Grp(F(A),G) \eqv (A \to G)$.
-\end{proposition}
-
-\todo{diagram}
-\todo{Plus/Minus}
-
-\begin{definition}
-  Let $A$ be a type and $R : A \to A \to \UU$ a binary relation on $A$. The group $G$ presented by
-  $\langle A ; R \rangle$ is given by the set-quotient of the free group $F(A)$ by the normal closure $N_{R}$ of $R$, or
-  equivalently, as the coequaliser
-  \[\begin{tikzcd}
-      FR && FA && G
-      \arrow[shift right=2, from=1-1, to=1-3]
-      \arrow[shift left=2, from=1-1, to=1-3]
-      \arrow[two heads, from=1-3, to=1-5]
-    \end{tikzcd}\]
-\end{definition}
-
-\todo{Universal property}
-\todo{Examples: empty relation, van Kampen of $\pi_{1}$}
-
 \subsubsection{Coxeter Relations}

### 2021-06-19 Pi+/paper/finite.tex
Rig reorganization

@@
 
-\subsection{Symmetric Monoidal structure}
-
-We describe the symmetric monoidal structure of the groupoid $\UFin$.
-
-First, we observe a few equivalences.
-
-\begin{proposition}
-  For any $n, m : \Nat$,
-  \begin{align*}
-    \Fin[0]                & \eqv \bot \\
-    \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m] \\
-  \end{align*}
-  and for any types $X, Y, Z$,
-  \begin{align*}
-    \bot \sqcup X          & \eqv X \\
-    X \sqcup \bot          & \eqv X \\
-    (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
-    X \sqcup Y             & \eqv Y \sqcup Y \\
-  \end{align*}
-\end{proposition}
-
-We lift these equivalences to $\UFin$ giving it the (additive) symmetric monoidal structure $(I, \oplus)$, with natural
-isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the symmetry isomorphism $\mathcal{B}_{X,Y}$. Note that
-types in $\UFin$ are $\hSet$s since they're merely equivalent to $\Fin[n]$ for some $n : \Nat$.
-
-\begin{definition}
-  \begin{align*}
-    I           & \defeq F_{0}    \\
-    X \oplus Y & \defeq X \sqcup Y \\
-    \lambda_{X} & : I \oplus X \eqv X \\
-    \rho_{X} & : X \oplus I \eqv X \\
-    \alpha_{X,Y,Z} & : (X \oplus Y) \oplus Z \eqv X \oplus (Y \oplus Z) \\
-    \mathcal{B}_{X,Y} & : X \oplus Y \eqv Y \oplus X
-  \end{align*}
-\end{definition}
-
-These isomorphisms satisfy the Mac Lane coherence conditions for symmetric monoidal categories, that is, the triangle,
-pentagon, and hexagon identities, and the syllepsis of the braiding, upto 2-paths in $\UFin$.
-
-\begin{proposition}
-  % https://q.uiver.app/?q=WzAsNCxbMCwwLCIoWCBcXG9wbHVzIEkpIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyAoSSBcXG9wbHVzIFkpIl0sWzEsMSwiWCBcXG9wbHVzIFkiXSxbMCwxXSxbMCwxLCJcXGFscGhhX3tYLEksWX0iXSxbMCwyLCJcXHJob197WH0gXFxvcGx1cyAxX3tZfSIsMl0sWzEsMiwiMV97WH0gXFxvcGx1cyBcXGxhbWJkYV97WX0iXSxbNSw2LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-  \[\begin{tikzcd}
-      {(X \oplus I) \oplus Y} && {X \oplus (I \oplus Y)} \\
-      {} & {X \oplus Y}
-      \arrow["{\alpha_{X,I,Y}}", from=1-1, to=1-3]
-      \arrow[""{name=0, anchor=center, inner sep=0}, "{\rho_{X} \oplus 1_{Y}}"', from=1-1, to=2-2]
-      \arrow[""{name=1, anchor=center, inner sep=0}, "{1_{X} \oplus \lambda_{Y}}", from=1-3, to=2-2]
-      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
-    \end{tikzcd}\]
-  % https://q.uiver.app/?q=WzAsNSxbMCwxLCIoKFcgXFxvcGx1cyBYKSBcXG9wbHVzIFkpIFxcb3BsdXMgWiJdLFsxLDAsIihXIFxcb3BsdXMgWCkgXFxvcGx1cyAoWSBcXG9wbHVzIFopIl0sWzIsMSwiVyBcXG9wbHVzIChYIFxcb3BsdXMgKFkgXFxvcGx1cyBaKSkiXSxbMiwzLCJXIFxcb3BsdXMgKChYIFxcb3BsdXMgWSkgXFxvcGx1cyBaKSJdLFswLDMsIihXIFxcb3BsdXMgKFggXFxvcGx1cyBZKSkgXFxvcGx1cyBaIl0sWzAsMSwiXFxhbHBoYV97VyBcXG9wbHVzIFgsIFksIFp9Il0sWzEsMiwiXFxhbHBoYV97VyxYLFkgXFxvcGx1cyBafSJdLFszLDIsIjFfe1d9IFxcb3BsdXMgXFxhbHBoYV97WCxZLFp9IiwyXSxbMCw0LCJcXGFscGhhX3tXLFgsWX0gXFxvcGx1cyAxX3tafSIsMl0sWzQsMywiXFxhbHBoYV97VyxYIFxcb3BsdXMgWSxafSIsMl0sWzAsMiwiXFxpZCIsMSx7Im9mZnNldCI6NSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-  \[\begin{tikzcd}
-      & {(W \oplus X) \oplus (Y \oplus Z)} \\
-      {((W \oplus X) \oplus Y) \oplus Z} && {W \oplus (X \oplus (Y \oplus Z))} \\
-      \\
-      {(W \oplus (X \oplus Y)) \oplus Z} && {W \oplus ((X \oplus Y) \oplus Z)}
-      \arrow["{\alpha_{W \oplus X, Y, Z}}", from=2-1, to=1-2]
-      \arrow["{\alpha_{W,X,Y \oplus Z}}", from=1-2, to=2-3]
-      \arrow["{1_{W} \oplus \alpha_{X,Y,Z}}"', from=4-3, to=2-3]
-      \arrow["{\alpha_{W,X,Y} \oplus 1_{Z}}"', from=2-1, to=4-1]
-      \arrow["{\alpha_{W,X \oplus Y,Z}}"', from=4-1, to=4-3]
-      \arrow["\id"{description}, shift right=5, draw=none, from=2-1, to=2-3]
-    \end{tikzcd}\]
-  % https://q.uiver.app/?q=WzAsNixbMSwwLCJYIFxcb3BsdXMgKFkgXFxvcGx1cyBaKSJdLFswLDEsIihYIFxcb3BsdXMgWSkgXFxvcGx1cyBaIl0sWzAsMiwiKFkgXFxvcGx1cyBYKSBcXG9wbHVzIFoiXSxbMSwzLCJZIFxcb3BsdXMgKFggXFxvcGx1cyBaKSJdLFsyLDIsIlkgXFxvcGx1cyAoWiBcXG9wbHVzIFgpIl0sWzIsMSwiKFkgXFxvcGx1cyBaKSBcXG9wbHVzIFgiXSxbMSwwLCJcXGFscGhhX3tYLFksWn0iXSxbMSwyLCJcXG1hdGhjYWx7Qn1fe1gsWX0gXFxvcGx1cyAxX3tafSIsMl0sWzIsMywiXFxhbHBoYV97WSxYLFp9IiwyXSxbMyw0LCIxX3tZfSBcXG9wbHVzIFxcbWF0aGNhbHtCfV97WCxafSIsMl0sWzUsNCwiXFxhbHBoYV97WSxaLFh9Il0sWzAsNSwiXFxtYXRoY2Fse0J9X3tYLFkgXFxvcGx1cyBafSJdLFs3LDEwLCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-  \[\begin{tikzcd}
-      & {X \oplus (Y \oplus Z)} \\
-      {(X \oplus Y) \oplus Z} && {(Y \oplus Z) \oplus X} \\
-      {(Y \oplus X) \oplus Z} && {Y \oplus (Z \oplus X)} \\
-      & {Y \oplus (X \oplus Z)}
-      \arrow["{\alpha_{X,Y,Z}}", from=2-1, to=1-2]
-      \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y} \oplus 1_{Z}}"', from=2-1, to=3-1]
-      \arrow["{\alpha_{Y,X,Z}}"', from=3-1, to=4-2]
-      \arrow["{1_{Y} \oplus \mathcal{B}_{X,Z}}"', from=4-2, to=3-3]
-      \arrow[""{name=1, anchor=center, inner sep=0}, "{\alpha_{Y,Z,X}}", from=2-3, to=3-3]
-      \arrow["{\mathcal{B}_{X,Y \oplus Z}}", from=1-2, to=2-3]
-      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
-    \end{tikzcd}\]
-  % https://q.uiver.app/?q=WzAsMyxbMCwwLCJYIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyBZIl0sWzEsMSwiWSBcXG9wbHVzIFgiXSxbMCwxLCIxX3tYIFxcb3BsdXMgWX0iLDAseyJsZXZlbCI6Miwic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsIlxcbWF0aGNhbHtCfV97WCxZfSIsMl0sWzIsMSwiXFxtYXRoY2Fse0J9X3tZLFh9IiwyXSxbNCw1LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
-  \[\begin{tikzcd}
-      {X \oplus Y} && {X \oplus Y} \\
-      & {Y \oplus X}
-      \arrow["{1_{X \oplus Y}}", Rightarrow, no head, from=1-1, to=1-3]
-      \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y}}"', from=1-1, to=2-2]
-      \arrow[""{name=1, anchor=center, inner sep=0}, "{\mathcal{B}_{Y,X}}"', from=2-2, to=1-3]
-      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
-    \end{tikzcd}\]
-\end{proposition}
-
 %%% Local Variables:

### 2021-06-19 Pi+/paper/reversible2.tex
Rig reorganization

@@
 
+An alternative point of view for the semantics in Sec.~\ref{reversibleone} is to consider it  as expressed in the
+category of finite sets and functions, $\SetFin$, which is the category freely generated by finite coproduct completion
+of the terminal category. Objects of $\SetFin$ can be identified with sets of fixed cardinality, that is,
+$\Fin[n] \defeq \Set{0,1,\ldots,n-1}$. $\SetFin$ has finite coproducts and products, which lets us interpret the types
+of $\PiLang$. Combinators are intepreted as morphisms in $\SetFin$, but we have to restrict to invertible morphisms,
+that is, isomorphisms. This gives the \emph{groupoid} of finite sets and bijections,
+$\BFin \defeq \mathsf{core}(\SetFin)$. The isomorphisms satisfied by coproducts and products in $\SetFin$ lift to
+$\BFin$, but they're no longer categorical coproducts and products. They give two symmetric monoidal tensor products on
+$\BFin$, the additive and multiplicative ones, with the multiplicative tensor distributing over the additive tensor.
+
 We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
@@
 
+\note{We use the rig structure of $\UFin$ in~\cref{subsec:rig} to interpret $\PiLang$.}
+
 %%% Local Variables:

### 2021-06-19 Pi+/paper/ufin.tex
Rig reorganization

@@
 
+\subsection{Rig structure}~\label{subsec:rig}
+
+\todo{rig}
+
+We describe the symmetric monoidal structure of the groupoid $\UFin$.
+
+First, we observe a few equivalences.
+
+\begin{proposition}
+  For any $n, m : \Nat$,
+  \begin{align*}
+    \Fin[0]                & \eqv \bot \\
+    \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m] \\
+  \end{align*}
+  and for any types $X, Y, Z$,
+  \begin{align*}
+    \bot \sqcup X          & \eqv X \\
+    X \sqcup \bot          & \eqv X \\
+    (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
+    X \sqcup Y             & \eqv Y \sqcup Y \\
+  \end{align*}
+\end{proposition}
+
+We lift these equivalences to $\UFin$ giving it the (additive) symmetric monoidal structure $(I, \oplus)$, with natural
+isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the symmetry isomorphism $\mathcal{B}_{X,Y}$. Note that
+types in $\UFin$ are $\hSet$s since they're merely equivalent to $\Fin[n]$ for some $n : \Nat$.
+
+\begin{definition}
+  \begin{align*}
+    I           & \defeq F_{0}    \\
+    X \oplus Y & \defeq X \sqcup Y \\
+    \lambda_{X} & : I \oplus X \eqv X \\
+    \rho_{X} & : X \oplus I \eqv X \\
+    \alpha_{X,Y,Z} & : (X \oplus Y) \oplus Z \eqv X \oplus (Y \oplus Z) \\
+    \mathcal{B}_{X,Y} & : X \oplus Y \eqv Y \oplus X
+  \end{align*}
+\end{definition}
+
+These isomorphisms satisfy the Mac Lane coherence conditions for symmetric monoidal categories, that is, the triangle,
+pentagon, and hexagon identities, and the syllepsis of the braiding, upto 2-paths in $\UFin$.
+
+\begin{proposition}
+  % https://q.uiver.app/?q=WzAsNCxbMCwwLCIoWCBcXG9wbHVzIEkpIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyAoSSBcXG9wbHVzIFkpIl0sWzEsMSwiWCBcXG9wbHVzIFkiXSxbMCwxXSxbMCwxLCJcXGFscGhhX3tYLEksWX0iXSxbMCwyLCJcXHJob197WH0gXFxvcGx1cyAxX3tZfSIsMl0sWzEsMiwiMV97WH0gXFxvcGx1cyBcXGxhbWJkYV97WX0iXSxbNSw2LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+  \[\begin{tikzcd}
+      {(X \oplus I) \oplus Y} && {X \oplus (I \oplus Y)} \\
+      {} & {X \oplus Y}
+      \arrow["{\alpha_{X,I,Y}}", from=1-1, to=1-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\rho_{X} \oplus 1_{Y}}"', from=1-1, to=2-2]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{1_{X} \oplus \lambda_{Y}}", from=1-3, to=2-2]
+      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
+    \end{tikzcd}\]
+  % https://q.uiver.app/?q=WzAsNSxbMCwxLCIoKFcgXFxvcGx1cyBYKSBcXG9wbHVzIFkpIFxcb3BsdXMgWiJdLFsxLDAsIihXIFxcb3BsdXMgWCkgXFxvcGx1cyAoWSBcXG9wbHVzIFopIl0sWzIsMSwiVyBcXG9wbHVzIChYIFxcb3BsdXMgKFkgXFxvcGx1cyBaKSkiXSxbMiwzLCJXIFxcb3BsdXMgKChYIFxcb3BsdXMgWSkgXFxvcGx1cyBaKSJdLFswLDMsIihXIFxcb3BsdXMgKFggXFxvcGx1cyBZKSkgXFxvcGx1cyBaIl0sWzAsMSwiXFxhbHBoYV97VyBcXG9wbHVzIFgsIFksIFp9Il0sWzEsMiwiXFxhbHBoYV97VyxYLFkgXFxvcGx1cyBafSJdLFszLDIsIjFfe1d9IFxcb3BsdXMgXFxhbHBoYV97WCxZLFp9IiwyXSxbMCw0LCJcXGFscGhhX3tXLFgsWX0gXFxvcGx1cyAxX3tafSIsMl0sWzQsMywiXFxhbHBoYV97VyxYIFxcb3BsdXMgWSxafSIsMl0sWzAsMiwiXFxpZCIsMSx7Im9mZnNldCI6NSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+  \[\begin{tikzcd}
+      & {(W \oplus X) \oplus (Y \oplus Z)} \\
+      {((W \oplus X) \oplus Y) \oplus Z} && {W \oplus (X \oplus (Y \oplus Z))} \\
+      \\
+      {(W \oplus (X \oplus Y)) \oplus Z} && {W \oplus ((X \oplus Y) \oplus Z)}
+      \arrow["{\alpha_{W \oplus X, Y, Z}}", from=2-1, to=1-2]
+      \arrow["{\alpha_{W,X,Y \oplus Z}}", from=1-2, to=2-3]
+      \arrow["{1_{W} \oplus \alpha_{X,Y,Z}}"', from=4-3, to=2-3]
+      \arrow["{\alpha_{W,X,Y} \oplus 1_{Z}}"', from=2-1, to=4-1]
+      \arrow["{\alpha_{W,X \oplus Y,Z}}"', from=4-1, to=4-3]
+      \arrow["\id"{description}, shift right=5, draw=none, from=2-1, to=2-3]
+    \end{tikzcd}\]
+  % https://q.uiver.app/?q=WzAsNixbMSwwLCJYIFxcb3BsdXMgKFkgXFxvcGx1cyBaKSJdLFswLDEsIihYIFxcb3BsdXMgWSkgXFxvcGx1cyBaIl0sWzAsMiwiKFkgXFxvcGx1cyBYKSBcXG9wbHVzIFoiXSxbMSwzLCJZIFxcb3BsdXMgKFggXFxvcGx1cyBaKSJdLFsyLDIsIlkgXFxvcGx1cyAoWiBcXG9wbHVzIFgpIl0sWzIsMSwiKFkgXFxvcGx1cyBaKSBcXG9wbHVzIFgiXSxbMSwwLCJcXGFscGhhX3tYLFksWn0iXSxbMSwyLCJcXG1hdGhjYWx7Qn1fe1gsWX0gXFxvcGx1cyAxX3tafSIsMl0sWzIsMywiXFxhbHBoYV97WSxYLFp9IiwyXSxbMyw0LCIxX3tZfSBcXG9wbHVzIFxcbWF0aGNhbHtCfV97WCxafSIsMl0sWzUsNCwiXFxhbHBoYV97WSxaLFh9Il0sWzAsNSwiXFxtYXRoY2Fse0J9X3tYLFkgXFxvcGx1cyBafSJdLFs3LDEwLCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+  \[\begin{tikzcd}
+      & {X \oplus (Y \oplus Z)} \\
+      {(X \oplus Y) \oplus Z} && {(Y \oplus Z) \oplus X} \\
+      {(Y \oplus X) \oplus Z} && {Y \oplus (Z \oplus X)} \\
+      & {Y \oplus (X \oplus Z)}
+      \arrow["{\alpha_{X,Y,Z}}", from=2-1, to=1-2]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y} \oplus 1_{Z}}"', from=2-1, to=3-1]
+      \arrow["{\alpha_{Y,X,Z}}"', from=3-1, to=4-2]
+      \arrow["{1_{Y} \oplus \mathcal{B}_{X,Z}}"', from=4-2, to=3-3]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{\alpha_{Y,Z,X}}", from=2-3, to=3-3]
+      \arrow["{\mathcal{B}_{X,Y \oplus Z}}", from=1-2, to=2-3]
+      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
+    \end{tikzcd}\]
+  % https://q.uiver.app/?q=WzAsMyxbMCwwLCJYIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyBZIl0sWzEsMSwiWSBcXG9wbHVzIFgiXSxbMCwxLCIxX3tYIFxcb3BsdXMgWX0iLDAseyJsZXZlbCI6Miwic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsIlxcbWF0aGNhbHtCfV97WCxZfSIsMl0sWzIsMSwiXFxtYXRoY2Fse0J9X3tZLFh9IiwyXSxbNCw1LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+  \[\begin{tikzcd}
+      {X \oplus Y} && {X \oplus Y} \\
+      & {Y \oplus X}
+      \arrow["{1_{X \oplus Y}}", Rightarrow, no head, from=1-1, to=1-3]
+      \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y}}"', from=1-1, to=2-2]
+      \arrow[""{name=1, anchor=center, inner sep=0}, "{\mathcal{B}_{Y,X}}"', from=2-2, to=1-3]
+      \arrow["\id", Rightarrow, draw=none, from=0, to=1]
+    \end{tikzcd}\]
+\end{proposition}
+
 %%% Local Variables:

### 2021-06-19 Pi+/paper/finite.tex
Sorting examples

@@
 
-Because $\Pi$ The group that we are interested in is the group of permutations on a
-fininte set. We can define it formally in two different ways:
-8a. Taking the type Aut (Fin n) and showing that it satisfies the group
-axiom.
-8b. Writing up a presentation - choosing generating set and equations that
-have to be satisfied.
+Because $\PiLang$ is used for reasoning about reversible functions on finite
+sets, the group that we are interested in is the group of permutations on a
+fininte set, which is classically know as the symmetric group on $n$ words,
+$\Sn$. We've already seen that we can define it formally by taking the
+automorphism group given by the type $\Aut[\Fin[n]]$, and showing that it has a
+group structure satisfying the group axioms. Alternatively, we can write a group
+presentation for $\Sn$ -- by defining a set of generators and a set of relations.
 
-9. By proving these two to be equivalent, we are bridging the gap between
-syntactic and semantic side in the whole proof - by assinging the meaning
-(a bijection) to a word (a program).
+The technical contribution of this section is a proof that these two
+descriptions are equivalent. By doing that, we bridge the gap between the
+syntactic and semantic notions in our completeness proof -- by establishing a
+correspondence between the meaning of a program (a bijection) and its syntax (a
+word).
 
-10. However, in the point 8b, there is an element of choice. One group can
-be presented in many different ways. For example, we could generate the
-permutations by using generators that:
- - swap i-th element with j-th element
- - swap i-th element with the first element
- - shift elements in a cyclic way: operation with label "i" leaves elements
-with numbers > i in place, and element j <= i goes to (j + 1) mod i.
+However, in giving a group presentation, there is an element of choice. A group
+can be presented in many different ways. For example, we could generate the
+permutation group on $\Fin[n]$ by using generators that:
 
-11. One way of thinking about these presentations is to compare them to
-sorting algorithms with different primitive operations. If using a set of
-operations, one can sort any list, then these operations generate the
-permutations group. One weird example: pancake sorting.
+\begin{itemize}
+\item swap the $i$-the element with the $(i+1)$-th element, that is, adjacent swaps, or
+\item swap the $i$-th element with the $j$-th element, for arbitrary $i$s and $j$s, or
+\item swap the $i$-th element with an element at a fixed position, or
+\item flip a prefix $\Fin[k]$ of $\Fin[n]$ for $k \leq n$, or
+\item cyclically shift any subset of $\Fin[n]$.
+\end{itemize}
+
+One way of thinking about these presentations is via sorting algorithms, which
+use different primitive operations. A sorting algorithm has to
+calculate a permutation of a list or a finite set, which satisfies the invariant
+of being a sorted sequence, which means, the primitive operations of a sorting
+algorithm should be able to generate all the permutations on a given list. If
+using a chosen set of reversible operations, one can write a sorting algorithm,
+then those operations generate the permutation group.
+
+For example, bubble sort uses the primitive operation of adjacent swaps,
+insertion sort and selection sort use the primitive operation of swapping the
+$i$-th element with the $j$-th element, cycle sort uses cyclical shifts of
+subsequences, pancake sort uses flips of prefixes of the list, et cetera.
+\todo{check!}
 

### 2021-06-19 Pi+/paper/finite.tex
Diagrams

@@
 \item swap the $i$-the element with the $(i+1)$-th element, that is, adjacent swaps, or
-\item swap the $i$-th element with the $j$-th element, for arbitrary $i$s and $j$s, or
+\item swap the $i$-th element with the $j$-th element, for arbitrary $i$-s and $j$-s, or
 \item swap the $i$-th element with an element at a fixed position, or
@@
 One way of thinking about these presentations is via sorting algorithms, which
-use different primitive operations. A sorting algorithm has to
-calculate a permutation of a list or a finite set, which satisfies the invariant
-of being a sorted sequence, which means, the primitive operations of a sorting
-algorithm should be able to generate all the permutations on a given list. If
-using a chosen set of reversible operations, one can write a sorting algorithm,
-then those operations generate the permutation group.
+use different primitive operations. A sorting algorithm has to calculate a
+permutation of a list or a finite set, which satisfies the invariant of being a
+sorted sequence, which means, the primitive operations of a sorting algorithm
+should be able to generate all the permutations on a given list. If using a
+chosen set of reversible operations, one can write a sorting algorithm, then
+those operations generate the permutation group.
 
@@
 
-12. Here, we are choosing a presentation based on the adjacent
-transpositions generators - i.e. primitive operations of the form "swap
-i-th and i+1 elements". In the language of sorting, this corresponds to
-what the insertion sort or bubble sort are doing.
-In the explanations below, we will use diagrams to draw the permutations.
+To do the proof, we are using a presentation based on the adjacent
+transpositions, i.e. where the primitive operations are adjacent swaps. There
+are three laws that these operations have to satisfy - it is easiest to
+visualize them on diagrams.
+\begin{itemize}
+\item Swapping the same two elements two times in a row is the same as doing nothing:
+    \begin{tikzpicture}
+        \pic[local bounding box=my braid,braid/.cd, 
+        number of strands = 2,
+        thick]
+        {braid={ s_1, s_1}};
+    \end{tikzpicture}
+    \hspace{1cm}
+    \begin{tikzpicture}
+        \draw [-stealth](0,0) -- node[anchor=south] {cancel} (2,0);
+    \end{tikzpicture}
+    \hspace{1cm}
+    \begin{tikzpicture}
+        \pic[local bounding box=my braid,braid/.cd, 
+        number of strands = 2,
+        thick]
+        {braid={1, 1}};
+    \end{tikzpicture}
+\item When swapping two pairs of elements on a positions far away, it does not
+matter in which order it is done (We can slide the wires).
+    \begin{tikzpicture}
+        \pic[local bounding box=my braid,braid/.cd, 
+        number of strands = 2,
+        thick]
+        {braid={ 1, s_1}};
+    \end{tikzpicture}
+    \hspace{0cm}
+    \dots
+    \hspace{0cm}
+    \begin{tikzpicture}
+        \pic[local bounding box=my braid,braid/.cd, 
+        number of strands = 2,
+        thick]
+        {braid={ s_1, 1 }};
+    \end{tikzpicture}
+    \hspace{1cm}
+    \begin{tikzpicture}
+        \draw [-stealth](0,0) -- node[anchor=south] {swap} (2,0);
+    \end{tikzpicture}
+    \hspace{1cm}
+    \begin{tikzpicture}
+        \pic[local bounding box=my braid,braid/.cd, 
+        number of strands = 2,
+        thick]
+        {braid={ s_1, 1}};
+    \end{tikzpicture}
+    \hspace{0cm}
+    \dots
+    \hspace{0cm}
+    \begin{tikzpicture}
+        \pic[local bounding box=my braid,braid/.cd, 
+        number of strands = 2,
+        thick]
+        {braid={ 1, s_1}};
+    \end{tikzpicture}
 
-12a. There two very obvious laws that such operation has to preserve:
- - Swapping the same two elements two times in a row does not change
-anything (cancel)
-TODO diagram.
- - When swapping two pairs of elements on a positions far away, it does not
-matter in which order one does it (swap). (We can slide the wires).
-TODO diagram.
-But there is also one more law that these operations have to satisfy, in
-the situation when we are swapping adjacent pairs of elements. It is best
-seen on a diagram (braid)
-TODO diagram.
+\item seen on a diagram (braid)
+    \begin{tikzpicture}
+        \pic[local bounding box=my braid,braid/.cd, 
+        number of strands = 3,
+        thick]
+        {braid={ s_2, s_1, s_2}};
+    \end{tikzpicture}
+    \hspace{1cm}
+    \begin{tikzpicture}
+        \draw [-stealth](0,0) -- node[anchor=south] {braid} (2,0);
+    \end{tikzpicture}
+    \hspace{1cm}
+    \begin{tikzpicture}
+        \pic[local bounding box=my braid,braid/.cd, 
+        number of strands = 3,
+        thick]
+        {braid={ s_1, s_2, s_1}};
+    \end{tikzpicture}
+
+\end{itemize}
+
+Diagram of long-braid:
+\begin{tikzpicture}
+    \pic[local bounding box=my braid,braid/.cd, 
+    number of strands = 2,
+    thick]
+    {braid={1, 1, 1, 1, s_1, 1}};
+\end{tikzpicture}
+\hspace{0cm}
+\dots
+\hspace{0cm}
+\begin{tikzpicture}
+    \pic[local bounding box=my braid,braid/.cd, 
+    number of strands = 5,
+    thick]
+    {braid={ s_4, s_3, s_2, s_1, 1, s_4}};
+\end{tikzpicture}
+\hspace{1cm}
+\begin{tikzpicture}
+    \draw [-stealth](0,0) -- node[anchor=south] {long-braid} (2,0);
+\end{tikzpicture}
+\hspace{1cm}
+\begin{tikzpicture}
+    \pic[local bounding box=my braid,braid/.cd, 
+    number of strands = 2,
+    thick]
+    {braid={1, 1, 1, 1, 1, s_1}};
+\end{tikzpicture}
+\hspace{0cm}
+\dots
+\hspace{0cm}
+\begin{tikzpicture}
+    \pic[local bounding box=my braid,braid/.cd, 
+    number of strands = 5,
+    thick]
+    {braid={ s_3, s_4, s_3, s_2, s_1, 1}};
+\end{tikzpicture}
+\hspace{1cm}
 

### 2021-06-19 Pi+/paper/finite.tex
Add outline of braid diagrams

@@
 
-To do the proof, we are using a presentation based on the adjacent
-transpositions, i.e. where the primitive operations are adjacent swaps. There
-are three laws that these operations have to satisfy - it is easiest to
-visualize them on diagrams.
-\begin{itemize}
-\item Swapping the same two elements two times in a row is the same as doing nothing:
+We use a presentation based on adjacent transpositions, which is called the
+Coxeter presentation. The primitive operations are adjacent swaps. Further,
+there are three laws that these operations have to satisfy -- it is easiest to
+visualise them as braid diagrams.
+
+Swapping the same two elements two times in a row is the same as doing nothing:
+
+\[
+  \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
+    \begin{center}
       \begin{tikzpicture}
@@
       \end{tikzpicture}
-    \hspace{1cm}
-    \begin{tikzpicture}
-        \draw [-stealth](0,0) -- node[anchor=south] {cancel} (2,0);
-    \end{tikzpicture}
-    \hspace{1cm}
+    \end{center}
+     &
+    \(\xlongrightarrow[]{\cancel}\)
+     &
+    \begin{center}
       \begin{tikzpicture}
@@
       \end{tikzpicture}
-\item When swapping two pairs of elements on a positions far away, it does not
-matter in which order it is done (We can slide the wires).
+    \end{center}
+  \end{tabular}
+\]
+
+When swapping two pairs of elements in positions far away, it does not matter in
+which order the swapping happens, that is, we can slide the wires.
+
+\[
+  \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
+    \begin{center}
+      \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
         \begin{tikzpicture}
@@
         \end{tikzpicture}
-    \hspace{0cm}
-    \dots
-    \hspace{0cm}
+         &
+        \(\cdots\)
+         &
         \begin{tikzpicture}
@@
         \end{tikzpicture}
-    \hspace{1cm}
-    \begin{tikzpicture}
-        \draw [-stealth](0,0) -- node[anchor=south] {swap} (2,0);
-    \end{tikzpicture}
-    \hspace{1cm}
+      \end{tabular}
+    \end{center}
+     &
+    \(\xlongrightarrow[]{\swap}\)
+     &
+    \begin{center}
+      \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
         \begin{tikzpicture}
@@
         \end{tikzpicture}
+      \end{tabular}
+    \end{center}
+  \end{tabular}
+\]
 
-\item seen on a diagram (braid)
+\[
+  \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
+    \begin{center}
       \begin{tikzpicture}
@@
       \end{tikzpicture}
-    \hspace{1cm}
-    \begin{tikzpicture}
-        \draw [-stealth](0,0) -- node[anchor=south] {braid} (2,0);
-    \end{tikzpicture}
-    \hspace{1cm}
+    \end{center}
+     &
+    \(\xlongrightarrow[]{\braid}\)
+     &
+    \begin{center}
       \begin{tikzpicture}
@@
       \end{tikzpicture}
-
-\end{itemize}
+    \end{center}
+  \end{tabular}
+\]
 
 Diagram of long-braid:
+
+\[
+  \begin{tabular}{m{0.45\linewidth}m{0.1\linewidth}m{0.45\linewidth}}
+    \begin{tabular}{m{0.15\linewidth}m{0.05\linewidth}m{0.1\linewidth}}
+      \begin{center}
         \begin{tikzpicture}
@@
         \end{tikzpicture}
-\hspace{0cm}
-\dots
-\hspace{0cm}
+      \end{center}
+       &
+      \(\cdots\)
+       &
+      \begin{center}
         \begin{tikzpicture}
@@
         \end{tikzpicture}
-\hspace{1cm}
-\begin{tikzpicture}
-    \draw [-stealth](0,0) -- node[anchor=south] {long-braid} (2,0);
-\end{tikzpicture}
-\hspace{1cm}
+      \end{center}
+    \end{tabular}
+     &
+    \(\xlongrightarrow[]{\longbraid}\)
+     &
+    \begin{tabular}{m{0.15\linewidth}m{0.05\linewidth}m{0.1\linewidth}}
+      \begin{center}
         \begin{tikzpicture}
@@
         \end{tikzpicture}
-\hspace{1cm}
+      \end{center}
+    \end{tabular}
+  \end{tabular}
+\]
 

### 2021-06-19 Pi+/paper/finite.tex
Change long braid diagram to a fixed set

@@
 \[
-  \begin{tabular}{m{0.45\linewidth}m{0.1\linewidth}m{0.45\linewidth}}
-    \begin{tabular}{m{0.15\linewidth}m{0.05\linewidth}m{0.1\linewidth}}
+  \begin{tabular}{m{0.4\linewidth}m{0.1\linewidth}m{0.4\linewidth}}
       \begin{center}
@@
           \pic[local bounding box=my braid,braid/.cd,
-            number of strands = 2,
+            number of strands = 6,
             thick]
-          {braid={1, 1, 1, 1, s_1, 1}};
+          {braid={s_5, s_4, s_3, s_2, s_1, s_5}};
         \end{tikzpicture}
@@
      &
-      \(\cdots\)
+    \(\xlongrightarrow[]{\longbraid}\)
      &
@@
           \pic[local bounding box=my braid,braid/.cd,
-            number of strands = 5,
+            number of strands = 6,
             thick]
-          {braid={ s_4, s_3, s_2, s_1, 1, s_4}};
+            {braid={s_4, s_5, s_4, s_3, s_2, s_1}};
         \end{tikzpicture}
       \end{center}
-    \end{tabular}
-     &
-    \(\xlongrightarrow[]{\longbraid}\)
-     &
-    \begin{tabular}{m{0.15\linewidth}m{0.05\linewidth}m{0.1\linewidth}}
-      \begin{center}
-        \begin{tikzpicture}
-          \pic[local bounding box=my braid,braid/.cd,
-            number of strands = 2,
-            thick]
-          {braid={1, 1, 1, 1, 1, s_1}};
-        \end{tikzpicture}
-      \end{center}
-       &
-      \(\cdots\)
-       &
-      \begin{center}
-        \begin{tikzpicture}
-          \pic[local bounding box=my braid,braid/.cd,
-            number of strands = 5,
-            thick]
-          {braid={ s_3, s_4, s_3, s_2, s_1, 1}};
-        \end{tikzpicture}
-      \end{center}
-    \end{tabular}
   \end{tabular}

### 2021-06-20 Pi+/paper/finite.tex
Outline the coxeter rewriting system

@@
 
-When swapping two pairs of elements in positions far away, it does not matter in
-which order the swapping happens, that is, we can slide the wires.
+When swapping two pairs of elements in positions "far-away", it does not matter
+in which order the swapping happens, that is, we can slide the wires freely.
 
@@
 
+\todo{explain braid in words, give some intution on why it holds}
+
 \[
@@
 \]
+
+\todo{an example/motivation from book.pdf}
+
+This construction is called a Coxeter presentation. Writing it formally, we get
+the following definition for $\Sn$. We define a binary relation $\cox$ on
+$\List[\Fin[n]]$.
+
+\begin{definition}[$\cox$]
+  \begin{align*}
+    \cancel
+     & : \forall n \to (n \cons n \cons \nil) \cox \nil                                                     \\
+    \swap
+     & : \forall k, n \to (\suc[k] < n) \to (n \cons k \cons \nil) \cox (k \cons n \cons \nil)              \\
+    \braid
+     & : \forall n \to (\suc[n] \cons n \cons \suc[n] \cons \nil) \cox (n \cons \suc[n] \cons n \cons \nil) \\
+  \end{align*}
+\end{definition}
+
+We define $\cox*$ as the congruence closure of $\cox$.
+
+\begin{definition}[$\cox*$]
+  \begin{align*}
+    \reflr{\cox}
+     & : \forall w \to w \cox* w                                                                                                           \\
+    \symr{\cox}
+     & : \forall w_{1}, w_{2} \to w_{1} \cox* w_{2} \to w_{2} \cox* w_{1}                                                                  \\
+    \transr{\cox}
+     & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \cox* w_{2} \to w_{2} \cox* w_{3} \to w_{1} \cox* w_{3}                                    \\
+    \congrf{\cox}{\append}
+     & : \forall w_{1}, w_{2}, w_{3}, w_{4} \to  w_{1} \cox* w_{2} \to w_{3} \cox* w_{4} \to w_{1} \append w_{3} \cox* w_{2} \append w_{4} \\
+    \relr{\cox}
+     & : \forall w_{1}, w_{2} \to w_{1} \cox w_{2} \to w_{1} \cox* w_{2}                                                                   \\
+  \end{align*}
+\end{definition}
+
+The choice of generators for this presentation is important. \todo{Think about this.}
+
+\begin{itemize}
+  \item It affects the difficulty of solving the word problem in $\Sn$ and
+        formalising the proof of its correctness.
+  \item It affects the proof of strong normalisation of reduction for the relations.
+  \item It dictates which words are normal forms in this presentation of $\Sn$.
+  \item It has to closely match the $\PiLang$ comobinators so that we can prove
+        completeness.
+  \item It dictates the choice of reversible gates in the synthesis and
+        normalisation of boolean circuits which are applications that we show in
+        the later sections.
+\end{itemize}
+
+We consider the generators for the Coxeter relations on $\List[\Fin[n]]$ as a
+rewriting system. Throwing out reflexivity, and symmetry, we observe that the
+right hand sides of the relations are strictly smaller than the left hand sides,
+in terms of the lexicographical ordering on words in $\Fin[n]$. By directing the
+relation from left to right, we would be able to prove a termination property
+out of the box.
+
+However, it is not at all obvious that the system would be confluent.
+There are many possible critical pairs, such as
+% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzQsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwxLCJcXGJyYWlkIiwxXSxbMCwyLCJcXGJyYWlkIiwxXV0=
+\[\begin{tikzcd}
+    && {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
+    {\tau_1\tau_2\tau_1\tau_1\tau_2} &&&& {\tau_2\tau_1\tau_1\tau_2\tau_1}
+    \arrow["\braid"{description}, from=1-3, to=2-1]
+    \arrow["\braid"{description}, from=1-3, to=2-5]
+  \end{tikzcd}\]
+or
+% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8yIl0sWzAsMSwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFs0LDEsIlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiXFxicmFpZCIsMV0sWzAsMiwiXFxjYW5jZWwiLDFdXQ==
+\[\begin{tikzcd}
+    && {\tau_2\tau_1\tau_2\tau_2} \\
+    {\tau_1\tau_2\tau_1\tau_2} &&&& {\tau_2\tau_1}
+    \arrow["\braid"{description}, from=1-3, to=2-1]
+    \arrow["\cancel"{description}, from=1-3, to=2-5]
+  \end{tikzcd}\]
+Although in the cases above, critical pairs resolve, there are some examples where they do not. One of the simplest ones is the following:
+% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8zXFx0YXVfMlxcdGF1XzNcXHRhdV8xIl0sWzAsMSwiXFx0YXVfMlxcdGF1XzNcXHRhdV8yXFx0YXVfMSJdLFs0LDEsIlxcdGF1XzNcXHRhdV8yXFx0YXVfMVxcdGF1XzMiXSxbMCwxLCJcXGJyYWlkIiwxXSxbMCwyLCJcXHN3YXAiLDFdXQ==
+\[\begin{tikzcd}
+    && {\tau_3\tau_2\tau_3\tau_1} \\
+    {\tau_2\tau_3\tau_2\tau_1} &&&& {\tau_3\tau_2\tau_1\tau_3}
+    \arrow["\braid"{description}, from=1-3, to=2-1]
+    \arrow["\swap"{description}, from=1-3, to=2-5]
+  \end{tikzcd}\]
+
+Because of this counterexample, the equations have to be fixed. We thus propose a new presentation, keeping the
+$\cancel$ and $\swap$ constructors, but replace $\braid$ with a more general $\longbraid$:
 
@@
 
-This construction is called a Coxeter presentation. Writing it formally, we
-get the following definition
-Definition: Coxeter presentation.
+\begin{definition}[$\longcox$]
+  \begin{align*}
+    \longcancel
+    & : \forall n, l, r \to (l \append n \cons n \cons r) \longcox (l \append r) \\
+    \longswap
+    & : \forall k, n, l, r \to (\suc[k] < n) \to (l \append n \cons k \cons r) \longcox (l \append k \cons n \append r) \\
+    \longbraid
+    & : \forall n, l, r \to (l \append (n \downarrow 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downarrow 2 + k) \append r) \\
+  \end{align*}
+\end{definition}
 
-13. There are two main considerations to this choice:
-13a. How well the presentation matches the syntactic side of Pi? This is,
-how closely does the operational semantics built from these particular
-primitive operations match the natural operational semantics of Pi?
-13b. How convenient is it to turn the presentation it into rewriting system
-and solve the word problem for it?
+\begin{definition}[$\longcox*$]
+  \begin{align*}
+    \reflr{\longcox}
+    & : \forall w \to w \longcox* w \\
+    \transr{\longcox}
+    & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcox* w_{3} \to w_{1} \longcox* w_{3} \\
+  \end{align*}
+\end{definition}
 
-Thus, the argument:
-14. TODO I don't know what to write here, we have to think.
+\review{%
+  In return for increased complexity of the equations, it satisfies the diamond property (if we direct the equation).
+  Thus, together with the (preserved) termination, it creates a confluent rewriting system. Proof: All the critical
+  pairs resolve, in the appendix. Definition: A normal form is a list that cannot be reduced any further.%
+}
 
-15a. The equations that define Coxeter presentation have a nice property:
-as it is written, the LHS is smaller (lexicographically) than the RHS.
-Thus, if we could direct them from right to left, it would have a
-termination property out of the box.
-15b. However, it is not at all obvious that the system would be confluent.
-There are many possible critical pairs, such as
-TODO diagram 1212 (braid, braid)
-or
-TODO diagram 3233 (braid, cancel)
-As it turns out, there are instances of critical pairs that do not resolve.
-A simplest example is the following:
-TODO diagram 3213
+Now, we state and prove that $(\List[\Fin[n]],\longcox*)$ is a confluent and strongly normalising rewriting system.
 
-16. The equations have to be changed. We thus propose a new presentation.
-We keep the (cancel) and (swap) equations, but replace (braid) with a more
-general:
-TODO diagram:
-
-Definition: Long Coxeter presentation.
-
-17. In return for increased complexity of the equations, it satisfies the
-diamond property (if we direct the equation). Thus, together with the
-(preserved) termination, it creates a confluent rewriting system.
-Proof: All the critical pairs resolve, in the appendix.
-Definition: A normal form is a list that cannot be reduced any further.
+\begin{proposition}
+  \leavevmode
+  \begin{enumerate}
+    \item $\longcox*$ is (locally) confluent. For every span $\longcoxspan{w_{1}}{w_{2}}{w_{3}}$, there is a matching
+          extended cospan $\longcoxcospan*{w_{2}}{w_{3}}{w}$.
+    \item $\longcox*$ is confluent. For every extended span $\longcoxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
+          extended cospan $\longcoxcospan*{w_{2}}{w_{3}}{w}$.
+    \item $\longcox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \longcox* v$.
+    \item $\cox*$ and $\longcox*$ are equivalent. For every $w$ and $v$, $w \cox* v$ iff $w \longcox* v$.
+  \end{enumerate}
+\end{proposition}
 
@@
 Proposition: Long Coxeter is equivalent to normal Coxeter.
+
+\todo{Long form of Coxeter relations and their equivalence.}
 
@@
 do these representatives look like exactly.
+
+We state and prove two desirable properties of this relation on $\List[\Fin[n]]$ from the point of view of an abstract
+rewriting system. We follow the terminology of~\cite{krausCoherenceWellFoundednessTaming2020}. \vc{Not completely
+  formalised. Is this with Coxeter or LongCoxeter?}
+
+The type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.
+
+\begin{definition}[$\Sn$]
+  \(\Sn \defeq \quot{\List(\Fin[n])}{\cox*}\)
+\end{definition}
+
+We will now prove the group structure of $\Sn$.
+
+\begin{proposition}
+  There is a group structure on $\Sn$, where the identity element is $\nil$, multiplication is given by list append, and
+  inverse is given by list reversal.
+\end{proposition}
+
+\begin{proposition}
+  $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to
+  $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $[\term{id},\term{id}] : A + A \to A$.
+\end{proposition}
 
@@
 
-\subsection{Symmetric groups}~\label{subsec:symmetric}
-
-There is an obvious group structure on $\Aut[\Fin[n]]$ given by identity,
-composition, and inverse. This is the symmetric group $S_n$ on $n$ symbols. In
-the rest of the section we will construct a convenient presentation of this
-group.
-
-\vc{this is just a rough draft for now}
-
-\todo{Reference T-algebra presentations as coequalisers (Mac Lane 6.7)}
-
-First, we formally define a presentation of a group.
-
-\todo{Free 1-group, needs to be truncated}
-
-\subsubsection{Coxeter Relations}
-
-We define a binary relation on $\List[\Fin[n]]$.
-
-\begin{definition}[$\cox$]
-  \begin{align*}
-    \cancel
-     & : \forall n \to (n \cons n \cons \nil) \cox \nil                                                     \\
-    \swap
-     & : \forall k, n \to (\suc[k] < n) \to (n \cons k \cons \nil) \cox (k \cons n \cons \nil)              \\
-    \braid
-     & : \forall n \to (\suc[n] \cons n \cons \suc[n] \cons \nil) \cox (n \cons \suc[n] \cons n \cons \nil) \\
-  \end{align*}
-\end{definition}
-
-We define $\cox*$ as the congruence closure of $\cox$.
-\vc{This could be simplified a bit depending on our formalisation. $\cox$ is directed, use a different symbol?}
-
-\begin{definition}[$\cox*$]
-  \begin{align*}
-    \reflr{\cox}
-     & : \forall w \to w \cox* w                                                                                                           \\
-    \symr{\cox}
-     & : \forall w_{1}, w_{2} \to w_{1} \cox* w_{2} \to w_{2} \cox* w_{1}                                                                  \\
-    \transr{\cox}
-     & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \cox* w_{2} \to w_{2} \cox* w_{3} \to w_{1} \cox* w_{3}                                    \\
-    \congrf{\cox}{\append}
-     & : \forall w_{1}, w_{2}, w_{3}, w_{4} \to  w_{1} \cox* w_{2} \to w_{3} \cox* w_{4} \to w_{1} \append w_{3} \cox* w_{2} \append w_{4} \\
-    \relr{\cox}
-     & : \forall w_{1}, w_{2} \to w_{1} \cox w_{2} \to w_{1} \cox* w_{2}                                                                   \\
-  \end{align*}
-\end{definition}
-
-\todo{Long form of Coxeter relations and their equivalence.}
-
-We state and prove two desirable properties of this relation on $\List[\Fin[n]]$ from the point of view of an abstract
-rewriting system. We follow the terminology of~\cite{krausCoherenceWellFoundednessTaming2020}. \vc{Not completely
-  formalised. Is this with Coxeter or LongCoxeter?}
-
-\begin{proposition}
-  \leavevmode
-  \begin{enumerate}
-    \item $\cox$ is (locally) confluent. For every span $w_{1} \cox w \cox w_{2}$, there is a matching extended cospan
-          $w_{1} \cox* v \cox* w_{2}$.
-    \item $\cox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \cox* v$.
-          \review{Equivalently, the type $\dsum{v:\List[\Fin[n]]}{w \cox* v}$ is contractible, and $w \cox* v$ is a
-            proposition for every $w$ and $v$.}
-  \end{enumerate}
-\end{proposition}
-
-The type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.
-
-\begin{definition}[$\Sn$]
-  \(\Sn \defeq \quot{\List(\Fin[n])}{\cox*}\)
-\end{definition}
-
-We will now prove the group structure of $\Sn$.
-
-\begin{proposition}
-  There is a group structure on $\Sn$, where the identity element is $\nil$, multiplication is given by list append, and
-  inverse is given by list reversal.
-\end{proposition}
-
-\begin{proposition}
-  $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to
-  $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $[\term{id},\term{id}] : A + A \to A$.
-\end{proposition}
-
 We will now show how to generate a Lehmer code from a word in $\Sn$ and back.

### 2021-06-20 Pi+/paper/finite.tex
Revise Coxeter section

@@
 
-We consider the generators for the Coxeter relations on $\List[\Fin[n]]$ as a
-rewriting system. Throwing out reflexivity, and symmetry, we observe that the
-right hand sides of the relations are strictly smaller than the left hand sides,
-in terms of the lexicographical ordering on words in $\Fin[n]$. By directing the
-relation from left to right, we would be able to prove a termination property
-out of the box.
+To consider solving the word problem for $\Sn$, we need to formalise the generators for the Coxeter relations on
+$\List[\Fin[n]]$ as a rewriting system $(\List[\Fin[n]],\cox*)$. The $\cox*$ relation as defined, does not produce a
+well-behaved rewriting system due to the symmetric closure. Throwing out reflexivity and symmetry, we observe that the
+right hand sides of the relations are strictly smaller than the left hand sides, in terms of the lexicographical
+ordering on words in $\Fin[n]$. By directing the relation from left to right, we would be able to prove a termination
+property out of the box.
 
-However, it is not at all obvious that the system would be confluent.
-There are many possible critical pairs, such as
-% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzQsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwxLCJcXGJyYWlkIiwxXSxbMCwyLCJcXGJyYWlkIiwxXV0=
-\[\begin{tikzcd}
-    && {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
-    {\tau_1\tau_2\tau_1\tau_1\tau_2} &&&& {\tau_2\tau_1\tau_1\tau_2\tau_1}
-    \arrow["\braid"{description}, from=1-3, to=2-1]
-    \arrow["\braid"{description}, from=1-3, to=2-5]
-  \end{tikzcd}\]
-or
-% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8yIl0sWzAsMSwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFs0LDEsIlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiXFxicmFpZCIsMV0sWzAsMiwiXFxjYW5jZWwiLDFdXQ==
-\[\begin{tikzcd}
-    && {\tau_2\tau_1\tau_2\tau_2} \\
-    {\tau_1\tau_2\tau_1\tau_2} &&&& {\tau_2\tau_1}
-    \arrow["\braid"{description}, from=1-3, to=2-1]
-    \arrow["\cancel"{description}, from=1-3, to=2-5]
-  \end{tikzcd}\]
-Although in the cases above, critical pairs resolve, there are some examples where they do not. One of the simplest ones is the following:
-% https://q.uiver.app/?q=WzAsMyxbMiwwLCJcXHRhdV8zXFx0YXVfMlxcdGF1XzNcXHRhdV8xIl0sWzAsMSwiXFx0YXVfMlxcdGF1XzNcXHRhdV8yXFx0YXVfMSJdLFs0LDEsIlxcdGF1XzNcXHRhdV8yXFx0YXVfMVxcdGF1XzMiXSxbMCwxLCJcXGJyYWlkIiwxXSxbMCwyLCJcXHN3YXAiLDFdXQ==
-\[\begin{tikzcd}
-    && {\tau_3\tau_2\tau_3\tau_1} \\
-    {\tau_2\tau_3\tau_2\tau_1} &&&& {\tau_3\tau_2\tau_1\tau_3}
-    \arrow["\braid"{description}, from=1-3, to=2-1]
-    \arrow["\swap"{description}, from=1-3, to=2-5]
-  \end{tikzcd}\]
+A well-behaved rewriting system should also be confluent, however, it is not at all obvious from this definition. Due to
+overlapping reduction rules, there are many possible critical pairs, such as
 
-Because of this counterexample, the equations have to be fixed. We thus propose a new presentation, keeping the
-$\cancel$ and $\swap$ constructors, but replace $\braid$ with a more general $\longbraid$:
+\[
+  \begin{array}{lcr}
+    \gspan[\braid][\braid]{\tau_2\tau_1\tau_2\tau_1\tau_2}{\tau_1\tau_2\tau_1\tau_1\tau_2}{\tau_2\tau_1\tau_1\tau_2\tau_1}
+     &
+    \text{or}
+     &
+    \gspan[\braid][\cancel]{\tau_2\tau_1\tau_2\tau_2}{\tau_1\tau_2\tau_1\tau_2}{\tau_2\tau_1}
+  \end{array}
+\]
 
-Diagram of long-braid:
+For confluence, all critical pairs should converge. The pairs above can be resolved, but there are many examples where
+they do not. The following is a simple example.
+
+\[
+  \gspan[\braid][\swap]{\tau_3\tau_2\tau_3\tau_1}{\tau_2\tau_3\tau_2\tau_1}{\tau_3\tau_2\tau_1\tau_3}
+\]
+
+Because of this counterexample, the relations have to be changed. We propose a long form of the relations as a new
+presentation, where we allow $\cancel$ and $\swap$ reductions anywhere inside the list, called $\longcancel$ and
+$\longswap$, but replace $\braid$ with a more general $\longbraid$ relation.
 
@@
 
+The directed $\longcox$ relation is defined formally with the following generators.
+
 \begin{definition}[$\longcox$]
@@
 
+Like before, we define the $\longcox*$ to be the closure of $\longcox$, but only under reflexivity and transitivity. The
+transitive extension is only performed on the right.
+
 \begin{definition}[$\longcox*$]
@@
 
-\review{%
-  In return for increased complexity of the equations, it satisfies the diamond property (if we direct the equation).
-  Thus, together with the (preserved) termination, it creates a confluent rewriting system. Proof: All the critical
-  pairs resolve, in the appendix. Definition: A normal form is a list that cannot be reduced any further.%
-}
-
-Now, we state and prove that $(\List[\Fin[n]],\longcox*)$ is a confluent and strongly normalising rewriting system.
+Despite the increased complexity of the relations, the rewriting system $(\List[\Fin[n]],\longcox*)$ can be shown to
+satisfy confluence, that is, the Church-Rosser (diamond) property. Further, we can show that this system is strongly
+normalising, by producing a unique normal form that doesn't reduce any further. We follow the terminology
+of~\cite{krausCoherenceWellFoundednessTaming2020} to state our results.
 
@@
   \begin{enumerate}
-    \item $\longcox*$ is (locally) confluent. For every span $\longcoxspan{w_{1}}{w_{2}}{w_{3}}$, there is a matching
-          extended cospan $\longcoxcospan*{w_{2}}{w_{3}}{w}$.
-    \item $\longcox*$ is confluent. For every extended span $\longcoxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
-          extended cospan $\longcoxcospan*{w_{2}}{w_{3}}{w}$.
+    \item $\longcox$ is (locally) confluent. For every span $\coxspan{w_{1}}{w_{2}}{w_{3}}$, there is a matching
+          extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
+    \item $\longcox*$ is confluent. For every extended span $\coxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
+          extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
     \item $\longcox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \longcox* v$.
-    \item $\cox*$ and $\longcox*$ are equivalent. For every $w$ and $v$, $w \cox* v$ iff $w \longcox* v$.
   \end{enumerate}
+\end{proposition}
+
+Another property we're interested in is well-foundedness of the relation $\longcox*$.
+\todo{lexicographic order, accessibility predicate and well-foundedness}
+
+In the theory of group presentations, the Coxeter relations are generally presented in their short form~\todo{citation},
+but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}. The long form of the
+Coxeter relations are unwieldy and difficult to prove properties about by induction, and have little connection to the
+broader scope of computational group theory. However, we can prove that the closures of the two forms of the relations
+are equivalent, and work up to this equivalence!
+
+\begin{proposition}
+  $\cox*$ and $\longcox*$ are equivalent. For every $w$ and $v$, $w \cox* v$ iff $w \longcox* v$.
 \end{proposition}
@@
 elements that have the same normal form.
-
-19. However, the Long Coxeter relations are unwieldly and have no
-connectino to the broader scope of computational group theory - while the
-usual Coxeter presentation is to-go choice in this area (cite all these
-papers in Coq, this guy from Sweden(?) etc.).
-Thus, we prove:
-Proposition: Long Coxeter is equivalent to normal Coxeter.
-
-\todo{Long form of Coxeter relations and their equivalence.}
-
-19a. (Note: here we try to convince the reader about reusability badge) So,
-the Long Coxeter can be regarded as a trick to get the proof to pass - but
-the external user-facing interface is still that of the usual Coxeter
-presentation.
 
@@
 do these representatives look like exactly.
-
-We state and prove two desirable properties of this relation on $\List[\Fin[n]]$ from the point of view of an abstract
-rewriting system. We follow the terminology of~\cite{krausCoherenceWellFoundednessTaming2020}. \vc{Not completely
-  formalised. Is this with Coxeter or LongCoxeter?}
 

### 2021-06-20 Pi+/paper/reversible1.tex
rev1

--- 2021-06-18 Pi+/paper/reversible1.tex
@@
-\section{A Reversible Programming Language: Set-Theoretic Semantics}~\label{sec:reversibleone}
+\section{A Reversible Programming Language: Set-Theoretic Semantics}
+\label{sec:reversibleone}
 
@@
 A natural candidate for a semantic foundation for reversible programming languages is the notion of type
-isomorphism. Indeed, the type isomorphisms among finite types represent a universal language for reversible boolean
-circuits~\cite{James:2012:IE:2103656.2103667} and the extension with recursive types and trace
-operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
+isomorphism. Indeed, the type isomorphisms among finite types are sound and complete for all permutations on finite
+types~\cite{Fiore:2004,fiore-remarks} and hence they are \emph{complete} for expressing reversible combinational
+circuits~\cite{fredkin1982conservative, James:2012:IE:2103656.2103667,Toffoli:1980} and the extension with recursive
+types and trace operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
 language~\cite{James:2012:IE:2103656.2103667,rc2011}.
 
-Focusing on finite types, the building blocks of type theory are: the
-empty type ($\mathbb{0}$), the unit type ($\mathbb{1}$) containing
-just one value $\Acon{tt}$, the sum type ($+$) containing values of
-the form $\inlv{v}$ and $\inrv{v}$, and the product ($\times$) type
-containing pairs of values $(v_1,v_2)$. One may view each type $A$
-as a collection of physical wires that can transmit $|A|$ distinct
-values where $|A|$ is a natural number that indicates the size of a
-type, computed as: $| \mathbb{0} | = 0$; $| \mathbb{1} | = 1$;
-$| A + B | = | A | + |B |$; and
-$| A \times B | = | A | * |B |$.  Thus the type
-$\mathbb{B} = \mathbb{1} + \mathbb{1}$ corresponds to a wire that can
-transmit one of two values, i.e., bits, with the convention that
-$\inlv{\Acon{tt}}$ represents \AgdaFunction{𝔽} and
-$\inrv{\Acon{tt}}$ represents \AgdaFunction{𝕋}.  The type
-$\mathbb{B} \times \mathbb{B} \times \mathbb{B}$ corresponds to a
-collection of wires that can transmit three bits. From that
-perspective, a type isomorphism between types $A$ and $B$
-(such that $|A|=|B|=n$) models a \emph{reversible}
-combinational circuit that \emph{permutes} the $n$ different
-values. These type isomorphisms are collected in Fig.~\ref{pi-terms}.
+Focusing on finite types, the building blocks of type theory are: the empty type ($\zerot$), the unit type
+($\onet$) containing just one value $\Acon{tt}$, the sum type ($+$) containing values of the form $\inlv{v}$ and
+$\inrv{v}$, and the product ($\times$) type containing pairs of values $(v_1,v_2)$. For this fragment of types, the
+syntax of the language $\Pi$ consists of the following sorts:
+
+{\scalebox{\scalef}{$%
+\begin{array}{lrcl}
+\textit{Value types} & A,B,C,D &::=& \zerot \alt \onet \alt A+B \alt A\times B \\
+\textit{Values}      & v,w,x,y &::=& \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
+\textit{Program types} &&& A \iso B \\
+\textit{Programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})
+\end{array}$}}
 \begin{figure}[t]
@@
 \\
-\identlp :&  \mathbb{0} + A & \iso & A &: \identrp \\
+\identlp :&  \zerot + A & \iso & A &: \identrp \\
 \swapp :&  A + B & \iso & B + A &: \swapp \\
 \assoclp :&  A + (B + C) & \iso & (A + B) + C &: \assocrp \\ [1.5ex]
-\identlt :&  \mathbb{1} \times A & \iso & A &: \identrt \\
+\identlt :&  \onet \times A & \iso & A &: \identrt \\
 \swapt :&  A \times B & \iso & B \times A &: \swapt \\
 \assoclt :&  A \times (B \times C) & \iso & (A \times B) \times C &: \assocrt \\ [1.5ex]
-\absorbr :&~ \mathbb{0} \times A & \iso & \mathbb{0} ~ &: \factorzl \\
+\absorbr :&~ \zerot \times A & \iso & \zerot ~ &: \factorzl \\
 \dist :&~ (A + B) \times C & \iso & (A \times C) + (B \times C)~ &: \factor
@@
 
-Each line in the top part of the figure introduces a pair of dual constants that witness the type isomorphism in the
-middle.  These are the \emph{base} (non-reducible) terms of $\Pi$. The isomorphisms are extended to form a congruence
-relation by adding the three constructors at the bottom of the figure that witness equivalence and compatible
-closure. Although austere, this combinator-based language has the advantage of being amenable to formal analysis for at
-least two reasons: (i) it is conceptually simple and small, and (ii) it has direct and evident connections to type
-theory and category theory and hence possesses a rich algebraic structure which we exploit in the remainder of the
-paper.
-
-% Specifically, the type isomorphisms of~$\Pi$ are sound and complete for all permutations on finite
-% types~\cite{Fiore:2004,fiore-remarks} and hence they are \emph{complete} for expressing reversible combinational
-% circuits~\cite{fredkin1982conservative, James:2012:IE:2103656.2103667,Toffoli:1980}. Algebraically, these types and
-% combinators form a \emph{commutative semiring} (up to type isomorphism). Logically, they form a superstructural logic
-% capturing space-time trade-offs~\cite{superstructural}. Categorically, they form a \emph{distributive bimonoidal
-%   category}~\cite{laplaza72}.
-
-% For the fragment over finite types, the syntax of the language $\Pi$ consists of several sorts:
-
-%  {\scalebox{\scalef}{$%
-% \begin{array}{lrcl}
-% \textit{Value types} & A,B,C,D &::=& \mathbb{0} \alt \mathbb{1} \alt A+B \alt A\times B \\
-% \textit{Values}      & v,w,x,y &::=& \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
-% \textit{Program types} &&& A \leftrightarrow B \\
-% \textit{Programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})
-% \end{array}$}}
-
-\begin{figure}[t]
-{\scalebox{\scalef}{%
-$\begin{array}{rclcrcl}
-\delta(\identlp,~\inrv{v}) & = & v & & \delta(\identrp,~v)        & = & \inrv{v}\\
-\delta(\swapp,~\inlv{v})     & = & \inrv{v}\\
-\delta(\swapp,~\inrv{v})     & = & \inlv{v}\\
-\delta(\assoclp,~\inlv{v})          & = & \inlv{(\inlv{v})} & & \delta(\assocrp,~\inlv{(\inlv{v})}) & = & \inlv{v}         \\
-\delta(\assoclp,~\inrv{(\inlv{v})}) & = & \inlv{(\inrv{v})} & & \delta(\assocrp,~\inlv{(\inrv{v})}) & = & \inrv{(\inlv{v})}\\
-\delta(\assoclp,~\inrv{(\inrv{v})}) & = & \inrv{v}          & & \delta(\assocrp,~\inrv{v})          & = & \inrv{(\inrv{v})}\\
-\delta(\identlt,~(\Acon{tt},v)) & = & v & & \delta(\identrt,~v)             & = & (\Acon{tt},v)\\
-\delta(\swapt,~(x,y))         & = & (y,x)    \\
-\delta(\assoclt,~(x,(y,z))) & = & ((x,y),z) & & \delta(\assocrt,~((x,y),z)) & = & (x,(y,z))\\
-\delta(\dist,~(\inlv{x},z)) & = & \inlv{(x,z)} & & \delta(\factor,~\inlv{(x,z)}) & = & (\inlv{x},z)\\
-\delta(\dist,~(\inrv{y},z)) & = & \inrv{(y,z)} & & \delta(\factor,~\inrv{(y,z)}) & = & (\inrv{y},z)
-\end{array}$}}
-\caption{Semantics of base combinators}\label{fig:delta}
-\end{figure}
+\noindent Intuitively, one may view each type $A$ as a collection of physical wires that can transmit $\sizet{A}$ distinct
+values where $\sizet{A}$ is a natural number that indicates the size of a type, computed as: $\sizet{\zerot} = 0$;
+$\sizet{\onet} = 1$; $\sizet{A + B} = \sizet{A} + \sizet{B}$; and $\sizet{A \times B} = \sizet{A} * \sizet{B}$.  Thus the type
+$\mathbb{B} = \onet + \onet$ corresponds to a wire that can transmit one of two values, i.e., bits, with the
+convention that $\inlv{\Acon{tt}}$ represents \AgdaFunction{𝔽} and $\inrv{\Acon{tt}}$ represents \AgdaFunction{𝕋}.  The
+type $\mathbb{B} \times \mathbb{B} \times \mathbb{B}$ corresponds to a collection of wires that can transmit three
+bits. From that perspective, a type isomorphism between types $A$ and $B$ (such that $\sizet{A}=\sizet{B}=n$) models a
+\emph{reversible} combinational circuit that \emph{permutes} the $n$ different values. These type isomorphisms form a
+\emph{commutative semiring} and are collected in Fig.~\ref{pi-terms}. Each line in the top part of the figure
+introduces a pair of dual constants that witness the type isomorphism in the middle.  These are the \emph{base}
+(non-reducible) terms of $\Pi$. The isomorphisms are extended to form a congruence relation by adding the three
+constructors at the bottom of the figure that witness equivalence and compatible closure. Although austere, this
+combinator-based language has the advantage of being amenable to formal analysis for at least two reasons: (i) it is
+conceptually simple and small, and (ii) it has direct and evident connections to type theory and category theory and
+hence possesses a rich algebraic structure which we exploit in the remainder of the paper.
 
@@
 
-\noindent A simple denotational semantics of $\Pi$ can be specified by mapping each type to a finite set and each
-combinator to a permutation between finite sets.
+\noindent A denotational semantics of $\Pi$ can be specified by mapping each type to a finite set and each combinator to
+a bijection between finite sets, that is, a permutation. In this section, we provide a simple semantics expressed in the
+metalanguage of set theory. This semantics is then extended to the metalanguage of homotopy type theory in
+Sec.~\ref{sec:reversibletwo}.
 
-Fig.~\ref{fig:delta} specifies the semantics of the base combinators
-using a function $\delta$. It is clear that this function has an inverse $\delta^\dagger$ such that
-$\delta^\dagger(c,\delta(c,v))=v$. The semantics of full programs is given by the interpreter in the second figure.
+As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
+we have that $\sizet{A} = \sizet{B}$. Hence the denotation of a type $A$ with $n$-elements will be a finite set
+$\Fin[n]$ and the denotation of a combinator $c : A \iso B$ such that $\sizet{A} = \sizet{B} = n$ will be a function
+from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. In more detail, we enumerate the values $v : A$ where
+$\sizet{A} = n$ in a canonical order such that each value $v$ is given an index in $\{ 0,\cdots,n-1 \}$. The enumeration
+is defined as follows:
+\[\begin{array}{rcl}
+    \mathit{enum}(\zerot) &=& \emptyset \\
+    \mathit{enum}(\onet) &=& \{ \Acon{tt} \} \\
+    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
+    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda w.\mathit{map}~(\lambda v. (v,w))~\mathit{enum}(A))~\mathit{enum}(B))
+\end{array}\]
+\noindent For example, the 3 values $\inlv{\Acon{tt}}$, $\inrv{(\inlv{\Acon{tt}})}$, and $\inrv{(\inrv{\Acon{tt}})}$ of
+type $\onet + (\onet + \onet)$ are given indices 0, 1, and 2 respectively. Given this canonical order, the combinator
+$\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$ would correspond to the permutation $(1 2 0)$ while the
+combinator $\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$ at the same type would correspond to the
+identity permutation.
 
-We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
-compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
-$\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:
-
-\[\begin{array}{rcll}
-    \sem{A} &=& \mathsf{Fin}~\sizet{A}
-\\
-    \sem{c : A \isot B} &:& \mathsf{Fin}~\sizet{A} \rightarrow \mathsf{Fin}~\sizet{B} \\
-    %% base combinators
-    \sem{c_1\oplus c_2}_{A+B \isot C+D}~\inlv{v} &=& \sem{c_1}_{A \isot C}~v \\
-    \sem{c_1\oplus c_2}_{A+B \isot C+D}~\inrv{v} &=& \sizet{A} + \sem{c_2}_{B \isot D}~v \\
-    \sem{\idc}~v &=& v \\
-    \sem{c_1\fatsemi c_2}~v &=& \sem{c_2} (\sem{c_1}~v) \\
-    \sem{c_1\otimes c_2}~(v_1,v_2) &=& (\sem{c_1}~v_1 , \sem{c_2}~v_2)
-\end{array}\]
+% Such functions are finitely supported, that is, their outputs can be tabulated using the canonical ordering on
+% $\Fin[n]$. For bijections, this gives a listed permutation. By observing the action of the combinators on the values of
+% the finite set, we can define a denotational semantics which constructs the bijection.
+% \note{\(\Fin[n] \to A \eqv \type{Vec_{n}}(A)\)}
 

### 2021-06-20 Pi+/paper/finite.tex
Normalisation lemmas

@@
 
-18. We define the equivalence classes on List (Fin n) to be precisely these
-elements that have the same normal form.
+By strong normalisation, we get a unique choice function that produces a normal form for terms of $\List[\Fin[n]]$.
+Equivalence classes on $\List[\Fin[n]]$ are precisely those terms that have the same normal form.
 

### 2021-06-20 Pi+/paper/reversible1.tex
enough for sec 3 for now?

@@
 \noindent A denotational semantics of $\Pi$ can be specified by mapping each type to a finite set and each combinator to
-a bijection between finite sets, that is, a permutation. In this section, we provide a simple semantics expressed in the
-metalanguage of set theory. This semantics is then extended to the metalanguage of homotopy type theory in
+a bijection between finite sets, that is, a permutation. In this section, we outline a simple semantics expressed in the
+metalanguage of set theory. This semantics is formalized and extended to the metalanguage of homotopy type theory in
 Sec.~\ref{sec:reversibletwo}.
@@
 type $\onet + (\onet + \onet)$ are given indices 0, 1, and 2 respectively. Given this canonical order, the combinator
-$\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$ would correspond to the permutation $(1 2 0)$ while the
+$\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$ would correspond to the permutation $(1~2~0)$ while the
 combinator $\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$ at the same type would correspond to the

### 2021-06-21 Pi+/paper/finite.tex
Finish Coxeter section

@@
 
-To do so, we will characterise the automorphisms on finite sets of cardinality
-$n$, and show them to be equivalent to the symmetric group $\Sn$, via the
-Coxeter presentation. We will do that in two steps,
+In the previous~\cref{sec:univalent}, we established that paths in $\UFin$ are equivalent to families of automorphisms
+of $\Fin{n}$ for every $n:\Nat$, that is, bijections on finite sets of size $n$. This is the extensional view of
+permutations. In the following sections, we will characterise these permutations, going through two intermediate steps.
+
+To do so, we will characterise the automorphisms on finite sets of cardinality $n$, and show them to be equivalent to
+the symmetric group $\Sn$, via the Coxeter presentation. We will do that in two steps,
 in~\cref*{subsec:permutations,subsec:lehmer,subsec:symmetric}.
@@
 \todo{Justification for why we need group theory.}
-
-%% Coxeter presentation of $\Sn$ $\eqv$ $\Lehmer[n]$ $\eqv$ $\Aut[\Fin[n]]$
 
@@
 
-This construction is called a Coxeter presentation. Writing it formally, we get
-the following definition for $\Sn$. We define a binary relation $\cox$ on
+This construction is called a Coxeter presentation. Writing it formally, we define a binary relation $\cox$ on
 $\List[\Fin[n]]$.
@@
 
+The type $\Sn$ is then defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.
+
+\begin{definition}[$\Sn$]
+  \(\Sn \defeq \quot{\List(\Fin[n])}{\cox*}\)
+\end{definition}
+
 The choice of generators for this presentation is important. \todo{Think about this.}
@@
 To consider solving the word problem for $\Sn$, we need to formalise the generators for the Coxeter relations on
-$\List[\Fin[n]]$ as a rewriting system $(\List[\Fin[n]],\cox*)$. The $\cox*$ relation as defined, does not produce a
-well-behaved rewriting system due to the symmetric closure. Throwing out reflexivity and symmetry, we observe that the
-right hand sides of the relations are strictly smaller than the left hand sides, in terms of the lexicographical
-ordering on words in $\Fin[n]$. By directing the relation from left to right, we would be able to prove a termination
-property out of the box.
-
-A well-behaved rewriting system should also be confluent, however, it is not at all obvious from this definition. Due to
-overlapping reduction rules, there are many possible critical pairs, such as
+$\List[\Fin[n]]$ as a rewriting system $(\List[\Fin[n]],\cox*)$. A well-behaved rewriting system has to be strongly
+normalising and confluent. Throwing out reflexivity and symmetry, we observe that the right hand sides of the relations
+are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$. Thus, by
+directing the relation from left to right, we would be able to prove a termination property out of the box. For a system
+to be confluent, all critical pairs, that is, terms with overlapping possible reduction rules, have to converge. For
+example, the pairs below converge.
 
@@
 
-For confluence, all critical pairs should converge. The pairs above can be resolved, but there are many examples where
-they do not. The following is a simple example.
+Unfortunately, in the system defined, this is not true for all critical pairs. One example is the following.
 
@@
 
-Because of this counterexample, the relations have to be changed. We propose a long form of the relations as a new
-presentation, where we allow $\cancel$ and $\swap$ reductions anywhere inside the list, called $\longcancel$ and
-$\longswap$, but replace $\braid$ with a more general $\longbraid$ relation.
+\subsection{Rewriting via Coxeter}
+
+Because of the counterexample discussed, the relations have to be changed. In this section, we will formally define a
+rewriting system partially based on the Coxeter relation, prove that it has good properties (confluence and strong
+normalisation).
+
+The reduction system $(\List[\Fin[n]], \longcox*)$ has generators corresponding to $\swap$, $\cancel$ and $\braid$. We
+fix the problem discussed previously by changing $\braid$ to be a slightly more general relation. However, we do not
+want to lose the connection to the theory of group presentations, where Coxeter relations are a standard
+notion~\todo{citation}. To make the result broadly applicable, even though we use the reduction system for computation,
+we prove that its reflexive-transitive closure is indeed equivalent to the conguence closure of the standard Coxeter
+relation $\cox$.
+
+\todo {but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
+
+Formally, the directed $\longcox$ relation is defined with the following generators.
+
+\begin{definition}[$\longcox$]
+  \begin{align*}
+    \longcancel
+     & : \forall n, l, r \to (l \append n \cons n \cons r) \longcox (l \append r)                                                                           \\
+    \longswap
+     & : \forall k, n, l, r \to (\suc[k] < n) \to (l \append n \cons k \cons r) \longcox (l \append k \cons n \append r)                                    \\
+    \longbraid
+     & : \forall n, l, r \to (l \append (n \downarrow 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downarrow 2 + k) \append r) \\
+  \end{align*}
+\end{definition}
+
+Like before, we define the $\longcox*$ to be the closure of $\longcox$, but only under reflexivity and transitivity. The
+transitive extension is only performed on the right.
+
+\begin{definition}[$\longcox*$]
+  \begin{align*}
+    \reflr{\longcox}
+     & : \forall w \to w \longcox* w                                                                               \\
+    \transr{\longcox}
+     & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcox* w_{3} \to w_{1} \longcox* w_{3} \\
+  \end{align*}
+\end{definition}
+
+Generators $\longcancel$ and $\longswap$ correspond directly to the generators of $\cox*$ and can be visualised in the
+same way as before. The generator $\longbraid$ uses a helper $n \downf k$, which produces the sequence
+\(k + n, k + n - 1 \ldots, n + 1\).
+
+\begin{definition}[$\downf : (n : \Nat) \to {\Fin[n]} \to {\List[\Fin[\suc[k + n]]]}$]
+  \begin{align*}
+    n \downf 0       & \defeq \nil                       \\
+    n \downf \suc[k] & \defeq (k + n) \cons (n \downf k)
+  \end{align*}
+\end{definition}
+
+As a sequence of transpositions, it moves the element $k - 1$ places. Expressed in terms of the braiding diagram, for
+$n = \dots$ and $k = \dots$, it has the following form:
+
+\[
+  \begin{tikzpicture}
+    \pic[local bounding box=my braid,braid/.cd,
+      number of strands = 6,
+      thick]
+    {braid={s_5, s_4, s_3, s_2, s_1}};
+  \end{tikzpicture}
+\]
+
+Now, the remaining constructor $\longbraid$ moves a transposition around such sequence, as in the diagram below.
 
@@
 
-The directed $\longcox$ relation is defined formally with the following generators.
+Note that the previous $\braid$ rule is a special case of $\longbraid$, with $k = 2$.
 
-\begin{definition}[$\longcox$]
-  \begin{align*}
-    \longcancel
-     & : \forall n, l, r \to (l \append n \cons n \cons r) \longcox (l \append r)                                                                           \\
-    \longswap
-     & : \forall k, n, l, r \to (\suc[k] < n) \to (l \append n \cons k \cons r) \longcox (l \append k \cons n \append r)                                    \\
-    \longbraid
-     & : \forall n, l, r \to (l \append (n \downarrow 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downarrow 2 + k) \append r) \\
-  \end{align*}
-\end{definition}
-
-Like before, we define the $\longcox*$ to be the closure of $\longcox$, but only under reflexivity and transitivity. The
-transitive extension is only performed on the right.
-
-\begin{definition}[$\longcox*$]
-  \begin{align*}
-    \reflr{\longcox}
-     & : \forall w \to w \longcox* w                                                                               \\
-    \transr{\longcox}
-     & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcox* w_{3} \to w_{1} \longcox* w_{3} \\
-  \end{align*}
-\end{definition}
-
-Despite the increased complexity of the relations, the rewriting system $(\List[\Fin[n]],\longcox*)$ can be shown to
-satisfy confluence, that is, the Church-Rosser (diamond) property. Further, we can show that this system is strongly
-normalising, by producing a unique normal form that doesn't reduce any further. We follow the terminology
-of~\cite{krausCoherenceWellFoundednessTaming2020} to state our results.
+Despite the increased complexity of the generators, the rewriting system $(\List[\Fin[n]],\longcox*)$ has the properties
+we desire. It satisfies confluence, that is, the Church-Rosser (diamond) property, and it is strongly normalising,
+because it produces a unique irreducible normal form. Existence of the normal form is implied by the well-foundedness of
+the relation -- informally, it does not contain any infinite decreasing sequences. We follow the terminology
+of~\cite{krausCoherenceWellFoundednessTaming2020} to state our results formally.~\vc{check!}
 
@@
           extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
+    \item $\longcox*$ is well-founded. With $<$ the lexicographic ordering on $\List[\Fin[n]]$, and $w$ being
+          $<$-accessible if every $v < w$ is $<$-accessible, we have that every $w$ is $<$-accesible.
     \item $\longcox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \longcox* v$.
@@
 
-Another property we're interested in is well-foundedness of the relation $\longcox*$.
-\todo{lexicographic order, accessibility predicate and well-foundedness}
-
-In the theory of group presentations, the Coxeter relations are generally presented in their short form~\todo{citation},
-but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}. The long form of the
-Coxeter relations are unwieldy and difficult to prove properties about by induction, and have little connection to the
-broader scope of computational group theory. However, we can prove that the closures of the two forms of the relations
-are equivalent, and work up to this equivalence!
+The long form of the Coxeter relations are unwieldy and difficult to prove properties about by induction, and have
+little connection to the broader scope of computational group theory. However, we can prove that the closures of the two
+forms of the relations are equivalent, and work up to this equivalence!
 
@@
 
-By strong normalisation, we get a unique choice function that produces a normal form for terms of $\List[\Fin[n]]$.
-Equivalence classes on $\List[\Fin[n]]$ are precisely those terms that have the same normal form.
+By strong normalisation, we get a unique choice function $\normf : {\List[\Fin[n]]} \to {\List[\Fin[n]]}$ that produces
+a normal form for terms of $\List[\Fin[n]]$. Equivalence classes on $\List[\Fin[n]]$ are precisely those terms that have
+the same normal form. We state a few important properties that $\normf$ satisfies.
 
-20. Now, to prove the equivalence of 9, on one side we have Aut(Fin n), and
-on the other, equivalence classes of List (Fin n) wrt to the usual Coxeter
-group, we would define two functions from and back. The easiest way to
-define a function out of the group presentation is to define them on the
-equivalence classes' representatives. But to do that, we have to know how
-do these representatives look like exactly.
+\begin{proposition}
+  \leavevmode
+  \begin{enumerate}
+    \item For all $l : \List[\Fin[n]]$, we have that $l \cox* \normf(l)$.
+    \item $\normf$ is idempotent, that is, $\normf \comp \normf \htpy \normf$.
+    \item $\normf$ splits, that is, we have ${\List[\Fin[n]]} \xrightarrow{s} \Sn[n] \xrightarrow{r} {\List[\Fin[n]]}$
+          such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
+  \end{enumerate}
+\end{proposition}
 
-The type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.
+\vc{Remark that $w \cox* v$ is not a proposition.}
 
-\begin{definition}[$\Sn$]
-  \(\Sn \defeq \quot{\List(\Fin[n])}{\cox*}\)
-\end{definition}
-
-We will now prove the group structure of $\Sn$.
+Finally, we prove the group structure of $\Sn$, and show that it is indeed a group presentation, given by a quotient of
+the free group $F(\Fin[n])$.
 
@@
   $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to
-  $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $[\term{id},\term{id}] : A + A \to A$.
+  $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $\nabla_{A} : A + A \to A$.
 \end{proposition}
 
-\subsection{Lehmer Codes}
+\subsection{Lehmer Codes}~\label{subsec:lehmer}
 
-21. Enter: Lehmer codes.
-TODO diagram with List (Fin n) being divided into equivalence classes, and
-Lehmer codes being an image of immersion, being the normal form
-(representative) in each class.
+To prove the equivalence between $\Aut[\Fin[n]]$ and $\Sn[n]$, we will need to define functions back and forth between
+the two types. The terms in $\Sn$ can be identified with equivalence classes of terms in $\List[\Fin[n]]$ with respect
+to the Coxeter relation $\cox*$. The easiest way to define a function out of this presentation is to define it on the
+representatives. We know that these are the unique normal forms in the set-quotient given by $q \comp \normf$, but now
+we will describe what these representatives look like exactly, using Lehmer codes.
 
-\subsection{Permutations}~\label{subsec:permutations}
-
-In the previous~\cref{sec:univalent}, we established that paths in $\UFin$ are
-equivalent to families of automorphisms of $\Fin{n}$ for every $n:\Nat$, that
-is, bijections on finite sets of size $n$. This is the extensional view of
-permutations. In the following sections, we will characterise these
-permutations, going through two intermediate steps.
-
-\vc{This is obvious, maybe add something more here.}
-
-\subsection{Lehmer codes}~\label{subsec:lehmer}
+\todo{diagram with List (Fin n) being divided into equivalence classes, and Lehmer codes being an image of immersion,
+  being the normal form (representative) in each class.}
 

### 2021-06-21 Pi+/paper/finite.tex
Text about prop-valued relation

@@
 
-\vc{Remark that $w \cox* v$ is not a proposition.}
+Note that this relation $\cox*$ is not $\hProp$-valued, that is, $w \cox* v$ is not a proposition because reductions are
+not unique. Because of this, the quotient $\Sn$ is not effective, that is, $\quotrel : w \cox* v \to q(w) \id q(v)$ is
+not an equivalence. We could truncate the relation to change this, but then the codomain of $\normf$ also needs to be
+truncated. Instead, using the $\normf$ function, we could define a new relation
+$(w \approx v) \defeq (\normf(w) \id \normf(v))$ which is $\hProp$-valued, and quotient $\List[\Fin[n]]$ by $\approx$.
+This is not necessary because our results hold for $\cox*$ using the properties of the rewriting system we proved, and
+we do not rely on the effectivity of the quotient.
 

### 2021-06-21 Pi+/paper/finite.tex
Revising Lehmer section

@@
   \begin{align*}
-    n \downf 0       & \defeq \nil                       \\
+    n \downf \zero   & \defeq \nil                       \\
     n \downf \suc[k] & \defeq (k + n) \cons (n \downf k)
@@
 to the Coxeter relation $\cox*$. The easiest way to define a function out of this presentation is to define it on the
-representatives. We know that these are the unique normal forms in the set-quotient given by $q \comp \normf$, but now
-we will describe what these representatives look like exactly, using Lehmer codes.
+representatives. We know that these are the unique normal forms in the set-quotient given by $\quotinc \comp \normf$,
+but now we will describe what these representatives exactly look like, using an encoding called Lehmer
+codes~\cite{lehmerTeachingCombinatorialTricks1960}.
 
@@
 
-From grade school combinatorics, we know that there are $\fac{n}$ permutations
-on a finite set with $n$ elements. The factorial function is defined by
-recursion on natural numbers. However, now, for every $n$, we want to produce a
-type, which is a finite set, with cardinality $\fac{n}$. And, to characterise
-$\Aut[\Fin[n]]$, we further need to construct a bijection between this type and
-$\Aut[\Fin[n]]$.
+From grade school combinatorics, we know that there are $\fac{n}$ permutations on a finite set with $n$ elements. For
+every $n$, we thus want to produce a finite set with cardinality $\fac{n}$, and a bijection with $\Aut[\Fin[n]]$. There
+are many ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer codes are known in
+Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960} where they are sometimes called "subexcedant sequences",
+or "factoriadics", which give the factorial number system. They are a particularly convenient way of representing
+permutations on a computer - since they have exactly the right cardinality, they are bitwise-optimal.
 
-First, let's define this type with $\fac{n}$ elements, we name this type family
-$\Lehmer : \Nat \to \UU$, which is defined by recursion on $\Nat$ as follows.
-This is the obvious definition of factorials by recursion, but categorified from
-natural numbers to sets.
+Formally, $\Lehmer[n]$ is an $n+1$-element tuple, where the position $k \leq n$ stores an element of $\Fin[k]$. Since
+the 0-th position is trivial, in practice, it is ignored, and the sequence starts at
+1~\cite{duboisTestsProofsCustom2018,vajnovszkiNewEulerMahonian2011}.\todo{There are more examples - listed
+  in~\cite{duboisTestsProofsCustom2018}}. We can define $\Lehmer$ in two equivalent ways, by a simple recursion on
+$\Nat$, or as a type family generated by two constructors.
 
-\begin{definition}
-  \begin{align*}
-    \Lehmer[0]       & \defeq \unit                           \\
-    \Lehmer[\suc[n]] & \defeq \Fin[\suc[n]] \times \Lehmer[n]
-  \end{align*}
+\begin{definition}[$\Lehmer : \Nat \to \UU$]
+  \begin{gather*}
+    \begin{aligned}
+      \Lehmer[\zero]   & \defeq \Fin[\suc[\zero]]                     \\
+      \Lehmer[\suc[n]] & \defeq \Fin[\suc[\suc[n]]] \times \Lehmer[n]
+    \end{aligned}
+    \qquad
+    \begin{aligned}
+      \lzero & : \Lehmer[\zero]                                                     \\
+      \lsuc  & : \forall n, r, (r \leq \suc[n]) \to \Lehmer[n] \to \Lehmer[\suc[n]]
+    \end{aligned}
+  \end{gather*}
 \end{definition}
 
-\todo{Subexcedant sequences and factorial definitions are equivalent, explain
-  this!}
+\todo{Explain why the two definitions are equivalent?}
+\note{immersion here?}
 
-The name Lehmer comes from Lehmer
-codes~\cite{lehmerTeachingCombinatorialTricks1960} which are known in
-Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960}. There are many
-ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer
-codes are a particularly convenient way to represent permutations on a
-computer,~\review{they are compact and have exactly the right cardinality.
-  $\Lehmer[n]$ is a $n+1$-element tuple, where the position $k \leq n$ has an
-  element of $\Fin[k]$. The 0-th position is trivial, so we ignore it, and in
-  both the example below and the Agda proof, consider only the remaining
-  $n$-element tuple.}
-
-\vc{This is just the classical algorithm to explain the example, not the actual
-  type-theoretic proof.}
-
-Suppose we have a permutation $p$ on an $n$-element set
-$\{\el{0}, \el{1}, \el{2}, \el{3}, \el{4}\}$, we encode it as follows.
-$\Lehmer[n]$ is a $n$-element tuple. At position $k$, we put the number of
-inversions of the element $\el{k}$ in $p$, i.e. the number of elements smaller
-than $\el{k}$ occurring after $\el{k}$.
-
-As an example, consider the following tabulated presentation of the permutation:
+Suppose we have a permutation $\sigma$ on an $n$-element set $\{\el{0}, \el{1}, \el{2}, \el{3}, \el{4}\}$, we encode it
+as follows. $\Lehmer[n]$ is a $n$-element tuple. At position $k$, we put the number of inversions of the element
+$\el{k}$ in $p$, i.e. the number of elements smaller than $\el{k}$ occurring after $\el{k}$. As an example, consider the
+following tabulated presentation of the permutation:
 
@@
 \[
-  p =
+  \sigma =
   \begin{array}{ccccccccccccccc}
@@
 
-The element $\el{0}$ has 0 inversions, because there are no elements smaller
-than $\el{0}$ occurring after it. In fact, there can be no elements smaller than
-$\el{0}$ at all, so the type at the first position of the Lehmer code tuple is
+The element $\el{0}$ has 0 inversions, because there are no elements smaller than $\el{0}$ occurring after it. In fact,
+there can be no elements smaller than $\el{0}$ at all, so the type at the first position of the Lehmer code tuple is
 $\unit$.
 
-The element $\el{1}$ has 0 inversions as well, since elements occurring after it
-in the permutation are $\el{4}$ and $\el{2}$. There is only one different case,
-if $\el{1}$ appeared before $\el{0}$, it would have 1 inversion. This is why the
-type of the second component of the Lehmer code is $\Fin[2]$.
+The element $\el{1}$ has 0 inversions as well, since elements occurring after it in the permutation are $\el{4}$ and
+$\el{2}$. There is only one different case, if $\el{1}$ appeared before $\el{0}$, it would have 1 inversion. This is why
+the type of the second component of the Lehmer code is $\Fin[2]$.
 
-The element $\el{2}$ has 2 inversions, because both $\el{0}$ and $\el{1}$ occur
-after it in the permutation. The element $\el{3}$ occurs as the last one, so it
-has 0 inversions. The element $\el{4}$ has 1 inversion, with the element
+The element $\el{2}$ has 2 inversions, because both $\el{0}$ and $\el{1}$ occur after it in the permutation. The element
+$\el{3}$ occurs as the last one, so it has 0 inversions. The element $\el{4}$ has 1 inversion, with the element
 $\el{3}$.
 
-Thus, the Lehmer code for the permutation $p$ is the 5-tuple
-$l = (0, 0, 2, 0, 1)$.
+Thus, the Lehmer code for the permutation $\sigma$ is the 5-tuple $l = (0, 0, 2, 0, 1)$.
 
-To reconstruct the tabulated presentation of the permutation from the Lehmer
-code, we perform an algorithm similar to \emph{insertion sort}. Starting from
-the left-most position of the tuple $l$, we'll read the value $v$, insert the
-new element at the end of the newly created list, and shift it backward $v$
-places.
+To reconstruct the tabulated presentation of the permutation from the Lehmer code, we perform an algorithm similar to
+\emph{insertion sort}. Starting from the left-most position of the tuple $l$, we'll read the value $v$, insert the new
+element at the end of the newly created list, and shift it backward $v$ places.
 
@@
 
-\begin{proposition}
-  For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$.
-\end{proposition}
+We state and prove a few auxiliary lemmas about how $\FinExcept{n}$ interacts with $\Fin$. \vc{These can be understood
+  using simple counting arguments using the pigeonhole principle, but we have to construct the equivalences.}
 
 \begin{proposition}
-  For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$.
-\end{proposition}
-
-\begin{proposition}
-  For any $n : \Nat$,
+  \leavevmode
+  \begin{enumerate}
+    \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$.
+    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$.
+    \item For any $n : \Nat$,
           \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}{k}} \).
+  \end{enumerate}
 \end{proposition}
@@
 \begin{proof}
-  For $n = 0$, note that $\Lehmer[0]$ is contractible, and so is $\Aut[\Fin[0]]$. For $n = \suc[m]$, we have the
-  following chain of equivalences.
-  \[\arraycolsep=0.5em\def\arraystretch{1.5}
+  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\zero]]$. For $n = \suc[m]$, we
+  compute a chain of equivalences.
+  \begin{gather*}
+    \arraycolsep=0.5em\def\arraystretch{1.5}
+    \begin{array}{rl}
+           & \Lehmer[\zero]    \\
+      \eqv & \unit             \\
+      \eqv & \Aut[\Fin[\zero]] \\
+    \end{array}
+    \qquad\qquad
     \begin{array}{rl}
@@
       \eqv & \Fin[\suc[m]] \times \Lehmer[m]                                                   \\
+      \eqv & \Lehmer[\suc[m]]
     \end{array}
-  \]
+  \end{gather*}
 \end{proof}

### 2021-06-23 Pi+/paper/finite.tex

@@
 
-\subsection{Permutation groups}
+\subsection{Presentation of permutation groups}
 
@@
     \longbraid
-     & : \forall n, l, r \to (l \append (n \downarrow 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downarrow 2 + k) \append r) \\
+     & : \forall n, l, r \to (l \append (n \downf 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downf 2 + k) \append r) \\
   \end{align*}
@@
 
-\begin{definition}[$\downf : (n : \Nat) \to {\Fin[n]} \to {\List[\Fin[\suc[k + n]]]}$]
+\begin{definition}[$\downf : (n : \Nat) \to (k : \Nat) \to {\List[\Fin[\suc[k + n]]]}$]
   \begin{align*}
@@
           such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
+    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \).
   \end{enumerate}
@@
 \todo{Explain why the two definitions are equivalent?}
-\note{immersion here?}
 
 Suppose we have a permutation $\sigma$ on an $n$-element set $\{\el{0}, \el{1}, \el{2}, \el{3}, \el{4}\}$, we encode it
-as follows. $\Lehmer[n]$ is a $n$-element tuple. At position $k$, we put the number of inversions of the element
+as follows. $\Lehmer[n]$ is a $n$-element tuple. Then, At position $k$, we put the number of inversions of the element
 $\el{k}$ in $p$, i.e. the number of elements smaller than $\el{k}$ occurring after $\el{k}$. As an example, consider the
@@
 
-Using this Lehmer encoding algorithm, we can now construct the equivalence between these types.
+Writing formally, to turn a code into a permutation written in the Coxeter presentation, we use $\immersion$:
 
-We define a type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except the one
-provided. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence an $\hSet$.
+\begin{definition}[$\immersion : (n : \Nat) \to {\Lehmer[n]} \to {\List[\Fin[n]]}$]
+  \begin{align*}
+    \immersion(\zero, \zero)    & \defeq \nil                                              \\
+    \immersion(\suc[n], (r, l)) & \defeq \immersion(n, l) \append ((\suc[n] - r) \downf r)
+  \end{align*}
+\end{definition}
+
+As described above, the number $r$ at the position $k$ in the tuple describes how many inversions the element $\el{k}$
+has. Thus, we need to perform $r$ many adjacent transpositions to get to the desired position. The function that
+constructs a sequence of $r$ adjacent transpositions $[\suc[n], \suc[n] - 1, \suc[n] - 2, \ldots \suc[n] - r]$ is
+$(\suc[n] - r) \downf r$.
+
+We can show that the function $\immersion$ gives an equivalence betweeen $\Lehmer[n]$ and $\im{\normf}$.
+
+\begin{proposition}
+  \leavevmode
+  \begin{enumerate}
+    \item For any Lehmer code $c$, $\immersion(c)$ is a normal form with resepct to $\cox*$, that is, $\immersion(c)$ is
+          in $\im{\normf}$.
+    \item Any element of $\im{\normf}$ can be constructed from a unique Lehmer code by $\immersion$, that is, the fibers
+          of $\immersion: \Lehmer[n] \to {\im{\normf}}$ are contractible.
+  \end{enumerate}
+  Therefore, there is an equivalence between $\Lehmer[n]$ and $\im{\normf}$.
+\end{proposition}
+
+We can also establish an equivalence between $\im{\normf}$ and $\Sn$, which gives the following.
+
+\begin{proposition}
+  For all $n : \Nat$, \( \Sn \eqv \im{\normf} \eqv \Lehmer[n] \).
+\end{proposition}
+
+\subsection{Running Lehmer codes}
+
+Finally, it is time to complete our goal of characterising the permutation groups. Having produced a Lehmer code by
+normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection
+between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
+between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$.
+
+To start, we define a type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except the
+one provided. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence an $\hSet$.
 
@@
 \begin{proposition}
-  For all $n:\Nat$, \( \Lehmer[n] \eqv \Aut[\Fin[n]] \).
+  For all $n:\Nat$, \( \Lehmer[n] \eqv \Aut[\Fin[\suc[n]]] \).
 \end{proposition}
@@
 \begin{proof}
-  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\zero]]$. For $n = \suc[m]$, we
-  compute a chain of equivalences.
+  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\suc[\zero]]]$. For $n = \suc[m]$,
+  we compute a chain of equivalences.
   \begin{gather*}
@@
       \eqv & \unit                   \\
-      \eqv & \Aut[\Fin[\zero]] \\
+      \eqv & \Aut[\Fin[\suc[\zero]]] \\
     \end{array}
@@
     \begin{array}{rl}
-           & \Aut[\Fin[\suc[m]]]                                                               \\
-      \eqv & \dsum{k : \Fin[\suc[m]]}{\FinExcept{\suc[m]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} \\
-      \eqv & \dsum{k : \Fin[\suc[m]]}{\FinExcept{\suc[m]}[\fzero] \eqv \Fin[m]}                \\
-      \eqv & \dsum{k : \Fin[\suc[m]]}{\Fin[m] \eqv \Fin[m]}                                    \\
-      \eqv & \Fin[\suc[m]] \times \Aut[\Fin[m]]                                                \\
-      \eqv & \Fin[\suc[m]] \times \Lehmer[m]                                                   \\
+           & \Aut[\Fin[\suc[\suc[m]]]]                                                                     \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \Fin[\suc[m]]}          \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                              \\
+      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                \\
+      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                         \\
       \eqv & \Lehmer[\suc[m]]
@@
 
-We will now show how to generate a Lehmer code from a word in $\Sn$ and back.
-
-\begin{definition}
-  \[
-    \encode{\List} : \List[\Fin[n]] \to \Lehmer[n]
-  \]
-\end{definition}
-
-\begin{proposition}
-  \leavevmode
-  \begin{enumerate}
-    \item \( l_{1} \cox* l_{2} \to \encode{\List}(l_{1}) \id \encode{\List}(l_{1}) \)
-  \end{enumerate}
-\end{proposition}
-
-\begin{definition}
-  \begin{align*}
-    \encodeSn & : \Sn[n] \to \Lehmer[n] \\
-    \decodeSn & : \Lehmer[n] \to \Sn[n]
-  \end{align*}
-\end{definition}
-
-\begin{proposition}
-  For all $n : \Nat$, $(\encodeSn, \decodeSn)$ is an equivalence.
-\end{proposition}
-
 \begin{corollary}
@@
   \(
-  \Sn \eqv \Lehmer[n] \eqv \Aut[\Fin[n]]
+  \Sn \eqv \Lehmer[n] \eqv \Aut[\Fin[\suc[n]]]
   \).

### 2021-06-23 Pi+/paper/finite.tex
Explain final Lehmer equivalence

@@
 
-To start, we define a type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except the
-one provided. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence an $\hSet$.
+To do so, we need to construct some equivalences by counting the elements of $\Fin[n]$ using its decidable equality.
+First, we define a helper type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except
+the one in the argument. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence an
+$\hSet$.
 
@@
 
-We state and prove a few auxiliary lemmas about how $\FinExcept{n}$ interacts with $\Fin$. \vc{These can be understood
-  using simple counting arguments using the pigeonhole principle, but we have to construct the equivalences.}
+We state and prove a few auxiliary lemmas about how $\FinExcept{n}$ interacts with $\Fin$.
 
@@
     \item For any $n : \Nat$,
-          \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}{k}} \).
+          \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}[k]} \).
   \end{enumerate}
 \end{proposition}
+
+\begin{proof}
+  The first and second propositions follow from simply constructing the bijections using the decidable equality of
+  $\Fin[n]$, and making sure to punch-in and punch-out the element $k$ at the right place.
+
+  The third proposition performs some interesting computations. On the left, we have the type of automorphisms of
+  $\Fin[\suc[n]]$. Assume a particular one $\phi : \Fin[\suc[n]] \xrightarrow{\sim} \Fin[\suc[n]]$. To construct an
+  element of the pair on the right, we pick the first component to be the image of the element $\zero$ under $\phi$.
+  After fixing $k \defeq \phi(\zero)$, we need to compute the bijection in the second component. We remove the two
+  elements $\zero$ and $k$ from $\Fin[\suc[n]]$ and construct an equivalence between the resulting types
+  $\FinExcept{\suc[n]}[\zero]$ and $\FinExcept{\suc[n]}[k]$ as follows:
+  $\FinExcept{\suc[n]}[\zero] \eqv \dsum{k:\Fin[\suc[n]]}{(\phi(\zero) \neq \phi(k))} \eqv \FinExcept{\suc[n]}[k]$
+
+  To go the other way, given a $k : \Fin[\suc[n]]$ and an equivalence
+  $\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}[k]$ we compute an equivalence as follows:
+  $\Fin[\suc[n]] \eqv \unit \sqcup \FinExcept{\suc[n]}[k] \eqv \unit \sqcup \Fin[n] \eqv \Fin[\suc[n]]$.
+\end{proof}
+
+To remind ourselves, in a Lehmer code, the element $\el{r}$ at postition $k$ counts the number of inversions of the
+element $\el{r}$. This corresponds exactly to what happens in the third proposition -- inserting element $\el{r}$ at the
+position $k$ shifts it by $k$ places.
 
@@
     \begin{array}{rl}
-           & \Aut[\Fin[\suc[\suc[m]]]]                                                                     \\
+           & \Lehmer[\suc[m]]                                                                              \\
+      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                         \\
+      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                              \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \Fin[\suc[m]]}          \\
       \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \Fin[\suc[m]]}          \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                              \\
-      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                \\
-      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                         \\
-      \eqv & \Lehmer[\suc[m]]
+      \eqv & \Aut[\Fin[\suc[\suc[m]]]]
     \end{array}

### 2021-06-23 Pi+/paper/reversible1.tex
map

@@
     \mathit{enum}(\onet) &=& \{ \Acon{tt} \} \\
-    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
-    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda w.\mathit{map}~(\lambda v. (v,w))~\mathit{enum}(A))~\mathit{enum}(B))
+    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₂}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₁}~\mathit{enum}(B) \\
+    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
 \end{array}\]
-\noindent For example, the 3 values $\inlv{\Acon{tt}}$, $\inrv{(\inlv{\Acon{tt}})}$, and $\inrv{(\inrv{\Acon{tt}})}$ of
+\noindent For example, the 3 values $\inrv{(\inrv{\Acon{tt}})}$, $\inrv{(\inlv{\Acon{tt}})}$, and $\inlv{\Acon{tt}}$ of
 type $\onet + (\onet + \onet)$ are given indices 0, 1, and 2 respectively. Given this canonical order, the combinator

### 2021-06-23 Pi+/paper/finite.tex
Revise Lehmer subsection

@@
 
-\begin{proposition}
+\begin{proposition}~\label{prop:sn-im-lehmer-equiv}
   For all $n : \Nat$, \( \Sn \eqv \im{\normf} \eqv \Lehmer[n] \).
@@
 
-\begin{proposition}
+\begin{proposition}~\label{prop:fin-finexcept}
   \leavevmode
   \begin{enumerate}
-    \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$.
-    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$.
+    \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$.~\label{prop:fin-finexcept-1}
+    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$.~\label{prop:fin-finexcept-2}
     \item For any $n : \Nat$,
-          \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}[k]} \).
+          \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}[k]} \).~\label{prop:fin-finexcept-3}
   \end{enumerate}
@@
   $\FinExcept{\suc[n]}[\zero]$ and $\FinExcept{\suc[n]}[k]$ as follows:
-  $\FinExcept{\suc[n]}[\zero] \eqv \dsum{k:\Fin[\suc[n]]}{(\phi(\zero) \neq \phi(k))} \eqv \FinExcept{\suc[n]}[k]$
+  $\FinExcept{\suc[n]}[\zero] \eqv \dsum{k:\Fin[\suc[n]]}{(\phi(\zero) \neq \phi(k))} \eqv \FinExcept{\suc[n]}[k]$.
 
@@
 
-To remind ourselves, in a Lehmer code, the element $\el{r}$ at postition $k$ counts the number of inversions of the
-element $\el{r}$. This corresponds exactly to what happens in the third proposition -- inserting element $\el{r}$ at the
-position $k$ shifts it by $k$ places.
+Using these facts, we can now prove the main result of this section.
 
-\begin{proposition}
+\begin{proposition}~\label{prop:lehmer-aut-equiv}
   For all $n:\Nat$, \( \Lehmer[n] \eqv \Aut[\Fin[\suc[n]]] \).
@@
     \qquad\qquad
-    \begin{array}{rl}
-           & \Lehmer[\suc[m]]                                                                              \\
-      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                         \\
-      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                              \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \Fin[\suc[m]]}          \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} \\
-      \eqv & \Aut[\Fin[\suc[\suc[m]]]]
+    \begin{array}{rlr}
+           & \Lehmer[\suc[m]]                                                                              &                                                                 \\
+      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                         & \text{by definition}                                            \\
+      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                & \text{induction hypothesis}                                     \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                              & \text{$\Sigma$ over a constant family}                          \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \Fin[\suc[m]]}          & \text{by~\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-2}} \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} & \text{by~\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-2}} \\
+      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                     & \text{by~\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-3}} \\
     \end{array}
@@
 \end{proof}
+
+\jk{TODO an example here}
+To remind ourselves, in a Lehmer code, the element $\el{r}$ at postition $k$ counts the number of inversions of the
+element $\el{r}$. This corresponds exactly to what happens in the \cref{prop:fin-finexcept-3}. Looking at the operation
+combinatorially, it is inserting element $\el{r}$ at the position $k$, meaning that it shifts the element
+$k$places.\jk{needs rewriting after the example is here}
+
+By composing~\cref{prop:lehmer-aut-equiv} and~\cref{prop:sn-im-lehmer-equiv}, we obtain the final
+equivalence.
 

### 2021-06-25 Pi+/paper/equivalence.tex
Move syntax
Outline for section 7

--- 2021-04-19 Pi+/paper/equivalence.tex
@@
 
-First, we normalise $\PiPlusLang$ to $\PiHatLang$, and show that they are equivalent.
+In this section, we use the semantics developed in the previous sections define a fragment $\PiPlusLang$ of $\PiLang$
+and a normalised fragment of it $\PiHatLang$, interpret to $\UFin$ and back, and translate between the languages.
 
-Then we establish the completeness of $\PiHatLang$ with respect to $\UFin$.
+We only present the types and 1-combinators in the language, and the full set of 2-combinators is listed in the appendix
+and the accompanying Agda code.
 
-We use this to establish completeness for $\PiPlusLang$.
+\vc{We can translate types to types, terms to terms, theorem: preserves types, preserves equations.}
 
-Finally, we show how to translate $\PiLang$ programs to $\PiPlusLang$ programs.
+\vc{Define maps quote/eval on 0,1,2-cells, and show commuting diagrams.}
+
+\subsection{$\PiHatLang$}
+
+We first present a language $\PiHatLang$ based on the theory of symmetric groups developed in~\cref{sec:finite}. The
+types in the language are unary natural numbers, and combinators are allowed to perform adjacent transpositions.
+
+\begin{figure}[t]
+  {\scalebox{\scalef}{$
+        \begin{array}{rrcll}
+          \idc :   & n             & \isoh & n             & : \idc   \\
+          \swapc : & \suc[\suc[n]] & \isoh & \suc[\suc[n]] & : \swapc \\
+        \end{array}
+      $}}
+
+  {\scalebox{\scalef}{
+      \Rule{}
+      {\jdg{}{}{c_1 : n \isoh m} \quad \vdash c_2 : m \isoh o}
+      {\jdg{}{}{c_1 \fatsemi c_2 : n \isoh o}}
+      {}
+
+      \Rule{}
+      {\jdg{}{}{c : n \isoh m}}
+      {\jdg{}{}{\oplus(c) : \suc[n] \isoh \suc[m]}}
+      {}
+    }}
+  \caption{$\PiHatLang$ syntax}
+  \label{fig:pihat}
+\end{figure}
+
+\begin{proposition}
+  We can form a weak 2-category $\PiHatCat$ with
+  \begin{itemize}
+    \item natural numbers for 0-cells,
+    \item for $n, m : \Nat$, a collection of 1-cells $n \isoh m$,
+    \item for $p, q : n \isoh m$, a collection of 2-cells $p \Isoh q$.
+  \end{itemize}
+\end{proposition}
+
+\begin{proposition}
+  There is a symmetric monoidal structure on $\PiHatCat$, with $0$ for the unit, and addition for the tensor.
+\end{proposition}
+
+Then we establish the completeness of $\PiHatLang$ with respect to $\UFin$. We define $\evalt$ and $\quotet$ for 0, 1,
+and 2-cells.
+
+\begin{definition}
+  \begin{align*}
+    \evalt_{0} & : \UHat \to \UFin                                                          \\
+    \evalt_{1} & : (c : t_{1} \isoh t_{2}) \to \evalh_{0}(t_{1}) \id \evalh_{0}(t_{2})      \\
+    \evalt_{2} & : (\alpha : c_{1} \Isoh c_{2}) \to \evalt_{1}(c_{1}) \id \evalt_{1}(c_{2}) \\
+  \end{align*}
+\end{definition}
+
+\begin{definition}
+  \begin{align*}
+    \quotet_{0} & : \UFin \to \UHat                                                            \\
+    \quotet_{1} & : (p : X_{1} \id X_{2}) \to \quoteh_{0}(X_{1}) \isoh \quoteh_{0}(X_{2})      \\
+    \quotet_{2} & : (\alpha : p_{1} \id p_{2}) \to \quoteh_{1}(p_{1}) \Isoh \quotet_{1}(p_{2}) \\
+  \end{align*}
+\end{definition}
+
+\begin{proposition}
+  $\evalt/\quotet$ give a symmetric monoidal biequivalence between $\PiHatCat$ and $\UFin$.
+\end{proposition}
+
+\subsection{$\PiPlusLang$}
+
+Now, we present the additive fragment of $\PiLang$, called $\PiPlusLang$.
+
+\begin{figure}[t]
+  {\scalebox{\scalef}{$
+        \begin{array}{rrcll}
+          \idc :     & A           & \iso & A           & : \idc     \\
+          \identlp : & \zerot + A  & \iso & A           & : \identrp \\
+          \swapp :   & A + B       & \iso & B + A       & : \swapp   \\
+          \assoclp : & A + (B + C) & \iso & (A + B) + C & : \assocrp \\ [1.5ex]
+        \end{array}$}}
+
+  {\scalebox{\scalef}{
+      \Rule{}
+      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : B \iso C}
+      {\jdg{}{}{c_1 \fatsemi c_2 : A \iso C}}
+      {}
+
+      \Rule{}
+      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : C \iso D}
+      {\jdg{}{}{c_1 \oplus c_2 : A + C \iso B + D}}
+      {}
+    }}
+  \caption{$\PiPlusLang$ syntax}
+  \label{fig:piplus}
+\end{figure}
+
+\begin{proposition}
+  We can form a weak 2-category $\PiPlusCat$ with
+  \begin{itemize}
+    \item $\PiPlusLang$ types for 0-cells,
+    \item for $X, Y : \UPlus$, a collection of 1-cells $X \iso Y$,
+    \item for $p, q : X \isoh Y$, a collection of 2-cells $p \Iso q$.
+  \end{itemize}
+\end{proposition}
+
+\begin{proposition}
+  There is a symmetric monoidal structure on $\PiPlusCat$, with $\zerot$ for the unit, and $+$ for the tensor.
+\end{proposition}
+
+We normalise $\PiPlusLang$ to $\PiHatLang$ and back, by giving $\evalh/\quoteh$ maps for 0, 1, and 2-cells.
+
+\begin{definition}
+  \begin{align*}
+    \evalh_{0} & : \UPlus \to \UHat                                                          \\
+    \evalh_{1} & : (c : t_{1} \iso t_{2}) \to \evalh_{0}(t_{1}) \isoh \evalh_{0}(t_{2})      \\
+    \evalh_{2} & : (\alpha : c_{1} \Iso c_{2}) \to \evalh_{1}(c_{1}) \Isoh \evalh_{1}(c_{2}) \\
+  \end{align*}
+\end{definition}
+
+\begin{definition}
+  \begin{align*}
+    \quoteh_{0} & : \UHat \to \UPlus                                                            \\
+    \quoteh_{1} & : (p : X_{1} \isoh X_{2}) \to \quoteh_{0}(X_{1}) \iso \quoteh_{0}(X_{2})      \\
+    \quoteh_{2} & : (\alpha : p_{1} \Isoh p_{2}) \to \quoteh_{1}(p_{1}) \Iso \quoteh_{1}(p_{2}) \\
+  \end{align*}
+\end{definition}
+
+\begin{proposition}
+  $\evalh/\quoteh$ give a symmetric monoidal biequivalence between $\PiPlusCat$ and $\PiHatCat$.
+\end{proposition}
+
+\subsection{$\PiLang$}
+
+Finally, We show how to translate $\PiLang$ programs to $\PiHatLang$ programs.
+
+\begin{definition}
+  \begin{align*}
+    \evalt_{0} & : U \to \Nat                                               \\
+    \evalt_{1} & : (c : X \iso Y) \to \evalt_{0}(X) \iso \evalt_{0}(Y)      \\
+    \evalt_{2} & : (\alpha : p \Iso q) \to \evalt_{1}(p) \Iso \evalt_{1}(q) \\
+  \end{align*}
+\end{definition}
+
+To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$ and quote it to $\PiPlusLang$.
+
+\begin{definition}
+  \begin{align*}
+    \normt_{0} & : U \to \UPlus                                        \\
+    \normt_{0} & = \quoteh_{0} \comp \evalt_{0}                        \\
+    \\
+    \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \iso \normt_{0}(Y) \\
+    \normt_{1} & = \quoteh_{1} \comp \evalt_{1}                        \\
+  \end{align*}
+\end{definition}
 

### 2021-06-25 Pi+/paper/finite.tex
Lehmer Aut equiv example

@@
 
-When swapping two pairs of elements in positions "far-away", it does not matter
-in which order the swapping happens, that is, we can slide the wires freely.
+When swapping two distinct pairs of elements (i.e. the indices of transpositions
+differ by at least 1), it does not matter in which order the swapping happens,
+that is, we can slide the wires freely.
 
@@
 
-The long form of the Coxeter relations are unwieldy and difficult to prove properties about by induction, and have
+The modified form of the Coxeter relations are unwieldy and difficult to prove properties about by induction, and have
 little connection to the broader scope of computational group theory. However, we can prove that the closures of the two
@@
 
-\jk{TODO an example here}
-To remind ourselves, in a Lehmer code, the element $\el{r}$ at postition $k$ counts the number of inversions of the
-element $\el{r}$. This corresponds exactly to what happens in the \cref{prop:fin-finexcept-3}. Looking at the operation
-combinatorially, it is inserting element $\el{r}$ at the position $k$, meaning that it shifts the element
-$k$places.\jk{needs rewriting after the example is here}
+To see how the proof works combinatorically, let us go through an example. We will start from a following bijection $p: \Aut[\Fin[4]]$:
+\begin{align*}
+    p(0) \mapsto & 1 \\
+    p(1) \mapsto & 3 \\
+    p(2) \mapsto & 0 \\
+    p(3) \mapsto & 2
+\end{align*}
+
+At first, we start from a full domain $\Fin[4]$ and full codomain $\Fin[4]$.
+\[
+    \begin{bmatrix}
+        0 \\
+        1 \\
+        2 \\
+        3
+    \end{bmatrix}
+    \mapsto
+    \begin{bmatrix}
+        0 \\
+        1 \\
+        2 \\
+        3
+    \end{bmatrix}
+\]
+
+As element $\el{0}$ is mapped to $\el{1}$, we remember that the last element of
+result tuple is $1$. The appropriate elements on both sides are removed,
+forming, respectively, $\FinExcept{4}[0]$ and $\FinExcept{4}[1]$. These are then
+both transformed to $\Fin[3]$ using
+\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-2}, as shown below.
+\[
+    \begin{bmatrix}
+        0 \\
+        1 \\
+        2
+    \end{bmatrix}
+    \equiv
+    \begin{bmatrix}
+        1 \\
+        2 \\
+        3
+    \end{bmatrix}
+    \mapsto
+    \begin{bmatrix}
+        0 \\
+        2 \\
+        3
+    \end{bmatrix}
+    \equiv
+    \begin{bmatrix}
+        0 \\
+        1 \\
+        2
+    \end{bmatrix}
+\]
+
+Then, the original permutation specified that the element $\el{1}$ has to be
+mapped to $\el{3}$. Since we performed renumbering, now it means that element is
+mapped to $\el{2}$ - the resulting tuple now looks like $[2, 1]$. We yet again remove these from the domain and codomain,
+forming $\FinExcept{0}[0]$ and $\FinExcept{0}[2]$.
+
+\[
+    \begin{bmatrix}
+        0 \\
+        1
+    \end{bmatrix}
+    \equiv
+    \begin{bmatrix}
+        1 \\
+        2
+    \end{bmatrix}
+    \mapsto
+    \begin{bmatrix}
+        0 \\
+        2
+    \end{bmatrix}
+    \equiv
+    \begin{bmatrix}
+        0 \\
+        1
+    \end{bmatrix}
+\]
+
+The situation repeats two more times: first, by assinging $\el{0}$ to $\el{1}$,
+giving the tuple $[0, 2, 1]$ and transforming the domain and codomain:
+\[
+    \begin{bmatrix}
+        0
+    \end{bmatrix}
+    \equiv
+    \begin{bmatrix}
+        1
+    \end{bmatrix}
+    \mapsto
+    \begin{bmatrix}
+        0
+    \end{bmatrix}
+    \equiv
+    \begin{bmatrix}
+        0
+    \end{bmatrix}
+\]
+
+And finally, by assinging $\el{0}$ to $\el{0}$, forming a tuple $[0, 0, 2, 1]$ and
+
+At this point we see that the renumbering of the domain is done in an orderly
+way (always shifting by 1), so it can be ignored. On the other hand, the
+ordering of the codomain is what matters. Each time we remove an element, all
+larger elemens are in effect decreased by one.
+
+To remind ourselves, when we introduced the combinatorial interpretation of
+Lehmer codes, the element $\el{r}$ at postition $k$ was said to count the number
+of inversions of the element $\el{r}$. Thus, this is exactly the meachanism in
+which the induction step works.
+
+This corresponds exactly to what happens in
+the \cref{prop:fin-finexcept-3}. Looking at the operation combinatorially, it is
+inserting element $\el{r}$ at the position $k$, meaning that it shifts the
+element $k$places.\jk{needs rewriting after the example is here}
+
+Viewing the codomain as a vector of four elements, element $\el{0}$ has zero
+inverses, element $\el{1}$ has 1 inverse, element $\el{2}$ has 1 inverse, and
+element $3$ has
 

### 2021-06-27 Pi+/paper/finite.tex
Proofs of crit pairs
Progress
Update finite.tex

@@
 
-When swapping two distinct pairs of elements (i.e. the indices of transpositions
+When swapping two distinct pairs of elements (i.e. when the indices of two transpositions
 differ by at least 1), it does not matter in which order the swapping happens,
@@
 
-Unfortunately, in the system defined, this is not true for all critical pairs. One example is the following.
+\begin{proof}
+\begin{array}{lcr}
+% https://q.uiver.app/?q=WzAsNixbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzQsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwyLCJcXHRhdV8xXFx0YXVfMlxcdGF1XzIiXSxbMiwzLCJcXHRhdV8xIl0sWzQsMiwiXFx0YXVfMlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFswLDIsImJyYWlkIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJjYW5jZWwiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFszLDQsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzIsNSwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbNSw0LCJjYW5jZWwiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dXQ==
+\[\begin{tikzcd}
+	&& {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
+	{\tau_1\tau_2\tau_1\tau_1\tau_2} &&&& {\tau_2\tau_1\tau_1\tau_2\tau_1} \\
+	{\tau_1\tau_2\tau_2} &&&& {\tau_2\tau_2\tau_1} \\
+	&& {\tau_1}
+	\arrow["braid"', squiggly, from=1-3, to=2-1]
+	\arrow["braid", squiggly, from=1-3, to=2-5]
+	\arrow["cancel"', squiggly, from=2-1, to=3-1]
+	\arrow["cancel"', squiggly, from=3-1, to=4-3]
+	\arrow["cancel", squiggly, from=2-5, to=3-5]
+	\arrow["cancel", squiggly, from=3-5, to=4-3]
+\end{tikzcd}\]
+&
+\text{and}
+&
+% https://q.uiver.app/?q=WzAsNSxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8yIl0sWzAsMSwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFsyLDQsIlxcdGF1XzJcXHRhdV8xIl0sWzAsMiwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFsxLDMsIlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwxLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzAsMiwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzMsNCwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFs0LDIsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV1d
+\[\begin{tikzcd}
+	&& {\tau_2\tau_1\tau_2\tau_2} \\
+	{\tau_1\tau_2\tau_1\tau_2} \\
+	{\tau_1\tau_2\tau_1\tau_2} \\
+	& {\tau_1\tau_1\tau_2\tau_1} \\
+	&& {\tau_2\tau_1}
+	\arrow["braid"', squiggly, from=1-3, to=2-1]
+	\arrow["cancel", squiggly, from=1-3, to=5-3]
+	\arrow["braid"', squiggly, from=2-1, to=3-1]
+	\arrow["braid"', squiggly, from=3-1, to=4-2]
+	\arrow["cancel"', squiggly, from=4-2, to=5-3]
+\end{tikzcd}\]
+\end{array}
+\end{proof}
+
+Unfortunately, in the system defined, this is not true for all critical pairs.
+One example is the following, where both endpoints are normal with respect to
+$\cox*$ relation.
 
@@
 
-As element $\el{0}$ is mapped to $\el{1}$, we remember that the last element of
-result tuple is $1$. The appropriate elements on both sides are removed,
+We remember that the last element of result tuple is $k - 1 = 2$. The appropriate elements on both sides are removed,
 forming, respectively, $\FinExcept{4}[0]$ and $\FinExcept{4}[1]$. These are then
@@
 \cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-2}, as shown below.
+
 \[

### 2021-06-27 Pi+/paper/reversible1.tex
reset+ makes sense now
working on sec 3

@@
 
-\noindent A denotational semantics of $\Pi$ can be specified by mapping each type to a finite set and each combinator to
-a bijection between finite sets, that is, a permutation. In this section, we outline a simple semantics expressed in the
-metalanguage of set theory. This semantics is formalized and extended to the metalanguage of homotopy type theory in
-Sec.~\ref{sec:reversibletwo}.
+\noindent The accompanying code includes a simple operational semantics for $\Pi$. A denotational semantics can be
+specified by mapping each type to a finite set and each combinator to a bijection between finite sets, that is, a
+permutation. In this section, we outline a simple semantics expressed in the metalanguage of set theory. This semantics
+is formalized and extended to the metalanguage of homotopy type theory in Sec.~\ref{sec:reversibletwo}.
 
@@
 \[\begin{array}{rcl}
-    \mathit{enum}(\zerot) &=& \emptyset \\
-    \mathit{enum}(\onet) &=& \{ \Acon{tt} \} \\
-    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₂}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₁}~\mathit{enum}(B) \\
+    \mathit{enum}(\zerot) &=& [~] \\
+    \mathit{enum}(\onet) &=& [ ~\Acon{tt}~ ] \\
+    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
     \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
 \end{array}\]
-\noindent For example, the 3 values $\inrv{(\inrv{\Acon{tt}})}$, $\inrv{(\inlv{\Acon{tt}})}$, and $\inlv{\Acon{tt}}$ of
-type $\onet + (\onet + \onet)$ are given indices 0, 1, and 2 respectively. Given this canonical order, the combinator
-$\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$ would correspond to the permutation $(1~2~0)$ while the
-combinator $\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$ at the same type would correspond to the
-identity permutation.
+\noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
+a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
+operation that appends all the subsequences in a sequence. For example, the 3 values $\inlv{\Acon{tt}}$,
+$\inrv{(\inlv{\Acon{tt}})}$, and $\inrv{(\inrv{\Acon{tt}})}$ of type $\onet + (\onet + \onet)$ are given indices 0, 1, and 2
+respectively. Given this canonical order, the combinator $\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$
+would correspond to the permutation $(1~2~0)$ while the combinator $\assoclp : \onet + (\onet + \onet) \iso (\onet +
+\onet) + \onet$ at the same type would correspond to the identity permutation.
+
+% \[\begin{array}{l}
+%     \Tree[ {\small a} [ {\small b} {\small c} ] ] \\
+%     2 \quad 1 \quad 0 \\
+%     0 \quad 2 \quad 1 \\
+%     \Tree[ [ {\small b}  {\small c} ] {\small a} ]
+% \end{array}\]
 
@@
 
-\note{Revisit to pick examples relevant to the rest of the paper; also unify notation}
+%% \note{Revisit to pick examples relevant to the rest of the paper; also unify notation}
 

### 2021-06-27 Pi+/paper/reversible1.tex
done with sec 3 for now

@@
 
-\noindent Intuitively, one may view each type $A$ as a collection of physical wires that can transmit $\sizet{A}$ distinct
-values where $\sizet{A}$ is a natural number that indicates the size of a type, computed as: $\sizet{\zerot} = 0$;
-$\sizet{\onet} = 1$; $\sizet{A + B} = \sizet{A} + \sizet{B}$; and $\sizet{A \times B} = \sizet{A} * \sizet{B}$.  Thus the type
-$\mathbb{B} = \onet + \onet$ corresponds to a wire that can transmit one of two values, i.e., bits, with the
-convention that $\inlv{\Acon{tt}}$ represents \AgdaFunction{𝔽} and $\inrv{\Acon{tt}}$ represents \AgdaFunction{𝕋}.  The
-type $\mathbb{B} \times \mathbb{B} \times \mathbb{B}$ corresponds to a collection of wires that can transmit three
-bits. From that perspective, a type isomorphism between types $A$ and $B$ (such that $\sizet{A}=\sizet{B}=n$) models a
-\emph{reversible} combinational circuit that \emph{permutes} the $n$ different values. These type isomorphisms form a
-\emph{commutative semiring} and are collected in Fig.~\ref{pi-terms}. Each line in the top part of the figure
-introduces a pair of dual constants that witness the type isomorphism in the middle.  These are the \emph{base}
-(non-reducible) terms of $\Pi$. The isomorphisms are extended to form a congruence relation by adding the three
-constructors at the bottom of the figure that witness equivalence and compatible closure. Although austere, this
-combinator-based language has the advantage of being amenable to formal analysis for at least two reasons: (i) it is
-conceptually simple and small, and (ii) it has direct and evident connections to type theory and category theory and
-hence possesses a rich algebraic structure which we exploit in the remainder of the paper.
+\noindent Intuitively, one may view each type $A$ as a collection of physical wires that can transmit $\sizet{A}$
+distinct values where $\sizet{A}$ is a natural number that indicates the size of a type, computed as:
+$\sizet{\zerot} = 0$; $\sizet{\onet} = 1$; $\sizet{A + B} = \sizet{A} + \sizet{B}$; and
+$\sizet{A \times B} = \sizet{A} * \sizet{B}$.  Thus the type $\mathbb{2} = \onet + \onet$ corresponds to a wire that can
+transmit one of two values, i.e., bits. The type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated
+$\mathbb{B}~3$) corresponds to a collection of wires that can transmit three bits. From that perspective, a type
+isomorphism between types $A$ and $B$ (such that $\sizet{A}=\sizet{B}=n$) models a \emph{reversible} combinational
+circuit that \emph{permutes} the $n$ different values. These type isomorphisms form a \emph{commutative semiring} and
+are collected in Fig.~\ref{pi-terms}. Each line in the top part of the figure introduces a pair of dual constants that
+witness the type isomorphism in the middle.  These are the \emph{base} (non-reducible) terms of $\Pi$. The isomorphisms
+are extended to form a congruence relation by adding the three constructors at the bottom of the figure that witness
+equivalence and compatible closure. Although austere, this combinator-based language has the advantage of being amenable
+to formal analysis for at least two reasons: (i) it is conceptually simple and small, and (ii) it has direct and evident
+connections to type theory and category theory and hence possesses a rich algebraic structure which we exploit in the
+remainder of the paper.
 
@@
 
-\paragraph*{Booleans}
-Let us start with encoding booleans. We use the type \ensuremath{1{\sumtype}1} to
-represent booleans with \ensuremath{\mathit{left} ~()} representing \ensuremath{\mathit{true}} and
-\ensuremath{\mathit{right}~()} representing \ensuremath{\mathit{false}}.
-Boolean negation is straightforward to define:
+In order to understand the examples of Sec.~\ref{sec:informal}, we review how to represent collections of booleans and
+conditionals in $\Pi$. We use the type $\onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing
+\textsf{false} and $\inrv{\Acon{tt}}$ representing $\textsf{true}$. Boolean negation is straightforward to define using
+the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values.
+For example, we can express a 3-bit word reversal operation as follows:
 
-\ensuremath{\mathit{not} : \mathit{bool} \leftrightarrow \mathit{bool}}
+$\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
 
-\ensuremath{\mathit{not} = \swapp}
+$\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
 
-\noindent
-It is easy to verify that \ensuremath{\mathit{not}} changes \ensuremath{\mathit{true}} to \ensuremath{\mathit{false}} and
-vice versa.
-
-\paragraph*{Bit Vectors.}
-We can represent $n$-bit words using an n-ary product of
-\ensuremath{\mathit{bool}}s. For example, we can represent a 3-bit word, \ensuremath{\mathit{word}_3},
-using the type \ensuremath{\mathit{bool} {\prodtype} (\mathit{bool} {\prodtype}  \mathit{bool})}.  We can perform various
-operations on these 3-bit words using combinators in \ensuremath{\Pi }. For
-instance the bitwise \ensuremath{\mathit{not}} operation is the parallel composition of
-three \ensuremath{\mathit{not}} operations:
-
-\ensuremath{\mathit{not}_{\mathit{word}_3} :: \mathit{word}_3 \leftrightarrow \mathit{word}_3}
-
-\ensuremath{\mathit{not}_{\mathit{word}_3} = \mathit{not}  {\prodtype}  (\mathit{not}  {\prodtype}  \mathit{not})}
-
-\noindent We can express a 3-bit word reversal operation as follows:
-
-\ensuremath{\mathit{reverse} : \mathit{word}_3 \leftrightarrow \mathit{word}_3}
-
-\ensuremath{\mathit{reverse} = \swapt \odot (\swapt  \otimes  \idc)~ \odot \assocrt}
-
-\noindent We can check that \ensuremath{\mathit{reverse}} does the right thing by
-applying it to a value \ensuremath{(v_1, (v_2, v_3))} and writing out the full
-derivation tree of the reduction.  The combinator \ensuremath{\mathit{reverse}}, like
-many others we will see in this paper, is formed by sequentially
-composing several simpler combinators. Instead of presenting the
-operation of \ensuremath{\mathit{reverse}} as a derivation tree, it is easier (purely
-for presentation reasons) to flatten the tree into a sequence of
-reductions as caused by each component. Such a sequence of reductions
-is given below:
+\noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
 \[\begin{array}{rlr}
@@
 
-\noindent On the first line is the initial value. On each subsequent
-line is a fragment of the \ensuremath{\mathit{reverse}} combinator and the value that
-results from applying this combinator to the value on the previous
-line. For example, \ensuremath{\swapt} transforms \ensuremath{(v_1, (v_2, v_3))} to
-\ensuremath{((v_2,v_3),v_1)}.  On the last line we see the expected result with
-the bits in reverse order.
+Even though \ensuremath{\Pi } lacks conditional expressions, they are expressible using the distributivity and factoring
+laws as shown below:
 
-We can also draw out the graphical representation of the 3-bit reverse
-combinator. In the graphical representation, it is clear that the
-combinator achieves the required shuffling.
+$\Afun{ifc}_{c_1,c_2} : \mathbb{2} \times A \iso \mathbb{2} \times B$
 
-% \inkscape{reverse-3-bit.pdf}
-
-\paragraph*{Conditionals.}
-Even though \ensuremath{\Pi } lacks conditional expressions, they are
-expressible using the distributivity and factoring laws. The
-diagrammatic representation of \ensuremath{\dist} shows that it redirects the flow
-of a value \ensuremath{v:b} based on the value of another one of type
-\ensuremath{b_1{\sumtype}b_2}. If we choose \ensuremath{1{\sumtype}1} to be
-\ensuremath{\mathit{bool}} and apply either \ensuremath{c_1:b_1\leftrightarrow
-b_2} or \ensuremath{c_2:b_1\leftrightarrow b_2} to the value \ensuremath{v},
-then we essentially have an `if' expression.
-
-\ensuremath{\mathit{if}_{c_1,c_2} : \mathit{bool}  {\prodtype}  b_1 \leftrightarrow \mathit{bool}  {\prodtype}  b_2}
-
-\ensuremath{\mathit{if}_{c_1,c_2} = \dist \odot ((\idc  \otimes\  c_1) {\sumtype} (\idc \otimes\  c_2)) \odot \factor}
+$\Afun{ifc}_{c_1,c_2} = \dist \fatsemi ((\idc  \otimes\  c_1) \oplus (\idc \otimes\  c_2)) \fatsemi \factor$
 
@@
 
-The diagram above shows the input value of type \ensuremath{(1{\sumtype}1) {\prodtype}  b_1}
-processed by the distribute operator \ensuremath{\dist}, which converts it into
-a value of type \ensuremath{(1 {\prodtype}  b_1){\sumtype}(1 {\prodtype}  b_1)}. In the
-\ensuremath{\mathit{left}} branch, which corresponds to the
-case when the boolean is \ensuremath{\mathit{true}} (i.e. the value was
-\ensuremath{\mathit{left} ~()}), the combinator~\ensuremath{c_1} is applied to
-the value of type~\ensuremath{b_1}. The right
-branch which corresponds to the boolean being \ensuremath{\mathit{false}} passes
-the value of type \ensuremath{b_1} through the combinator \ensuremath{c_2}.
-The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied
-to get the final result of type \ensuremath{(1{\sumtype}1) {\prodtype} b_2}.
-
-\paragraph*{Logic Gates}
-There are several universal primitives for conventional (irreversible)
-hardware circuits, such as \ensuremath{\mathit{nand}} and \ensuremath{\mathit{fanout}}. In the case
-of reversible hardware circuits, the canonical universal primitive is
-the Toffoli gate~\cite{Toffoli:1980}. The Toffoli gate takes three
-boolean inputs: if the first two inputs are \ensuremath{\mathit{true}} then the third
-bit is negated. In a traditional language, the Toffoli gate would be
-most conveniently expressed as a conditional expression like:
-
-\noindent
-\ensuremath{ \mathit{toffoli}(v_1,v_2,v_3) = \mathit{if} ~(v_1 ~\mathit{and} ~v_2) ~\mathit{then} ~(v_1, v_2, \mathit{not}(v_3)) ~\mathit{else} ~(v_1, v_2, v_3)}
-
-We will derive Toffoli gate in \ensuremath{\Pi } by first deriving a simpler
-logic gate called \ensuremath{\mathit{cnot}}.  Consider a one-armed version, \ensuremath{\mathit{if}_c},
-of the conditional derived above. If the \ensuremath{\mathit{bool}} is
-\ensuremath{\mathit{true}}, the value of type \ensuremath{b} is modified by the operator \ensuremath{c}.
-
-\begin{center}
-\scalebox{1.5}{
-%%subcode-line{pdfimage}[diagrams/if_c.pdf]
-% \includegraphics{inkscape/cnot.pdf}
-}
-\end{center}
-
-By choosing \ensuremath{b} to be \ensuremath{\mathit{bool}} and \ensuremath{c} to be \ensuremath{\mathit{not}}, we have the
-combinator \ensuremath{\mathit{if}_{\mathit{not}} : \mathit{bool} {\prodtype}  \mathit{bool}\leftrightarrow \mathit{bool} {\prodtype}  \mathit{bool}} which negates its
-second argument if the first argument is \ensuremath{\mathit{true}}. This gate
-\ensuremath{\mathit{if}_{\mathit{not}}} is often referred to as the \ensuremath{\mathit{cnot}} gate\cite{Toffoli:1980}.
-
-If we iterate this construction once more, the resulting combinator
-\ensuremath{\mathit{if}_{\mathit{cnot}}} has type \ensuremath{\mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})\leftrightarrow \mathit{bool} {\prodtype}  (\mathit{bool} {\prodtype}  \mathit{bool})}.
-The
-resulting gate checks the first argument and if it is \ensuremath{\mathit{true}},
-proceeds to check the second argument. If that is also \ensuremath{\mathit{true}} then
-it will negate the third argument. Thus \ensuremath{\mathit{if}_{\mathit{cnot}}} is the required
-Toffoli gate.
-
-\begin{center}
-\scalebox{1.6}{
-% \includegraphics{inkscape/toffoli.pdf}
-}
-\end{center}
+\noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
+which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the \ensuremath{\mathit{left}} branch,
+which corresponds to the case when the boolean is \ensuremath{\mathit{false}}, the combinator~\ensuremath{c_1} is
+applied to the value of type~\ensuremath{A}. The right branch which corresponds to the boolean being
+\ensuremath{\mathit{true}} passes the value of type $A$ through the combinator \ensuremath{c_2}.  The inverse of
+\ensuremath{\dist}, namely \ensuremath{\factor} is applied to get the final result. The full expressiveness of $\Pi$ is
+demonstrated by the following implementation of $\Afun{cnot}$ and $\Afun{toffoli}$ which are universal for reversible
+computation~\cite{Toffoli:1980}. The $\Afun{cnot}$ gates takes two input bits: a control bit and a target bit, and
+negates the target bit if the control bit is set. It is essentially a one-branch version of $\Afun{cif}$ above. The
+Toffoli gate generalizes the idea to a gate that takes 3 input bits: 2 control bits and one target bit: boolean inputs:
+it negates the target bit if both control bits are set.
 

### 2021-06-27 Pi+/paper/reversible2.tex
rev1
done with sec 3 for now

@@
 
-\noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \iso_2 q$, the denotation
+\noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \Iso q$, the denotation
 $\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.

### 2021-06-28 Pi+/paper/discussion.tex
Add notes and comments
Update references from zotero
Some text in discussion

@@
 
-Algebraic theories, 2-algebraic theories, Syntax of Coherence (Yanofsky)
+In this paper, we \ldots
+
+Our work lies at the intersection of programming language theory, category theory, group theory, rewriting theory, and
+formalised mathematics. We review related work in the literature for each topic.
+
+\paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe algebraic structures, such
+as groups or rings. A specific group or ring is a model of the appropriate algebraic theory. Algebraic theories are
+usually \emph{presented} in terms of logical syntax, that is, as first-order theories whose signatures allow only
+functional symbols, and whose axioms are universally quantified equations. In his seminal
+thesis~\cite{lawvereFUNCTORIALSEMANTICSALGEBRAIC1963}, Lawvere defined a presentation-free categorical notion of
+universal algebraic structure, called a Lawvere theory.
+
+Programming Languages, such as the $\lambda$-calculus, can be viewed as algebraic structures with variable-binding
+operators, which can be formalised using second-order algebraic theories~\cite{fioreSecondOrderAlgebraicTheories2010},
+or algebraic theories with closed structure~\cite{hylandClassicalLambdaCalculus2017}, called $\lambda$-theories, making
+the $\lambda$-calculus the presentation of the initial $\lambda$-theory $\Lambda$.
+
+Our family of reversible languages have been presented as first-order algebraic
+2-theories~\cite{cohenCoherenceRewriting2theories2009,bekeCategorificationTermRewriting2011,yanofskySyntaxCoherence2000},
+which are a categorification of algebraic theories. The types $\zerot$ and $\onet$ are nullary function symbols, the
+type formers $+$ and $\times$ are binary function symbols, the 1-combinators are invertible reduction rules, and the
+2-combinators are equations or coherence diagrams of compositions of reduction rules. Just like models of Lawvere
+theories are given by algebras of (finitary) monads on $\SetCat$, models of 2-theories are given by algebras of 2-monads
+on $\CatCat$. The particular one we're interested in here is the free symmetric monoidal completion 2-monad.
+
+\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of symmetric
+monoidal categories, strong symmetric monoidal functors, and symmetric monoidal natural transformations, to the
+2-category $\CatCat$, has a left adjoint.
+
+Curry-Howard-Lambek correspondence.
+
+Univalent Subuniverses, Univalent Typoids.
 

### 2021-06-28 Pi+/paper/ufin.tex
working on grpd fin

@@
-\section{The Universe of Finite Types}
+\section{The (Sub)Universe of Finite Types}
 \label{sec:ufin}
+
+In the previous section, we have picked a particular encoding of finite types: as sequences of elements in some
+canonical order.
+
+This is not a ``free'' presentation. Some properties of the encoding might pollute our semantics. We want to make that
+any theorems / properties of the semantics we prove apply to the source language (under various interpretations).
+
+This motivates us to use a different semantic representation: a univalent universe of finite types which contains the
+finite types and nothing but the finite types and is defined such as alternative representations of finite types are
+identified. This guarantees that we can get a proper, sound and complete, semantics for our language.
+
+\begin{center}
+\begin{tikzpicture}[scale=1,every node/.style={scale=1}]
+\usetikzlibrary{arrows}
+\usetikzlibrary{shapes}
+\node[ultra thick, draw=blue, ellipse, minimum width=300pt, minimum height=300pt, align=center] {};
+\node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (a) {$\{ 1 , 2 \} \uplus \{ 3 \}$};
+\node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (b) at ([shift=({70pt,-70pt})]a) {$\{ 1 \} \uplus \{ 2, 3 \}$};
+\node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (c) at
+([shift=({-70pt,-70pt})]a) {$\{ 10 \} \uplus \{ 20, 30 \}$};
+\node (d) at ([shift=({50pt,-10pt})]a) {$\cdots$};
+\draw[thick,-] (a) to node[draw=none,right] {$\equiv$} (b);
+\draw[thick,-] (b) to node[draw=none,below] {$\equiv$} (c);
+\draw[thick,-] (a) to node[draw=none,left] {$\equiv$} (c);
+\end{tikzpicture}
+\end{center}
+
+Given an operational model of how a type A reduces to a type B
+Given a denotational model of which types A and B are equivalent
+Want soundness and completeness results
+Guarantee that the two notions of equality are themselves “equivalent”
+Generalized “identity of indiscernibles”
 
@@
 groupoid semantics instead of set-theoretic.
-
-The lambda calculus is classically presented with types, terms, and
-$\beta\eta$-equations. This can be turned into a 1-category, with 0-cells for
-types, and 1-cells for programs quotiented by congruence closure of the
-$\beta\eta$-equations.
-
-Then there is the free CCC, which is a 1-category constructed using
-0-cells and 1-cells with universal properties that hold upto the notion
-of equality that from the ambient metatheory.
-
-Lambek/Scott show that these are equivalent. In modern language, this is
-a 1-equivalence of (1,1)-categories.
-
-Now, lets work without quotients in a proof-relevant metatheory.
-
-The lambda calculus can now be presented with 0-cells for types, 1-cells
-for terms, and 2-cells for equations/rewrites. This is a
-(2,1)-category. Then, you can give a description of the free 2-CCC,
-where universal properties hold upto canonical isomorphism.
-
-Then we can show that there is a 2-equivalence of these
-(2,1)-categories. (Actually, it can be weaker than (2,1)-categories, in
-fact it holds for bicategories.)
-
-Now, lets go to our result.
-
-For Pi/Pi+, we present it as a (2,0)-category (or 1-groupoid à la HoTT),
-with 0-cells for types, 1-cells for reversible programs, 2-cells for
-rewrites/equations. And we present different descriptions (Pi+, Ufin)
-of the free (2,0)-SMC (or 1-SMG) and show they're 2-equivalent.
-
-So this is just a difference of presentation, in going from the (1,r) to
-the (2,r) category. We have simply made the equations/rewrites
-first-class in the presentation.
 

### 2021-06-28 Pi+/paper/ufin.tex
Add a diagram about UFin and enumerations

@@
 \end{center}
+
+\[
+  \begin{tikzcd}
+    \Fin[3]
+    \arrow[""{name=0, anchor=center, inner sep=0}, "{f_{3}}", no head, loop, distance=4em, in=115, out=65]
+    \arrow[""{name=0, anchor=center, inner sep=0}, "{f_{2}}", no head, loop, distance=8em, in=125, out=55]
+    \arrow[""{name=1, anchor=center, inner sep=0}, "{f_{1}}"', no head, loop, distance=12em, in=135, out=45]
+    \arrow["", "{h}", shorten <=3pt, shorten >=3pt, Rightarrow, no head, from=0, to=1]
+  \end{tikzcd}
+\]
+
+We have $\Fin[3] & = \Set{0,1,2} \eqv \unit \sqcup (\unit \sqcup \unit)$ which fixes a particular enumeration of the
+elements. Suppose we have a set $X = (\unit \sqcup \unit) \sqcup \unit$, it has the same cardinality as $\Fin[3]$, so it
+is represented by the same 0-cell. But, $X$ can be made equivalent to $\Fin[3]$ in many different ways since there are
+many bijections between them. One bijection is
+$\Set{\inl(\inl(\ttt)) \mapsto 0, \inl(\inr(\ttt)) \mapsto 1, \inr(\ttt) \mapsto 2}$ which can be written in two
+different ways by composing more primitive operations, $f_{1} = \assocrp$, or
+$f_{2} = \swapp \compc \assoclp \compc \swapp$. Another bijection is
+$\Set{\inl(\inl(\ttt)) \mapsto 1, \inl(\inr(\ttt)) \mapsto 2, \inr(\ttt) \mapsto 0}$ which is given by $f_{3} = \ldots$.
+Since $f_{1}$ and $f_{2}$ produce the same enumeration of the elements of $X$, they are identified by a homotopy $h$
+which is encoded in the 2-cell between them.
 

### 2021-06-28 Pi+/paper/ufin.tex
More on UFin

@@
 which is encoded in the 2-cell between them.
+
+At level 0, all we know is that if $X : \UFin[3]$, then X is merely equal to $\Fin[3]$, that is
+$\Trunc[-1]{X \id \Fin[3]}$, and we don't have access to the bijection. At level 1, if we know that both $X$ and $Y$ are
+\emph{equal} in $\UFin[3]$, then we can extract an equivalence between them, that is, $(X \id Y) \to (X \eqv Y)$.
+$\UFin[3]$ being a univalent subuniverse asserts that there are as many elements (upto higher homotopy) in $X \id Y$ as
+there are $X \eqv Y$.
 

### 2021-06-29 Pi+/paper/equivalence.tex
Text in equivalence

@@
 
-Finally, We show how to translate $\PiLang$ programs to $\PiHatLang$ programs.
+Finally, we show how to translate $\PiLang$ programs to $\PiPlusLang$ programs. $\PiLang$ has two symmetric monoidal
+structure, the additive and the multiplicative one. In $\PiPlusLang$, the additive structure can be used to build the
+multiplicative structure, by defining multiplication as repeated addition.
 
@@
   \begin{align*}
-    \evalt_{0} & : U \to \Nat                                               \\
+    O \times Y           & \defeq O                       \\
+    I \times Y           & \defeq Y                       \\
+    (X_1 + X_2) \times Y & \defeq X_1 \times Y + X_2 \times Y
+  \end{align*}
+\end{definition}
+
+\begin{proposition}
+  $\PiPlusCat$ has a symmetric monoidal structure with $I$ for the unit and $\times$ for the tensor product. With the
+  two tensor products and the distributivity combinator, $\PiPlusCat$ has a rig structure.
+\end{proposition}
+
+Using this rig structure, we translate $\PiLang$ to $\PiPlusLang$.
+
+\begin{definition}
+  \begin{align*}
+    \evalt_{0} & : U \to \UPlus                                             \\
     \evalt_{1} & : (c : X \iso Y) \to \evalt_{0}(X) \iso \evalt_{0}(Y)      \\
@@
 
-To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$ and quote it to $\PiPlusLang$.
+To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$, compute a permutation, and quote back it to
+$\PiPlusLang$.
 

### 2021-06-29 Pi+/paper/reversible1.tex
denotational semantics in sec. 3.2

@@
 %%%%%%%%%
-\subsection{Denotational Semantics}
+\subsection{Denotational Semantics (Outline)}
 
-\noindent The accompanying code includes a simple operational semantics for $\Pi$. A denotational semantics can be
+\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be
 specified by mapping each type to a finite set and each combinator to a bijection between finite sets, that is, a
-permutation. In this section, we outline a simple semantics expressed in the metalanguage of set theory. This semantics
-is formalized and extended to the metalanguage of homotopy type theory in Sec.~\ref{sec:reversibletwo}.
+permutation. In this section, we outline a simple semantics expressed in the metalanguage of sets and functions. This
+semantics will be formalized and extended to the metalanguage of homotopy type theory in Sec.~\ref{sec:reversibletwo}.
 
 As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
-we have that $\sizet{A} = \sizet{B}$. Hence the denotation of a type $A$ with $n$-elements will be a finite set
-$\Fin[n]$ and the denotation of a combinator $c : A \iso B$ such that $\sizet{A} = \sizet{B} = n$ will be a function
-from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. In more detail, we enumerate the values $v : A$ where
-$\sizet{A} = n$ in a canonical order such that each value $v$ is given an index in $\{ 0,\cdots,n-1 \}$. The enumeration
-is defined as follows:
+we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
+set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
+the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
+$\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
+with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
+\[\begin{array}{rcl}
+    \denot{\onet + (\onet + \onet)} &=& \{ 0,1,2 \} \\
+    \denot{(\onet + \onet) + \onet} &=& \{ 0,1,2 \} \\
+    \\
+    \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (0~1~2) \\
+    \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (2~0~1)
+\end{array}\]
+where we have used the one-line notation for permutations with $(a~b~c)$ representing the
+permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
+enumeration of the elements of each type:
 \[\begin{array}{rcl}
@@
 a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
-operation that appends all the subsequences in a sequence. For example, the 3 values $\inlv{\Acon{tt}}$,
-$\inrv{(\inlv{\Acon{tt}})}$, and $\inrv{(\inrv{\Acon{tt}})}$ of type $\onet + (\onet + \onet)$ are given indices 0, 1, and 2
-respectively. Given this canonical order, the combinator $\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet$
-would correspond to the permutation $(1~2~0)$ while the combinator $\assoclp : \onet + (\onet + \onet) \iso (\onet +
-\onet) + \onet$ at the same type would correspond to the identity permutation.
+operation that appends all the subsequences in a sequence.
+
+Using the definition, we have:
+\[\begin{array}{rcl}
+    \mathit{enum}(\onet + (\onet + \onet)) &=& [ \inlv{\Acon{tt}},\inrv{(\inlv{\Acon{tt}})}, \inrv{(\inrv{\Acon{tt}})} ] \\
+    \mathit{enum}((\onet + \onet) + \onet) &=& [ \inlv{(\inlv{\Acon{tt}})},  \inlv{(\inrv{\Acon{tt}})}, \inrv{\Acon{tt}} ]
+  \end{array}\]
+Thus, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of $\inlv{\Acon{tt}}$ which is 0 and
+produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value $\inlv{(\inlv{\Acon{tt}})}$. Similarly,
+$\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of $\inlv{\Acon{tt}}$ which is 0 and produces
+index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value $\inrv{\Acon{tt}}$.
 

### 2021-06-29 Pi+/paper/discussion.tex
More in discussion

@@
 
-\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of symmetric
-monoidal categories, strong symmetric monoidal functors, and symmetric monoidal natural transformations, to the
-2-category $\CatCat$, has a left adjoint.
+\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of (small)
+symmetric monoidal categories, strong symmetric monoidal functors, and symmetric monoidal natural transformations, to
+the 2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
+$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, and its construction is known
+in the literature~\cite{abramskyAbstractScalarsLoops2005}. Concretely, the objects of $\FSM[\CCat{C}]$ are given by
+lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n] \to \CCat{C}_{0})$. An morphism between $(n,A)$ and
+$(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a permutation of $[n]$, and
+$\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the Grothendieck construction
+$\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and bijections to $\CatCat$, assigning
+each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on $[n]$ inducing an endofunctor on
+$C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one generator, $\FSM[\unit]$. The free
+symmetric monoidal category has been used to study concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri
+nets~\cite{baezCategoriesNets2021}, combinatorial structures~\cite{fioreCartesianClosedBicategory2008}, quantum
+mechanics~\cite{abramskyAbstractScalarsLoops2005}, bicategorical models of (differential) linear
+logic~\cite{melliesTemplateGamesDifferential2019}. In HoTT, the free symmetric monoidal groupoid $\FSM[A]$ over a
+groupoid $A$ can be given by $\dsum{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs.
+These HITs and the proof of their universal property have been considered
+in~\citet*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.
 
-Curry-Howard-Lambek correspondence.
+\paragraph{Curry-Howard-Lambek correspondence} In~\citet{curryCurryEssaysCombinatory1980}, Lambek extended the
+Curry-Howard correspondence to cartesian-closed categories. In this work, we have established a correspondence between a
+fragment of the $\PiLang$ family of languages and symmetric monoidal groupoids.
 

### 2021-06-29 Pi+/paper/finite.tex
Fix compilation
More text

@@
 of $\Fin{n}$ for every $n:\Nat$, that is, bijections on finite sets of size $n$. This is the extensional view of
-permutations. In the following sections, we will characterise these permutations, going through two intermediate steps.
+permutations. In the following sections, we will characterise these permutations syntactically- specifically, using a concept of \textit{a presentation} from the group theory.
 
-To do so, we will characterise the automorphisms on finite sets of cardinality $n$, and show them to be equivalent to
-the symmetric group $\Sn$, via the Coxeter presentation. We will do that in two steps,
-in~\cref*{subsec:permutations,subsec:lehmer,subsec:symmetric}.
+Although the meaning of a permutation - that is, the specific bijective function that it represents - is in some sense all there is, manipulating them syntactically still has its advantages.
+By writing reversible programs, we think of them in an intensional way.
+Comparing two programs for equality by evaluating them on all points in the domain is a very crude - and maybe even inefficient - way of doing that.
+
+It is crude, because there is no way of enforcing additional constraints on the process of transforming one program into another, and no way of inspecting what transformation occured.
+We can imagine a practical situation in which reversible circuts admit one kind of optimization \jk{FPGA?}, but do not admit another, even though they are equivalent in the extensional sense - or maybe one of these transformations is cheaper than another, or maybe it is important to know which transformation did occur for the \jk{producer(?)} to focus their resources of improving this kind of transformations.
+By comparing the circuits extensionally, the proof object = the path from one into the other - does not have any structure,  it is just a check that the values match.
+
+On the other hand, even though in general, the equality of circuits requires exponential time jk{triple check}, the proof object can be still very small.
+This creates structure that can possibly be exploited in a heurisic way.
+\jk{Additionally, if the goal is to convince third-party that two circuits are the same, we can just present the proof instead of showing the equality directly - for example, by choosing a particularly short one}.
+
+Taking all this into account, it is clear that we must seek a syntactic way of computing the meaning of the circuit, instead of just directly evaluating it to $\Aut[\Fin[n]]$ - since when we go back and quote the permutation as a circuit (in a normal form), there was no "trace" left to see what transformation (what sequence of 2-paths) maps the old one into the new one.
+
+Fortunately, group theory already has language in which such problems are expressed - we will be looking at solving the \textit{word problem}.
+Because $\PiLang$ is used for reasoning about reversible functions on finite sets, the group that we are interested in is the group of permutations on a fininte set, which is classically know as the symmetric group on $n$ words,
+$\Sn$.
+Putting it in this form allows us to connect to the broader scope of the computational group theory and reuse some of the ideas from there.
+For example, during the proof, we will use \emph{Lehmer codes} (introduced at \cref{subsec:lehmer}) - a concept used widely in this context, which makes for a very convenient and compact method of representing permutations on a computer.
+
+We begin the section with a brief introduction on doing group theory in HoTT, and state the main results using this language in the subsequent sections.
 
@@
 
-From universal algebra, a group is simply a set with a 0-ary constant, the neutral element, a binary operation for group
-multiplication, and a unary inverse operation. A simple example is $\mathbb{Z}/n\mathbb{Z}$, where the neutral element
-is 0, the inverse of $k$ is $-k$, and the group multiplication is given by addition modulo $n$. The neutral element has
+From universal algebra, a group is simply a set with a 0-ary constant (the neutral element), a binary operation for group multiplication, and a unary inverse operation. The neutral element has
 to satisfy unit and inverse laws, and the multiplication has to be associative.
+
+A very simple example of a group is $\mathbb{Z}$, where the neutral element is 0, the inverse of $k$ is $-k$, and the group multiplication is given by integer addition.
 
@@
 
-Given a generating set $A$, we can naively construct a \emph{free group on $A$},
+Given any set $A$, we can naively construct a \emph{free group on $A$},
 whose elements are drawn from the alphabet $A$, and closed under the group
 operations of multiplication and inverse, identified by the group axioms.
-For example, the singleton set generates the additive group of integers,
-$\mathbb{Z}$.
+For example, the singleton set generates the additive group of integers $\mathbb{Z}$ we introduced before. In this context, we will call $A$ the generating set of $F(A)$.
 
 Usually, there are are many equations, besides the group axioms, that hold for
-the elements of the group. For example, in the group $\mathbb{Z}/3\mathbb{Z}$,
-we have an equation $1 + 1 + 1 = 0$. A free group has the property that has no
-other relations than the ones directly implied by the group axioms.
+the elements of the group.
+For example, in the group of $\mathbb{Z}/3\mathbb{Z}$, where group multiplication is given by the addition modulo 3, we have an equation $1 + 1 + 1 = 0$, which is not a consequence of group axioms, but rather it is specific to the particular group $\mathbb{Z}/3\mathbb{Z}$.
+A free group can also be defined by its property that no other equations, except the ones directly implied by the group axioms, such as $g \dot g^[-1] = e$, hold for it.
+Thus, $\mathbb{Z}/3\mathbb{Z}$ is not a free group.
+
+Working in HoTT, as before, we can again use higher inductive types to define the free group.
 
@@
 this, we can state the universal property of free groups.
+\jk{In the HoTT version, which we now use, there is no underlying set}.
 
@@
 
-This definition of $F(A)$ however has lots of path constructors corresponding to
-each group axiom. Alternatively, the free group construction can be thought of
-as drawing letters from the generating set, while adding formal inverses, and
-constructing words which are sequences of these letters.
+We already mentioned that the structure of a free group $F(A)$ is fully specified by its generating set $A$.
+Thus, the univerasal property states that if we know how the mapping $f$ acts on the generating set $A$, we also know how does it act on every element of the group $F(A)$.
 
-Each element in the sequence can either be an element drawn from the generating
-set $a$, or an inverse of such an element $\inv{a}$. If we take the disjoint
-union of $A$ with itself, that is, $A + A$ as the underlying set, we can use
-$\inl/\inr$ to mark the elements. Hence, we can encode the free group using the
-free monoid, that is, lists.
+The HoTT definition of $F(A)$ has lots of path constructors corresponding to
+each group axiom. \jk{why did we introduced it? Why is it bad?}.
+
+Instead, in the proof, we will think about the free group as it was introduced at the beginning: elements of the free group as words over an alphabet of letters from generating set and the set of their formal inverses.
+If we take the disjoint union of $A$ with itself, that is, $A + A$ as the group's underlying set, we can use $\inl/\inr$ to mark the elements - for example, by stating that $\inl(a)}$ means $a$ and $\inr(a)$ means $\inv{a}$.
+Hence, we can encode the free group using the free monoid - which is to say, lists on $A + A$.
+Additionally, we need to ensure that group laws hold, so words identified by the group equations are equalt.
+Thus, elements of the group will actually be equivalence classes of lists on $A + A$.
 
@@
   empty list $\nil$ for the neutral element, multiplication given by list append
-  $\append$, and inverse given by list reversal. Further, $F(A)$ with $\eta_A :
+  $\append$, and inverse given by list reversal (\jk{THIS IS NOT TRUE}). Further, $F(A)$ with $\eta_A :
     A \to F(A) \defeq \inl(a) \cons \nil$ satisfies the universal property of free
-  groups.~\cref{prop:free-groups}
+  groups, as stated in the ~\cref{prop:free-groups}.
 \end{proposition}
 
-\subsubsection{Group presentations}
+\subsection{Group presentations}
 
-One useful way of defining groups is by their presentations. A presentation of a
-group builds it by starting from the free group $F(A)$ and introducing a
+We already talked about the semantic view of groups, but did not yet mention on how to syntactically represent them - which is the role of group presentations.
+
+A presentation of a group builds it by starting from the free group $F(A)$ and introducing a
 collection of equations that have to be satisfied in the resulting group. For
@@
   = 0$, the resulting group would be $\mathbb{Z}/3\mathbb{Z}$.
+Note that not all groups have finite (or even recursively enumerable) presentations, and that a group can have any number of presentations.
+\jk{citation}
 
@@
 
-Instead of opereating directly in the semantics of the group operation, here the
+As we mentioned, instead of opereating directly in the semantics of the group operation, here the
 focus is on the syntax. While before, the equality of elements (such as the
-result of multiplication of two elements) had to be computed externally, now it is
-reduced to a \emph{word problem}, i.e., deciding one word can be reduced to another
+result of multiplication of two elements) had to be computed in some external way, now it is
+reduced to a \emph{word problem}, i.e., deciding one word - representative of the group elements' equivalence class - can be reduced to another word,
 using group's equations.
@@
 However, because equations are not directed, is not always possible to construct a
-well-behaving rewriting system. In general, the word problem is proven to be undecidable.
+well-behaving rewriting system.
+In general, the word problem is proven to be undecidable.
+There exists an algorithm, due to Knuth and Bendix \jk{citation}, that, when succeeds, constructs a well-behaving rewriting system for an arbitrary finite set of (undirected) equations.
+However, this algorithm produced a system that is often large and hard to reason about \jk{definitely a citation if we can find it}.
 
@@
 
-Because $\PiLang$ is used for reasoning about reversible functions on finite
-sets, the group that we are interested in is the group of permutations on a
-fininte set, which is classically know as the symmetric group on $n$ words,
-$\Sn$. We've already seen that we can define it formally by taking the
-automorphism group given by the type $\Aut[\Fin[n]]$, and showing that it has a
-group structure satisfying the group axioms. Alternatively, we can write a group
-presentation for $\Sn$ -- by defining a set of generators and a set of relations.
+We have already seen that the group of permutations, symmetric group $\Sn$, can be defined by taking the type $\Aut[\Fin[n]]$, and showing that it has the appropriate
+group structure.
+Alternatively, we talked on how one can define a group by its presentation,defining a set of generators and a set of relations.
+However, until now, we did not show any specific presentationation for $\Sn$.
 
-The technical contribution of this section is a proof that these two
-descriptions are equivalent. By doing that, we bridge the gap between the
-syntactic and semantic notions in our completeness proof -- by establishing a
-correspondence between the meaning of a program (a bijection) and its syntax (a
-word).
-
-However, in giving a group presentation, there is an element of choice. A group
+In giving a group presentation, there is an element of choice - as we already mentioned, a group
 can be presented in many different ways. For example, we could generate the
@@
 
-\subsection{Coxeter Presentation}
+The choice of generators for a presentation is important. \todo{Think about this.}
 
-We use a presentation based on adjacent transpositions, which is called the
-Coxeter presentation. The primitive operations are adjacent swaps. Further,
-there are three laws that these operations have to satisfy -- it is easiest to
-visualise them as braid diagrams.
+\begin{itemize}
+  \item It affects the difficulty of solving the word problem in $\Sn$ and
+        formalising the proof of its correctness.
+  \item It affects the proof of strong normalisation of reduction for the relations.
+  \item It dictates which words are normal forms in this presentation of $\Sn$.
+  \item It has to closely match the $\PiLang$ comobinators so that we can prove
+        completeness.
+  \item It dictates the choice of reversible gates in the synthesis and
+        normalisation of boolean circuits which are applications that we show in
+        the later sections.
+\end{itemize}
+
+Connecting it all back to the language $\PiLang$ we are dealing with, it is possible to construct a simlar, yet simplified language $\PiLangHat$ (introduced at \cref{sec:equivalence}), which operates on adjacent transpositions, but still preservers much of $\PiLang$ structure.
+Thus, for the proof, a presentation based on adjacent transpositions is used, called \emph{Coxeter presentation}.
+Not only is this then, in some sense, the natural choice, but it is also a presentation for which we can solve the word problem in a relatively easy, and which is used broadely in such context \jk{by other people?
+get citations}.
+
+The technical contribution of this section is a proof that the two descriptions of permutation group - $\Aut[\Fin[n]]$ and the Coxeter presentation - are equivalent.
+By doing that, we bridge the gap between the syntactic and semantic notions in our completeness proof -- by establishing a
+correspondence between the meaning of a program (a bijection) and its syntax (a word).
+
+\subsubsection{Coxeter Presentation}
+
+We already mentioned that the Coxter presentation, which is the one will use, is based on adjacent transpositions, meaning that the primitive operations we use are adjacent swaps.
+However, $\Sn$ is not a free group, and so we also have to specify the relations of the presentaiton -- the laws that such generators have to satisfy.
+there are three of them -- and it is easiest to visualise them as braid diagrams.
 
@@
 
-\todo{explain braid in words, give some intution on why it holds}
+The previous two cases were for the transpositions that either completely overlap, or not overlap at all.
+The third case is thus left -- what happens if we perform transpositions that move the same elements.
+If the next transposition in the sequence is equal to the the first one, we can write a third law -- braiding relation.
 
@@
 
-The choice of generators for this presentation is important. \todo{Think about this.}
+The approach to solving the word problem for $\Sn$ we use is to turn the relations $\cox*$ on
+$\List[\Fin[n]]$ to a rewriting system $(\List[\Fin[n]],\cox*)$.
+A normal form of an element can then be computed by following the rewrite rules, and two words can be compared for $\cox*$-equality by comparing their normal forms.
 
-\begin{itemize}
-  \item It affects the difficulty of solving the word problem in $\Sn$ and
-        formalising the proof of its correctness.
-  \item It affects the proof of strong normalisation of reduction for the relations.
-  \item It dictates which words are normal forms in this presentation of $\Sn$.
-  \item It has to closely match the $\PiLang$ comobinators so that we can prove
-        completeness.
-  \item It dictates the choice of reversible gates in the synthesis and
-        normalisation of boolean circuits which are applications that we show in
-        the later sections.
-\end{itemize}
+A well-behaved rewriting system has to be strongly normalising and confluent. First, we observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
+are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$.
+Thus, by directing the relation from left to right, we would get the termination property out of the box.
 
-To consider solving the word problem for $\Sn$, we need to formalise the generators for the Coxeter relations on
-$\List[\Fin[n]]$ as a rewriting system $(\List[\Fin[n]],\cox*)$. A well-behaved rewriting system has to be strongly
-normalising and confluent. Throwing out reflexivity and symmetry, we observe that the right hand sides of the relations
-are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$. Thus, by
-directing the relation from left to right, we would be able to prove a termination property out of the box. For a system
-to be confluent, all critical pairs, that is, terms with overlapping possible reduction rules, have to converge. For
-example, the pairs below converge.
+The system also has to be confluent, meaning that all critical pairs, that is, terms with overlapping possible reduction rules, have to converge. For example, the pairs below converge.
 
@@
 \begin{proof}
+\[
     \begin{array}{lcr}
@@
     \end{array}
+\]
 \end{proof}
 
-Unfortunately, in the system defined, this is not true for all critical pairs.
+Unfortunately, in the system we defined, this is not true for all critical pairs.
 One example is the following, where both endpoints are normal with respect to
@@
 
-Because of the counterexample discussed, the relations have to be changed. In this section, we will formally define a
-rewriting system partially based on the Coxeter relation, prove that it has good properties (confluence and strong
-normalisation).
+Because of the counter-example discussed, the relations have to be changed.
+In this section, we will formally define a rewriting system partially based on the Coxeter relation, prove that it has the desired properties of confluence and strong normalisation.
 
-The reduction system $(\List[\Fin[n]], \longcox*)$ has generators corresponding to $\swap$, $\cancel$ and $\braid$. We
-fix the problem discussed previously by changing $\braid$ to be a slightly more general relation. However, we do not
-want to lose the connection to the theory of group presentations, where Coxeter relations are a standard
-notion~\todo{citation}. To make the result broadly applicable, even though we use the reduction system for computation,
-we prove that its reflexive-transitive closure is indeed equivalent to the conguence closure of the standard Coxeter
-relation $\cox$.
+Fortunately, the change from the rewriting system described before will be small.
+Thus, we will not lose the connection to the theory of group presentations, where Coxeter relations are a standard notion~\todo{citation} -- to make the result broadly applicable, even though we use the reduction system for computation, we prove that its reflexive-transitive closure is indeed equivalent to the conguence closure of the standard Coxeter relation $\cox$.
+
+The new reduction system $(\List[\Fin[n]], \longcox*)$ has generators corresponding to $\swap$, $\cancel$ and $\braid$.
+We fix the problem of the non-converging critical pair discussed previously by changing $\braid$ to be a slightly more general relation $\longbraid$.
 

### 2021-06-29 Pi+/paper/finite.tex
More progress on sec 6

@@
 We already mentioned that the Coxter presentation, which is the one will use, is based on adjacent transpositions, meaning that the primitive operations we use are adjacent swaps.
+When dealing with permutations on an $n + 1$-element set, there are $n$ adjacent transpositions, transposition number $k$ swapping elements $\el{k}$ and $\el{k+1}$.
+Thus, the generating set would be $\Fin[n]$.
+
 However, $\Sn$ is not a free group, and so we also have to specify the relations of the presentaiton -- the laws that such generators have to satisfy.
-there are three of them -- and it is easiest to visualise them as braid diagrams.
+there are three of them -- it is easiest to visualise them as braid diagrams.
 
-Swapping the same two elements two times in a row is the same as doing nothing:
+First, swapping the same two elements two times in a row is the same as doing nothing:
 
@@
 
-When swapping two distinct pairs of elements (i.e. when the indices of two transpositions
+Second, when swapping two distinct pairs of elements (i.e. when the indices of two transpositions
 differ by at least 1), it does not matter in which order the swapping happens,
@@
 
-This construction is called a Coxeter presentation. Writing it formally, we define a binary relation $\cox$ on
-$\List[\Fin[n]]$.
+This construction is called a Coxeter presentation.
+Writing it formally, we take the generating set to be $\Fin[n]$, where the element $k$ corresponds to an adjacent transpositions $\tau_k$, which swaps elements $\el{k}$ and $\el{k+1}$.
+Then, we define a binary relation $\cox$ on the $\List[\Fin[n]]$, which encodes the laws discussed above.
 
@@
 
+First, we need a helper function $n \downf k$.
+
+\begin{definition}[$\downf : (n : \Nat) \to (k : \Nat) \to {\List[\Fin[\suc[k + n]]]}$]
+  \begin{align*}
+    n \downf \zero   & \defeq \nil                       \\
+    n \downf \suc[k] & \defeq (k + n) \cons (n \downf k)
+  \end{align*}
+\end{definition}
+
+The reuslt of this function is the sequence \(k + n, k + n - 1, k + n - 2, \ldots, n + 1\).
+If we think of the element $m$ as the transposition which swaps elements $\el{m}$ and $\el{m+1}$, the role of this helper function is to produce a sequence of transpositions - a permutation - which moves element $n$ $k$ places right.
+
+As a sequence of transpositions, it moves the element $k - 1$ places. Expressed in terms of the braiding diagram, for
+$n = \dots$ and $k = \dots$, it has the following form:
+
+\[
+  \begin{tikzpicture}
+    \pic[local bounding box=my braid,braid/.cd,
+      number of strands = 6,
+      thick]
+    {braid={s_5, s_4, s_3, s_2, s_1}};
+  \end{tikzpicture}
+\]
+
 \todo {but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
@@
 Generators $\longcancel$ and $\longswap$ correspond directly to the generators of $\cox*$ and can be visualised in the
-same way as before. The generator $\longbraid$ uses a helper $n \downf k$, which produces the sequence
-\(k + n, k + n - 1 \ldots, n + 1\).
-
-\begin{definition}[$\downf : (n : \Nat) \to (k : \Nat) \to {\List[\Fin[\suc[k + n]]]}$]
-  \begin{align*}
-    n \downf \zero   & \defeq \nil                       \\
-    n \downf \suc[k] & \defeq (k + n) \cons (n \downf k)
-  \end{align*}
-\end{definition}
-
-As a sequence of transpositions, it moves the element $k - 1$ places. Expressed in terms of the braiding diagram, for
-$n = \dots$ and $k = \dots$, it has the following form:
-
-\[
-  \begin{tikzpicture}
-    \pic[local bounding box=my braid,braid/.cd,
-      number of strands = 6,
-      thick]
-    {braid={s_5, s_4, s_3, s_2, s_1}};
-  \end{tikzpicture}
-\]
+same way as before. The generator $\longbraid$
 

### 2021-06-29 Pi+/paper/finite.tex
Fix the proposition about cox equiv longcox

@@
 
-\begin{definition}
+\begin{definition}~\label{def:presentation}
   Let $A$ be an $\hSet$, and $\List[\blank]$ the free monoid. The free group
@@
 
-The type $\Sn$ is then defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.
+\jk{maybe change that?}
+Group persentation, as given in the ~\label{def:presentation}, requires the relation to be on the set of $A + A$, where the right copy corresponded to the set of formal inverses of the generators.
+However, in our case, we can observe that that the constructor $\cancel$ specifies that the inverse of each element is again the same element.
+Thus, the type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.
 
@@
 Fortunately, the change from the rewriting system described before will be small.
-Thus, we will not lose the connection to the theory of group presentations, where Coxeter relations are a standard notion~\todo{citation} -- to make the result broadly applicable, even though we use the reduction system for computation, we prove that its reflexive-transitive closure is indeed equivalent to the conguence closure of the standard Coxeter relation $\cox$.
+Thus, we will not lose the connection to the theory of group presentations, where Coxeter relations are a standard notion~\todo{citation} -- to make the result broadly applicable, even though we use the reduction system for computation, we prove that the \jk{relation it gives} is indeed equivalent to the conguence closure of the standard Coxeter relation $\cox$.
 
@@
 
-First, we need a helper function $n \downf k$.
+\todo {but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
+
+First, we need to define a helper function $n \downf k$.
 
@@
 
-The reuslt of this function is the sequence \(k + n, k + n - 1, k + n - 2, \ldots, n + 1\).
-If we think of the element $m$ as the transposition which swaps elements $\el{m}$ and $\el{m+1}$, the role of this helper function is to produce a sequence of transpositions - a permutation - which moves element $n$ $k$ places right.
-
-As a sequence of transpositions, it moves the element $k - 1$ places. Expressed in terms of the braiding diagram, for
-$n = \dots$ and $k = \dots$, it has the following form:
+The result of this function is the sequence \(k + n, k + n - 1, k + n - 2, \ldots, n + 1\).
+Since we think of the number $m$ as the transposition $\tau_m$ which swaps elements $\el{m}$ and $\el{m+1}$, the role of this helper function is to produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ by $k$ places left, shifting all the elements in between one place right.
+Expressed in terms of the braiding diagram, for
+$n = 0$ and $k = 4$, it has the following form:
 
@@
     \pic[local bounding box=my braid,braid/.cd,
-      number of strands = 6,
+      number of strands = 5,
       thick]
-    {braid={s_5, s_4, s_3, s_2, s_1}};
+    {braid={s_4, s_3, s_2, s_1}};
   \end{tikzpicture}
@@
 
-\todo {but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
-
-Formally, the directed $\longcox$ relation is defined with the following generators.
+Then, the directed relation $\longcox$ is defined with the following generators.
 
@@
 
-Like before, we define the $\longcox*$ to be the closure of $\longcox$, but only under reflexivity and transitivity. The
-transitive extension is only performed on the right.
-
-\begin{definition}[$\longcox*$]
-  \begin{align*}
-    \reflr{\longcox}
-     & : \forall w \to w \longcox* w                                                                               \\
-    \transr{\longcox}
-     & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcox* w_{3} \to w_{1} \longcox* w_{3} \\
-  \end{align*}
-\end{definition}
-
-Generators $\longcancel$ and $\longswap$ correspond directly to the generators of $\cox*$ and can be visualised in the
-same way as before. The generator $\longbraid$
-
-Now, the remaining constructor $\longbraid$ moves a transposition around such sequence, as in the diagram below.
+Constructors $\longcancel$ and $\longswap$ correspond directly to the approproiate constructors of $\cox$ and can be visualised in the same way as before.
+The remaining constructor uses the helper function to exchange the order of a long sequence of transpositions and a single transposition afterwards.
+For example, for $n = 2$ and $k = 4$, it allows for the reduction
+$[7, 6, 5, 4, 3, 7] \longcox* [6, 7, 6, 5, 4, 3]$. Visualized on the braid diagram, it looks as follows:
 
@@
 
-Note that the previous $\braid$ rule is a special case of $\longbraid$, with $k = 2$.
+Note that the previous $\braid$ rule is a special case of $\longbraid$, with $k = 0$.
+
+Like before, we define the $\longcox*$ to be the closure of $\longcox$, but only under reflexivity and transitivity.
+
+\begin{definition}[$\longcox*$]
+  \begin{align*}
+    \reflr{\longcox}
+     & : \forall w \to w \longcox* w                                                                               \\
+    \transr{\longcox}
+     & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcox* w_{3} \to w_{1} \longcox* w_{3} \\
+  \end{align*}
+\end{definition}
 
@@
 
-The modified form of the Coxeter relations are unwieldy and difficult to prove properties about by induction, and have
-little connection to the broader scope of computational group theory. However, we can prove that the closures of the two
-forms of the relations are equivalent, and work up to this equivalence!
+The modified form of the Coxeter relations are unwieldy and difficult to prove properties about by induction, and, as we mentioned, have little connection to the broader scope of computational group theory.
+However, we can prove that the closures of the two forms of the relations are equivalent, and work up to this equivalence!
 
-\begin{proposition}
-  $\cox*$ and $\longcox*$ are equivalent. For every $w$ and $v$, $w \cox* v$ iff $w \longcox* v$.
+\begin{proposition}~\label{prop:coxlongcox}
+  $\cox*$ and $\longcox*$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff there is a $u$ such that $w \longcox* u$ and $v \longcox* u$.
 \end{proposition}
+\jk{does the construction above has a name?
+It is analogous to the one we do in lambda calculus, when taking $\equiv_{\beta}$.
+We have to write something about this proof using the Church-Rosser idea.}
 

### 2021-06-29 Pi+/paper/finite.tex
Rework up to immersion

@@
 
-From grade school combinatorics, we know that there are $\fac{n}$ permutations on a finite set with $n$ elements. For
-every $n$, we thus want to produce a finite set with cardinality $\fac{n}$, and a bijection with $\Aut[\Fin[n]]$. There
-are many ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer codes are known in
+There are many ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer codes are known in
 Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960} where they are sometimes called "subexcedant sequences",
 or "factoriadics", which give the factorial number system. They are a particularly convenient way of representing
-permutations on a computer - since they have exactly the right cardinality, they are bitwise-optimal.
+permutations on a computer, partly because they are bitwise-optimal: for any $n : \Nat$, the type $\Lehmer[n]$ hass the cardinality $\fac{n}$, and has an easy to consttuct bijection with $\Aut[\Fin[n]]$.
+.
 
-Formally, $\Lehmer[n]$ is an $n+1$-element tuple, where the position $k \leq n$ stores an element of $\Fin[k]$. Since
+Formally, we define $\Lehmer[n]$ to be an $n+1$-element tuple, where the position $k \leq n$ stores an element of $\Fin[k]$. Since
 the 0-th position is trivial, in practice, it is ignored, and the sequence starts at
 1~\cite{duboisTestsProofsCustom2018,vajnovszkiNewEulerMahonian2011}.\todo{There are more examples - listed
-  in~\cite{duboisTestsProofsCustom2018}}. We can define $\Lehmer$ in two equivalent ways, by a simple recursion on
+  in~\cite{duboisTestsProofsCustom2018}}. We can then define $\Lehmer$ in two equivalent ways, by a simple recursion on
 $\Nat$, or as a type family generated by two constructors.
@@
 
-Suppose we have a permutation $\sigma$ on an $n$-element set $\{\el{0}, \el{1}, \el{2}, \el{3}, \el{4}\}$, we encode it
-as follows. $\Lehmer[n]$ is a $n$-element tuple. Then, At position $k$, we put the number of inversions of the element
-$\el{k}$ in $p$, i.e. the number of elements smaller than $\el{k}$ occurring after $\el{k}$. As an example, consider the
-following tabulated presentation of the permutation:
+For a permutation $\sigma : \Aut[\Fin[n]]$, for any element $i: \Fin[n]$, we can define the inversion count of $i$ as the number of elements smaller than it appearing after in the permutation.
+\todo{typify}
+\begin{definition}
+    Given $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by 
+    $|\{j < i : \sigma(j) > \sigma(i)\}|$.
+\end{definition}
+
+It turns out that knowing just the inversion counts for all the elements, one can reconstruct the starting permutation.
+Also, observe that the inversion count for element $i$ is guaranteed to be smaller than $i$, thus fitting in the $i$-th place of a Lehmer code tuple.
+As an example, consider the following tabulated presentation of the permutation:
 
@@
     \hline                                                             \\
-    | & \el{2} & | & \el{0} & | & \el{1} & | & \el{4} & | & \el{3} & | \\
+    | & \el{2} & | & \el{1} & | & \el{4} & | & \el{0} & | & \el{3} & | \\
     \hline                                                             \\
@@
 
-The element $\el{0}$ has 0 inversions, because there are no elements smaller than $\el{0}$ occurring after it. In fact,
-there can be no elements smaller than $\el{0}$ at all, so the type at the first position of the Lehmer code tuple is
-$\unit$.
+The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 0 (because of $\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it is last in the sequence), and for $\el{4}$ is 1 (because of $\el{3}$).
+Thus, the Lehmer code for the permutation $\sigma$ is the 5-tuple $l = (0, 1, 2, 0, 1)$.
 
-The element $\el{1}$ has 0 inversions as well, since elements occurring after it in the permutation are $\el{4}$ and
-$\el{2}$. There is only one different case, if $\el{1}$ appeared before $\el{0}$, it would have 1 inversion. This is why
-the type of the second component of the Lehmer code is $\Fin[2]$.
-
-The element $\el{2}$ has 2 inversions, because both $\el{0}$ and $\el{1}$ occur after it in the permutation. The element
-$\el{3}$ occurs as the last one, so it has 0 inversions. The element $\el{4}$ has 1 inversion, with the element
-$\el{3}$.
-
-Thus, the Lehmer code for the permutation $\sigma$ is the 5-tuple $l = (0, 0, 2, 0, 1)$.
-
-To reconstruct the tabulated presentation of the permutation from the Lehmer code, we perform an algorithm similar to
-\emph{insertion sort}. Starting from the left-most position of the tuple $l$, we'll read the value $v$, insert the new
+To construct the presentation of the permutation from the Lehmer code, written using adjacent transpositions, we perform an algorithm similar to \emph{insertion sort}.
+Starting from the left-most position of the tuple $l$, we'll read the value $v$, insert the new
 element at the end of the newly created list, and shift it backward $v$ places.
@@
 \end{proposition}
+
+\todo{So now, finally, give an explict example of a normal form, Lehmer immersion}.
 

### 2021-06-29 Pi+/paper/ufin.tex
Minor: fix build
Revise section 4.2

@@
 
-We have $\Fin[3] & = \Set{0,1,2} \eqv \unit \sqcup (\unit \sqcup \unit)$ which fixes a particular enumeration of the
+We have $\Fin[3] = \Set{0,1,2} \eqv \unit \sqcup (\unit \sqcup \unit)$ which fixes a particular enumeration of the
 elements. Suppose we have a set $X = (\unit \sqcup \unit) \sqcup \unit$, it has the same cardinality as $\Fin[3]$, so it
@@
 In order to define the semantics, we seek a mathematical structure satisfying the following properties: (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust enough to ensure that equivalent encodings of finite types are identified.
-\note{Is this the correct informal explanation of Ufin?}
 
@@
 
-We work in Homotopy Type Theory, in particular, intensional Martin-L\"{o}f Type Theory, with a univalent universe, and
-Higher Inductive Types (HITs) for propositional truncations and set quotients. We recall a few basic facts to
+We work in Homotopy Type Theory (HoTT), and use the language of the HoTT
+book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use intensional Martin-L\"{o}f Type Theory,
+with a univalent universe $\UU$, and we use Higher Inductive Types (HITs) for propositional truncations and set
+quotients. All arguments will hold in a cubical type theory as well, such
+as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019}. We recall a few basic facts to
 familiarise the reader with the notation we use. For more details about basics of HoTT, we refer the reader to the HoTT
@@
 
-\note{It is important that we work in HoTT using it as a metatheory, we care about proof relevance, constructivity, and
-  working with higher-dimensional algebraic structures, such as groupoids. Our results are also formalised using general
-  tools and principles from HoTT, and computer checked using a proof assistant, using Agda and the HoTT-Agda library.}
+Our denotational semantics is given using groupoids, and HoTT provides a rich internal language for describing (weak
+$\infty$-)groupoids. By formalising our results using general tools and principles from HoTT, we're able to use a proof
+assistant to computer-check our proofs. Our results are formalised in the Agda proof assistant using the HoTT-Agda
+library, though it could also be formalised using Cubical Agda~\cite{vezzosiCubicalAgdaDependently2019}. \review{Using
+  the formalisation of the semantics, we are also able to extract a computable procedure for performing normalisation of
+  circuits.}
 
@@
 
-In HoTT, the intensional identity type is the type of paths between two terms of the same type. Given two terms $x:A$
-and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the equality type between them. In book HoTT, the identity
-type is generated by reflexivity $\refl_{x}$, and the eliminator for the identity type is given by path induction or the
-$J$-rule. The identity type equips each type with the structure of an infinity groupoid, or a homotopy type.
+In HoTT, the identity type or the equality type is the type of paths or equalities between two terms of the same type.
+Given two terms $x:A$ and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the equality type between them. In
+book HoTT, the identity type is generated by reflexivity $\refl_{x} : x \id_{A} x$, and the eliminator for the identity
+type is given by path induction or the $J$-rule. The identity type equips each type with the structure of a weak
+$\infty$-groupoid, or a homotopy type. For example, given an arbitrary type (or groupoid) $A$, we list some groupoid
+laws that are provable using path induction.
 
-\todo{Say why it is a groupoid by listing some groupoid laws.}
+\begin{gather*}
+  \begin{aligned}
+    \term{sym} & : (x \id_{A} y) \to (y \id_{A} x) \\
+    \term{\comp} & : (x \id_{A} y) \to (y \id_{A} z) \to (x \id_{A} z)
+  \end{aligned}
+  \qquad
+  \begin{aligned}
+    \term{assoc} & : (p : x \id_{A} y)  (q : y \id_{A} z) (r : z \id_{A} w) \\
+    & \to (p \comp q) \comp r \id p \comp (q \comp r) \\
+    \term{invr} & : (p : x \id_{A} y) \to p \comp \term{sym}(p) \id \refl_{x}
+  \end{aligned}
+\end{gather*}
 
-Functions between types are functors between groupoids. Given a function $f : A \to B$, the functorial action is given
-by
+Functions between types are functors between groupoids. Given a function $f : A \to B$, the functorial action
+$\term{ap}_{f}(\blank)$ is given by path induction.
 
@@
 
-Type families are functions from a type to the universe, that is, an indexed family of groupoids. The $\term{transport}$
-operation gives the functorial action of paths in the indexing type, which is defined by path induction. If
-$P : A \to \UU$ is a type family, then for a path $x \id_{A} y$, we have
+Type families are functions from a type to the universe, such as $A \to \UU$, which is an $A$-indexed family of
+groupoids. The $\term{transport}$ operation gives the functorial action of paths in the indexing type, which is also
+defined by path induction. If $P : A \to \UU$ is a type family, then for a path $x \id_{A} y$, we have
 
@@
 
-\todo{Why is the topological viewpoint important, or saying fibrations? Explain the intuition using classifying spaces,
-  maybe draw some diagrams.}
+\todo{Diagram.}
 
-From the topological viewpoint, a type family can also be seen as a fibration. For a type family $P : A \to \UU$ and a
-point $x : A$, $P(x)$ gives the fiber over $a$. For a path $p : x \id_{A} y$, $\transport{P}{p}$ gives the path lifting
-operation. The total space is given by $\dsum{x:A}{P(x)}$ and the first projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ to
-the base space is the fibration. The lifting operation lifts paths in the base space to paths in the total space. If
-$p : x \id_{A} y$ is a path in the base space, and $u : P(x)$, we have
+A type family can also be seen as a fibration, from the topological viewpoint. For a type family $P : A \to \UU$ and a
+point $x : A$, the type $P(x)$ gives the fiber over $x$. The type $\dsum{x:A}{P(x)}$ is the collection of all the fibers
+and is called the total space of $P$. The first projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ from the total space to
+the base space $A$ is then a fibration. The lifting operation lifts paths in the base space to paths in the total space.
+Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have
 
@@
 
-Further, using the groupoid structure of $A$, we can show that transport lifts paths to equivalences, we define
+Further, using the groupoid structure of $A$, we can show that $\term{transport}$ lifts paths to equivalences, we define
 
@@
 
-\todo{Explain motivation.}
+\todo{Diagram.}
 

### 2021-06-30 Pi+/paper/reversible1.tex
diagrams for permutations in sec 3.2

@@
 \[\begin{array}{rcl}
-    \mathit{enum}(\onet + (\onet + \onet)) &=& [ \inlv{\Acon{tt}},\inrv{(\inlv{\Acon{tt}})}, \inrv{(\inrv{\Acon{tt}})} ] \\
-    \mathit{enum}((\onet + \onet) + \onet) &=& [ \inlv{(\inlv{\Acon{tt}})},  \inlv{(\inrv{\Acon{tt}})}, \inrv{\Acon{tt}} ]
+    \mathit{enum}(\onet + (\onet + \onet)) &=& [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
+    \mathit{enum}((\onet + \onet) + \onet) &=& [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
 \end{array}\]
-Thus, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of $\inlv{\Acon{tt}}$ which is 0 and
-produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value $\inlv{(\inlv{\Acon{tt}})}$. Similarly,
-$\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of $\inlv{\Acon{tt}}$ which is 0 and produces
-index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value $\inrv{\Acon{tt}}$.
+Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
+$\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
+$\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
+$\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
+$\inrv{\Acon{tt}}$.
+
+\begin{center}
+  \input{assoc-perm.tikz}
+  \qquad
+  \input{swap-perm.tikz}
+\end{center}
 

### 2021-06-30 Pi+/paper/equivalence.tex
Working on sec 7

--- 2021-06-25 Pi+/paper/equivalence.tex
@@
-\section{Equivalence between \texorpdfstring{$\PiPlusLang$}{Pi} and \texorpdfstring{$\UFin$}{UFin}}~\label{sec:equivalence}
+\section{Going from \texorpdfstring{$\PiLang$}{Pi} to \texorpdfstring{$\UFin$}{UFin} and back}~\label{sec:equivalence}
 
-In this section, we use the semantics developed in the previous sections define a fragment $\PiPlusLang$ of $\PiLang$
-and a normalised fragment of it $\PiHatLang$, interpret to $\UFin$ and back, and translate between the languages.
+% https://q.uiver.app/?q=WzAsNCxbMCwwLCJcXFBpTGFuZyJdLFsyLDAsIlxcUGlQbHVzTGFuZyJdLFs0LDAsIlxcUGlIYXRMYW5nIl0sWzYsMCwiXFxVRmluIl0sWzAsMSwiXFxldmFsdCJdLFsxLDIsIlxcZXZhbGgiLDAseyJjdXJ2ZSI6LTR9XSxbMiwzLCJcXGV2YWx0IiwwLHsiY3VydmUiOi00fV0sWzIsMSwiXFxxdW90ZWgiLDAseyJjdXJ2ZSI6LTR9XSxbMywyLCJcXHF1b3RldCIsMCx7ImN1cnZlIjotNH1dXQ==
+\[\begin{tikzcd}
+    \PiLang && \PiPlusLang && \PiHatLang && \UFin
+    \arrow["\evalt", from=1-1, to=1-3]
+    \arrow["\evalp", curve={height=-24pt}, from=1-3, to=1-5]
+    \arrow["\evalh", curve={height=-24pt}, from=1-5, to=1-7]
+    \arrow["\quotep", curve={height=-24pt}, from=1-5, to=1-3]
+    \arrow["\quoteh", curve={height=-24pt}, from=1-7, to=1-5]
+  \end{tikzcd}\]
 
-We only present the types and 1-combinators in the language, and the full set of 2-combinators is listed in the appendix
-and the accompanying Agda code.
+In this section, we consider the additive fragment $\PiPlusLang$ of $\PiLang$. This is the language that we interpret to
+$\UFin$ and back, using the tools developed in the previous sections. Further, we go through an intermediate step of the
+language $\PiHatLang$, which is a simplified variant of $\PiPlusLang$ that uses adjacent transpositions for combinators,
+but still preserving all the required structure.
 
-\vc{We can translate types to types, terms to terms, theorem: preserves types, preserves equations.}
+We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\cref*{fig:piplus,fig:pihat} respectively,
+eliding the 2-combinators for brevity. \note{We enforce that there is a unique 2-combinator between compatible
+  1-combinators, by introducing a level 3, which is trivial in the same way as in the case of $\PiLang$, as described
+  before.} They can be found in the appendix and in the accompanying Agda code.
 
-\vc{Define maps quote/eval on 0,1,2-cells, and show commuting diagrams.}
+The translations between the languages are defined separately on types, 1-combinators, and 2-combinators. Following the
+terminology of Normalisation by Evaluation, the translations from the left to the right, going from the syntax towards
+the semantics, are called $\evalt$ and the translations the other way are called $\quotet$.
 
-\subsection{$\PiHatLang$}
+To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
+category. Since the language have higher-dimensional structure, the syntactic categories here are actually
+(locally-strict) weak 2-categories. We define them formally in the appendix, and only state our results here.
 
-We first present a language $\PiHatLang$ based on the theory of symmetric groups developed in~\cref{sec:finite}. The
-types in the language are unary natural numbers, and combinators are allowed to perform adjacent transpositions.
+For each of the $\PiLang$, $\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$,
+$\PiPlusCat$ and $\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can
+show that these are $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are also locally-strict,
+or locally-posetal, because there is at most one 2-cell between compatible 1-cells.
+
+\begin{toappendix}
+  \todo{What is U?}
+  \begin{proposition}
+    We can form  weak 2-category $\PiCat$ with
+    \begin{itemize}
+      \item $\PiLang$ types for 0-cells,
+      \item for $X, Y : U$, a collection of 1-cells $X \iso Y$,
+      \item for $p, q : X \iso Y$, a collection of 2-cells $p \Iso q$.
+    \end{itemize}
+  \end{proposition}
+
+  \begin{proposition}
+    We can form  weak 2-category $\PiPlusCat$ with
+    \begin{itemize}
+      \item $\PiPlusLang$ types for 0-cells,
+      \item for $X, Y : \UPlus$, a collection of 1-cells $X \isop Y$,
+      \item for $p, q : X \isop Y$, a collection of 2-cells $p \Isop q$.
+    \end{itemize}
+  \end{proposition}
+
+  \begin{proposition}
+    We can form  weak 2-category $\PiHatCat$ with
+    \begin{itemize}
+      \item $\PiHatCat$ types for 0-cells,
+      \item for $X, Y : \UHat$, a collection of 1-cells $X \isoh Y$,
+      \item for $p, q : X \isoh Y$, a collection of 2-cells $p \Isoh q$.
+    \end{itemize}
+  \end{proposition}
+\end{toappendix}
+
+We use the $\evalt/\quotet$ translation maps to construct functors between these categories. We only name the maps on
+the 0, 1, and 2-cells -- the coherences hold by definition or by calculation, which is shown in our accompanying Agda
+code. We use these functors to state our results establishing the equivalences between the languages.
+
+\subsection{$\PiLang$ to $\PiPlusLang$}
+
+First, we show how to translate $\PiLang$ programs to $\PiPlusLang$, which is the additive fragment of $\PiLang$. The
+syntax is given in~\cref{fig:piplus}.
+
+\begin{figure}[t]
+  {\scalebox{\scalef}{$
+        \begin{array}{rrcll}
+          \idc :     & A           & \iso & A           & : \idc     \\
+          \identlp : & \zerot + A  & \iso & A           & : \identrp \\
+          \swapp :   & A + B       & \iso & B + A       & : \swapp   \\
+          \assoclp : & A + (B + C) & \iso & (A + B) + C & : \assocrp \\ [1.5ex]
+        \end{array}$}}
+
+  {\scalebox{\scalef}{
+      \Rule{}
+      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : B \iso C}
+      {\jdg{}{}{c_1 \fatsemi c_2 : A \iso C}}
+      {}
+
+      \Rule{}
+      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : C \iso D}
+      {\jdg{}{}{c_1 \oplus c_2 : A + C \iso B + D}}
+      {}
+    }}
+  \caption{$\PiPlusLang$ syntax}
+  \label{fig:piplus}
+\end{figure}
+
+$\PiLang$ has two 0-ary type constructors, and two binary type constructors -- the additive tensor product and the
+multiplicative one. $\PiPlusLang$ has all the type constructors of $\PiLang$ except multiplication. However, we will
+show how to recover the multiplicative structure, by defining multiplication as repeated addition. We encode $\times$ in
+terms of $+$ as follows.
+
+\begin{definition}[$\times : \UPlus \to \UPlus \to \UPlus$]
+  \begin{align*}
+    \zerot \times Y      & \defeq \zerot                      \\
+    \onet \times Y       & \defeq Y                           \\
+    (X_1 + X_2) \times Y & \defeq X_1 \times Y + X_2 \times Y
+  \end{align*}
+\end{definition}
+
+\begin{proposition}
+  There are two symmetric monoidal structures on $\PiPlusCat$, given by $(\zerot, +)$ and $(\onet, \times)$, with
+  $\times$ distributing over $+$, giving it a rig structure, as described in \todo{cite earlier section}.
+\end{proposition}
+\todo{either prove it or cite something}
+
+Using this rig structure, we translate $\PiLang$ to $\PiPlusLang$, constructing a rig functor from $\PiCat$ to
+$\PiPlusCat$.
+
+\begin{definition}
+  \begin{align*}
+    \evalt_{0} & : U \to \UPlus                                             \\
+    \evalt_{1} & : (c : X \iso Y) \to \evalt_{0}(X) \iso \evalt_{0}(Y)      \\
+    \evalt_{2} & : (\alpha : p \Iso q) \to \evalt_{1}(p) \Iso \evalt_{1}(q) \\
+  \end{align*}
+\end{definition}
+
+\subsection{$\PiPlusLang$ to $\PiHatLang$}
+
+Next, we show how to translate $\PiPlusLang$ programs to $\PiHatLang$ and back. $\PiHatLang$ is a simplified variant of
+$\PiPlusLang$, with (unary) natural numbers for types, constructed by $\zero$ and $\suc$, 1-combinators generated by
+adjacent transpositions, and an appropriate set of 2-combinators. The syntax, again omitting 2-combinators, is given
+in~\cref{fig:pihat}. \vc{Something about PROPs.}
+
+We normalise $\PiPlusLang$ to $\PiHatLang$ and back, by giving $\evalh/\quoteh$ maps for 0, 1, and 2-cells.
+
+\begin{definition}
+  \begin{align*}
+    \evalh_{0} & : \UPlus \to \UHat                                                          \\
+    \evalh_{1} & : (c : t_{1} \iso t_{2}) \to \evalh_{0}(t_{1}) \isoh \evalh_{0}(t_{2})      \\
+    \evalh_{2} & : (\alpha : c_{1} \Iso c_{2}) \to \evalh_{1}(c_{1}) \Isoh \evalh_{1}(c_{2}) \\
+  \end{align*}
+\end{definition}
+
+\begin{definition}
+  \begin{align*}
+    \quoteh_{0} & : \UHat \to \UPlus                                                            \\
+    \quoteh_{1} & : (p : X_{1} \isoh X_{2}) \to \quoteh_{0}(X_{1}) \iso \quoteh_{0}(X_{2})      \\
+    \quoteh_{2} & : (\alpha : p_{1} \Isoh p_{2}) \to \quoteh_{1}(p_{1}) \Iso \quoteh_{1}(p_{2}) \\
+  \end{align*}
+\end{definition}
+
+\begin{proposition}
+  $\evalh/\quoteh$ give a symmetric monoidal biequivalence between $\PiPlusCat$ and $\PiHatCat$.
+\end{proposition}
+
+\subsection{$\PiHatLang$ to $\UFin$}
+
+\vc{Define maps quote/eval on 0,,2-cells, and show commuting diagrams.}
 
@@
 
-\subsection{$\PiPlusLang$}
+\subsection{$\PiLang$ circuit normalization}
 
-Now, we present the additive fragment of $\PiLang$, called $\PiPlusLang$.
-
-\begin{figure}[t]
-  {\scalebox{\scalef}{$
-        \begin{array}{rrcll}
-          \idc :     & A           & \iso & A           & : \idc     \\
-          \identlp : & \zerot + A  & \iso & A           & : \identrp \\
-          \swapp :   & A + B       & \iso & B + A       & : \swapp   \\
-          \assoclp : & A + (B + C) & \iso & (A + B) + C & : \assocrp \\ [1.5ex]
-        \end{array}$}}
-
-  {\scalebox{\scalef}{
-      \Rule{}
-      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : B \iso C}
-      {\jdg{}{}{c_1 \fatsemi c_2 : A \iso C}}
-      {}
-
-      \Rule{}
-      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : C \iso D}
-      {\jdg{}{}{c_1 \oplus c_2 : A + C \iso B + D}}
-      {}
-    }}
-  \caption{$\PiPlusLang$ syntax}
-  \label{fig:piplus}
-\end{figure}
-
-\begin{proposition}
-  We can form a weak 2-category $\PiPlusCat$ with
-  \begin{itemize}
-    \item $\PiPlusLang$ types for 0-cells,
-    \item for $X, Y : \UPlus$, a collection of 1-cells $X \iso Y$,
-    \item for $p, q : X \isoh Y$, a collection of 2-cells $p \Iso q$.
-  \end{itemize}
-\end{proposition}
-
-\begin{proposition}
-  There is a symmetric monoidal structure on $\PiPlusCat$, with $\zerot$ for the unit, and $+$ for the tensor.
-\end{proposition}
-
-We normalise $\PiPlusLang$ to $\PiHatLang$ and back, by giving $\evalh/\quoteh$ maps for 0, 1, and 2-cells.
-
-\begin{definition}
-  \begin{align*}
-    \evalh_{0} & : \UPlus \to \UHat                                                          \\
-    \evalh_{1} & : (c : t_{1} \iso t_{2}) \to \evalh_{0}(t_{1}) \isoh \evalh_{0}(t_{2})      \\
-    \evalh_{2} & : (\alpha : c_{1} \Iso c_{2}) \to \evalh_{1}(c_{1}) \Isoh \evalh_{1}(c_{2}) \\
-  \end{align*}
-\end{definition}
-
-\begin{definition}
-  \begin{align*}
-    \quoteh_{0} & : \UHat \to \UPlus                                                            \\
-    \quoteh_{1} & : (p : X_{1} \isoh X_{2}) \to \quoteh_{0}(X_{1}) \iso \quoteh_{0}(X_{2})      \\
-    \quoteh_{2} & : (\alpha : p_{1} \Isoh p_{2}) \to \quoteh_{1}(p_{1}) \Iso \quoteh_{1}(p_{2}) \\
-  \end{align*}
-\end{definition}
-
-\begin{proposition}
-  $\evalh/\quoteh$ give a symmetric monoidal biequivalence between $\PiPlusCat$ and $\PiHatCat$.
-\end{proposition}
-
-\subsection{$\PiLang$}
-
-Finally, We show how to translate $\PiLang$ programs to $\PiHatLang$ programs.
-
-\begin{definition}
-  \begin{align*}
-    \evalt_{0} & : U \to \Nat                                               \\
-    \evalt_{1} & : (c : X \iso Y) \to \evalt_{0}(X) \iso \evalt_{0}(Y)      \\
-    \evalt_{2} & : (\alpha : p \Iso q) \to \evalt_{1}(p) \Iso \evalt_{1}(q) \\
-  \end{align*}
-\end{definition}
-
-To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$ and quote it to $\PiPlusLang$.
+To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$, compute a permutation, and quote back it to
+$\PiPlusLang$.
 

### 2021-06-30 Pi+/paper/ufin.tex
removed bad diagram in ufin

@@
 
-\begin{center}
-\begin{tikzpicture}[scale=1,every node/.style={scale=1}]
-\usetikzlibrary{arrows}
-\usetikzlibrary{shapes}
-\node[ultra thick, draw=blue, ellipse, minimum width=300pt, minimum height=300pt, align=center] {};
-\node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (a) {$\{ 1 , 2 \} \uplus \{ 3 \}$};
-\node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (b) at ([shift=({70pt,-70pt})]a) {$\{ 1 \} \uplus \{ 2, 3 \}$};
-\node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (c) at
-([shift=({-70pt,-70pt})]a) {$\{ 10 \} \uplus \{ 20, 30 \}$};
-\node (d) at ([shift=({50pt,-10pt})]a) {$\cdots$};
-\draw[thick,-] (a) to node[draw=none,right] {$\equiv$} (b);
-\draw[thick,-] (b) to node[draw=none,below] {$\equiv$} (c);
-\draw[thick,-] (a) to node[draw=none,left] {$\equiv$} (c);
-\end{tikzpicture}
-\end{center}
+% \begin{center}
+% \begin{tikzpicture}[scale=1,every node/.style={scale=1}]
+% \usetikzlibrary{arrows}
+% \usetikzlibrary{shapes}
+% \node[ultra thick, draw=blue, ellipse, minimum width=300pt, minimum height=300pt, align=center] {};
+% \node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (a) {$\{ 1 , 2 \} \uplus \{ 3 \}$};
+% \node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (b) at ([shift=({70pt,-70pt})]a) {$\{ 1 \} \uplus \{ 2, 3 \}$};
+% \node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (c) at
+% ([shift=({-70pt,-70pt})]a) {$\{ 10 \} \uplus \{ 20, 30 \}$};
+% \node (d) at ([shift=({50pt,-10pt})]a) {$\cdots$};
+% \draw[thick,-] (a) to node[draw=none,right] {$\equiv$} (b);
+% \draw[thick,-] (b) to node[draw=none,below] {$\equiv$} (c);
+% \draw[thick,-] (a) to node[draw=none,left] {$\equiv$} (c);
+% \end{tikzpicture}
+% \end{center}
 

### 2021-06-30 Pi+/paper/equivalence.tex
Outlining more in section 7

@@
   \begin{proposition}
-    We can form  weak 2-category $\PiCat$ with
+    We can form a weak 2-category $\PiCat$ with
     \begin{itemize}
@@
   \begin{proposition}
-    We can form  weak 2-category $\PiPlusCat$ with
+    We can form a weak 2-category $\PiPlusCat$ with
     \begin{itemize}
@@
   \begin{proposition}
-    We can form  weak 2-category $\PiHatCat$ with
+    We can form a weak 2-category $\PiHatCat$ with
     \begin{itemize}
@@
 $\PiPlusLang$, with (unary) natural numbers for types, constructed by $\zero$ and $\suc$, 1-combinators generated by
-adjacent transpositions, and an appropriate set of 2-combinators. The syntax, again omitting 2-combinators, is given
+adjacent transpositions, and an appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators,
 in~\cref{fig:pihat}. \vc{Something about PROPs.}
-
-We normalise $\PiPlusLang$ to $\PiHatLang$ and back, by giving $\evalh/\quoteh$ maps for 0, 1, and 2-cells.
-
-\begin{definition}
-  \begin{align*}
-    \evalh_{0} & : \UPlus \to \UHat                                                          \\
-    \evalh_{1} & : (c : t_{1} \iso t_{2}) \to \evalh_{0}(t_{1}) \isoh \evalh_{0}(t_{2})      \\
-    \evalh_{2} & : (\alpha : c_{1} \Iso c_{2}) \to \evalh_{1}(c_{1}) \Isoh \evalh_{1}(c_{2}) \\
-  \end{align*}
-\end{definition}
-
-\begin{definition}
-  \begin{align*}
-    \quoteh_{0} & : \UHat \to \UPlus                                                            \\
-    \quoteh_{1} & : (p : X_{1} \isoh X_{2}) \to \quoteh_{0}(X_{1}) \iso \quoteh_{0}(X_{2})      \\
-    \quoteh_{2} & : (\alpha : p_{1} \Isoh p_{2}) \to \quoteh_{1}(p_{1}) \Iso \quoteh_{1}(p_{2}) \\
-  \end{align*}
-\end{definition}
-
-\begin{proposition}
-  $\evalh/\quoteh$ give a symmetric monoidal biequivalence between $\PiPlusCat$ and $\PiHatCat$.
-\end{proposition}
-
-\subsection{$\PiHatLang$ to $\UFin$}
-
-\vc{Define maps quote/eval on 0,,2-cells, and show commuting diagrams.}
 
@@
 
+As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers.
+However, we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal strucuture.
+
 \begin{proposition}
-  We can form a weak 2-category $\PiHatCat$ with
-  \begin{itemize}
-    \item natural numbers for 0-cells,
-    \item for $n, m : \Nat$, a collection of 1-cells $n \isoh m$,
-    \item for $p, q : n \isoh m$, a collection of 2-cells $p \Isoh q$.
-  \end{itemize}
+  $\PiHatCat$ has a symmetric monoidal structure, with the unit given by 0 and the tensor product given by natural
+  number addition.
 \end{proposition}
 
+Using this symmetric monoidal structure, we translate from $\PiPlusLang$ to $\PiHatLang$.
+
+\begin{definition}
+  \begin{align*}
+    \evalh_{0} & : \UPlus \to \UHat                                                          \\
+    \evalh_{1} & : (c : t_{1} \iso t_{2}) \to \evalh_{0}(t_{1}) \isoh \evalh_{0}(t_{2})      \\
+    \evalh_{2} & : (\alpha : c_{1} \Iso c_{2}) \to \evalh_{1}(c_{1}) \Isoh \evalh_{1}(c_{2}) \\
+  \end{align*}
+\end{definition}
+
+To go back from $\PiHatLang$ to $\PiPlusLang$, we turn a natural number into a $\PiPlusLang$ type, using right-biased
+addition, that is, the natural number $n$ gets mapped to the type $\onet + (\onet + (\onet + \ldots + \zerot))$. The
+non-trivial part is coherently generating a $\PiPlusLang$ combinator from $\PiHatLang$ combinators, using adjacent swaps
+to generate big swaps.
+
+\begin{definition}
+  \begin{align*}
+    \quoteh_{0} & : \UHat \to \UPlus                                                            \\
+    \quoteh_{1} & : (p : X_{1} \isoh X_{2}) \to \quoteh_{0}(X_{1}) \iso \quoteh_{0}(X_{2})      \\
+    \quoteh_{2} & : (\alpha : p_{1} \Isoh p_{2}) \to \quoteh_{1}(p_{1}) \Iso \quoteh_{1}(p_{2}) \\
+  \end{align*}
+\end{definition}
+
 \begin{proposition}
-  There is a symmetric monoidal structure on $\PiHatCat$, with $0$ for the unit, and addition for the tensor.
+  $\evalh/\quoteh$ give a symmetric monoidal biequivalence between $\PiPlusCat$ and $\PiHatCat$.
 \end{proposition}
 
-Then we establish the completeness of $\PiHatLang$ with respect to $\UFin$. We define $\evalt$ and $\quotet$ for 0, 1,
-and 2-cells.
+\subsection{$\PiHatLang$ to $\UFin$}
+
+Finally, we show how to interpret $\PiHatLang$ to $\UFin$, and go back from $\UFin$ to $\PiHatLang$.
+
+\todo{Sn/Lehmer/Aut(Fin n) diagram here?}
+
+Types in $\PiHatLang$ are interpreted as 0-cells in $\UFin$, that is, a natural number $n$ is mapped to $\Fin[n]$. The
+1-combinators in $\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops $\Aut[\Fin[n]]$, or permutations, in
+each connected component. Finally, 2-combinators are mapped to 2-paths between loops in $\UFin$.
 
@@
 
+0-cells in $\UFin$ are mapped to their cardinalities in $\PiHatLang$, 1-loops are decoded to words in $\Sn$ to generate
+a sequence of adjacent transpositions, producing a 1-combinator in $\PiHatLang$. 2-paths are quoted back to
+2-combinators in $\PiHatLang$.
+
 \begin{definition}
@@
 
+$\UFin$ as a 1-groupoid, is a (locally strict) weak $(2,0)$-category.
+
 \begin{proposition}
@@
 
-To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$, compute a permutation, and quote back it to
-$\PiPlusLang$.
+To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$, compute a permutation, and quote it back to
+$\PiPlusLang$. Note that the normalisation happens in going from $\PiHatLang$ to $\UFin$ to $\PiHatLang$.
 

### 2021-06-30 Pi+/paper/ufin.tex
More in UFin

@@
 
+\[
+  \begin{tikzcd}
+    F_{0}
+    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
+    & F_{1}
+    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
+    & F_{2}
+    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
+    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=8em, in=125, out=55]
+    & \ldots
+    & F_{n}
+    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
+    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=8em, in=125, out=55]
+    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=12em, in=135, out=45]
+    & \ldots
+  \end{tikzcd}
+\]
+
 \subsection{Rig structure}~\label{subsec:rig}
@@
 
-We lift these equivalences to $\UFin$ giving it the (additive) symmetric monoidal structure $(I, \oplus)$, with natural
-isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the symmetry isomorphism $\mathcal{B}_{X,Y}$. Note that
-types in $\UFin$ are $\hSet$s since they're merely equivalent to $\Fin[n]$ for some $n : \Nat$.
+\begin{proposition}
+  For any $n, m : \Nat$,
+  \begin{align*}
+    \Fin[1]                & \eqv \top        \\
+    \Fin[n] \times \Fin[m] & \eqv \Fin[n * m] \\
+  \end{align*}
+  and for any types $X, Y, Z$,
+  \begin{align*}
+    \top \times X         & \eqv X                                \\
+    X \times \top         & \eqv X                                \\
+    (X \times Y) \times Z & \eqv X \times (Y \times Z)            \\
+    X \times Y            & \eqv Y \times X                       \\
+    X \times \bot         & \eqv \bot                             \\
+    X \times (Y \sqcup Z) & \eqv (X \times Y) \sqcup (X \times Z)
+  \end{align*}
+\end{proposition}
+
+We lift these equivalences to $\UFin$ giving it the additive and multiplicative symmetric monoidal structures,
+$(O, \oplus)$ and $(I, \otimes)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$,
+and the symmetry isomorphism $\mathcal{B}_{X,Y}$. Note that types in $\UFin$ are $\hSet$s since they're merely
+equivalent to $\Fin[n]$ for some $n : \Nat$.
+
+\todo{combine both symmetric monoidal structures}
 

### 2021-06-30 Pi+/paper/ufin.tex
Write a segue into section 4

@@
 
-In the previous section, we have picked a particular encoding of finite types: as sequences of elements in some
-canonical order.
+In the previous section, we gave a sound denotational semantics for $\PiLang$ using the groupoid of finite sets and
+bijections $\BFin$. Types are interpreted as objects in $\BFin$, and 1-combinators are interpreted as isomorphisms in
+$\BFin$. We also showed that if there is a 2-combinator between two 1-combinators, then their interpretations are
+extensionally equal in $\BFin$.
 
-This is not a ``free'' presentation. Some properties of the encoding might pollute our semantics. We want to make that
-any theorems / properties of the semantics we prove apply to the source language (under various interpretations).
+However, we want to go a step further, and claim completeness for this semantics, that is, we want to quote back from
+the semantics to the syntax. Given a morphism in $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two
+equal morphisms in $\BFin$, we want to produce a 2-combinator witnessing the equality of the quoted 1-combinators.
 
-This motivates us to use a different semantic representation: a univalent universe of finite types which contains the
-finite types and nothing but the finite types and is defined such as alternative representations of finite types are
-identified. This guarantees that we can get a proper, sound and complete, semantics for our language.
+The groupoid $\BFin$ that we presented, is strict, that is, two equal permutations are extensionally identified, without
+an explicit witness for the equality. Our first step is to weaken this groupoid $\BFin$ into a weak 2-category. In a
+weak category, equational axioms of the category are replaced by higher invertible cells. For example, there are higher
+invertible cells for the left and right units of composition and associativity. The higher cells have their own
+coherences as well, the 2-cells can be composed with 1-cells by whiskering, and so on.
+
+HoTT provides a rich internal language for working with weak higher groupoids, using the ``Types are weak
+$\infty$-groupoids'' correspondence. We will make this precise by observing that a 1-groupoid in HoTT gives a locally
+strict $(2,0)$-category, using the identity type for the higher cells. The groupoid $\BFin$ can be written in HoTT as a
+type $\UFin$, which is a 1-groupoid.
+
+Secondly, we would like to like to give a syntactic presentation for this groupoid, by describing exactly how the
+1-cells and 2-cells are generated. Note that, in $\BFin$, the set of bijections on a finite set of cardinality $n$ has a
+group structure, the identity function is the neutral element, group multiplication is given by function composition,
+and the group inverse is given by inverting the bijection. This is the automorphism group of the finite set, which is
+also known to be the finite symmetric group $\Sn[n]$.
+
+Groups can be defined by listing the carrier set and the operations of the group, or they can be described using a group
+presentation, which gives the generators and relations of a group. The last step of our technical development is to use
+a group presentation for $\Sn$ to give a presentation for the higher cells of the groupoid $\UFin$. Using this
+characterisation, we will obtain our completeness result.
+
+% In the previous section, we have picked a particular encoding of finite types: as sequences of elements in some
+% canonical order.
+
+% This is not a ``free'' presentation. Some properties of the encoding might pollute our semantics. We want to make that
+% any theorems / properties of the semantics we prove apply to the source language (under various interpretations).
+
+% This motivates us to use a different semantic representation: a univalent universe of finite types which contains the
+% finite types and nothing but the finite types and is defined such as alternative representations of finite types are
+% identified. This guarantees that we can get a proper, sound and complete, semantics for our language.
 

### 2021-06-30 Pi+/paper/reversible1.tex
merge

@@
 \label{sec:reversibleone}
+\label{sec:reversibletwo}
+\label{langeqeq}
 
@@
 % the finite set, we can define a denotational semantics which constructs the bijection.
-% \note{\(\Fin[n] \to A \eqv \type{Vec_{n}}(A)\)}
+% \note{\(\Fin[n] \to A \eqv <\type{Vec_{n}}(A)\)}
+
+An alternative point of view for the semantics in Sec.~\ref{reversibleone} is to consider it  as expressed in the
+category of finite sets and functions, $\SetFin$, which is the category freely generated by finite coproduct completion
+of the terminal category. Objects of $\SetFin$ can be identified with sets of fixed cardinality, that is,
+$\Fin[n] \defeq \Set{0,1,\ldots,n-1}$. $\SetFin$ has finite coproducts and products, which lets us interpret the types
+of $\PiLang$. Combinators are intepreted as morphisms in $\SetFin$, but we have to restrict to invertible morphisms,
+that is, isomorphisms. This gives the \emph{groupoid} of finite sets and bijections,
+$\BFin \defeq \mathsf{core}(\SetFin)$. The isomorphisms satisfied by coproducts and products in $\SetFin$ lift to
+$\BFin$, but they're no longer categorical coproducts and products. They give two symmetric monoidal tensor products on
+$\BFin$, the additive and multiplicative ones, with the multiplicative tensor distributing over the additive tensor.
+
+We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
+compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
+$\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:
+
+\[\begin{array}{rcll}
+\sem{A} &=& (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
+\end{array}\]
+
+\noindent The denotation $\sem{c}$ of a combinator $c : A \isot B$ is a path between $\sem{A}$ and $\sem{B}$. If the
+size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
+$(\mathsf{Fin}~n, [ n , \mathsf{refl} ])$. This path is directly constructed using $\mathit{ap}$ and the fact that $m=n$
+since combinators are always between types of the same size.
+
+\noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \Iso q$, the denotation
+$\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.
+
+\note{We use the rig structure of $\UFin$ in~\cref{subsec:rig} to interpret $\PiLang$.}
+
+We need a formal definition of normal form (canonical form)
+
+In the previous section, we examined equivalences between conventional data structures, i.e., structured trees of
+values. We now consider a richer but foundational notion of data: programs themselves. Indeed, universal computation
+models crucially rely on the fact that \emph{programs are (or can be encoded as) data}, e.g., a Turing machine can be
+encoded as a string that another Turing machine (or even the same machine) can manipulate. Similarly, first-class
+functions are the \emph{only} values in the $\lambda$-calculus.  In our setting, we ask whether the programs developed
+in the previous section can themselves be subject to (higher-level) equivalences?
+
+We will explain the ideas using two small exaamples. Consider the following two programs mapping between the types
+$A + B$ and $C+D$:
+
+\begin{center}
+\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+  \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
+  \draw[fill] (-2,-1.5) circle [radius=0.025];
+  \node[below] at (-2.1,-1.5) {$A$};
+  \node[below] at (-2.1,-1.9) {$+$};
+  \draw[fill] (-2,-2.5) circle [radius=0.025];
+  \node[below] at (-2.1,-2.5) {$B$};
+
+  \draw[fill] (6.5,-1.5) circle [radius=0.025];
+  \node[below] at (6.7,-1.5) {$C$};
+  \node[below] at (6.7,-1.9) {$+$};
+  \draw[fill] (6.5,-2.5) circle [radius=0.025];
+  \node[below] at (6.7,-2.5) {$D$};
+
+  \draw[<-] (-2,-1.5) to[bend left] (1,0.5) ;
+  \draw[<-] (-2,-2.5) to[bend left] (1,-0.5) ;
+  \draw[->] (3.5,0.5) to[bend left] (6.5,-1.45) ;
+  \draw[->] (3.5,-0.5) to[bend left] (6.5,-2.45) ;
+
+  \draw[<-] (-2,-1.5) to[bend right] (1,-3.5) ;
+  \draw[<-] (-2,-2.5) to[bend right] (1,-4.5) ;
+  \draw[->] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
+  \draw[->] (3.5,-4.5) to[bend right] (6.5,-2.55) ;
+
+  \draw     (2,0.5)  -- (2.5,0.5)  ;
+  \draw     (2,-0.5) -- (2.5,-0.5) ;
+
+  \draw     (2.5,0.5)  -- (3.5,-0.5)  ;
+  \draw     (2.5,-0.5) -- (3.5,0.5) ;
+
+  \draw     (1,-3.5)  -- (2,-4.5)    ;
+  \draw     (1,-4.5) -- (2,-3.5)   ;
+
+  \draw     (2,-3.5)  -- (2.5,-3.5)    ;
+  \draw     (2,-4.5) -- (2.5,-4.5)   ;
+
+  \path (1.5,0.5) node (tc1) [func] {$c_1$};
+  \path (1.5,-0.5) node (tc2) [func] {$c_2$};
+  \path (3,-4.5) node (bc1) [func] {$c_1$};
+  \path (3,-3.5) node (bc2) [func] {$c_2$};
+\end{tikzpicture}
+\end{center}
+The top path is the $\Pi$ program $(c_1~\oplus~c_2)~\odot~\swapp$ which acts on the type $A$ by $c_1$, acts on the type
+$B$ by $c_2$, and acts on the resulting value by a twist that exchanges the two injections into the sum type. The bottom
+path performs the twist first and then acts on the type $A$ by $c_1$ and on the type $B$ by $c_2$ as before. One could
+imagine the paths are physical \emph{elastic} wires in $3$ space, where the programs $c_1$ and $c_2$ as arbitrary
+deformations on these wires, and the twists do not touch but are in fact well-separated. Then, holding the points $A$,
+$B$, $C$, and $D$ fixed, it is possible to imagine sliding $c_1$ and $c_2$ from the top wire rightward past the twist,
+and then using the elasticity of the wires, pull the twist back to line up with that of the bottom --- thus making both
+parts of the diagram identical.  Each of these moves can be undone (reversed), and doing so would take the bottom part
+of the diagram into the top part.  In other words, there exists an equivalence of the program
+$(c_1~\oplus~c_2)~\odot~\swapp$ to the program $\swapp \odot (c_2~\oplus~c_1)$. We can also show that this means that,
+as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and $\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not all
+programs between the same types can be deformed into one another. The simplest example of inequivalent deformations are
+the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.
+
+As another example, consider consider a circuit that takes an input type consisting of three values \Tree [ {\small a} [
+{\small b} {\small c} ] ]~ and swaps the leftmost value with the rightmost value to produce \Tree [ {\small c} [ {\small
+  b} {\small a} ] ]~.  We can implement two such circuits using our Agda library for $\Pi$:
+
+\begin{code}%
+\>[0]\AgdaFunction{swap{-}fl1}\AgdaSpace{}%
+\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaSymbol{\{}\AgdaBound{a}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSpace{}%
+\AgdaSymbol{:}\AgdaSpace{}%
+\AgdaDatatype{U}\AgdaSymbol{\}}\AgdaSpace{}%
+\AgdaSymbol{→}\AgdaSpace{}%
+\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{a}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaDatatype{⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{c}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{PLUS}\AgdaSpace{}%
+\AgdaBound{b}\AgdaSpace{}%
+\AgdaBound{a}\AgdaSymbol{)}\<%
+\\
+\>[0]\AgdaFunction{swap{-}fl1}\AgdaSpace{}%
+\AgdaSymbol{=}\AgdaSpace{}%
+\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\AgdaSpace{}%
+\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\<%
+\\
+%
+\\[\AgdaEmptyExtraSkip]%
+\>[0]\AgdaFunction{swap{-}fl2}\AgdaSpace{}%
+\AgdaSymbol{=}%
+\>[52I]\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+\>[.]\<[52I]%
+\>[11]\AgdaInductiveConstructor{assocl₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaSymbol{(}\AgdaInductiveConstructor{swap₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{id⟷}\AgdaSymbol{)}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaInductiveConstructor{assocr₊}\AgdaSpace{}%
+\AgdaInductiveConstructor{◎}\<%
+\\
+%
+\>[11]\AgdaSymbol{(}\AgdaInductiveConstructor{id⟷}\AgdaSpace{}%
+\AgdaInductiveConstructor{⊕}\AgdaSpace{}%
+\AgdaInductiveConstructor{swap₊}\AgdaSymbol{)}\<%
+\end{code}
+
+\noindent The first implementation rewrites the incoming values as follows:
+\[
+\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+\Tree [ [ {\small a} {\small b} ] {\small c} ] ~\to~
+\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+\]
+\noindent
+The second implementation rewrites the incoming values as follows:
+\[
+\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+\Tree [ {\small a} [ {\small c} {\small b} ] ] ~\to~
+\Tree [ [ {\small a} {\small c} ] {\small b} ] ~\to~
+\Tree [ [ {\small c} {\small a} ] {\small b} ] ~\to~
+\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+\]
+\noindent The two circuits are extensionally equal. Using the level-2
+isomorphisms we can \emph{explicitly} construct a sequence of
+rewriting steps that transforms the second circuit to the first.
+
+Recalling that the $\lambda$-calculus arises as the internal language of Cartesian Closed Categories
+(Elliott~\cite{Elliott-2017} gives a particularly readable account of this), we can think of $\Pi$ in similar terms, but
+for symmetric Rig Groupoids instead. For example, we can ask what does the equivalence above represent? It is actually a
+``linear'' representation of a 2-categorial commutative diagram! In fact, it is a painfully verbose version thereof, as
+it includes many \emph{refocusing} steps because our language does not build associativity into its syntax. Categorical
+diagrams usually do.  Thus if we rewrite the example in diagrammatic form, eliding all uses of associativity, but
+keeping explicit uses of identity transformations, we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents
+
+\vspace*{3mm}
+\begin{tikzcd}[column sep=normal, row sep=normal]
+ && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
+ & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
+a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
+  \arrow [dr, "\assoclp"]
+  \arrow [ddr, swap, "\assoclp"]
+    & (a+b)+c \arrow [r, "\swapp"] &
+    c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}]
+    & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
+ & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
+ & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
+             \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] &&
+    c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
+ && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
+ \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
+ \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
+\end{tikzcd}
 

### 2021-07-01 Pi+/paper/reversible1.tex
clean up a bit early parts

@@
 %%%%%%%%%
-\subsection{Denotational Semantics (Outline)}
+\subsection{Informal Semantics and Examples}
+\label{sec:langRev-examples}
+\label{examples}
+
+%% \note{Revisit to pick examples relevant to the rest of the paper; also unify notation}
+
+In order to understand the examples of Sec.~\ref{sec:informal}, we review how to represent collections of booleans and
+conditionals in $\Pi$. We use the type $\onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing
+\textsf{false} and $\inrv{\Acon{tt}}$ representing $\textsf{true}$. Boolean negation is straightforward to define using
+the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values.
+For example, we can express a 3-bit word reversal operation as follows:
+
+$\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
+
+$\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
+
+\noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
+\[\begin{array}{rlr}
+ & (v_1, (v_2, v_3)) \\
+ \swapt & ((v_2, v_3), v_1) \\
+ \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
+ \assocrt & (v_3, (v_2, v_1)) \\
+ \end{array}\]
+%subcode source isomorphisms.tex:979
+
+Even though \ensuremath{\Pi } lacks conditional expressions, they are expressible using the distributivity and factoring
+laws as shown below:
+
+$\Afun{ifc}_{c_1,c_2} : \mathbb{2} \times A \iso \mathbb{2} \times B$
+
+$\Afun{ifc}_{c_1,c_2} = \dist \fatsemi ((\idc  \otimes\  c_1) \oplus (\idc \otimes\  c_2)) \fatsemi \factor$
+
+% \inkscape{if-c1-c2.pdf}
+
+\noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
+which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the \ensuremath{\mathit{left}} branch,
+which corresponds to the case when the boolean is \ensuremath{\mathit{false}}, the combinator~\ensuremath{c_1} is
+applied to the value of type~\ensuremath{A}. The right branch which corresponds to the boolean being
+\ensuremath{\mathit{true}} passes the value of type $A$ through the combinator \ensuremath{c_2}.  The inverse of
+\ensuremath{\dist}, namely \ensuremath{\factor} is applied to get the final result. The full expressiveness of $\Pi$ is
+demonstrated by an implementation of $\Afun{cnot}$ and $\Afun{toffoli}$ which are universal for reversible
+computation~\cite{Toffoli:1980}. The $\Afun{cnot}$ gates takes two input bits: a control bit and a target bit, and
+negates the target bit if the control bit is set. It is essentially a one-branch version of $\Afun{cif}$ above. The
+Toffoli gate generalizes the idea to a gate that takes 3 input bits: 2 control bits and one target bit: boolean inputs:
+it negates the target bit if both control bits are set.
+
+%%%%%%%%%
+\subsection{Denotational Semantics}
 
@@
 
-%%%%%%%%%
-\subsection{Examples}
-\label{sec:langRev-examples}
-\label{examples}
-
-%% \note{Revisit to pick examples relevant to the rest of the paper; also unify notation}
-
-In order to understand the examples of Sec.~\ref{sec:informal}, we review how to represent collections of booleans and
-conditionals in $\Pi$. We use the type $\onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing
-\textsf{false} and $\inrv{\Acon{tt}}$ representing $\textsf{true}$. Boolean negation is straightforward to define using
-the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values.
-For example, we can express a 3-bit word reversal operation as follows:
-
-$\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
-
-$\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
-
-\noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
-\[\begin{array}{rlr}
- & (v_1, (v_2, v_3)) \\
- \swapt & ((v_2, v_3), v_1) \\
- \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
- \assocrt & (v_3, (v_2, v_1)) \\
- \end{array}\]
-%subcode source isomorphisms.tex:979
-
-Even though \ensuremath{\Pi } lacks conditional expressions, they are expressible using the distributivity and factoring
-laws as shown below:
-
-$\Afun{ifc}_{c_1,c_2} : \mathbb{2} \times A \iso \mathbb{2} \times B$
-
-$\Afun{ifc}_{c_1,c_2} = \dist \fatsemi ((\idc  \otimes\  c_1) \oplus (\idc \otimes\  c_2)) \fatsemi \factor$
-
-% \inkscape{if-c1-c2.pdf}
-
-\noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
-which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the \ensuremath{\mathit{left}} branch,
-which corresponds to the case when the boolean is \ensuremath{\mathit{false}}, the combinator~\ensuremath{c_1} is
-applied to the value of type~\ensuremath{A}. The right branch which corresponds to the boolean being
-\ensuremath{\mathit{true}} passes the value of type $A$ through the combinator \ensuremath{c_2}.  The inverse of
-\ensuremath{\dist}, namely \ensuremath{\factor} is applied to get the final result. The full expressiveness of $\Pi$ is
-demonstrated by the following implementation of $\Afun{cnot}$ and $\Afun{toffoli}$ which are universal for reversible
-computation~\cite{Toffoli:1980}. The $\Afun{cnot}$ gates takes two input bits: a control bit and a target bit, and
-negates the target bit if the control bit is set. It is essentially a one-branch version of $\Afun{cif}$ above. The
-Toffoli gate generalizes the idea to a gate that takes 3 input bits: 2 control bits and one target bit: boolean inputs:
-it negates the target bit if both control bits are set.
-
 %%% Local Variables:

### 2021-07-01 Pi+/paper/ufin.tex

@@
 
-\todo{Diagram.}
-
 A type family can also be seen as a fibration, from the topological viewpoint. For a type family $P : A \to \UU$ and a
@@
 where $\tr{p}{u}$ is shorthand for $\transport{P,p}(u)$.
+
+\begin{center}
+  \begin{tikzpicture}[yscale=.5,xscale=2]
+    \draw (0,0) arc (-90:170:8ex) node[anchor=south east] {$A$} arc (170:270:8ex);
+    \draw (0,6) arc (-90:170:10ex) node[anchor=south east] {$\dsum{x:A}{P(x)}$} arc (170:270:10ex);
+    \draw[->] (0,5.8) -- node[auto] {$\fst$} (0,3.2);
+    \node[circle,fill,inner sep=1pt,label=left:{$x$}] (b1) at (-.5,1.4) {};
+    \node[circle,fill,inner sep=1pt,label=right:{$y$}] (b2) at (.5,1.4) {};
+    \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto,swap] {$p$} (b2);
+    \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
+    \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\tr{p}{u}}$}] (b2) at (.5,7.2) {};
+    \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
+  \end{tikzpicture}
+\end{center}
 

### 2021-07-01 Pi+/paper/reversible1.tex
denotation 1

--- 2021-06-30 Pi+/paper/reversible1.tex
@@
 %%%%%%%%%
-\subsection{Denotational Semantics (Outline)}
-
-\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be
-specified by mapping each type to a finite set and each combinator to a bijection between finite sets, that is, a
-permutation. In this section, we outline a simple semantics expressed in the metalanguage of sets and functions. This
-semantics will be formalized and extended to the metalanguage of homotopy type theory in Sec.~\ref{sec:reversibletwo}.
-
-As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
-we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
-set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
-the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
-$\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
-with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
-\[\begin{array}{rcl}
-    \denot{\onet + (\onet + \onet)} &=& \{ 0,1,2 \} \\
-    \denot{(\onet + \onet) + \onet} &=& \{ 0,1,2 \} \\
-    \\
-    \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (0~1~2) \\
-    \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (2~0~1)
-\end{array}\]
-where we have used the one-line notation for permutations with $(a~b~c)$ representing the
-permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
-enumeration of the elements of each type:
-\[\begin{array}{rcl}
-    \mathit{enum}(\zerot) &=& [~] \\
-    \mathit{enum}(\onet) &=& [ ~\Acon{tt}~ ] \\
-    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
-    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
-\end{array}\]
-\noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
-a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
-operation that appends all the subsequences in a sequence.
-
-Using the definition, we have:
-\[\begin{array}{rcl}
-    \mathit{enum}(\onet + (\onet + \onet)) &=& [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
-    \mathit{enum}((\onet + \onet) + \onet) &=& [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
-\end{array}\]
-Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
-$\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
-$\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
-$\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
-$\inrv{\Acon{tt}}$.
-
-\begin{center}
-  \input{assoc-perm.tikz}
-  \qquad
-  \input{swap-perm.tikz}
-\end{center}
-
-% \[\begin{array}{l}
-%     \Tree[ {\small a} [ {\small b} {\small c} ] ] \\
-%     2 \quad 1 \quad 0 \\
-%     0 \quad 2 \quad 1 \\
-%     \Tree[ [ {\small b}  {\small c} ] {\small a} ]
-% \end{array}\]
-
-% Such functions are finitely supported, that is, their outputs can be tabulated using the canonical ordering on
-% $\Fin[n]$. For bijections, this gives a listed permutation. By observing the action of the combinators on the values of
-% the finite set, we can define a denotational semantics which constructs the bijection.
-% \note{\(\Fin[n] \to A \eqv \type{Vec_{n}}(A)\)}
-
-%%%%%%%%%
-\subsection{Examples}
+\subsection{Informal Semantics and Examples}
 \label{sec:langRev-examples}
@@
 \ensuremath{\dist}, namely \ensuremath{\factor} is applied to get the final result. The full expressiveness of $\Pi$ is
-demonstrated by the following implementation of $\Afun{cnot}$ and $\Afun{toffoli}$ which are universal for reversible
+demonstrated by an implementation of $\Afun{cnot}$ and $\Afun{toffoli}$ which are universal for reversible
 computation~\cite{Toffoli:1980}. The $\Afun{cnot}$ gates takes two input bits: a control bit and a target bit, and
@@
 it negates the target bit if both control bits are set.
+
+%%%%%%%%%
+\subsection{Denotational Semantics}
+
+\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be
+specified by mapping each type to a finite set and each combinator to a bijection between finite sets, that is, a
+permutation. In this section, we outline a simple semantics expressed in the metalanguage of sets and functions. This
+semantics will be extended to the metalanguage of homotopy type theory in the remainder of the paper.
+
+The denotation of types is straightforward:
+\[\begin{array}{rcl}
+\denot{\zerot} &=& \emptyset \\
+\denot{\onet} &=& \{ \Acon{tt} \} \\
+\denot{A + B} &=& \denot{A} \uplus \denot{B} \\
+\denot{A \times B} &=& \denot{A} \times \denot{B}
+\end{array}\]
+where $\emptyset$ is the empty set, $\uplus$ is the disjoint union of sets, and $\times$ is the cartesian product of
+sets. The denotation of a combinator $c : A \isot B$ is a (bijective) function mapping $\denot{A}$ to $\denot{B}$:
+\[\begin{array}{rcl}
+\denot{\identlp} (\inlv{v}) &=& v \\
+\denot{\identrp} v &=& \inlv{v} \\
+\denot{\swapp} (\inlv{v}) &=& \inrv{v} \\
+\denot{\swapp} (\inrv{v}) &=& \inlv{v} \\
+\denot{\assoclp} (\inlv{v})  &=& \inlv{(\inlv{v})} \\
+\denot{\assoclp} (\inrv{(\inlv{v})}) &=& \inlv{(\inrv{v})} \\
+\denot{\assoclp} (\inrv{(\inrv{v})}) &=& \inrv{v} \\
+\denot{\assocrp} (\inlv{(\inlv{v})}) &=& \inlv{v} \\
+\denot{\assocrp} (\inlv{(\inrv{v})}) &=& \inrv{(\inlv{v})} \\
+\denot{\assocrp} (\inrv{v})  &=& \inrv{(\inrv{v})} \\
+\denot{\identlp} (\Acon{tt} , v)  &=& v \\
+\denot{\identrp} v &=& (\Acon{tt} , v) \\
+\denot{\swapt}  (v_1 , v_2) &=& (v_2 , v_1) \\
+\denot{\assoclt} (v_1 , (v_2 , v_3)) &=& ((v_1 , v_2) , v_3) \\
+\denot{\assocrt} ((v_1 , v_2) , v_3) &=& (v_1 , (v_2 , v_3)) \\
+\denot{\dist} (\inlv{v_1} , v_3)  &=& \inlv{(v_1 , v_3)} \\
+\denot{\dist} (\inrv{v_2 , v_3})  &=& \inrv{(v_2 , v_3)} \\
+\denot{\factor}  (\inlv{(v_1 , v_3)}) &=& (\inlv{v_1} , v_3) \\
+\denot{\factor}  (\inrv{(v_2 , v_3)}) &=& (\inrv{v_2} , v_3) \\
+\denot{\idc} v &=& v \\
+\denot{(c_1 \fatsemi c_2)} v &=& (\denot{c_2} \circ \denot{c_1}) v \\
+\denot{(c_1 \oplus c_2)} (\inlv{v}) &=& \inlv{(\denot{c_1}~v)} \\
+\denot{(c_1 \oplus c_2)} (\inrv{v}) &=& \inrv{(\denot{c_2}~v)} \\
+\denot{(c_1 \otimes c_2)} (v_1 , v_2) &=& (\denot{c_1} v_1 , \denot{c_2} v_2)
+  \end{array}\]
+
+% \begin{center}
+% \begin{tikzpicture}[scale=0.7,every node/.style={scale=0.8}]
+%   \draw[>=latex,<->,double,red,thick] (2.25,-1.2) -- (2.25,-2.9) ;
+%   \draw[fill] (-2,-1.5) circle [radius=0.025];
+%   \node[below] at (-2.1,-1.5) {$A$};
+%   \node[below] at (-2.1,-1.9) {$+$};
+%   \draw[fill] (-2,-2.5) circle [radius=0.025];
+%   \node[below] at (-2.1,-2.5) {$B$};
+
+%   \draw[fill] (6.5,-1.5) circle [radius=0.025];
+%   \node[below] at (6.7,-1.5) {$C$};
+%   \node[below] at (6.7,-1.9) {$+$};
+%   \draw[fill] (6.5,-2.5) circle [radius=0.025];
+%   \node[below] at (6.7,-2.5) {$D$};
+
+%   \draw[<-] (-2,-1.5) to[bend left] (1,0.5) ;
+%   \draw[<-] (-2,-2.5) to[bend left] (1,-0.5) ;
+%   \draw[->] (3.5,0.5) to[bend left] (6.5,-1.45) ;
+%   \draw[->] (3.5,-0.5) to[bend left] (6.5,-2.45) ;
+
+%   \draw[<-] (-2,-1.5) to[bend right] (1,-3.5) ;
+%   \draw[<-] (-2,-2.5) to[bend right] (1,-4.5) ;
+%   \draw[->] (3.5,-3.5) to[bend right] (6.5,-1.55) ;
+%   \draw[->] (3.5,-4.5) to[bend right] (6.5,-2.55) ;
+
+%   \draw     (2,0.5)  -- (2.5,0.5)  ;
+%   \draw     (2,-0.5) -- (2.5,-0.5) ;
+
+%   \draw     (2.5,0.5)  -- (3.5,-0.5)  ;
+%   \draw     (2.5,-0.5) -- (3.5,0.5) ;
+
+%   \draw     (1,-3.5)  -- (2,-4.5)    ;
+%   \draw     (1,-4.5) -- (2,-3.5)   ;
+
+%   \draw     (2,-3.5)  -- (2.5,-3.5)    ;
+%   \draw     (2,-4.5) -- (2.5,-4.5)   ;
+
+%   \path (1.5,0.5) node (tc1) [func] {$c_1$};
+%   \path (1.5,-0.5) node (tc2) [func] {$c_2$};
+%   \path (3,-4.5) node (bc1) [func] {$c_1$};
+%   \path (3,-3.5) node (bc2) [func] {$c_2$};
+% \end{tikzpicture}
+% \end{center}
+% The top path is the $\Pi$ program $(c_1~\oplus~c_2)~\odot~\swapp$ which acts on the type $A$ by $c_1$, acts on the type
+% $B$ by $c_2$, and acts on the resulting value by a twist that exchanges the two injections into the sum type. The bottom
+% path performs the twist first and then acts on the type $A$ by $c_1$ and on the type $B$ by $c_2$ as before. One could
+% imagine the paths are physical \emph{elastic} wires in $3$ space, where the programs $c_1$ and $c_2$ as arbitrary
+% deformations on these wires, and the twists do not touch but are in fact well-separated. Then, holding the points $A$,
+% $B$, $C$, and $D$ fixed, it is possible to imagine sliding $c_1$ and $c_2$ from the top wire rightward past the twist,
+% and then using the elasticity of the wires, pull the twist back to line up with that of the bottom --- thus making both
+% parts of the diagram identical.  Each of these moves can be undone (reversed), and doing so would take the bottom part
+% of the diagram into the top part.  In other words, there exists an equivalence of the program
+% $(c_1~\oplus~c_2)~\odot~\swapp$ to the program $\swapp \odot (c_2~\oplus~c_1)$. We can also show that this means that,
+% as permutations, $(c_1~\oplus~c_2)~\odot~\swapp$ and $\swapp \odot (c_2~\oplus~c_1)$ are equal. And, of course, not all
+% programs between the same types can be deformed into one another. The simplest example of inequivalent deformations are
+% the two automorphisms of $1+1$, namely $\idc$ and $\swapp$.
+
+The denotational semantics can be used to evaluate programs and to check equivalence of programs. For example, it is
+straightforward to verify that $\denot{(c_1~\oplus~c_2)~\fatsemi~\swapp} = \denot{\swapp \fatsemi (c_2~\oplus~c_1)}$. A
+slightly more involved example are the following two programs:
+
+\rotate{}
+
+\noindent The first program performs the following sequence of transformations:
+\[
+\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+\Tree [ [ {\small a} {\small b} ] {\small c} ] ~\to~
+\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+\]
+\noindent
+while the second evaluates as follows:
+\[
+\Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+\Tree [ {\small a} [ {\small c} {\small b} ] ] ~\to~
+\Tree [ [ {\small a} {\small c} ] {\small b} ] ~\to~
+\Tree [ [ {\small c} {\small a} ] {\small b} ] ~\to~
+\Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+\Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+\]
+
+The semantics above can be packaged in the category of finite sets and functions, $\SetFin$, which is the category
+freely generated by finite coproduct completion of the terminal category. Objects of $\SetFin$ can be identified with
+sets of fixed cardinality, that is, $\Fin[n] \defeq \Set{0,1,\ldots,n-1}$. $\SetFin$ has finite coproducts and products,
+which lets us interpret the types of $\PiLang$. Combinators are intepreted as morphisms in $\SetFin$, but we have to
+restrict to invertible morphisms, that is, isomorphisms. This gives the \emph{groupoid} of finite sets and bijections,
+$\BFin \defeq \mathsf{core}(\SetFin)$. The isomorphisms satisfied by coproducts and products in $\SetFin$ lift to
+$\BFin$, but they're no longer categorical coproducts and products. They give two symmetric monoidal tensor products on
+$\BFin$, the additive and multiplicative ones, with the multiplicative tensor distributing over the additive tensor.
+
+\begin{theorem}\label{thm:grp}
+  Let $\sim$ be the following equivalence relation on combinators $c_1 \sim c_2$ iff $\denot{c_1} = \denot{c_2}$
+  identifying combinators with the same denotation.  The notation $[c]_{\sim}$ refers to a representative combinator
+  from a $\sim$-equivalence class. Using this relation, we define the
+  category $\mathcal{C}$ as follows:
+  \begin{itemize}
+    \item $\mathit{Obj}(\mathcal{C})$ is the set of $\Pi$-types, and
+    \item $\mathit{Hom}(A,B) = [c : A \isot B]_{\sim}$
+    \end{itemize}
+  The category $\mathcal{C}$ is a groupoid.
+\end{theorem}
+\begin{proof}
+The relation $\sim$ identifies $c$, $\idc \circ c$, and $c \circ \idc$, and is associative. Furthermore, the denotation
+of each combinator is a reversible function thus making every morphism into an isomorphism.
+\end{proof}
+
+\note{Then say that $\mathcal{C}$ is ``the same'' as $\BFin$ We can say here that the relation $\sim$ also identifies
+ $(c_1~\oplus~c_2)~\fatsemi~\swapp$ and $\swapp \fatsemi (c_2~\oplus~c_1)$ and other properties. ?? }
+
+% \[\begin{array}{l}
+%     \Tree[ {\small a} [ {\small b} {\small c} ] ] \\
+%     2 \quad 1 \quad 0 \\
+%     0 \quad 2 \quad 1 \\
+%     \Tree[ [ {\small b}  {\small c} ] {\small a} ]
+% \end{array}\]
+
+% Such functions are finitely supported, that is, their outputs can be tabulated using the canonical ordering on
+% $\Fin[n]$. For bijections, this gives a listed permutation. By observing the action of the combinators on the values of
+% the finite set, we can define a denotational semantics which constructs the bijection.
+% \note{\(\Fin[n] \to A \eqv <\type{Vec_{n}}(A)\)}
+
+% In the previous section, we examined equivalences between conventional data structures, i.e., structured trees of
+% values. We now consider a richer but foundational notion of data: programs themselves. Indeed, universal computation
+% models crucially rely on the fact that \emph{programs are (or can be encoded as) data}, e.g., a Turing machine can be
+% encoded as a string that another Turing machine (or even the same machine) can manipulate. Similarly, first-class
+% functions are the \emph{only} values in the $\lambda$-calculus.  In our setting, we ask whether the programs developed
+% in the previous section can themselves be subject to (higher-level) equivalences?
 

### 2021-07-01 Pi+/paper/ufin.tex
denotation 1

@@
 on the choices made by the compiler writer, but we can think about that.
+
+%%%
+\subsection{To Put Somewhere or Delete}
+
+This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
+level~\cite{DBLP:conf/esop/CaretteS16}.
+As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
+we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
+set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
+the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
+$\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
+with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
+\[\begin{array}{rcl}
+    \denot{\onet + (\onet + \onet)} &=& \{ 0,1,2 \} \\
+    \denot{(\onet + \onet) + \onet} &=& \{ 0,1,2 \} \\
+    \\
+    \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (0~1~2) \\
+    \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (2~0~1)
+\end{array}\]
+where we have used the one-line notation for permutations with $(a~b~c)$ representing the
+permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
+enumeration of the elements of each type:
+\[\begin{array}{rcl}
+    \mathit{enum}(\zerot) &=& [~] \\
+    \mathit{enum}(\onet) &=& [ ~\Acon{tt}~ ] \\
+    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
+    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
+\end{array}\]
+\noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
+a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
+operation that appends all the subsequences in a sequence.
+
+Using the definition, we have:
+\[\begin{array}{rcl}
+    \mathit{enum}(\onet + (\onet + \onet)) &=& [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
+    \mathit{enum}((\onet + \onet) + \onet) &=& [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
+\end{array}\]
+Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
+$\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
+$\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
+$\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
+$\inrv{\Acon{tt}}$.
+
+\begin{center}
+  \input{assoc-perm.tikz}
+  \qquad
+  \input{swap-perm.tikz}
+\end{center}
+
+We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
+compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
+$\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:
+
+\[\begin{array}{rcll}
+\sem{A} &=& (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
+\end{array}\]
+
+\noindent The denotation $\sem{c}$ of a combinator $c : A \isot B$ is a path between $\sem{A}$ and $\sem{B}$. If the
+size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
+$(\mathsf{Fin}~n, [ n , \mathsf{refl} ])$. This path is directly constructed using $\mathit{ap}$ and the fact that $m=n$
+since combinators are always between types of the same size.
+
+\noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \Iso q$, the denotation
+$\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.
+
+\note{We use the rig structure of $\UFin$ in~\cref{subsec:rig} to interpret $\PiLang$.}
+
+We need a formal definition of normal form (canonical form)
+
+Recalling that the $\lambda$-calculus arises as the internal language of Cartesian Closed Categories
+(Elliott~\cite{Elliott-2017} gives a particularly readable account of this), we can think of $\Pi$ in similar terms, but
+for symmetric Rig Groupoids instead. For example, we can ask what does the equivalence above represent? It is actually a
+``linear'' representation of a 2-categorial commutative diagram! In fact, it is a painfully verbose version thereof, as
+it includes many \emph{refocusing} steps because our language does not build associativity into its syntax. Categorical
+diagrams usually do.  Thus if we rewrite the example in diagrammatic form, eliding all uses of associativity, but
+keeping explicit uses of identity transformations, we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents
+
+\vspace*{3mm}
+\begin{tikzcd}[column sep=normal, row sep=normal]
+ && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
+ & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
+a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
+  \arrow [dr, "\assoclp"]
+  \arrow [ddr, swap, "\assoclp"]
+    & (a+b)+c \arrow [r, "\swapp"] &
+    c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}]
+    & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
+ & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
+ & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
+             \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] &&
+    c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
+ && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
+ \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
+ \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
+\end{tikzcd}
 

### 2021-07-01 Pi+/paper/finite.tex
More text

@@
 Instead, in the proof, we will think about the free group as it was introduced at the beginning: elements of the free group as words over an alphabet of letters from generating set and the set of their formal inverses.
-If we take the disjoint union of $A$ with itself, that is, $A + A$ as the group's underlying set, we can use $\inl/\inr$ to mark the elements - for example, by stating that $\inl(a)}$ means $a$ and $\inr(a)$ means $\inv{a}$.
+If we take the disjoint union of $A$ with itself, that is, $A + A$ as the group's underlying set, we can use $\inl/\inr$ to mark the elements - for example, by stating that $\inl{a}$ means $a$ and $\inr{a}$ means $\inv{a}$.
 Hence, we can encode the free group using the free monoid - which is to say, lists on $A + A$.

### 2021-07-01 Pi+/paper/finite.tex
Fix latex

@@
 
-Connecting it all back to the language $\PiLang$ we are dealing with, it is possible to construct a simlar, yet simplified language $\PiLangHat$ (introduced at \cref{sec:equivalence}), which operates on adjacent transpositions, but still preservers much of $\PiLang$ structure.
+Connecting it all back to the language $\PiLang$ we are dealing with, it is possible to construct a simlar, yet simplified language $\PiHatLang$ (introduced at \cref{sec:equivalence}), which operates on adjacent transpositions, but still preservers much of $\PiLang$ structure.
 Thus, for the proof, a presentation based on adjacent transpositions is used, called \emph{Coxeter presentation}.
@@
   \[
-    \begin{array}{lcr}
+    \begin{array}{lr}
     % https://q.uiver.app/?q=WzAsNixbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzQsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwyLCJcXHRhdV8xXFx0YXVfMlxcdGF1XzIiXSxbMiwzLCJcXHRhdV8xIl0sWzQsMiwiXFx0YXVfMlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFswLDIsImJyYWlkIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJjYW5jZWwiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFszLDQsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzIsNSwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbNSw0LCJjYW5jZWwiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dXQ==
-    \[\begin{tikzcd}
+    \begin{tikzcd}
         && {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
@@
         \arrow["cancel", squiggly, from=3-5, to=4-3]
-    \end{tikzcd}\]
-    &
-    \text{and}
+    \end{tikzcd}
     &
     % https://q.uiver.app/?q=WzAsNSxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8yIl0sWzAsMSwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFsyLDQsIlxcdGF1XzJcXHRhdV8xIl0sWzAsMiwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFsxLDMsIlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwxLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzAsMiwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzMsNCwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFs0LDIsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV1d
-    \[\begin{tikzcd}
+    \begin{tikzcd}
         && {\tau_2\tau_1\tau_2\tau_2} \\
@@
         \arrow["cancel"', squiggly, from=4-2, to=5-3]
-    \end{tikzcd}\]
+    \end{tikzcd}
     \end{array}

### 2021-07-01 Pi+/paper/discussion.tex
Revise upto L360

@@
 
+\paragraph{Rewriting} There exists an algorithm, due to~\citet{knuthSimpleWordProblems1970}, that, when succeeds,
+constructs a well-behaved rewriting system for an arbitrary finite set of (undirected) equations. It did not work for
+us, there were too many equations, and proving correctness and termination was intractable.
+
+\paragraph{Computational group theory} Coxeter relations, Sn, other formalisations.
+
 Univalent Subuniverses, Univalent Typoids.
@@
 Coherence problems in type theory, coherence via Well-Foundedness.
-
-Computational group theory (in HoTT?)
 

### 2021-07-01 Pi+/paper/finite.tex
Revise upto L360

@@
 
+\todo{Move.}
+
 Although the meaning of a permutation - that is, the specific bijective function that it represents - is in some sense all there is, manipulating them syntactically still has its advantages.
@@
 Fortunately, group theory already has language in which such problems are expressed - we will be looking at solving the \textit{word problem}.
-Because $\PiLang$ is used for reasoning about reversible functions on finite sets, the group that we are interested in is the group of permutations on a fininte set, which is classically know as the symmetric group on $n$ words,
+Because $\PiLang$ is used for reasoning about reversible functions on finite sets, the group that we are interested in is the group of permutations on a finite set, which is classically known as the symmetric group,
 $\Sn$.
-Putting it in this form allows us to connect to the broader scope of the computational group theory and reuse some of the ideas from there.
-For example, during the proof, we will use \emph{Lehmer codes} (introduced at \cref{subsec:lehmer}) - a concept used widely in this context, which makes for a very convenient and compact method of representing permutations on a computer.
+Putting it in this form allows us to connect to the broader scope of computational group theory and reuse ideas from there.
+For example, during the proof, we will use \emph{Lehmer codes} (introduced in \cref{subsec:lehmer}) - a concept used widely in this context, which makes for a very convenient and compact method of representing permutations on a computer.
 
-We begin the section with a brief introduction on doing group theory in HoTT, and state the main results using this language in the subsequent sections.
+The goal of this section is to reconcile two different approaches to defining the symmetric group. We already defined
+$\UFin$ in the previous section, and showed how its loopspace encodes the automorphism group. Now, we're going to define
+the same group in a different way, in terms of the symmetric group presented using generators and relations, and show
+how the two points of view are equivalent.
 
@@
 
-However, more conveniently, in HoTT, we can instead use groupoids to talk about groups. A group can be identified with a
-1-object groupoid, using a technique called delooping. The delooping of a group $G$ is a groupoid $\B{G}$ given by a
-unique object $\pt$ with self-loops that are 1-paths $\pt \id_{\B{G}} \pt$ corresponding to the elements of $G$. Note
-that the group operations are automatically given by operations on the identity type, with $\refl_{\pt}$ for the neutral
-element, path composition for the group multiplication, and path inverse for the group inverse. These satisfy the group
-laws as well, up to the identity type, using the groupoid coherence laws. Moreover, for 1-groups which are supposed to
-be sets, these 2-paths should be propositions, so we have to restrict $\B{G}$ to be a 1-groupoid. Hence, a group is
-simply given by a pointed, connected 1-type~\cite*{buchholtzHigherGroupsHomotopy2018,symmetryBook2021}.
+However, more conveniently, in HoTT, we can instead use groupoids to talk about groups internally. A group can be
+identified with a 1-object groupoid, using a technique called delooping. The delooping of a group $G$ is a groupoid
+$\B{G}$ given by a unique object $\pt$ with self-loops that are 1-paths $\pt \id_{\B{G}} \pt$ corresponding to the
+elements of $G$. Note that the group operations are automatically given by operations on the identity type, with
+$\refl_{\pt}$ for the neutral element, path composition for the group multiplication, and path inverse for the group
+inverse. These satisfy the group laws as well, up to the identity type, using the groupoid coherence laws. Moreover, for
+1-groups which are supposed to be sets, these 2-paths should be propositions, so we have to restrict $\B{G}$ to be a
+1-groupoid. Hence, a group is simply given by a pointed, connected
+1-type~\cite*{buchholtzHigherGroupsHomotopy2018,symmetryBook2021}.
 
@@
 
-Given any set $A$, we can naively construct a \emph{free group on $A$},
-whose elements are drawn from the alphabet $A$, and closed under the group
-operations of multiplication and inverse, identified by the group axioms.
-For example, the singleton set generates the additive group of integers $\mathbb{Z}$ we introduced before. In this context, we will call $A$ the generating set of $F(A)$.
+Usually, there are are many equations, besides the group axioms, that hold for the elements of a group. For example, in
+the group $\mathbb{Z}/3\mathbb{Z}$, where group multiplication is given by addition modulo 3, we have an equation
+$1 + 1 + 1 = 0$, which is not a consequence of group axioms, but is specific to the particular group
+$\mathbb{Z}/3\mathbb{Z}$. A free group has the property that no other equations except the ones directly implied by the
+group axioms, such as $g \mult \inv{g} = e$, hold. Thus, $\mathbb{Z}/3\mathbb{Z}$ is not a free group. To be able to
+give an equational presentation of a group, we start by constructing the free group, then show how to add new equations
+to it.
 
-Usually, there are are many equations, besides the group axioms, that hold for
-the elements of the group.
-For example, in the group of $\mathbb{Z}/3\mathbb{Z}$, where group multiplication is given by the addition modulo 3, we have an equation $1 + 1 + 1 = 0$, which is not a consequence of group axioms, but rather it is specific to the particular group $\mathbb{Z}/3\mathbb{Z}$.
-A free group can also be defined by its property that no other equations, except the ones directly implied by the group axioms, such as $g \dot g^[-1] = e$, hold for it.
-Thus, $\mathbb{Z}/3\mathbb{Z}$ is not a free group.
+Given any set $A$, we can describe the free group $F(A)$ on $A$. We draw the elements from the alphabet $A$, close them
+under the group operations of multiplication and inverse, and identify them by the group axioms. For example, the
+singleton set generates the additive group of integers $\mathbb{Z}$ we introduced before. In this context, we will call
+$A$ the generating set of $F(A)$. Working in HoTT, as before, we can simply use higher inductive types to define the
+free group, by adding path constructors for the axioms we want.
 
-Working in HoTT, as before, we can again use higher inductive types to define the free group.
-
-\todo{What's the best notation for HITs?}
+\todo{Better notation for HITs?}
 \begin{definition}
@@
 
-A group homomorphism between groups is a function between the underlying sets
-that preserves the group structure, which we write as $G_1 \to^G G_2$. With
-this, we can state the universal property of free groups.
-\jk{In the HoTT version, which we now use, there is no underlying set}.
+A group homomorphism between groups is a function between the underlying sets that preserves the group structure, which
+we write as $G_1 \to^G G_2$. Using the induction principle for $F(A)$, we can easily prove the universal property of
+free groups, stemming from the free-forgetful adjucntion between the category of groups and sets. Informally, the
+universal property states that if we know how a map $f$ acts on the generating set $A$, we know exactly how it acts on
+every element of the group $F(A)$.
 
@@
 
-We already mentioned that the structure of a free group $F(A)$ is fully specified by its generating set $A$.
-Thus, the univerasal property states that if we know how the mapping $f$ acts on the generating set $A$, we also know how does it act on every element of the group $F(A)$.
+Although this definition is close to the intuitive universal-algebraic definition, it is not the most convenient one to
+work with. Besides being interested in the elements of the group, we also want to characterise the path space of the
+group. This HIT definition of $F(A)$ has lots of path constructors corresponding to each group axiom, making it
+difficult to describe its path space.
 
-The HoTT definition of $F(A)$ has lots of path constructors corresponding to
-each group axiom. \jk{why did we introduced it? Why is it bad?}.
-
-Instead, in the proof, we will think about the free group as it was introduced at the beginning: elements of the free group as words over an alphabet of letters from generating set and the set of their formal inverses.
-If we take the disjoint union of $A$ with itself, that is, $A + A$ as the group's underlying set, we can use $\inl/\inr$ to mark the elements - for example, by stating that $\inl{a}$ means $a$ and $\inr{a}$ means $\inv{a}$.
-Hence, we can encode the free group using the free monoid - which is to say, lists on $A + A$.
-Additionally, we need to ensure that group laws hold, so words identified by the group equations are equalt.
-Thus, elements of the group will actually be equivalence classes of lists on $A + A$.
+Instead, we will think about elements of the free group as words over an alphabet of letters drawn from the generating
+set \emph{and} the set of their formal inverses. If we take the disjoint union of $A$ with itself, that is, $A + A$ as
+the group's underlying set, we can use $\inl/\inr$ to mark the elements -- $\inl{a}$ means $a$ and $\inr{a}$ means
+$\inv{a}$. Hence, we can encode the free group using the free monoid, that is, lists of $A + A$. Additionally, we need
+to ensure that group laws hold, so words identified by the group equations are equal. Thus, elements of the free group
+will actually be equivalence classes of lists on $A + A$.
 
@@
 \begin{proposition}
-  $F(A) \defeq \quot{\List[A + A]}{\sim^{\ast}}$ has a group structure, with the
-  empty list $\nil$ for the neutral element, multiplication given by list append
-  $\append$, and inverse given by list reversal (\jk{THIS IS NOT TRUE}). Further, $F(A)$ with $\eta_A :
-    A \to F(A) \defeq \inl(a) \cons \nil$ satisfies the universal property of free
-  groups, as stated in the ~\cref{prop:free-groups}.
+  $F(A) \defeq \quot{\List[A + A]}{\sim^{\ast}}$ has a group structure, with the empty list $\nil$ for the neutral
+  element, multiplication given by list append $\append$, and inverse given by flipping $\inl$ and $\inr$, followed by
+  reversing the list. Further, $F(A)$ with $\eta_A : A \to F(A) \defeq \inl(a) \cons \nil$ satisfies the universal
+  property of free groups, as stated in~\cref{prop:free-groups}.
 \end{proposition}
@@
 
-We already talked about the semantic view of groups, but did not yet mention on how to syntactically represent them - which is the role of group presentations.
+\review{We have already seen many examples of groups, that is, sets carrying a group structure, but now we'll see how to
+  syntactically represent them -- which is the role of group presentations.}
 
-A presentation of a group builds it by starting from the free group $F(A)$ and introducing a
-collection of equations that have to be satisfied in the resulting group. For
-example, if we take $F(\unit) \defeq \mathbb{Z}$ and add an equation $1 + 1 + 1
-  = 0$, the resulting group would be $\mathbb{Z}/3\mathbb{Z}$.
-Note that not all groups have finite (or even recursively enumerable) presentations, and that a group can have any number of presentations.
-\jk{citation}
-
-The generators can be thought of as primitive operations in a (reversible)
-programming language, group structure gives the way of composing these
-operations and inverting them, and relations describe how these primitive
-operations interact with each other.
-
-As we mentioned, instead of opereating directly in the semantics of the group operation, here the
-focus is on the syntax. While before, the equality of elements (such as the
-result of multiplication of two elements) had to be computed in some external way, now it is
-reduced to a \emph{word problem}, i.e., deciding one word - representative of the group elements' equivalence class - can be reduced to another word,
-using group's equations.
-
-However, because equations are not directed, is not always possible to construct a
-well-behaving rewriting system.
-In general, the word problem is proven to be undecidable.
-There exists an algorithm, due to Knuth and Bendix \jk{citation}, that, when succeeds, constructs a well-behaving rewriting system for an arbitrary finite set of (undirected) equations.
-However, this algorithm produced a system that is often large and hard to reason about \jk{definitely a citation if we can find it}.
-
-\jk{Example?}
+A presentation of a group builds it by starting from the free group $F(A)$ and introducing a collection of equations
+that have to be satisfied in the resulting group. For example, if we take $F(\unit) \defeq \mathbb{Z}$ and add an
+equation $1 + 1 + 1 = 0$, the resulting group would be $\mathbb{Z}/3\mathbb{Z}$. Note that not all groups have finite
+(or computable) presentations, and, a group can have any number of presentations.
 
 \begin{definition}
-  Let $A$ be an $\hSet$ and $R : (A + A) \to (A + A) \to \UU$ a binary relation
-  on $A + A$. The group $G$ presented by $\langle A ; R \rangle$ is given by the
-  set-quotient of the free group $F(A)$ by the closure of $R$, or equivalently,
-  as the coequaliser
+  Let $A$ be an $\hSet$ and $R : (A + A) \to (A + A) \to \UU$ a binary relation on $A + A$. The group
+  $F(\langle A ; R \rangle)$ presented by $A$ and $R$, is given by the set-quotient of the free group $F(A)$ by the
+  closure of $R$, or equivalently, as the coequaliser
   \[\begin{tikzcd}
-      FR && FA && G
+      FR && FA && F(\langle A ; R \rangle)
       \arrow[shift right=2, from=1-1, to=1-3]
@@
 
-\vc{do we need this level of detail?}
-\todo{Universal property}
-\todo{Examples: empty relation, van Kampen of $\pi_{1}$}
+The universal property of the above definition allows for properly extending the relation on the generating set to the
+whole group.\vc{check!}
 
-The above definition is correct because the universal property of the free group allows
-for properly extending the relation on the generating set to the whole group.
-\jk{Turn that into a propostion?}
+\begin{proposition}[Universal property of $F(\langle A ; R \rangle)$]
+  Given a group $G$ and a map $f : A \to G$ that respects $R$, there is a unique group homomorphism
+  $\extend{f} : F(\langle A ; R \rangle) \to^G G$ such that $\extend{f} \comp \eta_A \htpy f$.
+\end{proposition}
 
-\subsection{Presentation of permutation groups}
+To relate this to our original problem, the generators of the group can be thought of as the primitive combinators in a
+(reversible) programming language, the group structure gives the composition and inversion operations, and the relations
+describe how these primitive operations interact with each other.
 
-We have already seen that the group of permutations, symmetric group $\Sn$, can be defined by taking the type $\Aut[\Fin[n]]$, and showing that it has the appropriate
-group structure.
-Alternatively, we talked on how one can define a group by its presentation,defining a set of generators and a set of relations.
-However, until now, we did not show any specific presentationation for $\Sn$.
+As we mentioned, instead of operating directly with a particular group, we focus on the syntactic presentation. While
+before, the only way to decide the equality of two elements in the group was to compute and check them on the nose, now
+it is reduced to a \emph{word problem}, that is, deciding whether one word -- a representative of the group elements'
+equivalence class, can be reduced to another word, using the group's relations. This can be thought of as a rewriting
+system. However, these equations are not directed, so it is not always possible to construct a well-behaved rewriting
+system. In general, the word problem for groups is proven to be undecidable.
 
-In giving a group presentation, there is an element of choice - as we already mentioned, a group
-can be presented in many different ways. For example, we could generate the
-permutation group on $\Fin[n]$ by using generators that:
+\jk{Example?}
+\todo{Examples: empty relation, full relation, van Kampen of $\pi_{1}$}
+
+\subsection{Presenting the permutation group}
+
+We have already seen that the group of permutations, or the symmetric group $\Sn$, can be defined by taking the type
+$\Aut[\Fin[n]]$ as the underlying set, and showing that it has the appropriate group structure. Now we're going to give
+a presentation for it, by defining a set of generators and relations.
+
+There is an element of choice here -- as we already mentioned, a group can be presented in many different ways. For
+example, we could generate the permutation group on $\Fin[n]$ by using generators that:
 
@@
 
-One way of thinking about these presentations is via sorting algorithms, which
-use different primitive operations. A sorting algorithm has to calculate a
-permutation of a list or a finite set, which satisfies the invariant of being a
-sorted sequence, which means, the primitive operations of a sorting algorithm
-should be able to generate all the permutations on a given list. If using a
-chosen set of reversible operations, one can write a sorting algorithm, then
-those operations generate the permutation group.
+One way of thinking about these presentations is via sorting algorithms, which use different primitive operations. A
+sorting algorithm has to calculate a permutation of a list or a finite set, which satisfies the invariant of being a
+sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the permutations
+on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate for the
+generators of a permutation group.
 
-For example, bubble sort uses the primitive operation of adjacent swaps,
-insertion sort and selection sort use the primitive operation of swapping the
-$i$-th element with the $j$-th element, cycle sort uses cyclical shifts of
-subsequences, pancake sort uses flips of prefixes of the list, et cetera.
-\todo{check!}
-
-The choice of generators for a presentation is important. \todo{Think about this.}
+For example, bubble sort uses the primitive operation of adjacent swaps, insertion sort and selection sort use the
+primitive operation of swapping the $i$-th element with the $j$-th element, cycle sort uses cyclical shifts of
+subsequences, pancake sort uses flips of prefixes of the list, et cetera~\todo{check!}. The choice of generators for our
+presentation is important for the following reasons.
 
 \begin{itemize}
-  \item It affects the difficulty of solving the word problem in $\Sn$ and
-        formalising the proof of its correctness.
-  \item It affects the proof of strong normalisation of reduction for the relations.
-  \item It dictates which words are normal forms in this presentation of $\Sn$.
-  \item It has to closely match the $\PiLang$ comobinators so that we can prove
-        completeness.
-  \item It dictates the choice of reversible gates in the synthesis and
-        normalisation of boolean circuits which are applications that we show in
-        the later sections.
+  \item It affects the difficulty of solving the word problem in $\Sn$ and formalising the proof of its correctness. In
+        the following subsections, we build a rewriting system using the generators and relations, and we need to prove
+        and formalise strong normalisation for this system.
+  \item The choice of generators dictates which words become normal forms in this presentation of $\Sn$. These normal
+        forms dictate the shape of the synthesised and normalised boolean circuits, which is the application we have in
+        mind.
+  \item Finally, the generators have to closely match the $\PiLang$ combinators so that we can quote back a permutation
+        to a $\PiLang$ program, for the proof of completeness.
 \end{itemize}
 
-Connecting it all back to the language $\PiLang$ we are dealing with, it is possible to construct a simlar, yet simplified language $\PiHatLang$ (introduced at \cref{sec:equivalence}), which operates on adjacent transpositions, but still preservers much of $\PiLang$ structure.
-Thus, for the proof, a presentation based on adjacent transpositions is used, called \emph{Coxeter presentation}.
-Not only is this then, in some sense, the natural choice, but it is also a presentation for which we can solve the word problem in a relatively easy, and which is used broadely in such context \jk{by other people?
-get citations}.
-
-The technical contribution of this section is a proof that the two descriptions of permutation group - $\Aut[\Fin[n]]$ and the Coxeter presentation - are equivalent.
-By doing that, we bridge the gap between the syntactic and semantic notions in our completeness proof -- by establishing a
-correspondence between the meaning of a program (a bijection) and its syntax (a word).
+We use a presentation based on adjacent transpositions for generators, called a~\emph{Coxeter presentation}, which we
+also use to solve the word problem for $\Sn$. Connecting it back to the language $\PiLang$, we will show that it is
+possible to encode all $\PiLang$ combinators using adjacent transpositions (in~\cref{sec:equivalence}).
 
@@
 
-We already mentioned that the Coxter presentation, which is the one will use, is based on adjacent transpositions, meaning that the primitive operations we use are adjacent swaps.
-When dealing with permutations on an $n + 1$-element set, there are $n$ adjacent transpositions, transposition number $k$ swapping elements $\el{k}$ and $\el{k+1}$.
-Thus, the generating set would be $\Fin[n]$.
+The primitive operations we use are going to be adjacent swaps. When dealing with permutations on an $n + 1$-element
+set, there are $n$ adjacent transpositions, transposition number $k$ swapping elements $\el{k}$ and $\el{k+1}$. Thus,
+the generating set would be $\Fin[n]$. There are three relations that we're going to specify for this presentation --
+these are the laws that these generators should satisfy. It is easiest to visualise them as braid diagrams.
 
-However, $\Sn$ is not a free group, and so we also have to specify the relations of the presentaiton -- the laws that such generators have to satisfy.
-there are three of them -- it is easiest to visualise them as braid diagrams.
-
-First, swapping the same two elements two times in a row is the same as doing nothing:
+First, swapping the same two elements two times in a row should be the same as doing nothing:
 
@@
   \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
+  \resizebox{0.5}{!}{
     \begin{center}
@@
      &
-    \(\xlongrightarrow[]{\cancel}\)
+    \resizebox{2.0}{!}{\(\xlongrightarrow[]{\cancel}\)}
      &
@@
     \end{center}
+       }
   \end{tabular}
@@
 
-Second, when swapping two distinct pairs of elements (i.e. when the indices of two transpositions
-differ by at least 1), it does not matter in which order the swapping happens,
-that is, we can slide the wires freely.
+Second, when swapping two distinct pairs of elements (i.e. when the indices of two transpositions differ by at least 1),
+it should not matter in which order the swapping happens, that is, we can slide the wires freely.
 
@@
 
-The previous two cases were for the transpositions that either completely overlap, or not overlap at all.
-The third case is thus left -- what happens if we perform transpositions that move the same elements.
-If the next transposition in the sequence is equal to the the first one, we can write a third law -- braiding relation.
+The previous two cases were for the transpositions that either completely overlap, or not overlap at all. The third case
+says what happens if we perform transpositions that move the same elements. If the next transposition in the sequence is
+equal to the the first one, we endorce a third law -- the braiding relation.
 
@@
 
-This construction is called a Coxeter presentation.
-Writing it formally, we take the generating set to be $\Fin[n]$, where the element $k$ corresponds to an adjacent transpositions $\tau_k$, which swaps elements $\el{k}$ and $\el{k+1}$.
-Then, we define a binary relation $\cox$ on the $\List[\Fin[n]]$, which encodes the laws discussed above.
+This construction is called a Coxeter presentation. Writing it formally, we take the generating set to be $\Fin[n]$,
+where the element $k$ corresponds to an adjacent transpositions $\tau_k$, which swaps elements $\el{k}$ and $\el{k+1}$.
+Then, we define a binary relation $\cox$ on $\List[\Fin[n]]$, which encodes the laws discussed above.
 

### 2021-07-02 Pi+/paper/finite.tex

@@
   \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
-  \resizebox{0.5}{!}{
     \begin{center}
@@
      &
-    \resizebox{2.0}{!}{\(\xlongrightarrow[]{\cancel}\)}
+    \(\xlongrightarrow[]{\cancel}\)
      &
@@
     \end{center}
-       }
   \end{tabular}
@@
 
-\jk{maybe change that?}
-Group persentation, as given in the ~\label{def:presentation}, requires the relation to be on the set of $A + A$, where the right copy corresponded to the set of formal inverses of the generators.
-However, in our case, we can observe that that the constructor $\cancel$ specifies that the inverse of each element is again the same element.
-Thus, the type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$.
+The type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$,
 
@@
 
-The approach to solving the word problem for $\Sn$ we use is to turn the relations $\cox*$ on
-$\List[\Fin[n]]$ to a rewriting system $(\List[\Fin[n]],\cox*)$.
-A normal form of an element can then be computed by following the rewrite rules, and two words can be compared for $\cox*$-equality by comparing their normal forms.
+A group presentation, as defined in~\label{def:presentation}, requires the relation to be on the set of words $A + A$,
+where the right copy corresponds to the set of formal inverses of the generators. We will resolve this mismatch later
+in~\cref{prop:lift-codiag}.
 
-A well-behaved rewriting system has to be strongly normalising and confluent. First, we observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
-are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$.
+\todo{check: However, in our case, the constructor $\cancel$ specifies that the inverse of each element is again the
+  same element.}
+
+To solve the word problem for $\Sn$, we turn the relations into a rewriting system $(\List[\Fin[n]],\cox*)$. A normal
+form of an element can then be computed by following the rewrite rules, and two words can be compared for
+$\cox*$-equality by comparing their normal forms. A well-behaved rewriting system has to be strongly normalising and
+confluent. First, we observe that after throwing out reflexivity and symmetry, the right hand sides of the relations
+$\cox*$ are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$.
 Thus, by directing the relation from left to right, we would get the termination property out of the box.
 
-The system also has to be confluent, meaning that all critical pairs, that is, terms with overlapping possible reduction rules, have to converge. For example, the pairs below converge.
+The system also has to be confluent, meaning that all critical pairs, that is, terms with overlapping possible reduction
+rules, have to converge. For example, the pairs below converge.
 
@@
 
-Unfortunately, in the system we defined, this is not true for all critical pairs.
-One example is the following, where both endpoints are normal with respect to
-$\cox*$ relation.
+Unfortunately, in the system we defined, this is not true for all critical pairs. One example is the following, where
+both endpoints are normal with respect to $\cox*$ relation.
 
@@
 
-Because of the counter-example discussed, the relations have to be changed.
-In this section, we will formally define a rewriting system partially based on the Coxeter relation, prove that it has the desired properties of confluence and strong normalisation.
+Because of the counter-example discussed, the relations have to be changed. In this section, we will formally define a
+rewriting system partially based on the Coxeter relations, and prove that it has the desired properties of confluence
+and strong normalisation.
 
-Fortunately, the change from the rewriting system described before will be small.
-Thus, we will not lose the connection to the theory of group presentations, where Coxeter relations are a standard notion~\todo{citation} -- to make the result broadly applicable, even though we use the reduction system for computation, we prove that the \jk{relation it gives} is indeed equivalent to the conguence closure of the standard Coxeter relation $\cox$.
+The Coxeter relations are a standard notion in the theory of group presentation~\todo{citation} -- even though we change
+the relations a bit, we will not lose the connection. In fact, we will prove that the new relations are equivalent, in a
+technical sense, to the standard Coxeter relations $\cox*$.
 
 The new reduction system $(\List[\Fin[n]], \longcox*)$ has generators corresponding to $\swap$, $\cancel$ and $\braid$.
-We fix the problem of the non-converging critical pair discussed previously by changing $\braid$ to be a slightly more general relation $\longbraid$.
+We fix the problem of the non-converging critical pairs discussed previously by changing $\braid$ to be a slightly more
+general relation $\longbraid$. \review{Further, we also inline the congruence closure in $\longcox$, allowing arbitrary
+  reductions inside the list.}
 
-\todo {but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
+\todo{but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
 
@@
 
-The result of this function is the sequence \(k + n, k + n - 1, k + n - 2, \ldots, n + 1\).
-Since we think of the number $m$ as the transposition $\tau_m$ which swaps elements $\el{m}$ and $\el{m+1}$, the role of this helper function is to produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ by $k$ places left, shifting all the elements in between one place right.
-Expressed in terms of the braiding diagram, for
-$n = 0$ and $k = 4$, it has the following form:
+The result of this function is the sequence \(k + n, k + n - 1, k + n - 2, \ldots, n + 1\). Since we think of the number
+$m$ as the transposition $\tau_m$ which swaps elements $\el{m}$ and $\el{m+1}$, the role of this helper function is to
+produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ $k$ places left, shifting all
+the elements in between one place right. Expressed in terms of the braiding diagram, for $n = 0$ and $k = 4$, it has the
+following form:
 
@@
 
-Constructors $\longcancel$ and $\longswap$ correspond directly to the approproiate constructors of $\cox$ and can be visualised in the same way as before.
-The remaining constructor uses the helper function to exchange the order of a long sequence of transpositions and a single transposition afterwards.
-For example, for $n = 2$ and $k = 4$, it allows for the reduction
-$[7, 6, 5, 4, 3, 7] \longcox* [6, 7, 6, 5, 4, 3]$. Visualized on the braid diagram, it looks as follows:
+Constructors $\longcancel$ and $\longswap$ correspond directly to the appropriate constructors of $\cox$ and can be
+visualised in the same way as before. The remaining constructor uses the helper function to exchange the order of a long
+sequence of transpositions and a single transposition afterwards. For example, for $n = 2$ and $k = 4$, it allows for
+the reduction $[7, 6, 5, 4, 3, 7] \longcox* [6, 7, 6, 5, 4, 3]$. Visualised on the braid diagram, it looks as follows:
 
@@
 
-Note that the previous $\braid$ rule is a special case of $\longbraid$, with $k = 0$.
+Note that the previous $\braid$ rule is a special case of $\longbraid$, with $k = 0$. Also, as before, the left-hand
+sides of the relation are strictly larger than the right-hand sides, in terms of the lexicographic order on
+$\List[\Fin[n]]$.
 
-Like before, we define the $\longcox*$ to be the closure of $\longcox$, but only under reflexivity and transitivity.
+We define the relation $\longcox*$ to be the closure of $\longcox$, under reflexivity and (right-step extended)
+transitivity.
 
@@
 
-The modified form of the Coxeter relations are unwieldy and difficult to prove properties about by induction, and, as we mentioned, have little connection to the broader scope of computational group theory.
-However, we can prove that the closures of the two forms of the relations are equivalent, and work up to this equivalence!
+The modified form of the Coxeter relations are unwieldy and difficult to prove properties about by induction. However,
+we can use the following observation to be able to use the more standard $\cox*$ relation.~\todo{Check Huet's paper for
+  this property.}
 
 \begin{proposition}~\label{prop:coxlongcox}
-  $\cox*$ and $\longcox*$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff there is a $u$ such that $w \longcox* u$ and $v \longcox* u$.
+  $\cox*$ and $\longcox*$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff there is a $u$
+  such that $w \longcox* u$ and $v \longcox* u$.
 \end{proposition}
-\jk{does the construction above has a name?
-It is analogous to the one we do in lambda calculus, when taking $\equiv_{\beta}$.
-We have to write something about this proof using the Church-Rosser idea.}
 
 By strong normalisation, we get a unique choice function $\normf : {\List[\Fin[n]]} \to {\List[\Fin[n]]}$ that produces
-a normal form for terms of $\List[\Fin[n]]$. Equivalence classes on $\List[\Fin[n]]$ are precisely those terms that have
-the same normal form. We state a few important properties that $\normf$ satisfies.
+a normal form for terms of $\List[\Fin[n]]$. Equivalence classes on $\List[\Fin[n]]$ are defined to be those terms that
+have the same normal form. We state a few important properties that $\normf$ satisfies.
 
@@
     \item $\normf$ is idempotent, that is, $\normf \comp \normf \htpy \normf$.
-    \item $\normf$ splits, that is, we have ${\List[\Fin[n]]} \xrightarrow{s} \Sn[n] \xrightarrow{r} {\List[\Fin[n]]}$
-          such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
+    \item $\normf$ splits into a section-retraction pair, that is, we have
+          ${\List[\Fin[n]]} \xrightarrow{s} \Sn[n] \xrightarrow{r} {\List[\Fin[n]]}$ such that $s \comp r \htpy \normf$
+          and $r \comp s \htpy \idfunc_{\Sn[n]}$.
     \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \).
@@
 \end{proposition}
+
+\todo{We set out to solve the word problem for $\Sn$ \ldots}
+\todo{Revise this section by introducing $\Sn$ here.}
 
@@
 
-\begin{proposition}
+\begin{proposition}~\label{prop:lift-codiag}
   $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to
@@
 
-For a permutation $\sigma : \Aut[\Fin[n]]$, for any element $i: \Fin[n]$, we can define the inversion count of $i$ as the number of elements smaller than it appearing after in the permutation.
+For a permutation $\sigma : \Aut[\Fin[n]]$, for any element $i: \Fin[n]$, we can define the inversion count of $i$ as
+the number of smaller elements appearing after it in the permutation.
+
 \todo{typify}
@@
   Given $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by
-    $|\{j < i : \sigma(j) > \sigma(i)\}|$.
+  $|\Set{j < i | \sigma(j) > \sigma(i)}|$.
 \end{definition}
@@
 It turns out that knowing just the inversion counts for all the elements, one can reconstruct the starting permutation.
-Also, observe that the inversion count for element $i$ is guaranteed to be smaller than $i$, thus fitting in the $i$-th place of a Lehmer code tuple.
-As an example, consider the following tabulated presentation of the permutation:
+Also, observe that the inversion count for element $i$ is guaranteed to be smaller than $i$, thus fitting in the $i$-th
+place of a Lehmer code tuple. As an example, consider the following tabulated presentation of the permutation:
 
@@
 
-The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 0 (because of $\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it is last in the sequence), and for $\el{4}$ is 1 (because of $\el{3}$).
-Thus, the Lehmer code for the permutation $\sigma$ is the 5-tuple $l = (0, 1, 2, 0, 1)$.
+The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 0 (because of
+$\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it comes
+last in the sequence), and for $\el{4}$ is 1 (because of $\el{3}$). Thus, the Lehmer code for the permutation $\sigma$
+is the 5-tuple $l = (0, 1, 2, 0, 1)$.
 
-To construct the presentation of the permutation from the Lehmer code, written using adjacent transpositions, we perform an algorithm similar to \emph{insertion sort}.
-Starting from the left-most position of the tuple $l$, we'll read the value $v$, insert the new
-element at the end of the newly created list, and shift it backward $v$ places.
+To construct the presentation of the permutation from the Lehmer code, written using adjacent transpositions, we perform
+an algorithm similar to \emph{insertion sort}. Starting from the left-most position of the tuple $l$, we'll read the
+value $v$, insert the new element at the end of the newly created list, and shift it backward $v$ places.
 
@@
 
-\todo{So now, finally, give an explict example of a normal form, Lehmer immersion}.
+\todo{So now, finally, give an explicit example of a normal form, Lehmer immersion}.
 

### 2021-07-02 Pi+/paper/reversible1.tex
note about changing thm
Minor: formatting denotation

@@
     \begin{array}{lrcl}
-\textit{Value types} & A,B,C,D &::=& \zerot \alt \onet \alt A+B \alt A\times B \\
-\textit{Values}      & v,w,x,y &::=& \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
-\textit{Program types} &&& A \iso B \\
-\textit{Programs} & c &::=& (\textrm{See Fig.~\ref{pi-terms}})
+      \textit{Value types}   & A,B,C,D & ::= & \zerot \alt \onet \alt A+B \alt A\times B        \\
+      \textit{Values}        & v,w,x,y & ::= & \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
+      \textit{Program types} &         &     & A \iso B                                         \\
+      \textit{Programs}      & c       & ::= & (\textrm{See Fig.~\ref{pi-terms}})
     \end{array}$}}
@@
         \begin{array}{rrcll}
-\idc :& A & \iso & A &: \idc \\
+          \idc :     & A                     & \iso & A                            & : \idc      \\
           \\
-\identlp :&  \zerot + A & \iso & A &: \identrp \\
-\swapp :&  A + B & \iso & B + A &: \swapp \\
-\assoclp :&  A + (B + C) & \iso & (A + B) + C &: \assocrp \\ [1.5ex]
-\identlt :&  \onet \times A & \iso & A &: \identrt \\
-\swapt :&  A \times B & \iso & B \times A &: \swapt \\
-\assoclt :&  A \times (B \times C) & \iso & (A \times B) \times C &: \assocrt \\ [1.5ex]
-\absorbr :&~ \zerot \times A & \iso & \zerot ~ &: \factorzl \\
-\dist :&~ (A + B) \times C & \iso & (A \times C) + (B \times C)~ &: \factor
+          \identlp : & \zerot + A            & \iso & A                            & : \identrp  \\
+          \swapp :   & A + B                 & \iso & B + A                        & : \swapp    \\
+          \assoclp : & A + (B + C)           & \iso & (A + B) + C                  & : \assocrp  \\ [1.5ex]
+          \identlt : & \onet \times A        & \iso & A                            & : \identrt  \\
+          \swapt :   & A \times B            & \iso & B \times A                   & : \swapt    \\
+          \assoclt : & A \times (B \times C) & \iso & (A \times B) \times C        & : \assocrt  \\ [1.5ex]
+          \absorbr : & ~ \zerot \times A     & \iso & \zerot ~                     & : \factorzl \\
+          \dist :    & ~ (A + B) \times C    & \iso & (A \times C) + (B \times C)~ & : \factor
         \end{array}$}}
@@
 The denotation of types is straightforward:
-\[\begin{array}{rcl}
-\denot{\zerot} &=& \emptyset \\
-\denot{\onet} &=& \{ \Acon{tt} \} \\
-\denot{A + B} &=& \denot{A} \uplus \denot{B} \\
-\denot{A \times B} &=& \denot{A} \times \denot{B}
-\end{array}\]
+
+\begin{align*}
+  \denot{\zerot}     & = \emptyset                  \\
+  \denot{\onet}      & = \{ \Acon{tt} \}            \\
+  \denot{A + B}      & = \denot{A} \uplus \denot{B} \\
+  \denot{A \times B} & = \denot{A} \times \denot{B}
+\end{align*}
+
 where $\emptyset$ is the empty set, $\uplus$ is the disjoint union of sets, and $\times$ is the cartesian product of
 sets. The denotation of a combinator $c : A \isot B$ is a (bijective) function mapping $\denot{A}$ to $\denot{B}$:
-\[\begin{array}{rcl}
-\denot{\identlp} (\inlv{v}) &=& v \\
-\denot{\identrp} v &=& \inlv{v} \\
-\denot{\swapp} (\inlv{v}) &=& \inrv{v} \\
-\denot{\swapp} (\inrv{v}) &=& \inlv{v} \\
-\denot{\assoclp} (\inlv{v})  &=& \inlv{(\inlv{v})} \\
-\denot{\assoclp} (\inrv{(\inlv{v})}) &=& \inlv{(\inrv{v})} \\
-\denot{\assoclp} (\inrv{(\inrv{v})}) &=& \inrv{v} \\
-\denot{\assocrp} (\inlv{(\inlv{v})}) &=& \inlv{v} \\
-\denot{\assocrp} (\inlv{(\inrv{v})}) &=& \inrv{(\inlv{v})} \\
-\denot{\assocrp} (\inrv{v})  &=& \inrv{(\inrv{v})} \\
-\denot{\identlp} (\Acon{tt} , v)  &=& v \\
-\denot{\identrp} v &=& (\Acon{tt} , v) \\
-\denot{\swapt}  (v_1 , v_2) &=& (v_2 , v_1) \\
-\denot{\assoclt} (v_1 , (v_2 , v_3)) &=& ((v_1 , v_2) , v_3) \\
-\denot{\assocrt} ((v_1 , v_2) , v_3) &=& (v_1 , (v_2 , v_3)) \\
-\denot{\dist} (\inlv{v_1} , v_3)  &=& \inlv{(v_1 , v_3)} \\
-\denot{\dist} (\inrv{v_2 , v_3})  &=& \inrv{(v_2 , v_3)} \\
-\denot{\factor}  (\inlv{(v_1 , v_3)}) &=& (\inlv{v_1} , v_3) \\
-\denot{\factor}  (\inrv{(v_2 , v_3)}) &=& (\inrv{v_2} , v_3) \\
-\denot{\idc} v &=& v \\
-\denot{(c_1 \fatsemi c_2)} v &=& (\denot{c_2} \circ \denot{c_1}) v \\
-\denot{(c_1 \oplus c_2)} (\inlv{v}) &=& \inlv{(\denot{c_1}~v)} \\
-\denot{(c_1 \oplus c_2)} (\inrv{v}) &=& \inrv{(\denot{c_2}~v)} \\
-\denot{(c_1 \otimes c_2)} (v_1 , v_2) &=& (\denot{c_1} v_1 , \denot{c_2} v_2)
-  \end{array}\]
+
+\begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
+  \denot{\identlp}           & (\inlv{v})           & = & v                                   \\
+  \denot{\identrp}           & v                    & = & \inlv{v}                            \\
+  \denot{\swapp}             & (\inlv{v})           & = & \inrv{v}                            \\
+  \denot{\swapp}             & (\inrv{v})           & = & \inlv{v}                            \\
+  \denot{\assoclp}           & (\inlv{v})           & = & \inlv{(\inlv{v})}                   \\
+  \denot{\assoclp}           & (\inrv{(\inlv{v})})  & = & \inlv{(\inrv{v})}                   \\
+  \denot{\assoclp}           & (\inrv{(\inrv{v})})  & = & \inrv{v}                            \\
+  \denot{\assocrp}           & (\inlv{(\inlv{v})})  & = & \inlv{v}                            \\
+  \denot{\assocrp}           & (\inlv{(\inrv{v})})  & = & \inrv{(\inlv{v})}                   \\
+  \denot{\assocrp}           & (\inrv{v})           & = & \inrv{(\inrv{v})}                   \\
+  \denot{\identlt}           & (\Acon{tt} , v)      & = & v                                   \\
+  \denot{\identrt}           & v                    & = & (\Acon{tt} , v)                     \\
+  \denot{\swapt}             & (v_1 , v_2)          & = & (v_2 , v_1)                         \\
+  \denot{\assoclt}           & (v_1 , (v_2 , v_3))  & = & ((v_1 , v_2) , v_3)                 \\
+  \denot{\assocrt}           & ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
+  \denot{\dist}              & (\inlv{v_1} , v_3)   & = & \inlv{(v_1 , v_3)}                  \\
+  \denot{\dist}              & (\inrv{v_2 , v_3})   & = & \inrv{(v_2 , v_3)}                  \\
+  \denot{\factor}            & (\inlv{(v_1 , v_3)}) & = & (\inlv{v_1} , v_3)                  \\
+  \denot{\factor}            & (\inrv{(v_2 , v_3)}) & = & (\inrv{v_2} , v_3)                  \\
+  \denot{\idc}               & v                    & = & v                                   \\
+  \denot{(c_1 \fatsemi c_2)} & v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
+  \denot{(c_1 \oplus c_2)}   & (\inlv{v})           & = & \inlv{(\denot{c_1}~v)}              \\
+  \denot{(c_1 \oplus c_2)}   & (\inrv{v})           & = & \inrv{(\denot{c_2}~v)}              \\
+  \denot{(c_1 \otimes c_2)}  & (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
+\end{longtable}
+
+% \begingroup
+% \allowdisplaybreaks
+% \begin{align*}
+% \end{align*}
+% \endgroup
 
@@
 
+\note{change thm:
+  1. Each function is a bijection. 2. If there is a 2-combinator, the denotations of the 1-combinators are equal
+}
+
 \begin{theorem}\label{thm:grp}

### 2021-07-02 Pi+/paper/finite.tex
Move the definition of Sn

@@
 
-The type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$,
-
-\begin{definition}[$\Sn$]
-  \(\Sn \defeq \quot{\List(\Fin[n])}{\cox*}\)
-\end{definition}
-
-A group presentation, as defined in~\label{def:presentation}, requires the relation to be on the set of words $A + A$,
-where the right copy corresponds to the set of formal inverses of the generators. We will resolve this mismatch later
-in~\cref{prop:lift-codiag}.
-
-\todo{check: However, in our case, the constructor $\cancel$ specifies that the inverse of each element is again the
-  same element.}
-
 To solve the word problem for $\Sn$, we turn the relations into a rewriting system $(\List[\Fin[n]],\cox*)$. A normal
@@
 By strong normalisation, we get a unique choice function $\normf : {\List[\Fin[n]]} \to {\List[\Fin[n]]}$ that produces
-a normal form for terms of $\List[\Fin[n]]$. Equivalence classes on $\List[\Fin[n]]$ are defined to be those terms that
-have the same normal form. We state a few important properties that $\normf$ satisfies.
+a normal form for terms of $\List[\Fin[n]]$. We state and prove a couple of important properties enjoyed by $\normf$.
 
@@
     \item $\normf$ is idempotent, that is, $\normf \comp \normf \htpy \normf$.
-    \item $\normf$ splits into a section-retraction pair, that is, we have
-          ${\List[\Fin[n]]} \xrightarrow{s} \Sn[n] \xrightarrow{r} {\List[\Fin[n]]}$ such that $s \comp r \htpy \normf$
-          and $r \comp s \htpy \idfunc_{\Sn[n]}$.
+  \end{enumerate}
+\end{proposition}
+
+The type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$,
+
+\begin{definition}[$\Sn$]
+  \(\Sn \defeq \quot{\List(\Fin[n])}{\cox*}\)
+\end{definition}
+
+Note that this relation $\cox*$ is not $\hProp$-valued, that is, $w \cox* v$ is not a proposition because reductions are
+not unique. Because of this, the quotient $\Sn$ is not effective, that is, $\quotrel : w \cox* v \to q(w) \id q(v)$ is
+not an equivalence. We could truncate the relation to change this, but then the codomain of $\normf$ also needs to be
+truncated.
+
+We could also define equivalence classes on $\List[\Fin[n]]$ to be those terms that have the same normal form. Using the
+$\normf$ function, we could define a new relation $(w \approx v) \defeq (\normf(w) \id \normf(v))$ which is
+$\hProp$-valued, and quotient $\List[\Fin[n]]$ by $\approx$. Using the properties of the rewriting system we proved, we
+make a few observations.
+
+\begin{proposition}
+  \leavevmode
+  \begin{enumerate}
+    \item $\normf$ splits into a section-retraction pair, that is, we have ${\List[\Fin[n]]} \xrightarrow{s} \Sn[n]
+            \xrightarrow{r} {\List[\Fin[n]]}$ such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
     \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \).
@@
 
-\todo{We set out to solve the word problem for $\Sn$ \ldots}
-\todo{Revise this section by introducing $\Sn$ here.}
-
-Note that this relation $\cox*$ is not $\hProp$-valued, that is, $w \cox* v$ is not a proposition because reductions are
-not unique. Because of this, the quotient $\Sn$ is not effective, that is, $\quotrel : w \cox* v \to q(w) \id q(v)$ is
-not an equivalence. We could truncate the relation to change this, but then the codomain of $\normf$ also needs to be
-truncated. Instead, using the $\normf$ function, we could define a new relation
-$(w \approx v) \defeq (\normf(w) \id \normf(v))$ which is $\hProp$-valued, and quotient $\List[\Fin[n]]$ by $\approx$.
-This is not necessary because our results hold for $\cox*$ using the properties of the rewriting system we proved, and
-we do not rely on the effectivity of the quotient.
-
-Finally, we prove the group structure of $\Sn$, and show that it is indeed a group presentation, given by a quotient of
-the free group $F(\Fin[n])$.
+Notice however, that a group presentation, as defined in~\cref{def:presentation}, requires the relation to be on the set
+of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The constructor
+$\cancel$ specifies that the inverse of each element is again the same element, using which we can show that our
+definition of $\Sn$ is equivalent to the definition of a presented group, by lifting the $\cox*$ relation.
 
 \begin{proposition}
-  There is a group structure on $\Sn$, where the identity element is $\nil$, multiplication is given by list append, and
-  inverse is given by list reversal.
+  \leavevmode
+  \begin{enumerate}
+    \item There is a group structure on $\Sn$, where the identity element is $\nil$, multiplication is given by list
+          append, and inverse is given by list reversal.
+    \item $\Sn$ is equivalent to the group presented by generators $\Fin[n] + \Fin[n]$ with the relations given by the
+          normal closure of $\cox*$ extended to $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $\nabla_{A} : A + A
+          \to A$.
+  \end{enumerate}
 \end{proposition}
 
-\begin{proposition}~\label{prop:lift-codiag}
-  $\Sn$ is equivalent to the generated group given by the normal closure of $\cox*$ extended to
-  $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $\nabla_{A} : A + A \to A$.
-\end{proposition}
+We set out to solve the word problem for $\Sn$. To decide if two words in $\List[\Fin[n]]$ are equal, we simply have to
+compute their normal forms using $\normf$ and check if they're equal on the nose. If they do, they correspond to the
+same permutation.
 

### 2021-07-02 Pi+/paper/finite.tex
Fix uses of downf function

@@
 
-\begin{definition}[$\downf : (n : \Nat) \to (k : \Nat) \to {\List[\Fin[\suc[k + n]]]}$]
+\begin{definition}[$\downf : (n : \Nat) \to (k : \Nat) \to {\List[\Fin[k + n]]}$]
   \begin{align*}
@@
 
-The result of this function is the sequence \(k + n, k + n - 1, k + n - 2, \ldots, n + 1\). Since we think of the number
-$m$ as the transposition $\tau_m$ which swaps elements $\el{m}$ and $\el{m+1}$, the role of this helper function is to
-produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ $k$ places left, shifting all
-the elements in between one place right. Expressed in terms of the braiding diagram, for $n = 0$ and $k = 4$, it has the
-following form:
+The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\). Since we think of the
+number $m$ as the transposition $\tau_m$ which swaps elements $\el{m}$ and $\el{m+1}$, the role of this helper function
+is to produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ $k$ places left,
+shifting all the elements in between one place right. Expressed in terms of the braiding diagram, for $n = 0$ and $k =
+  4$, it has the following form:
 
@@
 visualised in the same way as before. The remaining constructor uses the helper function to exchange the order of a long
-sequence of transpositions and a single transposition afterwards. For example, for $n = 2$ and $k = 4$, it allows for
-the reduction $[7, 6, 5, 4, 3, 7] \longcox* [6, 7, 6, 5, 4, 3]$. Visualised on the braid diagram, it looks as follows:
+sequence of transpositions and a single transposition afterwards. For example, for $n = 2$ and $k = 3$, it allows for
+the reduction $[6, 5, 4, 3, 2, 6] \longcox* [5, 6, 5, 4, 3, 2]$. Visualised on the braid diagram, it looks as follows:
 
@@
 
-\begin{definition}[$\immersion : (n : \Nat) \to {\Lehmer[n]} \to {\List[\Fin[n]]}$]
+\begin{definition}[$\immersion : (n : \Nat) \to {\Lehmer[n]} \to {\List[\Fin[\suc[n]]]}$]
   \begin{align*}
@@
 As described above, the number $r$ at the position $k$ in the tuple describes how many inversions the element $\el{k}$
-has. Thus, we need to perform $r$ many adjacent transpositions to get to the desired position. The function that
-constructs a sequence of $r$ adjacent transpositions $[\suc[n], \suc[n] - 1, \suc[n] - 2, \ldots \suc[n] - r]$ is
+has. Thus, we need to perform $r$ many adjacent transpositions to get to the desired position, which is given by
 $(\suc[n] - r) \downf r$.

### 2021-07-02 Pi+/paper/finite.tex
Explain how Lehmer codes are normal forms

@@
 
-\todo{diagram with List (Fin n) being divided into equivalence classes, and Lehmer codes being an image of immersion,
-  being the normal form (representative) in each class.}
-
 There are many ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer codes are known in
@@
 
-We can show that the function $\immersion$ gives an equivalence betweeen $\Lehmer[n]$ and $\im{\normf}$.
+As an example, let us look at the code $c = (4, 3, 2, 1, 0)$. The list $\immersion(c)$ is then going to be $[0, 1, 0, 2,
+1, 0, 3, 2, 1, 0, 4, 3, 2, 1, 0]$. It is a concatenation of a sequence of decreasing lists. Where could a reduction
+happen? First, it can't happen inside any of the decreasing components: $\longcancel$ requires repeating elements,
+$\longswap$ acts when a smaller number precedes a larger one, and $\longbraid$ has a non-monotone sequence on the left.
+This leaves the case of a reduction happening on a fragment that borders two subsequences. Again, $\longcancel$ requires
+two equal consecutive numbers, which have to be the last one in some decreasing sequence and the first one in the next
+one. But the first number in a sequence is always larger than every number in the sequence before it - which also shows
+why $\longswap$ cannot happen. The remaining case of $\longbraid$ is proven on similar grounds, since it requires the
+number appearing after the decreasing sequence to be equal to the first one in the sequence.
+
+Using this idea, we can show that the function $\immersion$ gives an equivalence betweeen $\Lehmer[n]$ and
+$\im{\normf}$.
 
@@
   \begin{enumerate}
-    \item For any Lehmer code $c$, $\immersion(c)$ is a normal form with resepct to $\cox*$, that is, $\immersion(c)$ is
+    \item For any Lehmer code $c$, $\immersion(c)$ is a normal form with resepct to $\longcox*$, that is, $\immersion(c)$ is
           in $\im{\normf}$.

### 2021-07-02 Pi+/paper/finite.tex
Redo Lehmer algorithm example

@@
 \begin{definition}
-  Given $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by
-  $|\Set{j < i | \sigma(j) > \sigma(i)}|$.
+  Given a permutation $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by
+  \[ |\Set{j < i | \sigma(j) > \sigma(i)}|. \]
 \end{definition}
 
-It turns out that knowing just the inversion counts for all the elements, one can reconstruct the starting permutation.
-Also, observe that the inversion count for element $i$ is guaranteed to be smaller than $i$, thus fitting in the $i$-th
-place of a Lehmer code tuple. As an example, consider the following tabulated presentation of the permutation:
-
-\todo{fix this figure}
+It turns out that from knowing just the inversion counts for all the elements, one can reconstruct the starting
+permutation. Also, observe that the inversion count for element $i$ is guaranteed to be smaller than $i$, thus fitting
+in the $i$-th place of a Lehmer code tuple. As an example, consider the following tabulated presentation of a
+permutation of $\Fin[5]$.
 
@@
   \sigma =
-  \begin{array}{ccccccccccccccc}
-    | & 0      & | & 1      & | & 2      & | & 3      & | & 4      & | \\
-    \hline                                                             \\
-    | & \el{2} & | & \el{1} & | & \el{4} & | & \el{0} & | & \el{3} & | \\
-    \hline                                                             \\
-  \end{array}
+  \begin{pmatrix}
+    0      & 1      & 2      & 3      & 4      \\
+    \el{2} & \el{1} & \el{4} & \el{0} & \el{3} \\
+  \end{pmatrix}
 \]
 
-%  0 1 2 3 4
-% -----------
-% |2|0|1|4|3|
-% -----------
+The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 1 (because of
+$\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it comes
+last in the sequence), and for $\el{4}$ is 2 (because of $\el{3}$ and $\el{0}$). Thus, the Lehmer code for the permutation $\sigma$
+is the 5-tuple $l = (0, 1, 2, 0, 2)$.
 
-The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 0 (because of
-$\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it comes
-last in the sequence), and for $\el{4}$ is 1 (because of $\el{3}$). Thus, the Lehmer code for the permutation $\sigma$
-is the 5-tuple $l = (0, 1, 2, 0, 1)$.
-
-To construct the presentation of the permutation from the Lehmer code, written using adjacent transpositions, we perform
-an algorithm similar to \emph{insertion sort}. Starting from the left-most position of the tuple $l$, we'll read the
-value $v$, insert the new element at the end of the newly created list, and shift it backward $v$ places.
+To decode the permutation back from this Lehmer code, we perform an algorithm similar to \emph{insertion sort}. The
+element of the Lehmer code being currently processed is highlighted in the left column of the table below. Starting from
+a sorted list, the element at index $k$ has to be given $l[k]$ inversions. Because of the the invariant that all the
+elements before newly processed one are smaller than it, the proper number of inversions is created by simply shifting
+the element $l[k]$ places left.
 
 \begin{center}
-  \begin{tabular}{c|p{0.75\linewidth}}
-    (0, 0, 2, 0, 1)               & We start from an empty list $[]$                                                                 \\
-    (\highlight{{0}}, 0, 2, 0, 1) & We read 0 as the left-most value from $l$. Thus, we append the element $\el{0}$ to our
-    list, getting $[\el{0}]$. The element is shifted $0$ places, so it remains in the
-    same place.                                                                                                                      \\
-    (0, \highlight{{0}}, 2, 0, 1) & Then, similarly, we read another 0 for the element $\el{1}$, append it to the
-    list getting $[\el{0}, \el{1}]$, and don't shift it either.                                                                      \\
-    (0, 0, \highlight{{2}}, 0, 1) & We read 2 for the next the element $\el{2}$ - we append $\el{2}$ to our list, getting
-    $[\el{0}, \el{1}, \el{2}]$, and shift it 2 places right, which results in a list $[\el{0}, \el{2}, \el{1}]$
-    \todo{Typeset it nicely, with arrows showing the shifting}.                                                                      \\
-    (0, 0, 2, \highlight{{0}}, 1) & Then we read 0 - appending $\el{3}$ and not shifting, getting $[\el{0}, \el{2}, \el{1}, \el{3}]$ \\
-    (0, 0, 2, 0, \highlight{{1}}) & Finally, reading 1 for element $\el{4}$ - appending $\el{4}$ to the list and shifting it
-    one place right results in the final list $[\el{0}, \el{2}, \el{1}, \el{4}, \el{3}]$                                             \\
+  \begin{tabular}{l|r}
+    (\highlight{{0}}, 1, 2, 0, 2) & $[\highlightAlt{\el{0}}, \el{1}, \el{2}, \el{3}, \el{4}]$ \\
+    (0, \highlight{{1}}, 2, 0, 2) & $[\highlightAlt{\el{1}}, \el{0}, \el{2}, \el{3}, \el{4}]$ \\
+    (0, 1, \highlight{{2}}, 0, 2) & $[\highlightAlt{\el{2}}, \el{1}, \el{0}, \el{3}, \el{4}]$ \\
+    (0, 1, 2, \highlight{{0}}, 2) & $[\el{2}, \el{1}, \el{0}, \highlightAlt{\el{3}}, \el{4}]$ \\
+    (0, 1, 2, 0, \highlight{{2}}) & $[\el{2}, \el{1}, \highlightAlt{\el{4}}, \el{0}, \el{3}]$ \\
   \end{tabular}
 \end{center}
-\todo{figure}
 
-Writing formally, to turn a code into a permutation written in the Coxeter presentation, we use $\immersion$:
+Writing formally, to turn a code into a permutation written in the Coxeter presentation, we define a function
+$\immersion$. As described above, the number $r$ at position $k$ in the tuple describes how many inversions the element
+$\el{k}$ has. Thus, we need to perform $r$ many adjacent transpositions to get to the desired position, which is given
+by $(\suc[n] - r) \downf r$.
 
@@
 \end{definition}
-
-As described above, the number $r$ at the position $k$ in the tuple describes how many inversions the element $\el{k}$
-has. Thus, we need to perform $r$ many adjacent transpositions to get to the desired position, which is given by
-$(\suc[n] - r) \downf r$.
 

### 2021-07-02 Pi+/paper/finite.tex
Fix text for the Lehmer-Fin equivalence proof

@@
 
-We can also establish an equivalence between $\im{\normf}$ and $\Sn$, which gives the following.
+We also established an equivalence between $\im{\normf}$ and $\Sn$, which gives the following.
 
@@
 \end{proposition}
-
-\todo{So now, finally, give an explicit example of a normal form, Lehmer immersion}.
 
@@
   \begin{enumerate}
-    \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$.~\label{prop:fin-finexcept-1}
-    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$.~\label{prop:fin-finexcept-2}
-    \item For any $n : \Nat$,
-          \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}[k]} \).~\label{prop:fin-finexcept-3}
+    % \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-1}
+    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-2}
+    \item For any $n : \Nat$, \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[n] \eqv
+            \FinExcept{\suc[n]}[n - k]} \). \label{prop:fin-finexcept-3}
   \end{enumerate}
@@
 
-  The third proposition performs some interesting computations. On the left, we have the type of automorphisms of
-  $\Fin[\suc[n]]$. Assume a particular one $\phi : \Fin[\suc[n]] \xrightarrow{\sim} \Fin[\suc[n]]$. To construct an
-  element of the pair on the right, we pick the first component to be the image of the element $\zero$ under $\phi$.
-  After fixing $k \defeq \phi(\zero)$, we need to compute the bijection in the second component. We remove the two
-  elements $\zero$ and $k$ from $\Fin[\suc[n]]$ and construct an equivalence between the resulting types
-  $\FinExcept{\suc[n]}[\zero]$ and $\FinExcept{\suc[n]}[k]$ as follows:
-  $\FinExcept{\suc[n]}[\zero] \eqv \dsum{k:\Fin[\suc[n]]}{(\phi(\zero) \neq \phi(k))} \eqv \FinExcept{\suc[n]}[k]$.
+  The third proposition performs some combinatorial tricks. On the left, we have the type of automorphisms of
+  $\Fin[\suc[n]]$. Assume a particular $\sigma : \Fin[\suc[n]] \xrightarrow{\sim} \Fin[\suc[n]]$. Pick $k$ to be the
+  inversion count of $n$, the largest element in $\Fin[\suc[n]]$. Then, the image of $n$ under $\phi$ has to be $n - k$,
+  since all other elements in the set are smaller. Removing those two from the domain and codomain of $\phi$, the rest
+  of the elements are fixed by $\sigma$, so we compute the bijection between the rest of the elements.
 
-  To go the other way, given a $k : \Fin[\suc[n]]$ and an equivalence
-  $\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}[k]$ we compute an equivalence as follows:
-  $\Fin[\suc[n]] \eqv \unit \sqcup \FinExcept{\suc[n]}[k] \eqv \unit \sqcup \Fin[n] \eqv \Fin[\suc[n]]$.
+  For the other direction, if we are given a $k$ and a bijection $\pi$ between $\FinExcept{\suc[n]}[n]$ and
+  $\FinExcept{\suc[n]}[n - k]$, we can extend $\pi$ to $\sigma : \Fin[\suc[n]] \eqv \Fin[\suc[n]]$ by inserting the
+  element $n$ at the position $n - k$, resulting in the element $n$ having inversion count $k$.
 \end{proof}
@@
       \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                          \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \Fin[\suc[m]]}          & \text{by~\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-2}} \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} & \text{by~\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-2}} \\
-      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                     & \text{by~\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-3}} \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
+      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                          & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}} \\
     \end{array}
@@
 
-To see how the proof works combinatorically, let us go through an example. We will start from a following bijection $p: \Aut[\Fin[4]]$:
-\begin{align*}
-  p(0) \mapsto & 1 \\
-  p(1) \mapsto & 3 \\
-  p(2) \mapsto & 0 \\
-  p(3) \mapsto & 2
-\end{align*}
-
-At first, we start from a full domain $\Fin[4]$ and full codomain $\Fin[4]$.
-\[
-  \begin{bmatrix}
-    0 \\
-    1 \\
-    2 \\
-    3
-  \end{bmatrix}
-  \mapsto
-  \begin{bmatrix}
-    0 \\
-    1 \\
-    2 \\
-    3
-  \end{bmatrix}
-\]
-
-We remember that the last element of result tuple is $k - 1 = 2$. The appropriate elements on both sides are removed,
-forming, respectively, $\FinExcept{4}[0]$ and $\FinExcept{4}[1]$. These are then
-both transformed to $\Fin[3]$ using
-\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-2}, as shown below.
-
-\[
-  \begin{bmatrix}
-    0 \\
-    1 \\
-    2
-  \end{bmatrix}
-  \equiv
-  \begin{bmatrix}
-    1 \\
-    2 \\
-    3
-  \end{bmatrix}
-  \mapsto
-  \begin{bmatrix}
-    0 \\
-    2 \\
-    3
-  \end{bmatrix}
-  \equiv
-  \begin{bmatrix}
-    0 \\
-    1 \\
-    2
-  \end{bmatrix}
-\]
-
-Then, the original permutation specified that the element $\el{1}$ has to be
-mapped to $\el{3}$. Since we performed renumbering, now it means that element is
-mapped to $\el{2}$ - the resulting tuple now looks like $[2, 1]$. We yet again remove these from the domain and codomain,
-forming $\FinExcept{0}[0]$ and $\FinExcept{0}[2]$.
-
-\[
-  \begin{bmatrix}
-    0 \\
-    1
-  \end{bmatrix}
-  \equiv
-  \begin{bmatrix}
-    1 \\
-    2
-  \end{bmatrix}
-  \mapsto
-  \begin{bmatrix}
-    0 \\
-    2
-  \end{bmatrix}
-  \equiv
-  \begin{bmatrix}
-    0 \\
-    1
-  \end{bmatrix}
-\]
-
-The situation repeats two more times: first, by assinging $\el{0}$ to $\el{1}$,
-giving the tuple $[0, 2, 1]$ and transforming the domain and codomain:
-\[
-  \begin{bmatrix}
-    0
-  \end{bmatrix}
-  \equiv
-  \begin{bmatrix}
-    1
-  \end{bmatrix}
-  \mapsto
-  \begin{bmatrix}
-    0
-  \end{bmatrix}
-  \equiv
-  \begin{bmatrix}
-    0
-  \end{bmatrix}
-\]
-
-And finally, by assinging $\el{0}$ to $\el{0}$, forming a tuple $[0, 0, 2, 1]$ and
-
-At this point we see that the renumbering of the domain is done in an orderly
-way (always shifting by 1), so it can be ignored. On the other hand, the
-ordering of the codomain is what matters. Each time we remove an element, all
-larger elemens are in effect decreased by one.
-
-To remind ourselves, when we introduced the combinatorial interpretation of
-Lehmer codes, the element $\el{r}$ at postition $k$ was said to count the number
-of inversions of the element $\el{r}$. Thus, this is exactly the meachanism in
-which the induction step works.
-
-This corresponds exactly to what happens in
-the \cref{prop:fin-finexcept-3}. Looking at the operation combinatorially, it is
-inserting element $\el{r}$ at the position $k$, meaning that it shifts the
-element $k$places.\jk{needs rewriting after the example is here}
-
-Viewing the codomain as a vector of four elements, element $\el{0}$ has zero
-inverses, element $\el{1}$ has 1 inverse, element $\el{2}$ has 1 inverse, and
-element $3$ has
-
-By composing~\cref{prop:lehmer-aut-equiv} and~\cref{prop:sn-im-lehmer-equiv}, we obtain the final
-equivalence.
+By composing~\cref{prop:lehmer-aut-equiv} and~\cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.
 

### 2021-07-02 Pi+/paper/equivalence.tex
Add some text in section 6

@@
 language $\PiHatLang$, which is a simplified variant of $\PiPlusLang$ that uses adjacent transpositions for combinators,
-but still preserving all the required structure.
+while preserving all the required structure.
 
@@
 Next, we show how to translate $\PiPlusLang$ programs to $\PiHatLang$ and back. $\PiHatLang$ is a simplified variant of
-$\PiPlusLang$, with (unary) natural numbers for types, constructed by $\zero$ and $\suc$, 1-combinators generated by
-adjacent transpositions, and an appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators,
-in~\cref{fig:pihat}. \vc{Something about PROPs.}
+$\PiPlusLang$, with (unary) natural numbers for 0-cells, 1-combinators generated by adjacent transpositions, and an
+appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators, in~\cref{fig:pihat}.
 
@@
 
-As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers.
-However, we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal strucuture.
+As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers, and,
+we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal strucuture. Since each object is a
+natural number generated by $\zero$ and $\suc$, this makes $\PiHatCat$ similar to a \emph{PROP}, that is, a products and
+permutations category.
 
@@
 non-trivial part is coherently generating a $\PiPlusLang$ combinator from $\PiHatLang$ combinators, using adjacent swaps
-to generate big swaps.
+to generate the braiding.
 
@@
 Types in $\PiHatLang$ are interpreted as 0-cells in $\UFin$, that is, a natural number $n$ is mapped to $\Fin[n]$. The
-1-combinators in $\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops $\Aut[\Fin[n]]$, or permutations, in
-each connected component. Finally, 2-combinators are mapped to 2-paths between loops in $\UFin$.
+1-combinators in $\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops $\Aut[\Fin[n]]$, in each connected
+component. In $\PiHatLang$, the 1-combinators are generated by adjacent transpositions, so these can be easily mapped to
+words in $\Sn$ and then to automorphisms using~\cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to
+2-paths between loops in $\UFin$.
 
@@
 0-cells in $\UFin$ are mapped to their cardinalities in $\PiHatLang$, 1-loops are decoded to words in $\Sn$ to generate
-a sequence of adjacent transpositions, producing a 1-combinator in $\PiHatLang$. 2-paths are quoted back to
+a sequence of adjacent transpositions, producing a 1-combinator in $\PiHatLang$. Finally, 2-paths are quoted back to
 2-combinators in $\PiHatLang$.
@@
 
-$\UFin$ as a 1-groupoid, is a (locally strict) weak $(2,0)$-category.
+$\UFin$ as a 1-groupoid, is a (locally strict) weak $(2,0)$-category, and we use this to state our final equivalence.
 
@@
 
-\subsection{$\PiLang$ circuit normalization}
+\subsection{$\PiLang$ circuit normalisation}
 
 To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$, compute a permutation, and quote it back to
-$\PiPlusLang$. Note that the normalisation happens in going from $\PiHatLang$ to $\UFin$ to $\PiHatLang$.
+$\PiPlusLang$. Note that the normalisation happens in going from $\PiHatLang$ to $\UFin$ and back to $\PiHatLang$.
 

### 2021-07-03 Pi+/paper/informalOLD.tex
restructure first few sections

\section{Informal Examples}
\label{sec:informal}

% \todo{Not the right title.}

% \note{This section should explain the main technical parts of the paper
%   informally, without using any technology. Use an example, such as, a
%   reversible language with $\leq 5$ bits, and examples of permutations and
%   transpositions, and when they're equal.}

Instead of working with circuits operating on sequences of bits, we work with circuits operating on algebraic data
types.

\paragraph*{Normalization by evaluation.} Consider the following reversible function:
$\Afun{RESET}(b,b_1,\ldots,b_n) = (b \; \underline{\vee} \; (\bigvee_{i=1}^n b_i),b_1,\ldots,b_n)$ where $\vee$ is
logical-or and $\underline{\vee}$ is exclusive-or. The function sets or resets the first bit depending on whether any of
the other bits is true or not. The natural definition for this function is recursive: modulo some re-shuffling of the
bits, the strategy is to examine the bits $b_i$ one-by-one: if we encounter a true value, we negate the first bit and
terminate; otherwise we continue making recursive calls until we reach the last bit at which point we return the
incoming value with no change:

%%\begin{minipage}{.65\textwidth}
%%  \PiRESET{}
%%\end{minipage}
%%\begin{minipage}{.30\textwidth}
%%  \begin{center}
%%  \resizebox{0.5\textwidth}{!}{\input{diagrams/reset.tikz}}
%%  \end{center}
%%\end{minipage}

\resettwo{}

\noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the
introduction. Our Agda infrastructure provides tools to normalize all programs to their normal form. For \Afun{reset 2}
we get:

\resetnormtwo{}

In principle, the normalized program can be produced following two strategies: (i) by repeatedly applying the rewrite
rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
using a normalization-by-evaluation (NBE) process that evaluates the program to a permutation on a finite set of 8
elements and reifies that permutation back to a program. The key idea of the NBE process is a systematic way to express
permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
left is compiled to the sequence of four adjacent transpositions on the right:

% \note{Motivation: There are two reversible circuits which describe the following permutation. They can be shown to be
%   equal using the 2-combinators.}

% \[
%   \begin{tikzpicture}
%     \begin{knot}[clip width=5]
%       \filldraw (0,5) circle (2pt) node[above] {0};
%       \filldraw (1,5) circle (2pt) node[above] {1};
%       \filldraw (2,5) circle (2pt) node[above] {2};
%       \filldraw (3,5) circle (2pt) node[above] {3};
%       \filldraw (4,5) circle (2pt) node[above] {4};
%       \filldraw (0,0) circle (2pt) node[below] {1};
%       \filldraw (1,0) circle (2pt) node[below] {4};
%       \filldraw (2,0) circle (2pt) node[below] {0};
%       \filldraw (3,0) circle (2pt) node[below] {3};
%       \filldraw (4,0) circle (2pt) node[below] {2};
%       \strand (0,5) .. controls (0.5,0.5) and (1.5,3.5) .. (2,0);
%       \strand (1,5) .. controls (0.75,0.5) and (0.25,3.5) .. (0,0);
%       \strand (2,5) .. controls (2.5,2.5) and (3.5,1.5) .. (4,0);
%       \strand (3,5) .. controls (4.5,2.5) and (4,1.5) .. (3,0);
%       \strand (4,5) .. controls (3.5,2.5) and (1.5,2.5) .. (1,0);
%       \flipcrossings{4,5};
%     \end{knot}
%   \end{tikzpicture}
% \]

\paragraph*{Program synthesis.} The NBE process embodies a quoting mechanism that synthesizes programs from
permutations. Indeed, instead of writing a program for \Afun{reset 2}, one might simply specify the desired permutation as:

\resetperm{}

\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Simply
``quoting'' this permutation generates the same normalized program \Afun{reset2Norm} above.

\paragraph*{Program equivalence.} The permutation above reveals another way to think about the desired program: it is a
special addition circuit that keeps 0 and 4 fixed but otherwise adds 4 modulo 8 to its input. From this specification,
one can use a standard synthesis algorithm for reversible circuits~\cite{10.1145/775832.775915} to generate the
following program:

\adder{}

\noindent The \Afun{adder3} program looks nothing like the original \Afun{reset 2} program and yet they both have the
same normal form thus establishing their equivalence. The reader can check that this circuit is the same as
\verb|reversibleOr1| from the introduction.

% Now imagine we want to write the following reversible function:

% 0 -> 0
% 8 -> 8
% n -> n + 8 `mod` 16

% \note{This might be followed by a section which explains the syntax of Pi.}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% fill-column: 120
%%% End:

### 2021-07-03 Pi+/paper/reversible1.tex
restructure first few sections

@@
 \label{langeqeq}
+\label{sec:informal}
+
+%%%%%%%%%%%%%%%%%
+\subsection{The $\Pi$ Family of Languages}
+
+\note{Start with conherence conditions and folklore result; cite ESOP paper as a syntactic presentation; then go into Pi details}
+
+%%%%%%%%%%%%%%%%%
+\subsection{The $\Pi$ Family of Languages}
+
+\note{Discuss level 2; even if briefly}
 
@@
   rc2011,James:2012:IE:2103656.2103667,Carette2016}.
-
-%%%%%%%%%%%%%%%%%
-\subsection{The $\Pi$ Family of Languages}
 
@@
 it negates the target bit if both control bits are set.
+
+% \todo{Not the right title.}
+
+% \note{This section should explain the main technical parts of the paper
+%   informally, without using any technology. Use an example, such as, a
+%   reversible language with $\leq 5$ bits, and examples of permutations and
+%   transpositions, and when they're equal.}
+
+We now revisit the examples in the introduction but instead of working with circuits operating on sequences of bits, we
+work with circuits operating on algebraic data types.
+
+\paragraph*{Normalization by evaluation.} Consider the following reversible function:
+$\Afun{RESET}(b,b_1,\ldots,b_n) = (b \; \underline{\vee} \; (\bigvee_{i=1}^n b_i),b_1,\ldots,b_n)$ where $\vee$ is
+logical-or and $\underline{\vee}$ is exclusive-or. The function sets or resets the first bit depending on whether any of
+the other bits is true or not. The natural definition for this function is recursive: modulo some re-shuffling of the
+bits, the strategy is to examine the bits $b_i$ one-by-one: if we encounter a true value, we negate the first bit and
+terminate; otherwise we continue making recursive calls until we reach the last bit at which point we return the
+incoming value with no change:
+
+%%\begin{minipage}{.65\textwidth}
+%%  \PiRESET{}
+%%\end{minipage}
+%%\begin{minipage}{.30\textwidth}
+%%  \begin{center}
+%%  \resizebox{0.5\textwidth}{!}{\input{diagrams/reset.tikz}}
+%%  \end{center}
+%%\end{minipage}
+
+\resettwo{}
+
+\noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
+provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
+reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the
+introduction. Our Agda infrastructure provides tools to normalize all programs to their normal form. For \Afun{reset 2}
+we get:
+
+\resetnormtwo{}
+
+In principle, the normalized program can be produced following two strategies: (i) by repeatedly applying the rewrite
+rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
+using a normalization-by-evaluation (NBE) process that evaluates the program to a permutation on a finite set of 8
+elements and reifies that permutation back to a program. The key idea of the NBE process is a systematic way to express
+permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
+left is compiled to the sequence of four adjacent transpositions on the right:
+
+% \note{Motivation: There are two reversible circuits which describe the following permutation. They can be shown to be
+%   equal using the 2-combinators.}
+
+% \[
+%   \begin{tikzpicture}
+%     \begin{knot}[clip width=5]
+%       \filldraw (0,5) circle (2pt) node[above] {0};
+%       \filldraw (1,5) circle (2pt) node[above] {1};
+%       \filldraw (2,5) circle (2pt) node[above] {2};
+%       \filldraw (3,5) circle (2pt) node[above] {3};
+%       \filldraw (4,5) circle (2pt) node[above] {4};
+%       \filldraw (0,0) circle (2pt) node[below] {1};
+%       \filldraw (1,0) circle (2pt) node[below] {4};
+%       \filldraw (2,0) circle (2pt) node[below] {0};
+%       \filldraw (3,0) circle (2pt) node[below] {3};
+%       \filldraw (4,0) circle (2pt) node[below] {2};
+%       \strand (0,5) .. controls (0.5,0.5) and (1.5,3.5) .. (2,0);
+%       \strand (1,5) .. controls (0.75,0.5) and (0.25,3.5) .. (0,0);
+%       \strand (2,5) .. controls (2.5,2.5) and (3.5,1.5) .. (4,0);
+%       \strand (3,5) .. controls (4.5,2.5) and (4,1.5) .. (3,0);
+%       \strand (4,5) .. controls (3.5,2.5) and (1.5,2.5) .. (1,0);
+%       \flipcrossings{4,5};
+%     \end{knot}
+%   \end{tikzpicture}
+% \]
+
+\paragraph*{Program synthesis.} The NBE process embodies a quoting mechanism that synthesizes programs from
+permutations. Indeed, instead of writing a program for \Afun{reset 2}, one might simply specify the desired permutation as:
+
+\resetperm{}
+
+\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
+\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
+which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Simply
+``quoting'' this permutation generates the same normalized program \Afun{reset2Norm} above.
+
+\paragraph*{Program equivalence.} The permutation above reveals another way to think about the desired program: it is a
+special addition circuit that keeps 0 and 4 fixed but otherwise adds 4 modulo 8 to its input. From this specification,
+one can use a standard synthesis algorithm for reversible circuits~\cite{10.1145/775832.775915} to generate the
+following program:
+
+\adder{}
+
+\noindent The \Afun{adder3} program looks nothing like the original \Afun{reset 2} program and yet they both have the
+same normal form thus establishing their equivalence. The reader can check that this circuit is the same as
+\verb|reversibleOr1| from the introduction.
+
+% Now imagine we want to write the following reversible function:
+
+% 0 -> 0
+% 8 -> 8
+% n -> n + 8 `mod` 16
+
+% \note{This might be followed by a section which explains the syntax of Pi.}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "main"
+%%% fill-column: 120
+%%% End:
 

### 2021-07-03 Pi+/paper/equivalence.tex
restructure

@@
 
+We now revisit the examples in the introduction but instead of working with circuits operating on sequences of bits, we
+work with circuits operating on algebraic data types.
+
+\paragraph*{Normalization by evaluation.} Consider the following reversible function:
+$\Afun{RESET}(b,b_1,\ldots,b_n) = (b \; \underline{\vee} \; (\bigvee_{i=1}^n b_i),b_1,\ldots,b_n)$ where $\vee$ is
+logical-or and $\underline{\vee}$ is exclusive-or. The function sets or resets the first bit depending on whether any of
+the other bits is true or not. The natural definition for this function is recursive: modulo some re-shuffling of the
+bits, the strategy is to examine the bits $b_i$ one-by-one: if we encounter a true value, we negate the first bit and
+terminate; otherwise we continue making recursive calls until we reach the last bit at which point we return the
+incoming value with no change:
+
+%%\begin{minipage}{.65\textwidth}
+%%  \PiRESET{}
+%%\end{minipage}
+%%\begin{minipage}{.30\textwidth}
+%%  \begin{center}
+%%  \resizebox{0.5\textwidth}{!}{\input{diagrams/reset.tikz}}
+%%  \end{center}
+%%\end{minipage}
+
+\resettwo{}
+
+\noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
+provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
+reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the
+introduction. Our Agda infrastructure provides tools to normalize all programs to their normal form. For \Afun{reset 2}
+we get:
+
+\resetnormtwo{}
+
+In principle, the normalized program can be produced following two strategies: (i) by repeatedly applying the rewrite
+rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
+using a normalization-by-evaluation (NBE) process that evaluates the program to a permutation on a finite set of 8
+elements and reifies that permutation back to a program. The key idea of the NBE process is a systematic way to express
+permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
+left is compiled to the sequence of four adjacent transpositions on the right:
+
+% \note{Motivation: There are two reversible circuits which describe the following permutation. They can be shown to be
+%   equal using the 2-combinators.}
+
+% \[
+%   \begin{tikzpicture}
+%     \begin{knot}[clip width=5]
+%       \filldraw (0,5) circle (2pt) node[above] {0};
+%       \filldraw (1,5) circle (2pt) node[above] {1};
+%       \filldraw (2,5) circle (2pt) node[above] {2};
+%       \filldraw (3,5) circle (2pt) node[above] {3};
+%       \filldraw (4,5) circle (2pt) node[above] {4};
+%       \filldraw (0,0) circle (2pt) node[below] {1};
+%       \filldraw (1,0) circle (2pt) node[below] {4};
+%       \filldraw (2,0) circle (2pt) node[below] {0};
+%       \filldraw (3,0) circle (2pt) node[below] {3};
+%       \filldraw (4,0) circle (2pt) node[below] {2};
+%       \strand (0,5) .. controls (0.5,0.5) and (1.5,3.5) .. (2,0);
+%       \strand (1,5) .. controls (0.75,0.5) and (0.25,3.5) .. (0,0);
+%       \strand (2,5) .. controls (2.5,2.5) and (3.5,1.5) .. (4,0);
+%       \strand (3,5) .. controls (4.5,2.5) and (4,1.5) .. (3,0);
+%       \strand (4,5) .. controls (3.5,2.5) and (1.5,2.5) .. (1,0);
+%       \flipcrossings{4,5};
+%     \end{knot}
+%   \end{tikzpicture}
+% \]
+
+\paragraph*{Program synthesis.} The NBE process embodies a quoting mechanism that synthesizes programs from
+permutations. Indeed, instead of writing a program for \Afun{reset 2}, one might simply specify the desired permutation as:
+
+\resetperm{}
+
+\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
+\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
+which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Simply
+``quoting'' this permutation generates the same normalized program \Afun{reset2Norm} above.
+
+\paragraph*{Program equivalence.} The permutation above reveals another way to think about the desired program: it is a
+special addition circuit that keeps 0 and 4 fixed but otherwise adds 4 modulo 8 to its input. From this specification,
+one can use a standard synthesis algorithm for reversible circuits~\cite{10.1145/775832.775915} to generate the
+following program:
+
+\adder{}
+
+\noindent The \Afun{adder3} program looks nothing like the original \Afun{reset 2} program and yet they both have the
+same normal form thus establishing their equivalence. The reader can check that this circuit is the same as
+\verb|reversibleOr1| from the introduction.
+
+% Now imagine we want to write the following reversible function:
+
+% 0 -> 0
+% 8 -> 8
+% n -> n + 8 `mod` 16
+
 %%% Local Variables:

### 2021-07-03 Pi+/paper/reversible1.tex
restructure first few sections
restructure

--- 2021-07-02 Pi+/paper/reversible1.tex
@@
 \label{langeqeq}
+\label{sec:informal}
+
+%%%%%%%%%%%%%%%%%
+\subsection{Coherence Conditions}
+
+\note{Start with conherence conditions and folklore result; cite ESOP paper as a syntactic presentation; then go into Pi details}
+
+%%%%%%%%%%%%%%%%%
+\subsection{The $\Pi$ Family of Languages}
+
+\note{Discuss level 2; even if briefly}
 
@@
 it negates the target bit if both control bits are set.
+
+Explain how Pi can be used to represent circuit; perhaps show how ifc corresponds to two cnots with negation surrounding
+the second one; talk about sequences of wires as a terrible representation
+
+% \todo{Not the right title.}
+
+% \note{This section should explain the main technical parts of the paper
+%   informally, without using any technology. Use an example, such as, a
+%   reversible language with $\leq 5$ bits, and examples of permutations and
+%   transpositions, and when they're equal.}
 

### 2021-07-04 Pi+/paper/equivalence.tex
Revise UFin section and fit into page budget

@@
 rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
-using a normalization-by-evaluation (NBE) process that evaluates the program to a permutation on a finite set of 8
-elements and reifies that permutation back to a program. The key idea of the NBE process is a systematic way to express
+using a normalization-by-evaluation (NbE) process that evaluates the program to a permutation on a finite set of 8
+elements and reifies that permutation back to a program. The key idea of the NbE process is a systematic way to express
 permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
@@
 
-\paragraph*{Program synthesis.} The NBE process embodies a quoting mechanism that synthesizes programs from
+\paragraph*{Program synthesis.} The NbE process embodies a quoting mechanism that synthesizes programs from
 permutations. Indeed, instead of writing a program for \Afun{reset 2}, one might simply specify the desired permutation as:

### 2021-07-04 Pi+/paper/finite.tex
Add some text in section 6
Revise UFin section and fit into page budget

@@
 \section{The groupoid of finite types}~\label{sec:finite}
+
+We would like to give a syntactic presentation for this groupoid. Note that, in $\BFin$, the set of bijections on a
+finite set of cardinality $n$ has a group structure -- the identity function is the neutral element, group
+multiplication is given by function composition, and the group inverse is given by inverting the bijection. This is the
+automorphism group of the finite set, which is also known to be the finite symmetric group $\Sn[n]$.
+
+Groups can be defined by listing the carrier set and the operations of the group, or they can be described using a group
+presentation, which gives the generators and relations of a group. The last step of our technical development is to use
+a group presentation for $\Sn$ to give a presentation for the higher cells of the groupoid $\UFin$. Using this
+characterisation, we will obtain our completeness result.
 
@@
 
-By composing~\cref{prop:lehmer-aut-equiv} and~\cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.
+By composing~\cref{prop:lehmer-aut-equiv} and~\cref{prop:sn-lehmer-fin-equiv}, we obtain the final equivalence.
 

### 2021-07-04 Pi+/paper/ufin.tex
Revise UFin section and fit into page budget

@@
 
-The groupoid $\BFin$ that we presented, is strict, that is, two equal permutations are extensionally identified, without
+The groupoid $\BFin$ that we presented is strict, that is, two equal permutations are extensionally identified, without
 an explicit witness for the equality. Our first step is to weaken this groupoid $\BFin$ into a weak 2-category. In a
 weak category, equational axioms of the category are replaced by higher invertible cells. For example, there are higher
-invertible cells for the left and right units of composition and associativity. The higher cells have their own
-coherences as well, the 2-cells can be composed with 1-cells by whiskering, and so on.
+invertible cells for the left and right units of composition, and associativity. The higher cells come with their own
+coherence data as well, such as vertical composition, and horizontal composition, or whiskering.\todo{define 2-category
+  in appendix.}
 
 HoTT provides a rich internal language for working with weak higher groupoids, using the ``Types are weak
-$\infty$-groupoids'' correspondence. We will make this precise by observing that a 1-groupoid in HoTT gives a locally
-strict $(2,0)$-category, using the identity type for the higher cells. The groupoid $\BFin$ can be written in HoTT as a
-type $\UFin$, which is a 1-groupoid.
+$\infty$-groupoids'' correspondence. Using the identity type for morphisms, a 1-groupoid in HoTT has points for 0-cells,
+1-paths for 1-cells, and 2-paths for 2-cells, but there is at most one 2-cell between compatible 1-cells.~\footnote{This
+  is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that
+  is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type $\UFin$, which is a 1-groupoid.
 
-Secondly, we would like to like to give a syntactic presentation for this groupoid, by describing exactly how the
-1-cells and 2-cells are generated. Note that, in $\BFin$, the set of bijections on a finite set of cardinality $n$ has a
-group structure, the identity function is the neutral element, group multiplication is given by function composition,
-and the group inverse is given by inverting the bijection. This is the automorphism group of the finite set, which is
-also known to be the finite symmetric group $\Sn[n]$.
-
-Groups can be defined by listing the carrier set and the operations of the group, or they can be described using a group
-presentation, which gives the generators and relations of a group. The last step of our technical development is to use
-a group presentation for $\Sn$ to give a presentation for the higher cells of the groupoid $\UFin$. Using this
-characterisation, we will obtain our completeness result.
-
-% In the previous section, we have picked a particular encoding of finite types: as sequences of elements in some
-% canonical order.
-
-% This is not a ``free'' presentation. Some properties of the encoding might pollute our semantics. We want to make that
-% any theorems / properties of the semantics we prove apply to the source language (under various interpretations).
-
-% This motivates us to use a different semantic representation: a univalent universe of finite types which contains the
-% finite types and nothing but the finite types and is defined such as alternative representations of finite types are
-% identified. This guarantees that we can get a proper, sound and complete, semantics for our language.
-
-% \begin{center}
-% \begin{tikzpicture}[scale=1,every node/.style={scale=1}]
-% \usetikzlibrary{arrows}
-% \usetikzlibrary{shapes}
-% \node[ultra thick, draw=blue, ellipse, minimum width=300pt, minimum height=300pt, align=center] {};
-% \node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (a) {$\{ 1 , 2 \} \uplus \{ 3 \}$};
-% \node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (b) at ([shift=({70pt,-70pt})]a) {$\{ 1 \} \uplus \{ 2, 3 \}$};
-% \node[ultra thick, draw=blue, ellipse, minimum width=70pt, minimum height=40pt, align=center] (c) at
-% ([shift=({-70pt,-70pt})]a) {$\{ 10 \} \uplus \{ 20, 30 \}$};
-% \node (d) at ([shift=({50pt,-10pt})]a) {$\cdots$};
-% \draw[thick,-] (a) to node[draw=none,right] {$\equiv$} (b);
-% \draw[thick,-] (b) to node[draw=none,below] {$\equiv$} (c);
-% \draw[thick,-] (a) to node[draw=none,left] {$\equiv$} (c);
-% \end{tikzpicture}
-% \end{center}
+\begin{toappendix}
+  We give an example of the groupoid structure on a 3-element set.
 
@@
 
-Given an operational model of how a type A reduces to a type B
-Given a denotational model of which types A and B are equivalent
-Want soundness and completeness results
-Guarantee that the two notions of equality are themselves “equivalent”
-Generalized “identity of indiscernibles”
+\end{toappendix}
 
-Previous semantics adequate for reasoning about evaluation and for checking the equivalence of two circuits.
+\begin{toappendix}
 
-Want NBE, want rewriting system of program equivalence, what else? So we need to move to proof-relevant higher-category
-groupoid semantics instead of set-theoretic.
-
-Now the question is, what does it mean for a compiler?
-
-My opinion is that it is still the same, what rewrites to apply depends
-on the choices made by the compiler writer, but we can think about that.
-
-%%%
   \subsection{To Put Somewhere or Delete}
@@
   \[\begin{array}{rcl}
-    \denot{\onet + (\onet + \onet)} &=& \{ 0,1,2 \} \\
-    \denot{(\onet + \onet) + \onet} &=& \{ 0,1,2 \} \\
+      \denot{\onet + (\onet + \onet)}                                         & = & \{ 0,1,2 \} \\
+      \denot{(\onet + \onet) + \onet}                                         & = & \{ 0,1,2 \} \\
       \\
-    \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (0~1~2) \\
-    \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} &=& (2~0~1)
+      \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} & = & (0~1~2)     \\
+      \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet}   & = & (2~0~1)
     \end{array}\]
@@
   \[\begin{array}{rcl}
-    \mathit{enum}(\zerot) &=& [~] \\
-    \mathit{enum}(\onet) &=& [ ~\Acon{tt}~ ] \\
-    \mathit{enum}(A + B) &=& \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B) \\
-    \mathit{enum}(A \times B) &=& \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
+      \mathit{enum}(\zerot)     & = & [~]                                                                                                          \\
+      \mathit{enum}(\onet)      & = & [ ~\Acon{tt}~ ]                                                                                              \\
+      \mathit{enum}(A + B)      & = & \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B)          \\
+      \mathit{enum}(A \times B) & = & \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
     \end{array}\]
@@
   \[\begin{array}{rcl}
-    \mathit{enum}(\onet + (\onet + \onet)) &=& [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
-    \mathit{enum}((\onet + \onet) + \onet) &=& [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
+      \mathit{enum}(\onet + (\onet + \onet)) & = & [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
+      \mathit{enum}((\onet + \onet) + \onet) & = & [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
     \end{array}\]
@@
   \[\begin{array}{rcll}
-\sem{A} &=& (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
+      \sem{A} & = & (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
     \end{array}\]
@@
 
+\end{toappendix}
+
 %%%
-\subsection{Univalent Subuniverses}
-\label{sec:univalent}
+\subsection{The Type Theory}~\label{subsec:type-theory}
+% \subsection{Univalent Subuniverses}
+% \label{sec:univalent}
 
-In order to define the semantics, we seek a mathematical structure satisfying the following properties: (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust enough to ensure that equivalent encodings of finite types are identified.
+% In order to define the weak groupoid we're after, we seek a mathematical structure satisfying the following properties:
+% (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust
+% enough to ensure that equivalent encodings of finite types are identified.
 
-In this section, we introduce some basic concepts and notation that we use in Homotopy Type Theory. Then, we define
-univalent subuniverses and discuss some specific examples.
-
-\subsection{The Type Theory}~\label{subsec:type-theory}
-
-We work in Homotopy Type Theory (HoTT), and use the language of the HoTT
-book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use intensional Martin-L\"{o}f Type Theory,
-with a univalent universe $\UU$, and we use Higher Inductive Types (HITs) for propositional truncations and set
-quotients. All arguments will hold in a cubical type theory as well, such
-as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019}. We recall a few basic facts to
-familiarise the reader with the notation we use. For more details about basics of HoTT, we refer the reader to the HoTT
-book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.
-
-Our denotational semantics is given using groupoids, and HoTT provides a rich internal language for describing (weak
-$\infty$-)groupoids. By formalising our results using general tools and principles from HoTT, we're able to use a proof
-assistant to computer-check our proofs. Our results are formalised in the Agda proof assistant using the HoTT-Agda
-library, though it could also be formalised using Cubical Agda~\cite{vezzosiCubicalAgdaDependently2019}. \review{Using
-  the formalisation of the semantics, we are also able to extract a computable procedure for performing normalisation of
-  circuits.}
+To define the weak groupoid we're after, we first introduce some basic concepts and notation from Homotopy Type Theory
+that we use.~\footnote{We use the language of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013},
+that is, we use intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive
+Types (HITs) for propositional truncation and set quotients. All arguments will hold in a Cubical Type Theory as well,
+such
+as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.}
+For more details about HoTT, we refer the reader to the HoTT
+book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}. By formalising our results in HoTT, we're also able to
+use a proof assistant to computer-check our proofs, and our results are formalised in the Agda proof assistant using the
+HoTT-Agda library. Using the formalisation of our semantics, we are able to extract a computable procedure for
+performing NbE and synthesis of reversible circuits, as motivated in~\cref{sec:introduction}.
 
@@
 
-\vc{I'm explaining everything starting from the categorical language. We do care more about the groupoid structure of
-  types, and how to encode groupoids using types, since that's what we use in~\cref{sec:finite}.}
+Given two terms $x:A$ and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the identity type, which is the type
+of equalities or identifications between them. The identity type is generated by reflexivity $\refl_{x} : x \id_{A} x$,
+and the eliminator for the identity type is given by path induction or the $J$-rule (\cref{def:path-induction}). This
+construction can be iterated, giving the identity type between two terms of an identity type, repeating ad infinitum.
+Using the iterated identity type for morphisms, each type is equipped with the structure of a weak $\infty$-groupoid,
+where each morphism satisfies groupoid laws only upto a higher one. Given an arbitrary type (or groupoid) $A$, we list
+some laws that are provable using path induction.
 
-In HoTT, the identity type or the equality type is the type of paths or equalities between two terms of the same type.
-Given two terms $x:A$ and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the equality type between them. In
-book HoTT, the identity type is generated by reflexivity $\refl_{x} : x \id_{A} x$, and the eliminator for the identity
-type is given by path induction or the $J$-rule. The identity type equips each type with the structure of a weak
-$\infty$-groupoid, or a homotopy type. For example, given an arbitrary type (or groupoid) $A$, we list some groupoid
-laws that are provable using path induction.
+\begin{toappendix}
+  \begin{definition}[Path Induction]
+    \label{def:path-induction}
+    Given a type family $C : \dfun{x,y:A}{(x \id_A y)} \to \UU$, and a function $c : \dfun{x:A}{C(x,x,\refl_x)}$, there is
+    a function $f : \dfun{x,y:A}{\dfun{p:x \id_A y}{C(x,y,p)}}$ such that $f(x,x,\refl_x) \defeq c(x)$.
+  \end{definition}
+\end{toappendix}
 
@@
 
-Functions between types are functors between groupoids. Given a function $f : A \to B$, the functorial action
-$\term{ap}_{f}(\blank)$ is given by path induction.
+Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action on
+the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to the
+universe, such as $A \to \UU$. For a type family $P : A \to \UU$ and a point $x : A$, the type $P(x)$ is called the
+fiber over $x$.  The $\term{transport}$ operation, named $\term{tr}$, lifts paths in the indexing type to functions
+between fibers.
+%which is an $A$-indexed family of groupoids.
+
+\begin{gather*}
+  \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
+  \qquad
+  \term{tr}_{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
+\end{gather*}
+
+The type $\dsum{x:A}{P(x)}$ is the collection of all the fibers and is called the total space of $P$. The first
+projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
+fibration, that is, there is a lifting operation (\cref{fig:lift}) which lifts paths in the base space to paths in the
+total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have
 
 \[
-  \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
+  \term{lift}(u,p) : (x , u) \id_{\dsum{x:A}{P(x)}} (y , \transport{P}{p}{u})
 \]
 
-Type families are functions from a type to the universe, such as $A \to \UU$, which is an $A$-indexed family of
-groupoids. The $\term{transport}$ operation gives the functorial action of paths in the indexing type, which is also
-defined by path induction. If $P : A \to \UU$ is a type family, then for a path $x \id_{A} y$, we have
+% where $\tr{p}{u}$ is shorthand for $$.
 
-\[
-  \transport{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
-\]
-
-A type family can also be seen as a fibration, from the topological viewpoint. For a type family $P : A \to \UU$ and a
-point $x : A$, the type $P(x)$ gives the fiber over $x$. The type $\dsum{x:A}{P(x)}$ is the collection of all the fibers
-and is called the total space of $P$. The first projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ from the total space to
-the base space $A$ is then a fibration. The lifting operation lifts paths in the base space to paths in the total space.
-Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have
-
-\[
-  \term{lift}(u,p) : (x , u) \id_{\dsum{x:A}{P(x)}} (y , \tr{p}{u})
-\]
-
-where $\tr{p}{u}$ is shorthand for $\transport{P,p}(u)$.
-
+\begin{figure}
   \begin{center}
@@
       \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
-    \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\tr{p}{u}}$}] (b2) at (.5,7.2) {};
+      \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\transport{P}{p}{u}}$}] (b2) at (.5,7.2) {};
       \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
@@
   \end{center}
+  \caption{Lifting operation in $P$}
+  \label{fig:lift}
+\end{figure}
 
-Further, using the groupoid structure of $A$, we can show that $\term{transport}$ lifts paths to equivalences, we define
+Further, using the groupoid structure of $A$, we can show that $\term{transport}$ lifts paths to equivalences of fibers.
 
@@
 \begin{definition}[Univalent Fibration]
-  $P$ is a univalent type family (or simply a univalent fibration) if $\tptEqv{P}$ is an equivalence.
+  $P$ is a univalent type family (or, $\fst : {\dsum{x:A}{P(x)}} \to A$ is a univalent fibration) if $\tptEqv{P}$ is an
+  equivalence.
 \end{definition}
@@
 
-Voevodsky's \emph{univalence} principle characterises paths in the universe. It says that equivalent types are equal, or
+Voevodsky's \emph{univalence} principle characterises paths in the universe. It says that equivalent types are equal, or,
 the following function is an equivalence.
@@
 
+\subsubsection{Homotopy Types}
+
+A type is \emph{contractible} (-2-type) if it has a unique element, that is, there is a center of contraction and every
+other point is equal to it. A type is a \emph{proposition} (-1-type) if its equality types are contractible, that is, it
+has at most one inhabitant. Iterating this, we can define sets or 0-types (whose equality types are propositions) and
+1-groupoids or 1-types (whose equality types are sets), and similarly, higher homotopy $n$-types.
+
+\begin{gather*}
+  \begin{aligned}
+    \isContr{A} & \defeq \dsum{x:A}{\dfun{y:A}{y \id x}} \\
+    \isProp{A}  & \defeq \dfun{x,y:A}{\isContr{x \id y}}
+  \end{aligned}
+  \qquad
+  \begin{aligned}
+    \isSet{A} & \defeq \dfun{x,y:A}{\isProp{x \id y}} \\
+    \isGpd{A} & \defeq \dfun{x,y:A}{\isSet{x \id y}}
+  \end{aligned}
+\end{gather*}
+
 \subsubsection{Higher Inductive Types}
 
-\todo{Explain homotopy types, $\hProp$, $\hSet$, etc.}
+Higher Inductive Types generalise Inductive Types, by allowing path constructors besides point constructors. While point
+constructors generate the elements of the type, path constructors generate equalities between points in the type. We
+describe a few basic HITs that we use.
 
-\vc{These are placeholder definitions, need informal explanations and references to the book.}
+Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, squashes the elements of $A$ turning it into a
+proposition. It is given by a HIT with a point constructor $\trunc{\blank} : A \to \Trunc{A}$, and a path constructor
+$\term{trunc}(x,y) : x \id_{\Trunc{A}} y$, which equates every pair of points in the truncation
+(see~\cref{def:prop-trunc}).
 
+\begin{toappendix}
   \begin{definition}[Propositional Truncation]
+    \label{def:prop-trunc}
     Given a type $A$, the propositional truncation $\Trunc[-1]{A}$, or simply $\Trunc{A}$, is a higher inductive type
@@
   \end{definition}
+\end{toappendix}
 
+Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes an $\hSet$ $A$ and a relation $R : A \to A \to
+  \UU$. It has an inclusion of points $q : A \to \quot{A}{R}$, and adds paths between related pairs of elements
+$\quotrel : R(x,y) \to q(x) \id_{\quot{A}{R}} q(y)$ (see~\cref{def:set-quot}).  We recall that the quotient is
+\emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff $(q(x) \id_{\quot{A}{R}}
+  q(y))$.
+
+\begin{toappendix}
   \begin{definition}[Set Quotient]
+    \label{def:set-quot}
     Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
@@
   \end{definition}
-
-We recall that quotients in HoTT are \emph{effective}, that is, if $R$ is an equivalence relation, we have
-$R(x,y) \eqv (q(x) \id_{\quot{A}{R}} q(y))$.
+\end{toappendix}
 
@@
 
-\todo{Explain motivation. Starting from a univalent universe which classifies all types, we want to define a subuniverse
-  which classifies only certain types, for example, types that satisfy some property that we want.}
+Starting from a univalent universe which classifies all types, we want to define a subuniverse which classifies only
+certain types, for example, types that satisfy some desired property. We use a prop-valued type family, that is, a
+predicate, which picks out only those types, and collect them into a univalent subuniverse. Being univalent ensures that
+the equality type of the ambient universe is reflected in the subuniverse.
 
@@
     \item a code $U : \UU$,
-    \item a decoding function $\El : U \to \UU$.
+    \item a decoding type family $\El : U \to \UU$.
   \end{itemize}
-  If $\El$ is a univalent fibration, $U$ is a univalent universe.
+  If $\El$ is univalent, $U$ is a univalent universe.
 \end{definition}
@@
 \begin{definition}[Subuniverse]
-  A subtype is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $\forall x, \isProp{P(x)}$. A
-  subuniverse generated by a subtype has $U \defeq \dsum{X:\UU}{P(X)}$ and $\El \defeq \fst$.
+  A subtype is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition for every
+  $X$. A subuniverse generated by a subtype has $U \defeq \dsum{X:\UU}{P(X)}$ and $\El \defeq \fst$.
 \end{definition}
@@
 
-\begin{proof}
+\begin{appendixproof}
   Suppose $(U, \El) \defeq (\dsum{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
@@
   inverse follows by calculation using the appropriate computation rules.
-\end{proof}
+\end{appendixproof}
 
-\begin{example}[$\BAut$]
-  The type of self-equivalences on a type $T$ is defined as $\Aut[T] \defeq T \eqv T$, which is an $\infty$-group. The
-  subuniverse of types that are merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the
-  classifying space of $\Aut[T]$, as justified by~\cref{lem:loop-deloop}. We write $T_{o} \defeq (T, \trunc{\refl_{T}})$
-  for the image of the inclusion of $T$.
-\end{example}
-
-\begin{example}
-  The univalent subuniverse of types merely equal to $\Bool$ is given by $\BAut[\Bool]$. This can be used to give the
-  denotational semantics of a 2-bit reversible language~\cite{caretteReversibleProgramsUnivalent2018}.
-\end{example}
+The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is subtle. We
+define a type to be (Bishop-) finite if it is equal to some finite set, and we setup the predicate $\isFin$ to do so.
 
 \begin{definition}[$\Fin$]
-  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by cardinality. It is defined equivalently in
-  two different ways,
+  The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by their cardinality. It is defined
+  equivalently in two different ways,
   \begin{gather*}
@@
     \end{aligned}
+    \qquad
     \begin{aligned}
@@
 \end{definition}
-
-\begin{example}
-  For any $n : \Nat$, we define the univalent subuniverse of types merely equal to $\Fin[n]$ as $\BAut[\Fin[n]]$. This
-  is the universe of finite sets with cardinality equal to $n$. We write $F_{n} \defeq (\Fin[n], \trunc{\refl})$ for the
-  image of the inclusion of $\Fin[n]$.
-\end{example}
 
@@
 
-\begin{proof}
+\begin{appendixproof}
   Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
@@
   applying the first projection.
-\end{proof}
+\end{appendixproof}
 
-\begin{example}
+\begin{definition}
   The univalent subuniverse of \emph{all finite types} is given by
@@
   \]
-  We write $F_{n} \defeq (\Fin[n], n, \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$. We use this as the
-  denotational semantics to interpret our reversible programming language $\PiPlusLang$.
-\end{example}
+  We write $F_{n} \defeq (\Fin[n], n, \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$.
+\end{definition}
 
-\review{We characterise the path space of univalent subuniverses.}
+We also consider a different way of thinking about $\UFin$, using the subuniverse $\BAut[T]$.
+
+\begin{definition}[$\BAut$]
+  The predicate $P(X) \defeq \Trunc[-1]{X \id T}$ picks out exactly those types that are merely equal to $T$, and this
+  generates the subuniverse
+  \[
+    \BAut[T] \defeq \Sub{T}.
+  \]
+  We write $T_0 \defeq (T, \trunc{\refl_{T}})$ for the image of the inclusion of $T$ in $\BAut[T]$
+\end{definition}
+
+Using $\BAut$, we can talk about types that have a specified cardinality, for example, the subuniverse of 2-element sets
+is given by $\BAut[\Bool]$. This can be used to give the denotational semantics of a 1-bit reversible
+language~\cite{caretteReversibleProgramsUnivalent2018}.
+
+\begin{definition}
+  For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.
+  Note that, $\UFin$ can be equivalently seen as the collection of all types of finite cardinality, that is, $\UFin \eqv
+    \displaystyle\sqcup_{n:\Nat} \UFin[n]$.
+\end{definition}
+
+Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
+has one point $T_0$, and 1-paths $T_0 \id_{\BAut[T]} T_0$, that is, loops, and higher paths between these loops. The
+type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$ is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
+the automorphism group of $T$.
 
 \begin{proposition}
-  If $T$ is an $n$-type, $\BAut[T]$ is an $(n+1)$-type.
+  \leavevmode
+  \begin{enumerate}
+    \item If $T$ is an $n$-type, $\BAut[T]$ is an $(n+1)$-type.
+    \item For any $T : \UU$, $\BAut[T]$ is 0-connected.
+    \item For any $T : \UU$, \( \loopspace[\BAut[T],T_{0}] \eqv \Aut[T] \). \label{lem:loop-deloop}
+  \end{enumerate}
 \end{proposition}
 
-\begin{proof}
+\begin{appendixproof}
   We need to show that the equality type of $\BAut[T]$ is an $n$-type. Assume $X, Y : \BAut[T]$. Since $\BAut[T]$ is a
@@
   $\fst(X) \eqv \fst(Y)$ is an $n$-type, and hence $X \id Y$ is an $n$-type.
-\end{proof}
 
-\begin{proposition}
-  For any $T : \UU$, $\BAut[T]$ is 0-connected.
-\end{proposition}
-
-\begin{corollary}
-  $\UFin$ is a pointed, connected, 1-type, that is, a 1-groupoid, or has h-level 3.
-\end{corollary}
-
-\begin{lemma}~\label{lem:loop-deloop}
-  \[
-    \loopspace[\BAut[T],T_{0}] \eqv \Aut[T]
-  \]
-\end{lemma}
-
-\begin{proof}
   Since $\BAut[T]$ is a univalent universe, it follows that
@@
   \]
-\end{proof}
+\end{appendixproof}
 
 \begin{corollary}
-  For every $n:\Nat$,
-  \[
-    \loopspace[\UFin[n],F_{n}] \eqv \Aut[\Fin[n]]
-  \]
+  $\UFin$ is a pointed, connected, 1-type, that is, a 1-groupoid. For every $n:\Nat$, \( \loopspace[\UFin[n],F_{n}] \eqv
+  \Aut[\Fin[n]] \).
 \end{corollary}
-
-\[
-  \begin{tikzcd}
-    F_{0}
-    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
-    & F_{1}
-    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
-    & F_{2}
-    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
-    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=8em, in=125, out=55]
-    & \ldots
-    & F_{n}
-    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
-    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=8em, in=125, out=55]
-    \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=12em, in=135, out=45]
-    & \ldots
-  \end{tikzcd}
-\]
+\note{Diagram below?}
+% \[
+%   \begin{tikzcd}
+%     F_{0}
+%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
+%     & F_{1}
+%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
+%     & F_{2}
+%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
+%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=8em, in=125, out=55]
+%     & \ldots
+%     & F_{n}
+%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=4em, in=115, out=65]
+%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=8em, in=125, out=55]
+%     \arrow[""{anchor=center, inner sep=0}, no head, loop, distance=12em, in=135, out=45]
+%     & \ldots
+%   \end{tikzcd}
+% \]
 
@@
 
-\todo{rig}
-
-We describe the symmetric monoidal structure of the groupoid $\UFin$.
-
-First, we observe a few equivalences.
+The groupoid $\UFin$ has two symmetric monoidal structures, the additive and the multiplicative one, and the
+multiplicative tensor product distributes over the additive one. To construct these, we first define some equivalences
+on $\Fin$, and some general type isomorphisms.
 
@@
   For any $n, m : \Nat$,
-  \begin{align*}
+  \begin{gather*}
+    \begin{aligned}
       \Fin[0]                & \eqv \bot        \\
       \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m] \\
-  \end{align*}
+    \end{aligned}
+    \qquad
+    \begin{aligned}
+      \Fin[1]                & \eqv \top        \\
+      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m] \\
+    \end{aligned}
+  \end{gather*}
   and for any types $X, Y, Z$,
-  \begin{align*}
+  \begin{gather*}
+    \begin{aligned}
       \bot \sqcup X         & \eqv X                     \\
@@
       X \sqcup Y            & \eqv Y \sqcup Y            \\
-  \end{align*}
-\end{proposition}
-
-\begin{proposition}
-  For any $n, m : \Nat$,
-  \begin{align*}
-    \Fin[1]                & \eqv \top        \\
-    \Fin[n] \times \Fin[m] & \eqv \Fin[n * m] \\
-  \end{align*}
-  and for any types $X, Y, Z$,
-  \begin{align*}
+    \end{aligned}
+    \qquad
+    \begin{aligned}
       \top \times X         & \eqv X                                \\
@@
       X \times (Y \sqcup Z) & \eqv (X \times Y) \sqcup (X \times Z)
-  \end{align*}
+    \end{aligned}
+  \end{gather*}
 \end{proposition}
 
-We lift these equivalences to $\UFin$ giving it the additive and multiplicative symmetric monoidal structures,
-$(O, \oplus)$ and $(I, \otimes)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$,
-and the symmetry isomorphism $\mathcal{B}_{X,Y}$. Note that types in $\UFin$ are $\hSet$s since they're merely
-equivalent to $\Fin[n]$ for some $n : \Nat$.
+We lift these equivalences to $\UFin$ giving it the additive and multiplicative symmetric monoidal structures, $(F_0,
+  \sqcup)$ and $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and
+the braiding isomorphism $\mathcal{B}_{X,Y}$. These isomorphisms satisfy the Mac Lane coherence conditions for symmetric
+monoidal categories, that is, the triangle, pentagon, and hexagon identities, and the symmetry of the braiding, upto
+2-paths in $\UFin$.
 
-\todo{combine both symmetric monoidal structures}
-
+\begin{toappendix}
   \begin{definition}
     \begin{align*}
-    I                 & \defeq F_{0}                                       \\
+      O                 & \defeq F_{0}                                       \\
       X \oplus Y        & \defeq X \sqcup Y                                  \\
-    \lambda_{X}       & : I \oplus X \eqv X                                \\
-    \rho_{X}          & : X \oplus I \eqv X                                \\
+      \lambda_{X}       & : O \oplus X \eqv X                                \\
+      \rho_{X}          & : X \oplus O \eqv X                                \\
       \alpha_{X,Y,Z}    & : (X \oplus Y) \oplus Z \eqv X \oplus (Y \oplus Z) \\
@@
   \end{definition}
-
-These isomorphisms satisfy the Mac Lane coherence conditions for symmetric monoidal categories, that is, the triangle,
-pentagon, and hexagon identities, and the syllepsis of the braiding, upto 2-paths in $\UFin$.
-
   \begin{proposition}
@@
   \end{proposition}
+\end{toappendix}
 

### 2021-07-04 Pi+/paper/extra.tex
Revising section 4

@@
 
+Although the meaning of a permutation - that is, the specific bijective function
+that it represents - is in some sense all there is, manipulating them
+syntactically still has its advantages. By writing reversible programs, we think
+of them in an intensional way. Comparing two programs for equality by evaluating
+them on all points in the domain is a very crude - and maybe even inefficient -
+way of doing that.
+
+It is crude, because there is no way of enforcing additional constraints on the
+process of transforming one program into another, and no way of inspecting what
+transformation occured. We can imagine a practical situation in which reversible
+circuts admit one kind of optimization \jk{FPGA?}, but do not admit another,
+even though they are equivalent in the extensional sense - or maybe one of these
+transformations is cheaper than another, or maybe it is important to know which
+transformation did occur for the \jk{producer(?)} to focus their resources of
+improving this kind of transformations. By comparing the circuits extensionally,
+the proof object = the path from one into the other - does not have any
+structure,  it is just a check that the values match.
+
+On the other hand, even though in general, the equality of circuits requires
+exponential time jk{triple check}, the proof object can be still very small.
+This creates structure that can possibly be exploited in a heurisic way.
+\jk{Additionally, if the goal is to convince third-party that two circuits are
+the same, we can just present the proof instead of showing the equality directly
+- for example, by choosing a particularly short one}.
+
+Taking all this into account, it is clear that we must seek a syntactic way of
+computing the meaning of the circuit, instead of just directly evaluating it to
+$\Aut[\Fin[n]]$ - since when we go back and quote the permutation as a circuit
+(in a normal form), there was no "trace" left to see what transformation (what
+sequence of 2-paths) maps the old one into the new one.
+
 \paragraph*{Categorical Connections.}

### 2021-07-04 Pi+/paper/finite.tex
Revising section 4

--- 2021-07-02 Pi+/paper/finite.tex
@@
 
-In this section, we describe \review{the algebraic structure} of the groupoid of
-finite types, and give \review{a computable presentation} for it.
+In the previous~\cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin{n}$
+for every $n:\Nat$, that is, automorphisms of finite sets of size $n$. The loopspace has a group structure -- the
+identity function is the neutral element, group multiplication is given by function composition, and the group inverse
+is given by inverting the bijection. This is also known to be the finite symmetric group $\Sn[n]$. In the following
+sections, we will characterise these permutations syntactically.
 
-\vc{The groupoid of finite types is the free symmetric monoidal groupoid on one
-  generator. This can be presented as an algebraic 2-theory, which is our syntax
-  for $\PiHatLang$. Vertical categorification of natural numbers as a free
-  commutative monoid. See groupoidification.}
+In order to study syntactic descriptions of permutations, we need to decide whether two descriptions refer to the same
+permutation. This problem is well known in group theory as the \emph{word problem} for $\Sn$. Putting it in this form
+also allows us to connect to the broader scope of computational group theory and combinatorics -- we borrow ideas such
+as Coxeter relations and Lehmer codes from there.
 
-\todo{Check Brent Yorgey's thesis?}
-
-In the previous~\cref{sec:univalent}, we established that paths in $\UFin$ are equivalent to families of automorphisms
-of $\Fin{n}$ for every $n:\Nat$, that is, bijections on finite sets of size $n$. This is the extensional view of
-permutations. In the following sections, we will characterise these permutations syntactically- specifically, using a concept of \textit{a presentation} from the group theory.
-
-\todo{Move.}
-
-Although the meaning of a permutation - that is, the specific bijective function that it represents - is in some sense all there is, manipulating them syntactically still has its advantages.
-By writing reversible programs, we think of them in an intensional way.
-Comparing two programs for equality by evaluating them on all points in the domain is a very crude - and maybe even inefficient - way of doing that.
-
-It is crude, because there is no way of enforcing additional constraints on the process of transforming one program into another, and no way of inspecting what transformation occured.
-We can imagine a practical situation in which reversible circuts admit one kind of optimization \jk{FPGA?}, but do not admit another, even though they are equivalent in the extensional sense - or maybe one of these transformations is cheaper than another, or maybe it is important to know which transformation did occur for the \jk{producer(?)} to focus their resources of improving this kind of transformations.
-By comparing the circuits extensionally, the proof object = the path from one into the other - does not have any structure,  it is just a check that the values match.
-
-On the other hand, even though in general, the equality of circuits requires exponential time jk{triple check}, the proof object can be still very small.
-This creates structure that can possibly be exploited in a heurisic way.
-\jk{Additionally, if the goal is to convince third-party that two circuits are the same, we can just present the proof instead of showing the equality directly - for example, by choosing a particularly short one}.
-
-Taking all this into account, it is clear that we must seek a syntactic way of computing the meaning of the circuit, instead of just directly evaluating it to $\Aut[\Fin[n]]$ - since when we go back and quote the permutation as a circuit (in a normal form), there was no "trace" left to see what transformation (what sequence of 2-paths) maps the old one into the new one.
-
-Fortunately, group theory already has language in which such problems are expressed - we will be looking at solving the \textit{word problem}.
-Because $\PiLang$ is used for reasoning about reversible functions on finite sets, the group that we are interested in is the group of permutations on a finite set, which is classically known as the symmetric group,
-$\Sn$.
-Putting it in this form allows us to connect to the broader scope of computational group theory and reuse ideas from there.
-For example, during the proof, we will use \emph{Lehmer codes} (introduced in \cref{subsec:lehmer}) - a concept used widely in this context, which makes for a very convenient and compact method of representing permutations on a computer.
-
-The goal of this section is to reconcile two different approaches to defining the symmetric group. We already defined
-$\UFin$ in the previous section, and showed how its loopspace encodes the automorphism group. Now, we're going to define
-the same group in a different way, in terms of the symmetric group presented using generators and relations, and show
-how the two points of view are equivalent.
-
-\todo{Big example: Start from a listed permutation, show Lehmer code, then adjacent swaps.}
-\todo{Justification for why we need group theory.}
+The goal of this section is to reconcile two different approaches to defining the symmetric group -- as an automorphism
+group, and as a group presented using generators and relations.
 
@@
 
-From universal algebra, a group is simply a set with a 0-ary constant (the neutral element), a binary operation for group multiplication, and a unary inverse operation. The neutral element has
-to satisfy unit and inverse laws, and the multiplication has to be associative.
+From universal algebra, a group is simply a set with a 0-ary constant $e$ (the neutral element), a binary operation
+$\blank\mult\blank$ for group multiplication, and a unary inverse operation $\inv{\blank}$. The neutral element has to
+satisfy unit and inverse laws, and the multiplication has to be associative (see~\cref{def:group}).
 
-A very simple example of a group is $\mathbb{Z}$, where the neutral element is 0, the inverse of $k$ is $-k$, and the group multiplication is given by integer addition.
+A very simple example of a group is $\mathbb{Z}$, where the neutral element is 0, the inverse of $k$ is $-k$, and the
+group multiplication is given by integer addition.
 
-In type theory, a group $G$ can be defined as an $\hSet$ $S$ with the following pieces of data:
+\begin{toappendix}
+  \begin{definition}~\label{def:group}
+    In type theory, a group $G$ can be defined as a set $S$ with the following pieces of data:
 
@@
     \end{enumerate}
+  \end{definition}
+\end{toappendix}
 
@@
 identified with a 1-object groupoid, using a technique called delooping. The delooping of a group $G$ is a groupoid
-$\B{G}$ given by a unique object $\pt$ with self-loops that are 1-paths $\pt \id_{\B{G}} \pt$ corresponding to the
-elements of $G$. Note that the group operations are automatically given by operations on the identity type, with
-$\refl_{\pt}$ for the neutral element, path composition for the group multiplication, and path inverse for the group
-inverse. These satisfy the group laws as well, up to the identity type, using the groupoid coherence laws. Moreover, for
-1-groups which are supposed to be sets, these 2-paths should be propositions, so we have to restrict $\B{G}$ to be a
-1-groupoid. Hence, a group is simply given by a pointed, connected
-1-type~\cite*{buchholtzHigherGroupsHomotopy2018,symmetryBook2021}.
+$\B{G}$, with a unique object $\pt$, where the group elements are encoded as self-loops $\pt \id_{\B{G}} \pt$.
 
-For example, given a pointed type $(A:\UU, a:A)$, the automorphism group structure at $a$ is given by $a \id_{A} a$. Of
-course, for 1-groups we will require that $a \id_{A} a$ is an $\hSet$, which is enforced by having $A$ be a groupoid. In
-our running example for the permutation group on finite sets, we have that $\Fin[n]$ is an $\hSet$, and hence,
-$\UFin[n] \defeq \BAut[\Fin[n]]$ is a pointed, connected 1-type, whose loopspace $\loopspace[\BAut[\Fin[n]],F_{n}]$ is
-equivalent to $\Aut[\Fin[n]] \defeq (\Fin[n] \eqv \Fin[n]) \eqv (\Fin[n] \id_{\UU} \Fin[n])$, which has the
-corresponding automorphism group structure.
+The group operations are automatically given by operations on the identity type, with $\refl_{\pt}$ for the neutral
+element, path composition for the group multiplication, and path inverse for the group inverse. These satisfy the group
+laws, up to the identity type, using the groupoid coherence laws.
+
+Since groups are sets, these 2-paths are propositions, making $\B{G}$ a 1-groupoid. Hence, a group is simply given by a
+pointed, connected 1-type~\cite*{buchholtzHigherGroupsHomotopy2018,symmetryBook2021}. \footnote{This technique also
+  allows defining higher groups in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}
+
+% For example, given a pointed type $(A:\UU, a:A)$, the automorphism group structure at $a$ is given by $a \id_{A} a$. Of
+% course, for 1-groups we will require that $a \id_{A} a$ is a set, which is enforced by having $A$ be a groupoid. In our
+% running example for the permutation group on finite sets, we have that $\Fin[n]$ is a set, and hence, $\UFin[n] \defeq
+% \BAut[\Fin[n]]$ is a pointed, connected 1-type. Its loopspace $\loopspace[\BAut[\Fin[n]],F_{n}]$ is equivalent to
+% $\Aut[\Fin[n]] \defeq (\Fin[n] \eqv \Fin[n]) \eqv (\Fin[n] \id_{\UU} \Fin[n])$, which has the corresponding automorphism
+% group structure.
 
@@
 Usually, there are are many equations, besides the group axioms, that hold for the elements of a group. For example, in
-the group $\mathbb{Z}/3\mathbb{Z}$, where group multiplication is given by addition modulo 3, we have an equation
-$1 + 1 + 1 = 0$, which is not a consequence of group axioms, but is specific to the particular group
-$\mathbb{Z}/3\mathbb{Z}$. A free group has the property that no other equations except the ones directly implied by the
-group axioms, such as $g \mult \inv{g} = e$, hold. Thus, $\mathbb{Z}/3\mathbb{Z}$ is not a free group. To be able to
-give an equational presentation of a group, we start by constructing the free group, then show how to add new equations
-to it.
+the group $\Aut[\Bool]$, or $\mathbb{Z}_2$, we have an equation $1 + 1 = 0$, which is not a consequence of the group
+axioms, but is specific to this particular group. A free group has the property that no other equations hold except the
+ones directly implied by the group axioms. To be able to give an equational presentation of a group, we start by
+constructing the free group, then show how to add new equations to it.
 
-Given any set $A$, we can describe the free group $F(A)$ on $A$. We draw the elements from the alphabet $A$, close them
-under the group operations of multiplication and inverse, and identify them by the group axioms. For example, the
-singleton set generates the additive group of integers $\mathbb{Z}$ we introduced before. In this context, we will call
-$A$ the generating set of $F(A)$. Working in HoTT, as before, we can simply use higher inductive types to define the
-free group, by adding path constructors for the axioms we want.
+Given any set $A$, we can describe the free group $F(A)$ -- we call $A$ the generating set of $F(A)$. We draw the
+elements from $A$, close them under the group operations, and identify them by the group axioms. For example, the
+additive group of integers $\mathbb{Z}$ is the free group on the singleton. In HoTT, we can use a higher inductive type
+to define the free group, by adding path constructors for the axioms we want (see~\cref{def:free-group}).
 
-\todo{Better notation for HITs?}
+\begin{toappendix}
   \begin{definition}
-  Given an $\hSet$ $A$, the free group $F(A)$ on it is given by a higher
-  inductive type with the following point and path constructors. Notice the
-  similarity with the definition of a group structure in~\ref{subsec:groups},
-  but note that each operation here is a generator for the type $F(A)$\dots.
+    \label{def:free-group}
+    Given a set $A$, the free group $F(A)$ on it is given by a higher inductive type with the following point and path
+    constructors. Notice the similarity with the definition of a group structure (\cref{def:group}), but note that each
+    operation here is a generator for $F(A)$.
     \begin{itemize}
@@
   \end{definition}
+\end{toappendix}
 
-A group homomorphism between groups is a function between the underlying sets that preserves the group structure, which
-we write as $G_1 \to^G G_2$. Using the induction principle for $F(A)$, we can easily prove the universal property of
-free groups, stemming from the free-forgetful adjucntion between the category of groups and sets. Informally, the
-universal property states that if we know how a map $f$ acts on the generating set $A$, we know exactly how it acts on
-every element of the group $F(A)$.
+A group homomorphism between groups $G$ and $H$, which we write as $G \to^G H$, is a function between the underlying
+sets that preserves the group structure. The universal property of free groups, stemming from the free-forgetful
+adjunction between the category of groups and sets, follows from the induction principle for $F(A)$. Informally, it
+states that if we know how a map $f$ acts on the generating set $A$, we know exactly how it acts on every element of the
+group $F(A)$.
 
@@
 
-Although this definition is close to the intuitive universal-algebraic definition, it is not the most convenient one to
-work with. Besides being interested in the elements of the group, we also want to characterise the path space of the
-group. This HIT definition of $F(A)$ has lots of path constructors corresponding to each group axiom, making it
-difficult to describe its path space.
+Although the description of $F(A)$ discussed above is close to the intuitive universal-algebraic definition, it is not
+the most convenient one to work with. Besides the elements of the group, we also want to characterise its the path
+space. This definition of $F(A)$ has lots of path constructors corresponding to each group axiom, making it difficult to
+describe its path space.
 
@@
 the group's underlying set, we can use $\inl/\inr$ to mark the elements -- $\inl{a}$ means $a$ and $\inr{a}$ means
-$\inv{a}$. Hence, we can encode the free group using the free monoid, that is, lists of $A + A$. Additionally, we need
-to ensure that group laws hold, so words identified by the group equations are equal. Thus, elements of the free group
-will actually be equivalence classes of lists on $A + A$.
+$\inv{a}$. Then, we can encode the free group using the free monoid, that is, lists of $A + A$. Additionally, we need to
+ensure that the inverse laws hold, so we have to coalesce adjacent occurences of $a$ and $\inv{a}$.
 
-\begin{definition}~\label{def:presentation}
-  Let $A$ be an $\hSet$, and $\List[\blank]$ the free monoid. The free group
-  $F(A)$ on $A$ is the set-quotient of $\List[A + A]$ by the congruence closure
-  of the relation $a \cons \inv{a} \cons \nil \sim \nil$.
+\begin{definition}
+  \label{def:presentation}
+  Let $A$ be a set, and $\List[\blank]$ the free monoid. The free group $F(A)$ on $A$ is the set-quotient of $\List[A +
+  A]$ by the congruence closure of the relation $a \cons \inv{a} \cons \nil \sim \nil$ and $\inv{a} \cons a \cons \nil
+  \sim \nil$.
 \end{definition}
@@
   element, multiplication given by list append $\append$, and inverse given by flipping $\inl$ and $\inr$, followed by
-  reversing the list. Further, $F(A)$ with $\eta_A : A \to F(A) \defeq \inl(a) \cons \nil$ satisfies the universal
-  property of free groups, as stated in~\cref{prop:free-groups}.
+  reversing the list. Further, $F(A)$ with $\eta_A(a) \defeq \inl(a) \cons \nil$ satisfies the universal property of
+  free groups, as stated in~\cref{prop:free-groups}.
 \end{proposition}
@@
 
-\review{We have already seen many examples of groups, that is, sets carrying a group structure, but now we'll see how to
-  syntactically represent them -- which is the role of group presentations.}
-
 A presentation of a group builds it by starting from the free group $F(A)$ and introducing a collection of equations
 that have to be satisfied in the resulting group. For example, if we take $F(\unit) \defeq \mathbb{Z}$ and add an
-equation $1 + 1 + 1 = 0$, the resulting group would be $\mathbb{Z}/3\mathbb{Z}$. Note that not all groups have finite
-(or computable) presentations, and, a group can have any number of presentations.
+equation $1 + 1 = 0$, the resulting group would be $\mathbb{Z}_2 \eqv \Aut[\Bool]$. Note that not all groups have finite
+(or computable) presentations, and, a group can have any number of different presentations.
 
 \begin{definition}
-  Let $A$ be an $\hSet$ and $R : (A + A) \to (A + A) \to \UU$ a binary relation on $A + A$. The group
+  Let $A$ be a set and $R : \List[A + A] \to \List[A + A] \to \UU$ a binary relation on $\List[A + A]$. The group
   $F(\langle A ; R \rangle)$ presented by $A$ and $R$, is given by the set-quotient of the free group $F(A)$ by the
-  closure of $R$, or equivalently, as the coequaliser
-  \[\begin{tikzcd}
-      FR && FA && F(\langle A ; R \rangle)
-      \arrow[shift right=2, from=1-1, to=1-3]
-      \arrow[shift left=2, from=1-1, to=1-3]
-      \arrow[two heads, from=1-3, to=1-5]
-    \end{tikzcd}\]
+  closure of $R$.
 \end{definition}

### 2021-07-04 Pi+/paper/reversible1.tex
sec 2 and a couple of macros etc

--- 2021-07-02 Pi+/paper/reversible1.tex
@@
+\begin{figure}[t]
+  {\scalebox{\scalef}{$%
+        %%\noindent\begin{minipage}{.7\linewidth}
+        \begin{array}{rrcll}
+          \idc :     & A                     & \isoone & A                            & : \idc      \\
+          \\
+          \identlp : & \zerot + A            & \isoone & A                            & : \identrp  \\
+          \swapp :   & A + B                 & \isoone & B + A                        & : \swapp    \\
+          \assoclp : & A + (B + C)           & \isoone & (A + B) + C                  & : \assocrp  \\ [1.5ex]
+          \identlt : & \onet \times A        & \isoone & A                            & : \identrt  \\
+          \swapt :   & A \times B            & \isoone & B \times A                   & : \swapt    \\
+          \assoclt : & A \times (B \times C) & \isoone & (A \times B) \times C        & : \assocrt  \\ [1.5ex]
+          \absorbr : & ~ \zerot \times A     & \isoone & \zerot ~                     & : \factorzl \\
+          \dist :    & ~ (A + B) \times C    & \isoone & (A \times C) + (B \times C)~ & : \factor
+        \end{array}$}}
+
+    \medskip
+
+    {\scalebox{\scalef}{%
+      \Rule{}
+      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : B \isoone C}
+      {\jdg{}{}{c_1 \fatsemi c_2 : A \isoone C}}
+      {}
+
+      \Rule{}
+      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : C \isoone D}
+      {\jdg{}{}{c_1 \oplus c_2 : A + C \isoone B + D}}
+      {}
+
+      \Rule{}
+      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : C \isoone D}
+      {\jdg{}{}{c_1 \otimes c_2 : A \times C \isoone B \times D}}
+      {}
+    }}
+  \caption{$\Pi$-terms, combinators, and their types.}
+  \label{fig:pi-terms}
+\end{figure}
+
 \section{A Reversible Programming Language: Set-Theoretic Semantics}
+%% ~3 pages
 \label{sec:reversibleone}
@@
 \label{langeqeq}
+\label{sec:informal}
 
-The practice of programming languages is replete with \emph{ad hoc} instances of reversible computations: database
-transactions, mechanisms for data provenance, checkpoints, stack and exception traces, logs, backups, rollback
-recoveries, version control systems, reverse engineering, software transactional memories, continuations, backtracking
-search, and multiple-level undo features in commercial applications. In the early nineties,
-\citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class, treatment of reversibility. But intensive
-research in full-fledged reversible models of computations and reversible programming languages was only sparked by the
-discovery of deep connections between physics and
-computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275, Hey:1999:FCE:304763,fredkin1982conservative}, and by the
-potential for efficient quantum computation~\cite{springerlink:10.1007/BF02650179}.
-
-The early developments of reversible programming languages started
-with a conventional programming language, e.g., an extended
-$\lambda$-calculus, and either:
-\begin{enumerate}
-  \item extended the language with a history
-        mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
-  \item imposed constraints on the control flow constructs to make them
-        reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.
-\end{enumerate}
-More foundational approaches recognize that reversible programming languages require a fresh approach and should be
-designed from first principles without the detour via conventional irreversible
-languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047,
-  rc2011,James:2012:IE:2103656.2103667,Carette2016}.
+The circuit model of reversible computation discussed in the introduction is a useful abstraction close to the hardware
+platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly''
+programming abstraction (e.g., \verb|qasm|). A mathematical model based on permutations of finite sets provides a richer
+algebraic model which we review in this section.
 
@@
 \subsection{The $\Pi$ Family of Languages}
+\label{sec:langRev-examples}
+\label{examples}
 
-A natural candidate for a semantic foundation for reversible programming languages is the notion of type
-isomorphism. Indeed, the type isomorphisms among finite types are sound and complete for all permutations on finite
-types~\cite{Fiore:2004,fiore-remarks} and hence they are \emph{complete} for expressing reversible combinational
-circuits~\cite{fredkin1982conservative, James:2012:IE:2103656.2103667,Toffoli:1980} and the extension with recursive
-types and trace operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
-language~\cite{James:2012:IE:2103656.2103667,rc2011}.
+A specialized language for programming with permutations on finite sets can be built using ideas going back to Mac
+Lane~\cite{XXX} that the commutative semiring (also known as the commutative rig) of natural numbers exactly
+characterizes permutations on finite sets. In programming parlance, as shown in Fig.~\ref{fig:pi-terms}, each primitive
+algebraic identity of commutative rigs becomes a constant in the programming language and the algebraic closure
+operators become composition operators in the programming language. Once equipped with a notion of values and types, we
+get the following syntactic presentation of a programming language for
+permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:
 
-Focusing on finite types, the building blocks of type theory are: the empty type ($\zerot$), the unit type
-($\onet$) containing just one value $\Acon{tt}$, the sum type ($+$) containing values of the form $\inlv{v}$ and
-$\inrv{v}$, and the product ($\times$) type containing pairs of values $(v_1,v_2)$. For this fragment of types, the
-syntax of the language $\Pi$ consists of the following sorts:
+% The practice of programming languages is replete with \emph{ad hoc} instances of reversible computations: database
+% transactions, mechanisms for data provenance, checkpoints, stack and exception traces, logs, backups, rollback
+% recoveries, version control systems, reverse engineering, software transactional memories, continuations, backtracking
+% search, and multiple-level undo features in commercial applications. In the early nineties,
+% \citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class, treatment of reversibility. But intensive
+% research in full-fledged reversible models of computations and reversible programming languages was only sparked by the
+% discovery of deep connections between physics and
+% computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275, Hey:1999:FCE:304763,fredkin1982conservative}, and by the
+% potential for efficient quantum computation~\cite{springerlink:10.1007/BF02650179}.
+
+% The early developments of reversible programming languages started
+% with a conventional programming language, e.g., an extended
+% $\lambda$-calculus, and either:
+% \begin{enumerate}
+%   \item extended the language with a history
+%         mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or
+%   \item imposed constraints on the control flow constructs to make them
+%         reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.
+% \end{enumerate}
+% More foundational approaches recognize that reversible programming languages require a fresh approach and should be
+% designed from first principles without the detour via conventional irreversible
+% languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047,
+%   rc2011, James:2012:IE:2103656.2103667, Carette2016}.
+
+\medskip
 
@@
       \textit{Values}        & v,w,x,y & ::= & \Acon{tt} \alt \inlv{v} \alt \inrv{v} \alt (v,w) \\
-      \textit{Program types} &         &     & A \iso B                                         \\
-      \textit{Programs}      & c       & ::= & (\textrm{See Fig.~\ref{pi-terms}})
-    \end{array}$}}
-\begin{figure}[t]
-  {\scalebox{\scalef}{$%
-        %%\noindent\begin{minipage}{.7\linewidth}
-        \begin{array}{rrcll}
-          \idc :     & A                     & \iso & A                            & : \idc      \\
-          \\
-          \identlp : & \zerot + A            & \iso & A                            & : \identrp  \\
-          \swapp :   & A + B                 & \iso & B + A                        & : \swapp    \\
-          \assoclp : & A + (B + C)           & \iso & (A + B) + C                  & : \assocrp  \\ [1.5ex]
-          \identlt : & \onet \times A        & \iso & A                            & : \identrt  \\
-          \swapt :   & A \times B            & \iso & B \times A                   & : \swapt    \\
-          \assoclt : & A \times (B \times C) & \iso & (A \times B) \times C        & : \assocrt  \\ [1.5ex]
-          \absorbr : & ~ \zerot \times A     & \iso & \zerot ~                     & : \factorzl \\
-          \dist :    & ~ (A + B) \times C    & \iso & (A \times C) + (B \times C)~ & : \factor
+      \textit{Program types} &         &     & A \isoone B                                         \\
+      \textit{Programs}      & c       & ::= & (\textrm{See Fig.~\ref{fig:pi-terms}})
     \end{array}$}}
 
-  {\scalebox{\scalef}{%
-      \Rule{}
-      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : B \iso C}
-      {\jdg{}{}{c_1 \fatsemi c_2 : A \iso C}}
-      {}
+\medskip\noindent The language of types is built from the empty type ($\zerot$), the unit type
+($\onet$) containing just one value~$\Acon{tt}$, the sum type ($+$) containing values of the form $\inlv{v}$ and
+$\inrv{v}$, and the product ($\times$) type containing pairs of values $(v_1,v_2)$.
+%
+% A natural candidate for a semantic foundation for reversible programming languages is the notion of type
+% isomorphism. Indeed, the type isomorphisms among finite types are sound and complete for all permutations on finite
+% types~\cite{Fiore:2004,fiore-remarks} and hence they are \emph{complete} for expressing reversible combinational
+% circuits~\cite{fredkin1982conservative, James:2012:IE:2103656.2103667,Toffoli:1980} and the extension with recursive
+% types and trace operators~\cite{Hasegawa:1997:RCS:645893.671607} is a Turing-complete reversible
+% language~\cite{James:2012:IE:2103656.2103667,rc2011}.
 
-      \Rule{}
-      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : C \iso D}
-      {\jdg{}{}{c_1 \oplus c_2 : A + C \iso B + D}}
-      {}
+To see how this language is expresses reversible circuits, we present a few examples. First it is possible to directly
+mimic the \verb|qasm|-perspective by defining types that describe sequences of booleans. We use the type
+$\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true} and
+$\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is straightforward to define using
+the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values, thus the
+type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$) corresponds to a collection of
+wires that can transmit three bits.
+%
+% For example, we can express a 3-bit word reversal operation as follows:
+% $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
+% $\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
+% \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
+% \[\begin{array}{rlr}
+%                          & (v_1, (v_2, v_3)) \\
+%     \swapt               & ((v_2, v_3), v_1) \\
+%     \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
+%     \assocrt             & (v_3, (v_2, v_1)) \\
+%   \end{array}\]
+%subcode source isomorphisms.tex:979
+%
+To express the \verb|cx| and \verb|ccx| gates, we need to encode a notion of conditional expression. Such conditionals
+turn out to be expressible using the distributivity and factoring identities of rigs as shown below:
 
-      \Rule{}
-      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : C \iso D}
-      {\jdg{}{}{c_1 \otimes c_2 : A \times C \iso B \times D}}
-      {}
-    }}
-  \caption{$\Pi$-terms, combinators, and their types.}
-  \label{pi-terms}
-\end{figure}
+\medskip
 
-\noindent Intuitively, one may view each type $A$ as a collection of physical wires that can transmit $\sizet{A}$
-distinct values where $\sizet{A}$ is a natural number that indicates the size of a type, computed as:
-$\sizet{\zerot} = 0$; $\sizet{\onet} = 1$; $\sizet{A + B} = \sizet{A} + \sizet{B}$; and
-$\sizet{A \times B} = \sizet{A} * \sizet{B}$.  Thus the type $\mathbb{2} = \onet + \onet$ corresponds to a wire that can
-transmit one of two values, i.e., bits. The type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated
-$\mathbb{B}~3$) corresponds to a collection of wires that can transmit three bits. From that perspective, a type
-isomorphism between types $A$ and $B$ (such that $\sizet{A}=\sizet{B}=n$) models a \emph{reversible} combinational
-circuit that \emph{permutes} the $n$ different values. These type isomorphisms form a \emph{commutative semiring} and
-are collected in Fig.~\ref{pi-terms}. Each line in the top part of the figure introduces a pair of dual constants that
-witness the type isomorphism in the middle.  These are the \emph{base} (non-reducible) terms of $\Pi$. The isomorphisms
-are extended to form a congruence relation by adding the three constructors at the bottom of the figure that witness
-equivalence and compatible closure. Although austere, this combinator-based language has the advantage of being amenable
-to formal analysis for at least two reasons: (i) it is conceptually simple and small, and (ii) it has direct and evident
-connections to type theory and category theory and hence possesses a rich algebraic structure which we exploit in the
-remainder of the paper.
-
-%%%%%%%%%
-\subsection{Informal Semantics and Examples}
-\label{sec:langRev-examples}
-\label{examples}
-
-%% \note{Revisit to pick examples relevant to the rest of the paper; also unify notation}
-
-In order to understand the examples of Sec.~\ref{sec:informal}, we review how to represent collections of booleans and
-conditionals in $\Pi$. We use the type $\onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing
-\textsf{false} and $\inrv{\Acon{tt}}$ representing $\textsf{true}$. Boolean negation is straightforward to define using
-the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values.
-For example, we can express a 3-bit word reversal operation as follows:
-
-$\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
-
-$\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
-
-\noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
-\[\begin{array}{rlr}
-                         & (v_1, (v_2, v_3)) \\
-    \swapt               & ((v_2, v_3), v_1) \\
-    \swapt \otimes  \idc & ((v_3, v_2), v_1) \\
-    \assocrt             & (v_3, (v_2, v_1)) \\
-  \end{array}\]
-%subcode source isomorphisms.tex:979
-
-Even though \ensuremath{\Pi } lacks conditional expressions, they are expressible using the distributivity and factoring
-laws as shown below:
-
-$\Afun{ifc}_{c_1,c_2} : \mathbb{2} \times A \iso \mathbb{2} \times B$
-
-$\Afun{ifc}_{c_1,c_2} = \dist \fatsemi ((\idc  \otimes\  c_1) \oplus (\idc \otimes\  c_2)) \fatsemi \factor$
-
-% \inkscape{if-c1-c2.pdf}
+\cif{}
 
 \noindent The input value of type $\mathbb{2} \times A$ is processed by the distribute operator \ensuremath{\dist},
-which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the \ensuremath{\mathit{left}} branch,
-which corresponds to the case when the boolean is \ensuremath{\mathit{false}}, the combinator~\ensuremath{c_1} is
-applied to the value of type~\ensuremath{A}. The right branch which corresponds to the boolean being
-\ensuremath{\mathit{true}} passes the value of type $A$ through the combinator \ensuremath{c_2}.  The inverse of
-\ensuremath{\dist}, namely \ensuremath{\factor} is applied to get the final result. The full expressiveness of $\Pi$ is
-demonstrated by an implementation of $\Afun{cnot}$ and $\Afun{toffoli}$ which are universal for reversible
-computation~\cite{Toffoli:1980}. The $\Afun{cnot}$ gates takes two input bits: a control bit and a target bit, and
-negates the target bit if the control bit is set. It is essentially a one-branch version of $\Afun{cif}$ above. The
-Toffoli gate generalizes the idea to a gate that takes 3 input bits: 2 control bits and one target bit: boolean inputs:
-it negates the target bit if both control bits are set.
+which converts it into a value of type $(\onet \times A) + (\onet \times A)$. In the left branch, which corresponds to
+the case when the boolean is \textsf{true}, the combinator~\ensuremath{c_1} is applied to the value of
+type~\ensuremath{A}. The right branch which corresponds to the boolean being \textsf{false} passes the value of type $A$
+through the combinator \ensuremath{c_2}.  The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied to
+get the final result. Using this conditional operator, \verb|cx| is defined as $\Afun{cif}~\verb|x|~\idc$ and
+\verb|ccx| is defined as $\Afun{cif}~\verb|cx|~\idc$. With these conventions, the first circuit in the introduction
+is transcribed as follows:
+
+\medskip
+
+\adder{}
+
+\noindent where we clearly see the sequences of the three operations \verb|ccx|, \verb|cx|, and \verb|cx| but, instead
+of using the indices in the sequence of wires to identify the relevant parameters, here we use structural isomorphisms
+(elided above) to re-shuffle the types. For the second circuit, instead of transcribing it directly, we express it using
+a slightly more abstract notation:
+
+\medskip
+
+\resettwo{}
+
+\noindent Like the original circuit, we examine the bit at index 1 (corresponding to the component $B$ in a tuple
+$(A,(B,C))$): if the bit is true, we perform an \verb|x| operation on component $A$, and otherwise we perform a
+\verb|cx| operation on $C,A$. The two uses of \verb|x| gates in the circuit are now unnecessary as they were only needed
+to encode a two-way conditional expression using a sequence of one-way conditional expressions (the only ones available in
+the linear circuit model).
+
+All of this is only half the story as the correspondence between the syntactic commutative rig (the syntax of $\PiLang$)
+and permutations on finite sets includes \emph{coherence conditions} that identify different syntactic representations
+of the same permutation~\cite{laplaza72,Carette2016}. These coherence conditions will be reflected in a second level of $\PiLang$
+syntax as level-2 combinators whose types are of the form $c_1 \isotwo c_2$ for appropriate $c_1$ and $c_2$ of the same
+level-1 type $A \isoone B$. For example, we have:
+
+\medskip
+
+\combtwo{}
+
+\noindent where the first level-2 combinator states the obvious fact that compositions with the identity can be
+optimized and the second level-2 combinator states the more interesting observation that swapping the two sides of a sum
+type and then applying $c_1$ to the left and $c_2$ to the right is equivalent to applying $c_2$ to the left and $c_1$ to
+the right and then swapping. The full set of level-2 combinators is large and is only listed in the accompanying code.
+
+% and that impose various coherence conditions on the level-1 combinators. In the remainder of this section, we show how
+% to use a subset of these level-2 combinators to show the equivalence of \Afun{reversibleOr1} and \Afun{reversibleOr2}.
+
+% \medskip
+
+% \orequiv{}
+
+% \note{This section should explain the main technical parts of the paper
+%   informally, without using any technology. Use an example, such as, a
+%   reversible language with $\leq 5$ bits, and examples of permutations and
+%   transpositions, and when they're equal.}
 
@@
 
-\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be
-specified by mapping each type to a finite set and each combinator to a bijection between finite sets, that is, a
-permutation. In this section, we outline a simple semantics expressed in the metalanguage of sets and functions. This
-semantics will be extended to the metalanguage of homotopy type theory in the remainder of the paper.
+\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be specified
+by mapping each type to a finite set and each combinator to a permutation between finite sets. In this section, we
+outline a simple denotational semantics expressed in the metalanguage of sets and functions that collapses the level-2
+combinators. This semantics will be extended in the next two sections to the metalanguage of homotopy type theory to
+properly model the level-2 combinators.
 
@@
 
-where $\emptyset$ is the empty set, $\uplus$ is the disjoint union of sets, and $\times$ is the cartesian product of
-sets. The denotation of a combinator $c : A \isot B$ is a (bijective) function mapping $\denot{A}$ to $\denot{B}$:
+\noindent where $\emptyset$ is the empty set, $\uplus$ is the disjoint union of sets, and $\times$ is the cartesian product of
+sets. The denotation of a combinator $c : A \isoone B$ is a (bijective) function mapping $\denot{A}$ to $\denot{B}$:
 
-\begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
-  \denot{\identlp}           & (\inlv{v})           & = & v                                   \\
-  \denot{\identrp}           & v                    & = & \inlv{v}                            \\
-  \denot{\swapp}             & (\inlv{v})           & = & \inrv{v}                            \\
-  \denot{\swapp}             & (\inrv{v})           & = & \inlv{v}                            \\
-  \denot{\assoclp}           & (\inlv{v})           & = & \inlv{(\inlv{v})}                   \\
-  \denot{\assoclp}           & (\inrv{(\inlv{v})})  & = & \inlv{(\inrv{v})}                   \\
-  \denot{\assoclp}           & (\inrv{(\inrv{v})})  & = & \inrv{v}                            \\
-  \denot{\assocrp}           & (\inlv{(\inlv{v})})  & = & \inlv{v}                            \\
-  \denot{\assocrp}           & (\inlv{(\inrv{v})})  & = & \inrv{(\inlv{v})}                   \\
-  \denot{\assocrp}           & (\inrv{v})           & = & \inrv{(\inrv{v})}                   \\
-  \denot{\identlt}           & (\Acon{tt} , v)      & = & v                                   \\
-  \denot{\identrt}           & v                    & = & (\Acon{tt} , v)                     \\
-  \denot{\swapt}             & (v_1 , v_2)          & = & (v_2 , v_1)                         \\
-  \denot{\assoclt}           & (v_1 , (v_2 , v_3))  & = & ((v_1 , v_2) , v_3)                 \\
-  \denot{\assocrt}           & ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
-  \denot{\dist}              & (\inlv{v_1} , v_3)   & = & \inlv{(v_1 , v_3)}                  \\
-  \denot{\dist}              & (\inrv{v_2 , v_3})   & = & \inrv{(v_2 , v_3)}                  \\
-  \denot{\factor}            & (\inlv{(v_1 , v_3)}) & = & (\inlv{v_1} , v_3)                  \\
-  \denot{\factor}            & (\inrv{(v_2 , v_3)}) & = & (\inrv{v_2} , v_3)                  \\
-  \denot{\idc}               & v                    & = & v                                   \\
-  \denot{(c_1 \fatsemi c_2)} & v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
-  \denot{(c_1 \oplus c_2)}   & (\inlv{v})           & = & \inlv{(\denot{c_1}~v)}              \\
-  \denot{(c_1 \oplus c_2)}   & (\inrv{v})           & = & \inrv{(\denot{c_2}~v)}              \\
-  \denot{(c_1 \otimes c_2)}  & (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
-\end{longtable}
+\begin{multicols}{2}
+%%  \begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
+\[\begin{array}[t]{rcl}
+  \denot{\identlp}     ~      (\inlv{v})           & = & v                                   \\
+  \denot{\identrp}      ~      v                    & = & \inlv{v}                            \\
+  \denot{\swapp}        ~      (\inlv{v})           & = & \inrv{v}                            \\
+  \denot{\swapp}        ~      (\inrv{v})           & = & \inlv{v}                            \\
+  \denot{\assoclp}      ~      (\inlv{v})           & = & \inlv{(\inlv{v})}                   \\
+  \denot{\assoclp}      ~      (\inrv{(\inlv{v})})  & = & \inlv{(\inrv{v})}                   \\
+  \denot{\assoclp}      ~      (\inrv{(\inrv{v})})  & = & \inrv{v}                            \\
+  \denot{\assocrp}      ~      (\inlv{(\inlv{v})})  & = & \inlv{v}                            \\
+  \denot{\assocrp}      ~      (\inlv{(\inrv{v})})  & = & \inrv{(\inlv{v})}                   \\
+  \denot{\assocrp}      ~      (\inrv{v})           & = & \inrv{(\inrv{v})}
+\end{array}\]
+
+\[\begin{array}[t]{rcl}
+  \denot{\identlt}      ~      (\Acon{tt} , v)      & = & v                                   \\
+  \denot{\identrt}      ~      v                    & = & (\Acon{tt} , v)                     \\
+  \denot{\swapt}        ~      (v_1 , v_2)          & = & (v_2 , v_1)                         \\
+  \denot{\assoclt}      ~      (v_1 , (v_2 , v_3))  & = & ((v_1 , v_2) , v_3)                 \\
+  \denot{\assocrt}      ~      ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
+  \denot{\dist}         ~      (\inlv{v_1} , v_3)   & = & \inlv{(v_1 , v_3)}                  \\
+  \denot{\dist}         ~      (\inrv{v_2 , v_3})   & = & \inrv{(v_2 , v_3)}                  \\
+  \denot{\factor}       ~      (\inlv{(v_1 , v_3)}) & = & (\inlv{v_1} , v_3)                  \\
+  \denot{\factor}       ~      (\inrv{(v_2 , v_3)}) & = & (\inrv{v_2} , v_3)
+\end{array}\]
+
+%%\end{longtable}
+\end{multicols}
+
+  \[\begin{array}{rcl}
+  \denot{\idc}          ~      v                    & = & v                                   \\
+  \denot{(c_1 \fatsemi c_2)} ~ v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
+  \denot{(c_1 \oplus c_2)}   ~ (\inlv{v})           & = & \inlv{(\denot{c_1}~v)}              \\
+  \denot{(c_1 \oplus c_2)}   ~ (\inrv{v})           & = & \inrv{(\denot{c_2}~v)}              \\
+  \denot{(c_1 \otimes c_2)}  ~ (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
+  \end{array}\]
 
@@
 
-The denotational semantics can be used to evaluate programs and to check equivalence of programs. For example, it is
-straightforward to verify that $\denot{(c_1~\oplus~c_2)~\fatsemi~\swapp} = \denot{\swapp \fatsemi (c_2~\oplus~c_1)}$. A
-slightly more involved example are the following two programs:
+% The denotational semantics can be used to evaluate programs and to check equivalence of programs. For example, it is
+% straightforward to verify that $\denot{(c_1~\oplus~c_2)~\fatsemi~\swapp} = \denot{\swapp \fatsemi (c_2~\oplus~c_1)}$. A
+% slightly more involved example are the following two programs:
 
-\rotate{}
+% \rotate{}
 
-\noindent The first program performs the following sequence of transformations:
-\[
-  \Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
-  \Tree [ [ {\small a} {\small b} ] {\small c} ] ~\to~
-  \Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
-  \Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
-\]
-\noindent
-while the second evaluates as follows:
-\[
-  \Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
-  \Tree [ {\small a} [ {\small c} {\small b} ] ] ~\to~
-  \Tree [ [ {\small a} {\small c} ] {\small b} ] ~\to~
-  \Tree [ [ {\small c} {\small a} ] {\small b} ] ~\to~
-  \Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
-  \Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
-\]
+% \noindent The first program performs the following sequence of transformations:
+% \[
+%   \Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+%   \Tree [ [ {\small a} {\small b} ] {\small c} ] ~\to~
+%   \Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+%   \Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+% \]
+% \noindent
+% while the second evaluates as follows:
+% \[
+%   \Tree [ {\small a} [ {\small b} {\small c} ] ] ~\to~
+%   \Tree [ {\small a} [ {\small c} {\small b} ] ] ~\to~
+%   \Tree [ [ {\small a} {\small c} ] {\small b} ] ~\to~
+%   \Tree [ [ {\small c} {\small a} ] {\small b} ] ~\to~
+%   \Tree [ {\small c} [ {\small a} {\small b} ] ] ~\to~
+%   \Tree [ {\small c} [ {\small b} {\small a} ] ] ~.
+% \]
 
-The semantics above can be packaged in the category of finite sets and functions, $\SetFin$, which is the category
-freely generated by finite coproduct completion of the terminal category. Objects of $\SetFin$ can be identified with
-sets of fixed cardinality, that is, $\Fin[n] \defeq \Set{0,1,\ldots,n-1}$. $\SetFin$ has finite coproducts and products,
-which lets us interpret the types of $\PiLang$. Combinators are intepreted as morphisms in $\SetFin$, but we have to
-restrict to invertible morphisms, that is, isomorphisms. This gives the \emph{groupoid} of finite sets and bijections,
-$\BFin \defeq \mathsf{core}(\SetFin)$. The isomorphisms satisfied by coproducts and products in $\SetFin$ lift to
-$\BFin$, but they're no longer categorical coproducts and products. They give two symmetric monoidal tensor products on
-$\BFin$, the additive and multiplicative ones, with the multiplicative tensor distributing over the additive tensor.
+% The semantics above can be packaged in the category of finite sets and functions, $\SetFin$, which is the category
+% freely generated by finite coproduct completion of the terminal category. Objects of $\SetFin$ can be identified with
+% sets of fixed cardinality, that is, $\Fin[n] \defeq \Set{0,1,\ldots,n-1}$. $\SetFin$ has finite coproducts and products,
+% which lets us interpret the types of $\PiLang$. Combinators are interpreted as morphisms in $\SetFin$, but we have to
+% restrict to invertible morphisms, that is, isomorphisms. This gives the \emph{groupoid} of finite sets and bijections,
+% $\BFin \defeq \mathsf{core}(\SetFin)$. The isomorphisms satisfied by coproducts and products in $\SetFin$ lift to
+% $\BFin$, but they're no longer categorical coproducts and products. They give two symmetric monoidal tensor products on
+% $\BFin$, the additive and multiplicative ones, with the multiplicative tensor distributing over the additive tensor.
 
-\note{change thm:
-  1. Each function is a bijection. 2. If there is a 2-combinator, the denotations of the 1-combinators are equal
-}
+% \note{change thm:
+%   1. Each function is a bijection. 2. If there is a 2-combinator, the denotations of the 1-combinators are equal
+% }
 
-\begin{theorem}\label{thm:grp}
-  Let $\sim$ be the following equivalence relation on combinators $c_1 \sim c_2$ iff $\denot{c_1} = \denot{c_2}$
-  identifying combinators with the same denotation.  The notation $[c]_{\sim}$ refers to a representative combinator
-  from a $\sim$-equivalence class. Using this relation, we define the
-  category $\mathcal{C}$ as follows:
+\begin{theorem}\label{thm:semone}
+  The denotational semantics is sound in the following sense:
   \begin{itemize}
-    \item $\mathit{Obj}(\mathcal{C})$ is the set of $\Pi$-types, and
-    \item $\mathit{Hom}(A,B) = [c : A \isot B]_{\sim}$
+    \item For every level-1 combinator $c : A \isoone B$, we have that $\denot{c}$ is a bijection between $\denot{A}$ and $\denot{B}$.
+      \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
+        combinator $\alpha$ such that $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using the
+        extensional equivalence of functions.
   \end{itemize}
-  The category $\mathcal{C}$ is a groupoid.
 \end{theorem}
-\begin{proof}
-  The relation $\sim$ identifies $c$, $\idc \circ c$, and $c \circ \idc$, and is associative. Furthermore, the denotation
-  of each combinator is a reversible function thus making every morphism into an isomorphism.
-\end{proof}
+%% proof?
 
-\note{Then say that $\mathcal{C}$ is ``the same'' as $\BFin$ We can say here that the relation $\sim$ also identifies
-  $(c_1~\oplus~c_2)~\fatsemi~\swapp$ and $\swapp \fatsemi (c_2~\oplus~c_1)$ and other properties. ?? }
+%   Let $\sim$ be the following equivalence relation on combinators $c_1 \sim c_2$ iff $\denot{c_1} = \denot{c_2}$
+%   identifying combinators with the same denotation.  The notation $[c]_{\sim}$ refers to a representative combinator
+%   from a $\sim$-equivalence class. Using this relation, we define the
+%   category $\mathcal{C}$ as follows:
+%   \begin{itemize}
+%     \item $\mathit{Obj}(\mathcal{C})$ is the set of $\Pi$-types, and
+%     \item $\mathit{Hom}(A,B) = [c : A \isot B]_{\sim}$
+%   \end{itemize}
+%   The category $\mathcal{C}$ is a groupoid.
+% \end{theorem}
+% \begin{proof}
+%   The relation $\sim$ identifies $c$, $\idc \circ c$, and $c \circ \idc$, and is associative. Furthermore, the denotation
+%   of each combinator is a reversible function thus making every morphism into an isomorphism.
+% \end{proof}
+
+% \note{Then say that $\mathcal{C}$ is ``the same'' as $\BFin$ We can say here that the relation $\sim$ also identifies
+%   $(c_1~\oplus~c_2)~\fatsemi~\swapp$ and $\swapp \fatsemi (c_2~\oplus~c_1)$ and other properties. ?? }
 

### 2021-07-04 Pi+/paper/ufin.tex
Revising section 4
sec 2 and a couple of macros etc

--- 2021-07-01 Pi+/paper/ufin.tex
@@
 
-\noindent The denotation $\sem{c}$ of a combinator $c : A \isot B$ is a path between $\sem{A}$ and $\sem{B}$. If the
+\noindent The denotation $\sem{c}$ of a combinator $c : A \isoone B$ is a path between $\sem{A}$ and $\sem{B}$. If the
 size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
@@
 
-\noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \Iso q$, the denotation
+\noindent Finally, given two combinators $p , q : A \isoone B$ and a 2-combinator $\alpha : p \isotwo q$, the denotation
 $\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.

### 2021-07-04 Pi+/paper/ufin.tex
did something go wrong when merging ufin?

--- 2021-07-04 Pi+/paper/ufin.tex
@@
 
-  \noindent The denotation $\sem{c}$ of a combinator $c : A \isot B$ is a path between $\sem{A}$ and $\sem{B}$. If the
+  \noindent The denotation $\sem{c}$ of a combinator $c : A \isoone B$ is a path between $\sem{A}$ and $\sem{B}$. If the
   size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
@@
 
-  \noindent Finally, given two combinators $p , q : A \isot_1 B$ and a 2-combinator $\alpha : p \Iso q$, the denotation
+  \noindent Finally, given two combinators $p , q : A \isoone B$ and a 2-combinator $\alpha : p \isotwo q$, the denotation
   $\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.
@@
 
+\todo{Cite Brent Yorgey's thesis?}
+
 \begin{definition}[$\Fin$]

### 2021-07-04 Pi+/paper/reversible1.tex
scale math
minor fixes in sec 2

@@
 ($\onet$) containing just one value~$\Acon{tt}$, the sum type ($+$) containing values of the form $\inlv{v}$ and
-$\inrv{v}$, and the product ($\times$) type containing pairs of values $(v_1,v_2)$.
+$\inrv{v}$, and the product type ($\times$) containing pairs of values $(v_1,v_2)$.
 %
@@
 
-To see how this language is expresses reversible circuits, we present a few examples. First it is possible to directly
+To see how this language expresses reversible circuits, we present a few examples. First it is possible to directly
 mimic the \verb|qasm|-perspective by defining types that describe sequences of booleans. We use the type
@@
 
-All of this is only half the story as the correspondence between the syntactic commutative rig (the syntax of $\PiLang$)
+All of this is only half the story however as the correspondence between the syntactic commutative rig (the syntax of $\PiLang$)
 and permutations on finite sets includes \emph{coherence conditions} that identify different syntactic representations
-of the same permutation~\cite{laplaza72,Carette2016}. These coherence conditions will be reflected in a second level of $\PiLang$
+of the same permutation~\cite{laplaza72,Carette2016}. These coherence conditions are collected in a second level of $\PiLang$
 syntax as level-2 combinators whose types are of the form $c_1 \isotwo c_2$ for appropriate $c_1$ and $c_2$ of the same
@@
 
-\begin{align*}
-  \denot{\zerot}     & = \emptyset                  \\
-  \denot{\onet}      & = \{ \Acon{tt} \}            \\
-  \denot{A + B}      & = \denot{A} \uplus \denot{B} \\
-  \denot{A \times B} & = \denot{A} \times \denot{B}
-\end{align*}
+\begin{center}
+{\scalebox{\scalef}{$%
+\begin{array}{rcl}
+  \denot{\zerot}     & = & \emptyset                  \\
+  \denot{\onet}      & = & \{ \Acon{tt} \}            \\
+  \denot{A + B}      & = & \denot{A} \uplus \denot{B} \\
+  \denot{A \times B} & = & \denot{A} \times \denot{B}
+\end{array}$}}
+\end{center}
 
@@
 %%  \begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
-\[\begin{array}[t]{rcl}
-  \denot{\identlp}     ~      (\inlv{v})           & = & v                                   \\
-  \denot{\identrp}      ~      v                    & = & \inlv{v}                            \\
-  \denot{\swapp}        ~      (\inlv{v})           & = & \inrv{v}                            \\
-  \denot{\swapp}        ~      (\inrv{v})           & = & \inlv{v}                            \\
-  \denot{\assoclp}      ~      (\inlv{v})           & = & \inlv{(\inlv{v})}                   \\
-  \denot{\assoclp}      ~      (\inrv{(\inlv{v})})  & = & \inlv{(\inrv{v})}                   \\
-  \denot{\assoclp}      ~      (\inrv{(\inrv{v})})  & = & \inrv{v}                            \\
-  \denot{\assocrp}      ~      (\inlv{(\inlv{v})})  & = & \inlv{v}                            \\
-  \denot{\assocrp}      ~      (\inlv{(\inrv{v})})  & = & \inrv{(\inlv{v})}                   \\
-  \denot{\assocrp}      ~      (\inrv{v})           & = & \inrv{(\inrv{v})}
-\end{array}\]
+{\scalebox{\scalef}{$%
+\begin{array}[t]{rlcl}
+  \denot{\identlp}     &      (\inlv{v})           & = & v                                   \\
+  \denot{\identrp}     &      v                    & = & \inlv{v}                            \\
+  \denot{\swapp}       &     (\inlv{v})           & = & \inrv{v}                            \\
+  \denot{\swapp}       &      (\inrv{v})           & = & \inlv{v}                            \\
+  \denot{\assoclp}     &      (\inlv{v})           & = & \inlv{(\inlv{v})}                   \\
+  \denot{\assoclp}     &      (\inrv{(\inlv{v})})  & = & \inlv{(\inrv{v})}                   \\
+  \denot{\assoclp}     &      (\inrv{(\inrv{v})})  & = & \inrv{v}                            \\
+  \denot{\assocrp}     &      (\inlv{(\inlv{v})})  & = & \inlv{v}                            \\
+  \denot{\assocrp}     &      (\inlv{(\inrv{v})})  & = & \inrv{(\inlv{v})}                   \\
+  \denot{\assocrp}     &      (\inrv{v})           & = & \inrv{(\inrv{v})}
+\end{array}$}}
 
-\[\begin{array}[t]{rcl}
-  \denot{\identlt}      ~      (\Acon{tt} , v)      & = & v                                   \\
-  \denot{\identrt}      ~      v                    & = & (\Acon{tt} , v)                     \\
-  \denot{\swapt}        ~      (v_1 , v_2)          & = & (v_2 , v_1)                         \\
-  \denot{\assoclt}      ~      (v_1 , (v_2 , v_3))  & = & ((v_1 , v_2) , v_3)                 \\
-  \denot{\assocrt}      ~      ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
-  \denot{\dist}         ~      (\inlv{v_1} , v_3)   & = & \inlv{(v_1 , v_3)}                  \\
-  \denot{\dist}         ~      (\inrv{v_2 , v_3})   & = & \inrv{(v_2 , v_3)}                  \\
-  \denot{\factor}       ~      (\inlv{(v_1 , v_3)}) & = & (\inlv{v_1} , v_3)                  \\
-  \denot{\factor}       ~      (\inrv{(v_2 , v_3)}) & = & (\inrv{v_2} , v_3)
-\end{array}\]
+{\scalebox{\scalef}{$%
+\begin{array}[t]{rlcl}
+  \denot{\identlt}      &      (\Acon{tt} , v)      & = & v                                   \\
+  \denot{\identrt}      &      v                    & = & (\Acon{tt} , v)                     \\
+  \denot{\swapt}        &      (v_1 , v_2)          & = & (v_2 , v_1)                         \\
+  \denot{\assoclt}     &      (v_1 , (v_2 , v_3))  & = & ((v_1 , v_2) , v_3)                 \\
+  \denot{\assocrt}     &      ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
+  \denot{\dist}       &      (\inlv{v_1} , v_3)   & = & \inlv{(v_1 , v_3)}                  \\
+  \denot{\dist}        &      (\inrv{v_2 , v_3})   & = & \inrv{(v_2 , v_3)}                  \\
+  \denot{\factor}       &      (\inlv{(v_1 , v_3)}) & = & (\inlv{v_1} , v_3)                  \\
+  \denot{\factor}      &      (\inrv{(v_2 , v_3)}) & = & (\inrv{v_2} , v_3)
+\end{array}$}}
 
@@
 
-  \[\begin{array}{rcl}
-  \denot{\idc}          ~      v                    & = & v                                   \\
-  \denot{(c_1 \fatsemi c_2)} ~ v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
-  \denot{(c_1 \oplus c_2)}   ~ (\inlv{v})           & = & \inlv{(\denot{c_1}~v)}              \\
-  \denot{(c_1 \oplus c_2)}   ~ (\inrv{v})           & = & \inrv{(\denot{c_2}~v)}              \\
-  \denot{(c_1 \otimes c_2)}  ~ (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
-  \end{array}\]
+\begin{center}
+{\scalebox{\scalef}{$%
+\begin{array}{rlcl}
+  \denot{\idc}         &      v                    & = & v                                   \\
+  \denot{(c_1 \fatsemi c_2)} & v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
+  \denot{(c_1 \oplus c_2)}   & (\inlv{v})           & = & \inlv{(\denot{c_1}~v)}              \\
+  \denot{(c_1 \oplus c_2)}   & (\inrv{v})           & = & \inrv{(\denot{c_2}~v)}              \\
+  \denot{(c_1 \otimes c_2)}  & (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
+  \end{array}
+$}}
+\end{center}
 
@@
       \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
-        combinator $\alpha$ such that $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using the
+        combinator $\alpha$ such that $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using
         extensional equivalence of functions.

### 2021-07-04 Pi+/paper/ufin.tex
Text about homotopies and equivalences

@@
 The groupoid $\BFin$ that we presented is strict, that is, two equal permutations are extensionally identified, without
-an explicit witness for the equality. Our first step is to weaken this groupoid $\BFin$ into a weak 2-category. In a
-weak category, equational axioms of the category are replaced by higher invertible cells. For example, there are higher
-invertible cells for the left and right units of composition, and associativity. The higher cells come with their own
-coherence data as well, such as vertical composition, and horizontal composition, or whiskering.\todo{define 2-category
-  in appendix.}
+an explicit witness for the equality. Our first step is to weaken this groupoid $\BFin$. In a weak category, equational
+axioms of the category are replaced by higher invertible cells. For example, there are higher invertible cells for the
+left and right units of composition, and associativity. The higher cells come with their own coherence data as well,
+such as vertical composition, and horizontal composition, or whiskering.\todo{define 2-category in appendix.}
 
@@
   \begin{aligned}
-    \term{sym}   & : (x \id_{A} y) \to (y \id_{A} x)                   \\
-    \term{\comp} & : (x \id_{A} y) \to (y \id_{A} z) \to (x \id_{A} z)
+    \term{\inv{\blank}}      & : (x \id_{A} y) \to (y \id_{A} x)                   \\
+    \term{\blank\comp\blank} & : (x \id_{A} y) \to (y \id_{A} z) \to (x \id_{A} z)
   \end{aligned}
@@
                  & \to (p \comp q) \comp r \id p \comp (q \comp r)          \\
-    \term{invr}  & : (p : x \id_{A} y) \to p \comp \term{sym}(p) \id \refl_{x}
+    \term{invr}  & : (p : x \id_{A} y) \to p \comp \inv{p} \id \refl_{x}
   \end{aligned}
 \end{gather*}
+
+A homotopy between functions $f \htpy g$ is given by pointwise equality between them $\dfun{x:A}{f(x) \id_{B} g(x)}$.
+The identity type for functions is equivalent to homotopies between them ${(f \id_{A \to B} g)} \eqv {(f \htpy g)}$, by
+function extensionality. An equivalent between types $A \eqv B$ is given by a pair of functions between them which
+compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy \idfunc_{A}$. This is equivalent to the
+identity type for the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.
 
@@
 
-Further, using the groupoid structure of $A$, we can show that $\term{transport}$ lifts paths to equivalences of fibers.
+Further, using the groupoid structure of $A$, we can show that $\transport{P}{p}{\blank}$ and
+$\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name the map $\tptEqv{P}$ which lifts paths to equivalences
+of fibers.
 
@@
 \todo{Diagram.}
+The type families (or fibrations) we're interested in are the ones where paths in the base space completely determine
+the equivalences in the fibers -- these are called univalent fibrations.
 
@@
 
-\subsubsection{Univalence}
-
-Voevodsky's \emph{univalence} principle characterises paths in the universe. It says that equivalent types are equal, or,
-the following function is an equivalence.
-
-\[
-  \ua : A \id_{\UU} B \to A \eqv B
-\]
-
-\todo{Revise.}
-
-Alternatively, one can say that the identity type family $\term{id} : \UU \to \UU$ is univalent.
+The use of the word \emph{univalent} here is a reference to Voevodsky's \emph{univalence} principle. Indeed, univalence
+characterises paths in the universe as equivalences between types, which follows from the canonical fibration $\idfunc :
+\UU \to \UU$ being univalent.
 

### 2021-07-04 Pi+/paper/finite.tex
Cut more in section 3

@@
 
-However, more conveniently, in HoTT, we can instead use groupoids to talk about groups internally. A group can be
-identified with a 1-object groupoid, using a technique called delooping. The delooping of a group $G$ is a groupoid
-$\B{G}$, with a unique object $\pt$, where the group elements are encoded as self-loops $\pt \id_{\B{G}} \pt$.
+% However, more conveniently, in HoTT, we can instead use groupoids to talk about groups internally. A group can be
+% identified with a 1-object groupoid, using a technique called delooping. The delooping of a group $G$ is a groupoid
+% $\B{G}$, with a unique object $\pt$, where the group elements are encoded as self-loops $\pt \id_{\B{G}} \pt$.
 
-The group operations are automatically given by operations on the identity type, with $\refl_{\pt}$ for the neutral
-element, path composition for the group multiplication, and path inverse for the group inverse. These satisfy the group
-laws, up to the identity type, using the groupoid coherence laws.
+% The group operations are automatically given by operations on the identity type, with $\refl_{\pt}$ for the neutral
+% element, path composition for the group multiplication, and path inverse for the group inverse. These satisfy the group
+% laws, up to the identity type, using the groupoid coherence laws.
 
-Since groups are sets, these 2-paths are propositions, making $\B{G}$ a 1-groupoid. Hence, a group is simply given by a
-pointed, connected 1-type~\cite*{buchholtzHigherGroupsHomotopy2018,symmetryBook2021}. \footnote{This technique also
-  allows defining higher groups in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}
+% Since groups are sets, these 2-paths are propositions, making $\B{G}$ a 1-groupoid. Hence, a group is simply given by a
+% pointed, connected 1-type~\cite*{buchholtzHigherGroupsHomotopy2018,symmetryBook2021}. \footnote{This technique also
+%   allows defining higher groups in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}
 

### 2021-07-04 Pi+/paper/ufin.tex
Cut more in section 3

@@
 is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that
-  is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type $\UFin$, which is a 1-groupoid.
+is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type $\UFin$, which is a 1-groupoid. In this
+section, we develop the tools necessary to construct $\UFin$, starting with a quick review of HoTT.
 
@@
 
+The weak groupoid structure of types is given by the identity type in HoTT, where points are given by terms of the type,
+and higher morphisms are given by the iterated identity type. Functions between types are functors between groupoids,
+and type families, or functions to the universe, are indexed families of groupoids. A type family $P : A \to \UU$ comes
+equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$ which has a lifting operation giving it the structure of a
+fibration (see the appendix for details).
+
+\begin{toappendix}
 \subsubsection{Identity Types}
@@
 
-\begin{toappendix}
   \begin{definition}[Path Induction]
@@
   \end{definition}
-\end{toappendix}
 
@@
 
-Further, using the groupoid structure of $A$, we can show that $\transport{P}{p}{\blank}$ and
-$\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name the map $\tptEqv{P}$ which lifts paths to equivalences
-of fibers.
+\end{toappendix}
+
+The $\term{transport}$ operation lifts paths in the base space to maps between the fibers. Using the groupoid structure
+of $A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name
+this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
 
@@
 
-\todo{Diagram.}
 The type families (or fibrations) we're interested in are the ones where paths in the base space completely determine
@@
 \UU \to \UU$ being univalent.
+
+\begin{toappendix}
 
@@
 
-\begin{toappendix}
   \begin{definition}[Propositional Truncation]
@@
 
-\begin{toappendix}
   \begin{definition}[Set Quotient]
@@
   \end{definition}
+
 \end{toappendix}
@@
 certain types, for example, types that satisfy some desired property. We use a prop-valued type family, that is, a
-predicate, which picks out only those types, and collect them into a univalent subuniverse. Being univalent ensures that
-the equality type of the ambient universe is reflected in the subuniverse.
+predicate on the universe, which picks out only those types, and collect them into a univalent subuniverse. Being
+univalent ensures that the equality type of the ambient universe is reflected in the subuniverse.
 
@@
 \begin{definition}[Subuniverse]
-  A subtype is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition for every
-  $X$. A subuniverse generated by a subtype has $U \defeq \dsum{X:\UU}{P(X)}$ and $\El \defeq \fst$.
+  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition
+  for every $X$. A subuniverse generated by $P$ is given by its total space $U \defeq \dsum{X:\UU}{P(X)}$ and $\El
+  \defeq \fst$.
 \end{definition}
@@
 \begin{proposition}[Univalent Subuniverse]
-  Subuniverses generated by subtypes are univalent.
+  Subuniverses generated by predicates are univalent.
 \end{proposition}
@@
 The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is subtle. We
-define a type to be (Bishop-) finite if it is equal to some finite set, and we setup the predicate $\isFin$ to do so.
-
+define a type to be (Bishop) finite if it is equal to some finite set, and we setup the predicate $\isFin$ to do so.
 \todo{Cite Brent Yorgey's thesis?}
@@
 Using $\BAut$, we can talk about types that have a specified cardinality, for example, the subuniverse of 2-element sets
-is given by $\BAut[\Bool]$. This can be used to give the denotational semantics of a 1-bit reversible
+is given by $\BAut[\Bool]$. This has been used to construct the real projective spaces in
+HoTT~\cite{buchholtzRealProjectiveSpaces2017}, and also to give the denotational semantics for a 1-bit reversible
 language~\cite{caretteReversibleProgramsUnivalent2018}.
@@
 \begin{corollary}
-  $\UFin$ is a pointed, connected, 1-type, that is, a 1-groupoid. For every $n:\Nat$, \( \loopspace[\UFin[n],F_{n}] \eqv
-  \Aut[\Fin[n]] \).
+  $\UFin$ is a pointed, connected, 1-groupoid. For every $n:\Nat$, \( \loopspace[\UFin[n],F_{n}] \eqv \Aut[\Fin[n]] \).
 \end{corollary}
-\note{Diagram below?}
+
+The loopspace of a pointed type automatically has the structure of a group, that is, with $\refl_{\pt}$ for the neutral
+element, path composition for the group multiplication, and path inverse for the group inverse operation. The group
+axioms are then given by the higher paths corresponding to groupoid laws. We have shown that loops in $\UFin$ exactly
+encode the automorphism group $\Aut[\Fin[n]]$ for every $n$. This is a general technique called delooping, where a group
+can be identified with a 1-object groupoid, internally in HoTT.~\footnote{This technique also allows defining higher
+groups in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}
+
 % \[

### 2021-07-04 Pi+/paper/ufin.tex
Update section titles
Move some text in UFin

@@
-\section{The (Sub)Universe of Finite Types}
+\section{The Groupoid of Finite Types}
 \label{sec:ufin}
@@
 
-To define the weak groupoid we're after, we first introduce some basic concepts and notation from Homotopy Type Theory
-that we use.~\footnote{We use the language of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013},
-that is, we use intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive
-Types (HITs) for propositional truncation and set quotients. All arguments will hold in a Cubical Type Theory as well,
-such
+We review some basic concepts and notation from Homotopy Type Theory that we use in the appendix.~\footnote{We use the
+language of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use intensional
+Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for propositional
+truncation and set quotients. All arguments will hold in a Cubical Type Theory as well, such
 as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.}
 For more details about HoTT, we refer the reader to the HoTT
-book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}. By formalising our results in HoTT, we're also able to
-use a proof assistant to computer-check our proofs, and our results are formalised in the Agda proof assistant using the
-HoTT-Agda library. Using the formalisation of our semantics, we are able to extract a computable procedure for
-performing NbE and synthesis of reversible circuits, as motivated in~\cref{sec:introduction}.
+book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.
 
-The weak groupoid structure of types is given by the identity type in HoTT, where points are given by terms of the type,
-and higher morphisms are given by the iterated identity type. Functions between types are functors between groupoids,
-and type families, or functions to the universe, are indexed families of groupoids. A type family $P : A \to \UU$ comes
-equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$ which has a lifting operation giving it the structure of a
-fibration (see the appendix for details).
+To recap, types in HoTT have a weak groupoid structure given by the identity type, where points are given by terms of
+the type, and higher morphisms are given by the iterated identity type. Functions between types are functors between
+groupoids, and type families, or functions to the universe, are indexed families of groupoids. A type family $P : A \to
+\UU$ comes equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$ which has a lifting operation giving it the
+structure of a fibration (see the appendix for details).
 
@@
 The identity type for functions is equivalent to homotopies between them ${(f \id_{A \to B} g)} \eqv {(f \htpy g)}$, by
-function extensionality. An equivalent between types $A \eqv B$ is given by a pair of functions between them which
-compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy \idfunc_{A}$. This is equivalent to the
+function extensionality. An equivalence between types $A \eqv B$ is given by a pair of functions between them which
+compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy \idfunc_{A}$, and this is equivalent to the
 identity type for the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.
@@
 \end{toappendix}
+
+\subsection{Univalent Fibrations}
 

### 2021-07-04 Pi+/paper/reversible1.tex
Cut more in section 3
proof outline for thm1

@@
 
-\section{A Reversible Programming Language: Set-Theoretic Semantics}
+\section{A Reversible Programming Language}
 %% ~3 pages
@@
 $(A,(B,C))$): if the bit is true, we perform an \verb|x| operation on component $A$, and otherwise we perform a
-\verb|cx| operation on $C,A$. The two uses of \verb|x| gates in the circuit are now unnecessary as they were only needed
+\verb|cx| operation on $(C,A)$. The two uses of \verb|x| gates in the circuit are now unnecessary as they were only needed
 to encode a two-way conditional expression using a sequence of one-way conditional expressions (the only ones available in
@@
 \end{theorem}
-%% proof?
+\begin{proof}
+  For every primitive combinator $c$ listed on one side of Fig.~\ref{fig:pi-terms}, let $!c$ be the combinator listed on
+  the other side. Thus $! \assoclp$ is $\assocrp$ and $! \swapp$ is $\swapp$ itself. Then we have that $\denot{c}$ and
+  $\denot{!c}$ is an equivalence. For the level-2 combinator \Afun{idr◎l}, we check
+  $\denot{\AgdaBound{c}~\AgdaOperator{\AgdaInductiveConstructor{◎}}~\AgdaInductiveConstructor{id⟷₁}}
+  = \mathit{id} \circ \denot{c} = \denot{c}$. The other cases are only slightly more involved calculations.
+\end{proof}
 

### 2021-07-05 Pi+/paper/finite.tex
Update section titles
Revising sec 4

--- 2021-07-04 Pi+/paper/finite.tex
@@
-\section{The groupoid of finite types}~\label{sec:finite}
+\section{The Group of Permutations}~\label{sec:finite}
 
-In the previous~\cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin{n}$
-for every $n:\Nat$, that is, automorphisms of finite sets of size $n$. The loopspace has a group structure -- the
-identity function is the neutral element, group multiplication is given by function composition, and the group inverse
-is given by inverting the bijection. This is also known to be the finite symmetric group $\Sn[n]$. In the following
-sections, we will characterise these permutations syntactically.
+In~\cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
+$n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group.This is
+also known to be the finite symmetric group $\Sn[n]$. In the following sections, we will describe this group
+syntactically.
 
 In order to study syntactic descriptions of permutations, we need to decide whether two descriptions refer to the same
-permutation. This problem is well known in group theory as the \emph{word problem} for $\Sn$. Putting it in this form
-also allows us to connect to the broader scope of computational group theory and combinatorics -- we borrow ideas such
-as Coxeter relations and Lehmer codes from there.
+permutation. This is the \emph{word problem} for $\Sn$. Putting it in this form also allows us to connect it to the
+broader scope of computational group theory and combinatorics -- we borrow ideas such as Coxeter relations and Lehmer
+codes from there.
 
 The goal of this section is to reconcile two different approaches to defining the symmetric group -- as an automorphism
-group, and as a group presented using generators and relations.
+group, and as a group presented using generators and relations. The generators of the group can be thought of as the
+primitive combinators in a (reversible) programming language, the group structure gives the composition and inverse
+operations, and the relations describe how these primitive combinators interact with each other.
 
+\subsection{Presenting the permutation group}
+
+One way of thinking about presentations of $\Sn$ is via sorting algorithms, which use different primitive operations. A
+sorting algorithm has to calculate a permutation of a list or a finite set, which satisfies the invariant of being a
+sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the permutations
+on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate for the
+generators of a permutation group. For example, we could generate the permutation group on $\Fin[n]$ by using generators
+(primitive operations) that:
+
+\begin{itemize}
+  \item swap the $i$-th element with the $(i+1)$-th element, that is, adjacent swaps, or
+  \item swap the $i$-th element with the $j$-th element, for arbitrary $i$-s and $j$-s, or
+  \item swap the $i$-th element with an element at a fixed position, or
+  \item flip a prefix $\Fin[k]$ of $\Fin[n]$ for $k \leq n$, or
+  \item cyclically shift any subset of $\Fin[n]$.
+\end{itemize}
+
+Bubble sort uses the primitive operation of adjacent swaps, insertion sort and selection sort use the primitive
+operation of swapping the $i$-th element with the $j$-th element, cycle sort uses cyclical shifts of subsequences,
+pancake sort uses flips of prefixes of the list, et cetera~\todo{check!}. The choice of generators for our presentation
+is important for the following reasons.
+
+\begin{itemize}
+  \item It affects the difficulty of solving the word problem in $\Sn$ and formalising the proof of its correctness. In
+        the following subsections, we build a rewriting system using the generators and relations, and we need to prove
+        and formalise strong normalisation for this system.
+  \item The choice of generators dictates which words become normal forms in this presentation of $\Sn$. These normal
+        forms dictate the shape of the synthesised and normalised boolean circuits, which is the application we have in
+        mind.
+  \item Finally, the generators have to closely match the $\PiLang$ combinators so that we can quote back a permutation
+        to a $\PiLang$ program, for the proof of completeness.
+\end{itemize}
+
+We use a presentation based on adjacent transpositions for generators, called a~\emph{Coxeter presentation}, which we
+also use to solve the word problem for $\Sn$. Connecting it back to the language $\PiLang$, we will show that it is
+possible to encode all $\PiLang$ combinators using adjacent transpositions (in~\cref{sec:equivalence}). The rest of this
+section develops these intuitions formally. Group presentations are built by adding equations to a \emph{free group},
+which we define first.
+
+\begin{toappendix}
   \subsection{Groups}
@@
 
-\begin{toappendix}
   \begin{definition}~\label{def:group}
@@
 
-However, more conveniently, in HoTT, we can instead use groupoids to talk about groups internally. A group can be
-identified with a 1-object groupoid, using a technique called delooping. The delooping of a group $G$ is a groupoid
-$\B{G}$, with a unique object $\pt$, where the group elements are encoded as self-loops $\pt \id_{\B{G}} \pt$.
-
-The group operations are automatically given by operations on the identity type, with $\refl_{\pt}$ for the neutral
-element, path composition for the group multiplication, and path inverse for the group inverse. These satisfy the group
-laws, up to the identity type, using the groupoid coherence laws.
-
-Since groups are sets, these 2-paths are propositions, making $\B{G}$ a 1-groupoid. Hence, a group is simply given by a
-pointed, connected 1-type~\cite*{buchholtzHigherGroupsHomotopy2018,symmetryBook2021}. \footnote{This technique also
-  allows defining higher groups in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}
-
-% For example, given a pointed type $(A:\UU, a:A)$, the automorphism group structure at $a$ is given by $a \id_{A} a$. Of
-% course, for 1-groups we will require that $a \id_{A} a$ is a set, which is enforced by having $A$ be a groupoid. In our
-% running example for the permutation group on finite sets, we have that $\Fin[n]$ is a set, and hence, $\UFin[n] \defeq
-% \BAut[\Fin[n]]$ is a pointed, connected 1-type. Its loopspace $\loopspace[\BAut[\Fin[n]],F_{n}]$ is equivalent to
-% $\Aut[\Fin[n]] \defeq (\Fin[n] \eqv \Fin[n]) \eqv (\Fin[n] \id_{\UU} \Fin[n])$, which has the corresponding automorphism
-% group structure.
-
 \subsection{Free groups}
@@
 axioms, but is specific to this particular group. A free group has the property that no other equations hold except the
-ones directly implied by the group axioms. To be able to give an equational presentation of a group, we start by
-constructing the free group, then show how to add new equations to it.
-
-Given any set $A$, we can describe the free group $F(A)$ -- we call $A$ the generating set of $F(A)$. We draw the
-elements from $A$, close them under the group operations, and identify them by the group axioms. For example, the
-additive group of integers $\mathbb{Z}$ is the free group on the singleton. In HoTT, we can use a higher inductive type
-to define the free group, by adding path constructors for the axioms we want (see~\cref{def:free-group}).
+ones directly implied by the group axioms. For example, the additive group of integers $\mathbb{Z}$ is the free group on
+the singleton set.
 
@@
 A group homomorphism between groups $G$ and $H$, which we write as $G \to^G H$, is a function between the underlying
-sets that preserves the group structure. The universal property of free groups, stemming from the free-forgetful
-adjunction between the category of groups and sets, follows from the induction principle for $F(A)$. Informally, it
-states that if we know how a map $f$ acts on the generating set $A$, we know exactly how it acts on every element of the
-group $F(A)$.
+sets that preserves the group structure. Giving a group homomorphism out of the free group is equivalent to giving a
+function out of the generating set. This is the universal property of free groups, stemming from the free-forgetful
+adjunction between the category of groups and sets.
 
-\begin{proposition}[Universal Property of $F(A)$]~\label{prop:free-groups}
+% Informally, it states that if we know how a map $f$ acts on the
+% generating set $A$, we know exactly how it acts on every element of the group $F(A)$.
+
+\begin{proposition}[Universal Property of $F(A)$]
+  \label{prop:free-groups}
   Given a group $G$ and a map $f : A \to G$, there is a unique group
@@
 
-Although the description of $F(A)$ discussed above is close to the intuitive universal-algebraic definition, it is not
-the most convenient one to work with. Besides the elements of the group, we also want to characterise its the path
-space. This definition of $F(A)$ has lots of path constructors corresponding to each group axiom, making it difficult to
-describe its path space.
+Following the universal-algebraic definition, in HoTT, we could use a naive higher inductive type to define the free
+group, which enforces the group axioms by adding path constructors (see~\cref{def:free-group} for the full definition).
+Using the induction principle, we can easily verify the universal property. However, since this definition of $F(A)$ has
+lots of path constructors corresponding to each group axiom, characterising its path space is difficult.
 
@@
 
-A presentation of a group builds it by starting from the free group $F(A)$ and introducing a collection of equations
-that have to be satisfied in the resulting group. For example, if we take $F(\unit) \defeq \mathbb{Z}$ and add an
-equation $1 + 1 = 0$, the resulting group would be $\mathbb{Z}_2 \eqv \Aut[\Bool]$. Note that not all groups have finite
-(or computable) presentations, and, a group can have any number of different presentations.
+A presentation of a group builds it by starting from the free group $F(A)$ and introducing additional equations that are
+satisfied in the resulting group. For example, if we take $F(\unit) \defeq \mathbb{Z}$ and add an equation $1 + 1 = 0$,
+the resulting group would be $\mathbb{Z}_2 \eqv \Aut[\Bool]$. Note that not all groups have finite (or computable)
+presentations, and, a group can have any number of different presentations.
 
@@
   $F(\langle A ; R \rangle)$ presented by $A$ and $R$, is given by the set-quotient of the free group $F(A)$ by the
-  closure of $R$.
+  congruence closure of $R$.
 \end{definition}
 
-The universal property of the above definition allows for properly extending the relation on the generating set to the
-whole group.\vc{check!}
+The universal property of the above definition is similar to~\cref{prop:free-groups} except the relation has to be
+preserved by the function mapping out of the generating set.
 
 \begin{proposition}[Universal property of $F(\langle A ; R \rangle)$]
-  Given a group $G$ and a map $f : A \to G$ that respects $R$, there is a unique group homomorphism 
-  $\extend{f} : F(\langle A ; R \rangle) \to^G G$ such that $\extend{f} \comp \eta_A \htpy f$.
+  Given a group $G$ and a map $f : A \to G$, such that $f$ extended to $F(A)$ respects $R$, there is a unique group
+  homomorphism $\extend{f} : F(\langle A ; R \rangle) \to^G G$ such that $\extend{f} \comp \eta_A \htpy f$.
 \end{proposition}
 
-To relate this to our original problem, the generators of the group can be thought of as the primitive combinators in a
-(reversible) programming language, the group structure gives the composition and inversion operations, and the relations
-describe how these primitive operations interact with each other.
+Before, the only way to decide the equality of two elements in a group was to compute and check them on the nose, but in
+a group presentation, this is reduced to a \emph{word problem}, that is, deciding whether one word -- a representative
+of the equivalence class of the group's elements, can be reduced to another word, using the group's relations. However,
+these equations are not directed, so it is not always possible to construct a well-behaved rewriting system. In general,
+the word problem for groups is known to be undecidable.\jk{Example?}\todo{Examples: empty relation, full relation, van
+  Kampen of $\pi_{1}$}
 
-As we mentioned, instead of operating directly with a particular group, we focus on the syntactic presentation. While
-before, the only way to decide the equality of two elements in the group was to compute and check them on the nose, now
-it is reduced to a \emph{word problem}, that is, deciding whether one word -- a representative of the group elements'
-equivalence class, can be reduced to another word, using the group's relations. This can be thought of as a rewriting
-system. However, these equations are not directed, so it is not always possible to construct a well-behaved rewriting
-system. In general, the word problem for groups is proven to be undecidable.
-
-\jk{Example?}
-\todo{Examples: empty relation, full relation, van Kampen of $\pi_{1}$}
-
-\subsection{Presenting the permutation group}
-
-We have already seen that the group of permutations, or the symmetric group $\Sn$, can be defined by taking the type
-$\Aut[\Fin[n]]$ as the underlying set, and showing that it has the appropriate group structure. Now we're going to give
-a presentation for it, by defining a set of generators and relations.
-
-There is an element of choice here -- as we already mentioned, a group can be presented in many different ways. For
-example, we could generate the permutation group on $\Fin[n]$ by using generators that:
-
-\begin{itemize}
-  \item swap the $i$-the element with the $(i+1)$-th element, that is, adjacent swaps, or
-  \item swap the $i$-th element with the $j$-th element, for arbitrary $i$-s and $j$-s, or
-  \item swap the $i$-th element with an element at a fixed position, or
-  \item flip a prefix $\Fin[k]$ of $\Fin[n]$ for $k \leq n$, or
-  \item cyclically shift any subset of $\Fin[n]$.
-\end{itemize}
-
-One way of thinking about these presentations is via sorting algorithms, which use different primitive operations. A
-sorting algorithm has to calculate a permutation of a list or a finite set, which satisfies the invariant of being a
-sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the permutations
-on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate for the
-generators of a permutation group.
-
-For example, bubble sort uses the primitive operation of adjacent swaps, insertion sort and selection sort use the
-primitive operation of swapping the $i$-th element with the $j$-th element, cycle sort uses cyclical shifts of
-subsequences, pancake sort uses flips of prefixes of the list, et cetera~\todo{check!}. The choice of generators for our
-presentation is important for the following reasons.
-
-\begin{itemize}
-  \item It affects the difficulty of solving the word problem in $\Sn$ and formalising the proof of its correctness. In
-        the following subsections, we build a rewriting system using the generators and relations, and we need to prove
-        and formalise strong normalisation for this system.
-  \item The choice of generators dictates which words become normal forms in this presentation of $\Sn$. These normal
-        forms dictate the shape of the synthesised and normalised boolean circuits, which is the application we have in
-        mind.
-  \item Finally, the generators have to closely match the $\PiLang$ combinators so that we can quote back a permutation
-        to a $\PiLang$ program, for the proof of completeness.
-\end{itemize}
-
-We use a presentation based on adjacent transpositions for generators, called a~\emph{Coxeter presentation}, which we
-also use to solve the word problem for $\Sn$. Connecting it back to the language $\PiLang$, we will show that it is
-possible to encode all $\PiLang$ combinators using adjacent transpositions (in~\cref{sec:equivalence}).
+% This can be thought of as word reductions in a rewriting system.
 
@@
 
-The primitive operations we use are going to be adjacent swaps. When dealing with permutations on an $n + 1$-element
-set, there are $n$ adjacent transpositions, transposition number $k$ swapping elements $\el{k}$ and $\el{k+1}$. Thus,
-the generating set would be $\Fin[n]$. There are three relations that we're going to specify for this presentation --
-these are the laws that these generators should satisfy. It is easiest to visualise them as braid diagrams.
+To present the group $\Sn$, the primitive operations we are going to use will be adjacent swaps. When dealing with
+permutations on an $n + 1$-element set, there are $n$ adjacent transpositions, transposition number $k$ swapping
+elements $(\el{k}, \el{k+1})$. Thus, the generating set would be $\Fin[n]$. There are three relations that we're going
+to specify for this presentation -- these are the equations that these generators should satisfy. We visualise them as
+braid diagrams in~\cref{fig:coxeter-braid}.
 
-First, swapping the same two elements two times in a row should be the same as doing nothing:
+\begin{enumerate}
+  \item[\labelcref{fig:coxeter-a}] Swapping the same two elements two times in a row should be the same as doing nothing
+  \item[\labelcref{fig:coxeter-b}] When swapping two distinct pairs of elements, it should not matter in which order the
+    swapping happens, that is, we can slide the wires freely.
+  \item[\labelcref{fig:coxeter-c}] If the next transposition in the sequence is equal to the the first one, the order of
+    operations can be reversed.
+\end{enumerate}
 
-\[
-  \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
-    \begin{center}
+\begin{figure}
+  \centering
+  \begin{subfigure}[b]{0.3\textwidth}
+    \centering
     \begin{tikzpicture}
@@
         number of strands = 2,
-          thick]
+        width = 0.6cm,
+        height = 0.4cm,
+        border height = 0.3cm,
+        thick] at (0, 0)
       {braid={ s_1, s_1}};
+      \node[font=\Huge] at (2, 0.7) {\(\rightarrow\)};
+      \pic[local bounding box=my braid,braid/.cd,
+        number of strands = 2,
+        width = 0.6cm,
+        height = 0.4cm,
+        border height = 0.3cm,
+        thick] at (3, 0)
+      {braid={ 1, 1}};
     \end{tikzpicture}
-    \end{center}
-     &
-    \(\xlongrightarrow[]{\cancel}\)
-     &
-    \begin{center}
+    \caption{$\cancel$}
+    \label{fig:coxeter-a}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{0.5\textwidth}
+    \centering
     \begin{tikzpicture}
@@
       number of strands = 2,
-          thick]
-        {braid={1, 1}};
-      \end{tikzpicture}
-    \end{center}
-  \end{tabular}
-\]
-
-Second, when swapping two distinct pairs of elements (i.e. when the indices of two transpositions differ by at least 1),
-it should not matter in which order the swapping happens, that is, we can slide the wires freely.
-
-\[
-  \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
-    \begin{center}
-      \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
-        \begin{tikzpicture}
+      width = 0.6cm,
+      height = 0.4cm,
+      border height = 0.3cm,
+      thick] at (0, 0)
+      {braid={ 1, s_1}};
+      \node[] at (1, 0.7) {\(\dots\)};
       \pic[local bounding box=my braid,braid/.cd, 
       number of strands = 2,
-            thick]
-          {braid={ 1, s_1}};
-        \end{tikzpicture}
-         &
-        \(\cdots\)
-         &
-        \begin{tikzpicture}
+      width = 0.6cm,
+      height = 0.4cm,
+      border height = 0.3cm,
+      thick] at (1.5, 0)
+      {braid={ s_1, 1}};
+      \node[font=\Huge] at (3, 0.7) {\(\rightarrow\)};
       \pic[local bounding box=my braid,braid/.cd, 
       number of strands = 2,
-            thick]
-          {braid={ s_1, 1 }};
-        \end{tikzpicture}
-      \end{tabular}
-    \end{center}
-     &
-    \(\xlongrightarrow[]{\swap}\)
-     &
-    \begin{center}
-      \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
-        \begin{tikzpicture}
+      width = 0.6cm,
+      height = 0.4cm,
+      border height = 0.3cm,
+      thick] at (4, 0)
+      {braid={ 1, s_1}};
+      \node[] at (5.2, 0.7) {\(\dots\)};
       \pic[local bounding box=my braid,braid/.cd, 
       number of strands = 2,
-            thick]
+      width = 0.6cm,
+      height = 0.4cm,
+      border height = 0.3cm,
+      thick] at (5.5, 0)
       {braid={ s_1, 1}};
   \end{tikzpicture}
-         &
-        \(\cdots\)
-         &
-        \begin{tikzpicture}
-          \pic[local bounding box=my braid,braid/.cd,
-            number of strands = 2,
-            thick]
-          {braid={ 1, s_1}};
-        \end{tikzpicture}
-      \end{tabular}
-    \end{center}
-  \end{tabular}
-\]
-
-The previous two cases were for the transpositions that either completely overlap, or not overlap at all. The third case
-says what happens if we perform transpositions that move the same elements. If the next transposition in the sequence is
-equal to the the first one, we endorce a third law -- the braiding relation.
-
-\[
-  \begin{tabular}{m{0.3\linewidth}m{0.1\linewidth}m{0.3\linewidth}}
-    \begin{center}
+    \caption{$\swap$}
+    \label{fig:coxeter-b}
+  \end{subfigure}
+  \hfill
+  \begin{subfigure}[b]{\textwidth}
+    \centering
     \begin{tikzpicture}
@@
         number of strands = 3,
-          thick]
+        width = 0.6cm,
+        height = 0.4cm,
+        border height = 0.1cm,
+        thick] at (0, 0)
       {braid={ s_2, s_1, s_2}};
-      \end{tikzpicture}
-    \end{center}
-     &
-    \(\xlongrightarrow[]{\braid}\)
-     &
-    \begin{center}
-      \begin{tikzpicture}
+      \node[font=\Huge] at (2, 0.7) {\(\rightarrow\)};
       \pic[local bounding box=my braid,braid/.cd,
         number of strands = 3,
-          thick]
+        width = 0.6cm,
+        height = 0.4cm,
+        border height = 0.1cm,
+        thick] at (3, 0)
       {braid={ s_1, s_2, s_1}};
     \end{tikzpicture}
-    \end{center}
-  \end{tabular}
-\]
-
+    \caption{$\braid$}
+    \label{fig:coxeter-c}
+  \end{subfigure}
+  \caption{Braiding diarams for Coxeter relations.}
+  \label{fig:coxeter-braid}
+\end{figure}
 \todo{an example/motivation from book.pdf}
 
-This construction is called a Coxeter presentation. Writing it formally, we take the generating set to be $\Fin[n]$,
-where the element $k$ corresponds to an adjacent transpositions $\tau_k$, which swaps elements $\el{k}$ and $\el{k+1}$.
-Then, we define a binary relation $\cox$ on $\List[\Fin[n]]$, which encodes the laws discussed above.
+This construction is called a Coxeter presentation of $\Sn$. Writing it formally, we encode the rules discussed above
+using a relation $\cox$ on $\List[\Fin[n]]$, and then take its congruence closure $\cox*$.
 
-\begin{definition}[$\cox$]
+\begin{definition}[$\cox : {\List[\Fin[n]]} \to {\List[\Fin[n]]} \to {\UU}$]
   \begin{align*}
@@
 \end{definition}
-
-We define $\cox*$ as the congruence closure of $\cox$.
-
-\begin{definition}[$\cox*$]
+\begin{definition}[$\cox* : {\List[\Fin[n]]} \to {\List[\Fin[n]]} \to {\UU}$]
   \begin{align*}
@@
 
-To solve the word problem for $\Sn$, we turn the relations into a rewriting system $(\List[\Fin[n]],\cox*)$. A normal
-form of an element can then be computed by following the rewrite rules, and two words can be compared for
-$\cox*$-equality by comparing their normal forms. A well-behaved rewriting system has to be strongly normalising and
-confluent. First, we observe that after throwing out reflexivity and symmetry, the right hand sides of the relations
-$\cox*$ are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$.
-Thus, by directing the relation from left to right, we would get the termination property out of the box.
+The idea for solving the word problem for $\Sn$ is to turn these undirected relations into a terminating rewriting
+system $(\List[\Fin[n]],\cox*)$, so that, by repeatedly applying the reduction rules as long as possible, any two
+$\cox*$-equal terms would eventually converge to the same normal form.
 
-The system also has to be confluent, meaning that all critical pairs, that is, terms with overlapping possible reduction
-rules, have to converge. For example, the pairs below converge.
+For this to work, we first need the system to have the strong normalisation property, meaning that there are no infinite
+reductions. We observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
+are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$ (which is
+well-founded). Thus, by directing the relation from left to right, we would get the termination property out of the box.
+Second, we need the normal forms to be unique, so that we can get a normalisation function. This will be true if the
+rewriting system is confluent -- meaning that all critical pairs, that is, terms with overlapping possible reduction
+rules, have to converge. For example, in our system, the pairs in~\cref{fig:critical-pairs-converge} converge.
+Unfortunately, this is not true for all critical pairs -- an  example is in~\cref{fig:critical-pairs-non-converging},
+where both endpoints are normal with respect to the $\cox*$ relation.
 
-\[
-  \begin{array}{lcr}
-    \gspan[\braid][\braid]{\tau_2\tau_1\tau_2\tau_1\tau_2}{\tau_1\tau_2\tau_1\tau_1\tau_2}{\tau_2\tau_1\tau_1\tau_2\tau_1}
-     &
-    \text{or}
-     &
-    \gspan[\braid][\cancel]{\tau_2\tau_1\tau_2\tau_2}{\tau_1\tau_2\tau_1\tau_2}{\tau_2\tau_1}
-  \end{array}
-\]
+% \begin{figure}
+% \[
+%   \begin{array}{lcr}
+%     \gspan[\braid][\braid]{\tau_2\tau_1\tau_2\tau_1\tau_2}{\tau_1\tau_2\tau_1\tau_1\tau_2}{\tau_2\tau_1\tau_1\tau_2\tau_1}
+%      &
+%     \text{or}
+%      &
+%     \gspan[\braid][\cancel]{\tau_2\tau_1\tau_2\tau_2}{\tau_1\tau_2\tau_1\tau_2}{\tau_2\tau_1}
+%   \end{array}
+% \]
+% \caption{Critical pairs}
+% \label{fig:critical-pairs}
+% \end{figure}
 
-\begin{proof}
-  \[
-    \begin{array}{lr}
+\begin{figure}
+  \begin{subfigure}[b]{0.45\textwidth}
     % https://q.uiver.app/?q=WzAsNixbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzQsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwyLCJcXHRhdV8xXFx0YXVfMlxcdGF1XzIiXSxbMiwzLCJcXHRhdV8xIl0sWzQsMiwiXFx0YXVfMlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFswLDIsImJyYWlkIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJjYW5jZWwiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFszLDQsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzIsNSwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbNSw0LCJjYW5jZWwiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dXQ==
+    \adjustbox{scale=\scalef,center}{
       \begin{tikzcd}
@@
         && {\tau_1}
-        \arrow["braid"', squiggly, from=1-3, to=2-1]
-        \arrow["braid", squiggly, from=1-3, to=2-5]
-        \arrow["cancel"', squiggly, from=2-1, to=3-1]
-        \arrow["cancel"', squiggly, from=3-1, to=4-3]
-        \arrow["cancel", squiggly, from=2-5, to=3-5]
-        \arrow["cancel", squiggly, from=3-5, to=4-3]
+        \arrow["\braid"', squiggly, from=1-3, to=2-1]
+        \arrow["\braid", squiggly, from=1-3, to=2-5]
+        \arrow["\cancel"', squiggly, from=2-1, to=3-1]
+        \arrow["\cancel"', squiggly, from=3-1, to=4-3]
+        \arrow["\cancel", squiggly, from=2-5, to=3-5]
+        \arrow["\cancel", squiggly, from=3-5, to=4-3]
       \end{tikzcd}
-       &
+    }
+  \end{subfigure}
+  \begin{subfigure}[]{0.45\textwidth}
     % https://q.uiver.app/?q=WzAsNSxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8yIl0sWzAsMSwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFsyLDQsIlxcdGF1XzJcXHRhdV8xIl0sWzAsMiwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFsxLDMsIlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwxLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzAsMiwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzMsNCwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFs0LDIsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV1d
+    \adjustbox{scale=\scalef,center}{
       \begin{tikzcd}
@@
         && {\tau_2\tau_1}
-        \arrow["braid"', squiggly, from=1-3, to=2-1]
-        \arrow["cancel", squiggly, from=1-3, to=5-3]
-        \arrow["braid"', squiggly, from=2-1, to=3-1]
-        \arrow["braid"', squiggly, from=3-1, to=4-2]
-        \arrow["cancel"', squiggly, from=4-2, to=5-3]
+        \arrow["\braid"', squiggly, from=1-3, to=2-1]
+        \arrow["\cancel", squiggly, from=1-3, to=5-3]
+        \arrow["\braid"', squiggly, from=2-1, to=3-1]
+        \arrow["\braid"', squiggly, from=3-1, to=4-2]
+        \arrow["\cancel"', squiggly, from=4-2, to=5-3]
       \end{tikzcd}
-    \end{array}
-  \]
-\end{proof}
+    }
+  \end{subfigure}
+  \caption{Converging critical pairs}
+  \label{fig:critical-pairs-converge}
+\end{figure}
 
-Unfortunately, in the system we defined, this is not true for all critical pairs. One example is the following, where
-both endpoints are normal with respect to $\cox*$ relation.
-
-\[
-  \gspan[\braid][\swap]{\tau_3\tau_2\tau_3\tau_1}{\tau_2\tau_3\tau_2\tau_1}{\tau_3\tau_2\tau_1\tau_3}
-\]
+\begin{figure}
+  $\gspan[\braid][\swap]{\tau_3\tau_2\tau_3\tau_1}{\tau_2\tau_3\tau_2\tau_1}{\tau_3\tau_2\tau_1\tau_3}$
+  \caption{Non-converging critical pairs}
+  \label{fig:critical-pairs-non-converging}
+\end{figure}
 
@@
 
-Because of the counter-example discussed, the relations have to be changed. In this section, we will formally define a
-rewriting system partially based on the Coxeter relations, and prove that it has the desired properties of confluence
-and strong normalisation.
+Because of this counter-example, the relations have to be changed. In this section, we formally define a rewriting
+system $(\List[\Fin[n]], \longcox*)$, partially based on the Coxeter relations, and prove that it has the desired
+properties of confluence and strong normalisation. We prove that the new relation defined by this system is equivalent,
+in a technical sense, to the standard Coxeter relation $\cox*$.
 
-The Coxeter relations are a standard notion in the theory of group presentation~\todo{citation} -- even though we change
-the relations a bit, we will not lose the connection. In fact, we will prove that the new relations are equivalent, in a
-technical sense, to the standard Coxeter relations $\cox*$.
+% The Coxeter relations are a standard notion in the theory of group presentation~\todo{citation} -- even though we change
+% the relations a bit, we will not lose the connection. In fact, we will prove that the new relations are equivalent, in a
+% technical sense, to the standard Coxeter relations $\cox*$.
+% The system has generators corresponding to $\swap$, $\cancel$
+% and $\braid$, the problem of the non-converging critical pairs discussed previously is fixed by changing $\braid$ to be
+% a slightly more general relation $\longbraid$.
 
-The new reduction system $(\List[\Fin[n]], \longcox*)$ has generators corresponding to $\swap$, $\cancel$ and $\braid$.
-We fix the problem of the non-converging critical pairs discussed previously by changing $\braid$ to be a slightly more
-general relation $\longbraid$. \review{Further, we also inline the congruence closure in $\longcox$, allowing arbitrary
-  reductions inside the list.}
+\review{Further, we also inline the congruence closure in $\longcox$, allowing arbitrary reductions inside the list.}
 
@@
 The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\). Since we think of the
-number $m$ as the transposition $\tau_m$ which swaps elements $\el{m}$ and $\el{m+1}$, the role of this helper function
-is to produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ $k$ places left,
-shifting all the elements in between one place right. Expressed in terms of the braiding diagram, for $n = 0$ and $k =
-  4$, it has the following form:
+number $m$ as the transposition swapping elements $(\el{m}, \el{m+1})$, the role of this helper function
+is to produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ by $k$ places left,
+shifting all the elements in between one place right (see~\cref{fig:downf}).
 
+\begin{figure}
   \[
@@
   \]
+  \caption{Braiding diagram for $0 \downf 4$}
+  \label{fig:downf}
+\end{figure}
 
@@
 sequence of transpositions and a single transposition afterwards. For example, for $n = 2$ and $k = 3$, it allows for
-the reduction $[6, 5, 4, 3, 2, 6] \longcox* [5, 6, 5, 4, 3, 2]$. Visualised on the braid diagram, it looks as follows:
-
+the reduction $[6, 5, 4, 3, 2, 6] \longcox* [5, 6, 5, 4, 3, 2]$ (for the appropriate braid diagram,
+see~\cref{fig:longcox}).
+\begin{figure}
   \[
@@
   \]
+  \caption{Braiding diagram for $\longcox$ with $n = 0$ and $k = 3$}
+  \label{fig:longcox}
+\end{figure}
 
@@
 we desire. It satisfies confluence, that is, the Church-Rosser (diamond) property, and it is strongly normalising,
-because it produces a unique irreducible normal form. Existence of the normal form is implied by the well-foundedness of
-the relation -- informally, it does not contain any infinite decreasing sequences. We follow the terminology
+because it produces a unique irreducible normal form. We follow the terminology
 of~\cite{krausCoherenceWellFoundednessTaming2020} to state our results formally.~\vc{check!}

### 2021-07-05 Pi+/paper/finite.tex
Revise section 4 and cut text

@@
 
-Before, the only way to decide the equality of two elements in a group was to compute and check them on the nose, but in
-a group presentation, this is reduced to a \emph{word problem}, that is, deciding whether one word -- a representative
-of the equivalence class of the group's elements, can be reduced to another word, using the group's relations. However,
-these equations are not directed, so it is not always possible to construct a well-behaved rewriting system. In general,
-the word problem for groups is known to be undecidable.\jk{Example?}\todo{Examples: empty relation, full relation, van
-  Kampen of $\pi_{1}$}
+Before, the only way to decide the equality of two elements in a group was to evaluate and check them on the nose, but
+in a group presentation, this is reduced to a \emph{word problem}, that is, deciding whether one word -- a
+representative of the equivalence class of the group's elements, can be reduced to another word, using the group's
+relations. However, these equations are not directed, so it is not always possible to construct a well-behaved rewriting
+system. In general, the word problem for groups is known to be undecidable.\jk{Example?}\todo{Examples: empty relation,
+  full relation, van Kampen of $\pi_{1}$}
 
@@
 This construction is called a Coxeter presentation of $\Sn$. Writing it formally, we encode the rules discussed above
-using a relation $\cox$ on $\List[\Fin[n]]$, and then take its congruence closure $\cox*$.
+using a relation $\cox$ on $\List[\Fin[n]]$ (\cref{def:cox}), and then take its congruence closure $\cox*$
+(\cref{def:coxstar}).
 
 \begin{definition}[$\cox : {\List[\Fin[n]]} \to {\List[\Fin[n]]} \to {\UU}$]
+  \label{def:cox}
   \begin{align*}
@@
 \end{definition}
+
+\begin{toappendix}
   \begin{definition}[$\cox* : {\List[\Fin[n]]} \to {\List[\Fin[n]]} \to {\UU}$]
+    \label{def:coxstar}
     \begin{align*}
@@
   \end{definition}
+\end{toappendix}
 
@@
         number of strands = 5,
+        width = 0.6cm,
+        height = 0.4cm,
+        border height = 0.1cm,
         thick]
@@
 sequence of transpositions and a single transposition afterwards. For example, for $n = 2$ and $k = 3$, it allows for
-the reduction $[6, 5, 4, 3, 2, 6] \longcox* [5, 6, 5, 4, 3, 2]$ (for the appropriate braid diagram,
-see~\cref{fig:longcox}).
+the reduction $[6, 5, 4, 3, 2, 6] \longcox* [5, 6, 5, 4, 3, 2]$ (see~\cref{fig:longcox} for the braid diagram).
+
 \begin{figure}
-  \[
-    \begin{tabular}{m{0.4\linewidth}m{0.1\linewidth}m{0.4\linewidth}}
-      \begin{center}
   \begin{tikzpicture}
@@
       number of strands = 6,
-            thick]
+      width = 0.6cm,
+      height = 0.4cm,
+      border height = 0.1cm,
+      thick] at (0, 0)
     {braid={s_5, s_4, s_3, s_2, s_1, s_5}};
-        \end{tikzpicture}
-      \end{center}
-       &
-      \(\xlongrightarrow[]{\longbraid}\)
-       &
-      \begin{center}
-        \begin{tikzpicture}
+    \node[font=\Huge] at (4, 1.3) {\(\rightarrow\)};
     \pic[local bounding box=my braid,braid/.cd,
       number of strands = 6,
-            thick]
+      width = 0.6cm,
+      height = 0.4cm,
+      border height = 0.1cm,
+      thick] at (5, 0)
     {braid={s_4, s_5, s_4, s_3, s_2, s_1}};
   \end{tikzpicture}
-      \end{center}
-    \end{tabular}
-  \]
   \caption{Braiding diagram for $\longcox$ with $n = 0$ and $k = 3$}
@@
 Despite the increased complexity of the generators, the rewriting system $(\List[\Fin[n]],\longcox*)$ has the properties
-we desire. It satisfies confluence, that is, the Church-Rosser (diamond) property, and it is strongly normalising,
-because it produces a unique irreducible normal form. We follow the terminology
-of~\cite{krausCoherenceWellFoundednessTaming2020} to state our results formally.~\vc{check!}
+we desire. It satisfies confluence, that is, the Church-Rosser (diamond) property, and it is strongly normalising, since
+it produces a unique normal form. We follow the terminology
+of~\cite*{huetConfluentReductionsAbstract1980a,krausCoherenceWellFoundednessTaming2020} to state our results
+formally.~\vc{check!}
 
@@
 The modified form of the Coxeter relations are unwieldy and difficult to prove properties about by induction. However,
-we can use the following observation to be able to use the more standard $\cox*$ relation.~\todo{Check Huet's paper for
-  this property.}
+we can use the following observation relating it to $\cox*$.~\todo{Check Huet's paper for this property.}
 
@@
 
-The type $\Sn$ is defined as the set-quotient of $\List[\Fin[n]]$ by $\cox*$,
+Finally, we define the type $\Sn$ as the set-quotient of $\List[\Fin[n]]$ by $\cox*$,
 
@@
 
-Note that this relation $\cox*$ is not $\hProp$-valued, that is, $w \cox* v$ is not a proposition because reductions are
-not unique. Because of this, the quotient $\Sn$ is not effective, that is, $\quotrel : w \cox* v \to q(w) \id q(v)$ is
-not an equivalence. We could truncate the relation to change this, but then the codomain of $\normf$ also needs to be
-truncated.
+% Note that this relation $\cox*$ is not prop-valued, that is, $w \cox* v$ is not a proposition because reductions need
+% not be unique. So, the quotient $\Sn$ is not effective, that is, $\quotrel : w \cox* v \to q(w) \id q(v)$ is not an
+% equivalence. We could instead define equivalence classes on $\List[\Fin[n]]$ to be those terms that have the same normal
+% form, using the $\normf$ function. Defining a new relation $(w \approx v) \defeq (\normf(w) \id \normf(v))$ which is
+% prop-valued, we could quotient $\List[\Fin[n]]$ by $\approx$. This is an equivalent definition for $\Sn$.
 
-We could also define equivalence classes on $\List[\Fin[n]]$ to be those terms that have the same normal form. Using the
-$\normf$ function, we could define a new relation $(w \approx v) \defeq (\normf(w) \id \normf(v))$ which is
-$\hProp$-valued, and quotient $\List[\Fin[n]]$ by $\approx$. Using the properties of the rewriting system we proved, we
-make a few observations.
+Note that this relation $\cox*$ is not prop-valued, that is, $w \cox* v$ is not a proposition because reductions need
+not be unique. So, the quotient $\Sn$ is not effective, that is, $\quotrel : w \cox* v \to q(w) \id q(v)$ is not an
+equivalence. Using the $\normf$ function, we could instead define a new relation $\approx$ to equate those terms that
+have the same normal form, $(w \approx v) \defeq (\normf(w) \id \normf(v))$. This relation is prop-valued, and we could
+quotient $\List[\Fin[n]]$ by $\approx$, obtaining an equivalent definition for $\Sn$.
 
@@
             \xrightarrow{r} {\List[\Fin[n]]}$ such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
-    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \).
+    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \), where $\quotinc$ is the inclusion into the quotient.
   \end{enumerate}
@@
           append, and inverse is given by list reversal.
-    \item $\Sn$ is equivalent to the group presented by generators $\Fin[n] + \Fin[n]$ with the relations given by the
-          normal closure of $\cox*$ extended to $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $\nabla_{A} : A + A
-            \to A$.
+    \item $\Sn$ is equivalent to the group presented by generators $\Fin[n]$ with the relations given by $\cox*$
+          extended to $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $\nabla_{A} : A + A \to A$.
   \end{enumerate}
@@
 
-We set out to solve the word problem for $\Sn$. To decide if two words in $\List[\Fin[n]]$ are equal, we simply have to
-compute their normal forms using $\normf$ and check if they're equal on the nose. If they do, they correspond to the
-same permutation.
+We set out to solve the word problem for $\Sn$. To decide if two words in $\List[\Fin[n]]$ are $\cox*$-equal, we simply
+have to compute their normal forms using $\normf$. They correspond to the same permutation if and only if these normal
+forms are equal.
 
@@
 representatives. We know that these are the unique normal forms in the set-quotient given by $\quotinc \comp \normf$,
-but now we will describe what these representatives exactly look like, using an encoding called Lehmer
+but now we will explicitly describe what these representatives look like, using an encoding called Lehmer
 codes~\cite{lehmerTeachingCombinatorialTricks1960}.
 
-There are many ways to represent permutations, e.g. inversions, or cycles, or matrices. Lehmer codes are known in
-Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960} where they are sometimes called "subexcedant sequences",
-or "factoriadics", which give the factorial number system. They are a particularly convenient way of representing
-permutations on a computer, partly because they are bitwise-optimal: for any $n : \Nat$, the type $\Lehmer[n]$ hass the cardinality $\fac{n}$, and has an easy to consttuct bijection with $\Aut[\Fin[n]]$.
-.
+Lehmer codes are known in Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960} where they are sometimes called
+"subexcedant sequences", or "factoriadics", which give the factorial number system. They are a convenient way of
+representing permutations on a computer, partly because they are
+bitwise-optimal~\cite{bergerTeachingOrdinalPatterns2019a}.
 
-Formally, we define $\Lehmer[n]$ to be an $n+1$-element tuple, where the position $k \leq n$ stores an element of $\Fin[k]$. Since
-the 0-th position is trivial, in practice, it is ignored, and the sequence starts at
-1~\cite{duboisTestsProofsCustom2018,vajnovszkiNewEulerMahonian2011}.\todo{There are more examples - listed
-  in~\cite{duboisTestsProofsCustom2018}}. We can then define $\Lehmer$ in two equivalent ways, by a simple recursion on
-$\Nat$, or as a type family generated by two constructors.
+Formally, we define $\Lehmer[n]$ to be an $n+1$-element tuple, where the position $k \leq n$ stores an element of
+$\Fin[k]$. Since the 0-th position is trivial, in practice it is ignored, and the sequence starts at
+1~\cite{duboisTestsProofsCustom2018,vajnovszkiNewEulerMahonian2011}.
 
 \begin{definition}[$\Lehmer : \Nat \to \UU$]
-  \begin{gather*}
-    \begin{aligned}
+  \begin{align*}
     \Lehmer[\zero]   & \defeq \Fin[\suc[\zero]]                     \\
     \Lehmer[\suc[n]] & \defeq \Fin[\suc[\suc[n]]] \times \Lehmer[n]
-    \end{aligned}
-    \qquad
-    \begin{aligned}
-      \lzero & : \Lehmer[\zero]                                                     \\
-      \lsuc  & : \forall n, r, (r \leq \suc[n]) \to \Lehmer[n] \to \Lehmer[\suc[n]]
-    \end{aligned}
-  \end{gather*}
+  \end{align*}
+  % \begin{aligned}
+  %   \lzero & : \Lehmer[\zero]                                                     \\
+  %   \lsuc  & : \forall n, r, (r \leq \suc[n]) \to \Lehmer[n] \to \Lehmer[\suc[n]]
+  % \end{aligned}
 \end{definition}
 
-\todo{Explain why the two definitions are equivalent?}
-
-For a permutation $\sigma : \Aut[\Fin[n]]$, for any element $i: \Fin[n]$, we can define the inversion count of $i$ as
+Given a permutation $\sigma : \Aut[\Fin[n]]$, for any element $i: \Fin[n]$, we can define the inversion count of $i$ as
 the number of smaller elements appearing after it in the permutation.
 
-\todo{typify}
 \begin{definition}
   Given a permutation $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by
-  \[ |\Set{j < i | \sigma(j) > \sigma(i)}|. \]
+  \[ \invcount{i} = \#\Set{j < i | \sigma(j) > \sigma(i)}. \]
 \end{definition}
 
-It turns out that from knowing just the inversion counts for all the elements, one can reconstruct the starting
-permutation. Also, observe that the inversion count for element $i$ is guaranteed to be smaller than $i$, thus fitting
-in the $i$-th place of a Lehmer code tuple. As an example, consider the following tabulated presentation of a
-permutation of $\Fin[5]$.
+From knowing just the inversion counts for all the elements, one can reconstruct the starting permutation. Also, observe
+that $\invcount{i} < i$, thus fitting in the $i$-th place of a Lehmer code tuple. As an example, consider the following
+tabulated presentation of a permutation of $\Fin[5]$.
 
@@
 
-The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 1 (because of
-$\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it comes
-last in the sequence), and for $\el{4}$ is 2 (because of $\el{3}$ and $\el{0}$). Thus, the Lehmer code for the permutation $\sigma$
-is the 5-tuple $l = (0, 1, 2, 0, 2)$.
+% The inversion count for $\el{0}$ is 0 (because there are no smaller elements at all), for $\el{1}$ is 1 (because of
+% $\el{0}$ appearing after), for $\el{2}$ is 2 (because of $\el{0}$ and $\el{1}$), for $\el{3}$ is 0 (because it comes
+% last in the sequence), and for $\el{4}$ is 2 (because of $\el{3}$ and $\el{0}$).
+
+The Lehmer code for the permutation $\sigma$ is then the 5-tuple
+\[
+  l = (\invcount{0}, \invcount{1}, \invcount{2}, \invcount{3}, \invcount{4}) = (0, 1, 2, 0, 2)
+\]
 
@@
 
-Writing formally, to turn a code into a permutation written in the Coxeter presentation, we define a function
-$\immersion$. As described above, the number $r$ at position $k$ in the tuple describes how many inversions the element
-$\el{k}$ has. Thus, we need to perform $r$ many adjacent transpositions to get to the desired position, which is given
-by $(\suc[n] - r) \downf r$.
+Writing formally, to turn a Lehmer code into a word in $\Sn$, we define a function $\immersion$. As described above, the
+number $r$ at position $k$ in the tuple describes how many inversions the element $\el{k}$ has. Thus, we need to perform
+$r$ many adjacent transpositions to get to the desired position, which is given by $(\suc[n] - r) \downf r$.
 
@@
 
-As an example, let us look at the code $c = (4, 3, 2, 1, 0)$. The list $\immersion(c)$ is then going to be $[0, 1, 0, 2,
-      1, 0, 3, 2, 1, 0, 4, 3, 2, 1, 0]$. It is a concatenation of a sequence of decreasing lists. Where could a reduction
-happen? First, it can't happen inside any of the decreasing components: $\longcancel$ requires repeating elements,
-$\longswap$ acts when a smaller number precedes a larger one, and $\longbraid$ has a non-monotone sequence on the left.
-This leaves the case of a reduction happening on a fragment that borders two subsequences. Again, $\longcancel$ requires
-two equal consecutive numbers, which have to be the last one in some decreasing sequence and the first one in the next
-one. But the first number in a sequence is always larger than every number in the sequence before it - which also shows
-why $\longswap$ cannot happen. The remaining case of $\longbraid$ is proven on similar grounds, since it requires the
-number appearing after the decreasing sequence to be equal to the first one in the sequence.
-
-Using this idea, we can show that the function $\immersion$ gives an equivalence betweeen $\Lehmer[n]$ and
-$\im{\normf}$.
+We can show that the function $\immersion$ gives an equivalence betweeen $\Lehmer[n]$ and $\im{\normf}$.
 
@@
 
-We also established an equivalence between $\im{\normf}$ and $\Sn$, which gives the following.
+\jk{Review:}
+To see that Lehmer codes under $\immersion$ are normal with respect to $\longcox*$, let us look at the example code $c =
+  (4, 3, 2, 1, 0)$. The list $\immersion(c)$ is then going to be $[0, 1, 0, 2, 1, 0, 3, 2, 1, 0, 4, 3, 2, 1, 0]$. It is a
+concatenation of five decreasing lists. Where could a reduction happen?
 
-\begin{proposition}~\label{prop:sn-im-lehmer-equiv}
+First, it can't happen inside any of the decreasing components: $\longcancel$ requires repeating elements, $\longswap$
+acts when a smaller number precedes a larger one, and $\longbraid$ has a non-monotone sequence on the left. This leaves
+the case of a reduction happening on a fragment that borders two subsequences. Again, $\longcancel$ requires two equal
+consecutive numbers, which have to be the last one in some decreasing sequence, and the first one in the next sequence.
+But the first number in a sequence is always larger than every number in the sequence before it -- which also shows why
+$\longswap$ cannot happen. The remaining case of $\longbraid$ follows similarly, since it requires the number appearing
+after the decreasing sequence to be equal to the first one in the sequence.
+
+% We also established an equivalence between $\im{\normf}$ and $\Sn$, which gives the following.
+\begin{corollary}
+  \label{prop:sn-im-lehmer-equiv}
   For all $n : \Nat$, \( \Sn \eqv \im{\normf} \eqv \Lehmer[n] \).
-\end{proposition}
+\end{corollary}
 
@@
 between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
-between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$.
+between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$.~\footnote{Note that the permutations are off-by-one, because
+we ignored the 0-position in the Lehmer code.}
 
@@
 
-\begin{proof}
+\begin{appendixproof}
   The first and second propositions follow from simply constructing the bijections using the decidable equality of
@@
   element $n$ at the position $n - k$, resulting in the element $n$ having inversion count $k$.
-\end{proof}
+\end{appendixproof}
 
@@
 
-By composing~\cref{prop:lehmer-aut-equiv} and~\cref{prop:sn-lehmer-fin-equiv}, we obtain the final equivalence.
+By composing~\cref{prop:lehmer-aut-equiv} and~\cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.
 

### 2021-07-05 Pi+/paper/reversible1.tex
intro

@@
 platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly''
-programming abstraction (e.g., \verb|qasm|). A mathematical model based on permutations of finite sets provides a richer
-algebraic model which we review in this section.
+programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical model based on
+permutations of finite sets provides a richer algebraic model which we review in this section.
 
@@
 
-A specialized language for programming with permutations on finite sets can be built using ideas going back to Mac
-Lane~\cite{XXX} that the commutative semiring (also known as the commutative rig) of natural numbers exactly
-characterizes permutations on finite sets. In programming parlance, as shown in Fig.~\ref{fig:pi-terms}, each primitive
-algebraic identity of commutative rigs becomes a constant in the programming language and the algebraic closure
-operators become composition operators in the programming language. Once equipped with a notion of values and types, we
-get the following syntactic presentation of a programming language for
+A specialized language for programming with permutations on finite sets can be built using ideas going back to Kelly,
+Laplaza, and Mac Lane~\cite{laplaza72,kelly74,KELLY197197} that the commutative semiring (also known as the commutative
+rig) of natural numbers exactly characterizes permutations on finite sets. In programming parlance, as shown in
+Fig.~\ref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes a constant in the programming
+language and the algebraic closure operators become composition operators in the programming language. Once equipped
+with a notion of values and types, we get the following syntactic presentation of a programming language for
 permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:

### 2021-07-05 Pi+/paper/finite.tex
Change the coxeter relation symbols

@@
 
-\begin{proposition}~\label{prop:coxlongcox}
-  $\cox*$ and $\longcox*$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff there is a $u$
-  such that $w \longcox* u$ and $v \longcox* u$.
+\begin{proposition}
+  \label{prop:coxlongcox}
+  The relations $\cox*$ and $\longcox*$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff
+  there is a $u$ such that $w \longcox* u$ and $v \longcox* u$.
 \end{proposition}

### 2021-07-06 Pi+/paper/reversible1.tex
unifying notation between sec 2 and sec 3; working on small changes to sec 3

@@
 outline a simple denotational semantics expressed in the metalanguage of sets and functions that collapses the level-2
-combinators. This semantics will be extended in the next two sections to the metalanguage of homotopy type theory to
-properly model the level-2 combinators.
+combinators. This semantics will be extended in the next two sections to the HoTT metalanguage to properly model the
+level-2 combinators.
 
@@
 \begin{array}{rcl}
-  \denot{\zerot}     & = & \emptyset                  \\
-  \denot{\onet}      & = & \{ \Acon{tt} \}            \\
-  \denot{A + B}      & = & \denot{A} \uplus \denot{B} \\
+  \denot{\zerot}     & = & \bot                  \\
+  \denot{\onet}      & = & \top \\
+  \denot{A + B}      & = & \denot{A} \sqcup \denot{B} \\
   \denot{A \times B} & = & \denot{A} \times \denot{B}
@@
 
-\noindent where $\emptyset$ is the empty set, $\uplus$ is the disjoint union of sets, and $\times$ is the cartesian product of
-sets. The denotation of a combinator $c : A \isoone B$ is a (bijective) function mapping $\denot{A}$ to $\denot{B}$:
+\noindent where $\bot$ is the empty set, $\top$ is a singleton set, $\sqcup$ is the disjoint union of sets, and $\times$
+is the cartesian product of sets. The denotation of a combinator $c : A \isoone B$ is a (bijective) function mapping
+$\denot{A}$ to $\denot{B}$:
 
@@
 \begin{array}[t]{rlcl}
-  \denot{\identlp}     &      (\inlv{v})           & = & v                                   \\
-  \denot{\identrp}     &      v                    & = & \inlv{v}                            \\
-  \denot{\swapp}       &     (\inlv{v})           & = & \inrv{v}                            \\
-  \denot{\swapp}       &      (\inrv{v})           & = & \inlv{v}                            \\
-  \denot{\assoclp}     &      (\inlv{v})           & = & \inlv{(\inlv{v})}                   \\
-  \denot{\assoclp}     &      (\inrv{(\inlv{v})})  & = & \inlv{(\inrv{v})}                   \\
-  \denot{\assoclp}     &      (\inrv{(\inrv{v})})  & = & \inrv{v}                            \\
-  \denot{\assocrp}     &      (\inlv{(\inlv{v})})  & = & \inlv{v}                            \\
-  \denot{\assocrp}     &      (\inlv{(\inrv{v})})  & = & \inrv{(\inlv{v})}                   \\
-  \denot{\assocrp}     &      (\inrv{v})           & = & \inrv{(\inrv{v})}
+  \denot{\identlp}     &      (\inl{v})           & = & v                                   \\
+  \denot{\identrp}     &      v                    & = & \inl{v}                            \\
+  \denot{\swapp}       &     (\inl{v})           & = & \inr{v}                            \\
+  \denot{\swapp}       &      (\inr{v})           & = & \inl{v}                            \\
+  \denot{\assoclp}     &      (\inl{v})           & = & \inl{(\inl{v})}                   \\
+  \denot{\assoclp}     &      (\inr{(\inl{v})})  & = & \inl{(\inr{v})}                   \\
+  \denot{\assoclp}     &      (\inr{(\inr{v})})  & = & \inr{v}                            \\
+  \denot{\assocrp}     &      (\inl{(\inl{v})})  & = & \inl{v}                            \\
+  \denot{\assocrp}     &      (\inl{(\inr{v})})  & = & \inr{(\inl{v})}                   \\
+  \denot{\assocrp}     &      (\inr{v})           & = & \inr{(\inr{v})}
 \end{array}$}}
@@
 \begin{array}[t]{rlcl}
-  \denot{\identlt}      &      (\Acon{tt} , v)      & = & v                                   \\
-  \denot{\identrt}      &      v                    & = & (\Acon{tt} , v)                     \\
+  \denot{\identlt}      &      (\ttt , v)      & = & v                                   \\
+  \denot{\identrt}      &      v                    & = & (\ttt , v)                     \\
   \denot{\swapt}        &      (v_1 , v_2)          & = & (v_2 , v_1)                         \\
@@
   \denot{\assocrt}     &      ((v_1 , v_2) , v_3)  & = & (v_1 , (v_2 , v_3))                 \\
-  \denot{\dist}       &      (\inlv{v_1} , v_3)   & = & \inlv{(v_1 , v_3)}                  \\
-  \denot{\dist}        &      (\inrv{v_2 , v_3})   & = & \inrv{(v_2 , v_3)}                  \\
-  \denot{\factor}       &      (\inlv{(v_1 , v_3)}) & = & (\inlv{v_1} , v_3)                  \\
-  \denot{\factor}      &      (\inrv{(v_2 , v_3)}) & = & (\inrv{v_2} , v_3)
+  \denot{\dist}       &      (\inl{v_1} , v_3)   & = & \inl{(v_1 , v_3)}                  \\
+  \denot{\dist}        &      (\inr{v_2 , v_3})   & = & \inr{(v_2 , v_3)}                  \\
+  \denot{\factor}       &      (\inl{(v_1 , v_3)}) & = & (\inl{v_1} , v_3)                  \\
+  \denot{\factor}      &      (\inr{(v_2 , v_3)}) & = & (\inr{v_2} , v_3)
 \end{array}$}}
@@
   \denot{(c_1 \fatsemi c_2)} & v                    & = & (\denot{c_2} \circ \denot{c_1}) v   \\
-  \denot{(c_1 \oplus c_2)}   & (\inlv{v})           & = & \inlv{(\denot{c_1}~v)}              \\
-  \denot{(c_1 \oplus c_2)}   & (\inrv{v})           & = & \inrv{(\denot{c_2}~v)}              \\
+  \denot{(c_1 \oplus c_2)}   & (\inl{v})           & = & \inl{(\denot{c_1}~v)}              \\
+  \denot{(c_1 \oplus c_2)}   & (\inr{v})           & = & \inr{(\denot{c_2}~v)}              \\
   \denot{(c_1 \otimes c_2)}  & (v_1 , v_2)          & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
@@
   the other side. Thus $! \assoclp$ is $\assocrp$ and $! \swapp$ is $\swapp$ itself. Then we have that $\denot{c}$ and
-  $\denot{!c}$ is an equivalence. For the level-2 combinator \Afun{idr◎l}, we check
+  $\denot{!c}$ form an equivalence. For the level-2 combinator \Afun{idr◎l}, we check
   $\denot{\AgdaBound{c}~\AgdaOperator{\AgdaInductiveConstructor{◎}}~\AgdaInductiveConstructor{id⟷₁}}

### 2021-07-06 Pi+/paper/ufin.tex
messing with packages; missing font problem first reported around p.12 grep code *.tex
unifying notation between sec 2 and sec 3; working on small changes to sec 3

@@
 
-In the previous section, we gave a sound denotational semantics for $\PiLang$ using the groupoid of finite sets and
-bijections $\BFin$. Types are interpreted as objects in $\BFin$, and 1-combinators are interpreted as isomorphisms in
-$\BFin$. We also showed that if there is a 2-combinator between two 1-combinators, then their interpretations are
-extensionally equal in $\BFin$.
+The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are sets and
+whose (iso)morphisms as bijections. The semantics interprets types as objects in $\BFin$, 1-combinators as isomorphisms
+in $\BFin$, and for every pair of 1-combinators related by a 2-combinator, their interpretations are extensionally equal
+in $\BFin$. This groupoid is strict: two equal permutations are extensionally identified, without an explicit witness
+for the equality. In order to establish completeness, we want to quote back from the semantics to the
+syntax. Specifically, given a morphism in $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two
+extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator witnessing the equality of the quoted
+1-combinators. Our first step is therefore to \emph{weaken} this groupoid $\BFin$.
 
-However, we want to go a step further, and claim completeness for this semantics, that is, we want to quote back from
-the semantics to the syntax. Given a morphism in $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two
-equal morphisms in $\BFin$, we want to produce a 2-combinator witnessing the equality of the quoted 1-combinators.
-
-The groupoid $\BFin$ that we presented is strict, that is, two equal permutations are extensionally identified, without
-an explicit witness for the equality. Our first step is to weaken this groupoid $\BFin$. In a weak category, equational
-axioms of the category are replaced by higher invertible cells. For example, there are higher invertible cells for the
-left and right units of composition, and associativity. The higher cells come with their own coherence data as well,
-such as vertical composition, and horizontal composition, or whiskering.\todo{define 2-category in appendix.}
-
-HoTT provides a rich internal language for working with weak higher groupoids, using the ``Types are weak
-$\infty$-groupoids'' correspondence. Using the identity type for morphisms, a 1-groupoid in HoTT has points for 0-cells,
-1-paths for 1-cells, and 2-paths for 2-cells, but there is at most one 2-cell between compatible 1-cells.~\footnote{This
-is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that
-is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type $\UFin$, which is a 1-groupoid. In this
-section, we develop the tools necessary to construct $\UFin$, starting with a quick review of HoTT.
+In a weak category, equational axioms of the category are replaced by higher invertible cells. For example, there are
+higher invertible cells for the left and right units of composition, and associativity. The higher cells come with their
+own coherence data as well, such as vertical composition, and horizontal composition, or whiskering.\todo{define
+  2-category in appendix.} HoTT provides a rich internal language for working with weak higher groupoids, using the
+``Types are weak $\infty$-groupoids'' correspondence. Using the identity type for morphisms, a 1-groupoid in HoTT has
+points for 0-cells, 1-paths for 1-cells, and 2-paths for 2-cells, but there is at most one 2-cell between compatible
+1-cells.~\footnote{This is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the
+  hom-categories are posets, that is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type
+$\UFin$, which is a 1-groupoid. In this section, we develop the tools necessary to construct $\UFin$, starting with a
+quick review of HoTT.
 
@@
   This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
-  level~\cite{DBLP:conf/esop/CaretteS16}.
+  level~\cite{carette2016}.
   As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$

### 2021-07-06 Pi+/paper/ufin.tex
working on ufin

@@
 
-The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are sets and
-whose (iso)morphisms as bijections. The semantics interprets types as objects in $\BFin$, 1-combinators as isomorphisms
-in $\BFin$, and for every pair of 1-combinators related by a 2-combinator, their interpretations are extensionally equal
-in $\BFin$. This groupoid is strict: two equal permutations are extensionally identified, without an explicit witness
-for the equality. In order to establish completeness, we want to quote back from the semantics to the
-syntax. Specifically, given a morphism in $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two
-extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator witnessing the equality of the quoted
-1-combinators. Our first step is therefore to \emph{weaken} this groupoid $\BFin$.
+The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are finite sets
+and whose (iso)morphisms as bijections between finite sets. The semantics interprets types as objects in~$\BFin$,
+1-combinators as isomorphisms in~$\BFin$, and for every pair of 1-combinators related by a 2-combinator, their
+interpretations are extensionally equal in~$\BFin$. In other words, this groupoid is strict: two equal permutations are
+extensionally identified, without an explicit witness for the equality. In order to establish completeness, we want to
+quote back from the semantics to the syntax. Specifically, given a morphism in $\BFin$, we want to produce a
+1-combinator in $\PiLang$, and given two extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator
+witnessing the equality of the quoted 1-combinators. Our first step is therefore to \emph{weaken} this groupoid $\BFin$
+by replacing the equational axioms of the category by higher invertible cells. In a weak category, the identity and
+associativity equational conditions are replaced by higher invertible cells for the left and right units of composition
+and for associativity; these higher cells come with their own coherence data as well, such as vertical composition, and
+horizontal composition, or whiskering as detailed in Appendix~\ref{app:twocat}.
 
-In a weak category, equational axioms of the category are replaced by higher invertible cells. For example, there are
-higher invertible cells for the left and right units of composition, and associativity. The higher cells come with their
-own coherence data as well, such as vertical composition, and horizontal composition, or whiskering.\todo{define
-  2-category in appendix.} HoTT provides a rich internal language for working with weak higher groupoids, using the
-``Types are weak $\infty$-groupoids'' correspondence. Using the identity type for morphisms, a 1-groupoid in HoTT has
-points for 0-cells, 1-paths for 1-cells, and 2-paths for 2-cells, but there is at most one 2-cell between compatible
-1-cells.~\footnote{This is a locally-strict $(2,0)$-category, since every cell above 0 is invertible, and the
-  hom-categories are posets, that is, truth-value enriched.} The groupoid $\BFin$ can be written in HoTT as a type
-$\UFin$, which is a 1-groupoid. In this section, we develop the tools necessary to construct $\UFin$, starting with a
-quick review of HoTT.
+HoTT provides a rich internal language for working with weak higher groupoids, using the ``types are weak
+$\infty$-groupoids'' correspondence. In this section, we show that the weakened version of the groupoid $\BFin$ can be
+written in HoTT as a type $\UFin$, which is a 1-groupoid that has points for 0-cells, 1-paths for 1-cells, 2-paths for
+2-cells with at most one 2-cell between compatible 1-cells.
+
+%% In this section, we develop the tools necessary to construct $\UFin$, starting with a quick review of HoTT.
+%% Using the identity type for morphisms, a 1-groupoid in HoTT
+
+IS THAT FOOTNOTE NECESSARY ??? We should avoid using footnotes as much as possible.\footnote{This is a locally-strict
+  $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that is, truth-value
+  enriched.}
 

### 2021-07-06 Pi+/paper/ufin.tex
intro to ufin

@@
 The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are finite sets
-and whose (iso)morphisms as bijections between finite sets. The semantics interprets types as objects in~$\BFin$,
+and whose morphisms as bijections between finite sets. The semantics interprets types as objects in~$\BFin$,
 1-combinators as isomorphisms in~$\BFin$, and for every pair of 1-combinators related by a 2-combinator, their
@@
 1-combinator in $\PiLang$, and given two extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator
-witnessing the equality of the quoted 1-combinators. Our first step is therefore to \emph{weaken} this groupoid $\BFin$
-by replacing the equational axioms of the category by higher invertible cells. In a weak category, the identity and
+witnessing the equality of the quoted 1-combinators.
+
+Our first step is therefore to expose the implicit equalities by \emph{weakening} the groupoid $\BFin$, i.e., by
+replacing the equational axioms of the category by higher invertible cells. In a weak category, the identity and
 associativity equational conditions are replaced by higher invertible cells for the left and right units of composition
 and for associativity; these higher cells come with their own coherence data as well, such as vertical composition, and
-horizontal composition, or whiskering as detailed in Appendix~\ref{app:twocat}.
+horizontal composition, or whiskering as detailed in Appendix~\ref{app:twocat}. However, although weakening the groupoid
+exposes some implicit equalities, it does not, by itself, give us a handle to characterize such equalities. One elegant
+way to acquire such an ability is to ensure that the groupoid is ``univalent,'' as this provides a way to characterize
+the semantic equivalences using the elements of the identity type, i.e., using objects whose equality is equipped with
+an induction principle. We will therefore work within a HoTT metatheory which provides an ambient univalent universe of
+types and introduce the novel idea of a \emph{univalent subuniverse} to carve out $\UFin$: a weakened, univalent,
+version of the groupoid $\BFin$.
 
-HoTT provides a rich internal language for working with weak higher groupoids, using the ``types are weak
-$\infty$-groupoids'' correspondence. In this section, we show that the weakened version of the groupoid $\BFin$ can be
-written in HoTT as a type $\UFin$, which is a 1-groupoid that has points for 0-cells, 1-paths for 1-cells, 2-paths for
-2-cells with at most one 2-cell between compatible 1-cells.
+% HoTT provides a rich internal language for working with weak higher groupoids, using the ``types are weak
+% $\infty$-groupoids'' correspondence. In this section, we show that the weakened version of the groupoid $\BFin$ can be
+% written in HoTT as a type $\UFin$, which is a 1-groupoid that has points for 0-cells, 1-paths for 1-cells, 2-paths for
+% 2-cells with at most one 2-cell between compatible 1-cells.
 

### 2021-07-06 Pi+/paper/ufin.tex
intro to ufin

@@
 
-The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are finite sets
-and whose morphisms as bijections between finite sets. The semantics interprets types as objects in~$\BFin$,
-1-combinators as isomorphisms in~$\BFin$, and for every pair of 1-combinators related by a 2-combinator, their
-interpretations are extensionally equal in~$\BFin$. In other words, this groupoid is strict: two equal permutations are
-extensionally identified, without an explicit witness for the equality. In order to establish completeness, we want to
-quote back from the semantics to the syntax. Specifically, given a morphism in $\BFin$, we want to produce a
-1-combinator in $\PiLang$, and given two extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator
-witnessing the equality of the quoted 1-combinators.
+The denotational semantics for $\PiLang$ in the previous section forms a commutative rig groupoid $\BFin$ whose objects
+are finite sets, whose morphisms as bijections between finite sets, and whose monoidal structures are induced by the
+properties $\sqcup$ and $\times$. The semantics interprets types as objects in~$\BFin$, 1-combinators as isomorphisms
+in~$\BFin$, and for every pair of 1-combinators related by a 2-combinator, their interpretations are extensionally equal
+in~$\BFin$. In other words, this groupoid is strict: two equal permutations are extensionally identified, without an
+explicit witness for the equality. In order to establish completeness, we want to quote back from the semantics to the
+syntax. Specifically, given a morphism in $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two
+extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator witnessing the equality of the quoted
+1-combinators.
 
@@
 types and introduce the novel idea of a \emph{univalent subuniverse} to carve out $\UFin$: a weakened, univalent,
-version of the groupoid $\BFin$.
+version of the groupoid $\BFin$ which is then extended with the appropriate commutative rig structure. The main
+technical result of this section that is exploited in the next section is that, for each finite type $T$, we have
+transferred the structure of the space of equivalences $T \eqv T$ to the structure of paths in $\UFin$. Since the space
+equivalences forms the automorphism group of $T$, we can now use the tools of computational group theory together with
+the induction principle of paths and term rewriting theory to describe the paths in $\UFin$.
 

### 2021-07-06 Pi+/paper/finite.tex
What is novel in Sn

@@
 In~\cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
-$n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group.This is
+$n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group. This is
 also known to be the finite symmetric group $\Sn[n]$. In the following sections, we will describe this group

### 2021-07-06 Pi+/paper/finite.tex
Fix diagrams

@@
       {braid={ s_1, s_1}};
-      \node[font=\Huge] at (2, 0.7) {\(\rightarrow\)};
+      \node[font=\Huge] at (1.75, 0.7) {\(\rightarrow\)};
       \pic[local bounding box=my braid,braid/.cd,
@@
         thick] at (0, 0)
-      {braid={ 1, s_1}};
-      \node[] at (1, 0.7) {\(\dots\)};
+      {braid={ s_1, 1}};
+      \node[] at (1.1, 0.7) {\(\dots\)};
       \pic[local bounding box=my braid,braid/.cd,
@@
         thick] at (1.5, 0)
-      {braid={ s_1, 1}};
+      {braid={ 1, s_1 }};
       \node[font=\Huge] at (3, 0.7) {\(\rightarrow\)};
@@
       {braid={ 1, s_1}};
-      \node[] at (5.2, 0.7) {\(\dots\)};
+      \node[] at (5.1, 0.7) {\(\dots\)};
       \pic[local bounding box=my braid,braid/.cd,
@@
       {braid={ s_2, s_1, s_2}};
-      \node[font=\Huge] at (2, 0.7) {\(\rightarrow\)};
+      \node[font=\Huge] at (2.15, 0.7) {\(\rightarrow\)};
       \pic[local bounding box=my braid,braid/.cd,

### 2021-07-06 Pi+/paper/finite.tex
Fix second braid diagram

@@
         thick]
-      {braid={s_4, s_3, s_2, s_1}};
+      {braid={s_1, s_2, s_3, s_4}};
     \end{tikzpicture}
@@
 visualised in the same way as before. The remaining constructor uses the helper function to exchange the order of a long
-sequence of transpositions and a single transposition afterwards. For example, for $n = 2$ and $k = 3$, it allows for
-the reduction $[6, 5, 4, 3, 2, 6] \longcox* [5, 6, 5, 4, 3, 2]$ (see~\cref{fig:longcox} for the braid diagram).
+sequence of transpositions and a single transposition afterwards. For example, for $n = 0$ and $k = 3$, it allows for
+the reduction $[4, 3, 2, 1, 0, 4] \longcox* [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox}).
 
@@
       thick] at (0, 0)
-    {braid={s_5, s_4, s_3, s_2, s_1, s_5}};
+    {braid={s_5, s_1, s_2, s_3, s_4, s_5}};
     \node[font=\Huge] at (4, 1.3) {\(\rightarrow\)};
@@
       thick] at (5, 0)
-    {braid={s_4, s_5, s_4, s_3, s_2, s_1}};
+    {braid={s_1, s_2, s_3, s_4, s_5, s_4}};
   \end{tikzpicture}

### 2021-07-06 Pi+/paper/reversible1.tex
Revise section 3

@@
 
+Using categorical language, the denotational semantics for $\PiLang$ is using the category of finite sets and functions
+$\SetFin$. However, we only use the bijective functions for the semantics, which means, we use the groupoid $\BFin =
+\term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and finite
+products $(\unit, \times)$, which we use to interpret our types. In $\BFin$, the coproducts and products restrict to
+additive and multiplicative symmetric monoidal structures, respectively, making $\BFin$ a symmetric rig groupoid.
+
 %%% Local Variables:

### 2021-07-06 Pi+/paper/ufin.tex
Revise section 3

--- 2021-07-06 Pi+/paper/ufin.tex
@@
 
-The denotational semantics for $\PiLang$ in the previous section forms a groupoid $\BFin$ whose objects are finite sets
-and whose morphisms as bijections between finite sets. The semantics interprets types as objects in~$\BFin$,
-1-combinators as isomorphisms in~$\BFin$, and for every pair of 1-combinators related by a 2-combinator, their
-interpretations are extensionally equal in~$\BFin$. In other words, this groupoid is strict: two equal permutations are
-extensionally identified, without an explicit witness for the equality. In order to establish completeness, we want to
-quote back from the semantics to the syntax. Specifically, given a morphism in $\BFin$, we want to produce a
-1-combinator in $\PiLang$, and given two extensionally equal morphisms in $\BFin$, we want to produce a 2-combinator
-witnessing the equality of the quoted 1-combinators.
+% First - recap the previous section
+% Second - say what was wrong
+To recap, the denotational semantics of $\PiLang$ interprets types as objects in $\BFin$, 1-combinators as isomorphisms,
+and we prove that for every pair of 1-combinators related by a 2-combinator, their interpretations in~$\BFin$ are equal.
+The groupoid $\BFin$ is strict, since the collection of isomorphisms is a set, that is, a discrete category. There is no
+explicit witness for the equality of two isomorphisms, since we can decide by evaluating two bijections whether they're
+equal. To be able to establish completeness for $\PiLang$, we want a witness for this equality, so that we can quote
+back to the syntax and produce a 2-combinator witnessing the equality of the corresponding 1-combinators.
 
-Our first step is therefore to expose the implicit equalities by \emph{weakening} the groupoid $\BFin$, i.e., by
-replacing the equational axioms of the category by higher invertible cells. In a weak category, the identity and
-associativity equational conditions are replaced by higher invertible cells for the left and right units of composition
-and for associativity; these higher cells come with their own coherence data as well, such as vertical composition, and
-horizontal composition, or whiskering as detailed in Appendix~\ref{app:twocat}. However, although weakening the groupoid
-exposes some implicit equalities, it does not, by itself, give us a handle to characterize such equalities. One elegant
-way to acquire such an ability is to ensure that the groupoid is ``univalent,'' as this provides a way to characterize
-the semantic equivalences using the elements of the identity type, i.e., using objects whose equality is equipped with
-an induction principle. We will therefore work within a HoTT metatheory which provides an ambient univalent universe of
-types and introduce the novel idea of a \emph{univalent subuniverse} to carve out $\UFin$: a weakened, univalent,
-version of the groupoid $\BFin$.
+% The denotational semantics for $\PiLang$ in the previous section uses the symmetric rig groupoid structure of $\BFin$,
+% whose objects are finite sets, whose morphisms as bijections between finite sets, and whose symmetric monoidal
+% structures are induced by the operations $\sqcup$ and $\times$. The
+
+% two equal permutations are extensionally identified, without an explicit witness for the equality. In order to
+% establish completeness, we want to quote back from the semantics to the syntax. Specifically, given a morphism in
+% $\BFin$, we want to produce a 1-combinator in $\PiLang$, and given two extensionally equal morphisms in $\BFin$, we
+% want to produce a 2-combinator witnessing the equality of the quoted 1-combinators.
+
+% Third - say how are we fixing it
+First, we observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
+homotopies -- we \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells for
+them. We work in Homotopy Type Theory (Univalent Foundations) -- it provides a proof-relevant, constructive metatheory
+to get a handle on these equalities, and, it provides a rich internal language for describing weak groupoids, using the
+"types are weak $\infty$-groupoids" correspondence.
+
+% To get a handle on these equalities, we will work in a proof-relevant metatheory, that is, HoTT, and then we will
+% be able to write functions out of them back into the syntax.
+
+% HoTT provides a rich internal language for describing weak groupoids.
+
+% The idea is that we move from (discrete) sets to proof-relevant $\hSet$s, where the type of equalities between two
+% points in a set is a proposition, that is, a prop-enriched groupoid.
+
+% Our first step is therefore to expose the implicit equalities by \emph{weakening} the groupoid $\BFin$, i.e., by
+% replacing the equational axioms of the category by higher invertible cells. In a weak category, the identity and
+% associativity equational conditions are replaced by higher cells for the left and right units of composition and for
+% associativity; these higher cells come with their own coherence data as well, such as vertical composition, and
+% horizontal composition, or whiskering as detailed in appendix~\ref{app:twocat}.
+
+% Fourth - expain the tool we use for fixing it
+Every type in HoTT is a weak $\infty$-groupoid whose points are the terms of the type, and the (iterated) identity type
+gives the (higher) morphisms. The groupoid we're interested in has types as points, type equivalences for 1-cells, and
+higher homotopies for higher cells. This is the groupoid structure for the universe type $\UU$, since the identity type
+on types can be characterised as type equivalences (by \emph{univalence}). But, we only want to carve out a
+\emph{subuniverse} of \emph{finite types}, still satisfying univalence, to get the groupoid structure. In this section,
+we formally define \emph{univalent subuniverses}, and proceed to construct the particular instance for finite types,
+$\UFin$ (\cref{def:ufin}).
+
+% The way to describe such groupoid in HoTT is to construct a type which has the types themselves as points, i.e. a
+% universe. Then, equivalences between those types can be characterized using elements of the equality type. For this to
+% work, the universe in question has be a \emph{univalent subuniverse}.
+
+% In HoTT, every type carries a groupoid structure, with terms for points, and the (iterated) identity type for (higher)
+% morphisms. We're interested in the groupoid structure of the universe, which has types for points, equivalences for
+% identities between them (by univalence), and higher homotopies for higher identitifications. We want a similar
+% universe, whose only points are the finite types, that is, a subuniverse, and, equalities between them should
+% correspond to equivalences of finite types -- the subuniverse should be univalent. To construct a univalent
+% subuniverse starting from a univalent universe, we will introduce the concept of univalent fibrations.
+
+% However, although weakening the groupoid exposes some implicit equalities, it does not, by itself, give us a handle to
+% characterize such equalities. One elegant way to acquire such an ability is to ensure that the groupoid is
+% ``univalent,'' as this provides a way to characterize the semantic equivalences using the elements of the identity
+% type, i.e., using objects whose equality is equipped with an induction principle. We will therefore work within a HoTT
+% metatheory which provides an ambient univalent universe of types and introduce the novel idea of a \emph{univalent
+% subuniverse} to carve out $\UFin$: a weakened, univalent, version of the groupoid $\BFin$ which is then extended with
+% the appropriate commutative rig structure. The main technical result of this section that is exploited in the next
+% section is that, for each finite type $T$, we have transferred the structure of the space of equivalences $T \eqv T$
+% to the structure of paths in $\UFin$. Since the space equivalences forms the automorphism group of $T$, we can now use
+% the tools of computational group theory together with the induction principle of paths and term rewriting theory to
+% describe the paths in $\UFin$.
 
@@
 
-%% In this section, we develop the tools necessary to construct $\UFin$, starting with a quick review of HoTT.
-%% Using the identity type for morphisms, a 1-groupoid in HoTT
+%% In this section, we develop the tools necessary to construct $\UFin$, starting with a quick review of HoTT. % Using
+%the identity type for morphisms, a 1-groupoid in HoTT
 
-IS THAT FOOTNOTE NECESSARY ??? We should avoid using footnotes as much as possible.\footnote{This is a locally-strict
-  $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that is, truth-value
-  enriched.}
+% IS THAT FOOTNOTE NECESSARY ??? We should avoid using footnotes as much as possible.\footnote{This is a locally-strict
+%   $(2,0)$-category, since every cell above 0 is invertible, and the hom-categories are posets, that is, truth-value
+%   enriched.}
 
@@
 
-We review some basic concepts and notation from Homotopy Type Theory that we use in the appendix.~\footnote{We use the
-language of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use intensional
-Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for propositional
-truncation and set quotients. All arguments will hold in a Cubical Type Theory as well, such
-as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.}
-For more details about HoTT, we refer the reader to the HoTT
-book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.
+We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
+intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
+propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type Theory as well, such
+as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.
+In the appendix, we review some basic concepts and notation from the HoTT book that we use. For more details, we refer
+the reader to the book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.
 
-To recap, types in HoTT have a weak groupoid structure given by the identity type, where points are given by terms of
-the type, and higher morphisms are given by the iterated identity type. Functions between types are functors between
-groupoids, and type families, or functions to the universe, are indexed families of groupoids. A type family $P : A \to
-\UU$ comes equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$ which has a lifting operation giving it the
-structure of a fibration (see the appendix for details).
+% To recap, types in HoTT have a weak groupoid structure given by the identity type, where points are given by terms of
+% the type, and higher morphisms are given by the iterated identity type. Functions between types are functors between
+% groupoids, and type families, or functions to the universe, are indexed families of groupoids.
 
@@
 
-The $\term{transport}$ operation lifts paths in the base space to maps between the fibers. Using the groupoid structure
-of $A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name
+Functions between types are functors between groupoids, and type families, or functions to the universe, are indexed
+families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$
+which has a lifting operation giving it the structure of a fibration (see the appendix for details). The
+$\term{transport}$ operation lifts paths in the base space to maps between the fibers. Using the groupoid structure of
+$A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name
 this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
@@
 The type families (or fibrations) we're interested in are the ones where paths in the base space completely determine
-the equivalences in the fibers -- these are called univalent fibrations.
+the equivalences in the fibers -- these are called univalent
+fibrations~\cite*{kapulkinUnivalenceSimplicialSets2018,kapulkinSimplicialModelUnivalent2021,christensenCharacterizationUnivalentFibrations2015}.
 
@@
   \end{itemize}
-  If $\El$ is univalent, $U$ is a univalent universe.
-\end{definition}
-
-\begin{definition}[Subuniverse]
-  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition
-  for every $X$. A subuniverse generated by $P$ is given by its total space $U \defeq \dsum{X:\UU}{P(X)}$ and $\El
-  \defeq \fst$.
+  If $\El$ is univalent, we call $(U,\El)$ a \emph{univalent} universe.
 \end{definition}
@@
 \begin{proposition}[Univalent Subuniverse]
-  Subuniverses generated by predicates are univalent.
+  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition
+  for every $X$. The fibration ${\fst : \dsum{X:\UU}{P(X)} \to \UU}$ is univalent. This generates a univalent
+  subuniverse
+  \[
+    \UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)
+  \]
 \end{proposition}
@@
 
-The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is subtle. We
-define a type to be (Bishop) finite if it is equal to some finite set, and we setup the predicate $\isFin$ to do so.
-\todo{Cite Brent Yorgey's thesis?}
+The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is
+subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness -- a type is finite if it is
+merely equivalent to a finite set (\cref{def:finite-set}), and we define the predicate $\isFin$ (\cref{def:isfin}) to do
+so.
 
 \begin{definition}[$\Fin$]
+  \label{def:finite-set}
   The type family $\Fin : \Nat \to \UU$ is the type of finite sets indexed by their cardinality. It is defined
@@
     \end{aligned}
-    \qquad
+    \qquad\qquad
     \begin{aligned}
@@
   \end{gather*}
-  Note that $\Fin[n]$ is a $\hSet$, and we use the definitions interchangeably.
+  Note that $\Fin[n]$ is a set, and we use both definitions interchangeably.
 \end{definition}
@@
 \begin{definition}[$\isFin$]
+  \label{def:isfin}
   We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n$.
@@
 
+Since $\isFin[X]$ is a predicate on the universe $\UU$, we easily get our univalent subuniverse $\UFin$.
+
 \begin{definition}
+  \label{def:ufin}
   The univalent subuniverse of \emph{all finite types} is given by
@@
 
-We also consider a different way of thinking about $\UFin$, using the subuniverse $\BAut[T]$.
+While $\UFin$ has \emph{all} the finite types, we're also interested in constructing a subuniverse of finite types of a
+specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$.
 
@@
   \]
-  We write $T_0 \defeq (T, \trunc{\refl_{T}})$ for the image of the inclusion of $T$ in $\BAut[T]$
+  We write $T_0 \defeq (T, \trunc{\refl_{T}})$ for the image of the inclusion of $T$ in $\BAut[T]$.
 \end{definition}
 
-Using $\BAut$, we can talk about types that have a specified cardinality, for example, the subuniverse of 2-element sets
-is given by $\BAut[\Bool]$. This has been used to construct the real projective spaces in
+Using $\BAut$, we can talk about types that are equivalent to a finite set of specified cardinality, for example, the
+subuniverse of 2-element sets is given by $\BAut[\Bool]$. This has been used to construct the real projective spaces in
 HoTT~\cite{buchholtzRealProjectiveSpaces2017}, and also to give the denotational semantics for a 1-bit reversible
-language~\cite{caretteReversibleProgramsUnivalent2018}.
+programming language~\cite{caretteReversibleProgramsUnivalent2018}.
 
@@
   Note that, $\UFin$ can be equivalently seen as the collection of all types of finite cardinality, that is, $\UFin \eqv
-    \displaystyle\sqcup_{n:\Nat} \UFin[n]$.
+    \dsum{n:\Nat}{\UFin[n]}$.
 \end{definition}
@@
 type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$ is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
-the automorphism group of $T$.
+the group of automorphisms of $T$.
 
@@
 \begin{corollary}
-  $\UFin$ is a pointed, connected, 1-groupoid. For every $n:\Nat$, \( \loopspace[\UFin[n],F_{n}] \eqv \Aut[\Fin[n]] \).
+  $\UFin[n]$ is a pointed, connected, 1-groupoid for every $n:\Nat$, and \( \loopspace[\UFin[n],F_{n}] \eqv
+  \Aut[\Fin[n]] \). $\UFin$ is a 1-groupoid with connected components for every $n:\Nat$.
 \end{corollary}
 
-The loopspace of a pointed type automatically has the structure of a group, that is, with $\refl_{\pt}$ for the neutral
-element, path composition for the group multiplication, and path inverse for the group inverse operation. The group
-axioms are then given by the higher paths corresponding to groupoid laws. We have shown that loops in $\UFin$ exactly
-encode the automorphism group $\Aut[\Fin[n]]$ for every $n$. This is a general technique called delooping, where a group
-can be identified with a 1-object groupoid, internally in HoTT.~\footnote{This technique also allows defining higher
-groups in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}
+The loopspace of a pointed type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element,
+path composition for the group multiplication, and path inverse for the group inverse operation. The group axioms are
+then given by the higher paths corresponding to groupoid laws. We have shown that loops in $\UFin$ exactly encode the
+automorphism group $\Aut[\Fin[n]]$ for every $n$. This is a general technique called \emph{delooping}, where a group can
+be identified with a 1-object groupoid, internally in HoTT.~\footnote{This technique also allows defining higher groups
+  in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}
 
@@
 
-The groupoid $\UFin$ has two symmetric monoidal structures, the additive and the multiplicative one, and the
-multiplicative tensor product distributes over the additive one. To construct these, we first define some equivalences
-on $\Fin$, and some general type isomorphisms.
+Similar to $\BFin$, the groupoid $\UFin$ has two symmetric monoidal structures, the additive and the multiplicative one,
+and the multiplicative tensor product distributes over the additive one. To construct these, we first state and prove
+some equivalences on $\Fin$, and some general type isomorphisms. Then we simply lift these equivalences to $\UFin$, by
+the univalence principle.
 
@@
 
-We lift these equivalences to $\UFin$ giving it the additive and multiplicative symmetric monoidal structures, $(F_0,
-  \sqcup)$ and $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and
-the braiding isomorphism $\mathcal{B}_{X,Y}$. These isomorphisms satisfy the Mac Lane coherence conditions for symmetric
-monoidal categories, that is, the triangle, pentagon, and hexagon identities, and the symmetry of the braiding, upto
-2-paths in $\UFin$.
+\begin{proposition}
+  $\UFin$ has a symmetric rig structure, with the additive and multiplicative symmetric monoidal structures given by
+  $(F_0, \sqcup)$ and $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$,
+  $\alpha_{X,Y,Z}$, and the braiding isomorphism $\mathcal{B}_{X,Y}$ upto 1-paths in $\UFin$. These isomorphisms satisfy
+  the Mac Lane coherence conditions for symmetric monoidal categories, that is, the triangle, pentagon, and hexagon
+  identities, and the symmetry of the braiding, upto 2-paths in $\UFin$.
+\end{proposition}
 

### 2021-07-06 Pi+/paper/finite.tex
Revise section 4

@@
 
+% Recap of the previous section
 In~\cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
@@
 
-In order to study syntactic descriptions of permutations, we need to decide whether two descriptions refer to the same
-permutation. This is the \emph{word problem} for $\Sn$. Putting it in this form also allows us to connect it to the
-broader scope of computational group theory and combinatorics -- we borrow ideas such as Coxeter relations and Lehmer
-codes from there.
+% What now?
+In order to study syntactic descriptions of permutations, we will hit the problem of deciding whether two descriptions
+refer to the same permutation -- in group theory, this is the \emph{word problem} for $\Sn$. Putting it in this form
+allows us to connect it to the broader scope of computational group theory and combinatorics -- we can borrow ideas such
+as Coxeter relations and Lehmer codes.
 
-The goal of this section is to reconcile two different approaches to defining the symmetric group -- as an automorphism
-group, and as a group presented using generators and relations. The generators of the group can be thought of as the
-primitive combinators in a (reversible) programming language, the group structure gives the composition and inverse
-operations, and the relations describe how these primitive combinators interact with each other.
+% The summary of what we'll do
+Thus, the goal of this section is to reconcile two different approaches to defining the symmetric group -- as an
+automorphism group, and as a group syntactically presented using \emph{generators} and \emph{relations}. The generators
+of the group are similar to the primitive combinators in a (reversible) programming language -- the group structure
+gives the composition and inverse operations, and the relations describe how these primitive combinators interact with
+each other.
+
+% High-level explanation of technical devices
+First, we will define the required notions of free groups and group presentations, and state some of their most
+important properties. Then, we introduce our chosen \emph{Coxeter} presentation for $\Sn$. To solve the word problem for
+$\Sn$, we will use a rewriting system, with a suitable, well behaved collection of reduction rules corresponding to the
+Coxeter presentation. Finally, we describe the normal forms in this rewriting system, using Lehmer codes, and prove the
+correspondence between them and the type $\Aut[\Fin[n]]$ of automorphisms on a finite set. The generators and relations
+we use here will be used to quote back to 1 and 2-combinators in $\PiLang$ (see~\cref{sec:equivalence}).
 
@@
   \item swap the $i$-th element with an element at a fixed position, or
-  \item flip a prefix $\Fin[k]$ of $\Fin[n]$ for $k \leq n$, or
+  \item reverses a prefix $\Fin[k]$ of $\Fin[n]$ for $k \leq n$, or
   \item cyclically shift any subset of $\Fin[n]$.
@@
 operation of swapping the $i$-th element with the $j$-th element, cycle sort uses cyclical shifts of subsequences,
-pancake sort uses flips of prefixes of the list, et cetera~\todo{check!}. The choice of generators for our presentation
-is important for the following reasons.
+pancake sort uses reversals of the prefixes of the list, et cetera. The choice of generators for our presentation is
+important for the following reasons.
 
 \begin{itemize}
-  \item It affects the difficulty of solving the word problem in $\Sn$ and formalising the proof of its correctness. In
-        the following subsections, we build a rewriting system using the generators and relations, and we need to prove
-        and formalise strong normalisation for this system.
+  \item It affects the difficulty of solving the word problem in $\Sn$ and formalising the proof of its correctness.
   \item The choice of generators dictates which words become normal forms in this presentation of $\Sn$. These normal
@@
     F(A) \to^G G$ satisfying $h \comp \eta_A \htpy f$ is contractible.
-
+  \begin{toappendix}
     % https://q.uiver.app/?q=WzAsMyxbMCwyLCJBIl0sWzAsMCwiRihBKSJdLFsyLDAsIkciXSxbMCwxLCJcXGV0YV9BIl0sWzAsMiwiZiIsMl0sWzEsMiwiXFxleHRlbmR7Zn0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkYXNoZWQifX19XSxbMyw0LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
@@
       \end{tikzcd}\]
+  \end{toappendix}
 \end{proposition}
@@
 
-The idea for solving the word problem for $\Sn$ is to turn these undirected relations into a terminating rewriting
+The idea for solving the word problem for $\Sn$ is to turn these undirected relations into a rewriting
 system $(\List[\Fin[n]],\cox*)$, so that, by repeatedly applying the reduction rules as long as possible, any two
@@
 
-For this to work, we first need the system to have the strong normalisation property, meaning that there are no infinite
+For this to work, we first need the system to have the termination property, meaning that there are no infinite
 reductions. We observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
@@
 rewriting system is confluent -- meaning that all critical pairs, that is, terms with overlapping possible reduction
-rules, have to converge. For example, in our system, the pairs in~\cref{fig:critical-pairs-converge} converge.
+rules, have to converge. For example, in our system, the pairs in~\cref{fig:critical-pairs-converging} converge.
 Unfortunately, this is not true for all critical pairs -- an  example is in~\cref{fig:critical-pairs-non-converging},
@@
 % \end{figure}
+%  \begin{subfigure}[]{0.45\textwidth}
+%    % https://q.uiver.app/?q=WzAsNSxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8yIl0sWzAsMSwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFsyLDQsIlxcdGF1XzJcXHRhdV8xIl0sWzAsMiwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFsxLDMsIlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwxLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzAsMiwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzMsNCwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFs0LDIsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV1d
+%    \adjustbox{scale=\scalef,center}{
+%      \begin{tikzcd}
+%        && {\tau_2\tau_1\tau_2\tau_2} \\
+%        {\tau_1\tau_2\tau_1\tau_2} \\
+%        {\tau_1\tau_2\tau_1\tau_2} \\
+%        & {\tau_1\tau_1\tau_2\tau_1} \\
+%        && {\tau_2\tau_1}
+%        \arrow["\braid"', squiggly, from=1-3, to=2-1]
+%        \arrow["\cancel", squiggly, from=1-3, to=5-3]
+%        \arrow["\braid"', squiggly, from=2-1, to=3-1]
+%        \arrow["\braid"', squiggly, from=3-1, to=4-2]
+%        \arrow["\cancel"', squiggly, from=4-2, to=5-3]
+%      \end{tikzcd}
+%    }
+%  \end{subfigure}
+%   \begin{subfigure}[]{0.45\textwidth}
+%    $\gspan[\braid][\swap]{\tau_3\tau_2\tau_3\tau_1}{\tau_2\tau_3\tau_2\tau_1}{\tau_3\tau_2\tau_1\tau_3}$
+%    \caption{Non-converging critical pairs}
+%    \label{fig:critical-pairs-non-converging}
+%  \end{subfigure}
 
 \begin{figure}
+  \centering
   \begin{subfigure}[b]{0.45\textwidth}
-    % https://q.uiver.app/?q=WzAsNixbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzQsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwyLCJcXHRhdV8xXFx0YXVfMlxcdGF1XzIiXSxbMiwzLCJcXHRhdV8xIl0sWzQsMiwiXFx0YXVfMlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFswLDIsImJyYWlkIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJjYW5jZWwiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFszLDQsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzIsNSwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbNSw0LCJjYW5jZWwiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dXQ==
+    % https://q.uiver.app/?q=WzAsNixbMSwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzIsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwyLCJcXHRhdV8xXFx0YXVfMlxcdGF1XzIiXSxbMSwzLCJcXHRhdV8xIl0sWzIsMiwiXFx0YXVfMlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFswLDIsImJyYWlkIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJjYW5jZWwiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFszLDQsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzIsNSwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbNSw0LCJjYW5jZWwiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dXQ==
     \adjustbox{scale=\scalef,center}{
       \begin{tikzcd}
-        && {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
-        {\tau_1\tau_2\tau_1\tau_1\tau_2} &&&& {\tau_2\tau_1\tau_1\tau_2\tau_1} \\
-        {\tau_1\tau_2\tau_2} &&&& {\tau_2\tau_2\tau_1} \\
-        && {\tau_1}
-        \arrow["\braid"', squiggly, from=1-3, to=2-1]
-        \arrow["\braid", squiggly, from=1-3, to=2-5]
+        & {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
+        {\tau_1\tau_2\tau_1\tau_1\tau_2} && {\tau_2\tau_1\tau_1\tau_2\tau_1} \\
+        {\tau_1\tau_2\tau_2} && {\tau_2\tau_2\tau_1} \\
+        & {\tau_1}
+        \arrow["\braid"', squiggly, from=1-2, to=2-1]
+        \arrow["\braid", squiggly, from=1-2, to=2-3]
         \arrow["\cancel"', squiggly, from=2-1, to=3-1]
-        \arrow["\cancel"', squiggly, from=3-1, to=4-3]
-        \arrow["\cancel", squiggly, from=2-5, to=3-5]
-        \arrow["\cancel", squiggly, from=3-5, to=4-3]
+        \arrow["\cancel"', squiggly, from=3-1, to=4-2]
+        \arrow["\cancel", squiggly, from=2-3, to=3-3]
+        \arrow["\cancel", squiggly, from=3-3, to=4-2]
       \end{tikzcd}
     }
+    \caption{Converging critical pairs}
+    \label{fig:critical-pairs-converging}
   \end{subfigure}
-  \begin{subfigure}[]{0.45\textwidth}
-    % https://q.uiver.app/?q=WzAsNSxbMiwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8yIl0sWzAsMSwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFsyLDQsIlxcdGF1XzJcXHRhdV8xIl0sWzAsMiwiXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFsxLDMsIlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwxLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzAsMiwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzMsNCwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFs0LDIsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV1d
+  \begin{subfigure}[b]{0.50\textwidth}
     \adjustbox{scale=\scalef,center}{
+      % https://q.uiver.app/?q=WzAsMyxbMCwwLCJcXHRhdV8zXFx0YXVfMlxcdGF1XzNcXHRhdV8xIl0sWzAsMiwiXFx0YXVfMlxcdGF1XzNcXHRhdV8yXFx0YXVfMSJdLFsyLDEsIlxcdGF1XzNcXHRhdV8yXFx0YXVfMVxcdGF1XzMiXSxbMCwxLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzAsMiwiYnJhaWQiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFsyLDEsIlxcYnJhaWQqIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XV0=
       \begin{tikzcd}
-        && {\tau_2\tau_1\tau_2\tau_2} \\
-        {\tau_1\tau_2\tau_1\tau_2} \\
-        {\tau_1\tau_2\tau_1\tau_2} \\
-        & {\tau_1\tau_1\tau_2\tau_1} \\
-        && {\tau_2\tau_1}
-        \arrow["\braid"', squiggly, from=1-3, to=2-1]
-        \arrow["\cancel", squiggly, from=1-3, to=5-3]
-        \arrow["\braid"', squiggly, from=2-1, to=3-1]
-        \arrow["\braid"', squiggly, from=3-1, to=4-2]
-        \arrow["\cancel"', squiggly, from=4-2, to=5-3]
+        {\tau_3\tau_2\tau_3\tau_1} \\
+        && {\tau_3\tau_2\tau_1\tau_3} \\
+        {\tau_2\tau_3\tau_2\tau_1}
+        \arrow["\braid"', squiggly, from=1-1, to=3-1]
+        \arrow["\braid", squiggly, from=1-1, to=2-3]
+        \arrow["{\longbraid}", squiggly, from=2-3, to=3-1]
       \end{tikzcd}
     }
+    \caption{Non-converging critical pairs converging in $\longcox*$}
+    \label{fig:critical-pairs-non-converging}
   \end{subfigure}
-  \caption{Converging critical pairs}
-  \label{fig:critical-pairs-converge}
-\end{figure}
 
-\begin{figure}
-  $\gspan[\braid][\swap]{\tau_3\tau_2\tau_3\tau_1}{\tau_2\tau_3\tau_2\tau_1}{\tau_3\tau_2\tau_1\tau_3}$
-  \caption{Non-converging critical pairs}
-  \label{fig:critical-pairs-non-converging}
+  \caption{Critical Pairs}
+  \label{fig:critical-pairs}
 \end{figure}
@@
 system $(\List[\Fin[n]], \longcox*)$, partially based on the Coxeter relations, and prove that it has the desired
-properties of confluence and strong normalisation. We prove that the new relation defined by this system is equivalent,
+properties of confluence and termination. We prove that the new relation defined by this system is equivalent,
 in a technical sense, to the standard Coxeter relation $\cox*$.
@@
   \end{tikzpicture}
-  \caption{Braiding diagram for $\longcox$ with $n = 0$ and $k = 3$}
+  \caption{Braiding diagram for $\longbraid$ with $n = 0$ and $k = 3$}
   \label{fig:longcox}
@@
 We define the relation $\longcox*$ to be the closure of $\longcox$, under reflexivity and (right-step extended)
-transitivity.
+transitivity (see~\cref{def:longcox-star}).
 
+\begin{toappendix}
   \begin{definition}[$\longcox*$]
+    \label{def:longcox-star}
     \begin{align*}
@@
   \end{definition}
+\end{toappendix}
 
 Despite the increased complexity of the generators, the rewriting system $(\List[\Fin[n]],\longcox*)$ has the properties
-we desire. It satisfies confluence, that is, the Church-Rosser (diamond) property, and it is strongly normalising, since
-it produces a unique normal form. We follow the terminology
+we desire. It satisfies local confluence, that is, the Church-Rosser (diamond) property, and it is terminating, so, by
+Newman's lemma, it produces a unique normal form. We follow the terminology
 of~\cite*{huetConfluentReductionsAbstract1980a,krausCoherenceWellFoundednessTaming2020} to state our results
@@
 \begin{proposition}
+  \label{prop:nf}
   \leavevmode
@@
           extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
+    \item $\longcox*$ is terminating. For every $w \longcox v$, it holds that $v < w$, where $<$ is the (well-founded)
+          lexicographic ordering on $\List[\Fin[n]]$.
     \item $\longcox*$ is confluent. For every extended span $\coxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
           extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
-    \item $\longcox*$ is well-founded. With $<$ the lexicographic ordering on $\List[\Fin[n]]$, and $w$ being
-          $<$-accessible if every $v < w$ is $<$-accessible, we have that every $w$ is $<$-accesible.
-    \item $\longcox*$ is strongly normalising. For every $w$, there exists a unique $v$ such that $w \longcox* v$.
+    \item For every $w$, there exists a unique normal form $v$ such that $w \longcox* v$.~\label{prop:nf-uniqueness}
   \end{enumerate}
@@
 The modified form of the Coxeter relations are unwieldy and difficult to prove properties about by induction. However,
-we can use the following observation relating it to $\cox*$.~\todo{Check Huet's paper for this property.}
+we can make the following observation relating it to $\cox*$.
 
@@
 
-By strong normalisation, we get a unique choice function $\normf : {\List[\Fin[n]]} \to {\List[\Fin[n]]}$ that produces
-a normal form for terms of $\List[\Fin[n]]$. We state and prove a couple of important properties enjoyed by $\normf$.
+By~\cref{prop:nf}~\ref{prop:nf-uniqueness}, we get a unique choice function $\normf : {\List[\Fin[n]]} \to
+  {\List[\Fin[n]]}$ that produces a normal form for terms of $\List[\Fin[n]]$. We state and prove a couple of important
+properties enjoyed by $\normf$.
 

### 2021-07-06 Pi+/paper/finite.tex
Remove dangling newlines
Fix the example longcox

@@
     \braid
-     & : \forall n \to (\suc[n] \cons n \cons \suc[n] \cons \nil) \cox (n \cons \suc[n] \cons n \cons \nil) \\
+     & : \forall n \to (\suc[n] \cons n \cons \suc[n] \cons \nil) \cox (n \cons \suc[n] \cons n \cons \nil)
   \end{align*}
@@
       \relr{\cox}
-       & : \forall w_{1}, w_{2} \to w_{1} \cox w_{2} \to w_{1} \cox* w_{2}                                                                   \\
+       & : \forall w_{1}, w_{2} \to w_{1} \cox w_{2} \to w_{1} \cox* w_{2}
     \end{align*}
@@
     \longbraid
-     & : \forall n, l, r \to (l \append (n \downf 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downf 2 + k) \append r) \\
+     & : \forall n, l, r \to (l \append (n \downf 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downf 2 + k) \append r)
   \end{align*}
@@
 sequence of transpositions and a single transposition afterwards. For example, for $n = 0$ and $k = 3$, it allows for
-the reduction $[4, 3, 2, 1, 0, 4] \longcox* [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox}).
+the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox}).
 
@@
       \transr{\longcox}
-       & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcox* w_{3} \to w_{1} \longcox* w_{3} \\
+       & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcox* w_{3} \to w_{1} \longcox* w_{3}
     \end{align*}
@@
       \eqv & \unit                   \\
-      \eqv & \Aut[\Fin[\suc[\zero]]] \\
+      \eqv & \Aut[\Fin[\suc[\zero]]] 
     \end{array}
@@
       \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
-      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                          & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}} \\
+      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                          & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}}
     \end{array}

### 2021-07-06 Pi+/paper/equivalence.tex
Remove dangling newlines
working on intro

@@
-\section{Going from \texorpdfstring{$\PiLang$}{Pi} to \texorpdfstring{$\UFin$}{UFin} and back}~\label{sec:equivalence}
+\section{Going from \texorpdfstring{$\PiLang$}{Pi} to \texorpdfstring{$\UFin$}{UFin} and back}
+\label{sec:equivalence}
 
@@
   \end{tikzcd}\]
+
+\vc{ The groupoid of finite types is the free symmetric monoidal groupoid on one generator. This can be presented as an
+  algebraic 2-theory, which is our syntax for $\PiHatLang$. Vertical categorification of natural numbers as a free
+  commutative monoid. See groupoidification. }
 

### 2021-07-06 Pi+/paper/examples.tex
working on intro

\section{Reversible Circuits in Qiskit}
\label{sec:qiskit}

To make the connection to reversible computing explicit, we use the Agda
formalisation of our technical result to extract various computational
procedures to synthesise, normalise, optimise, and verify reversible
circuits. In this introduction, we use the popular IBM Qiskit platform for
quantum computing to illustrate these ideas. First, most current quantum
algorithms start with generating superpositions, evolving them using a unitary
transformation, and then projecting them with a measurement operator. The middle
stage is essentially a reversible classical computation (executed in a
quantum-parallel fashion). As a simple example for that middle stage, we
synthesise a reversible circuit implementing boolean disjunction
($\vee$). Following~\citet{Toffoli:1980}, the first step is to write a
specification for the desired reversible function:
\[
\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)
\]
where $\underline{\vee}$ is the exclusive-or operation. From the definition of $\underline{\vee}$ it is evident that setting $h=0$, we can compute the desired disjunction by observing the first component of the result. The $\mathit{reversibleOr}$ function has the following truth table (in binary on the left and in a more convenient decimal notation on the right):

\begin{center}\begin{tabular}{|ccc|ccc|@{\qquad\qquad}|c|c|}
0 & 0 & 0 &     0 & 0 & 0     & 0 & 0 \\
0 & 0 & 1 &     1 & 0 & 1     & 1 & 5 \\
0 & 1 & 0 &     1 & 1 & 0    & 2 & 6 \\
0 & 1 & 1 &     1 & 1 & 1    & 3 & 7 \\
1 & 0 & 0 &     1 & 0 & 0    & 4 & 4 \\
1 & 0 & 1 &     0 & 0 & 1    & 5 & 1 \\
1 & 1 & 0 &     0 & 1 & 0    & 6 & 2 \\
1 & 1 & 1 &     0 & 1 & 1    & 7 & 3
\end{tabular}\end{center}

\noindent where it is evident that it is a bijective function, i.e., reversible.

The above embedding of an irreversible function into a reversible function with additional inputs and outputs is
completely general and is the starting point for specifications of quantum circuits. The challenge is to synthesise a
program / circuit from this specification. Of course, writing this program in a conventional (irreversible) language
defeats the purpose. The challenge is to construct the desired program / circuit exclusively using reversible
primitives, e.g., the standard set of universal reversible gates used in frameworks like Qiskit which consists of the
computational gates \textsf{not} (boolean negation, called \verb|x|), \textsf{cnot} (conditional negation of the second
input if the first is true; called \verb|cx|), and \textsf{toffoli} (conditional negation of the third input if both the
first two inputs are true; called \verb|ccx|) gates, and the ability re-arrange the layout of wires. For concreteness,
here is a possible implementation of the desired function in Qiskit:

\begin{center}
  \begin{minipage}[c]{0.4\linewidth}
\begin{verbatim}
reversibleOr.qasm:

  // setup
  ccx q[1], q[2], q[0];
  cx  q[1], q[0];
  cx  q[2], q[0];
  // measure

% ./qasm -t reversibleOr.qasm
+-------+-------+
| 0 0 0 | 0 0 0 |
| 0 0 1 | 1 0 1 |
| 0 1 0 | 1 1 0 |
| 0 1 1 | 1 1 1 |
| 1 0 0 | 1 0 0 |
| 1 0 1 | 0 0 1 |
| 1 1 0 | 0 1 0 |
| 1 1 1 | 0 1 1 |
+-------+-------+
  \end{verbatim}
  \end{minipage}
  \qquad
  \includegraphics[scale=0.7]{reversibleOr.png}
\end{center}

\noindent This implementation was manually produced using a standard synthesis algorithm for reversible
circuits~\cite{10.1145/775832.775915}. To gain some intuition, we trace the evaluation of the circuit for input
\verb|011|. In this context, the most significant bit is at index 0. Thus the first \verb|ccx| gate negates \verb|q[0]|
since both \verb|q[1]| and \verb|q[2]| are true producing \verb|111|; the following \verb|cx| gate produces \verb|011|; finally the last \verb|cx| produces the result \verb|111|.

There is wealth of manual and algorithmic approaches for such synthesis problems each optimizing along  different dimensions~\cite{maslov:2003:rls:1087512,1201583}. Here is the circuit produced using an approach that analyzes the recursive structure of the circuit (and would generalise to computing the disjunction of more than two inputs):

\begin{center}
  \begin{minipage}[c]{0.4\linewidth}
\begin{verbatim}
reversibleOr2.qasm:

  // setup
  cx  q[1], q[0];
  x   q[1];
  ccx q[1], q[2], q[0];
  x   q[1];
  // measure
  \end{verbatim}
  \end{minipage}
  \qquad
  \includegraphics[scale=0.7]{reversibleOr2.png}
\end{center}

\noindent The evaluation of this circuit prints the same truth table as above confirming their equivalence. Tracing the
evaluation on the same input value \verb|011| goes through the stages \verb|111|, \verb|101|, \verb|101|, and finally \verb|111|.

The situation for Shor's algorithm is just as described above but with a more involved function of the form $f(r) = a^{r} \mod N$ for fixed $a$ and $N$. The specification of the circuit is relatively straightforward to calculate. Here it is for $a=11$ and $N=15$:
\[\begin{array}{rcll}
g(r,h) &=& \left\{ \begin{array}{ll}
                     (r,h+1) & \mbox{when~$r$~even~and~$h$~even} \\
                     (r,h-1) & \mbox{when~$r$~even~and~$h$~odd} \\
                     (r,11-h) & \mbox{when~$r$~odd~and~$4 > h \geq 0$~or~$12 > h \geq 8$} \\
                     (r,19-h) & \mbox{when~$r$~odd~and~$8 > h \geq 4$~or~$16 > h \geq 12$}
                                \end{array}\right.
\end{array}\]

\noindent However, as explained in standard accounts of the algorithm (e.g., the Qiskit implementation), producing an efficient
modular exponentiation circuit from this specification  is not straightforward and is actually the bottleneck in Shor’s algorithm. Typical
derivations of the circuit start from elementary gates, build a circuit for reversible disjunction and conjunction of
booleans, a circuit for a half-adder, a circuit for computing the carry, progressing to a circuit for modular addition,
which is used to build a circuit for modular multiplication, and then finally a circuit for modular exponentiation
taking care at each step to avoid the exponential blowup (e.g., by implementing exponentiation by squaring instead of repeated
multiplication)~\cite{shorefficient}.

### 2021-07-06 Pi+/paper/finite.tex
working on intro

--- 2021-07-06 Pi+/paper/finite.tex
@@
-\section{The Group of Permutations}~\label{sec:finite}
+\section{The Group of Permutations}
+\label{sec:finite}
 
@@
         thick]
-      {braid={s_1, s_2, s_3, s_4}};
+      {braid={s_4, s_3, s_2, s_1}};
     \end{tikzpicture}
@@
 visualised in the same way as before. The remaining constructor uses the helper function to exchange the order of a long
-sequence of transpositions and a single transposition afterwards. For example, for $n = 0$ and $k = 3$, it allows for
-the reduction $[4, 3, 2, 1, 0, 4] \longcox* [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox}).
+sequence of transpositions and a single transposition afterwards. For example, for $n = 2$ and $k = 3$, it allows for
+the reduction $[6, 5, 4, 3, 2, 6] \longcox* [5, 6, 5, 4, 3, 2]$ (see~\cref{fig:longcox} for the braid diagram).
 
@@
       thick] at (0, 0)
-    {braid={s_5, s_1, s_2, s_3, s_4, s_5}};
+    {braid={s_5, s_4, s_3, s_2, s_1, s_5}};
     \node[font=\Huge] at (4, 1.3) {\(\rightarrow\)};
@@
       thick] at (5, 0)
-    {braid={s_1, s_2, s_3, s_4, s_5, s_4}};
+    {braid={s_4, s_5, s_4, s_3, s_2, s_1}};
   \end{tikzpicture}

### 2021-07-06 Pi+/paper/finite.tex
Some text in section 3.1

--- 2021-07-06 Pi+/paper/finite.tex
@@
 First, we will define the required notions of free groups and group presentations, and state some of their most
-important properties. Then, we introduce our chosen \emph{Coxeter} presentation for $\Sn$. To solve the word problem for
+important properties. Then, we introduce our chosen \emph{Coxeter presentation} for $\Sn$. To solve the word problem for
 $\Sn$, we will use a rewriting system, with a suitable, well behaved collection of reduction rules corresponding to the
@@
 Following the universal-algebraic definition, in HoTT, we could use a naive higher inductive type to define the free
-group, which enforces the group axioms by adding path constructors (see~\cref{def:free-group} for the full definition).
+group, which enforces the group axioms by adding path constructors (see~\cref{def:free-group}).
 Using the induction principle, we can easily verify the universal property. However, since this definition of $F(A)$ has
@@
     \adjustbox{scale=\scalef,center}{
-      % https://q.uiver.app/?q=WzAsMyxbMCwwLCJcXHRhdV8zXFx0YXVfMlxcdGF1XzNcXHRhdV8xIl0sWzAsMiwiXFx0YXVfMlxcdGF1XzNcXHRhdV8yXFx0YXVfMSJdLFsyLDEsIlxcdGF1XzNcXHRhdV8yXFx0YXVfMVxcdGF1XzMiXSxbMCwxLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzAsMiwiYnJhaWQiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFsyLDEsIlxcYnJhaWQqIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XV0=
+    % https://q.uiver.app/?q=WzAsMyxbMSwwLCJcXHRhdV8zXFx0YXVfMlxcdGF1XzNcXHRhdV8xIl0sWzAsMiwiXFx0YXVfMlxcdGF1XzNcXHRhdV8yXFx0YXVfMSJdLFszLDEsIlxcdGF1XzNcXHRhdV8yXFx0YXVfMVxcdGF1XzMiXSxbMCwxLCJicmFpZCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzAsMiwiYnJhaWQiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFsyLDEsIlxcbG9uZ2JyYWlkIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XV0=
     \begin{tikzcd}
-        {\tau_3\tau_2\tau_3\tau_1} \\
-        && {\tau_3\tau_2\tau_1\tau_3} \\
+      & {\tau_3\tau_2\tau_3\tau_1} \\
+      &&& {\tau_3\tau_2\tau_1\tau_3} \\
       {\tau_2\tau_3\tau_2\tau_1}
-        \arrow["\braid"', squiggly, from=1-1, to=3-1]
-        \arrow["\braid", squiggly, from=1-1, to=2-3]
-        \arrow["{\longbraid}", squiggly, from=2-3, to=3-1]
+      \arrow["braid"', squiggly, from=1-2, to=3-1]
+      \arrow["braid", squiggly, from=1-2, to=2-4]
+      \arrow["\longbraid", squiggly, from=2-4, to=3-1]
     \end{tikzcd}
     }
-    \caption{Non-converging critical pairs converging in $\longcox*$}
+    \caption{A non-converging critical pair in $\cox*$ converging in $\longcox*$}
     \label{fig:critical-pairs-non-converging}

### 2021-07-06 Pi+/paper/reversible1.tex
working on intro
Some text in section 3.1

@@
 %% ~3 pages
+\label{sec:pi}
 \label{sec:reversibleone}
@@
 The circuit model of reversible computation discussed in the introduction is a useful abstraction close to the hardware
-platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly''
+platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly-level''
 programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical model based on
-permutations of finite sets provides a richer algebraic model which we review in this section.
+permutations of finite sets provides a richer algebraic structure which we review in this section.
 
@@
 
-A specialized language for programming with permutations on finite sets can be built using ideas going back to Kelly,
-Laplaza, and Mac Lane~\cite{laplaza72,kelly74,KELLY197197} that the commutative semiring (also known as the commutative
-rig) of natural numbers exactly characterizes permutations on finite sets. In programming parlance, as shown in
-Fig.~\ref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes a constant in the programming
-language and the algebraic closure operators become composition operators in the programming language. Once equipped
-with a notion of values and types, we get the following syntactic presentation of a programming language for
-permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:
+In a reversible boolean circuit, the number of input bits matches the number of output bits. Circuits can be composed
+sequentially, vertically, or horizontally, respectively preserving the number of bits, adding up the number of bits or
+multiplying them. So, to program with reversible circuits, all we need to do is make sure that each primitive operation
+preserves the number of bits, which is just a natural number. Natural numbers are the free commutative semiring (or,
+commutative rig) on one generator, with $(0,+)$ for addition, and $(1,\times)$ for multiplication, so the commutative
+semirig identities can be used to design a logic for reversible programming. This inspired the syntax for various
+first-order reversible logics and reversible programming
+languages~\cite*{sparksSuperstructuralReversibleLogic2014,jamesInformationEffects2012}.
+
+Natural number identities have no computational content. To obtain a computational interpretation of the commutative rig
+structure, the logic needs to be equipped with values and types, and a notion of operational semantics and contextual
+equivalence~\cite{jamesInformationEffects2012}. This is a programming language which embodies the computational content
+of isomorphisms of finite types, or permutations. On the semantic side, we need to consider the groupoidification of a
+commutative rig, that is, a symmetric rig groupoid. A sound semantics for $\PiLang$ in weak rig groupoids was
+established in~\cite{caretteComputingSemiringsWeak2016}, and conjectured to be complete.
+
+% In programming parlance, as shown in~\cref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes
+% a reversible combinator in the programming language and the algebraic closure operators become composition operators in
+% the programming language. Once equipped with a notion of values and types, we get the following syntactic presentation
+% of a programming language for permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:
+
+% A specialised language for programming with permutations on finite sets can be built using ideas going back to Kelly,
+% Laplaza, and Mac Lane~\cite{laplaza72,kelly74,KELLY197197} that the commutative semiring (also known as the commutative
+% rig) of natural numbers exactly characterizes permutations on finite sets. In programming parlance, as shown in
+% Fig.~\ref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes a constant in the programming
+% language and the algebraic closure operators become composition operators in the programming language. Once equipped
+% with a notion of values and types, we get the following syntactic presentation of a programming language for
+% permutations~\cite{James:2012:IE:2103656.2103667,Carette2016}:
 

### 2021-07-07 Pi+/paper/equivalence.tex
Revise equivalence section

@@
     \evalt_{1} & : (c : X \iso Y) \to \evalt_{0}(X) \iso \evalt_{0}(Y)      \\
-    \evalt_{2} & : (\alpha : p \Iso q) \to \evalt_{1}(p) \Iso \evalt_{1}(q) \\
+    \evalt_{2} & : (\alpha : p \Iso q) \to \evalt_{1}(p) \Iso \evalt_{1}(q)
   \end{align*}
@@
   \begin{align*}
-    \evalh_{0} & : \UPlus \to \UHat                                                          \\
-    \evalh_{1} & : (c : t_{1} \iso t_{2}) \to \evalh_{0}(t_{1}) \isoh \evalh_{0}(t_{2})      \\
-    \evalh_{2} & : (\alpha : c_{1} \Iso c_{2}) \to \evalh_{1}(c_{1}) \Isoh \evalh_{1}(c_{2}) \\
+    \evalp_{0} & : \UPlus \to \UHat                                                          \\
+    \evalp_{1} & : (c : t_{1} \iso t_{2}) \to \evalp_{0}(t_{1}) \isoh \evalp_{0}(t_{2})      \\
+    \evalp_{2} & : (\alpha : c_{1} \Iso c_{2}) \to \evalp_{1}(c_{1}) \Isoh \evalp_{1}(c_{2})
   \end{align*}
@@
   \begin{align*}
-    \quoteh_{0} & : \UHat \to \UPlus                                                            \\
-    \quoteh_{1} & : (p : X_{1} \isoh X_{2}) \to \quoteh_{0}(X_{1}) \iso \quoteh_{0}(X_{2})      \\
-    \quoteh_{2} & : (\alpha : p_{1} \Isoh p_{2}) \to \quoteh_{1}(p_{1}) \Iso \quoteh_{1}(p_{2}) \\
+    \quotep_{0} & : \UHat \to \UPlus                                                            \\
+    \quotep_{1} & : (p : X_{1} \isoh X_{2}) \to \quotep_{0}(X_{1}) \iso \quotep_{0}(X_{2})      \\
+    \quotep_{2} & : (\alpha : p_{1} \Isoh p_{2}) \to \quotep_{1}(p_{1}) \Iso \quotep_{1}(p_{2})
   \end{align*}
@@
 \begin{proposition}
-  $\evalh/\quoteh$ give a symmetric monoidal biequivalence between $\PiPlusCat$ and $\PiHatCat$.
+  $\evalp/\quotep$ give a symmetric monoidal biequivalence between $\PiPlusCat$ and $\PiHatCat$.
 \end{proposition}
@@
   \begin{align*}
-    \evalt_{0} & : \UHat \to \UFin                                                          \\
-    \evalt_{1} & : (c : t_{1} \isoh t_{2}) \to \evalh_{0}(t_{1}) \id \evalh_{0}(t_{2})      \\
-    \evalt_{2} & : (\alpha : c_{1} \Isoh c_{2}) \to \evalt_{1}(c_{1}) \id \evalt_{1}(c_{2}) \\
+    \evalh_{0} & : \UHat \to \UFin                                                          \\
+    \evalh_{1} & : (c : t_{1} \isoh t_{2}) \to \evalh_{0}(t_{1}) \id \evalh_{0}(t_{2})      \\
+    \evalh_{2} & : (\alpha : c_{1} \Isoh c_{2}) \to \evalh_{1}(c_{1}) \id \evalh_{1}(c_{2})
   \end{align*}
@@
   \begin{align*}
-    \quotet_{0} & : \UFin \to \UHat                                                            \\
-    \quotet_{1} & : (p : X_{1} \id X_{2}) \to \quoteh_{0}(X_{1}) \isoh \quoteh_{0}(X_{2})      \\
-    \quotet_{2} & : (\alpha : p_{1} \id p_{2}) \to \quoteh_{1}(p_{1}) \Isoh \quotet_{1}(p_{2}) \\
+    \quoteh_{0} & : \UFin \to \UHat                                                            \\
+    \quoteh_{1} & : (p : X_{1} \id X_{2}) \to \quoteh_{0}(X_{1}) \isoh \quoteh_{0}(X_{2})      \\
+    \quoteh_{2} & : (\alpha : p_{1} \id p_{2}) \to \quoteh_{1}(p_{1}) \Isoh \quotet_{1}(p_{2})
   \end{align*}
@@
 \begin{proposition}
-  $\evalt/\quotet$ give a symmetric monoidal biequivalence between $\PiHatCat$ and $\UFin$.
+  $\evalh/\quoteh$ give a symmetric monoidal biequivalence between $\PiHatCat$ and $\UFin$.
+\end{proposition}
+
+The semantics that we presented here takes a different route to constructing the permutation from a $\PiLang$
+combinator, compared to the direct interpretation given using the big-step interpreter in~\cref{subsec:denotational}. We
+verify that the two semantics agree, establishing that the semantics is adequate and fully abstract.
+
+\begin{definition}[${\gdenot{\blank}} : U \to {\UFin}$]
+  \begin{align*}
+    \gdenot{\blank}_{0} & = \evalh_{0} \comp \evalp_{0} \comp \evalt_{0} \\
+    \gdenot{\blank}_{1} & = \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
+  \end{align*}
+\end{definition}
+
+\begin{proposition}
+  If $c_1, c_2 : X \iso Y$, then $\denot{c_1} = \denot{c_2}$ iff $\gdenot{c_1}_{1} = \gdenot{c_2}_{1}$.
 \end{proposition}
@@
     \normt_{0} & : U \to \UPlus                                            \\
-    \normt_{0} & = \quoteh_{0} \comp \evalt_{0}                        \\
+    \normt_{0} & = \quotep_{0} \comp \quoteh_{0} \comp \gdenot{\blank}_{0} \\
     \\
     \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \iso \normt_{0}(Y)     \\
-    \normt_{1} & = \quoteh_{1} \comp \evalt_{1}                        \\
+    \normt_{1} & = \quotep_{1} \comp \quoteh_{1} \comp \gdenot{\blank}_{1}
   \end{align*}

### 2021-07-07 Pi+/paper/discussion.tex
done with intro draft for now
Add more text in discussion

--- 2021-06-29 Pi+/paper/discussion.tex
@@
 In this paper, we \ldots
+
+Pi types -- Natural number -- Finite sets
+1-combinator -- Generators of Sn -- 1-paths
+2-combinator -- Relations of Sn -- 2-paths
+
+In HoTT, univalence characterises the path type in the universe as equivalences of types. The map $\term{idtoeqv} : A
+\id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term $\term{ua} : A \eqv B \to A \id_{\UU}
+B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its
+extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p == \term{ua}(\term{idtoeqv(p)})$ are generally added as
+postulates when formalising in Agda. Together, $\term{ua}$ and $\term{ua}-\beta$ give the full univalence axiom $(A \eqv
+B) \eqv (A \id_{\UU} B)$.
+
+% Let's think of the $\PiLang$ combinators as describing the inhabitants of the dentity type of finite types.
+
+By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
+syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse. 
+% By the property of being univalent, this subuniverse gives a model of the univalence axiom. 
+The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
+The $\term{ua}$ map corresponds to synthesing a program from an equivalence (which, in general, is of course
+undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
+we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
+a constructive argument. Then, the computation rule $\term{ua-\beta}$ expresses the fact that program synthesis is
+sound, while $\term{ua-\eta}$ corresponds to the soundess of the equational theory ($\PiLang$ 2-combinators). Thus, we
+this suggests a new computational interpretation of the univalence principle, which provides an intution on why certain
+constructions are hard (or impossible in general case).
+
+We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
+$\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
+doesn't have one, but we could show externally that it satisfies univalence.
 
@@
 \paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of (small)
-symmetric monoidal categories, strong symmetric monoidal functors, and symmetric monoidal natural transformations, to
-the 2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
-$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, and its construction is known
-in the literature~\cite{abramskyAbstractScalarsLoops2005}. Concretely, the objects of $\FSM[\CCat{C}]$ are given by
-lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n] \to \CCat{C}_{0})$. An morphism between $(n,A)$ and
-$(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a permutation of $[n]$, and
-$\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the Grothendieck construction
-$\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and bijections to $\CatCat$, assigning
-each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on $[n]$ inducing an endofunctor on
-$C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one generator, $\FSM[\unit]$. The free
-symmetric monoidal category has been used to study concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri
-nets~\cite{baezCategoriesNets2021}, combinatorial structures~\cite{fioreCartesianClosedBicategory2008}, quantum
-mechanics~\cite{abramskyAbstractScalarsLoops2005}, bicategorical models of (differential) linear
-logic~\cite{melliesTemplateGamesDifferential2019}. In HoTT, the free symmetric monoidal groupoid $\FSM[A]$ over a
-groupoid $A$ can be given by $\dsum{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs.
-These HITs and the proof of their universal property have been considered
-in~\citet*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.
+symmetric monoidal categories, strong symmetric monoidal functors, and monoidal natural transformations, to the
+2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
+$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict)
+symmetric monoidal categories. Its construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.
+Concretely, the objects of $\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n]
+\to \CCat{C}_{0})$. An morphism between $(n,A)$ and $(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a
+permutation of $[n]$, and $\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the
+Grothendieck construction $\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and
+bijections to $\CatCat$, assigning each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on
+$[n]$ inducing an endofunctor on $C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one
+generator, $\FSM[\unit]$. The free symmetric monoidal category has been used to study
+concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri nets~\cite{baezCategoriesNets2021}, combinatorial
+structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005},
+bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}.
+
+Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories was
+studied in~\cite{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a groupoid
+and the proof of its universal property has been considered in~\cite{piceghelloCoherenceMonoidalGroupoids2020}.
+
+Free commutative monoids in type theory have been studied in~\cite{gylterudMultisetsTypeTheory2020}, and using HoTT
+in~\cite{choudhuryFinitemultisetConstructionHoTT2019}. The free symmetric monoidal groupoid $\FSM[A]$ over a groupoid
+$A$ can be given by $\dsum{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs. These HITs
+and the proof of their universal property have been considered
+in~\cite*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.
+
+The proof of the universal property of the $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric
+monoidal categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite
+set. The existence of the proof is folklore, and we have produced a new proof of it while working in constructive type
+theory.
 
@@
 Curry-Howard correspondence to cartesian-closed categories. In this work, we have established a correspondence between a
-fragment of the $\PiLang$ family of languages and symmetric monoidal groupoids.
+fragment of the $\PiLang$ family of reversible programming languages and symmetric monoidal groupoids. The Curry-Howard
+part of this correspondence with reversible logic was established in~\cite{sparksSuperstructuralReversibleLogic2014}.
 
-Univalent Subuniverses, Univalent Typoids.
+\paragraph{Rewriting} We presented a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.
 
-Higher order rewriting theory in constructive type theory.
+There exists an algorithm, due to~\citet{knuthSimpleWordProblems1970}, that, when succeeds, constructs a well-behaved
+rewriting system for an arbitrary finite set of (undirected) equations. It did not work for us, producing too many
+equations, and proving correctness and termination was intractable.
 
-Free monoids, free commutative monoids, normalisation of monoids.
+We chose to encode permutations as adjacent transpositions corresponding to words in $\Sn$, since that closely
+corresponds to the syntax of $\PiLang$ combinators. Other representations of permutations would give a different
+denotational semantics for $\PiLang$ combinators. Permutations can be encoded as listed vectors or matrices, inductively
+generated trees (Motzkin trees), Young diagrams, or String diagrams, but the difficulty of formalising them in type
+theory varies depending on the encoding.
+
+Rewriting systems and word problems have a long history of being formalised in proof assistants. In the recent past,
+higher order rewriting systems have been formalised in proof assistants like {homotopy.io} and Lean and Coq. The use of
+HoTT to study rewriting has been considered in~\cite{krausCoherenceWellFoundednessTaming2020}.
+
+\paragraph{Computational group theory} Coxeter relations are used in computational group theory to study XXX problems.
+
+\paragraph{Univalent Fibrations} Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
+to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
+\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
+$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
+subuniverses.
 
@@
 
-Computational group theory (in HoTT?)
-
-Presentation of the free symmetric monoidal groupoid on one generator, proof of Mac Lane's coherence theorem in HoTT.
-
-Alternatively, we can describe them using HITs, prove they have the universal property and hence equivalent.
-The full Pi is FSMG 1, and the normalised fragment of Pi is M1.
+Formalised proofs of Mac Lane's coherence theorem.
 

### 2021-07-07 Pi+/paper/discussion.tex
Add some ideas for discussion

@@
 
-Pi types -- Natural number -- Finite sets
-1-combinator -- Generators of Sn -- 1-paths
-2-combinator -- Relations of Sn -- 2-paths
-
-In HoTT, univalence characterises the path type in the universe as equivalences of types. The map $\term{idtoeqv} : A
-\id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term $\term{ua} : A \eqv B \to A \id_{\UU}
-B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its
-extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p == \term{ua}(\term{idtoeqv(p)})$ are generally added as
-postulates when formalising in Agda. Together, $\term{ua}$ and $\term{ua}-\beta$ give the full univalence axiom $(A \eqv
-B) \eqv (A \id_{\UU} B)$.
+\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
+types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
+$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to
+    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p ==
+    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
+$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
 
@@
 in~\cite{choudhuryFinitemultisetConstructionHoTT2019}. The free symmetric monoidal groupoid $\FSM[A]$ over a groupoid
-$A$ can be given by $\dsum{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs. These HITs
+$A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs. These HITs
 and the proof of their universal property have been considered
@@
 part of this correspondence with reversible logic was established in~\cite{sparksSuperstructuralReversibleLogic2014}.
+
+\begin{center}
+    \begin{tabular}{c c c}
+        \hline
+        $\PiLang$     & $\Sn$               & $\UFin$     \\
+        \hline
+        types         & Natural numbers     & Finite sets \\
+        \hline
+        1-combinators & Generators of $\Sn$ & 1-paths     \\
+        \hline
+        2-combinators & Relations of $\Sn$  & 2-paths     \\
+        \hline
+    \end{tabular}
+\end{center}
+
+\begin{center}
+    \begin{tabular}{c c c}
+        \hline
+        Logic                              & Program                       & Algebra                     \\
+        \hline
+        Propositions                       & Types                         & Objects                     \\
+        \hline
+        Proofs                             & Terms                         & Morphisms                   \\
+        \hline
+        $\blank$                           & Equations                     & Homotopies                  \\
+        \hline
+        Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
+        \hline
+        Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
+        \hline
+        Connectives                        & Type constructors             & Bifunctors                  \\
+        \hline
+        ?                                  & Type normalization            & ?                           \\
+        \hline
+        ?                                  & Program normalization         & ?                           \\
+        \hline
+        ?                                  & Soundness of 2-combinators    & ?                           \\
+    \end{tabular}
+\end{center}
+
+\begin{center}
+    \begin{tabular}{c c c}
+        \hline
+        Reversible Logic             & Reversible Programming                  & Algebra                  \\
+        \hline
+        Free commutative rig         & Free symmetric rig groupoid             & $\UFin$                  \\
+        \hline
+        Propositions                 & Types                                   & Objects                  \\
+        \hline
+        Entailment?                  & Reversible programs                     & Isomorphisms             \\
+        \hline
+        $\blank$                     & 2-combinators                           & Homotopies               \\
+        \hline
+        Conjunction?                 & $\times$                                & $\otimes$                \\
+        \hline
+        Disjunction?                 & $+$                                     & $\oplus$                 \\
+        \hline
+        $A \land B \vdash B \land A$ & $\swapt$                                & $\mathcal{B}_{A,B}$      \\
+        \hline
+        True proposition             & $\onet$                                 & $F_1$                    \\
+        \hline
+        Decidability                 & Program Synthesis                       & Reachability in a graph? \\
+        \hline
+        Soundness/Completeness       & Soundness/Completeness of 1-combinators & ?                        \\
+    \end{tabular}
+\end{center}
 

### 2021-07-07 Pi+/paper/discussion.tex
intro

--- 2021-07-01 Pi+/paper/discussion.tex
@@
 In this paper, we \ldots
+
+Pi types -- Natural number -- Finite sets
+1-combinator -- Generators of Sn -- 1-paths
+2-combinator -- Relations of Sn -- 2-paths
 
@@
 
+\paragraph{Extensional vs. Intensional View of Permutations.}
+Although the meaning of a permutation - that is, the specific bijective function
+that it represents - is in some sense all there is, manipulating them
+syntactically still has its advantages. By writing reversible programs, we think
+of them in an intensional way. Comparing two programs for equality by evaluating
+them on all points in the domain is a very crude - and maybe even inefficient -
+way of doing that. ~\cite{Kuehlmann:2006:RBR:2298470.2300327,10.1007/978-3-540-24605-3_4,Yamashita:2010:FEQ:1835957.1835965}.
+
+It is crude, because there is no way of enforcing additional constraints on the
+process of transforming one program into another, and no way of inspecting what
+transformation occurred. We can imagine a practical situation in which reversible
+circuits admit one kind of optimization FPGA?, but do not admit another,
+even though they are equivalent in the extensional sense - or maybe one of these
+transformations is cheaper than another, or maybe it is important to know which
+transformation did occur for the producer(?) to focus their resources of
+improving this kind of transformations. By comparing the circuits extensionally,
+the proof object = the path from one into the other - does not have any
+structure,  it is just a check that the values match.
+
+On the other hand, even though in general, the equality of circuits requires
+exponential time jk{triple check}, the proof object can be still very small.
+This creates structure that can possibly be exploited in a heuristic way.
+Additionally, if the goal is to convince third-party that two circuits are
+the same, we can just present the proof instead of showing the equality directly
+- for example, by choosing a particularly short one.
+
+Taking all this into account, it is clear that we must seek a syntactic way of
+computing the meaning of the circuit, instead of just directly evaluating it to
+$\Aut[\Fin[n]]$ - since when we go back and quote the permutation as a circuit
+(in a normal form), there was no "trace" left to see what transformation (what
+sequence of 2-paths) maps the old one into the new one.
+
+We are additionally experimenting with user annotations that can guide
+the search. Each level-2 combinator can be annotated with various
+``cost'' annotations indicating whether it reduces the number of
+gates, reduces the number of choice points, or other cost
+functions. Then one can ask for a proof that takes no more than a
+certain number of steps or a proof that does not create more than a
+certain number of additional wires etc. We illustrate these ideas by
+defining a simple cost function and using it to annotate level-2
+combinators.
+
+We define the \emph{length} $L(c)$ of a composite circuit $c$ as
+follows: the length of a sequential composition of circuits is the sum
+of the lengths of the subcircuits $L(f \odot g) = L(f) + L(g)$; and
+the length of choice or parallel composition is the maximum of either
+branch $L(f \oplus g) = L(f \otimes g) = \max(L(f),L(g))$. For
+primitive gates, the length needs to be postulated to reflect the
+``length'' of the computation involved in applying that primitive. As
+examples, consider the following two level-2 combinators:
+
+% \begin{code}
+% linv◎l'  :  {t₁ t₂ : U} {c : t₁ ⟷ t₂} → (c ◎ ! c) ⇔ id⟷
+% idl◎l'   :  {t₁ t₂ : U} {c : t₁ ⟷ t₂} → (id⟷ ◎ c) ⇔ c
+% \end{code}
+% \AgdaHide{
+% \begin{code}
+% linv◎l'  = ?
+% idl◎l'   = ?
+% \end{code}
+% }
+
+\noindent Assuming that $\idc$ takes a unit length of computation, the
+first can be annotated with $L(c)*2 \isotwo 1$ and the second with $L(c)+1 \isotwo
+L(c)$ indicating that the first combinator reduces the length of the
+circuit from twice the length of $c$ to 1 and the second combinator
+reduces the length of the circuit by 1. Such annotations can then be
+used to constrain or guide the search for transformations between
+circuits.
+
 %%% Local Variables:

### 2021-07-07 Pi+/paper/examples.tex
sec 2

@@
 
-To make the connection to reversible computing explicit, we use the Agda
-formalisation of our technical result to extract various computational
-procedures to synthesise, normalise, optimise, and verify reversible
-circuits. In this introduction, we use the popular IBM Qiskit platform for
-quantum computing to illustrate these ideas. First, most current quantum
-algorithms start with generating superpositions, evolving them using a unitary
-transformation, and then projecting them with a measurement operator. The middle
-stage is essentially a reversible classical computation (executed in a
-quantum-parallel fashion). As a simple example for that middle stage, we
-synthesise a reversible circuit implementing boolean disjunction
-($\vee$). Following~\citet{Toffoli:1980}, the first step is to write a
-specification for the desired reversible function:
-\[
-\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)
-\]
-where $\underline{\vee}$ is the exclusive-or operation.
-From the definition of $\underline{\vee}$ it is evident that setting $h=0$, we can compute the desired disjunction by observing the first component of the result.
-The $\mathit{reversibleOr}$ function has the following truth table (in binary on the left and in a more convenient decimal notation on the right):
+Classical reversible boolean circuits are at the core of most quantum algorithms and
+hence are supported by popular platforms for quantum computing such as IBM
+Qiskit. Specifically, the Qiskit framework provides the following universal set
+of gates for reversible computing: \textsf{not} (boolean negation, called
+\verb|x|), \textsf{cnot} (conditional negation of the second input if the first
+is true; called \verb|cx|), and \textsf{toffoli} (conditional negation of the
+third input if both the first two inputs are true; called \verb|ccx|)
+gates. Additionally, Qiskit allows implicit re-shuffling of bits by allowing
+each operation to specify the indices of its input bits.
 
-\begin{center}\begin{tabular}{|ccc|ccc|@{\qquad\qquad}|c|c|}
-0 & 0 & 0 &     0 & 0 & 0     & 0 & 0 \\
-0 & 0 & 1 &     1 & 0 & 1     & 1 & 5 \\
-0 & 1 & 0 &     1 & 1 & 0    & 2 & 6 \\
-0 & 1 & 1 &     1 & 1 & 1    & 3 & 7 \\
-1 & 0 & 0 &     1 & 0 & 0    & 4 & 4 \\
-1 & 0 & 1 &     0 & 0 & 1    & 5 & 1 \\
-1 & 1 & 0 &     0 & 1 & 0    & 6 & 2 \\
-1 & 1 & 1 &     0 & 1 & 1    & 7 & 3
-\end{tabular}\end{center}
+For concreteness, we demonstrate two different circuits that implement the
+following reversible function specification
+$\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2),
+~b_1, ~b_2)$ where $\vee$ is boolean disjunction and $\underline{\vee}$ is the
+exclusive-or operation. The circuits are presented in both the textual interface
+\verb|qasm| and the graphical interface:
 
-\noindent where it is evident that it is a bijective function, i.e., reversible.
-
-The above embedding of an irreversible function into a reversible function with additional inputs and outputs is
-completely general and is the starting point for specifications of quantum circuits. The challenge is to synthesise a
-program / circuit from this specification. Of course, writing this program in a conventional (irreversible) language
-defeats the purpose. The challenge is to construct the desired program / circuit exclusively using reversible
-primitives, e.g., the standard set of universal reversible gates used in frameworks like Qiskit which consists of the
-computational gates \textsf{not} (boolean negation, called \verb|x|), \textsf{cnot} (conditional negation of the second
-input if the first is true; called \verb|cx|), and \textsf{toffoli} (conditional negation of the third input if both the
-first two inputs are true; called \verb|ccx|) gates, and the ability re-arrange the layout of wires. For concreteness,
-here is a possible implementation of the desired function in Qiskit:
-
-\begin{center}
-  \begin{minipage}[c]{0.4\linewidth}
+\medskip
+\begin{tabular}{c@{\qquad}c}
+\begin{minipage}[t]{0.42\linewidth}
   \begin{verbatim}
-reversibleOr.qasm:
-
-  // setup
   ccx q[1], q[2], q[0];
@@
   cx  q[2], q[0];
-  // measure
-
-% ./qasm -t reversibleOr.qasm
-+-------+-------+
-| 0 0 0 | 0 0 0 |
-| 0 0 1 | 1 0 1 |
-| 0 1 0 | 1 1 0 |
-| 0 1 1 | 1 1 1 |
-| 1 0 0 | 1 0 0 |
-| 1 0 1 | 0 0 1 |
-| 1 1 0 | 0 1 0 |
-| 1 1 1 | 0 1 1 |
-+-------+-------+
   \end{verbatim}
+  \includegraphics[scale=0.7]{reversibleOr.png}
   \end{minipage}
-  \qquad
-  \includegraphics[scale=0.7]{reversibleOr.png}
-\end{center}
-
-\noindent This implementation was manually produced using a standard synthesis algorithm for reversible
-circuits~\cite{10.1145/775832.775915}. To gain some intuition, we trace the evaluation of the circuit for input
-\verb|011|. In this context, the most significant bit is at index 0. Thus the first \verb|ccx| gate negates \verb|q[0]|
-since both \verb|q[1]| and \verb|q[2]| are true producing \verb|111|; the following \verb|cx| gate produces \verb|011|; finally the last \verb|cx| produces the result \verb|111|.
-
-There is wealth of manual and algorithmic approaches for such synthesis problems each optimizing along  different dimensions~\cite{maslov:2003:rls:1087512,1201583}.
-Here is the circuit produced using an approach that analyzes the recursive structure of the circuit (and would generalise to computing the disjunction of more than two inputs):
-
-\begin{center}
-  \begin{minipage}[c]{0.4\linewidth}
+&
+\begin{minipage}[t]{0.43\linewidth}
   \begin{verbatim}
-reversibleOr2.qasm:
-
-  // setup
   cx  q[1], q[0];
@@
   x   q[1];
-  // measure
   \end{verbatim}
+  \includegraphics[scale=0.6]{reversibleOr2.png}
   \end{minipage}
-  \qquad
-  \includegraphics[scale=0.7]{reversibleOr2.png}
-\end{center}
+\end{tabular}
 
-\noindent The evaluation of this circuit prints the same truth table as above confirming their equivalence. Tracing the
-evaluation on the same input value \verb|011| goes through the stages \verb|111|, \verb|101|, \verb|101|, and finally \verb|111|.
+\medskip
 
-The situation for Shor's algorithm is just as described above but with a more involved function of the form $f(r) = a^{r} \mod N$ for fixed $a$ and $N$.
-The specification of the circuit is relatively straightforward to calculate.
-Here it is for $a=11$ and $N=15$:
+There is wealth of manual and algorithmic approaches for producing circuits such
+as the two above~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the left
+was manually produced using a standard synthesis algorithm for reversible
+circuits~\cite{10.1145/775832.775915}. The circuit on the right was produced
+using an approach that analyzes the recursive structure of the circuit (and
+would generalise to computing the disjunction of more than two inputs):
+
+From the specification of the circuit, we expect input \verb|011| to be mapped
+to \verb|111|. To gain some intuition, we trace the evaluation of each circuit for
+input \verb|011|. In this context, the most significant bit is at index 0. In
+the first circuit, the \verb|ccx| gate negates \verb|q[0]| since both \verb|q[1]| and
+\verb|q[2]| are true producing \verb|111|; the following \verb|cx| gate produces
+\verb|011|; finally the last \verb|cx| produces the result \verb|111|. For the
+second circuit, the trace of the execution on the same input value \verb|011| goes through the stages \verb|111|, \verb|101|, \verb|101|, and finally \verb|111|.
+
+Although rather trivial, the examples above illustrate the general idea. A more
+interesting example would be the classical core of Shor's algorithm which
+requires a circuit implementing $f(r) = a^{r} \mod N$ for fixed $a$ and $N$. The
+specification of the circuit is relatively straightforward to calculate. Here it
+is for $a=11$ and $N=15$:
 \[\begin{array}{rcll}
@@
 
-\noindent However, as explained in standard accounts of the algorithm (e.g., the Qiskit implementation), producing an efficient
-modular exponentiation circuit from this specification  is not straightforward and is actually the bottleneck in Shor’s algorithm. Typical
-derivations of the circuit start from elementary gates, build a circuit for reversible disjunction and conjunction of
-booleans, a circuit for a half-adder, a circuit for computing the carry, progressing to a circuit for modular addition,
-which is used to build a circuit for modular multiplication, and then finally a circuit for modular exponentiation
-taking care at each step to avoid the exponential blowup (e.g., by implementing exponentiation by squaring instead of repeated
+\noindent However, as explained in standard accounts of the algorithm (e.g., the
+Qiskit implementation), producing an efficient modular exponentiation circuit
+from this specification is not straightforward and is actually the bottleneck in
+Shor’s algorithm. Typical derivations of the circuit start from elementary
+gates, build a circuit for reversible disjunction (like the two circuits above),
+reversible conjunction, a circuit for a half-adder, a circuit for computing the
+carry, progressing to a circuit for modular addition, which is used to build a
+circuit for modular multiplication, and then finally a circuit for modular
+exponentiation taking care at each step to avoid the exponential blowup (e.g.,
+by implementing exponentiation by squaring instead of repeated
 multiplication)~\cite{shorefficient}.
+
+% To make the connection to reversible computing explicit, we use the Agda
+% formalisation of our technical result to extract various computational
+% procedures to synthesise, normalise, optimise, and verify reversible
+% circuits. In this introduction,
+
+% From the definition of $\underline{\vee}$ it is evident that setting $h=0$, we can compute the desired disjunction by observing the first component of the result.
+The $\mathit{reversibleOr}$ function has the following truth table (in binary on the left and in a more convenient decimal notation on the right):
+
+% \begin{center}\begin{tabular}{|ccc|ccc|@{\qquad\qquad}|c|c|}
+% 0 & 0 & 0 &     0 & 0 & 0     & 0 & 0 \\
+% 0 & 0 & 1 &     1 & 0 & 1     & 1 & 5 \\
+% 0 & 1 & 0 &     1 & 1 & 0    & 2 & 6 \\
+% 0 & 1 & 1 &     1 & 1 & 1    & 3 & 7 \\
+% 1 & 0 & 0 &     1 & 0 & 0    & 4 & 4 \\
+% 1 & 0 & 1 &     0 & 0 & 1    & 5 & 1 \\
+% 1 & 1 & 0 &     0 & 1 & 0    & 6 & 2 \\
+% 1 & 1 & 1 &     0 & 1 & 1    & 7 & 3
+% \end{tabular}\end{center}
+
+% \noindent where it is evident that it is a bijective function, i.e., reversible.
+
+% As a simple example for that middle stage, we synthesise a reversible circuit
+% implementing boolean disjunction ($\vee$). Following~\citet{Toffoli:1980}, the
+% first step is to write a specification for the desired reversible function:
+
+% The above embedding of an irreversible function into a reversible function with
+% additional inputs and outputs is completely general and is the starting point
+% for specifications of quantum circuits. The challenge is to synthesise a program
+% / circuit from this specification. Of course, writing this program in a
+% conventional (irreversible) language defeats the purpose. The challenge is to
+% construct the desired program / circuit exclusively using reversible primitives,
+% e.g.,
+
+% the standard set of universal reversible gates used in frameworks like
+% Qiskit which consists of the computational gates \textsf{not} (boolean negation,
+% called \verb|x|), \textsf{cnot} (conditional negation of the second input if the
+% first is true; called \verb|cx|), and \textsf{toffoli} (conditional negation of
+% the third input if both the first two inputs are true; called \verb|ccx|) gates,
+% and the ability re-arrange the layout of wires.

### 2021-07-07 Pi+/paper/equivalence.tex
Change and fix references

@@
 
-We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\cref*{fig:piplus,fig:pihat} respectively,
+We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\Cref*{fig:piplus,fig:pihat} respectively,
 eliding the 2-combinators for brevity. \note{We enforce that there is a unique 2-combinator between compatible
@@
 First, we show how to translate $\PiLang$ programs to $\PiPlusLang$, which is the additive fragment of $\PiLang$. The
-syntax is given in~\cref{fig:piplus}.
+syntax is given in~\Cref{fig:piplus}.
 
@@
 $\PiPlusLang$, with (unary) natural numbers for 0-cells, 1-combinators generated by adjacent transpositions, and an
-appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators, in~\cref{fig:pihat}.
+appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators, in~\Cref{fig:pihat}.
 
@@
 component. In $\PiHatLang$, the 1-combinators are generated by adjacent transpositions, so these can be easily mapped to
-words in $\Sn$ and then to automorphisms using~\cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to
+words in $\Sn$ and then to automorphisms using~\Cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to
 2-paths between loops in $\UFin$.
@@
 The semantics that we presented here takes a different route to constructing the permutation from a $\PiLang$
-combinator, compared to the direct interpretation given using the big-step interpreter in~\cref{subsec:denotational}. We
+combinator, compared to the direct interpretation given using the big-step interpreter in~\Cref{subsec:denotational}. We
 verify that the two semantics agree, establishing that the semantics is adequate and fully abstract.

### 2021-07-07 Pi+/paper/finite.tex
Change and fix references

@@
 % Recap of the previous section
-In~\cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
+In~\Cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
 $n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group. This is
@@
 correspondence between them and the type $\Aut[\Fin[n]]$ of automorphisms on a finite set. The generators and relations
-we use here will be used to quote back to 1 and 2-combinators in $\PiLang$ (see~\cref{sec:equivalence}).
+we use here will be used to quote back to 1 and 2-combinators in $\PiLang$ (see~\Cref{sec:equivalence}).
 
@@
 also use to solve the word problem for $\Sn$. Connecting it back to the language $\PiLang$, we will show that it is
-possible to encode all $\PiLang$ combinators using adjacent transpositions (in~\cref{sec:equivalence}). The rest of this
+possible to encode all $\PiLang$ combinators using adjacent transpositions (in~\Cref{sec:equivalence}). The rest of this
 section develops these intuitions formally. Group presentations are built by adding equations to a \emph{free group},
@@
   $\blank\mult\blank$ for group multiplication, and a unary inverse operation $\inv{\blank}$. The neutral element has to
-  satisfy unit and inverse laws, and the multiplication has to be associative (see~\cref{def:group}).
+  satisfy unit and inverse laws, and the multiplication has to be associative (see~\Cref{def:group}).
 
@@
     Given a set $A$, the free group $F(A)$ on it is given by a higher inductive type with the following point and path
-    constructors. Notice the similarity with the definition of a group structure (\cref{def:group}), but note that each
+    constructors. Notice the similarity with the definition of a group structure (\Cref{def:group}), but note that each
     operation here is a generator for $F(A)$.
@@
 Following the universal-algebraic definition, in HoTT, we could use a naive higher inductive type to define the free
-group, which enforces the group axioms by adding path constructors (see~\cref{def:free-group}).
+group, which enforces the group axioms by adding path constructors (see~\Cref{def:free-group}).
 Using the induction principle, we can easily verify the universal property. However, since this definition of $F(A)$ has
@@
   reversing the list. Further, $F(A)$ with $\eta_A(a) \defeq \inl(a) \cons \nil$ satisfies the universal property of
-  free groups, as stated in~\cref{prop:free-groups}.
+  free groups, as stated in~\Cref{prop:free-groups}.
 \end{proposition}
@@
 
-The universal property of the above definition is similar to~\cref{prop:free-groups} except the relation has to be
+The universal property of the above definition is similar to~\Cref{prop:free-groups} except the relation has to be
 preserved by the function mapping out of the generating set.
@@
 to specify for this presentation -- these are the equations that these generators should satisfy. We visualise them as
-braid diagrams in~\cref{fig:coxeter-braid}.
+braid diagrams in~\Cref{fig:coxeter-braid}.
 
@@
 This construction is called a Coxeter presentation of $\Sn$. Writing it formally, we encode the rules discussed above
-using a relation $\cox$ on $\List[\Fin[n]]$ (\cref{def:cox}), and then take its congruence closure $\cox*$
-(\cref{def:coxstar}).
+using a relation $\cox$ on $\List[\Fin[n]]$ (\Cref{def:cox}), and then take its congruence closure $\cox*$
+(\Cref{def:coxstar}).
 
@@
 rewriting system is confluent -- meaning that all critical pairs, that is, terms with overlapping possible reduction
-rules, have to converge. For example, in our system, the pairs in~\cref{fig:critical-pairs-converging} converge.
-Unfortunately, this is not true for all critical pairs -- an  example is in~\cref{fig:critical-pairs-non-converging},
+rules, have to converge. For example, in our system, the pairs in~\Cref{fig:critical-pairs-converging} converge.
+Unfortunately, this is not true for all critical pairs -- an  example is in~\Cref{fig:critical-pairs-non-converging},
 where both endpoints are normal with respect to the $\cox*$ relation.
@@
 is to produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ by $k$ places left,
-shifting all the elements in between one place right (see~\cref{fig:downf}).
+shifting all the elements in between one place right (see~\Cref{fig:downf}).
 
@@
 sequence of transpositions and a single transposition afterwards. For example, for $n = 0$ and $k = 3$, it allows for
-the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox}).
+the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\Cref{fig:longcox}).
 
@@
 We define the relation $\longcox*$ to be the closure of $\longcox$, under reflexivity and (right-step extended)
-transitivity (see~\cref{def:longcox-star}).
+transitivity (see~\Cref{def:longcox-star}).
 
@@
 
-By~\cref{prop:nf}~\ref{prop:nf-uniqueness}, we get a unique choice function $\normf : {\List[\Fin[n]]} \to
+By~\Cref{prop:nf}~\ref{prop:nf-uniqueness}, we get a unique choice function $\normf : {\List[\Fin[n]]} \to
   {\List[\Fin[n]]}$ that produces a normal form for terms of $\List[\Fin[n]]$. We state and prove a couple of important
@@
 
-Notice however, that a group presentation, as defined in~\cref{def:presentation}, requires the relation to be on the set
+Notice however, that a group presentation, as defined in~\Cref{def:presentation}, requires the relation to be on the set
 of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The constructor
@@
       \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                          \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
-      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                          & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}}
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\Cref{prop:fin-finexcept}~\Cref{prop:fin-finexcept-2}} \\
+      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\Cref{prop:fin-finexcept}~\Cref{prop:fin-finexcept-2}} \\
+      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                          & \text{by~\Cref{prop:fin-finexcept}~\Cref{prop:fin-finexcept-3}}
     \end{array}
@@
 
-By composing~\cref{prop:lehmer-aut-equiv} and~\cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.
+By composing~\Cref{prop:lehmer-aut-equiv} and~\Cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.
 
 \begin{corollary}
+  \label{prop:sn-lehmer-fin-equiv}
   For all $n : \Nat$,

### 2021-07-07 Pi+/paper/ufin.tex
Change and fix references

@@
 we formally define \emph{univalent subuniverses}, and proceed to construct the particular instance for finite types,
-$\UFin$ (\cref{def:ufin}).
+$\UFin$ (\Cref{def:ufin}).
 
@@
 
-  \note{We use the rig structure of $\UFin$ in~\cref{subsec:rig} to interpret $\PiLang$.}
+  \note{We use the rig structure of $\UFin$ in~\Cref{subsec:rig} to interpret $\PiLang$.}
 
@@
   of equalities or identifications between them. The identity type is generated by reflexivity $\refl_{x} : x \id_{A} x$,
-  and the eliminator for the identity type is given by path induction or the $J$-rule (\cref{def:path-induction}). This
+  and the eliminator for the identity type is given by path induction or the $J$-rule (\Cref{def:path-induction}). This
   construction can be iterated, giving the identity type between two terms of an identity type, repeating ad infinitum.
@@
   projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
-  fibration, that is, there is a lifting operation (\cref{fig:lift}) which lifts paths in the base space to paths in the
+  fibration, that is, there is a lifting operation (\Cref{fig:lift}) which lifts paths in the base space to paths in the
   total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have
@@
   $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$, which equates every pair of points in the truncation
-  (see~\cref{def:prop-trunc}).
+  (see~\Cref{def:prop-trunc}).
 
@@
     \UU$. It has an inclusion of points $q : A \to \quot{A}{R}$, and adds paths between related pairs of elements
-  $\quotrel : R(x,y) \to q(x) \id_{\quot{A}{R}} q(y)$ (see~\cref{def:set-quot}).  We recall that the quotient is
+  $\quotrel : R(x,y) \to q(x) \id_{\quot{A}{R}} q(y)$ (see~\Cref{def:set-quot}).  We recall that the quotient is
   \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff $(q(x) \id_{\quot{A}{R}}
@@
 subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness -- a type is finite if it is
-merely equivalent to a finite set (\cref{def:finite-set}), and we define the predicate $\isFin$ (\cref{def:isfin}) to do
+merely equivalent to a finite set (\Cref{def:finite-set}), and we define the predicate $\isFin$ (\Cref{def:isfin}) to do
 so.
@@
       \Fin[0]                & \eqv \bot        \\
-      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m] \\
+      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]
     \end{aligned}
@@
       \Fin[1]                & \eqv \top        \\
-      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m] \\
+      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]
     \end{aligned}
@@
       (X \sqcup Y) \sqcup Z & \eqv X \sqcup (Y \sqcup Z) \\
-      X \sqcup Y            & \eqv Y \sqcup Y            \\
+      X \sqcup Y            & \eqv Y \sqcup Y
     \end{aligned}

### 2021-07-07 Pi+/paper/ufin.tex
Squeeze some space in equations
Change inline sigma notation

@@
 
-  The type $\dsum{x:A}{P(x)}$ is the collection of all the fibers and is called the total space of $P$. The first
-  projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
+  The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is called the total space of $P$. The first
+  projection ${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
   fibration, that is, there is a lifting operation (\Cref{fig:lift}) which lifts paths in the base space to paths in the
@@
 Functions between types are functors between groupoids, and type families, or functions to the universe, are indexed
-families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum{x:A}{P(x)} \to \UU$
+families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to \UU$
 which has a lifting operation giving it the structure of a fibration (see the appendix for details). The
@@
 \begin{definition}[Univalent Fibration]
-  $P$ is a univalent type family (or, $\fst : {\dsum{x:A}{P(x)}} \to A$ is a univalent fibration) if $\tptEqv{P}$ is an
+  $P$ is a univalent type family (or, $\fst : {\dsum*{x:A}{P(x)}} \to A$ is a univalent fibration) if $\tptEqv{P}$ is an
   equivalence.
@@
 The use of the word \emph{univalent} here is a reference to Voevodsky's \emph{univalence} principle. Indeed, univalence
-characterises paths in the universe as equivalences between types, which follows from the canonical fibration $\idfunc :
-  \UU \to \UU$ being univalent.
+characterises paths in the universe as equivalences between types, which follows from the canonical fibration
+$\idfunc : \UU \to \UU$ being univalent.
 
@@
 \begin{proposition}[Univalent Subuniverse]
-  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are prop-valued, that is, $P(X)$ is a proposition
-  for every $X$. The fibration ${\fst : \dsum{X:\UU}{P(X)} \to \UU}$ is univalent. This generates a univalent
-  subuniverse
+  A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
+  proposition for every $X$. The fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent. This generates a
+  univalent subuniverse
   \[
@@
 \begin{appendixproof}
-  Suppose $(U, \El) \defeq (\dsum{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
+  Suppose $(U, \El) \defeq (\dsum*{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
   any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
@@
   For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.
-  Note that, $\UFin$ can be equivalently seen as the collection of all types of finite cardinality, that is, $\UFin \eqv
-    \dsum{n:\Nat}{\UFin[n]}$.
+  Note that, $\UFin$ can be equivalently seen as the collection of all types of finite cardinality, that is,
+  $\UFin \eqv \dsum*{n:\Nat}{\UFin[n]}$.
 \end{definition}
@@
       (X \sqcup Y) \sqcup Z & \eqv X \sqcup (Y \sqcup Z) \\
-      X \sqcup Y            & \eqv Y \sqcup Y
+      X \sqcup Y            & \eqv Y \sqcup Y            \\
+      X \times \bot         & \eqv \bot
     \end{aligned}
@@
       X \times Y            & \eqv Y \times X                       \\
-      X \times \bot         & \eqv \bot                             \\
       X \times (Y \sqcup Z) & \eqv (X \times Y) \sqcup (X \times Z)

### 2021-07-07 Pi+/paper/finite.tex
Align fig 3
Resize braid diagrams

@@
   \centering
-  \begin{subfigure}[b]{0.3\textwidth}
+  \begin{subfigure}[b]{0.25\textwidth}
     \centering
@@
         number of strands = 2,
-        width = 0.6cm,
+        width = 0.4cm,
         height = 0.4cm,
@@
       {braid={ s_1, s_1}};
-      \node[font=\Huge] at (1.75, 0.7) {\(\rightarrow\)};
+      \node[font=\large] at (1, 0.7) {\(\rightarrow\)};
       \pic[local bounding box=my braid,braid/.cd,
         number of strands = 2,
-        width = 0.6cm,
+        width = 0.4cm,
         height = 0.4cm,
         border height = 0.3cm,
-        thick] at (3, 0)
+        thick] at (1.6, 0)
       {braid={ 1, 1}};
@@
   \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{0.5\textwidth}
+  \begin{subfigure}[b]{0.4\textwidth}
     \centering
@@
       {braid={ s_1, 1}};
-      \node[] at (1.1, 0.7) {\(\dots\)};
+      \node[] at (0.7, 0.7) {\(\dots\)};
       \pic[local bounding box=my braid,braid/.cd,
         number of strands = 2,
-        width = 0.6cm,
+        width = 0.4cm,
         height = 0.4cm,
         border height = 0.3cm,
-        thick] at (1.5, 0)
+        thick] at (1, 0)
       {braid={ 1, s_1 }};
-      \node[font=\Huge] at (3, 0.7) {\(\rightarrow\)};
+      \node[font=\large] at (1.9, 0.7) {\(\rightarrow\)};
       \pic[local bounding box=my braid,braid/.cd,
         number of strands = 2,
-        width = 0.6cm,
+        width = 0.4cm,
         height = 0.4cm,
         border height = 0.3cm,
-        thick] at (4, 0)
+        thick] at (2.4, 0)
       {braid={ 1, s_1}};
-      \node[] at (5.1, 0.7) {\(\dots\)};
+      \node[] at (3.1, 0.7) {\(\dots\)};
       \pic[local bounding box=my braid,braid/.cd,
         number of strands = 2,
-        width = 0.6cm,
+        width = 0.4cm,
         height = 0.4cm,
         border height = 0.3cm,
-        thick] at (5.5, 0)
+        thick] at (3.4, 0)
       {braid={ s_1, 1}};
@@
   \end{subfigure}
-  \hfill
-  \begin{subfigure}[b]{\textwidth}
+  \begin{subfigure}[b]{0.25\textwidth}
     \centering
@@
         number of strands = 3,
-        width = 0.6cm,
+        width = 0.4cm,
         height = 0.4cm,
@@
       {braid={ s_2, s_1, s_2}};
-      \node[font=\Huge] at (2.15, 0.7) {\(\rightarrow\)};
+      \node[font=\large] at (1.4, 0.7) {\(\rightarrow\)};
       \pic[local bounding box=my braid,braid/.cd,
         number of strands = 3,
-        width = 0.6cm,
+        width = 0.4cm,
         height = 0.4cm,
         border height = 0.1cm,
-        thick] at (3, 0)
+        thick] at (2, 0)
       {braid={ s_1, s_2, s_1}};
@@
   \centering
-  \begin{subfigure}[b]{0.45\textwidth}
+  \begin{subfigure}[b]{0.35\textwidth}
     % https://q.uiver.app/?q=WzAsNixbMSwwLCJcXHRhdV8yXFx0YXVfMVxcdGF1XzJcXHRhdV8xXFx0YXVfMiJdLFswLDEsIlxcdGF1XzFcXHRhdV8yXFx0YXVfMVxcdGF1XzFcXHRhdV8yIl0sWzIsMSwiXFx0YXVfMlxcdGF1XzFcXHRhdV8xXFx0YXVfMlxcdGF1XzEiXSxbMCwyLCJcXHRhdV8xXFx0YXVfMlxcdGF1XzIiXSxbMSwzLCJcXHRhdV8xIl0sWzIsMiwiXFx0YXVfMlxcdGF1XzJcXHRhdV8xIl0sWzAsMSwiYnJhaWQiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFswLDIsImJyYWlkIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbMSwzLCJjYW5jZWwiLDIseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dLFszLDQsImNhbmNlbCIsMix7InN0eWxlIjp7ImJvZHkiOnsibmFtZSI6InNxdWlnZ2x5In19fV0sWzIsNSwiY2FuY2VsIiwwLHsic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoic3F1aWdnbHkifX19XSxbNSw0LCJjYW5jZWwiLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJzcXVpZ2dseSJ9fX1dXQ==
@@
     }
-    \caption{Converging critical pairs}
+    \caption{A converging critical pair in $\cox*$}
     \label{fig:critical-pairs-converging}
   \end{subfigure}
-  \begin{subfigure}[b]{0.50\textwidth}
+  \begin{subfigure}[b]{0.55\textwidth}
     \adjustbox{scale=\scalef,center}{
@@
 The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\). Since we think of the
-number $m$ as the transposition swapping elements $(\el{m}, \el{m+1})$, the role of this helper function
-is to produce a sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ by $k$ places left,
-shifting all the elements in between one place right (see~\Cref{fig:downf}).
-
-\begin{figure}
-  \[
-    \begin{tikzpicture}
-      \pic[local bounding box=my braid,braid/.cd,
-        number of strands = 5,
-        width = 0.6cm,
-        height = 0.4cm,
-        border height = 0.1cm,
-        thick]
-      {braid={s_1, s_2, s_3, s_4}};
-    \end{tikzpicture}
-  \]
-  \caption{Braiding diagram for $0 \downf 4$}
-  \label{fig:downf}
-\end{figure}
-
-Then, the directed relation $\longcox$ is defined with the following generators.
+number $m$ as the transposition swapping elements $(\el{m}, \el{m+1})$, the role of this helper function is to produce a
+sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ by $k$ places left, shifting all the
+elements in between one place right (see~\Cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the
+following generators.
 
@@
     \longswap
-     & : \forall k, n, l, r \to (\suc[k] < n) \to (l \append n \cons k \cons r) \longcox (l \append k \cons n \append r)                            \\
+     & : \forall n, k, l, r \to (\suc[k] < n) \to (l \append n \cons k \cons r) \longcox (l \append k \cons n \append r)                               \\
     \longbraid
-     & : \forall n, l, r \to (l \append (n \downf 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downf 2 + k) \append r)
+     & : \forall n, k, l, r \to (l \append (n \downf 2 + k) \append (1 + k + n) \cons r) \longcox (l \append (k + n) \cons (n \downf 2 + k) \append r)
   \end{align*}
@@
 \begin{figure}
+  \begin{subfigure}[b]{0.3\textwidth}
+    \centering
+    \begin{tikzpicture}
+      \pic[local bounding box=my braid,braid/.cd,
+        number of strands = 7,
+        width = 0.4cm,
+        height = 0.4cm,
+        border height = 0.1cm,
+        thick]
+      {braid={s_1, s_2, s_3, s_4, s_5, s_6}};
+    \end{tikzpicture}
+    \caption{$0 \downf 6$}
+    \label{fig:downf}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.6\textwidth}
+    \centering
     \begin{tikzpicture}
@@
         number of strands = 6,
-      width = 0.6cm,
+        width = 0.4cm,
         height = 0.4cm,
@@
       {braid={s_5, s_1, s_2, s_3, s_4, s_5}};
-    \node[font=\Huge] at (4, 1.3) {\(\rightarrow\)};
+      \node[font=\large] at (3, 1.3) {\(\rightarrow\)};
       \pic[local bounding box=my braid,braid/.cd,
         number of strands = 6,
-      width = 0.6cm,
+        width = 0.4cm,
         height = 0.4cm,
         border height = 0.1cm,
-      thick] at (5, 0)
+        thick] at (4, 0)
       {braid={s_1, s_2, s_3, s_4, s_5, s_4}};
     \end{tikzpicture}
-  \caption{Braiding diagram for $\longbraid$ with $n = 0$ and $k = 3$}
+    \caption{$\longbraid$ with $n = 0$ and $k = 3$}
     \label{fig:longcox}
+  \end{subfigure}
+  \caption{Braiding diarams for modified Coxeter relations.}
+  \label{fig:coxeter-braid-mod}
 \end{figure}

### 2021-07-07 Pi+/paper/finite.tex
Add labels to braid diagram

@@
 Unfortunately, this is not true for all critical pairs -- an  example is in~\Cref{fig:critical-pairs-non-converging},
-where both endpoints are normal with respect to the $\cox*$ relation.
+where left and right endpoints are normal with respect to the $\cox*$ relation.
 
@@
     }
-    \caption{A non-converging critical pair in $\cox*$ converging in $\longcox*$}
+    \caption{A non-converging critical pair in $\cox*$ converges with $\longbraid$}
     \label{fig:critical-pairs-non-converging}
@@
     \begin{tikzpicture}
+      \def\nstrandsdf{7}
       \pic[local bounding box=my braid,braid/.cd,
-        number of strands = 7,
+        number of strands = \nstrandsdf,
         width = 0.4cm,
@@
         border height = 0.1cm,
-        thick]
+        thick,
+        name prefix=braid]
       {braid={s_1, s_2, s_3, s_4, s_5, s_6}};
+      % \draw[thick] % draws the top/bottom bars
+      % ([xshift=-1ex]my braid.north west) --  ([xshift=1ex]my braid.north east)
+      % ([xshift=-1ex]my braid.south west) --  ([xshift=1ex]my braid.south east);
+      \foreach \n in {1,...,\nstrandsdf}{
+          \pgfmathtruncatemacro{\nminusone}{\n - 1}
+          \node at (braid-\n-s)[yshift = 2.8cm] {\nminusone};
+        }
+      % labels the bottom bar
+      \node at (braid-1-e)[yshift = -2.8cm] {5};
+      \node at (braid-2-e)[yshift = -2.8cm] {6};
+      \node at (braid-3-e)[yshift = -2.8cm] {0};
+      \node at (braid-4-e)[yshift = -2.8cm] {1};
+      \node at (braid-5-e)[yshift = -2.8cm] {2};
+      \node at (braid-6-e)[yshift = -2.8cm] {3};
+      \node at (braid-7-e)[yshift = -2.8cm] {4};
     \end{tikzpicture}
@@
   \end{subfigure}
+  \hfill
   \begin{subfigure}[b]{0.6\textwidth}
@@
     \begin{tikzpicture}
+      \def\nstrandsbl{6}
       \pic[local bounding box=my braid,braid/.cd,
-        number of strands = 6,
+        number of strands = \nstrandsbl, % number of  strands
         width = 0.4cm,
@@
         border height = 0.1cm,
-        thick] at (0, 0)
+        thick,
+        name prefix=braid] at (0, 0)
       {braid={s_5, s_1, s_2, s_3, s_4, s_5}};
       \node[font=\large] at (3, 1.3) {\(\rightarrow\)};
+      % \draw[thick] % draws the top/bottom bars
+      % ([xshift=-1ex]my braid.north west) --  ([xshift=1ex]my braid.north east)
+      % ([xshift=-1ex]my braid.south west) --  ([xshift=1ex]my braid.south east);
+      \foreach \n in {1,...,\nstrandsbl}{
+          \pgfmathtruncatemacro{\nminusone}{\n - 1}
+          \node at (braid-\n-s)[yshift = 2.8cm] {\nminusone};
+        }
+      % labels the bottom bar
+      \node at (braid-1-e)[yshift = -2.8cm] {3};
+      \node at (braid-2-e)[yshift = -2.8cm] {5};
+      \node at (braid-3-e)[yshift = -2.8cm] {0};
+      \node at (braid-4-e)[yshift = -2.8cm] {1};
+      \node at (braid-5-e)[yshift = -2.8cm] {4};
+      \node at (braid-6-e)[yshift = -2.8cm] {2};
+
+      \def\nstrandsbr{6}
       \pic[local bounding box=my braid,braid/.cd,
-        number of strands = 6,
+        number of strands = \nstrandsbr,
         width = 0.4cm,
@@
         border height = 0.1cm,
-        thick] at (4, 0)
+        thick,
+        name prefix=braid] at (4, 0)
       {braid={s_1, s_2, s_3, s_4, s_5, s_4}};
+      % \draw[thick] % draws the top/bottom bars
+      % ([xshift=-1ex]my braid.north west) --  ([xshift=1ex]my braid.north east)
+      % ([xshift=-1ex]my braid.south west) --  ([xshift=1ex]my braid.south east);
+      \foreach \n in {1,...,\nstrandsbl}{
+          \pgfmathtruncatemacro{\nminusone}{\n - 1}
+          \node at (braid-\n-s)[yshift = 2.8cm] {\nminusone};
+        }
+      % labels the bottom bar
+      \node at (braid-1-e)[yshift = -2.8cm] {3};
+      \node at (braid-2-e)[yshift = -2.8cm] {5};
+      \node at (braid-3-e)[yshift = -2.8cm] {0};
+      \node at (braid-4-e)[yshift = -2.8cm] {1};
+      \node at (braid-5-e)[yshift = -2.8cm] {4};
+      \node at (braid-6-e)[yshift = -2.8cm] {2};
     \end{tikzpicture}
-    \caption{$\longbraid$ with $n = 0$ and $k = 3$}
+    \caption{ $\longbraid$ with $n = 0$ and $k = 3$}
     \label{fig:longcox}

### 2021-07-07 Pi+/paper/finite.tex
Start revising introduction

@@
 sequence of transpositions and a single transposition afterwards. For example, for $n = 0$ and $k = 3$, it allows for
-the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\Cref{fig:longcox}).
+the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\Cref{fig:longcox} -- note the distinction between
+wires, where numbers represent the values, and transpositions, where numbers represent which wires are crossing).
 

### 2021-07-08 Pi+/paper/reversible1.tex
sec 2
Change and fix references
sec 2 and sec 3

@@
 
-The circuit model of reversible computation discussed in the introduction is a useful abstraction close to the hardware
+The circuit model of reversible computation discussed in the previous section is a useful abstraction close to the hardware
 platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly-level''
@@
 
-In a reversible boolean circuit, the number of input bits matches the number of output bits. Circuits can be composed
-sequentially, vertically, or horizontally, respectively preserving the number of bits, adding up the number of bits or
-multiplying them. So, to program with reversible circuits, all we need to do is make sure that each primitive operation
-preserves the number of bits, which is just a natural number. Natural numbers are the free commutative semiring (or,
-commutative rig) on one generator, with $(0,+)$ for addition, and $(1,\times)$ for multiplication, so the commutative
-semirig identities can be used to design a logic for reversible programming. This inspired the syntax for various
-first-order reversible logics and reversible programming
-languages~\cite*{sparksSuperstructuralReversibleLogic2014,jamesInformationEffects2012}.
+In reversible boolean circuits, the number of input bits matches the number of output bits. Thus, a key insight for a
+programming language of reversible circuits is to ensure that each primitive operation preserves the number of bits,
+which is just a natural number. The algebraic structure of natural numbers as the free commutative semiring (or,
+commutative rig) on one generator, with $(0,+)$ for addition, and $(1,\times)$ for multiplication then provides
+sequential, vertical, and horizontal circuit composition operators.
 
-Natural number identities have no computational content. To obtain a computational interpretation of the commutative rig
-structure, the logic needs to be equipped with values and types, and a notion of operational semantics and contextual
-equivalence~\cite{jamesInformationEffects2012}. This is a programming language which embodies the computational content
-of isomorphisms of finite types, or permutations. On the semantic side, we need to consider the groupoidification of a
-commutative rig, that is, a symmetric rig groupoid. A sound semantics for $\PiLang$ in weak rig groupoids was
-established in~\cite{caretteComputingSemiringsWeak2016}, and conjectured to be complete.
+These commutative semirig identities can be used to design a logic for reversible
+programming~\cite*{sparksSuperstructuralReversibleLogic2014}. But as natural number identities have no computational
+content, the logic needs to be equipped with values and types, and a notion of operational semantics and contextual
+equivalence to obtain a computational interpretation of the commutative rig
+structure~\cite{jamesInformationEffects2012}.  On the semantic side, the natural space to consider is the
+groupoidification of a commutative rig, that is, a symmetric rig groupoid.
 
-% In programming parlance, as shown in~\cref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes
+Putting these ideas together, the programming language $\PiLang$, whose syntax is given below, embodies the
+computational content of isomorphisms of finite types, or permutations.
+
+% In programming parlance, as shown in~\Cref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes
 % a reversible combinator in the programming language and the algebraic closure operators become composition operators in
@@
 the case when the boolean is \textsf{true}, the combinator~\ensuremath{c_1} is applied to the value of
-type~\ensuremath{A}. The right branch which corresponds to the boolean being \textsf{false} passes the value of type $A$
+type~\ensuremath{A}. The right branch, which corresponds to the boolean being \textsf{false}, passes the value of type $A$
 through the combinator \ensuremath{c_2}.  The inverse of \ensuremath{\dist}, namely \ensuremath{\factor} is applied to
 get the final result. Using this conditional operator, \verb|cx| is defined as $\Afun{cif}~\verb|x|~\idc$ and
-\verb|ccx| is defined as $\Afun{cif}~\verb|cx|~\idc$. With these conventions, the first circuit in the introduction
+\verb|ccx| is defined as $\Afun{cif}~\verb|cx|~\idc$. With these conventions, the first circuit in the previous section
 is transcribed as follows:
@@
 
+\addertwo{}
+
 \noindent where we clearly see the sequences of the three operations \verb|ccx|, \verb|cx|, and \verb|cx| but, instead
 of using the indices in the sequence of wires to identify the relevant parameters, here we use structural isomorphisms
-(elided above) to re-shuffle the types. For the second circuit, instead of transcribing it directly, we express it using
+to re-shuffle the types. We only show one of these re-shuffling isomorphisms and elide the others. For the second circuit, instead of transcribing it directly, we express it using
 a slightly more abstract notation:
@@
 
-All of this is only half the story however as the correspondence between the syntactic commutative rig (the syntax of $\PiLang$)
-and permutations on finite sets includes \emph{coherence conditions} that identify different syntactic representations
-of the same permutation~\cite{laplaza72,Carette2016}. These coherence conditions are collected in a second level of $\PiLang$
-syntax as level-2 combinators whose types are of the form $c_1 \isotwo c_2$ for appropriate $c_1$ and $c_2$ of the same
-level-1 type $A \isoone B$. For example, we have:
+All of this is only half the story, however. A sound semantics for $\PiLang$ in weak rig
+groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
+semantics, rig groupoid \emph{coherence conditions} that identify different syntactic representations of the same
+permutation~\cite{laplaza72,Carette2016} were collected  in a
+second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
+$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For example, we have:
 
@@
 %%%%%%%%%
-\subsection{Denotational Semantics}
+\subsection{Semantics}
+\label{subsec:denotational}
 
-\noindent The accompanying code includes an operational semantics for $\Pi$. A denotational semantics can be specified
-by mapping each type to a finite set and each combinator to a permutation between finite sets. In this section, we
-outline a simple denotational semantics expressed in the metalanguage of sets and functions that collapses the level-2
-combinators. This semantics will be extended in the next two sections to the HoTT metalanguage to properly model the
-level-2 combinators.
+\noindent Below we present a semantic map that dually serves as an operational semantics and a denotational
+semantics. As an operational semantics, each $\PiLang$ type $A$ is mapped to an Agda type $\denot{A}$ and each
+combinator $c : A \isoone B$ is mapped to an Agda function $\denot{A} \to \denot{B}$. As a denotational semantics, each
+$\PiLang$ type $A$ is mapped to a finite set $\denot{A}$ and each combinator $c : A \isoone B$ is mapped to bijection
+between $\denot{A} \to \denot{B}$.
 
-The denotation of types is straightforward:
+The semantics of types is straightforward:
 
@@
 \noindent where $\bot$ is the empty set, $\top$ is a singleton set, $\sqcup$ is the disjoint union of sets, and $\times$
-is the cartesian product of sets. The denotation of a combinator $c : A \isoone B$ is a (bijective) function mapping
+is the cartesian product of sets. The semantics of a combinator $c : A \isoone B$ is a (bijective) function mapping
 $\denot{A}$ to $\denot{B}$:
@@
 \begin{theorem}\label{thm:semone}
-  The denotational semantics is sound in the following sense:
+  The semantics is sound in the following sense:
   \begin{itemize}
@@
 
-Using categorical language, the denotational semantics for $\PiLang$ is using the category of finite sets and functions
-$\SetFin$. However, we only use the bijective functions for the semantics, which means, we use the groupoid $\BFin =
+Using categorical language, the denotations of types and combinators are objects and morphisms in the category of finite sets and functions
+$\SetFin$. However, as we only use the bijective functions for the semantics, we use the groupoid $\BFin =
   \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and finite

### 2021-07-08 Pi+/paper/reversible1.tex
working on ufin

@@
 
-Using categorical language, the denotations of types and combinators are objects and morphisms in the category of finite sets and functions
-$\SetFin$. However, as we only use the bijective functions for the semantics, we use the groupoid $\BFin =
-  \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and finite
-products $(\unit, \times)$, which we use to interpret our types. In $\BFin$, the coproducts and products restrict to
-additive and multiplicative symmetric monoidal structures, respectively, making $\BFin$ a symmetric rig groupoid.
-
 %%% Local Variables:

### 2021-07-08 Pi+/paper/ufin.tex
working on ufin

@@
 % Second - say what was wrong
-To recap, the denotational semantics of $\PiLang$ interprets types as objects in $\BFin$, 1-combinators as isomorphisms,
-and we prove that for every pair of 1-combinators related by a 2-combinator, their interpretations in~$\BFin$ are equal.
-The groupoid $\BFin$ is strict, since the collection of isomorphisms is a set, that is, a discrete category. There is no
-explicit witness for the equality of two isomorphisms, since we can decide by evaluating two bijections whether they're
-equal. To be able to establish completeness for $\PiLang$, we want a witness for this equality, so that we can quote
-back to the syntax and produce a 2-combinator witnessing the equality of the corresponding 1-combinators.
+
+Using categorical language, the setting for the semantics in the previous section is the category of finite sets and
+functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
+$\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
+finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
+structures, respectively, making $\BFin$ a symmetric rig groupoid.
+
+The semantics interprets $\PiLang$-types as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of 1-combinators
+related by a 2-combinator, their interpretations in~$\BFin$ are equal.  The groupoid $\BFin$ is strict, since the
+collection of isomorphisms is a set, that is, a discrete category. There is no explicit witness for the equality of two
+isomorphisms, since we can decide by evaluating two bijections whether they are equal. To be able to establish
+completeness for $\PiLang$, we want a witness for this equality, so that we can quote back to the syntax and produce a
+2-combinator witnessing the equality of the corresponding 1-combinators.
 
@@
 % Third - say how are we fixing it
-First, we observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
-homotopies -- we \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells for
-them. We work in Homotopy Type Theory (Univalent Foundations) -- it provides a proof-relevant, constructive metatheory
-to get a handle on these equalities, and, it provides a rich internal language for describing weak groupoids, using the
-"types are weak $\infty$-groupoids" correspondence.
+
+We observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
+homotopies. We therefore \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells
+and work in HoTT (Univalent Foundations) as it provides a proof-relevant, constructive metatheory to get a handle
+on these equalities and provides a rich internal language for describing weak groupoids, using the ``types are weak
+$\infty$-groupoids'' correspondence.
 
@@
 
-% Fourth - expain the tool we use for fixing it
+% Fourth - explain the tool we use for fixing it
 Every type in HoTT is a weak $\infty$-groupoid whose points are the terms of the type, and the (iterated) identity type
-gives the (higher) morphisms. The groupoid we're interested in has types as points, type equivalences for 1-cells, and
-higher homotopies for higher cells. This is the groupoid structure for the universe type $\UU$, since the identity type
-on types can be characterised as type equivalences (by \emph{univalence}). But, we only want to carve out a
-\emph{subuniverse} of \emph{finite types}, still satisfying univalence, to get the groupoid structure. In this section,
-we formally define \emph{univalent subuniverses}, and proceed to construct the particular instance for finite types,
-$\UFin$ (\Cref{def:ufin}).
+gives the (higher) morphisms. The groupoid we are interested in has types as points, type equivalences for 1-cells, and
+higher homotopies for higher cells. (See App.~\ref{app:grpdexample} for an example on a 3-element set.) This is the
+groupoid structure for the universe type $\UU$, since the identity type on types can be characterised as type
+equivalences (by \emph{univalence}). But, we only want to carve out a \emph{subuniverse} of \emph{finite types}, still
+satisfying univalence, to get the groupoid structure. In this section, we formally define \emph{univalent subuniverses},
+and proceed to construct the particular instance for finite types, $\UFin$ (\Cref{def:ufin}).
 
@@
 % morphisms. We're interested in the groupoid structure of the universe, which has types for points, equivalences for
-% identities between them (by univalence), and higher homotopies for higher identitifications. We want a similar
+% identities between them (by univalence), and higher homotopies for higher identifications. We want a similar
 % universe, whose only points are the finite types, that is, a subuniverse, and, equalities between them should
@@
 \begin{toappendix}
+  \label{app:grpdexample}
   We give an example of the groupoid structure on a 3-element set.
@@
 
-  We have $\Fin[3] = \Set{0,1,2} \eqv \unit \sqcup (\unit \sqcup \unit)$ which fixes a particular enumeration of the
+  \noindent We have $\Fin[3] = \Set{0,1,2} \eqv \unit \sqcup (\unit \sqcup \unit)$ which fixes a particular enumeration of the
   elements. Suppose we have a set $X = (\unit \sqcup \unit) \sqcup \unit$, it has the same cardinality as $\Fin[3]$, so it
@@
   there are $X \eqv Y$.
-
 \end{toappendix}
 
-\begin{toappendix}
+% \begin{toappendix}
 
-  \subsection{To Put Somewhere or Delete}
+%   \subsection{To Put Somewhere or Delete}
 
-  This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
-  level~\cite{carette2016}.
-  As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
-  we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
-  set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
-  the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
-  $\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
-  with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
-  \[\begin{array}{rcl}
-      \denot{\onet + (\onet + \onet)}                                         & = & \{ 0,1,2 \} \\
-      \denot{(\onet + \onet) + \onet}                                         & = & \{ 0,1,2 \} \\
-      \\
-      \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} & = & (0~1~2)     \\
-      \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet}   & = & (2~0~1)
-    \end{array}\]
-  where we have used the one-line notation for permutations with $(a~b~c)$ representing the
-  permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
-  enumeration of the elements of each type:
-  \[\begin{array}{rcl}
-      \mathit{enum}(\zerot)     & = & [~]                                                                                                          \\
-      \mathit{enum}(\onet)      & = & [ ~\Acon{tt}~ ]                                                                                              \\
-      \mathit{enum}(A + B)      & = & \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B)          \\
-      \mathit{enum}(A \times B) & = & \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
-    \end{array}\]
-  \noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
-  a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
-  operation that appends all the subsequences in a sequence.
+%   This approach is sufficient to prove the semantics forms a 1-category but ignores the rich structure at the next
+%   level~\cite{carette2016}.
+%   As explained in the previous section, a $\Pi$-type $A$ has $\sizet{A}$-elements and for all combinators $c : A \iso B$
+%   we have that $\sizet{A} = \sizet{B}$. Hence, the denotation $\denot{A}$ of a type $A$ with $n$-elements can be the finite
+%   set $\Fin[n] = \{ 0, 1, \cdots, n-1\}$; the denotation of a value $v : A$ such that $\sizet{A}=n$ will be an index in
+%   the range $[0,n-1]$, and the denotation $\denot{c}$ of a combinator $c : A \iso B$ such that
+%   $\sizet{A} = \sizet{B} = n$ will be a function from $\Fin[n]$ to $\Fin[n]$ that permutes the elements. Thus, all types
+%   with 3 elements will denote $\Fin[3]$ and combinators between them will denote permutations on $\Fin[3]$, e.g.:
+%   \[\begin{array}{rcl}
+%       \denot{\onet + (\onet + \onet)}                                         & = & \{ 0,1,2 \} \\
+%       \denot{(\onet + \onet) + \onet}                                         & = & \{ 0,1,2 \} \\
+%       \\
+%       \denot{\assoclp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet} & = & (0~1~2)     \\
+%       \denot{\swapp : \onet + (\onet + \onet) \iso (\onet + \onet) + \onet}   & = & (2~0~1)
+%     \end{array}\]
+%   where we have used the one-line notation for permutations with $(a~b~c)$ representing the
+%   permutation that maps 0 to $a$, 1 to $b$, and 2 to $c$. To make the denotation of values precise, we compute a canonical
+%   enumeration of the elements of each type:
+%   \[\begin{array}{rcl}
+%       \mathit{enum}(\zerot)     & = & [~]                                                                                                          \\
+%       \mathit{enum}(\onet)      & = & [ ~\Acon{tt}~ ]                                                                                              \\
+%       \mathit{enum}(A + B)      & = & \mathit{map}~\Acon{inj₁}~\mathit{enum}(A) ~\textsf{+\!+}~ \mathit{map}~\Acon{inj₂}~\mathit{enum}(B)          \\
+%       \mathit{enum}(A \times B) & = & \mathit{concat}~(\mathit{map}~(\lambda v.\mathit{map}~(\lambda w. (v,w))~\mathit{enum}(B))~\mathit{enum}(A))
+%     \end{array}\]
+%   \noindent The specification uses a Haskell-like notation for sequences with $\mathit{map}$ as the operation that applies
+%   a function to each element of a sequence, \textsf{+\!+} as the binary append operation, and $\mathit{concat}$ the
+%   operation that appends all the subsequences in a sequence.
 
-  Using the definition, we have:
-  \[\begin{array}{rcl}
-      \mathit{enum}(\onet + (\onet + \onet)) & = & [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
-      \mathit{enum}((\onet + \onet) + \onet) & = & [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
-    \end{array}\]
-  Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
-  $\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
-  $\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
-  $\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
-  $\inrv{\Acon{tt}}$.
+%   Using the definition, we have:
+%   \[\begin{array}{rcl}
+%       \mathit{enum}(\onet + (\onet + \onet)) & = & [ \inlv{\Acon{tt}},~\inrv{(\inlv{\Acon{tt}})},~\inrv{(\inrv{\Acon{tt}})} ] \\
+%       \mathit{enum}((\onet + \onet) + \onet) & = & [ \inlv{(\inlv{\Acon{tt}})},~\inlv{(\inrv{\Acon{tt}})},~\inrv{\Acon{tt}} ]
+%     \end{array}\]
+%   Thus, as shown in the diagrams below, $\assoclp~(\inlv{\Acon{tt}})$ applies the permutation $(0~1~2)$ to the index of
+%   $\inlv{\Acon{tt}}$ which is 0 and produces index 0 in the $(\onet + \onet) + \onet$ type corresponding to value
+%   $\inlv{(\inlv{\Acon{tt}})}$. Similarly, $\swapp~(\inlv{\Acon{tt}})$ applies the permutation $(2~0~1)$ to the index of
+%   $\inlv{\Acon{tt}}$ which is 0 and produces index 2 in the $(\onet + \onet) + \onet)$ type corresponding to value
+%   $\inrv{\Acon{tt}}$.
 
-  \begin{center}
-    \input{assoc-perm.tikz}
-    \qquad
-    \input{swap-perm.tikz}
-  \end{center}
+%   \begin{center}
+%     \input{assoc-perm.tikz}
+%     \qquad
+%     \input{swap-perm.tikz}
+%   \end{center}
 
-  We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
-  compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
-  $\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:
+%   We choose a canonical set of size $n$, called $\mathsf{Fin}~n$, whose elements are natural numbers less than $n$. To
+%   compute the denotation of a type $A$, we first calculate its size $n = \sizet{A}$. We then construct the canonical set
+%   $\mathsf{Fin}~n$ and provide the (trivial) evidence that this set is identical to $(\mathsf{Fin}~n)$:
 
-  \[\begin{array}{rcll}
-      \sem{A} & = & (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
-    \end{array}\]
+%   \[\begin{array}{rcll}
+%       \sem{A} & = & (\mathsf{Fin}~n, [ n , \mathsf{refl} ]) & \mbox{where}~\sizet{A} = n
+%     \end{array}\]
 
-  \noindent The denotation $\sem{c}$ of a combinator $c : A \isoone B$ is a path between $\sem{A}$ and $\sem{B}$. If the
-  size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
-  $(\mathsf{Fin}~n, [ n , \mathsf{refl} ])$. This path is directly constructed using $\mathit{ap}$ and the fact that $m=n$
-  since combinators are always between types of the same size.
+%   \noindent The denotation $\sem{c}$ of a combinator $c : A \isoone B$ is a path between $\sem{A}$ and $\sem{B}$. If the
+%   size of $A$ is $m$ and the size of $B$ is $n$, the desired path is between $(\mathsf{Fin}~m, [ m , \mathsf{refl} ])$ and
+%   $(\mathsf{Fin}~n, [ n , \mathsf{refl} ])$. This path is directly constructed using $\mathit{ap}$ and the fact that $m=n$
+%   since combinators are always between types of the same size.
 
-  \noindent Finally, given two combinators $p , q : A \isoone B$ and a 2-combinator $\alpha : p \isotwo q$, the denotation
-  $\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.
+%   \noindent Finally, given two combinators $p , q : A \isoone B$ and a 2-combinator $\alpha : p \isotwo q$, the denotation
+%   $\sem{\alpha}$ of $\alpha$ is a path between $\sem{p}$ and $\sem{q}$.
 
-  \note{We use the rig structure of $\UFin$ in~\Cref{subsec:rig} to interpret $\PiLang$.}
+%   \note{We use the rig structure of $\UFin$ in~\Cref{subsec:rig} to interpret $\PiLang$.}
 
-  We need a formal definition of normal form (canonical form)
+%   We need a formal definition of normal form (canonical form)
 
-  Recalling that the $\lambda$-calculus arises as the internal language of Cartesian Closed Categories
-  (Elliott~\cite{Elliott-2017} gives a particularly readable account of this), we can think of $\Pi$ in similar terms, but
-  for symmetric Rig Groupoids instead. For example, we can ask what does the equivalence above represent? It is actually a
-  ``linear'' representation of a 2-categorial commutative diagram! In fact, it is a painfully verbose version thereof, as
-  it includes many \emph{refocusing} steps because our language does not build associativity into its syntax. Categorical
-  diagrams usually do.  Thus if we rewrite the example in diagrammatic form, eliding all uses of associativity, but
-  keeping explicit uses of identity transformations, we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents
+%   Recalling that the $\lambda$-calculus arises as the internal language of Cartesian Closed Categories
+%   (Elliott~\cite{Elliott-2017} gives a particularly readable account of this), we can think of $\Pi$ in similar terms, but
+%   for symmetric Rig Groupoids instead. For example, we can ask what does the equivalence above represent? It is actually a
+%   ``linear'' representation of a 2-categorial commutative diagram! In fact, it is a painfully verbose version thereof, as
+%   it includes many \emph{refocusing} steps because our language does not build associativity into its syntax. Categorical
+%   diagrams usually do.  Thus if we rewrite the example in diagrammatic form, eliding all uses of associativity, but
+%   keeping explicit uses of identity transformations, we get that \AgdaFunction{swap{-}fl2⇔swap{-}fl1} represents
 
-  \vspace*{3mm}
-  \begin{tikzcd}[column sep=normal, row sep=normal]
-    && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
-    & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
-    a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
-    \arrow [dr, "\assoclp"]
-    \arrow [ddr, swap, "\assoclp"]
-    & (a+b)+c \arrow [r, "\swapp"] &
-    c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}]
-    & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
-    & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
-    & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
-    \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] &&
-    c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
-    && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
-    \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
-    \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
-  \end{tikzcd}
+%   \vspace*{3mm}
+%   \begin{tikzcd}[column sep=normal, row sep=normal]
+%     && (a+c)+b \arrow [r, "\swapp \oplus\idd", ""{name=U, below}] & (c+a)+b \arrow [dr, "\assocrp"] && \\
+%     & a+(c+b) \arrow [ur, "\assoclp"] & & & c+(a+b) \arrow [dr, "\idd\oplus\swapp"] &  \\
+%     a+(b+c) \arrow [ur, "\idd\oplus\swapp"] \arrow [r, "\assoclp"]
+%     \arrow [dr, "\assoclp"]
+%     \arrow [ddr, swap, "\assoclp"]
+%     & (a+b)+c \arrow [r, "\swapp"] &
+%     c+(a+b) \arrow [r, swap, "\assoclp", ""{name=D, above}]
+%     & |[alias=Z]| (c+a)+b \arrow [r, "\assocrp"] &c+(a+b) \arrow [r, "\idd\oplus\swapp"] & c+(b+a) \\
+%     & (a+b)+c \arrow [dr, "\swapp"] &&&& \\
+%     & (a+b)+c \arrow [dr, swap, "\swapp"] & c+(a+b) \arrow [rr, swap, "\idd", ""{name=DD, above}]
+%     \arrow [d, Rightarrow, "\idf\, \mathit{idl}\odot{l}"] &&
+%     c+(a+b) \arrow [ruu, "\idd\oplus\swapp"] & \\
+%     && c+(a+b) \arrow [rrruuu, bend right = 40, swap, "\idd\oplus\swapp"] && \\
+%     \arrow[Rightarrow, from=U, to=D, "\mathit{hexagon}\oplus{r}\, \boxdot\, \idf"]
+%     \arrow[Rightarrow, from=Z, to=DD, swap, "\idf\boxdot\mathit{linv}\odot{l}\,\boxdot\,\idf"]
+%   \end{tikzcd}
 
-\end{toappendix}
+% \end{toappendix}
 
@@
 intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
-propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type Theory as well, such
-as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.
-In the appendix, we review some basic concepts and notation from the HoTT book that we use. For more details, we refer
-the reader to the book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.
+propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type
+Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
+as well.  In App.~\ref{app:identitytypes}, App.~\ref{app:homotopytypes}, and App~\ref{app:hits}, we review the basics of
+identity types, homotopy types, and HITs and refer the reader to the book for more details.
 
@@
 \begin{toappendix}
-  \subsubsection{Identity Types}
+  \subsection{Identity Types}
+  \label{app:identitytypes}
 
@@
 
-  A homotopy between functions $f \htpy g$ is given by pointwise equality between them $\dfun{x:A}{f(x) \id_{B} g(x)}$.
+  \noindent A homotopy between functions $f \htpy g$ is given by pointwise equality between them $\dfun{x:A}{f(x) \id_{B} g(x)}$.
   The identity type for functions is equivalent to homotopies between them ${(f \id_{A \to B} g)} \eqv {(f \htpy g)}$, by
@@
   fibration, that is, there is a lifting operation (\Cref{fig:lift}) which lifts paths in the base space to paths in the
-  total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have
+  total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
 
@@
 
-Functions between types are functors between groupoids, and type families, or functions to the universe, are indexed
+Functions between types are functors between groupoids, and type families (or functions to the universe) are indexed
 families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to \UU$
-which has a lifting operation giving it the structure of a fibration (see the appendix for details). The
-$\term{transport}$ operation lifts paths in the base space to maps between the fibers. Using the groupoid structure of
-$A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence -- we name
-this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
-
+which has a lifting operation giving it the structure of a fibration. The
+$\term{transport}$ operation $\transport{P}{p}{\blank}$ lifts paths in the base space to equivalences between the fibers:
+% Using the groupoid structure of
+% $A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence.
+% We name
+% this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
 \[
@@
 
-The type families (or fibrations) we're interested in are the ones where paths in the base space completely determine
+\noindent The type families (or fibrations) we are interested in are the ones where paths in the base space completely determine
 the equivalences in the fibers -- these are called univalent
@@
 
-  \subsubsection{Homotopy Types}
+  \subsection{Homotopy Types}
+  \label{app:homotopytypes}
 
@@
 
-  \subsubsection{Higher Inductive Types}
+  \subsection{Higher Inductive Types}
+  \label{app:hits}
 
@@
     \end{itemize}
-    with \review{an appropriate induction principle.}
+    with an appropriate induction principle.
   \end{definition}
@@
 \begin{proposition}[Univalent Subuniverse]
+  \label{prop:univsub}
   A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
-  proposition for every $X$. The fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent. This generates a
-  univalent subuniverse
-  \[
-    \UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)
-  \]
+  proposition for every $X$. Given such a predicate, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent and generates a
+  univalent subuniverse $\UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)$.
 \end{proposition}
@@
 \begin{appendixproof}
+  (of Prop.~\ref{prop:univsub})
   Suppose $(U, \El) \defeq (\dsum*{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
@@
 
-The types we're interested in are the finite types. In constructive mathematics, the notion of finiteness is
-subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness -- a type is finite if it is
-merely equivalent to a finite set (\Cref{def:finite-set}), and we define the predicate $\isFin$ (\Cref{def:isfin}) to do
-so.
+The types we are interested in are the finite types. In constructive mathematics, the notion of finiteness is
+subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness: a type is finite if it is
+merely equivalent to a finite set (\Cref{def:finite-set} and \Cref{def:isfin}).
 
@@
     \end{aligned}
-    \qquad\qquad
+    \qquad\qquad \qquad\qquad
     \begin{aligned}
@@
 \begin{proposition}
+  \label{prop:isFin}
   For any type $X$, $\isFin[X]$ is a proposition.
@@
 \begin{appendixproof}
+  (of Prop.~\ref{prop.isFin})
   Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
@@
   The univalent subuniverse of \emph{all finite types} is given by
-  \[
+  $
     \UFin \defeq \dsum{X:\UU}{\isFin[X]}.
-  \]
+  $
   We write $F_{n} \defeq (\Fin[n], n, \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$.
@@
 
-While $\UFin$ has \emph{all} the finite types, we're also interested in constructing a subuniverse of finite types of a
+While $\UFin$ has \emph{all} the finite types, we are also interested in constructing a subuniverse of finite types of a
 specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$.
@@
   generates the subuniverse
-  \[
+  $
     \BAut[T] \defeq \Sub{T}.
-  \]
+  $
   We write $T_0 \defeq (T, \trunc{\refl_{T}})$ for the image of the inclusion of $T$ in $\BAut[T]$.
@@
 
-\begin{proposition}
+\begin{proprep}
   \leavevmode
@@
   \end{enumerate}
-\end{proposition}
+\end{proprep}
+\begin{proofsketch}
+  See Appendix.
+\end{proofsketch}
 
@@
 
-The loopspace of a pointed type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element,
-path composition for the group multiplication, and path inverse for the group inverse operation. The group axioms are
-then given by the higher paths corresponding to groupoid laws. We have shown that loops in $\UFin$ exactly encode the
-automorphism group $\Aut[\Fin[n]]$ for every $n$. This is a general technique called \emph{delooping}, where a group can
-be identified with a 1-object groupoid, internally in HoTT.~\footnote{This technique also allows defining higher groups
-  in HoTT, as done in~\cite{buchholtzHigherGroupsHomotopy2018}.}
+We have shown that loops in $\UFin$ exactly encode the automorphism group $\Aut[\Fin[n]]$ for every~$n$. This is a
+general technique called \emph{delooping}, where a group can be identified with a 1-object groupoid, internally in
+HoTT. This technique also allows defining higher groups in HoTT~\cite{buchholtzHigherGroupsHomotopy2018}. The loopspace
+of a pointed type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element, path
+composition for the group multiplication, and path inverse for the group inverse operation. The group axioms are then
+given by the higher paths corresponding to groupoid laws.
 
@@
   \begin{definition}
+    \label{app:defone}
     \begin{align*}
@@
   \begin{proposition}
+    \label{app:propone}
     % https://q.uiver.app/?q=WzAsNCxbMCwwLCIoWCBcXG9wbHVzIEkpIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyAoSSBcXG9wbHVzIFkpIl0sWzEsMSwiWCBcXG9wbHVzIFkiXSxbMCwxXSxbMCwxLCJcXGFscGhhX3tYLEksWX0iXSxbMCwyLCJcXHJob197WH0gXFxvcGx1cyAxX3tZfSIsMl0sWzEsMiwiMV97WH0gXFxvcGx1cyBcXGxhbWJkYV97WX0iXSxbNSw2LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d

### 2021-07-08 Pi+/paper/reversible1.tex
Revise section 3.2

@@
 
-\noindent Below we present a semantic map that dually serves as an operational semantics and a denotational
-semantics. As an operational semantics, each $\PiLang$ type $A$ is mapped to an Agda type $\denot{A}$ and each
-combinator $c : A \isoone B$ is mapped to an Agda function $\denot{A} \to \denot{B}$. As a denotational semantics, each
-$\PiLang$ type $A$ is mapped to a finite set $\denot{A}$ and each combinator $c : A \isoone B$ is mapped to bijection
-between $\denot{A} \to \denot{B}$.
+\noindent Below we present a simple denotational semantics for our language, using finite types and type isomorphisms.
+Each $\PiLang$ type $A$ is mapped to a (finite) set $\denot{A}$ and each combinator $c : A \isoone B$ is mapped to a
+(bijective) function $\denot{A} \to \denot{B}$. We describe this semantics by writing an interpreter in Agda. First, we
+state the semantics for types, where $\bot$ is the empty set, $\top$ is the singleton set, $\sqcup$ is the disjoint
+union of sets, and $\times$ is the cartesian product of sets.
 
-The semantics of types is straightforward:
+% semantic map that dually serves as an operational semantics and a denotational semantics. As an operational semantics,
+% As a denotational semantics, each $\PiLang$ type $A$ is mapped to a finite set $\denot{A}$ and each combinator $c : A
+% \isoone B$ is mapped to bijection between $\denot{A} \to \denot{B}$.
 
@@
 
-\noindent where $\bot$ is the empty set, $\top$ is a singleton set, $\sqcup$ is the disjoint union of sets, and $\times$
-is the cartesian product of sets. The semantics of a combinator $c : A \isoone B$ is a (bijective) function mapping
-$\denot{A}$ to $\denot{B}$:
+For combinators, we show explcitly how values reduce along each combinator, similar to a big-step operational
+semantics~\cite{chenComputationalInterpretationCompact2021,theseus}.
+
+% The semantics of a combinator $c : A \isoone B$ is a (bijective) function mapping $\denot{A}$ to $\denot{B}$:
 
@@
           \denot{\factor}  & (\inl{(v_1 , v_3)}) & = & (\inl{v_1} , v_3)   \\
-          \denot{\factor}  & (\inr{(v_2 , v_3)}) & = & (\inr{v_2} , v_3)
+          \denot{\factor}  & (\inr{(v_2 , v_3)}) & = & (\inr{v_2} , v_3)   \\
+          \denot{\idc}     & v                   & = & v
         \end{array}$}}
-
   %%\end{longtable}
@@
         \begin{array}{rlcl}
-          \denot{\idc}               & v           & = & v                                   \\
           \denot{(c_1 \fatsemi c_2)} & v           & = & (\denot{c_2} \circ \denot{c_1}) v   \\
@@
 \end{theorem}
-\begin{proof}
+\begin{appendixproof}
   For every primitive combinator $c$ listed on one side of Fig.~\ref{fig:pi-terms}, let $!c$ be the combinator listed on
@@
     = \mathit{id} \circ \denot{c} = \denot{c}$. The other cases are only slightly more involved calculations.
-\end{proof}
+\end{appendixproof}
 

### 2021-07-08 Pi+/paper/ufin.tex
Switch theorems to theoremreps in sections 3 and 4

@@
 
-\begin{proposition}[Univalent Subuniverse]
+\begin{propositionrep}[Univalent Subuniverse]
   \label{prop:univsub}
@@
   univalent subuniverse $\UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)$.
-\end{proposition}
+\end{propositionrep}
 
-\begin{appendixproof}
-  (of Prop.~\ref{prop:univsub})
+\begin{proof}
   Suppose $(U, \El) \defeq (\dsum*{X:\UU}{P(X)}, \fst)$ is a subuniverse generated by a subtype $P : \UU \to \UU$. For
-  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that
-  $\tptEqv{\fst} : (X,\phi) \id (Y,\psi) \to X \eqv Y$ is an equivalence. We construct
-  $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the fact that $P(\blank)$ is a proposition. That it is an
-  inverse follows by calculation using the appropriate computation rules.
-\end{appendixproof}
+  any $X, Y : \UU$ such that $\phi : P(X)$ and $\psi : P(Y)$, we want to show that $\tptEqv{\fst} : (X,\phi) \id
+    (Y,\psi) \to X \eqv Y$ is an equivalence. We construct $X \eqv Y \to (X,\phi) \id (Y,\psi)$ by $\ua$ and using the
+  fact that $P(\blank)$ is a proposition. That it is an inverse follows by calculation using the appropriate computation
+  rules.
+\end{proof}
 
 The types we are interested in are the finite types. In constructive mathematics, the notion of finiteness is
-subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness: a type is finite if it is
-merely equivalent to a finite set (\Cref{def:finite-set} and \Cref{def:isfin}).
+subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness: a type is finite if it is merely
+equivalent to a finite set (\Cref{def:finite-set} and \Cref{def:isfin}).
 
@@
 
-\begin{proposition}
+\begin{propositionrep}
 \label{prop:isFin}
 For any type $X$, $\isFin[X]$ is a proposition.
-\end{proposition}
+\end{propositionrep}
 
-\begin{appendixproof}
+\begin{proof}
   (of Prop.~\ref{prop.isFin})
@@
   applying the first projection.
-\end{appendixproof}
+\end{proof}
 
@@
 
-\begin{proprep}
+\begin{propositionrep}
   \leavevmode
@@
   \end{enumerate}
-\end{proprep}
-\begin{proofsketch}
-  See Appendix.
-\end{proofsketch}
+\end{propositionrep}
 
-\begin{appendixproof}
+\begin{proof}
   We need to show that the equality type of $\BAut[T]$ is an $n$-type. Assume $X, Y : \BAut[T]$. Since $\BAut[T]$ is a
@@
   \]
-\end{appendixproof}
+\end{proof}
 

### 2021-07-08 Pi+/paper/finite.tex
Say horizontal and vertical categorification

@@
 In~\Cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
-$n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group. This is
-also known to be the finite symmetric group $\Sn[n]$. In the following sections, we will describe this group
-syntactically.
+$n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group.  This
+is also known to be the finite symmetric group $\Sn[n]$, making $\UFin$ the \emph{horizontal categorification} of $\Sn$
+for every $n$. In this section, we will describe this group syntactically.
 

### 2021-07-08 Pi+/paper/equivalence.tex
Squeeze some space in equations
Revise sections 6 and 7 and copy-editing

@@
   \end{tikzcd}\]
-
-\vc{ The groupoid of finite types is the free symmetric monoidal groupoid on one generator. This can be presented as an
-  algebraic 2-theory, which is our syntax for $\PiHatLang$. Vertical categorification of natural numbers as a free
-  commutative monoid. See groupoidification. }
 
@@
 To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
-category. Since the language have higher-dimensional structure, the syntactic categories here are actually
+category. Since the languages have higher-dimensional structure, the syntactic categories here are actually
 (locally-strict) weak 2-categories. We define them formally in the appendix, and only state our results here.
@@
 
-Using this rig structure, we translate $\PiLang$ to $\PiPlusLang$, constructing a rig functor from $\PiCat$ to
+Using this rig structure, we translate $\PiLang$ to $\PiPlusLang$, constructing a rig equivalence from $\PiCat$ to
 $\PiPlusCat$.
@@
 \end{definition}
+
+\begin{proposition}
+  $\evalt$ gives a rig biequivalence between $\PiCat$ and $\PiPlusCat$.
+\end{proposition}
 
@@
 
-\todo{Sn/Lehmer/Aut(Fin n) diagram here?}
-
 Types in $\PiHatLang$ are interpreted as 0-cells in $\UFin$, that is, a natural number $n$ is mapped to $\Fin[n]$. The
@@
 \begin{definition}[${\gdenot{\blank}} : U \to {\UFin}$]
-  \begin{align*}
-    \gdenot{\blank}_{0} & = \evalh_{0} \comp \evalp_{0} \comp \evalt_{0} \\
-    \gdenot{\blank}_{1} & = \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
-  \end{align*}
+  \begin{gather*}
+    \begin{aligned}
+      \gdenot{\blank}_{0} &\defeq \evalh_{0} \comp \evalp_{0} \comp \evalt_{0}
+    \end{aligned}
+    \qquad
+    \begin{aligned}
+      \gdenot{\blank}_{1} &\defeq \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
+    \end{aligned}
+  \end{gather*}
 \end{definition}
@@
 
-\subsection{$\PiLang$ circuit normalisation}
+\section{Normalisation of reversible circuits}
+\label{sec:applications}
 
-To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$, compute a permutation, and quote it back to
-$\PiPlusLang$. Note that the normalisation happens in going from $\PiHatLang$ to $\UFin$ and back to $\PiHatLang$.
+Using our semantics, we can normalise a circuit writen in $\PiLang$ -- we first translate it to $\PiHatLang$, then
+compute a permutation, and quote it back to $\PiPlusLang$. Note that the normalisation happens in going from
+$\PiHatLang$ to $\UFin$ and back to $\PiHatLang$.
 
 \begin{definition}
-  \begin{align*}
+  \begin{gather*}
+    \begin{aligned}
       \normt_{0} & : U \to \UPlus                                            \\
-    \normt_{0} & = \quotep_{0} \comp \quoteh_{0} \comp \gdenot{\blank}_{0} \\
-    \\
+      \normt_{0} & = \quotep_{0} \comp \quoteh_{0} \comp \gdenot{\blank}_{0}
+    \end{aligned}
+    \qquad
+    \begin{aligned}
       \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \iso \normt_{0}(Y)     \\
       \normt_{1} & = \quotep_{1} \comp \quoteh_{1} \comp \gdenot{\blank}_{1}
-  \end{align*}
+    \end{aligned}
+  \end{gather*}
 \end{definition}
 
-We now revisit the examples in the introduction but instead of working with circuits operating on sequences of bits, we
-work with circuits operating on algebraic data types.
+We revisit the examples in~\Cref{sec:examples} but instead of working with circuits operating on sequences of bits, we
+work with circuits operating on algebraic data types in $\PiLang$.
 
-\paragraph*{Normalization by evaluation.} Consider the following reversible function:
-$\Afun{RESET}(b,b_1,\ldots,b_n) = (b \; \underline{\vee} \; (\bigvee_{i=1}^n b_i),b_1,\ldots,b_n)$ where $\vee$ is
-logical-or and $\underline{\vee}$ is exclusive-or. The function sets or resets the first bit depending on whether any of
-the other bits is true or not. The natural definition for this function is recursive: modulo some re-shuffling of the
-bits, the strategy is to examine the bits $b_i$ one-by-one: if we encounter a true value, we negate the first bit and
-terminate; otherwise we continue making recursive calls until we reach the last bit at which point we return the
-incoming value with no change:
+\paragraph*{Normalisation by evaluation.} Consider the following reversible function: $\Afun{RESET}(b,b_1,\ldots,b_n) =
+(b \; \underline{\vee} \; (\bigvee_{i=1}^n b_i),b_1,\ldots,b_n)$ where $\vee$ is logical-or and $\underline{\vee}$ is
+exclusive-or. The function sets or resets the first bit depending on whether any of the other bits is true or not. The
+natural definition for this function is recursive: modulo some re-shuffling of the bits, the strategy is to examine the
+bits $b_i$ one-by-one: if we encounter a true value, we negate the first bit and terminate; otherwise we continue making
+recursive calls until we reach the last bit at which point we return the incoming value with no change:
 
@@
 provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
-reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the
-introduction. Our Agda infrastructure provides tools to normalize all programs to their normal form. For \Afun{reset 2}
-we get:
+reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the introduction.
+Our Agda infrastructure provides tools to normalise all programs to their normal form. For \Afun{reset 2} we get:
 
@@
 
-In principle, the normalized program can be produced following two strategies: (i) by repeatedly applying the rewrite
+In principle, the normalised program can be produced following two strategies: (i) by repeatedly applying the rewrite
 rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
-using a normalization-by-evaluation (NbE) process that evaluates the program to a permutation on a finite set of 8
+using a normalisation-by-evaluation (NbE) process that evaluates the program to a permutation on a finite set of 8
 elements and reifies that permutation back to a program. The key idea of the NbE process is a systematic way to express
@@
 \paragraph*{Program synthesis.} The NbE process embodies a quoting mechanism that synthesizes programs from
-permutations. Indeed, instead of writing a program for \Afun{reset 2}, one might simply specify the desired permutation as:
+permutations. Indeed, instead of writing a program for \Afun{reset 2}, one could simply specify the desired permutation
+as:
 
@@
 which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Simply
-``quoting'' this permutation generates the same normalized program \Afun{reset2Norm} above.
+``quoting'' this permutation generates the same normalised program \Afun{reset2Norm} above.
 

### 2021-07-08 Pi+/paper/finite.tex
Fix longcoxstar and longcoxplus

--- 2021-07-07 Pi+/paper/finite.tex
@@
 Because of this counter-example, the relations have to be changed. In this section, we formally define a rewriting
-system $(\List[\Fin[n]], \longcox*)$, partially based on the Coxeter relations, and prove that it has the desired
+system $(\List[\Fin[n]], \longcox)$, partially based on the Coxeter relations, and prove that it has the desired
 properties of confluence and termination. We prove that the new relation defined by this system is equivalent,
@@
 
-We define the relation $\longcox*$ to be the closure of $\longcox$, under reflexivity and (right-step extended)
-transitivity (see~\Cref{def:longcox-star}).
-
 \begin{toappendix}
-  \begin{definition}[$\longcox*$]
+  \begin{definition}[$\longcoxstar$]
     \label{def:longcox-star}
@@
       \reflr{\longcox}
-       & : \forall w \to w \longcox* w                                                                               \\
+       & : \forall w \to w \longcoxstar w                                                                               \\
       \transr{\longcox}
-       & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcox* w_{3} \to w_{1} \longcox* w_{3}
+       & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcoxstar w_{3} \to w_{1} \longcoxstar w_{3}
+    \end{align*}
+  \end{definition}
+  \begin{definition}[$\longcoxplus$]
+    \label{def:longcox-plus}
+    \begin{align*}
+      \relr{\longcoxplus}
+       & : \forall w_{1}, w_{2} \to w_{1} \longcox w_{2} \to w_{1} \longcoxplus w_{2}
+      \transr{\longcoxplus}
+       & : \forall w_{1}, w_{2}, w_{3} \to  w_{1} \longcox w_{2} \to w_{2} \longcoxplus w_{3} \to w_{1} \longcoxplus w_{3}
     \end{align*}
@@
 
-Despite the increased complexity of the generators, the rewriting system $(\List[\Fin[n]],\longcox*)$ has the properties
+We define $\longcoxplus$ to be the transitive closure (\Cref{def:longcox-star}) and $\longcoxstar$ to be
+reflexive-transitive closure (\Cref{def:longcox-plus})) of $\longcox$.
+
+Despite the increased complexity of the generators, the rewriting system $(\List[\Fin[n]],\longcox)$ has the properties
 we desire. It satisfies local confluence, that is, the Church-Rosser (diamond) property, and it is terminating, so, by
@@
           extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
-    \item $\longcox*$ is terminating. For every $w \longcox v$, it holds that $v < w$, where $<$ is the (well-founded)
+    \item $\longcoxplus$ is terminating. For every $w \longcoxplus v$, it holds that $v < w$, where $<$ is the (well-founded)
           lexicographic ordering on $\List[\Fin[n]]$.
-    \item $\longcox*$ is confluent. For every extended span $\coxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
+    \item $\longcoxstar$ is confluent. For every extended span $\coxspan*{w_{1}}{w_{2}}{w_{3}}$, there is a matching
           extended cospan $\coxcospan*{w_{2}}{w_{3}}{w}$.
-    \item For every $w$, there exists a unique normal form $v$ such that $w \longcox* v$.~\label{prop:nf-uniqueness}
+    \item For every $w$, there exists a unique normal form $v$ such that $w \longcoxstar v$.~\label{prop:nf-uniqueness}
   \end{enumerate}
@@
   \label{prop:coxlongcox}
-  The relations $\cox*$ and $\longcox*$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff
-  there is a $u$ such that $w \longcox* u$ and $v \longcox* u$.
+  The relations $\cox*$ and $\longcoxstar$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff
+  there is a $u$ such that $w \longcoxstar u$ and $v \longcoxstar u$.
 \end{proposition}
@@
   \begin{enumerate}
-    \item For any Lehmer code $c$, $\immersion(c)$ is a normal form with resepct to $\longcox*$, that is, $\immersion(c)$ is
+    \item For any Lehmer code $c$, $\immersion(c)$ is a normal form with resepct to $\longcoxstar$, that is, $\immersion(c)$ is
           in $\im{\normf}$.
@@
 \jk{Review:}
-To see that Lehmer codes under $\immersion$ are normal with respect to $\longcox*$, let us look at the example code $c =
+To see that Lehmer codes under $\immersion$ are normal with respect to $\longcoxstar$, let us look at the example code $c =
   (4, 3, 2, 1, 0)$. The list $\immersion(c)$ is then going to be $[0, 1, 0, 2, 1, 0, 3, 2, 1, 0, 4, 3, 2, 1, 0]$. It is a

### 2021-07-08 Pi+/paper/reversible1.tex

@@
 
-\begin{theorem}\label{thm:semone}
+\begin{theoremrep}\label{thm:semone}
   The semantics is sound in the following sense:
@@
   \end{itemize}
-\end{theorem}
-\begin{appendixproof}
+\end{theoremrep}
+\begin{proof}
   For every primitive combinator $c$ listed on one side of Fig.~\ref{fig:pi-terms}, let $!c$ be the combinator listed on
@@
     = \mathit{id} \circ \denot{c} = \denot{c}$. The other cases are only slightly more involved calculations.
-\end{appendixproof}
+\end{proof}
 

### 2021-07-08 Pi+/paper/ufin.tex

@@
 finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
-structures, respectively, making $\BFin$ a symmetric rig groupoid.
+structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
+commutative rig of natural numbers $\Nat$.
 

### 2021-07-08 Pi+/paper/finite.tex
Revise some things in the appendix
Fix immersion normal form example-proof

@@
     \Lehmer[\zero]   & \defeq \Fin[\suc[\zero]]                     \\
-    \Lehmer[\suc[n]] & \defeq \Fin[\suc[\suc[n]]] \times \Lehmer[n]
+    \Lehmer[\suc[n]] & \defeq \Lehmer[n] \times \Fin[\suc[\suc[n]]]
   \end{align*}
@@
 
-\jk{Review:}
-To see that Lehmer codes under $\immersion$ are normal with respect to $\longcoxstar$, let us look at the example code $c =
-  (4, 3, 2, 1, 0)$. The list $\immersion(c)$ is then going to be $[0, 1, 0, 2, 1, 0, 3, 2, 1, 0, 4, 3, 2, 1, 0]$. It is a
-concatenation of five decreasing lists. Where could a reduction happen?
+\begin{proof}
+\jk{Only for part 1 of the proposition!}
+For any code $c : \Lehmer[n]$, we have that 
+\[
+  \immersion(n, c) = (r_0 \downf k_0) \append (r_1 \downf k_1) \append \dots \append (r_{m-1} \downf k_{m-1})
+\]
+is concatenation of $m \geq 0$ non-empty strictly decreasing lists. Reduction $\longcox$ cannot happen inside any of the strictly
+decreasing $(r \downf k)$: $\longcancel$ requires repeating elements, $\longswap$ acts when consecutive numbers differ
+by at least 2, and $\longbraid$ acts on a non-monotone sequence. This leaves the case of reduction happening on a
+fragment that borders two (or more) subsequences. Again, $\longcancel$ requires two equal consecutive numbers, which
+would then have to be the last one in some $(r_i \downf k_i)$ sequence, and the first one in the next $(r_{i+1} \downf k_{i+1})$.
+But the first number in a sequence $(r_{i+1} \downf k_{i+1})$ is larger than every number in $(r_{i} \downf k_i)$ -- which also
+shows why $\longswap$ cannot happen. The remaining case of $\longbraid$ follows similarly, since its argument is a
+decreasing sequence followed immediately by a number equal to the first element of this sequence.
+\end{proof}
+% \jk{Review:}
+% To see that Lehmer codes under $\immersion$ are normal with respect to $\longcoxstar$, let us look at the example code
+% $c = (0, 1, 2, 3, 4)$ -- for the full proof, see accompanying Agda code. We have that 
+% \[
+%   \immersion(c) = (0 \downf 1) \append (0 \downf 2) \append (0 \downf 3) \append (0 \downf 4) = [0, 1, 0, 2, 1, 0, 3, 2, 1, 0]
+% \]
+% is concatenation of four decreasing lists . Where could a reduction happen?
 
-First, it can't happen inside any of the decreasing components: $\longcancel$ requires repeating elements, $\longswap$
-acts when a smaller number precedes a larger one, and $\longbraid$ has a non-monotone sequence on the left. This leaves
-the case of a reduction happening on a fragment that borders two subsequences. Again, $\longcancel$ requires two equal
-consecutive numbers, which have to be the last one in some decreasing sequence, and the first one in the next sequence.
-But the first number in a sequence is always larger than every number in the sequence before it -- which also shows why
-$\longswap$ cannot happen. The remaining case of $\longbraid$ follows similarly, since it requires the number appearing
-after the decreasing sequence to be equal to the first one in the sequence.
+% First, it can't happen inside any of the strictly decreasing $(n \downf k)$: $\longcancel$ requires repeating elements,
+% $\longswap$ acts when consecutive numbers differ by at least 2, and $\longbraid$ acts on a non-monotone
+% sequence. This leaves the case of reduction happening on a fragment that borders two (or more) subsequences. Again,
+% $\longcancel$ requires two equal consecutive numbers, which would then have to be the last one in some $(0 \downf k)$
+% sequence, and the first one in the next $(0 \downf \suc[k])$. But the first number in a sequence $(0 \downf \suc[k])$ is
+% larger than every number in $(0 \downf k)$ -- which also shows why $\longswap$ cannot happen. The remaining case of
+% $\longbraid$ follows similarly, since its argument is a decreasing sequence followed immediately by a number equal to
+% the first element of the list.
 

### 2021-07-08 Pi+/paper/finite.tex
Fixes in sec finite

@@
 correspondence between them and the type $\Aut[\Fin[n]]$ of automorphisms on a finite set. The generators and relations
-we use here will be used to quote back to 1 and 2-combinators in $\PiLang$ (see~\Cref{sec:equivalence}).
+we use here will be used to quote back to 1 and 2-combinators in $\PiLang$ (see~\cref{sec:equivalence}).
 
@@
   \subsection{Groups}
+  \label{subsec:groups}
 
@@
   $\blank\mult\blank$ for group multiplication, and a unary inverse operation $\inv{\blank}$. The neutral element has to
-  satisfy unit and inverse laws, and the multiplication has to be associative (see~\Cref{def:group}).
+  satisfy unit and inverse laws, and the multiplication has to be associative (see~\cref{def:group}).
 
@@
     Given a set $A$, the free group $F(A)$ on it is given by a higher inductive type with the following point and path
-    constructors. Notice the similarity with the definition of a group structure (\Cref{def:group}), but note that each
+    constructors. Notice the similarity with the definition of a group structure (\cref{def:group}), but note that each
     operation here is a generator for $F(A)$.
@@
 
-\begin{proposition}[Universal Property of $F(A)$]
+\begin{propositionrep}[Universal Property of $F(A)$]
   \label{prop:free-groups}
@@
     F(A) \to^G G$ satisfying $h \comp \eta_A \htpy f$ is contractible.
+\end{propositionrep}
+
 \begin{toappendix}
@@
 \end{toappendix}
-\end{proposition}
 
 Following the universal-algebraic definition, in HoTT, we could use a naive higher inductive type to define the free
-group, which enforces the group axioms by adding path constructors (see~\Cref{def:free-group}).
-Using the induction principle, we can easily verify the universal property. However, since this definition of $F(A)$ has
-lots of path constructors corresponding to each group axiom, characterising its path space is difficult.
+group, which enforces the group axioms by adding path constructors (see~\cref{def:free-group}). Using the induction
+principle, we can easily verify the universal property. However, since this definition of $F(A)$ has lots of path
+constructors corresponding to each group axiom, characterising its path space is difficult.
 
@@
 relations. However, these equations are not directed, so it is not always possible to construct a well-behaved rewriting
-system. In general, the word problem for groups is known to be undecidable.\jk{Example?}\todo{Examples: empty relation,
-  full relation, van Kampen of $\pi_{1}$}
-
-% This can be thought of as word reductions in a rewriting system.
+system. In general, the word problem for groups is known to be undecidable.
 
@@
 \end{figure}
-\todo{an example/motivation from book.pdf}
 
 This construction is called a Coxeter presentation of $\Sn$. Writing it formally, we encode the rules discussed above
-using a relation $\cox$ on $\List[\Fin[n]]$ (\Cref{def:cox}), and then take its congruence closure $\cox*$
-(\Cref{def:coxstar}).
+using a relation $\cox$ on $\List[\Fin[n]]$ (\cref{def:cox}), and then take its congruence closure $\cox*$
+(\cref{def:coxstar}).
 
@@
 rewriting system is confluent -- meaning that all critical pairs, that is, terms with overlapping possible reduction
-rules, have to converge. For example, in our system, the pairs in~\Cref{fig:critical-pairs-converging} converge.
+rules, have to converge. For example, in our system, the pair in~\Cref{fig:critical-pairs-converging} converges.
 Unfortunately, this is not true for all critical pairs -- an  example is in~\Cref{fig:critical-pairs-non-converging},
@@
 
+% a + \UOLoverbrace{b +}[c + d]ˆx \UOLunderbrace{+ e}_y + f
+
 \begin{figure}
@@
       \begin{tikzcd}
-        & {\tau_2\tau_1\tau_2\tau_1\tau_2} \\
-        {\tau_1\tau_2\tau_1\tau_1\tau_2} && {\tau_2\tau_1\tau_1\tau_2\tau_1} \\
-        {\tau_1\tau_2\tau_2} && {\tau_2\tau_2\tau_1} \\
+        & {\UOLoverline{\tau_2\tau_1}[\tau_2]\UOLunderline{\tau_1\tau_2}} \\
+        {\tau_1\tau_2\overline{\tau_1\tau_1}\tau_2} && {\tau_2\underline{\tau_1\tau_1}\tau_2\tau_1} \\
+        {\tau_1\overline{\tau_2\tau_2}} && {\underline{\tau_2\tau_2}\tau_1} \\
         & {\tau_1}
@@
     }
-    \caption{A converging critical pair in $\cox*$}
+    \caption{Converging critical pair}
     \label{fig:critical-pairs-converging}
@@
       \begin{tikzcd}
-        & {\tau_3\tau_2\tau_3\tau_1} \\
-        &&& {\tau_3\tau_2\tau_1\tau_3} \\
+        & {\UOLoverline{\tau_3\tau_2}[\tau_3]\UOLunderline{\tau_1}} \\
+        &&& {\underline{\tau_3\tau_2\tau_1\tau_3}} \\
         {\tau_2\tau_3\tau_2\tau_1}
-        \arrow["braid"', squiggly, from=1-2, to=3-1]
-        \arrow["braid", squiggly, from=1-2, to=2-4]
+        \arrow["\braid"', squiggly, from=1-2, to=3-1]
+        \arrow["\swap", squiggly, from=1-2, to=2-4]
         \arrow["\longbraid", squiggly, from=2-4, to=3-1]
@@
     }
-    \caption{A non-converging critical pair in $\cox*$ converges with $\longbraid$}
+    \caption{Non-converging critical pair that converges with $\longbraid$}
     \label{fig:critical-pairs-non-converging}
@@
 
-  \caption{Critical Pairs}
+  \caption{Critical Pairs in $\cox*$}
   \label{fig:critical-pairs}
@@
 
-% The Coxeter relations are a standard notion in the theory of group presentation~\todo{citation} -- even though we change
-% the relations a bit, we will not lose the connection. In fact, we will prove that the new relations are equivalent, in a
-% technical sense, to the standard Coxeter relations $\cox*$.
-% The system has generators corresponding to $\swap$, $\cancel$
-% and $\braid$, the problem of the non-converging critical pairs discussed previously is fixed by changing $\braid$ to be
-% a slightly more general relation $\longbraid$.
-
-\review{Further, we also inline the congruence closure in $\longcox$, allowing arbitrary reductions inside the list.}
-
-\todo{but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
-
-First, we need to define a helper function $n \downf k$.
+First, we need to define a function $n \downf k$.
 
@@
 The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\). Since we think of the
-number $m$ as the transposition swapping elements $(\el{m}, \el{m+1})$, the role of this helper function is to produce a
-sequence of transpositions -- a permutation -- which moves element $\el{k + n}$ by $k$ places left, shifting all the
-elements in between one place right (see~\Cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the
-following generators.
+number $m$ as the transposition swapping elements $(\el{m}, \el{m+1})$, the role of this function is to produce a
+sequence of transpositions which moves element $\el{k + n}$ left by $k$ places, shifting all the elements in between one
+place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the following generators,
+inlining the congruence closure in $\longcox$, allowing arbitrary reductions inside the list.
 
@@
 Constructors $\longcancel$ and $\longswap$ correspond directly to the appropriate constructors of $\cox$ and can be
-visualised in the same way as before. The remaining constructor uses the helper function to exchange the order of a long
-sequence of transpositions and a single transposition afterwards. For example, for $n = 0$ and $k = 3$, it allows for
-the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\Cref{fig:longcox} -- note the distinction between
-wires, where numbers represent the values, and transpositions, where numbers represent which wires are crossing).
+visualised in the same way as before. The remaining constructor $\longbraid$ uses the $\downf$ function to exchange the
+order of a long sequence of transpositions and a single transposition afterwards. For example, for $n = 0$ and $k = 3$,
+it allows for the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox} -- note the
+distinction between wires, where numbers represent the values, and transpositions, where numbers represent which wires
+are crossing).
 
@@
   \end{subfigure}
-  \caption{Braiding diarams for modified Coxeter relations.}
+  \caption{Braiding diagrams for modified Coxeter relations.}
   \label{fig:coxeter-braid-mod}
@@
   \end{definition}
+
   \begin{definition}[$\longcoxplus$]
@@
       \relr{\longcoxplus}
-       & : \forall w_{1}, w_{2} \to w_{1} \longcox w_{2} \to w_{1} \longcoxplus w_{2}
+       & : \forall w_{1}, w_{2} \to w_{1} \longcox w_{2} \to w_{1} \longcoxplus w_{2}                                      \\
       \transr{\longcoxplus}
@@
 
-We define $\longcoxplus$ to be the transitive closure (\Cref{def:longcox-star}) and $\longcoxstar$ to be
-reflexive-transitive closure (\Cref{def:longcox-plus})) of $\longcox$.
+For the relation $\longcox$, we define its transitive closure to be $\longcoxplus$ (\cref{def:longcox-plus}) and its
+reflexive-transitive closure to be $\longcoxstar$ (\cref{def:longcox-star}).
 
@@
 Newman's lemma, it produces a unique normal form. We follow the terminology
-of~\cite*{huetConfluentReductionsAbstract1980a,krausCoherenceWellFoundednessTaming2020} to state our results
-formally.~\vc{check!}
+of~\cite*{huetConfluentReductionsAbstract1980a,krausCoherenceWellFoundednessTaming2020} to state our results formally.
 
@@
 
-% Note that this relation $\cox*$ is not prop-valued, that is, $w \cox* v$ is not a proposition because reductions need
-% not be unique. So, the quotient $\Sn$ is not effective, that is, $\quotrel : w \cox* v \to q(w) \id q(v)$ is not an
-% equivalence. We could instead define equivalence classes on $\List[\Fin[n]]$ to be those terms that have the same normal
-% form, using the $\normf$ function. Defining a new relation $(w \approx v) \defeq (\normf(w) \id \normf(v))$ which is
-% prop-valued, we could quotient $\List[\Fin[n]]$ by $\approx$. This is an equivalent definition for $\Sn$.
-
 Note that this relation $\cox*$ is not prop-valued, that is, $w \cox* v$ is not a proposition because reductions need
@@
 have to compute their normal forms using $\normf$. They correspond to the same permutation if and only if these normal
-forms are equal.
+forms are equal, which is decidable for $\List[\Fin[n]]$.
 
@@
 Lehmer codes are known in Combinatorial Analysis~\cite{bellmanCombinatorialAnalysis1960} where they are sometimes called
-"subexcedant sequences", or "factoriadics", which give the factorial number system. They are a convenient way of
-representing permutations on a computer, partly because they are
+"subexcedant sequences", or "factoriadics". They can be written as a decimal number in the factorial number system, or
+as a tuple encoding the digits~\cite*{knuthArtComputerProgramming1997,laisantNumerationFactorielleApplication1888}. This
+gives a convenient way of representing permutations on a computer, partly because they are
 bitwise-optimal~\cite{bergerTeachingOrdinalPatterns2019a}.
@@
   \begin{align*}
-    \immersion(\zero, \zero)    & \defeq \nil                                              \\
-    \immersion(\suc[n], (r, l)) & \defeq \immersion(n, l) \append ((\suc[n] - r) \downf r)
+    \immersion_{\zero}(\zero)    & \defeq \nil                                              \\
+    \immersion_{\suc[n]}((r, l)) & \defeq \immersion_{n}(l) \append ((\suc[n] - r) \downf r)
   \end{align*}
@@
 
-We can show that the function $\immersion$ gives an equivalence betweeen $\Lehmer[n]$ and $\im{\normf}$.
+We can show that the function $\immersion_{n}$ gives an equivalence betweeen $\Lehmer[n]$ and $\im{\normf}$.
 
-\begin{proposition}
+\begin{propositionrep}
   \leavevmode
   \begin{enumerate}
-    \item For any Lehmer code $c$, $\immersion(c)$ is a normal form with resepct to $\longcoxstar$, that is, $\immersion(c)$ is
-          in $\im{\normf}$.
+    \item For any Lehmer code $c$, $\immersion_{n}(c)$ is a normal form with resepct to $\longcoxstar$, that is,
+          $\immersion_{n}(c)$ is in $\im{\normf}$.
     \item Any element of $\im{\normf}$ can be constructed from a unique Lehmer code by $\immersion$, that is, the fibers
-          of $\immersion: \Lehmer[n] \to {\im{\normf}}$ are contractible.
+          of $\immersion_{n} : \Lehmer[n] \to {\im{\normf}}$ are contractible.
   \end{enumerate}
   Therefore, there is an equivalence between $\Lehmer[n]$ and $\im{\normf}$.
-\end{proposition}
+\end{propositionrep}
 
 \begin{proof}
-\jk{Only for part 1 of the proposition!}
   For any code $c : \Lehmer[n]$, we have that
   \[
-  \immersion(n, c) = (r_0 \downf k_0) \append (r_1 \downf k_1) \append \dots \append (r_{m-1} \downf k_{m-1})
+    \immersion_{n}(c) = (r_0 \downf k_0) \append (r_1 \downf k_1) \append \dots \append (r_{m-1} \downf k_{m-1})
   \]
-is concatenation of $m \geq 0$ non-empty strictly decreasing lists. Reduction $\longcox$ cannot happen inside any of the strictly
-decreasing $(r \downf k)$: $\longcancel$ requires repeating elements, $\longswap$ acts when consecutive numbers differ
-by at least 2, and $\longbraid$ acts on a non-monotone sequence. This leaves the case of reduction happening on a
-fragment that borders two (or more) subsequences. Again, $\longcancel$ requires two equal consecutive numbers, which
-would then have to be the last one in some $(r_i \downf k_i)$ sequence, and the first one in the next $(r_{i+1} \downf k_{i+1})$.
-But the first number in a sequence $(r_{i+1} \downf k_{i+1})$ is larger than every number in $(r_{i} \downf k_i)$ -- which also
-shows why $\longswap$ cannot happen. The remaining case of $\longbraid$ follows similarly, since its argument is a
-decreasing sequence followed immediately by a number equal to the first element of this sequence.
+  is a concatenation of $n \geq m \geq 0$ non-empty strictly decreasing lists. Reductions in $\longcox$ cannot happen
+  inside any of the strictly decreasing $(r \downf k)$: $\longcancel$ requires repeating elements, $\longswap$ acts when
+  consecutive numbers differ by at least 2, and $\longbraid$ acts on a non-monotone sequence. This leaves the case of
+  reduction happening on a fragment that borders two (or more) subsequences. Again, $\longcancel$ requires two equal
+  consecutive numbers, which would then have to be the last one in some $(r_i \downf k_i)$ sequence, and the first one
+  in the next $(r_{i+1} \downf k_{i+1})$. But the first number in a sequence $(r_{i+1} \downf k_{i+1})$ is larger than
+  every number in $(r_{i} \downf k_i)$ -- which also shows why $\longswap$ cannot happen. The remaining case of
+  $\longbraid$ follows similarly, since its argument is a decreasing sequence followed immediately by a number equal to
+  the first element of this sequence.
 \end{proof}
@@
 To do so, we need to construct some equivalences by counting the elements of $\Fin[n]$ using its decidable equality.
-First, we define a helper type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except
+First, we define a type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except
 the one in the argument. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence an
@@
 
-We state and prove a few auxiliary lemmas about how $\FinExcept{n}$ interacts with $\Fin$.
+We state and prove a few auxiliary lemmas about how $\FinExcept{n}$ interacts with $\Fin$, these are obtained by
+counting arguments using the fact that $\Fin[n]$ and $\FinExcept{n}$ both have decidable equality.
 
-\begin{proposition}~\label{prop:fin-finexcept}
+\begin{propositionrep}~\label{prop:fin-finexcept}
   \leavevmode
@@
   \end{enumerate}
-\end{proposition}
+\end{propositionrep}
 
-\begin{appendixproof}
+\begin{proof}
   The first and second propositions follow from simply constructing the bijections using the decidable equality of
@@
   element $n$ at the position $n - k$, resulting in the element $n$ having inversion count $k$.
-\end{appendixproof}
+\end{proof}
 

### 2021-07-08 Pi+/paper/equivalence.tex
Revise some things in the appendix

@@
 As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers, and,
-we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal strucuture. Since each object is a
+we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal structure. Since each object is a
 natural number generated by $\zero$ and $\suc$, this makes $\PiHatCat$ similar to a \emph{PROP}, that is, a products and
@@
 
-\section{Normalisation of reversible circuits}
+\section{Normalisation of Reversible Circuits}
 \label{sec:applications}
 
-Using our semantics, we can normalise a circuit writen in $\PiLang$ -- we first translate it to $\PiHatLang$, then
-compute a permutation, and quote it back to $\PiPlusLang$. Note that the normalisation happens in going from
-$\PiHatLang$ to $\UFin$ and back to $\PiHatLang$.
+Using our semantics, we can normalise, synthesise, prove equivalence, and generally reason about~$\PiLang$
+programs. The two key definitions are presented below.
 
-\begin{definition}
+\begin{definition}[Normalisation of $\PiLang$ programs]
   \begin{gather*}
@@
 
-We revisit the examples in~\Cref{sec:examples} but instead of working with circuits operating on sequences of bits, we
-work with circuits operating on algebraic data types in $\PiLang$.
+\noindent Normalisation consists of translating $\PiLang$ programs to $\PiHatLang$, computing a permutation, and quoting
+back to $\PiPlusLang$. Note that the normalisation happens in the step from $\PiHatLang$ to $\UFin$ and back to
+$\PiHatLang$ and that normalisation also provides a decision procedure for program equivalence. Synthesis consists of
+quoting permutations. More general, user-guided, reasoning can be done using the sound and complete level-2
+combinators to rewrite $\PiLang$ programs.
 
-\paragraph*{Normalisation by evaluation.} Consider the following reversible function: $\Afun{RESET}(b,b_1,\ldots,b_n) =
-(b \; \underline{\vee} \; (\bigvee_{i=1}^n b_i),b_1,\ldots,b_n)$ where $\vee$ is logical-or and $\underline{\vee}$ is
-exclusive-or. The function sets or resets the first bit depending on whether any of the other bits is true or not. The
-natural definition for this function is recursive: modulo some re-shuffling of the bits, the strategy is to examine the
-bits $b_i$ one-by-one: if we encounter a true value, we negate the first bit and terminate; otherwise we continue making
-recursive calls until we reach the last bit at which point we return the incoming value with no change:
+Recall the specification of reversible disjunction from \Cref{sec:examples}, the two Qiskit circuits for implementing it, and the
+corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The normal for both circuit is
+computed to be, establishing their equivalence:
+
+\medskip
+\resetnormtwo{}
+
+Instead of manually producing $\Pi$-programs to implement the reversible disjunction specification, it is also possible
+to simply quote the desired permutation:
+
+\medskip
+\resetperm{}
+
+\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
+\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
+which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Quoting this
+permutation generates the same normalised program which can then be composed with a map from $\PiLang$ to
+$\PiPlusLang$ to produce a program matching the desired structured types.
+
+% \paragraph*{Program synthesis.} The NbE process embodies a quoting mechanism that synthesizes programs from
+% permutations. Indeed, instead of writing a program for \Afun{reset 2}, one could simply specify the desired permutation
+% as:
 
@@
 
-\resettwo{}
+%% \resettwo{}
 
-\noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
-provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
-reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the introduction.
-Our Agda infrastructure provides tools to normalise all programs to their normal form. For \Afun{reset 2} we get:
+% \noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
+% provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
+% reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the introduction.
 
-\resetnormtwo{}
-
-In principle, the normalised program can be produced following two strategies: (i) by repeatedly applying the rewrite
-rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
-using a normalisation-by-evaluation (NbE) process that evaluates the program to a permutation on a finite set of 8
-elements and reifies that permutation back to a program. The key idea of the NbE process is a systematic way to express
-permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
-left is compiled to the sequence of four adjacent transpositions on the right:
+% In principle, the normalised program can be produced following two strategies: (i) by repeatedly applying the rewrite
+% rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
+% using a NbE process that evaluates the program to a permutation on a finite set of 8
+% elements and reifies that permutation back to a program. The key idea of the NbE process is a systematic way to express
+% permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
+% left is compiled to the sequence of four adjacent transpositions on the right:
 
@@
 
-\paragraph*{Program synthesis.} The NbE process embodies a quoting mechanism that synthesizes programs from
-permutations. Indeed, instead of writing a program for \Afun{reset 2}, one could simply specify the desired permutation
-as:
+% \paragraph*{Program equivalence.} The permutation above reveals another way to think about the desired program: it is a
+% special addition circuit that keeps 0 and 4 fixed but otherwise adds 4 modulo 8 to its input. From this specification,
+% one can use a standard synthesis algorithm for reversible circuits~\cite{10.1145/775832.775915} to generate the
+% following program:
 
-\resetperm{}
+% \adder{}
 
-\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
-\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
-which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Simply
-``quoting'' this permutation generates the same normalised program \Afun{reset2Norm} above.
-
-\paragraph*{Program equivalence.} The permutation above reveals another way to think about the desired program: it is a
-special addition circuit that keeps 0 and 4 fixed but otherwise adds 4 modulo 8 to its input. From this specification,
-one can use a standard synthesis algorithm for reversible circuits~\cite{10.1145/775832.775915} to generate the
-following program:
-
-\adder{}
-
-\noindent The \Afun{adder3} program looks nothing like the original \Afun{reset 2} program and yet they both have the
-same normal form thus establishing their equivalence. The reader can check that this circuit is the same as
-\verb|reversibleOr1| from the introduction.
+% \noindent The \Afun{adder3} program looks nothing like the original \Afun{reset 2} program and yet they both have the
+% same normal form thus establishing their equivalence. The reader can check that this circuit is the same as
+% \verb|reversibleOr1| from the introduction.
 

### 2021-07-08 Pi+/paper/discussion.tex
Fixes in sec finite
working on sec 8

@@
-\section{Discussion \& Related Work}~\label{sec:discussion}
+\section{Discussion \& Related Work}
+\label{sec:discussion}
 
-In this paper, we \ldots
+% In this paper, we \ldots
 
@@
 
-% Let's think of the $\PiLang$ combinators as describing the inhabitants of the dentity type of finite types.
+% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
 
@@
 The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
-The $\term{ua}$ map corresponds to synthesing a program from an equivalence (which, in general, is of course
+The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which, in general, is of course
 undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
@@
 a constructive argument. Then, the computation rule $\term{ua-\beta}$ expresses the fact that program synthesis is
-sound, while $\term{ua-\eta}$ corresponds to the soundess of the equational theory ($\PiLang$ 2-combinators). Thus, we
-this suggests a new computational interpretation of the univalence principle, which provides an intution on why certain
+sound, while $\term{ua-\eta}$ corresponds to the soundness of the equational theory ($\PiLang$ 2-combinators). Thus, we
+this suggests a new computational interpretation of the univalence principle, which provides an intuition on why certain
 constructions are hard (or impossible in general case).
@@
 
-\begin{center}
-    \begin{tabular}{c c c}
-        \hline
-        $\PiLang$     & $\Sn$               & $\UFin$     \\
-        \hline
-        types         & Natural numbers     & Finite sets \\
-        \hline
-        1-combinators & Generators of $\Sn$ & 1-paths     \\
-        \hline
-        2-combinators & Relations of $\Sn$  & 2-paths     \\
-        \hline
-    \end{tabular}
-\end{center}
+% \begin{center}
+%     \begin{tabular}{c c c}
+%         \hline
+%         $\PiLang$     & $\Sn$               & $\UFin$     \\
+%         \hline
+%         types         & Natural numbers     & Finite sets \\
+%         \hline
+%         1-combinators & Generators of $\Sn$ & 1-paths     \\
+%         \hline
+%         2-combinators & Relations of $\Sn$  & 2-paths     \\
+%         \hline
+%     \end{tabular}
+% \end{center}
 
-\begin{center}
-    \begin{tabular}{c c c}
-        \hline
-        Logic                              & Program                       & Algebra                     \\
-        \hline
-        Propositions                       & Types                         & Objects                     \\
-        \hline
-        Proofs                             & Terms                         & Morphisms                   \\
-        \hline
-        $\blank$                           & Equations                     & Homotopies                  \\
-        \hline
-        Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
-        \hline
-        Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
-        \hline
-        Connectives                        & Type constructors             & Bifunctors                  \\
-        \hline
-        ?                                  & Type normalization            & ?                           \\
-        \hline
-        ?                                  & Program normalization         & ?                           \\
-        \hline
-        ?                                  & Soundness of 2-combinators    & ?                           \\
-    \end{tabular}
-\end{center}
+% \begin{center}
+%     \begin{tabular}{c c c}
+%         \hline
+%         Logic                              & Program                       & Algebra                     \\
+%         \hline
+%         Propositions                       & Types                         & Objects                     \\
+%         \hline
+%         Proofs                             & Terms                         & Morphisms                   \\
+%         \hline
+%         $\blank$                           & Equations                     & Homotopies                  \\
+%         \hline
+%         Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
+%         \hline
+%         Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
+%         \hline
+%         Connectives                        & Type constructors             & Bifunctors                  \\
+%         \hline
+%         ?                                  & Type normalisation            & ?                           \\
+%         \hline
+%         ?                                  & Program normalisation         & ?                           \\
+%         \hline
+%         ?                                  & Soundness of 2-combinators    & ?                           \\
+%     \end{tabular}
+% \end{center}
 
-\begin{center}
-    \begin{tabular}{c c c}
-        \hline
-        Reversible Logic             & Reversible Programming                  & Algebra                  \\
-        \hline
-        Free commutative rig         & Free symmetric rig groupoid             & $\UFin$                  \\
-        \hline
-        Propositions                 & Types                                   & Objects                  \\
-        \hline
-        Entailment?                  & Reversible programs                     & Isomorphisms             \\
-        \hline
-        $\blank$                     & 2-combinators                           & Homotopies               \\
-        \hline
-        Conjunction?                 & $\times$                                & $\otimes$                \\
-        \hline
-        Disjunction?                 & $+$                                     & $\oplus$                 \\
-        \hline
-        $A \land B \vdash B \land A$ & $\swapt$                                & $\mathcal{B}_{A,B}$      \\
-        \hline
-        True proposition             & $\onet$                                 & $F_1$                    \\
-        \hline
-        Decidability                 & Program Synthesis                       & Reachability in a graph? \\
-        \hline
-        Soundness/Completeness       & Soundness/Completeness of 1-combinators & ?                        \\
-    \end{tabular}
-\end{center}
+% \begin{center}
+%     \begin{tabular}{c c c}
+%         \hline
+%         Reversible Logic             & Reversible Programming                  & Algebra                  \\
+%         \hline
+%         Free commutative rig         & Free symmetric rig groupoid             & $\UFin$                  \\
+%         \hline
+%         Propositions                 & Types                                   & Objects                  \\
+%         \hline
+%         Entailment?                  & Reversible programs                     & Isomorphisms             \\
+%         \hline
+%         $\blank$                     & 2-combinators                           & Homotopies               \\
+%         \hline
+%         Conjunction?                 & $\times$                                & $\otimes$                \\
+%         \hline
+%         Disjunction?                 & $+$                                     & $\oplus$                 \\
+%         \hline
+%         $A \land B \vdash B \land A$ & $\swapt$                                & $\mathcal{B}_{A,B}$      \\
+%         \hline
+%         True proposition             & $\onet$                                 & $F_1$                    \\
+%         \hline
+%         Decidability                 & Program Synthesis                       & Reachability in a graph? \\
+%         \hline
+%         Soundness/Completeness       & Soundness/Completeness of 1-combinators & ?                        \\
+%     \end{tabular}
+% \end{center}
 
@@
 
+\todo{Why is this better than evaluating the permutation and checking each bit?}
+\todo{Why care about an equational theory? Compilers, optimisation.}
+
 %%% Local Variables:

### 2021-07-08 Pi+/paper/finite.tex
Fix group def

@@
 
-A group homomorphism between groups $G$ and $H$, which we write as $G \to^G H$, is a function between the underlying
-sets that preserves the group structure. Giving a group homomorphism out of the free group is equivalent to giving a
-function out of the generating set. This is the universal property of free groups, stemming from the free-forgetful
-adjunction between the category of groups and sets.
+A group homomorphism between groups $G$ and $H$ is a function $f : G \to H$ between the underlying sets that preserves
+the group structure. Giving a group homomorphism out of the free group is equivalent to giving a function out of the
+generating set. This is the universal property of free groups, stemming from the free-forgetful adjunction between the
+category of groups and sets.
 
@@
   \label{prop:free-groups}
-  Given a group $G$ and a map $f : A \to G$, there is a unique group
-  homomorphism $\extend{f} : F(A) \to^G G$ such that $\extend{f} \comp \eta_A
-    \htpy f$. Equivalently, composition with $\eta_A$ gives an equivalence $F(A)
-    \to^G G \eqv A \to G$. Alternatively, the type of group homomorphisms $h :
-    F(A) \to^G G$ satisfying $h \comp \eta_A \htpy f$ is contractible.
+  Given a group $G$ and a map $f : A \to G$, there is a unique group homomorphism $\extend{f} : \term{Hom}(F(A), G)$
+  such that $\extend{f} \comp \eta_A \htpy f$. Equivalently, composition with $\eta_A$ gives an equivalence
+  $\term{Hom}(F(A),G) \eqv A \to G$.
 \end{propositionrep}
@@
 \begin{toappendix}
+ Alternatively, the type of group homomorphisms $h : \term{Hom}(F(A),G)$ satisfying $h \comp \eta_A \htpy f$ is
+ contractible.
+
   % https://q.uiver.app/?q=WzAsMyxbMCwyLCJBIl0sWzAsMCwiRihBKSJdLFsyLDAsIkciXSxbMCwxLCJcXGV0YV9BIl0sWzAsMiwiZiIsMl0sWzEsMiwiXFxleHRlbmR7Zn0iLDAseyJzdHlsZSI6eyJib2R5Ijp7Im5hbWUiOiJkYXNoZWQifX19XSxbMyw0LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d

### 2021-07-08 Pi+/paper/discussion.tex
working on sec 8

@@
 
-\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
-types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
-$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to
-    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p ==
-    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
-$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
+The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
+presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
+semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible
+languages. Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
+cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
+correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence is
+established in this work.
 
-% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
-
-By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
-syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse.
-% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
-The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
-The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which, in general, is of course
-undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
-we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
-a constructive argument. Then, the computation rule $\term{ua-\beta}$ expresses the fact that program synthesis is
-sound, while $\term{ua-\eta}$ corresponds to the soundness of the equational theory ($\PiLang$ 2-combinators). Thus, we
-this suggests a new computational interpretation of the univalence principle, which provides an intuition on why certain
-constructions are hard (or impossible in general case).
-
-We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
-$\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
-doesn't have one, but we could show externally that it satisfies univalence.
-
-Our work lies at the intersection of programming language theory, category theory, group theory, rewriting theory, and
-formalised mathematics. We review related work in the literature for each topic.
-
-\paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe algebraic structures, such
-as groups or rings. A specific group or ring is a model of the appropriate algebraic theory. Algebraic theories are
-usually \emph{presented} in terms of logical syntax, that is, as first-order theories whose signatures allow only
-functional symbols, and whose axioms are universally quantified equations. In his seminal
-thesis~\cite{lawvereFUNCTORIALSEMANTICSALGEBRAIC1963}, Lawvere defined a presentation-free categorical notion of
-universal algebraic structure, called a Lawvere theory.
-
-Programming Languages, such as the $\lambda$-calculus, can be viewed as algebraic structures with variable-binding
-operators, which can be formalised using second-order algebraic theories~\cite{fioreSecondOrderAlgebraicTheories2010},
-or algebraic theories with closed structure~\cite{hylandClassicalLambdaCalculus2017}, called $\lambda$-theories, making
-the $\lambda$-calculus the presentation of the initial $\lambda$-theory $\Lambda$.
-
-Our family of reversible languages have been presented as first-order algebraic
-2-theories~\cite{cohenCoherenceRewriting2theories2009,bekeCategorificationTermRewriting2011,yanofskySyntaxCoherence2000},
-which are a categorification of algebraic theories. The types $\zerot$ and $\onet$ are nullary function symbols, the
-type formers $+$ and $\times$ are binary function symbols, the 1-combinators are invertible reduction rules, and the
-2-combinators are equations or coherence diagrams of compositions of reduction rules. Just like models of Lawvere
-theories are given by algebras of (finitary) monads on $\SetCat$, models of 2-theories are given by algebras of 2-monads
-on $\CatCat$. The particular one we're interested in here is the free symmetric monoidal completion 2-monad.
-
-\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of (small)
-symmetric monoidal categories, strong symmetric monoidal functors, and monoidal natural transformations, to the
-2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
-$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict)
-symmetric monoidal categories. Its construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.
-Concretely, the objects of $\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n]
-    \to \CCat{C}_{0})$. An morphism between $(n,A)$ and $(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a
-permutation of $[n]$, and $\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the
-Grothendieck construction $\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and
-bijections to $\CatCat$, assigning each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on
-$[n]$ inducing an endofunctor on $C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one
-generator, $\FSM[\unit]$. The free symmetric monoidal category has been used to study
-concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri nets~\cite{baezCategoriesNets2021}, combinatorial
-structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005},
-bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}.
-
-Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories was
-studied in~\cite{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a groupoid
-and the proof of its universal property has been considered in~\cite{piceghelloCoherenceMonoidalGroupoids2020}.
-
-Free commutative monoids in type theory have been studied in~\cite{gylterudMultisetsTypeTheory2020}, and using HoTT
-in~\cite{choudhuryFinitemultisetConstructionHoTT2019}. The free symmetric monoidal groupoid $\FSM[A]$ over a groupoid
-$A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs. These HITs
-and the proof of their universal property have been considered
-in~\cite*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.
-
-The proof of the universal property of the $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric
-monoidal categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite
-set. The existence of the proof is folklore, and we have produced a new proof of it while working in constructive type
-theory.
-
-\paragraph{Curry-Howard-Lambek correspondence} In~\citet{curryCurryEssaysCombinatory1980}, Lambek extended the
-Curry-Howard correspondence to cartesian-closed categories. In this work, we have established a correspondence between a
-fragment of the $\PiLang$ family of reversible programming languages and symmetric monoidal groupoids. The Curry-Howard
-part of this correspondence with reversible logic was established in~\cite{sparksSuperstructuralReversibleLogic2014}.
+In the remainder of this section, we discuss some broader related work.
 
@@
 
+%%%%%%%%%%%%%%
 \paragraph{Rewriting} We presented a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.
@@
 
-\paragraph{Computational group theory} Coxeter relations are used in computational group theory to study XXX problems.
-
-\paragraph{Univalent Fibrations} Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
-to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
-\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
-$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
-subuniverses.
-
-Coherence problems in type theory, coherence via Well-Foundedness.
-
-Formalised proofs of Mac Lane's coherence theorem.
-
-Applications of FSMG and history of the coherence theorems.
-
-Other proofs of coherence theorems, Joyal-Street.
-
-Pi has other extensions (fractional/negative/recursive types).
-What are the free X monoidal structures they're describing?
-
-Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
-should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.
-
-Other applications of symmetric groups.
-
-Actions of symmetric groups, permutation groupoids.
-
+%%%%%%%%%%%%%%
 \paragraph{Extensional vs. Intensional View of Permutations.}
@@
 
+%%%%%%%%%%%%%%
+\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of (small)
+symmetric monoidal categories, strong symmetric monoidal functors, and monoidal natural transformations, to the
+2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
+$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict)
+symmetric monoidal categories. Its construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.
+Concretely, the objects of $\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n]
+    \to \CCat{C}_{0})$. An morphism between $(n,A)$ and $(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a
+permutation of $[n]$, and $\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the
+Grothendieck construction $\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and
+bijections to $\CatCat$, assigning each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on
+$[n]$ inducing an endofunctor on $C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one
+generator, $\FSM[\unit]$. The free symmetric monoidal category has been used to study
+concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri nets~\cite{baezCategoriesNets2021}, combinatorial
+structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005},
+bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}.
+
+Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories was
+studied in~\cite{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a groupoid
+and the proof of its universal property has been considered in~\cite{piceghelloCoherenceMonoidalGroupoids2020}.
+
+Free commutative monoids in type theory have been studied in~\cite{gylterudMultisetsTypeTheory2020}, and using HoTT
+in~\cite{choudhuryFinitemultisetConstructionHoTT2019}. The free symmetric monoidal groupoid $\FSM[A]$ over a groupoid
+$A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs. These HITs
+and the proof of their universal property have been considered
+in~\cite*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.
+
+The proof of the universal property of the $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric
+monoidal categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite
+set. The existence of the proof is folklore, and we have produced a new proof of it while working in constructive type
+theory.
+
+%%%%%%%%%%%%%%
+\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
+types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
+$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to
+    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p ==
+    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
+$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
+
+% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
+
+By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
+syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse.
+% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
+The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
+The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which, in general, is of course
+undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
+we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
+a constructive argument. Then, the computation rule $\term{ua-\beta}$ expresses the fact that program synthesis is
+sound, while $\term{ua-\eta}$ corresponds to the soundness of the equational theory ($\PiLang$ 2-combinators). Thus, we
+this suggests a new computational interpretation of the univalence principle, which provides an intuition on why certain
+constructions are hard (or impossible in general case).
+
+We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
+$\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
+doesn't have one, but we could show externally that it satisfies univalence.
+
+%%%%%%%%%%%%%%
+\paragraph{Univalent Fibrations} Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
+to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
+\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
+$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
+subuniverses.
+
+Coherence problems in type theory, coherence via Well-Foundedness.
+
+Formalised proofs of Mac Lane's coherence theorem.
+
+Applications of FSMG and history of the coherence theorems.
+
+Other proofs of coherence theorems, Joyal-Street.
+
+Pi has other extensions (fractional/negative/recursive types).
+What are the free X monoidal structures they're describing?
+
+Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
+should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.
+
+Other applications of symmetric groups.
+
+Actions of symmetric groups, permutation groupoids.
+
+%%%%%%%%%%%%%%
+\paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe algebraic structures, such
+as groups or rings. A specific group or ring is a model of the appropriate algebraic theory. Algebraic theories are
+usually \emph{presented} in terms of logical syntax, that is, as first-order theories whose signatures allow only
+functional symbols, and whose axioms are universally quantified equations. In his seminal
+thesis~\cite{lawvereFUNCTORIALSEMANTICSALGEBRAIC1963}, Lawvere defined a presentation-free categorical notion of
+universal algebraic structure, called a Lawvere theory.
+
+Programming Languages, such as the $\lambda$-calculus, can be viewed as algebraic structures with variable-binding
+operators, which can be formalised using second-order algebraic theories~\cite{fioreSecondOrderAlgebraicTheories2010},
+or algebraic theories with closed structure~\cite{hylandClassicalLambdaCalculus2017}, called $\lambda$-theories, making
+the $\lambda$-calculus the presentation of the initial $\lambda$-theory $\Lambda$.
+
+Our family of reversible languages have been presented as first-order algebraic
+2-theories~\cite{cohenCoherenceRewriting2theories2009,bekeCategorificationTermRewriting2011,yanofskySyntaxCoherence2000},
+which are a categorification of algebraic theories. The types $\zerot$ and $\onet$ are nullary function symbols, the
+type formers $+$ and $\times$ are binary function symbols, the 1-combinators are invertible reduction rules, and the
+2-combinators are equations or coherence diagrams of compositions of reduction rules. Just like models of Lawvere
+theories are given by algebras of (finitary) monads on $\SetCat$, models of 2-theories are given by algebras of 2-monads
+on $\CatCat$. The particular one we're interested in here is the free symmetric monoidal completion 2-monad.
+
+% %%%%%%%%%%%%%%
+% \paragraph{Computational group theory} Coxeter relations are used in computational group theory to study XXX problems.
+
+% \todo{but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
+
 %%% Local Variables:

### 2021-07-08 Pi+/paper/discussion.tex
Abstract, some fixes

--- 2021-07-07 Pi+/paper/discussion.tex
@@
 types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
-$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to
-    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p ==
+$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
+    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
     \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
-$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
+$\term{ua_\beta}$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
 
@@
 we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
-a constructive argument. Then, the computation rule $\term{ua-\beta}$ expresses the fact that program synthesis is
-sound, while $\term{ua-\eta}$ corresponds to the soundess of the equational theory ($\PiLang$ 2-combinators). Thus, we
+a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact that program synthesis is
+sound, while $\term{ua_\eta}$ corresponds to the soundess of the equational theory ($\PiLang$ 2-combinators). Thus, we
 this suggests a new computational interpretation of the univalence principle, which provides an intution on why certain
-constructions are hard (or impossible in general case).
+constructions are hard (or impossible in the general case).
 
@@
         \hline
-        ?                                  & Type normalization            & ?                           \\
+        ?                                  & Type normalisation            & ?                           \\
         \hline
-        ?                                  & Program normalization         & ?                           \\
+        ?                                  & Program normalisation         & ?                           \\
         \hline
@@
 \paragraph{Computational group theory} Coxeter relations are used in computational group theory to study XXX problems.
+
+\todo{but computation is usually done by using Coxeter matrices~\cite{davisGeometryTopologyCoxeter2008}.}
 

### 2021-07-08 Pi+/paper/discussion.tex

--- 2021-07-08 Pi+/paper/discussion.tex
@@
 types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
-$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua-\beta} : (e : A \eqv B) \to
-    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua-\eta} : (p : A \id_{\UU} B) \to p ==
+$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
+    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
     \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
@@
 we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
-a constructive argument. Then, the computation rule $\term{ua-\beta}$ expresses the fact that program synthesis is
-sound, while $\term{ua-\eta}$ corresponds to the soundness of the equational theory ($\PiLang$ 2-combinators). Thus, we
+a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact that program synthesis is
+sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$ 2-combinators). Thus, we
 this suggests a new computational interpretation of the univalence principle, which provides an intuition on why certain
-constructions are hard (or impossible in general case).
+constructions are hard (or impossible in the general case).
 

### 2021-07-09 Pi+/paper/discussion.tex
working on sec 8

@@
 %%%%%%%%%%%%%%
-\paragraph{Rewriting} We presented a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.
-
-There exists an algorithm, due to~\citet{knuthSimpleWordProblems1970}, that, when succeeds, constructs a well-behaved
-rewriting system for an arbitrary finite set of (undirected) equations. It did not work for us, producing too many
-equations, and proving correctness and termination was intractable.
-
-We chose to encode permutations as adjacent transpositions corresponding to words in $\Sn$, since that closely
-corresponds to the syntax of $\PiLang$ combinators. Other representations of permutations would give a different
-denotational semantics for $\PiLang$ combinators. Permutations can be encoded as listed vectors or matrices, inductively
-generated trees (Motzkin trees), Young diagrams, or String diagrams, but the difficulty of formalising them in type
-theory varies depending on the encoding.
-
-Rewriting systems and word problems have a long history of being formalised in proof assistants. In the recent past,
-higher order rewriting systems have been formalised in proof assistants like {homotopy.io} and Lean and Coq. The use of
-HoTT to study rewriting has been considered in~\cite{krausCoherenceWellFoundednessTaming2020}.
+\paragraph{Rewriting} Term-rewriting systems and word problems have a long history of being formalised in proof
+assistants. In the recent past, higher order rewriting systems have been formalised in {homotopy.io}, Lean, Coq, and
+HoTT~\citet{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main result, we developed a rewriting
+system for the Coxeter relations for $\Sn$ to solve its word problem and formalised its correctness and termination in a
+proof assistant. Our design based on encoding permutations as adjacent transpositions led to an intuitive whose
+correctness was straightforward to establish and whose termination only required slightly modified variations on
+Church-Rosser theorems. Other encodings of permutations as listed vectors, matrices, inductively generated trees
+(Motzkin trees), Young diagrams, or string diagrams, proved either difficult formalise in type theory or difficult to
+relate directly to the primitive type isomorphisms of $\PiLang$. The automatic \citet{knuthSimpleWordProblems1970}
+produced too many equations making proving correctness and termination intractable.
 

### 2021-07-09 Pi+/paper/discussion.tex

@@
 %%%%%%%%%%%%%%
-\paragraph{Rewriting} Term-rewriting systems and word problems have a long history of being formalised in proof
-assistants. In the recent past, higher order rewriting systems have been formalised in {homotopy.io}, Lean, Coq, and
-HoTT~\citet{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main result, we developed a rewriting
-system for the Coxeter relations for $\Sn$ to solve its word problem and formalised its correctness and termination in a
-proof assistant. Our design based on encoding permutations as adjacent transpositions led to an intuitive whose
-correctness was straightforward to establish and whose termination only required slightly modified variations on
-Church-Rosser theorems. Other encodings of permutations as listed vectors, matrices, inductively generated trees
-(Motzkin trees), Young diagrams, or string diagrams, proved either difficult formalise in type theory or difficult to
-relate directly to the primitive type isomorphisms of $\PiLang$. The automatic \citet{knuthSimpleWordProblems1970}
-produced too many equations making proving correctness and termination intractable.
+\paragraph{Rewriting} Higher-order term-rewriting systems and word problems have a long history of being formalised in
+proof assistants like {homotopy.io}, Lean, Coq, and HoTT~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the
+proof of our main result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem
+and formalised its correctness and termination in Agda. Our design based on encoding permutations as
+adjacent transpositions led to an intuitive system whose correctness was straightforward to establish and whose termination
+only required slightly modified variations on Church-Rosser theorems. Other encodings of permutations as listed vectors,
+matrices, inductively generated trees (Motzkin trees), Young diagrams, or string diagrams, proved either difficult
+formalise in type theory or difficult to relate directly to the primitive type isomorphisms of $\PiLang$. The automatic
+\citet{knuthSimpleWordProblems1970} completion produced too many equations making proving correctness and termination
+intractable.
 

### 2021-07-09 Pi+/paper/ufin.tex

@@
   $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$, which equates every pair of points in the truncation
-  (see~\Cref{def:prop-trunc}).
+  (see~\cref{def:prop-trunc}).
 
@@
 
+  \begin{definition}[${\fib_{f}} : B \to \UU$]
+    \label{def:fib}
+    The fiber of $f : A \to B$ at $b : B$ is
+    \[
+      \fib_{f}(b) \defeq \dsum{a:A}{f(a) \id_{B} b}.
+    \]
+  \end{definition}
+
+  \begin{definition}[${\term{im}} : (f : A \to B) \to \UU$]
+    \label{def:im}
+    The image of $f$ is the (-1)-truncation of its fiber.
+    \[
+      \im{f} \defeq \dsum{b:B}{\Trunc[-1]{\fib_{f}(b)}}
+    \]
+  \end{definition}
+
+  \begin{proposition}
+    The following are equivalent.
+    \begin{enumerate}
+      \item $f : A \to B$ is an equivalence.
+      \item $f$ has a left and right inverse.
+      \item $f$ has contractible fibers.
+    \end{enumerate}
+  \end{proposition}
+
   Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes an $\hSet$ $A$ and a relation $R : A \to A \to
     \UU$. It has an inclusion of points $q : A \to \quot{A}{R}$, and adds paths between related pairs of elements
-  $\quotrel : R(x,y) \to q(x) \id_{\quot{A}{R}} q(y)$ (see~\Cref{def:set-quot}).  We recall that the quotient is
+  $\quotrel : R(x,y) \to q(x) \id_{\quot{A}{R}} q(y)$ (see~\cref{def:set-quot}).  We recall that the quotient is
   \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff $(q(x) \id_{\quot{A}{R}}
@@
 \begin{toappendix}
-  \begin{definition}
-    \label{app:defone}
+  \begin{definition}[Additive symmetric monoidal structure]
+    \label{def:additive}
     \begin{align*}
@@
   \begin{proposition}
-    \label{app:propone}
+    \label{prop:additive}
     % https://q.uiver.app/?q=WzAsNCxbMCwwLCIoWCBcXG9wbHVzIEkpIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyAoSSBcXG9wbHVzIFkpIl0sWzEsMSwiWCBcXG9wbHVzIFkiXSxbMCwxXSxbMCwxLCJcXGFscGhhX3tYLEksWX0iXSxbMCwyLCJcXHJob197WH0gXFxvcGx1cyAxX3tZfSIsMl0sWzEsMiwiMV97WH0gXFxvcGx1cyBcXGxhbWJkYV97WX0iXSxbNSw2LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
@@
 
+\begin{toappendix}
+  \begin{definition}[Multiplicative symmetric monoidal structure]
+    \label{def:multiplicative}
+    \begin{align*}
+      I                 & \defeq F_{1}                                           \\
+      X \otimes Y       & \defeq X \times Y                                      \\
+      \lambda_{X}       & : I \times X \eqv X                                    \\
+      \rho_{X}          & : X \times I \eqv X                                    \\
+      \alpha_{X,Y,Z}    & : (X \otimes Y) \otimes Z \eqv X \otimes (Y \otimes Z) \\
+      \mathcal{B}_{X,Y} & : X \otimes Y \eqv Y \otimes X
+    \end{align*}
+  \end{definition}
+  \begin{proposition}
+    \label{prop:multiplicative}
+    % https://q.uiver.app/?q=WzAsNCxbMCwwLCIoWCBcXG9wbHVzIEkpIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyAoSSBcXG9wbHVzIFkpIl0sWzEsMSwiWCBcXG9wbHVzIFkiXSxbMCwxXSxbMCwxLCJcXGFscGhhX3tYLEksWX0iXSxbMCwyLCJcXHJob197WH0gXFxvcGx1cyAxX3tZfSIsMl0sWzEsMiwiMV97WH0gXFxvcGx1cyBcXGxhbWJkYV97WX0iXSxbNSw2LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    \[\begin{tikzcd}
+        {(X \otimes I) \otimes Y} && {X \otimes (I \otimes Y)} \\
+        {} & {X \otimes Y}
+        \arrow["{\alpha_{X,I,Y}}", from=1-1, to=1-3]
+        \arrow[""{name=0, anchor=center, inner sep=0}, "{\rho_{X} \otimes 1_{Y}}"', from=1-1, to=2-2]
+        \arrow[""{name=1, anchor=center, inner sep=0}, "{1_{X} \otimes \lambda_{Y}}", from=1-3, to=2-2]
+        \arrow["\id", Rightarrow, draw=none, from=0, to=1]
+      \end{tikzcd}\]
+    % https://q.uiver.app/?q=WzAsNSxbMCwxLCIoKFcgXFxvcGx1cyBYKSBcXG9wbHVzIFkpIFxcb3BsdXMgWiJdLFsxLDAsIihXIFxcb3BsdXMgWCkgXFxvcGx1cyAoWSBcXG9wbHVzIFopIl0sWzIsMSwiVyBcXG9wbHVzIChYIFxcb3BsdXMgKFkgXFxvcGx1cyBaKSkiXSxbMiwzLCJXIFxcb3BsdXMgKChYIFxcb3BsdXMgWSkgXFxvcGx1cyBaKSJdLFswLDMsIihXIFxcb3BsdXMgKFggXFxvcGx1cyBZKSkgXFxvcGx1cyBaIl0sWzAsMSwiXFxhbHBoYV97VyBcXG9wbHVzIFgsIFksIFp9Il0sWzEsMiwiXFxhbHBoYV97VyxYLFkgXFxvcGx1cyBafSJdLFszLDIsIjFfe1d9IFxcb3BsdXMgXFxhbHBoYV97WCxZLFp9IiwyXSxbMCw0LCJcXGFscGhhX3tXLFgsWX0gXFxvcGx1cyAxX3tafSIsMl0sWzQsMywiXFxhbHBoYV97VyxYIFxcb3BsdXMgWSxafSIsMl0sWzAsMiwiXFxpZCIsMSx7Im9mZnNldCI6NSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    \[\begin{tikzcd}
+        & {(W \otimes X) \otimes (Y \otimes Z)} \\
+        {((W \otimes X) \otimes Y) \otimes Z} && {W \otimes (X \otimes (Y \otimes Z))} \\
+        \\
+        {(W \otimes (X \otimes Y)) \otimes Z} && {W \otimes ((X \otimes Y) \otimes Z)}
+        \arrow["{\alpha_{W \otimes X, Y, Z}}", from=2-1, to=1-2]
+        \arrow["{\alpha_{W,X,Y \otimes Z}}", from=1-2, to=2-3]
+        \arrow["{1_{W} \otimes \alpha_{X,Y,Z}}"', from=4-3, to=2-3]
+        \arrow["{\alpha_{W,X,Y} \otimes 1_{Z}}"', from=2-1, to=4-1]
+        \arrow["{\alpha_{W,X \otimes Y,Z}}"', from=4-1, to=4-3]
+        \arrow["\id"{description}, shift right=5, draw=none, from=2-1, to=2-3]
+      \end{tikzcd}\]
+    % https://q.uiver.app/?q=WzAsNixbMSwwLCJYIFxcb3BsdXMgKFkgXFxvcGx1cyBaKSJdLFswLDEsIihYIFxcb3BsdXMgWSkgXFxvcGx1cyBaIl0sWzAsMiwiKFkgXFxvcGx1cyBYKSBcXG9wbHVzIFoiXSxbMSwzLCJZIFxcb3BsdXMgKFggXFxvcGx1cyBaKSJdLFsyLDIsIlkgXFxvcGx1cyAoWiBcXG9wbHVzIFgpIl0sWzIsMSwiKFkgXFxvcGx1cyBaKSBcXG9wbHVzIFgiXSxbMSwwLCJcXGFscGhhX3tYLFksWn0iXSxbMSwyLCJcXG1hdGhjYWx7Qn1fe1gsWX0gXFxvcGx1cyAxX3tafSIsMl0sWzIsMywiXFxhbHBoYV97WSxYLFp9IiwyXSxbMyw0LCIxX3tZfSBcXG9wbHVzIFxcbWF0aGNhbHtCfV97WCxafSIsMl0sWzUsNCwiXFxhbHBoYV97WSxaLFh9Il0sWzAsNSwiXFxtYXRoY2Fse0J9X3tYLFkgXFxvcGx1cyBafSJdLFs3LDEwLCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    \[\begin{tikzcd}
+        & {X \otimes (Y \otimes Z)} \\
+        {(X \otimes Y) \otimes Z} && {(Y \otimes Z) \otimes X} \\
+        {(Y \otimes X) \otimes Z} && {Y \otimes (Z \otimes X)} \\
+        & {Y \otimes (X \otimes Z)}
+        \arrow["{\alpha_{X,Y,Z}}", from=2-1, to=1-2]
+        \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y} \otimes 1_{Z}}"', from=2-1, to=3-1]
+        \arrow["{\alpha_{Y,X,Z}}"', from=3-1, to=4-2]
+        \arrow["{1_{Y} \otimes \mathcal{B}_{X,Z}}"', from=4-2, to=3-3]
+        \arrow[""{name=1, anchor=center, inner sep=0}, "{\alpha_{Y,Z,X}}", from=2-3, to=3-3]
+        \arrow["{\mathcal{B}_{X,Y \otimes Z}}", from=1-2, to=2-3]
+        \arrow["\id", Rightarrow, draw=none, from=0, to=1]
+      \end{tikzcd}\]
+    % https://q.uiver.app/?q=WzAsMyxbMCwwLCJYIFxcb3BsdXMgWSJdLFsyLDAsIlggXFxvcGx1cyBZIl0sWzEsMSwiWSBcXG9wbHVzIFgiXSxbMCwxLCIxX3tYIFxcb3BsdXMgWX0iLDAseyJsZXZlbCI6Miwic3R5bGUiOnsiaGVhZCI6eyJuYW1lIjoibm9uZSJ9fX1dLFswLDIsIlxcbWF0aGNhbHtCfV97WCxZfSIsMl0sWzIsMSwiXFxtYXRoY2Fse0J9X3tZLFh9IiwyXSxbNCw1LCJcXGlkIiwwLHsic2hvcnRlbiI6eyJzb3VyY2UiOjIwLCJ0YXJnZXQiOjIwfSwic3R5bGUiOnsiYm9keSI6eyJuYW1lIjoibm9uZSJ9LCJoZWFkIjp7Im5hbWUiOiJub25lIn19fV1d
+    \[\begin{tikzcd}
+        {X \otimes Y} && {X \otimes Y} \\
+        & {Y \otimes X}
+        \arrow["{1_{X \otimes Y}}", Rightarrow, no head, from=1-1, to=1-3]
+        \arrow[""{name=0, anchor=center, inner sep=0}, "{\mathcal{B}_{X,Y}}"', from=1-1, to=2-2]
+        \arrow[""{name=1, anchor=center, inner sep=0}, "{\mathcal{B}_{Y,X}}"', from=2-2, to=1-3]
+        \arrow["\id", Rightarrow, draw=none, from=0, to=1]
+      \end{tikzcd}\]
+  \end{proposition}
+\end{toappendix}
+
+\begin{toappendix}
+  \begin{proposition}[Distributivity]
+    \label{prop:distributivity}
+    \begin{gather*}
+      \begin{aligned}
+        \delta_{l} : X \otimes (Y \oplus Z) & \eqv (X \otimes Y) \oplus (X \otimes Z) \\
+        \delta_{r} : (X \oplus Y) \otimes Z & \eqv (X \otimes Z) \oplus (Y \otimes Z)
+      \end{aligned}
+      \qquad
+      \begin{aligned}
+        a_{l} : X \otimes O \eqv O \\
+        a_{r} : O \otimes X \eqv O
+      \end{aligned}
+    \end{gather*}
+  \end{proposition}
+\end{toappendix}
+
 %%% Local Variables:

### 2021-07-09 Pi+/paper/finite.tex
Remove word problem

@@
 $\Sn$, we will use a rewriting system, with a suitable, well behaved collection of reduction rules corresponding to the
-Coxeter presentation. Finally, we describe the normal forms in this rewriting system, using Lehmer codes, and prove the
+Coxeter presentation equations. Finally, we describe the normal forms in this rewriting system, using Lehmer codes, and prove the
 correspondence between them and the type $\Aut[\Fin[n]]$ of automorphisms on a finite set. The generators and relations
@@
 
-We use a presentation based on adjacent transpositions for generators, called a~\emph{Coxeter presentation}, which we
-also use to solve the word problem for $\Sn$. Connecting it back to the language $\PiLang$, we will show that it is
-possible to encode all $\PiLang$ combinators using adjacent transpositions (in~\Cref{sec:equivalence}). The rest of this
-section develops these intuitions formally. Group presentations are built by adding equations to a \emph{free group},
-which we define first.
+We will show that it is possible to encode all $\PiLang$ combinators using adjacent transpositions
+(in~\Cref{sec:equivalence}). Group presentations are built by adding equations to a \emph{free group}, which we define
+first.
 
@@
 Before, the only way to decide the equality of two elements in a group was to evaluate and check them on the nose, but
-in a group presentation, this is reduced to a \emph{word problem}, that is, deciding whether one word -- a
-representative of the equivalence class of the group's elements, can be reduced to another word, using the group's
-relations. However, these equations are not directed, so it is not always possible to construct a well-behaved rewriting
-system. In general, the word problem for groups is known to be undecidable.
+in a group presentation, this is reduced to deciding whether one word -- a representative of the equivalence class of
+the group's elements, can be reduced to another word, using the group's relations. However, these equations are not
+directed, so it is not always possible to construct a well-behaved rewriting system. In general, the word problem for
+groups is known to be undecidable.
 
@@
             \xrightarrow{r} {\List[\Fin[n]]}$ such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
-    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \), where $\quotinc$ is the inclusion into the quotient.
+    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \), where $\quotinc$ is the inclusion into the quotient, and
+    $\im{f}$ denotes the image of $f$ (see~\cref{def:im}).
   \end{enumerate}
@@
 
-We set out to solve the word problem for $\Sn$. To decide if two words in $\List[\Fin[n]]$ are $\cox*$-equal, we simply
-have to compute their normal forms using $\normf$. They correspond to the same permutation if and only if these normal
-forms are equal, which is decidable for $\List[\Fin[n]]$.
+To decide if two words in $\List[\Fin[n]]$ are $\cox*$-equal, we simply have to compute their normal forms using
+$\normf$. They correspond to the same permutation if and only if these normal forms are equal, which is decidable for
+$\List[\Fin[n]]$.
 

### 2021-07-09 Pi+/paper/ufin.tex
Fix crefs

@@
 satisfying univalence, to get the groupoid structure. In this section, we formally define \emph{univalent subuniverses},
-and proceed to construct the particular instance for finite types, $\UFin$ (\Cref{def:ufin}).
+and proceed to construct the particular instance for finite types, $\UFin$ (\cref{def:ufin}).
 
@@
   of equalities or identifications between them. The identity type is generated by reflexivity $\refl_{x} : x \id_{A} x$,
-  and the eliminator for the identity type is given by path induction or the $J$-rule (\Cref{def:path-induction}). This
+  and the eliminator for the identity type is given by path induction or the $J$-rule (\cref{def:path-induction}). This
   construction can be iterated, giving the identity type between two terms of an identity type, repeating ad infinitum.
@@
 subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness: a type is finite if it is merely
-equivalent to a finite set (\Cref{def:finite-set} and \Cref{def:isfin}).
+equivalent to a finite set (\cref{def:finite-set} and \cref{def:isfin}).
 

### 2021-07-09 Pi+/paper/discussion.tex
more on sec 8

@@
 cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
-correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence is
-established in this work.
+correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
+suggested by~\citet{Carette2016} and~\citet{threemodels} and established in this work.
 
@@
 intractable.
+
+%%%%%%%%%%%%%%
+\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
+types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
+$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
+    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
+    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
+$\term{ua}_\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
+
+% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
+
+By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
+via a complete equational axiomatisation of the equivalences between types in the subuniverse.
+% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
+In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
+which is easily done by induction.  The $\term{ua}$ map corresponds to synthesising a program from an equivalence
+(which, in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible
+boolean circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the
+existence of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses
+the fact that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory
+($\PiLang$ 2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
+provides an intuition on why certain constructions are hard (or impossible in the general case).
+
+% We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
+% $\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
+% doesn't have one, but we could show externally that it satisfies univalence.
+
+Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
+to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
+\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
+$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
+subuniverses.
+
+%%%%%%%%%%%%%%
+% \paragraph{Univalent Fibrations}
+
+% Coherence problems in type theory, coherence via Well-Foundedness.
+
+% Formalised proofs of Mac Lane's coherence theorem.
+
+% Applications of FSMG and history of the coherence theorems.
+
+% Other proofs of coherence theorems, Joyal-Street.
+
+% Pi has other extensions (fractional/negative/recursive types).
+% What are the free X monoidal structures they're describing?
+
+% Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
+% should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.
+
+% Other applications of symmetric groups.
+
+% Actions of symmetric groups, permutation groupoids.
+
+%%%%%%%%%%%%%%
+\paragraph{Reversible Programming Languages}
+
+Pi has other extensions (fractional/negative/recursive types).
+What are the free X monoidal structures they're describing?
 
@@
 %%%%%%%%%%%%%%
-\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
-types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
-$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
-    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
-    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
-$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
-
-% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
-
-By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
-syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse.
-% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
-The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
-The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which, in general, is of course
-undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
-we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
-a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact that program synthesis is
-sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$ 2-combinators). Thus, we
-this suggests a new computational interpretation of the univalence principle, which provides an intuition on why certain
-constructions are hard (or impossible in the general case).
-
-We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
-$\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
-doesn't have one, but we could show externally that it satisfies univalence.
-
-%%%%%%%%%%%%%%
-\paragraph{Univalent Fibrations} Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
-to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
-\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
-$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
-subuniverses.
-
-Coherence problems in type theory, coherence via Well-Foundedness.
-
-Formalised proofs of Mac Lane's coherence theorem.
-
-Applications of FSMG and history of the coherence theorems.
-
-Other proofs of coherence theorems, Joyal-Street.
-
-Pi has other extensions (fractional/negative/recursive types).
-What are the free X monoidal structures they're describing?
-
-Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
-should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.
-
-Other applications of symmetric groups.
-
-Actions of symmetric groups, permutation groupoids.
-
-%%%%%%%%%%%%%%
 \paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe algebraic structures, such

### 2021-07-09 Pi+/paper/discussion.tex
clean up univalence para in sec 8

@@
 $\term{ua}_\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
-
 % Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
-
 By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
@@
 
-Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
-to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
-\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
-$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
-subuniverses.
+% Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
+% to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
+% \citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
+% $\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
+% subuniverses.
 

### 2021-07-09 Pi+/paper/discussion.tex
more on sec 8

@@
 %%%%%%%%%%%%%%
+\paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe structures such as groups
+or rings. A specific group or ring is a model of the appropriate algebraic theory. Algebraic theories are usually
+\emph{presented} in terms of logical syntax, that is, as first-order theories whose signatures allow only functional
+symbols, and whose axioms are universally quantified equations. In his seminal thesis,
+\citet{lawvereFUNCTORIALSEMANTICSALGEBRAIC1963} defined a presentation-free categorical notion of universal algebraic
+structure, called a Lawvere theory. Programming Languages, such as the $\lambda$-calculus, can be viewed as algebraic
+structures with variable-binding operators, which can be formalised using second-order algebraic
+theories~\cite{fioreSecondOrderAlgebraicTheories2010}, or algebraic theories with closed
+structure~\cite{hylandClassicalLambdaCalculus2017}, called $\lambda$-theories, making the $\lambda$-calculus the
+presentation of the initial $\lambda$-theory $\Lambda$. Our family of reversible languages $\PiLang$ have been presented
+as first-order algebraic
+2-theories~\cite{cohenCoherenceRewriting2theories2009,bekeCategorificationTermRewriting2011,yanofskySyntaxCoherence2000},
+which are a categorification of algebraic theories. The types $\zerot$ and $\onet$ are nullary function symbols, the
+type formers $+$ and $\times$ are binary function symbols, the 1-combinators are invertible reduction rules, and the
+2-combinators are equations or coherence diagrams of compositions of reduction rules. Just like models of Lawvere
+theories are given by algebras of (finitary) monads on $\SetCat$, models of 2-theories are given by algebras of 2-monads
+on $\CatCat$. The particular one related to our development is the free symmetric monoidal completion 2-monad.
+
+%%%%%%%%%%%%%%
 \paragraph{Reversible Programming Languages}
@@
 
-%%%%%%%%%%%%%%
-\paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe algebraic structures, such
-as groups or rings. A specific group or ring is a model of the appropriate algebraic theory. Algebraic theories are
-usually \emph{presented} in terms of logical syntax, that is, as first-order theories whose signatures allow only
-functional symbols, and whose axioms are universally quantified equations. In his seminal
-thesis~\cite{lawvereFUNCTORIALSEMANTICSALGEBRAIC1963}, Lawvere defined a presentation-free categorical notion of
-universal algebraic structure, called a Lawvere theory.
-
-Programming Languages, such as the $\lambda$-calculus, can be viewed as algebraic structures with variable-binding
-operators, which can be formalised using second-order algebraic theories~\cite{fioreSecondOrderAlgebraicTheories2010},
-or algebraic theories with closed structure~\cite{hylandClassicalLambdaCalculus2017}, called $\lambda$-theories, making
-the $\lambda$-calculus the presentation of the initial $\lambda$-theory $\Lambda$.
-
-Our family of reversible languages have been presented as first-order algebraic
-2-theories~\cite{cohenCoherenceRewriting2theories2009,bekeCategorificationTermRewriting2011,yanofskySyntaxCoherence2000},
-which are a categorification of algebraic theories. The types $\zerot$ and $\onet$ are nullary function symbols, the
-type formers $+$ and $\times$ are binary function symbols, the 1-combinators are invertible reduction rules, and the
-2-combinators are equations or coherence diagrams of compositions of reduction rules. Just like models of Lawvere
-theories are given by algebras of (finitary) monads on $\SetCat$, models of 2-theories are given by algebras of 2-monads
-on $\CatCat$. The particular one we're interested in here is the free symmetric monoidal completion 2-monad.
-
 % %%%%%%%%%%%%%%

### 2021-07-09 Pi+/paper/discussion.tex
Fixes and fixes

--- 2021-07-09 Pi+/paper/discussion.tex
@@
 cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
-correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
-suggested by~\citet{Carette2016} and~\citet{threemodels} and established in this work.
+correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence is
+established in this work.
 
@@
 intractable.
-
-%%%%%%%%%%%%%%
-\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
-types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
-$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
-    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
-    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
-$\term{ua}_\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
-
-% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
-
-By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
-via a complete equational axiomatisation of the equivalences between types in the subuniverse.
-% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
-In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
-which is easily done by induction.  The $\term{ua}$ map corresponds to synthesising a program from an equivalence
-(which, in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible
-boolean circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the
-existence of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses
-the fact that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory
-($\PiLang$ 2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
-provides an intuition on why certain constructions are hard (or impossible in the general case).
-
-% We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
-% $\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
-% doesn't have one, but we could show externally that it satisfies univalence.
-
-Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
-to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
-\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
-$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
-subuniverses.
-
-%%%%%%%%%%%%%%
-% \paragraph{Univalent Fibrations}
-
-% Coherence problems in type theory, coherence via Well-Foundedness.
-
-% Formalised proofs of Mac Lane's coherence theorem.
-
-% Applications of FSMG and history of the coherence theorems.
-
-% Other proofs of coherence theorems, Joyal-Street.
-
-% Pi has other extensions (fractional/negative/recursive types).
-% What are the free X monoidal structures they're describing?
-
-% Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
-% should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.
-
-% Other applications of symmetric groups.
-
-% Actions of symmetric groups, permutation groupoids.
-
-%%%%%%%%%%%%%%
-\paragraph{Reversible Programming Languages}
-
-Pi has other extensions (fractional/negative/recursive types).
-What are the free X monoidal structures they're describing?
 
@@
 %%%%%%%%%%%%%%
+\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
+types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
+$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
+    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
+    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
+$\term{ua}-\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
+
+% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
+
+By giving a computable presentation for a univalent subuniverse, we are able to describe the path space of it
+syntactically, by giving a complete equational axiomatisation of the equivalences between types in the subuniverse.
+% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
+The $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator), which is easily done by induction.
+The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which, in general, is of course
+undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean circuits, it is decidable, as
+we have shown, but still far from trivial, which matches the need to assert the existence of $\term{ua}$ without giving
+a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact that program synthesis is
+sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$ 2-combinators). Thus, we
+this suggests a new computational interpretation of the univalence principle, which provides an intuition on why certain
+constructions are hard (or impossible in the general case).
+
+We could present a dependent type theory for the topos $\SetFin$, with an identity type for terms (generated by
+$\refl$), and one for types (generated by $\PiLang$ combinators). We can't talk about universes in $\SetFin$ since it
+doesn't have one, but we could show externally that it satisfies univalence.
+
+%%%%%%%%%%%%%%
+
+Coherence problems in type theory, coherence via Well-Foundedness.
+
+Formalised proofs of Mac Lane's coherence theorem.
+
+Applications of FSMG and history of the coherence theorems.
+
+Other proofs of coherence theorems, Joyal-Street.
+
+Pi has other extensions (fractional/negative/recursive types).
+What are the free X monoidal structures they're describing?
+
+Using our presentation of $S_{n}$, we can construct the Eilenberg-Maclane space (using a HIT) $K(S_{n},1)$. Then, it
+should be true that $\UFin \eqv \sqcup_{n:\Nat} K(S_{n},1)$. This is future work.
+
+Other applications of symmetric groups.
+
+Actions of symmetric groups, permutation groupoids.
+
+%%%%%%%%%%%%%%
 \paragraph{Algebraic Theories} In universal algebra, algebraic theories are used to describe algebraic structures, such

### 2021-07-09 Pi+/paper/finite.tex
Fix crefs
Fixes and fixes

@@
 We will show that it is possible to encode all $\PiLang$ combinators using adjacent transpositions
-(in~\Cref{sec:equivalence}). Group presentations are built by adding equations to a \emph{free group}, which we define
+(in~\cref{sec:equivalence}). Group presentations are built by adding equations to a \emph{free group}, which we define
 first.
@@
 
-  \begin{definition}~\label{def:group}
+  \begin{definition}
+  \label{def:group}
     In type theory, a group $G$ can be defined as a set $S$ with the following pieces of data:
@@
 
-\begin{propositionrep}~\label{prop:fin-finexcept}
+\begin{propositionrep}
+  \label{prop:fin-finexcept}
   \leavevmode
@@
 
-\begin{proposition}~\label{prop:lehmer-aut-equiv}
+\begin{proposition}
+  \label{prop:lehmer-aut-equiv}
   For all $n:\Nat$, \( \Lehmer[n] \eqv \Aut[\Fin[\suc[n]]] \).

### 2021-07-09 Pi+/paper/ufin.tex
Fixes and fixes

@@
 gives the (higher) morphisms. The groupoid we are interested in has types as points, type equivalences for 1-cells, and
-higher homotopies for higher cells. (See App.~\ref{app:grpdexample} for an example on a 3-element set.) This is the
+higher homotopies for higher cells. (See~\cref{app:grpdexample} for an example on a 3-element set.) This is the
 groupoid structure for the universe type $\UU$, since the identity type on types can be characterised as type
@@
 %%%
-\subsection{The Type Theory}~\label{subsec:type-theory}
+\subsection{The Type Theory}
+\label{subsec:type-theory}
 % \subsection{Univalent Subuniverses}
@@
 Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
-as well.  In App.~\ref{app:identitytypes}, App.~\ref{app:homotopytypes}, and App~\ref{app:hits}, we review the basics of
-identity types, homotopy types, and HITs and refer the reader to the book for more details.
+as well.  In~\cref{app:identitytypes},~\cref{app:homotopytypes}, and~\cref{app:hits}, we review the basics of identity
+types, homotopy types, and HITs and refer the reader to the book for more details.
 
@@
 
-  Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action on
-  the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to the
-  universe, such as $A \to \UU$. For a type family $P : A \to \UU$ and a point $x : A$, the type $P(x)$ is called the
-  fiber over $x$.  The $\term{transport}$ operation, named $\term{tr}$, lifts paths in the indexing type to functions
-  between fibers.
+  Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action
+  on the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to
+  the universe, such as $A \to \UU$. For a type family $P : A \to \UU$ and a point $x : A$, the type $P(x)$ is called
+  the fiber over $x$.  The $\term{transport}$ operation, named $\term{transport}/term{tr}$, lifts paths in the indexing
+  type to functions between fibers.
   %which is an $A$-indexed family of groupoids.
@@
     \qquad
-    \term{tr}_{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
+    \term{transport}_{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
   \end{gather*}
@@
         \node[circle,fill,inner sep=1pt,label=left:{$\pair{x,u}$}] (b1) at (-.5,7.2) {};
-        \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\transport{P}{p}{u}}$}] (b2) at (.5,7.2) {};
+        \node[circle,fill,inner sep=1pt,label=right:{$\pair{y,\transport*{P}{p}{u}}$}] (b2) at (.5,7.2) {};
         \draw[decorate,decoration={snake,amplitude=1}] (b1) -- node[auto] {$\term{lift}(u,p)$} (b2);
@@
 families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to \UU$
-which has a lifting operation giving it the structure of a fibration. The
-$\term{transport}$ operation $\transport{P}{p}{\blank}$ lifts paths in the base space to equivalences between the fibers:
-% Using the groupoid structure of
-% $A$, we can show that $\transport{P}{p}{\blank}$ and $\transport{P}{\inv{p}}{\blank}$ form an equivalence.
-% We name
-% this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
+which has a lifting operation giving it the structure of a fibration. The $\term{transport}$ operation lifts paths in
+the base space to functions between the fibers.  Using the groupoid structure of $A$, for a path $p : x \id_{A} y$,
+$\term{transport}(p)$ and $\term{transport}(\inv{p})$ form an equivalence.
+% We name this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
 \[
-  \tptEqv{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \eqv P(y)}
+  \tptEqv{P} : \dfun*{x,y:A}{(x \id_{A} y) \to (P(x) \eqv P(y))}
 \]
@@
 
-The use of the word \emph{univalent} here is a reference to Voevodsky's \emph{univalence} principle. Indeed, univalence
-characterises paths in the universe as equivalences between types, which follows from the canonical fibration
-$\idfunc : \UU \to \UU$ being univalent.
+Univalent fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018}, to build a model of Voevodsky's
+\emph{univalence} principle in simplicial sets. Indeed, univalence characterises paths in the universe as equivalences
+between types, which follows from the canonical fibration $\idfunc : \UU \to \UU$ being univalent.
 
@@
   A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
-  proposition for every $X$. Given such a predicate, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent and generates a
-  univalent subuniverse $\UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)$.
+  proposition for every $X$. Given such a predicate, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent
+  and generates a univalent subuniverse $\UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)$.~\footnote{Univalent
+    typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent subuniverses.}
 \end{propositionrep}
@@
 subtle~\cite{spiwackConstructivelyFinite2010}. We use the notion of Bishop-finiteness: a type is finite if it is merely
-equivalent to a finite set (\cref{def:finite-set} and \cref{def:isfin}).
+equivalent to a finite set (\cref*{def:finite-set,def:isfin}).
 
@@
 specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$.
+\footnote{Characterisations of univalent fibrations using the $\BAut$ construction have been studied
+  by~\citet{christensenCharacterizationUnivalentFibrations2015}.}
 
@@
 
-\subsection{Rig structure}~\label{subsec:rig}
+\subsection{Rig structure}
+\label{subsec:rig}
 
@@
 \begin{proposition}
-  For any $n, m : \Nat$,
+  For any $n, m : \Nat$, and for any types $X, Y, Z$,
+
   \begin{gather*}
@@
       \Fin[0]                & \eqv \bot                  \\
-      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]
-    \end{aligned}
-    \qquad
-    \begin{aligned}
-      \Fin[1]                & \eqv \top        \\
-      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]
-    \end{aligned}
-  \end{gather*}
-  and for any types $X, Y, Z$,
-  \begin{gather*}
-    \begin{aligned}
+      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]           \\
+      \\
       \bot \sqcup X          & \eqv X                     \\
@@
     \begin{aligned}
+      \Fin[1]                & \eqv \top                             \\
+      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]                      \\
+      \\
       \top \times X          & \eqv X                                \\

### 2021-07-09 Pi+/paper/discussion.tex

--- 2021-07-09 Pi+/paper/discussion.tex
@@
 %%%%%%%%%%%%%%
+\paragraph{Free Symmetric Monoidal Category} The free symmetric monoidal category has been used to study
+concurrency~\cite{hylandSymmetricMonoidalSketches2004}, Petri nets~\cite{baezCategoriesNets2021}, combinatorial
+structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005}, and
+bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}. The forgetful functor
+from $\SymMonCat$, the 2-category of (small) symmetric monoidal categories, strong symmetric monoidal functors, and
+monoidal natural transformations, to the 2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal
+category $\FSM[\CCat{C}]$ on a category $\CCat{C}$. This is a 2-monad on
+$\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict) symmetric monoidal categories. Its
+construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.  Concretely, the objects of
+$\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n] \to \CCat{C}_{0})$;
+morphisms between $(n,A)$ and $(n,B)$ are pairs $(\pi,\lambda)$ where $\pi$ is a permutation of $[n]$, and
+$\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the Grothendieck construction
+$\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and bijections to $\CatCat$, assigning
+each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on $[n]$ inducing an endofunctor on
+$C^{n}$ by action. The groupoid $\BFin$ is the free symmetric monoidal category (groupoid) on one generator, $\FSM[\unit]$.
+
+Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories
+were studied by \citet{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a
+groupoid and the proof of its universal property has been considered
+by~\citet{piceghelloCoherenceMonoidalGroupoids2020}. Free commutative monoids in type theory have been studied by
+\citet{gylterudMultisetsTypeTheory2020}, and using HoTT by \citet{choudhuryFinitemultisetConstructionHoTT2019}. The free
+symmetric monoidal groupoid $\FSM[A]$ over a groupoid $A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be
+presented as an algebraic 2-theory using 1-HITs. These HITs and the proof of their universal property have been
+considered by \citet{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}. The proof of
+the universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal
+categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The
+existence of the proof is folklore. We have produced a proof and formalised it in constructive type theory.
+
+%%%%%%%%%%%%%%
 \paragraph{Reversible Programming Languages}
@@
 
-%%%%%%%%%%%%%%
-\paragraph{Free Symmetric Monoidal Category} The forgetful functor from $\SymMonCat$, the 2-category of (small)
-symmetric monoidal categories, strong symmetric monoidal functors, and monoidal natural transformations, to the
-2-category $\CatCat$, has a left adjoint giving the free symmetric monoidal category $\FSM[\CCat{C}]$ on a category
-$\CCat{C}$. This is a 2-monad on $\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict)
-symmetric monoidal categories. Its construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.
-Concretely, the objects of $\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n]
-    \to \CCat{C}_{0})$. An morphism between $(n,A)$ and $(n,B)$ is given by a pair $(\pi,\lambda)$ where $\pi$ is a
-permutation of $[n]$, and $\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the
-Grothendieck construction $\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and
-bijections to $\CatCat$, assigning each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on
-$[n]$ inducing an endofunctor on $C^{n}$ by action. $\BFin$ is the free symmetric monoidal category (groupoid) on one
-generator, $\FSM[\unit]$. The free symmetric monoidal category has been used to study
-concurrency~\cite{hylandSymmetricMonoidalSketches2004}, petri nets~\cite{baezCategoriesNets2021}, combinatorial
-structures~\cite{fioreCartesianClosedBicategory2008}, quantum mechanics~\cite{abramskyAbstractScalarsLoops2005},
-bicategorical models of (differential) linear logic~\cite{melliesTemplateGamesDifferential2019}.
-
-Coherence and normalisation problems for monoids in constructive type theory using coherence for monoidal categories was
-studied in~\cite{beylinExtractingProofCoherence1996}. In HoTT, coherence for the free monoidal groupoid over a groupoid
-and the proof of its universal property has been considered in~\cite{piceghelloCoherenceMonoidalGroupoids2020}.
-
-Free commutative monoids in type theory have been studied in~\cite{gylterudMultisetsTypeTheory2020}, and using HoTT
-in~\cite{choudhuryFinitemultisetConstructionHoTT2019}. The free symmetric monoidal groupoid $\FSM[A]$ over a groupoid
-$A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be presented as an algebraic 2-theory using 1-HITs. These HITs
-and the proof of their universal property have been considered
-in~\cite*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}.
-
-The proof of the universal property of the $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric
-monoidal categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite
-set. The existence of the proof is folklore, and we have produced a new proof of it while working in constructive type
-theory.
-
 % %%%%%%%%%%%%%%

### 2021-07-09 Pi+/paper/ufin.tex
Even more fixes

@@
 structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
-commutative rig of natural numbers $\Nat$.
+commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA}.
 
@@
 
-  Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes an $\hSet$ $A$ and a relation $R : A \to A \to
-    \UU$. It has an inclusion of points $q : A \to \quot{A}{R}$, and adds paths between related pairs of elements
-  $\quotrel : R(x,y) \to q(x) \id_{\quot{A}{R}} q(y)$ (see~\cref{def:set-quot}).  We recall that the quotient is
-  \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff $(q(x) \id_{\quot{A}{R}}
-    q(y))$.
+  Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes an set $A$ and a relation $R : A \to A \to \UU$.
+  It has an inclusion of points $\quotinc : A \to \quot{A}{R}$, and adds paths between related pairs of elements
+  $\quotrel : R(x,y) \to \quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$ (see~\cref{def:set-quot}).  We recall that the
+  quotient is \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff $(q(x)
+  \id_{\quot{A}{R}} q(y))$.
 
@@
     \label{def:set-quot}
-    Given a type $A$ which is an $\hSet$, and a relation $R : A \to A \to \hProp$, the set-quotient $\quot{A}{R}$ is the
-    higher inductive type generated by
+    Given a type $A$ which is a set, and a relation $R : A \to A \to \UU$, the set-quotient $\quot{A}{R}$ is the higher
+    inductive type generated by
     \begin{itemize}
-      \item an inclusion function $q : A \to \quot{A}{R}$,
-      \item for each $x, y : A$ such that $R(x,y)$, a path $q(x) \id_{\quot{A}{R}} q(y)$,
+      \item an inclusion function $\quotinc : A \to \quot{A}{R}$,
+      \item for each $x, y : A$ such that $R(x,y)$, a path $\quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$,
       \item a set truncation, for each $x, y : \quot{A}{R}$ and $r, s : x \id_{\quot{A}{R}} y$, we have $r \id s$,

### 2021-07-09 Pi+/paper/discussion.tex

@@
 %%%%%%%%%%%%%%
-\paragraph{Reversible Programming Languages}
-
-Pi has other extensions (fractional/negative/recursive types).
-What are the free X monoidal structures they're describing?
+\paragraph{Reversible Programming Languages} The practice of programming languages is replete with \emph{ad hoc}
+instances of reversible computations: database transactions, mechanisms for data provenance, checkpoints, stack and
+exception traces, logs, backups, rollback recoveries, version control systems, reverse engineering, software
+transactional memories, continuations, backtracking search, and multiple-level undo features in commercial
+applications. In the early nineties, \citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
+treatment of reversibility. But intensive research in full-fledged reversible models of computations and reversible
+programming languages was only sparked by the discovery of deep connections between physics and
+computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275,Hey:1999:FCE:304763,fredkin1982conservative},
+and by the potential for efficient quantum computation~\cite{springerlink:10.1007/BF02650179}. The early developments of
+reversible programming languages started with a conventional programming language, e.g., an extended $\lambda$-calculus,
+and either (i) extended the language with a history
+mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or (ii) imposed constraints on the control
+flow constructs to make them reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.  More modern approaches recognize that
+reversible programming languages require a fresh approach and should be designed from first principles without the
+detour via conventional irreversible
+languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}. The version of
+$\PiLang$ studied in this paper is restricted to finite types and terminating total computations. It would be
+interesting to understand which versions of free monoidal structures correspond to extensions of $\Pi$ with recursive
+types~\cite{James:2012:IE:2103656.2103667,rc2011} and negative/fractional
+types~\cite{chenComputationalInterpretationCompact2021}.
 

### 2021-07-09 Pi+/paper/equivalence.tex

@@
-\section{Going from \texorpdfstring{$\PiLang$}{Pi} to \texorpdfstring{$\UFin$}{UFin} and back}
+\section{Correspondence between \texorpdfstring{$\PiLang$}{Pi} and \texorpdfstring{$\UFin$}{UFin}}
 \label{sec:equivalence}
@@
 
-In this section, we consider the additive fragment $\PiPlusLang$ of $\PiLang$. This is the language that we interpret to
-$\UFin$ and back, using the tools developed in the previous sections. Further, we go through an intermediate step of the
-language $\PiHatLang$, which is a simplified variant of $\PiPlusLang$ that uses adjacent transpositions for combinators,
-while preserving all the required structure.
+In this section, we first translate $\PiLang$ to its additive fragment $\PiPlusLang$. This is the language that we
+interpret to $\UFin$ and back, using the tools developed in the previous sections. Further, we go through an
+intermediate step of the language $\PiHatLang$, which is a simplified variant of $\PiPlusLang$ that uses adjacent
+transpositions for combinators, while preserving all the required structure.
 
 We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\Cref*{fig:piplus,fig:pihat} respectively,
-eliding the 2-combinators for brevity. \note{We enforce that there is a unique 2-combinator between compatible
-1-combinators, by introducing a level 3, which is trivial in the same way as in the case of $\PiLang$, as described
-before.} They can be found in the appendix and in the accompanying Agda code.
+eliding the 2-combinators for brevity. We enforce that there is a unique 2-combinator between compatible 1-combinators,
+by relating them with a truncation. These can be found in the appendix and in the accompanying Agda code
+(~\cref{app:level-two}).
 
@@
 To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
-category. Since the languages have higher-dimensional structure, the syntactic categories here are actually
-(locally-strict) weak 2-categories. We define them formally in the appendix, and only state our results here.
-
-For each of the $\PiLang$, $\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$,
-$\PiPlusCat$ and $\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can
-show that these are $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are also locally-strict,
-or locally-posetal, because there is at most one 2-cell between compatible 1-cells.
+category. We define them formally in the appendix, and only state our results here. For each of the $\PiLang$,
+$\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and
+$\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can shoe that these
+syntactic categories here are actually $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are
+also locally-strict, or locally-posetal, because there is at most one 2-cell between compatible 1-cells.
 
@@
   There are two symmetric monoidal structures on $\PiPlusCat$, given by $(\zerot, +)$ and $(\onet, \times)$, with
-  $\times$ distributing over $+$, giving it a rig structure, as described in \todo{cite earlier section}.
+  $\times$ distributing over $+$, giving it a rig structure.
 \end{proposition}
-\todo{either prove it or cite something}
 
@@
 \begin{proposition}
-  $\evalt$ gives a rig biequivalence between $\PiCat$ and $\PiPlusCat$.
+  $\evalt$ gives a rig equivalence between $\PiCat$ and $\PiPlusCat$.
 \end{proposition}
@@
 \begin{proposition}
-  $\evalp/\quotep$ give a symmetric monoidal biequivalence between $\PiPlusCat$ and $\PiHatCat$.
+  $\evalp/\quotep$ give a symmetric monoidal equivalence between $\PiPlusCat$ and $\PiHatCat$.
 \end{proposition}
@@
 \begin{proposition}
-  $\evalh/\quoteh$ give a symmetric monoidal biequivalence between $\PiHatCat$ and $\UFin$.
+  $\evalh/\quoteh$ give a symmetric monoidal equivalence between $\PiHatCat$ and $\UFin$.
 \end{proposition}

### 2021-07-09 Pi+/paper/discussion.tex

@@
 %%%%%%%%%%%%%%
-\paragraph{Extensional vs. Intensional View of Permutations.}
-Although the meaning of a permutation - that is, the specific bijective function
-that it represents - is in some sense all there is, manipulating them
-syntactically still has its advantages. By writing reversible programs, we think
-of them in an intensional way. Comparing two programs for equality by evaluating
-them on all points in the domain is a very crude - and maybe even inefficient -
-way of doing that. ~\cite{Kuehlmann:2006:RBR:2298470.2300327,10.1007/978-3-540-24605-3_4,Yamashita:2010:FEQ:1835957.1835965}.
+\paragraph{Permutations} One might argue that finding formal systems for expressing various flavors of computable
+functions has been a major focus of logic and computer science since its inception. Permutations, being at the core of
+reversible computing, are an interesting class of functions, the bijective ones, for which there is no sound and
+complete formal system. Our work develops such a formal system bringing in all the associated benefits of
+syntactic calculi, most notably, their calculational flavor. Thus instead of comparing two reversible programs by
+extensional equality of the underlying bijective functions, a calculus offers more nuanced techniques that can enforce
+additional intensional constraints on the desired equality relation.
 
-It is crude, because there is no way of enforcing additional constraints on the
-process of transforming one program into another, and no way of inspecting what
-transformation occurred. We can imagine a practical situation in which reversible
-circuits admit one kind of optimization FPGA?, but do not admit another,
-even though they are equivalent in the extensional sense - or maybe one of these
-transformations is cheaper than another, or maybe it is important to know which
-transformation did occur for the producer(?) to focus their resources of
-improving this kind of transformations. By comparing the circuits extensionally,
-the proof object = the path from one into the other - does not have any
-structure,  it is just a check that the values match.
+% additional constraints on the
+% process of transforming one program into another, and no way of inspecting what
+% transformation occurred. We can imagine a practical situation in which reversible
+% circuits admit one kind of optimization FPGA?, but do not admit another,
+% even though they are equivalent in the extensional sense - or maybe one of these
+% transformations is cheaper than another, or maybe it is important to know which
+% transformation did occur for the producer(?) to focus their resources of
+% improving this kind of transformations. By comparing the circuits extensionally,
+% the proof object = the path from one into the other - does not have any
+% structure,  it is just a check that the values match.
 
-On the other hand, even though in general, the equality of circuits requires
-exponential time jk{triple check}, the proof object can be still very small.
-This creates structure that can possibly be exploited in a heuristic way.
-Additionally, if the goal is to convince third-party that two circuits are
-the same, we can just present the proof instead of showing the equality directly
-- for example, by choosing a particularly short one.
+% On the other hand, even though in general, the equality of circuits requires
+% exponential time jk{triple check}, the proof object can be still very small.
+% This creates structure that can possibly be exploited in a heuristic way.
+% Additionally, if the goal is to convince third-party that two circuits are
+% the same, we can just present the proof instead of showing the equality directly
+% - for example, by choosing a particularly short one.
 
-Taking all this into account, it is clear that we must seek a syntactic way of
-computing the meaning of the circuit, instead of just directly evaluating it to
-$\Aut[\Fin[n]]$ - since when we go back and quote the permutation as a circuit
-(in a normal form), there was no "trace" left to see what transformation (what
-sequence of 2-paths) maps the old one into the new one.
+% Taking all this into account, it is clear that we must seek a syntactic way of
+% computing the meaning of the circuit, instead of just directly evaluating it to
+% $\Aut[\Fin[n]]$ - since when we go back and quote the permutation as a circuit
+% (in a normal form), there was no "trace" left to see what transformation (what
+% sequence of 2-paths) maps the old one into the new one.
 
-We are additionally experimenting with user annotations that can guide
-the search. Each level-2 combinator can be annotated with various
-``cost'' annotations indicating whether it reduces the number of
-gates, reduces the number of choice points, or other cost
-functions. Then one can ask for a proof that takes no more than a
-certain number of steps or a proof that does not create more than a
-certain number of additional wires etc. We illustrate these ideas by
-defining a simple cost function and using it to annotate level-2
-combinators.
+% We are additionally experimenting with user annotations that can guide
+% the search. Each level-2 combinator can be annotated with various
+% ``cost'' annotations indicating whether it reduces the number of
+% gates, reduces the number of choice points, or other cost
+% functions. Then one can ask for a proof that takes no more than a
+% certain number of steps or a proof that does not create more than a
+% certain number of additional wires etc. We illustrate these ideas by
+% defining a simple cost function and using it to annotate level-2
+% combinators.
 
-We define the \emph{length} $L(c)$ of a composite circuit $c$ as
-follows: the length of a sequential composition of circuits is the sum
-of the lengths of the subcircuits $L(f \odot g) = L(f) + L(g)$; and
-the length of choice or parallel composition is the maximum of either
-branch $L(f \oplus g) = L(f \otimes g) = \max(L(f),L(g))$. For
-primitive gates, the length needs to be postulated to reflect the
-``length'' of the computation involved in applying that primitive. As
-examples, consider the following two level-2 combinators:
+% We define the \emph{length} $L(c)$ of a composite circuit $c$ as
+% follows: the length of a sequential composition of circuits is the sum
+% of the lengths of the subcircuits $L(f \odot g) = L(f) + L(g)$; and
+% the length of choice or parallel composition is the maximum of either
+% branch $L(f \oplus g) = L(f \otimes g) = \max(L(f),L(g))$. For
+% primitive gates, the length needs to be postulated to reflect the
+% ``length'' of the computation involved in applying that primitive. As
+% examples, consider the following two level-2 combinators:
 
-% \begin{code}
-% linv◎l'  :  {t₁ t₂ : U} {c : t₁ ⟷ t₂} → (c ◎ ! c) ⇔ id⟷
-% idl◎l'   :  {t₁ t₂ : U} {c : t₁ ⟷ t₂} → (id⟷ ◎ c) ⇔ c
-% \end{code}
-% \AgdaHide{
-% \begin{code}
-% linv◎l'  = ?
-% idl◎l'   = ?
-% \end{code}
-% }
+% % \begin{code}
+% % linv◎l'  :  {t₁ t₂ : U} {c : t₁ ⟷ t₂} → (c ◎ ! c) ⇔ id⟷
+% % idl◎l'   :  {t₁ t₂ : U} {c : t₁ ⟷ t₂} → (id⟷ ◎ c) ⇔ c
+% % \end{code}
+% % \AgdaHide{
+% % \begin{code}
+% % linv◎l'  = ?
+% % idl◎l'   = ?
+% % \end{code}
+% % }
 
-\noindent Assuming that $\idc$ takes a unit length of computation, the
-first can be annotated with $L(c)*2 \isotwo 1$ and the second with $L(c)+1 \isotwo
-    L(c)$ indicating that the first combinator reduces the length of the
-circuit from twice the length of $c$ to 1 and the second combinator
-reduces the length of the circuit by 1. Such annotations can then be
-used to constrain or guide the search for transformations between
-circuits.
+% \noindent Assuming that $\idc$ takes a unit length of computation, the
+% first can be annotated with $L(c)*2 \isotwo 1$ and the second with $L(c)+1 \isotwo
+%     L(c)$ indicating that the first combinator reduces the length of the
+% circuit from twice the length of $c$ to 1 and the second combinator
+% reduces the length of the circuit by 1. Such annotations can then be
+% used to constrain or guide the search for transformations between
+% circuits.
 
-\todo{Why is this better than evaluating the permutation and checking each bit?}
-\todo{Why care about an equational theory? Compilers, optimisation.}
+% \todo{Why is this better than evaluating the permutation and checking each bit?}
+% \todo{Why care about an equational theory? Compilers, optimisation.}
 

### 2021-07-09 Pi+/paper/discussion.tex
table

@@
 
+\begin{center}
+    \begin{tabular}{c c c}
+        \hline
+        $\PiLang$     & $\Sn$               & $\UFin$     \\
+        \hline
+        types         & Natural numbers     & Finite sets \\
+        \hline
+        1-combinators & Generators of $\Sn$ & 1-paths     \\
+        \hline
+        2-combinators & Relations of $\Sn$  & 2-paths     \\
+        \hline
+    \end{tabular}
+\end{center}
+
+\begin{center}
+    \begin{tabular}{c c c}
+        \hline
+        Logic                              & Program                       & Algebra                     \\
+        \hline
+        Propositions                       & Types                         & Objects                     \\
+        \hline
+        Proofs                             & Terms                         & Morphisms                   \\
+        \hline
+        $\blank$                           & Equations                     & Homotopies                  \\
+        \hline
+        Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
+        \hline
+        Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
+        \hline
+        Connectives                        & Type constructors             & Bifunctors                  \\
+        \hline
+        ?                                  & Type normalisation            & ?                           \\
+        \hline
+        ?                                  & Program normalisation         & ?                           \\
+        \hline
+        ?                                  & Soundness of 2-combinators    & ?                           \\
+    \end{tabular}
+\end{center}
+
+\begin{center}
+    \begin{tabular}{c c c}
+        \hline
+        Reversible Logic             & Reversible Programming                  & Algebra                  \\
+        \hline
+        Free commutative rig         & Free symmetric rig groupoid             & $\UFin$                  \\
+        \hline
+        Propositions                 & Types                                   & Objects                  \\
+        \hline
+        Entailment?                  & Reversible programs                     & Isomorphisms             \\
+        \hline
+        $\blank$                     & 2-combinators                           & Homotopies               \\
+        \hline
+        Conjunction?                 & $\times$                                & $\otimes$                \\
+        \hline
+        Disjunction?                 & $+$                                     & $\oplus$                 \\
+        \hline
+        $A \land B \vdash B \land A$ & $\swapt$                                & $\mathcal{B}_{A,B}$      \\
+        \hline
+        True proposition             & $\onet$                                 & $F_1$                    \\
+        \hline
+        Decidability                 & Program Synthesis                       & Reachability in a graph? \\
+        \hline
+        Soundness/Completeness       & Soundness/Completeness of 1-combinators & ?                        \\
+    \end{tabular}
+\end{center}
+
 In the remainder of this section, we discuss some broader related work.
-
-% \begin{center}
-%     \begin{tabular}{c c c}
-%         \hline
-%         $\PiLang$     & $\Sn$               & $\UFin$     \\
-%         \hline
-%         types         & Natural numbers     & Finite sets \\
-%         \hline
-%         1-combinators & Generators of $\Sn$ & 1-paths     \\
-%         \hline
-%         2-combinators & Relations of $\Sn$  & 2-paths     \\
-%         \hline
-%     \end{tabular}
-% \end{center}
-
-% \begin{center}
-%     \begin{tabular}{c c c}
-%         \hline
-%         Logic                              & Program                       & Algebra                     \\
-%         \hline
-%         Propositions                       & Types                         & Objects                     \\
-%         \hline
-%         Proofs                             & Terms                         & Morphisms                   \\
-%         \hline
-%         $\blank$                           & Equations                     & Homotopies                  \\
-%         \hline
-%         Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
-%         \hline
-%         Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
-%         \hline
-%         Connectives                        & Type constructors             & Bifunctors                  \\
-%         \hline
-%         ?                                  & Type normalisation            & ?                           \\
-%         \hline
-%         ?                                  & Program normalisation         & ?                           \\
-%         \hline
-%         ?                                  & Soundness of 2-combinators    & ?                           \\
-%     \end{tabular}
-% \end{center}
-
-% \begin{center}
-%     \begin{tabular}{c c c}
-%         \hline
-%         Reversible Logic             & Reversible Programming                  & Algebra                  \\
-%         \hline
-%         Free commutative rig         & Free symmetric rig groupoid             & $\UFin$                  \\
-%         \hline
-%         Propositions                 & Types                                   & Objects                  \\
-%         \hline
-%         Entailment?                  & Reversible programs                     & Isomorphisms             \\
-%         \hline
-%         $\blank$                     & 2-combinators                           & Homotopies               \\
-%         \hline
-%         Conjunction?                 & $\times$                                & $\otimes$                \\
-%         \hline
-%         Disjunction?                 & $+$                                     & $\oplus$                 \\
-%         \hline
-%         $A \land B \vdash B \land A$ & $\swapt$                                & $\mathcal{B}_{A,B}$      \\
-%         \hline
-%         True proposition             & $\onet$                                 & $F_1$                    \\
-%         \hline
-%         Decidability                 & Program Synthesis                       & Reachability in a graph? \\
-%         \hline
-%         Soundness/Completeness       & Soundness/Completeness of 1-combinators & ?                        \\
-%     \end{tabular}
-% \end{center}
 

### 2021-07-09 Pi+/paper/finite.tex
Even more fixes

@@
 between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
-between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$.~\footnote{Note that the permutations are off-by-one, because
-  we ignored the 0-position in the Lehmer code.}
+between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$.~\footnote{Note that the indices for the type of permutations
+are off-by-one, because we chose $\Fin[n-1]$ to represent generators for permutations on $\Fin[n]$.}
 
@@
 First, we define a type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except
-the one in the argument. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence an
-$\hSet$.
+the one in the argument. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence a
+set.
 

### 2021-07-09 Pi+/paper/finite.tex
Citations and fixes

@@
 between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
-between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$.~\footnote{Note that the indices for the type of permutations
-are off-by-one, because we chose $\Fin[n-1]$ to represent generators for permutations on $\Fin[n]$.}
+between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$. The idea of this proof is borrowed from~\cite{Molzer-cubical}
+~\footnote{Note that the indices for the type of permutations are off-by-one, because we chose $\Fin[n-1]$ to represent
+  generators for permutations on $\Fin[n]$.}
 

### 2021-07-09 Pi+/paper/reversible1.tex
Citations and fixes

@@
 
-These commutative semirig identities can be used to design a logic for reversible
+These commutative rig identities can be used to design a logic for reversible
 programming~\cite*{sparksSuperstructuralReversibleLogic2014}. But as natural number identities have no computational
@@
 semantics, rig groupoid \emph{coherence conditions} that identify different syntactic representations of the same
-permutation~\cite{laplaza72,Carette2016} were collected  in a
+permutation~\cite{laplaza72,caretteComputingSemiringsWeak2016} were collected  in a
 second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate

### 2021-07-09 Pi+/paper/ufin.tex
Citations and fixes

@@
 structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
-commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA}.
+commutative rig of natural numbers $\Nat$~\cite{baezHIGHERDIMENSIONALALGEBRA2010}.
 
@@
 \begin{proof}
-  (of Prop.~\ref{prop.isFin})
   Suppose we have $(n,\phi) : \isFin[X]$ and $(m,\psi) : \isFin[X]$, we need to show that $(n,\phi) \id (m,\psi)$. It is
@@
 
-While $\UFin$ has \emph{all} the finite types, we are also interested in constructing a subuniverse of finite types of a
-specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$.
-\footnote{Characterisations of univalent fibrations using the $\BAut$ construction have been studied
+This definition of the groupoid of finite types has also been considered
+in~\cite{yorgeyCombinatorialSpeciesLabelled2014}. While $\UFin$ has \emph{all} the finite types, we are also interested
+in constructing a subuniverse of finite types of a specified cardinality. To do so, we will start with the subuniverse
+$\BAut[T]$. \footnote{Characterisations of univalent fibrations using the $\BAut$ construction have been studied
   by~\citet{christensenCharacterizationUnivalentFibrations2015}.}

### 2021-07-09 Pi+/paper/reversible1.tex
intro and level 2

--- 2021-07-08 Pi+/paper/reversible1.tex
@@
 
-To see how this language expresses reversible circuits, we present a few examples. First it is possible to directly
-mimic the \verb|qasm|-perspective by defining types that describe sequences of booleans. We use the type
-$\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true} and
-$\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is straightforward to define using
-the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values, thus the
-type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$) corresponds to a collection of
-wires that can transmit three bits.
+To see how this language expresses reversible circuits, we present a few examples in the Agda embedding of the
+language. The Agda syntax is almost identical to the syntax in Fig.~\ref{fig:pi-terms} except that sequential
+composition is written using $\circledcirc$. First it is possible to directly mimic the \verb|qasm|-perspective by
+defining types that describe sequences of booleans. We use the type $\mathbb{2} = \onet + \onet$ to represent booleans
+with $\inlv{\Acon{tt}}$ representing \textsf{true} and $\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean
+negation (the \verb|x|-gate) is straightforward to define using the primitive combinator $\swapp$. We can represent
+$n$-bit words using an n-ary product of boolean values, thus the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$
+(abbreviated $\mathbb{B}~3$) corresponds to a collection of wires that can transmit three bits.
 %
@@
 second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
-$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For example, we have:
+$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For
+example, we have the following level-2 combinators dealing with associativity:
 
@@
 
-\combtwo{}
+\leveltwoblockone{}
 
-\noindent where the first level-2 combinator states the obvious fact that compositions with the identity can be
-optimized and the second level-2 combinator states the more interesting observation that swapping the two sides of a sum
-type and then applying $c_1$ to the left and $c_2$ to the right is equivalent to applying $c_2$ to the left and $c_1$ to
-the right and then swapping. The full set of level-2 combinators is large and is only listed in the accompanying code.
+\noindent The full set of level-2 combinators is large; the remaining combinators are listed in App.~\ref{app:leveltwo}.
+
+\begin{toappendix}
+  \label{app:leveltwo}
+
+  The additional level-2 combinators:
+
+  \medskip
+  \leveltwoblocktwo{}
+\end{toappendix}
 

### 2021-07-09 Pi+/paper/finite.tex
General formatting

--- 2021-07-09 Pi+/paper/finite.tex
@@
 permutations on an $n + 1$-element set, there are $n$ adjacent transpositions, transposition number $k$ swapping
-elements $(\el{k}, \el{k+1})$. Thus, the generating set would be $\Fin[n]$. There are three relations that we're going
+elements at indices $k$ and $k+1$. Thus, the generating set would be $\Fin[n]$. There are three relations that we're going
 to specify for this presentation -- these are the equations that these generators should satisfy. We visualise them as
@@
         swapping happens, that is, we can slide the wires freely.
-  \item[\labelcref{fig:coxeter-c}] If the next transposition in the sequence is equal to the the first one, the order of
-        operations can be reversed.
+  \item[\labelcref{fig:coxeter-c}] \jk{If the next transposition in the sequence is equal to the the first one, the order of
+        operations can be reversed}.
 \end{enumerate}
@@
 
-The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\). Since we think of the
-number $m$ as the transposition swapping elements $(\el{m}, \el{m+1})$, the role of this function is to produce a
-sequence of transpositions which moves element $\el{k + n}$ left by $k$ places, shifting all the elements in between one
-place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the following generators,
-inlining the congruence closure in $\longcox$, allowing arbitrary reductions inside the list.
+The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\). The role of this function
+is to produce a sequence of transpositions which moves element at index $k + n$ left by $k$ places, shifting all the
+elements in between one place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the
+following generators, inlining the congruence closure in $\longcox$, allowing arbitrary reductions inside the list.
 
-\begin{definition}[$\longcox$]
+\begin{definition}[$\longcox : {\List[\Fin[n]]} \to {\List[\Fin[n]]} \to {\UU}$]
   \begin{align*}
@@
 \begin{toappendix}
-  \begin{definition}[$\longcoxstar$]
+  \begin{definition}[$\longcoxstar : {\List[\Fin[n]]} \to {\List[\Fin[n]]} \to {\UU}$]
     \label{def:longcox-star}
@@
 
-  \begin{definition}[$\longcoxplus$]
+  \begin{definition}[$\longcoxplus : {\List[\Fin[n]]} \to {\List[\Fin[n]]} \to {\UU}$]
     \label{def:longcox-plus}
@@
 Despite the increased complexity of the generators, the rewriting system $(\List[\Fin[n]],\longcox)$ has the properties
-we desire. It satisfies local confluence, that is, the Church-Rosser (diamond) property, and it is terminating, so, by
-Newman's lemma, it produces a unique normal form. We follow the terminology
+we desire. It satisfies local confluence, that is, the Church-Rosser (diamond) property (for example, using
+$\longbraid$, we can now show the problematic critical pair \cref{fig:critical-pairs-non-converging} to converge), and it is
+terminating, so, Newman's lemma, it produces a unique normal form. We follow the terminology
 of~\cite*{huetConfluentReductionsAbstract1980a,krausCoherenceWellFoundednessTaming2020} to state our results formally.
@@
 tabulated presentation of a permutation of $\Fin[5]$.
-
 \[

### 2021-07-09 Pi+/paper/ufin.tex
General formatting

--- 2021-07-09 Pi+/paper/ufin.tex
@@
   \begin{gather*}
-    \begin{aligned}
-      \Fin[n] & \defeq \dsum{k:\Nat}{k < n}
-    \end{aligned}
-    \qquad\qquad \qquad\qquad
+    \Fin[n] \defeq \dsum{k:\Nat}{k < n}
+    \qquad\qquad \text{or} \qquad\qquad
     \begin{aligned}
@@
   \label{def:isfin}
-  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n$.
+  We say that a type is finite if it is merely equal to $\Fin[n]$ for some $n : \Nat$.
   \[
@@
 has one point $T_0$, and 1-paths $T_0 \id_{\BAut[T]} T_0$, that is, loops, and higher paths between these loops. The
-type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$ is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
+type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$, is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
 the group of automorphisms of $T$.
@@
   For any $n, m : \Nat$, and for any types $X, Y, Z$,
-
   \begin{gather*}

### 2021-07-09 Pi+/paper/discussion.tex

@@
 presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
-semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible
-languages. Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
+semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
+Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
 cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
 correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
-suggested by~\citet{Carette2016} and~\citet{threemodels} and established in this work.
+suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work.
 
@@
 %%%%%%%%%%%%%%
-\paragraph{Rewriting} Higher-order term-rewriting systems and word problems have a long history of being formalised in
-proof assistants like {homotopy.io}, Lean, Coq, and HoTT~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the
-proof of our main result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem
-and formalised its correctness and termination in Agda. Our design based on encoding permutations as
-adjacent transpositions led to an intuitive system whose correctness was straightforward to establish and whose termination
-only required slightly modified variations on Church-Rosser theorems. Other encodings of permutations as listed vectors,
-matrices, inductively generated trees (Motzkin trees), Young diagrams, or string diagrams, proved either difficult
-formalise in type theory or difficult to relate directly to the primitive type isomorphisms of $\PiLang$. The automatic
+\paragraph{Coherence and Rewriting}
+
+Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
+{homotopy.io}, Lean, Coq, and HoTT~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
+result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem and formalised its
+correctness and termination. \citet{Hiver-coq} describes an explicit algorithm for producing normal forms. could provide
+an alternative to our rewriting system. Other encodings of permutations as listed vectors, matrices, inductively
+generated trees (Motzkin trees), Young diagrams, or string diagrams, proved either difficult formalise in type theory or
+difficult to relate directly to the primitive type isomorphisms of $\PiLang$. The automatic
 \citet{knuthSimpleWordProblems1970} completion produced too many equations making proving correctness and termination
 intractable.
+
+Coherence theorems are famous problems in (higher) category theory, and Mac Lane's coherence
+theorem~\cite{maclaneNaturalAssociativityCommutativity1963,joyalBraidedTensorCategories1993,gurskiInfiniteLoopSpaces2013}
+for monoidal categories is a particular one. The use of rewriting theory and proof assistants to establish coherence
+theorems for various higher categorical structures has a long history as well,
+see~\cite*{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.
 
@@
 types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
-$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
-    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
-    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
-$\term{ua}_\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
+$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
+$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
+$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
+formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
+$(A \eqv B) \eqv (A \id_{\UU} B)$.
 % Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
@@
 In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
-which is easily done by induction.  The $\term{ua}$ map corresponds to synthesising a program from an equivalence
-(which, in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible
-boolean circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the
-existence of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses
-the fact that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory
-($\PiLang$ 2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
+which is easily done by induction. The $\term{ua}$ map corresponds to synthesising a program from an equivalence (which,
+in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible boolean
+circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the existence
+of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses the fact
+that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
+2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
 provides an intuition on why certain constructions are hard (or impossible in the general case).
+
+There are other, different approaches to computing with univalence,
+in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
+Cubical Type
+Theory\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019},
+univalence computes.
 
@@
 
+A stronger universal property that $\BFin$ is biinitial in the 2-category of rig categories, was established
+in~\cite{elguetaGroupoidFiniteSets2021}.
+
 %%%%%%%%%%%%%%
@@
 interesting to understand which versions of free monoidal structures correspond to extensions of $\Pi$ with recursive
-types~\cite{James:2012:IE:2103656.2103667,rc2011} and negative/fractional
+types~\cite{jamesInformationEffects2012,rc2011} and negative/fractional
 types~\cite{chenComputationalInterpretationCompact2021}.

### 2021-07-09 Pi+/paper/finite.tex

--- 2021-07-09 Pi+/paper/finite.tex
@@
       \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                      & \text{induction hypothesis}                                  \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                          \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\Cref{prop:fin-finexcept}~\Cref{prop:fin-finexcept-2}} \\
-      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\Cref{prop:fin-finexcept}~\Cref{prop:fin-finexcept-2}} \\
-      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                          & \text{by~\Cref{prop:fin-finexcept}~\Cref{prop:fin-finexcept-3}}
+      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                       \\
+      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
+      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
+      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                           & \text{\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}}
     \end{array}

### 2021-07-09 Pi+/paper/finite.tex

--- 2021-07-09 Pi+/paper/finite.tex
@@
         operations can be reversed}.
+
+>>>>>>> ee83d6f6650c383d20e046a4ada0de4cb9c42dca
 \end{enumerate}

### 2021-07-09 Pi+/paper/reversible1.tex

--- 2021-07-09 Pi+/paper/reversible1.tex
@@
 \noindent where the first level-2 combinator states the obvious fact that compositions with the identity can be
-optimized and the second level-2 combinator states the more interesting observation that swapping the two sides of a sum
+optimised and the second level-2 combinator states the more interesting observation that swapping the two sides of a sum
 type and then applying $c_1$ to the left and $c_2$ to the right is equivalent to applying $c_2$ to the left and $c_1$ to

### 2021-07-09 Pi+/paper/ufin.tex

--- 2021-07-09 Pi+/paper/ufin.tex
@@
 % The way to describe such groupoid in HoTT is to construct a type which has the types themselves as points, i.e. a
-% universe. Then, equivalences between those types can be characterized using elements of the equality type. For this to
+% universe. Then, equivalences between those types can be characterised using elements of the equality type. For this to
 % work, the universe in question has be a \emph{univalent subuniverse}.

### 2021-07-09 Pi+/paper/equivalence.tex
General formatting
Update

@@
 \begin{toappendix}
-  \todo{What is U?}
   \begin{proposition}
@@
     \begin{itemize}
-      \item $\PiLang$ types for 0-cells,
+      \item $\PiLang$ types ($U$) for 0-cells,
       \item for $X, Y : U$, a collection of 1-cells $X \iso Y$,
@@
     \begin{itemize}
-      \item $\PiPlusLang$ types for 0-cells,
+      \item $\PiPlusLang$ types ($\UPlus$) for 0-cells,
       \item for $X, Y : \UPlus$, a collection of 1-cells $X \isop Y$,
@@
     \begin{itemize}
-      \item $\PiHatCat$ types for 0-cells,
+      \item $\PiHatCat$ types ($\UHat$) for 0-cells,
       \item for $X, Y : \UHat$, a collection of 1-cells $X \isoh Y$,
@@
 
-\begin{proposition}
+\begin{lemma}
   There are two symmetric monoidal structures on $\PiPlusCat$, given by $(\zerot, +)$ and $(\onet, \times)$, with
   $\times$ distributing over $+$, giving it a rig structure.
-\end{proposition}
+\end{lemma}
 
@@
 
-\begin{definition}
+\begin{definition}[$\evalt$]
   \begin{align*}
@@
 
-\begin{proposition}
+\begin{theorem}
   $\evalt$ gives a rig equivalence between $\PiCat$ and $\PiPlusCat$.
-\end{proposition}
+\end{theorem}
 
@@
 
-\begin{proposition}
+\begin{lemma}
   $\PiHatCat$ has a symmetric monoidal structure, with the unit given by 0 and the tensor product given by natural
   number addition.
-\end{proposition}
+\end{lemma}
 
@@
 
-\begin{definition}
+\begin{definition}[$\evalp$]
   \begin{align*}
     \evalp_{0} & : \UPlus \to \UHat                                                          \\
-    \evalp_{1} & : (c : t_{1} \iso t_{2}) \to \evalp_{0}(t_{1}) \isoh \evalp_{0}(t_{2})      \\
-    \evalp_{2} & : (\alpha : c_{1} \Iso c_{2}) \to \evalp_{1}(c_{1}) \Isoh \evalp_{1}(c_{2})
+    \evalp_{1} & : (c : t_{1} \isop t_{2}) \to \evalp_{0}(t_{1}) \isoh \evalp_{0}(t_{2})      \\
+    \evalp_{2} & : (\alpha : c_{1} \Isop c_{2}) \to \evalp_{1}(c_{1}) \Isoh \evalp_{1}(c_{2})
   \end{align*}
@@
 
-\begin{definition}
+\begin{definition}[$\quotep$]
   \begin{align*}
@@
 
-\begin{proposition}
+\begin{theorem}
   $\evalp/\quotep$ give a symmetric monoidal equivalence between $\PiPlusCat$ and $\PiHatCat$.
-\end{proposition}
+\end{theorem}
 
@@
 
-\begin{definition}
+\begin{definition}[$\evalh$]
   \begin{align*}
@@
 
-\begin{definition}
+\begin{definition}[$\quoteh$]
   \begin{align*}
@@
 
-\begin{proposition}
+\begin{theorem}
   $\evalh/\quoteh$ give a symmetric monoidal equivalence between $\PiHatCat$ and $\UFin$.
-\end{proposition}
+\end{theorem}
 
@@
 
-\begin{definition}[${\gdenot{\blank}} : U \to {\UFin}$]
+\begin{definition}
   \begin{gather*}
@@
 
-\begin{proposition}
+\begin{theorem}[Full Abstraction and Adequacy]
   If $c_1, c_2 : X \iso Y$, then $\denot{c_1} = \denot{c_2}$ iff $\gdenot{c_1}_{1} = \gdenot{c_2}_{1}$.
-\end{proposition}
+\end{theorem}
 

### 2021-07-09 Pi+/paper/finite.tex
Update

@@
 
-  \begin{definition}
+  \begin{definition}[Group]
     \label{def:group}
@@
 \begin{toappendix}
-  \begin{definition}
+  \begin{definition}[Free group]
     \label{def:free-group}
@@
 
-\begin{definition}
+\begin{definition}[Free group]
   \label{def:presentation}
@@
 
-\begin{definition}
+\begin{definition}[Group presentation]
   Let $A$ be a set and $R : \List[A + A] \to \List[A + A] \to \UU$ a binary relation on $\List[A + A]$. The group
@@
         swapping happens, that is, we can slide the wires freely.
-  \item[\labelcref{fig:coxeter-c}] \jk{If the next transposition in the sequence is equal to the the first one, the order of
-        operations can be reversed}.
-
->>>>>>> ee83d6f6650c383d20e046a4ada0de4cb9c42dca
+  \item[\labelcref{fig:coxeter-c}] There are two equivalent ways of swapping the first and the last element in a
+  three-element sequence.
 \end{enumerate}
@@
       {braid={ s_1, s_1}};
-      \node[font=\large] at (1, 0.7) {\(\rightarrow\)};
+      \node[font=\large] at (1, 0.7) {\(=\)};
       \pic[local bounding box=my braid,braid/.cd,
@@
       {braid={ 1, s_1 }};
-      \node[font=\large] at (1.9, 0.7) {\(\rightarrow\)};
+      \node[font=\large] at (1.9, 0.7) {\(=\)};
       \pic[local bounding box=my braid,braid/.cd,
@@
       {braid={ s_2, s_1, s_2}};
-      \node[font=\large] at (1.4, 0.7) {\(\rightarrow\)};
+      \node[font=\large] at (1.4, 0.7) {\(=\)};
       \pic[local bounding box=my braid,braid/.cd,
@@
 
-\begin{proposition}
+\begin{theorem}
   \label{prop:nf}
@@
   \end{enumerate}
-\end{proposition}
+\end{theorem}
 
@@
 
-\begin{proposition}
+\begin{theorem}
   \leavevmode
@@
 
-\begin{definition}
+\begin{definition}[Inversion count]
   Given a permutation $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by
@@
 
-\begin{propositionrep}
+\begin{theoremrep}
   \leavevmode
@@
   Therefore, there is an equivalence between $\Lehmer[n]$ and $\im{\normf}$.
-\end{propositionrep}
+\end{theoremrep}
 
@@
 
-To do so, we need to construct some equivalences by counting the elements of $\Fin[n]$ using its decidable equality.
-First, we define a type family $\FinExcept{n} : \Fin[n] \to \UU$ which picks out all elements in $\Fin[n]$ except
-the one in the argument. Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence a
-set.
-
-\begin{definition}
-  \( \FinExcept{n}[i] \defeq \dsum{j : \Fin[n]}{i \neq j} \).
+\begin{definition}[${\FinExcept{n}} : {\Fin[n]} \to \UU$]
+  For $n : \Nat$, the type family $\FinExcept{n}$ picks out all elements in $\Fin[n]$ except the one in the argument.
+  \[
+    \FinExcept{n}[i] \defeq \dsum{j : \Fin[n]}{i \neq j} \).
+  \]
 \end{definition}
 
-We state and prove a few auxiliary lemmas about how $\FinExcept{n}$ interacts with $\Fin$, these are obtained by
-counting arguments using the fact that $\Fin[n]$ and $\FinExcept{n}$ both have decidable equality.
+Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence a set. We state and prove a few
+auxiliary lemmas about how $\FinExcept{n}$ interacts with $\Fin$ -- these are obtained by counting arguments using the
+fact that $\Fin[n]$ and $\FinExcept{n}$ both have decidable equality.
 
-\begin{propositionrep}
+\begin{lemmarep}
   \label{prop:fin-finexcept}
@@
   \end{enumerate}
-\end{propositionrep}
+\end{lemmarep}
 
@@
 
-\begin{proposition}
+\begin{theorem}
   \label{prop:lehmer-aut-equiv}
   For all $n:\Nat$, \( \Lehmer[n] \eqv \Aut[\Fin[\suc[n]]] \).
-\end{proposition}
+\end{theorem}
 

### 2021-07-09 Pi+/paper/ufin.tex
Update

@@
   projection ${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
-  fibration, that is, there is a lifting operation (\Cref{fig:lift}) which lifts paths in the base space to paths in the
+  fibration, that is, there is a lifting operation (\cref{fig:lift}) which lifts paths in the base space to paths in the
   total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
-
   \[
@@
 
-  \begin{proposition}
+  \begin{lemma}
     The following are equivalent.
@@
     \end{enumerate}
-  \end{proposition}
+  \end{lemma}
 
@@
 
-\begin{propositionrep}
+\begin{lemmarep}
   \label{prop:isFin}
   For any type $X$, $\isFin[X]$ is a proposition.
-\end{propositionrep}
+\end{lemmarep}
 
@@
 
-Since $\isFin[X]$ is a predicate on the universe $\UU$, we easily get our univalent subuniverse $\UFin$.
+Since $\isFin$ is a predicate on the universe $\UU$, we easily get our univalent subuniverse $\UFin$.
 
@@
   generates the subuniverse
-  $
+  \[
     \BAut[T] \defeq \Sub{T}.
-  $
+  \]
   We write $T_0 \defeq (T, \trunc{\refl_{T}})$ for the image of the inclusion of $T$ in $\BAut[T]$.
@@
 
-\begin{definition}
+\begin{definition}[$\UFin[n]$]
   For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.
@@
 
-\begin{propositionrep}
+\begin{lemmarep}
   \leavevmode
@@
 
-\begin{corollary}
+\begin{theorem}
   $\UFin[n]$ is a pointed, connected, 1-groupoid for every $n:\Nat$, and \( \loopspace[\UFin[n],F_{n}] \eqv
   \Aut[\Fin[n]] \). $\UFin$ is a 1-groupoid with connected components for every $n:\Nat$.
-\end{corollary}
+\end{theorem}
 
@@
 
-\begin{proposition}
+\begin{theorem}
   $\UFin$ has a symmetric rig structure, with the additive and multiplicative symmetric monoidal structures given by
@@
   the Mac Lane coherence conditions for symmetric monoidal categories, that is, the triangle, pentagon, and hexagon
-  identities, and the symmetry of the braiding, upto 2-paths in $\UFin$.
-\end{proposition}
+  identities, and the symmetry of the braiding, upto 2-paths in $\UFin$. \todo{Add rig}
+\end{theorem}
 

### 2021-07-09 Pi+/paper/equivalence.tex
Fixes
Fix appendix ref

@@
 by relating them with a truncation. These can be found in the appendix and in the accompanying Agda code
-(~\cref{app:level-two}).
+(~\cref{app:leveltwo}).
 
@@
 
-\begin{definition}
+\begin{definition}[${\gdenot{\blank}}$]
   \begin{gather*}
     \begin{aligned}
+      \gdenot{\blank}_{0} &: U \to \UFin \\
       \gdenot{\blank}_{0} &\defeq \evalh_{0} \comp \evalp_{0} \comp \evalt_{0}
@@
     \begin{aligned}
+      \gdenot{\blank}_{0} &: X \iso Y \to \gdenot{X}_{0} \id_{\UFin} \gdenot{Y}_{0} \\
       \gdenot{\blank}_{1} &\defeq \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
@@
 \begin{theorem}[Full Abstraction and Adequacy]
-  If $c_1, c_2 : X \iso Y$, then $\denot{c_1} = \denot{c_2}$ iff $\gdenot{c_1}_{1} = \gdenot{c_2}_{1}$.
+  For any $c_1, c_2 : X \iso Y$, we have that
+  \[
+    \denot{c_1} = \denot{c_2} \text{ if and only if } \gdenot{c_1}_{1} = \gdenot{c_2}_{1}
+  \]
 \end{theorem}

### 2021-07-09 Pi+/paper/discussion.tex
Change sigma notation

@@
 
-\begin{center}
-    \begin{tabular}{c c c}
-        \hline
-        $\PiLang$     & $\Sn$               & $\UFin$     \\
-        \hline
-        types         & Natural numbers     & Finite sets \\
-        \hline
-        1-combinators & Generators of $\Sn$ & 1-paths     \\
-        \hline
-        2-combinators & Relations of $\Sn$  & 2-paths     \\
-        \hline
-    \end{tabular}
-\end{center}
+% \begin{center}
+%     \begin{tabular}{c c c}
+%         \hline
+%         $\PiLang$     & $\Sn$               & $\UFin$     \\
+%         \hline
+%         types         & Natural numbers     & Finite sets \\
+%         \hline
+%         1-combinators & Generators of $\Sn$ & 1-paths     \\
+%         \hline
+%         2-combinators & Relations of $\Sn$  & 2-paths     \\
+%         \hline
+%     \end{tabular}
+% \end{center}
 
-\begin{center}
-    \begin{tabular}{c c c}
-        \hline
-        Logic                              & Program                       & Algebra                     \\
-        \hline
-        Propositions                       & Types                         & Objects                     \\
-        \hline
-        Proofs                             & Terms                         & Morphisms                   \\
-        \hline
-        $\blank$                           & Equations                     & Homotopies                  \\
-        \hline
-        Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
-        \hline
-        Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
-        \hline
-        Connectives                        & Type constructors             & Bifunctors                  \\
-        \hline
-        ?                                  & Type normalisation            & ?                           \\
-        \hline
-        ?                                  & Program normalisation         & ?                           \\
-        \hline
-        ?                                  & Soundness of 2-combinators    & ?                           \\
-    \end{tabular}
-\end{center}
+% \begin{center}
+%     \begin{tabular}{c c c}
+%         \hline
+%         Logic                              & Program                       & Algebra                     \\
+%         \hline
+%         Propositions                       & Types                         & Objects                     \\
+%         \hline
+%         Proofs                             & Terms                         & Morphisms                   \\
+%         \hline
+%         $\blank$                           & Equations                     & Homotopies                  \\
+%         \hline
+%         Intuitionistic propositional logic & Simply typed Lambda Calculus  & Cartesian Closed Categories \\
+%         \hline
+%         Superstructural Reversible logic   & Reversible Language $\PiLang$ & Symmetric Rig Groupoids     \\
+%         \hline
+%         Connectives                        & Type constructors             & Bifunctors                  \\
+%         \hline
+%         ?                                  & Type normalisation            & ?                           \\
+%         \hline
+%         ?                                  & Program normalisation         & ?                           \\
+%         \hline
+%         ?                                  & Soundness of 2-combinators    & ?                           \\
+%     \end{tabular}
+% \end{center}
 
-\begin{center}
-    \begin{tabular}{c c c}
-        \hline
-        Reversible Logic             & Reversible Programming                  & Algebra                  \\
-        \hline
-        Free commutative rig         & Free symmetric rig groupoid             & $\UFin$                  \\
-        \hline
-        Propositions                 & Types                                   & Objects                  \\
-        \hline
-        Entailment?                  & Reversible programs                     & Isomorphisms             \\
-        \hline
-        $\blank$                     & 2-combinators                           & Homotopies               \\
-        \hline
-        Conjunction?                 & $\times$                                & $\otimes$                \\
-        \hline
-        Disjunction?                 & $+$                                     & $\oplus$                 \\
-        \hline
-        $A \land B \vdash B \land A$ & $\swapt$                                & $\mathcal{B}_{A,B}$      \\
-        \hline
-        True proposition             & $\onet$                                 & $F_1$                    \\
-        \hline
-        Decidability                 & Program Synthesis                       & Reachability in a graph? \\
-        \hline
-        Soundness/Completeness       & Soundness/Completeness of 1-combinators & ?                        \\
-    \end{tabular}
-\end{center}
+% \begin{center}
+%     \begin{tabular}{c c c}
+%         \hline
+%         Reversible Logic             & Reversible Programming                  & Algebra                  \\
+%         \hline
+%         Free commutative rig         & Free symmetric rig groupoid             & $\UFin$                  \\
+%         \hline
+%         Propositions                 & Types                                   & Objects                  \\
+%         \hline
+%         Entailment?                  & Reversible programs                     & Isomorphisms             \\
+%         \hline
+%         $\blank$                     & 2-combinators                           & Homotopies               \\
+%         \hline
+%         Conjunction?                 & $\times$                                & $\otimes$                \\
+%         \hline
+%         Disjunction?                 & $+$                                     & $\oplus$                 \\
+%         \hline
+%         $A \land B \vdash B \land A$ & $\swapt$                                & $\mathcal{B}_{A,B}$      \\
+%         \hline
+%         True proposition             & $\onet$                                 & $F_1$                    \\
+%         \hline
+%         Decidability                 & Program Synthesis                       & Reachability in a graph? \\
+%         \hline
+%         Soundness/Completeness       & Soundness/Completeness of 1-combinators & ?                        \\
+%     \end{tabular}
+% \end{center}
 

### 2021-07-09 Pi+/paper/ufin.tex
Fixes
Change sigma notation

@@
   A universe predicate is a type family $P : \UU \to \UU$ whose fibers are propositions, that is, $P(X)$ is a
-  proposition for every $X$. Given such a predicate, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is univalent
-  and generates a univalent subuniverse $\UU_{P} \defeq (\dsum{X:\UU}{P(X)}, \fst)$.~\footnote{Univalent
+  proposition for every $X$. Given such a predicate $P$, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is
+  univalent and generates a univalent subuniverse ${\UU_{P} \defeq (\dsum*{X:\UU}{P(X)}, \fst)}$.~\footnote{Univalent
     typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent subuniverses.}
@@
   \begin{gather*}
-    \Fin[n] \defeq \dsum{k:\Nat}{k < n}
+    \Fin[n] \defeq \dsum*{k:\Nat}{k < n}
     \qquad\qquad \text{or} \qquad\qquad
@@
   \[
-    \isFin[X] \defeq \dsum{n:\Nat}{\SubP{X}{\Fin[n]}}
+    \isFin[X] \defeq \dsum*{n:\Nat}{\SubP{X}{\Fin[n]}}
   \]
@@
   $
-    \UFin \defeq \dsum{X:\UU}{\isFin[X]}.
+    \UFin \defeq \dsum*{X:\UU}{\isFin[X]}.
   $
@@
 
-\begin{definition}[$\UFin[n]$]
+\begin{definition}[${\UFin[n]}$]
   For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.

### 2021-07-09 Pi+/paper/equivalence.tex
Futher work

@@
 \label{sec:equivalence}
+
+In this section, we first translate $\PiLang$ to its additive fragment $\PiPlusLang$. This is the language that we
+interpret to $\UFin$ and back, using the tools developed in the previous sections. Further, we go through an
+intermediate step of the language $\PiHatLang$, which is a simplified variant of $\PiPlusLang$ that uses adjacent
+transpositions for combinators, while preserving all the required structure.
 
@@
 
-In this section, we first translate $\PiLang$ to its additive fragment $\PiPlusLang$. This is the language that we
-interpret to $\UFin$ and back, using the tools developed in the previous sections. Further, we go through an
-intermediate step of the language $\PiHatLang$, which is a simplified variant of $\PiPlusLang$ that uses adjacent
-transpositions for combinators, while preserving all the required structure.
-
 We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\Cref*{fig:piplus,fig:pihat} respectively,
 eliding the 2-combinators for brevity. We enforce that there is a unique 2-combinator between compatible 1-combinators,
-by relating them with a truncation. These can be found in the appendix and in the accompanying Agda code
-(~\cref{app:leveltwo}).
+by relating them with a truncation. These can be found in~\cref{app:leveltwo} and in the accompanying Agda code.
 
@@
 First, we show how to translate $\PiLang$ programs to $\PiPlusLang$, which is the additive fragment of $\PiLang$. The
-syntax is given in~\Cref{fig:piplus}.
+syntax for 1-combinators is given in~\Cref{fig:piplus}.
 
@@
 
-Using this rig structure, we translate $\PiLang$ to $\PiPlusLang$, constructing a rig equivalence from $\PiCat$ to
-$\PiPlusCat$.
+\noindent Using this rig structure, we translate $\PiLang$ to $\PiPlusLang$, constructing a rig equivalence from
+$\PiCat$ to $\PiPlusCat$.
 
@@
 As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers, and,
-we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal structure. Since each object is a
-natural number generated by $\zero$ and $\suc$, this makes $\PiHatCat$ similar to a \emph{PROP}, that is, a products and
-permutations category.
+we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal structure.~\footnote{Since each object
+is a natural number, this makes $\PiHatCat$ a \emph{PROP}, that is, a products and permutations category.}
+
+To produce a braiding $n + m \isoh m + n$ from adjacent tranpositions, we recursively traverse the left subexpression,
+swapping each element using adjacent transpositions along the elements on the right, placing it in the right position.
+The computational content of this translation can be visualised using tree transformations, for the recursive case
+in~\cref{fig:plusplusswap}. The challenging part is showing that these moves are coherent with respect to 2-combinators.
+
+\begin{figure}
+   \[
+      \Tree [ [ {\tiny 1} {\tiny n} ] {\tiny m} ] ~\xrightarrow{}~
+      \Tree [ {\tiny 1} [ {\tiny n} {\tiny m} ] ] ~\xrightarrow{}~
+      \Tree [ [ {\tiny n} {\tiny m} ] {\tiny 1} ] ~\xrightarrow{}~
+      \Tree [ [ {\tiny m} {\tiny n} ] {\tiny 1} ] ~\xrightarrow{}~
+      \Tree [ {\tiny m} [ {\tiny n} {\tiny 1} ] ] ~\xrightarrow{}~
+      \Tree [ {\tiny m} [ {\tiny 1} {\tiny n} ] ] ~
+    \]
+  \label{fig:plusplusswap}
+  \caption{Braiding from transpositions, recursive case}
+\end{figure}
 
@@
 
-Using this symmetric monoidal structure, we translate from $\PiPlusLang$ to $\PiHatLang$.
+\noindent Using this symmetric monoidal structure, we translate from $\PiPlusLang$ to $\PiHatLang$.
 
@@
 To go back from $\PiHatLang$ to $\PiPlusLang$, we turn a natural number into a $\PiPlusLang$ type, using right-biased
-addition, that is, the natural number $n$ gets mapped to the type $\onet + (\onet + (\onet + \ldots + \zerot))$. The
-non-trivial part is coherently generating a $\PiPlusLang$ combinator from $\PiHatLang$ combinators, using adjacent swaps
-to generate the braiding.
+addition, that is, the natural number $n$ gets mapped to the type $\onet + (\onet + (\onet + \ldots + \zerot))$. Since
+the types are already right-biased, an adjacent transposition in $\PiHatLang$ is easily encoded by using the braiding in
+$\PiPlusLang$.
+
+\begin{figure}
+  \[
+     \Tree [ {\tiny I} [ {\tiny I} {\tiny n} ] ] ~\xrightarrow{\assoclp}~
+     \Tree [ [ {\tiny I} {\tiny I} ] {\tiny n} ] ~\xrightarrow[\phantom{xx}\swapp]{\idc\phantom{xx}}~
+     \Tree [ [ {\tiny I} {\tiny I} ] {\tiny n} ] ~\xrightarrow{\assocrp}~
+     \Tree [ {\tiny I} [ {\tiny I} {\tiny n} ] ] ~
+   \]
+ \label{fig:plusplusswap}
+ \caption{Transpositions from braiding}
+\end{figure}
 
@@
     \begin{aligned}
-      \gdenot{\blank}_{0} &: U \to \UFin \\
-      \gdenot{\blank}_{0} &\defeq \evalh_{0} \comp \evalp_{0} \comp \evalt_{0}
+      \gdenot{\blank}_{0} & : U \to \UFin                                       \\
+      \gdenot{\blank}_{0} & \defeq \evalh_{0} \comp \evalp_{0} \comp \evalt_{0}
     \end{aligned}
@@
     \begin{aligned}
-      \gdenot{\blank}_{0} &: X \iso Y \to \gdenot{X}_{0} \id_{\UFin} \gdenot{Y}_{0} \\
-      \gdenot{\blank}_{1} &\defeq \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
+      \gdenot{\blank}_{0} & : X \iso Y \to \gdenot{X}_{0} \id_{\UFin} \gdenot{Y}_{0} \\
+      \gdenot{\blank}_{1} & \defeq \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
     \end{aligned}

### 2021-07-09 Pi+/paper/examples.tex
Revise sections 6 and 7 and copy-editing
Futher work

@@
 \label{sec:qiskit}
+\label{sec:examples}
 
-Classical reversible boolean circuits are at the core of most quantum algorithms and
-hence are supported by popular platforms for quantum computing such as IBM
-Qiskit. Specifically, the Qiskit framework provides the following universal set
-of gates for reversible computing: \textsf{not} (boolean negation, called
-\verb|x|), \textsf{cnot} (conditional negation of the second input if the first
-is true; called \verb|cx|), and \textsf{toffoli} (conditional negation of the
-third input if both the first two inputs are true; called \verb|ccx|)
-gates. Additionally, Qiskit allows implicit re-shuffling of bits by allowing
-each operation to specify the indices of its input bits.
+Classical reversible boolean circuits are at the core of most quantum algorithms and hence are supported by popular
+platforms for quantum computing such as IBM Qiskit~\cite{aleksandrowiczQiskitOpensourceFramework2019}. Specifically, the
+Qiskit framework provides the following universal set of gates for reversible computing: \textsf{not} (boolean negation,
+called \verb|x|), \textsf{cnot} (conditional negation of the second input if the first is true; called \verb|cx|), and
+\textsf{toffoli} (conditional negation of the third input if both the first two inputs are true; called \verb|ccx|)
+gates. Additionally, Qiskit allows implicit re-shuffling of bits by allowing each operation to specify the indices of
+its input bits.
 

### 2021-07-09 Pi+/paper/finite.tex
Fixes
Futher work

@@
 We will show that it is possible to encode all $\PiLang$ combinators using adjacent transpositions
-(in~\cref{sec:equivalence}). Group presentations are built by adding equations to a \emph{free group}, which we define
-first.
+(in~\cref{sec:equivalence}). Group presentations are built by adding equations to a \emph{free group}.
 
@@
 directed, so it is not always possible to construct a well-behaved rewriting system. In general, the word problem for
-groups is known to be undecidable.
+groups is undecidable.
 
-\subsubsection{Coxeter Presentation}
+\paragraph{Coxeter Presentation} To present the group $\Sn$, the primitive operations we use will be adjacent swaps.
+When dealing with permutations on an $n + 1$-element set, there are $n$ adjacent transpositions -- transposition number
+$k$ swaps elements at indices $k$ and $k+1$. Thus, the generating set is $\Fin[n]$. There are three relations that we're
+going to specify for this presentation -- we visualise them as braid diagrams in~\Cref{fig:coxeter-braid}.
 
-To present the group $\Sn$, the primitive operations we are going to use will be adjacent swaps. When dealing with
-permutations on an $n + 1$-element set, there are $n$ adjacent transpositions, transposition number $k$ swapping
-elements at indices $k$ and $k+1$. Thus, the generating set would be $\Fin[n]$. There are three relations that we're going
-to specify for this presentation -- these are the equations that these generators should satisfy. We visualise them as
-braid diagrams in~\Cref{fig:coxeter-braid}.
-
-\begin{enumerate}
-  \item[\labelcref{fig:coxeter-a}] Swapping the same two elements two times in a row should be the same as doing nothing
-  \item[\labelcref{fig:coxeter-b}] When swapping two distinct pairs of elements, it should not matter in which order the
-        swapping happens, that is, we can slide the wires freely.
-  \item[\labelcref{fig:coxeter-c}] There are two equivalent ways of swapping the first and the last element in a
-  three-element sequence.
+\begin{enumerate}[leftmargin=*]
+  \item[\labelcref{fig:coxeter-a}] Swapping the same two elements twice in a row should be the same as doing nothing.
+  \item[\labelcref{fig:coxeter-b}] When swapping two distinct pairs of elements, the order in which swapping happens
+        should not matter, that is, we can slide the wires freely.
+  \item[\labelcref{fig:coxeter-c}] There are two equivalent ways of swapping the first and last elements in a sequence
+        of three elements.
 \end{enumerate}
@@
 system $(\List[\Fin[n]], \longcox)$, partially based on the Coxeter relations, and prove that it has the desired
-properties of confluence and termination. We prove that the new relation defined by this system is equivalent,
-in a technical sense, to the standard Coxeter relation $\cox*$.
-
-First, we need to define a function $n \downf k$.
+properties of confluence and termination. We prove that the new relation defined by this system is equivalent, in a
+technical sense, to the standard Coxeter relation $\cox*$. First, we need to define a function $n \downf k$.
 
@@
 
-The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\). The role of this function
-is to produce a sequence of transpositions which moves element at index $k + n$ left by $k$ places, shifting all the
-elements in between one place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the
-following generators, inlining the congruence closure in $\longcox$, allowing arbitrary reductions inside the list.
+\noindent The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\), which is a
+sequence of transpositions that moves the element at index $k + n$ left by $k$ places, shifting all the elements in
+between one place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the following
+generators (inlining the congruence closure in $\longcox$, allowing arbitrary reductions inside the list).
 
@@
 
-Constructors $\longcancel$ and $\longswap$ correspond directly to the appropriate constructors of $\cox$ and can be
-visualised in the same way as before. The remaining constructor $\longbraid$ uses the $\downf$ function to exchange the
-order of a long sequence of transpositions and a single transposition afterwards. For example, for $n = 0$ and $k = 3$,
-it allows for the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox} -- note the
-distinction between wires, where numbers represent the values, and transpositions, where numbers represent which wires
-are crossing).
+\noindent Constructors $\longcancel$ and $\longswap$ correspond directly to the appropriate constructors of $\cox$ and
+can be visualised in the same way as before. The remaining constructor $\longbraid$ uses the $\downf$ function to
+exchange the order of a long sequence of transpositions and a single transposition afterwards. For example, for $n = 0$
+and $k = 3$, it allows for the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox} --
+note the distinction between wires, where numbers represent the values, and transpositions, where numbers represent
+which wires are crossing).
 
@@
 \end{figure}
-
-Note that the previous $\braid$ rule is a special case of $\longbraid$, with $k = 0$. Also, as before, the left-hand
-sides of the relation are strictly larger than the right-hand sides, in terms of the lexicographic order on
-$\List[\Fin[n]]$.
+% in terms of the lexicographic order on $\List[\Fin[n]]$.
 
@@
 
-For the relation $\longcox$, we define its transitive closure to be $\longcoxplus$ (\cref{def:longcox-plus}) and its
-reflexive-transitive closure to be $\longcoxstar$ (\cref{def:longcox-star}).
+Note that the previous $\braid$ rule is a special case of $\longbraid$, with $k = 0$. As before, the left-hand sides of
+the relation are (lexicographically) strictly larger than the right-hand sides. We define the transitive closure of
+$\longcox$ to be $\longcoxplus$ (\cref{def:longcox-plus}) and its reflexive-transitive closure to be $\longcoxstar$
+(\cref{def:longcox-star}).
 
 Despite the increased complexity of the generators, the rewriting system $(\List[\Fin[n]],\longcox)$ has the properties
-we desire. It satisfies local confluence, that is, the Church-Rosser (diamond) property (for example, using
-$\longbraid$, we can now show the problematic critical pair \cref{fig:critical-pairs-non-converging} to converge), and it is
-terminating, so, Newman's lemma, it produces a unique normal form. We follow the terminology
-of~\cite*{huetConfluentReductionsAbstract1980a,krausCoherenceWellFoundednessTaming2020} to state our results formally.
+we desire. It satisfies local confluence, that is, the Church-Rosser (diamond) property -- for example, using
+$\longbraid$, we can now show the problematic critical pair \cref{fig:critical-pairs-non-converging} converges, and, it
+is terminating, so by Newman's lemma, it produces a unique normal form. We follow the terminology
+of~\citet*{huetConfluentReductionsAbstract1980a,krausCoherenceWellFoundednessTaming2020} to state our results formally.
 
@@
 By~\Cref{prop:nf}~\ref{prop:nf-uniqueness}, we get a unique choice function $\normf : {\List[\Fin[n]]} \to
-  {\List[\Fin[n]]}$ that produces a normal form for terms of $\List[\Fin[n]]$. We state and prove a couple of important
-properties enjoyed by $\normf$.
+  {\List[\Fin[n]]}$ that produces a normal form for terms of $\List[\Fin[n]]$. We state two important properties enjoyed
+by $\normf$.
 
@@
 
-Finally, we define the type $\Sn$ as the set-quotient of $\List[\Fin[n]]$ by $\cox*$,
+\noindent Finally, we define the type $\Sn$ as the set-quotient of $\List[\Fin[n]]$ by $\cox*$,
 
@@
 
-Note that this relation $\cox*$ is not prop-valued, that is, $w \cox* v$ is not a proposition because reductions need
-not be unique. So, the quotient $\Sn$ is not effective, that is, $\quotrel : w \cox* v \to q(w) \id q(v)$ is not an
-equivalence. Using the $\normf$ function, we could instead define a new relation $\approx$ to equate those terms that
-have the same normal form, $(w \approx v) \defeq (\normf(w) \id \normf(v))$. This relation is prop-valued, and we could
-quotient $\List[\Fin[n]]$ by $\approx$, obtaining an equivalent definition for $\Sn$.
+\noindent Note that reductions need not be unique, hence $w \cox* v$ is not necessarily a proposition. So, the quotient
+$\Sn$ is not effective, that is, $\quotrel : w \cox* v \to q(w) \id q(v)$ is not an equivalence. Using the $\normf$
+function, we could instead define a new relation $\approx$ to equate those terms that have the same normal form, $(w
+  \approx v) \defeq (\normf(w) \id \normf(v))$. This relation is prop-valued, and we could quotient $\List[\Fin[n]]$ by
+$\approx$, obtaining an equivalent definition for $\Sn$.
 
@@
 
-\subsection{Lehmer Codes}~\label{subsec:lehmer}
+\subsection{Lehmer Codes}
+\label{subsec:lehmer}
 
@@
 Formally, we define $\Lehmer[n]$ to be an $n+1$-element tuple, where the position $k \leq n$ stores an element of
-$\Fin[k]$. Since the 0-th position is trivial, in practice it is ignored, and the sequence starts at
-1~\cite{duboisTestsProofsCustom2018,vajnovszkiNewEulerMahonian2011}.
+$\Fin[k]$. Since the 0-th position is trivial, in practice it is
+ignored~\cite{duboisTestsProofsCustom2018,vajnovszkiNewEulerMahonian2011}.
 
@@
 
-From knowing just the inversion counts for all the elements, one can reconstruct the starting permutation. Also, observe
-that $\invcount{i} < i$, thus fitting in the $i$-th place of a Lehmer code tuple. As an example, consider the following
+Just knowing the inversion counts for all the elements, one can reconstruct the starting permutation. Also, observe that
+$\invcount{i} < i$, thus fitting in the $i$-th place of a Lehmer code tuple. As an example, consider the following
 tabulated presentation of a permutation of $\Fin[5]$.
@@
 
-The Lehmer code for the permutation $\sigma$ is then the 5-tuple
+\noindent The Lehmer code for the permutation $\sigma$ is then the 5-tuple
 \[
@@
 
-To decode the permutation back from this Lehmer code, we perform an algorithm similar to \emph{insertion sort}. The
-element of the Lehmer code being currently processed is highlighted in the left column of the table below. Starting from
-a sorted list, the element at index $k$ has to be given $l[k]$ inversions. Because of the the invariant that all the
-elements before newly processed one are smaller than it, the proper number of inversions is created by simply shifting
-the element $l[k]$ places left.
+\noindent To decode the permutation back from this Lehmer code, we perform an algorithm similar to \emph{insertion
+  sort}. The element of the Lehmer code being currently processed is highlighted in the left column of the table below.
+Starting from a sorted list, the element at index $k$ has to be given $l[k]$ inversions. Because of the the invariant
+that all the elements before newly processed one are smaller than it, the proper number of inversions is created by
+simply shifting the element $l[k]$ places left.
 
@@
 
-We can show that the function $\immersion_{n}$ gives an equivalence betweeen $\Lehmer[n]$ and $\im{\normf}$.
+\noindent We can show that the function $\immersion_{n}$ gives an equivalence betweeen $\Lehmer[n]$ and $\im{\normf}$.
 
@@
 between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
-between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$. The idea of this proof is borrowed from~\cite{Molzer-cubical}
-~\footnote{Note that the indices for the type of permutations are off-by-one, because we chose $\Fin[n-1]$ to represent
-  generators for permutations on $\Fin[n]$.}
+between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$. The idea for this proof is borrowed
+from~\cite{Molzer-cubical}. ~\footnote{Note that the indices for the type of permutations are off-by-one, because we
+  chose $\Fin[n]$ to represent generators for permutations on $\Fin[\suc[n]]$.}
 
@@
   \[
-    \FinExcept{n}[i] \defeq \dsum{j : \Fin[n]}{i \neq j} \).
+    \FinExcept{n}[i] \defeq \dsum*{j : \Fin[n]}{i \neq j}
   \]
@@
 
-Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence a set. We state and prove a few
-auxiliary lemmas about how $\FinExcept{n}$ interacts with $\Fin$ -- these are obtained by counting arguments using the
-fact that $\Fin[n]$ and $\FinExcept{n}$ both have decidable equality.
+\noindent Note that $\FinExcept{n}[i]$ for $i : \Fin[n]$ is a subtype of $\Fin[n]$ and is hence a set. We state and
+prove a few auxiliary lemmas about how $\FinExcept{n}$ interacts with $\Fin$ -- these are obtained by counting arguments
+using the fact that $\Fin[n]$ and $\FinExcept{n}$ both have decidable equality.
 
@@
     % \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-1}
-    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-2}
-    \item For any $n : \Nat$, \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[n] \eqv
+    \item For any $k : \Fin[\suc[n]]$, we have $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-2}
+    \item For any $n : \Nat$, we have \( \Aut[\Fin[\suc[n]]] \eqv \dsum*{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[n] \eqv
             \FinExcept{\suc[n]}[n - k]} \). \label{prop:fin-finexcept-3}
@@
 
-Using these facts, we can now prove the main result of this section.
+\noindent Using these facts, we can now prove the main result of this section.
 
@@
       \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                          \\
-      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
-      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
-      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                           & \text{\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}}
+      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
+      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
+      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                           & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}}
     \end{array}
@@
 
-By composing~\Cref{prop:lehmer-aut-equiv} and~\Cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.
+\noindent By composing~\Cref{prop:lehmer-aut-equiv} and~\Cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.
 

### 2021-07-09 Pi+/paper/ufin.tex
Futher work

@@
 We have shown that loops in $\UFin$ exactly encode the automorphism group $\Aut[\Fin[n]]$ for every~$n$. This is a
-general technique called \emph{delooping}, where a group can be identified with a 1-object groupoid, internally in
-HoTT. This technique also allows defining higher groups in HoTT~\cite{buchholtzHigherGroupsHomotopy2018}. The loopspace
-of a pointed type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element, path
-composition for the group multiplication, and path inverse for the group inverse operation. The group axioms are then
-given by the higher paths corresponding to groupoid laws.
+general technique called \emph{delooping}, where a group can be identified with a 1-object groupoid, internally in HoTT.
+This technique also allows defining higher groups~\cite{buchholtzHigherGroupsHomotopy2018}. The loopspace of a pointed
+type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element, path composition for the
+group multiplication, and path inverse for the group inverse operation. The group axioms are given by the higher paths
+corresponding to groupoid laws.
 
@@
 \begin{theorem}
-  $\UFin$ has a symmetric rig structure, with the additive and multiplicative symmetric monoidal structures given by
-  $(F_0, \sqcup)$ and $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$,
-  $\alpha_{X,Y,Z}$, and the braiding isomorphism $\mathcal{B}_{X,Y}$ upto 1-paths in $\UFin$. These isomorphisms satisfy
-  the Mac Lane coherence conditions for symmetric monoidal categories, that is, the triangle, pentagon, and hexagon
-  identities, and the symmetry of the braiding, upto 2-paths in $\UFin$. \todo{Add rig}
+  $\UFin$ has two symmetric monoidal structures, the additive and multiplicative ones, given by $(F_0, \sqcup)$ and
+  $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the braiding
+  isomorphism $\mathcal{B}_{X,Y}$ upto 1-paths in $\UFin$. These isomorphisms satisfy the Mac Lane coherence conditions
+  for symmetric monoidal categories~\cite{maclaneNaturalAssociativityCommutativity1963}, that is, the triangle,
+  pentagon, and hexagon identities, and the symmetry of the braiding, upto 2-paths in $\UFin$. The multiplicative
+  structure distributes over the additive structure and satisfies the Laplaza coherence conditions for rig
+  categories~\cite{laplaza72}.
 \end{theorem}

### 2021-07-09 Pi+/paper/equivalence.tex
Rest of styling

@@
 the types are already right-biased, an adjacent transposition in $\PiHatLang$ is easily encoded by using the braiding in
-$\PiPlusLang$.
+$\PiPlusLang$, as shown in~\cref{fig:transpfrombraid}. Again, these are shown to be coherent.
 
@@
      \Tree [ {\tiny I} [ {\tiny I} {\tiny n} ] ] ~\xrightarrow{\assoclp}~
-     \Tree [ [ {\tiny I} {\tiny I} ] {\tiny n} ] ~\xrightarrow[\phantom{xx}\swapp]{\idc\phantom{xx}}~
+     \Tree [ [ {\tiny I} {\tiny I} ] {\tiny n} ] ~\xrightarrow[\swapp\phantom{xx}]{\phantom{xx}\idc}~
      \Tree [ [ {\tiny I} {\tiny I} ] {\tiny n} ] ~\xrightarrow{\assocrp}~
@@
    \]
- \label{fig:plusplusswap}
+ \label{fig:transpfrombraid}
  \caption{Transpositions from braiding}
@@
 
-Finally, we show how to interpret $\PiHatLang$ to $\UFin$, and go back from $\UFin$ to $\PiHatLang$.
-
-Types in $\PiHatLang$ are interpreted as 0-cells in $\UFin$, that is, a natural number $n$ is mapped to $\Fin[n]$. The
-1-combinators in $\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops $\Aut[\Fin[n]]$, in each connected
-component. In $\PiHatLang$, the 1-combinators are generated by adjacent transpositions, so these can be easily mapped to
-words in $\Sn$ and then to automorphisms using~\Cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to
-2-paths between loops in $\UFin$.
+Finally, we show how to interpret $\PiHatLang$ to $\UFin$, and back from $\UFin$ to $\PiHatLang$. Types in $\PiHatLang$
+are interpreted as 0-cells in $\UFin$, that is, a natural number $n$ is mapped to $\Fin[n]$. The 1-combinators in
+$\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops $\Aut[\Fin[n]]$, in each connected component. In
+$\PiHatLang$, the 1-combinators are generated by adjacent transpositions, so these can be mapped to words in $\Sn$ and
+then to automorphisms using~\Cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to 2-paths between loops
+in $\UFin$.
 
@@
 
-0-cells in $\UFin$ are mapped to their cardinalities in $\PiHatLang$, 1-loops are decoded to words in $\Sn$ to generate
-a sequence of adjacent transpositions, producing a 1-combinator in $\PiHatLang$. Finally, 2-paths are quoted back to
-2-combinators in $\PiHatLang$.
+\noindent 0-cells in $\UFin$ are mapped to their cardinalities in $\PiHatLang$, 1-loops are decoded to words in $\Sn$ to
+generate a sequence of adjacent transpositions, producing a 1-combinator in $\PiHatLang$. Finally, 2-paths are quoted
+back to 2-combinators in $\PiHatLang$.
 
@@
 
-$\UFin$ as a 1-groupoid, is a (locally strict) weak $(2,0)$-category, and we use this to state our final equivalence.
+% $\UFin$ as a 1-groupoid, is a (locally strict) weak $(2,0)$-category, and we use this to state our final equivalence.
 
@@
 
-Using our semantics, we can normalise, synthesise, prove equivalence, and generally reason about~$\PiLang$
-programs. The two key definitions are presented below.
+Using our semantics, we can normalise, synthesise, prove equivalence, and generally reason about~$\PiLang$ programs. The
+two key definitions are presented below.
 

### 2021-07-09 Pi+/paper/finite.tex
Footnote

@@
 allows us to connect it to the broader scope of computational group theory and combinatorics -- we can borrow ideas such
-as Coxeter relations and Lehmer codes.
+as Coxeter relations and Lehmer codes.~\footnote{The concepts we use can be found in any standard textbook on group
+theory, or see~\cite{symmetryBook2021} for a univalent point of view.}
 

### 2021-07-09 Pi+/paper/discussion.tex
Rest of styling
Small fixes to discussion

@@
 correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
-suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work.
+suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work. In the
+remainder of this section, we discuss some broader related work.
 
@@
 
-In the remainder of this section, we discuss some broader related work.
-
 %%%%%%%%%%%%%%
@@
 Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
-{homotopy.io}, Lean, Coq, and HoTT~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
-result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem and formalised its
-correctness and termination. \citet{Hiver-coq} describes an explicit algorithm for producing normal forms. could provide
-an alternative to our rewriting system. Other encodings of permutations as listed vectors, matrices, inductively
-generated trees (Motzkin trees), Young diagrams, or string diagrams, proved either difficult formalise in type theory or
-difficult to relate directly to the primitive type isomorphisms of $\PiLang$. The automatic
-\citet{knuthSimpleWordProblems1970} completion produced too many equations making proving correctness and termination
-intractable.
+{homotopy.io}, Agda, Coq and Lean~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
+result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem. \citet{Hiver-coq}
+describes an explicit algorithm for producing normal forms. could provide an alternative to our rewriting system. Other
+encodings of permutations as listed vectors, matrices, inductively generated trees (Motzkin trees), Young diagrams, or
+string diagrams, proved either difficult formalise in type theory or difficult to relate directly to the primitive type
+isomorphisms of $\PiLang$. The automatic \citet{knuthSimpleWordProblems1970} completion produced too many equations
+making proving correctness and termination intractable.
 
@@
 Cubical Type
-Theory\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019},
-univalence computes.
+Theory\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.
 

### 2021-07-09 Pi+/paper/equivalence.tex

@@
 $\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and
-$\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can shoe that these
+$\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can show that these
 syntactic categories here are actually $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are
@@
     \]
+    \caption{Braiding from transpositions, recursive case}
     \label{fig:plusplusswap}
-  \caption{Braiding from transpositions, recursive case}
 \end{figure}
@@
   \[
-     \Tree [ {\tiny I} [ {\tiny I} {\tiny n} ] ] ~\xrightarrow{\assoclp}~
-     \Tree [ [ {\tiny I} {\tiny I} ] {\tiny n} ] ~\xrightarrow[\swapp\phantom{xx}]{\phantom{xx}\idc}~
-     \Tree [ [ {\tiny I} {\tiny I} ] {\tiny n} ] ~\xrightarrow{\assocrp}~
-     \Tree [ {\tiny I} [ {\tiny I} {\tiny n} ] ] ~
+    \Tree [ {\tiny $\onet$} [ {\tiny $\onet$} {\tiny X} ] ] ~\xrightarrow{\assoclp}~
+    \Tree [ [ {\tiny $\onet$} {\tiny $\onet$} ] {\tiny X} ] ~\xrightarrow[\swapp\phantom{xx}]{\phantom{xx}\idc}~
+    \Tree [ [ {\tiny $\onet$} {\tiny $\onet$} ] {\tiny X} ] ~\xrightarrow{\assocrp}~
+    \Tree [ {\tiny $\onet$} [ {\tiny $\onet$} {\tiny X} ] ] ~
   \]
+  \caption{Transpositions from braiding}
   \label{fig:transpfrombraid}
- \caption{Transpositions from braiding}
 \end{figure}
@@
     \begin{aligned}
-      \gdenot{\blank}_{0} & : X \iso Y \to \gdenot{X}_{0} \id_{\UFin} \gdenot{Y}_{0} \\
+      \gdenot{\blank}_{1} & : (c : X \iso Y) \to \gdenot{X}_{0} \id_{\UFin} \gdenot{Y}_{0} \\
       \gdenot{\blank}_{1} & \defeq \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
@@
 Recall the specification of reversible disjunction from \Cref{sec:examples}, the two Qiskit circuits for implementing it, and the
-corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The normal for both circuit is
-computed to be, establishing their equivalence:
+corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The normal forms for both circuit is
+computes to the following, establishing their equivalence:
 

### 2021-07-09 Pi+/paper/finite.tex

@@
   The relations $\cox*$ and $\longcoxstar$ are equivalent in the following sense: for every $w$ and $v$, $w \cox* v$ iff
-  there is a $u$ such that $w \longcoxstar u$ and $v \longcoxstar u$.
+  there is a $u$ such that $\coxcospan*{w}{v}{u}$.
 \end{proposition}

### 2021-07-09 Pi+/paper/reversible1.tex
Fixes

--- 2021-07-09 Pi+/paper/reversible1.tex
@@
 
-The circuit model of reversible computation discussed in the previous section is a useful abstraction close to the hardware
-platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an ``assembly-level''
-programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical model based on
-permutations of finite sets provides a richer algebraic structure which we review in this section.
+The circuit model of reversible computation discussed in the previous section is a useful abstraction close to the
+hardware platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an
+``assembly-level'' programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical
+model based on permutations of finite sets provides a richer algebraic structure, using which we describe a reversible
+programming language in this section.
 
@@
 These commutative rig identities can be used to design a logic for reversible
-programming~\cite*{sparksSuperstructuralReversibleLogic2014}. But as natural number identities have no computational
-content, the logic needs to be equipped with values and types, and a notion of operational semantics and contextual
-equivalence to obtain a computational interpretation of the commutative rig
-structure~\cite{jamesInformationEffects2012}.  On the semantic side, the natural space to consider is the
-groupoidification of a commutative rig, that is, a symmetric rig groupoid.
+programming~\cite*{sparksSuperstructuralReversibleLogic2014}. To interpret natural number identities as reversible programs,
+the logic needs to be equipped with values and types, and a notion of operational semantics and contextual equivalence,
+giving a computational interpretation of the commutative rig structure~\cite{jamesInformationEffects2012}. On the
+semantic side, the natural space to consider is the groupoidification of a commutative rig, that is, a symmetric rig
+groupoid.
 
@@
 groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
-semantics, rig groupoid \emph{coherence conditions} that identify different syntactic representations of the same
-permutation~\cite{laplaza72,Carette2016} were collected  in a
+semantics, \emph{coherence conditions} for symmetric rig groupoids that identify different syntactic representations of the same
+permutation~\cite{laplaza72,caretteComputingSemiringsWeak2016}, were collected  in a
 second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
@@
 
-\noindent The full set of level-2 combinators is large; the remaining combinators are listed in App.~\ref{app:leveltwo}.
+\noindent The full set of level-2 combinators is large; the remaining combinators are listed in~\cref{app:leveltwo}.
 
 \begin{toappendix}
-  \label{app:leveltwo}
+  \subsection{2-combinators}
 
   The additional level-2 combinators:
+
+  \label{app:leveltwo}
 

### 2021-07-09 Pi+/paper/ufin.tex

@@
 
-Using categorical language, the setting for the semantics in the previous section is the category of finite sets and
+In categorical language, the setting for the semantics in the previous section is the category of finite sets and
 functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
@@
 We observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
-homotopies. We therefore \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells
-and work in HoTT (Univalent Foundations) as it provides a proof-relevant, constructive metatheory to get a handle
+homotopies. We therefore \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells.
+We work in HoTT (Univalent Foundations) as it provides a proof-relevant, constructive metatheory to get a handle
 on these equalities and provides a rich internal language for describing weak groupoids, using the ``types are weak
@@
 Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
-has one point $T_0$, and 1-paths $T_0 \id_{\BAut[T]} T_0$, that is, loops, and higher paths between these loops. The
+has one point $T_0$, and 1-paths $T_0 \id T_0$, that is, loops, and higher paths between these loops. The
 type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$, is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is

### 2021-07-09 Pi+/paper/discussion.tex
Finalise

@@
 
-Coherence theorems are famous problems in (higher) category theory, and Mac Lane's coherence
+Coherence theorems are famous problems in category theory, and Mac Lane's coherence
 theorem~\cite{maclaneNaturalAssociativityCommutativity1963,joyalBraidedTensorCategories1993,gurskiInfiniteLoopSpaces2013}
-for monoidal categories is a particular one. The use of rewriting theory and proof assistants to establish coherence
-theorems for various higher categorical structures has a long history as well,
+for monoidal categories is a particular one. The use of rewriting and proof assistants to prove coherence theorems
+for higher categorical structures has a long history
 see~\cite*{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.
@@
 2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
-provides an intuition on why certain constructions are hard (or impossible in the general case).
-
-There are other, different approaches to computing with univalence,
+provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,
+different approaches to computing with univalence,
 in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
 Cubical Type
-Theory\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.
+Theory\cite*{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.
 
@@
 theories are given by algebras of (finitary) monads on $\SetCat$, models of 2-theories are given by algebras of 2-monads
-on $\CatCat$. The particular one related to our development is the free symmetric monoidal completion 2-monad.
+on $\CatCat$. Our development is related to the free symmetric monoidal completion 2-monad.
 
@@
 categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The
-existence of the proof is folklore. We have produced a proof and formalised it in constructive type theory.
-
-A stronger universal property that $\BFin$ is biinitial in the 2-category of rig categories, was established
-in~\cite{elguetaGroupoidFiniteSets2021}.
+existence of the proof is folklore. We have produced a proof and formalised it in constructive type theory. A stronger
+universal property that $\BFin$ is biinitial in $\term{RigCat}$, is in~\cite{elguetaGroupoidFiniteSets2021}.
 
@@
 %%%%%%%%%%%%%%
-\paragraph{Permutations} One might argue that finding formal systems for expressing various flavors of computable
-functions has been a major focus of logic and computer science since its inception. Permutations, being at the core of
-reversible computing, are an interesting class of functions, the bijective ones, for which there is no sound and
-complete formal system. Our work develops such a formal system bringing in all the associated benefits of
-syntactic calculi, most notably, their calculational flavor. Thus instead of comparing two reversible programs by
-extensional equality of the underlying bijective functions, a calculus offers more nuanced techniques that can enforce
-additional intensional constraints on the desired equality relation.
+\paragraph{Permutations} Finding formal systems for expressing various flavors of computable functions has been a major
+focus of logic and computer science since its inception. Permutations, being at the core of reversible computing, are an
+interesting class of functions, for which there are few formal systems. We develop such a system bringing in all the
+associated benefits of syntactic calculi, notably, their calculational flavor. Instead of comparing two reversible
+programs by extensional equality of the underlying bijective functions, a calculus offers more nuanced techniques that
+can enforce additional intensional constraints on the desired equality relation.
 

### 2021-09-05 Pi+/paper/finite.tex
Fix part of the typos

@@
             \xrightarrow{r} {\List[\Fin[n]]}$ such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
-    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \), where $\quotinc$ is the inclusion into the quotient, and
+    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \), where $\quotinc$ is the mapping into the quotient, and
           $\im{f}$ denotes the image of $f$ (see~\cref{def:im}).
@@
 \begin{definition}[Inversion count]
-  Given a permutation $\sigma : \Aut[\Fin[n]]$, the inversion count of $k: \Fin[n]$ is given by
+  Given a permutation $\sigma : \Aut[\Fin[n]]$, the inversion count of $i: \Fin[n]$ is given by
   \[ \invcount{i} = \#\Set{j < i | \sigma(j) > \sigma(i)}. \]
@@
   sort}. The element of the Lehmer code being currently processed is highlighted in the left column of the table below.
-Starting from a sorted list, the element at index $k$ has to be given $l[k]$ inversions. Because of the the invariant
+Starting from a sorted list, the element at index $k$ has to be given $l[k]$ inversions. Because of the invariant
 that all the elements before newly processed one are smaller than it, the proper number of inversions is created by
@@
   \begin{enumerate}
-    \item For any Lehmer code $c$, $\immersion_{n}(c)$ is a normal form with resepct to $\longcoxstar$, that is,
+    \item For any Lehmer code $c$, $\immersion_{n}(c)$ is a normal form with respect to $\longcoxstar$, that is,
           $\immersion_{n}(c)$ is in $\im{\normf}$.

### 2021-09-05 Pi+/paper/ufin.tex
Fix part of the typos

@@
 Functions between types are functors between groupoids, and type families (or functions to the universe) are indexed
-families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to \UU$
+families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to A$
 which has a lifting operation giving it the structure of a fibration. The $\term{transport}$ operation lifts paths in
@@
 in constructing a subuniverse of finite types of a specified cardinality. To do so, we will start with the subuniverse
-$\BAut[T]$. \footnote{Characterisations of univalent fibrations using the $\BAut$ construction have been studied
-  by~\citet{christensenCharacterizationUnivalentFibrations2015}.}
+$\BAut[T]$, for any type $T : \UU$. \footnote{Characterisations of univalent fibrations using the $\BAut$ construction
+have been studied by~\citet{christensenCharacterizationUnivalentFibrations2015}.}
 
@@
 
-Similar to $\BFin$, the groupoid $\UFin$ has two symmetric monoidal structures, the additive and the multiplicative one,
+Similar to $\BFin$, the groupoid $\UFin$ has two symmetric monoidal structures, the additive and the multiplicative ones,
 and the multiplicative tensor product distributes over the additive one. To construct these, we first state and prove
@@
       (X \sqcup Y) \sqcup Z  & \eqv X \sqcup (Y \sqcup Z) \\
-      X \sqcup Y             & \eqv Y \sqcup Y            \\
+      X \sqcup Y             & \eqv Y \sqcup X            \\
       X \times \bot          & \eqv \bot

### 2021-09-05 Pi+/paper/ufin.tex
Fix Pi-types

@@
 
-The semantics interprets $\PiLang$-types as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of 1-combinators
+The semantics interprets types of $\PiLang$ as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of 1-combinators
 related by a 2-combinator, their interpretations in~$\BFin$ are equal.  The groupoid $\BFin$ is strict, since the

### 2021-09-05 Pi+/paper/equivalence.tex
Fix part of the typos
Consists of

@@
 
-\noindent Normalisation consists of translating $\PiLang$ programs to $\PiHatLang$, computing a permutation, and quoting
+\noindent Normalisation involves translating $\PiLang$ programs to $\PiHatLang$, computing a permutation, and quoting
 back to $\PiPlusLang$. Note that the normalisation happens in the step from $\PiHatLang$ to $\UFin$ and back to
-$\PiHatLang$ and that normalisation also provides a decision procedure for program equivalence. Synthesis consists of
+$\PiHatLang$ and that normalisation also provides a decision procedure for program equivalence. Synthesis happens by
 quoting permutations. More general, user-guided, reasoning can be done using the sound and complete level-2
@@
 
-Recall the specification of reversible disjunction from \Cref{sec:examples}, the two Qiskit circuits for implementing it, and the
-corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The normal forms for both circuit is
-computes to the following, establishing their equivalence:
+Recall the specification of reversible disjunction from \Cref{sec:examples}, the two Qiskit circuits for implementing
+it, and the corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The normal forms for both
+circuit compute to the following, establishing their equivalence:
 

### 2021-09-05 Pi+/paper/finite.tex

@@
   Given a group $G$ and a map $f : A \to G$, such that $f$ extended to $F(A)$ respects $R$, there is a unique group
-  homomorphism $\extend{f} : F(\langle A ; R \rangle) \to^G G$ such that $\extend{f} \comp \eta_A \htpy f$.
+  homomorphism $\extend{f} : \term{Hom}(F(\langle A ; R \rangle), G)$ such that $\extend{f} \comp \eta_A \htpy f$.
 \end{proposition}

### 2021-09-05 Pi+/paper/discussion.tex

@@
 result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem. \citet{Hiver-coq}
-describes an explicit algorithm for producing normal forms. could provide an alternative to our rewriting system. Other
+describes an explicit algorithm for producing normal forms. It could provide an alternative to our rewriting system. Other
 encodings of permutations as listed vectors, matrices, inductively generated trees (Motzkin trees), Young diagrams, or
-string diagrams, proved either difficult formalise in type theory or difficult to relate directly to the primitive type
+string diagrams, proved either difficult to formalise in type theory or difficult to relate directly to the primitive type
 isomorphisms of $\PiLang$. The automatic \citet{knuthSimpleWordProblems1970} completion produced too many equations
@@
 for monoidal categories is a particular one. The use of rewriting and proof assistants to prove coherence theorems
-for higher categorical structures has a long history
-see~\cite*{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.
+for higher categorical structures has a long history, see~\cite*{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.
 
@@
 that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
-2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
+2-combinators). Thus, our results suggest a new computational interpretation of the univalence principle, which
 provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,

### 2021-09-05 Pi+/paper/ufin.tex

@@
 which has a lifting operation giving it the structure of a fibration. The $\term{transport}$ operation lifts paths in
-the base space to functions between the fibers.  Using the groupoid structure of $A$, for a path $p : x \id_{A} y$,
-$\term{transport}(p)$ and $\term{transport}(\inv{p})$ form an equivalence.
+the base space to functions between the fibers.  Using the groupoid structure of $A$, for any $x, y : A$ and a path 
+$p : x \id_{A} y$, $\term{transport}(p)$ and $\term{transport}(\inv{p})$ form an equivalence.
 % We name this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
 \[
-  \tptEqv{P} : \dfun*{x,y:A}{(x \id_{A} y) \to (P(x) \eqv P(y))}
+  \tptEqv{P} : (x \id_{A} y) \to (P(x) \eqv P(y))
 \]

### 2021-09-06 Pi/paper/informalOLD.tex
Rename Pi+ to Pi

--- 2021-07-04 Pi+/paper/informalOLD.tex
@@
 
-\paragraph*{Normalization by evaluation.} Consider the following reversible function:
+\paragraph*{Normalisation by evaluation.} Consider the following reversible function:
 $\Afun{RESET}(b,b_1,\ldots,b_n) = (b \; \underline{\vee} \; (\bigvee_{i=1}^n b_i),b_1,\ldots,b_n)$ where $\vee$ is
@@
 reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the
-introduction. Our Agda infrastructure provides tools to normalize all programs to their normal form. For \Afun{reset 2}
+introduction. Our Agda infrastructure provides tools to normalise all programs to their normal form. For \Afun{reset 2}
 we get:

### 2021-09-06 Pi/paper/introduction.tex
Rename Pi+ to Pi

\section{Introduction}
\label{sec:introduction}
% Consider two programs that implement the type isomorphism {$A + (B + C) \isoone C + (B + A)$}, written as two different
% sequences of primitive isomorphisms:

Consider two programs that implement the same permutation using different sequences of type isomorphisms:
\begin{align*}
      A + (B + C) \isoone
      (A + B) + C & \isoone
      C + (A + B) \isoone
      C + (B + A)
      \\
      A + (B + C) \isoone
      A + (C + B) \isoone
      (A + C) + B & \isoone
      (C + A) + B \isoone
      C + (A + B) \isoone
      C + (B + A)
\end{align*}

\noindent These permutations are written as sequences of primitive operations: associativity, symmetry, and
composition. Can we find necessary and sufficient equations to identify \emph{all} such equivalent sequences of type
isomorphisms?

Such sequences of type isomorphisms are pervasive in reversible boolean circuits, which are at the core of quantum
computing. Typically, they might be formalised as permutations $\{0,1\}^k \to \{0,1\}^k$ on bit strings of some length
$k$~\cite{aaronson_et_al:LIPIcs:2017:8173,1201583}. But from the perspective of programming languages, permutations and
reversible circuits can be more conveniently expressed as isomorphisms over algebraic datatypes as proposed in the
$\PiLang$ family of reversible languages~\cite*{jamesInformationEffects2012,theseus} whose syntax is inspired by the
sound and complete axiomatisation of type isomorphisms~\cite{fioreRemarksIsomorphismsTyped2002}, with respect to the
bicartesian structure, that is, coproducts and products. Since $\PiLang$ programs correspond to type isomorphisms of
finite types, our observation is that the syntax of $\PiLang$ is a presentation of the free symmetric rig groupoid on
one generator.

It is folklore that the groupoid of finite sets and permutations is the free symmetric rig groupoid on one
generator~\cite{laplaza72,kelly74,baez2000finite}. Our main result formally establishes this correspondence by giving an
equational theory for $\PiLang$ that exactly includes all the necessary equations to decide equivalence of $\PiLang$
programs. As conjectured by \citet{caretteComputingSemiringsWeak2016}, these equations correspond to the \emph{coherence
  conditions of symmetric rig groupoids}, answering the conjecture in the positive.

\paragraph{Equivalence by Example} Without going into proofs, let us try to answer the question for the original
pair of examples.

\begin{figure}
      \[
            \Tree [ {\tiny A} [ {\tiny B} {\tiny C} ] ] ~\xrightarrow{\assoclp}~
            \Tree [ [ {\tiny A} {\tiny B} ] {\tiny C} ] ~\xrightarrow{\swapp}~
            \Tree [ {\tiny C} [ {\tiny A} {\tiny B} ] ] ~\xrightarrow[\phantom{xx}\swapp]{\idc\phantom{xx}}~
            \Tree [ {\tiny C} [ {\tiny B} {\tiny A} ] ] ~
      \]

      \[
            \Tree [ {\tiny A} [ {\tiny B} {\tiny C} ] ] ~\xrightarrow[\phantom{xx}\swapp]{\idc\phantom{xx}}~
            \Tree [ {\tiny A} [ {\tiny C} {\tiny B} ] ] ~\xrightarrow{\assoclp}~
            \Tree [ [ {\tiny A} {\tiny C} ] {\tiny B} ] ~\xrightarrow[\swapp\phantom{x}]{\phantom{x}\idc}~
            \Tree [ [ {\tiny C} {\tiny A} ] {\tiny B} ] ~\xrightarrow{\assocrp}~
            \Tree [ {\tiny C} [ {\tiny A} {\tiny B} ] ] ~\xrightarrow[\phantom{xx}\swapp]{\idc\phantom{xx}}~
            \Tree [ {\tiny C} [ {\tiny B} {\tiny A} ] ] ~
      \]
      \label{fig:example-first-stage}
      \caption{Type isomorphisms as tree transformations}
\end{figure}

\begin{figure}
      \begin{subfigure}[b]{0.3\textwidth}
            \centering
            \begin{tikzpicture}
                  \def\nstrandsdf{3}
                  \pic[local bounding box=my braid,braid/.cd,
                        number of strands = \nstrandsdf,
                        width = 0.8cm,
                        height = 0.3cm,
                        border height = 0.1cm,
                        thick,
                        name prefix=braid]
                  {braid={s_2, s_2, s_2, s_1, s_2}};
                  %   \draw[thick] % draws the top/bottom bars
                  % ([xshift=-1ex]my braid.north west) --  ([xshift=1ex]my braid.north east)
                  % ([xshift=-1ex]my braid.south west) --  ([xshift=1ex]my braid.south east);
                  \node at (braid-1-s)[yshift = 2.0cm] {\tiny A};
                  \node at (braid-2-s)[yshift = 2.0cm] {\tiny B};
                  \node at (braid-3-s)[yshift = 2.0cm] {\tiny C};

                  % labels the bottom bar
                  \node at (braid-1-e)[yshift = -2.0cm] {\tiny A};
                  \node at (braid-2-e)[yshift = -2.0cm] {\tiny B};
                  \node at (braid-3-e)[yshift = -2.0cm] {\tiny C};
            \end{tikzpicture}
            \caption{}
            \label{fig:cone-braid}
      \end{subfigure}
      \begin{subfigure}[b]{0.3\textwidth}
            \centering
            \begin{tikzpicture}
                  \def\nstrandsdf{3}
                  \pic[local bounding box=my braid,braid/.cd,
                        number of strands = \nstrandsdf,
                        width = 0.8cm,
                        height = 0.3cm,
                        border height = 0.1cm,
                        thick,
                        name prefix=braid]
                  {braid={1, s_2,, s_1, s_2, 1}};
                  %   \draw[thick] % draws the top/bottom bars
                  % ([xshift=-1ex]my braid.north west) --  ([xshift=1ex]my braid.north east)
                  % ([xshift=-1ex]my braid.south west) --  ([xshift=1ex]my braid.south east);
                  \node at (braid-1-s)[yshift = 2.0cm] {\tiny A};
                  \node at (braid-2-s)[yshift = 2.0cm] {\tiny B};
                  \node at (braid-3-s)[yshift = 2.0cm] {\tiny C};

                  % labels the bottom bar
                  \node at (braid-1-e)[yshift = -2.0cm] {\tiny A};
                  \node at (braid-2-e)[yshift = -2.0cm] {\tiny B};
                  \node at (braid-3-e)[yshift = -2.0cm] {\tiny C};
            \end{tikzpicture}
            \caption{}
            \label{fig:ctwo-braid}
      \end{subfigure}
      \begin{subfigure}[b]{0.3\textwidth}
            \centering
            \begin{tikzpicture}
                  \def\nstrandsdf{3}
                  \pic[local bounding box=my braid,braid/.cd,
                        number of strands = \nstrandsdf,
                        width = 0.8cm,
                        height = 0.3cm,
                        border height = 0.1cm,
                        thick,
                        name prefix=braid]
                  {braid={1, s_1,, s_2, s_1, 1}};
                  %   \draw[thick] % draws the top/bottom bars
                  % ([xshift=-1ex]my braid.north west) --  ([xshift=1ex]my braid.north east)
                  % ([xshift=-1ex]my braid.south west) --  ([xshift=1ex]my braid.south east);
                  \node at (braid-1-s)[yshift = 2.0cm] {\tiny A};
                  \node at (braid-2-s)[yshift = 2.0cm] {\tiny B};
                  \node at (braid-3-s)[yshift = 2.0cm] {\tiny C};

                  % labels the bottom bar
                  \node at (braid-1-e)[yshift = -2.0cm] {\tiny A};
                  \node at (braid-2-e)[yshift = -2.0cm] {\tiny B};
                  \node at (braid-3-e)[yshift = -2.0cm] {\tiny C};
            \end{tikzpicture}
            \caption{}
            \label{fig:norm-braid}
      \end{subfigure}

      \label{fig:example-braid}
      \caption{Type isomorphisms as braid diagrams}
\end{figure}

The first step is to realise that the use of associativity is uninteresting, and the only steps with interesting
computational content are the swaps. The swaps can be either big or small -- they can happen between leaf nodes, or
between bigger subtrees, respectively.

First, we normalise the types to a list-like representation $A + (B + (C + \zerot))$, where $\zerot$ is the empty type.
In this representation, each type is identified with its index, $A$ has index 0, $B$ has index 1, and $C$ has index 2.
Then, we need to compile each primitive isomorphism to a list of adjacent transpositions, and then compose by appending
the lists. A small swap is trivially implemented as a single adjacent transposition. To compile a big swap into adjacent
transpositions, we need to traverse the subtrees in-order and recursively swap elements from the left by transposing
them across the ones in the middle -- this generates a large number of adjacent transpositions. For the two programs, we
get the following two lists: $[1,0,1,1,1]$ (\cref{fig:cone-braid}) and $[1,0,1]$ (\cref{fig:ctwo-braid}), where the
number $k$ encodes a transposition of elements at indices $k$ and $k+1$.

Swapping is symmetric, that is, swapping two elements and immediately swapping them back produces the same permutation.
The first list contains such redundant operations, so to eliminate these, we have to normalise the lists, which we will
do by setting up reduction relations and an appropriate rewriting system. This system will normalise both lists to
$[0,1,0]$ (\cref{fig:norm-braid}), which is lexicographically the smallest list that corresponds to this permutation.
Since both programs have the same normal forms, they're equivalent!

The crux of designing an equational theory for $\PiLang$ relies on the choice of relations we use to design this
rewriting system. First, there should be enough equations relating the programs to their normalised forms, and second,
there should be enough equations corresponding to the reduction rules. We will show that these correspond to the
coherence conditions for symmetric rig groupoids.

From the normalised list $[0,1,0]$, we can construct a permutation on a list of 3 elements $[a,b,c]$ as follows. We think
of it as insertion-sorting the elements of the list. Starting from the list $[a, b, c]$, we first insert $b$ at the
right place -- by applying transposition $0$, we get the list $[b, a, c]$. Then, element $c$ is inserted in the right
place by applying transpositions $1$ and $0$ -- as a result, we get the desired permutation $[c, b, a]$. Notice that we
could specify a more compact way of describing this process, since the key information was only how many shifts we
needed to apply to an element. We could describe this procedure using a code $(0, 1, 2)$, which says how many inversions
to apply to elements $a$, $b$, and $c$, respectively.

Using this algorithm, we can turn a type isomorphism into a permutation of finite sets, relating the operational
semantics of $\PiLang$ as permutations of finite sets of bits, to our denotational semantics. This allows us to
establish full abstraction and adequacy.

\paragraph{Outline and Contributions} Our main result is a proof of the soundness and completeness of $\PiLang$ with
respect to its semantics in the weak symmetric rig groupoid of finite sets and permutations. We state our result as a
Curry-Howard-Lambek correspondence for Reversible Logic, Reversible Programming Languages, and Symmetric Rig Groupoids,
using which we can build a toolbox of technical devices for reasoning about reversible circuits.

\begin{itemize}[leftmargin=*]
      \item We start in~\Cref{sec:qiskit} by presenting a few reversible circuits in the popular IBM Qiskit framework to
            serve as running examples throughout the paper.
      \item In~\Cref{sec:pi}, we introduce the two-level language
            $\PiLang$~\cite{jamesInformationEffects2012,caretteComputingSemiringsWeak2016} and illustrate how to write
            reversible circuits and their equivalences using 1-combinators and 2-combinators respectively. We give a
            semantic account of the language by translating each level-1 program to a bijection between finite sets, and
            verifying that programs identified by level-2 constructs denote the same bijection.
      \item ~\Cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in
            Homotopy Type Theory. We define and characterise the notion of a univalent subuniverse, and construct
            $\UFin$ as a univalent subuniverse which classifies all finite types. We establish that paths in $\UFin$ are
            families of loops on finite sets of specified cardinality, given by $\Aut[\Fin[n]]$, which produces the
            permutation group on $\Fin[n]$.
      \item In~\Cref{sec:finite}, we proceed to give a presentation of the permutation group, as the symmetric group
            $\Sn$ with generators and relations, and solve its word problem. In particular, we present $\Sn$ as a
            Coxeter group, build a rewriting system based on Coxeter relations, and prove confluence and termination.
            Using our rewriting system, we establish that normal forms for words in $\Sn$ are equivalent to Lehmer
            codes~\cite{lehmerTeachingCombinatorialTricks1960}, which are a convenient and compact representation of
            permutations. Finally, we show that there is an equivalence between Lehmer codes and permutations
            $\Aut[\Fin[n]]$ given by the Lehmer encode-decode algorithm.
      \item In~\Cref{sec:equivalence}, we show how to interpret the language $\PiLang$ into the groupoid $\UFin$, in
            stages. First we define a subset $\PiPlusLang$ of the language which only includes the additive monoidal
            structure, and show how to translate $\PiLang$ programs to $\PiPlusLang$ programs. Then, we further define a
            normalised form for this language called $\PiHatLang$, which has normalised 1-combinators and 2-combinators
            corresponding to adjacent transpositions. We show that $\PiPlusLang$ can be translated to $\PiHatLang$ and
            back. Then, we show how to interpret this language $\PiHatLang$ into $\UFin$ -- the 1-combinators are
            translated into permutations via words in $\Sn$, and 2-combinators are interpreted as 2-paths in $\UFin$. We
            further show how to quote back a permutation in $\UFin$ into a 1-combinator using the normal forms for words
            in $\Sn$. The main result of this section is a symmetric monoidal equivalence between the syntactic
            groupoids of $\PiPlusLang$ and $\PiHatLang$, with $\UFin$. Finally, we also establish full abstraction and
            adequacy of this model with respect to the operational semantics.
      \item In~\Cref{sec:applications}, we show applications of our results to reversible circuits, using our
            formalisation. Our results are stated using HoTT~\cite{univalentfoundationsprogramHomotopyTypeTheory2013},
            and formalised using the HoTT-Agda library (around 7,500 lines of code). Using the formalisation, we are
            able to extract procedures for:
            \begin{enumerate*}
                  \item the synthesis of a reversible circuit from a permutation on a finite set,
                  \item the verification that a reversible circuit realises a given permutation on finite sets,
                  \item a normalisation-by-evaluation (NbE) procedure that reduces reversible circuits to canonical normal forms,
                  \item a sound and complete calculus for reasoning about reversible circuits and their equivalences, and
                  \item the transfer of theorems about permutations and reversible circuits from one representation to the other.
            \end{enumerate*}
\end{itemize}

\noindent
The proofs of some of our lemmas and propositions and theorems, as well as additional material, are given in the
supplementary appendices, and we refer to them in the text. Our accompanying Agda code contains the formalisation of the
full syntax, and most of our proofs.

% Throughout the paper, we defer longer proofs to a technical appendix and refer to the accompanying Agda code for some of
% the elided (routine) definitions.

% Not only can we compare the equivalence of two programs, but using the normalisation procedure, we can synthesise a
% normalised program that uses a smaller set of primitive isomorphisms.

% Sn and normal forms gives us the equational theory.

% Lehmer and Aut(Fin n) give us adequacy and full abstraction?

% Not only can we compare the equivalence of two programs, but also synthesise programs from a permutation.

% Thus, we can get a handle on the normal forms of lists, by putting them in an equivalence with such codes.

% In technical terms, the main result is that we define \emph{eval / quote} maps that give a symmetric monoidal
% equivalence between a programming language $\Pi$ and a space of denotations $\UFin$. The language $\Pi$ is specialised
% for expressing reversible circuits using type isomorphisms and constitutes an algebraic presentation of the free
% symmetric rig groupoid. The space $\UFin$ is the weak symmetric rig groupoid of finite sets and bijections. The
% \emph{eval / quote} maps are composed from smaller maps that we trace through with a small representative example before
% giving a more technical outline.

% The above presentation rests on two key technical contributions:

% \begin{itemize}
%       \item First, the definition of the weak symmetric rig groupoid $\UFin$ of finite sets and bijections in
%             Sec.~\ref{sec:ufin} builds on a notion of \emph{univalent subuniverse} expressed in the HoTT metalanguage, and
%             whose definition and properties are our first technical contribution.
%       \item The definition of $\UFin$ as a univalent subuniverse implies that, for each finite type $T$ of $n$-elements, the
%             structure of the symmetric group $S_n$ is reflected in the elements of the identity type (i.e., the 1-paths).
%             This observation reduces the bulk of the definition of $\mathit{quote}$ to the problem of describing $S_n$
%             syntactically using 1-paths. We do this in a novel way in Sec.~\ref{sec:finite} by first observing that $S_n$ is
%             a \emph{Coxeter group} and hence admits a formal description which we formulate using the explicit collection of
%             reduction rules alluded to in the example above. These rules organize the words of the permutation group as a
%             confluent, strongly-normalising rewriting system whose normal forms are the \emph{Lehmer code} description of
%             permutations. We note that there have been already some attempts to solve this problem. In
%             particular~\citet{LAFONT2003257} follows a similar approach, writing the Coxeter equations as a rewriting
%             system, but chooses to make the problematic commutation rule implicit, which, being undirected, would cause our
%             system to be non-terminating. There is also a formulation by~\citet{Hiver-coq}, which, instead of using a
%             rewriting system, describes an explicit algorithm for producing normal forms. That approach could provide an
%             alternative to our rewriting system.
% \end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% ***
%% Various older fragments

% Boolean circuits are fundamental to the study of computation -- they encode
% boolean functions $\{0,1\}^{n}\to\{0,1\}$ which are logic gates used in models
% of computation. Motivated by connections between physics and computation,
% reversible and quantum boolean circuits are the fundamental building blocks used
% in reversible computing and quantum computing. Reversible boolean circuits, by
% themselves are an interesting object of study, since quantum circuits also are
% also built from a reversible core.

% Reversible circuits encode reversible logic gates which are bijective functions
% $\{0,1\}^{n} \to \{0,1\}^{n}$, on a fixed number of bits $n$. Bijective
% functions between finite sets of the same cardinality are simply permutations,
% and it is known that reversible logic gates can be studied using permutation
% groups.

% Motivated by the idea of logical reversibility,
% \citet{jamesInformationEffects2012} proposed the $\PiLang$ family of first-order
% reversible languages, which is a high-level typed programming language for
% constructing reversible boolean circuits and reasoning about their
% equivalences. The language supports $\zerot$ and $\onet$ types corresponding to
% the bits 0 and 1, and allows type constructors $+$ and $\times$ for composing
% bits in sequential or parallel order.  Corresponding to bijective functions
% between bits, the language provides 1-combinators for building reversible
% functions, and 2-combinators for their equivalences. The syntax for the language
% is motivated by type isomorphisms of finite types with sums and products, for
% which a complete axiomatisation is known.

% \jk{Mediating between the syntactic and the semantic side is done in a novel way
%   by introducing an explicit collection of reduction rules, that make the words
%   in the permutation group into a confluent, strongly-terminating rewriting
%   system, which allows us to extract the normalisation function.  There has been
%   already some attempts to solve this problem. In
%   particular~\cite{LAFONT2003257} give a similar treatment, writing the
%   \emph{Coxeter equations} as a rewriting system, but chooses to make the
%   problematic commutation rule implicit, which, by being un-directed, would
%   cause the system to be non-terminating. There is also a formulation
%   by~\cite{Hiver-coq}, which, instead of using a rewriting system, describes an
%   explicit procedure for inserting a new transposition into a sequence of
%   already normalised ones.  Ultimately, in all cases, the normal form ends up
%   being closely related to \emph{Lehmer code} description of permutations.  To
%   show its equivalence to the type of bijections on a finite set, we describe a
%   (slightly modified version of) unpublished proof due
%   to~\cite{Molzer-cubical}.}

% First, although intuitive in some sense, the folklore result we prove is
% actually non-trivial to fully formalise. To get a taste of the involved
% complexity, consider the obvious fact that, semantically, there is only one
% bijection from the empty set to itself. However, in a syntactic presentation of
% permutations based on type isomorphisms, there would be an infinite number of
% programs from the empty type to itself that go through arbitrary complex
% subtypes, e.g., letting $\isoone$ be the type constructor for type isomorphisms
% we can have a sequence of syntactic equivalences
% $\zerot \isoone \zerot \times A \isoone \zerot \times (A + \zerot) \isoone
% (\zerot \times A) + (\zerot \times \zerot) \isoone (\zerot \times \zerot) +
% (\zerot \times A) \isoone \zerot + (\zerot \times A) \isoone \zerot + \zerot
% \isoone \zerot$ and a powerful proof technique would be needed to establish that
% all programs of type $\zerot \isoone \zerot$ are reducible to the identity.

% The base case for proving that the semantic groupoid is
% equivalent to the syntactic one requires enough coherence conditions to identify
% this infinite collection of programs from the empty type to itself with the
% unique bijection on the empty set.

%% The most interesting thing: a proof of the base case, which we have been
%% fighting with for quite long. (it's in Equiv2Hat.agda)

% c₊⟷₂id⟷₁ : (c : O ⟷₁ O) → c ⟷₂ id⟷₁
% c₊⟷₂id⟷₁ c =
%         let x = quote^₂ (eval^₂-O c)
%         in  ((idr◎r ■ idl◎r) ■ !⟷₂ (quote-eval^₁ c)) ■ x

% \url{https://mathoverflow.net/questions/350289/mac-lanes-proof-of-coherence-for-symmetric-monoidal-categories}

% And this is the same result that is implicitly assumed in the proof of the
% universal property (in Abramsky)

% Our contribution is a proof of this folklore result, while working in
% constructive type theory, by giving an algebraic presentation of the free
% symmetric rig groupoid (Pi), the free symmetric monoidal groupoid (PiPlus), and
% the weak groupoid of sets and bijections, and establishing a correspondence
% between them.

% "The objects are in their right place anyway, so we might as well just pretend
% that they are equal" -- this is what happens when we strictify and go to PiHat

% (The tricks in that answer are the tricks we're doing in UFin to get the
% monoidal structure from the coproduct)

% The fact that the free smc can be presented as an algebraic 2-theory is known,
% the fact that the free smc is finite sets and bijections is generally asserted
% to be true by MacLane's coherence, but now proving the two presentations are
% equivalent is new.

% \emph{the groupoid of finite sets and bijections
%   equipped with the canonical structure given by the disjoint union and the
%   cartesian product of finite sets}. This groupoid, characterizing permutations
% among finite sets closed under sums and products, has connections to quantum
% mechanics, combinatorial species and linear
% logic~\cite{brent,catalgqm,catalgqm2}.

% Groupoids are natural algebraic gadgets for studying the semantics of a
% typed-reversible language, since they horizontally categorify groups. The
% observation made in~\cite{caretteComputingSemiringsWeak2016} that $\PiLang$ can
% be presented as a weak rig groupoid, that is, a groupoid with interacting
% additive and multiplicative symmetric monoidal structures. In this paper, we
% strengthen this correspondence by relating it to the groupoid of finite sets and
% bijections, which is the free symmetric monoidal groupoid on one generator.

% The relatively recent discovery of deep connections between physics and
% computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275,
%   Hey:1999:FCE:304763,fredkin1982conservative, springerlink:10.1007/BF02650179}
% has renewed interest in connections between categorial accounts of symmetry and
% syntactic accounts of reversible programs. On the categorical (semantic) side,
% the baseline account of symmetry is \emph{the groupoid of finite sets and
%   bijections equipped with the canonical structure given by the disjoint union
%   and the cartesian product of finite sets}. This groupoid, characterizing
% permutations among finite sets closed under sums and products, has connections
% to quantum mechanics, combinatorial species and linear
% logic~\cite{brent,catalgqm,catalgqm2}. On the syntactic side, we have \emph{the
%   free commutative rig groupoid} which exploits the computational content of
% isomorphisms to induce a programming language for expressing and reasoning about
% reversible programs and their
% equivalences~\cite{James:2012:IE:2103656.2103667,Carette2016}.

% It is a folklore result that these two groupoids, the semantic one and the
% syntactic one, are equivalent~\cite{baez2000finite,math/9802029}. Although
% intuitive in some sense, this folklore result is actually non-trivial to fully
% formalise. To get a taste of the involved complexity, consider the obvious fact
% that, semantically, there is only one bijection from the empty set to
% itself. However, in the syntactic groupoid, there are an infinite number of
% programs from the empty type to itself that go through arbitrary complex
% subtypes, e.g., letting $\isoone$ be the type constructor for type isomorphisms
% we can have a sequence of syntactic equivalences
% $\zerot \isoone \zerot \times A \isoone \zerot \times (A + \zerot) \isoone
% (\zerot \times A) + (\zerot \times \zerot) \isoone (\zerot \times \zerot) +
% (\zerot \times A) \isoone \zerot + (\zerot \times A) \isoone \zerot + \zerot
% \isoone \zerot$. The base case for proving that the semantic groupoid is
% equivalent to the syntactic one requires enough coherence conditions to identify
% this infinite collection of programs from the empty type to itself with the
% unique bijection on the empty set.

% Our main technical result is a proof, in the homotopy type theory (HoTT)
% metatheory, of this equivalence between the semantic and syntactic
% groupoids. The proof itself is novel XXX. Most in Agda but not all.
% The steps in formalising the main
% theorem are as follows:
% \begin{itemize}
% \item Define the syntactic groupoid by giving inductive definitions of the types, isomorphisms between types, and
%   coherence conditions for these isomorphisms.
% \item Define the semantic groupoid as a univalent subuniverse in HoTT (about 300 lines of code).
% \item Mediate between the syntactic presentation of the groupoid and the semantic one using ideas inspired by computational
%   group theory, sorting algorithms, and term rewriting systems (about 4,500 lines of code).
% \item Compose the maps from syntax to semantics and back and establish that they form an equivalence (about 2,500 lines
%   of code).
% \end{itemize}

% The technical result implies several immediate applications to reversible
% circuits: (i) a reversible circuit expressed in $\PiPlusLang$ can be
% automatically generated from a permutation in $\UFin$ \emph{and the generation
%   comes equipped with a proof of correctness establishing its equivalence
%   between the circuit and the original permutation}; (ii) circuits in $\PiLang$
% or $\PiPlusLang$ can be reduced to a circuit normal form using a normalisation
% by evaluation process that evaluates them to a permutation in $\UFin$ and quotes
% it back; (iii) equivalence of circuits in either $\PiLang$ or $\PiPlusLang$ can
% be decided by reducing them to normal forms; (i) a circuit can be verified
% against a given permutation by evaluating it; and (v) the induced circuit
% equivalences in $\PiLang$ and $\PiPlusLang$ form a sound and complete calculus
% for reasoning about and optimizing circuits. (See Sec.~\ref{sec:informal} for
% more details.)

% Natural numbers under addition form the free commutative monoid on one
% generator. Categorifying this, we get the free symmetric monoidal groupoid on
% generators. This is the groupoid of finite sets and bijections.

% More generally, we can construct the free symmetric monoidal groupoid on a
% groupoid, by taking the action of the groupoid of finite sets and bijections. We
% can construct this in HoTT by using the classifying space of the symmetric
% group.

% The free symmetric monoidal category on a category is a 2-monad on
% Cat~\cite{blackwellTwodimensionalMonadTheory1989,abramskyAbstractScalarsLoops2005,leinsterHigherOperadsHigher2004}.
% Other applications are Fock Spaces, Generalised Species, Abstract Syntax.

% We establish a Curry-Howard-Lambek correspondence for Pi, with 0, 1,
% +. Categorically, the syntactic groupoid of Pi is the free symmetric monoidal
% groupoid on one generator. The groupoid of finite sets and bijections, with
% coproducts, is equivalent to the free symmetric monoidal groupoid on one
% generator, making Pi fully adequate with respect to this semantics.

% The logic part of this is in superstructural reversible logic, which is the free
% commutative semirig, since there are no equations.

% perhaps fig. on p.30 of 4 different ways of going from a + (b + c) to c + (b + a) ??

% \paragraph*{Our Technical Results.} The main result of the paper is a proof, formalised in the HoTT-Agda library, that
% the category of finite sets and bijections is the free rig groupoid; its structure is illustrated diagrammatically below:

% \[\begin{tikzcd}
%     \PiLang && \PiPlusLang && \PiHatLang && \UFin
%     \arrow["\evalt", from=1-1, to=1-3]
%     \arrow["\evalp", curve={height=-24pt}, from=1-3, to=1-5]
%     \arrow["\evalh", curve={height=-24pt}, from=1-5, to=1-7]
%     \arrow["\quotep", curve={height=-24pt}, from=1-5, to=1-3]
%     \arrow["\quoteh", curve={height=-24pt}, from=1-7, to=1-5]
%   \end{tikzcd}\]

% \noindent The nodes $\PiLang$, $\PiPlusLang$, and $\PiHatLang$ in the diagram
% each represent a syntactic weak 2-category (groupoid actually as all morphisms
% are isomorphisms) where the 0-cells represent types, the 1-cells represent
% reversible circuits, and the 2-cells represent circuit equivalences. In
% $\PiLang$, the circuits represent arbitrary permutations among finite sets
% with products and coproducts; in $\PiPlusLang$, the circuits represent
% arbitrary permutations among finite sets with only coproducts; and in
% $\PiHatLang$, the permutations are expressed using adjacent transpositions.

% The syntactic groupoid $\PiPlusLang$ is the free rig groupoid; $\UFin$ is the
% groupoid of finite sets and bijections represented as the \emph{univalent
% subuniverse of all finite types}. The $\evalp/\quotep$ and $\evalh/\quoteh$
% arrows establish a \emph{symmetric monoidal equivalence} between these two
% groupoids. To get a taste of the involved complexity of this result, consider
% the obvious fact that, semantically, i.e., in $\UFin$, there is only one
% bijection from the empty set to itself. However, in the free groupoid, there
% are an infinite number of isomorphisms from the empty type to itself that go
% through arbitrary complex subtypes, e.g., letting $\isoone$ be the type
% constructor for type isomorphisms we can have a sequence of syntactic
% equivalences
% $\zerot \isoone \zerot \times A \isoone \zerot \times (A + \zerot) \isoone
% (\zerot \times A) + (\zerot \times \zerot) \isoone (\zerot \times \zerot) +
% (\zerot \times A) \isoone \zerot + (\zerot \times A) \isoone \zerot + \zerot
% \isoone \zerot$, and all such isomorphisms must be identified using
% appropriate coherence laws.

% Below we reduce $\mathsf{swap} : 2 + 2 \leftrightarrow 2 + 2$ to a sequence of adjacent swaps. This is an example of
% a translation from $\PiPlusLang$ to $\PiHatLang$.

% \begin{align*}
%   \begin{tikzpicture}[scale=0.4,every node/.style={scale=0.4}]
%     \begin{knot}[clip width=3]
%       \filldraw (0,4) circle (2pt) node[above] {0};
%       \filldraw (1,4) circle (2pt) node[above] {1};
%       \filldraw (2,4) circle (2pt) node[above] {2};
%       \filldraw (3,4) circle (2pt) node[above] {3};
%       \filldraw (0,0) circle (2pt) node[below] {2};
%       \filldraw (1,0) circle (2pt) node[below] {3};
%       \filldraw (2,0) circle (2pt) node[below] {0};
%       \filldraw (3,0) circle (2pt) node[below] {1};
%       \strand (0,4) .. controls (0.5,1.5) and (1.5,2.5) .. (2,0);
%       \strand (1,4) .. controls (1.5,1.5) and (2.5,2.5) .. (3,0);
%       \strand (2,4) .. controls (1.5,1.5) and (1.5,2.5) .. (0,0);
%       \strand (3,4) .. controls (2.5,1.5) and (2.5,2.5) .. (1,0);
%     \end{knot}
%   \end{tikzpicture}
% \quad=\quad
%   \begin{tikzpicture}[scale=0.4,every node/.style={scale=0.4}]
%     \begin{knot}[clip width=3]
%       \filldraw (0,4) circle (2pt) node[above] {0};
%       \filldraw (1,4) circle (2pt) node[above] {1};
%       \filldraw (2,4) circle (2pt) node[above] {2};
%       \filldraw (3,4) circle (2pt) node[above] {3};
%       \filldraw (0,0) circle (2pt) node[below] {0};
%       \filldraw (1,0) circle (2pt) node[below] {2};
%       \filldraw (2,0) circle (2pt) node[below] {1};
%       \filldraw (3,0) circle (2pt) node[below] {3};
%       \strand (0,4) to (0,0);
%       \strand (1,4) .. controls (0.5,2) and (2.5,2) .. (2,0);
%       \strand (2,4) .. controls (2.5,2) and (0.5,2) .. (1,0);
%       \strand (3,4) to (3,0);
%     \end{knot}
%   \end{tikzpicture}
%   &&
%     \begin{tikzpicture}[scale=0.4,every node/.style={scale=0.4}]
%       \begin{knot}[clip width=3]
%         \filldraw (0,4) circle (2pt) node[above] {0};
%         \filldraw (1,4) circle (2pt) node[above] {2};
%         \filldraw (2,4) circle (2pt) node[above] {1};
%         \filldraw (3,4) circle (2pt) node[above] {3};
%         \filldraw (0,0) circle (2pt) node[below] {2};
%         \filldraw (1,0) circle (2pt) node[below] {0};
%         \filldraw (2,0) circle (2pt) node[below] {1};
%         \filldraw (3,0) circle (2pt) node[below] {3};
%         \strand (0,4) .. controls (-0.5,2) and (1.5,2) .. (1,0);
%         \strand (1,4) .. controls (1.5,2) and (-0.5,2) .. (0,0);
%         \strand (2,4) to (2,0);
%         \strand (3,4) to (3,0);
%       \end{knot}
%     \end{tikzpicture}
%   &&
%   \begin{tikzpicture}[scale=0.4,every node/.style={scale=0.4}]
%     \begin{knot}[clip width=3]
%       \filldraw (0,4) circle (2pt) node[above] {2};
%       \filldraw (1,4) circle (2pt) node[above] {0};
%       \filldraw (2,4) circle (2pt) node[above] {1};
%       \filldraw (3,4) circle (2pt) node[above] {3};
%       \filldraw (0,0) circle (2pt) node[below] {2};
%       \filldraw (1,0) circle (2pt) node[below] {0};
%       \filldraw (2,0) circle (2pt) node[below] {3};
%       \filldraw (3,0) circle (2pt) node[below] {1};
%       \strand (0,4) to (0,0);
%       \strand (1,4) to (1,0);
%       \strand (2,4) .. controls (1.5,2) and (3.5,2) .. (3,0);
%       \strand (3,4) .. controls (3.5,2) and (1.5,2) .. (2,0);
%     \end{knot}
%   \end{tikzpicture}
%   &&
%     \begin{tikzpicture}[scale=0.4,every node/.style={scale=0.4}]
%       \begin{knot}[clip width=3]
%         \filldraw (0,4) circle (2pt) node[above] {2};
%         \filldraw (1,4) circle (2pt) node[above] {0};
%         \filldraw (2,4) circle (2pt) node[above] {3};
%         \filldraw (3,4) circle (2pt) node[above] {1};
%         \filldraw (0,0) circle (2pt) node[below] {2};
%         \filldraw (1,0) circle (2pt) node[below] {3};
%         \filldraw (2,0) circle (2pt) node[below] {0};
%         \filldraw (3,0) circle (2pt) node[below] {1};
%         \strand (0,4) to (0,0);
%         \strand (1,4) .. controls (0.5,2) and (2.5,2) .. (2,0);
%         \strand (2,4) .. controls (2.5,2) and (0.5,2) .. (1,0);
%         \strand (3,4) to (3,0);
%       \end{knot}
%     \end{tikzpicture}
% \end{align*}

% \note{Show codes; normalise; explain the dense paragraph above using that example}

% \note{perhaps a note about the result being folklore; assumed in many places;
% but no proof; and certainly none
% formalised in a proof assistant. In some sense, what we have done is to take
% MacLane's coherence theorem and
% formalised it in HoTT, and given presentations for it using Pi's syntax. Also
% relevant is that the equivalence
% result hides implicit isomorphisms that have computational relevance. For
% example, transporting properties across
% equivalences of finite types can be done via executing permutations, something
% which has a clear computational cost
% and which itself depends on the choice of representations of the
% permutations.}

% Our results are formalised in the proof assistant Agda using the HoTT-Agda library.

% \begin{center}
%   \includegraphics[scale=0.3]{outline.png}
% \end{center}

% \note{Redo this figure or delete ??}

%% \note{Spiel about reversible computing and logical reversibility ??}

%% \note{Spiel about using groupoids for denotational semantics ??}

%% \note{Could we include a short paragraph about entropy and bits and logical
%% reversibility ??}

%% \note{Novel interpretation of the univalence axiom, operational and
%% denotational semantics and adequacy.}

% * STLC(first) is the type theory for CCC
%   HoTT is type theory for weak infty groupoids(first)
%   Correspondence between TT and categories are fruitful:
%   answer hard questions about syntax without worrying about presentation
%   normalisation of lcal with strong sums: if you have specific syntax; no obvious induction on syntax
%   Lawvere thesis

% * Rig groupoids are interesting! WHY????
%   Add monoidal structure to weak groupoids. Easy. Lists
%   How symmetry is going to act on higher dimension
%   Higher-dim all symmetric; topology higher things abelian groups

%   * What is the corresponding type theory ?  We give it for the free symmetric
%   groupoid; for other groupoids build on it and add more

% The technical device to achieve this normalisation is as follows. First, we
% observe that 1-paths in $\UFin$ are permutations on finite sets with a fixed
% cardinality $n$, given by $\Aut[\Fin[n]]$, which produce the permutation group
% on $\Fin[n]$, or the symmetric group $\Sn$. By giving a presentation of $\Sn$
% using generators and relations, we build a rewriting system using the Coxeter
% relations~\cite{XXX} on the set of words $\List[\Fin[n]]$, and show that it is
% (locally) confluent and strongly normalising. We then establish that the
% symmetric group $\Sn$ is the set-quotient of $\List[\Fin[n]]$ by the Coxeter
% relations, and show that it produces a group presentation, as a quotient of
% the free group. Using this strongly normalising rewriting system, we establish
% that normal forms for words in $\Sn$ are Lehmer
% codes~, which are a convenient and
% compact representation of permutations. Finally, we show that there is an
% equivalence between Lehmer codes and permutations $\Aut[\Fin[n]]$ given by the
% Lehmer encode-decode algorithm.

%% ***
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% fill-column: 120
%%% End:

### 2021-09-06 Pi/paper/univalentOLD.tex
Rename Pi+ to Pi

--- 2021-05-26 Pi+/paper/univalent.tex
@@
 \vc{I'm explaining everything starting from the categorical language. We do care more about the groupoid structure of
-  types, and how to encode groupoids using types, since that's what we use in~\cref{sec:finite}.}
+  types, and how to encode groupoids using types, since that's what we use in~\Cref{sec:finite}.}
 
@@
   subuniverse of types that are merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the
-  classifying space of $\Aut[T]$, as justified by~\cref{lem:loop-deloop}. We write $T_{o} \defeq (T, \trunc{\refl_{T}})$
+  classifying space of $\Aut[T]$, as justified by~\Cref{lem:loop-deloop}. We write $T_{o} \defeq (T, \trunc{\refl_{T}})$
   for the image of the inclusion of $T$.
@@
 \begin{proposition}
-  $\isFin$ is a proposition.
+  For any type $X$, $\isFin[X]$ is a proposition.
 \end{proposition}
@@
   \]
-  We write $F_{n} \defeq (n, \Fin[n], \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$. We use this as the
+  We write $F_{n} \defeq (\Fin[n], n, \trunc{\refl})$, for the image of the inclusion of $\Fin[n]$. We use this as the
   denotational semantics to interpret our reversible programming language $\PiPlusLang$.
@@
 
+\begin{proposition}
+  For any $T : \UU$, $\BAut[T]$ is 0-connected.
+\end{proposition}
+
 \begin{corollary}
-  $\UFin$ is a 1-type, that is, a 1-groupoid, or has h-level 3.
+  $\UFin$ is a pointed, connected, 1-type, that is, a 1-groupoid, or has h-level 3.
 \end{corollary}

### 2021-09-06 Pi/paper/reversible1.tex

--- 2021-07-09 Pi+/paper/reversible1.tex
@@
 
-For combinators, we show explcitly how values reduce along each combinator, similar to a big-step operational
+For combinators, we show explicitly how values reduce along each combinator, similar to a big-step operational
 semantics~\cite{chenComputationalInterpretationCompact2021,theseus}.

### 2021-09-11 Pi/paper/introduction.tex
Fix "one generator" for free symmetric rig
Fixes #128

@@
 bicartesian structure, that is, coproducts and products. Since $\PiLang$ programs correspond to type isomorphisms of
-finite types, our observation is that the syntax of $\PiLang$ is a presentation of the free symmetric rig groupoid on
-one generator.
+finite types, our observation is that the syntax of $\PiLang$ is a presentation of the free symmetric rig groupoid (on
+zero generators).
 
-It is folklore that the groupoid of finite sets and permutations is the free symmetric rig groupoid on one
-generator~\cite{laplaza72,kelly74,baez2000finite}. Our main result formally establishes this correspondence by giving an
+It is folklore that the groupoid of finite sets and permutations is the free symmetric rig groupoid on zero
+generators~\cite{laplaza72,kelly74,baez2000finite}. Our main result formally establishes this correspondence by giving an
 equational theory for $\PiLang$ that exactly includes all the necessary equations to decide equivalence of $\PiLang$

### 2021-09-11 Pi/paper/reversible1.tex
Fix "one generator" for free symmetric rig
Fixes #128

@@
 which is just a natural number. The algebraic structure of natural numbers as the free commutative semiring (or,
-commutative rig) on one generator, with $(0,+)$ for addition, and $(1,\times)$ for multiplication then provides
-sequential, vertical, and horizontal circuit composition operators.
+commutative rig), with $(0,+)$ for addition, and $(1,\times)$ for multiplication then provides sequential, vertical, and
+horizontal circuit composition operators.
 

### 2021-09-11 Pi/paper/equivalence.tex
Fix typo on l1025
Fixes #137

--- 2021-09-05 Pi+/paper/equivalence.tex
@@
 are interpreted as 0-cells in $\UFin$, that is, a natural number $n$ is mapped to $\Fin[n]$. The 1-combinators in
-$\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops $\Aut[\Fin[n]]$, in each connected component. In
-$\PiHatLang$, the 1-combinators are generated by adjacent transpositions, so these can be mapped to words in $\Sn$ and
-then to automorphisms using~\Cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to 2-paths between loops
-in $\UFin$.
+$\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops in each connected component, equivalent to
+$\Aut[\Fin[n]]$. In $\PiHatLang$, the 1-combinators are generated by adjacent transpositions, so these can be mapped to
+words in $\Sn$ and then to automorphisms using~\Cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to
+2-paths between loops in $\UFin$.
 
@@
 it, and the corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The normal forms for both
-circuit compute to the following, establishing their equivalence:
+circuits compute to the following, establishing their equivalence:
 

### 2021-10-24 Pi/paper/examples.tex
fixed example in Sec. 2

--- 2021-07-09 Pi+/paper/examples.tex
@@
 $\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2),
-~b_1, ~b_2)$ where $\vee$ is boolean disjunction and $\underline{\vee}$ is the
-exclusive-or operation. The circuits are presented in both the textual interface
+~b_1, ~b_2)$ where~$\vee$ is disjunction and~$\underline{\vee}$ is exclusive-disjunction. The circuits are presented in both the textual interface
 \verb|qasm| and the graphical interface:
@@
 
-There is wealth of manual and algorithmic approaches for producing circuits such
-as the two above~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the left
-was manually produced using a standard synthesis algorithm for reversible
+There is a wealth of manual and algorithmic approaches for producing circuits
+such as the two above~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the
+left was manually produced using a standard synthesis algorithm for reversible
 circuits~\cite{10.1145/775832.775915}. The circuit on the right was produced
-using an approach that analyzes the recursive structure of the circuit (and
-would generalise to computing the disjunction of more than two inputs):
+using an approach that analyzes the recursive structure of
+$\mathit{reversibleOr}$ (and would generalise to computing the disjunction of
+more than two inputs):
 
@@
 interesting example would be the classical core of Shor's algorithm which
-requires a circuit implementing $f(r) = a^{r} \mod N$ for fixed $a$ and $N$. The
-specification of the circuit is relatively straightforward to calculate. Here it
-is for $a=11$ and $N=15$:
-\[\begin{array}{rcll}
-g(r,h) &=& \left\{ \begin{array}{ll}
-                     (r,h+1) & \mbox{when~$r$~even~and~$h$~even} \\
-                     (r,h-1) & \mbox{when~$r$~even~and~$h$~odd} \\
-                     (r,11-h) & \mbox{when~$r$~odd~and~$4 > h \geq 0$~or~$12 > h \geq 8$} \\
-                     (r,19-h) & \mbox{when~$r$~odd~and~$8 > h \geq 4$~or~$16 > h \geq 12$}
-                                \end{array}\right.
-\end{array}\]
-
-\noindent However, as explained in standard accounts of the algorithm (e.g., the
+requires a circuit implementing modular exponentiation $f(r) = a^{r} \mod N$ for
+fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{Toffoli:1980}, a
+reversible specification of this function is $f^R(r,h) = (r, f(r)
+~\underline{\vee}~ h)$ where the exclusive disjunction~$\underline{\vee}$ is
+applied bitwise. When $h=0$ the second component of the result of $f^R$ is the
+result of the original modular exponentiation $f$.
+%%
+%% the specification of the circuit is relatively straightforward to
+%% calculate. Here it is for $a=11$ and $N=15$:
+%% \[\begin{array}{rcll}
+%% g(r,h) &=& \left\{ \begin{array}{ll}
+%%                      (r,h+1) & \mbox{when~$r$~even~and~$h$~even} \\
+%%                      (r,h-1) & \mbox{when~$r$~even~and~$h$~odd} \\
+%%                      (r,11-h) & \mbox{when~$r$~odd~and~$4 > h \geq 0$~or~$12 > h \geq 8$} \\
+%%                      (r,19-h) & \mbox{when~$r$~odd~and~$8 > h \geq 4$~or~$16 > h \geq 12$}
+%%                                 \end{array}\right.
+%% \end{array}\]
+%%
+However, as explained in standard accounts of the algorithm (e.g., the
 Qiskit implementation), producing an efficient modular exponentiation circuit

### 2021-10-24 Pi/paper/reversible1.tex
Fix the composition symbol

@@
       {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : B \isoone C}
-      {\jdg{}{}{c_1 \fatsemi c_2 : A \isoone C}}
+      {\jdg{}{}{c_1 \circledcirc c_2 : A \isoone C}}
       {}
@@
 
-To see how this language expresses reversible circuits, we present a few examples in the Agda embedding of the
-language. The Agda syntax is almost identical to the syntax in Fig.~\ref{fig:pi-terms} except that sequential
-composition is written using $\circledcirc$. First it is possible to directly mimic the \verb|qasm|-perspective by
-defining types that describe sequences of booleans. We use the type $\mathbb{2} = \onet + \onet$ to represent booleans
-with $\inlv{\Acon{tt}}$ representing \textsf{true} and $\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean
-negation (the \verb|x|-gate) is straightforward to define using the primitive combinator $\swapp$. We can represent
-$n$-bit words using an n-ary product of boolean values, thus the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$
-(abbreviated $\mathbb{B}~3$) corresponds to a collection of wires that can transmit three bits.
+To see how this language expresses reversible circuits, we present a few examples in the Agda embedding of the language.
+First it is possible to directly mimic the \verb|qasm|-perspective by defining types that describe sequences of
+booleans. We use the type $\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing
+\textsf{true} and $\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is
+straightforward to define using the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product
+of boolean values, thus the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$)
+corresponds to a collection of wires that can transmit three bits.
 %
@@
 % $\Afun{reverse} : \mathbb{B}~3 \iso \mathbb{B}~3$
-% $\Afun{reverse} = \swapt \fatsemi (\swapt  \otimes  \idc)~ \fatsemi \assocrt$
+% $\Afun{reverse} = \swapt \circledcirc (\swapt  \otimes  \idc)~ \circledcirc \assocrt$
 % \noindent The manual trace of $\Afun{reverse}$ below confirms that it indeed reverses the three bits:
@@
         \begin{array}{rlcl}
-          \denot{(c_1 \fatsemi c_2)} & v           & = & (\denot{c_2} \circ \denot{c_1}) v   \\
+          \denot{(c_1 \circledcirc c_2)} & v           & = & (\denot{c_2} \circ \denot{c_1}) v   \\
           \denot{(c_1 \oplus c_2)}   & (\inl{v})   & = & \inl{(\denot{c_1}~v)}               \\
@@
 % The denotational semantics can be used to evaluate programs and to check equivalence of programs. For example, it is
-% straightforward to verify that $\denot{(c_1~\oplus~c_2)~\fatsemi~\swapp} = \denot{\swapp \fatsemi (c_2~\oplus~c_1)}$. A
+% straightforward to verify that $\denot{(c_1~\oplus~c_2)~\circledcirc~\swapp} = \denot{\swapp \circledcirc (c_2~\oplus~c_1)}$. A
 % slightly more involved example are the following two programs:
@@
 % \note{Then say that $\mathcal{C}$ is ``the same'' as $\BFin$ We can say here that the relation $\sim$ also identifies
-%   $(c_1~\oplus~c_2)~\fatsemi~\swapp$ and $\swapp \fatsemi (c_2~\oplus~c_1)$ and other properties. ?? }
+%   $(c_1~\oplus~c_2)~\circledcirc~\swapp$ and $\swapp \circledcirc (c_2~\oplus~c_1)$ and other properties. ?? }
 

### 2021-10-26 Pi/paper/equivalence.tex
Fix the composition symbol

@@
 Recall the specification of reversible disjunction from \Cref{sec:examples}, the two Qiskit circuits for implementing
-it, and the corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The normal forms for both
-circuits compute to the following, establishing their equivalence:
+it, and the corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The type $\mathbb{B}\ 3$
+normalises to $\mathbb{8}+ = \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} +
+(\mathbb{1} + \mathbb{0})))))))$, and normal forms for both circuits compute to the following, establishing their equivalence:
 
@@
 
-Instead of manually producing $\Pi$-programs to implement the reversible disjunction specification, it is also possible
-to simply quote the desired permutation:
+Instead of manually producing $\Pi$-programs to implement the reversible disjunction specification, it is possible to
+simply quote the desired permutation (using a function $\term{lookup}$ to read a tabulated version for readability):
 

### 2021-10-26 Pi/paper/finite.tex
one pass addressing things from review A
Explain how the relation is lifted to presentation
Fixes #142

Co-authored-by: Jacek Karwowski <inexxt@users.noreply.github.com>

--- 2021-09-05 Pi+/paper/finite.tex
@@
 \begin{definition}[Free group]
-  \label{def:presentation}
   Let $A$ be a set, and $\List[\blank]$ the free monoid. The free group $F(A)$ on $A$ is the set-quotient of $\List[A +
@@
 \begin{definition}[Group presentation]
+  \label{def:presentation}
   Let $A$ be a set and $R : \List[A + A] \to \List[A + A] \to \UU$ a binary relation on $\List[A + A]$. The group
@@
 
-Notice however, that a group presentation, as defined in~\Cref{def:presentation}, requires the relation to be on the set
-of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The constructor
-$\cancel$ specifies that the inverse of each element is again the same element, using which we can show that our
-definition of $\Sn$ is equivalent to the definition of a presented group, by lifting the $\cox*$ relation.
+\noindent
+Finally, we need to show that $\Sn$ and is indeed a group presentation.
 
-\begin{theorem}
-  \leavevmode
-  \begin{enumerate}
-    \item There is a group structure on $\Sn$, where the identity element is $\nil$, multiplication is given by list
-          append, and inverse is given by list reversal.
-    \item $\Sn$ is equivalent to the group presented by generators $\Fin[n]$ with the relations given by $\cox*$
-          extended to $\List(\Fin[n] + \Fin[n])$ along the codiagonal map $\nabla_{A} : A + A \to A$.
-  \end{enumerate}
-\end{theorem}
+\begin{propositionrep}
+  \label{prop:group-structure-Sn}
+  There is a group structure on $\Sn$, where the identity element is $\nil$, multiplication is given by list append, and
+  inverse is given by list reversal.
+\end{propositionrep}
 
-To decide if two words in $\List[\Fin[n]]$ are $\cox*$-equal, we simply have to compute their normal forms using
+\begin{proof}
+   The operation $\append$ is associative, with $\nil$ as the neutral element, and the constructor $\cancel$ makes each
+   element an inverse of itself, so the inverse in the group is indeed given by list reversal.
+\end{proof}
+
+\noindent
+Notice however, that for a group presentation, as defined in~\Cref{def:presentation}, the relation needs to be on the
+set of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The following
+proposition lets us lift a group structure on $\List[A]$ to a group presentation.
+
+\begin{propositionrep}
+  \label{prop:generated-group-generalised}
+  Given a set $A$ and a congruence relation $R$ on $\List[A]$ with the group structure given by append and reversal,
+  there is a generated group structure $F(\langle A ; \liftrel{R} \rangle)$, where $\liftrel{R}$ is the relation $R$
+  lifted to $\List[A+A]$ along the codiagonal map $\nabla_{A} : A + A \to A$.
+\end{propositionrep}
+
+\begin{proof}
+  Let $G$ be the group on $\List[A]$. We construct the group homomorphism from $F(\langle A ; \liftrel{R} \rangle)$ to
+  $G$ using the universal property of the group presentation, where the map on generators is given by taking the
+  singleton list. To construct the group homomorphism from $G$, we use the universal property of the set quotient, where
+  the map on the underlying set is given by mapping $\inl$ on the list. Using that the relation is a congruence, we show
+  that this consttitutes a group isomorphism.
+\end{proof}
+
+\begin{corollary}
+  The group $\Sn$ is isomorphic to the group presentation given by ${F(\langle \Fin[n] ; \liftrel{\cox*} \rangle)}$.
+\end{corollary}
+
+\noindent To decide if two words in $\List[\Fin[n]]$ are $\cox*$-equal, we simply have to compute their normal forms using
 $\normf$. They correspond to the same permutation if and only if these normal forms are equal, which is decidable for
@@
     \item For any $k : \Fin[\suc[n]]$, we have $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-2}
-    \item For any $n : \Nat$, we have \( \Aut[\Fin[\suc[n]]] \eqv \dsum*{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[n] \eqv
-            \FinExcept{\suc[n]}[n - k]} \). \label{prop:fin-finexcept-3}
+    \item For any $n : \Nat$, we have \( \Aut[\Fin[\suc[n]]] \eqv \dsum*{k : \Fin[\suc[n]]}{(\FinExcept{\suc[n]}[n] \eqv
+            \FinExcept{\suc[n]}[n - k])} \). \label{prop:fin-finexcept-3}
   \end{enumerate}
@@
     \arraycolsep=0.5em\def\arraystretch{1.5}
-    \begin{array}{rl}
-           & \Lehmer[\zero]          \\
-      \eqv & \unit                   \\
-      \eqv & \Aut[\Fin[\suc[\zero]]]
-    \end{array}
-    \qquad\qquad
+    %% \begin{array}{rl}
+    %%        & \Lehmer[\zero]          \\
+    %%   \eqv & \unit                   \\
+    %%   \eqv & \Aut[\Fin[\suc[\zero]]]
+    %% \end{array}
+    %% \qquad\qquad
     \begin{array}{rlr}

### 2021-10-26 Pi/paper/equivalence.tex
Explain U, U+, U^
Fixes #143

@@
 $\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and
-$\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can show that these
-syntactic categories here are actually $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are
-also locally-strict, or locally-posetal, because there is at most one 2-cell between compatible 1-cells.
+$\PiHatCat$, have 0-cells for types ($U$, $\UPlus$, and $\UHat$), 1-cells for 1-combinators ($\iso$, $\isop$, and
+$\isoh$), and 2-cells for 2-combinators ($\Iso$, $\Isop$, and $\Isoh$). We can show that these syntactic categories here
+are actually $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are also locally-strict, or
+locally-posetal, because there is at most one 2-cell between compatible 1-cells.
 
@@
         \begin{array}{rrcll}
-          \idc :     & A           & \iso & A           & : \idc     \\
-          \identlp : & \zerot + A  & \iso & A           & : \identrp \\
-          \swapp :   & A + B       & \iso & B + A       & : \swapp   \\
-          \assoclp : & A + (B + C) & \iso & (A + B) + C & : \assocrp \\ [1.5ex]
+          \idc :     & A           & \isop & A           & : \idc     \\
+          \identlp : & \zerot + A  & \isop & A           & : \identrp \\
+          \swapp :   & A + B       & \isop & B + A       & : \swapp   \\
+          \assoclp : & A + (B + C) & \isop & (A + B) + C & : \assocrp \\ [1.5ex]
         \end{array}$}}
@@
       \Rule{}
-      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : B \iso C}
-      {\jdg{}{}{c_1 \fatsemi c_2 : A \iso C}}
+      {\jdg{}{}{c_1 : A \isop B} \quad \vdash c_2 : B \isop C}
+      {\jdg{}{}{c_1 \circledcirc c_2 : A \isop C}}
       {}
@@
       \Rule{}
-      {\jdg{}{}{c_1 : A \iso B} \quad \vdash c_2 : C \iso D}
-      {\jdg{}{}{c_1 \oplus c_2 : A + C \iso B + D}}
+      {\jdg{}{}{c_1 : A \isop B} \quad \vdash c_2 : C \isop D}
+      {\jdg{}{}{c_1 \oplus c_2 : A + C \isop B + D}}
       {}
@@
       {\jdg{}{}{c_1 : n \isoh m} \quad \vdash c_2 : m \isoh o}
-      {\jdg{}{}{c_1 \fatsemi c_2 : n \isoh o}}
+      {\jdg{}{}{c_1 \circledcirc c_2 : n \isoh o}}
       {}

### 2021-10-27 Pi/paper/ufin.tex
Restore some HoTT text from appendix
Fixes #141

--- 2021-09-05 Pi+/paper/ufin.tex
@@
 Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
-as well.  In~\cref{app:identitytypes},~\cref{app:homotopytypes}, and~\cref{app:hits}, we review the basics of identity
-types, homotopy types, and HITs and refer the reader to the book for more details.
+as well. We review the basics of identity types, homotopy types, and some HITs that we use, and refer the reader to the
+book for more details.
 
@@
 
-\begin{toappendix}
 \subsection{Identity Types}
@@
 Given two terms $x:A$ and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the identity type, which is the type
-  of equalities or identifications between them. The identity type is generated by reflexivity $\refl_{x} : x \id_{A} x$,
-  and the eliminator for the identity type is given by path induction or the $J$-rule (\cref{def:path-induction}). This
-  construction can be iterated, giving the identity type between two terms of an identity type, repeating ad infinitum.
-  Using the iterated identity type for morphisms, each type is equipped with the structure of a weak $\infty$-groupoid,
-  where each morphism satisfies groupoid laws only upto a higher one. Given an arbitrary type (or groupoid) $A$, we list
-  some laws that are provable using path induction.
+of equalities or identifications between them. The identity type is generated by reflexivity $\refl_{x} : x \id_{A}
+  x$, and the eliminator for the identity type is given by path induction or the $J$-rule (see~\cref{def:path-induction}
+in the appendix). This construction can be iterated, giving the identity type between two terms of an identity type,
+repeating ad infinitum. Using the iterated identity type for morphisms, each type is equipped with the structure of a
+weak $\infty$-groupoid, where each morphism satisfies groupoid laws only upto a higher one.
+
+\begin{toappendix}
+  Given an arbitrary type (or groupoid) $A$, we list some laws that are provable using path induction.
 
@@
   \end{gather*}
+\end{toappendix}
 
-  \noindent A homotopy between functions $f \htpy g$ is given by pointwise equality between them $\dfun{x:A}{f(x) \id_{B} g(x)}$.
-  The identity type for functions is equivalent to homotopies between them ${(f \id_{A \to B} g)} \eqv {(f \htpy g)}$, by
-  function extensionality. An equivalence between types $A \eqv B$ is given by a pair of functions between them which
-  compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy \idfunc_{A}$, and this is equivalent to the
-  identity type for the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.
+A homotopy between functions $f, g : A \to B$, written $f \htpy g$, is given by pointwise equality between them,
+$\dfun*{x:A}{f(x) \id_{B} g(x)}$. The identity type for functions is equivalent to homotopies between them ${(f \id_{A
+        \to B} g)} \eqv {(f \htpy g)}$, by $\term{funext}$ and $\term{happly}$. An equivalence between types $A \eqv B$ is given
+by a pair of functions between them which compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy
+  \idfunc_{A}$ (see~\cref{prop:three-chars-equiv} in the appendix), and this is equivalent to the identity type for the
+universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.
 
-  Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action
-  on the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to
-  the universe, such as $A \to \UU$. For a type family $P : A \to \UU$ and a point $x : A$, the type $P(x)$ is called
-  the fiber over $x$.  The $\term{transport}$ operation, named $\term{transport}/term{tr}$, lifts paths in the indexing
-  type to functions between fibers.
-  %which is an $A$-indexed family of groupoids.
-
+Functions between types are functors between groupoids. Given a function $f : A \to B$, there is a functorial action on
+the paths given by $\term{ap}$. Type families, that is, types indexed by terms, are simply functions from a type to the
+universe, such as $A \to \UU$, which is an $A$-indexed family of groupoids. For a type family $P : A \to \UU$ and a
+point $x : A$, the type $P(x)$ is the fiber over $x$.  The $\term{transport}$ operation, named
+$\term{transport}/\term{tr}$, lifts paths in the indexing type to functions between fibers.
 \begin{gather*}
-    \term{ap}_{f} : \dfun{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
+  \term{ap}_{f} : \dfun*{x,y:A}{x \id_{A} y \to f(x) \id_{B} f(y)}
   \qquad
-    \term{transport}_{P} : \dfun{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
+  \term{transport}_{P} : \dfun*{x,y:A}{x \id_{A} y \to P(x) \to P(y)}
 \end{gather*}
 
-  The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is called the total space of $P$. The first
-  projection ${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a
-  fibration, that is, there is a lifting operation (\cref{fig:lift}) which lifts paths in the base space to paths in the
+The type $\dsum*{x:A}{P(x)}$ is the collection of all the fibers and is the total space of $P$. The first projection
+${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a fibration, that
+is, there is a lifting operation (\cref{fig:lift} in the appendix) which lifts paths in the base space to paths in the
 total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
 \[
-    \term{lift}(u,p) : (x , u) \id_{\dsum{x:A}{P(x)}} (y , \transport{P}{p}{u})
+  \term{lift}(u,p) : (x , u) \id_{\dsum*{x:A}{P(x)}} (y , \transport{P}{p}{u})
 \]
@@
 % where $\tr{p}{u}$ is shorthand for $$.
-
+\begin{toappendix}
   \begin{figure}
@@
   \end{figure}
-
 \end{toappendix}
@@
 
-Functions between types are functors between groupoids, and type families (or functions to the universe) are indexed
-families of groupoids. A type family $P : A \to \UU$ comes equipped with a functor $\pi_1 : \dsum*{x:A}{P(x)} \to A$
-which has a lifting operation giving it the structure of a fibration. The $\term{transport}$ operation lifts paths in
-the base space to functions between the fibers.  Using the groupoid structure of $A$, for any $x, y : A$ and a path 
-$p : x \id_{A} y$, $\term{transport}(p)$ and $\term{transport}(\inv{p})$ form an equivalence.
-% We name this map $\tptEqv{P}$ which lifts paths to equivalences of fibers.
+Using the groupoid structure of $A$, for any $x, y : A$ and a path $p : x \id_{A} y$, $\term{transport}(p)$ and
+$\term{transport}(\inv{p})$ form an equivalence.
 \[
@@
 
-\noindent The type families (or fibrations) we are interested in are the ones where paths in the base space completely determine
-the equivalences in the fibers -- these are called univalent
-fibrations~\cite*{kapulkinUnivalenceSimplicialSets2018,kapulkinSimplicialModelUnivalent2021,christensenCharacterizationUnivalentFibrations2015}.
+\noindent The type families (or fibrations) we are interested in are the ones where paths in the base space completely
+determine the equivalences in the fibers -- these are called univalent
+fibrations~\cite*{kapulkinUnivalenceSimplicialSets2012,kapulkinSimplicialModelUnivalent2021,christensenCharacterizationUnivalentFibrations2015}.
 
@@
 
-Univalent fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018}, to build a model of Voevodsky's
+Univalent fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2012}, to build a model of Voevodsky's
 \emph{univalence} principle in simplicial sets. Indeed, univalence characterises paths in the universe as equivalences
 between types, which follows from the canonical fibration $\idfunc : \UU \to \UU$ being univalent.
-
-\begin{toappendix}
 
@@
 1-groupoids or 1-types (whose equality types are sets), and similarly, higher homotopy $n$-types.
-
 \begin{gather*}
   \begin{aligned}
-      \isContr{A} & \defeq \dsum{x:A}{\dfun{y:A}{y \id x}} \\
-      \isProp{A}  & \defeq \dfun{x,y:A}{\isContr{x \id y}}
+    \isContr{A} & \defeq \dsum*{x:A}{\dfun*{y:A}{y \id x}} \\
+    \isProp{A}  & \defeq \dfun*{x,y:A}{\isContr{x \id y}}
   \end{aligned}
@@
   \begin{aligned}
-      \isSet{A} & \defeq \dfun{x,y:A}{\isProp{x \id y}} \\
-      \isGpd{A} & \defeq \dfun{x,y:A}{\isSet{x \id y}}
+    \isSet{A} & \defeq \dfun*{x,y:A}{\isProp{x \id y}} \\
+    \isGpd{A} & \defeq \dfun*{x,y:A}{\isSet{x \id y}}
   \end{aligned}
@@
 
+\begin{toappendix}
   \begin{definition}[Propositional Truncation]
@@
     \begin{itemize}
-      \item an inclusion function $\trunc{\blank} : A \to \Trunc{A}$,
+      \item a function $\trunc{\blank} : A \to \Trunc{A}$,
       \item for each $x, y : \Trunc{A}$, a path $\term{trunc}(x,y) : x \id_{\Trunc{A}} y$,
@@
   \end{definition}
+\end{toappendix}
 
+\begin{toappendix}
   \begin{definition}[${\fib_{f}} : B \to \UU$]
@@
     \[
-      \fib_{f}(b) \defeq \dsum{a:A}{f(a) \id_{B} b}.
+      \fib_{f}(b) \defeq \dsum*{a:A}{f(a) \id_{B} b}.
     \]
@@
     \[
-      \im{f} \defeq \dsum{b:B}{\Trunc[-1]{\fib_{f}(b)}}
+      \im{f} \defeq \dsum*{b:B}{\Trunc[-1]{\fib_{f}(b)}}
     \]
@@
 
-  \begin{lemma}
+  \begin{proposition}
+    \label{prop:three-chars-equiv}
     The following are equivalent.
@@
     \end{enumerate}
-  \end{lemma}
+  \end{proposition}
+\end{toappendix}
 
-  Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes an set $A$ and a relation $R : A \to A \to \UU$.
-  It has an inclusion of points $\quotinc : A \to \quot{A}{R}$, and adds paths between related pairs of elements
-  $\quotrel : R(x,y) \to \quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$ (see~\cref{def:set-quot}).  We recall that the
-  quotient is \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff $(q(x)
-  \id_{\quot{A}{R}} q(y))$.
+Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes a set $A$ and a binary relation $R : {A \to A \to
+\UU}$. It has a mapping of points $\quotinc : A \to \quot{A}{R}$, and a constructor that adds paths between related
+pairs of elements $\quotrel : R(x,y) \to \quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$ (see~\cref{def:set-quot}).  We
+recall that the quotient is \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff
+$(q(x) \id_{\quot{A}{R}} q(y))$.
 
+\begin{toappendix}
   \begin{definition}[Set Quotient]
@@
   \end{definition}
-
 \end{toappendix}
@@
   univalent and generates a univalent subuniverse ${\UU_{P} \defeq (\dsum*{X:\UU}{P(X)}, \fst)}$.~\footnote{Univalent
-    typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent subuniverses.}
+  typoids~\cite{petrakisUnivalentTypoids2019} are a different presentation of univalent subuniverses.}
 \end{propositionrep}

### 2021-10-27 Pi/paper/ufin.tex
Resolve comments

@@
 We use the type theory of the HoTT book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use
-intensional Martin-L\"{o}f Type Theory, with a (univalent) universe $\UU$, and a few Higher Inductive Types (HITs) for
-propositional and set truncation, and set-quotients. All arguments will hold in a Cubical Type
+intensional Martin-L\"{o}f Type Theory, with a hierarchy of (univalent) universes $\UU_0 : \UU_1 : \ldots$ (though we
+will just write $\UU$ since we mostly use one universe), and a few Higher Inductive Types (HITs) for truncations and
+quotients. We write $\dfun*{x:A}{B(x)}$ or simply $(x:A) \to B(x)$ for dependent function types, and $\dsum*{x:A}{B(x)}$
+or simply $(x:A) \times B(x)$ for dependent sum types.
+
+All arguments will hold in a Cubical Type
 Theory~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}
@@
 by a pair of functions between them which compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy
-  \idfunc_{A}$ (see~\cref{prop:three-chars-equiv} in the appendix), and this is equivalent to the identity type for the
-universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, by univalence.
+  \idfunc_{A}$ (also see~\cref{prop:three-chars-equiv} in the appendix), and this is equivalent to the identity type for
+the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, which is the univalence principle.
 
@@
 is, there is a lifting operation (\cref{fig:lift} in the appendix) which lifts paths in the base space to paths in the
-total space. Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
+total space.
+
+\begin{toappendix}
+  Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have:
   \[
@@
   % where $\tr{p}{u}$ is shorthand for $$.
-\begin{toappendix}
+
   \begin{figure}
@@
 Using the groupoid structure of $A$, for any $x, y : A$ and a path $p : x \id_{A} y$, $\term{transport}(p)$ and
-$\term{transport}(\inv{p})$ form an equivalence.
+$\term{transport}(\inv{p})$ form an equivalence, lifting paths in the base space to equivalences between fibers.
 \[

### 2021-10-27 Pi/paper/ufin.tex
Expand homotopy types and add fib/im
Fixes #149

@@
 by a pair of functions between them which compose to the identity, $f \comp g \htpy \idfunc_{B}$ and $g \comp f \htpy
-  \idfunc_{A}$ (also see~\cref{prop:three-chars-equiv} in the appendix), and this is equivalent to the identity type for
-the universe, $(A \id_{\UU} B) \eqv (A \eqv B)$, which is the univalence principle.
+\idfunc_{A}$ (also see~\cref{prop:three-chars-equiv}), and this is equivalent to the identity type for the universe, $(A
+\id_{\UU} B) \eqv (A \eqv B)$, which is the univalence principle.
 
@@
 
-\subsection{Homotopy Types}
-\label{app:homotopytypes}
-
-A type is \emph{contractible} (-2-type) if it has a unique element, that is, there is a center of contraction and every
-other point is equal to it. A type is a \emph{proposition} (-1-type) if its equality types are contractible, that is, it
-has at most one inhabitant. Iterating this, we can define sets or 0-types (whose equality types are propositions) and
-1-groupoids or 1-types (whose equality types are sets), and similarly, higher homotopy $n$-types.
-\begin{gather*}
-  \begin{aligned}
-    \isContr{A} & \defeq \dsum*{x:A}{\dfun*{y:A}{y \id x}} \\
-    \isProp{A}  & \defeq \dfun*{x,y:A}{\isContr{x \id y}}
-  \end{aligned}
-  \qquad
-  \begin{aligned}
-    \isSet{A} & \defeq \dfun*{x,y:A}{\isProp{x \id y}} \\
-    \isGpd{A} & \defeq \dfun*{x,y:A}{\isSet{x \id y}}
-  \end{aligned}
-\end{gather*}
-
 \subsection{Higher Inductive Types}
@@
 
-\begin{toappendix}
-  \begin{definition}[${\fib_{f}} : B \to \UU$]
-    \label{def:fib}
-    The fiber of $f : A \to B$ at $b : B$ is
-    \[
-      \fib_{f}(b) \defeq \dsum*{a:A}{f(a) \id_{B} b}.
-    \]
-  \end{definition}
-
-  \begin{definition}[${\term{im}} : (f : A \to B) \to \UU$]
-    \label{def:im}
-    The image of $f$ is the (-1)-truncation of its fiber.
-    \[
-      \im{f} \defeq \dsum*{b:B}{\Trunc[-1]{\fib_{f}(b)}}
-    \]
-  \end{definition}
-
-  \begin{proposition}
-    \label{prop:three-chars-equiv}
-    The following are equivalent.
-    \begin{enumerate}
-      \item $f : A \to B$ is an equivalence.
-      \item $f$ has a left and right inverse.
-      \item $f$ has contractible fibers.
-    \end{enumerate}
-  \end{proposition}
-\end{toappendix}
-
 Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes a set $A$ and a binary relation $R : {A \to A \to
@@
 \end{toappendix}
+
+\subsection{Homotopy Types}
+\label{app:homotopytypes}
+
+A type is \emph{contractible} (-2-type) if it has a unique element, that is, there is a center of contraction and every
+other point is equal to it. A type is a \emph{proposition} (-1-type) if its equality types are contractible, that is, it
+has at most one inhabitant. Iterating this, we can define sets or 0-types (whose equality types are propositions) and
+1-groupoids or 1-types (whose equality types are sets), and similarly, higher homotopy $n$-types.
+\begin{gather*}
+  \begin{aligned}
+    \isContr{A} & \defeq \dsum*{x:A}{\dfun*{y:A}{y \id x}} \\
+    \isProp{A}  & \defeq \dfun*{x,y:A}{\isContr{x \id y}}
+  \end{aligned}
+  \qquad
+  \begin{aligned}
+    \isSet{A} & \defeq \dfun*{x,y:A}{\isProp{x \id y}} \\
+    \isGpd{A} & \defeq \dfun*{x,y:A}{\isSet{x \id y}}
+  \end{aligned}
+\end{gather*}
+
+\noindent
+Given a function $f : A \to B$, there is an associated fibration over $B$, given by the homotopy fiber. The image of $f$
+is defined to be the (-1)-truncation of its fiber.
+\begin{gather*}
+  \fib_{f}(b) \defeq \dsum*{a:A}{f(a) \id_{B} b}
+  \qquad
+  \im{f} \defeq \dsum*{b:B}{\Trunc[-1]{\fib_{f}(b)}}
+\end{gather*}
+
+\begin{proposition}
+  \label{prop:three-chars-equiv}
+  The following propositions are equivalent.
+  \begin{enumerate}
+    \item $f : A \to B$ is an equivalence.
+    \item $f$ has a left and right inverse.
+    \item $f$ has contractible fibers.
+  \end{enumerate}
+\end{proposition}
 

### 2021-10-27 Pi/paper/finite.tex
Restore some HoTT text from appendix
Fixes #141
Expand homotopy types and add fib/im
Fixes #149
Fix broken references

@@
 is terminating, so by Newman's lemma, it produces a unique normal form. We follow the terminology
-of~\citet*{huetConfluentReductionsAbstract1980a,krausCoherenceWellFoundednessTaming2020} to state our results formally.
+of~\citet*{huetConfluentReductionsAbstract1980,krausCoherenceWellFoundednessTaming2020} to state our results formally.
 
@@
             \xrightarrow{r} {\List[\Fin[n]]}$ such that $s \comp r \htpy \normf$ and $r \comp s \htpy \idfunc_{\Sn[n]}$.
-    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \), where $\quotinc$ is the mapping into the quotient, and
-          $\im{f}$ denotes the image of $f$ (see~\cref{def:im}).
+    \item \(\im{\quotinc} \eqv \Sn \eqv \im{\normf} \), where $\quotinc : \List[\Fin[n]] \to \Sn$ is the mapping into
+    the quotient.
   \end{enumerate}
@@
   \end{enumerate}
-  Therefore, there is an equivalence between $\Lehmer[n]$ and $\im{\normf}$.
+  Therefore, there is an equivalence between $\Lehmer[n]$ and $\im{\normf}$ (see~\cref{prop:three-chars-equiv}).
 \end{theoremrep}

### 2021-10-27 Pi/paper/reversible1.tex
Adjust spacing in semantics notation

@@
   {\scalebox{\scalef}{$%
-        \begin{array}[t]{rlcl}
-          \denot{\identlp} & (\inl{v})         & = & v               \\
-          \denot{\identrp} & v                 & = & \inl{v}         \\
-          \denot{\swapp}   & (\inl{v})         & = & \inr{v}         \\
-          \denot{\swapp}   & (\inr{v})         & = & \inl{v}         \\
-          \denot{\assoclp} & (\inl{v})         & = & \inl{(\inl{v})} \\
-          \denot{\assoclp} & (\inr{(\inl{v})}) & = & \inl{(\inr{v})} \\
-          \denot{\assoclp} & (\inr{(\inr{v})}) & = & \inr{v}         \\
-          \denot{\assocrp} & (\inl{(\inl{v})}) & = & \inl{v}         \\
-          \denot{\assocrp} & (\inl{(\inr{v})}) & = & \inr{(\inl{v})} \\
-          \denot{\assocrp} & (\inr{v})         & = & \inr{(\inr{v})}
-        \end{array}$}}
+        \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
+          \pdenot{\identlp} & (\inl{v})         & = & v               \\
+          \pdenot{\identrp} & v                 & = & \inl{v}         \\
+          \pdenot{\swapp}   & (\inl{v})         & = & \inr{v}         \\
+          \pdenot{\swapp}   & (\inr{v})         & = & \inl{v}         \\
+          \pdenot{\assoclp} & (\inl{v})         & = & \inl{(\inl{v})} \\
+          \pdenot{\assoclp} & (\inr{(\inl{v})}) & = & \inl{(\inr{v})} \\
+          \pdenot{\assoclp} & (\inr{(\inr{v})}) & = & \inr{v}         \\
+          \pdenot{\assocrp} & (\inl{(\inl{v})}) & = & \inl{v}         \\
+          \pdenot{\assocrp} & (\inl{(\inr{v})}) & = & \inr{(\inl{v})} \\
+          \pdenot{\assocrp} & (\inr{v})         & = & \inr{(\inr{v})}
+        \end{tabular}$}}
 
   {\scalebox{\scalef}{$%
-        \begin{array}[t]{rlcl}
-          \denot{\identlt} & (\ttt , v)          & = & v                   \\
-          \denot{\identrt} & v                   & = & (\ttt , v)          \\
-          \denot{\swapt}   & (v_1 , v_2)         & = & (v_2 , v_1)         \\
-          \denot{\assoclt} & (v_1 , (v_2 , v_3)) & = & ((v_1 , v_2) , v_3) \\
-          \denot{\assocrt} & ((v_1 , v_2) , v_3) & = & (v_1 , (v_2 , v_3)) \\
-          \denot{\dist}    & (\inl{v_1} , v_3)   & = & \inl{(v_1 , v_3)}   \\
-          \denot{\dist}    & (\inr{v_2 , v_3})   & = & \inr{(v_2 , v_3)}   \\
-          \denot{\factor}  & (\inl{(v_1 , v_3)}) & = & (\inl{v_1} , v_3)   \\
-          \denot{\factor}  & (\inr{(v_2 , v_3)}) & = & (\inr{v_2} , v_3)   \\
-          \denot{\idc}     & v                   & = & v
-        \end{array}$}}
+        \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
+          \pdenot{\identlt} & (\ttt , v)          & = & v                   \\
+          \pdenot{\identrt} & v                   & = & (\ttt , v)          \\
+          \pdenot{\swapt}   & (v_1 , v_2)         & = & (v_2 , v_1)         \\
+          \pdenot{\assoclt} & (v_1 , (v_2 , v_3)) & = & ((v_1 , v_2) , v_3) \\
+          \pdenot{\assocrt} & ((v_1 , v_2) , v_3) & = & (v_1 , (v_2 , v_3)) \\
+          \pdenot{\dist}    & (\inl{v_1} , v_3)   & = & \inl{(v_1 , v_3)}   \\
+          \pdenot{\dist}    & (\inr{v_2 , v_3})   & = & \inr{(v_2 , v_3)}   \\
+          \pdenot{\factor}  & (\inl{(v_1 , v_3)}) & = & (\inl{v_1} , v_3)   \\
+          \pdenot{\factor}  & (\inr{(v_2 , v_3)}) & = & (\inr{v_2} , v_3)   \\
+          \pdenot{\idc}     & v                   & = & v
+        \end{tabular}$}}
   %%\end{longtable}
@@
   {\scalebox{\scalef}{$%
-        \begin{array}{rlcl}
-          \denot{(c_1 \circledcirc c_2)} & v           & = & (\denot{c_2} \circ \denot{c_1}) v   \\
-          \denot{(c_1 \oplus c_2)}   & (\inl{v})   & = & \inl{(\denot{c_1}~v)}               \\
-          \denot{(c_1 \oplus c_2)}   & (\inr{v})   & = & \inr{(\denot{c_2}~v)}               \\
-          \denot{(c_1 \otimes c_2)}  & (v_1 , v_2) & = & (\denot{c_1} v_1 , \denot{c_2} v_2)
-        \end{array}
+        \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
+          \pdenot{(c_1 \circledcirc c_2)} & v           & = & \denot{c_2} (\denot{c_1}~v)         \\
+          \pdenot{(c_1 \oplus c_2)}       & (\inl{v})   & = & \inl{(\denot{c_1}~v)}               \\
+          \pdenot{(c_1 \oplus c_2)}       & (\inr{v})   & = & \inr{(\denot{c_2}~v)}               \\
+          \pdenot{(c_1 \otimes c_2)}      & (v_1 , v_2) & = & (\denot{c_1}~v_1 , \denot{c_2}~v_2)
+        \end{tabular}
       $}}
@@
   \begin{itemize}
-    \item For every level-1 combinator $c : A \isoone B$, we have that $\denot{c}$ is a bijection between $\denot{A}$ and $\denot{B}$.
+    \item For every level-1 combinator $c : A \isoone B$, we have a bijection $\denot{c}$ between $\denot{A}$ and $\denot{B}$.
     \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
-          combinator $\alpha$ such that $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using
-          extensional equivalence of functions.
+          combinator $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using extensional equivalence of
+          functions.
   \end{itemize}

### 2021-10-27 Pi/paper/introduction.tex
Citation updates

@@
 The proofs of some of our lemmas and propositions and theorems, as well as additional material, are given in the
-supplementary appendices, and we refer to them in the text. Our accompanying Agda code contains the formalisation of the
-full syntax, and most of our proofs.
+supplementary appendices in the extended version~\cite{choudhurySymmetriesReversibleProgramming2021}, and we refer to
+them in the text. Our accompanying Agda code~\cite{choudhuryArtifactSymmetriesReversible2021} contains the formalisation
+of the full syntax, and most of our proofs.
 

### 2021-10-27 Pi/paper/examples.tex
Move qiskit example to a figure

@@
 
-For concreteness, we demonstrate two different circuits that implement the
-following reversible function specification
-$\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2),
-~b_1, ~b_2)$ where~$\vee$ is disjunction and~$\underline{\vee}$ is exclusive-disjunction. The circuits are presented in both the textual interface
-\verb|qasm| and the graphical interface:
+For concreteness, we demonstrate two different circuits that implement the following reversible function specification
+$\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
+and~$\underline{\vee}$ is exclusive-disjunction. The circuits are presented in both the textual interface \verb|qasm|
+and the graphical interface in~\cref{fig:qiskit}.
 
-\medskip
+\begin{figure}
   \begin{tabular}{c@{\qquad}c}
@@
   \end{tabular}
+  \caption{$\mathit{reversibleOr}$ in Qiskit}
+  \label{fig:qiskit}
+\end{figure}
 
-\medskip
-
-There is a wealth of manual and algorithmic approaches for producing circuits
-such as the two above~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the
-left was manually produced using a standard synthesis algorithm for reversible
-circuits~\cite{10.1145/775832.775915}. The circuit on the right was produced
-using an approach that analyzes the recursive structure of
-$\mathit{reversibleOr}$ (and would generalise to computing the disjunction of
+There is a wealth of manual and algorithmic approaches for producing circuits such as these
+two~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the left was manually produced using a standard synthesis
+algorithm for reversible circuits~\cite{10.1145/775832.775915}. The circuit on the right was produced using an approach
+that analyzes the recursive structure of $\mathit{reversibleOr}$ (and would generalise to computing the disjunction of
 more than two inputs):
@@
 
-Although rather trivial, the examples above illustrate the general idea. A more
-interesting example would be the classical core of Shor's algorithm which
-requires a circuit implementing modular exponentiation $f(r) = a^{r} \mod N$ for
-fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{Toffoli:1980}, a
-reversible specification of this function is $f^R(r,h) = (r, f(r)
-~\underline{\vee}~ h)$ where the exclusive disjunction~$\underline{\vee}$ is
-applied bitwise. When $h=0$ the second component of the result of $f^R$ is the
-result of the original modular exponentiation $f$.
+Although rather trivial, the examples above illustrate the general idea. A more interesting example would be the
+classical core of Shor's algorithm which requires a circuit implementing modular exponentiation $f(r) = a^{r} \mod N$
+for fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{Toffoli:1980}, a reversible specification of this
+function is $f^R(r,h) = (r, f(r)~\underline{\vee}~ h)$ where the exclusive disjunction~$\underline{\vee}$ is applied
+bitwise. When $h=0$ the second component of the result of $f^R$ is the result of the original modular exponentiation
+$f$.
 %%

### 2021-10-27 Pi/paper/equivalence.tex
Use uniform style for references

@@
 
-We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\Cref*{fig:piplus,fig:pihat} respectively,
+We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\cref*{fig:piplus,fig:pihat} respectively,
 eliding the 2-combinators for brevity. We enforce that there is a unique 2-combinator between compatible 1-combinators,
@@
 First, we show how to translate $\PiLang$ programs to $\PiPlusLang$, which is the additive fragment of $\PiLang$. The
-syntax for 1-combinators is given in~\Cref{fig:piplus}.
+syntax for 1-combinators is given in~\cref{fig:piplus}.
 
@@
 $\PiPlusLang$, with (unary) natural numbers for 0-cells, 1-combinators generated by adjacent transpositions, and an
-appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators, in~\Cref{fig:pihat}.
+appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators, in~\cref{fig:pihat}.
 
@@
 $\Aut[\Fin[n]]$. In $\PiHatLang$, the 1-combinators are generated by adjacent transpositions, so these can be mapped to
-words in $\Sn$ and then to automorphisms using~\Cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to
+words in $\Sn$ and then to automorphisms using~\cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to
 2-paths between loops in $\UFin$.
@@
 The semantics that we presented here takes a different route to constructing the permutation from a $\PiLang$
-combinator, compared to the direct interpretation given using the big-step interpreter in~\Cref{subsec:denotational}. We
+combinator, compared to the direct interpretation given using the big-step interpreter in~\cref{subsec:denotational}. We
 verify that the two semantics agree, establishing that the semantics is adequate and fully abstract.
@@
 
-Recall the specification of reversible disjunction from \Cref{sec:examples}, the two Qiskit circuits for implementing
+Recall the specification of reversible disjunction from \cref{sec:examples}, the two Qiskit circuits for implementing
 it, and the corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The type $\mathbb{B}\ 3$

### 2021-10-27 Pi/paper/finite.tex
Use uniform style for references

@@
 % Recap of the previous section
-In~\Cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
+In~\cref{sec:ufin}, we established that paths in $\UFin$ are equivalent to families of loops on $\Fin[n]$ for every
 $n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group.  This
@@
   reversing the list. Further, $F(A)$ with $\eta_A(a) \defeq \inl(a) \cons \nil$ satisfies the universal property of
-  free groups, as stated in~\Cref{prop:free-groups}.
+  free groups, as stated in~\cref{prop:free-groups}.
 \end{proposition}
@@
 
-The universal property of the above definition is similar to~\Cref{prop:free-groups} except the relation has to be
+The universal property of the above definition is similar to~\cref{prop:free-groups} except the relation has to be
 preserved by the function mapping out of the generating set.
@@
 $k$ swaps elements at indices $k$ and $k+1$. Thus, the generating set is $\Fin[n]$. There are three relations that we're
-going to specify for this presentation -- we visualise them as braid diagrams in~\Cref{fig:coxeter-braid}.
+going to specify for this presentation -- we visualise them as braid diagrams in~\cref{fig:coxeter-braid}.
 
@@
 rewriting system is confluent -- meaning that all critical pairs, that is, terms with overlapping possible reduction
-rules, have to converge. For example, in our system, the pair in~\Cref{fig:critical-pairs-converging} converges.
-Unfortunately, this is not true for all critical pairs -- an  example is in~\Cref{fig:critical-pairs-non-converging},
+rules, have to converge. For example, in our system, the pair in~\cref{fig:critical-pairs-converging} converges.
+Unfortunately, this is not true for all critical pairs -- an  example is in~\cref{fig:critical-pairs-non-converging},
 where left and right endpoints are normal with respect to the $\cox*$ relation.
@@
 
-By~\Cref{prop:nf}~\ref{prop:nf-uniqueness}, we get a unique choice function $\normf : {\List[\Fin[n]]} \to
+By~\cref{prop:nf}~\ref{prop:nf-uniqueness}, we get a unique choice function $\normf : {\List[\Fin[n]]} \to
   {\List[\Fin[n]]}$ that produces a normal form for terms of $\List[\Fin[n]]$. We state two important properties enjoyed
@@
 \noindent
-Notice however, that for a group presentation, as defined in~\Cref{def:presentation}, the relation needs to be on the
+Notice however, that for a group presentation, as defined in~\cref{def:presentation}, the relation needs to be on the
 set of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The following
@@
 
-\noindent By composing~\Cref{prop:lehmer-aut-equiv} and~\Cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.
+\noindent By composing~\cref{prop:lehmer-aut-equiv} and~\cref{prop:sn-im-lehmer-equiv}, we obtain the final equivalence.
 

### 2021-10-27 Pi/paper/introduction.tex
Use uniform style for references

@@
 \begin{itemize}[leftmargin=*]
-      \item We start in~\Cref{sec:qiskit} by presenting a few reversible circuits in the popular IBM Qiskit framework to
+      \item We start in~\cref{sec:qiskit} by presenting a few reversible circuits in the popular IBM Qiskit framework to
             serve as running examples throughout the paper.
-      \item In~\Cref{sec:pi}, we introduce the two-level language
+      \item In~\cref{sec:pi}, we introduce the two-level language
             $\PiLang$~\cite{jamesInformationEffects2012,caretteComputingSemiringsWeak2016} and illustrate how to write
@@
             verifying that programs identified by level-2 constructs denote the same bijection.
-      \item ~\Cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in
+      \item ~\cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in
             Homotopy Type Theory. We define and characterise the notion of a univalent subuniverse, and construct
@@
             permutation group on $\Fin[n]$.
-      \item In~\Cref{sec:finite}, we proceed to give a presentation of the permutation group, as the symmetric group
+      \item In~\cref{sec:finite}, we proceed to give a presentation of the permutation group, as the symmetric group
             $\Sn$ with generators and relations, and solve its word problem. In particular, we present $\Sn$ as a
@@
             $\Aut[\Fin[n]]$ given by the Lehmer encode-decode algorithm.
-      \item In~\Cref{sec:equivalence}, we show how to interpret the language $\PiLang$ into the groupoid $\UFin$, in
+      \item In~\cref{sec:equivalence}, we show how to interpret the language $\PiLang$ into the groupoid $\UFin$, in
             stages. First we define a subset $\PiPlusLang$ of the language which only includes the additive monoidal
@@
             adequacy of this model with respect to the operational semantics.
-      \item In~\Cref{sec:applications}, we show applications of our results to reversible circuits, using our
+      \item In~\cref{sec:applications}, we show applications of our results to reversible circuits, using our
             formalisation. Our results are stated using HoTT~\cite{univalentfoundationsprogramHomotopyTypeTheory2013},

### 2021-10-27 Pi/paper/univalentOLD.tex
Use uniform style for references

@@
 \vc{I'm explaining everything starting from the categorical language. We do care more about the groupoid structure of
-  types, and how to encode groupoids using types, since that's what we use in~\Cref{sec:finite}.}
+  types, and how to encode groupoids using types, since that's what we use in~\cref{sec:finite}.}
 
@@
   subuniverse of types that are merely equal to $T$ is given by $\BAut[T] \defeq \Sub{T}$. This turns out to be the
-  classifying space of $\Aut[T]$, as justified by~\Cref{lem:loop-deloop}. We write $T_{o} \defeq (T, \trunc{\refl_{T}})$
+  classifying space of $\Aut[T]$, as justified by~\cref{lem:loop-deloop}. We write $T_{o} \defeq (T, \trunc{\refl_{T}})$
   for the image of the inclusion of $T$.

### 2021-10-27 Pi/paper/discussion.tex
Add citation for full univalence from naive
Fixes #144

--- 2021-09-05 Pi+/paper/discussion.tex
@@
 formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
-$(A \eqv B) \eqv (A \id_{\UU} B)$.
+$(A \eqv B) \eqv (A \id_{\UU} B)$~\cite[Theorem 3.5]{ortonDecomposingUnivalenceAxiom2018}.
 % Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
@@
 by~\citet{piceghelloCoherenceMonoidalGroupoids2020}. Free commutative monoids in type theory have been studied by
-\citet{gylterudMultisetsTypeTheory2020}, and using HoTT by \citet{choudhuryFinitemultisetConstructionHoTT2019}. The free
-symmetric monoidal groupoid $\FSM[A]$ over a groupoid $A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be
-presented as an algebraic 2-theory using 1-HITs. These HITs and the proof of their universal property have been
-considered by \citet{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}. The proof of
-the universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal
-categories, and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The
-existence of the proof is folklore. We have produced a proof and formalised it in constructive type theory. A stronger
-universal property that $\BFin$ is biinitial in $\term{RigCat}$, is in~\cite{elguetaGroupoidFiniteSets2021}.
+\citet{gylterudMultisetsTypeTheory2020}, and in HoTT by
+\citet*{choudhuryFinitemultisetConstructionHoTT2019,choudhuryFreeCommutativeMonoids2021}. The free symmetric monoidal
+groupoid $\FSM[A]$ over a groupoid $A$ can be given by $\dsum*{X:\UFin}{A^{X}}$, or it can be presented as an algebraic
+2-theory using 1-HITs. These HITs and the proof of their universal property have been considered by
+\citet*{piceghelloCoherenceSymmetricMonoidal2019,choudhuryFinitemultisetConstructionHoTT2019}. The proof of the
+universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal categories,
+and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The existence of
+the proof is folklore. We have produced a proof and formalised it in constructive type theory. A stronger universal
+property that $\BFin$ is biinitial in $\term{RigCat}$, is in~\cite{elguetaGroupoidFiniteSets2021}.
 

### 2021-10-28 Pi/paper/ufin.tex
Citation updates
Use uniform style for references
Add signatures for funext/ua
Add a line about equalities

@@
 
-%   \note{We use the rig structure of $\UFin$ in~\Cref{subsec:rig} to interpret $\PiLang$.}
+%   \note{We use the rig structure of $\UFin$ in~\cref{subsec:rig} to interpret $\PiLang$.}
 
@@
 quotients. We write $\dfun*{x:A}{B(x)}$ or simply $(x:A) \to B(x)$ for dependent function types, and $\dsum*{x:A}{B(x)}$
-or simply $(x:A) \times B(x)$ for dependent sum types.
+or simply $(x:A) \times B(x)$ for dependent sum types. We use $\defeq$ for definitions, $\equiv$ for definitional
+equalities, and $\id$ for the identity type.
 
@@
 \id_{\UU} B) \eqv (A \eqv B)$, which is the univalence principle.
+\begin{gather*}
+  \begin{aligned}
+    \term{happly} & : (f \id g) \to (f \htpy g) \\
+    \term{funext} & : (f \htpy g) \to (f \id g)
+  \end{aligned}
+  \qquad
+  \begin{aligned}
+    \term{idtoeqv} & : (A \id B) \to (A \eqv B) \\
+    \term{ua}      & : (A \eqv B) \to (A \id B)
+  \end{aligned}
+\end{gather*}
 
@@
 point $x : A$, the type $P(x)$ is the fiber over $x$.  The $\term{transport}$ operation, named
-$\term{transport}/\term{tr}$, lifts paths in the indexing type to functions between fibers.
+$\term{transport}$, lifts paths in the indexing type to functions between fibers.
 \begin{gather*}
@@
   \]
+  shown in~\cref{fig:lift}.
 

### 2021-10-28 Pi/paper/discussion.tex
Miscellaneous changes

@@
 theorem~\cite{maclaneNaturalAssociativityCommutativity1963,joyalBraidedTensorCategories1993,gurskiInfiniteLoopSpaces2013}
-for monoidal categories is a particular one. The use of rewriting and proof assistants to prove coherence theorems
-for higher categorical structures has a long history, see~\cite*{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.
+for monoidal categories is a particular one. The use of rewriting and proof assistants to prove coherence theorems for
+higher categorical structures has a long history,
+see~\cite{forestCoherenceGrayCategories2018,beylinExtractingProofCoherence1996}.
 
@@
 By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
-via a complete equational axiomatisation of the equivalences between types in the subuniverse.
+via a complete equational axiomatisation of the equivalences between types in the subuniverse. Univalent
+typoids~\cite{petrakisUnivalentTypoids2019} are another way of axiomatising univalent subuniverses.
+
 % By the property of being univalent, this subuniverse gives a model of the univalence axiom.
@@
 that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
-2-combinators). Thus, our results suggest a new computational interpretation of the univalence principle, which
-provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,
-different approaches to computing with univalence,
-in~\cite*{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in
-Cubical Type
-Theory\cite*{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.
+2-combinators). Thus, our results suggest a new computational interpretation of the univalence principle, which provides
+an intuition on why certain constructions are hard (or impossible in the general case). There are other, different
+approaches to computing with univalence,
+in~\cite{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in Cubical
+Type Theory\cite{angiuliComputationalSemanticsCartesianCubical2019,vezzosiCubicalAgdaDependently2019}.
 

### 2021-10-28 Pi/paper/examples.tex
Change size of qiskit figures
Use qiskit svg figures
Miscellaneous changes

@@
 and~$\underline{\vee}$ is exclusive-disjunction. The circuits are presented in both the textual interface \verb|qasm|
-and the graphical interface in~\cref{fig:qiskit}.
+and the graphical interface, in~\cref{fig:qiskit}.
 
@@
   \begin{tabular}{c@{\qquad}c}
-    \begin{minipage}[t]{0.42\linewidth}
-      \begin{verbatim}
+    \begin{minipage}[t]{0.45\linewidth}
+      \begin{BVerbatim}[fontsize=\small]
         ccx q[1], q[2], q[0];
@@
         cx  q[2], q[0];
-  \end{verbatim}
-      \includegraphics[scale=0.7]{reversibleOr.png}
+  \end{BVerbatim}
+      \inkscape[0.5]{qis1}
     \end{minipage}
      &
-    \begin{minipage}[t]{0.43\linewidth}
-      \begin{verbatim}
+    \begin{minipage}[t]{0.45\linewidth}
+      \begin{BVerbatim}[fontsize=\small]
         cx  q[1], q[0];
@@
         x   q[1];
-  \end{verbatim}
-      \includegraphics[scale=0.6]{reversibleOr2.png}
+  \end{BVerbatim}
+      \inkscape[0.5]{qis2}
     \end{minipage}
@@
 that analyzes the recursive structure of $\mathit{reversibleOr}$ (and would generalise to computing the disjunction of
-more than two inputs):
+more than two inputs).
 
@@
 Although rather trivial, the examples above illustrate the general idea. A more interesting example would be the
-classical core of Shor's algorithm which requires a circuit implementing modular exponentiation $f(r) = a^{r} \mod N$
-for fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{Toffoli:1980}, a reversible specification of this
-function is $f^R(r,h) = (r, f(r)~\underline{\vee}~ h)$ where the exclusive disjunction~$\underline{\vee}$ is applied
-bitwise. When $h=0$ the second component of the result of $f^R$ is the result of the original modular exponentiation
-$f$.
+classical core of Shor's algorithm which requires a circuit implementing modular exponentiation $f(r) =
+a^{r}~\mathrm{mod}~N$ for fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{Toffoli:1980}, a reversible
+specification of this function is $f^R(r,h) = (r, f(r)~\underline{\vee}~ h)$ where the exclusive
+disjunction~$\underline{\vee}$ is applied bitwise. When $h=0$ the second component of the result of $f^R$ is the result
+of the original modular exponentiation $f$.
 %%

### 2021-10-28 Pi/paper/ufin.tex
Miscellaneous changes

@@
 
-In categorical language, the setting for the semantics in the previous section is the category of finite sets and
+In categorical language, the target for the semantics in the previous section is the category of finite sets and
 functions $\SetFin$. However, as $\PiLang$ only refers to bijective functions, a more precise setting is the groupoid
@@
 
-The semantics interprets types of $\PiLang$ as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of 1-combinators
-related by a 2-combinator, their interpretations in~$\BFin$ are equal.  The groupoid $\BFin$ is strict, since the
-collection of isomorphisms is a set, that is, a discrete category. There is no explicit witness for the equality of two
-isomorphisms, since we can decide by evaluating two bijections whether they are equal. To be able to establish
-completeness for $\PiLang$, we want a witness for this equality, so that we can quote back to the syntax and produce a
-2-combinator witnessing the equality of the corresponding 1-combinators.
+The semantics interprets types of $\PiLang$ as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of
+1-combinators related by a 2-combinator, their interpretations in~$\BFin$ are equal. $\BFin$ is a 1-groupoid, where the
+collection of isomorphisms is a set. We can evaluate two bijections on their inputs to decide if they are equal, and
+there is no higher cell witnessing the equality of two isomorphisms. To be able to establish completeness for $\PiLang$,
+we want a witness for this equality, so that we can quote back to the syntax and produce a 2-combinator witnessing the
+equality of the corresponding 1-combinator.
 
@@
 We observe that the implicit equalities between the isomorphisms are pointwise equalities of functions, that is,
-homotopies. We therefore \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible cells.
-We work in HoTT (Univalent Foundations) as it provides a proof-relevant, constructive metatheory to get a handle
-on these equalities and provides a rich internal language for describing weak groupoids, using the ``types are weak
-$\infty$-groupoids'' correspondence.
+homotopies. We therefore \emph{weaken} the groupoid $\BFin$, exposing these homotopies, by using higher invertible
+cells. We work in HoTT/UF as it provides a proof-relevant, constructive metatheory to get a handle on these equalities
+and provides a rich internal language for describing weak groupoids, using the ``types are weak $\infty$-groupoids''
+correspondence.
 
@@
 ${\pi_1 : \dsum*{x:A}{P(x)} \to A}$ from the total space to the base space $A$ has the structure of a fibration, that
-is, there is a lifting operation (\cref{fig:lift} in the appendix) which lifts paths in the base space to paths in the
-total space.
+is, there is a lifting operation (see~\cref{fig:lift} in the appendix) which lifts paths in the base space to paths in
+the total space.
 
@@
 
-\noindent The type families (or fibrations) we are interested in are the ones where paths in the base space completely
-determine the equivalences in the fibers -- these are called univalent
+\noindent
+The type families (or fibrations) we are interested in are the ones where paths in the base space completely determine
+the equivalences in the fibers -- these are called univalent
 fibrations~\cite*{kapulkinUnivalenceSimplicialSets2012,kapulkinSimplicialModelUnivalent2021,christensenCharacterizationUnivalentFibrations2015}.
@@
 
-Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes a set $A$ and a binary relation $R : {A \to A \to
-  \UU}$. It has a mapping of points $\quotinc : A \to \quot{A}{R}$, and a constructor that adds paths between related
-pairs of elements $\quotrel : R(x,y) \to \quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$ (see~\cref{def:set-quot}).  We
-recall that the quotient is \emph{effective} if $R$ is a prop-valued equivalence relation, that is, $R(x,y)$ holds iff
-$(q(x) \id_{\quot{A}{R}} q(y))$.
+Another HIT that we use is the set-quotient $\quot{A}{R}$ which takes a set $A$ and a binary relation
+$R : {A \to A \to \UU}$. It has a mapping of points $\quotinc : A \to \quot{A}{R}$, and a constructor that adds paths
+between related pairs of elements $\quotrel : R(x,y) \to \quotinc(x) \id_{\quot{A}{R}} \quotinc(y)$
+(see~\cref{def:set-quot}). We recall that if $R$ is a prop-valued equivalence relation, then the quotient is
+\emph{effective}, that is, $R(x,y)$ holds iff $(q(x) \id_{\quot{A}{R}} q(y))$.
 
@@
 
-A type is \emph{contractible} (-2-type) if it has a unique element, that is, there is a center of contraction and every
-other point is equal to it. A type is a \emph{proposition} (-1-type) if its equality types are contractible, that is, it
-has at most one inhabitant. Iterating this, we can define sets or 0-types (whose equality types are propositions) and
-1-groupoids or 1-types (whose equality types are sets), and similarly, higher homotopy $n$-types.
+A type is \emph{contractible} (a~-2-type) if it has a unique element, that is, there is a center of contraction and
+every other point is equal to it. A type is a \emph{proposition} (-1-type) if its equality types are contractible, that
+is, it has at most one inhabitant. Iterating this, we can define~\emph{sets} or 0-types (whose equality types are
+propositions) and~\emph{1-groupoids} or 1-types (whose equality types are sets), and similarly, higher homotopy
+$n$-types.
 \begin{gather*}
@@
 \noindent
-Given a function $f : A \to B$, there is an associated fibration over $B$, given by the homotopy fiber. The image of $f$
+Given a function $f : A \to B$, there is an associated fibration over $B$, given by its homotopy fiber. The image of $f$
 is defined to be the (-1)-truncation of its fiber.
@@
 
+\noindent
+Finally, we recall a few characterisations of equivalences of homotopy types.
 \begin{proposition}
   \label{prop:three-chars-equiv}
-  The following propositions are equivalent.
+  The following are equivalent.
   \begin{enumerate}
@@
   proposition for every $X$. Given such a predicate $P$, the fibration ${\fst : \dsum*{X:\UU}{P(X)} \to \UU}$ is
-  univalent and generates a univalent subuniverse ${\UU_{P} \defeq (\dsum*{X:\UU}{P(X)}, \fst)}$.~\footnote{Univalent
-    typoids~\cite{petrakisUnivalentTypoids2019} are a different presentation of univalent subuniverses.}
+  univalent and generates a univalent subuniverse ${\UU_{P} \defeq (\dsum*{X:\UU}{P(X)}, \fst)}$.
 \end{propositionrep}
@@
   \end{gather*}
+\end{definition}
+
+\noindent
 Note that $\Fin[n]$ is a set, and we use both definitions interchangeably.
-\end{definition}
 
@@
   \]
+\end{definition}
+
+\noindent
 Note that the natural number $n$ need not be truncated, as justified below.
-\end{definition}
 
@@
 
-Since $\isFin$ is a predicate on the universe $\UU$, we easily get our univalent subuniverse $\UFin$.
+\noindent
+Since $\isFin$ is a predicate on the universe $\UU$, we easily get our univalent subuniverse $\UFin$. This definition of
+the groupoid of finite types has also been considered in~\cite{yorgeyCombinatorialSpeciesLabelled2014}.
 
@@
 
-This definition of the groupoid of finite types has also been considered
-in~\cite{yorgeyCombinatorialSpeciesLabelled2014}. While $\UFin$ has \emph{all} the finite types, we are also interested
-in constructing a subuniverse of finite types of a specified cardinality. To do so, we will start with the subuniverse
-$\BAut[T]$, for any type $T : \UU$. \footnote{Characterisations of univalent fibrations using the $\BAut$ construction
-  have been studied by~\citet{christensenCharacterizationUnivalentFibrations2015}.}
+While $\UFin$ has \emph{all} the finite types, we are also interested in constructing a subuniverse of finite types of a
+specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$, for any type $T : \UU$.
+Characterisations of univalent fibrations using the $\BAut$ construction have also been studied
+by~\citet{christensenCharacterizationUnivalentFibrations2015}.
 
@@
 subuniverse of 2-element sets is given by $\BAut[\Bool]$. This has been used to construct the real projective spaces in
-HoTT~\cite{buchholtzRealProjectiveSpaces2017}, and also to give the denotational semantics for a 1-bit reversible
-programming language~\cite{caretteReversibleProgramsUnivalent2018}.
+HoTT~\cite{buchholtzRealProjectiveSpaces2017}, and also to give sound and complete denotational semantics to a 1-bit
+fragment of $\PiLang$~\cite*{caretteReversibleProgramsUnivalent2018}.
 
@@
 We have shown that loops in $\UFin$ exactly encode the automorphism group $\Aut[\Fin[n]]$ for every~$n$. This is a
-general technique called \emph{delooping}, where a group can be identified with a 1-object groupoid, internally in HoTT.
-This technique also allows defining higher groups~\cite{buchholtzHigherGroupsHomotopy2018}. The loopspace of a pointed
-type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element, path composition for the
-group multiplication, and path inverse for the group inverse operation. The group axioms are given by the higher paths
-corresponding to groupoid laws.
+general technique called \emph{delooping}, where a group can be identified with a 1-object groupoid, that is, a
+0-connected type in HoTT. This technique also allows defining higher groups~\cite{buchholtzHigherGroupsHomotopy2018}.
+The loopspace of a pointed type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element,
+path composition for the group multiplication, and path inverse for the group inverse operation. The group axioms are
+given by the higher paths corresponding to groupoid laws.
 
@@
       \Fin[0]                & \eqv \bot                  \\
-      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]           \\
-      \\
+      \Fin[n] \sqcup \Fin[m] & \eqv \Fin[n + m]           \\[1.5ex]
       \bot \sqcup X          & \eqv X                     \\
@@
       \Fin[1]                & \eqv \top                             \\
-      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]                      \\
-      \\
+      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]                      \\[1.5ex]
       \top \times X          & \eqv X                                \\

### 2021-10-28 Pi/paper/finite.tex

@@
 allows us to connect it to the broader scope of computational group theory and combinatorics -- we can borrow ideas such
-as Coxeter relations and Lehmer codes.~\footnote{The concepts we use can be found in any standard textbook on group
-theory, or see~\cite{symmetryBook2021} for a univalent point of view.}
+as Coxeter relations and Lehmer codes. The concepts we use can be found in any standard textbook on group theory, or see
+the Symmetry book~\cite{symmetryBook2021} for a univalent point of view.
 
@@
 
+\noindent
 This construction is called a Coxeter presentation of $\Sn$. Writing it formally, we encode the rules discussed above
@@
 
-Given a permutation $\sigma : \Aut[\Fin[n]]$, for any element $i: \Fin[n]$, we can define the inversion count of $i$ as
-the number of smaller elements appearing after it in the permutation.
+The inversion count of an element is the number of smaller elements appearing after it in the permutation.
 
 \begin{definition}[Inversion count]
-  Given a permutation $\sigma : \Aut[\Fin[n]]$, the inversion count of $i: \Fin[n]$ is given by
-  \[ \invcount{i} = \#\Set{j < i | \sigma(j) > \sigma(i)}. \]
+  Given a permutation $\sigma : \Aut[\Fin[n]]$, the inversion count of an element $i: \Fin[n]$ is given by
+  \[ \invcount{i} \defeq \#\Set{j < i | \sigma(j) > \sigma(i)}. \]
 \end{definition}
 
-Just knowing the inversion counts for all the elements, one can reconstruct the starting permutation. Also, observe that
+Knowing the inversion counts for all the elements, one can reconstruct the starting permutation. Also, observe that
 $\invcount{i} < i$, thus fitting in the $i$-th place of a Lehmer code tuple. As an example, consider the following
@@
 \[
-  \sigma =
+  \sigma \defeq
   \begin{pmatrix}
@@
 \[
-  l = (\invcount{0}, \invcount{1}, \invcount{2}, \invcount{3}, \invcount{4}) = (0, 1, 2, 0, 2)
+  l \equiv (\invcount{0}, \invcount{1}, \invcount{2}, \invcount{3}, \invcount{4}) \equiv (0, 1, 2, 0, 2)
 \]
@@
 between finite sets, we need to encode it as a Lehmer code. We will prove that these maps construct an equivalence
-between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$. The idea for this proof is borrowed
-from~\cite{Molzer-cubical}. ~\footnote{Note that the indices for the type of permutations are off-by-one, because we
-  chose $\Fin[n]$ to represent generators for permutations on $\Fin[\suc[n]]$.}
+between the types $\Lehmer[n]$ and $\Aut[\Fin[\suc[n]]]$~\footnote{Note that the indices for the type of permutations
+  are off-by-one, because we chose $\Fin[n]$ to represent generators for permutations on $\Fin[\suc[n]]$.}. The idea for
+this proof is borrowed from~\citet{Molzer-cubical}.
 
@@
 \begin{proof}
-  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\suc[\zero]]]$. For $n = \suc[m]$,
-  we compute a chain of equivalences.
-  \begin{gather*}
-    \arraycolsep=0.5em\def\arraystretch{1.5}
-    %% \begin{array}{rl}
-    %%        & \Lehmer[\zero]          \\
-    %%   \eqv & \unit                   \\
-    %%   \eqv & \Aut[\Fin[\suc[\zero]]]
-    %% \end{array}
-    %% \qquad\qquad
+  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\suc[\zero]]]$, hence we have that
+  ${\Lehmer[\zero] \eqv \unit \eqv \Aut[\Fin[\suc[\zero]]]}$. For $n = \suc[m]$, we compute the following chain of equivalences.
+  \[
+    \arraycolsep=0.5em\def\arraystretch{1.3}
     \begin{array}{rlr}
@@
       \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                              \\
-      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
-      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
-      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                           & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}}
+      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\cref{prop:fin-finexcept}~part~\ref{prop:fin-finexcept-2}} \\
+      \eqv & \dsum*{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~part~\ref{prop:fin-finexcept-2}} \\
+      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                           & \text{by~\cref{prop:fin-finexcept}~part~\ref{prop:fin-finexcept-3}}
     \end{array}
-  \end{gather*}
+  \]
 \end{proof}

### 2021-10-28 Pi/paper/introduction.tex

@@
             verifying that programs identified by level-2 constructs denote the same bijection.
-      \item ~\cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in
-            Homotopy Type Theory. We define and characterise the notion of a univalent subuniverse, and construct
+      \item \cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in
+            Homotopy Type Theory and Univalent Foundations. We define and characterise the notion of a univalent subuniverse, and construct
             $\UFin$ as a univalent subuniverse which classifies all finite types. We establish that paths in $\UFin$ are
@@
       \item In~\cref{sec:applications}, we show applications of our results to reversible circuits, using our
-            formalisation. Our results are stated using HoTT~\cite{univalentfoundationsprogramHomotopyTypeTheory2013},
+            formalisation. Our results are stated using HoTT/UF~\cite{univalentfoundationsprogramHomotopyTypeTheory2013},
             and formalised using the HoTT-Agda library (around 7,500 lines of code). Using the formalisation, we are
@@
 The proofs of some of our lemmas and propositions and theorems, as well as additional material, are given in the
-supplementary appendices in the extended version~\cite{choudhurySymmetriesReversibleProgramming2021}, and we refer to
-them in the text. Our accompanying Agda code~\cite{choudhuryArtifactSymmetriesReversible2021} contains the formalisation
-of the full syntax, and most of our proofs.
+supplementary appendices in the extended version~\cite*{choudhurySymmetriesReversibleProgramming2021}, and we refer to
+them in the text. Our accompanying Agda code~\cite*{choudhuryArtifactSymmetriesReversible2021} contains the
+formalisation of the full syntax, and most of our proofs.
 

### 2021-10-28 Pi/paper/reversible1.tex
Use uniform style for references
Move macros to syntax.sty

@@
-\begin{figure}[t]
-  {\scalebox{\scalef}{$%
-        %%\noindent\begin{minipage}{.7\linewidth}
-        \begin{array}{rrcll}
-          \idc :     & A                     & \isoone & A                            & : \idc      \\
-          \\
-          \identlp : & \zerot + A            & \isoone & A                            & : \identrp  \\
-          \swapp :   & A + B                 & \isoone & B + A                        & : \swapp    \\
-          \assoclp : & A + (B + C)           & \isoone & (A + B) + C                  & : \assocrp  \\ [1.5ex]
-          \identlt : & \onet \times A        & \isoone & A                            & : \identrt  \\
-          \swapt :   & A \times B            & \isoone & B \times A                   & : \swapt    \\
-          \assoclt : & A \times (B \times C) & \isoone & (A \times B) \times C        & : \assocrt  \\ [1.5ex]
-          \absorbr : & ~ \zerot \times A     & \isoone & \zerot ~                     & : \factorzl \\
-          \dist :    & ~ (A + B) \times C    & \isoone & (A \times C) + (B \times C)~ & : \factor
-        \end{array}$}}
-
-  \medskip
-
-  {\scalebox{\scalef}{%
-      \Rule{}
-      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : B \isoone C}
-      {\jdg{}{}{c_1 \circledcirc c_2 : A \isoone C}}
-      {}
-
-      \Rule{}
-      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : C \isoone D}
-      {\jdg{}{}{c_1 \oplus c_2 : A + C \isoone B + D}}
-      {}
-
-      \Rule{}
-      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : C \isoone D}
-      {\jdg{}{}{c_1 \otimes c_2 : A \times C \isoone B \times D}}
-      {}
-    }}
-  \caption{$\Pi$-terms, combinators, and their types.}
-  \label{fig:pi-terms}
-\end{figure}
-
 \section{A Reversible Programming Language}
@@
 commutative rig), with $(0,+)$ for addition, and $(1,\times)$ for multiplication then provides sequential, vertical, and
-horizontal circuit composition operators.
+horizontal circuit composition.
 
 These commutative rig identities can be used to design a logic for reversible
-programming~\cite*{sparksSuperstructuralReversibleLogic2014}. To interpret natural number identities as reversible programs,
-the logic needs to be equipped with values and types, and a notion of operational semantics and contextual equivalence,
-giving a computational interpretation of the commutative rig structure~\cite{jamesInformationEffects2012}. On the
-semantic side, the natural space to consider is the groupoidification of a commutative rig, that is, a symmetric rig
-groupoid.
+programming~\cite*{jamesInformationEffects2012,sparksSuperstructuralReversibleLogic2014}. To interpret natural number
+identities as reversible programs, the logic needs to be equipped with values and types, and a notion of operational
+semantics and contextual equivalence, giving a computational interpretation of the commutative rig
+structure~\cite{jamesInformationEffects2012}. On the semantic side, the natural space to consider is the
+groupoidification of a commutative rig, that is, a symmetric rig groupoid.
 
@@
 
-% In programming parlance, as shown in~\Cref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes
+% In programming parlance, as shown in~\cref{fig:pi-terms}, each primitive algebraic identity of commutative rigs becomes
 % a reversible combinator in the programming language and the algebraic closure operators become composition operators in
@@
         \textit{Program types} &         &     & A \isoone B                                      \\
-        \textit{Programs}      & c       & ::= & (\textrm{See Fig.~\ref{fig:pi-terms}})
+        \textit{Programs}      & c       & ::= & (\textrm{See~\cref{fig:pi-terms}})
       \end{array}$}}
+
+\begin{figure}
+  {\scalebox{\scalef}{$%
+        %%\noindent\begin{minipage}{.7\linewidth}
+        \begin{array}{rrcll}
+          \idc :     & A                     & \isoone & A                            & : \idc      \\ [1.5ex]
+          \identlp : & \zerot + A            & \isoone & A                            & : \identrp  \\
+          \swapp :   & A + B                 & \isoone & B + A                        & : \swapp    \\
+          \assoclp : & A + (B + C)           & \isoone & (A + B) + C                  & : \assocrp  \\ [1.5ex]
+          \identlt : & \onet \times A        & \isoone & A                            & : \identrt  \\
+          \swapt :   & A \times B            & \isoone & B \times A                   & : \swapt    \\
+          \assoclt : & A \times (B \times C) & \isoone & (A \times B) \times C        & : \assocrt  \\ [1.5ex]
+          \absorbr : & ~ \zerot \times A     & \isoone & \zerot ~                     & : \factorzl \\
+          \dist :    & ~ (A + B) \times C    & \isoone & (A \times C) + (B \times C)~ & : \factor
+        \end{array}$}}
+
+  \medskip
+
+  {\scalebox{\scalef}{%
+      \Rule{}
+      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : B \isoone C}
+      {\jdg{}{}{c_1 \circledcirc c_2 : A \isoone C}}
+      {}
+
+      \Rule{}
+      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : C \isoone D}
+      {\jdg{}{}{c_1 \oplus c_2 : A + C \isoone B + D}}
+      {}
+
+      \Rule{}
+      {\jdg{}{}{c_1 : A \isoone B} \quad \vdash c_2 : C \isoone D}
+      {\jdg{}{}{c_1 \otimes c_2 : A \times C \isoone B \times D}}
+      {}
+    }}
+  \caption{$\Pi$-terms, combinators, and their types.}
+  \label{fig:pi-terms}
+\end{figure}
 
@@
 
-To see how this language expresses reversible circuits, we present a few examples in the Agda embedding of the language.
-First it is possible to directly mimic the \verb|qasm|-perspective by defining types that describe sequences of
-booleans. We use the type $\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing
-\textsf{true} and $\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is
-straightforward to define using the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product
-of boolean values, thus the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$)
-corresponds to a collection of wires that can transmit three bits.
+To see how this language expresses reversible circuits, we present a few examples using the Agda embedding of the
+language~\cite{choudhuryArtifactSymmetriesReversible2021}. First it is possible to directly mimic the
+\verb|qasm|-perspective by defining types that describe sequences of booleans ($\mathbb{2}^{n}$). We use the type
+$\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true}, and
+$\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is straightforward to define
+using the primitive combinator $\swapp$. We can represent $n$-bit words using an n-ary product of boolean values, thus
+the type $\mathbb{2} \times (\mathbb{2} \times \mathbb{2})$ (abbreviated $\mathbb{B}~3$) corresponds to a collection of
+wires that can transmit three bits.
 %
@@
 of using the indices in the sequence of wires to identify the relevant parameters, here we use structural isomorphisms
-to re-shuffle the types. We only show one of these re-shuffling isomorphisms and elide the others. For the second circuit, instead of transcribing it directly, we express it using
-a slightly more abstract notation:
+to re-shuffle the types. We only show one of these re-shuffling isomorphisms and elide the others. For the second
+circuit, instead of transcribing it directly, we express it using $\Afun{cif}$.
 
@@
 
-\noindent Like the original circuit, we examine the bit at index 1 (corresponding to the component $B$ in a tuple
-$(A,(B,C))$): if the bit is true, we perform an \verb|x| operation on component $A$, and otherwise we perform a
-\verb|cx| operation on $(C,A)$. The two uses of \verb|x| gates in the circuit are now unnecessary as they were only needed
-to encode a two-way conditional expression using a sequence of one-way conditional expressions (the only ones available in
-the linear circuit model).
+\noindent
+Like the original circuit, we examine the bit at index 1 (corresponding to the component $B$ in a tuple $(A,(B,C))$): if
+the bit is true, we perform an \verb|x| operation on component $A$, and otherwise we perform a \verb|cx| operation on
+$(C,A)$. The two uses of \verb|x| gates in the circuit are now unnecessary as they were only needed to encode a two-way
+conditional expression using a sequence of one-way conditional expressions (the only ones available in the linear
+circuit model).
 
-All of this is only half the story, however. A sound semantics for $\PiLang$ in weak rig
-groupoids was established by \citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
+All of this is only half the story, however. A sound semantics for $\PiLang$ using weak rig
+groupoids was established by~\citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
 semantics, \emph{coherence conditions} for symmetric rig groupoids that identify different syntactic representations of the same
@@
 
-\begin{multicols}{2}
-  %%  \begin{longtable}{>{$}r<{$} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
-  {\scalebox{\scalef}{$%
+\begin{center}
+\begin{tabular}[t]{c @{\hspace{5em}} c}
+  {\scalebox{\scalef}{
       \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
-          \pdenot{\identlp} & (\inl{v})         & = & v               \\
-          \pdenot{\identrp} & v                 & = & \inl{v}         \\
-          \pdenot{\swapp}   & (\inl{v})         & = & \inr{v}         \\
-          \pdenot{\swapp}   & (\inr{v})         & = & \inl{v}         \\
-          \pdenot{\assoclp} & (\inl{v})         & = & \inl{(\inl{v})} \\
-          \pdenot{\assoclp} & (\inr{(\inl{v})}) & = & \inl{(\inr{v})} \\
-          \pdenot{\assoclp} & (\inr{(\inr{v})}) & = & \inr{v}         \\
-          \pdenot{\assocrp} & (\inl{(\inl{v})}) & = & \inl{v}         \\
-          \pdenot{\assocrp} & (\inl{(\inr{v})}) & = & \inr{(\inl{v})} \\
-          \pdenot{\assocrp} & (\inr{v})         & = & \inr{(\inr{v})}
-        \end{tabular}$}}
-
-  {\scalebox{\scalef}{$%
+        \denotbox{\identlp} & (\inl{v})         & = & v               \\
+        \denotbox{\identrp} & v                 & = & \inl{v}         \\
+        \denotbox{\swapp}   & (\inl{v})         & = & \inr{v}         \\
+        \denotbox{\swapp}   & (\inr{v})         & = & \inl{v}         \\
+        \denotbox{\assoclp} & (\inl{v})         & = & \inl{(\inl{v})} \\
+        \denotbox{\assoclp} & (\inr{(\inl{v})}) & = & \inl{(\inr{v})} \\
+        \denotbox{\assoclp} & (\inr{(\inr{v})}) & = & \inr{v}         \\
+        \denotbox{\assocrp} & (\inl{(\inl{v})}) & = & \inl{v}         \\
+        \denotbox{\assocrp} & (\inl{(\inr{v})}) & = & \inr{(\inl{v})} \\
+        \denotbox{\assocrp} & (\inr{v})         & = & \inr{(\inr{v})}
+      \end{tabular}}}
+  &
+  {\scalebox{\scalef}{
       \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
-          \pdenot{\identlt} & (\ttt , v)          & = & v                   \\
-          \pdenot{\identrt} & v                   & = & (\ttt , v)          \\
-          \pdenot{\swapt}   & (v_1 , v_2)         & = & (v_2 , v_1)         \\
-          \pdenot{\assoclt} & (v_1 , (v_2 , v_3)) & = & ((v_1 , v_2) , v_3) \\
-          \pdenot{\assocrt} & ((v_1 , v_2) , v_3) & = & (v_1 , (v_2 , v_3)) \\
-          \pdenot{\dist}    & (\inl{v_1} , v_3)   & = & \inl{(v_1 , v_3)}   \\
-          \pdenot{\dist}    & (\inr{v_2 , v_3})   & = & \inr{(v_2 , v_3)}   \\
-          \pdenot{\factor}  & (\inl{(v_1 , v_3)}) & = & (\inl{v_1} , v_3)   \\
-          \pdenot{\factor}  & (\inr{(v_2 , v_3)}) & = & (\inr{v_2} , v_3)   \\
-          \pdenot{\idc}     & v                   & = & v
-        \end{tabular}$}}
-  %%\end{longtable}
-\end{multicols}
+        \denotbox{\identlt} & (\ttt , v)          & = & v                   \\
+        \denotbox{\identrt} & v                   & = & (\ttt , v)          \\
+        \denotbox{\swapt}   & (v_1 , v_2)         & = & (v_2 , v_1)         \\
+        \denotbox{\assoclt} & (v_1 , (v_2 , v_3)) & = & ((v_1 , v_2) , v_3) \\
+        \denotbox{\assocrt} & ((v_1 , v_2) , v_3) & = & (v_1 , (v_2 , v_3)) \\
+        \denotbox{\dist}    & (\inl{v_1} , v_3)   & = & \inl{(v_1 , v_3)}   \\
+        \denotbox{\dist}    & (\inr{v_2 , v_3})   & = & \inr{(v_2 , v_3)}   \\
+        \denotbox{\factor}  & (\inl{(v_1 , v_3)}) & = & (\inl{v_1} , v_3)   \\
+        \denotbox{\factor}  & (\inr{(v_2 , v_3)}) & = & (\inr{v_2} , v_3)   \\
+        \denotbox{\idc}     & v                   & = & v
+      \end{tabular}}}
+\end{tabular}
+\end{center}
 
 \begin{center}
-  {\scalebox{\scalef}{$%
+  {\scalebox{\scalef}{
       \begin{tabular}[t]{>{$}r<{$} @{\hspace{4pt}} >{$}l<{$} >{$}c<{$} >{$}l<{$}}
-          \pdenot{(c_1 \circledcirc c_2)} & v           & = & \denot{c_2} (\denot{c_1}~v)         \\
-          \pdenot{(c_1 \oplus c_2)}       & (\inl{v})   & = & \inl{(\denot{c_1}~v)}               \\
-          \pdenot{(c_1 \oplus c_2)}       & (\inr{v})   & = & \inr{(\denot{c_2}~v)}               \\
-          \pdenot{(c_1 \otimes c_2)}      & (v_1 , v_2) & = & (\denot{c_1}~v_1 , \denot{c_2}~v_2)
-        \end{tabular}
-      $}}
+        \denotbox{(c_1 \circledcirc c_2)} & v           & = & \denot{c_2} (\denot{c_1}~v)         \\
+        \denotbox{(c_1 \oplus c_2)}       & (\inl{v})   & = & \inl{(\denot{c_1}~v)}               \\
+        \denotbox{(c_1 \oplus c_2)}       & (\inr{v})   & = & \inr{(\denot{c_2}~v)}               \\
+        \denotbox{(c_1 \otimes c_2)}      & (v_1 , v_2) & = & (\denot{c_1}~v_1 , \denot{c_2}~v_2)
+      \end{tabular}}}
 \end{center}
@@
     \item For every pair of combinators $c_1$ and $c_2$ of the same type $A \isoone B$, if there exists a level-2
-          combinator $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$ using extensional equivalence of
+          combinator $\alpha : c_1 \isotwo c_2$, then $\denot{c_1} = \denot{c_2}$, using extensional equivalence of
           functions.
@@
 \begin{proof}
-  For every primitive combinator $c$ listed on one side of Fig.~\ref{fig:pi-terms}, let $!c$ be the combinator listed on
+  For every primitive combinator $c$ listed on one side of~\cref{fig:pi-terms}, let $!c$ be the combinator listed on
   the other side. Thus $! \assoclp$ is $\assocrp$ and $! \swapp$ is $\swapp$ itself. Then we have that $\denot{c}$ and

### 2021-10-29 Pi/paper/applications.tex
Reorganise applications section

--- 2021-10-27 Pi/paper/equivalence.tex
@@
-\section{Correspondence between \texorpdfstring{$\PiLang$}{Pi} and \texorpdfstring{$\UFin$}{UFin}}
-\label{sec:equivalence}
-
-In this section, we first translate $\PiLang$ to its additive fragment $\PiPlusLang$. This is the language that we
-interpret to $\UFin$ and back, using the tools developed in the previous sections. Further, we go through an
-intermediate step of the language $\PiHatLang$, which is a simplified variant of $\PiPlusLang$ that uses adjacent
-transpositions for combinators, while preserving all the required structure.
-
-% https://q.uiver.app/?q=WzAsNCxbMCwwLCJcXFBpTGFuZyJdLFsyLDAsIlxcUGlQbHVzTGFuZyJdLFs0LDAsIlxcUGlIYXRMYW5nIl0sWzYsMCwiXFxVRmluIl0sWzAsMSwiXFxldmFsdCJdLFsxLDIsIlxcZXZhbGgiLDAseyJjdXJ2ZSI6LTR9XSxbMiwzLCJcXGV2YWx0IiwwLHsiY3VydmUiOi00fV0sWzIsMSwiXFxxdW90ZWgiLDAseyJjdXJ2ZSI6LTR9XSxbMywyLCJcXHF1b3RldCIsMCx7ImN1cnZlIjotNH1dXQ==
-\[\begin{tikzcd}
-    \PiLang && \PiPlusLang && \PiHatLang && \UFin
-    \arrow["\evalt", from=1-1, to=1-3]
-    \arrow["\evalp", curve={height=-24pt}, from=1-3, to=1-5]
-    \arrow["\evalh", curve={height=-24pt}, from=1-5, to=1-7]
-    \arrow["\quotep", curve={height=-24pt}, from=1-5, to=1-3]
-    \arrow["\quoteh", curve={height=-24pt}, from=1-7, to=1-5]
-  \end{tikzcd}\]
-
-We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\cref*{fig:piplus,fig:pihat} respectively,
-eliding the 2-combinators for brevity. We enforce that there is a unique 2-combinator between compatible 1-combinators,
-by relating them with a truncation. These can be found in~\cref{app:leveltwo} and in the accompanying Agda code.
-
-The translations between the languages are defined separately on types, 1-combinators, and 2-combinators. Following the
-terminology of Normalisation by Evaluation, the translations from the left to the right, going from the syntax towards
-the semantics, are called $\evalt$ and the translations the other way are called $\quotet$.
-
-To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
-category. We define them formally in the appendix, and only state our results here. For each of the $\PiLang$,
-$\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and
-$\PiHatCat$, have 0-cells for types ($U$, $\UPlus$, and $\UHat$), 1-cells for 1-combinators ($\iso$, $\isop$, and
-$\isoh$), and 2-cells for 2-combinators ($\Iso$, $\Isop$, and $\Isoh$). We can show that these syntactic categories here
-are actually $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are also locally-strict, or
-locally-posetal, because there is at most one 2-cell between compatible 1-cells.
-
-\begin{toappendix}
-  \begin{proposition}
-    We can form a weak 2-category $\PiCat$ with
-    \begin{itemize}
-      \item $\PiLang$ types ($U$) for 0-cells,
-      \item for $X, Y : U$, a collection of 1-cells $X \iso Y$,
-      \item for $p, q : X \iso Y$, a collection of 2-cells $p \Iso q$.
-    \end{itemize}
-  \end{proposition}
-
-  \begin{proposition}
-    We can form a weak 2-category $\PiPlusCat$ with
-    \begin{itemize}
-      \item $\PiPlusLang$ types ($\UPlus$) for 0-cells,
-      \item for $X, Y : \UPlus$, a collection of 1-cells $X \isop Y$,
-      \item for $p, q : X \isop Y$, a collection of 2-cells $p \Isop q$.
-    \end{itemize}
-  \end{proposition}
-
-  \begin{proposition}
-    We can form a weak 2-category $\PiHatCat$ with
-    \begin{itemize}
-      \item $\PiHatCat$ types ($\UHat$) for 0-cells,
-      \item for $X, Y : \UHat$, a collection of 1-cells $X \isoh Y$,
-      \item for $p, q : X \isoh Y$, a collection of 2-cells $p \Isoh q$.
-    \end{itemize}
-  \end{proposition}
-\end{toappendix}
-
-We use the $\evalt/\quotet$ translation maps to construct functors between these categories. We only name the maps on
-the 0, 1, and 2-cells -- the coherences hold by definition or by calculation, which is shown in our accompanying Agda
-code. We use these functors to state our results establishing the equivalences between the languages.
-
-\subsection{$\PiLang$ to $\PiPlusLang$}
-
-First, we show how to translate $\PiLang$ programs to $\PiPlusLang$, which is the additive fragment of $\PiLang$. The
-syntax for 1-combinators is given in~\cref{fig:piplus}.
-
-\begin{figure}[t]
-  {\scalebox{\scalef}{$
-        \begin{array}{rrcll}
-          \idc :     & A           & \isop & A           & : \idc     \\
-          \identlp : & \zerot + A  & \isop & A           & : \identrp \\
-          \swapp :   & A + B       & \isop & B + A       & : \swapp   \\
-          \assoclp : & A + (B + C) & \isop & (A + B) + C & : \assocrp \\ [1.5ex]
-        \end{array}$}}
-
-  {\scalebox{\scalef}{
-      \Rule{}
-      {\jdg{}{}{c_1 : A \isop B} \quad \vdash c_2 : B \isop C}
-      {\jdg{}{}{c_1 \circledcirc c_2 : A \isop C}}
-      {}
-
-      \Rule{}
-      {\jdg{}{}{c_1 : A \isop B} \quad \vdash c_2 : C \isop D}
-      {\jdg{}{}{c_1 \oplus c_2 : A + C \isop B + D}}
-      {}
-    }}
-  \caption{$\PiPlusLang$ syntax}
-  \label{fig:piplus}
-\end{figure}
-
-$\PiLang$ has two 0-ary type constructors, and two binary type constructors -- the additive tensor product and the
-multiplicative one. $\PiPlusLang$ has all the type constructors of $\PiLang$ except multiplication. However, we will
-show how to recover the multiplicative structure, by defining multiplication as repeated addition. We encode $\times$ in
-terms of $+$ as follows.
-
-\begin{definition}[$\times : \UPlus \to \UPlus \to \UPlus$]
-  \begin{align*}
-    \zerot \times Y      & \defeq \zerot                      \\
-    \onet \times Y       & \defeq Y                           \\
-    (X_1 + X_2) \times Y & \defeq X_1 \times Y + X_2 \times Y
-  \end{align*}
-\end{definition}
-
-\begin{lemma}
-  There are two symmetric monoidal structures on $\PiPlusCat$, given by $(\zerot, +)$ and $(\onet, \times)$, with
-  $\times$ distributing over $+$, giving it a rig structure.
-\end{lemma}
-
-\noindent Using this rig structure, we translate $\PiLang$ to $\PiPlusLang$, constructing a rig equivalence from
-$\PiCat$ to $\PiPlusCat$.
-
-\begin{definition}[$\evalt$]
-  \begin{align*}
-    \evalt_{0} & : U \to \UPlus                                             \\
-    \evalt_{1} & : (c : X \iso Y) \to \evalt_{0}(X) \iso \evalt_{0}(Y)      \\
-    \evalt_{2} & : (\alpha : p \Iso q) \to \evalt_{1}(p) \Iso \evalt_{1}(q)
-  \end{align*}
-\end{definition}
-
-\begin{theorem}
-  $\evalt$ gives a rig equivalence between $\PiCat$ and $\PiPlusCat$.
-\end{theorem}
-
-\subsection{$\PiPlusLang$ to $\PiHatLang$}
-
-Next, we show how to translate $\PiPlusLang$ programs to $\PiHatLang$ and back. $\PiHatLang$ is a simplified variant of
-$\PiPlusLang$, with (unary) natural numbers for 0-cells, 1-combinators generated by adjacent transpositions, and an
-appropriate set of 2-combinators. We give the syntax, again omitting 2-combinators, in~\cref{fig:pihat}.
-
-\begin{figure}[t]
-  {\scalebox{\scalef}{$
-        \begin{array}{rrcll}
-          \idc :   & n             & \isoh & n             & : \idc   \\
-          \swapc : & \suc[\suc[n]] & \isoh & \suc[\suc[n]] & : \swapc \\
-        \end{array}
-      $}}
-
-  {\scalebox{\scalef}{
-      \Rule{}
-      {\jdg{}{}{c_1 : n \isoh m} \quad \vdash c_2 : m \isoh o}
-      {\jdg{}{}{c_1 \circledcirc c_2 : n \isoh o}}
-      {}
-
-      \Rule{}
-      {\jdg{}{}{c : n \isoh m}}
-      {\jdg{}{}{\oplus(c) : \suc[n] \isoh \suc[m]}}
-      {}
-    }}
-  \caption{$\PiHatLang$ syntax}
-  \label{fig:pihat}
-\end{figure}
-
-As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers, and,
-we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal structure.~\footnote{Since each object
-is a natural number, this makes $\PiHatCat$ a \emph{PROP}, that is, a products and permutations category.}
-
-To produce a braiding $n + m \isoh m + n$ from adjacent tranpositions, we recursively traverse the left subexpression,
-swapping each element using adjacent transpositions along the elements on the right, placing it in the right position.
-The computational content of this translation can be visualised using tree transformations, for the recursive case
-in~\cref{fig:plusplusswap}. The challenging part is showing that these moves are coherent with respect to 2-combinators.
-
-\begin{figure}
-   \[
-      \Tree [ [ {\tiny 1} {\tiny n} ] {\tiny m} ] ~\xrightarrow{}~
-      \Tree [ {\tiny 1} [ {\tiny n} {\tiny m} ] ] ~\xrightarrow{}~
-      \Tree [ [ {\tiny n} {\tiny m} ] {\tiny 1} ] ~\xrightarrow{}~
-      \Tree [ [ {\tiny m} {\tiny n} ] {\tiny 1} ] ~\xrightarrow{}~
-      \Tree [ {\tiny m} [ {\tiny n} {\tiny 1} ] ] ~\xrightarrow{}~
-      \Tree [ {\tiny m} [ {\tiny 1} {\tiny n} ] ] ~
-    \]
-    \caption{Braiding from transpositions, recursive case}
-    \label{fig:plusplusswap}
-\end{figure}
-
-\begin{lemma}
-  $\PiHatCat$ has a symmetric monoidal structure, with the unit given by 0 and the tensor product given by natural
-  number addition.
-\end{lemma}
-
-\noindent Using this symmetric monoidal structure, we translate from $\PiPlusLang$ to $\PiHatLang$.
-
-\begin{definition}[$\evalp$]
-  \begin{align*}
-    \evalp_{0} & : \UPlus \to \UHat                                                           \\
-    \evalp_{1} & : (c : t_{1} \isop t_{2}) \to \evalp_{0}(t_{1}) \isoh \evalp_{0}(t_{2})      \\
-    \evalp_{2} & : (\alpha : c_{1} \Isop c_{2}) \to \evalp_{1}(c_{1}) \Isoh \evalp_{1}(c_{2})
-  \end{align*}
-\end{definition}
-
-To go back from $\PiHatLang$ to $\PiPlusLang$, we turn a natural number into a $\PiPlusLang$ type, using right-biased
-addition, that is, the natural number $n$ gets mapped to the type $\onet + (\onet + (\onet + \ldots + \zerot))$. Since
-the types are already right-biased, an adjacent transposition in $\PiHatLang$ is easily encoded by using the braiding in
-$\PiPlusLang$, as shown in~\cref{fig:transpfrombraid}. Again, these are shown to be coherent.
-
-\begin{figure}
-  \[
-    \Tree [ {\tiny $\onet$} [ {\tiny $\onet$} {\tiny X} ] ] ~\xrightarrow{\assoclp}~
-    \Tree [ [ {\tiny $\onet$} {\tiny $\onet$} ] {\tiny X} ] ~\xrightarrow[\swapp\phantom{xx}]{\phantom{xx}\idc}~
-    \Tree [ [ {\tiny $\onet$} {\tiny $\onet$} ] {\tiny X} ] ~\xrightarrow{\assocrp}~
-    \Tree [ {\tiny $\onet$} [ {\tiny $\onet$} {\tiny X} ] ] ~
-  \]
-  \caption{Transpositions from braiding}
-  \label{fig:transpfrombraid}
-\end{figure}
-
-\begin{definition}[$\quotep$]
-  \begin{align*}
-    \quotep_{0} & : \UHat \to \UPlus                                                            \\
-    \quotep_{1} & : (p : X_{1} \isoh X_{2}) \to \quotep_{0}(X_{1}) \iso \quotep_{0}(X_{2})      \\
-    \quotep_{2} & : (\alpha : p_{1} \Isoh p_{2}) \to \quotep_{1}(p_{1}) \Iso \quotep_{1}(p_{2})
-  \end{align*}
-\end{definition}
-
-\begin{theorem}
-  $\evalp/\quotep$ give a symmetric monoidal equivalence between $\PiPlusCat$ and $\PiHatCat$.
-\end{theorem}
-
-\subsection{$\PiHatLang$ to $\UFin$}
-
-Finally, we show how to interpret $\PiHatLang$ to $\UFin$, and back from $\UFin$ to $\PiHatLang$. Types in $\PiHatLang$
-are interpreted as 0-cells in $\UFin$, that is, a natural number $n$ is mapped to $\Fin[n]$. The 1-combinators in
-$\PiHatLang$ are mapped to 1-paths in $\UFin$, that is, 1-loops in each connected component, equivalent to
-$\Aut[\Fin[n]]$. In $\PiHatLang$, the 1-combinators are generated by adjacent transpositions, so these can be mapped to
-words in $\Sn$ and then to automorphisms using~\cref{prop:sn-lehmer-fin-equiv}. Finally, 2-combinators are mapped to
-2-paths between loops in $\UFin$.
-
-\begin{definition}[$\evalh$]
-  \begin{align*}
-    \evalh_{0} & : \UHat \to \UFin                                                          \\
-    \evalh_{1} & : (c : t_{1} \isoh t_{2}) \to \evalh_{0}(t_{1}) \id \evalh_{0}(t_{2})      \\
-    \evalh_{2} & : (\alpha : c_{1} \Isoh c_{2}) \to \evalh_{1}(c_{1}) \id \evalh_{1}(c_{2})
-  \end{align*}
-\end{definition}
-
-\noindent 0-cells in $\UFin$ are mapped to their cardinalities in $\PiHatLang$, 1-loops are decoded to words in $\Sn$ to
-generate a sequence of adjacent transpositions, producing a 1-combinator in $\PiHatLang$. Finally, 2-paths are quoted
-back to 2-combinators in $\PiHatLang$.
-
-\begin{definition}[$\quoteh$]
-  \begin{align*}
-    \quoteh_{0} & : \UFin \to \UHat                                                            \\
-    \quoteh_{1} & : (p : X_{1} \id X_{2}) \to \quoteh_{0}(X_{1}) \isoh \quoteh_{0}(X_{2})      \\
-    \quoteh_{2} & : (\alpha : p_{1} \id p_{2}) \to \quoteh_{1}(p_{1}) \Isoh \quotet_{1}(p_{2})
-  \end{align*}
-\end{definition}
-
-% $\UFin$ as a 1-groupoid, is a (locally strict) weak $(2,0)$-category, and we use this to state our final equivalence.
-
-\begin{theorem}
-  $\evalh/\quoteh$ give a symmetric monoidal equivalence between $\PiHatCat$ and $\UFin$.
-\end{theorem}
-
-The semantics that we presented here takes a different route to constructing the permutation from a $\PiLang$
-combinator, compared to the direct interpretation given using the big-step interpreter in~\cref{subsec:denotational}. We
-verify that the two semantics agree, establishing that the semantics is adequate and fully abstract.
-
-\begin{definition}[${\gdenot{\blank}}$]
-  \begin{gather*}
-    \begin{aligned}
-      \gdenot{\blank}_{0} & : U \to \UFin                                       \\
-      \gdenot{\blank}_{0} & \defeq \evalh_{0} \comp \evalp_{0} \comp \evalt_{0}
-    \end{aligned}
-    \qquad
-    \begin{aligned}
-      \gdenot{\blank}_{1} & : (c : X \iso Y) \to \gdenot{X}_{0} \id_{\UFin} \gdenot{Y}_{0} \\
-      \gdenot{\blank}_{1} & \defeq \evalh_{1} \comp \evalp_{1} \comp \evalt_{1}
-    \end{aligned}
-  \end{gather*}
-\end{definition}
-
-\begin{theorem}[Full Abstraction and Adequacy]
-  For any $c_1, c_2 : X \iso Y$, we have that
-  \[
-    \denot{c_1} = \denot{c_2} \text{ if and only if } \gdenot{c_1}_{1} = \gdenot{c_2}_{1}
-  \]
-\end{theorem}
-
-\section{Normalisation of Reversible Circuits}
+\section{Applications to Reversible Circuits}
 \label{sec:applications}
 
-Using our semantics, we can normalise, synthesise, prove equivalence, and generally reason about~$\PiLang$ programs. The
-two key definitions are presented below.
+Using our semantics, we can normalise, synthesise, prove equivalence, and reason about~$\PiLang$ programs.
+
+\[
+  \begin{tikzpicture}[->,>=stealth',auto,thick,node distance=1em]
+    % nodes
+    \node (A) at (0, 2) {\Afun{reversibleOr1}};
+    \node (B) at (0, 0) {\Afun{reversibleOr2}};
+    \node (C) at (4, 1) {\Afun{reversibleOrNorm}};
+    \node (D) at (9, 1) {\AgdaSymbol{(}\AgdaNumber{0 5 6 7 4 1 2 3}\AgdaSymbol{)}};
+    % arrows
+    \path
+    (A) edge[bend left] node [above left] { $\strut\normt_{1}$ } (C)
+    (B) edge[bend right] node [below left] { $\strut\normt_{1}$ } (C)
+    (D) edge node[above] { $\strut\term{quote}$ } (C);
+  \end{tikzpicture}
+\]
+
+% \[
+%   \begin{tabular}{>{$}c<{$} @{\hspace{5em}} >{$}c<{$}}
+%     \tikzmarknode{A}{\Afun{reversibleOr1}} & \\
+%                                            & \tikzmarknode{C}{\Afun{reversibleOrNorm}} \\
+%     \tikzmarknode{B}{\Afun{reversibleOr2}} &
+%   \end{tabular}
+%   % \tikzmarknode{D}{4}
+%   \begin{tikzpicture}[overlay, remember picture, shorten <=1mm,
+%     nodes={inner sep=1pt, align=center, font=\footnotesize}]
+%     \draw (A.south) -- ++ (-1,-1) node[below] {2 bits};
+%     \draw (B.south) -- ++ (-.5,-1) node[below] {2 bits};
+%     % \draw (D.south) -- ++ (.5,-1) node[below] {3 bits};
+%   \end{tikzpicture}
+% \]
+
+\subsection{Normalisation}
+
+Normalisation involves translating $\PiLang$ programs to $\PiHatLang$, computing a permutation, and quoting back to
+$\PiPlusLang$. This happends in several steps -- the evaluation step goes from $\PiLang$ to $\PiPlusLang$ reducing
+$\times$ to $+$ and translating the corresponding combinators, then from $\PiPlusLang$ to $\PiHatLang$ reducing to
+adjacent transpositions, then finally to $\UFin$ through words in $\Sn$ and Lehmer codes, and then, the quoting step
+goes all the way back until $\PiPlusLang$. We define this $\normt$ function below using our previous evaluation
+$\gdenot{\blank}$.
 
@@
 
-\noindent Normalisation involves translating $\PiLang$ programs to $\PiHatLang$, computing a permutation, and quoting
-back to $\PiPlusLang$. Note that the normalisation happens in the step from $\PiHatLang$ to $\UFin$ and back to
-$\PiHatLang$ and that normalisation also provides a decision procedure for program equivalence. Synthesis happens by
-quoting permutations. More general, user-guided, reasoning can be done using the sound and complete level-2
-combinators to rewrite $\PiLang$ programs.
+\subsection{Equivalence}
 
-Recall the specification of reversible disjunction from \cref{sec:examples}, the two Qiskit circuits for implementing
-it, and the corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The type $\mathbb{B}\ 3$
-normalises to $\mathbb{8}+ = \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} +
-(\mathbb{1} + \mathbb{0})))))))$, and normal forms for both circuits compute to the following, establishing their equivalence:
+By normalisation, we also get a decision procedure for program equivalence -- two circuits are equivalent iff they
+represent the same permutation, or equivalently, they have the same normal form. Recall the specification of reversible
+disjunction from \cref{sec:examples}, the two Qiskit circuits for implementing it, and the corresponding $\PiLang$
+definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The type $\mathbb{B}\ 3$ normalises to
+$\mathbb{8}+ \equiv \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + \mathbb{0})))))))$,
+and normal forms for both circuits compute to the following, establishing their equivalence:
 
@@
 
+\subsection{Synthesis}
+
 Instead of manually producing $\Pi$-programs to implement the reversible disjunction specification, it is possible to
-simply quote the desired permutation (using a function $\term{lookup}$ to read a tabulated version for readability):
+simply quote the desired permutation, synthesising a $\PiLang$ program. We use a function $\term{lookup}$ to write
+tabulated permutations for readability.
 
@@
 
-\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
-\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
-which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Quoting this
-permutation generates the same normalised program which can then be composed with a map from $\PiLang$ to
-$\PiPlusLang$ to produce a program matching the desired structured types.
+\noindent
+The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
+\textsf{true},\textsf{true}) is encoded as 011 or 3). The second entry maps index 1 (= 001) to the value 5 (= 101) which
+states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Quoting this
+permutation generates the same normalised program matching the desired structured types.
+
+\subsection{Reasoning}
+
+More general, user-guided, reasoning can be done using the sound and complete level-2 combinators to rewrite $\PiLang$
+programs.
 

### 2021-10-30 Pi/paper/applications.tex
Work on section 7

@@
 
-Normalisation involves translating $\PiLang$ programs to $\PiHatLang$, computing a permutation, and quoting back to
-$\PiPlusLang$. This happends in several steps -- the evaluation step goes from $\PiLang$ to $\PiPlusLang$ reducing
-$\times$ to $+$ and translating the corresponding combinators, then from $\PiPlusLang$ to $\PiHatLang$ reducing to
-adjacent transpositions, then finally to $\UFin$ through words in $\Sn$ and Lehmer codes, and then, the quoting step
-goes all the way back until $\PiPlusLang$. We define this $\normt$ function below using our previous evaluation
-$\gdenot{\blank}$.
+Normalisation happens by evaluating a combinator to a permutation and then reifying it back. The evaluation step goes
+from $\PiLang$ to $\PiPlusLang$ reducing $\times$ to $+$ and translating the corresponding combinators, then from
+$\PiPlusLang$ to $\PiHatLang$ reducing them to natural numbers and adjacent transpositions, then finally to $\UFin$
+through words in $\Sn$ and Lehmer codes. Then, the quotation step goes all the way back until $\PiPlusLang$. We define
+this composite $\normt$ function below using our previous evaluation $\gdenot{\blank}$.
 
@@
 
-\subsection{Equivalence}
-
-By normalisation, we also get a decision procedure for program equivalence -- two circuits are equivalent iff they
-represent the same permutation, or equivalently, they have the same normal form. Recall the specification of reversible
-disjunction from \cref{sec:examples}, the two Qiskit circuits for implementing it, and the corresponding $\PiLang$
-definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The type $\mathbb{B}\ 3$ normalises to
-$\mathbb{8}+ \equiv \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + \mathbb{0})))))))$,
-and normal forms for both circuits compute to the following, establishing their equivalence:
+As an example, consider the \Afun{reversibleOr1} circuit introduced in~\cref{sec:pi}. After applying $\normt_{1}$, the
+type $\mathbb{B}\ 3$ normalises to $\mathbb{8}+ \equiv \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} +
+  (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + \mathbb{0})))))))$, and the circuit reduces to the following
+normal form:
 
@@
 
+\subsection{Equivalence}
+
+By normalisation, we also get a decision procedure for program equivalence -- two circuits are equivalent iff they
+represent the same permutation, or equivalently, they have the same normal form. Recall that in~\cref{sec:examples} we
+introduced two different ways of computing reversible disjunction using two Qiskit circuit implementations, and then
+in~\cref{sec:pi} showed the corresponding $\PiLang$ definitions: \Afun{reversibleOr1} and \Afun{reversibleOr2}.
+
+After applying $\normt_{1}$ to each circuit, we verify that they both reduce to the same normal form
+\Afun{reversibleOrNorm}, along with a 2-combinator witnessing their equivalence with the normal form -- by composing
+these, we get the 2-combinator witnessing their equivalence.
+
 \subsection{Synthesis}
 
-Instead of manually producing $\Pi$-programs to implement the reversible disjunction specification, it is possible to
-simply quote the desired permutation, synthesising a $\PiLang$ program. We use a function $\term{lookup}$ to write
-tabulated permutations for readability.
+Instead of manually producing $\PiLang$-programs to implement the reversible disjunction specification, it is possible
+to simply write down the permutation directly and then quote it, synthesising a $\PiLang$ program. We use a function
+$\term{lookup}$ to write tabulated permutations for readability.
 
@@
 \noindent
-The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
-\textsf{true},\textsf{true}) is encoded as 011 or 3). The second entry maps index 1 (= 001) to the value 5 (= 101) which
-states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Quoting this
-permutation generates the same normalised program matching the desired structured types.
+The permutation uses the canonical encoding of sequences of bits as natural numbers, e.g., {(\textsf{false},
+\textsf{true},\textsf{true})} is encoded as 011 or 3. The second entry in the $\term{lookup}$ table maps index 1 (= 001)
+to the value 5 (= 101), following the reversible disjunction specification (recall that
+$\mathsf{reversibleOr}(h,b_1,b_2) = (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
+and~$\underline{\vee}$ is exclusive-disjunction). Quoting this permutation generates the same normalised program
+\Afun{reversibleOrNorm}, matching the desired structured type of a 1-combinator on $\mathbb{8}+$.
 
@@
 
-More general, user-guided, reasoning can be done using the sound and complete level-2 combinators to rewrite $\PiLang$
-programs.
+In our development, we described a number of different representations of permutations: $\PiLang$ programs, lists of
+transpositions, Lehmer codes, and automorphisms of finite sets. The equivalences between these representations allows us
+to transfer theorems about permutations, proving them on the representations most suited for the particular task and
+getting others for free.
+
+As an example, consider the following problem. Toffoli gate on 3 bits can be extended to 4 bits, inserting a wire in
+four possible positions. The function of such gate is then identity on the newly inserted bit, and controlled-controlled
+not on the three remaining ones.
+
+\medskip
+\extendedToffoli3{}
+
+On the other hand, it is possible to construct a proper 4-bit toffoli gate:
+
+\medskip
+\toffoli4{}
+
+Now, the question we want to asnwer is: is it possible to compose some number of (extended) 3-bit toffoli gates in some way, so as to replicate the
+behaviour of a 4-bit toffoli gate?
+
+Solving this problem using the circuit representation is difficult. However, we can observe that 3-bit toffoli gate
+implements an even permutation, that is, a permutation consisting of an even number of transpositions. On the other
+hand, 4-bit toffoli gate is an odd permutation. It is an easily proven lemma that composition of even permutations
+always produces an even permutation, which answers the question to the negative.
+
+In our case, we define the $\term{parity}$ function:
+parity : {n m : ℕ} → (c : S n Pi^.⟷₁^ S m) → ⊤ ⊔ ⊤
+parity {X} {Y} c = list-len-parity (pi^2list c)
+
+Thus, we formulate the theorem using the representation on lists of transpositions, and transfer it to circuits using
+the equivalences proven in the previous sections.
+
+% More general, user-guided, reasoning can be done using the sound and complete level-2 combinators to rewrite $\PiLang$
+% programs.
 

### 2021-10-30 Pi/paper/examples.tex
Work on section 7

@@
 For concreteness, we demonstrate two different circuits that implement the following reversible function specification
-$\mathit{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
+$\mathsf{reversibleOr}(h,b_1,b_2) ~=~ (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
 and~$\underline{\vee}$ is exclusive-disjunction. The circuits are presented in both the textual interface \verb|qasm|

### 2021-10-30 Pi/paper/applications.tex
Progress

@@
 
-\[
-  \begin{tikzpicture}[->,>=stealth',auto,thick,node distance=1em]
-    % nodes
-    \node (A) at (0, 2) {\Afun{reversibleOr1}};
-    \node (B) at (0, 0) {\Afun{reversibleOr2}};
-    \node (C) at (4, 1) {\Afun{reversibleOrNorm}};
-    \node (D) at (9, 1) {\AgdaSymbol{(}\AgdaNumber{0 5 6 7 4 1 2 3}\AgdaSymbol{)}};
-    % arrows
-    \path
-    (A) edge[bend left] node [above left] { $\strut\normt_{1}$ } (C)
-    (B) edge[bend right] node [below left] { $\strut\normt_{1}$ } (C)
-    (D) edge node[above] { $\strut\term{quote}$ } (C);
-  \end{tikzpicture}
-\]
+% \[
+%   \begin{tikzpicture}[->,>=stealth',auto,thick,node distance=1em]
+%     % nodes
+%     \node (A) at (0, 2) {\Afun{reversibleOr1}};
+%     \node (B) at (0, 0) {\Afun{reversibleOr2}};
+%     \node (C) at (4, 1) {\Afun{reversibleOrNorm}};
+%     \node (D) at (9, 1) {\AgdaSymbol{(}\AgdaNumber{0 5 6 7 4 1 2 3}\AgdaSymbol{)}};
+%     % arrows
+%     \path
+%     (A) edge[bend left] node [above left] { $\strut\normt_{1}$ } (C)
+%     (B) edge[bend right] node [below left] { $\strut\normt_{1}$ } (C)
+%     (D) edge node[above] { $\strut\term{quote}$ } (C);
+%   \end{tikzpicture}
+% \]
 
@@
 not on the three remaining ones.
-
 \medskip
-\extendedToffoli3{}
-
+\extendedToffoli{}
+\baselineskip
 On the other hand, it is possible to construct a proper 4-bit toffoli gate:
-
 \medskip
-\toffoli4{}
-
-Now, the question we want to asnwer is: is it possible to compose some number of (extended) 3-bit toffoli gates in some way, so as to replicate the
+\toffoli{}
+\baselineskip
+Now, the question we want to answer is: is it possible to compose some number of (extended) 3-bit toffoli gates in some way, so as to replicate the
 behaviour of a 4-bit toffoli gate?
@@
 
-In our case, we define the $\term{parity}$ function:
-parity : {n m : ℕ} → (c : S n Pi^.⟷₁^ S m) → ⊤ ⊔ ⊤
-parity {X} {Y} c = list-len-parity (pi^2list c)
+We define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a combinator
+$c$. It is implemented as the transformation from a $\PiLang$ program into a list of transposition, using $\evalh_{1}$
+defined in~\cref{sec:equivalence}, and computing the parity on this representation, which means simply checking the
+parity of the length of the list.
 
-Thus, we formulate the theorem using the representation on lists of transpositions, and transfer it to circuits using
+\begin{propositionrep}
+  \label{prop:parity-preserved}
+  For two $\PiLang$ circuits $c_1, c_2$, if there exists a 2-combinator between them, then $\parity c_1 = \term{parity} c_2$. 
+\end{propositionrep}
+\begin{proof}
+  First, it is easy to check that Coxeter relations preserve parity of the list. Second, if $c_1, c_2$ are related by a
+  2-combinator, then their list-of-transpositions representations are related in the Coxeter relation. Composing these
+  two facts completes the proof.
+\end{proof}
+
+Using ~\cref{prop:parity-preserved} and showing that any composition of even cicuits yields an even ciruit, we can
+answer the question by computing the parity of the \Afun{toffoli₄} and all of the \Afun{toffoli₃} gates.
+
+Thus, we formulated the theorem using the representation on lists of transpositions, and transferred it to circuits using
 the equivalences proven in the previous sections.

### 2021-10-30 Pi/paper/discussion.tex
Progress

@@
 correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
-suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work. In the
+suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work, as in~\cref{fig:table-summary}. In the
 remainder of this section, we discuss some broader related work.
 
-% \begin{center}
-%     \begin{tabular}{c c c}
-%         \hline
-%         $\PiLang$     & $\Sn$               & $\UFin$     \\
-%         \hline
-%         types         & Natural numbers     & Finite sets \\
-%         \hline
-%         1-combinators & Generators of $\Sn$ & 1-paths     \\
-%         \hline
-%         2-combinators & Relations of $\Sn$  & 2-paths     \\
-%         \hline
-%     \end{tabular}
-% \end{center}
+\begin{figure}
+    \begin{center}
+        \begin{tabular}{|c|c|c|c|}
+            \hline
+            \rowcolor{lightgray} $\PiLang$     & $\bigsqcup_n \mathcal{B}\Sn $ & $\BFin$     & $\UFin$ \tabularnewline
+            \hline
+            types         & Natural numbers                & Finite sets & 0-cells \tabularnewline
+            \hline
+            1-combinators & Generators of $\Sn$            & Bijections  & 1-paths \tabularnewline
+            \hline
+            2-combinators & Relations of $\Sn$             & Homotopies  & 2-paths \tabularnewline
+            \hline
+        \end{tabular}
+    \end{center}
+    \caption{Summary of the established correspondences}
+    \label{fig:table-summary}        
+\end{figure}
 

### 2021-10-30 Pi/paper/applications.tex
Formatting

@@
 
-\subsection{Synthesis}
+\subsection{Synthesis and verification}
 
@@
 
+In a similar way we can also verify whether a circuit is correctly implemented: by running the equivalence described
+above in the other direction, we can evaluate the circuit to a bijection and compare it extensionally with the intended
+one.
+
 \subsection{Reasoning}
@@
 \extendedToffoli{}
-\baselineskip
+
+\noindent
 On the other hand, it is possible to construct a proper 4-bit toffoli gate:
@@
 \toffoli{}
-\baselineskip
+
+\noindent
 Now, the question we want to answer is: is it possible to compose some number of (extended) 3-bit toffoli gates in some way, so as to replicate the

### 2021-10-30 Pi/paper/equivalence.tex

@@
 The translations between the languages are defined separately on types, 1-combinators, and 2-combinators. Following the
-terminology of Normalisation by Evaluation, the translations from the left to the right, going from the syntax towards
-the semantics, are called $\evalt$ and the translations the other way are called $\quotet$.
+terminology of Normalisation by Evaluation (NbE), the translations from the left to the right, going from the syntax
+towards the semantics, are called $\evalt$ and the translations the other way are called $\quotet$.
 
@@
 
-\section{Normalisation of Reversible Circuits}
-\label{sec:applications}
-
-Using our semantics, we can normalise, synthesise, prove equivalence, and generally reason about~$\PiLang$ programs. The
-two key definitions are presented below.
-
-\begin{definition}[Normalisation of $\PiLang$ programs]
-  \begin{gather*}
-    \begin{aligned}
-      \normt_{0} & : U \to \UPlus                                            \\
-      \normt_{0} & = \quotep_{0} \comp \quoteh_{0} \comp \gdenot{\blank}_{0}
-    \end{aligned}
-    \qquad
-    \begin{aligned}
-      \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \iso \normt_{0}(Y)     \\
-      \normt_{1} & = \quotep_{1} \comp \quoteh_{1} \comp \gdenot{\blank}_{1}
-    \end{aligned}
-  \end{gather*}
-\end{definition}
-
-\noindent Normalisation involves translating $\PiLang$ programs to $\PiHatLang$, computing a permutation, and quoting
-back to $\PiPlusLang$. Note that the normalisation happens in the step from $\PiHatLang$ to $\UFin$ and back to
-$\PiHatLang$ and that normalisation also provides a decision procedure for program equivalence. Synthesis happens by
-quoting permutations. More general, user-guided, reasoning can be done using the sound and complete level-2
-combinators to rewrite $\PiLang$ programs.
-
-Recall the specification of reversible disjunction from \cref{sec:examples}, the two Qiskit circuits for implementing
-it, and the corresponding $\PiLang$ definitions \Afun{reversibleOr1} and \Afun{reversibleOr2}. The type $\mathbb{B}\ 3$
-normalises to $\mathbb{8}+ = \mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} + (\mathbb{1} +
-(\mathbb{1} + \mathbb{0})))))))$, and normal forms for both circuits compute to the following, establishing their equivalence:
-
-\medskip
-\resetnormtwo{}
-
-Instead of manually producing $\Pi$-programs to implement the reversible disjunction specification, it is possible to
-simply quote the desired permutation (using a function $\term{lookup}$ to read a tabulated version for readability):
-
-\medskip
-\resetperm{}
-
-\noindent The permutation uses the canonical encoding of sequences of bits as natural numbers (e.g., (\textsf{false},
-\textsf{true},\textsf{true}) is encoded as 011 or 3).  The second entry maps index 1 (= 001) to the value 5 (= 101)
-which states that since one of the right bits is set in 001 then the leftmost bit in the output is set. Quoting this
-permutation generates the same normalised program which can then be composed with a map from $\PiLang$ to
-$\PiPlusLang$ to produce a program matching the desired structured types.
-
-% \paragraph*{Program synthesis.} The NbE process embodies a quoting mechanism that synthesizes programs from
-% permutations. Indeed, instead of writing a program for \Afun{reset 2}, one could simply specify the desired permutation
-% as:
-
-%%\begin{minipage}{.65\textwidth}
-%%  \PiRESET{}
-%%\end{minipage}
-%%\begin{minipage}{.30\textwidth}
-%%  \begin{center}
-%%  \resizebox{0.5\textwidth}{!}{\input{diagrams/reset.tikz}}
-%%  \end{center}
-%%\end{minipage}
-
-%% \resettwo{}
-
-% \noindent The syntax will be explained in detail in the next section and the full definitions of the helpers are
-% provided in the supplementary material. For now, it is sufficient to know that there is some program that implements the
-% reversible function of interest and that applying \Afun{reset} to 2 produces \verb|reversibleOr2| from the introduction.
-
-% In principle, the normalised program can be produced following two strategies: (i) by repeatedly applying the rewrite
-% rules of our calculus of reversible functions (explained in Sec.~\ref{sec:reversibletwo}), or (ii) in the case above by
-% using a NbE process that evaluates the program to a permutation on a finite set of 8
-% elements and reifies that permutation back to a program. The key idea of the NbE process is a systematic way to express
-% permutations as sequences of adjacent swaps as illustrated in the following small example where the permutation on the
-% left is compiled to the sequence of four adjacent transpositions on the right:
-
-% \note{Motivation: There are two reversible circuits which describe the following permutation. They can be shown to be
-%   equal using the 2-combinators.}
-
-% \[
-%   \begin{tikzpicture}
-%     \begin{knot}[clip width=5]
-%       \filldraw (0,5) circle (2pt) node[above] {0};
-%       \filldraw (1,5) circle (2pt) node[above] {1};
-%       \filldraw (2,5) circle (2pt) node[above] {2};
-%       \filldraw (3,5) circle (2pt) node[above] {3};
-%       \filldraw (4,5) circle (2pt) node[above] {4};
-%       \filldraw (0,0) circle (2pt) node[below] {1};
-%       \filldraw (1,0) circle (2pt) node[below] {4};
-%       \filldraw (2,0) circle (2pt) node[below] {0};
-%       \filldraw (3,0) circle (2pt) node[below] {3};
-%       \filldraw (4,0) circle (2pt) node[below] {2};
-%       \strand (0,5) .. controls (0.5,0.5) and (1.5,3.5) .. (2,0);
-%       \strand (1,5) .. controls (0.75,0.5) and (0.25,3.5) .. (0,0);
-%       \strand (2,5) .. controls (2.5,2.5) and (3.5,1.5) .. (4,0);
-%       \strand (3,5) .. controls (4.5,2.5) and (4,1.5) .. (3,0);
-%       \strand (4,5) .. controls (3.5,2.5) and (1.5,2.5) .. (1,0);
-%       \flipcrossings{4,5};
-%     \end{knot}
-%   \end{tikzpicture}
-% \]
-
-% \paragraph*{Program equivalence.} The permutation above reveals another way to think about the desired program: it is a
-% special addition circuit that keeps 0 and 4 fixed but otherwise adds 4 modulo 8 to its input. From this specification,
-% one can use a standard synthesis algorithm for reversible circuits~\cite{10.1145/775832.775915} to generate the
-% following program:
-
-% \adder{}
-
-% \noindent The \Afun{adder3} program looks nothing like the original \Afun{reset 2} program and yet they both have the
-% same normal form thus establishing their equivalence. The reader can check that this circuit is the same as
-% \verb|reversibleOr1| from the introduction.
-
-% Now imagine we want to write the following reversible function:
-
-% 0 -> 0
-% 8 -> 8
-% n -> n + 8 `mod` 16
-
 %%% Local Variables:

### 2021-10-30 Pi/paper/introduction.tex

@@
 \paragraph{Outline and Contributions} Our main result is a proof of the soundness and completeness of $\PiLang$ with
-respect to its semantics in the weak symmetric rig groupoid of finite sets and permutations. We state our result as a
-Curry-Howard-Lambek correspondence for Reversible Logic, Reversible Programming Languages, and Symmetric Rig Groupoids,
-using which we can build a toolbox of technical devices for reasoning about reversible circuits.
+respect to its semantics in the weak symmetric rig groupoid of finite sets and permutations, formalised in HoTT/UF. We
+state our result as a Curry-Howard-Lambek correspondence for Reversible Logic, Reversible Programming Languages, and
+Symmetric Rig Groupoids, using which we can build a toolbox of technical devices for reasoning about reversible
+circuits.
 

### 2021-10-30 Pi/paper/reversible1.tex
Progress
Formatting
Cite pi-dual

@@
 To see how this language expresses reversible circuits, we present a few examples using the Agda embedding of the
-language~\cite{choudhuryArtifactSymmetriesReversible2021}. First it is possible to directly mimic the
-\verb|qasm|-perspective by defining types that describe sequences of booleans ($\mathbb{2}^{n}$). We use the type
-$\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true}, and
+language~\cite{choudhuryArtifactSymmetriesReversible2021,sabryJacquesCarettePidualSecond2021}. First it is possible to
+directly mimic the \verb|qasm|-perspective by defining types that describe sequences of booleans ($\mathbb{2}^{n}$). We
+use the type $\mathbb{2} = \onet + \onet$ to represent booleans with $\inlv{\Acon{tt}}$ representing \textsf{true}, and
 $\inrv{\Acon{tt}}$ representing $\textsf{false}$. Boolean negation (the \verb|x|-gate) is straightforward to define
@@
 turn out to be expressible using the distributivity and factoring identities of rigs as shown below:
-
 \medskip
-
 \cif{}
@@
 is transcribed as follows:
-
 \medskip
-
 \adder{}
@@
 circuit, instead of transcribing it directly, we express it using $\Afun{cif}$.
-
 \medskip
-
 \resettwo{}
@@
 example, we have the following level-2 combinators dealing with associativity:
-
 \medskip
-
 \leveltwoblockone{}

### 2021-10-30 Pi/paper/applications.tex
Say symmetric group instead of permutation group

@@
 
-In our development, we described a number of different representations of permutations: $\PiLang$ programs, lists of
+In our development, we described a number of different representations of permutations: $\PiLang$, $\PiPlusLang$ and $\PiHatLang$ programs, lists of
 transpositions, Lehmer codes, and automorphisms of finite sets. The equivalences between these representations allows us
@@
 \noindent
-Now, the question we want to answer is: is it possible to compose some number of (extended) 3-bit toffoli gates in some way, so as to replicate the
-behaviour of a 4-bit toffoli gate?
+Now, the question we want to answer is: is it possible to implement \Afun{toffoli₄} gate only using \Afun{toffoli₃} gates?
 
-Solving this problem using the circuit representation is difficult. However, we can observe that 3-bit toffoli gate
-implements an even permutation, that is, a permutation consisting of an even number of transpositions. On the other
-hand, 4-bit toffoli gate is an odd permutation. It is an easily proven lemma that composition of even permutations
+Solving this problem using the circuit representation is difficult. However, we can observe that all \Afun{toffoli₃} gates
+implements even permutations, that is, permutations consisting of an even number of transpositions. On the other
+hand, \Afun{toffoli₄} implements an odd permutation. It is an easily proven lemma that any composition of even permutations
 always produces an even permutation, which answers the question to the negative.
 
-We define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a combinator
-$c$. It is implemented as the transformation from a $\PiLang$ program into a list of transposition, using $\evalh_{1}$
-defined in~\cref{sec:equivalence}, and computing the parity on this representation, which means simply checking the
-parity of the length of the list.
+Thus, define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a combinator
+$c$. It is implemented first tranforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$
+defined in~\cref{sec:equivalence}), and then computing the parity on this representation, which amounts to a simple check of the length of the list.
 

### 2021-10-30 Pi/paper/discussion.tex
Say symmetric group instead of permutation group

@@
 universal property of $\FSM$ is asserted by appealing to Mac Lane's coherence theorem for symmetric monoidal categories,
-and using the fact that the finite symmetric group $\Sn$ encodes the permutation group on a finite set. The existence of
+and using the fact that the finite symmetric group $\Sn$ encodes the group of permutations on a finite set. The existence of
 the proof is folklore. We have produced a proof and formalised it in constructive type theory. A stronger universal

### 2021-10-30 Pi/paper/finite.tex
Say symmetric group instead of permutation group

@@
 
-\subsection{Presenting the permutation group}
+\subsection{Presenting the symmetric group}
 
@@
 on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate for the
-generators of a permutation group. For example, we could generate the permutation group on $\Fin[n]$ by using generators
+generators of a permutation group. For example, we could generate the symmetric group on $\Fin[n]$ by using generators
 (primitive operations) that:
@@
 
-Finally, it is time to complete our goal of characterising the permutation groups. Having produced a Lehmer code by
+Finally, it is time to complete our goal of characterising the symmetric group. Having produced a Lehmer code by
 normalising words in $\Sn$, we need to run it to produce a concrete bijection of finite sets, and, given a bijection

### 2021-10-30 Pi/paper/finite.tex
Copy-editing finite section

@@
 One way of thinking about presentations of $\Sn$ is via sorting algorithms, which use different primitive operations. A
-sorting algorithm has to calculate a permutation of a list or a finite set, which satisfies the invariant of being a
-sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the permutations
-on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate for the
-generators of a permutation group. For example, we could generate the symmetric group on $\Fin[n]$ by using generators
-(primitive operations) that:
+sorting algorithm has to calculate a permutation of a list or a (listed) finite set, which satisfies the invariant of
+being a sorted sequence, which means, the primitive operations of a sorting algorithm are able to generate all the
+permutations on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate
+for the generators of a permutation group. For example, we could generate the symmetric group on $\Fin[n]$ by using
+generators (primitive operations) that:
 
@@
 We will show that it is possible to encode all $\PiLang$ combinators using adjacent transpositions
-(in~\cref{sec:equivalence}). Group presentations are built by adding equations to a \emph{free group}.
+(in~\cref{sec:equivalence}).
+%Group presentations are built by adding equations to a \emph{free group}.
 
@@
 generating set. This is the universal property of free groups, stemming from the free-forgetful adjunction between the
-category of groups and sets.
+category of groups and sets. We say that $\eta_{A} : A \to F(A)$ is the free group on $A$ if it satisfies the following
+universal property.
 
@@
 
-Following the universal-algebraic definition, in HoTT, we could use a naive higher inductive type to define the free
+Following the naive universal-algebraic definition, in HoTT, we could use a higher inductive type to define the free
 group, which enforces the group axioms by adding path constructors (see~\cref{def:free-group}). Using the induction
@@
 the resulting group would be $\mathbb{Z}_2 \eqv \Aut[\Bool]$. Note that not all groups have finite (or computable)
-presentations, and, a group can have any number of different presentations.
+presentations, and, a group can have a number of different presentations.
 
@@
 
-The universal property of the above definition is similar to~\cref{prop:free-groups} except the relation has to be
-preserved by the function mapping out of the generating set.
+The universal property of the above definition of a presented group is similar to~\cref{prop:free-groups}, except the
+function mapping out of the generating set also has to respect the relation.
 
@@
 in a group presentation, this is reduced to deciding whether one word -- a representative of the equivalence class of
-the group's elements, can be reduced to another word, using the group's relations. However, these equations are not
+the group's elements, can be reduced to another word, using the group's relations. However, these relations are not
 directed, so it is not always possible to construct a well-behaved rewriting system. In general, the word problem for
@@
 $k$ swaps elements at indices $k$ and $k+1$. Thus, the generating set is $\Fin[n]$. There are three relations that we're
-going to specify for this presentation -- we visualise them as braid diagrams in~\cref{fig:coxeter-braid}.
+going to specify for this presentation -- we visualise them as braiding diagrams in~\cref{fig:coxeter-braid}.
 
@@
 
-For this to work, we first need the system to have the termination property, meaning that there are no infinite
+For this to work, we first need the system to have the termination property, meaning that there should be no infinite
 reductions. We observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
@@
 well-founded). Thus, by directing the relation from left to right, we would get the termination property out of the box.
-Second, we need the normal forms to be unique, so that we can get a normalisation function. This will be true if the
-rewriting system is confluent -- meaning that all critical pairs, that is, terms with overlapping possible reduction
-rules, have to converge. For example, in our system, the pair in~\cref{fig:critical-pairs-converging} converges.
-Unfortunately, this is not true for all critical pairs -- an  example is in~\cref{fig:critical-pairs-non-converging},
-where left and right endpoints are normal with respect to the $\cox*$ relation.
+Second, we need the normal forms to be unique, so that we can extract a normalisation function. This will be true if the
+rewriting system is confluent -- all critical pairs, that is, terms with overlapping possible reduction rules, have to
+converge. For example, in our system, the pair in~\cref{fig:critical-pairs-converging} converges. Unfortunately, this is
+not true for all critical pairs -- an example is in~\cref{fig:critical-pairs-non-converging}, where left and right
+endpoints are normal with respect to the $\cox*$ relation.
 
@@
 properties of confluence and termination. We prove that the new relation defined by this system is equivalent, in a
-technical sense, to the standard Coxeter relation $\cox*$. First, we need to define a function $n \downf k$.
+technical sense (\cref{prop:coxlongcox}), to the standard Coxeter relation $\cox*$. First, we need to define a function
+$\downf$.
 
@@
 
-\noindent The result of this function is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\), which is a
-sequence of transpositions that moves the element at index $k + n$ left by $k$ places, shifting all the elements in
+\noindent The result of computing $n \downf k$ is the sequence \([k + n - 1, k + n - 2, k + n - 3, \ldots, n]\), which
+is a sequence of transpositions that moves the element at index $k + n$ left by $k$ places, shifting all the elements in
 between one place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the following
-generators (inlining the congruence closure in $\longcox$, allowing arbitrary reductions inside the list).
+generators (inlining the congruence closure in $\longcox$ allows arbitrary reductions inside the list).
 
@@
 and $k = 3$, it allows for the reduction $[4, 3, 2, 1, 0, 4] \longcox [3, 4, 3, 2, 1, 0]$ (see~\cref{fig:longcox} --
-note the distinction between wires, where numbers represent the values, and transpositions, where numbers represent
-which wires are crossing).
+there is a distinction between wires, where numbers represent the values, and transpositions, where numbers represent
+crossing wires).
 

### 2021-10-30 Pi/paper/applications.tex
Copy-editing applications section

@@
 
-Using our semantics, we can normalise, synthesise, prove equivalence, and reason about~$\PiLang$ programs.
+Using our semantics, we can normalise, synthesise, prove equivalence, and reason about~$\PiLang$ programs. We describe a
+few applications in this section, and a more comprehensive list of applications and examples along with their
+foramlisation is available in our artifact~\cite{choudhuryArtifactSymmetriesReversible2021}.
 
@@
 $\PiPlusLang$ to $\PiHatLang$ reducing them to natural numbers and adjacent transpositions, then finally to $\UFin$
-through words in $\Sn$ and Lehmer codes. Then, the quotation step goes all the way back until $\PiPlusLang$. We define
-this composite $\normt$ function below using our previous evaluation $\gdenot{\blank}$.
+through words in $\Sn$ and Lehmer codes. Finally, the quotation step goes all the way back until $\PiPlusLang$. We
+define this composite $\normt$ function below using our previous evaluation function $\gdenot{\blank}$.
 
@@
 
+\noindent
 As an example, consider the \Afun{reversibleOr1} circuit introduced in~\cref{sec:pi}. After applying $\normt_{1}$, the
@@
 \Afun{reversibleOrNorm}, along with a 2-combinator witnessing their equivalence with the normal form -- by composing
-these, we get the 2-combinator witnessing their equivalence.
+these, we get the 2-combinator witnessing their equivalence. More general, user-guided, reasoning can be done using the
+sound and complete level-2 combinators to rewrite $\PiLang$ programs.
 
@@
 
-In our development, we described a number of different representations of permutations: $\PiLang$, $\PiPlusLang$ and $\PiHatLang$ programs, lists of
-transpositions, Lehmer codes, and automorphisms of finite sets. The equivalences between these representations allows us
-to transfer theorems about permutations, proving them on the representations most suited for the particular task and
-getting others for free.
+In our development, we described a number of different representations of permutations: $\PiLang$, $\PiPlusLang$ and
+$\PiHatLang$ programs, lists of transpositions, Lehmer codes, and automorphisms of finite sets. The equivalences between
+these representations allows us to transfer theorems about permutations, proving them on the representations most suited
+for the particular task and getting others for free.
 
@@
 
-Solving this problem using the circuit representation is difficult. However, we can observe that all \Afun{toffoli₃} gates
-implements even permutations, that is, permutations consisting of an even number of transpositions. On the other
-hand, \Afun{toffoli₄} implements an odd permutation. It is an easily proven lemma that any composition of even permutations
-always produces an even permutation, which answers the question to the negative.
+Solving this problem using the circuit representation is difficult. However, we can observe that all \Afun{toffoli₃}
+gates implements even permutations, that is, permutations consisting of an even number of transpositions. On the other
+hand, \Afun{toffoli₄} implements an odd permutation. It is an easily proven lemma that any composition of even
+permutations always produces an even permutation, which answers the question to the negative. Thus, we want to formulate
+the theorem using the representation on lists of transpositions, and transfer it to circuits using the equivalences
+proven in the previous sections.
 
-Thus, define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a combinator
-$c$. It is implemented first tranforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$
-defined in~\cref{sec:equivalence}), and then computing the parity on this representation, which amounts to a simple check of the length of the list.
+Thus, define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a
+combinator $c$, by first tranforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$ defined
+in~\cref{sec:equivalence}), and then computing the parity on this representation, which amounts to a simple check of the
+partiy of the length of the list.
 
@@
   \label{prop:parity-preserved}
-  For two $\PiLang$ circuits $c_1, c_2$, if there exists a 2-combinator between them, then $\parity c_1 = \term{parity} c_2$. 
+  For two $\PiLang$ circuits $c_1, c_2$, if $c_1 \Iso c_2$, then $\parity(c_1) = \term{parity}(c_2)$.
 \end{propositionrep}
 \begin{proof}
-  First, it is easy to check that Coxeter relations preserve parity of the list. Second, if $c_1, c_2$ are related by a
-  2-combinator, then their list-of-transpositions representations are related in the Coxeter relation. Composing these
-  two facts completes the proof.
+  First, it is an easy check that Coxeter relations preserve parity of the list. Second, if $c_1$ and $c_2$ are related
+  by a 2-combinator, then their Lehmer codes are the same, so their list-of-transpositions representations have to be
+  related by the Coxeter relation. Composing these two facts completes the proof.
 \end{proof}
 
-Using ~\cref{prop:parity-preserved} and showing that any composition of even cicuits yields an even ciruit, we can
-answer the question by computing the parity of the \Afun{toffoli₄} and all of the \Afun{toffoli₃} gates.
-
-Thus, we formulated the theorem using the representation on lists of transpositions, and transferred it to circuits using
-the equivalences proven in the previous sections.
-
-% More general, user-guided, reasoning can be done using the sound and complete level-2 combinators to rewrite $\PiLang$
-% programs.
+Using ~\cref{prop:parity-preserved} and already mentioned lemma that any compostion of even circuits is even, we can
+finish the proof by running $\parity$ on \Afun{toffoli₄} and on all of the \Afun{toffoli₃}.
 

### 2021-10-30 Pi/paper/applications.tex
Many more edits

@@
 few applications in this section, and a more comprehensive list of applications and examples along with their
-foramlisation is available in our artifact~\cite{choudhuryArtifactSymmetriesReversible2021}.
+formalisation is available in our artifact~\cite{choudhuryArtifactSymmetriesReversible2021}.
 
@@
     \begin{aligned}
-      \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \iso \normt_{0}(Y)     \\
+      \normt_{1} & : (c : X \iso Y) \to \normt_{0}(X) \isop \normt_{0}(Y)    \\
       \normt_{1} & = \quotep_{1} \comp \quoteh_{1} \comp \gdenot{\blank}_{1}
@@
 
-After applying $\normt_{1}$ to each circuit, we verify that they both reduce to the same normal form
-\Afun{reversibleOrNorm}, along with a 2-combinator witnessing their equivalence with the normal form -- by composing
-these, we get the 2-combinator witnessing their equivalence. More general, user-guided, reasoning can be done using the
-sound and complete level-2 combinators to rewrite $\PiLang$ programs.
+After applying $\normt_{1}$ to each circuit, we can verify that they both reduce to the same normal form
+\Afun{reversibleOrNorm}, and there exist 2-combinators between each of them and the normal form. More general,
+user-guided, reasoning can be done using the sound and complete level-2 combinators to rewrite $\PiLang$ programs.
+% as an example, we can manually produce a proof of equivalence of ?? and ?? -- a 2-combinator presented below: 
+% \medskip
+% \manualTwoCombinator{}
 
@@
 
-Instead of manually producing $\PiLang$-programs to implement the reversible disjunction specification, it is possible
-to simply write down the permutation directly and then quote it, synthesising a $\PiLang$ program. We use a function
-$\term{lookup}$ to write tabulated permutations for readability.
+Instead of manually writing $\PiLang$ programs to implement the reversible disjunction specification, it is possible to
+simply write down the permutation directly and then quote it, synthesising a $\PiLang$ program. We show this below,
+using a function $\term{lookup}$ to write tabulated permutations for readability.
 
@@
 \noindent
-The permutation uses the canonical encoding of sequences of bits as natural numbers, e.g., {(\textsf{false},
-    \textsf{true},\textsf{true})} is encoded as 011 or 3. The second entry in the $\term{lookup}$ table maps index 1 (= 001)
-to the value 5 (= 101), following the reversible disjunction specification (recall that
+The permutation uses the canonical encoding of sequences of bits as natural numbers, e.g.,
+${(\mathsf{false},\mathsf{true},\mathsf{true})}$ is encoded as 011 or 3. The second entry in the $\term{lookup}$ table
+maps index 1 (= 001) to the value 5 (= 101), following the reversible disjunction specification (recall that
 $\mathsf{reversibleOr}(h,b_1,b_2) = (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
@@
 
-In a similar way we can also verify whether a circuit is correctly implemented: by running the equivalence described
-above in the other direction, we can evaluate the circuit to a bijection and compare it extensionally with the intended
-one.
+Similarly, we can also verify whether a circuit is correctly implemented -- by running the equivalence described above
+in the other direction, we can evaluate the circuit to a bijection and compare it extensionally with the intended one.
 
@@
 $\PiHatLang$ programs, lists of transpositions, Lehmer codes, and automorphisms of finite sets. The equivalences between
-these representations allows us to transfer theorems about permutations, proving them on the representations most suited
-for the particular task and getting others for free.
+these representations allow us to transfer theorems about permutations -- we prove them on the most suitable
+representation, and then transport them to a different representation ``for free''.
 
-As an example, consider the following problem. Toffoli gate on 3 bits can be extended to 4 bits, inserting a wire in
-four possible positions. The function of such gate is then identity on the newly inserted bit, and controlled-controlled
-not on the three remaining ones.
+As an example, consider the following problem. The Toffoli gate on 3 bits can be extended to a reversible gate on 4
+bits, by inserting a wire in four possible positions. This gate performs identity on the newly inserted bit, and
+controlled-controlled not on the three remaining ones. We show how to write these gates using \Afun{cif}.
 \medskip
@@
 \noindent
-On the other hand, it is possible to construct a proper 4-bit toffoli gate:
+On the other hand, it is possible to construct a proper 4-bit toffoli gate.
 \medskip
@@
 \noindent
-Now, the question we want to answer is: is it possible to implement \Afun{toffoli₄} gate only using \Afun{toffoli₃} gates?
+Now, we can ask: is it possible to implement \Afun{toffoli₄} gate using only \Afun{toffoli₃} gates?
+\medskip
 
-Solving this problem using the circuit representation is difficult. However, we can observe that all \Afun{toffoli₃}
-gates implements even permutations, that is, permutations consisting of an even number of transpositions. On the other
-hand, \Afun{toffoli₄} implements an odd permutation. It is an easily proven lemma that any composition of even
-permutations always produces an even permutation, which answers the question to the negative. Thus, we want to formulate
-the theorem using the representation on lists of transpositions, and transfer it to circuits using the equivalences
-proven in the previous sections.
+Solving this problem in the circuit representation is difficult. However, we can observe that all \Afun{toffoli₃} gates
+implement even permutations, that is, permutations that perform an even number of transpositions. On the other hand,
+\Afun{toffoli₄} implements an odd permutation. It can be easily proved that any composition of even permutations always
+produces an even permutation, which answers the question to the negative. Thus, we formulate the theorem using the
+representation of lists of transpositions, and transfer it to circuits using the equivalences proven in the previous
+sections.
 
-Thus, define a function $\parity : (c : X \iso Y) \to \Bool$ to compute a parity of a permutation defined by a
-combinator $c$, by first tranforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$ defined
-in~\cref{sec:equivalence}), and then computing the parity on this representation, which amounts to a simple check of the
-partiy of the length of the list.
+We define a function $\parity : (X \iso Y) \to \Bool$ to compute the parity of a permutation defined by a combinator
+$c$, by first transforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$ defined
+in~\cref{sec:equivalence}), and then computing the parity of this representation, which amounts to a simple checking of
+the parity of the length of the list.
 
@@
 \begin{proof}
-  First, it is an easy check that Coxeter relations preserve parity of the list. Second, if $c_1$ and $c_2$ are related
+  First, it is easy to check that Coxeter relations preserve parity of the list. Second, if $c_1$ and $c_2$ are related
   by a 2-combinator, then their Lehmer codes are the same, so their list-of-transpositions representations have to be
@@
 
-Using ~\cref{prop:parity-preserved} and already mentioned lemma that any compostion of even circuits is even, we can
-finish the proof by running $\parity$ on \Afun{toffoli₄} and on all of the \Afun{toffoli₃}.
+\noindent
+Using~\cref{prop:parity-preserved} and the fact that any composition of even circuits is even, we complete the proof by
+computing the $\parity$ of \Afun{toffoli₄}, and all of the \Afun{toffoli₃} combinations.
 

### 2021-10-30 Pi/paper/equivalence.tex
Many more edits

@@
 
-To produce a braiding $n + m \isoh m + n$ from adjacent tranpositions, we recursively traverse the left subexpression,
+To produce a braiding $n + m \isoh m + n$ from adjacent transpositions, we recursively traverse the left subexpression,
 swapping each element using adjacent transpositions along the elements on the right, placing it in the right position.

### 2021-10-30 Pi/paper/finite.tex
Many more edits

@@
 
+\noindent
 Bubble sort uses the primitive operation of adjacent swaps, insertion sort and selection sort use the primitive
@@
 
-We will show that it is possible to encode all $\PiLang$ combinators using adjacent transpositions
-(in~\cref{sec:equivalence}).
+\noindent
+We will show in~\cref{sec:equivalence} that all $\PiLang$ combinators can be encoded using adjacent transpositions.
 %Group presentations are built by adding equations to a \emph{free group}.
@@
 
-Usually, there are are many equations, besides the group axioms, that hold for the elements of a group. For example, in
+Usually, there are many equations, besides the group axioms, that hold for the elements of a group. For example, in
 the group $\Aut[\Bool]$, or $\mathbb{Z}_2$, we have an equation $1 + 1 = 0$, which is not a consequence of the group
@@
 $\inv{a}$. Then, we can encode the free group using the free monoid, that is, lists of $A + A$. Additionally, we need to
-ensure that the inverse laws hold, so we have to coalesce adjacent occurences of $a$ and $\inv{a}$.
+ensure that the inverse laws hold, so we have to coalesce adjacent occurrences of $a$ and $\inv{a}$.
 
@@
   \end{subfigure}
-  \caption{Braiding diarams for Coxeter relations.}
+  \caption{Braiding diagrams for Coxeter relations.}
   \label{fig:coxeter-braid}
@@
 For this to work, we first need the system to have the termination property, meaning that there should be no infinite
-reductions. We observe that after throwing out reflexivity and symmetry, the right hand sides of the relations $\cox*$
-are strictly smaller than the left hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$ (which is
+reductions. We observe that after throwing out reflexivity and symmetry, the right-hand sides of the relations $\cox*$
+are strictly smaller than the left-hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$ (which is
 well-founded). Thus, by directing the relation from left to right, we would get the termination property out of the box.
@@
 between one place right (see~\cref{fig:downf}). Then, the directed relation $\longcox$ is defined with the following
-generators (inlining the congruence closure in $\longcox$ allows arbitrary reductions inside the list).
+generators (inlining the congruence closure in the relation $\longcox$ allows arbitrary reductions inside the list).
 
@@
 the relation are (lexicographically) strictly larger than the right-hand sides. We define the transitive closure of
-$\longcox$ to be $\longcoxplus$ (\cref{def:longcox-plus}) and its reflexive-transitive closure to be $\longcoxstar$
-(\cref{def:longcox-star}).
+$\longcox$ to be $\longcoxplus$ (see~\cref{def:longcox-plus} in the appendix) and its reflexive-transitive closure to be
+$\longcoxstar$ (see~\cref{def:longcox-star}).
 
@@
 
+\noindent
 The modified form of the Coxeter relations are unwieldy and difficult to prove properties about by induction. However,
@@
 
-By~\cref{prop:nf}~\ref{prop:nf-uniqueness}, we get a unique choice function $\normf : {\List[\Fin[n]]} \to
+\noindent
+By~\cref{prop:nf}~part~\ref{prop:nf-uniqueness}, we get a unique choice function $\normf : {\List[\Fin[n]]} \to
 {\List[\Fin[n]]}$ that produces a normal form for terms of $\List[\Fin[n]]$. We state two important properties enjoyed
@@
 set of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The following
-proposition lets us lift a group structure on $\List[A]$ to a group presentation.
+proposition lets us lift a group structure on $\List[A]$ to a group presentation, with underlying set $\List[A+A]$.
 
@@
   the map on the underlying set is given by mapping $\inl$ on the list. Using that the relation is a congruence, we show
-  that this consttitutes a group isomorphism.
+  that this constitutes a group isomorphism.
 \end{proof}
@@
 
-The inversion count of an element is the number of smaller elements appearing after it in the permutation.
+The inversion count of an element is given by the number of smaller elements that appear after it in the permutation.
 
@@
 
-\noindent We can show that the function $\immersion_{n}$ gives an equivalence betweeen $\Lehmer[n]$ and $\im{\normf}$.
+\noindent We can show that the function $\immersion_{n}$ gives an equivalence between $\Lehmer[n]$ and $\im{\normf}$.
 

### 2021-10-30 Pi/paper/applications.tex
More edits

@@
 Solving this problem in the circuit representation is difficult. However, we can observe that all \Afun{toffoli₃} gates
-implement even permutations, that is, permutations that perform an even number of transpositions. On the other hand,
+implement even permutations, that is, permutations that contain an even number of transpositions. On the other hand,
 \Afun{toffoli₄} implements an odd permutation. It can be easily proved that any composition of even permutations always

### 2021-10-30 Pi/paper/reversible1.tex
More edits

@@
 
-\noindent The full set of level-2 combinators is large; the remaining combinators are listed in~\cref{app:leveltwo}.
+\noindent The full set of level-2 combinators is large; the remaining combinators are listed in~\cref{app:leveltwo} in
+the appendix of the extended version~\cite{choudhurySymmetriesReversibleProgramming2021}, and also in the accompanying
+Agda formalisation~\cite{choudhuryArtifactSymmetriesReversible2021}.
 

### 2021-10-30 Pi/paper/discussion.tex

@@
 The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
-presentation of reversible programming based on algebraic types matches---exactly---the categorified group-theoretic
+presentation of reversible computing based on algebraic types matches---exactly---the categorified group-theoretic
 semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
@@
 cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
-correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014} and the Lambek correspondence
-suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels} and established in this work, as in~\cref{fig:table-summary}. In the
-remainder of this section, we discuss some broader related work.
+correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014}, and the Lambek correspondence was
+suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels}. In this work, we have established this
+correspondence, as summarised in~\cref{fig:table-summary}. In the remainder of this section, we discuss some broader
+related work.
 
@@
             \hline
-            types         & Natural numbers                & Finite sets & 0-cells \tabularnewline
+            Types                          & Natural numbers               & Finite sets & 0-cells \tabularnewline
             \hline
@@
 exception traces, logs, backups, rollback recoveries, version control systems, reverse engineering, software
-transactional memories, continuations, backtracking search, and multiple-level undo features in commercial
-applications. In the early nineties, \citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
-treatment of reversibility. But intensive research in full-fledged reversible models of computations and reversible
-programming languages was only sparked by the discovery of deep connections between physics and
+transactional memories, continuations, backtracking search, and multiple-level undo features in commercial applications.
+In the early nineties, \citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class, treatment of
+reversibility. But intensive research in full-fledged reversible models of computations and reversible programming
+languages was only sparked by the discovery of deep connections between physics and
 computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275,Hey:1999:FCE:304763,fredkin1982conservative},
@@
 interesting to understand which versions of free monoidal structures correspond to extensions of $\Pi$ with recursive
-types~\cite{jamesInformationEffects2012,rc2011} and negative/fractional
-types~\cite{chenComputationalInterpretationCompact2021}.
+types~\cite{jamesInformationEffects2012,rc2011}, or negative and fractional
+types~\cite*{chenFractionalTypes2020,chenComputationalInterpretationCompact2021}.
 
@@
 focus of logic and computer science since its inception. Permutations, being at the core of reversible computing, are an
-interesting class of functions, for which there are few formal systems. We develop such a system bringing in all the
+interesting class of functions, for which there are few formal systems. The system we have developed brings in all the
 associated benefits of syntactic calculi, notably, their calculational flavor. Instead of comparing two reversible

### 2021-11-11 Pi/paper/discussion.tex
Fix headlines and captions
Citations

@@
 correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014}, and the Lambek correspondence was
-suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{threemodels}. In this work, we have established this
-correspondence, as summarised in~\cref{fig:table-summary}. In the remainder of this section, we discuss some broader
-related work.
+suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{caretteEmbracingLawsPhysics2021}. In this work, we
+have established this correspondence, as summarised in~\cref{fig:table-summary}. In the remainder of this section, we
+discuss some broader related work.
 
@@
     \end{center}
-    \caption{Summary of the established correspondences}
+    \caption{Summary of the established correspondences TODO: Move above}
     \label{fig:table-summary}
@@
 {homotopy.io}, Agda, Coq and Lean~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
-result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem. \citet{Hiver-coq}
-describes an explicit algorithm for producing normal forms. It could provide an alternative to our rewriting system. Other
-encodings of permutations as listed vectors, matrices, inductively generated trees (Motzkin trees), Young diagrams, or
-string diagrams, proved either difficult to formalise in type theory or difficult to relate directly to the primitive type
-isomorphisms of $\PiLang$. The automatic \citet{knuthSimpleWordProblems1970} completion produced too many equations
-making proving correctness and termination intractable.
+result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.
+\citet{hiverCoqCombiCoxeterPresentation2021} describes an explicit algorithm for producing normal forms. It could
+provide an alternative to our rewriting system. Other encodings of permutations as listed vectors, matrices, inductively
+generated trees (Motzkin trees), Young diagrams, or string diagrams, proved either difficult to formalise in type theory
+or difficult to relate directly to the primitive type isomorphisms of $\PiLang$. The automatic
+\citet{knuthSimpleWordProblems1970} completion produced too many equations making proving correctness and termination
+intractable.
 
@@
 transactional memories, continuations, backtracking search, and multiple-level undo features in commercial applications.
-In the early nineties, \citet{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class, treatment of
-reversibility. But intensive research in full-fledged reversible models of computations and reversible programming
-languages was only sparked by the discovery of deep connections between physics and
-computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275,Hey:1999:FCE:304763,fredkin1982conservative},
-and by the potential for efficient quantum computation~\cite{springerlink:10.1007/BF02650179}. The early developments of
-reversible programming languages started with a conventional programming language, e.g., an extended $\lambda$-calculus,
-and either (i) extended the language with a history
-mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible}, or (ii) imposed constraints on the control
-flow constructs to make them reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}.  More modern approaches recognize that
-reversible programming languages require a fresh approach and should be designed from first principles without the
-detour via conventional irreversible
-languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}. The version of
-$\PiLang$ studied in this paper is restricted to finite types and terminating total computations. It would be
-interesting to understand which versions of free monoidal structures correspond to extensions of $\Pi$ with recursive
-types~\cite{jamesInformationEffects2012,rc2011}, or negative and fractional
+In the early nineties, \citet{bakerNREVERSALFortuneThermodynamics1992,bakerLivelyLinearLisp1992} argued for a
+systematic, first-class, treatment of reversibility. But intensive research in full-fledged reversible models of
+computations and reversible programming languages was only sparked by the discovery of deep connections between physics
+and computation~\cite{landauerIrreversibilityHeatGeneration1961, peresReversibleLogicQuantum1985,
+    toffoliReversibleComputing1980, bennettFundamentalPhysicalLimits1985, frankReversibilityEfficientComputing1999,
+    heyFeynmanComputationExploring1999, fredkinConservativeLogic1982}, and by the potential for efficient quantum
+computation~\cite{feynmanSimulatingPhysicsComputers1982}. The early developments of reversible programming languages
+started with a conventional programming language, e.g., an extended $\lambda$-calculus, and either (i) extended the
+language with a history mechanism~\cite{vantonderLambdaCalculusQuantum2004,klugeReversibleSECD2000,
+    huelsbergenLogicallyReversibleEvaluator1996,danosReversibleCommunicatingSystems2004}, or (ii) imposed constraints on the
+control flow constructs to make them reversible~\cite{yokoyamaReversibleProgrammingLanguage2007}. More modern approaches
+recognize that reversible programming languages require a fresh approach and should be designed from first principles
+without the detour via conventional irreversible languages~\cite{yokoyamaPrinciplesReversibleProgramming2008,
+    muInjectiveLanguageReversible2004, abramskyStructuralApproachReversible2005, dipierroReversibleCombinatoryLogic2006}.
+The version of $\PiLang$ studied in this paper is restricted to finite types and terminating total computations. It
+would be interesting to understand which versions of free monoidal structures correspond to extensions of $\Pi$ with
+recursive types~\cite{jamesInformationEffects2012,bowmanDaggerTracedSymmetric2011}, or negative and fractional
 types~\cite*{chenFractionalTypes2020,chenComputationalInterpretationCompact2021}.

### 2021-11-11 Pi/paper/finite.tex
Fix headlines and captions
Citations

@@
 
-\subsection{Presenting the symmetric group}
+\subsection{Presenting the Symmetric Group}
 
@@
 
-\subsection{Free groups}
+\subsection{Free Groups}
 
@@
 
-\subsection{Group presentations}
+\subsection{Group Presentations}
 
@@
 gives a convenient way of representing permutations on a computer, partly because they are
-bitwise-optimal~\cite{bergerTeachingOrdinalPatterns2019a}.
+bitwise-optimal~\cite{bergerTeachingOrdinalPatterns2019}.
 
@@
 
-\subsection{Running Lehmer codes}
+\subsection{Running Lehmer Codes}
 
@@
 are off-by-one, because we chose $\Fin[n]$ to represent generators for permutations on $\Fin[\suc[n]]$.}. The idea for
-this proof is borrowed from~\citet{Molzer-cubical}.
+this proof is borrowed from~\citet{molzerCubicalAgdaSimple2021}.
 

### 2021-11-11 Pi/paper/introduction.tex
Say symmetric group instead of permutation group
More edits
Citations

@@
 
-Consider two programs that implement the same permutation using different sequences of type isomorphisms:
+Consider two programs that implement the \emph{same} permutation using different sequences of type isomorphisms:
 \begin{align*}
-      A + (B + C) \isoone
-      (A + B) + C & \isoone
-      C + (A + B) \isoone
+      A + (B + C) \leftrightarrow
+      (A + B) + C & \leftrightarrow
+      C + (A + B) \leftrightarrow
       C + (B + A)
       \\
-      A + (B + C) \isoone
-      A + (C + B) \isoone
-      (A + C) + B & \isoone
-      (C + A) + B \isoone
-      C + (A + B) \isoone
+      A + (B + C) \leftrightarrow
+      A + (C + B) \leftrightarrow
+      (A + C) + B & \leftrightarrow
+      (C + A) + B \leftrightarrow
+      C + (A + B) \leftrightarrow
       C + (B + A)
@@
 computing. Typically, they might be formalised as permutations $\{0,1\}^k \to \{0,1\}^k$ on bit strings of some length
-$k$~\cite{aaronson_et_al:LIPIcs:2017:8173,1201583}. But from the perspective of programming languages, permutations and
-reversible circuits can be more conveniently expressed as isomorphisms over algebraic datatypes as proposed in the
-$\PiLang$ family of reversible languages~\cite*{jamesInformationEffects2012,theseus} whose syntax is inspired by the
-sound and complete axiomatisation of type isomorphisms~\cite{fioreRemarksIsomorphismsTyped2002}, with respect to the
-bicartesian structure, that is, coproducts and products. Since $\PiLang$ programs correspond to type isomorphisms of
-finite types, our observation is that the syntax of $\PiLang$ is a presentation of the free symmetric rig groupoid (on
-zero generators).
+$k$~\cite{aaronsonClassificationReversibleBit2017}. But from the perspective of programming languages, permutations and
+reversible circuits can be more conveniently expressed as isomorphisms over algebraic data types, as proposed in the
+$\PiLang$ family of reversible languages~\cite*{jamesInformationEffects2012, jamesTheseusHighLevelLanguage2014}. The
+syntax of $\PiLang$ is inspired by the sound and complete axiomatisation of type isomorphisms of lambda
+calculi~\cite{fioreRemarksIsomorphismsTyped2002}, with respect to the bicartesian structure, that is, product and sum
+types. $\PiLang$ programs correspond to type isomorphisms of finite types, and our observation is that the syntax of
+$\PiLang$ is a presentation of the free symmetric rig groupoid (on zero generators).
 
-It is folklore that the groupoid of finite sets and permutations is the free symmetric rig groupoid on zero
-generators~\cite{laplaza72,kelly74,baez2000finite}. Our main result formally establishes this correspondence by giving an
-equational theory for $\PiLang$ that exactly includes all the necessary equations to decide equivalence of $\PiLang$
-programs. As conjectured by \citet{caretteComputingSemiringsWeak2016}, these equations correspond to the \emph{coherence
-      conditions of symmetric rig groupoids}, answering the conjecture in the positive.
+It is folklore that the groupoid of finite sets and bijections is the free symmetric rig groupoid on zero
+generators~\cite{laplazaCoherenceDistributivity1972, kellyCoherenceTheoremsLax1974, baezFiniteSetsFeynman2001}. Our main
+result formally establishes this correspondence by giving an equational theory for $\PiLang$ that includes exactly all
+the necessary equations to decide equivalence of $\PiLang$ programs. As conjectured
+by~\citet{caretteComputingSemiringsWeak2016}, these equations correspond to the \emph{coherence conditions of symmetric
+      rig groupoids}, answering the conjecture in the positive.
 
-\paragraph{Equivalence by Example} Without going into proofs, let us try to answer the question for the original
-pair of examples.
+\paragraph{Equivalence by Example} Without going into details of proofs, let us try to answer the question for the
+original pair of examples.
 
@@
       \]
+      \caption{Type isomorphisms as tree transformations}
       \label{fig:example-first-stage}
-      \caption{Type isomorphisms as tree transformations}
 \end{figure}
@@
       \label{fig:example-braid}
-      \caption{Type isomorphisms as braid diagrams}
+      \caption{Type isomorphisms as braiding diagrams}
 \end{figure}
@@
 The first step is to realise that the use of associativity is uninteresting, and the only steps with interesting
-computational content are the swaps. The swaps can be either big or small -- they can happen between leaf nodes, or
-between bigger subtrees, respectively.
+computational content are the swaps. The swaps can be either \emph{small} or \emph{big} -- they can happen between leaf
+nodes, or between bigger subtrees, respectively, as shown in~\cref{fig:example-first-stage}.
 
@@
 In this representation, each type is identified with its index, $A$ has index 0, $B$ has index 1, and $C$ has index 2.
-Then, we need to compile each primitive isomorphism to a list of adjacent transpositions, and then compose by appending
-the lists. A small swap is trivially implemented as a single adjacent transposition. To compile a big swap into adjacent
-transpositions, we need to traverse the subtrees in-order and recursively swap elements from the left by transposing
-them across the ones in the middle -- this generates a large number of adjacent transpositions. For the two programs, we
-get the following two lists: $[1,0,1,1,1]$ (\cref{fig:cone-braid}) and $[1,0,1]$ (\cref{fig:ctwo-braid}), where the
-number $k$ encodes a transposition of elements at indices $k$ and $k+1$.
+Then, we compile each primitive isomorphism to a list of adjacent transpositions, which are composed by appending the
+lists. A small swap is trivially implemented as a single adjacent transposition. To compile a big swap into adjacent
+transpositions, we traverse the subtrees in-order and recursively swap elements from the left by transposing them across
+the ones in the middle -- this generates a large number of adjacent transpositions. For the two programs, we get the
+following two lists: $[1,0,1,1,1]$ (\cref{fig:cone-braid}) and $[1,0,1]$ (\cref{fig:ctwo-braid}), where the number $k$
+encodes a transposition of elements at indices $k$ and $k+1$.
 
 Swapping is symmetric, that is, swapping two elements and immediately swapping them back produces the same permutation.
-The first list contains such redundant operations, so to eliminate these, we have to normalise the lists, which we will
-do by setting up reduction relations and an appropriate rewriting system. This system will normalise both lists to
-$[0,1,0]$ (\cref{fig:norm-braid}), which is lexicographically the smallest list that corresponds to this permutation.
-Since both programs have the same normal forms, they're equivalent!
+The first list contains such redundant operations, so to eliminate these, we will normalise the lists, by setting up
+reduction relations and an appropriate rewriting system. This system normalises both lists to $[0,1,0]$
+(\cref{fig:norm-braid}), which is lexicographically the smallest list that corresponds to this permutation. Since both
+programs have the same normal form, they're equivalent!
 
 The crux of designing an equational theory for $\PiLang$ relies on the choice of relations we use to design this
-rewriting system. First, there should be enough equations relating the programs to their normalised forms, and second,
-there should be enough equations corresponding to the reduction rules. We will show that these correspond to the
-coherence conditions for symmetric rig groupoids.
+rewriting system. First, there should be enough equations to compile programs to their normalised forms, that is,
+sequences of adjacent transpositions, and second, there should be enough equations to reduce them to a normal form, that
+is, the reduced list of transpositions. We will show that these equations correspond to the coherence conditions for
+symmetric rig groupoids.
 
-From the normalised list $[0,1,0]$, we can construct a permutation on a list of 3 elements $[a,b,c]$ as follows. We think
+From the normalised list $[0,1,0]$, we can recover a permutation on a list of 3 elements $[a,b,c]$ as follows. We think
 of it as insertion-sorting the elements of the list. Starting from the list $[a, b, c]$, we first insert $b$ at the
@@
 respect to its semantics in the weak symmetric rig groupoid of finite sets and permutations, formalised in HoTT/UF. We
-state our result as a Curry-Howard-Lambek correspondence for Reversible Logic, Reversible Programming Languages, and
-Symmetric Rig Groupoids, using which we can build a toolbox of technical devices for reasoning about reversible
-circuits.
+state our result as a Curry-Howard-Lambek correspondence for Reversible
+Logic~\cite{sparksSuperstructuralReversibleLogic2014}, Reversible Programming
+Languages~\cite{jamesInformationEffects2012}, and Symmetric Rig Groupoids~\cite{laplazaCoherenceDistributivity1972,
+kellyCoherenceTheoremsLax1974, caretteComputingSemiringsWeak2016}, using which we can build a toolbox of technical
+devices for reasoning about reversible circuits.
 
 \begin{itemize}[leftmargin=*]
-      \item We start in~\cref{sec:qiskit} by presenting a few reversible circuits in the popular IBM Qiskit framework to
-            serve as running examples throughout the paper.
-      \item In~\cref{sec:pi}, we introduce the two-level language
-            $\PiLang$~\cite{jamesInformationEffects2012,caretteComputingSemiringsWeak2016} and illustrate how to write
-            reversible circuits and their equivalences using 1-combinators and 2-combinators respectively. We give a
-            semantic account of the language by translating each level-1 program to a bijection between finite sets, and
-            verifying that programs identified by level-2 constructs denote the same bijection.
-      \item \cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and permutations, in
-            Homotopy Type Theory and Univalent Foundations. We define and characterise the notion of a univalent subuniverse, and construct
-            $\UFin$ as a univalent subuniverse which classifies all finite types. We establish that paths in $\UFin$ are
-            families of loops on finite sets of specified cardinality, given by $\Aut[\Fin[n]]$, which produces the
-            permutation group on $\Fin[n]$.
-      \item In~\cref{sec:finite}, we proceed to give a presentation of the permutation group, as the symmetric group
+      \item We start in~\cref{sec:qiskit} by presenting a few reversible circuits in the popular IBM Qiskit
+            framework~\cite{aleksandrowiczQiskitOpensourceFramework2019} to serve as running examples throughout the
+            paper.
+      \item In~\cref{sec:pi}, we introduce the two-level language $\PiLang$~\cite{caretteComputingSemiringsWeak2016} and
+            illustrate how to write reversible circuits and their equivalences using 1-combinators and 2-combinators
+            respectively. We give a semantic account of the language by translating each level-1 program to a bijection
+            between finite sets, and verifying that programs identified by level-2 constructs denote the same bijection.
+      \item \cref{sec:ufin} describes the construction of $\UFin$, the groupoid of finite sets and bijections, in
+            Homotopy Type Theory and Univalent Foundations (HoTT/UF). We define and characterise the notion of a
+            univalent subuniverse, and construct $\UFin$ as a univalent subuniverse which classifies all finite types.
+            We establish that paths in $\UFin$ are families of loops on finite sets of specified cardinality, given by
+            $\Aut[\Fin[n]]$, which produces the group of permutations on $\Fin[n]$.
+      \item In~\cref{sec:finite}, we proceed to give a presentation of this group, as the symmetric group
             $\Sn$ with generators and relations, and solve its word problem. In particular, we present $\Sn$ as a
@@
       \item In~\cref{sec:equivalence}, we show how to interpret the language $\PiLang$ into the groupoid $\UFin$, in
-            stages. First we define a subset $\PiPlusLang$ of the language which only includes the additive monoidal
+            stages. First, we define a subset $\PiPlusLang$ of the language which  includes the additive monoidal
             structure, and show how to translate $\PiLang$ programs to $\PiPlusLang$ programs. Then, we further define a

### 2021-11-11 Pi/paper/reversible1.tex
Citations

@@
 hardware platform. However, since its main data abstraction is a \emph{sequence of wires}, it only provides an
-``assembly-level'' programming abstraction (e.g., \verb|qasm|). As motivated by \citet{LAFONT2003257}, a mathematical
-model based on permutations of finite sets provides a richer algebraic structure, using which we describe a reversible
-programming language in this section.
+``assembly-level'' programming abstraction (e.g., \verb|qasm|). As motivated by
+\citet{lafontAlgebraicTheoryBoolean2003}, a mathematical model based on permutations of finite sets provides a richer
+algebraic structure, using which we describe a reversible programming language in this section.
 
@@
 % A specialised language for programming with permutations on finite sets can be built using ideas going back to Kelly,
-% Laplaza, and Mac Lane~\cite{laplaza72,kelly74,KELLY197197} that the commutative semiring (also known as the commutative
+% Laplaza, and Mac Lane~\cite{laplazaCoherenceDistributivity1972,kelly74,KELLY197197} that the commutative semiring (also known as the commutative
 % rig) of natural numbers exactly characterizes permutations on finite sets. In programming parlance, as shown in
@@
 
-All of this is only half the story, however. A sound semantics for $\PiLang$ using weak rig
-groupoids was established by~\citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this
-semantics, \emph{coherence conditions} for symmetric rig groupoids that identify different syntactic representations of the same
-permutation~\cite{laplaza72,caretteComputingSemiringsWeak2016}, were collected  in a
-second level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate
-$c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For
-example, we have the following level-2 combinators dealing with associativity:
+All of this is only half the story, however. A sound semantics for $\PiLang$ using weak rig groupoids was established
+by~\citet{caretteComputingSemiringsWeak2016}, and conjectured to be complete. For this semantics, \emph{coherence
+conditions} for symmetric rig groupoids that identify different syntactic representations of the same
+permutation~\cite{laplazaCoherenceDistributivity1972,caretteComputingSemiringsWeak2016}, were collected in a second
+level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for
+appropriate $c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same
+bijection. For example, we have the following level-2 combinators dealing with associativity:
 \medskip
@@
 For combinators, we show explicitly how values reduce along each combinator, similar to a big-step operational
-semantics~\cite{chenComputationalInterpretationCompact2021,theseus}.
+semantics~\cite{chenComputationalInterpretationCompact2021,jamesTheseusHighLevelLanguage2014}.
 

### 2021-11-11 Pi/paper/ufin.tex
Fix headlines and captions
Citations

@@
 
-\subsection{Rig structure}
+\subsection{Rig Structure}
 \label{subsec:rig}
@@
   structure distributes over the additive structure and satisfies the Laplaza coherence conditions for rig
-  categories~\cite{laplaza72}.
+  categories~\cite{laplazaCoherenceDistributivity1972}.
 \end{theorem}

### 2021-11-11 Pi/paper/discussion.tex
Fix table formatting
Co-authored-by: Jacek Karwowski <inexxt@users.noreply.github.com>
Final revisions

@@
 
-\begin{figure}
+\begin{table}
+    \caption{Summary of the established correspondences}
     \begin{center}
-        \begin{tabular}{|c|c|c|c|}
+        \begin{tabular}{cccc}
             \hline
-            \rowcolor{lightgray} $\PiLang$ & $\bigsqcup_n \mathcal{B}\Sn $ & $\BFin$     & $\UFin$ \tabularnewline
-            \hline
+            $\PiLang$     & $\bigsqcup_n \mathcal{B}\Sn $ & $\BFin$     & $\UFin$ \tabularnewline
+            \hline\hline
             Types         & Natural numbers               & Finite sets & 0-cells \tabularnewline
-            \hline
             1-combinators & Generators of $\Sn$           & Bijections  & 1-paths \tabularnewline
-            \hline
             2-combinators & Relations of $\Sn$            & Homotopies  & 2-paths \tabularnewline
@@
     \end{center}
-    \caption{Summary of the established correspondences TODO: Move above}
     \label{fig:table-summary}
-\end{figure}
+\end{table}
 
@@
 Higher-order term-rewriting systems and word problems have a long history of being formalised in proof assistants like
-{homotopy.io}, Agda, Coq and Lean~\cite{krausCoherenceWellFoundednessTaming2020}. As part of the proof of our main
+    {homotopy.io}, Agda, Coq and Lean. As part of the proof of our main
 result, we developed a rewriting system for the Coxeter relations for $\Sn$ to solve its word problem.
-\citet{hiverCoqCombiCoxeterPresentation2021} describes an explicit algorithm for producing normal forms. It could
+\citet{hiverCoqCombiCoxeterPresentation2021} describes an explicit algorithm for producing normal forms, which could
 provide an alternative to our rewriting system. Other encodings of permutations as listed vectors, matrices, inductively
@@
 \paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
-types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
-$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule
-$\term{ua_\beta} : (e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule
-$\term{ua_\eta} : (p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when
-formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full univalence axiom
-$(A \eqv B) \eqv (A \id_{\UU} B)$~\cite[Theorem 3.5]{ortonDecomposingUnivalenceAxiom2018}.
+types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ is constructed using path induction. The term $\term{ua} :
+{A \eqv B \to A \id_{\UU} B}$, its computation rule $\term{ua_\beta} : {(e : A \eqv B) \to \term{idtoeqv}(\term{ua}(e))
+\id e}$, and its extensionality rule $\term{ua_\eta} : {(p : A \id_{\UU} B) \to p \id \term{ua}(\term{idtoeqv(p)})}$ are
+generally added as postulates when formalising in Agda. Together, $\term{ua}$ and $\term{ua}_\beta$ give the full
+univalence axiom ${(A \eqv B) \eqv (A \id_{\UU} B)}$ \cite[Theorem 3.5]{ortonDecomposingUnivalenceAxiom2018}.
 % Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
@@
 that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory ($\PiLang$
-2-combinators). Thus, our results suggest a new computational interpretation of the univalence principle, which provides
-an intuition on why certain constructions are hard (or impossible in the general case). There are other, different
-approaches to computing with univalence,
+2-combinators). Thus, our results suggest a new way of thinking about computing using the univalence principle, which
+provides an intuition on why certain constructions are hard (or impossible in the general case). There are other,
+different approaches to computing with univalence,
 in~\cite{angiuliInternalizingRepresentationIndependence2021,tabareauMarriageUnivalenceParametricity2021}, and in Cubical
@@
 category $\FSM[\CCat{C}]$ on a category $\CCat{C}$. This is a 2-monad on
-$\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict) symmetric monoidal categories. Its
-construction is known in the literature~\cite{abramskyAbstractScalarsLoops2005}.  Concretely, the objects of
-$\FSM[\CCat{C}]$ are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n] \to \CCat{C}_{0})$;
-morphisms between $(n,A)$ and $(n,B)$ are pairs $(\pi,\lambda)$ where $\pi$ is a permutation of $[n]$, and
-$\lambda_{i} : A_{i} \to B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the Grothendieck construction
-$\int F$ of the functor $F : \BFin \to \CatCat$ from the groupoid of finite sets and bijections to $\CatCat$, assigning
-each natural number $n$ to the $n$-power $C^{n}$ of $C$, and each permutation on $[n]$ inducing an endofunctor on
-$C^{n}$ by action. The groupoid $\BFin$ is the free symmetric monoidal category (groupoid) on one generator, $\FSM[\unit]$.
+$\CatCat$~\cite{blackwellTwodimensionalMonadTheory1989}, whose algebras are (strict) symmetric monoidal categories. A
+construction of $\FSM$ is given by~\citet{abramskyAbstractScalarsLoops2005}. Concretely, the objects of $\FSM[\CCat{C}]$
+are given by lists of objects of $\CCat{C}$, that is, a pair $(n:\Nat, A:[n] \to \CCat{C}_{0})$; morphisms between
+$(n,A)$ and $(n,B)$ are pairs $(\pi,\lambda)$ where $\pi$ is a permutation of $[n]$, and $\lambda_{i} : A_{i} \to
+B_{\pi(i)}$ for $1 \leq i \leq n$. Abstractly, this is given by the Grothendieck construction $\int F$ of the functor $F
+: \BFin \to \CatCat$ from the groupoid of finite sets and bijections to $\CatCat$, assigning each natural number $n$ to
+the $n$-power $C^{n}$ of $C$, and each permutation on $[n]$ inducing an endofunctor on $C^{n}$ by action. The groupoid
+$\BFin$ is the free symmetric monoidal category (groupoid) on one generator, $\FSM[\unit]$.
 

### 2021-11-11 Pi/paper/equivalence.tex
Final revisions

--- 2021-10-30 Pi/paper/equivalence.tex
@@
 
-To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
-category. We define them formally in the appendix, and only state our results here. For each of the $\PiLang$,
-$\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and
-$\PiHatCat$, have 0-cells for types ($U$, $\UPlus$, and $\UHat$), 1-cells for 1-combinators ($\iso$, $\isop$, and
-$\isoh$), and 2-cells for 2-combinators ($\Iso$, $\Isop$, and $\Isoh$). We can show that these syntactic categories here
-are actually $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are also locally-strict, or
-locally-posetal, because there is at most one 2-cell between compatible 1-cells.
+To state our results formally, we organise the syntax for each language using its syntactic category. We define them
+formally in the appendix, and only state our results here. For each of the $\PiLang$, $\PiPlusLang$ and $\PiHatLang$
+languages, their syntactic categories, respectively $\PiCat$, $\PiPlusCat$ and $\PiHatCat$, have 0-cells for types ($U$,
+$\UPlus$, and $\UHat$), 1-cells for 1-combinators ($\iso$, $\isop$, and $\isoh$), and 2-cells for 2-combinators ($\Iso$,
+$\Isop$, and $\Isoh$). These syntactic categories here are actually $(2,0)$-categories, since all the 1-cells and
+2-cells are invertible. They are also locally-strict, or locally-posetal, because there is at most one 2-cell between
+compatible 1-cells.
 
@@
 As described, $\PiHatLang$ doesn't have a tensor product, but we can build it simply by adding up natural numbers, and,
-we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal structure.~\footnote{Since each object
-is a natural number, this makes $\PiHatCat$ a \emph{PROP}, that is, a products and permutations category.}
+we need to verify that this indeed equips $\PiHatCat$ with a symmetric monoidal structure. Since each object is a
+natural number, this makes $\PiHatCat$ a \emph{PROP}, that is, a products and permutations category.
 
-To produce a braiding $n + m \isoh m + n$ from adjacent tranpositions, we recursively traverse the left subexpression,
-swapping each element using adjacent transpositions along the elements on the right, placing it in the right position.
+To produce a braiding $n + m \isoh m + n$ from adjacent transpositions, we recursively traverse the left subexpression,
+swapping each element using adjacent transpositions along the elements on the right, placing it in the correct position.
 The computational content of this translation can be visualised using tree transformations, for the recursive case
@@
 \begin{definition}[$\evalh$]
+  \label{def:evalh}
   \begin{align*}

### 2021-11-11 Pi/paper/examples.tex
Final revisions

@@
   \end{tabular}
-  \caption{$\mathit{reversibleOr}$ in Qiskit}
+  \caption{$\mathsf{reversibleOr}$ in Qiskit}
   \label{fig:qiskit}
@@
 There is a wealth of manual and algorithmic approaches for producing circuits such as these
-two~\cite{maslov:2003:rls:1087512,1201583}. The circuit on the left was manually produced using a standard synthesis
-algorithm for reversible circuits~\cite{10.1145/775832.775915}. The circuit on the right was produced using an approach
-that analyzes the recursive structure of $\mathit{reversibleOr}$ (and would generalise to computing the disjunction of
-more than two inputs).
+two~\cite{maslovReversibleLogicSynthesis2003,shendeSynthesisReversibleLogic2003}. The circuit on the left was manually
+produced using a standard synthesis algorithm for reversible circuits~\cite{millerTransformationBasedAlgorithm2003}. The
+circuit on the right was produced using an approach that analyzes the recursive structure of $\mathsf{reversibleOr}$
+(and would generalise to computing the disjunction of more than two inputs).
 
-From the specification of the circuit, we expect input \verb|011| to be mapped
-to \verb|111|. To gain some intuition, we trace the evaluation of each circuit for
-input \verb|011|. In this context, the most significant bit is at index 0. In
-the first circuit, the \verb|ccx| gate negates \verb|q[0]| since both \verb|q[1]| and
-\verb|q[2]| are true producing \verb|111|; the following \verb|cx| gate produces
-\verb|011|; finally the last \verb|cx| produces the result \verb|111|. For the
-second circuit, the trace of the execution on the same input value \verb|011| goes through the stages \verb|111|, \verb|101|, \verb|101|, and finally \verb|111|.
+From the specification of the circuit, we expect input \verb|011| to be mapped to \verb|111|. To gain some intuition, we
+trace the evaluation of each circuit for input \verb|011|. In this context, the most significant bit is at index 0. In
+the first circuit, the \verb|ccx| gate negates \verb|q[0]| since both \verb|q[1]| and \verb|q[2]| are true producing
+\verb|111|; the following \verb|cx| gate produces \verb|011|; finally the last \verb|cx| produces the result \verb|111|.
+For the second circuit, the trace of the execution on the same input value \verb|011| goes through the stages
+\verb|111|, \verb|101|, \verb|101|, and finally \verb|111|.
 
@@
 classical core of Shor's algorithm which requires a circuit implementing modular exponentiation $f(r) =
-a^{r}~\mathrm{mod}~N$ for fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{Toffoli:1980}, a reversible
-specification of this function is $f^R(r,h) = (r, f(r)~\underline{\vee}~ h)$ where the exclusive
+a^{r}~\mathrm{mod}~N$ for fixed $a$ and $N$. Using Toffoli's construction~\citeyearpar{toffoliReversibleComputing1980},
+a reversible specification of this function is $f^R(r,h) = (r, f(r)~\underline{\vee}~ h)$ where the exclusive
 disjunction~$\underline{\vee}$ is applied bitwise. When $h=0$ the second component of the result of $f^R$ is the result
@@
 %%
-However, as explained in standard accounts of the algorithm (e.g., the
-Qiskit implementation), producing an efficient modular exponentiation circuit
-from this specification is not straightforward and is actually the bottleneck in
-Shor’s algorithm. Typical derivations of the circuit start from elementary
-gates, build a circuit for reversible disjunction (like the two circuits above),
-reversible conjunction, a circuit for a half-adder, a circuit for computing the
-carry, progressing to a circuit for modular addition, which is used to build a
-circuit for modular multiplication, and then finally a circuit for modular
-exponentiation taking care at each step to avoid the exponential blowup (e.g.,
-by implementing exponentiation by squaring instead of repeated
-multiplication)~\cite{shorefficient}.
+However, as explained in standard accounts of the algorithm (e.g., the Qiskit implementation), producing an efficient
+modular exponentiation circuit from this specification is not straightforward and is actually the bottleneck in Shor’s
+algorithm. Typical derivations of the circuit start from elementary gates, build a circuit for reversible disjunction
+(like the two circuits above), reversible conjunction, a circuit for a half-adder, a circuit for computing the carry,
+progressing to a circuit for modular addition, which is used to build a circuit for modular multiplication, and then
+finally a circuit for modular exponentiation taking care at each step to avoid the exponential blowup (e.g., by
+implementing exponentiation by squaring instead of repeated multiplication)~\cite{beauregardCircuitShorAlgorithm2003}.
 
@@
 % From the definition of $\underline{\vee}$ it is evident that setting $h=0$, we can compute the desired disjunction by observing the first component of the result.
-The $\mathit{reversibleOr}$ function has the following truth table (in binary on the left and in a more convenient decimal notation on the right):
+The $\mathsf{reversibleOr}$ function has the following truth table (in binary on the left and in a more convenient decimal notation on the right):
 

### 2021-11-11 Pi/paper/finite.tex
Final revisions

@@
 $n:\Nat$, that is, automorphisms of finite sets of size $n$, with the loopspace encoding the automorphism group.  This
-is also known to be the finite symmetric group $\Sn[n]$, making $\UFin$ the \emph{horizontal categorification} of $\Sn$
-for every $n$. In this section, we will describe this group syntactically.
+is also the finite symmetric group $\Sn[n]$, making $\UFin$ the \emph{horizontal categorification} of $\Sn$ for every
+$n$. In this section, we will describe this group syntactically.
 
@@
 permutations on a given list. So, a chosen set of reversible operations in a sorting algorithm can be a good candidate
-for the generators of a permutation group. For example, we could generate the symmetric group on $\Fin[n]$ by using
+for the generators of symmetric groups. For example, we could generate the symmetric group on $\Fin[n]$ by using
 generators (primitive operations) that:
@@
   \item swap the $i$-th element with an element at a fixed position, or
-  \item reverses a prefix $\Fin[k]$ of $\Fin[n]$ for $k \leq n$, or
+  \item reverse a prefix $\Fin[k]$ of $\Fin[n]$ for $k \leq n$, or
   \item cyclically shift any subset of $\Fin[n]$.
@@
 generating set. This is the universal property of free groups, stemming from the free-forgetful adjunction between the
-category of groups and sets. We say that $\eta_{A} : A \to F(A)$ is the free group on $A$ if it satisfies the following
-universal property.
+categories of groups and sets. We say that $F(A)$ with $\eta_{A} : A \to F(A)$ is the free group on $A$ if it satisfies
+the following.
 
@@
 \begin{definition}[Free group]
-  Let $A$ be a set, and $\List[\blank]$ the free monoid. The free group $F(A)$ on $A$ is the set-quotient of $\List[A +
-      A]$ by the congruence closure of the relation $a \cons \inv{a} \cons \nil \sim \nil$ and $\inv{a} \cons a \cons \nil
-    \sim \nil$.
+  Let $A$ be a set, and $\List[\blank]$ the free monoid monad. The free group $F(A)$ on $A$ is the set-quotient of
+      $\List[A + A]$ by the congruence closure of the relation $a \cons \inv{a} \cons \nil \sim \nil$ and $\inv{a} \cons
+      a \cons \nil \sim \nil$.
 \end{definition}
@@
 For this to work, we first need the system to have the termination property, meaning that there should be no infinite
-reductions. We observe that after throwing out reflexivity and symmetry, the right-hand sides of the relations $\cox*$
+reductions. We observe that, after throwing out reflexivity and symmetry, the right-hand sides of the relations $\cox*$
 are strictly smaller than the left-hand sides, in terms of the lexicographical ordering on words in $\Fin[n]$ (which is
@@
 \noindent
-Finally, we need to show that $\Sn$ and is indeed a group presentation.
+Finally, we need to show that $\Sn$ is indeed a group presentation.
 
@@
 set of words $A + A$, where the right copy corresponds to the set of formal inverses of the generators. The following
-proposition lets us lift a group structure on $\List[A]$ to a group presentation, with underlying set $\List[A+A]$.
+proposition lets us lift a group structure on $\List[A]$ to a group presentation, whose underlying set is $\List[A+A]$.
 
@@
 
-To prove the equivalence between $\Aut[\Fin[n]]$ and $\Sn[n]$, we will need to define functions back and forth between
-the two types. The terms in $\Sn$ can be identified with equivalence classes of terms in $\List[\Fin[n]]$ with respect
-to the Coxeter relation $\cox*$. The easiest way to define a function out of this presentation is to define it on the
+To prove the equivalence between $\Aut[\Fin[n]]$ and $\Sn[n]$, we will define functions back and forth between the two
+types. The terms in $\Sn$ can be identified with equivalence classes of terms in $\List[\Fin[n]]$ with respect to the
+Coxeter relation $\cox*$. The easiest way to define a function out of this presentation is to define it on the
 representatives. We know that these are the unique normal forms in the set-quotient given by $\quotinc \comp \normf$,

### 2021-11-11 Pi/paper/introduction.tex
Final revisions

@@
 calculi~\cite{fioreRemarksIsomorphismsTyped2002}, with respect to the bicartesian structure, that is, product and sum
-types. $\PiLang$ programs correspond to type isomorphisms of finite types, and our observation is that the syntax of
-$\PiLang$ is a presentation of the free symmetric rig groupoid (on zero generators).
+types. $\PiLang$ programs correspond to type isomorphisms of finite types, and in this paper, we make the observation
+that the syntax of $\PiLang$ is a presentation of the free symmetric rig groupoid (on zero generators).
 
@@
 
-Using this algorithm, we can turn a type isomorphism into a permutation of finite sets, relating the operational
-semantics of $\PiLang$ as permutations of finite sets of bits, to our denotational semantics. This allows us to
-establish full abstraction and adequacy.
+Using this algorithm, we can turn a type isomorphism into a bijection of finite sets, relating the operational semantics
+of $\PiLang$ as permutations of finite sets of bits, to our denotational semantics. This allows us to establish full
+abstraction and adequacy.
 
 \paragraph{Outline and Contributions} Our main result is a proof of the soundness and completeness of $\PiLang$ with
-respect to its semantics in the weak symmetric rig groupoid of finite sets and permutations, formalised in HoTT/UF. We
-state our result as a Curry-Howard-Lambek correspondence for Reversible
+respect to its semantics in the weak symmetric rig groupoid of finite sets, bijections, and homotopies, formalised in
+HoTT/UF. We state our result as a Curry-Howard-Lambek correspondence for Reversible
 Logic~\cite{sparksSuperstructuralReversibleLogic2014}, Reversible Programming

### 2021-11-11 Pi/paper/reversible1.tex
Final revisions

@@
 conditions} for symmetric rig groupoids that identify different syntactic representations of the same
-permutation~\cite{laplazaCoherenceDistributivity1972,caretteComputingSemiringsWeak2016}, were collected in a second
-level of $\PiLang$ syntax as level-2 combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for
-appropriate $c_1$ and $c_2$ of the same level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same
-bijection. For example, we have the following level-2 combinators dealing with associativity:
+permutation~\cite{laplazaCoherenceDistributivity1972}, were collected in a second level of $\PiLang$ syntax as level-2
+combinators.  Each level-2 combinator is of the form $c_1 \isotwo c_2$ for appropriate $c_1$ and $c_2$ of the same
+level-1 type $A \isoone B$ and asserts that $c_1$ and $c_2$ denote the same bijection. For example, we have the
+following level-2 combinators dealing with associativity:
 \medskip
@@
 \noindent The full set of level-2 combinators is large; the remaining combinators are listed in~\cref{app:leveltwo} in
-the appendix of the extended version~\cite{choudhurySymmetriesReversibleProgramming2021}, and also in the accompanying
-Agda formalisation~\cite{choudhuryArtifactSymmetriesReversible2021}.
+the extended version~\cite{choudhurySymmetriesReversibleProgramming2021}, and also in the accompanying Agda
+formalisation~\cite{choudhuryArtifactSymmetriesReversible2021}.
 

### 2021-11-11 Pi/paper/ufin.tex
Final revisions

@@
 $\BFin = \term{core}(\SetFin)$ of finite sets and bijections. $\SetFin$ has finite coproducts $(\emptyt, \sqcup)$ and
-finite products $(\unit, \times)$ and in $\BFin$ these restrict to additive and multiplicative symmetric monoidal
+finite products $(\unit, \times)$, and in $\BFin$, these restrict to additive and multiplicative symmetric monoidal
 structures, respectively, making $\BFin$ a symmetric rig groupoid -- the \emph{vertical categorification} of the
@@
 The semantics interprets types of $\PiLang$ as objects in $\BFin$, 1-combinators as isomorphisms, and for every pair of
-1-combinators related by a 2-combinator, their interpretations in~$\BFin$ are equal. $\BFin$ is a 1-groupoid, where the
-collection of isomorphisms is a set. We can evaluate two bijections on their inputs to decide if they are equal, and
-there is no higher cell witnessing the equality of two isomorphisms. To be able to establish completeness for $\PiLang$,
-we want a witness for this equality, so that we can quote back to the syntax and produce a 2-combinator witnessing the
-equality of the corresponding 1-combinator.
+1-combinators related by a 2-combinator, their interpretations in~$\BFin$ are equal. $\BFin$ is a (strict) 1-groupoid
+where isomorphisms are identified upto extensional equality. We can evaluate two bijections on their inputs to decide if
+they are equal, and there is no higher cell witnessing the equality of two isomorphisms. To be able to establish
+completeness for $\PiLang$, we want a witness for this equality, so that we can quote back to the syntax and produce a
+2-combinator witnessing the equality of the corresponding 1-combinator.
 
@@
 gives the (higher) morphisms. The groupoid we are interested in has types as points, type equivalences for 1-cells, and
-higher homotopies for higher cells. (See~\cref{app:grpdexample} for an example on a 3-element set.) This is the
-groupoid structure for the universe type $\UU$, since the identity type on types can be characterised as type
-equivalences (by \emph{univalence}). But, we only want to carve out a \emph{subuniverse} of \emph{finite types}, still
-satisfying univalence, to get the groupoid structure. In this section, we formally define \emph{univalent subuniverses},
-and proceed to construct the particular instance for finite types, $\UFin$ (\cref{def:ufin}).
+higher homotopies for higher cells. (See~\cref{app:grpdexample} for an example on a 3-element set.) This is the groupoid
+structure for the universe type $\UU$, since the identity type on types can be characterised as type equivalences (by
+\emph{univalence}). But, we only want to carve out a \emph{subuniverse} of \emph{finite types}, still satisfying
+univalence, to get the groupoid structure. In this section, we formally define \emph{univalent subuniverses}, and
+proceed to construct the particular instance for finite types, $\UFin$ (see~\cref{def:ufin}).
 
@@
 universe, such as $A \to \UU$, which is an $A$-indexed family of groupoids. For a type family $P : A \to \UU$ and a
-point $x : A$, the type $P(x)$ is the fiber over $x$.  The $\term{transport}$ operation, named
-$\term{transport}$, lifts paths in the indexing type to functions between fibers.
+point $x : A$, the type $P(x)$ is the fiber over $x$.  The $\term{transport}$ operation lifts paths in the indexing type
+to functions between fibers.
 \begin{gather*}
@@
 While $\UFin$ has \emph{all} the finite types, we are also interested in constructing a subuniverse of finite types of a
-specified cardinality. To do so, we will start with the subuniverse $\BAut[T]$, for any type $T : \UU$.
+\emph{specified} cardinality. To do so, we will start with the subuniverse $\BAut[T]$, for any type $T : \UU$.
 Characterisations of univalent fibrations using the $\BAut$ construction have also been studied
@@
 \begin{definition}[$\BAut$]
-  The predicate $P(X) \defeq \Trunc[-1]{X \id T}$ picks out exactly those types that are merely equal to $T$, and this
-  generates the subuniverse
+  The predicate $P(X) \defeq \Trunc[-1]{X \eqv T}$, or equivalently, $\Trunc[-1]{X \id T}$, picks out exactly those
+  types that are merely equivalent to $T$, and this generates the subuniverse
   \[
@@
   For any $n : \Nat$, we define $\UFin[n] \defeq \BAut[\Fin[n]]$ to be the univalent subuniverse of $n$-element sets.
-  Note that, $\UFin$ can be equivalently seen as the collection of all types of finite cardinality, that is,
-  $\UFin \eqv \dsum*{n:\Nat}{\UFin[n]}$.
+  Note that, $\UFin$ can be equivalently seen as the collection of all finite types, that is, $\UFin \eqv
+    \dsum*{n:\Nat}{\UFin[n]}$.
 \end{definition}
@@
 Since $\BAut[T]$ is a univalent subuniverse, we can characterise its path space. The intuition is that $\BAut[T]$ only
-has one point $T_0$, and 1-paths $T_0 \id T_0$, that is, loops, and higher paths between these loops. The
-type of loops on $T_0$, $\loopspace[\BAut[T],T_{0}]$, is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which is
-the group of automorphisms of $T$.
+has one point or 0-cell $T_0$, and 1-paths $T_0 \id T_0$, that is, loops, and higher paths between these loops. The type
+of loops on $T_0$, that is, $\loopspace[\BAut[T],T_{0}]$, is shown to be equivalent to $\Aut[T] \defeq T \eqv T$, which
+is the group of automorphisms of $T$.
 
@@
 \begin{theorem}
-  $\UFin[n]$ is a pointed, connected, 1-groupoid for every $n:\Nat$, and \( \loopspace[\UFin[n],F_{n}] \eqv
-  \Aut[\Fin[n]] \). $\UFin$ is a 1-groupoid with connected components for every $n:\Nat$.
+  \leavevmode
+  \begin{enumerate}
+    \item $\UFin[n]$ is a pointed, connected, 1-groupoid for every $n:\Nat$, and \( {\loopspace[\UFin[n],F_{n}] \eqv
+          \Aut[\Fin[n]]} \).
+    \item $\UFin$ is a 1-groupoid with connected components for every $n:\Nat$.
+  \end{enumerate}
 \end{theorem}
@@
 The loopspace of a pointed type automatically has the structure of a group, with $\refl_{\pt}$ for the neutral element,
-path composition for the group multiplication, and path inverse for the group inverse operation. The group axioms are
-given by the higher paths corresponding to groupoid laws.
+path composition for the group multiplication, and path inverse for the group inverse. The group axioms follow from the
+higher paths corresponding to groupoid laws.
 
@@
       \Fin[1]                & \eqv \top                             \\
-      \Fin[n] \times \Fin[m] & \eqv \Fin[n * m]                      \\[1.5ex]
+      \Fin[n] \times \Fin[m] & \eqv \Fin[n \cdot m]                  \\[1.5ex]
       \top \times X          & \eqv X                                \\
@@
   $(F_1, \times)$, with corresponding natural isomorphisms $\lambda_{X}$, $\rho_{X}$, $\alpha_{X,Y,Z}$, and the braiding
-  isomorphism $\mathcal{B}_{X,Y}$ upto 1-paths in $\UFin$. These isomorphisms satisfy the Mac Lane coherence conditions
+  isomorphism $\mathcal{B}_{X,Y}$, upto 1-paths in $\UFin$. These isomorphisms satisfy the Mac Lane coherence conditions
   for symmetric monoidal categories~\cite{maclaneNaturalAssociativityCommutativity1963}, that is, the triangle,

### 2021-11-11 Pi/paper/finite.tex
Final final revisions

@@
 Thus, the goal of this section is to reconcile two different approaches to defining the symmetric group -- as an
-automorphism group, and as a group syntactically presented using \emph{generators} and \emph{relations}. The generators
-of the group are similar to the primitive combinators in a (reversible) programming language -- the group structure
-gives the composition and inverse operations, and the relations describe how these primitive combinators interact with
-each other.
+automorphism group, and as a group syntactically presented using \emph{generators} and
+\emph{relations}~\cite{matsumoto1964generateurs}. The generators of the group are similar to the primitive combinators
+in a (reversible) programming language -- the group structure gives the composition and inverse operations, and the
+relations describe how these primitive combinators interact with each other.
 

### 2021-11-12 Pi/paper/applications.tex
Fix headlines and captions

@@
 
-\subsection{Synthesis and verification}
+\subsection{Synthesis and Verification}
 
@@
 ${(\mathsf{false},\mathsf{true},\mathsf{true})}$ is encoded as 011 or 3. The second entry in the $\term{lookup}$ table
-maps index 1 (= 001) to the value 5 (= 101), following the reversible disjunction specification (recall that
+maps index 1 (${= 001}$) to the value 5 (${= 101}$), following the reversible disjunction specification (recall the definition
 $\mathsf{reversibleOr}(h,b_1,b_2) = (h \,\underline{\vee}\, (b_1 \vee b_2), ~b_1, ~b_2)$ where~$\vee$ is disjunction
@@
 
-\subsection{Reasoning}
+\subsection{Reasoning and Transfer of Theorems}
 
@@
 We define a function $\parity : (X \iso Y) \to \Bool$ to compute the parity of a permutation defined by a combinator
-$c$, by first transforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$ defined
-in~\cref{sec:equivalence}), and then computing the parity of this representation, which amounts to a simple checking of
-the parity of the length of the list.
+$c$, by first transforming a $\PiLang$ program into a list of transpositions (using $\evalh_{1}$ from~\cref{def:evalh}),
+and then computing the parity of this representation, which amounts to a simple checking of the parity of the length of
+the list.
 
@@
 \noindent
-Using~\cref{prop:parity-preserved} and the fact that any composition of even circuits is even, we complete the proof by
-computing the $\parity$ of \Afun{toffoli₄}, and all of the \Afun{toffoli₃} combinations.
+Using~\cref{prop:parity-preserved}, we complete the proof by computing the $\parity$ of \Afun{toffoli₄}, and all
+combinations of the \Afun{toffoli₃} gates.
 

### 2021-11-12 Pi/paper/discussion.tex

@@
 The main theme of our work is the semantic foundation of reversible languages. We prove that a programming language
-presentation of reversible computing based on algebraic types matches---exactly---the categorified group-theoretic
+presentation of reversible computing based on algebraic types matches \emph{exactly} the categorified group-theoretic
 semantics, thereby closing the circle on a complete Curry-Howard-Lambek correspondence for reversible languages.
-Historically, the first such correspondence was between the $\lambda$-calculus, intuitionistic logic, and
-cartesian-closed categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard
-correspondence was established by~\citet{sparksSuperstructuralReversibleLogic2014}, and the Lambek correspondence was
-suggested by~\citet{caretteComputingSemiringsWeak2016} and~\citet{caretteEmbracingLawsPhysics2021}. In this work, we
-have established this correspondence, as summarised in~\cref{fig:table-summary}. In the remainder of this section, we
-discuss some broader related work.
+Historically, the first such correspondence related the $\lambda$-calculus, intuitionistic logic, and cartesian-closed
+categories~\cite{curryCurryEssaysCombinatory1980}. For reversible languages, the Curry-Howard correspondence was
+suggested by~\citet{sparksSuperstructuralReversibleLogic2014}, and the Lambek correspondence was suggested
+by~\citet*{caretteComputingSemiringsWeak2016} and~\citet*{caretteEmbracingLawsPhysics2021}. For a 1-bit fragment of
+$\PiLang$, this was established by~\citet*{caretteReversibleProgramsUnivalent2018}. In this work, we have completely
+established this correspondence, as summarised in~\cref{fig:table-summary}. In the remainder of this section, we discuss
+some broader related work.
 
